### 2022-07-29 archived/v1/BiSig.tex
[ refactor ] Tidy up archived files

%% For double-blind review submission, w/o CCS and ACM Reference (max submission space)
\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
%\documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
% For final camera-ready submission, w/ required CCS and ACM Reference
%\documentclass[acmsmall]{acmart}\settopmatter{}

%% Journal information
%% Supplied to authors by publisher for camera-ready submission;
%% use defaults for review submission.
\acmJournal{PACMPL}
\acmVolume{1}
\acmNumber{POPL} % CONF = POPL or ICFP or OOPSLA
\acmArticle{1}
\acmYear{2023}
\acmMonth{1}
\acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
\startPage{1}

%% Copyright information
%% Supplied to authors (based on authors' rights management selection;
%% see authors.acm.org) by publisher for camera-ready submission;
%% use 'none' for review submission.
\setcopyright{none}
%\copyrightyear{2018}           %% If different from \acmYear

%% Bibliography style
\bibliographystyle{ACM-Reference-Format}
\citestyle{acmauthoryear}   %% For author/year citations

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Note: Authors migrating a paper from PACMPL format to traditional
%% SIGPLAN proceedings format must update the '\documentclass' and
%% topmatter commands above; see 'acmart-sigplanproc-template.tex'.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% Some recommended packages.
\usepackage[utf8]{inputenc}
\usepackage[british]{babel}
\usepackage[hyperpageref]{backref}
\usepackage[capitalise,noabbrev]{cleveref}
\usepackage{xifthen,xspace}
\usepackage[color=yellow,textsize=footnotesize,prependcaption,obeyFinal]{todonotes}
\usepackage[inline]{enumitem} % for environment enumerate*
%\definecolor{addition}{RGB}{204,255,216}
%\usepackage{subcaption}
\usepackage{bussproofs}
\EnableBpAbbreviations
\usepackage[all,cmtip]{xy}
%\CompileMatrices

%\usepackage{mathtools}
%\usepackage[euler]{textgreek}

\usepackage{newunicodechar}

%% citation 
\newcommand{\varcitet}[3][]{\citeauthor{#2}#3~[\ifthenelse{\isempty{#1}}{\citeyear{#2}}{\citeyear[#1]{#2}}]}
\newcommand{\NoPeriod}[1]{\,}
\newcommand{\arXiv}[1]{\href{http://arxiv.org/abs/#1}{arXiv:\nolinkurl{#1}}}

%% set up unicode symbols
\newunicodechar{δ}{\ensuremath{\delta}}
\newunicodechar{ƛ}{\ensuremath{\lambdabar}}
\newunicodechar{Λ}{\ensuremath{\Lambda}}
\newunicodechar{⁺}{\ensuremath{{}^+}}

%% set up todonotes
\setlength {\marginparwidth }{2cm}
\newcommand{\LT}[2][]{%
  \ifthenelse{\isempty{#1}}%
  {\todo[author=LT,inline]{#2}}%
  {\todo[author=LT,inline,caption={#1}]{#2}}%
}
\newcommand{\Josh}[2][]{%
  \ifthenelse{\isempty{#1}}%
  {\todo[author=Josh,inline]{#2}}%
  {\todo[author=Josh,inline,caption={#1}]{#2}}%
}

\AtEndPreamble{%
 \theoremstyle{acmdefinition}
 \newtheorem{remark}[theorem]{Remark}
 \newtheorem{fact}[theorem]{Fact}
 \newtheorem*{remark*}{Remark}
}

\input{macros.tex}

\begin{document}

\title{From Binding Signature to Bidirectional Type Checking}
\author{Liang-Ting Chen}
\email{liang.ting.chen.tw@gmail.com}
\orcid{0000-0002-3250-1331}
\author{Hsiang-Shang Ko}
\email{joshko@iis.sinica.edu.tw}
\orcid{0000-0002-2439-1048}
\affiliation{%
  \department{Institute of Information Science}
  \institution{Academia Sinica}
  \streetaddress{128 Academia Road, Section 2, Nankang}
  \city{Taipei}
  \country{Taiwan}
  \postcode{115201}
}
\begin{abstract}

\end{abstract}

%% 2012 ACM Computing Classification System (CSS) concepts
%% Generate at 'http://dl.acm.org/ccs/ccs.cfm'.
\begin{CCSXML}
<ccs2012>
   <concept>
       <concept_id>10011007.10011006.10011039.10011040</concept_id>
       <concept_desc>Software and its engineering~Syntax</concept_desc>
       <concept_significance>500</concept_significance>
       </concept>
   <concept>
       <concept_id>10011007.10011006.10011008.10011024.10011028</concept_id>
       <concept_desc>Software and its engineering~Data types and structures</concept_desc>
       <concept_significance>300</concept_significance>
       </concept>
 </ccs2012>
\end{CCSXML}
\ccsdesc[500]{Software and its engineering~Syntax}
\ccsdesc[300]{Software and its engineering~Data types and structures}
%% End of generated code

\keywords{abstract syntax, binding signature, $F$-algebra, bidirectional type checking}
%% \keywords are mandatory in final camera-ready submission

\maketitle

\todototoc
\listoftodos[List of Todos]\\

\section{Introduction}
Context cannot be finite if we consider the transformation from an abstract syntax where the type of identifiers is not finite.

\subsection{Related Work}

\paragraph{Formal theory of abstract syntax with binding} \citep{Altenkirch1999a,Fiore1999,Fiore2008,Matthes2004,Tanaka2006,Ahrens2018b,Ahrens2018}
\paragraph{Formalisation and implementation} \citep{Allais2018,Allais2021}\citep{Ahrens2022}\citep{Fiore2022}
\paragraph{Bidirectional typing} \citep{Dunfield2021}

\subsection{Synopsis and Contributions}

\section{Mathematical Foundations of Substitution}
We assume the reader is familiar with basic category theory such as category, functor, natural transformation, limit, monad, and monoidal category in the classic textbook by \citet{MacLane1978}.

\paragraph{Terminology} A \emph{monad} in this paper refers to a \emph{functor} $T\colon \catC \to \catC$ with two natural transformations $\eta\colon \Id \to T$ and $\mu\colon T\fcomp T \to T$ satisfying monad laws.
An equivalent notion is \emph{Kleisli triple} which refers to a \emph{function} $T\colon \Obj{\catC} \to \Obj{\catC}$, a morphism $\eta_X \colon X \to T X$ for $X \in \Obj{X}$ and $\subst{f} \colon T X \to TY$ for $f\colon X \to Y$, satisfying monad laws.
The natural transformations $\eta$, $\mu$, and the mapping $(\blank)^\star$ are called \texttt{return}, \texttt{join}, and \texttt{bind} (with its two arguments flipped), for $T$ being strong, respectively in functional programming.

\subsection{Prelude: \texorpdfstring{$F$}{F}-Algebra and Initial Semantics}

In this section, we start with a recap of initial semantics using $F$-algebras, also fixing our notations.
Examples are carefully walked through to motivate our definitions and later developments.

\begin{definition}
  An algebra $(A, \alpha)$ for an endofunctor $F\colon \catC \to \catC$ is an object $A$ of $\catC$ and a morphism $\alpha \colon FA \to A$ where $A$ is called the \emph{carrier} and $\alpha$ the \emph{structure map}
  of the algebra, and $F$ the \emph{signature} functor. 
  An $F$-homomorphism from $(A, \alpha)$ to $(B, \beta)$ is a morphism $f\colon A \to B$ satisfying $f \comp \alpha = \beta \comp Ff$.
  The category of $F$-algebras and $F$-homomorphisms is denoted by $\mathsf{Alg}(F)$. 
\end{definition}
One simplistic example for signature functor is $(\Unit + \blank)\colon \set \to \set$ which maps any set $X$ to the coproduct the disjoint union $\Unit + X$ of the singleton $\Unit$ and $X$ and a function $f\colon X \to Y$ to $\idfunc[\Unit] + f$ defined as the mediating function $[\inl_{\Unit, X} \circ \idfunc[\Unit], \inr_{\Unit, X} \circ f]\colon \Unit + X \to \Unit + Y$ where $\inl_{1, X}$ (and $\inr_{1, X}$) is the left (and right) injection to the coproduct. 
For brevity, we follow the convention that $\idfunc[\Unit] + f$ is denoted $\Unit + f$.
The functoriality of $\Unit + (\blank)$ can be verified directly, or derived \emph{compositionally}, following from that (co-)limits of functors from $\catD$ to $\catC$ are calculated pointwise in~$\catC$ provided the pointwise (co-)limits exist.
That is, $\Unit + (\blank)$ is the coproduct of the constant functor $\underline{\Unit_{\set}}$ and the identity functor~$\Id[\set]$.
Since $\Unit$ is a singleton, $\alpha_1$ is viewed interchangeably as its value $\alpha_1(\ttt)$.

The structural map of a $(\Unit + \blank)$-algebra $(A, \alpha)$ is equal to the mediating morphism $[\alpha_1, \alpha_2]$ of $\alpha_1 = \alpha \comp \inl_{\Unit, A}$ and $\alpha_2 = \alpha \comp \inr_{\Unit, A} \colon A \to A$.
Hence, a $(\Unit + \blank)$-algebra can be regarded equivalently as a triple $(A, \alpha_1, \alpha_2)$. 
For example, the set $\Nat$ of natural numbers with the constant $0$ and the successor function $s\colon \Nat \to \Nat$ forms an $F$-algebra $(\Nat, [\underline{0}, s])$.
Moreover, this algebra $(\Nat, [\underline{0}, s])$ is initial:
For every $(\Unit + \blank)$-algebra $(A, \alpha)$, there is a unique homomorphism $f\colon (\Nat, [\underline{0}, s]) \to (A, \alpha)$.
Equivalently, every triple $(A, \underline{z}, \alpha_2)$ associates with
a unique function $f\colon \Nat \to A$ such that $f(0) = z$ and $f(s(n)) = \alpha_2 (f(n))$, modelling the structural recursion over the inductive type $\Nat$.
In terms of functional programming, this is exactly $\mathtt{fold}_{\Nat}\colon A \to (A \to A) \to \Nat \to A$ for natural numbers.

The next is one of our leading example: \citet{Bird1999}'s formulation of $\lambda$-terms as a nested type.
\begin{example}\label{ex:untyped-lambda-calculus}
  Define $\lambda$-terms up to $\alpha$-equivalence over an \emph{arbitrary} set $X$ of variables by
\begin{figure}[H]
  \begin{minipage}[b]{.3\textwidth}
    \begin{prooftree}
      \AXC{$x \in X$}
      \RightLabel{$(\mathsf{var})$}
      \UIC{$`x \in \Lambda X$}
    \end{prooftree}
  \end{minipage}
  \begin{minipage}[b]{.3\textwidth}
    \begin{prooftree}
      \AXC{$t \in \Lambda X$}
      \AXC{$u \in \Lambda X$}
      \RightLabel{$(\mathsf{app})$}
      \BIC{$t \cdot u \in \Lambda X$}
    \end{prooftree}
  \end{minipage}
  \begin{minipage}[b]{.3\textwidth}
    \begin{prooftree}
      \AXC{$t \in \Lambda (\Unit + X)$}
      \RightLabel{$(\mathsf{abs})$}
      \UIC{$\lambda t \in \Lambda X$}
    \end{prooftree}
  \end{minipage}
%  \caption{Untyped lambda terms up to $\alpha$-equivalence}
%  \label{fig:lambda-calculus}
\end{figure}%
\noindent also a function named after each rule---$\mathsf{var}_X(x) = `x$, $\mathsf{app}_X(t, u) = t \cdot u$, and $\mathsf{abs}_{X}(t) = \lambda\,t$.
  This formulation is a variant of \emph{de Bruijn representation} in the sense each bound variable in $t \in \Lambda X$ is of the form $(\inr^n \circ \inl)(\ttt) \in \Unit + (\cdots + (\Unit + X))$ for some natural number $n$;
  free variables are $\inr^{n+1}(x)$ for some $x$ in $X$.
  For brevity, we denote $\inl_{\Unit, X}(\ttt)$ and $\inr_{\Unit, X} \colon X \to \Unit + X$ by $z_X$ and $s_X$ respectively, e.g.,\ $\lambda\,\lambda `s(z)$ is the first projection (where the subscripts are omitted).
  This formulation resembles the \emph{locally nameless representation} by \citet{Chargueraud2012} where bound variables are represented by de Bruijn indices but free variables, introduced by a separate rule, are from another set.

  Simultaneous renaming $\Lambda\rho\colon \Lambda X \to \Lambda Y$ for a renaming rule  $\rho\colon X \to Y$  
  defined by
  \begin{align*}
    \Lambda\rho (x) & =  `\rho(x) & \Lambda\rho (t \cdot u)  & =  \Lambda\rho(t) \cdot \Lambda\rho(u) & \Lambda\rho (\lambda\,t) & =  \lambda\left(\Lambda(\Unit + \rho)(t)\right)
  \end{align*}
  replaces variables $x \in X$ by $\rho(x)$ but leaves bound variables intact, illustrating three diagrams
  \[
    \xymatrix@C+1em@R-1em{
      X \ar[r]^-{\mathsf{var}_X} \ar[d]_{\rho} & \Lambda X \ar[d]^{\Lambda(\rho)} \\
      Y \ar[r]_-{\mathsf{var}_Y} & \Lambda Y 
    }
    \qquad
    \xymatrix@C+1em@R-1em{
      \Lambda X\times\Lambda X \ar[r]^-{\mathsf{app}_X} \ar[d]_{\Lambda\rho \times \Lambda\rho} & \Lambda X \ar[d]^{\Lambda\rho} \\
      \Lambda Y\times\Lambda Y \ar[r]_-{\mathsf{app}_Y} & \Lambda Y 
    }
    \qquad
    \xymatrix@C+1em@R-1em{
      \Lambda(\Unit + X) \ar[r]^-{\mathsf{abs}_X} \ar[d]_{\Lambda(\Unit + \rho)} & \Lambda X \ar[d]^{\Lambda\rho} \\
      \Lambda(\Unit + Y) \ar[r]_-{\mathsf{abs}_Y} & \Lambda Y
    }
  \]
  where $\Lambda(\Unit + X)$ is the composed functor $\Lambda \fcomp (\Unit + \blank)$ applied to $X$.
  It is straightforward to check that with renaming $\Lambda\colon\set\to\set$ forms a functor, i.e.
  \[
    \Lambda\rho(\idfunc[X]) = \idfunc[\Lambda X] \quad\text{and}\quad \Lambda(\rho_2 \comp \rho_1)  = \Lambda\rho_2 \comp \Lambda\rho_1
  \]
  for any $\rho_1\colon X \to Y$ and $\rho_2\colon Y \to Z$.
  Therefore, each family of functions above is a natural transformation to $\Lambda$. 
  Moreover, each construct can be seen as an algebra for a suitable endofunctor:
  \begin{enumerate*}
    \item the constant functor $\underline{\Id[\set]}$, 
    \item the product $\Id \times \Id$, and
    \item the (context) extension $\delta$ defined by $\delta T \defid T \fcomp (\Unit + \blank)$ and $\delta\tau \defid \tau \fcomp (\Unit + \blank)$ for $\tau \colon T \to T'$, respectively.
  \end{enumerate*}
  In short, the three constructs of $\lambda$-terms exhibit an algebra structure on $\Lambda$:
  \[
    [\mathsf{var}, \mathsf{app}, \mathsf{abs}] \colon \Lambda + \LC(\Lambda) \to \Lambda
  \]
  for the endofunctor $\underline{\Id} + \LC$ of $\Endo{\set}$ where $\LC \defid \Id \times \Id + \delta$.
  This algebra is even the \emph{initial} algebra, as we will show later .
\end{example}

\begin{remark*}
We do not require the context set $X$ being finite or, in particular, a list of variables to make sense of context operations, contrast to the frameworks proposed by \citet{Fiore1999,Allais2021}.
For example, the \emph{weakening} operation is (still) defined solely via functorality and the injection $s \colon X \to \Unit + X$
\begin{align}\label{eq:weakening}
  \Lambda s      \colon \Lambda X \to \Lambda(\Unit + X)
\end{align}
where each variable $x$ in $t \in \Lambda X$ is `shifted' to $s(x)$, resulting a term $\Lambda s(t)$ in the context $\Unit + X$.

\end{remark*}

Formulating $\lambda$-terms as an algebra would be useless if this was the only algebra for $(\underline{\Id} + \LC)$.
Indeed, there are other examples and such an algebra needs not be syntactic at all.
\begin{example}\label{ex:free-variables}
  The set $\fv_X(t)$ of free variables in $t \in \Lambda X$ is defined for arbitrary $X$ by
  \begin{align*}
    \fv_X(`x) & = \{x\} & \fv_X(t \cdot u) & = \fv_X(t) \cup \fv_X(u) & \fv_X(\lambda\,t) = \fv_{\Unit + X}(t) \cap X
  \end{align*}
  Put differently, the definition of $\fv_X$ can be illustrated by three commutative diagrams 
  \begin{equation}\label{eq:free-variables}
    \vcenter{
    \xymatrix@C+1em@R-1em{
      X \ar[r]^-{\mathsf{var}_X} \ar[d]_{\idfunc} & \Lambda X \ar[d]^{\fv_X} \\
      X \ar[r]_-{\{\blank\}_X} & \power X 
    }}
    \qquad
    \vcenter{
    \xymatrix@C+1em@R-1em{
      \Lambda X\times\Lambda X \ar[r]^-{\mathsf{app}_X} \ar[d]_{\fv_X \times \fv_X} & \Lambda X \ar[d]^{\fv_X} \\
      \power X\times\power X \ar[r]_-{\blank \cup_X \blank} & \power X 
    }}
    \qquad
    \vcenter{
    \xymatrix@C+1em@R-1em{
      \Lambda(\Unit + X) \ar[r]^-{\mathsf{abs}_X} \ar[d]_{\fv_{\Unit +X}} & \Lambda X \ar[d]^{\fv_X} \\
      \power(\Unit + X) \ar[r]_-{\blank \cap X} & \power X 
    }}
  \end{equation}
  where $\power$ is the powerset functor and metaphorically $(\blank)\cap X$ strengthens indices in $S \in \power(\Unit + X)$ by removing $z_X$ and decrementing $s_X(x)$ to $x$.
  We can readily check that functions of the lower legs in~\eqref{eq:free-variables} are natural in $X$, forming a $(\underline{\Id} + \LC)$-algebra on the functor $\power$. 
  Renaming variables does not make a free variable bound or vice versa, so $\power\rho  \comp \fv_X = \fv_Y \comp \Lambda\rho$ holds for any $\rho$, showing that $\fv_X\colon \Lambda X \to \power X$ is natural in~$X$.
  Thus, by \eqref{eq:free-variables}, the natural transformation $\fv$ is a homomorphism.
  Indeed, this is the unique $\LC$-homomorphism from $(\Lambda, [\mathsf{var}, \mathsf{app}, \mathsf{abs}])$ as shown later.
\end{example}

We may wonder if \emph{simultaneous substitution} $\subst{\sigma}\colon \Lambda X \to \Lambda Y$ for a substitution rule $\sigma\colon X \to \Lambda Y$ is a homomorphism.
However, $\subst{\sigma}(x) = \sigma(x)$ is not necessarily a variable, so it is not homomorphic.
\subsection{Algebra with Substitution}\label{sec:algebra-with-substitution}
In this section, we extend algebras by substitution that is compatible with its algebra structure. 
Simultaneous substitution $\subst{\sigma}\colon \Lambda X \to \Lambda Y$ for a substitution rule $\sigma\colon X \to \Lambda Y$ is defined by
\begin{align*}
  \subst{\sigma}`x & = \sigma(x)  & \subst{\sigma}(t \cdot u) & = \subst{\sigma}t \cdot \subst{\sigma}u
                     &  \subst{\sigma}\lambda\,t & = \lambda \subst{[\mathsf{var}_{\Unit + Y} \circ \underline{z_Y}, \Lambda s \circ \sigma]}t
\end{align*}
where the substitution rule $[\mathsf{var}_{\Unit + Y} \comp \underline{z_Y}, \Lambda s \circ \sigma]$ in the third rule amounts to two functions
\[
  \mathsf{var}_{1 + Y} \circ \underline{z_Y} \colon \Unit \to \Lambda(\Unit + Y)
  \quad\text{and}\quad
  \Lambda s \circ \sigma \colon X \to \Lambda(\Unit + Y)
\]
specifying that the variable $z$ bound to $\lambda$ remains the same and every other variable $x \in X$ is replaced by $\sigma(x)$ but weakened using \eqref{eq:weakening}. 
It is is well-known \citep{Altenkirch1999a} that $(\Lambda, \mathsf{var}, \subst{\blank})$ forms a Kleisli triple, so
by the equivalence between Kleisli triple and monad, substitution $\subst{\sigma}$ boils down to---$\mu_Y \comp \Lambda \sigma$ with $\mu_Y = \subst{\idfunc[TY]}$---simultaneous renaming for the rule $\sigma$ followed by flattening a $\lambda$-term over $\lambda$-terms.
It turns out that $\subst{[\mathsf{var}_{\Unit + Y}\comp \underline{z_Y}, \Lambda s \circ \sigma]}$ decomposes into
\begin{align*}
      & \mu_{\Unit + Y} \comp \Lambda[\mathsf{var} \circ \underline{z}, \Lambda s \circ \sigma] \\
  ={} & \mu_{\Unit + Y} \comp {\color{red}\Lambda[\mathsf{var} \circ \underline{z}, \Lambda s]} \comp {\color{blue}\Lambda(\idfunc[\Unit] + \sigma)}
      && \{\,\text{by functorality and the `fusion law' for coproduct} \,\} \\
  ={} & (\delta\mu)_Y \comp {\color{red}\theta^{\mathsf{abs}}} \comp {\color{blue}(\delta\Lambda)\sigma}.
      && \{\,\text{by $\theta^{\mathsf{abs}} \defid \Lambda[\mathsf{var}_{\Unit + Y} \circ \underline{z_Y}, \Lambda s]$ and by the definition of $\delta$} \,\} 
\end{align*}
(with some subscripts omitted) so that the following diagram illustrates the third rule
\[
  \vcenter{
    \xymatrix@C+1em@R-1em{
      (\delta \Lambda) X \ar[r]^{\color{blue}(\delta\Lambda) \sigma} \ar[d]_{\mathsf{abs}_X} & (\delta \Lambda)\Lambda Y \ar[r]^{\color{red}\theta^{\mathsf{abs}}} \ar[d]_{\mathsf{abs}_{\Lambda Y}} &
      \left(\delta (\Lambda \fcomp \Lambda)\right) Y \ar[r]^-{(\delta\mu)_Y}& (\delta \Lambda) Y \ar[d]^{\mathsf{abs}_Y} \\
      \Lambda X \ar[r]_{\Lambda\sigma} \ar@<-.1ex> `d[r] `[rrr]_{\subst{\sigma}} [rrr] & \Lambda \Lambda Y \ar[rr]_{\mu_Y} & & \Lambda Y
    }
  }
\]
%\[
%  \vcenter{
%    \xymatrix{
%      (\Lambda \times \Lambda) X \ar[r]^-{\mathsf{app}_X} \ar[d]_{(\Lambda \times \Lambda)\sigma} & \Lambda X \ar[d]^{\Lambda\sigma} \\
%      (\Lambda\times\Lambda) \Lambda Y  \ar@{=}[d]_{\theta^{\mathsf{app}}_Y} \ar[r]^-{\mathsf{app}_{\Lambda Y}} & \Lambda\Lambda Y \ar[dd]^{\mu_Y} \\ 
%      (\Lambda \fcomp \Lambda \times \Lambda \fcomp \Lambda) Y \ar[d]_{(\mu\times\mu)_Y}&  \\ 
%      (\Lambda \times \Lambda) Y \ar[r]_{\mathsf{app}_{Y}} & \Lambda Y 
%    }
%  }
%  \quad
%  \vcenter{
%    \xymatrix{
%      (\delta \Lambda)X \ar[r]^{\mathsf{abs}_X} \ar[d]_{(\delta\Lambda)\sigma} & \Lambda X \ar[d]^{\Lambda\sigma} \\
%      (\delta \Lambda) \Lambda Y \ar[d]_{\theta^{\mathsf{abs}}_Y} \ar[r]^{\mathsf{abs}_{\Lambda Y}} &  \Lambda\Lambda Y \ar[dd]^{\mu_Y} \\ 
%      (\delta(\Lambda\fcomp \Lambda)) Y \ar[d]_{(\delta\mu)_Y} &  \\ 
%      (\delta\Lambda) Y \ar[r]_{\mathsf{abs}_{Y}} & \Lambda Y 
%    }
%  }
%\]
By examining the diagram carefully, we find that the only datum required to have substitution compatible with $\mathsf{abs}$ is $\theta^{\mathsf{abs}}$, motivating \Cref{def:strength,def:substitution-algebra}.
In order to define $\theta^{\mathsf{abs}}$ independent of $\Lambda$, we work with pointed endofunctors $(Z, e)$ representing structures with a (semantic) variable rule.
A \emph{pointed functor} consists of an endofunctor $Z$ of $\catC$ and a natural transformation $e\colon \Id \to Z$ and morphisms $f\colon (Z, e) \to (Z', e')$ between them are natural transformations $f\colon Z \to Z'$ satisfying $e' = f \comp e$.
We denote the category of pointed functors of $\catC$ by $\Pt{\catC}$ and its forgetful functor $(Z, e) \mapsto Z$ by $U\colon \Pt{\catC}\to \Endo{\catC}$.
  \begin{definition}\label{def:strength}
  A (right) \emph{strength} $\theta$ for $H\colon \Endo{\catC} \to\Endo{\catC}$ relative to $U\colon \Pt{\catC} \to \Endo{\catC}$ is a family $\theta$ of natural transformations
  \[
    \theta_{T, (Z, e)} \colon H T \fcomp Z \to H(T \fcomp Z)
  \]
  \emph{natural} in $T\colon \catC \to \catC$ and the pointed functor $(Z, e)$ such that 
  \begin{enumerate}
    \item $\theta_{T, (\Id, \idfunc)} = \idfunc[HT]$ for each $T$ and
    \item $\theta_{T, (Z' \fcomp Z, e' \fcomp e)} = \theta_{T \fcomp Z', (Z, e)} \comp \theta_{T, (Z', e')} \fcomp Z$ for any $T$ and pointed functors $(Z, e)$ and $(Z', e')$.
%      \[
%      \xymatrix@C+5em{
%        HT \fcomp (Z' \fcomp Z)
%        \ar[rr]^{\theta_{T, (Z' \fcomp Z, e' \fcomp e)}}
%        \ar@{=}[d] & & H(T \fcomp (Z' \fcomp Z)) \\
%        (HT \fcomp Z') \fcomp Z \ar[r]_{\theta_{T, (Z', e')} \fcomp Z} &  H(T \fcomp Z') \fcomp Z
%        \ar[r]_{\theta_{T\fcomp Z', (Z, e)}} & H((T \fcomp Z') \fcomp Z) \ar@{=}[u] 
%      }
%    \]
  \end{enumerate}
  For brevity, we simply call $H$ a \emph{functor with strength}, if the associated strength is clear from context.
\end{definition}
There are three naturality conditions, i.e.\ for $T$, $(Z, e)$ and objects $X$ in $\catC$, imposed on the strength but we shall not worry about checking them.
In practice, we will work with a presentation for $(H, \theta)$, called binding signatures, for a class of functors with strength so that the condition will be checked once for all.
\begin{definition}\label{def:substitution-algebra}
  An \emph{$H$-algebra with substitution} $(T, \eta, \alpha, \mu)$ consists of a monad $(T, \eta, \mu)$ and an $H$-algebra $(T, \alpha)$ satisfying
  \begin{equation}\label{dig:substitution}
    \vcenter{
      \xymatrix@C+2em{
        HT\fcomp T \ar[d]_{\alpha \fcomp T} \ar[r]^-{\theta_{T, (T, \eta)}} & H(T\fcomp T) \ar[r]^-{H\mu} & HT \ar[d]^{\alpha} \\
        T\fcomp T \ar[rr]_{\mu} & & T
      }
    }
  \end{equation}
  We may write $(T, \eta, \alpha, \blank^\star)$ for an $H$-algebra with substitution and call $\blank^\star$ \emph{(semantic) substitution} where $(T, \eta, \blank^\star)$ is the Kleisli triple in bijection with $(T, \eta, \mu)$.
\end{definition}
In other words, an $(H, \theta)$-algebra with substitution consists of a monad structure and an $H$-algebra on the same functor $T$ such that the algebra map commutes with the `semantic' substitution operation $(\blank)^\star$ in the way determined by the strength $\theta$.

\begin{remark*}
  This notion is a special case of \emph{$\Sigma$-monoids}~\citep{Fiore2008} for an endofunctor $\Sigma$ with a $(I/\catC)$-strength $\theta$ on a monoidal category~$(\catC, I, \otimes)$ defined as a $\Sigma$-algebra $(A, \alpha\colon \Sigma A \to A)$ for $A$ in $\catC$ equipped with a monoid structure $(A, \eta, \mu)$ satisfying \eqref{dig:substitution} while $\Sigma = H$ and $\catC$ is the monoidal category of endofunctors with the functor composition $\fcomp$ as the monoidal tensor $\otimes$.
\end{remark*}

Two previous examples are indeed algebras with substitution.
\begin{example}[\Cref{ex:untyped-lambda-calculus}, continued]
  Based on the discussion at the beginning of this section, define a strength $\theta^{\LC}$ for $\LC$ as the coproduct
  $\theta^{\times} + \theta^{\delta} $ of another two strengthens for $\Id \times \Id$ and $\delta$ respectively given by
  \begin{align*}
    \theta_{T, (Z, e)}^{\times} &\colon (T \times T) \fcomp Z \to TZ \times TZ  &
    \theta_{T, (Z, e)}^{\delta} &\colon \delta T \fcomp Z \to \delta (T \fcomp Z) \\
    \left(\theta_{T, (Z, e)}^{\times}\right)_Y & = \idfunc[TZY] \times \idfunc[TZY] &
    \left(\theta_{T, (Z, e)}^{\delta}\right)_Y & = T[e_{\Unit + Y} \comp \underline{z_Y}, Zs_Y].
  \end{align*}
  By instantiating $\theta^\delta$ with $\Lambda$ and $(\Lambda, \mathsf{var})$, we derive $\theta^{\mathsf{abs}}$ exactly as desired, so by construction $(\Lambda, \mathsf{var}, [\mathsf{app}, \mathsf{abs}], \subst{\blank})$ is an $\LC$-algebra with substitution.
\end{example}

\begin{example}[\Cref{ex:free-variables}, continued] \label{ex:free-variable-alg-subst}
  By instantiating $\theta$ with $\power$ and $(\power, \subset{\blank})$, the strength becomes a family of morphisms $\power[\subst{\blank} \comp \underline{z_X}, \power s]\colon \power(\Unit + \power X) \to \power^2(\Unit + X)$ natural in $X$ where 
  \[
    [\subst{\blank} \comp \underline{z_X}, \power s] \colon \Unit + \power X \to \power(\Unit + X)
  \]
  maps $z_X$ to the singleton $\{z\}$ and every subset $U \subseteq X$ to $s[U] \subseteq \Unit + X$, i.e.\ every element $x \in U$ is weakened.
  Then, it is routine to check the condition \eqref{dig:substitution}, so we omit it.
  It follows that $(\power, \{\blank\}, [\blank\cup_X\blank, \blank \cap X]_X, \bigcup)$ is indeed an $\LC$-algebra with substitution.
  %The case of $\mu^\power(U \cup V) = \mu^\power(U) \cup \mu^\power(V)$ for $U, V \in \power^2(X)$ is routine to check.
\end{example}

\subsection{Morphism of Algebra with Substitution}
We observe that the homomorphism $\fv$ satisfies a nice property called \emph{semantic substitution lemma}:
\begin{lemma}\label{lem:substitution-lemma-fv}
  Substitution for any rule $\sigma \colon X \to \Lambda Y$ commutes with $\fv$ in the sense that the diagram
  \begin{equation}\label{eq:substitution-lemma}
    \vcenter{
      \xymatrix@C+1em@R-1em{
        \Lambda X \ar[r]^{\subst{\sigma}} \ar[d]_{\fv_X} & \Lambda Y \ar[d]^{\fv_Y} \\
        \power  X \ar[r]_{\left(\fv_Y \comp \sigma\right)^\star} & \power Y
      }
    }
    \;\text{or, equivalently }\;
    \vcenter{
      \xymatrix@C+.5em@R-1em{
        \Lambda X \ar[d]_{\fv_X} \ar[r]^{\Lambda \sigma} & \Lambda \Lambda Y \ar[rr]^{\mu_Y} \ar[d]^{\fv_{\Lambda Y}} & & \Lambda Y \ar[d]^{\fv_Y} \\
        \power  X \ar[r]_{\power \sigma} & \power  \Lambda Y \ar[r]_{\power(\fv_Y)} & \power \power Y \ar[r]_{\bigcup} & \power Y
      }
    }
  \end{equation}
  commutes where $(\blank)^\star$ is the semantic substitution for the powerset monad $(\power, \{\blank\}, \bigcup)$. 
\end{lemma}
From a computational viewpoint, calculating the set $\fv_Y(\subst{\sigma} t)$ of free variables can be performed priori to substitution without building the intermediate term $\subst{\sigma}t$. 
Moreover, we only need to traverse the term $t$ once while collecting free variables $x_i$'s and compute the free variables of $\sigma(x_i)$ once even if $x_i$ may occur more than once in $t$, saving us not only space but also time.

This property holds even for every $(\underline{\Id}+\LC)$-morphism from the $(\Lambda, [\mathsf{var}, \mathsf{app}, \mathsf{abs}])$ if the target is an $\LC$-algebra with substitution.
Accordingly we revise the notion of morphisms for algebras with substitution.
Note that for \eqref{eq:substitution-lemma} to commute, it suffice to consider the right rectangle of the diagram on the right. 
Therefore the condition boils down to the following definition.
\begin{definition}
An \emph{$H$-homomorphism} $f$ between algebras with substitution from $(T, \eta, \alpha, \mu)$ to $(T', \eta', \alpha', \mu')$ is a natural transformation $f\colon T \to T'$ such that $f$ is an $H$-algebra homomorphism from $(T, \alpha)$ to $(T', \alpha')$ and $f$ is a monad morphism from $(T, \eta, \mu)$ to $(T', \eta', \mu')$, i.e.\ satisfying
\[
  \vcenter{
    \xymatrix{
      \Id \ar[r]^{\eta} \ar[rd]_{\eta'} & T \ar[d]^{f} \\
        & T'
    }
  }
  \qquad
  \vcenter{
    \xymatrix{
      HT  \ar[d]_{Hf} \ar[r]^{\alpha} & T \ar[d]^{f} \\
      HT' \ar[r]_{\alpha'} & T'
    }
  }
  \qquad
  \vcenter{
    \xymatrix{
      T \fcomp T \ar[d]_{\mu} \ar[r]^{f \fcomp T} & T' \fcomp T \ar[r]^{T' \fcomp f} & T' \fcomp T' \ar[d]^{\mu'} \\
      T \ar[rr]_{f} & & T'
    }
  }
\]
Let $\AlgS{H}$ denote the category formed by $H$-algebras with substitution and their homomorphisms.
\end{definition}
That is, an $H$-homomorphism between algebras with substitution commutes not only with $H$-algebra structure but also with the monad structure.
In \Cref{sec:binding-signature} we will show that the functor with strength associated with a binding signature, including $\LC$, has an initial algebra with substitution and $(\LC, \mathsf{var}, [\mathsf{app}, \mathsf{abs}], \subst{\blank})$ is indeed the initial algebra.
\Cref{lem:substitution-lemma-fv} follows from initiality immediately once substitution is proved compatible with the $\LC$-algebra map.

\subsection{Initial \texorpdfstring{$H$}{H}-algebra with substitution}

\section{From Signatures to Representations and Scope Checking}

\subsection{Binding Signature}\label{sec:binding-signature}
\LT{Define a binding signature as a sum of products for untyped language with weakening}
\LT{Derive a strong functor along $U$ from a binding signature}

\begin{example}
  \LT{An example of binding signature}

\end{example}

\begin{theorem}\label{thm:initial-algebra}
  \LT{Show that the strong functor associated with a signature gives rise to an initial algebra with substitution}
\end{theorem}

\subsection{Morphism between Signatures}
For two (mere) endofunctors $F$ and $F'$ on the same category~$\catC$, a natural transformation $\tau$ may be rightfully called a morphism between signature functors, as it gives rise to a functor turning every $F'$-algebra $(A, \alpha)$ to an $F$-algebra $(A, \alpha \comp \tau_A)$.
If we consider endofunctors on \emph{different} categories, say $\catC$ and $\catC'$, connected by a functor $E\colon \catC' \to \catC$ then a suitable notion of morphism from $F\colon \catC \to \catC$ to $F'\colon \catC' \to \catC'$, defined as a natural transformation $\tau$ from $F\fcomp E$ to $E\fcomp F'$, also gives rise to a functor from $\Alg{F'}$ to $\Alg{F}$.

\begin{definition}[Strong $J$-relative functor]
  Let $(J, e^J, m^J)$ be a monoidal functor from $\Endo{\catC'}$ to $\Endo{\catC}$ where $e^J\colon \Id[\catC] \to J\Id[\catC']$ and $m_{F, G}\colon JF \cdot JG \to J(F \cdot G)$.
\end{definition}

\begin{definition}
  \LT{Define morphism between signatures}
%  \[
%    \xymatrix@C+3em{
%      H(E T) \fcomp JZ \ar[r]^{\theta_{ET, (JZ, e \comp e_J)}} \ar[d]_{\tau_T \fcomp JZ} & H(E T \fcomp JZ) \ar[d]^{Ht_{T, Z}} \\
%      E(H'T) \fcomp JZ \ar[d]_{t_{H'T, Z}} & H (E(T \fcomp Z)) \ar[d]^{\tau_{TZ}} \\
%      E (H'T \fcomp Z) \ar[r]_{E \theta'_{T, (Z, e)}} & E (H'(T \fcomp Z)) \\
%    }
%  \]
\end{definition}
%\begin{theorem}
%  Every morphism of\, $U$-strong signature functor along a strong $J$-relative functor from $(\Endo{\catC}, H, \theta)$ to $(\Endo{\catC'}, H', \theta')$ defines a functor from $\HSS{H, \theta}$ to $\HSS{H', \theta'}$.   
%\end{theorem}

\begin{example}[Type erasure]

\end{example}

\LT{Show that each signature morphism defines a morphism between strong functors}
\subsection{Representing Substitution Systems}

\LT[Representations for initial HSSs]{1. Named variable representation; 2. Locally nameless representation; 3. Finitely-scoped representation}

\LT{Translation from named variable to locally nameless representations}
\LT{Translation from locally nameless to finitely-scoped representations}

\subsection{Scope Checking by Initiality}

\LT{Scope checking by initiality}

\section{From Bidirectional Typing Signatures to Type Inference}
\LT{Organise this section clearly}
$H\colon \Endo{\set^{\bool \times \catS}} \to \Endo{\set^{\bool \times \catS}}$ 
where $\bool$ is the free category over the graph $\bfalse, \btrue$ with edges $\bfalse \to \btrue$ and $\btrue \to \bfalse$.
There is a natural \emph{embedding} 
$J\colon \set^{\catS} \to \set^{\bool \times \catS}$ defined by
$(J \Gamma)(\btrue, s) = \Gamma(s)$ and $(J \Gamma)(\bfalse, s) = \Empty$.

\subsection{Signature for Bidirectional Typing}
Mode-Correctness and Annotatability

\subsection{Signature Functor for Bidirectional Typing}

\begin{example}[Direction erasure]

\end{example}

\subsection{Type Inference by Initiality}
\LT{Define bidirectional type inference by initiality}

\section{More Examples}

\section{Discussion and Future Work}

\paragraph{Relative monad}
For the sake of clarity, we only consider ordinary monads as the carrier of a heterogeneous substitution system, but however it is nature to use a category of contexts with a different chosen context extension than the one stated in this paper. 
The view motivates the use of relative monad proposed by \citet{Altenkirch2015}.

\paragraph{Comparing frameworks for scoped- and typed-safe syntaxes}

\begin{acks}
The work is supported by the \grantsponsor{MOST}{Ministry of Science and Technology of Taiwan}{https://www.most.gov.tw/} under grant \grantnum{MOST}{MOST 109-2222-E-001-002-MY3}.
\end{acks}

\bibliography{./library}

\appendix
%\section{Appendix}

\end{document}

### 2022-07-29 tex/BiSig.tex
[ new ] Tex file

\documentclass[envcountsect,runningheads,a4paper]{llncs}

\usepackage[utf8]{inputenc}
\usepackage[british]{babel}
\usepackage{cite}
\usepackage{amsmath}
\usepackage{xifthen,xspace,xcolor}
\usepackage[colorlinks=true,allcolors=black]{hyperref}
%\usepackage[hyperpageref]{backref}
\usepackage[capitalise,noabbrev]{cleveref}
\usepackage[color=yellow,textsize=footnotesize,prependcaption,obeyFinal]{todonotes}
\usepackage[inline]{enumitem} % for environment enumerate*
\usepackage{bussproofs}
\EnableBpAbbreviations
\usepackage[all,cmtip]{xy}
\usepackage{mathtools}
\usepackage{microtype}

\input{macros}

\title{From Binding Signatures to Bidirectional Typing}
\author{Liang-Ting Chen\orcidID{0000-0002-3250-1331} \and Hsiang-Shang Ko\orcidID{0000-0002-2439-1048}}
\authorrunning{L.-T.\ Chen and H.-S.\ Ko}
\institute{Institute of Information Science, Academia Sinica, Taiwan}

\begin{document}

\maketitle

\begin{abstract}
  Formalising an abstract (typed) syntax with variable binding to reason about typically starts from the boilerplate of definitions for substitution along with its metatheory, stimulating a number of language-formalisation frameworks:
  Given the presentation of a syntax in such a framework, a type of terms with recursion principle(s) and its metatheory will be instantiated, removing the burden of repeating the boilerplate.
  However, the metatheory of substitution is only the tip of the iceberg.

  In this paper, we extend the scope to bidirectional typing which is used for designing type checking algorithms.
  First, we adopt the notion of binding signatures to bidirectional type systems which allows us to formally specify and reason about criteria such as soundness, annotatability, and mode-correctness.
  Contrast to many frameworks, we emphasise the use of the extrinsic representation and implement a proof-relevant type checker for mode-correct bidirectional type systems in \Agda.
  This extension amounts to a type checker generator for simple type systems, 
  filling the gap between parser generator and language formalisation.
  \keywords{binding signature \and bidirectional typing \and initial semantics \and abstract syntax \and type checking \and language formalisation \and \Agda}
\end{abstract}

\section{Introduction}
\todo{abbreviate \emph{abstract (typed) syntax with variable binding} to \emph{(typed) language})}
\section{Initial Semantics of Syntax with Variable Binding}
\section{Extrinsically-Typed Representation}
\section{Binding Signature for Bidirectional Typing}
\section{Bidirectional Type Checking}
\section{Discussion}

\paragraph{Bidirectional typing}
\cite{Pierce2000a,Dunfield2021}

\paragraph{Universal theories of abstract syntax with variable binding} have been studied and proposed via various approaches over a few decades.
Herein we are interested in the category-theoretic approaches---particuarly \emph{the initisl semantics}---since they appear to be the most flexible and free from design choices if implemented properly (see below).
The initial semantics~\cite{Fiore1999} initiates from the observation that the family of scoped terms as an indexed set equipped with renaming has a $\Sigma$-monoid structure: substitution is monoidal and syntax constructs form an algebra for an endo-functor $\Sigma$ in the category of presheaves on contexts.

\todo{Heterogeneous substitution systems\cite{Matthes2004,Ahrens2018b}}
\todo{Modules over monads \cite{Hirschowitz2010}}
\todo{\cite{Hirschowitz2022}}

The idea of including the presentation as well as the language itself appeared earlier~\cite{Hamana2011a,Fiore2013a,Arkor2020} and is essential for our development, since we need to recognise type variables in the binding signature for defining mode-correctness.

\paragraph{Language-formalisation frameworks} have been implemented in proof assistants and vary in design choices including the binding representation paradigm, the language presentation, how the formal syntax is instantiated, and more.
Due to the limited space we will only discuss frameworks that are somewhat of different characteristics.
Nevertheless, the interested reader is referred to the discussion by Gheri and Popescu~\cite[Section 8]{Gheri2020} for a more comprehensive list of frameworks.

Among the plethora of choices, the formalisation of the category-theoretic approach based on (coalgebraic) $\Sigma$-monoids with metavariable~\cite{Fiore2008} by Fiore and Szamozvancev~\cite{Fiore2022} provides a metatheory independent of the aforementioned design choices whereas, instead of having the formal metatheory bolted on to a set of chosen paradigms, they only presume the property of a language to satisfy, i.e.\ the \emph{initiality}, offering a greater flexibility.
Another frameworks based on the category-theoretic notion of heterogeneous substitution systems are implemented by Ahrens et al.~\cite{Ahrens2018,Ahrens2022}
As we focus on the presentation of bidirectional typing, the category-theoretic approach is orthogonal to our interest.
Having said that, we have organised the underlying theory using category theory, hoping to integrate our development with the category-theoretic frameworks in future.

Our work is especially inspired by Allais et al.~\cite{Allais2021}
They demonstrated a number of generic programs and `meta-meta-properties' including term printing, scope checking, simulation lemma (strong normaliation via logical relation as an instance), and fusion lemma (Substitution Lemma~\cite{Mitchell1991a} as an instance) for languages expressible in their framework.
They also implemented a bidirectional type checker but for simply typed lambda calculus only.

\subsubsection*{Acknowledgements}
We thank Nathanael Arkor for the useful conversion over Twitter and to thank the Programming Languages and Formal Methods Laboratory at Academia Sinica for the opportunity of sharing earlier ideas.

The work is supported by the Ministry of Science and Technology of Taiwan under grant MOST 109-2222-E-001-002-MY3.

\bibliographystyle{splncs04}
\bibliography{ref}
%\bibliography{library}
\end{document}

### 2022-09-19 tex/BiSig.tex
[ update ] Bidirectionally typed terms with type variables

@@
 
-The idea of including the presentation as well as the language itself appeared earlier~\cite{Hamana2011a,Fiore2013a,Arkor2020} and is essential for our development, since we need to recognise type variables in the binding signature for defining mode-correctness.
+The idea of including the presentation of types as well as the language itself appeared earlier~\cite{Hamana2011a,Fiore2013a,Arkor2020} and is essential for our development, since we need to recognise type variables in the binding signature for defining mode-correctness.
 
-\paragraph{Language-formalisation frameworks} have been implemented in proof assistants and vary in design choices including the binding representation paradigm, the language presentation, how the formal syntax is instantiated, and more.
-Due to the limited space we will only discuss frameworks that are somewhat of different characteristics.
-Nevertheless, the interested reader is referred to the discussion by Gheri and Popescu~\cite[Section 8]{Gheri2020} for a more comprehensive list of frameworks.
+\paragraph{Language-formalisation frameworks} have been implemented in proof assistants and vary in design choices including binding representation paradigms, language presentations, the instantiation of formal syntax, and more.
+Due to the limited space the interested reader is referred to the discussion by Gheri and Popescu~\cite[Section 8]{Gheri2020} for a more comprehensive list of frameworks.
 

### 2023-05-23 tex/BiSig.tex
[ new ] Outline of the paper

@@
-\documentclass[envcountsect,runningheads,a4paper]{llncs}
-
+\documentclass[acmsmall]{acmart}
+\usepackage[T1]{fontenc}
 \usepackage[utf8]{inputenc}
 \usepackage[british]{babel}
-\usepackage{cite}
-\usepackage{amsmath}
-\usepackage{xifthen,xspace,xcolor}
-\usepackage[colorlinks=true,allcolors=black]{hyperref}
-%\usepackage[hyperpageref]{backref}
-\usepackage[capitalise,noabbrev]{cleveref}
-\usepackage[color=yellow,textsize=footnotesize,prependcaption,obeyFinal]{todonotes}
-\usepackage[inline]{enumitem} % for environment enumerate*
+
+\usepackage{mathtools}
+\usepackage{mathpartir}
+\usepackage[all,cmtip]{xy}
 \usepackage{bussproofs}
 \EnableBpAbbreviations
-\usepackage[all,cmtip]{xy}
-\usepackage{mathtools}
+
+\usepackage{xfrac,xspace}
+\usepackage{xcolor}
+\usepackage{subcaption}
+\usepackage{hyperref}
+\usepackage[capitalise,noabbrev]{cleveref}
+\usepackage[inline]{enumitem} % for environment enumerate*
+
+\usepackage[draft]{fixme}
+\usepackage{subfiles}
+
 \usepackage{microtype}
@@
 
-\title{From Binding Signatures to Bidirectional Typing}
-\author{Liang-Ting Chen\orcidID{0000-0002-3250-1331} \and Hsiang-Shang Ko\orcidID{0000-0002-2439-1048}}
-\authorrunning{L.-T.\ Chen and H.-S.\ Ko}
-\institute{Institute of Information Science, Academia Sinica, Taiwan}
+\FXRegisterAuthor{LT}{LTan}{L-T}
+\FXRegisterAuthor{Josh}{Joshan}{Josh}
+
+\citestyle{acmauthoryear}
 
@@
 
-\maketitle
+\author{Liang-Ting Chen}
+\email{liang.ting.chen.tw@gmail.com}
+\orcid{0000-0002-3250-1331}
+\author{Hsiang-Shang Ko}
+\orcid{0000-0002-2439-1048}
+\email{joshko@iis.sinica.edu.tw}
+
+\affiliation{%
+  \institution{Academia Sinica}
+  \streetaddress{128 Academia Road}
+  \city{Taipei}
+  \country{Taiwan}
+  \postcode{115}
+}
+
+\title{A Formal Theory of Bidirectional Typing for Simple Types}
 
 \begin{abstract}
-  Formalising an abstract (typed) syntax with variable binding to reason about typically starts from the boilerplate of definitions for substitution along with its metatheory, stimulating a number of language-formalisation frameworks:
-  Given the presentation of a syntax in such a framework, a type of terms with recursion principle(s) and its metatheory will be instantiated, removing the burden of repeating the boilerplate.
-  However, the metatheory of substitution is only the tip of the iceberg.
-
-  In this paper, we extend the scope to bidirectional typing which is used for designing type checking algorithms.
-  First, we adopt the notion of binding signatures to bidirectional type systems which allows us to formally specify and reason about criteria such as soundness, annotatability, and mode-correctness.
-  Contrast to many frameworks, we emphasise the use of the extrinsic representation and implement a proof-relevant type checker for mode-correct bidirectional type systems in \Agda.
-  This extension amounts to a type checker generator for simple type systems, 
-  filling the gap between parser generator and language formalisation.
-  \keywords{binding signature \and bidirectional typing \and initial semantics \and abstract syntax \and type checking \and language formalisation \and \Agda}
+  This paper proposes a formal theory of bidirectional typing.
 \end{abstract}
 
+\maketitle
 \section{Introduction}
-\todo{abbreviate \emph{abstract (typed) syntax with variable binding} to \emph{(typed) language})}
-\section{Initial Semantics of Syntax with Variable Binding}
-\section{Extrinsically-Typed Representation}
-\section{Binding Signature for Bidirectional Typing}
-\section{Bidirectional Type Checking}
-\section{Discussion}
 
-\paragraph{Bidirectional typing}
-\cite{Pierce2000a,Dunfield2021}
+\subsection{Related work}
 
-\paragraph{Universal theories of abstract syntax with variable binding} have been studied and proposed via various approaches over a few decades.
-Herein we are interested in the category-theoretic approaches---particuarly \emph{the initisl semantics}---since they appear to be the most flexible and free from design choices if implemented properly (see below).
-The initial semantics~\cite{Fiore1999} initiates from the observation that the family of scoped terms as an indexed set equipped with renaming has a $\Sigma$-monoid structure: substitution is monoidal and syntax constructs form an algebra for an endo-functor $\Sigma$ in the category of presheaves on contexts.
+\subsubsection{Language formalization and its frameworks}
+\cite{Wadler2022}
+PoplMark Challenge~\citep{Aydemir2005}
 
-\todo{Heterogeneous substitution systems\cite{Matthes2004,Ahrens2018b}}
-\todo{Modules over monads \cite{Hirschowitz2010}}
-\todo{\cite{Hirschowitz2022}}
+\cite{Cimini2020,Cimini2022}
+\citep{Fiore2022}
+\cite{Allais2021,Gheri2020,Ahrens2022}
 
-The idea of including the presentation of types as well as the language itself appeared earlier~\cite{Hamana2011a,Fiore2013a,Arkor2020} and is essential for our development, since we need to recognise type variables in the binding signature for defining mode-correctness.
+\subsubsection{Algebraic theory of type theories}
+\cite{Fiore1999,Fiore2022,Hamana2022,Ahrens2021,Arkor2020}
 
-\paragraph{Language-formalisation frameworks} have been implemented in proof assistants and vary in design choices including binding representation paradigms, language presentations, the instantiation of formal syntax, and more.
-Due to the limited space the interested reader is referred to the discussion by Gheri and Popescu~\cite[Section 8]{Gheri2020} for a more comprehensive list of frameworks.
+\subsubsection{Type checker generation}
+\cite{Gast2004,Grewe2015,Pacak2020,Cimini2020}
 
-Among the plethora of choices, the formalisation of the category-theoretic approach based on (coalgebraic) $\Sigma$-monoids with metavariable~\cite{Fiore2008} by Fiore and Szamozvancev~\cite{Fiore2022} provides a metatheory independent of the aforementioned design choices whereas, instead of having the formal metatheory bolted on to a set of chosen paradigms, they only presume the property of a language to satisfy, i.e.\ the \emph{initiality}, offering a greater flexibility.
-Another frameworks based on the category-theoretic notion of heterogeneous substitution systems are implemented by Ahrens et al.~\cite{Ahrens2018,Ahrens2022}
-As we focus on the presentation of bidirectional typing, the category-theoretic approach is orthogonal to our interest.
-Having said that, we have organised the underlying theory using category theory, hoping to integrate our development with the category-theoretic frameworks in future.
+\subsubsection{Bidirectional typing}
+\cite{Dunfield2021}
 
-Our work is especially inspired by Allais et al.~\cite{Allais2021}
-They demonstrated a number of generic programs and `meta-meta-properties' including term printing, scope checking, simulation lemma (strong normaliation via logical relation as an instance), and fusion lemma (Substitution Lemma~\cite{Mitchell1991a} as an instance) for languages expressible in their framework.
-They also implemented a bidirectional type checker but for simply typed lambda calculus only.
+\subsection{Plan and Contributions of this paper}
 
-\subsubsection*{Acknowledgements}
-We thank Nathanael Arkor for the useful conversion over Twitter and to thank the Programming Languages and Formal Methods Laboratory at Academia Sinica for the opportunity of sharing earlier ideas.
+\section{Bidirectional simply typed \texorpdfstring{$\lambda$}{lambda}-calculus}
+\subsection{Type Checking via Bidirectional Typing}
+
+\section{A Definition of Simple Type Theories}
+\subsection{Simple Types}
+\begin{definition}
+  A \emph{(simple) type signature} is a set $I$ with an \emph{arity} $\mathit{ar}_i : \mathbb{N}$ associated to each $i : I$.
+  An element $i : I$ is meant to be the \emph{$i$-th operation} $\tyop{i}$ and $\mathit{ar}(i)$ the \emph{number} of arguments of $\tyop{i}$.
+\end{definition}
+\LTnote{Define $\mathsf{Type}$}
+\begin{figure}
+  \begin{subfigure}{.5\textwidth}
+    \begin{mathpar}
+      \boxed{\Theta\vdash A} \\
+      \inferrule{\Theta \ni X_i}{\Theta \vdash X_i} \and
+      \inferrule{\Theta \vdash A_1 \\ \cdots \\ \Theta \vdash A_n}{\Theta \vdash \tyop{i}(A_1, \ldots, A_n)}
+    \end{mathpar}
+  \end{subfigure}
+  \begin{subfigure}{.4\textwidth}
+    \begin{mathpar}
+      \boxed{\Theta \vdash \Gamma} \\
+      \inferrule{ }{\Theta \vdash \emptyset}\and
+      \inferrule{\Theta \vdash A \\ \Theta\vdash \Gamma}{\Theta \vdash \Gamma, A}
+    \end{mathpar}
+  \end{subfigure}
+  \caption{Simple types and contexts}
+\end{figure}
+
+\begin{example}[STLC]
+
+\end{example}
+
+\subsection{Binding Signatures and Typed Terms}
+\begin{definition}
+  For a type signature, a \emph{term signature} consists of a set $J$ meant to be the set of term constructs $\op{j}$ with a triple $\left(\Xi_j, A_{0, j}, \left[\left(\Delta_{1, j}, A_{1, j}\right), \ldots, \left(\Delta_{n_j, j}, A_{n_j, j}\right) \right]\right)$ for each $j:J$ where
+  \begin{enumerate}
+    \item $A_{0, j} : \mathsf{Type}(\Xi_j)$ is the type of its target and
+    \item $\left[\left(\Delta_{1, j}, A_{1, j}\right), \ldots, \left(\Delta_{n_j, j}, A_{n_j, j}\right) \right]$ is a finite sequence of pairs of a context $\Delta_{k, j} : \mathsf{Cxt}(\Xi_j)$ for types of its binding variables and its type $A_{k, j} : \mathsf{Type}(\Xi_j)$.
+  \end{enumerate}
+\end{definition}
+\begin{figure}
+  \centering
+  \begin{mathpar}
+    \boxed{\Theta \mid \Gamma \vdash t : A} \\
+    \inferrule{\Gamma \ni (x : A)}{\Theta \mid \Gamma \vdash \term{x} : A}\;(\text{var})
+    \and
+    \inferrule{\Theta \mid \Gamma, \simsub{\Delta_{1,j}}{\rho} \vdash \term{t_1} : \simsub{A_{1,j}}{\rho} \\ \cdots \\ \Theta \mid \Gamma, \simsub{\Delta_{n_j, j}}{\rho} \vdash \term{t_{n}} : \simsub{A_{n, j}}{\rho}} {\Theta \mid \Gamma \vdash \term{\op{j, \rho}(t_1, \ldots, t_{n_j})} : \simsub{A_{0,j}}{\rho}}\;(\text{op})
+    \quad \text{where $\rho \colon \Xi_j \to \Theta$}
+  \end{mathpar}
+
+  \caption{Typing rules}
+\end{figure}
+
+\begin{example}[STLC]
+
+\end{example}
+
+\section{Definitions of Bidirectional Simple Type Systems}
+
+\subsection{Bidirectional Binding Signatures}
+
+\begin{definition}
+
+\end{definition}
+
+\begin{figure}
+  \centering
+  \begin{mathpar}
+    \boxed{\Theta \mid \Gamma \vdash \term{t} \Rightarrow A} \quad\text{and}\quad\boxed{\Theta \mid \Gamma \vdash \term{t} \Leftarrow A} \\
+    \inferrule{\Gamma \ni (\term{x} : A)}{\Theta \mid \Gamma \vdash \term{x} \mathrel{\color{red}\Rightarrow} A}\;(\text{var}{\Rightarrow})
+    \and
+    \inferrule{\Theta\mid \Gamma \vdash \term{t} \Rightarrow B \\ A = B}{\Theta \mid \Gamma \vdash (\term{t} : B) \Rightarrow A}\;(\textcolor{red}{anno})
+    \and
+    \inferrule{\Theta\mid \Gamma \vdash \term{t} \Rightarrow B \\ A = B}{\Theta \mid \Gamma \vdash \term{t\uparrow} \Leftarrow A}\;(\textcolor{red}{sub})
+    \and
+    \inferrule{\Theta \mid \Gamma, \simsub{\vec{x}_{1,j} : \Delta_{1, j}}{\rho} \vdash \term{t_{1, j}} \mathrel{\color{red}\leftrightarrows_{1, j}} \simsub{A_{1, j}}{\rho} \\ \cdots \\ \Theta \mid \Gamma, \simsub{\vec{x}_{n_j} : \Delta_{n_j, j}}{\rho} \vdash \term{t_{n_j, j}} \mathrel{\color{red}\leftrightarrows_{n_j}} \simsub{A_{n_j ,j}}{\rho}}
+    {\Theta \mid \Gamma \vdash \term{\op{j, \rho}(\vec{x}_1 .\, t_1, \ldots, \vec{x}_{n_j}.\, t_{n_j})} \mathrel{\color{red}\leftrightarrows_{0, j}} \simsub{A_{0, j}}{\rho}} \; (\text{op}{\leftrightarrows})
+    \quad \text{where $\rho \colon \Xi_i \to \Theta$}
+  \end{mathpar}
+  \caption{Bidirectional typing rules}
+\end{figure}
+
+\begin{example}
+
+\end{example}
+
+\subsection{Bidirectionally Typed Terms: Intrinsic Typing vs.\ Extrinsic Typing}
+\subsubsection{Intrinsically Typed Terms}
+
+\subsubsection{Raw Terms}
+
+\begin{figure}
+  \caption{Bidirectional raw terms}
+\end{figure}
+\subsubsection{Extrinsically Typed Terms}
+
+\begin{example}
+
+\end{example}
+
+\subsection{Binding Signatures vs.\ their Bidirectional Variants}
+\subsubsection{Soundness}
+\subsubsection{Completeness (Annotatability)}
+\subsection{Mode Correctness}
+
+\section{Bidirectional Type Synthesis and Checking}
+\subsection{Substitution Properties}
+\subsection{Generic First-Order Unification}
+\cite{McBride2003,McBride2003a}
+\subsection{Uniqueness of Synthesis}
+
+\subsection{Type Synthesis and Checking with Unification}
+
+\section{Formal Implementation and Examples}
+
+\citep{Xie2018}
+
+\section{Concluding Remarks and Future Work}
+\begin{acks}
+We thank Nathanael Arkor for the useful conversation and to thank the Programming Languages and Formal Methods Laboratory at Academia Sinica for the opportunity of sharing earlier ideas.
 
 The work is supported by the Ministry of Science and Technology of Taiwan under grant MOST 109-2222-E-001-002-MY3.
+\end{acks}
 
-\bibliographystyle{splncs04}
-\bibliography{ref}
-%\bibliography{library}
+\bibliographystyle{ACM-Reference-Format}
+
+\IfFileExists{reference.bib}{%
+  \bibliography{reference}
+}{\bibliography{ref}}
+
 \end{document}

### 2023-05-23 tex/BiSig.tex
[ update ] Fine-tune the paper outline

@@
 \subsection{Type Checking via Bidirectional Typing}
+\subsection{Formalisation in \Agda}
 
@@
 \begin{definition}
-  A \emph{(simple) type signature} is a set $I$ with an \emph{arity} $\mathit{ar}_i : \mathbb{N}$ associated to each $i : I$.
+  A \emph{(simple type) signature} is a set $I$ with an \emph{arity} $\mathit{ar}_i : \mathbb{N}$ associated to each $i : I$.
   An element $i : I$ is meant to be the \emph{$i$-th operation} $\tyop{i}$ and $\mathit{ar}(i)$ the \emph{number} of arguments of $\tyop{i}$.
@@
 
-\subsection{Binding Signatures and Typed Terms}
+\subsection{Binding Signatures and Intrinsically Typed Terms}
 \begin{definition}
-  For a type signature, a \emph{term signature} consists of a set $J$ meant to be the set of term constructs $\op{j}$ with a triple $\left(\Xi_j, A_{0, j}, \left[\left(\Delta_{1, j}, A_{1, j}\right), \ldots, \left(\Delta_{n_j, j}, A_{n_j, j}\right) \right]\right)$ for each $j:J$ where
+  For a type signature, a \emph{binding signature} consists of a set $J$ meant to be the set of term constructs $\op{j}$ with a triple $\left(\Xi_j, A_{0, j}, \left[\left(\Delta_{1, j}, A_{1, j}\right), \ldots, \left(\Delta_{n_j, j}, A_{n_j, j}\right) \right]\right)$ for each $j:J$ where
   \begin{enumerate}
-    \item $A_{0, j} : \mathsf{Type}(\Xi_j)$ is the type of its target and
+    \item $\Xi_j : \mathbb{N}$ is the number of variables, 
+    \item $A_{0, j} : \mathsf{Type}(\Xi_j)$ is the type of its target,  and
     \item $\left[\left(\Delta_{1, j}, A_{1, j}\right), \ldots, \left(\Delta_{n_j, j}, A_{n_j, j}\right) \right]$ is a finite sequence of pairs of a context $\Delta_{k, j} : \mathsf{Cxt}(\Xi_j)$ for types of its binding variables and its type $A_{k, j} : \mathsf{Type}(\Xi_j)$.
@@
   \begin{mathpar}
-    \boxed{\Theta \mid \Gamma \vdash t : A} \\
+    \boxed{\Theta \mid \Gamma \vdash t : A} 
+    \quad\text{where $\Theta \vdash \Gamma$ and $\Theta \vdash A$} \\
     \inferrule{\Gamma \ni (x : A)}{\Theta \mid \Gamma \vdash \term{x} : A}\;(\text{var})
     \and
-    \inferrule{\Theta \mid \Gamma, \simsub{\Delta_{1,j}}{\rho} \vdash \term{t_1} : \simsub{A_{1,j}}{\rho} \\ \cdots \\ \Theta \mid \Gamma, \simsub{\Delta_{n_j, j}}{\rho} \vdash \term{t_{n}} : \simsub{A_{n, j}}{\rho}} {\Theta \mid \Gamma \vdash \term{\op{j, \rho}(t_1, \ldots, t_{n_j})} : \simsub{A_{0,j}}{\rho}}\;(\text{op})
+    \inferrule{\Theta \mid \Gamma, \simsub{\Delta_{1,j}}{\rho} \vdash \term{t_1} : \simsub{A_{1,j}}{\rho} \\ \cdots \\ \Theta \mid \Gamma, \simsub{\Delta_{n_j, j}}{\rho} \vdash \term{t_{n}} : \simsub{A_{n, j}}{\rho}} {\Theta \mid \Gamma \vdash \term{\op{j}(\rho; t_1, \ldots, t_{n_j})} : \simsub{A_{0,j}}{\rho}}\;(\text{op})
     \quad \text{where $\rho \colon \Xi_j \to \Theta$}
@@
 
-\begin{figure}
-  \centering
-  \begin{mathpar}
-    \boxed{\Theta \mid \Gamma \vdash \term{t} \Rightarrow A} \quad\text{and}\quad\boxed{\Theta \mid \Gamma \vdash \term{t} \Leftarrow A} \\
-    \inferrule{\Gamma \ni (\term{x} : A)}{\Theta \mid \Gamma \vdash \term{x} \mathrel{\color{red}\Rightarrow} A}\;(\text{var}{\Rightarrow})
-    \and
-    \inferrule{\Theta\mid \Gamma \vdash \term{t} \Rightarrow B \\ A = B}{\Theta \mid \Gamma \vdash (\term{t} : B) \Rightarrow A}\;(\textcolor{red}{anno})
-    \and
-    \inferrule{\Theta\mid \Gamma \vdash \term{t} \Rightarrow B \\ A = B}{\Theta \mid \Gamma \vdash \term{t\uparrow} \Leftarrow A}\;(\textcolor{red}{sub})
-    \and
-    \inferrule{\Theta \mid \Gamma, \simsub{\vec{x}_{1,j} : \Delta_{1, j}}{\rho} \vdash \term{t_{1, j}} \mathrel{\color{red}\leftrightarrows_{1, j}} \simsub{A_{1, j}}{\rho} \\ \cdots \\ \Theta \mid \Gamma, \simsub{\vec{x}_{n_j} : \Delta_{n_j, j}}{\rho} \vdash \term{t_{n_j, j}} \mathrel{\color{red}\leftrightarrows_{n_j}} \simsub{A_{n_j ,j}}{\rho}}
-    {\Theta \mid \Gamma \vdash \term{\op{j, \rho}(\vec{x}_1 .\, t_1, \ldots, \vec{x}_{n_j}.\, t_{n_j})} \mathrel{\color{red}\leftrightarrows_{0, j}} \simsub{A_{0, j}}{\rho}} \; (\text{op}{\leftrightarrows})
-    \quad \text{where $\rho \colon \Xi_i \to \Theta$}
-  \end{mathpar}
-  \caption{Bidirectional typing rules}
-\end{figure}
-
 \begin{example}
@@
 \subsection{Bidirectionally Typed Terms: Intrinsic Typing vs.\ Extrinsic Typing}
+
 \subsubsection{Intrinsically Typed Terms}
@@
 \end{figure}
+
 \subsubsection{Extrinsically Typed Terms}
+
+\begin{figure}
+  \centering
+  \begin{mathpar}
+    \boxed{\Theta \mid \Gamma \vdash \term{t} \Rightarrow A} \quad\text{and}\quad\boxed{\Theta \mid \Gamma \vdash \term{t} \Leftarrow A} \quad
+    \text{where $\Theta \vdash \Gamma$ and $\Theta \vdash A$}\\
+    \inferrule{\Gamma \ni (\term{x} : A)}{\Theta \mid \Gamma \vdash \term{x} \mathrel{\color{red}\Rightarrow} A}\;(\text{var}{\Rightarrow})
+    \and
+    \inferrule{\Theta\mid \Gamma \vdash \term{t} \Rightarrow B \\ A = B}{\Theta \mid \Gamma \vdash (\term{t} : B) \Rightarrow A}\;(\textcolor{red}{anno})
+    \and
+    \inferrule{\Theta\mid \Gamma \vdash \term{t} \Rightarrow B \\ A = B}{\Theta \mid \Gamma \vdash \term{t\uparrow} \Leftarrow A}\;(\textcolor{red}{sub})
+    \and
+    \inferrule{\Theta \mid \Gamma, \simsub{\vec{x}_{1,j} : \Delta_{1, j}}{\rho} \vdash \term{t_{1, j}} \mathrel{\color{red}\leftrightarrows_{1, j}} \simsub{A_{1, j}}{\rho} \\ \cdots \\ \Theta \mid \Gamma, \simsub{\vec{x}_{n_j, j} : \Delta_{n_j, j}}{\rho} \vdash \term{t_{n_j, j}} \mathrel{\color{red}\leftrightarrows_{n_j, j}} \simsub{A_{n_j ,j}}{\rho}}
+    {\Theta \mid \Gamma \vdash \term{\op{j}(\rho; \vec{x}_{1, j} .\, t_{1, j}, \ldots, \vec{x}_{n_j, j}.\, t_{n_j, j})} \mathrel{\color{red}\leftrightarrows_{0, j}} \simsub{A_{0, j}}{\rho}} \; (\text{op}{\leftrightarrows})
+    \quad \text{where $\rho \colon \Xi_i \to \Theta$}
+  \end{mathpar}
+  \caption{Bidirectional typing rules}
+\end{figure}
 
@@
 
-\section{Formal Implementation and Examples}
+\section{Formal Implementation with Examples}
 
@@
 
-\section{Concluding Remarks and Future Work}
+\section{Future Work}
 \begin{acks}

### 2023-05-24 tex/BiSig.tex
[ fix ] The premise of the annotation rule
[ update ] Simplified notations

--- 2023-05-23 tex/BiSig.tex
@@
 
+\usepackage{bm}
+\usepackage{braket}
 \usepackage{mathtools}
@@
 
-\input{macros}
-
 \FXRegisterAuthor{LT}{LTan}{L-T}
@@
 \citestyle{acmauthoryear}
+
+\AtEndPreamble{%
+\theoremstyle{acmdefinition}
+\newtheorem{remark}[theorem]{Remark}}
+
+\input{type-notation} %% for type theory notation copied from HoTT Book
+\input{macros}
 
@@
 \maketitle
-\section{Introduction}
+\section{Introduction}\label{sec:intro}
 
@@
 
-\section{Bidirectional simply typed \texorpdfstring{$\lambda$}{lambda}-calculus}
-\subsection{Type Checking via Bidirectional Typing}
+\section{Bidirectional simply typed \texorpdfstring{$\lambda$}{lambda}-calculus}\label{sec:bi-stlc}
+\begin{figure}
+  \begin{mathpar}
+    \boxed{\Gamma \vdash \isTerm{t} \Rightarrow A}
+    \quad\text{and}\quad \boxed{\Gamma \vdash \isTerm{t} \Leftarrow A}
+    \\
+    \inferrule{(x : A) \in \Gamma}{\Gamma \vdash \isTerm{x} \Rightarrow A}\;(\text{var}^\Rightarrow)
+    \and
+    \inferrule{\Gamma \vdash \isTerm{t} \Leftarrow B \\ A = B}{\Gamma \vdash \isTerm{t : B}\Rightarrow A}\;(\text{anno})
+    \and
+    \inferrule{\Gamma \vdash \isTerm{t} \Rightarrow B \\ A = B}{\Gamma \vdash \isTerm{t\uparrow} \Leftarrow A}\;(\text{sub})
+    \and
+    \inferrule{\Gamma \vdash \isTerm{t} \Rightarrow A \bto B \\ \Gamma \vdash \isTerm{u} \Leftarrow A}{\Gamma \vdash \isTerm{t\;u} \Rightarrow B}
+    \and
+    \inferrule{\Gamma, \isTerm{x} : A \vdash \isTerm{t} \Rightarrow B}{\Gamma \vdash \isTerm{\lam{x}t} \Rightarrow A \bto B}
+  \end{mathpar}
+  \caption{The extrinsic style of bidirectional simply typed $\lambda$-calculus}
+  \label{fig:bi-stlc}
+\end{figure}
+\begin{remark}
+  We avoid using the term `function type' and its conventional notation $\to$ at the object level on purpose, as it may be confused with function types in our type-theoretic meta-language.
+\end{remark}
 
-\section{A Definition of Simple Type Theories}
+\subsection{Type Synthesis and Checking via Bidirectional Typing}
+\subsection{Formalisation in \Agda}
+
+\section{A Definition of Simple Type Theories}\label{sec:simple-type-theories}
 \subsection{Simple Types}
 \begin{definition}
-  A \emph{(simple) type signature} is a set $I$ with an \emph{arity} $\mathit{ar}_i : \mathbb{N}$ associated to each $i : I$.
-  An element $i : I$ is meant to be the \emph{$i$-th operation} $\tyop{i}$ and $\mathit{ar}(i)$ the \emph{number} of arguments of $\tyop{i}$.
+  A \emph{(simple type) signature} $\Sigma$ consists of a set $I$ and an \emph{arity} function $\arity\colon I \to \mathbb{N}$.
+  An inhabitant $i : I$ is meant to represent the \emph{$i$-th operation} $\tyOp_i$ and its \emph{number} $ \arity(i)$ of arguments.
+
+  The judgements of \emph{$\Sigma$-types} and \emph{$\Sigma$-contexts} over a variable set $\Theta$ are defined inductively by \Cref{fig:simple-type,fig:simple-context} respectively.
+We write $A : \Type_{\Sigma}(\Theta)$ for $\Theta \vdash_{\Sigma} A$ and $\Gamma \colon \Cxt_{\Sigma}(\Theta)$ for $\Theta \vdash_{\Sigma} \Gamma$ to disambiguate $\Sigma$-types and $\Sigma$-contexts in the case of any confusion.
 \end{definition}
-\LTnote{Define $\mathsf{Type}$}
+
 \begin{figure}
-  \begin{subfigure}{.5\textwidth}
+  \begin{minipage}[b]{.55\textwidth}
+    \centering
+    \small
     \begin{mathpar}
-      \boxed{\Theta\vdash A} \\
-      \inferrule{\Theta \ni X_i}{\Theta \vdash X_i} \and
-      \inferrule{\Theta \vdash A_1 \\ \cdots \\ \Theta \vdash A_n}{\Theta \vdash \tyop{i}(A_1, \ldots, A_n)}
+      \boxed{\Theta\vdash_{\Sigma} A} \\
+      \inferrule{\Theta \ni X_i}{\Theta \vdash_\Sigma X_i} \and
+      \inferrule{\Theta \vdash_{\Sigma} A_1 \\ \cdots \\ \Theta \vdash_{\Sigma} A_n}{\Theta \vdash_{\Sigma} \tyOp_i(A_1, \ldots, A_n)}
     \end{mathpar}
-  \end{subfigure}
-  \begin{subfigure}{.4\textwidth}
+    where $n = \arity_i$
+    \caption{Type formation}
+    \label{fig:simple-type}
+  \end{minipage}
+  \begin{minipage}[b]{.4\textwidth}
+    \centering
+    \small
     \begin{mathpar}
-      \boxed{\Theta \vdash \Gamma} \\
-      \inferrule{ }{\Theta \vdash \emptyset}\and
-      \inferrule{\Theta \vdash A \\ \Theta\vdash \Gamma}{\Theta \vdash \Gamma, A}
+      \boxed{\Theta \vdash_{\Sigma} \Gamma} \\
+      \inferrule{ }{\Theta \vdash_{\Sigma} \cdot }\and
+      \inferrule{\Theta \vdash_{\Sigma} A \\ \Theta\vdash_{\Sigma} \Gamma}{\Theta \vdash_{\Sigma} \Gamma, A}
     \end{mathpar}
-  \end{subfigure}
-  \caption{Simple types and contexts}
+    \caption{Context formation}
+  \label{fig:simple-context}
+  \end{minipage}
 \end{figure}
 
-\begin{example}[STLC]
+\begin{example}\label{ex:implication}
+  Simply typed $\lambda$-calculus typically has a binary implication type $\bto$ and some base types so that the set of all types is at least inhabited.
+  In our setting, the set of $\Sigma$-types over $\Theta$ may have variables from $\Theta$, so $\Type_{\Sigma}(\Theta)$ is inhabited if $\Theta$ is.
+  Hence, there is no need to introduce base types which are not introduced by any term construct.
+  The type signature $\Sigma_{\bto}$ of simply typed $\lambda$-calculus is thus given solely by an operation $\bto$ with $\arity(\bto) = 2$, i.e.\ a binary operation.
+  For simply typed $\lambda$-calculus with binary products, we can augment
+  $\Sigma_{\bto}$ by adding a binary operation $\btimes$ for the binary product type.
+  We denote this augmented signature by $\Sigma_{\bto, \btimes}$.
 
@@
 
-\subsection{Binding Signatures and Typed Terms}
-\begin{definition}
-  For a type signature, a \emph{term signature} consists of a set $J$ meant to be the set of term constructs $\op{j}$ with a triple $\left(\Xi_j, A_{0, j}, \left[\left(\Delta_{1, j}, A_{1, j}\right), \ldots, \left(\Delta_{n_j, j}, A_{n_j, j}\right) \right]\right)$ for each $j:J$ where
+\subsection{Binding Signatures and Intrinsically Typed Terms}
+\begin{definition}\label{def:binding-signature}
+  For a type signature $\Sigma$, a \emph{binding signature} $\Omega$ consists of a set $O$ and a function
+  \[
+    \mathit{ar}\colon O \to \sum_{\Xi : \UU} \left(\Cxt_\Sigma(\Xi) \times \Type_\Sigma(\Xi)\right)^* \times \Type_\Sigma(\Xi).
+  \]
+  Each inhabitant $o: O$ is meant to represent a term construct $\termOp_o$ in a simple type theory with a triple $\left(\Xi, \left[\left(\Delta_1; A_1\right), \ldots, \left(\Delta_{n}; A_{n}\right) \right], A\right)$
+  instead of a number as its arity $\arity_o$ where
   \begin{enumerate}
-    \item $A_{0, j} : \mathsf{Type}(\Xi_j)$ is the type of its target and
-    \item $\left[\left(\Delta_{1, j}, A_{1, j}\right), \ldots, \left(\Delta_{n_j, j}, A_{n_j, j}\right) \right]$ is a finite sequence of pairs of a context $\Delta_{k, j} : \mathsf{Cxt}(\Xi_j)$ for types of its binding variables and its type $A_{k, j} : \mathsf{Type}(\Xi_j)$.
+    \item $A : \Type_\Sigma(\Xi)$ is the target type of $\termOp_o$ and
+    \item $\left[\left(\Delta_1; A_{1}\right), \ldots, \left(\Delta_{n}; A_{n}\right) \right]$ is a list of pairs where
+    \item the $i$-th pair $(\Delta_i, A_i) :\Cxt_{\Sigma}(\Xi) \times \Term_{\Sigma}(\Xi)$ are types of the binding variables and the type of the $i$-th argument of $\termOp_{o}$.
   \end{enumerate}
+  For brevity, we write $o \colon \Xi \vdash (\Delta_1)A_{1}, \ldots, \left[\Delta_{n}\right] A_{n} \to A$ to indicate an operation $o$ with its arity. 
 \end{definition}
+\begin{remark}[Terminology on binding signature]
+  \cite{Aczel1978,Fiore2010}
+\end{remark}
+
+\begin{example} \label{ex:STLC-sig}
+  Given the type signature $\Sigma_{\bto}$ for the implication type, the term signature $\Lambda_{\bto}$ for simply typed $\lambda$-calculus can be described by operations
+  \begin{align*}
+    \mathsf{app}\colon A, B \vdash (A \bto B), A \to B && \mathsf{abs}\colon A , B \vdash [A]B \to (A \bto B)
+  \end{align*}
+    or, verbosely, the signature $\Lambda_{\bto}$ consists of the type $O_\Lambda = \{\app, \abs\}$ with arities
+  \begin{align*}
+    \arity(\app) = (\{A, B\}, [(\cdot; A \bto B), (\cdot; A)], B)
+    && 
+    \arity(\abs) = (\{A, B\}, [(\cdot, A; B)], A \bto B)
+  \end{align*}
+\end{example}
+
+\begin{example}\label{ex:STLCx-sig}
+  For the type signature $\Sigma_{\bto, \btimes}$, the term signature $\Omega_{\bto, \btimes}$ for simply typed $\lambda$-calculus with finite products has three additional operations
+to $\Omega_{\bto}$:
+  \begin{align*}
+    \mathsf{pair}\colon A, B \vdash A, B \to A \btimes B
+    && \fst \colon A, B \vdash A \btimes B \to A
+    && \snd \colon A, B \vdash A \btimes B \to B
+  \end{align*}
+\end{example}
+
+\begin{figure}
+  \centering
+  \small
+  \begin{mathpar}
+    \boxed{\Theta \mid \Gamma_{\Sigma, \Omega} \vdash A} \quad \text{where $\Theta \vdash_{\Sigma} \Gamma$ and $\Theta \vdash_{\Sigma} A$} \\
+    \inferrule{A \in \Gamma}{\Theta \mid \Gamma \vdash_{\Sigma, \Omega} A}\;(\var)
+    \and
+    \inferrule{\rho : \Xi \to \Theta \\\Theta \mid \Gamma, \simsub{\Delta_{1}}{\rho} \vdash_{\Sigma, \Omega} \simsub{A_{1}}{\rho} \\ \cdots \\ \Theta \mid \Gamma, \simsub{\Delta_{n}}{\rho} \vdash_{\Sigma, \Omega} \simsub{A_{n}}{\rho}}
+    {\Theta \mid \Gamma \vdash_{\Sigma, \Omega} \simsub{A}{\rho}}\;(\termOp_o)
+  \end{mathpar}
+  for $o\colon \Xi \vdash [\Delta_1]A_1, \ldots, [\Delta_{n}]A_{n} \to A$ in $\Omega$
+  \caption{Intrinsic typing rules}
+\end{figure}
+
+\subsection{Raw Terms and Extrinsically Typed Terms?}
+\begin{LTannote}{no extrinsic typing}
+Argue that we don't need extrinsically typed terms but raw terms only as the output of some parser.
+\end{LTannote}
+
 \begin{figure}
@@
   \begin{mathpar}
-    \boxed{\Theta \mid \Gamma \vdash t : A} \\
-    \inferrule{\Gamma \ni (x : A)}{\Theta \mid \Gamma \vdash \term{x} : A}\;(\text{var})
-    \and
-    \inferrule{\Theta \mid \Gamma, \simsub{\Delta_{1,j}}{\rho} \vdash \term{t_1} : \simsub{A_{1,j}}{\rho} \\ \cdots \\ \Theta \mid \Gamma, \simsub{\Delta_{n_j, j}}{\rho} \vdash \term{t_{n}} : \simsub{A_{n, j}}{\rho}} {\Theta \mid \Gamma \vdash \term{\op{j, \rho}(t_1, \ldots, t_{n_j})} : \simsub{A_{0,j}}{\rho}}\;(\text{op})
-    \quad \text{where $\rho \colon \Xi_j \to \Theta$}
+    \boxed{\Theta \vdash_{\Sigma, \Omega} t}
   \end{mathpar}
 
-  \caption{Typing rules}
+  \caption{Raw terms}
 \end{figure}
-
-\begin{example}[STLC]
-
-\end{example}
 
@@
 \begin{definition}
+  For a type signature $\Sigma$, a \emph{bidirectional binding signature} $\Omega$ is a set $O$ with a function
+  \[
+    \mathit{ar}\colon O \to \sum_{\Xi : \UU} \left(\Cxt_{\Sigma}(\Xi) \times \Mode \times \Type_{\Sigma}(\Xi)\right)^* \times \Mode \times \Type_{\Sigma}(\Xi).
+  \]
+  Bidirectional binding signatures are just binding signatures (\Cref{def:binding-signature}) augmented with a mode for each argument and its target of a construct $\termOp_o$ in a bidirectional system, i.e.
+  an arity is a $4$-tuple
+  \[
+    \left(\Xi, \left[\left(\Delta_1, \leftrightarrows_1, A_1\right), \ldots, \left(\Delta_{n}, \leftrightarrows_n, A_{n}\right) \right], \leftrightarrows, A\right)
+  \]
+  where ${\leftrightarrows_i}$ and ${\leftrightarrows}$ are of type $\Mode$ indicating the modes of a construct and its arguments respectively.
+  For brevity, we write $o \colon \Xi \vdash [\Delta_1]A_{1}^{\leftrightarrows_1}, \ldots, [\Delta_{n}] A^{\leftrightarrows_n}_{n} \to A^{\leftrightarrows}$ to indicate an operation $o$ with its arity.
+\end{definition}
 
-\end{definition}
+\begin{example}
+  With the type signature $\Sigma_{\bto}$, the term signature $\Lambda_{\bto}^{\leftrightarrows}$ for bidirectional simply typed $\lambda$-calculus introduced in \Cref{sec:bi-stlc} can be specified by operations
+  \begin{align*}
+    \mathsf{app}\colon A, B \vdash (A \bto B)^{\Rightarrow}, A^{\Leftarrow} \to B^{\Rightarrow} &&
+    \mathsf{abs}\colon A , B \vdash [A]B^{\Leftarrow} \to (A \bto B)^{\Leftarrow}
+  \end{align*}
+  like those in \Cref{ex:STLC-sig} but augmented with the mode information.
+\end{example}
+
+\subsection{Bidirectionally Typed Terms: Intrinsic Typing vs.\ Extrinsic Typing}
+
+\subsubsection{Intrinsically Typed Terms}
+
+\subsubsection{Bidirectional Raw Terms}
 
@@
   \begin{mathpar}
-    \boxed{\Theta \mid \Gamma \vdash \term{t} \Rightarrow A} \quad\text{and}\quad\boxed{\Theta \mid \Gamma \vdash \term{t} \Leftarrow A} \\
-    \inferrule{\Gamma \ni (\term{x} : A)}{\Theta \mid \Gamma \vdash \term{x} \mathrel{\color{red}\Rightarrow} A}\;(\text{var}{\Rightarrow})
-    \and
-    \inferrule{\Theta\mid \Gamma \vdash \term{t} \Rightarrow B \\ A = B}{\Theta \mid \Gamma \vdash (\term{t} : B) \Rightarrow A}\;(\textcolor{red}{anno})
-    \and
-    \inferrule{\Theta\mid \Gamma \vdash \term{t} \Rightarrow B \\ A = B}{\Theta \mid \Gamma \vdash \term{t\uparrow} \Leftarrow A}\;(\textcolor{red}{sub})
-    \and
-    \inferrule{\Theta \mid \Gamma, \simsub{\vec{x}_{1,j} : \Delta_{1, j}}{\rho} \vdash \term{t_{1, j}} \mathrel{\color{red}\leftrightarrows_{1, j}} \simsub{A_{1, j}}{\rho} \\ \cdots \\ \Theta \mid \Gamma, \simsub{\vec{x}_{n_j} : \Delta_{n_j, j}}{\rho} \vdash \term{t_{n_j, j}} \mathrel{\color{red}\leftrightarrows_{n_j}} \simsub{A_{n_j ,j}}{\rho}}
-    {\Theta \mid \Gamma \vdash \term{\op{j, \rho}(\vec{x}_1 .\, t_1, \ldots, \vec{x}_{n_j}.\, t_{n_j})} \mathrel{\color{red}\leftrightarrows_{0, j}} \simsub{A_{0, j}}{\rho}} \; (\text{op}{\leftrightarrows})
-    \quad \text{where $\rho \colon \Xi_i \to \Theta$}
+    \boxed{\Theta \vdash_{\Sigma, \Omega} t \Rightarrow}
+    \quad\text{and}\quad
+    \boxed{\Theta \vdash_{\Sigma, \Omega} t \Leftarrow}
   \end{mathpar}
-  \caption{Bidirectional typing rules}
+
+  \caption{Bidirectional raw terms}
 \end{figure}
 
-\begin{example}
-
-\end{example}
-
-\subsection{Bidirectionally Typed Terms: Intrinsic Typing vs.\ Extrinsic Typing}
-\subsubsection{Intrinsically Typed Terms}
-
-\subsubsection{Raw Terms}
+\subsubsection{Extrinsically Typed Terms}
 
 \begin{figure}
-  \caption{Bidirectional raw terms}
+  \centering
+  \small
+  \begin{mathpar}
+    \boxed{\Theta \mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Rightarrow A}\quad\text{and}\quad\boxed{\Theta \mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Leftarrow A} 
+    \quad \text{where $\Theta \vdash_{\Sigma} \Gamma$, $\Theta \vdash_{\Sigma} A$, and $\Theta \vdash_{\Sigma, \Omega} t \leftrightarrows$} 
+    \\
+    \inferrule{(x : A) \in \Gamma}{\Theta \mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{x} \mathrel{\color{red}\Rightarrow} A}\;(\var^{\Rightarrow})
+    \and
+    \inferrule{\Theta\mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Leftarrow B \\ A = B}{\Theta \mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t : B}\Rightarrow A}\;(\textcolor{red}{anno})
+    \and
+    \inferrule{\Theta\mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Rightarrow B \\ A = B}{\Theta \mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t\uparrow} \Leftarrow A}\;(\textcolor{red}{sub})
+    \and
+    \inferrule{\rho\colon \Xi \to \Theta \\
+      \Theta \mid \Gamma, \simsub{\isTerm{\vec{x}_{1}} : \Delta_{1}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_{1}} \mathrel{\color{red}\leftrightarrows_{1}} \simsub{A_{1}}{\rho} \\
+      \cdots \\
+    \Theta \mid \Gamma, \simsub{\isTerm{\vec{x}_{n}} : \Delta_{n}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_n} \mathrel{\color{red}\leftrightarrows_{n}} \simsub{A_{n}}{\rho}}
+    {\Theta \mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{\termOp_o(\vec{x}_1 .\, t_{1}; \ldots; \vec{x}_{n}.\, t_{n})} \mathrel{\color{red}\leftrightarrows} \simsub{A}{\rho}} \; (\termOp^{\leftrightarrows}_o)
+  \end{mathpar}
+  for $o \colon \Xi \vdash [\Delta_1]A_{1}^{\leftrightarrows_1}, \ldots, [\Delta_{n}] A^{\leftrightarrows_n}_{n} \to A^{\leftrightarrows}$ in $\Omega$
+  \caption{Bidirectional extrinsic typing rules}
 \end{figure}
-\subsubsection{Extrinsically Typed Terms}
 
@@
 \section{Bidirectional Type Synthesis and Checking}
+
 \subsection{Substitution Properties}
@@
 \cite{McBride2003,McBride2003a}
+\begin{definition}
+A \emph{decidable equality} for a type $X$ is a dependent function $\prod_{x, y : X} \left(\id{x}{y}\right) + \neg \left(\id{x}{y}\right)$.
+\end{definition}
+
 \subsection{Uniqueness of Synthesis}

### 2023-05-24 tex/BiSig.tex
[ fix ] Substitution in (op) and the rule scheme

@@
   \begin{mathpar}
-    \boxed{\Gamma \vdash \isTerm{t} \Rightarrow A}
-    \quad\text{and}\quad \boxed{\Gamma \vdash \isTerm{t} \Leftarrow A}
+    \boxed{\Theta \mid \Gamma \vdash \isTerm{t} \Rightarrow A}
+    \quad\text{and}\quad \boxed{\Theta \mid \Gamma \vdash \isTerm{t} \Leftarrow A}
+    \quad\text{where $\Theta \vdash \Gamma$ and $\Theta \vdash A$} 
     \\
-    \inferrule{(x : A) \in \Gamma}{\Gamma \vdash \isTerm{x} \Rightarrow A}\;(\text{var}^\Rightarrow)
+    \inferrule{(x : A) \in \Gamma}{\Theta \mid \Gamma \vdash \isTerm{x} \Rightarrow A}\;(\text{var}^\Rightarrow)
     \and
-    \inferrule{\Gamma \vdash \isTerm{t} \Leftarrow B \\ A = B}{\Gamma \vdash \isTerm{t : B}\Rightarrow A}\;(\text{anno})
+    \inferrule{\Theta \mid \Gamma \vdash \isTerm{t} \Leftarrow B \\ A = B}{\Theta \mid \Gamma \vdash \isTerm{t : B}\Rightarrow A}\;(\text{anno})
     \and
-    \inferrule{\Gamma \vdash \isTerm{t} \Rightarrow B \\ A = B}{\Gamma \vdash \isTerm{t\uparrow} \Leftarrow A}\;(\text{sub})
+    \inferrule{\Theta \mid \Gamma \vdash \isTerm{t} \Rightarrow B \\ A = B}{\Theta \mid \Gamma \vdash \isTerm{t\uparrow} \Leftarrow A}\;(\text{sub})
     \and
-    \inferrule{\Gamma \vdash \isTerm{t} \Rightarrow A \bto B \\ \Gamma \vdash \isTerm{u} \Leftarrow A}{\Gamma \vdash \isTerm{t\;u} \Rightarrow B}
+    \inferrule{\Theta \mid \Gamma \vdash \isTerm{t} \Rightarrow A \bto B \\ \Theta \mid \Gamma \vdash \isTerm{u} \Leftarrow A}{\Theta \mid \Gamma \vdash \isTerm{t\;u} \Rightarrow B}
     \and
-    \inferrule{\Gamma, \isTerm{x} : A \vdash \isTerm{t} \Rightarrow B}{\Gamma \vdash \isTerm{\lam{x}t} \Rightarrow A \bto B}
+    \inferrule{\Theta \mid \Gamma, \isTerm{x} : A \vdash \isTerm{t} \Rightarrow B}{\Theta \mid \Gamma \vdash \isTerm{\lam{x}t} \Rightarrow A \bto B}
   \end{mathpar}
-  \caption{The extrinsic style of bidirectional simply typed $\lambda$-calculus}
+  \caption{The extrinsic style of bidirectional simply typed $\lambda$-calculus with type variables}
   \label{fig:bi-stlc}
@@
 \begin{example}\label{ex:implication}
-  Simply typed $\lambda$-calculus typically has a binary implication type $\bto$ and some base types so that the set of all types is at least inhabited.
-  In our setting, the set of $\Sigma$-types over $\Theta$ may have variables from $\Theta$, so $\Type_{\Sigma}(\Theta)$ is inhabited if $\Theta$ is.
-  Hence, there is no need to introduce base types which are not introduced by any term construct.
-  The type signature $\Sigma_{\bto}$ of simply typed $\lambda$-calculus is thus given solely by an operation $\bto$ with $\arity(\bto) = 2$, i.e.\ a binary operation.
-  For simply typed $\lambda$-calculus with binary products, we can augment
-  $\Sigma_{\bto}$ by adding a binary operation $\btimes$ for the binary product type.
-  We denote this augmented signature by $\Sigma_{\bto, \btimes}$.
-
+  Simply typed $\lambda$-calculus typically includes a binary implication type denoted by $\bto$ and some base types to ensure that the set of all types is non-empty.
+  In our setting, the set of $\Sigma$-types over $\Theta$ may contain variables from $\Theta$, so $\Type_{\Sigma}(\Theta)$ is non-empty if $\Theta$ is.
+  Therefore, there is no need to introduce base types that are not used by any term construct.
+  The type signature $\Sigma_{\bto}$ of simply typed $\lambda$-calculus consists solely of a binary operation $\bto$, where $\arity(\bto) = 2$.
+  In the case of simply typed $\lambda$-calculus with binary products, we can extend $\Sigma_{\bto}$ by adding a binary operation $\btimes$ to represent the binary product type.
+  This extended signature is denoted as $\Sigma_{\bto, \btimes}$.
 \end{example}
 
-\subsection{Binding Signatures and Intrinsically Typed Terms}
+\subsection{Binding Signatures}
 \begin{definition}\label{def:binding-signature}
@@
   \end{align*}
-\end{example}
-
-\begin{example}\label{ex:STLCx-sig}
   For the type signature $\Sigma_{\bto, \btimes}$, the term signature $\Omega_{\bto, \btimes}$ for simply typed $\lambda$-calculus with finite products has three additional operations
@@
 
+First, it is noteworthy that the set of term constructs in a type theory need not be finite.
+For instance, a type theory that includes ground types like arbitrary-bit integers can be specified by a nullary operation, i.e.\ a constant, for each integer.
+
+More importantly, the inclusion of the set of variables used in an operation is a salient feature of our binding signatures.
+Instead of treating the application construct $\app_{A, B}$ as indexed by all types $A$ and $B$, as done by \citet{Fiore2022}, we are able to identify them as a single construct $\app$.
+This is not only for brevity but also necessary to incorporate the global context of type variables available, which is crucial for unification to solve constraints during type synthesis.
+
+In contrast to the set $\Theta$ representing the context of global type variables for a derivation of a typing judgement $\Theta \mid \Gamma \vdash_{\Sigma, \Omega} A$ introduced below, the set $\Xi$ associated with an operation can be referred to as the \emph{local context} for that operation.
+
+\subsection{Intrinsically Typed Terms}
+
+Intrinsically typed terms for a type signature $\Sigma$ and a term signature $\Omega$ are nothing more than derivations of the intrinsic typing judgement $\boxed{\Theta \mid \Gamma \vdash_{\Sigma, \Omega} A}$ constructed with only the rule $(\var)$ and the rule scheme $(\termOp_o)$ for every operation $o$ in $\Omega$ displayed in \Cref{fig:intrinsic-typing}.
+
+The substitution $\rho\colon \Xi \to \Theta$ is used to instantiate explicitly variables in the local context $\Xi$ with global type variables in $\Theta$.
+Accordingly, $\rho$ has to be applied to types that appear in the arity to construct a term by $\termOp_o$ to ensure that all types are will-formed under $\Theta$.
+
 \begin{figure}
@@
   \begin{mathpar}
-    \boxed{\Theta \mid \Gamma_{\Sigma, \Omega} \vdash A} \quad \text{where $\Theta \vdash_{\Sigma} \Gamma$ and $\Theta \vdash_{\Sigma} A$} \\
+    \boxed{\Theta \mid \Gamma \vdash_{\Sigma, \Omega} A} \quad \text{where $\Theta \vdash_{\Sigma} \Gamma$ and $\Theta \vdash_{\Sigma} A$} \\
     \inferrule{A \in \Gamma}{\Theta \mid \Gamma \vdash_{\Sigma, \Omega} A}\;(\var)
     \and
-    \inferrule{\rho : \Xi \to \Theta \\\Theta \mid \Gamma, \simsub{\Delta_{1}}{\rho} \vdash_{\Sigma, \Omega} \simsub{A_{1}}{\rho} \\ \cdots \\ \Theta \mid \Gamma, \simsub{\Delta_{n}}{\rho} \vdash_{\Sigma, \Omega} \simsub{A_{n}}{\rho}}
+    \inferrule*{\rho : \Sub{\Xi}{\Theta} \\ \Theta \mid \Gamma, \simsub{\Delta_{1}}{\rho} \vdash_{\Sigma, \Omega} \simsub{A_{1}}{\rho} \\ \cdots \\ \Theta \mid \Gamma, \simsub{\Delta_{n}}{\rho} \vdash_{\Sigma, \Omega} \simsub{A_{n}}{\rho}}
     {\Theta \mid \Gamma \vdash_{\Sigma, \Omega} \simsub{A}{\rho}}\;(\termOp_o)
@@
   \caption{Intrinsic typing rules}
+  \label{fig:intrinsic-typing}
 \end{figure}
-
 \subsection{Raw Terms and Extrinsically Typed Terms?}
@@
     \and
-    \inferrule{\rho\colon \Xi \to \Theta \\
+    \inferrule{\rho\colon \Sub{\Xi}{\Theta} \\
       \Theta \mid \Gamma, \simsub{\isTerm{\vec{x}_{1}} : \Delta_{1}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_{1}} \mathrel{\color{red}\leftrightarrows_{1}} \simsub{A_{1}}{\rho} \\

### 2023-05-25 tex/BiSig.tex
[ fix ] Use mode-indexing judgement instead of the conventional mutually defined judgements

@@
 \begin{figure}
+  \centering
+  \small
   \begin{mathpar}
@@
 
-\section{A Definition of Simple Type Theories}\label{sec:simple-type-theories}
+\section{Definitions of Simple Type Theories}\label{sec:simple-type-theories}
 \subsection{Simple Types}
@@
 
-Intrinsically typed terms for a type signature $\Sigma$ and a term signature $\Omega$ are nothing more than derivations of the intrinsic typing judgement $\boxed{\Theta \mid \Gamma \vdash_{\Sigma, \Omega} A}$ constructed with only the rule $(\var)$ and the rule scheme $(\termOp_o)$ for every operation $o$ in $\Omega$ displayed in \Cref{fig:intrinsic-typing}.
+Intrinsically typed terms for a type signature $\Sigma$ and a term signature $\Omega$ are nothing more than derivations of the intrinsic typing judgement $\boxed{\Theta \mid \Gamma \vdash_{\Sigma, \Omega} A}$ constructed with only the rule $(\var)$ and the rule scheme $(\termOp)$ displayed in \Cref{fig:intrinsic-typing}.
 
-The substitution $\rho\colon \Xi \to \Theta$ is used to instantiate explicitly variables in the local context $\Xi$ with global type variables in $\Theta$.
-Accordingly, $\rho$ has to be applied to types that appear in the arity to construct a term by $\termOp_o$ to ensure that all types are will-formed under $\Theta$.
+The substitution $\rho\colon \Sub{\Xi}{\Theta}$ is used to instantiate variables in the local context $\Xi$ with global type variables in $\Theta$.
+Accordingly, $\rho$ has to be applied to types that appear in the arity to construct a term by $\termOp_o$ to ensure all types are will-formed under $\Theta$.
 
@@
     \inferrule*{\rho : \Sub{\Xi}{\Theta} \\ \Theta \mid \Gamma, \simsub{\Delta_{1}}{\rho} \vdash_{\Sigma, \Omega} \simsub{A_{1}}{\rho} \\ \cdots \\ \Theta \mid \Gamma, \simsub{\Delta_{n}}{\rho} \vdash_{\Sigma, \Omega} \simsub{A_{n}}{\rho}}
-    {\Theta \mid \Gamma \vdash_{\Sigma, \Omega} \simsub{A}{\rho}}\;(\termOp_o)
+    {\Theta \mid \Gamma \vdash_{\Sigma, \Omega} \simsub{A}{\rho}}\;(\termOp)
   \end{mathpar}
@@
   \centering
+  \small
   \begin{mathpar}
@@
   \]
+  where $\Mode$ consists of $\chk$ for checking and $\syn$ for synthesis.
   Bidirectional binding signatures are just binding signatures (\Cref{def:binding-signature}) augmented with a mode for each argument and its target of a construct $\termOp_o$ in a bidirectional system, i.e.
@@
   \[
-    \left(\Xi, \left[\left(\Delta_1, \leftrightarrows_1, A_1\right), \ldots, \left(\Delta_{n}, \leftrightarrows_n, A_{n}\right) \right], \leftrightarrows, A\right)
+    \left(\Xi, \left[\left(\Delta_1, d_1, A_1\right), \ldots, \left(\Delta_{n}, d_n, A_{n}\right) \right], d, A\right)
   \]
-  where ${\leftrightarrows_i}$ and ${\leftrightarrows}$ are of type $\Mode$ indicating the modes of a construct and its arguments respectively.
-  For brevity, we write $o \colon \Xi \vdash [\Delta_1]A_{1}^{\leftrightarrows_1}, \ldots, [\Delta_{n}] A^{\leftrightarrows_n}_{n} \to A^{\leftrightarrows}$ to indicate an operation $o$ with its arity.
+  where $d$ and $d_i$'s indicate the modes of a construct and its arguments respectively.
+  For brevity, we write $o \colon \Xi \vdash [\Delta_1]A_{1}^{d_1}, \ldots, [\Delta_{n}] A^{d_n}_{n} \to A^{d}$ to indicate an operation $o$ with its arity. 
 \end{definition}
@@
 \begin{example}
-  With the type signature $\Sigma_{\bto}$, the term signature $\Lambda_{\bto}^{\leftrightarrows}$ for bidirectional simply typed $\lambda$-calculus introduced in \Cref{sec:bi-stlc} can be specified by operations
+  The term signature $\Lambda_{\bto}^{\leftrightarrows}$ for bidirectional simply typed $\lambda$-calculus introduced in \Cref{sec:bi-stlc} can be specified by operations 
   \begin{align*}
@@
   \end{align*}
-  like those in \Cref{ex:STLC-sig} but augmented with the mode information.
+  extending $\Lambda_{\bto}$ (\Cref{ex:STLC-sig}) with the mode information.
 \end{example}
@@
   \centering
+  \small
   \begin{mathpar}
-    \boxed{\Theta \vdash_{\Sigma, \Omega} t \Rightarrow}
-    \quad\text{and}\quad
-    \boxed{\Theta \vdash_{\Sigma, \Omega} t \Leftarrow}
+    \boxed{\Theta \vdash_{\Sigma, \Omega} t^d}
   \end{mathpar}
@@
   \begin{mathpar}
-    \boxed{\Theta \mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Rightarrow A}\quad\text{and}\quad\boxed{\Theta \mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Leftarrow A} 
-    \quad \text{where $\Theta \vdash_{\Sigma} \Gamma$, $\Theta \vdash_{\Sigma} A$, and $\Theta \vdash_{\Sigma, \Omega} t \leftrightarrows$} 
+    \boxed{\Theta\mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A^\isDir{d}}
+    \quad \text{where $d$ is either $\chk$ or $\syn$, $\Theta \vdash_{\Sigma} \Gamma$, $\Theta \vdash_{\Sigma} A$, and $\Theta \vdash_{\Sigma, \Omega} t^d$} 
      \\
-    \inferrule{(x : A) \in \Gamma}{\Theta \mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{x} \mathrel{\color{red}\Rightarrow} A}\;(\var^{\Rightarrow})
+    \inferrule{(x : A) \in \Gamma}{\Theta \mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{x} : A^\syn}\;(\var)
     \and
-    \inferrule{\Theta\mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Leftarrow B \\ A = B}{\Theta \mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t : B}\Rightarrow A}\;(\textcolor{red}{anno})
+    \inferrule{\Theta\mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : B^\chk \\ A = B}{\Theta \mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t : B}:  A^\syn}\;(\text{anno})
     \and
-    \inferrule{\Theta\mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Rightarrow B \\ A = B}{\Theta \mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t\uparrow} \Leftarrow A}\;(\textcolor{red}{sub})
+    \inferrule{\Theta\mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : B^{\syn} \\ A = B}{\Theta \mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t\uparrow} : A^\chk}\;(\text{sub})
     \and
+  \end{mathpar}
+  \begin{mathpar}
     \inferrule{\rho\colon \Sub{\Xi}{\Theta} \\
-      \Theta \mid \Gamma, \simsub{\isTerm{\vec{x}_{1}} : \Delta_{1}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_{1}} \mathrel{\color{red}\leftrightarrows_{1}} \simsub{A_{1}}{\rho} \\
+      \Theta \mid \Gamma, \simsub{\isTerm{\vec{x}_{1}} : \Delta_{1}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_{1}} : \simsub{A_{1}}{\rho}^{\isDir{d_1}} \\
       \cdots \\
-    \Theta \mid \Gamma, \simsub{\isTerm{\vec{x}_{n}} : \Delta_{n}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_n} \mathrel{\color{red}\leftrightarrows_{n}} \simsub{A_{n}}{\rho}}
-    {\Theta \mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{\termOp_o(\vec{x}_1 .\, t_{1}; \ldots; \vec{x}_{n}.\, t_{n})} \mathrel{\color{red}\leftrightarrows} \simsub{A}{\rho}} \; (\termOp^{\leftrightarrows}_o)
+    \Theta \mid \Gamma, \simsub{\isTerm{\vec{x}_{n}} : \Delta_{n}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_n} : \simsub{A_{n}}{\rho}^{\isDir{d_n}}}
+  {\Theta \mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{\termOp_o(\vec{x}_1 .\, t_{1}; \ldots; \vec{x}_{n}.\, t_{n})} : \simsub{A}{\rho}^{\isDir{d}}} \; (\termOp)
+    \\ \text{for $o \colon \Xi \vdash [\Delta_1]A_{1}^{d_1}, \ldots, [\Delta_{n}] A^{d_n}_{n} \to A^{d}$ in $\Omega$}
   \end{mathpar}
-  for $o \colon \Xi \vdash [\Delta_1]A_{1}^{\leftrightarrows_1}, \ldots, [\Delta_{n}] A^{\leftrightarrows_n}_{n} \to A^{\leftrightarrows}$ in $\Omega$
+  Abbreviations: $\boxed{\Theta\mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Rightarrow A} \defeq \boxed{\Theta\mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A^\syn}$ and $\boxed{\Theta\mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Leftarrow A} \defeq \boxed{\Theta\mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A^\chk}$
   \caption{Bidirectional extrinsic typing rules}
@@
 
-\subsection{Binding Signatures vs.\ their Bidirectional Variants}
+\subsection{Between Binding Signatures and their Bidirectional Variants}
 \subsubsection{Soundness}
@@
 \section{Formal Implementation and Examples}
+\begin{LTannote}{de Bruijn}
+  In our Agda formalisation, we use de Bruijn indices to represent variables, so we don't use an arbitrary type $\Theta$ but the number of variables as the index of typing judgements.
+\end{LTannote}
 

### 2023-05-25 tex/BiSig.tex
[ update ] Fine-tune typing rules

@@
-\documentclass[acmsmall]{acmart}
+\documentclass[acmsmall,screen]{acmart}
 \usepackage[T1]{fontenc}
@@
 \subsubsection{Algebraic theory of type theories}
-\cite{Fiore1999,Fiore2022,Hamana2022,Ahrens2021,Arkor2020}
+\cite{Fiore1999,Fiore2022,Ahrens2021,Arkor2020,Hamana2011}
 
@@
 More importantly, the inclusion of the set of variables used in an operation is a salient feature of our binding signatures.
-Instead of treating the application construct $\app_{A, B}$ as indexed by all types $A$ and $B$, as done by \citet{Fiore2022}, we are able to identify them as a single construct $\app$.
-This is not only for brevity but also necessary to incorporate the global context of type variables available, which is crucial for unification to solve constraints during type synthesis.
-
-In contrast to the set $\Theta$ representing the context of global type variables for a derivation of a typing judgement $\Theta \mid \Gamma \vdash_{\Sigma, \Omega} A$ introduced below, the set $\Xi$ associated with an operation can be referred to as the \emph{local context} for that operation.
+Instead of treating application as a family of constructs $\app_{A, B}$ indexed by all types $A$ and $B$, as done by \citet{Fiore2022}, we are able to identify them as a single construct $\app$.
+This is not only for brevity but also necessary to compare the type equality during type synthesis and checking.
 
@@
 
-The substitution $\rho\colon \Sub{\Xi}{\Theta}$ is used to instantiate variables in the local context $\Xi$ with global type variables in $\Theta$.
-Accordingly, $\rho$ has to be applied to types that appear in the arity to construct a term by $\termOp_o$ to ensure all types are will-formed under $\Theta$.
+The substitution $\rho\colon \Sub{\Xi}{\Theta}$ is used to instantiate variables in the local context $\Xi$ with types within the context of type variables in $\Theta$.
+Accordingly, $\rho$ has to be applied to types that appear in the arity to construct a term by $\termOp$ to ensure all types are well-formed under $\Theta$.
 
@@
     \and
-    \inferrule*{\rho : \Sub{\Xi}{\Theta} \\ \Theta \mid \Gamma, \simsub{\Delta_{1}}{\rho} \vdash_{\Sigma, \Omega} \simsub{A_{1}}{\rho} \\ \cdots \\ \Theta \mid \Gamma, \simsub{\Delta_{n}}{\rho} \vdash_{\Sigma, \Omega} \simsub{A_{n}}{\rho}}
+    \inferrule{\rho : \Sub{\Xi}{\Theta} \\ \Theta \mid \Gamma, \simsub{\Delta_{1}}{\rho} \vdash_{\Sigma, \Omega} \simsub{A_{1}}{\rho} \quad\cdots\quad \Theta \mid \Gamma, \simsub{\Delta_{n}}{\rho} \vdash_{\Sigma, \Omega} \simsub{A_{n}}{\rho}}
     {\Theta \mid \Gamma \vdash_{\Sigma, \Omega} \simsub{A}{\rho}}\;(\termOp)
+    \and \text{for $o\colon \Xi \vdash [\Delta_1]A_1, \ldots, [\Delta_{n}]A_{n} \to A$ in $\Omega$}
   \end{mathpar}
-  for $o\colon \Xi \vdash [\Delta_1]A_1, \ldots, [\Delta_{n}]A_{n} \to A$ in $\Omega$
   \caption{Intrinsic typing rules}
@@
   where $d$ and $d_i$'s indicate the modes of a construct and its arguments respectively.
+
   For brevity, we write $o \colon \Xi \vdash [\Delta_1]A_{1}^{d_1}, \ldots, [\Delta_{n}] A^{d_n}_{n} \to A^{d}$ to indicate an operation $o$ with its arity. 
@@
   \begin{mathpar}
-    \boxed{\Theta \vdash_{\Sigma, \Omega} t^d}
+    \boxed{\Theta \vdash_{\Sigma, \Omega} \isTerm{t}^\isDir{d}}
   \end{mathpar}
@@
     \inferrule{\Theta\mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : B^{\syn} \\ A = B}{\Theta \mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t\uparrow} : A^\chk}\;(\text{sub})
-    \and
   \end{mathpar}
@@
   {\Theta \mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{\termOp_o(\vec{x}_1 .\, t_{1}; \ldots; \vec{x}_{n}.\, t_{n})} : \simsub{A}{\rho}^{\isDir{d}}} \; (\termOp)
-    \\ \text{for $o \colon \Xi \vdash [\Delta_1]A_{1}^{d_1}, \ldots, [\Delta_{n}] A^{d_n}_{n} \to A^{d}$ in $\Omega$}
+    \and \text{for $o \colon \Xi \vdash [\Delta_1]A_{1}^{d_1}, \ldots, [\Delta_{n}] A^{d_n}_{n} \to A^{d}$ in $\Omega$}
+    \\
+    \text{Abbreviations: $\boxed{\Theta\mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Rightarrow A} \defeq \boxed{\Theta\mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A^\syn}$ and $\boxed{\Theta\mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Leftarrow A} \defeq \boxed{\Theta\mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A^\chk}$}
   \end{mathpar}
-  Abbreviations: $\boxed{\Theta\mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Rightarrow A} \defeq \boxed{\Theta\mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A^\syn}$ and $\boxed{\Theta\mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Leftarrow A} \defeq \boxed{\Theta\mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A^\chk}$
   \caption{Bidirectional extrinsic typing rules}

### 2023-05-26 tex/BiSig.tex
[ tex ] Attempt to fix bidirectional pre-typed terms

@@
 
-\usepackage{bm}
-\usepackage{braket}
-\usepackage{mathtools}
-\usepackage{mathpartir}
+\usepackage{bm,braket,mathtools,mathpartir,bussproofs}
+\EnableBpAbbreviations
 \usepackage[all,cmtip]{xy}
-\usepackage{bussproofs}
-\EnableBpAbbreviations
 
-\usepackage{xfrac,xspace}
-\usepackage{xcolor}
-\usepackage{subcaption}
+\usepackage{xfrac,xspace,xcolor,subcaption}
 \usepackage{hyperref}
@@
     \and
-    \inferrule{\Theta \mid \Gamma, \isTerm{x} : A \vdash \isTerm{t} \Rightarrow B}{\Theta \mid \Gamma \vdash \isTerm{\lam{x}t} \Rightarrow A \bto B}
+    \inferrule{\Theta \mid \Gamma, \isTerm{x} : A \vdash \isTerm{t} \Leftarrow B}{\Theta \mid \Gamma \vdash \isTerm{\lam{x}t} \Leftarrow A \bto B}
   \end{mathpar}
@@
   \]
-  Each inhabitant $o: O$ is meant to represent a term construct $\termOp_o$ in a simple type theory with a triple $\left(\Xi, \left[\left(\Delta_1; A_1\right), \ldots, \left(\Delta_{n}; A_{n}\right) \right], A\right)$
+  Each inhabitant $o: O$ is meant to represent a term construct $\tmOp_o$ in a simple type theory with a triple $\left(\Xi, \left[\left(\Delta_1; A_1\right), \ldots, \left(\Delta_{n}; A_{n}\right) \right], A\right)$
   instead of a number as its arity $\arity_o$ where
   \begin{enumerate}
-    \item $A : \Type_\Sigma(\Xi)$ is the target type of $\termOp_o$ and
+    \item $A : \Type_\Sigma(\Xi)$ is the target type of $\tmOp_o$ and
     \item $\left[\left(\Delta_1; A_{1}\right), \ldots, \left(\Delta_{n}; A_{n}\right) \right]$ is a list of pairs where
-    \item the $i$-th pair $(\Delta_i, A_i) :\Cxt_{\Sigma}(\Xi) \times \Term_{\Sigma}(\Xi)$ are types of the binding variables and the type of the $i$-th argument of $\termOp_{o}$.
+    \item the $i$-th pair $(\Delta_i, A_i) :\Cxt_{\Sigma}(\Xi) \times \Term_{\Sigma}(\Xi)$ are types of the binding variables and the type of the $i$-th argument of $\tmOp_{o}$.
   \end{enumerate}
@@
 
-Intrinsically typed terms for a type signature $\Sigma$ and a term signature $\Omega$ are nothing more than derivations of the intrinsic typing judgement $\boxed{\Theta \mid \Gamma \vdash_{\Sigma, \Omega} A}$ constructed with only the rule $(\var)$ and the rule scheme $(\termOp)$ displayed in \Cref{fig:intrinsic-typing}.
+Intrinsically typed terms for a type signature $\Sigma$ and a term signature $\Omega$ are nothing more than derivations of the intrinsic typing judgement $\boxed{\Theta \mid \Gamma \vdash_{\Sigma, \Omega} A}$ constructed with only the rule $(\var)$ and the rule scheme $(\tmOp)$ displayed in \Cref{fig:intrinsic-typing}.
 
 The substitution $\rho\colon \Sub{\Xi}{\Theta}$ is used to instantiate variables in the local context $\Xi$ with types within the context of type variables in $\Theta$.
-Accordingly, $\rho$ has to be applied to types that appear in the arity to construct a term by $\termOp$ to ensure all types are well-formed under $\Theta$.
+Accordingly, $\rho$ has to be applied to types that appear in the arity to construct a term by $\tmOp$ to ensure all types are well-formed under $\Theta$.
 
@@
     \and
-    \inferrule{\rho : \Sub{\Xi}{\Theta} \\ \Theta \mid \Gamma, \simsub{\Delta_{1}}{\rho} \vdash_{\Sigma, \Omega} \simsub{A_{1}}{\rho} \quad\cdots\quad \Theta \mid \Gamma, \simsub{\Delta_{n}}{\rho} \vdash_{\Sigma, \Omega} \simsub{A_{n}}{\rho}}
-    {\Theta \mid \Gamma \vdash_{\Sigma, \Omega} \simsub{A}{\rho}}\;(\termOp)
-    \and \text{for $o\colon \Xi \vdash [\Delta_1]A_1, \ldots, [\Delta_{n}]A_{n} \to A$ in $\Omega$}
+    \inferrule{\Theta \mid \Gamma, \simsub{\Delta_{1}}{\rho} \vdash_{\Sigma, \Omega} \simsub{A_{1}}{\rho} \quad\cdots\quad \Theta \mid \Gamma, \simsub{\Delta_{n}}{\rho} \vdash_{\Sigma, \Omega} \simsub{A_{n}}{\rho}}
+    {\Theta \mid \Gamma \vdash_{\Sigma, \Omega} \simsub{A}{\rho}}\;(\tmOp)
+    \and \text{for $o\colon \Xi \vdash [\Delta_1]A_1, \ldots, [\Delta_{n}]A_{n} \to A$ in $\Omega$
+    and $\rho : \Sub{\Xi}{\Theta}$}
   \end{mathpar}
@@
 \end{figure}
-\subsection{Raw Terms and Extrinsically Typed Terms?}
-\begin{LTannote}{no extrinsic typing}
-Argue that we don't need extrinsically typed terms but raw terms only as the output of some parser.
-\end{LTannote}
+
+\subsection{Raw Terms and Extrinsically Typed Terms}
 
@@
   \begin{mathpar}
-    \boxed{\Theta \vdash_{\Sigma, \Omega} t}
+    \boxed{\Theta \vdash_{\Sigma, \Omega} \isTerm{t}}
+    \\
+    \inferrule{x : \Identifier}{\Theta \vdash_{\Sigma, \Omega} \isTerm{x}}\;(\text{var})
+    \and
+    \inferrule{\Theta \vdash A \\ \Theta \vdash_{\Sigma, \Omega}\isTerm{t}}{\Theta \vdash_{\Sigma, \Theta} \isTerm{t \annote A}}\;(\text{anno})
+    \\
+    \inferrule{\Theta \vdash_{\Sigma, \Omega} \isTerm{t_1} \quad \cdots \quad \Theta \vdash_{\Sigma, \Omega} \isTerm{t_n}}
+    {\Theta \vdash_{\Sigma, \Omega} \isTerm{\tmOp_o(t_1, \ldots, t_n)}}\;(\text{op}) 
+    \and \text{for $o \colon \Xi \vdash [\Delta_1]A_{1}^{d_1}, \ldots, [\Delta_{n}] A^{d_n}_{n} \to A^{d}$ in $\Omega$}
   \end{mathpar}
+  \caption{Annotatable raw terms}
+\end{figure}
 
-  \caption{Raw terms}
+\begin{figure}
+  \centering
+  \small
+  \begin{mathpar}
+    \boxed{\Theta \mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A} \quad \text{where $\Theta \vdash_{\Sigma} \Gamma$, $\Theta \vdash_{\Sigma} A$, and $\Theta \vdash_{\Sigma, \Omega} t$} \\
+    \inferrule{(x : A) \in \Gamma}{\Theta \mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{x} : A}\;(\var)
+    \and
+    \inferrule{\Theta \mid \Gamma \vdash \isTerm{t} : A}{\Theta \mid \Gamma \vdash (\isTerm{t \annote A}) : A}\;(\text{anno})
+    \and
+    \inferrule{\Theta \mid \Gamma, \isTerm{\vec{x}_1} : \simsub{\Delta_{1}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_1} : \simsub{A_{1}}{\rho} \quad\cdots\quad \Theta \mid \Gamma, \isTerm{\vec{x}_n} : \simsub{\Delta_{n}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_n} : \simsub{A_{n}}{\rho}}
+    {\Theta \mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{\tmOp_o(\vec{x}_1.\,t_1; \ldots; \vec{x}_n.\,t_n)} : \simsub{A}{\rho}}\;(\tmOp)
+    \and \text{for $o\colon \Xi \vdash [\Delta_1]A_1, \ldots, [\Delta_{n}]A_{n} \to A$ in $\Omega$
+    and $\rho : \Sub{\Xi}{\Theta}$}
+  \end{mathpar}
+  \caption{Extrinsic typing rules}
+  \label{fig:extrinsic-typing}
 \end{figure}
@@
   where $\Mode$ consists of $\chk$ for checking and $\syn$ for synthesis.
-  Bidirectional binding signatures are just binding signatures (\Cref{def:binding-signature}) augmented with a mode for each argument and its target of a construct $\termOp_o$ in a bidirectional system, i.e.
+  Bidirectional binding signatures are just binding signatures (\Cref{def:binding-signature}) augmented with a mode for each argument and its target of a construct $\tmOp_o$ in a bidirectional system, i.e.
   an arity is a $4$-tuple
@@
 
-\subsubsection{Bidirectional Raw Terms}
-
-\begin{figure}
-  \centering
-  \small
-  \begin{mathpar}
-    \boxed{\Theta \vdash_{\Sigma, \Omega} \isTerm{t}^\isDir{d}}
-  \end{mathpar}
-
-  \caption{Bidirectional raw terms}
-\end{figure}
-
 \subsubsection{Extrinsically Typed Terms}
@@
   \begin{mathpar}
+    \boxed{\Theta \vdash_{\Sigma, \Omega} \isTerm{t}\;\isDir{d}}
+    \\
+    \inferrule{x : \Identifier}{\Theta \vdash_{\Sigma, \Omega} \isTerm{x}\;\isDir{\syn}}\;(\text{var})
+    \and
+    \inferrule{\Theta \vdash A \\ \Theta \vdash_{\Sigma, \Omega}\isTerm{t}\;\isDir{\chk}}{\Theta \vdash_{\Sigma, \Theta} \isTerm{t \annote A}\;\isDir{\syn}}\;(\text{anno})
+    \and
+%    \inferrule{\Theta \vdash_{\Sigma, \Omega}\isTerm{t}\;\isDir{\chk}}{\Theta, X \vdash_{\Sigma, \Theta} \isTerm{t \downarrow}\;\isDir{\syn}}\;(\text{anno}')
+%    \and
+    \inferrule{\Theta \vdash_{\Sigma, \Omega} \isTerm{t}\;\isDir{\syn}}{\Theta \vdash_{\Sigma, \Omega} (\isTerm{t \uparrow})\;\isDir{\chk}}
+  \end{mathpar}
+  \begin{mathpar}
+    \inferrule{\rho \colon \Sub{\Xi}{\Theta} \\ \Theta \vdash_{\Sigma, \Omega} \isTerm{t_1}\;\isDir{d_1} \quad \cdots \quad \Theta \vdash_{\Sigma, \Omega} \isTerm{t_n}\;\isDir{d_n}}
+    {\Theta \vdash_{\Sigma, \Omega} \isTerm{\tmOp_o(\rho; \vec{x}_1.\, t_1; \ldots;\vec{x}_n.\, t_n)}\;\isDir{d}}\;(\text{op})
+    \and \text{for $o \colon \Xi \vdash [\Delta_1]A_{1}^{d_1}, \ldots, [\Delta_{n}] A^{d_n}_{n} \to A^{d}$ in $\Omega$}
+  \end{mathpar}
+
+  \caption{Bidirectional pre-typed terms}
+\end{figure}
+\begin{figure}
+  \centering
+  \small
+  \begin{mathpar}
     \boxed{\Theta\mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A^\isDir{d}}
-    \quad \text{where $d$ is either $\chk$ or $\syn$, $\Theta \vdash_{\Sigma} \Gamma$, $\Theta \vdash_{\Sigma} A$, and $\Theta \vdash_{\Sigma, \Omega} t^d$} 
+    \quad \text{where $d$ is either $\chk$ or $\syn$, $\Theta \vdash_{\Sigma} \Gamma$, $\Theta \vdash_{\Sigma} A$, and $\Theta \vdash_{\Sigma, \Omega} t\;d$} 
     \\
@@
     \and
-    \inferrule{\Theta\mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : B^\chk \\ A = B}{\Theta \mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t : B}:  A^\syn}\;(\text{anno})
+    \inferrule{\Theta\mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : B^\chk \\ A = B}{\Theta \mid \Gamma \vdash_{\Sigma, \Omega} (\isTerm{t \annote B}):  A^\syn}\;(\text{anno})
     \and
     \inferrule{\Theta\mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : B^{\syn} \\ A = B}{\Theta \mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t\uparrow} : A^\chk}\;(\text{sub})
-  \end{mathpar}
-  \begin{mathpar}
-    \inferrule{\rho\colon \Sub{\Xi}{\Theta} \\
-      \Theta \mid \Gamma, \simsub{\isTerm{\vec{x}_{1}} : \Delta_{1}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_{1}} : \simsub{A_{1}}{\rho}^{\isDir{d_1}} \\
+    \\
+    \inferrule{\rho\colon \Sub{\Xi}{\Theta} \\ \Theta \mid \Gamma, \simsub{\isTerm{\vec{x}_{1}} : \Delta_{1}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_{1}} : \simsub{A_{1}}{\rho}^{\isDir{d_1}} \\
       \cdots \\
     \Theta \mid \Gamma, \simsub{\isTerm{\vec{x}_{n}} : \Delta_{n}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_n} : \simsub{A_{n}}{\rho}^{\isDir{d_n}}}
-  {\Theta \mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{\termOp_o(\vec{x}_1 .\, t_{1}; \ldots; \vec{x}_{n}.\, t_{n})} : \simsub{A}{\rho}^{\isDir{d}}} \; (\termOp)
+  {\Theta \mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{\tmOp_o(\rho; \vec{x}_1 .\, t_{1}; \ldots; \vec{x}_{n}.\, t_{n})} : \simsub{A}{\rho}^{\isDir{d}}} \; (\tmOp)
     \and \text{for $o \colon \Xi \vdash [\Delta_1]A_{1}^{d_1}, \ldots, [\Delta_{n}] A^{d_n}_{n} \to A^{d}$ in $\Omega$}

### 2023-05-28 tex/BiSig.tex
[ update ] Remove the context of type variables from typing rules

--- 2023-05-25 tex/BiSig.tex
@@
 \subsubsection{Bidirectional typing}
+\cite{Pierce2000}
 \cite{Dunfield2021}
@@
   \begin{mathpar}
-    \boxed{\Theta \mid \Gamma \vdash \isTerm{t} \Rightarrow A}
-    \quad\text{and}\quad \boxed{\Theta \mid \Gamma \vdash \isTerm{t} \Leftarrow A}
-    \quad\text{where $\Theta \vdash \Gamma$ and $\Theta \vdash A$} 
+    \boxed{\Gamma \vdash \isTerm{t} \Rightarrow A}
+    \quad\text{and}\quad \boxed{\Gamma \vdash \isTerm{t} \Leftarrow A}
     \\
-    \inferrule{(x : A) \in \Gamma}{\Theta \mid \Gamma \vdash \isTerm{x} \Rightarrow A}\;(\text{var}^\Rightarrow)
+    \inferrule{(x : A) \in \Gamma}{\Gamma \vdash \isTerm{x} \Rightarrow A}\;(\text{var}^\Rightarrow)
     \and
-    \inferrule{\Theta \mid \Gamma \vdash \isTerm{t} \Leftarrow B \\ A = B}{\Theta \mid \Gamma \vdash \isTerm{t : B}\Rightarrow A}\;(\text{anno})
+    \inferrule{\Gamma \vdash \isTerm{t} \Leftarrow B \\ A = B}{\Gamma \vdash \isTerm{t : B}\Rightarrow A}\;(\text{anno})
     \and
-    \inferrule{\Theta \mid \Gamma \vdash \isTerm{t} \Rightarrow B \\ A = B}{\Theta \mid \Gamma \vdash \isTerm{t\uparrow} \Leftarrow A}\;(\text{sub})
+    \inferrule{\Gamma \vdash \isTerm{t} \Rightarrow B \\ A = B}{\Gamma \vdash \isTerm{t\uparrow} \Leftarrow A}\;(\text{sub})
     \and
-    \inferrule{\Theta \mid \Gamma \vdash \isTerm{t} \Rightarrow A \bto B \\ \Theta \mid \Gamma \vdash \isTerm{u} \Leftarrow A}{\Theta \mid \Gamma \vdash \isTerm{t\;u} \Rightarrow B}
+    \inferrule{\Gamma \vdash \isTerm{t} \Rightarrow A \bto B \\ \Gamma \vdash \isTerm{u} \Leftarrow A}{\Gamma \vdash \isTerm{t\;u} \Rightarrow B}
     \and
-    \inferrule{\Theta \mid \Gamma, \isTerm{x} : A \vdash \isTerm{t} \Rightarrow B}{\Theta \mid \Gamma \vdash \isTerm{\lam{x}t} \Rightarrow A \bto B}
+    \inferrule{\Gamma, \isTerm{x} : A \vdash \isTerm{t} \Leftarrow B}{\Gamma \vdash \isTerm{\lam{x}t} \Leftarrow A \bto B}
   \end{mathpar}
@@
 
+\begin{remark}
+  \begin{LTannote}{$\Xi$ is just a number}
+    We will just use $\Xi : \mathbb{N}$ for the set of variables.
+  \end{LTannote}
+
+\end{remark}
+
 \subsection{Binding Signatures}
@@
   \[
-    \mathit{ar}\colon O \to \sum_{\Xi : \UU} \left(\Cxt_\Sigma(\Xi) \times \Type_\Sigma(\Xi)\right)^* \times \Type_\Sigma(\Xi).
+    \mathit{ar}\colon O \to \sum_{\Xi : \N} \left(\Cxt_\Sigma(\Xi) \times \Type_\Sigma(\Xi)\right)^* \times \Type_\Sigma(\Xi).
   \]
-  Each inhabitant $o: O$ is meant to represent a term construct $\termOp_o$ in a simple type theory with a triple $\left(\Xi, \left[\left(\Delta_1; A_1\right), \ldots, \left(\Delta_{n}; A_{n}\right) \right], A\right)$
+  Each inhabitant $o: O$ is meant to represent a term construct $\tmOp_o$ in a simple type theory with a triple $\left(\Xi, \left[\left(\Delta_1; A_1\right), \ldots, \left(\Delta_{n}; A_{n}\right) \right], A\right)$
   instead of a number as its arity $\arity_o$ where
   \begin{enumerate}
-    \item $A : \Type_\Sigma(\Xi)$ is the target type of $\termOp_o$ and
+    \item $\Xi$ is the number of type variables, 
+    \item $A : \Type_\Sigma(\Xi)$ is the target type of $\tmOp_o$, and
     \item $\left[\left(\Delta_1; A_{1}\right), \ldots, \left(\Delta_{n}; A_{n}\right) \right]$ is a list of pairs where
-    \item the $i$-th pair $(\Delta_i, A_i) :\Cxt_{\Sigma}(\Xi) \times \Term_{\Sigma}(\Xi)$ are types of the binding variables and the type of the $i$-th argument of $\termOp_{o}$.
+    \item the $i$-th pair $(\Delta_i, A_i) :\Cxt_{\Sigma}(\Xi) \times \Term_{\Sigma}(\Xi)$ are types of the binding variables and the type of the $i$-th argument of $\tmOp_{o}$.
   \end{enumerate}
@@
 First, it is noteworthy that the set of term constructs in a type theory need not be finite.
-For instance, a type theory that includes ground types like arbitrary-bit integers can be specified by a nullary operation, i.e.\ a constant, for each integer.
+For instance, a type theory may adopt a spine application which takes indefinitely many arguments---for each number $n$ of arguments an $(n+1)$-ary application construct can be introduced, so even if each operation has a definite number of arguments by definition a spine application is still expressible.
 
@@
 
-Intrinsically typed terms for a type signature $\Sigma$ and a term signature $\Omega$ are nothing more than derivations of the intrinsic typing judgement $\boxed{\Theta \mid \Gamma \vdash_{\Sigma, \Omega} A}$ constructed with only the rule $(\var)$ and the rule scheme $(\termOp)$ displayed in \Cref{fig:intrinsic-typing}.
+Intrinsically typed terms for a type signature $\Sigma$ and a term signature $\Omega$ are nothing more than derivations of the intrinsic typing judgement $\Gamma \vdash_{\Sigma, \Omega} A$ constructed with only the rule $(\var)$ and the rule scheme $(\tmOp)$ displayed in \Cref{fig:intrinsic-typing}.
 
-The substitution $\rho\colon \Sub{\Xi}{\Theta}$ is used to instantiate variables in the local context $\Xi$ with types within the context of type variables in $\Theta$.
-Accordingly, $\rho$ has to be applied to types that appear in the arity to construct a term by $\termOp$ to ensure all types are well-formed under $\Theta$.
+The substitution $\rho\colon \Sub{\Xi}{\emptyset}$ is used to instantiate variables in the local context $\Xi$ with concrete types.
+Accordingly, $\rho$ has to be applied to types that appear in the arity to construct a term by $\tmOp$ to ensure all types are well-formed without any use of type variables.
 
@@
   \begin{mathpar}
-    \boxed{\Theta \mid \Gamma \vdash_{\Sigma, \Omega} A} \quad \text{where $\Theta \vdash_{\Sigma} \Gamma$ and $\Theta \vdash_{\Sigma} A$} \\
-    \inferrule{A \in \Gamma}{\Theta \mid \Gamma \vdash_{\Sigma, \Omega} A}\;(\var)
+    \boxed{\Gamma \vdash_{\Sigma, \Omega} A} \\
+    \inferrule{A \in \Gamma}{\Gamma \vdash_{\Sigma, \Omega} A}\;(\var)
     \and
-    \inferrule{\rho : \Sub{\Xi}{\Theta} \\ \Theta \mid \Gamma, \simsub{\Delta_{1}}{\rho} \vdash_{\Sigma, \Omega} \simsub{A_{1}}{\rho} \quad\cdots\quad \Theta \mid \Gamma, \simsub{\Delta_{n}}{\rho} \vdash_{\Sigma, \Omega} \simsub{A_{n}}{\rho}}
-    {\Theta \mid \Gamma \vdash_{\Sigma, \Omega} \simsub{A}{\rho}}\;(\termOp)
-    \and \text{for $o\colon \Xi \vdash [\Delta_1]A_1, \ldots, [\Delta_{n}]A_{n} \to A$ in $\Omega$}
+    \inferrule{\Gamma, \simsub{\Delta_{1}}{\rho} \vdash_{\Sigma, \Omega} \simsub{A_{1}}{\rho} \quad\cdots\quad \mid \Gamma, \simsub{\Delta_{n}}{\rho} \vdash_{\Sigma, \Omega} \simsub{A_{n}}{\rho}}
+    {\Gamma \vdash_{\Sigma, \Omega} \simsub{A}{\rho}}\;(\tmOp)
+    \and \text{for $o\colon \Xi \vdash [\Delta_1]A_1, \ldots, [\Delta_{n}]A_{n} \to A$ in $\Omega$
+    and $\rho : \Sub{\Xi}{\emptyset}$}
   \end{mathpar}
-  \caption{Intrinsic typing rules}
+  \caption{Intrinsic typing rules for a simple type theory $(\Sigma, \Omega)$}
   \label{fig:intrinsic-typing}
 \end{figure}
-\subsection{Raw Terms and Extrinsically Typed Terms?}
-\begin{LTannote}{no extrinsic typing}
-Argue that we don't need extrinsically typed terms but raw terms only as the output of some parser.
-\end{LTannote}
+
+\subsection{Raw Terms and Extrinsically Typed Terms}
 
@@
   \begin{mathpar}
-    \boxed{\Theta \vdash_{\Sigma, \Omega} t}
+    \boxed{\vdash_{\Sigma, \Omega} \isTerm{t}}
+    \\
+    \inferrule{x : \Identifier}{\vdash_{\Sigma, \Omega} \isTerm{x}}\;(\text{var})
+    \and
+    \inferrule{\cdot \vdash_{\Sigma} A \\ \vdash_{\Sigma, \Omega}\isTerm{t}}{\vdash_{\Sigma, \Omega} \isTerm{t \annote A}}\;(\text{anno})
+    \\
+    \inferrule{\vdash_{\Sigma, \Omega} \isTerm{t_1} \quad \cdots \quad \vdash_{\Sigma, \Omega} \isTerm{t_n}}
+    {\vdash_{\Sigma, \Omega} \isTerm{\tmOp_o(t_1, \ldots, t_n)}}\;(\text{op}) 
+    \and \text{for $o \colon \Xi \vdash [\Delta_1]A_{1}^{d_1}, \ldots, [\Delta_{n}] A^{d_n}_{n} \to A^{d}$ in $\Omega$}
   \end{mathpar}
+  \caption{Annotatable raw terms for $(\Sigma, \Omega)$}
+\end{figure}
 
-  \caption{Raw terms}
+\begin{figure}
+  \centering
+  \small
+  \begin{mathpar}
+    \boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A} \quad \text{where $\vdash_{\Sigma, \Omega} t$} \\
+    \inferrule{(x : A) \in \Gamma}{\Gamma \vdash_{\Sigma, \Omega} \isTerm{x} : A}\;(\var)
+    \and
+    \inferrule{\Gamma \vdash \isTerm{t} : A}{\Gamma \vdash (\isTerm{t \annote A}) : A}\;(\text{anno})
+    \and
+    \inferrule{\Gamma, \isTerm{\vec{x}_1} : \simsub{\Delta_{1}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_1} : \simsub{A_{1}}{\rho} \quad\cdots\quad \Gamma, \isTerm{\vec{x}_n} : \simsub{\Delta_{n}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_n} : \simsub{A_{n}}{\rho}}
+    {\Gamma \vdash_{\Sigma, \Omega} \isTerm{\tmOp_o(\vec{x}_1.\,t_1; \ldots; \vec{x}_n.\,t_n)} : \simsub{A}{\rho}}\;(\tmOp)
+    \and \text{for $o\colon \Xi \vdash [\Delta_1]A_1, \ldots, [\Delta_{n}]A_{n} \to A$ in $\Omega$
+    and $\rho : \Sub{\Xi}{\emptyset}$}
+  \end{mathpar}
+  \caption{Extrinsic typing rules for a simple type theory $(\Sigma, \Omega)$ with annotation}
+  \label{fig:extrinsic-typing}
 \end{figure}
@@
   \[
-    \mathit{ar}\colon O \to \sum_{\Xi : \UU} \left(\Cxt_{\Sigma}(\Xi) \times \Mode \times \Type_{\Sigma}(\Xi)\right)^* \times \Mode \times \Type_{\Sigma}(\Xi).
+    \mathit{ar}\colon O \to \sum_{\Xi : \N} \left(\Cxt_{\Sigma}(\Xi) \times \Mode \times \Type_{\Sigma}(\Xi)\right)^* \times \Mode \times \Type_{\Sigma}(\Xi).
   \]
   where $\Mode$ consists of $\chk$ for checking and $\syn$ for synthesis.
-  Bidirectional binding signatures are just binding signatures (\Cref{def:binding-signature}) augmented with a mode for each argument and its target of a construct $\termOp_o$ in a bidirectional system, i.e.
+  Bidirectional binding signatures are just binding signatures (\Cref{def:binding-signature}) augmented with a mode for each argument and its target of a construct $\tmOp_o$ in a bidirectional system, i.e.
   an arity is a $4$-tuple
@@
   \begin{mathpar}
-    \boxed{\Theta\mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A^\isDir{d}}
-    \quad \text{where $d$ is either $\chk$ or $\syn$, $\Theta \vdash_{\Sigma} \Gamma$, $\Theta \vdash_{\Sigma} A$, and $\Theta \vdash_{\Sigma, \Omega} t^d$} 
+    \boxed{\vdash_{\Sigma, \Omega} \isTerm{t}\;\isDir{d}}
     \\
-    \inferrule{(x : A) \in \Gamma}{\Theta \mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{x} : A^\syn}\;(\var)
+    \inferrule{x : \Identifier}{\vdash_{\Sigma, \Omega} \isTerm{x}\;\isDir{\syn}}\;(\text{var})
     \and
-    \inferrule{\Theta\mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : B^\chk \\ A = B}{\Theta \mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t : B}:  A^\syn}\;(\text{anno})
+    \inferrule{\cdot \vdash_{\Sigma} A \\ \vdash_{\Sigma, \Omega}\isTerm{t}\;\isDir{\chk}}{\vdash_{\Sigma, \Omega} \isTerm{t \annote A}\;\isDir{\syn}}\;(\text{anno})
     \and
-    \inferrule{\Theta\mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : B^{\syn} \\ A = B}{\Theta \mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t\uparrow} : A^\chk}\;(\text{sub})
+    \inferrule{\vdash_{\Sigma, \Omega} \isTerm{t}\;\isDir{\syn}}{\vdash_{\Sigma, \Omega} (\isTerm{t \uparrow})\;\isDir{\chk}}
   \end{mathpar}
   \begin{mathpar}
-    \inferrule{\rho\colon \Sub{\Xi}{\Theta} \\
-      \Theta \mid \Gamma, \simsub{\isTerm{\vec{x}_{1}} : \Delta_{1}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_{1}} : \simsub{A_{1}}{\rho}^{\isDir{d_1}} \\
+    \inferrule{\vdash_{\Sigma, \Omega} \isTerm{t_1}\;\isDir{d_1} \quad \cdots \quad \vdash_{\Sigma, \Omega} \isTerm{t_n}\;\isDir{d_n}}
+    {\vdash_{\Sigma, \Omega} \isTerm{\tmOp_o(\vec{x}_1.\, t_1; \ldots;\vec{x}_n.\, t_n)}\;\isDir{d}}\;(\text{op})
+    \and \text{for $o \colon \Xi \vdash [\Delta_1]A_{1}^{d_1}, \ldots, [\Delta_{n}] A^{d_n}_{n} \to A^{d}$ in $\Omega$}
+  \end{mathpar}
+
+  \caption{Raw terms for a bidirectional system $(\Sigma, \Omega)$}
+\end{figure}
+\begin{figure}
+  \centering
+  \small
+  \begin{mathpar}
+    \boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A^\isDir{d}}
+    \quad \text{where $\vdash_{\Sigma, \Omega} \isTerm{t}\;d$} 
+    \\
+    \inferrule{(x : A) \in \Gamma}{\Gamma \vdash_{\Sigma, \Omega} \isTerm{x} : A^\syn}\;(\var)
+    \and
+    \inferrule{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : B^\chk \\ A = B}{\mid \Gamma \vdash_{\Sigma, \Omega} (\isTerm{t \annote B}):  A^\syn}\;(\text{anno})
+    \and
+    \inferrule{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : B^{\syn} \\ A = B}{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t\uparrow} : A^\chk}\;(\text{sub})
+    \\
+    \inferrule{\rho\colon \Sub{\Xi}{\emptyset} \\ \Gamma, \simsub{\isTerm{\vec{x}_{1}} : \Delta_{1}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_{1}} : \simsub{A_{1}}{\rho}^{\isDir{d_1}} \\
       \cdots \\
-    \Theta \mid \Gamma, \simsub{\isTerm{\vec{x}_{n}} : \Delta_{n}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_n} : \simsub{A_{n}}{\rho}^{\isDir{d_n}}}
-  {\Theta \mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{\termOp_o(\vec{x}_1 .\, t_{1}; \ldots; \vec{x}_{n}.\, t_{n})} : \simsub{A}{\rho}^{\isDir{d}}} \; (\termOp)
+    \Gamma, \simsub{\isTerm{\vec{x}_{n}} : \Delta_{n}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_n} : \simsub{A_{n}}{\rho}^{\isDir{d_n}}}
+  {\Gamma \vdash_{\Sigma, \Omega} \isTerm{\tmOp_o(\vec{x}_1 .\, t_{1}; \ldots; \vec{x}_{n}.\, t_{n})} : \simsub{A}{\rho}^{\isDir{d}}} \; (\tmOp)
     \and \text{for $o \colon \Xi \vdash [\Delta_1]A_{1}^{d_1}, \ldots, [\Delta_{n}] A^{d_n}_{n} \to A^{d}$ in $\Omega$}
     \\
-    \text{Abbreviations: $\boxed{\Theta\mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Rightarrow A} \defeq \boxed{\Theta\mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A^\syn}$ and $\boxed{\Theta\mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Leftarrow A} \defeq \boxed{\Theta\mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A^\chk}$}
+    \text{Abbreviations: $\boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Rightarrow A} \defeq \boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A^\syn}$ and $\boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Leftarrow A} \defeq \boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A^\chk}$}
   \end{mathpar}
-  \caption{Bidirectional extrinsic typing rules}
+  \caption{Extrinsic typing rules for a bidirectional system $(\Sigma, \Omega)$}
 \end{figure}

### 2023-05-30 tex/BiSig.tex
[ update ] A seperate subsection on Annotatability

@@
 
+\citep{Xie2018}
 \subsubsection{Language formalization and its frameworks}
@@
 
-\section{Bidirectional simply typed \texorpdfstring{$\lambda$}{lambda}-calculus}\label{sec:bi-stlc}
+\section{A Case Study: Bidirectional Simply Typed \texorpdfstring{$\lambda$}{lambda}-Calculus}\label{sec:bi-stlc}
 \begin{figure}
@@
 \end{remark}
-
-\subsection{Type Synthesis and Checking via Bidirectional Typing}
-\subsection{Formalisation in \Agda}
 
@@
   Simply typed $\lambda$-calculus typically includes a binary implication type denoted by $\bto$ and some base types to ensure that the set of all types is non-empty.
-  In our setting, the set of $\Sigma$-types over $\Theta$ may contain variables from $\Theta$, so $\Type_{\Sigma}(\Theta)$ is non-empty if $\Theta$ is.
-  Therefore, there is no need to introduce base types that are not used by any term construct.
-  The type signature $\Sigma_{\bto}$ of simply typed $\lambda$-calculus consists solely of a binary operation $\bto$, where $\arity(\bto) = 2$.
+  The type signature $\Sigma_{\bto}$ of simply typed $\lambda$-calculus consists of a binary operation $\bto$, where $\arity(\bto) = 2$ and an operation $N$ of arity $0$.
   In the case of simply typed $\lambda$-calculus with binary products, we can extend $\Sigma_{\bto}$ by adding a binary operation $\btimes$ to represent the binary product type.
@@
   \end{enumerate}
-  For brevity, we write $o \colon \Xi \vdash (\Delta_1)A_{1}, \ldots, \left[\Delta_{n}\right] A_{n} \to A$ to indicate an operation $o$ with its arity. 
+  For brevity, we write $o \colon \Xi \rhd (\Delta_1)A_{1}, \ldots, \left[\Delta_{n}\right] A_{n} \to A$ to indicate an operation $o$ with its arity. 
 \end{definition}
@@
   \begin{align*}
-    \mathsf{app}\colon A, B \vdash (A \bto B), A \to B && \mathsf{abs}\colon A , B \vdash [A]B \to (A \bto B)
+    \mathsf{app}\colon A, B \rhd (A \bto B), A \to B && \mathsf{abs}\colon A , B \rhd [A]B \to (A \bto B)
   \end{align*}
@@
   \begin{align*}
-    \mathsf{pair}\colon A, B \vdash A, B \to A \btimes B
-    && \fst \colon A, B \vdash A \btimes B \to A
-    && \snd \colon A, B \vdash A \btimes B \to B
+    \mathsf{pair}\colon A, B \rhd A, B \to A \btimes B
+    && \fst \colon A, B \rhd A \btimes B \to A
+    && \snd \colon A, B \rhd A \btimes B \to B
   \end{align*}
@@
     \and
-    \inferrule{\Gamma, \simsub{\Delta_{1}}{\rho} \vdash_{\Sigma, \Omega} \simsub{A_{1}}{\rho} \quad\cdots\quad \mid \Gamma, \simsub{\Delta_{n}}{\rho} \vdash_{\Sigma, \Omega} \simsub{A_{n}}{\rho}}
+    \inferrule{\rho\colon \Sub{\Xi}{\emptyset}  \\ \Gamma, \simsub{\Delta_{1}}{\rho} \vdash_{\Sigma, \Omega} \simsub{A_{1}}{\rho} \quad\cdots\quad \Gamma, \simsub{\Delta_{n}}{\rho} \vdash_{\Sigma, \Omega} \simsub{A_{n}}{\rho}}
     {\Gamma \vdash_{\Sigma, \Omega} \simsub{A}{\rho}}\;(\tmOp)
-    \and \text{for $o\colon \Xi \vdash [\Delta_1]A_1, \ldots, [\Delta_{n}]A_{n} \to A$ in $\Omega$
-    and $\rho : \Sub{\Xi}{\emptyset}$}
+    \and \text{for $o\colon \Xi \rhd [\Delta_1]A_1, \ldots, [\Delta_{n}]A_{n} \to A$ in $\Omega$}
   \end{mathpar}
@@
     {\vdash_{\Sigma, \Omega} \isTerm{\tmOp_o(t_1, \ldots, t_n)}}\;(\text{op}) 
-    \and \text{for $o \colon \Xi \vdash [\Delta_1]A_{1}^{d_1}, \ldots, [\Delta_{n}] A^{d_n}_{n} \to A^{d}$ in $\Omega$}
+    \and \text{for $o \colon \Xi \rhd [\Delta_1]A_{1}^{d_1}, \ldots, [\Delta_{n}] A^{d_n}_{n} \to A^{d}$ in $\Omega$}
   \end{mathpar}
@@
     {\Gamma \vdash_{\Sigma, \Omega} \isTerm{\tmOp_o(\vec{x}_1.\,t_1; \ldots; \vec{x}_n.\,t_n)} : \simsub{A}{\rho}}\;(\tmOp)
-    \and \text{for $o\colon \Xi \vdash [\Delta_1]A_1, \ldots, [\Delta_{n}]A_{n} \to A$ in $\Omega$
+    \and \text{for $o\colon \Xi \rhd [\Delta_1]A_1, \ldots, [\Delta_{n}]A_{n} \to A$ in $\Omega$
     and $\rho : \Sub{\Xi}{\emptyset}$}
@@
   \[
-    \mathit{ar}\colon O \to \sum_{\Xi : \N} \left(\Cxt_{\Sigma}(\Xi) \times \Mode \times \Type_{\Sigma}(\Xi)\right)^* \times \Mode \times \Type_{\Sigma}(\Xi).
+    \mathit{ar}\colon O \to \sum_{\Xi : \N} \left(\Cxt_{\Sigma}(\Xi) \times \isDir{\Mode} \times \Type_{\Sigma}(\Xi)\right)^* \times \isDir{\Mode} \times \Type_{\Sigma}(\Xi).
   \]
@@
 
-  For brevity, we write $o \colon \Xi \vdash [\Delta_1]A_{1}^{d_1}, \ldots, [\Delta_{n}] A^{d_n}_{n} \to A^{d}$ to indicate an operation $o$ with its arity. 
+  For brevity, we write $o \colon \Xi \rhd [\Delta_1]A_{1}^{d_1}, \ldots, [\Delta_{n}] A^{d_n}_{n} \to A^{d}$ to indicate an operation $o$ with its arity. 
 \end{definition}
@@
   \begin{align*}
-    \mathsf{app}\colon A, B \vdash (A \bto B)^{\Rightarrow}, A^{\Leftarrow} \to B^{\Rightarrow} &&
-    \mathsf{abs}\colon A , B \vdash [A]B^{\Leftarrow} \to (A \bto B)^{\Leftarrow}
+    \mathsf{app}\colon A, B \rhd (A \bto B)^{\Rightarrow}, A^{\Leftarrow} \to B^{\Rightarrow} &&
+    \mathsf{abs}\colon A , B \rhd [A]B^{\Leftarrow} \to (A \bto B)^{\Leftarrow}
   \end{align*}
@@
 
-\subsection{Bidirectionally Typed Terms: Intrinsic Typing vs.\ Extrinsic Typing}
+\begin{LTannote}{deriving from annotation}
+  deriving bidirectional binding signature from annotation
+\end{LTannote}
+\begin{theorem}[Signature annotation]
 
-\subsubsection{Intrinsically Typed Terms}
+\end{theorem}
 
-\subsubsection{Extrinsically Typed Terms}
+\subsection{Extrinsically Typed Terms for Bidirectional Type Systems}
 
@@
     {\vdash_{\Sigma, \Omega} \isTerm{\tmOp_o(\vec{x}_1.\, t_1; \ldots;\vec{x}_n.\, t_n)}\;\isDir{d}}\;(\text{op})
-    \and \text{for $o \colon \Xi \vdash [\Delta_1]A_{1}^{d_1}, \ldots, [\Delta_{n}] A^{d_n}_{n} \to A^{d}$ in $\Omega$}
+    \and \text{for $o \colon \Xi \rhd [\Delta_1]A_{1}^{d_1}, \ldots, [\Delta_{n}] A^{d_n}_{n} \to A^{d}$ in $\Omega$}
   \end{mathpar}
-
   \caption{Raw terms for a bidirectional system $(\Sigma, \Omega)$}
 \end{figure}
+
 \begin{figure}
@@
   {\Gamma \vdash_{\Sigma, \Omega} \isTerm{\tmOp_o(\vec{x}_1 .\, t_{1}; \ldots; \vec{x}_{n}.\, t_{n})} : \simsub{A}{\rho}^{\isDir{d}}} \; (\tmOp)
-    \and \text{for $o \colon \Xi \vdash [\Delta_1]A_{1}^{d_1}, \ldots, [\Delta_{n}] A^{d_n}_{n} \to A^{d}$ in $\Omega$}
+    \and \text{for $o \colon \Xi \rhd [\Delta_1]A_{1}^{d_1}, \ldots, [\Delta_{n}] A^{d_n}_{n} \to A^{d}$ in $\Omega$}
     \\
@@
 
-\subsection{Between Binding Signatures and their Bidirectional Variants}
-\subsubsection{Soundness}
-\subsubsection{Completeness (Annotatability)}
-\subsection{Mode Correctness}
+\subsection{Soundness and Completeness with respect to Binding Signatures}
+
+\subsection{Annotatability}
 
@@
 
-\subsection{Substitution Properties}
-\subsection{Generic First-Order Unification}
-\cite{McBride2003,McBride2003a}
-\begin{definition}
-A \emph{decidable equality} for a type $X$ is a dependent function $\prod_{x, y : X} \left(\id{x}{y}\right) + \neg \left(\id{x}{y}\right)$.
-\end{definition}
+\subsection{Mode Correctness}
+%\subsection{Substitution Properties}
+%\subsection{Generic First-Order Unification}
+%\cite{McBride2003,McBride2003a}
 
@@
 
-\subsection{Type Synthesis and Checking with Unification}
+\subsection{Type Synthesis and Checking}
 
-\section{Formal Implementation and Examples}
-\begin{LTannote}{de Bruijn}
-  In our Agda formalisation, we use de Bruijn indices to represent variables, so we don't use an arbitrary type $\Theta$ but the number of variables as the index of typing judgements.
-\end{LTannote}
-
-\citep{Xie2018}
+\section{Formalisation in \Agda with Examples}
 

### 2023-06-09 tex/BiSig.tex
[ tex ] Statements of main results

@@
 \usepackage[inline]{enumitem} % for environment enumerate*
+\setlist[enumerate]{mode=unboxed}
 
-\usepackage[draft]{fixme}
+\setlength {\marginparwidth }{2cm}
+\usepackage[color=yellow,textsize=small]{todonotes}
+\newcommand{\LT}[1]{\todo[author=LT,inline,color=green!40,caption={}]{{#1}}}
+\newcommand{\Josh}[1]{\todo[author=Josh,inline,color=green!40,caption={}]{{#1}}}
+
 \usepackage{subfiles}
@@
 \usepackage{microtype}
-
-\FXRegisterAuthor{LT}{LTan}{L-T}
-\FXRegisterAuthor{Josh}{Joshan}{Josh}
 
@@
 
-\title{A Formal Theory of Bidirectional Typing for Simple Types}
+\title{A Formal Practical Foundation of Bidirectional Typing for Simple Types}
 
@@
 \maketitle
+
 \section{Introduction}\label{sec:intro}
@@
 
-\section{A Case Study: Bidirectional Simply Typed \texorpdfstring{$\lambda$}{lambda}-Calculus}\label{sec:bi-stlc}
+Our assumptions and contributions:
+\begin{enumerate}
+  \item Syntax-directed typing rules
+  \item Simple types
+  \item Fully formalised in \Agda 
+  \item Generic.
+\end{enumerate}
+
+\section{Using Bidirectional Simply Typed \texorpdfstring{$\lambda$}{lambda}-Calculus as an Outline} \label{sec:bi-stlc}
 \begin{figure}
@@
     \and
-    \inferrule{\Gamma \vdash \isTerm{t} \Rightarrow A \bto B \\ \Gamma \vdash \isTerm{u} \Leftarrow A}{\Gamma \vdash \isTerm{t\;u} \Rightarrow B}
+    \inferrule{\Gamma \vdash \isTerm{t} \Rightarrow A \bto B \\ \Gamma \vdash \isTerm{u} \Leftarrow A}{\Gamma \vdash \isTerm{t\;u} \Rightarrow B}\;(\text{app})
     \and
-    \inferrule{\Gamma, \isTerm{x} : A \vdash \isTerm{t} \Leftarrow B}{\Gamma \vdash \isTerm{\lam{x}t} \Leftarrow A \bto B}
+    \inferrule{\Gamma, \isTerm{x} : A \vdash \isTerm{t} \Leftarrow B}{\Gamma \vdash \isTerm{\lam{x}t} \Leftarrow A \bto B}\;(\text{abs})
   \end{mathpar}
@@
 \end{example}
-
-\begin{remark}
-  \begin{LTannote}{$\Xi$ is just a number}
-    We will just use $\Xi : \mathbb{N}$ for the set of variables.
-  \end{LTannote}
-
-\end{remark}
 
@@
     \and
-    \inferrule{\Gamma, \isTerm{\vec{x}_1} : \simsub{\Delta_{1}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_1} : \simsub{A_{1}}{\rho} \quad\cdots\quad \Gamma, \isTerm{\vec{x}_n} : \simsub{\Delta_{n}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_n} : \simsub{A_{n}}{\rho}}
+    \inferrule{{\color{red}\rho : \Sub{\Xi}{\emptyset}} \\ \Gamma, \isTerm{\vec{x}_1} : \simsub{\Delta_{1}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_1} : \simsub{A_{1}}{\rho} \quad\cdots\quad \Gamma, \isTerm{\vec{x}_n} : \simsub{\Delta_{n}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_n} : \simsub{A_{n}}{\rho}}
     {\Gamma \vdash_{\Sigma, \Omega} \isTerm{\tmOp_o(\vec{x}_1.\,t_1; \ldots; \vec{x}_n.\,t_n)} : \simsub{A}{\rho}}\;(\tmOp)
-    \and \text{for $o\colon \Xi \rhd [\Delta_1]A_1, \ldots, [\Delta_{n}]A_{n} \to A$ in $\Omega$
-    and $\rho : \Sub{\Xi}{\emptyset}$}
+    \and \text{for $o\colon \Xi \rhd [\Delta_1]A_1, \ldots, [\Delta_{n}]A_{n} \to A$ in $\Omega$}
   \end{mathpar}
@@
 
-\begin{LTannote}{deriving from annotation}
-  deriving bidirectional binding signature from annotation
-\end{LTannote}
+\LT{deriving bidirectional binding signature from annotation}
+
 \begin{theorem}[Signature annotation]
@@
   \begin{mathpar}
-    \boxed{\vdash_{\Sigma, \Omega} \isTerm{t}\;\isDir{d}}
+    \boxed{\vdash_{\Sigma, \Omega} \isTerm{t}^\isDir{d}}
     \\
-    \inferrule{x : \Identifier}{\vdash_{\Sigma, \Omega} \isTerm{x}\;\isDir{\syn}}\;(\text{var})
+    \inferrule{x : \Identifier}{\vdash_{\Sigma, \Omega} \isTerm{x}^\isDir{\syn}}\;(\text{var})
     \and
-    \inferrule{\cdot \vdash_{\Sigma} A \\ \vdash_{\Sigma, \Omega}\isTerm{t}\;\isDir{\chk}}{\vdash_{\Sigma, \Omega} \isTerm{t \annote A}\;\isDir{\syn}}\;(\text{anno})
+    \inferrule{\cdot \vdash_{\Sigma} A \\ \vdash_{\Sigma, \Omega}\isTerm{t}^\isDir{\chk}}{\vdash_{\Sigma, \Omega} \isTerm{t \annote A}^\isDir{\syn}}\;(\text{anno})
     \and
-    \inferrule{\vdash_{\Sigma, \Omega} \isTerm{t}\;\isDir{\syn}}{\vdash_{\Sigma, \Omega} (\isTerm{t \uparrow})\;\isDir{\chk}}
+    \inferrule{\vdash_{\Sigma, \Omega} \isTerm{t}^\isDir{\syn}}{\vdash_{\Sigma, \Omega} (\isTerm{t \uparrow})^\isDir{\chk}}\;(\text{sub})
   \end{mathpar}
   \begin{mathpar}
-    \inferrule{\vdash_{\Sigma, \Omega} \isTerm{t_1}\;\isDir{d_1} \quad \cdots \quad \vdash_{\Sigma, \Omega} \isTerm{t_n}\;\isDir{d_n}}
-    {\vdash_{\Sigma, \Omega} \isTerm{\tmOp_o(\vec{x}_1.\, t_1; \ldots;\vec{x}_n.\, t_n)}\;\isDir{d}}\;(\text{op})
+    \inferrule{\vdash_{\Sigma, \Omega} \isTerm{t_1}^\isDir{d_1} \quad \cdots \quad \vdash_{\Sigma, \Omega} \isTerm{t_n}^\isDir{d_n}}
+    {\vdash_{\Sigma, \Omega} \isTerm{\tmOp_o(\vec{x}_1.\, t_1; \ldots;\vec{x}_n.\, t_n)}^\isDir{d}}\;(\text{op})
     \and \text{for $o \colon \Xi \rhd [\Delta_1]A_{1}^{d_1}, \ldots, [\Delta_{n}] A^{d_n}_{n} \to A^{d}$ in $\Omega$}
@@
     \boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A^\isDir{d}}
-    \quad \text{where $\vdash_{\Sigma, \Omega} \isTerm{t}\;d$} 
+    \quad \text{where $\vdash_{\Sigma, \Omega} \isTerm{t}^\isDir{d}$} 
     \\
@@
 
-\section{Bidirectional Type Synthesis and Checking}
+\begin{theorem}[Decidability for Bidirectionalisation]
+  For every annotatable raw $t$, it is decidable that $t$ has a mode $d$, i.e.\ $\Gamma \vdash_{\Sigma, \Omega} \isTerm{t}\,\isDir{d}$ is derivable.
+\end{theorem}
+
+\section{Bidirectional Type Synthesis}
 
@@
 
-\subsection{Uniqueness of Synthesis}
+\subsection{Decidability of Type Synthesis}
 
-\subsection{Type Synthesis and Checking}
+\begin{theorem}[Decidability for Bidirectional Typing Judgements]
+  For every type signature $\Sigma$ and bidirectional binding signature $\Omega$ for $\Sigma$ that is mode correct, the following two statements hold constructively.
+  \begin{enumerate}
+    \item For every context $\Gamma$ and a pre-synthesised term $\isTerm{t}$, it is decidable that there exists a type $A$ and a bidirectional typing derivation of
+      \[
+        \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Rightarrow A.
+      \]
+    \item For every context $\Gamma$, a pre-checked term $\isTerm{t}$, and a type $A$, it is decidable that there is a typing derivation of
+      \[
+        \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Leftarrow A.
+      \]
+  \end{enumerate}
+\end{theorem}
+
+\begin{theorem}[Decidability of Type Inference]
+  \LT{combine the above two steps}
+  Let $\Sigma$ be a type signature and $\Omega$ a binding signature with a bidirectional annotation $\Omega^{\leftrightarrows}$.
+  Then, for every context $\Gamma$ and raw term $t$, one and only one of the following statements holds:
+  \begin{enumerate}
+    \item $t$ has typing derivation $\Gamma \vdash_{\Sigma, \Omega} t : A$ for some type $\vdash_{\Sigma} A$.
+    \item $t$ is not typable under $\Gamma$.
+    \item there exists $t'$ such that $t$ needs more type annotations marked in $t'$ to be inferred.
+  \end{enumerate}
+\end{theorem}
+
+\subsection{Decidability of Bidirectional Type Synthesis and Checking}
+
+\begin{theorem}{Uniqueness of Synthesised Types}\label{thm:unique-syn}
+
+\end{theorem}
+
+\begin{theorem}
+  For every type signature $\Sigma$ and bidirectional binding signature $\Omega$ for $\Sigma$ that is mode correct, the following statements hold constructively.
+  \begin{enumerate}
+    \item For every context $\Gamma$ and a raw term $t^d$ with a mode $d$, 
+      \begin{enumerate}
+        \item if $d$ is $\Rightarrow$, then it is decidable that there exists a type $A$ and a typing derivation of
+          \[\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Rightarrow A;
+          \]
+        \item if $d$ is $\Leftarrow$, then for every type $A$ it is decidable that there is a typing derivation of
+          \[
+            \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Leftarrow A.
+          \]
+      \end{enumerate}
+
+    \item For every operation $o \colon \Xi \rhd [\Delta_1]A_{1}^{d_1}, \ldots, [\Delta_{n}] A^{d_n}_{n} \to A^{d}$ in $\Omega$, a context $\Gamma$, and a raw term
+      \[
+        \isTerm{\tmOp_{o}(\vec{x}_1.t_1,\ldots, \vec{x}_n.t_n)}, 
+      \]
+      that has a mode $d$, it is decidable that there exists a substitution $\rho\colon \Sub{\Xi}{\emptyset}$ and a typing derivation of
+      \[
+        \Gamma \vdash_{\Sigma, \Omega} \isTerm{\tmOp_{o}(\vec{x}_1.t_1,\ldots, \vec{x}_n.t_n)} : \simsub{A}{\rho}^\isDir{d}.
+      \]
+    \item For every set $\Xi$ of type variables, a list
+      \[
+        \left[\left(\Delta_1, d_1, A_1\right), \ldots, \left(\Delta_n, d_n, A_n\right)\right]
+      \]
+      of triples in $\Cxt_{\Sigma}(\Xi) \times \Mode \times \Type_{\Sigma}(\Xi)$, a \emph{partial substitution} $\rho\colon\PSub{\Xi}{\emptyset}$, a context $\Gamma$, and a list of raw terms $t_i$'s with mode $d_i$ for $i = 1, \ldots, n$, there is either a minimal extension $\rho'$ of $\rho$ such that there is typing derivations of
+      \[
+        \Gamma, \simsub{\Delta_i}{\rho'} \vdash_{\Sigma, \Omega} t_i \colon \simsub{A_i}{\rho'}^{d_i}
+      \]
+      for each $i$, or there is no such extension $\rho'$ such that each raw term $t_i^{d_i}$ has a typing derivation.
+
+    \item For every set $\Xi$ of type variables, a list $\Delta$ of types with variables in $\Xi$, a partial substitution $\rho\colon \PSub{\Xi, \emptyset}$ defined for all types in $\Delta$, and a raw term with mode $d$ then
+      \begin{enumerate}
+        \item if $d$ is $\Rightarrow$, then it is decidable that there exists a type $A$ and a typing derivation of
+          \[
+            \Gamma, \simsub{\Delta}{\rho} \vdash_{\Sigma, \Omega} t \Rightarrow A 
+          \]
+        \item if $d$ is $\Leftarrow$, then for every type $A$ it is decidable that there is a typing derivation of 
+          \[
+            \Gamma, \simsub{\Delta}{\rho} \vdash_{\Sigma, \Omega} t \Leftarrow A.
+          \]
+      \end{enumerate}
+  \end{enumerate}
+\end{theorem}
+By mutual induction on raw terms $t$ with a mode. 
+Moreover, if the type $A$ of $t$ is synthesised but checked against $B$, i.e.\ $t\uparrow$, but $A \neq B$, then \Cref{thm:unique-syn} is applied to derive a contradiction.
+\begin{proof}
+\end{proof}
 

### 2023-06-10 tex/BiSig.tex
[ tex ] Fix some typos

@@
   For every type signature $\Sigma$ and bidirectional binding signature $\Omega$ for $\Sigma$ that is mode correct, the following two statements hold constructively.
+\LT{emphasise that LEM is not allowed.}
   \begin{enumerate}
@@
 
-\begin{theorem}{Uniqueness of Synthesised Types}\label{thm:unique-syn}
+\begin{theorem}[Uniqueness of Synthesised Types]\label{thm:unique-syn}
 
@@
 
-    \item For every set $\Xi$ of type variables, a list $\Delta$ of types with variables in $\Xi$, a partial substitution $\rho\colon \PSub{\Xi, \emptyset}$ defined for all types in $\Delta$, and a raw term with mode $d$ then
+    \item For every set $\Xi$ of type variables, a list $\Delta$ of types with variables in $\Xi$, a partial substitution $\rho\colon \PSub{\Xi, \emptyset}$ defined at least for all types in $\Delta$, a raw term $t$ with mode $d$, and a list of variables $\vec{x}$ then
       \begin{enumerate}
@@
           \[
-            \Gamma, \simsub{\Delta}{\rho} \vdash_{\Sigma, \Omega} t \Rightarrow A 
+            \Gamma, \vec{x} : \simsub{\Delta}{\rho} \vdash_{\Sigma, \Omega} t \Rightarrow A 
           \]
@@
           \[
-            \Gamma, \simsub{\Delta}{\rho} \vdash_{\Sigma, \Omega} t \Leftarrow A.
+            \Gamma, \vec{x} : \simsub{\Delta}{\rho} \vdash_{\Sigma, \Omega} t \Leftarrow A.
           \]

### 2023-06-12 tex/BiSig.tex
[ text ] Update the structure of paper

--- 2023-06-09 tex/BiSig.tex
@@
 
-\title{A Formal Practical Foundation of Bidirectional Typing for Simple Types}
+\title{A Formal and Practical Metatheory of Bidirectional Type Synthesis for Simple Types}
 
@@
 \section{Introduction}\label{sec:intro}
+
+\begin{enumerate}
+  \item Type checking as a transition from (parsed) abstract syntax trees to well-typed terms.
+  \item Bidirectional typing is the only solution that works that scales well.
+  \item Not yet a general theory but by now a design principle.
+    No `type checker generator' as each type checker is still designed and implemented on a case-by-case basis, but can be derived from a syntax-directed type theory `easily'.
+  \item The existing work of bidirectional typing does not take account of missing annotations carefully enough. 
+  \item A meta-problem: existing work is based on (or at least left unspecified) classical logic.
+    One has to state soundness of completeness of their type checking algorithm in a less straightforward way.
+  \item A review of what has been done towards this line of research.
+  \item Finally, our responses (contributions of this paper).
+\end{enumerate}
 
@@
 \subsubsection{Bidirectional typing}
+
 \cite{Pierce2000}
@@
 \begin{enumerate}
-  \item Syntax-directed typing rules
-  \item Simple types
-  \item Fully formalised in \Agda 
-  \item Generic.
+  \item Assumption: Syntax-directed typing rules
+  \item Practical: bidirectionalisation over annotatability, decidability over soundness and completeness.
+  \item Generic: Type checking once and for all bidirectional type systems for simple types.
+  \item Correct and Constructive: Fully formalised in Martin-Löf type theory (with Axiom K).
+    LEM is not assumed, so a decidable statement has both computational and non-trivial logical readings.
 \end{enumerate}
 
-\section{Using Bidirectional Simply Typed \texorpdfstring{$\lambda$}{lambda}-Calculus as an Outline} \label{sec:bi-stlc}
+Plan of this paper:
+\begin{enumerate}
+  \item Key ideas (\Cref{sec:key-ideas})
+  \item Definitions for simple type theories and for bidirectional type systems (signature erasure and mode annotation) (\Cref{sec:defs})
+  \item Soundness, completeness, bidirectionalisation, and annotatability (\Cref{sec:annotatability})
+    \Josh{Erasure from bidirectional typing derivation to raw terms with a mode}
+  \item Bidirectional type inference (\Cref{sec:type-synthesis})
+  \item Formalisation with further examples (\Cref{sec:formalisation})
+  \item Future work (\Cref{sec:future})
+\end{enumerate}
+  \LT{better title?}
+
+\section{Core Ideas with Bidirectional Simply Typed \texorpdfstring{$\lambda$}{lambda}-Calculus} \label{sec:key-ideas}
 \begin{figure}
@@
 
-\section{Definitions of Simple Type Theories}\label{sec:simple-type-theories}
+\section{Definitions for Simple Type Theories and Bidirectional Type Systems}\label{sec:defs}
+
 \subsection{Simple Types}
@@
   Simply typed $\lambda$-calculus typically includes a binary implication type denoted by $\bto$ and some base types to ensure that the set of all types is non-empty.
-  The type signature $\Sigma_{\bto}$ of simply typed $\lambda$-calculus consists of a binary operation $\bto$, where $\arity(\bto) = 2$ and an operation $N$ of arity $0$.
+  The type signature $\Sigma_{\bto}$ of simply typed $\lambda$-calculus consists of a binary operation $\bto$ and a base type (nullary operation) $N$.
   In the case of simply typed $\lambda$-calculus with binary products, we can extend $\Sigma_{\bto}$ by adding a binary operation $\btimes$ to represent the binary product type.
@@
 
-\subsection{Binding Signatures}
+\subsection{Binding Signatures} \label{subsec:binding-sig}
 \begin{definition}\label{def:binding-signature}
@@
 
-\subsection{Intrinsically Typed Terms}
+%\subsection{Intrinsically Typed Terms}
+%
+%Intrinsically typed terms for a type signature $\Sigma$ and a term signature $\Omega$ are nothing more than derivations of the intrinsic typing judgement $\Gamma \vdash_{\Sigma, \Omega} A$ constructed with only the rule $(\var)$ and the rule scheme $(\tmOp)$ displayed in \Cref{fig:intrinsic-typing}.
+%
+%The substitution $\rho\colon \Sub{\Xi}{\emptyset}$ is used to instantiate variables in the local context $\Xi$ with concrete types.
+%Accordingly, $\rho$ has to be applied to types that appear in the arity to construct a term by $\tmOp$ to ensure all types are well-formed without any use of type variables.
+%
+%\begin{figure}
+%  \centering
+%  \small
+%  \begin{mathpar}
+%    \boxed{\Gamma \vdash_{\Sigma, \Omega} A} \\
+%    \inferrule{A \in \Gamma}{\Gamma \vdash_{\Sigma, \Omega} A}\;(\var)
+%    \and
+%    \inferrule{\rho\colon \Sub{\Xi}{\emptyset}  \\ \Gamma, \simsub{\Delta_{1}}{\rho} \vdash_{\Sigma, \Omega} \simsub{A_{1}}{\rho} \quad\cdots\quad \Gamma, \simsub{\Delta_{n}}{\rho} \vdash_{\Sigma, \Omega} \simsub{A_{n}}{\rho}}
+%    {\Gamma \vdash_{\Sigma, \Omega} \simsub{A}{\rho}}\;(\tmOp)
+%    \and \text{for $o\colon \Xi \rhd [\Delta_1]A_1, \ldots, [\Delta_{n}]A_{n} \to A$ in $\Omega$}
+%  \end{mathpar}
+%  \caption{Intrinsic typing rules for a simple type theory $(\Sigma, \Omega)$}
+%  \label{fig:intrinsic-typing}
+%\end{figure}
 
-Intrinsically typed terms for a type signature $\Sigma$ and a term signature $\Omega$ are nothing more than derivations of the intrinsic typing judgement $\Gamma \vdash_{\Sigma, \Omega} A$ constructed with only the rule $(\var)$ and the rule scheme $(\tmOp)$ displayed in \Cref{fig:intrinsic-typing}.
-
-The substitution $\rho\colon \Sub{\Xi}{\emptyset}$ is used to instantiate variables in the local context $\Xi$ with concrete types.
-Accordingly, $\rho$ has to be applied to types that appear in the arity to construct a term by $\tmOp$ to ensure all types are well-formed without any use of type variables.
-
-\begin{figure}
-  \centering
-  \small
-  \begin{mathpar}
-    \boxed{\Gamma \vdash_{\Sigma, \Omega} A} \\
-    \inferrule{A \in \Gamma}{\Gamma \vdash_{\Sigma, \Omega} A}\;(\var)
-    \and
-    \inferrule{\rho\colon \Sub{\Xi}{\emptyset}  \\ \Gamma, \simsub{\Delta_{1}}{\rho} \vdash_{\Sigma, \Omega} \simsub{A_{1}}{\rho} \quad\cdots\quad \Gamma, \simsub{\Delta_{n}}{\rho} \vdash_{\Sigma, \Omega} \simsub{A_{n}}{\rho}}
-    {\Gamma \vdash_{\Sigma, \Omega} \simsub{A}{\rho}}\;(\tmOp)
-    \and \text{for $o\colon \Xi \rhd [\Delta_1]A_1, \ldots, [\Delta_{n}]A_{n} \to A$ in $\Omega$}
-  \end{mathpar}
-  \caption{Intrinsic typing rules for a simple type theory $(\Sigma, \Omega)$}
-  \label{fig:intrinsic-typing}
-\end{figure}
-
-\subsection{Raw Terms and Extrinsically Typed Terms}
+\subsection{Simple Type Theories}
 
@@
     \and
-    \inferrule{{\color{red}\rho : \Sub{\Xi}{\emptyset}} \\ \Gamma, \isTerm{\vec{x}_1} : \simsub{\Delta_{1}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_1} : \simsub{A_{1}}{\rho} \quad\cdots\quad \Gamma, \isTerm{\vec{x}_n} : \simsub{\Delta_{n}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_n} : \simsub{A_{n}}{\rho}}
+    \inferrule{\rho : \Sub{\Xi}{\emptyset} \\ \Gamma, \isTerm{\vec{x}_1} : \simsub{\Delta_{1}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_1} : \simsub{A_{1}}{\rho} \quad\cdots\quad \Gamma, \isTerm{\vec{x}_n} : \simsub{\Delta_{n}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_n} : \simsub{A_{n}}{\rho}}
     {\Gamma \vdash_{\Sigma, \Omega} \isTerm{\tmOp_o(\vec{x}_1.\,t_1; \ldots; \vec{x}_n.\,t_n)} : \simsub{A}{\rho}}\;(\tmOp)
@@
   \end{mathpar}
-  \caption{Extrinsic typing rules for a simple type theory $(\Sigma, \Omega)$ with annotation}
+  \caption{Typing rules for a simple type theory $(\Sigma, \Omega)$ with annotation}
   \label{fig:extrinsic-typing}
@@
 
-\section{Definitions of Bidirectional Simple Type Systems}
-
-\subsection{Bidirectional Binding Signatures}
+\subsection{Bidirectional Binding Signatures and Bidirectional Type Systems}
 
@@
   \[
-    \mathit{ar}\colon O \to \sum_{\Xi : \N} \left(\Cxt_{\Sigma}(\Xi) \times \isDir{\Mode} \times \Type_{\Sigma}(\Xi)\right)^* \times \isDir{\Mode} \times \Type_{\Sigma}(\Xi).
+    \mathit{ar}\colon O \to \sum_{\Xi : \N} \left(\Cxt_{\Sigma}(\Xi) \times \Type_{\Sigma}(\Xi) \times \isDir{\Mode}\right)^* \times \Type_{\Sigma}(\Xi) \times \isDir{\Mode}.
   \]
@@
   \[
-    \left(\Xi, \left[\left(\Delta_1, d_1, A_1\right), \ldots, \left(\Delta_{n}, d_n, A_{n}\right) \right], d, A\right)
+    \left(\Xi, \left[\left(\Delta_1, A_1, d_1\right), \ldots, \left(\Delta_{n}, A_{n}, d_n\right) \right], A, d\right)
   \]
@@
 \begin{example}
-  The term signature $\Lambda_{\bto}^{\leftrightarrows}$ for bidirectional simply typed $\lambda$-calculus introduced in \Cref{sec:bi-stlc} can be specified by operations 
+  The term signature $\Lambda_{\bto}^{\leftrightarrows}$ for bidirectional simply typed $\lambda$-calculus introduced in \Cref{subsec:binding-sig} can be specified by operations 
   \begin{align*}
@@
 
+\LT{Signature Erasure}
 \begin{theorem}[Signature annotation]
@@
 \end{theorem}
-
-\subsection{Extrinsically Typed Terms for Bidirectional Type Systems}
 
@@
   \end{mathpar}
-  \caption{Raw terms for a bidirectional system $(\Sigma, \Omega)$}
+  \caption{Raw terms with a mode $d$ for a bidirectional type system $(\Sigma, \Omega)$}
 \end{figure}
@@
   \end{mathpar}
-  \caption{Extrinsic typing rules for a bidirectional system $(\Sigma, \Omega)$}
+  \caption{Typing rules for a bidirectional system $(\Sigma, \Omega)$}
 \end{figure}
@@
 
-\subsection{Soundness and Completeness with respect to Binding Signatures}
+\section{Soundness, Completeness, and Bidirectionalisation} \label{sec:annotatability}
 
-\subsection{Annotatability}
+\subsection{Soundness and Completeness}
 
-\begin{theorem}[Decidability for Bidirectionalisation]
-  For every annotatable raw $t$, it is decidable that $t$ has a mode $d$, i.e.\ $\Gamma \vdash_{\Sigma, \Omega} \isTerm{t}\,\isDir{d}$ is derivable.
+\subsection{Bidirectionalisation}
+\begin{theorem}[Decidability for Bidirectionalisation] \label{thm:bidirectionalisation}
+  For every raw term $t$ and a mode $d$, it is decidable that $t$ has a mode $d$, i.e.\ $\Gamma \vdash_{\Sigma, \Omega} \isTerm{t}\,\isDir{d}$ is derivable.
 \end{theorem}
 
-\section{Bidirectional Type Synthesis}
+\subsection{Completeness versus Annotatability}
+
+\section{Bidirectional Type Synthesis and Checking} \label{sec:type-synthesis}
 
 \subsection{Mode Correctness}
-%\subsection{Substitution Properties}
-%\subsection{Generic First-Order Unification}
-%\cite{McBride2003,McBride2003a}
+
+\begin{theorem}[Uniqueness of Synthesised Types]\label{thm:unique-syn}
+  Let $(\Sigma, \Omega)$ be a bidirectional type system where $\Omega$ is mode-correct.
+  Then, for every raw $t$ with a synthesis mode $\Rightarrow$, and any derivations of 
+  \[
+    \Gamma \vdash t \Rightarrow A
+    \quad\text{and}\quad
+    \Gamma \vdash t \Rightarrow B
+  \]
+  for some types $A$ and $B$, the synthesised types must be equal, i.e.\ $A = B$.
+\end{theorem}
 
@@
 
-\begin{theorem}[Decidability for Bidirectional Typing Judgements]
-  For every type signature $\Sigma$ and bidirectional binding signature $\Omega$ for $\Sigma$ that is mode correct, the following two statements hold constructively.
+\begin{theorem}[Decidability for Bidirectional Typing Synthesis] \label{thm:bidirectional-type-synthesis}
+  For every type signature $\Sigma$ and bidirectional binding signature $\Omega$ for $\Sigma$ that is mode-correct, the following two statements hold.
+
   \begin{enumerate}
-    \item For every context $\Gamma$ and a pre-synthesised term $\isTerm{t}$, it is decidable that there exists a type $A$ and a bidirectional typing derivation of
+    \item For every context $\Gamma$ and a raw term $\isTerm{t}$ with a synthesis mode, it is decidable that there exists a type $A$ and a derivation of
       \[
@@
       \]
-    \item For every context $\Gamma$, a pre-checked term $\isTerm{t}$, and a type $A$, it is decidable that there is a typing derivation of
+    \item For every context $\Gamma$, a raw term $\isTerm{t}$ with a checking mode, and a type $A$, it is decidable that there is a derivation of
       \[
@@
 
-\begin{theorem}[Decidability of Type Inference]
-  \LT{combine the above two steps}
-  Let $\Sigma$ be a type signature and $\Omega$ a binding signature with a bidirectional annotation $\Omega^{\leftrightarrows}$.
+\begin{corollary}[`Decidability' of Type Synthesis]
+  \LT{Still call it `decidability'?}
+  Let $\Sigma$ be a type signature and $\Omega$ a bidirectional binding signature.
   Then, for every context $\Gamma$ and raw term $t$, one and only one of the following statements holds:
   \begin{enumerate}
-    \item $t$ has typing derivation $\Gamma \vdash_{\Sigma, \Omega} t : A$ for some type $\vdash_{\Sigma} A$.
-    \item $t$ is not typable under $\Gamma$.
-    \item there exists $t'$ such that $t$ needs more type annotations marked in $t'$ to be inferred.
+    \item there is a typing derivation of $\Gamma \vdash_{\Sigma, \Omega} t : A$ for some type $A$.
+    \item there is no typing derivation of $\Gamma \vdash_{\Sigma, \Omega} t : A$ for any type $A$.
+    \item $t$ cannot have a synthesis mode.
   \end{enumerate}
-\end{theorem}
+\end{corollary}
+\begin{proof}
+  Combine \Cref{thm:bidirectionalisation} and \Cref{thm:bidirectional-type-synthesis}.
 
-\subsection{Decidability of Bidirectional Type Synthesis and Checking}
+\end{proof}
 
-\begin{theorem}{Uniqueness of Synthesised Types}\label{thm:unique-syn}
+\subsection{Proof for Decidability of Bidirectional Type Synthesis}
 
-\end{theorem}
-
-\begin{theorem}
-  For every type signature $\Sigma$ and bidirectional binding signature $\Omega$ for $\Sigma$ that is mode correct, the following statements hold constructively.
+\begin{theorem}[Bidirectional Type Synthesis and Checking]
+  For every type signature $\Sigma$ and bidirectional binding signature $\Omega$ for $\Sigma$ that is mode-correct, each of the following statements holds.
   \begin{enumerate}
@@
 
-    \item For every operation $o \colon \Xi \rhd [\Delta_1]A_{1}^{d_1}, \ldots, [\Delta_{n}] A^{d_n}_{n} \to A^{d}$ in $\Omega$, a context $\Gamma$, and a raw term
+%    \item For every mode-correct operation $o \colon \Xi \rhd [\Delta_1]A_{1}^{d_1}, \ldots, [\Delta_{n}] A^{d_n}_{n} \to A^{d}$, a context $\Gamma$, and a raw term
+%      \[
+%        \isTerm{\tmOp_{o}(\vec{x}_1.t_1,\ldots, \vec{x}_n.t_n)}, 
+%      \]
+%      with a mode $d$, it is decidable that there exists a substitution $\rho$ from $\Xi$ to $\emptyset$ and a typing derivation of
+%      \[
+%        \Gamma \vdash_{\Sigma, \Omega} \isTerm{\tmOp_{o}(\vec{x}_1.t_1,\ldots, \vec{x}_n.t_n)} : \simsub{A}{\rho}^\isDir{d}.
+%      \]
+    \item For every set $\Xi$ of type variables, a list of
       \[
-        \isTerm{\tmOp_{o}(\vec{x}_1.t_1,\ldots, \vec{x}_n.t_n)}, 
+        \left(\Delta_i, A_i, d_i\right) : \Cxt_{\Sigma}(\Xi) \times \Type_{\Sigma}(\Xi) \times \Mode, 
       \]
-      that has a mode $d$, it is decidable that there exists a substitution $\rho\colon \Sub{\Xi}{\emptyset}$ and a typing derivation of
+      a \emph{partial} substitution $\rho$ from $\Xi$ to $\emptyset$, a context $\Gamma$, and raw terms $t_i$'s with a mode $d_i$, either one of the following statements holds:
+      \begin{enumerate}
+        \item there exists a minimal extension $\ext{\rho}$ of $\rho$ such that there are typing derivations of
           \[
-        \Gamma \vdash_{\Sigma, \Omega} \isTerm{\tmOp_{o}(\vec{x}_1.t_1,\ldots, \vec{x}_n.t_n)} : \simsub{A}{\rho}^\isDir{d}.
+            \Gamma, \simsub{\Delta_i}{\ext{\rho}} \vdash_{\Sigma, \Omega} t_i \colon \simsub{A_i}{\ext{\rho}}^{d_i}
           \]
-    \item For every set $\Xi$ of type variables, a list
-      \[
-        \left[\left(\Delta_1, d_1, A_1\right), \ldots, \left(\Delta_n, d_n, A_n\right)\right]
-      \]
-      of triples in $\Cxt_{\Sigma}(\Xi) \times \Mode \times \Type_{\Sigma}(\Xi)$, a \emph{partial substitution} $\rho\colon\PSub{\Xi}{\emptyset}$, a context $\Gamma$, and a list of raw terms $t_i$'s with mode $d_i$ for $i = 1, \ldots, n$, there is either a minimal extension $\rho'$ of $\rho$ such that there is typing derivations of
-      \[
-        \Gamma, \simsub{\Delta_i}{\rho'} \vdash_{\Sigma, \Omega} t_i \colon \simsub{A_i}{\rho'}^{d_i}
-      \]
-      for each $i$, or there is no such extension $\rho'$ such that each raw term $t_i^{d_i}$ has a typing derivation.
+          for each $i$;
 
-    \item For every set $\Xi$ of type variables, a list $\Delta$ of types with variables in $\Xi$, a partial substitution $\rho\colon \PSub{\Xi, \emptyset}$ defined for all types in $\Delta$, and a raw term with mode $d$ then
-      \begin{enumerate}
-        \item if $d$ is $\Rightarrow$, then it is decidable that there exists a type $A$ and a typing derivation of
-          \[
-            \Gamma, \simsub{\Delta}{\rho} \vdash_{\Sigma, \Omega} t \Rightarrow A 
-          \]
-        \item if $d$ is $\Leftarrow$, then for every type $A$ it is decidable that there is a typing derivation of 
-          \[
-            \Gamma, \simsub{\Delta}{\rho} \vdash_{\Sigma, \Omega} t \Leftarrow A.
-          \]
+        \item there is no extension $\ext{\rho}$ such that all $\Gamma, \simsub{\Delta_i}{\ext{\rho}} \vdash_{\Sigma, \Omega} t_i \colon \simsub{A_i}{\ext{\rho}}^{d_i}$ has a typing derivation. 
       \end{enumerate}
@@
 \end{theorem}
-By mutual induction on raw terms $t$ with a mode. 
-Moreover, if the type $A$ of $t$ is synthesised but checked against $B$, i.e.\ $t\uparrow$, but $A \neq B$, then \Cref{thm:unique-syn} is applied to derive a contradiction.
+By induction on raw terms $t$ with a mode and on the list of arguments for operations. 
+Moreover, if the type $A$ of $t$ is synthesised and checked against $B$ but $A \neq B$, then \Cref{thm:unique-syn} is applied to derive a contradiction.
 \begin{proof}
+  \begin{enumerate}
+
+  \end{enumerate}
 \end{proof}
 
-\section{Formalisation in \Agda with Examples}
+\section{Formalisation and More Examples} \label{sec:formalisation}
 
-\section{Future Work}
+\begin{enumerate}
+  \item Bidirectional binding signature, functor, and terms (extrinsic typing, raw terms, raw terms with a mode, i.e.\ pre-typed terms).
+  \item type synthesis and checking.
+  \item examples. 
+\end{enumerate}
+
+\section{Future Work} \label{sec:future}
+
+\Josh{Getting rid of the assumption of being syntax-directed (the triangular picture)}
+
 \begin{acks}

### 2023-06-12 tex/BiSig.tex
[ tex ] Fix typos and rename `Decidability of Type Synthesis` to `Trichotomy of Annotatable Terms`

@@
 
-\title{A Formal and Practical Metatheory of Bidirectional Type Synthesis for Simple Types}
+\title{A Practical Metatheory of Bidirectional Type Synthesis for Simple Types}
 
@@
 
-\subsection{Simple Type Theories}
+\subsection{Simple Type Theory of a Signature}
 
@@
     {\vdash_{\Sigma, \Omega} \isTerm{\tmOp_o(t_1, \ldots, t_n)}}\;(\text{op}) 
-    \and \text{for $o \colon \Xi \rhd [\Delta_1]A_{1}^{d_1}, \ldots, [\Delta_{n}] A^{d_n}_{n} \to A^{d}$ in $\Omega$}
+    \and \text{for $o \colon \Xi \rhd [\Delta_1]A_{1}, \ldots, [\Delta_{n}] A_{n} \to A$ in $\Omega$}
   \end{mathpar}
@@
 
-\subsection{Bidirectional Binding Signatures and Bidirectional Type Systems}
+\subsection{Bidirectional Binding Signatures and their Bidirectional Type Systems}
 
@@
   \]
-  where $\Mode$ consists of $\chk$ for checking and $\syn$ for synthesis.
+  where $\Mode$ consists of two inhabitants $\chk$ for checking and $\syn$ for synthesis.
   Bidirectional binding signatures are just binding signatures (\Cref{def:binding-signature}) augmented with a mode for each argument and its target of a construct $\tmOp_o$ in a bidirectional system, i.e.
@@
     \and
-    \inferrule{\vdash_{\Sigma, \Omega} \isTerm{t}^\isDir{\syn}}{\vdash_{\Sigma, \Omega} (\isTerm{t \uparrow})^\isDir{\chk}}\;(\text{sub})
+    \inferrule{\vdash_{\Sigma, \Omega} \isTerm{t}^\isDir{\syn}}{\vdash_{\Sigma, \Omega} {\isTerm{t\mathord{\uparrow}}}^\isDir{\chk}}\;(\text{sub})
   \end{mathpar}
@@
 \begin{theorem}[Decidability for Bidirectionalisation] \label{thm:bidirectionalisation}
-  For every raw term $t$ and a mode $d$, it is decidable that $t$ has a mode $d$, i.e.\ $\Gamma \vdash_{\Sigma, \Omega} \isTerm{t}\,\isDir{d}$ is derivable.
+  For every raw term $t$ and a mode $d$, it is decidable that $t$ has a mode $d$, i.e.\ $\vdash_{\Sigma, \Omega} \isTerm{t}^\isDir{d}$ is derivable.
 \end{theorem}
@@
   Let $(\Sigma, \Omega)$ be a bidirectional type system where $\Omega$ is mode-correct.
-  Then, for every raw $t$ with a synthesis mode $\Rightarrow$, and any derivations of 
+  Then, for every raw $t$ with a synthesis mode and two derivations of 
   \[
-    \Gamma \vdash t \Rightarrow A
+    \Gamma \vdash_{\Sigma, \Omega} t \Rightarrow A
     \quad\text{and}\quad
-    \Gamma \vdash t \Rightarrow B
+    \Gamma \vdash_{\Sigma, \Omega} t \Rightarrow B
   \]
-  for some types $A$ and $B$, the synthesised types must be equal, i.e.\ $A = B$.
+  for some types $A$ and $B$ respectively, the synthesised types must be equal, i.e.\ $A = B$.
 \end{theorem}
 
-\subsection{Decidability of Type Synthesis}
+\subsection{Trichotomy of Annotatable Raw Terms}
 
-\begin{theorem}[Decidability for Bidirectional Typing Synthesis] \label{thm:bidirectional-type-synthesis}
+\begin{theorem}[Decidability of Bidirectional Typing Synthesis] \label{thm:bidirectional-type-synthesis}
   For every type signature $\Sigma$ and bidirectional binding signature $\Omega$ for $\Sigma$ that is mode-correct, the following two statements hold.
@@
   \begin{enumerate}
-    \item For every context $\Gamma$ and a raw term $\isTerm{t}$ with a synthesis mode, it is decidable that there exists a type $A$ and a derivation of
+    \item For every context $\Gamma$ and raw term $\isTerm{t}$ with a synthesis mode, it is decidable that there exists a type $A$ and a derivation of
       \[
@@
       \]
-    \item For every context $\Gamma$, a raw term $\isTerm{t}$ with a checking mode, and a type $A$, it is decidable that there is a derivation of
+    \item For every context $\Gamma$, raw term $\isTerm{t}$ with a checking mode, and type $A$, it is decidable that there exists a derivation of
       \[
@@
 
-\begin{corollary}[`Decidability' of Type Synthesis]
-  \LT{Still call it `decidability'?}
+\begin{corollary}[Trichotomy of Annotatable Raw Terms]
+  \LT{trichotomy over decidability?}
   Let $\Sigma$ be a type signature and $\Omega$ a bidirectional binding signature.
-  Then, for every context $\Gamma$ and raw term $t$, one and only one of the following statements holds:
+  Then, for every context $\Gamma$ and raw term $t$, exactly one of the following statements holds:
   \begin{enumerate}
-    \item there is a typing derivation of $\Gamma \vdash_{\Sigma, \Omega} t : A$ for some type $A$.
+    \item there exists a typing derivation of $\Gamma \vdash_{\Sigma, \Omega} t : A$ for some type $A$.
     \item there is no typing derivation of $\Gamma \vdash_{\Sigma, \Omega} t : A$ for any type $A$.
@@
 
-\subsection{Proof for Decidability of Bidirectional Type Synthesis}
+\subsection{Decidability of Bidirectional Type Synthesis}
 
-\begin{theorem}[Bidirectional Type Synthesis and Checking]
-  For every type signature $\Sigma$ and bidirectional binding signature $\Omega$ for $\Sigma$ that is mode-correct, each of the following statements holds.
+\begin{proof}[Proof of {\Cref{thm:bidirectional-type-synthesis}}]
+  By induction on raw terms $t$ with a mode and for the (op) rule on the list of arguments.
+Moreover, if the type $A$ of $t$ is synthesised and checked against $B$ but $A \neq B$, then \Cref{thm:unique-syn} is applied to derive a contradiction.
+
   \begin{enumerate}
-    \item For every context $\Gamma$ and a raw term $t^d$ with a mode $d$, 
-      \begin{enumerate}
-        \item if $d$ is $\Rightarrow$, then it is decidable that there exists a type $A$ and a typing derivation of
-          \[\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Rightarrow A;
-          \]
-        \item if $d$ is $\Leftarrow$, then for every type $A$ it is decidable that there is a typing derivation of
-          \[
-            \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Leftarrow A.
-          \]
-      \end{enumerate}
-
+    \item 
+        \LT{explain all cases but operation here}
 %    \item For every mode-correct operation $o \colon \Xi \rhd [\Delta_1]A_{1}^{d_1}, \ldots, [\Delta_{n}] A^{d_n}_{n} \to A^{d}$, a context $\Gamma$, and a raw term
@@
 %      \]
-    \item For every set $\Xi$ of type variables, a list of
+    \item For every set $\Xi$ of type variables, list of
       \[
@@
       \]
-      a \emph{partial} substitution $\rho$ from $\Xi$ to $\emptyset$, a context $\Gamma$, and raw terms $t_i$'s with a mode $d_i$, either one of the following statements holds:
+      \emph{partial} substitution $\rho$ from $\Xi$ to $\emptyset$, context $\Gamma$, and raw terms $t_i$'s with a mode $d_i$,
       \begin{enumerate}
-        \item there exists a minimal extension $\ext{\rho}$ of $\rho$ such that there are typing derivations of
+        \item either there is a minimal extension $\ext{\rho}$ of $\rho$ such that all     of following judgements are derivable
           \[
@@
           \]
-          for each $i$;
 
-        \item there is no extension $\ext{\rho}$ such that all $\Gamma, \simsub{\Delta_i}{\ext{\rho}} \vdash_{\Sigma, \Omega} t_i \colon \simsub{A_i}{\ext{\rho}}^{d_i}$ has a typing derivation. 
+        \item or there is no extension $\ext{\rho}$ such that all $\Gamma, \simsub{\Delta_i}{\ext{\rho}} \vdash_{\Sigma, \Omega} t_i \colon \simsub{A_i}{\ext{\rho}}^{d_i}$ has a typing derivation. 
       \end{enumerate}
-  \end{enumerate}
-\end{theorem}
-By induction on raw terms $t$ with a mode and on the list of arguments for operations. 
-Moreover, if the type $A$ of $t$ is synthesised and checked against $B$ but $A \neq B$, then \Cref{thm:unique-syn} is applied to derive a contradiction.
-\begin{proof}
-  \begin{enumerate}
-
   \end{enumerate}
@@
 \section{Formalisation and More Examples} \label{sec:formalisation}
+\LT{how to type Agda code without lhs2tex?}
 

### 2023-06-12 tex/BiSig.tex
[ tex ] Use “Trichotomy of Raw Terms” (without annotatable)

@@
 \subsection{Soundness and Completeness}
+\begin{theorem}[Soundness]\label{thm:term-soundness}
+
+\end{theorem}
+
+\begin{theorem}[Completeness]\label{thm:term-completeness}
+
+\end{theorem}
 
@@
 
-\subsection{Trichotomy of Annotatable Raw Terms}
+\subsection{Trichotomy of Raw Terms by Type Synthesis}
 
@@
 
-\begin{corollary}[Trichotomy of Annotatable Raw Terms]
+\begin{corollary}[Trichotomy of Raw Terms]
   \LT{trichotomy over decidability?}
@@
 \begin{proof}
-  Combine \Cref{thm:bidirectionalisation} and \Cref{thm:bidirectional-type-synthesis}.
+  Combine  \Cref{thm:term-soundness,thm:term-completeness,thm:bidirectionalisation} with \Cref{thm:bidirectional-type-synthesis},.
 

### 2023-06-12 tex/BiSig.tex
[ tex ] Fix Corollary 5.3 (add the condition that `t` has a synthesise mode).

@@
   \begin{enumerate}
-    \item there exists a typing derivation of $\Gamma \vdash_{\Sigma, \Omega} t : A$ for some type $A$.
-    \item there is no typing derivation of $\Gamma \vdash_{\Sigma, \Omega} t : A$ for any type $A$.
-    \item $t$ cannot have a synthesis mode.
+    \item $t$ has a synthesis mode and there exists a typing derivation of $\Gamma \vdash_{\Sigma, \Omega} t : A$ for some type $A$.
+    \item $t$ has a synthesis mode but there is no typing derivation of $\Gamma \vdash_{\Sigma, \Omega} t : A$ for any type $A$.
+    \item $t$ does not have a synthesis mode.
   \end{enumerate}

### 2023-06-13 tex/BiSig.tex
[ tex ] Improve wording (with a mode => in some mode or in synthesising / checking mode)

--- 2023-06-12 tex/BiSig.tex
@@
 
-\usepackage{xfrac,xspace,xcolor,subcaption}
+\usepackage{subcaption,xfrac,xspace,xcolor,xspace}
 \usepackage{hyperref}
@@
 
-\title{A Practical Metatheory of Bidirectional Type Synthesis for Simple Types}
+\title{A Constructive Metatheory of Bidirectional Type Synthesis for Simple Types}
 
@@
     \and
-    \inferrule{\Gamma \vdash \isTerm{t} \Rightarrow B \\ A = B}{\Gamma \vdash \isTerm{t\uparrow} \Leftarrow A}\;(\text{sub})
+    \inferrule{\Gamma \vdash \isTerm{t} \Rightarrow B \\ A = B}{\Gamma \vdash \isTerm{t\subsum} \Leftarrow A}\;(\text{sub})
     \and
@@
   \end{mathpar}
-  \caption{The extrinsic style of bidirectional simply typed $\lambda$-calculus with type variables}
+  \caption{Bidirectional simply typed $\lambda$-calculus $\Lambda^{\Leftrightarrow}_{\bto}$}
   \label{fig:bi-stlc}
@@
       \inferrule{\Theta \ni X_i}{\Theta \vdash_\Sigma X_i} \and
-      \inferrule{\Theta \vdash_{\Sigma} A_1 \\ \cdots \\ \Theta \vdash_{\Sigma} A_n}{\Theta \vdash_{\Sigma} \tyOp_i(A_1, \ldots, A_n)}
+    \inferrule{\Theta \vdash_{\Sigma} A_1 \\ \cdots \\ \Theta \vdash_{\Sigma} A_{\arity(o)}}{\Theta \vdash_{\Sigma} \tyOp_o(A_1, \ldots, A_{\arity(o)})}
     \end{mathpar}
-    where $n = \arity_i$
     \caption{Type formation}
@@
 \begin{example}\label{ex:implication}
-  Simply typed $\lambda$-calculus typically includes a binary implication type denoted by $\bto$ and some base types to ensure that the set of all types is non-empty.
+  Simply typed $\lambda$-calculus $\Lambda_{\bto}$ typically includes a binary implication type denoted by $\bto$ and some base types to ensure that the set of all types is non-empty.
   The type signature $\Sigma_{\bto}$ of simply typed $\lambda$-calculus consists of a binary operation $\bto$ and a base type (nullary operation) $N$.
-  In the case of simply typed $\lambda$-calculus with binary products, we can extend $\Sigma_{\bto}$ by adding a binary operation $\btimes$ to represent the binary product type.
-  This extended signature is denoted as $\Sigma_{\bto, \btimes}$.
+  In the case of simply typed $\lambda$-calculus $\Lambda_{\bto, \btimes}$ with binary products, we can extend $\Sigma_{\bto}$ by adding a binary operation $\btimes$ to represent the binary product type, denoted as $\Sigma_{\bto, \btimes}$.
 \end{example}
@@
   \end{enumerate}
-  For brevity, we write $o \colon \Xi \rhd (\Delta_1)A_{1}, \ldots, \left[\Delta_{n}\right] A_{n} \to A$ to indicate an operation $o$ with its arity. 
+  For brevity, we write $o \colon \Xi \rhd [\Delta_1]A_{1}, \ldots, \left[\Delta_{n}\right] A_{n} \to A$ to indicate an operation with its arity. 
 \end{definition}
@@
 
-\subsection{Simple Type Theory of a Signature}
+\subsection{Simple Type Theory for a Signature}
 
@@
   \end{mathpar}
-  \caption{Annotatable raw terms for $(\Sigma, \Omega)$}
+  \caption{Raw terms for $(\Sigma, \Omega)$ with annotation}
 \end{figure}
@@
   \[
-    \left(\Xi, \left[\left(\Delta_1, A_1, d_1\right), \ldots, \left(\Delta_{n}, A_{n}, d_n\right) \right], A, d\right)
+    \left(\Xi, \left[\left(\Delta_1; A_1; d_1\right), \ldots, \left(\Delta_{n}; A_{n}; d_n\right) \right], A, d\right)
   \]
@@
 
-  For brevity, we write $o \colon \Xi \rhd [\Delta_1]A_{1}^{d_1}, \ldots, [\Delta_{n}] A^{d_n}_{n} \to A^{d}$ to indicate an operation $o$ with its arity. 
+  For brevity, we write $o \colon \Xi \rhd [\Delta_1]A_{1}^{d_1}, \ldots, [\Delta_{n}] A^{d_n}_{n} \to A^{d}$ to indicate an operation with its arity. 
 \end{definition}
@@
 
-\LT{deriving bidirectional binding signature from annotation}
-
-\LT{Signature Erasure}
-\begin{theorem}[Signature annotation]
-
-\end{theorem}
+\LT{Some remark about signature erasure and annotation; introduce the notation $\erase{\Omega}$}
 
@@
     \boxed{\vdash_{\Sigma, \Omega} \isTerm{t}^\isDir{d}}
+    \quad\text{where $\vdash_{\Sigma, \erase{\Omega}} \isTerm{t}$}
     \\
@@
     \and
-    \inferrule{\cdot \vdash_{\Sigma} A \\ \vdash_{\Sigma, \Omega}\isTerm{t}^\isDir{\chk}}{\vdash_{\Sigma, \Omega} \isTerm{t \annote A}^\isDir{\syn}}\;(\text{anno})
+    \inferrule{\cdot \vdash_{\Sigma} A \\ \vdash_{\Sigma, \Omega}\isTerm{t}^\isDir{\chk}}{\vdash_{\Sigma, \Omega} (\isTerm{t \annote A})^\isDir{\syn}}\;(\text{anno})
     \and
-    \inferrule{\vdash_{\Sigma, \Omega} \isTerm{t}^\isDir{\syn}}{\vdash_{\Sigma, \Omega} {\isTerm{t\mathord{\uparrow}}}^\isDir{\chk}}\;(\text{sub})
+    \inferrule{\vdash_{\Sigma, \Omega} \isTerm{t}^\isDir{\syn}}{\vdash_{\Sigma, \Omega} {\isTerm{t\subsum}}^\isDir{\chk}}\;(\text{ sub})
   \end{mathpar}
@@
   \end{mathpar}
-  \caption{Raw terms with a mode $d$ for a bidirectional type system $(\Sigma, \Omega)$}
+  \caption{Raw terms in some mode $d$ for a bidirectional type system $(\Sigma, \Omega)$}
 \end{figure}
@@
     \boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A^\isDir{d}}
-    \quad \text{where $\vdash_{\Sigma, \Omega} \isTerm{t}^\isDir{d}$} 
+    \quad \text{where $\vdash_{\Sigma, \erase{\Omega}} \isTerm{t}$ and $\vdash_{\Sigma, \Omega} \isTerm{t}^\isDir{d}$} 
     \\
-    \inferrule{(x : A) \in \Gamma}{\Gamma \vdash_{\Sigma, \Omega} \isTerm{x} : A^\syn}\;(\var)
+    \inferrule{(x : A) \in \Gamma}{\Gamma \vdash_{\Sigma, \Omega} \isTerm{x} : A^\isDir{\syn}}\;(\var)
     \and
-    \inferrule{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : B^\chk \\ A = B}{\mid \Gamma \vdash_{\Sigma, \Omega} (\isTerm{t \annote B}):  A^\syn}\;(\text{anno})
+    \inferrule{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A^\isDir{\chk}}{\Gamma \vdash_{\Sigma, \Omega} (\isTerm{t \annote B}):  A^\isDir{\syn}}\;(\text{anno})
     \and
-    \inferrule{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : B^{\syn} \\ A = B}{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t\uparrow} : A^\chk}\;(\text{sub})
+    \inferrule{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : B^{\isDir{\syn}} \\ A = B}{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t\subsum} : A^\isDir{\chk}}\;(\text{sub})
     \\
@@
     \\
-    \text{Abbreviations: $\boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Rightarrow A} \defeq \boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A^\syn}$ and $\boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Leftarrow A} \defeq \boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A^\chk}$}
+    \text{Abbreviations: $\boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Rightarrow A} \defeq \boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A^\isDir{\syn}}$ and $\boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Leftarrow A} \defeq \boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A^\isDir{\chk}}$}
   \end{mathpar}
@@
 \subsection{Bidirectionalisation}
-\begin{theorem}[Decidability for Bidirectionalisation] \label{thm:bidirectionalisation}
-  For every raw term $t$ and a mode $d$, it is decidable that $t$ has a mode $d$, i.e.\ $\vdash_{\Sigma, \Omega} \isTerm{t}^\isDir{d}$ is derivable.
-\end{theorem}
+\begin{proposition}[Decidability for Bidirectionalisation] \label{thm:bidirectionalisation}
+  For every raw term $t$ and a mode $d$, it is decidable that $t$ is in mode $d$, i.e.\ $\vdash_{\Sigma, \Omega} \isTerm{t}^\isDir{d}$ is derivable.
+\end{proposition}
 
@@
   Let $(\Sigma, \Omega)$ be a bidirectional type system where $\Omega$ is mode-correct.
-  Then, for every raw $t$ with a synthesis mode and two derivations of 
+  Then, for every raw term $t$ in synthesising mode and two derivations of 
   \[
@@
   \begin{enumerate}
-    \item For every context $\Gamma$ and raw term $\isTerm{t}$ with a synthesis mode, it is decidable that there exists a type $A$ and a derivation of
+    \item For every context $\Gamma$ and raw term $\isTerm{t}$ in synthesising mode, it is decidable that there is a type~$A$ and a derivation of
       \[
@@
       \]
-    \item For every context $\Gamma$, raw term $\isTerm{t}$ with a checking mode, and type $A$, it is decidable that there exists a derivation of
+    \item For every context $\Gamma$, raw term $\isTerm{t}$ in checking mode, and type $A$, it is decidable that there is a derivation of
       \[
@@
 \begin{corollary}[Trichotomy of Raw Terms]
-  \LT{trichotomy over decidability?}
   Let $\Sigma$ be a type signature and $\Omega$ a bidirectional binding signature.
@@
   \begin{enumerate}
-    \item there exists a typing derivation of $\Gamma \vdash_{\Sigma, \Omega} t : A$ for some type $A$.
-    \item there is no typing derivation of $\Gamma \vdash_{\Sigma, \Omega} t : A$ for any type $A$.
-    \item $t$ cannot have a synthesis mode.
+    \item $t$ is synthesising and there exists a typing derivation of\, $\Gamma \vdash_{\Sigma, \Omega} t : A$ for some type $A$.
+    \item $t$ is synthesising but there is no typing derivation of\, $\Gamma \vdash_{\Sigma, \Omega} t : A$ for any type $A$.
+    \item $t$ is not synthesising.
   \end{enumerate}
@@
 \begin{proof}
-  Combine  \Cref{thm:term-soundness,thm:term-completeness,thm:bidirectionalisation} with \Cref{thm:bidirectional-type-synthesis},.
+  Combine  \Cref{thm:term-soundness,thm:term-completeness,thm:bidirectionalisation} with \Cref{thm:bidirectional-type-synthesis}.
 
@@
 \begin{proof}[Proof of {\Cref{thm:bidirectional-type-synthesis}}]
-  By induction on raw terms $t$ with a mode and for the (op) rule on the list of arguments.
+  By induction on raw terms $t$ in some mode and for the $(\text{op})$ rule on the list of arguments.
   Moreover, if the type $A$ of $t$ is synthesised and checked against $B$ but $A \neq B$, then \Cref{thm:unique-syn} is applied to derive a contradiction.
@@
       \]
-      \emph{partial} substitution $\rho$ from $\Xi$ to $\emptyset$, context $\Gamma$, and raw terms $t_i$'s with a mode $d_i$,
+      \emph{partial} substitution $\rho$ from $\Xi$ to $\emptyset$, context $\Gamma$, and raw terms $t_i$'s in mode $d_i$,
       \begin{enumerate}

### 2023-06-13 tex/sec1-intro.tex
[ tex ] Split BiSig.tex into smaller files

--- 2023-06-12 tex/BiSig.tex
@@
-\documentclass[acmsmall,screen]{acmart}
-\usepackage[T1]{fontenc}
-\usepackage[utf8]{inputenc}
-\usepackage[british]{babel}
-
-\usepackage{bm,braket,mathtools,mathpartir,bussproofs}
-\EnableBpAbbreviations
-\usepackage[all,cmtip]{xy}
-
-\usepackage{xfrac,xspace,xcolor,subcaption}
-\usepackage{hyperref}
-\usepackage[capitalise,noabbrev]{cleveref}
-\usepackage[inline]{enumitem} % for environment enumerate*
-\setlist[enumerate]{mode=unboxed}
-
-\setlength {\marginparwidth }{2cm}
-\usepackage[color=yellow,textsize=small]{todonotes}
-\newcommand{\LT}[1]{\todo[author=LT,inline,color=green!40,caption={}]{{#1}}}
-\newcommand{\Josh}[1]{\todo[author=Josh,inline,color=green!40,caption={}]{{#1}}}
-
-\usepackage{subfiles}
-
-\usepackage{microtype}
-
-\citestyle{acmauthoryear}
-
-\AtEndPreamble{%
-\theoremstyle{acmdefinition}
-\newtheorem{remark}[theorem]{Remark}}
-
-\input{type-notation} %% for type theory notation copied from HoTT Book
-\input{macros}
-
-\begin{document}
-
-\author{Liang-Ting Chen}
-\email{liang.ting.chen.tw@gmail.com}
-\orcid{0000-0002-3250-1331}
-\author{Hsiang-Shang Ko}
-\orcid{0000-0002-2439-1048}
-\email{joshko@iis.sinica.edu.tw}
-
-\affiliation{%
-  \institution{Academia Sinica}
-  \streetaddress{128 Academia Road}
-  \city{Taipei}
-  \country{Taiwan}
-  \postcode{115}
-}
-
-\title{A Practical Metatheory of Bidirectional Type Synthesis for Simple Types}
-
-\begin{abstract}
-  This paper proposes a formal theory of bidirectional typing.
-\end{abstract}
-
-\maketitle
+%! TEX root = BiSig.tex
 
@@
 \cite{Cimini2020,Cimini2022}
-\citep{Fiore2022}
-\cite{Allais2021,Gheri2020,Ahrens2022}
 
-\subsubsection{Algebraic theory of type theories}
-\cite{Fiore1999,Fiore2022,Ahrens2021,Arkor2020,Hamana2011}
+\citep{Ahrens2018,Fiore2022,Gheri2020,Ahrens2022}
+\cite{Allais2021}
+
+\subsubsection{Theories of abstract syntax with variable binding}
+\cite{Fiore1999,Hirschowitz2010,Ahrens2018,Fiore2022,Ahrens2021,Arkor2020,Hirschowitz2022}
+\cite{Fiore2013,Hamana2011,Hamana2022}
 
@@
   \LT{better title?}
-
-\section{Core Ideas with Bidirectional Simply Typed \texorpdfstring{$\lambda$}{lambda}-Calculus} \label{sec:key-ideas}
-\begin{figure}
-  \centering
-  \small
-  \begin{mathpar}
-    \boxed{\Gamma \vdash \isTerm{t} \Rightarrow A}
-    \quad\text{and}\quad \boxed{\Gamma \vdash \isTerm{t} \Leftarrow A}
-    \\
-    \inferrule{(x : A) \in \Gamma}{\Gamma \vdash \isTerm{x} \Rightarrow A}\;(\text{var}^\Rightarrow)
-    \and
-    \inferrule{\Gamma \vdash \isTerm{t} \Leftarrow B \\ A = B}{\Gamma \vdash \isTerm{t : B}\Rightarrow A}\;(\text{anno})
-    \and
-    \inferrule{\Gamma \vdash \isTerm{t} \Rightarrow B \\ A = B}{\Gamma \vdash \isTerm{t\uparrow} \Leftarrow A}\;(\text{sub})
-    \and
-    \inferrule{\Gamma \vdash \isTerm{t} \Rightarrow A \bto B \\ \Gamma \vdash \isTerm{u} \Leftarrow A}{\Gamma \vdash \isTerm{t\;u} \Rightarrow B}\;(\text{app})
-    \and
-    \inferrule{\Gamma, \isTerm{x} : A \vdash \isTerm{t} \Leftarrow B}{\Gamma \vdash \isTerm{\lam{x}t} \Leftarrow A \bto B}\;(\text{abs})
-  \end{mathpar}
-  \caption{The extrinsic style of bidirectional simply typed $\lambda$-calculus with type variables}
-  \label{fig:bi-stlc}
-\end{figure}
-\begin{remark}
-  We avoid using the term `function type' and its conventional notation $\to$ at the object level on purpose, as it may be confused with function types in our type-theoretic meta-language.
-\end{remark}
-
-\section{Definitions for Simple Type Theories and Bidirectional Type Systems}\label{sec:defs}
-
-\subsection{Simple Types}
-\begin{definition}
-  A \emph{(simple type) signature} $\Sigma$ consists of a set $I$ and an \emph{arity} function $\arity\colon I \to \mathbb{N}$.
-  An inhabitant $i : I$ is meant to represent the \emph{$i$-th operation} $\tyOp_i$ and its \emph{number} $ \arity(i)$ of arguments.
-
-  The judgements of \emph{$\Sigma$-types} and \emph{$\Sigma$-contexts} over a variable set $\Theta$ are defined inductively by \Cref{fig:simple-type,fig:simple-context} respectively.
-We write $A : \Type_{\Sigma}(\Theta)$ for $\Theta \vdash_{\Sigma} A$ and $\Gamma \colon \Cxt_{\Sigma}(\Theta)$ for $\Theta \vdash_{\Sigma} \Gamma$ to disambiguate $\Sigma$-types and $\Sigma$-contexts in the case of any confusion.
-\end{definition}
-
-\begin{figure}
-  \begin{minipage}[b]{.55\textwidth}
-    \centering
-    \small
-    \begin{mathpar}
-      \boxed{\Theta\vdash_{\Sigma} A} \\
-      \inferrule{\Theta \ni X_i}{\Theta \vdash_\Sigma X_i} \and
-      \inferrule{\Theta \vdash_{\Sigma} A_1 \\ \cdots \\ \Theta \vdash_{\Sigma} A_n}{\Theta \vdash_{\Sigma} \tyOp_i(A_1, \ldots, A_n)}
-    \end{mathpar}
-    where $n = \arity_i$
-    \caption{Type formation}
-    \label{fig:simple-type}
-  \end{minipage}
-  \begin{minipage}[b]{.4\textwidth}
-    \centering
-    \small
-    \begin{mathpar}
-      \boxed{\Theta \vdash_{\Sigma} \Gamma} \\
-      \inferrule{ }{\Theta \vdash_{\Sigma} \cdot }\and
-      \inferrule{\Theta \vdash_{\Sigma} A \\ \Theta\vdash_{\Sigma} \Gamma}{\Theta \vdash_{\Sigma} \Gamma, A}
-    \end{mathpar}
-    \caption{Context formation}
-  \label{fig:simple-context}
-  \end{minipage}
-\end{figure}
-
-\begin{example}\label{ex:implication}
-  Simply typed $\lambda$-calculus typically includes a binary implication type denoted by $\bto$ and some base types to ensure that the set of all types is non-empty.
-  The type signature $\Sigma_{\bto}$ of simply typed $\lambda$-calculus consists of a binary operation $\bto$ and a base type (nullary operation) $N$.
-  In the case of simply typed $\lambda$-calculus with binary products, we can extend $\Sigma_{\bto}$ by adding a binary operation $\btimes$ to represent the binary product type.
-  This extended signature is denoted as $\Sigma_{\bto, \btimes}$.
-\end{example}
-
-\subsection{Binding Signatures} \label{subsec:binding-sig}
-\begin{definition}\label{def:binding-signature}
-  For a type signature $\Sigma$, a \emph{binding signature} $\Omega$ consists of a set $O$ and a function
-  \[
-    \mathit{ar}\colon O \to \sum_{\Xi : \N} \left(\Cxt_\Sigma(\Xi) \times \Type_\Sigma(\Xi)\right)^* \times \Type_\Sigma(\Xi).
-  \]
-  Each inhabitant $o: O$ is meant to represent a term construct $\tmOp_o$ in a simple type theory with a triple $\left(\Xi, \left[\left(\Delta_1; A_1\right), \ldots, \left(\Delta_{n}; A_{n}\right) \right], A\right)$
-  instead of a number as its arity $\arity_o$ where
-  \begin{enumerate}
-    \item $\Xi$ is the number of type variables, 
-    \item $A : \Type_\Sigma(\Xi)$ is the target type of $\tmOp_o$, and
-    \item $\left[\left(\Delta_1; A_{1}\right), \ldots, \left(\Delta_{n}; A_{n}\right) \right]$ is a list of pairs where
-    \item the $i$-th pair $(\Delta_i, A_i) :\Cxt_{\Sigma}(\Xi) \times \Term_{\Sigma}(\Xi)$ are types of the binding variables and the type of the $i$-th argument of $\tmOp_{o}$.
-  \end{enumerate}
-  For brevity, we write $o \colon \Xi \rhd (\Delta_1)A_{1}, \ldots, \left[\Delta_{n}\right] A_{n} \to A$ to indicate an operation $o$ with its arity. 
-\end{definition}
-\begin{remark}[Terminology on binding signature]
-  \cite{Aczel1978,Fiore2010}
-\end{remark}
-
-\begin{example} \label{ex:STLC-sig}
-  Given the type signature $\Sigma_{\bto}$ for the implication type, the term signature $\Lambda_{\bto}$ for simply typed $\lambda$-calculus can be described by operations
-  \begin{align*}
-    \mathsf{app}\colon A, B \rhd (A \bto B), A \to B && \mathsf{abs}\colon A , B \rhd [A]B \to (A \bto B)
-  \end{align*}
-    or, verbosely, the signature $\Lambda_{\bto}$ consists of the type $O_\Lambda = \{\app, \abs\}$ with arities
-  \begin{align*}
-    \arity(\app) = (\{A, B\}, [(\cdot; A \bto B), (\cdot; A)], B)
-    && 
-    \arity(\abs) = (\{A, B\}, [(\cdot, A; B)], A \bto B)
-  \end{align*}
-  For the type signature $\Sigma_{\bto, \btimes}$, the term signature $\Omega_{\bto, \btimes}$ for simply typed $\lambda$-calculus with finite products has three additional operations
-to $\Omega_{\bto}$:
-  \begin{align*}
-    \mathsf{pair}\colon A, B \rhd A, B \to A \btimes B
-    && \fst \colon A, B \rhd A \btimes B \to A
-    && \snd \colon A, B \rhd A \btimes B \to B
-  \end{align*}
-\end{example}
-
-First, it is noteworthy that the set of term constructs in a type theory need not be finite.
-For instance, a type theory may adopt a spine application which takes indefinitely many arguments---for each number $n$ of arguments an $(n+1)$-ary application construct can be introduced, so even if each operation has a definite number of arguments by definition a spine application is still expressible.
-
-More importantly, the inclusion of the set of variables used in an operation is a salient feature of our binding signatures.
-Instead of treating application as a family of constructs $\app_{A, B}$ indexed by all types $A$ and $B$, as done by \citet{Fiore2022}, we are able to identify them as a single construct $\app$.
-This is not only for brevity but also necessary to compare the type equality during type synthesis and checking.
-
-%\subsection{Intrinsically Typed Terms}
-%
-%Intrinsically typed terms for a type signature $\Sigma$ and a term signature $\Omega$ are nothing more than derivations of the intrinsic typing judgement $\Gamma \vdash_{\Sigma, \Omega} A$ constructed with only the rule $(\var)$ and the rule scheme $(\tmOp)$ displayed in \Cref{fig:intrinsic-typing}.
-%
-%The substitution $\rho\colon \Sub{\Xi}{\emptyset}$ is used to instantiate variables in the local context $\Xi$ with concrete types.
-%Accordingly, $\rho$ has to be applied to types that appear in the arity to construct a term by $\tmOp$ to ensure all types are well-formed without any use of type variables.
-%
-%\begin{figure}
-%  \centering
-%  \small
-%  \begin{mathpar}
-%    \boxed{\Gamma \vdash_{\Sigma, \Omega} A} \\
-%    \inferrule{A \in \Gamma}{\Gamma \vdash_{\Sigma, \Omega} A}\;(\var)
-%    \and
-%    \inferrule{\rho\colon \Sub{\Xi}{\emptyset}  \\ \Gamma, \simsub{\Delta_{1}}{\rho} \vdash_{\Sigma, \Omega} \simsub{A_{1}}{\rho} \quad\cdots\quad \Gamma, \simsub{\Delta_{n}}{\rho} \vdash_{\Sigma, \Omega} \simsub{A_{n}}{\rho}}
-%    {\Gamma \vdash_{\Sigma, \Omega} \simsub{A}{\rho}}\;(\tmOp)
-%    \and \text{for $o\colon \Xi \rhd [\Delta_1]A_1, \ldots, [\Delta_{n}]A_{n} \to A$ in $\Omega$}
-%  \end{mathpar}
-%  \caption{Intrinsic typing rules for a simple type theory $(\Sigma, \Omega)$}
-%  \label{fig:intrinsic-typing}
-%\end{figure}
-
-\subsection{Simple Type Theory of a Signature}
-
-\begin{figure}
-  \centering
-  \small
-  \begin{mathpar}
-    \boxed{\vdash_{\Sigma, \Omega} \isTerm{t}}
-    \\
-    \inferrule{x : \Identifier}{\vdash_{\Sigma, \Omega} \isTerm{x}}\;(\text{var})
-    \and
-    \inferrule{\cdot \vdash_{\Sigma} A \\ \vdash_{\Sigma, \Omega}\isTerm{t}}{\vdash_{\Sigma, \Omega} \isTerm{t \annote A}}\;(\text{anno})
-    \\
-    \inferrule{\vdash_{\Sigma, \Omega} \isTerm{t_1} \quad \cdots \quad \vdash_{\Sigma, \Omega} \isTerm{t_n}}
-    {\vdash_{\Sigma, \Omega} \isTerm{\tmOp_o(t_1, \ldots, t_n)}}\;(\text{op}) 
-    \and \text{for $o \colon \Xi \rhd [\Delta_1]A_{1}, \ldots, [\Delta_{n}] A_{n} \to A$ in $\Omega$}
-  \end{mathpar}
-  \caption{Annotatable raw terms for $(\Sigma, \Omega)$}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \small
-  \begin{mathpar}
-    \boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A} \quad \text{where $\vdash_{\Sigma, \Omega} t$} \\
-    \inferrule{(x : A) \in \Gamma}{\Gamma \vdash_{\Sigma, \Omega} \isTerm{x} : A}\;(\var)
-    \and
-    \inferrule{\Gamma \vdash \isTerm{t} : A}{\Gamma \vdash (\isTerm{t \annote A}) : A}\;(\text{anno})
-    \and
-    \inferrule{\rho : \Sub{\Xi}{\emptyset} \\ \Gamma, \isTerm{\vec{x}_1} : \simsub{\Delta_{1}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_1} : \simsub{A_{1}}{\rho} \quad\cdots\quad \Gamma, \isTerm{\vec{x}_n} : \simsub{\Delta_{n}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_n} : \simsub{A_{n}}{\rho}}
-    {\Gamma \vdash_{\Sigma, \Omega} \isTerm{\tmOp_o(\vec{x}_1.\,t_1; \ldots; \vec{x}_n.\,t_n)} : \simsub{A}{\rho}}\;(\tmOp)
-    \and \text{for $o\colon \Xi \rhd [\Delta_1]A_1, \ldots, [\Delta_{n}]A_{n} \to A$ in $\Omega$}
-  \end{mathpar}
-  \caption{Typing rules for a simple type theory $(\Sigma, \Omega)$ with annotation}
-  \label{fig:extrinsic-typing}
-\end{figure}
-
-\subsection{Bidirectional Binding Signatures and their Bidirectional Type Systems}
-
-\begin{definition}
-  For a type signature $\Sigma$, a \emph{bidirectional binding signature} $\Omega$ is a set $O$ with a function
-  \[
-    \mathit{ar}\colon O \to \sum_{\Xi : \N} \left(\Cxt_{\Sigma}(\Xi) \times \Type_{\Sigma}(\Xi) \times \isDir{\Mode}\right)^* \times \Type_{\Sigma}(\Xi) \times \isDir{\Mode}.
-  \]
-  where $\Mode$ consists of two inhabitants $\chk$ for checking and $\syn$ for synthesis.
-  Bidirectional binding signatures are just binding signatures (\Cref{def:binding-signature}) augmented with a mode for each argument and its target of a construct $\tmOp_o$ in a bidirectional system, i.e.
-  an arity is a $4$-tuple
-  \[
-    \left(\Xi, \left[\left(\Delta_1, A_1, d_1\right), \ldots, \left(\Delta_{n}, A_{n}, d_n\right) \right], A, d\right)
-  \]
-  where $d$ and $d_i$'s indicate the modes of a construct and its arguments respectively.
-
-  For brevity, we write $o \colon \Xi \rhd [\Delta_1]A_{1}^{d_1}, \ldots, [\Delta_{n}] A^{d_n}_{n} \to A^{d}$ to indicate an operation $o$ with its arity. 
-\end{definition}
-
-\begin{example}
-  The term signature $\Lambda_{\bto}^{\leftrightarrows}$ for bidirectional simply typed $\lambda$-calculus introduced in \Cref{subsec:binding-sig} can be specified by operations 
-  \begin{align*}
-    \mathsf{app}\colon A, B \rhd (A \bto B)^{\Rightarrow}, A^{\Leftarrow} \to B^{\Rightarrow} &&
-    \mathsf{abs}\colon A , B \rhd [A]B^{\Leftarrow} \to (A \bto B)^{\Leftarrow}
-  \end{align*}
-  extending $\Lambda_{\bto}$ (\Cref{ex:STLC-sig}) with the mode information.
-\end{example}
-
-\LT{deriving bidirectional binding signature from annotation}
-
-\LT{Signature Erasure}
-\begin{theorem}[Signature annotation]
-
-\end{theorem}
-
-\begin{figure}
-  \centering
-  \small
-  \begin{mathpar}
-    \boxed{\vdash_{\Sigma, \Omega} \isTerm{t}^\isDir{d}}
-    \\
-    \inferrule{x : \Identifier}{\vdash_{\Sigma, \Omega} \isTerm{x}^\isDir{\syn}}\;(\text{var})
-    \and
-    \inferrule{\cdot \vdash_{\Sigma} A \\ \vdash_{\Sigma, \Omega}\isTerm{t}^\isDir{\chk}}{\vdash_{\Sigma, \Omega} \isTerm{t \annote A}^\isDir{\syn}}\;(\text{anno})
-    \and
-    \inferrule{\vdash_{\Sigma, \Omega} \isTerm{t}^\isDir{\syn}}{\vdash_{\Sigma, \Omega} {\isTerm{t\mathord{\uparrow}}}^\isDir{\chk}}\;(\text{sub})
-  \end{mathpar}
-  \begin{mathpar}
-    \inferrule{\vdash_{\Sigma, \Omega} \isTerm{t_1}^\isDir{d_1} \quad \cdots \quad \vdash_{\Sigma, \Omega} \isTerm{t_n}^\isDir{d_n}}
-    {\vdash_{\Sigma, \Omega} \isTerm{\tmOp_o(\vec{x}_1.\, t_1; \ldots;\vec{x}_n.\, t_n)}^\isDir{d}}\;(\text{op})
-    \and \text{for $o \colon \Xi \rhd [\Delta_1]A_{1}^{d_1}, \ldots, [\Delta_{n}] A^{d_n}_{n} \to A^{d}$ in $\Omega$}
-  \end{mathpar}
-  \caption{Raw terms with a mode $d$ for a bidirectional type system $(\Sigma, \Omega)$}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \small
-  \begin{mathpar}
-    \boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A^\isDir{d}}
-    \quad \text{where $\vdash_{\Sigma, \Omega} \isTerm{t}^\isDir{d}$} 
-    \\
-    \inferrule{(x : A) \in \Gamma}{\Gamma \vdash_{\Sigma, \Omega} \isTerm{x} : A^\syn}\;(\var)
-    \and
-    \inferrule{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : B^\chk \\ A = B}{\mid \Gamma \vdash_{\Sigma, \Omega} (\isTerm{t \annote B}):  A^\syn}\;(\text{anno})
-    \and
-    \inferrule{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : B^{\syn} \\ A = B}{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t\uparrow} : A^\chk}\;(\text{sub})
-    \\
-    \inferrule{\rho\colon \Sub{\Xi}{\emptyset} \\ \Gamma, \simsub{\isTerm{\vec{x}_{1}} : \Delta_{1}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_{1}} : \simsub{A_{1}}{\rho}^{\isDir{d_1}} \\
-      \cdots \\
-    \Gamma, \simsub{\isTerm{\vec{x}_{n}} : \Delta_{n}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_n} : \simsub{A_{n}}{\rho}^{\isDir{d_n}}}
-  {\Gamma \vdash_{\Sigma, \Omega} \isTerm{\tmOp_o(\vec{x}_1 .\, t_{1}; \ldots; \vec{x}_{n}.\, t_{n})} : \simsub{A}{\rho}^{\isDir{d}}} \; (\tmOp)
-    \and \text{for $o \colon \Xi \rhd [\Delta_1]A_{1}^{d_1}, \ldots, [\Delta_{n}] A^{d_n}_{n} \to A^{d}$ in $\Omega$}
-    \\
-    \text{Abbreviations: $\boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Rightarrow A} \defeq \boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A^\syn}$ and $\boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Leftarrow A} \defeq \boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A^\chk}$}
-  \end{mathpar}
-  \caption{Typing rules for a bidirectional system $(\Sigma, \Omega)$}
-\end{figure}
-
-\begin{example}
-
-\end{example}
-
-\section{Soundness, Completeness, and Bidirectionalisation} \label{sec:annotatability}
-
-\subsection{Soundness and Completeness}
-\begin{theorem}[Soundness]\label{thm:term-soundness}
-
-\end{theorem}
-
-\begin{theorem}[Completeness]\label{thm:term-completeness}
-
-\end{theorem}
-
-\subsection{Bidirectionalisation}
-\begin{theorem}[Decidability for Bidirectionalisation] \label{thm:bidirectionalisation}
-  For every raw term $t$ and a mode $d$, it is decidable that $t$ has a mode $d$, i.e.\ $\vdash_{\Sigma, \Omega} \isTerm{t}^\isDir{d}$ is derivable.
-\end{theorem}
-
-\subsection{Completeness versus Annotatability}
-
-\section{Bidirectional Type Synthesis and Checking} \label{sec:type-synthesis}
-
-\subsection{Mode Correctness}
-
-\begin{theorem}[Uniqueness of Synthesised Types]\label{thm:unique-syn}
-  Let $(\Sigma, \Omega)$ be a bidirectional type system where $\Omega$ is mode-correct.
-  Then, for every raw $t$ with a synthesis mode and two derivations of 
-  \[
-    \Gamma \vdash_{\Sigma, \Omega} t \Rightarrow A
-    \quad\text{and}\quad
-    \Gamma \vdash_{\Sigma, \Omega} t \Rightarrow B
-  \]
-  for some types $A$ and $B$ respectively, the synthesised types must be equal, i.e.\ $A = B$.
-\end{theorem}
-
-\subsection{Trichotomy of Raw Terms by Type Synthesis}
-
-\begin{theorem}[Decidability of Bidirectional Typing Synthesis] \label{thm:bidirectional-type-synthesis}
-  For every type signature $\Sigma$ and bidirectional binding signature $\Omega$ for $\Sigma$ that is mode-correct, the following two statements hold.
-
-  \begin{enumerate}
-    \item For every context $\Gamma$ and raw term $\isTerm{t}$ with a synthesis mode, it is decidable that there exists a type $A$ and a derivation of
-      \[
-        \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Rightarrow A.
-      \]
-    \item For every context $\Gamma$, raw term $\isTerm{t}$ with a checking mode, and type $A$, it is decidable that there exists a derivation of
-      \[
-        \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Leftarrow A.
-      \]
-  \end{enumerate}
-\end{theorem}
-
-\begin{corollary}[Trichotomy of Raw Terms]
-  \LT{trichotomy over decidability?}
-  Let $\Sigma$ be a type signature and $\Omega$ a bidirectional binding signature.
-  Then, for every context $\Gamma$ and raw term $t$, exactly one of the following statements holds:
-  \begin{enumerate}
-    \item there exists a typing derivation of $\Gamma \vdash_{\Sigma, \Omega} t : A$ for some type $A$.
-    \item there is no typing derivation of $\Gamma \vdash_{\Sigma, \Omega} t : A$ for any type $A$.
-    \item $t$ cannot have a synthesis mode.
-  \end{enumerate}
-\end{corollary}
-\begin{proof}
-  Combine  \Cref{thm:term-soundness,thm:term-completeness,thm:bidirectionalisation} with \Cref{thm:bidirectional-type-synthesis},.
-
-\end{proof}
-
-\subsection{Decidability of Bidirectional Type Synthesis}
-
-\begin{proof}[Proof of {\Cref{thm:bidirectional-type-synthesis}}]
-  By induction on raw terms $t$ with a mode and for the (op) rule on the list of arguments.
-Moreover, if the type $A$ of $t$ is synthesised and checked against $B$ but $A \neq B$, then \Cref{thm:unique-syn} is applied to derive a contradiction.
-
-  \begin{enumerate}
-    \item 
-        \LT{explain all cases but operation here}
-%    \item For every mode-correct operation $o \colon \Xi \rhd [\Delta_1]A_{1}^{d_1}, \ldots, [\Delta_{n}] A^{d_n}_{n} \to A^{d}$, a context $\Gamma$, and a raw term
-%      \[
-%        \isTerm{\tmOp_{o}(\vec{x}_1.t_1,\ldots, \vec{x}_n.t_n)}, 
-%      \]
-%      with a mode $d$, it is decidable that there exists a substitution $\rho$ from $\Xi$ to $\emptyset$ and a typing derivation of
-%      \[
-%        \Gamma \vdash_{\Sigma, \Omega} \isTerm{\tmOp_{o}(\vec{x}_1.t_1,\ldots, \vec{x}_n.t_n)} : \simsub{A}{\rho}^\isDir{d}.
-%      \]
-    \item For every set $\Xi$ of type variables, list of
-      \[
-        \left(\Delta_i, A_i, d_i\right) : \Cxt_{\Sigma}(\Xi) \times \Type_{\Sigma}(\Xi) \times \Mode, 
-      \]
-      \emph{partial} substitution $\rho$ from $\Xi$ to $\emptyset$, context $\Gamma$, and raw terms $t_i$'s with a mode $d_i$,
-      \begin{enumerate}
-        \item either there is a minimal extension $\ext{\rho}$ of $\rho$ such that all     of following judgements are derivable
-          \[
-            \Gamma, \simsub{\Delta_i}{\ext{\rho}} \vdash_{\Sigma, \Omega} t_i \colon \simsub{A_i}{\ext{\rho}}^{d_i}
-          \]
-
-        \item or there is no extension $\ext{\rho}$ such that all $\Gamma, \simsub{\Delta_i}{\ext{\rho}} \vdash_{\Sigma, \Omega} t_i \colon \simsub{A_i}{\ext{\rho}}^{d_i}$ has a typing derivation. 
-      \end{enumerate}
-  \end{enumerate}
-\end{proof}
-
-\section{Formalisation and More Examples} \label{sec:formalisation}
-\LT{how to type Agda code without lhs2tex?}
-
-\begin{enumerate}
-  \item Bidirectional binding signature, functor, and terms (extrinsic typing, raw terms, raw terms with a mode, i.e.\ pre-typed terms).
-  \item type synthesis and checking.
-  \item examples. 
-\end{enumerate}
-
-\section{Future Work} \label{sec:future}
-
-\Josh{Getting rid of the assumption of being syntax-directed (the triangular picture)}
-
-\begin{acks}
-We thank Nathanael Arkor for the useful conversation and to thank the Programming Languages and Formal Methods Laboratory at Academia Sinica for the opportunity of sharing earlier ideas.
-
-The work is supported by the Ministry of Science and Technology of Taiwan under grant MOST 109-2222-E-001-002-MY3.
-\end{acks}
-
-\bibliographystyle{ACM-Reference-Format}
-
-\IfFileExists{reference.bib}{%
-  \bibliography{reference}
-}{\bibliography{ref}}
-
-\end{document}

### 2023-06-13 tex/sec3-defs.tex
[ tex ] Split BiSig.tex into smaller files

--- 2023-06-13 tex/BiSig.tex
@@
-\documentclass[acmsmall,screen]{acmart}
-\usepackage[T1]{fontenc}
-\usepackage[utf8]{inputenc}
-\usepackage[british]{babel}
-
-\usepackage{bm,braket,mathtools,mathpartir,bussproofs}
-\EnableBpAbbreviations
-\usepackage[all,cmtip]{xy}
-
-\usepackage{subcaption,xfrac,xspace,xcolor,xspace}
-\usepackage{hyperref}
-\usepackage[capitalise,noabbrev]{cleveref}
-\usepackage[inline]{enumitem} % for environment enumerate*
-\setlist[enumerate]{mode=unboxed}
-
-\setlength {\marginparwidth }{2cm}
-\usepackage[color=yellow,textsize=small]{todonotes}
-\newcommand{\LT}[1]{\todo[author=LT,inline,color=green!40,caption={}]{{#1}}}
-\newcommand{\Josh}[1]{\todo[author=Josh,inline,color=green!40,caption={}]{{#1}}}
-
-\usepackage{subfiles}
-
-\usepackage{microtype}
-
-\citestyle{acmauthoryear}
-
-\AtEndPreamble{%
-\theoremstyle{acmdefinition}
-\newtheorem{remark}[theorem]{Remark}}
-
-\input{type-notation} %% for type theory notation copied from HoTT Book
-\input{macros}
-
-\begin{document}
-
-\author{Liang-Ting Chen}
-\email{liang.ting.chen.tw@gmail.com}
-\orcid{0000-0002-3250-1331}
-\author{Hsiang-Shang Ko}
-\orcid{0000-0002-2439-1048}
-\email{joshko@iis.sinica.edu.tw}
-
-\affiliation{%
-  \institution{Academia Sinica}
-  \streetaddress{128 Academia Road}
-  \city{Taipei}
-  \country{Taiwan}
-  \postcode{115}
-}
-
-\title{A Constructive Metatheory of Bidirectional Type Synthesis for Simple Types}
-
-\begin{abstract}
-  This paper proposes a formal theory of bidirectional typing.
-\end{abstract}
-
-\maketitle
-
-\section{Introduction}\label{sec:intro}
-
-\begin{enumerate}
-  \item Type checking as a transition from (parsed) abstract syntax trees to well-typed terms.
-  \item Bidirectional typing is the only solution that works that scales well.
-  \item Not yet a general theory but by now a design principle.
-    No `type checker generator' as each type checker is still designed and implemented on a case-by-case basis, but can be derived from a syntax-directed type theory `easily'.
-  \item The existing work of bidirectional typing does not take account of missing annotations carefully enough. 
-  \item A meta-problem: existing work is based on (or at least left unspecified) classical logic.
-    One has to state soundness of completeness of their type checking algorithm in a less straightforward way.
-  \item A review of what has been done towards this line of research.
-  \item Finally, our responses (contributions of this paper).
-\end{enumerate}
-
-\subsection{Related work}
-
-\citep{Xie2018}
-\subsubsection{Language formalization and its frameworks}
-\cite{Wadler2022}
-PoplMark Challenge~\citep{Aydemir2005}
-
-\cite{Cimini2020,Cimini2022}
-\citep{Fiore2022}
-\cite{Allais2021,Gheri2020,Ahrens2022}
-
-\subsubsection{Algebraic theory of type theories}
-\cite{Fiore1999,Fiore2022,Ahrens2021,Arkor2020,Hamana2011}
-
-\subsubsection{Type checker generation}
-\cite{Gast2004,Grewe2015,Pacak2020,Cimini2020}
-
-\subsubsection{Bidirectional typing}
-
-\cite{Pierce2000}
-\cite{Dunfield2021}
-
-\subsection{Plan and Contributions of this paper}
-
-Our assumptions and contributions:
-\begin{enumerate}
-  \item Assumption: Syntax-directed typing rules
-  \item Practical: bidirectionalisation over annotatability, decidability over soundness and completeness.
-  \item Generic: Type checking once and for all bidirectional type systems for simple types.
-  \item Correct and Constructive: Fully formalised in Martin-Löf type theory (with Axiom K).
-    LEM is not assumed, so a decidable statement has both computational and non-trivial logical readings.
-\end{enumerate}
-
-Plan of this paper:
-\begin{enumerate}
-  \item Key ideas (\Cref{sec:key-ideas})
-  \item Definitions for simple type theories and for bidirectional type systems (signature erasure and mode annotation) (\Cref{sec:defs})
-  \item Soundness, completeness, bidirectionalisation, and annotatability (\Cref{sec:annotatability})
-    \Josh{Erasure from bidirectional typing derivation to raw terms with a mode}
-  \item Bidirectional type inference (\Cref{sec:type-synthesis})
-  \item Formalisation with further examples (\Cref{sec:formalisation})
-  \item Future work (\Cref{sec:future})
-\end{enumerate}
-  \LT{better title?}
-
-\section{Core Ideas with Bidirectional Simply Typed \texorpdfstring{$\lambda$}{lambda}-Calculus} \label{sec:key-ideas}
-\begin{figure}
-  \centering
-  \small
-  \begin{mathpar}
-    \boxed{\Gamma \vdash \isTerm{t} \Rightarrow A}
-    \quad\text{and}\quad \boxed{\Gamma \vdash \isTerm{t} \Leftarrow A}
-    \\
-    \inferrule{(x : A) \in \Gamma}{\Gamma \vdash \isTerm{x} \Rightarrow A}\;(\text{var}^\Rightarrow)
-    \and
-    \inferrule{\Gamma \vdash \isTerm{t} \Leftarrow B \\ A = B}{\Gamma \vdash \isTerm{t : B}\Rightarrow A}\;(\text{anno})
-    \and
-    \inferrule{\Gamma \vdash \isTerm{t} \Rightarrow B \\ A = B}{\Gamma \vdash \isTerm{t\subsum} \Leftarrow A}\;(\text{sub})
-    \and
-    \inferrule{\Gamma \vdash \isTerm{t} \Rightarrow A \bto B \\ \Gamma \vdash \isTerm{u} \Leftarrow A}{\Gamma \vdash \isTerm{t\;u} \Rightarrow B}\;(\text{app})
-    \and
-    \inferrule{\Gamma, \isTerm{x} : A \vdash \isTerm{t} \Leftarrow B}{\Gamma \vdash \isTerm{\lam{x}t} \Leftarrow A \bto B}\;(\text{abs})
-  \end{mathpar}
-  \caption{Bidirectional simply typed $\lambda$-calculus $\Lambda^{\Leftrightarrow}_{\bto}$}
-  \label{fig:bi-stlc}
-\end{figure}
-\begin{remark}
-  We avoid using the term `function type' and its conventional notation $\to$ at the object level on purpose, as it may be confused with function types in our type-theoretic meta-language.
-\end{remark}
+%! TEX root = BiSig.tex
 
@@
     \and
-    \inferrule{\vdash_{\Sigma, \Omega} \isTerm{t}^\isDir{\syn}}{\vdash_{\Sigma, \Omega} {\isTerm{t\subsum}}^\isDir{\chk}}\;(\text{ sub})
+    \inferrule{\vdash_{\Sigma, \Omega} \isTerm{t}^\isDir{\syn}}{\vdash_{\Sigma, \Omega} {\isTerm{t\subsum}}^\isDir{\chk}}\;(\text{sub})
   \end{mathpar}
@@
 \end{example}
-
-\section{Soundness, Completeness, and Bidirectionalisation} \label{sec:annotatability}
-
-\subsection{Soundness and Completeness}
-\begin{theorem}[Soundness]\label{thm:term-soundness}
-
-\end{theorem}
-
-\begin{theorem}[Completeness]\label{thm:term-completeness}
-
-\end{theorem}
-
-\subsection{Bidirectionalisation}
-\begin{proposition}[Decidability for Bidirectionalisation] \label{thm:bidirectionalisation}
-  For every raw term $t$ and a mode $d$, it is decidable that $t$ is in mode $d$, i.e.\ $\vdash_{\Sigma, \Omega} \isTerm{t}^\isDir{d}$ is derivable.
-\end{proposition}
-
-\subsection{Completeness versus Annotatability}
-
-\section{Bidirectional Type Synthesis and Checking} \label{sec:type-synthesis}
-
-\subsection{Mode Correctness}
-
-\begin{theorem}[Uniqueness of Synthesised Types]\label{thm:unique-syn}
-  Let $(\Sigma, \Omega)$ be a bidirectional type system where $\Omega$ is mode-correct.
-  Then, for every raw term $t$ in synthesising mode and two derivations of 
-  \[
-    \Gamma \vdash_{\Sigma, \Omega} t \Rightarrow A
-    \quad\text{and}\quad
-    \Gamma \vdash_{\Sigma, \Omega} t \Rightarrow B
-  \]
-  for some types $A$ and $B$ respectively, the synthesised types must be equal, i.e.\ $A = B$.
-\end{theorem}
-
-\subsection{Trichotomy of Raw Terms by Type Synthesis}
-
-\begin{theorem}[Decidability of Bidirectional Typing Synthesis] \label{thm:bidirectional-type-synthesis}
-  For every type signature $\Sigma$ and bidirectional binding signature $\Omega$ for $\Sigma$ that is mode-correct, the following two statements hold.
-
-  \begin{enumerate}
-    \item For every context $\Gamma$ and raw term $\isTerm{t}$ in synthesising mode, it is decidable that there is a type~$A$ and a derivation of
-      \[
-        \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Rightarrow A.
-      \]
-    \item For every context $\Gamma$, raw term $\isTerm{t}$ in checking mode, and type $A$, it is decidable that there is a derivation of
-      \[
-        \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Leftarrow A.
-      \]
-  \end{enumerate}
-\end{theorem}
-
-\begin{corollary}[Trichotomy of Raw Terms]
-  Let $\Sigma$ be a type signature and $\Omega$ a bidirectional binding signature.
-  Then, for every context $\Gamma$ and raw term $t$, exactly one of the following statements holds:
-  \begin{enumerate}
-    \item $t$ is synthesising and there exists a typing derivation of\, $\Gamma \vdash_{\Sigma, \Omega} t : A$ for some type $A$.
-    \item $t$ is synthesising but there is no typing derivation of\, $\Gamma \vdash_{\Sigma, \Omega} t : A$ for any type $A$.
-    \item $t$ is not synthesising.
-  \end{enumerate}
-\end{corollary}
-\begin{proof}
-  Combine  \Cref{thm:term-soundness,thm:term-completeness,thm:bidirectionalisation} with \Cref{thm:bidirectional-type-synthesis}.
-
-\end{proof}
-
-\subsection{Decidability of Bidirectional Type Synthesis}
-
-\begin{proof}[Proof of {\Cref{thm:bidirectional-type-synthesis}}]
-  By induction on raw terms $t$ in some mode and for the $(\text{op})$ rule on the list of arguments.
-  Moreover, if the type $A$ of $t$ is synthesised and checked against $B$ but $A \neq B$, then \Cref{thm:unique-syn} is applied to derive a contradiction.
-
-  \begin{enumerate}
-    \item 
-        \LT{explain all cases but operation here}
-%    \item For every mode-correct operation $o \colon \Xi \rhd [\Delta_1]A_{1}^{d_1}, \ldots, [\Delta_{n}] A^{d_n}_{n} \to A^{d}$, a context $\Gamma$, and a raw term
-%      \[
-%        \isTerm{\tmOp_{o}(\vec{x}_1.t_1,\ldots, \vec{x}_n.t_n)}, 
-%      \]
-%      with a mode $d$, it is decidable that there exists a substitution $\rho$ from $\Xi$ to $\emptyset$ and a typing derivation of
-%      \[
-%        \Gamma \vdash_{\Sigma, \Omega} \isTerm{\tmOp_{o}(\vec{x}_1.t_1,\ldots, \vec{x}_n.t_n)} : \simsub{A}{\rho}^\isDir{d}.
-%      \]
-    \item For every set $\Xi$ of type variables, list of
-      \[
-        \left(\Delta_i, A_i, d_i\right) : \Cxt_{\Sigma}(\Xi) \times \Type_{\Sigma}(\Xi) \times \Mode, 
-      \]
-      \emph{partial} substitution $\rho$ from $\Xi$ to $\emptyset$, context $\Gamma$, and raw terms $t_i$'s in mode $d_i$,
-      \begin{enumerate}
-        \item either there is a minimal extension $\ext{\rho}$ of $\rho$ such that all     of following judgements are derivable
-          \[
-            \Gamma, \simsub{\Delta_i}{\ext{\rho}} \vdash_{\Sigma, \Omega} t_i \colon \simsub{A_i}{\ext{\rho}}^{d_i}
-          \]
-
-        \item or there is no extension $\ext{\rho}$ such that all $\Gamma, \simsub{\Delta_i}{\ext{\rho}} \vdash_{\Sigma, \Omega} t_i \colon \simsub{A_i}{\ext{\rho}}^{d_i}$ has a typing derivation. 
-      \end{enumerate}
-  \end{enumerate}
-\end{proof}
-
-\section{Formalisation and More Examples} \label{sec:formalisation}
-\LT{how to type Agda code without lhs2tex?}
-
-\begin{enumerate}
-  \item Bidirectional binding signature, functor, and terms (extrinsic typing, raw terms, raw terms with a mode, i.e.\ pre-typed terms).
-  \item type synthesis and checking.
-  \item examples. 
-\end{enumerate}
-
-\section{Future Work} \label{sec:future}
-
-\Josh{Getting rid of the assumption of being syntax-directed (the triangular picture)}
-
-\begin{acks}
-We thank Nathanael Arkor for the useful conversation and to thank the Programming Languages and Formal Methods Laboratory at Academia Sinica for the opportunity of sharing earlier ideas.
-
-The work is supported by the Ministry of Science and Technology of Taiwan under grant MOST 109-2222-E-001-002-MY3.
-\end{acks}
-
-\bibliographystyle{ACM-Reference-Format}
-
-\IfFileExists{reference.bib}{%
-  \bibliography{reference}
-}{\bibliography{ref}}
-
-\end{document}

### 2023-06-13 tex/sec4-bidirectionalisation.tex
[ tex ] Split BiSig.tex into smaller files

--- 2023-06-13 tex/BiSig.tex
@@
-\documentclass[acmsmall,screen]{acmart}
-\usepackage[T1]{fontenc}
-\usepackage[utf8]{inputenc}
-\usepackage[british]{babel}
-
-\usepackage{bm,braket,mathtools,mathpartir,bussproofs}
-\EnableBpAbbreviations
-\usepackage[all,cmtip]{xy}
-
-\usepackage{subcaption,xfrac,xspace,xcolor,xspace}
-\usepackage{hyperref}
-\usepackage[capitalise,noabbrev]{cleveref}
-\usepackage[inline]{enumitem} % for environment enumerate*
-\setlist[enumerate]{mode=unboxed}
-
-\setlength {\marginparwidth }{2cm}
-\usepackage[color=yellow,textsize=small]{todonotes}
-\newcommand{\LT}[1]{\todo[author=LT,inline,color=green!40,caption={}]{{#1}}}
-\newcommand{\Josh}[1]{\todo[author=Josh,inline,color=green!40,caption={}]{{#1}}}
-
-\usepackage{subfiles}
-
-\usepackage{microtype}
-
-\citestyle{acmauthoryear}
-
-\AtEndPreamble{%
-\theoremstyle{acmdefinition}
-\newtheorem{remark}[theorem]{Remark}}
-
-\input{type-notation} %% for type theory notation copied from HoTT Book
-\input{macros}
-
-\begin{document}
-
-\author{Liang-Ting Chen}
-\email{liang.ting.chen.tw@gmail.com}
-\orcid{0000-0002-3250-1331}
-\author{Hsiang-Shang Ko}
-\orcid{0000-0002-2439-1048}
-\email{joshko@iis.sinica.edu.tw}
-
-\affiliation{%
-  \institution{Academia Sinica}
-  \streetaddress{128 Academia Road}
-  \city{Taipei}
-  \country{Taiwan}
-  \postcode{115}
-}
-
-\title{A Constructive Metatheory of Bidirectional Type Synthesis for Simple Types}
-
-\begin{abstract}
-  This paper proposes a formal theory of bidirectional typing.
-\end{abstract}
-
-\maketitle
-
-\section{Introduction}\label{sec:intro}
-
-\begin{enumerate}
-  \item Type checking as a transition from (parsed) abstract syntax trees to well-typed terms.
-  \item Bidirectional typing is the only solution that works that scales well.
-  \item Not yet a general theory but by now a design principle.
-    No `type checker generator' as each type checker is still designed and implemented on a case-by-case basis, but can be derived from a syntax-directed type theory `easily'.
-  \item The existing work of bidirectional typing does not take account of missing annotations carefully enough. 
-  \item A meta-problem: existing work is based on (or at least left unspecified) classical logic.
-    One has to state soundness of completeness of their type checking algorithm in a less straightforward way.
-  \item A review of what has been done towards this line of research.
-  \item Finally, our responses (contributions of this paper).
-\end{enumerate}
-
-\subsection{Related work}
-
-\citep{Xie2018}
-\subsubsection{Language formalization and its frameworks}
-\cite{Wadler2022}
-PoplMark Challenge~\citep{Aydemir2005}
-
-\cite{Cimini2020,Cimini2022}
-\citep{Fiore2022}
-\cite{Allais2021,Gheri2020,Ahrens2022}
-
-\subsubsection{Algebraic theory of type theories}
-\cite{Fiore1999,Fiore2022,Ahrens2021,Arkor2020,Hamana2011}
-
-\subsubsection{Type checker generation}
-\cite{Gast2004,Grewe2015,Pacak2020,Cimini2020}
-
-\subsubsection{Bidirectional typing}
-
-\cite{Pierce2000}
-\cite{Dunfield2021}
-
-\subsection{Plan and Contributions of this paper}
-
-Our assumptions and contributions:
-\begin{enumerate}
-  \item Assumption: Syntax-directed typing rules
-  \item Practical: bidirectionalisation over annotatability, decidability over soundness and completeness.
-  \item Generic: Type checking once and for all bidirectional type systems for simple types.
-  \item Correct and Constructive: Fully formalised in Martin-Löf type theory (with Axiom K).
-    LEM is not assumed, so a decidable statement has both computational and non-trivial logical readings.
-\end{enumerate}
-
-Plan of this paper:
-\begin{enumerate}
-  \item Key ideas (\Cref{sec:key-ideas})
-  \item Definitions for simple type theories and for bidirectional type systems (signature erasure and mode annotation) (\Cref{sec:defs})
-  \item Soundness, completeness, bidirectionalisation, and annotatability (\Cref{sec:annotatability})
-    \Josh{Erasure from bidirectional typing derivation to raw terms with a mode}
-  \item Bidirectional type inference (\Cref{sec:type-synthesis})
-  \item Formalisation with further examples (\Cref{sec:formalisation})
-  \item Future work (\Cref{sec:future})
-\end{enumerate}
-  \LT{better title?}
-
-\section{Core Ideas with Bidirectional Simply Typed \texorpdfstring{$\lambda$}{lambda}-Calculus} \label{sec:key-ideas}
-\begin{figure}
-  \centering
-  \small
-  \begin{mathpar}
-    \boxed{\Gamma \vdash \isTerm{t} \Rightarrow A}
-    \quad\text{and}\quad \boxed{\Gamma \vdash \isTerm{t} \Leftarrow A}
-    \\
-    \inferrule{(x : A) \in \Gamma}{\Gamma \vdash \isTerm{x} \Rightarrow A}\;(\text{var}^\Rightarrow)
-    \and
-    \inferrule{\Gamma \vdash \isTerm{t} \Leftarrow B \\ A = B}{\Gamma \vdash \isTerm{t : B}\Rightarrow A}\;(\text{anno})
-    \and
-    \inferrule{\Gamma \vdash \isTerm{t} \Rightarrow B \\ A = B}{\Gamma \vdash \isTerm{t\subsum} \Leftarrow A}\;(\text{sub})
-    \and
-    \inferrule{\Gamma \vdash \isTerm{t} \Rightarrow A \bto B \\ \Gamma \vdash \isTerm{u} \Leftarrow A}{\Gamma \vdash \isTerm{t\;u} \Rightarrow B}\;(\text{app})
-    \and
-    \inferrule{\Gamma, \isTerm{x} : A \vdash \isTerm{t} \Leftarrow B}{\Gamma \vdash \isTerm{\lam{x}t} \Leftarrow A \bto B}\;(\text{abs})
-  \end{mathpar}
-  \caption{Bidirectional simply typed $\lambda$-calculus $\Lambda^{\Leftrightarrow}_{\bto}$}
-  \label{fig:bi-stlc}
-\end{figure}
-\begin{remark}
-  We avoid using the term `function type' and its conventional notation $\to$ at the object level on purpose, as it may be confused with function types in our type-theoretic meta-language.
-\end{remark}
-
-\section{Definitions for Simple Type Theories and Bidirectional Type Systems}\label{sec:defs}
-
-\subsection{Simple Types}
-\begin{definition}
-  A \emph{(simple type) signature} $\Sigma$ consists of a set $I$ and an \emph{arity} function $\arity\colon I \to \mathbb{N}$.
-  An inhabitant $i : I$ is meant to represent the \emph{$i$-th operation} $\tyOp_i$ and its \emph{number} $ \arity(i)$ of arguments.
-
-  The judgements of \emph{$\Sigma$-types} and \emph{$\Sigma$-contexts} over a variable set $\Theta$ are defined inductively by \Cref{fig:simple-type,fig:simple-context} respectively.
-We write $A : \Type_{\Sigma}(\Theta)$ for $\Theta \vdash_{\Sigma} A$ and $\Gamma \colon \Cxt_{\Sigma}(\Theta)$ for $\Theta \vdash_{\Sigma} \Gamma$ to disambiguate $\Sigma$-types and $\Sigma$-contexts in the case of any confusion.
-\end{definition}
-
-\begin{figure}
-  \begin{minipage}[b]{.55\textwidth}
-    \centering
-    \small
-    \begin{mathpar}
-      \boxed{\Theta\vdash_{\Sigma} A} \\
-      \inferrule{\Theta \ni X_i}{\Theta \vdash_\Sigma X_i} \and
-    \inferrule{\Theta \vdash_{\Sigma} A_1 \\ \cdots \\ \Theta \vdash_{\Sigma} A_{\arity(o)}}{\Theta \vdash_{\Sigma} \tyOp_o(A_1, \ldots, A_{\arity(o)})}
-    \end{mathpar}
-    \caption{Type formation}
-    \label{fig:simple-type}
-  \end{minipage}
-  \begin{minipage}[b]{.4\textwidth}
-    \centering
-    \small
-    \begin{mathpar}
-      \boxed{\Theta \vdash_{\Sigma} \Gamma} \\
-      \inferrule{ }{\Theta \vdash_{\Sigma} \cdot }\and
-      \inferrule{\Theta \vdash_{\Sigma} A \\ \Theta\vdash_{\Sigma} \Gamma}{\Theta \vdash_{\Sigma} \Gamma, A}
-    \end{mathpar}
-    \caption{Context formation}
-  \label{fig:simple-context}
-  \end{minipage}
-\end{figure}
-
-\begin{example}\label{ex:implication}
-  Simply typed $\lambda$-calculus $\Lambda_{\bto}$ typically includes a binary implication type denoted by $\bto$ and some base types to ensure that the set of all types is non-empty.
-  The type signature $\Sigma_{\bto}$ of simply typed $\lambda$-calculus consists of a binary operation $\bto$ and a base type (nullary operation) $N$.
-  In the case of simply typed $\lambda$-calculus $\Lambda_{\bto, \btimes}$ with binary products, we can extend $\Sigma_{\bto}$ by adding a binary operation $\btimes$ to represent the binary product type, denoted as $\Sigma_{\bto, \btimes}$.
-\end{example}
-
-\subsection{Binding Signatures} \label{subsec:binding-sig}
-\begin{definition}\label{def:binding-signature}
-  For a type signature $\Sigma$, a \emph{binding signature} $\Omega$ consists of a set $O$ and a function
-  \[
-    \mathit{ar}\colon O \to \sum_{\Xi : \N} \left(\Cxt_\Sigma(\Xi) \times \Type_\Sigma(\Xi)\right)^* \times \Type_\Sigma(\Xi).
-  \]
-  Each inhabitant $o: O$ is meant to represent a term construct $\tmOp_o$ in a simple type theory with a triple $\left(\Xi, \left[\left(\Delta_1; A_1\right), \ldots, \left(\Delta_{n}; A_{n}\right) \right], A\right)$
-  instead of a number as its arity $\arity_o$ where
-  \begin{enumerate}
-    \item $\Xi$ is the number of type variables, 
-    \item $A : \Type_\Sigma(\Xi)$ is the target type of $\tmOp_o$, and
-    \item $\left[\left(\Delta_1; A_{1}\right), \ldots, \left(\Delta_{n}; A_{n}\right) \right]$ is a list of pairs where
-    \item the $i$-th pair $(\Delta_i, A_i) :\Cxt_{\Sigma}(\Xi) \times \Term_{\Sigma}(\Xi)$ are types of the binding variables and the type of the $i$-th argument of $\tmOp_{o}$.
-  \end{enumerate}
-  For brevity, we write $o \colon \Xi \rhd [\Delta_1]A_{1}, \ldots, \left[\Delta_{n}\right] A_{n} \to A$ to indicate an operation with its arity. 
-\end{definition}
-\begin{remark}[Terminology on binding signature]
-  \cite{Aczel1978,Fiore2010}
-\end{remark}
-
-\begin{example} \label{ex:STLC-sig}
-  Given the type signature $\Sigma_{\bto}$ for the implication type, the term signature $\Lambda_{\bto}$ for simply typed $\lambda$-calculus can be described by operations
-  \begin{align*}
-    \mathsf{app}\colon A, B \rhd (A \bto B), A \to B && \mathsf{abs}\colon A , B \rhd [A]B \to (A \bto B)
-  \end{align*}
-    or, verbosely, the signature $\Lambda_{\bto}$ consists of the type $O_\Lambda = \{\app, \abs\}$ with arities
-  \begin{align*}
-    \arity(\app) = (\{A, B\}, [(\cdot; A \bto B), (\cdot; A)], B)
-    && 
-    \arity(\abs) = (\{A, B\}, [(\cdot, A; B)], A \bto B)
-  \end{align*}
-  For the type signature $\Sigma_{\bto, \btimes}$, the term signature $\Omega_{\bto, \btimes}$ for simply typed $\lambda$-calculus with finite products has three additional operations
-to $\Omega_{\bto}$:
-  \begin{align*}
-    \mathsf{pair}\colon A, B \rhd A, B \to A \btimes B
-    && \fst \colon A, B \rhd A \btimes B \to A
-    && \snd \colon A, B \rhd A \btimes B \to B
-  \end{align*}
-\end{example}
-
-First, it is noteworthy that the set of term constructs in a type theory need not be finite.
-For instance, a type theory may adopt a spine application which takes indefinitely many arguments---for each number $n$ of arguments an $(n+1)$-ary application construct can be introduced, so even if each operation has a definite number of arguments by definition a spine application is still expressible.
-
-More importantly, the inclusion of the set of variables used in an operation is a salient feature of our binding signatures.
-Instead of treating application as a family of constructs $\app_{A, B}$ indexed by all types $A$ and $B$, as done by \citet{Fiore2022}, we are able to identify them as a single construct $\app$.
-This is not only for brevity but also necessary to compare the type equality during type synthesis and checking.
-
-%\subsection{Intrinsically Typed Terms}
-%
-%Intrinsically typed terms for a type signature $\Sigma$ and a term signature $\Omega$ are nothing more than derivations of the intrinsic typing judgement $\Gamma \vdash_{\Sigma, \Omega} A$ constructed with only the rule $(\var)$ and the rule scheme $(\tmOp)$ displayed in \Cref{fig:intrinsic-typing}.
-%
-%The substitution $\rho\colon \Sub{\Xi}{\emptyset}$ is used to instantiate variables in the local context $\Xi$ with concrete types.
-%Accordingly, $\rho$ has to be applied to types that appear in the arity to construct a term by $\tmOp$ to ensure all types are well-formed without any use of type variables.
-%
-%\begin{figure}
-%  \centering
-%  \small
-%  \begin{mathpar}
-%    \boxed{\Gamma \vdash_{\Sigma, \Omega} A} \\
-%    \inferrule{A \in \Gamma}{\Gamma \vdash_{\Sigma, \Omega} A}\;(\var)
-%    \and
-%    \inferrule{\rho\colon \Sub{\Xi}{\emptyset}  \\ \Gamma, \simsub{\Delta_{1}}{\rho} \vdash_{\Sigma, \Omega} \simsub{A_{1}}{\rho} \quad\cdots\quad \Gamma, \simsub{\Delta_{n}}{\rho} \vdash_{\Sigma, \Omega} \simsub{A_{n}}{\rho}}
-%    {\Gamma \vdash_{\Sigma, \Omega} \simsub{A}{\rho}}\;(\tmOp)
-%    \and \text{for $o\colon \Xi \rhd [\Delta_1]A_1, \ldots, [\Delta_{n}]A_{n} \to A$ in $\Omega$}
-%  \end{mathpar}
-%  \caption{Intrinsic typing rules for a simple type theory $(\Sigma, \Omega)$}
-%  \label{fig:intrinsic-typing}
-%\end{figure}
-
-\subsection{Simple Type Theory for a Signature}
-
-\begin{figure}
-  \centering
-  \small
-  \begin{mathpar}
-    \boxed{\vdash_{\Sigma, \Omega} \isTerm{t}}
-    \\
-    \inferrule{x : \Identifier}{\vdash_{\Sigma, \Omega} \isTerm{x}}\;(\text{var})
-    \and
-    \inferrule{\cdot \vdash_{\Sigma} A \\ \vdash_{\Sigma, \Omega}\isTerm{t}}{\vdash_{\Sigma, \Omega} \isTerm{t \annote A}}\;(\text{anno})
-    \\
-    \inferrule{\vdash_{\Sigma, \Omega} \isTerm{t_1} \quad \cdots \quad \vdash_{\Sigma, \Omega} \isTerm{t_n}}
-    {\vdash_{\Sigma, \Omega} \isTerm{\tmOp_o(t_1, \ldots, t_n)}}\;(\text{op}) 
-    \and \text{for $o \colon \Xi \rhd [\Delta_1]A_{1}, \ldots, [\Delta_{n}] A_{n} \to A$ in $\Omega$}
-  \end{mathpar}
-  \caption{Raw terms for $(\Sigma, \Omega)$ with annotation}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \small
-  \begin{mathpar}
-    \boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A} \quad \text{where $\vdash_{\Sigma, \Omega} t$} \\
-    \inferrule{(x : A) \in \Gamma}{\Gamma \vdash_{\Sigma, \Omega} \isTerm{x} : A}\;(\var)
-    \and
-    \inferrule{\Gamma \vdash \isTerm{t} : A}{\Gamma \vdash (\isTerm{t \annote A}) : A}\;(\text{anno})
-    \and
-    \inferrule{\rho : \Sub{\Xi}{\emptyset} \\ \Gamma, \isTerm{\vec{x}_1} : \simsub{\Delta_{1}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_1} : \simsub{A_{1}}{\rho} \quad\cdots\quad \Gamma, \isTerm{\vec{x}_n} : \simsub{\Delta_{n}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_n} : \simsub{A_{n}}{\rho}}
-    {\Gamma \vdash_{\Sigma, \Omega} \isTerm{\tmOp_o(\vec{x}_1.\,t_1; \ldots; \vec{x}_n.\,t_n)} : \simsub{A}{\rho}}\;(\tmOp)
-    \and \text{for $o\colon \Xi \rhd [\Delta_1]A_1, \ldots, [\Delta_{n}]A_{n} \to A$ in $\Omega$}
-  \end{mathpar}
-  \caption{Typing rules for a simple type theory $(\Sigma, \Omega)$ with annotation}
-  \label{fig:extrinsic-typing}
-\end{figure}
-
-\subsection{Bidirectional Binding Signatures and their Bidirectional Type Systems}
-
-\begin{definition}
-  For a type signature $\Sigma$, a \emph{bidirectional binding signature} $\Omega$ is a set $O$ with a function
-  \[
-    \mathit{ar}\colon O \to \sum_{\Xi : \N} \left(\Cxt_{\Sigma}(\Xi) \times \Type_{\Sigma}(\Xi) \times \isDir{\Mode}\right)^* \times \Type_{\Sigma}(\Xi) \times \isDir{\Mode}.
-  \]
-  where $\Mode$ consists of two inhabitants $\chk$ for checking and $\syn$ for synthesis.
-  Bidirectional binding signatures are just binding signatures (\Cref{def:binding-signature}) augmented with a mode for each argument and its target of a construct $\tmOp_o$ in a bidirectional system, i.e.
-  an arity is a $4$-tuple
-  \[
-    \left(\Xi, \left[\left(\Delta_1; A_1; d_1\right), \ldots, \left(\Delta_{n}; A_{n}; d_n\right) \right], A, d\right)
-  \]
-  where $d$ and $d_i$'s indicate the modes of a construct and its arguments respectively.
-
-  For brevity, we write $o \colon \Xi \rhd [\Delta_1]A_{1}^{d_1}, \ldots, [\Delta_{n}] A^{d_n}_{n} \to A^{d}$ to indicate an operation with its arity. 
-\end{definition}
-
-\begin{example}
-  The term signature $\Lambda_{\bto}^{\leftrightarrows}$ for bidirectional simply typed $\lambda$-calculus introduced in \Cref{subsec:binding-sig} can be specified by operations 
-  \begin{align*}
-    \mathsf{app}\colon A, B \rhd (A \bto B)^{\Rightarrow}, A^{\Leftarrow} \to B^{\Rightarrow} &&
-    \mathsf{abs}\colon A , B \rhd [A]B^{\Leftarrow} \to (A \bto B)^{\Leftarrow}
-  \end{align*}
-  extending $\Lambda_{\bto}$ (\Cref{ex:STLC-sig}) with the mode information.
-\end{example}
-
-\LT{Some remark about signature erasure and annotation; introduce the notation $\erase{\Omega}$}
-
-\begin{figure}
-  \centering
-  \small
-  \begin{mathpar}
-    \boxed{\vdash_{\Sigma, \Omega} \isTerm{t}^\isDir{d}}
-    \quad\text{where $\vdash_{\Sigma, \erase{\Omega}} \isTerm{t}$}
-    \\
-    \inferrule{x : \Identifier}{\vdash_{\Sigma, \Omega} \isTerm{x}^\isDir{\syn}}\;(\text{var})
-    \and
-    \inferrule{\cdot \vdash_{\Sigma} A \\ \vdash_{\Sigma, \Omega}\isTerm{t}^\isDir{\chk}}{\vdash_{\Sigma, \Omega} (\isTerm{t \annote A})^\isDir{\syn}}\;(\text{anno})
-    \and
-    \inferrule{\vdash_{\Sigma, \Omega} \isTerm{t}^\isDir{\syn}}{\vdash_{\Sigma, \Omega} {\isTerm{t\subsum}}^\isDir{\chk}}\;(\text{ sub})
-  \end{mathpar}
-  \begin{mathpar}
-    \inferrule{\vdash_{\Sigma, \Omega} \isTerm{t_1}^\isDir{d_1} \quad \cdots \quad \vdash_{\Sigma, \Omega} \isTerm{t_n}^\isDir{d_n}}
-    {\vdash_{\Sigma, \Omega} \isTerm{\tmOp_o(\vec{x}_1.\, t_1; \ldots;\vec{x}_n.\, t_n)}^\isDir{d}}\;(\text{op})
-    \and \text{for $o \colon \Xi \rhd [\Delta_1]A_{1}^{d_1}, \ldots, [\Delta_{n}] A^{d_n}_{n} \to A^{d}$ in $\Omega$}
-  \end{mathpar}
-  \caption{Raw terms in some mode $d$ for a bidirectional type system $(\Sigma, \Omega)$}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \small
-  \begin{mathpar}
-    \boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A^\isDir{d}}
-    \quad \text{where $\vdash_{\Sigma, \erase{\Omega}} \isTerm{t}$ and $\vdash_{\Sigma, \Omega} \isTerm{t}^\isDir{d}$} 
-    \\
-    \inferrule{(x : A) \in \Gamma}{\Gamma \vdash_{\Sigma, \Omega} \isTerm{x} : A^\isDir{\syn}}\;(\var)
-    \and
-    \inferrule{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A^\isDir{\chk}}{\Gamma \vdash_{\Sigma, \Omega} (\isTerm{t \annote B}):  A^\isDir{\syn}}\;(\text{anno})
-    \and
-    \inferrule{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : B^{\isDir{\syn}} \\ A = B}{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t\subsum} : A^\isDir{\chk}}\;(\text{sub})
-    \\
-    \inferrule{\rho\colon \Sub{\Xi}{\emptyset} \\ \Gamma, \simsub{\isTerm{\vec{x}_{1}} : \Delta_{1}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_{1}} : \simsub{A_{1}}{\rho}^{\isDir{d_1}} \\
-      \cdots \\
-    \Gamma, \simsub{\isTerm{\vec{x}_{n}} : \Delta_{n}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_n} : \simsub{A_{n}}{\rho}^{\isDir{d_n}}}
-  {\Gamma \vdash_{\Sigma, \Omega} \isTerm{\tmOp_o(\vec{x}_1 .\, t_{1}; \ldots; \vec{x}_{n}.\, t_{n})} : \simsub{A}{\rho}^{\isDir{d}}} \; (\tmOp)
-    \and \text{for $o \colon \Xi \rhd [\Delta_1]A_{1}^{d_1}, \ldots, [\Delta_{n}] A^{d_n}_{n} \to A^{d}$ in $\Omega$}
-    \\
-    \text{Abbreviations: $\boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Rightarrow A} \defeq \boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A^\isDir{\syn}}$ and $\boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Leftarrow A} \defeq \boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A^\isDir{\chk}}$}
-  \end{mathpar}
-  \caption{Typing rules for a bidirectional system $(\Sigma, \Omega)$}
-\end{figure}
-
-\begin{example}
-
-\end{example}
+%! TEX root = BiSig.tex
 
@@
 \subsection{Completeness versus Annotatability}
-
-\section{Bidirectional Type Synthesis and Checking} \label{sec:type-synthesis}
-
-\subsection{Mode Correctness}
-
-\begin{theorem}[Uniqueness of Synthesised Types]\label{thm:unique-syn}
-  Let $(\Sigma, \Omega)$ be a bidirectional type system where $\Omega$ is mode-correct.
-  Then, for every raw term $t$ in synthesising mode and two derivations of 
-  \[
-    \Gamma \vdash_{\Sigma, \Omega} t \Rightarrow A
-    \quad\text{and}\quad
-    \Gamma \vdash_{\Sigma, \Omega} t \Rightarrow B
-  \]
-  for some types $A$ and $B$ respectively, the synthesised types must be equal, i.e.\ $A = B$.
-\end{theorem}
-
-\subsection{Trichotomy of Raw Terms by Type Synthesis}
-
-\begin{theorem}[Decidability of Bidirectional Typing Synthesis] \label{thm:bidirectional-type-synthesis}
-  For every type signature $\Sigma$ and bidirectional binding signature $\Omega$ for $\Sigma$ that is mode-correct, the following two statements hold.
-
-  \begin{enumerate}
-    \item For every context $\Gamma$ and raw term $\isTerm{t}$ in synthesising mode, it is decidable that there is a type~$A$ and a derivation of
-      \[
-        \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Rightarrow A.
-      \]
-    \item For every context $\Gamma$, raw term $\isTerm{t}$ in checking mode, and type $A$, it is decidable that there is a derivation of
-      \[
-        \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Leftarrow A.
-      \]
-  \end{enumerate}
-\end{theorem}
-
-\begin{corollary}[Trichotomy of Raw Terms]
-  Let $\Sigma$ be a type signature and $\Omega$ a bidirectional binding signature.
-  Then, for every context $\Gamma$ and raw term $t$, exactly one of the following statements holds:
-  \begin{enumerate}
-    \item $t$ is synthesising and there exists a typing derivation of\, $\Gamma \vdash_{\Sigma, \Omega} t : A$ for some type $A$.
-    \item $t$ is synthesising but there is no typing derivation of\, $\Gamma \vdash_{\Sigma, \Omega} t : A$ for any type $A$.
-    \item $t$ is not synthesising.
-  \end{enumerate}
-\end{corollary}
-\begin{proof}
-  Combine  \Cref{thm:term-soundness,thm:term-completeness,thm:bidirectionalisation} with \Cref{thm:bidirectional-type-synthesis}.
-
-\end{proof}
-
-\subsection{Decidability of Bidirectional Type Synthesis}
-
-\begin{proof}[Proof of {\Cref{thm:bidirectional-type-synthesis}}]
-  By induction on raw terms $t$ in some mode and for the $(\text{op})$ rule on the list of arguments.
-  Moreover, if the type $A$ of $t$ is synthesised and checked against $B$ but $A \neq B$, then \Cref{thm:unique-syn} is applied to derive a contradiction.
-
-  \begin{enumerate}
-    \item 
-        \LT{explain all cases but operation here}
-%    \item For every mode-correct operation $o \colon \Xi \rhd [\Delta_1]A_{1}^{d_1}, \ldots, [\Delta_{n}] A^{d_n}_{n} \to A^{d}$, a context $\Gamma$, and a raw term
-%      \[
-%        \isTerm{\tmOp_{o}(\vec{x}_1.t_1,\ldots, \vec{x}_n.t_n)}, 
-%      \]
-%      with a mode $d$, it is decidable that there exists a substitution $\rho$ from $\Xi$ to $\emptyset$ and a typing derivation of
-%      \[
-%        \Gamma \vdash_{\Sigma, \Omega} \isTerm{\tmOp_{o}(\vec{x}_1.t_1,\ldots, \vec{x}_n.t_n)} : \simsub{A}{\rho}^\isDir{d}.
-%      \]
-    \item For every set $\Xi$ of type variables, list of
-      \[
-        \left(\Delta_i, A_i, d_i\right) : \Cxt_{\Sigma}(\Xi) \times \Type_{\Sigma}(\Xi) \times \Mode, 
-      \]
-      \emph{partial} substitution $\rho$ from $\Xi$ to $\emptyset$, context $\Gamma$, and raw terms $t_i$'s in mode $d_i$,
-      \begin{enumerate}
-        \item either there is a minimal extension $\ext{\rho}$ of $\rho$ such that all     of following judgements are derivable
-          \[
-            \Gamma, \simsub{\Delta_i}{\ext{\rho}} \vdash_{\Sigma, \Omega} t_i \colon \simsub{A_i}{\ext{\rho}}^{d_i}
-          \]
-
-        \item or there is no extension $\ext{\rho}$ such that all $\Gamma, \simsub{\Delta_i}{\ext{\rho}} \vdash_{\Sigma, \Omega} t_i \colon \simsub{A_i}{\ext{\rho}}^{d_i}$ has a typing derivation. 
-      \end{enumerate}
-  \end{enumerate}
-\end{proof}
-
-\section{Formalisation and More Examples} \label{sec:formalisation}
-\LT{how to type Agda code without lhs2tex?}
-
-\begin{enumerate}
-  \item Bidirectional binding signature, functor, and terms (extrinsic typing, raw terms, raw terms with a mode, i.e.\ pre-typed terms).
-  \item type synthesis and checking.
-  \item examples. 
-\end{enumerate}
-
-\section{Future Work} \label{sec:future}
-
-\Josh{Getting rid of the assumption of being syntax-directed (the triangular picture)}
-
-\begin{acks}
-We thank Nathanael Arkor for the useful conversation and to thank the Programming Languages and Formal Methods Laboratory at Academia Sinica for the opportunity of sharing earlier ideas.
-
-The work is supported by the Ministry of Science and Technology of Taiwan under grant MOST 109-2222-E-001-002-MY3.
-\end{acks}
-
-\bibliographystyle{ACM-Reference-Format}
-
-\IfFileExists{reference.bib}{%
-  \bibliography{reference}
-}{\bibliography{ref}}
-
-\end{document}

### 2023-06-13 tex/sec5-synthesis.tex
[ tex ] Split BiSig.tex into smaller files

--- 2023-06-13 tex/BiSig.tex
@@
-\documentclass[acmsmall,screen]{acmart}
-\usepackage[T1]{fontenc}
-\usepackage[utf8]{inputenc}
-\usepackage[british]{babel}
-
-\usepackage{bm,braket,mathtools,mathpartir,bussproofs}
-\EnableBpAbbreviations
-\usepackage[all,cmtip]{xy}
-
-\usepackage{subcaption,xfrac,xspace,xcolor,xspace}
-\usepackage{hyperref}
-\usepackage[capitalise,noabbrev]{cleveref}
-\usepackage[inline]{enumitem} % for environment enumerate*
-\setlist[enumerate]{mode=unboxed}
-
-\setlength {\marginparwidth }{2cm}
-\usepackage[color=yellow,textsize=small]{todonotes}
-\newcommand{\LT}[1]{\todo[author=LT,inline,color=green!40,caption={}]{{#1}}}
-\newcommand{\Josh}[1]{\todo[author=Josh,inline,color=green!40,caption={}]{{#1}}}
-
-\usepackage{subfiles}
-
-\usepackage{microtype}
-
-\citestyle{acmauthoryear}
-
-\AtEndPreamble{%
-\theoremstyle{acmdefinition}
-\newtheorem{remark}[theorem]{Remark}}
-
-\input{type-notation} %% for type theory notation copied from HoTT Book
-\input{macros}
-
-\begin{document}
-
-\author{Liang-Ting Chen}
-\email{liang.ting.chen.tw@gmail.com}
-\orcid{0000-0002-3250-1331}
-\author{Hsiang-Shang Ko}
-\orcid{0000-0002-2439-1048}
-\email{joshko@iis.sinica.edu.tw}
-
-\affiliation{%
-  \institution{Academia Sinica}
-  \streetaddress{128 Academia Road}
-  \city{Taipei}
-  \country{Taiwan}
-  \postcode{115}
-}
-
-\title{A Constructive Metatheory of Bidirectional Type Synthesis for Simple Types}
-
-\begin{abstract}
-  This paper proposes a formal theory of bidirectional typing.
-\end{abstract}
-
-\maketitle
-
-\section{Introduction}\label{sec:intro}
-
-\begin{enumerate}
-  \item Type checking as a transition from (parsed) abstract syntax trees to well-typed terms.
-  \item Bidirectional typing is the only solution that works that scales well.
-  \item Not yet a general theory but by now a design principle.
-    No `type checker generator' as each type checker is still designed and implemented on a case-by-case basis, but can be derived from a syntax-directed type theory `easily'.
-  \item The existing work of bidirectional typing does not take account of missing annotations carefully enough. 
-  \item A meta-problem: existing work is based on (or at least left unspecified) classical logic.
-    One has to state soundness of completeness of their type checking algorithm in a less straightforward way.
-  \item A review of what has been done towards this line of research.
-  \item Finally, our responses (contributions of this paper).
-\end{enumerate}
-
-\subsection{Related work}
-
-\citep{Xie2018}
-\subsubsection{Language formalization and its frameworks}
-\cite{Wadler2022}
-PoplMark Challenge~\citep{Aydemir2005}
-
-\cite{Cimini2020,Cimini2022}
-\citep{Fiore2022}
-\cite{Allais2021,Gheri2020,Ahrens2022}
-
-\subsubsection{Algebraic theory of type theories}
-\cite{Fiore1999,Fiore2022,Ahrens2021,Arkor2020,Hamana2011}
-
-\subsubsection{Type checker generation}
-\cite{Gast2004,Grewe2015,Pacak2020,Cimini2020}
-
-\subsubsection{Bidirectional typing}
-
-\cite{Pierce2000}
-\cite{Dunfield2021}
-
-\subsection{Plan and Contributions of this paper}
-
-Our assumptions and contributions:
-\begin{enumerate}
-  \item Assumption: Syntax-directed typing rules
-  \item Practical: bidirectionalisation over annotatability, decidability over soundness and completeness.
-  \item Generic: Type checking once and for all bidirectional type systems for simple types.
-  \item Correct and Constructive: Fully formalised in Martin-Löf type theory (with Axiom K).
-    LEM is not assumed, so a decidable statement has both computational and non-trivial logical readings.
-\end{enumerate}
-
-Plan of this paper:
-\begin{enumerate}
-  \item Key ideas (\Cref{sec:key-ideas})
-  \item Definitions for simple type theories and for bidirectional type systems (signature erasure and mode annotation) (\Cref{sec:defs})
-  \item Soundness, completeness, bidirectionalisation, and annotatability (\Cref{sec:annotatability})
-    \Josh{Erasure from bidirectional typing derivation to raw terms with a mode}
-  \item Bidirectional type inference (\Cref{sec:type-synthesis})
-  \item Formalisation with further examples (\Cref{sec:formalisation})
-  \item Future work (\Cref{sec:future})
-\end{enumerate}
-  \LT{better title?}
-
-\section{Core Ideas with Bidirectional Simply Typed \texorpdfstring{$\lambda$}{lambda}-Calculus} \label{sec:key-ideas}
-\begin{figure}
-  \centering
-  \small
-  \begin{mathpar}
-    \boxed{\Gamma \vdash \isTerm{t} \Rightarrow A}
-    \quad\text{and}\quad \boxed{\Gamma \vdash \isTerm{t} \Leftarrow A}
-    \\
-    \inferrule{(x : A) \in \Gamma}{\Gamma \vdash \isTerm{x} \Rightarrow A}\;(\text{var}^\Rightarrow)
-    \and
-    \inferrule{\Gamma \vdash \isTerm{t} \Leftarrow B \\ A = B}{\Gamma \vdash \isTerm{t : B}\Rightarrow A}\;(\text{anno})
-    \and
-    \inferrule{\Gamma \vdash \isTerm{t} \Rightarrow B \\ A = B}{\Gamma \vdash \isTerm{t\subsum} \Leftarrow A}\;(\text{sub})
-    \and
-    \inferrule{\Gamma \vdash \isTerm{t} \Rightarrow A \bto B \\ \Gamma \vdash \isTerm{u} \Leftarrow A}{\Gamma \vdash \isTerm{t\;u} \Rightarrow B}\;(\text{app})
-    \and
-    \inferrule{\Gamma, \isTerm{x} : A \vdash \isTerm{t} \Leftarrow B}{\Gamma \vdash \isTerm{\lam{x}t} \Leftarrow A \bto B}\;(\text{abs})
-  \end{mathpar}
-  \caption{Bidirectional simply typed $\lambda$-calculus $\Lambda^{\Leftrightarrow}_{\bto}$}
-  \label{fig:bi-stlc}
-\end{figure}
-\begin{remark}
-  We avoid using the term `function type' and its conventional notation $\to$ at the object level on purpose, as it may be confused with function types in our type-theoretic meta-language.
-\end{remark}
-
-\section{Definitions for Simple Type Theories and Bidirectional Type Systems}\label{sec:defs}
-
-\subsection{Simple Types}
-\begin{definition}
-  A \emph{(simple type) signature} $\Sigma$ consists of a set $I$ and an \emph{arity} function $\arity\colon I \to \mathbb{N}$.
-  An inhabitant $i : I$ is meant to represent the \emph{$i$-th operation} $\tyOp_i$ and its \emph{number} $ \arity(i)$ of arguments.
-
-  The judgements of \emph{$\Sigma$-types} and \emph{$\Sigma$-contexts} over a variable set $\Theta$ are defined inductively by \Cref{fig:simple-type,fig:simple-context} respectively.
-We write $A : \Type_{\Sigma}(\Theta)$ for $\Theta \vdash_{\Sigma} A$ and $\Gamma \colon \Cxt_{\Sigma}(\Theta)$ for $\Theta \vdash_{\Sigma} \Gamma$ to disambiguate $\Sigma$-types and $\Sigma$-contexts in the case of any confusion.
-\end{definition}
-
-\begin{figure}
-  \begin{minipage}[b]{.55\textwidth}
-    \centering
-    \small
-    \begin{mathpar}
-      \boxed{\Theta\vdash_{\Sigma} A} \\
-      \inferrule{\Theta \ni X_i}{\Theta \vdash_\Sigma X_i} \and
-    \inferrule{\Theta \vdash_{\Sigma} A_1 \\ \cdots \\ \Theta \vdash_{\Sigma} A_{\arity(o)}}{\Theta \vdash_{\Sigma} \tyOp_o(A_1, \ldots, A_{\arity(o)})}
-    \end{mathpar}
-    \caption{Type formation}
-    \label{fig:simple-type}
-  \end{minipage}
-  \begin{minipage}[b]{.4\textwidth}
-    \centering
-    \small
-    \begin{mathpar}
-      \boxed{\Theta \vdash_{\Sigma} \Gamma} \\
-      \inferrule{ }{\Theta \vdash_{\Sigma} \cdot }\and
-      \inferrule{\Theta \vdash_{\Sigma} A \\ \Theta\vdash_{\Sigma} \Gamma}{\Theta \vdash_{\Sigma} \Gamma, A}
-    \end{mathpar}
-    \caption{Context formation}
-  \label{fig:simple-context}
-  \end{minipage}
-\end{figure}
-
-\begin{example}\label{ex:implication}
-  Simply typed $\lambda$-calculus $\Lambda_{\bto}$ typically includes a binary implication type denoted by $\bto$ and some base types to ensure that the set of all types is non-empty.
-  The type signature $\Sigma_{\bto}$ of simply typed $\lambda$-calculus consists of a binary operation $\bto$ and a base type (nullary operation) $N$.
-  In the case of simply typed $\lambda$-calculus $\Lambda_{\bto, \btimes}$ with binary products, we can extend $\Sigma_{\bto}$ by adding a binary operation $\btimes$ to represent the binary product type, denoted as $\Sigma_{\bto, \btimes}$.
-\end{example}
-
-\subsection{Binding Signatures} \label{subsec:binding-sig}
-\begin{definition}\label{def:binding-signature}
-  For a type signature $\Sigma$, a \emph{binding signature} $\Omega$ consists of a set $O$ and a function
-  \[
-    \mathit{ar}\colon O \to \sum_{\Xi : \N} \left(\Cxt_\Sigma(\Xi) \times \Type_\Sigma(\Xi)\right)^* \times \Type_\Sigma(\Xi).
-  \]
-  Each inhabitant $o: O$ is meant to represent a term construct $\tmOp_o$ in a simple type theory with a triple $\left(\Xi, \left[\left(\Delta_1; A_1\right), \ldots, \left(\Delta_{n}; A_{n}\right) \right], A\right)$
-  instead of a number as its arity $\arity_o$ where
-  \begin{enumerate}
-    \item $\Xi$ is the number of type variables, 
-    \item $A : \Type_\Sigma(\Xi)$ is the target type of $\tmOp_o$, and
-    \item $\left[\left(\Delta_1; A_{1}\right), \ldots, \left(\Delta_{n}; A_{n}\right) \right]$ is a list of pairs where
-    \item the $i$-th pair $(\Delta_i, A_i) :\Cxt_{\Sigma}(\Xi) \times \Term_{\Sigma}(\Xi)$ are types of the binding variables and the type of the $i$-th argument of $\tmOp_{o}$.
-  \end{enumerate}
-  For brevity, we write $o \colon \Xi \rhd [\Delta_1]A_{1}, \ldots, \left[\Delta_{n}\right] A_{n} \to A$ to indicate an operation with its arity. 
-\end{definition}
-\begin{remark}[Terminology on binding signature]
-  \cite{Aczel1978,Fiore2010}
-\end{remark}
-
-\begin{example} \label{ex:STLC-sig}
-  Given the type signature $\Sigma_{\bto}$ for the implication type, the term signature $\Lambda_{\bto}$ for simply typed $\lambda$-calculus can be described by operations
-  \begin{align*}
-    \mathsf{app}\colon A, B \rhd (A \bto B), A \to B && \mathsf{abs}\colon A , B \rhd [A]B \to (A \bto B)
-  \end{align*}
-    or, verbosely, the signature $\Lambda_{\bto}$ consists of the type $O_\Lambda = \{\app, \abs\}$ with arities
-  \begin{align*}
-    \arity(\app) = (\{A, B\}, [(\cdot; A \bto B), (\cdot; A)], B)
-    && 
-    \arity(\abs) = (\{A, B\}, [(\cdot, A; B)], A \bto B)
-  \end{align*}
-  For the type signature $\Sigma_{\bto, \btimes}$, the term signature $\Omega_{\bto, \btimes}$ for simply typed $\lambda$-calculus with finite products has three additional operations
-to $\Omega_{\bto}$:
-  \begin{align*}
-    \mathsf{pair}\colon A, B \rhd A, B \to A \btimes B
-    && \fst \colon A, B \rhd A \btimes B \to A
-    && \snd \colon A, B \rhd A \btimes B \to B
-  \end{align*}
-\end{example}
-
-First, it is noteworthy that the set of term constructs in a type theory need not be finite.
-For instance, a type theory may adopt a spine application which takes indefinitely many arguments---for each number $n$ of arguments an $(n+1)$-ary application construct can be introduced, so even if each operation has a definite number of arguments by definition a spine application is still expressible.
-
-More importantly, the inclusion of the set of variables used in an operation is a salient feature of our binding signatures.
-Instead of treating application as a family of constructs $\app_{A, B}$ indexed by all types $A$ and $B$, as done by \citet{Fiore2022}, we are able to identify them as a single construct $\app$.
-This is not only for brevity but also necessary to compare the type equality during type synthesis and checking.
-
-%\subsection{Intrinsically Typed Terms}
-%
-%Intrinsically typed terms for a type signature $\Sigma$ and a term signature $\Omega$ are nothing more than derivations of the intrinsic typing judgement $\Gamma \vdash_{\Sigma, \Omega} A$ constructed with only the rule $(\var)$ and the rule scheme $(\tmOp)$ displayed in \Cref{fig:intrinsic-typing}.
-%
-%The substitution $\rho\colon \Sub{\Xi}{\emptyset}$ is used to instantiate variables in the local context $\Xi$ with concrete types.
-%Accordingly, $\rho$ has to be applied to types that appear in the arity to construct a term by $\tmOp$ to ensure all types are well-formed without any use of type variables.
-%
-%\begin{figure}
-%  \centering
-%  \small
-%  \begin{mathpar}
-%    \boxed{\Gamma \vdash_{\Sigma, \Omega} A} \\
-%    \inferrule{A \in \Gamma}{\Gamma \vdash_{\Sigma, \Omega} A}\;(\var)
-%    \and
-%    \inferrule{\rho\colon \Sub{\Xi}{\emptyset}  \\ \Gamma, \simsub{\Delta_{1}}{\rho} \vdash_{\Sigma, \Omega} \simsub{A_{1}}{\rho} \quad\cdots\quad \Gamma, \simsub{\Delta_{n}}{\rho} \vdash_{\Sigma, \Omega} \simsub{A_{n}}{\rho}}
-%    {\Gamma \vdash_{\Sigma, \Omega} \simsub{A}{\rho}}\;(\tmOp)
-%    \and \text{for $o\colon \Xi \rhd [\Delta_1]A_1, \ldots, [\Delta_{n}]A_{n} \to A$ in $\Omega$}
-%  \end{mathpar}
-%  \caption{Intrinsic typing rules for a simple type theory $(\Sigma, \Omega)$}
-%  \label{fig:intrinsic-typing}
-%\end{figure}
-
-\subsection{Simple Type Theory for a Signature}
-
-\begin{figure}
-  \centering
-  \small
-  \begin{mathpar}
-    \boxed{\vdash_{\Sigma, \Omega} \isTerm{t}}
-    \\
-    \inferrule{x : \Identifier}{\vdash_{\Sigma, \Omega} \isTerm{x}}\;(\text{var})
-    \and
-    \inferrule{\cdot \vdash_{\Sigma} A \\ \vdash_{\Sigma, \Omega}\isTerm{t}}{\vdash_{\Sigma, \Omega} \isTerm{t \annote A}}\;(\text{anno})
-    \\
-    \inferrule{\vdash_{\Sigma, \Omega} \isTerm{t_1} \quad \cdots \quad \vdash_{\Sigma, \Omega} \isTerm{t_n}}
-    {\vdash_{\Sigma, \Omega} \isTerm{\tmOp_o(t_1, \ldots, t_n)}}\;(\text{op}) 
-    \and \text{for $o \colon \Xi \rhd [\Delta_1]A_{1}, \ldots, [\Delta_{n}] A_{n} \to A$ in $\Omega$}
-  \end{mathpar}
-  \caption{Raw terms for $(\Sigma, \Omega)$ with annotation}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \small
-  \begin{mathpar}
-    \boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A} \quad \text{where $\vdash_{\Sigma, \Omega} t$} \\
-    \inferrule{(x : A) \in \Gamma}{\Gamma \vdash_{\Sigma, \Omega} \isTerm{x} : A}\;(\var)
-    \and
-    \inferrule{\Gamma \vdash \isTerm{t} : A}{\Gamma \vdash (\isTerm{t \annote A}) : A}\;(\text{anno})
-    \and
-    \inferrule{\rho : \Sub{\Xi}{\emptyset} \\ \Gamma, \isTerm{\vec{x}_1} : \simsub{\Delta_{1}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_1} : \simsub{A_{1}}{\rho} \quad\cdots\quad \Gamma, \isTerm{\vec{x}_n} : \simsub{\Delta_{n}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_n} : \simsub{A_{n}}{\rho}}
-    {\Gamma \vdash_{\Sigma, \Omega} \isTerm{\tmOp_o(\vec{x}_1.\,t_1; \ldots; \vec{x}_n.\,t_n)} : \simsub{A}{\rho}}\;(\tmOp)
-    \and \text{for $o\colon \Xi \rhd [\Delta_1]A_1, \ldots, [\Delta_{n}]A_{n} \to A$ in $\Omega$}
-  \end{mathpar}
-  \caption{Typing rules for a simple type theory $(\Sigma, \Omega)$ with annotation}
-  \label{fig:extrinsic-typing}
-\end{figure}
-
-\subsection{Bidirectional Binding Signatures and their Bidirectional Type Systems}
-
-\begin{definition}
-  For a type signature $\Sigma$, a \emph{bidirectional binding signature} $\Omega$ is a set $O$ with a function
-  \[
-    \mathit{ar}\colon O \to \sum_{\Xi : \N} \left(\Cxt_{\Sigma}(\Xi) \times \Type_{\Sigma}(\Xi) \times \isDir{\Mode}\right)^* \times \Type_{\Sigma}(\Xi) \times \isDir{\Mode}.
-  \]
-  where $\Mode$ consists of two inhabitants $\chk$ for checking and $\syn$ for synthesis.
-  Bidirectional binding signatures are just binding signatures (\Cref{def:binding-signature}) augmented with a mode for each argument and its target of a construct $\tmOp_o$ in a bidirectional system, i.e.
-  an arity is a $4$-tuple
-  \[
-    \left(\Xi, \left[\left(\Delta_1; A_1; d_1\right), \ldots, \left(\Delta_{n}; A_{n}; d_n\right) \right], A, d\right)
-  \]
-  where $d$ and $d_i$'s indicate the modes of a construct and its arguments respectively.
-
-  For brevity, we write $o \colon \Xi \rhd [\Delta_1]A_{1}^{d_1}, \ldots, [\Delta_{n}] A^{d_n}_{n} \to A^{d}$ to indicate an operation with its arity. 
-\end{definition}
-
-\begin{example}
-  The term signature $\Lambda_{\bto}^{\leftrightarrows}$ for bidirectional simply typed $\lambda$-calculus introduced in \Cref{subsec:binding-sig} can be specified by operations 
-  \begin{align*}
-    \mathsf{app}\colon A, B \rhd (A \bto B)^{\Rightarrow}, A^{\Leftarrow} \to B^{\Rightarrow} &&
-    \mathsf{abs}\colon A , B \rhd [A]B^{\Leftarrow} \to (A \bto B)^{\Leftarrow}
-  \end{align*}
-  extending $\Lambda_{\bto}$ (\Cref{ex:STLC-sig}) with the mode information.
-\end{example}
-
-\LT{Some remark about signature erasure and annotation; introduce the notation $\erase{\Omega}$}
-
-\begin{figure}
-  \centering
-  \small
-  \begin{mathpar}
-    \boxed{\vdash_{\Sigma, \Omega} \isTerm{t}^\isDir{d}}
-    \quad\text{where $\vdash_{\Sigma, \erase{\Omega}} \isTerm{t}$}
-    \\
-    \inferrule{x : \Identifier}{\vdash_{\Sigma, \Omega} \isTerm{x}^\isDir{\syn}}\;(\text{var})
-    \and
-    \inferrule{\cdot \vdash_{\Sigma} A \\ \vdash_{\Sigma, \Omega}\isTerm{t}^\isDir{\chk}}{\vdash_{\Sigma, \Omega} (\isTerm{t \annote A})^\isDir{\syn}}\;(\text{anno})
-    \and
-    \inferrule{\vdash_{\Sigma, \Omega} \isTerm{t}^\isDir{\syn}}{\vdash_{\Sigma, \Omega} {\isTerm{t\subsum}}^\isDir{\chk}}\;(\text{ sub})
-  \end{mathpar}
-  \begin{mathpar}
-    \inferrule{\vdash_{\Sigma, \Omega} \isTerm{t_1}^\isDir{d_1} \quad \cdots \quad \vdash_{\Sigma, \Omega} \isTerm{t_n}^\isDir{d_n}}
-    {\vdash_{\Sigma, \Omega} \isTerm{\tmOp_o(\vec{x}_1.\, t_1; \ldots;\vec{x}_n.\, t_n)}^\isDir{d}}\;(\text{op})
-    \and \text{for $o \colon \Xi \rhd [\Delta_1]A_{1}^{d_1}, \ldots, [\Delta_{n}] A^{d_n}_{n} \to A^{d}$ in $\Omega$}
-  \end{mathpar}
-  \caption{Raw terms in some mode $d$ for a bidirectional type system $(\Sigma, \Omega)$}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \small
-  \begin{mathpar}
-    \boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A^\isDir{d}}
-    \quad \text{where $\vdash_{\Sigma, \erase{\Omega}} \isTerm{t}$ and $\vdash_{\Sigma, \Omega} \isTerm{t}^\isDir{d}$} 
-    \\
-    \inferrule{(x : A) \in \Gamma}{\Gamma \vdash_{\Sigma, \Omega} \isTerm{x} : A^\isDir{\syn}}\;(\var)
-    \and
-    \inferrule{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A^\isDir{\chk}}{\Gamma \vdash_{\Sigma, \Omega} (\isTerm{t \annote B}):  A^\isDir{\syn}}\;(\text{anno})
-    \and
-    \inferrule{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : B^{\isDir{\syn}} \\ A = B}{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t\subsum} : A^\isDir{\chk}}\;(\text{sub})
-    \\
-    \inferrule{\rho\colon \Sub{\Xi}{\emptyset} \\ \Gamma, \simsub{\isTerm{\vec{x}_{1}} : \Delta_{1}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_{1}} : \simsub{A_{1}}{\rho}^{\isDir{d_1}} \\
-      \cdots \\
-    \Gamma, \simsub{\isTerm{\vec{x}_{n}} : \Delta_{n}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_n} : \simsub{A_{n}}{\rho}^{\isDir{d_n}}}
-  {\Gamma \vdash_{\Sigma, \Omega} \isTerm{\tmOp_o(\vec{x}_1 .\, t_{1}; \ldots; \vec{x}_{n}.\, t_{n})} : \simsub{A}{\rho}^{\isDir{d}}} \; (\tmOp)
-    \and \text{for $o \colon \Xi \rhd [\Delta_1]A_{1}^{d_1}, \ldots, [\Delta_{n}] A^{d_n}_{n} \to A^{d}$ in $\Omega$}
-    \\
-    \text{Abbreviations: $\boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Rightarrow A} \defeq \boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A^\isDir{\syn}}$ and $\boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Leftarrow A} \defeq \boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A^\isDir{\chk}}$}
-  \end{mathpar}
-  \caption{Typing rules for a bidirectional system $(\Sigma, \Omega)$}
-\end{figure}
-
-\begin{example}
-
-\end{example}
-
-\section{Soundness, Completeness, and Bidirectionalisation} \label{sec:annotatability}
-
-\subsection{Soundness and Completeness}
-\begin{theorem}[Soundness]\label{thm:term-soundness}
-
-\end{theorem}
-
-\begin{theorem}[Completeness]\label{thm:term-completeness}
-
-\end{theorem}
-
-\subsection{Bidirectionalisation}
-\begin{proposition}[Decidability for Bidirectionalisation] \label{thm:bidirectionalisation}
-  For every raw term $t$ and a mode $d$, it is decidable that $t$ is in mode $d$, i.e.\ $\vdash_{\Sigma, \Omega} \isTerm{t}^\isDir{d}$ is derivable.
-\end{proposition}
-
-\subsection{Completeness versus Annotatability}
+%! TEX root = BiSig.tex
 
@@
 \begin{proof}[Proof of {\Cref{thm:bidirectional-type-synthesis}}]
-  By induction on raw terms $t$ in some mode and for the $(\text{op})$ rule on the list of arguments.
-  Moreover, if the type $A$ of $t$ is synthesised and checked against $B$ but $A \neq B$, then \Cref{thm:unique-syn} is applied to derive a contradiction.
+  We prove this statement by induction on $t$.
 
-  \begin{enumerate}
-    \item 
-        \LT{explain all cases but operation here}
+  \begin{description}
+    \item[(var)] If $x : A$ is in the context $\Gamma$, then there exists a type $A$ such that $\Gamma \vdash_{\Sigma, \Omega} x \Rightarrow A$.
+      Otherwise, if $\Gamma \vdash_{\Sigma, \Omega} x \Rightarrow A$ is derivable, then by inversion $(x : A)$ must be in $\Gamma$, a contradiction.
+    \item[(anno)]
+    \item[(sub)]
 %    \item For every mode-correct operation $o \colon \Xi \rhd [\Delta_1]A_{1}^{d_1}, \ldots, [\Delta_{n}] A^{d_n}_{n} \to A^{d}$, a context $\Gamma$, and a raw term
@@
       \end{enumerate}
-  \end{enumerate}
+  \end{description}
 \end{proof}
-
-\section{Formalisation and More Examples} \label{sec:formalisation}
-\LT{how to type Agda code without lhs2tex?}
-
-\begin{enumerate}
-  \item Bidirectional binding signature, functor, and terms (extrinsic typing, raw terms, raw terms with a mode, i.e.\ pre-typed terms).
-  \item type synthesis and checking.
-  \item examples. 
-\end{enumerate}
-
-\section{Future Work} \label{sec:future}
-
-\Josh{Getting rid of the assumption of being syntax-directed (the triangular picture)}
-
-\begin{acks}
-We thank Nathanael Arkor for the useful conversation and to thank the Programming Languages and Formal Methods Laboratory at Academia Sinica for the opportunity of sharing earlier ideas.
-
-The work is supported by the Ministry of Science and Technology of Taiwan under grant MOST 109-2222-E-001-002-MY3.
-\end{acks}
-
-\bibliographystyle{ACM-Reference-Format}
-
-\IfFileExists{reference.bib}{%
-  \bibliography{reference}
-}{\bibliography{ref}}
-
-\end{document}

### 2023-06-13 tex/sec6-formalisation.tex
[ tex ] Split BiSig.tex into smaller files

--- 2023-06-12 tex/BiSig.tex
@@
-\documentclass[acmsmall,screen]{acmart}
-\usepackage[T1]{fontenc}
-\usepackage[utf8]{inputenc}
-\usepackage[british]{babel}
-
-\usepackage{bm,braket,mathtools,mathpartir,bussproofs}
-\EnableBpAbbreviations
-\usepackage[all,cmtip]{xy}
-
-\usepackage{xfrac,xspace,xcolor,subcaption}
-\usepackage{hyperref}
-\usepackage[capitalise,noabbrev]{cleveref}
-\usepackage[inline]{enumitem} % for environment enumerate*
-\setlist[enumerate]{mode=unboxed}
-
-\setlength {\marginparwidth }{2cm}
-\usepackage[color=yellow,textsize=small]{todonotes}
-\newcommand{\LT}[1]{\todo[author=LT,inline,color=green!40,caption={}]{{#1}}}
-\newcommand{\Josh}[1]{\todo[author=Josh,inline,color=green!40,caption={}]{{#1}}}
-
-\usepackage{subfiles}
-
-\usepackage{microtype}
-
-\citestyle{acmauthoryear}
-
-\AtEndPreamble{%
-\theoremstyle{acmdefinition}
-\newtheorem{remark}[theorem]{Remark}}
-
-\input{type-notation} %% for type theory notation copied from HoTT Book
-\input{macros}
-
-\begin{document}
-
-\author{Liang-Ting Chen}
-\email{liang.ting.chen.tw@gmail.com}
-\orcid{0000-0002-3250-1331}
-\author{Hsiang-Shang Ko}
-\orcid{0000-0002-2439-1048}
-\email{joshko@iis.sinica.edu.tw}
-
-\affiliation{%
-  \institution{Academia Sinica}
-  \streetaddress{128 Academia Road}
-  \city{Taipei}
-  \country{Taiwan}
-  \postcode{115}
-}
-
-\title{A Practical Metatheory of Bidirectional Type Synthesis for Simple Types}
-
-\begin{abstract}
-  This paper proposes a formal theory of bidirectional typing.
-\end{abstract}
-
-\maketitle
-
-\section{Introduction}\label{sec:intro}
-
-\begin{enumerate}
-  \item Type checking as a transition from (parsed) abstract syntax trees to well-typed terms.
-  \item Bidirectional typing is the only solution that works that scales well.
-  \item Not yet a general theory but by now a design principle.
-    No `type checker generator' as each type checker is still designed and implemented on a case-by-case basis, but can be derived from a syntax-directed type theory `easily'.
-  \item The existing work of bidirectional typing does not take account of missing annotations carefully enough. 
-  \item A meta-problem: existing work is based on (or at least left unspecified) classical logic.
-    One has to state soundness of completeness of their type checking algorithm in a less straightforward way.
-  \item A review of what has been done towards this line of research.
-  \item Finally, our responses (contributions of this paper).
-\end{enumerate}
-
-\subsection{Related work}
-
-\citep{Xie2018}
-\subsubsection{Language formalization and its frameworks}
-\cite{Wadler2022}
-PoplMark Challenge~\citep{Aydemir2005}
-
-\cite{Cimini2020,Cimini2022}
-\citep{Fiore2022}
-\cite{Allais2021,Gheri2020,Ahrens2022}
-
-\subsubsection{Algebraic theory of type theories}
-\cite{Fiore1999,Fiore2022,Ahrens2021,Arkor2020,Hamana2011}
-
-\subsubsection{Type checker generation}
-\cite{Gast2004,Grewe2015,Pacak2020,Cimini2020}
-
-\subsubsection{Bidirectional typing}
-
-\cite{Pierce2000}
-\cite{Dunfield2021}
-
-\subsection{Plan and Contributions of this paper}
-
-Our assumptions and contributions:
-\begin{enumerate}
-  \item Assumption: Syntax-directed typing rules
-  \item Practical: bidirectionalisation over annotatability, decidability over soundness and completeness.
-  \item Generic: Type checking once and for all bidirectional type systems for simple types.
-  \item Correct and Constructive: Fully formalised in Martin-Löf type theory (with Axiom K).
-    LEM is not assumed, so a decidable statement has both computational and non-trivial logical readings.
-\end{enumerate}
-
-Plan of this paper:
-\begin{enumerate}
-  \item Key ideas (\Cref{sec:key-ideas})
-  \item Definitions for simple type theories and for bidirectional type systems (signature erasure and mode annotation) (\Cref{sec:defs})
-  \item Soundness, completeness, bidirectionalisation, and annotatability (\Cref{sec:annotatability})
-    \Josh{Erasure from bidirectional typing derivation to raw terms with a mode}
-  \item Bidirectional type inference (\Cref{sec:type-synthesis})
-  \item Formalisation with further examples (\Cref{sec:formalisation})
-  \item Future work (\Cref{sec:future})
-\end{enumerate}
-  \LT{better title?}
-
-\section{Core Ideas with Bidirectional Simply Typed \texorpdfstring{$\lambda$}{lambda}-Calculus} \label{sec:key-ideas}
-\begin{figure}
-  \centering
-  \small
-  \begin{mathpar}
-    \boxed{\Gamma \vdash \isTerm{t} \Rightarrow A}
-    \quad\text{and}\quad \boxed{\Gamma \vdash \isTerm{t} \Leftarrow A}
-    \\
-    \inferrule{(x : A) \in \Gamma}{\Gamma \vdash \isTerm{x} \Rightarrow A}\;(\text{var}^\Rightarrow)
-    \and
-    \inferrule{\Gamma \vdash \isTerm{t} \Leftarrow B \\ A = B}{\Gamma \vdash \isTerm{t : B}\Rightarrow A}\;(\text{anno})
-    \and
-    \inferrule{\Gamma \vdash \isTerm{t} \Rightarrow B \\ A = B}{\Gamma \vdash \isTerm{t\uparrow} \Leftarrow A}\;(\text{sub})
-    \and
-    \inferrule{\Gamma \vdash \isTerm{t} \Rightarrow A \bto B \\ \Gamma \vdash \isTerm{u} \Leftarrow A}{\Gamma \vdash \isTerm{t\;u} \Rightarrow B}\;(\text{app})
-    \and
-    \inferrule{\Gamma, \isTerm{x} : A \vdash \isTerm{t} \Leftarrow B}{\Gamma \vdash \isTerm{\lam{x}t} \Leftarrow A \bto B}\;(\text{abs})
-  \end{mathpar}
-  \caption{The extrinsic style of bidirectional simply typed $\lambda$-calculus with type variables}
-  \label{fig:bi-stlc}
-\end{figure}
-\begin{remark}
-  We avoid using the term `function type' and its conventional notation $\to$ at the object level on purpose, as it may be confused with function types in our type-theoretic meta-language.
-\end{remark}
-
-\section{Definitions for Simple Type Theories and Bidirectional Type Systems}\label{sec:defs}
-
-\subsection{Simple Types}
-\begin{definition}
-  A \emph{(simple type) signature} $\Sigma$ consists of a set $I$ and an \emph{arity} function $\arity\colon I \to \mathbb{N}$.
-  An inhabitant $i : I$ is meant to represent the \emph{$i$-th operation} $\tyOp_i$ and its \emph{number} $ \arity(i)$ of arguments.
-
-  The judgements of \emph{$\Sigma$-types} and \emph{$\Sigma$-contexts} over a variable set $\Theta$ are defined inductively by \Cref{fig:simple-type,fig:simple-context} respectively.
-We write $A : \Type_{\Sigma}(\Theta)$ for $\Theta \vdash_{\Sigma} A$ and $\Gamma \colon \Cxt_{\Sigma}(\Theta)$ for $\Theta \vdash_{\Sigma} \Gamma$ to disambiguate $\Sigma$-types and $\Sigma$-contexts in the case of any confusion.
-\end{definition}
-
-\begin{figure}
-  \begin{minipage}[b]{.55\textwidth}
-    \centering
-    \small
-    \begin{mathpar}
-      \boxed{\Theta\vdash_{\Sigma} A} \\
-      \inferrule{\Theta \ni X_i}{\Theta \vdash_\Sigma X_i} \and
-      \inferrule{\Theta \vdash_{\Sigma} A_1 \\ \cdots \\ \Theta \vdash_{\Sigma} A_n}{\Theta \vdash_{\Sigma} \tyOp_i(A_1, \ldots, A_n)}
-    \end{mathpar}
-    where $n = \arity_i$
-    \caption{Type formation}
-    \label{fig:simple-type}
-  \end{minipage}
-  \begin{minipage}[b]{.4\textwidth}
-    \centering
-    \small
-    \begin{mathpar}
-      \boxed{\Theta \vdash_{\Sigma} \Gamma} \\
-      \inferrule{ }{\Theta \vdash_{\Sigma} \cdot }\and
-      \inferrule{\Theta \vdash_{\Sigma} A \\ \Theta\vdash_{\Sigma} \Gamma}{\Theta \vdash_{\Sigma} \Gamma, A}
-    \end{mathpar}
-    \caption{Context formation}
-  \label{fig:simple-context}
-  \end{minipage}
-\end{figure}
-
-\begin{example}\label{ex:implication}
-  Simply typed $\lambda$-calculus typically includes a binary implication type denoted by $\bto$ and some base types to ensure that the set of all types is non-empty.
-  The type signature $\Sigma_{\bto}$ of simply typed $\lambda$-calculus consists of a binary operation $\bto$ and a base type (nullary operation) $N$.
-  In the case of simply typed $\lambda$-calculus with binary products, we can extend $\Sigma_{\bto}$ by adding a binary operation $\btimes$ to represent the binary product type.
-  This extended signature is denoted as $\Sigma_{\bto, \btimes}$.
-\end{example}
-
-\subsection{Binding Signatures} \label{subsec:binding-sig}
-\begin{definition}\label{def:binding-signature}
-  For a type signature $\Sigma$, a \emph{binding signature} $\Omega$ consists of a set $O$ and a function
-  \[
-    \mathit{ar}\colon O \to \sum_{\Xi : \N} \left(\Cxt_\Sigma(\Xi) \times \Type_\Sigma(\Xi)\right)^* \times \Type_\Sigma(\Xi).
-  \]
-  Each inhabitant $o: O$ is meant to represent a term construct $\tmOp_o$ in a simple type theory with a triple $\left(\Xi, \left[\left(\Delta_1; A_1\right), \ldots, \left(\Delta_{n}; A_{n}\right) \right], A\right)$
-  instead of a number as its arity $\arity_o$ where
-  \begin{enumerate}
-    \item $\Xi$ is the number of type variables, 
-    \item $A : \Type_\Sigma(\Xi)$ is the target type of $\tmOp_o$, and
-    \item $\left[\left(\Delta_1; A_{1}\right), \ldots, \left(\Delta_{n}; A_{n}\right) \right]$ is a list of pairs where
-    \item the $i$-th pair $(\Delta_i, A_i) :\Cxt_{\Sigma}(\Xi) \times \Term_{\Sigma}(\Xi)$ are types of the binding variables and the type of the $i$-th argument of $\tmOp_{o}$.
-  \end{enumerate}
-  For brevity, we write $o \colon \Xi \rhd (\Delta_1)A_{1}, \ldots, \left[\Delta_{n}\right] A_{n} \to A$ to indicate an operation $o$ with its arity. 
-\end{definition}
-\begin{remark}[Terminology on binding signature]
-  \cite{Aczel1978,Fiore2010}
-\end{remark}
-
-\begin{example} \label{ex:STLC-sig}
-  Given the type signature $\Sigma_{\bto}$ for the implication type, the term signature $\Lambda_{\bto}$ for simply typed $\lambda$-calculus can be described by operations
-  \begin{align*}
-    \mathsf{app}\colon A, B \rhd (A \bto B), A \to B && \mathsf{abs}\colon A , B \rhd [A]B \to (A \bto B)
-  \end{align*}
-    or, verbosely, the signature $\Lambda_{\bto}$ consists of the type $O_\Lambda = \{\app, \abs\}$ with arities
-  \begin{align*}
-    \arity(\app) = (\{A, B\}, [(\cdot; A \bto B), (\cdot; A)], B)
-    && 
-    \arity(\abs) = (\{A, B\}, [(\cdot, A; B)], A \bto B)
-  \end{align*}
-  For the type signature $\Sigma_{\bto, \btimes}$, the term signature $\Omega_{\bto, \btimes}$ for simply typed $\lambda$-calculus with finite products has three additional operations
-to $\Omega_{\bto}$:
-  \begin{align*}
-    \mathsf{pair}\colon A, B \rhd A, B \to A \btimes B
-    && \fst \colon A, B \rhd A \btimes B \to A
-    && \snd \colon A, B \rhd A \btimes B \to B
-  \end{align*}
-\end{example}
-
-First, it is noteworthy that the set of term constructs in a type theory need not be finite.
-For instance, a type theory may adopt a spine application which takes indefinitely many arguments---for each number $n$ of arguments an $(n+1)$-ary application construct can be introduced, so even if each operation has a definite number of arguments by definition a spine application is still expressible.
-
-More importantly, the inclusion of the set of variables used in an operation is a salient feature of our binding signatures.
-Instead of treating application as a family of constructs $\app_{A, B}$ indexed by all types $A$ and $B$, as done by \citet{Fiore2022}, we are able to identify them as a single construct $\app$.
-This is not only for brevity but also necessary to compare the type equality during type synthesis and checking.
-
-%\subsection{Intrinsically Typed Terms}
-%
-%Intrinsically typed terms for a type signature $\Sigma$ and a term signature $\Omega$ are nothing more than derivations of the intrinsic typing judgement $\Gamma \vdash_{\Sigma, \Omega} A$ constructed with only the rule $(\var)$ and the rule scheme $(\tmOp)$ displayed in \Cref{fig:intrinsic-typing}.
-%
-%The substitution $\rho\colon \Sub{\Xi}{\emptyset}$ is used to instantiate variables in the local context $\Xi$ with concrete types.
-%Accordingly, $\rho$ has to be applied to types that appear in the arity to construct a term by $\tmOp$ to ensure all types are well-formed without any use of type variables.
-%
-%\begin{figure}
-%  \centering
-%  \small
-%  \begin{mathpar}
-%    \boxed{\Gamma \vdash_{\Sigma, \Omega} A} \\
-%    \inferrule{A \in \Gamma}{\Gamma \vdash_{\Sigma, \Omega} A}\;(\var)
-%    \and
-%    \inferrule{\rho\colon \Sub{\Xi}{\emptyset}  \\ \Gamma, \simsub{\Delta_{1}}{\rho} \vdash_{\Sigma, \Omega} \simsub{A_{1}}{\rho} \quad\cdots\quad \Gamma, \simsub{\Delta_{n}}{\rho} \vdash_{\Sigma, \Omega} \simsub{A_{n}}{\rho}}
-%    {\Gamma \vdash_{\Sigma, \Omega} \simsub{A}{\rho}}\;(\tmOp)
-%    \and \text{for $o\colon \Xi \rhd [\Delta_1]A_1, \ldots, [\Delta_{n}]A_{n} \to A$ in $\Omega$}
-%  \end{mathpar}
-%  \caption{Intrinsic typing rules for a simple type theory $(\Sigma, \Omega)$}
-%  \label{fig:intrinsic-typing}
-%\end{figure}
-
-\subsection{Simple Type Theory of a Signature}
-
-\begin{figure}
-  \centering
-  \small
-  \begin{mathpar}
-    \boxed{\vdash_{\Sigma, \Omega} \isTerm{t}}
-    \\
-    \inferrule{x : \Identifier}{\vdash_{\Sigma, \Omega} \isTerm{x}}\;(\text{var})
-    \and
-    \inferrule{\cdot \vdash_{\Sigma} A \\ \vdash_{\Sigma, \Omega}\isTerm{t}}{\vdash_{\Sigma, \Omega} \isTerm{t \annote A}}\;(\text{anno})
-    \\
-    \inferrule{\vdash_{\Sigma, \Omega} \isTerm{t_1} \quad \cdots \quad \vdash_{\Sigma, \Omega} \isTerm{t_n}}
-    {\vdash_{\Sigma, \Omega} \isTerm{\tmOp_o(t_1, \ldots, t_n)}}\;(\text{op}) 
-    \and \text{for $o \colon \Xi \rhd [\Delta_1]A_{1}, \ldots, [\Delta_{n}] A_{n} \to A$ in $\Omega$}
-  \end{mathpar}
-  \caption{Annotatable raw terms for $(\Sigma, \Omega)$}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \small
-  \begin{mathpar}
-    \boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A} \quad \text{where $\vdash_{\Sigma, \Omega} t$} \\
-    \inferrule{(x : A) \in \Gamma}{\Gamma \vdash_{\Sigma, \Omega} \isTerm{x} : A}\;(\var)
-    \and
-    \inferrule{\Gamma \vdash \isTerm{t} : A}{\Gamma \vdash (\isTerm{t \annote A}) : A}\;(\text{anno})
-    \and
-    \inferrule{\rho : \Sub{\Xi}{\emptyset} \\ \Gamma, \isTerm{\vec{x}_1} : \simsub{\Delta_{1}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_1} : \simsub{A_{1}}{\rho} \quad\cdots\quad \Gamma, \isTerm{\vec{x}_n} : \simsub{\Delta_{n}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_n} : \simsub{A_{n}}{\rho}}
-    {\Gamma \vdash_{\Sigma, \Omega} \isTerm{\tmOp_o(\vec{x}_1.\,t_1; \ldots; \vec{x}_n.\,t_n)} : \simsub{A}{\rho}}\;(\tmOp)
-    \and \text{for $o\colon \Xi \rhd [\Delta_1]A_1, \ldots, [\Delta_{n}]A_{n} \to A$ in $\Omega$}
-  \end{mathpar}
-  \caption{Typing rules for a simple type theory $(\Sigma, \Omega)$ with annotation}
-  \label{fig:extrinsic-typing}
-\end{figure}
-
-\subsection{Bidirectional Binding Signatures and their Bidirectional Type Systems}
-
-\begin{definition}
-  For a type signature $\Sigma$, a \emph{bidirectional binding signature} $\Omega$ is a set $O$ with a function
-  \[
-    \mathit{ar}\colon O \to \sum_{\Xi : \N} \left(\Cxt_{\Sigma}(\Xi) \times \Type_{\Sigma}(\Xi) \times \isDir{\Mode}\right)^* \times \Type_{\Sigma}(\Xi) \times \isDir{\Mode}.
-  \]
-  where $\Mode$ consists of two inhabitants $\chk$ for checking and $\syn$ for synthesis.
-  Bidirectional binding signatures are just binding signatures (\Cref{def:binding-signature}) augmented with a mode for each argument and its target of a construct $\tmOp_o$ in a bidirectional system, i.e.
-  an arity is a $4$-tuple
-  \[
-    \left(\Xi, \left[\left(\Delta_1, A_1, d_1\right), \ldots, \left(\Delta_{n}, A_{n}, d_n\right) \right], A, d\right)
-  \]
-  where $d$ and $d_i$'s indicate the modes of a construct and its arguments respectively.
-
-  For brevity, we write $o \colon \Xi \rhd [\Delta_1]A_{1}^{d_1}, \ldots, [\Delta_{n}] A^{d_n}_{n} \to A^{d}$ to indicate an operation $o$ with its arity. 
-\end{definition}
-
-\begin{example}
-  The term signature $\Lambda_{\bto}^{\leftrightarrows}$ for bidirectional simply typed $\lambda$-calculus introduced in \Cref{subsec:binding-sig} can be specified by operations 
-  \begin{align*}
-    \mathsf{app}\colon A, B \rhd (A \bto B)^{\Rightarrow}, A^{\Leftarrow} \to B^{\Rightarrow} &&
-    \mathsf{abs}\colon A , B \rhd [A]B^{\Leftarrow} \to (A \bto B)^{\Leftarrow}
-  \end{align*}
-  extending $\Lambda_{\bto}$ (\Cref{ex:STLC-sig}) with the mode information.
-\end{example}
-
-\LT{deriving bidirectional binding signature from annotation}
-
-\LT{Signature Erasure}
-\begin{theorem}[Signature annotation]
-
-\end{theorem}
-
-\begin{figure}
-  \centering
-  \small
-  \begin{mathpar}
-    \boxed{\vdash_{\Sigma, \Omega} \isTerm{t}^\isDir{d}}
-    \\
-    \inferrule{x : \Identifier}{\vdash_{\Sigma, \Omega} \isTerm{x}^\isDir{\syn}}\;(\text{var})
-    \and
-    \inferrule{\cdot \vdash_{\Sigma} A \\ \vdash_{\Sigma, \Omega}\isTerm{t}^\isDir{\chk}}{\vdash_{\Sigma, \Omega} \isTerm{t \annote A}^\isDir{\syn}}\;(\text{anno})
-    \and
-    \inferrule{\vdash_{\Sigma, \Omega} \isTerm{t}^\isDir{\syn}}{\vdash_{\Sigma, \Omega} {\isTerm{t\mathord{\uparrow}}}^\isDir{\chk}}\;(\text{sub})
-  \end{mathpar}
-  \begin{mathpar}
-    \inferrule{\vdash_{\Sigma, \Omega} \isTerm{t_1}^\isDir{d_1} \quad \cdots \quad \vdash_{\Sigma, \Omega} \isTerm{t_n}^\isDir{d_n}}
-    {\vdash_{\Sigma, \Omega} \isTerm{\tmOp_o(\vec{x}_1.\, t_1; \ldots;\vec{x}_n.\, t_n)}^\isDir{d}}\;(\text{op})
-    \and \text{for $o \colon \Xi \rhd [\Delta_1]A_{1}^{d_1}, \ldots, [\Delta_{n}] A^{d_n}_{n} \to A^{d}$ in $\Omega$}
-  \end{mathpar}
-  \caption{Raw terms with a mode $d$ for a bidirectional type system $(\Sigma, \Omega)$}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \small
-  \begin{mathpar}
-    \boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A^\isDir{d}}
-    \quad \text{where $\vdash_{\Sigma, \Omega} \isTerm{t}^\isDir{d}$} 
-    \\
-    \inferrule{(x : A) \in \Gamma}{\Gamma \vdash_{\Sigma, \Omega} \isTerm{x} : A^\syn}\;(\var)
-    \and
-    \inferrule{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : B^\chk \\ A = B}{\mid \Gamma \vdash_{\Sigma, \Omega} (\isTerm{t \annote B}):  A^\syn}\;(\text{anno})
-    \and
-    \inferrule{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : B^{\syn} \\ A = B}{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t\uparrow} : A^\chk}\;(\text{sub})
-    \\
-    \inferrule{\rho\colon \Sub{\Xi}{\emptyset} \\ \Gamma, \simsub{\isTerm{\vec{x}_{1}} : \Delta_{1}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_{1}} : \simsub{A_{1}}{\rho}^{\isDir{d_1}} \\
-      \cdots \\
-    \Gamma, \simsub{\isTerm{\vec{x}_{n}} : \Delta_{n}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_n} : \simsub{A_{n}}{\rho}^{\isDir{d_n}}}
-  {\Gamma \vdash_{\Sigma, \Omega} \isTerm{\tmOp_o(\vec{x}_1 .\, t_{1}; \ldots; \vec{x}_{n}.\, t_{n})} : \simsub{A}{\rho}^{\isDir{d}}} \; (\tmOp)
-    \and \text{for $o \colon \Xi \rhd [\Delta_1]A_{1}^{d_1}, \ldots, [\Delta_{n}] A^{d_n}_{n} \to A^{d}$ in $\Omega$}
-    \\
-    \text{Abbreviations: $\boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Rightarrow A} \defeq \boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A^\syn}$ and $\boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Leftarrow A} \defeq \boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A^\chk}$}
-  \end{mathpar}
-  \caption{Typing rules for a bidirectional system $(\Sigma, \Omega)$}
-\end{figure}
-
-\begin{example}
-
-\end{example}
-
-\section{Soundness, Completeness, and Bidirectionalisation} \label{sec:annotatability}
-
-\subsection{Soundness and Completeness}
-\begin{theorem}[Soundness]\label{thm:term-soundness}
-
-\end{theorem}
-
-\begin{theorem}[Completeness]\label{thm:term-completeness}
-
-\end{theorem}
-
-\subsection{Bidirectionalisation}
-\begin{theorem}[Decidability for Bidirectionalisation] \label{thm:bidirectionalisation}
-  For every raw term $t$ and a mode $d$, it is decidable that $t$ has a mode $d$, i.e.\ $\vdash_{\Sigma, \Omega} \isTerm{t}^\isDir{d}$ is derivable.
-\end{theorem}
-
-\subsection{Completeness versus Annotatability}
-
-\section{Bidirectional Type Synthesis and Checking} \label{sec:type-synthesis}
-
-\subsection{Mode Correctness}
-
-\begin{theorem}[Uniqueness of Synthesised Types]\label{thm:unique-syn}
-  Let $(\Sigma, \Omega)$ be a bidirectional type system where $\Omega$ is mode-correct.
-  Then, for every raw $t$ with a synthesis mode and two derivations of 
-  \[
-    \Gamma \vdash_{\Sigma, \Omega} t \Rightarrow A
-    \quad\text{and}\quad
-    \Gamma \vdash_{\Sigma, \Omega} t \Rightarrow B
-  \]
-  for some types $A$ and $B$ respectively, the synthesised types must be equal, i.e.\ $A = B$.
-\end{theorem}
-
-\subsection{Trichotomy of Raw Terms by Type Synthesis}
-
-\begin{theorem}[Decidability of Bidirectional Typing Synthesis] \label{thm:bidirectional-type-synthesis}
-  For every type signature $\Sigma$ and bidirectional binding signature $\Omega$ for $\Sigma$ that is mode-correct, the following two statements hold.
-
-  \begin{enumerate}
-    \item For every context $\Gamma$ and raw term $\isTerm{t}$ with a synthesis mode, it is decidable that there exists a type $A$ and a derivation of
-      \[
-        \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Rightarrow A.
-      \]
-    \item For every context $\Gamma$, raw term $\isTerm{t}$ with a checking mode, and type $A$, it is decidable that there exists a derivation of
-      \[
-        \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Leftarrow A.
-      \]
-  \end{enumerate}
-\end{theorem}
-
-\begin{corollary}[Trichotomy of Raw Terms]
-  \LT{trichotomy over decidability?}
-  Let $\Sigma$ be a type signature and $\Omega$ a bidirectional binding signature.
-  Then, for every context $\Gamma$ and raw term $t$, exactly one of the following statements holds:
-  \begin{enumerate}
-    \item there exists a typing derivation of $\Gamma \vdash_{\Sigma, \Omega} t : A$ for some type $A$.
-    \item there is no typing derivation of $\Gamma \vdash_{\Sigma, \Omega} t : A$ for any type $A$.
-    \item $t$ cannot have a synthesis mode.
-  \end{enumerate}
-\end{corollary}
-\begin{proof}
-  Combine  \Cref{thm:term-soundness,thm:term-completeness,thm:bidirectionalisation} with \Cref{thm:bidirectional-type-synthesis},.
-
-\end{proof}
-
-\subsection{Decidability of Bidirectional Type Synthesis}
-
-\begin{proof}[Proof of {\Cref{thm:bidirectional-type-synthesis}}]
-  By induction on raw terms $t$ with a mode and for the (op) rule on the list of arguments.
-Moreover, if the type $A$ of $t$ is synthesised and checked against $B$ but $A \neq B$, then \Cref{thm:unique-syn} is applied to derive a contradiction.
-
-  \begin{enumerate}
-    \item 
-        \LT{explain all cases but operation here}
-%    \item For every mode-correct operation $o \colon \Xi \rhd [\Delta_1]A_{1}^{d_1}, \ldots, [\Delta_{n}] A^{d_n}_{n} \to A^{d}$, a context $\Gamma$, and a raw term
-%      \[
-%        \isTerm{\tmOp_{o}(\vec{x}_1.t_1,\ldots, \vec{x}_n.t_n)}, 
-%      \]
-%      with a mode $d$, it is decidable that there exists a substitution $\rho$ from $\Xi$ to $\emptyset$ and a typing derivation of
-%      \[
-%        \Gamma \vdash_{\Sigma, \Omega} \isTerm{\tmOp_{o}(\vec{x}_1.t_1,\ldots, \vec{x}_n.t_n)} : \simsub{A}{\rho}^\isDir{d}.
-%      \]
-    \item For every set $\Xi$ of type variables, list of
-      \[
-        \left(\Delta_i, A_i, d_i\right) : \Cxt_{\Sigma}(\Xi) \times \Type_{\Sigma}(\Xi) \times \Mode, 
-      \]
-      \emph{partial} substitution $\rho$ from $\Xi$ to $\emptyset$, context $\Gamma$, and raw terms $t_i$'s with a mode $d_i$,
-      \begin{enumerate}
-        \item either there is a minimal extension $\ext{\rho}$ of $\rho$ such that all     of following judgements are derivable
-          \[
-            \Gamma, \simsub{\Delta_i}{\ext{\rho}} \vdash_{\Sigma, \Omega} t_i \colon \simsub{A_i}{\ext{\rho}}^{d_i}
-          \]
-
-        \item or there is no extension $\ext{\rho}$ such that all $\Gamma, \simsub{\Delta_i}{\ext{\rho}} \vdash_{\Sigma, \Omega} t_i \colon \simsub{A_i}{\ext{\rho}}^{d_i}$ has a typing derivation. 
-      \end{enumerate}
-  \end{enumerate}
-\end{proof}
+%! TEX root = BiSig.tex
 
@@
 \LT{how to type Agda code without lhs2tex?}
-
 \begin{enumerate}
-  \item Bidirectional binding signature, functor, and terms (extrinsic typing, raw terms, raw terms with a mode, i.e.\ pre-typed terms).
+  \item Bidirectional binding signature, functor, and terms (extrinsic typing, raw terms, raw terms in some mode).
   \item type synthesis and checking.
-  \item examples. 
+  \item examples including STLC, STLC with products, PCF, Computational Calculi, Partial Differentiation
 \end{enumerate}
-
-\section{Future Work} \label{sec:future}
-
-\Josh{Getting rid of the assumption of being syntax-directed (the triangular picture)}
-
-\begin{acks}
-We thank Nathanael Arkor for the useful conversation and to thank the Programming Languages and Formal Methods Laboratory at Academia Sinica for the opportunity of sharing earlier ideas.
-
-The work is supported by the Ministry of Science and Technology of Taiwan under grant MOST 109-2222-E-001-002-MY3.
-\end{acks}
-
-\bibliographystyle{ACM-Reference-Format}
-
-\IfFileExists{reference.bib}{%
-  \bibliography{reference}
-}{\bibliography{ref}}
-
-\end{document}

### 2023-06-13 tex/sec7-future.tex
[ tex ] Split BiSig.tex into smaller files

--- 2023-06-12 tex/BiSig.tex
@@
-\documentclass[acmsmall,screen]{acmart}
-\usepackage[T1]{fontenc}
-\usepackage[utf8]{inputenc}
-\usepackage[british]{babel}
-
-\usepackage{bm,braket,mathtools,mathpartir,bussproofs}
-\EnableBpAbbreviations
-\usepackage[all,cmtip]{xy}
-
-\usepackage{xfrac,xspace,xcolor,subcaption}
-\usepackage{hyperref}
-\usepackage[capitalise,noabbrev]{cleveref}
-\usepackage[inline]{enumitem} % for environment enumerate*
-\setlist[enumerate]{mode=unboxed}
-
-\setlength {\marginparwidth }{2cm}
-\usepackage[color=yellow,textsize=small]{todonotes}
-\newcommand{\LT}[1]{\todo[author=LT,inline,color=green!40,caption={}]{{#1}}}
-\newcommand{\Josh}[1]{\todo[author=Josh,inline,color=green!40,caption={}]{{#1}}}
-
-\usepackage{subfiles}
-
-\usepackage{microtype}
-
-\citestyle{acmauthoryear}
-
-\AtEndPreamble{%
-\theoremstyle{acmdefinition}
-\newtheorem{remark}[theorem]{Remark}}
-
-\input{type-notation} %% for type theory notation copied from HoTT Book
-\input{macros}
-
-\begin{document}
-
-\author{Liang-Ting Chen}
-\email{liang.ting.chen.tw@gmail.com}
-\orcid{0000-0002-3250-1331}
-\author{Hsiang-Shang Ko}
-\orcid{0000-0002-2439-1048}
-\email{joshko@iis.sinica.edu.tw}
-
-\affiliation{%
-  \institution{Academia Sinica}
-  \streetaddress{128 Academia Road}
-  \city{Taipei}
-  \country{Taiwan}
-  \postcode{115}
-}
-
-\title{A Practical Metatheory of Bidirectional Type Synthesis for Simple Types}
-
-\begin{abstract}
-  This paper proposes a formal theory of bidirectional typing.
-\end{abstract}
-
-\maketitle
-
-\section{Introduction}\label{sec:intro}
-
-\begin{enumerate}
-  \item Type checking as a transition from (parsed) abstract syntax trees to well-typed terms.
-  \item Bidirectional typing is the only solution that works that scales well.
-  \item Not yet a general theory but by now a design principle.
-    No `type checker generator' as each type checker is still designed and implemented on a case-by-case basis, but can be derived from a syntax-directed type theory `easily'.
-  \item The existing work of bidirectional typing does not take account of missing annotations carefully enough. 
-  \item A meta-problem: existing work is based on (or at least left unspecified) classical logic.
-    One has to state soundness of completeness of their type checking algorithm in a less straightforward way.
-  \item A review of what has been done towards this line of research.
-  \item Finally, our responses (contributions of this paper).
-\end{enumerate}
-
-\subsection{Related work}
-
-\citep{Xie2018}
-\subsubsection{Language formalization and its frameworks}
-\cite{Wadler2022}
-PoplMark Challenge~\citep{Aydemir2005}
-
-\cite{Cimini2020,Cimini2022}
-\citep{Fiore2022}
-\cite{Allais2021,Gheri2020,Ahrens2022}
-
-\subsubsection{Algebraic theory of type theories}
-\cite{Fiore1999,Fiore2022,Ahrens2021,Arkor2020,Hamana2011}
-
-\subsubsection{Type checker generation}
-\cite{Gast2004,Grewe2015,Pacak2020,Cimini2020}
-
-\subsubsection{Bidirectional typing}
-
-\cite{Pierce2000}
-\cite{Dunfield2021}
-
-\subsection{Plan and Contributions of this paper}
-
-Our assumptions and contributions:
-\begin{enumerate}
-  \item Assumption: Syntax-directed typing rules
-  \item Practical: bidirectionalisation over annotatability, decidability over soundness and completeness.
-  \item Generic: Type checking once and for all bidirectional type systems for simple types.
-  \item Correct and Constructive: Fully formalised in Martin-Löf type theory (with Axiom K).
-    LEM is not assumed, so a decidable statement has both computational and non-trivial logical readings.
-\end{enumerate}
-
-Plan of this paper:
-\begin{enumerate}
-  \item Key ideas (\Cref{sec:key-ideas})
-  \item Definitions for simple type theories and for bidirectional type systems (signature erasure and mode annotation) (\Cref{sec:defs})
-  \item Soundness, completeness, bidirectionalisation, and annotatability (\Cref{sec:annotatability})
-    \Josh{Erasure from bidirectional typing derivation to raw terms with a mode}
-  \item Bidirectional type inference (\Cref{sec:type-synthesis})
-  \item Formalisation with further examples (\Cref{sec:formalisation})
-  \item Future work (\Cref{sec:future})
-\end{enumerate}
-  \LT{better title?}
-
-\section{Core Ideas with Bidirectional Simply Typed \texorpdfstring{$\lambda$}{lambda}-Calculus} \label{sec:key-ideas}
-\begin{figure}
-  \centering
-  \small
-  \begin{mathpar}
-    \boxed{\Gamma \vdash \isTerm{t} \Rightarrow A}
-    \quad\text{and}\quad \boxed{\Gamma \vdash \isTerm{t} \Leftarrow A}
-    \\
-    \inferrule{(x : A) \in \Gamma}{\Gamma \vdash \isTerm{x} \Rightarrow A}\;(\text{var}^\Rightarrow)
-    \and
-    \inferrule{\Gamma \vdash \isTerm{t} \Leftarrow B \\ A = B}{\Gamma \vdash \isTerm{t : B}\Rightarrow A}\;(\text{anno})
-    \and
-    \inferrule{\Gamma \vdash \isTerm{t} \Rightarrow B \\ A = B}{\Gamma \vdash \isTerm{t\uparrow} \Leftarrow A}\;(\text{sub})
-    \and
-    \inferrule{\Gamma \vdash \isTerm{t} \Rightarrow A \bto B \\ \Gamma \vdash \isTerm{u} \Leftarrow A}{\Gamma \vdash \isTerm{t\;u} \Rightarrow B}\;(\text{app})
-    \and
-    \inferrule{\Gamma, \isTerm{x} : A \vdash \isTerm{t} \Leftarrow B}{\Gamma \vdash \isTerm{\lam{x}t} \Leftarrow A \bto B}\;(\text{abs})
-  \end{mathpar}
-  \caption{The extrinsic style of bidirectional simply typed $\lambda$-calculus with type variables}
-  \label{fig:bi-stlc}
-\end{figure}
-\begin{remark}
-  We avoid using the term `function type' and its conventional notation $\to$ at the object level on purpose, as it may be confused with function types in our type-theoretic meta-language.
-\end{remark}
-
-\section{Definitions for Simple Type Theories and Bidirectional Type Systems}\label{sec:defs}
-
-\subsection{Simple Types}
-\begin{definition}
-  A \emph{(simple type) signature} $\Sigma$ consists of a set $I$ and an \emph{arity} function $\arity\colon I \to \mathbb{N}$.
-  An inhabitant $i : I$ is meant to represent the \emph{$i$-th operation} $\tyOp_i$ and its \emph{number} $ \arity(i)$ of arguments.
-
-  The judgements of \emph{$\Sigma$-types} and \emph{$\Sigma$-contexts} over a variable set $\Theta$ are defined inductively by \Cref{fig:simple-type,fig:simple-context} respectively.
-We write $A : \Type_{\Sigma}(\Theta)$ for $\Theta \vdash_{\Sigma} A$ and $\Gamma \colon \Cxt_{\Sigma}(\Theta)$ for $\Theta \vdash_{\Sigma} \Gamma$ to disambiguate $\Sigma$-types and $\Sigma$-contexts in the case of any confusion.
-\end{definition}
-
-\begin{figure}
-  \begin{minipage}[b]{.55\textwidth}
-    \centering
-    \small
-    \begin{mathpar}
-      \boxed{\Theta\vdash_{\Sigma} A} \\
-      \inferrule{\Theta \ni X_i}{\Theta \vdash_\Sigma X_i} \and
-      \inferrule{\Theta \vdash_{\Sigma} A_1 \\ \cdots \\ \Theta \vdash_{\Sigma} A_n}{\Theta \vdash_{\Sigma} \tyOp_i(A_1, \ldots, A_n)}
-    \end{mathpar}
-    where $n = \arity_i$
-    \caption{Type formation}
-    \label{fig:simple-type}
-  \end{minipage}
-  \begin{minipage}[b]{.4\textwidth}
-    \centering
-    \small
-    \begin{mathpar}
-      \boxed{\Theta \vdash_{\Sigma} \Gamma} \\
-      \inferrule{ }{\Theta \vdash_{\Sigma} \cdot }\and
-      \inferrule{\Theta \vdash_{\Sigma} A \\ \Theta\vdash_{\Sigma} \Gamma}{\Theta \vdash_{\Sigma} \Gamma, A}
-    \end{mathpar}
-    \caption{Context formation}
-  \label{fig:simple-context}
-  \end{minipage}
-\end{figure}
-
-\begin{example}\label{ex:implication}
-  Simply typed $\lambda$-calculus typically includes a binary implication type denoted by $\bto$ and some base types to ensure that the set of all types is non-empty.
-  The type signature $\Sigma_{\bto}$ of simply typed $\lambda$-calculus consists of a binary operation $\bto$ and a base type (nullary operation) $N$.
-  In the case of simply typed $\lambda$-calculus with binary products, we can extend $\Sigma_{\bto}$ by adding a binary operation $\btimes$ to represent the binary product type.
-  This extended signature is denoted as $\Sigma_{\bto, \btimes}$.
-\end{example}
-
-\subsection{Binding Signatures} \label{subsec:binding-sig}
-\begin{definition}\label{def:binding-signature}
-  For a type signature $\Sigma$, a \emph{binding signature} $\Omega$ consists of a set $O$ and a function
-  \[
-    \mathit{ar}\colon O \to \sum_{\Xi : \N} \left(\Cxt_\Sigma(\Xi) \times \Type_\Sigma(\Xi)\right)^* \times \Type_\Sigma(\Xi).
-  \]
-  Each inhabitant $o: O$ is meant to represent a term construct $\tmOp_o$ in a simple type theory with a triple $\left(\Xi, \left[\left(\Delta_1; A_1\right), \ldots, \left(\Delta_{n}; A_{n}\right) \right], A\right)$
-  instead of a number as its arity $\arity_o$ where
-  \begin{enumerate}
-    \item $\Xi$ is the number of type variables, 
-    \item $A : \Type_\Sigma(\Xi)$ is the target type of $\tmOp_o$, and
-    \item $\left[\left(\Delta_1; A_{1}\right), \ldots, \left(\Delta_{n}; A_{n}\right) \right]$ is a list of pairs where
-    \item the $i$-th pair $(\Delta_i, A_i) :\Cxt_{\Sigma}(\Xi) \times \Term_{\Sigma}(\Xi)$ are types of the binding variables and the type of the $i$-th argument of $\tmOp_{o}$.
-  \end{enumerate}
-  For brevity, we write $o \colon \Xi \rhd (\Delta_1)A_{1}, \ldots, \left[\Delta_{n}\right] A_{n} \to A$ to indicate an operation $o$ with its arity. 
-\end{definition}
-\begin{remark}[Terminology on binding signature]
-  \cite{Aczel1978,Fiore2010}
-\end{remark}
-
-\begin{example} \label{ex:STLC-sig}
-  Given the type signature $\Sigma_{\bto}$ for the implication type, the term signature $\Lambda_{\bto}$ for simply typed $\lambda$-calculus can be described by operations
-  \begin{align*}
-    \mathsf{app}\colon A, B \rhd (A \bto B), A \to B && \mathsf{abs}\colon A , B \rhd [A]B \to (A \bto B)
-  \end{align*}
-    or, verbosely, the signature $\Lambda_{\bto}$ consists of the type $O_\Lambda = \{\app, \abs\}$ with arities
-  \begin{align*}
-    \arity(\app) = (\{A, B\}, [(\cdot; A \bto B), (\cdot; A)], B)
-    && 
-    \arity(\abs) = (\{A, B\}, [(\cdot, A; B)], A \bto B)
-  \end{align*}
-  For the type signature $\Sigma_{\bto, \btimes}$, the term signature $\Omega_{\bto, \btimes}$ for simply typed $\lambda$-calculus with finite products has three additional operations
-to $\Omega_{\bto}$:
-  \begin{align*}
-    \mathsf{pair}\colon A, B \rhd A, B \to A \btimes B
-    && \fst \colon A, B \rhd A \btimes B \to A
-    && \snd \colon A, B \rhd A \btimes B \to B
-  \end{align*}
-\end{example}
-
-First, it is noteworthy that the set of term constructs in a type theory need not be finite.
-For instance, a type theory may adopt a spine application which takes indefinitely many arguments---for each number $n$ of arguments an $(n+1)$-ary application construct can be introduced, so even if each operation has a definite number of arguments by definition a spine application is still expressible.
-
-More importantly, the inclusion of the set of variables used in an operation is a salient feature of our binding signatures.
-Instead of treating application as a family of constructs $\app_{A, B}$ indexed by all types $A$ and $B$, as done by \citet{Fiore2022}, we are able to identify them as a single construct $\app$.
-This is not only for brevity but also necessary to compare the type equality during type synthesis and checking.
-
-%\subsection{Intrinsically Typed Terms}
-%
-%Intrinsically typed terms for a type signature $\Sigma$ and a term signature $\Omega$ are nothing more than derivations of the intrinsic typing judgement $\Gamma \vdash_{\Sigma, \Omega} A$ constructed with only the rule $(\var)$ and the rule scheme $(\tmOp)$ displayed in \Cref{fig:intrinsic-typing}.
-%
-%The substitution $\rho\colon \Sub{\Xi}{\emptyset}$ is used to instantiate variables in the local context $\Xi$ with concrete types.
-%Accordingly, $\rho$ has to be applied to types that appear in the arity to construct a term by $\tmOp$ to ensure all types are well-formed without any use of type variables.
-%
-%\begin{figure}
-%  \centering
-%  \small
-%  \begin{mathpar}
-%    \boxed{\Gamma \vdash_{\Sigma, \Omega} A} \\
-%    \inferrule{A \in \Gamma}{\Gamma \vdash_{\Sigma, \Omega} A}\;(\var)
-%    \and
-%    \inferrule{\rho\colon \Sub{\Xi}{\emptyset}  \\ \Gamma, \simsub{\Delta_{1}}{\rho} \vdash_{\Sigma, \Omega} \simsub{A_{1}}{\rho} \quad\cdots\quad \Gamma, \simsub{\Delta_{n}}{\rho} \vdash_{\Sigma, \Omega} \simsub{A_{n}}{\rho}}
-%    {\Gamma \vdash_{\Sigma, \Omega} \simsub{A}{\rho}}\;(\tmOp)
-%    \and \text{for $o\colon \Xi \rhd [\Delta_1]A_1, \ldots, [\Delta_{n}]A_{n} \to A$ in $\Omega$}
-%  \end{mathpar}
-%  \caption{Intrinsic typing rules for a simple type theory $(\Sigma, \Omega)$}
-%  \label{fig:intrinsic-typing}
-%\end{figure}
-
-\subsection{Simple Type Theory of a Signature}
-
-\begin{figure}
-  \centering
-  \small
-  \begin{mathpar}
-    \boxed{\vdash_{\Sigma, \Omega} \isTerm{t}}
-    \\
-    \inferrule{x : \Identifier}{\vdash_{\Sigma, \Omega} \isTerm{x}}\;(\text{var})
-    \and
-    \inferrule{\cdot \vdash_{\Sigma} A \\ \vdash_{\Sigma, \Omega}\isTerm{t}}{\vdash_{\Sigma, \Omega} \isTerm{t \annote A}}\;(\text{anno})
-    \\
-    \inferrule{\vdash_{\Sigma, \Omega} \isTerm{t_1} \quad \cdots \quad \vdash_{\Sigma, \Omega} \isTerm{t_n}}
-    {\vdash_{\Sigma, \Omega} \isTerm{\tmOp_o(t_1, \ldots, t_n)}}\;(\text{op}) 
-    \and \text{for $o \colon \Xi \rhd [\Delta_1]A_{1}, \ldots, [\Delta_{n}] A_{n} \to A$ in $\Omega$}
-  \end{mathpar}
-  \caption{Annotatable raw terms for $(\Sigma, \Omega)$}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \small
-  \begin{mathpar}
-    \boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A} \quad \text{where $\vdash_{\Sigma, \Omega} t$} \\
-    \inferrule{(x : A) \in \Gamma}{\Gamma \vdash_{\Sigma, \Omega} \isTerm{x} : A}\;(\var)
-    \and
-    \inferrule{\Gamma \vdash \isTerm{t} : A}{\Gamma \vdash (\isTerm{t \annote A}) : A}\;(\text{anno})
-    \and
-    \inferrule{\rho : \Sub{\Xi}{\emptyset} \\ \Gamma, \isTerm{\vec{x}_1} : \simsub{\Delta_{1}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_1} : \simsub{A_{1}}{\rho} \quad\cdots\quad \Gamma, \isTerm{\vec{x}_n} : \simsub{\Delta_{n}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_n} : \simsub{A_{n}}{\rho}}
-    {\Gamma \vdash_{\Sigma, \Omega} \isTerm{\tmOp_o(\vec{x}_1.\,t_1; \ldots; \vec{x}_n.\,t_n)} : \simsub{A}{\rho}}\;(\tmOp)
-    \and \text{for $o\colon \Xi \rhd [\Delta_1]A_1, \ldots, [\Delta_{n}]A_{n} \to A$ in $\Omega$}
-  \end{mathpar}
-  \caption{Typing rules for a simple type theory $(\Sigma, \Omega)$ with annotation}
-  \label{fig:extrinsic-typing}
-\end{figure}
-
-\subsection{Bidirectional Binding Signatures and their Bidirectional Type Systems}
-
-\begin{definition}
-  For a type signature $\Sigma$, a \emph{bidirectional binding signature} $\Omega$ is a set $O$ with a function
-  \[
-    \mathit{ar}\colon O \to \sum_{\Xi : \N} \left(\Cxt_{\Sigma}(\Xi) \times \Type_{\Sigma}(\Xi) \times \isDir{\Mode}\right)^* \times \Type_{\Sigma}(\Xi) \times \isDir{\Mode}.
-  \]
-  where $\Mode$ consists of two inhabitants $\chk$ for checking and $\syn$ for synthesis.
-  Bidirectional binding signatures are just binding signatures (\Cref{def:binding-signature}) augmented with a mode for each argument and its target of a construct $\tmOp_o$ in a bidirectional system, i.e.
-  an arity is a $4$-tuple
-  \[
-    \left(\Xi, \left[\left(\Delta_1, A_1, d_1\right), \ldots, \left(\Delta_{n}, A_{n}, d_n\right) \right], A, d\right)
-  \]
-  where $d$ and $d_i$'s indicate the modes of a construct and its arguments respectively.
-
-  For brevity, we write $o \colon \Xi \rhd [\Delta_1]A_{1}^{d_1}, \ldots, [\Delta_{n}] A^{d_n}_{n} \to A^{d}$ to indicate an operation $o$ with its arity. 
-\end{definition}
-
-\begin{example}
-  The term signature $\Lambda_{\bto}^{\leftrightarrows}$ for bidirectional simply typed $\lambda$-calculus introduced in \Cref{subsec:binding-sig} can be specified by operations 
-  \begin{align*}
-    \mathsf{app}\colon A, B \rhd (A \bto B)^{\Rightarrow}, A^{\Leftarrow} \to B^{\Rightarrow} &&
-    \mathsf{abs}\colon A , B \rhd [A]B^{\Leftarrow} \to (A \bto B)^{\Leftarrow}
-  \end{align*}
-  extending $\Lambda_{\bto}$ (\Cref{ex:STLC-sig}) with the mode information.
-\end{example}
-
-\LT{deriving bidirectional binding signature from annotation}
-
-\LT{Signature Erasure}
-\begin{theorem}[Signature annotation]
-
-\end{theorem}
-
-\begin{figure}
-  \centering
-  \small
-  \begin{mathpar}
-    \boxed{\vdash_{\Sigma, \Omega} \isTerm{t}^\isDir{d}}
-    \\
-    \inferrule{x : \Identifier}{\vdash_{\Sigma, \Omega} \isTerm{x}^\isDir{\syn}}\;(\text{var})
-    \and
-    \inferrule{\cdot \vdash_{\Sigma} A \\ \vdash_{\Sigma, \Omega}\isTerm{t}^\isDir{\chk}}{\vdash_{\Sigma, \Omega} \isTerm{t \annote A}^\isDir{\syn}}\;(\text{anno})
-    \and
-    \inferrule{\vdash_{\Sigma, \Omega} \isTerm{t}^\isDir{\syn}}{\vdash_{\Sigma, \Omega} {\isTerm{t\mathord{\uparrow}}}^\isDir{\chk}}\;(\text{sub})
-  \end{mathpar}
-  \begin{mathpar}
-    \inferrule{\vdash_{\Sigma, \Omega} \isTerm{t_1}^\isDir{d_1} \quad \cdots \quad \vdash_{\Sigma, \Omega} \isTerm{t_n}^\isDir{d_n}}
-    {\vdash_{\Sigma, \Omega} \isTerm{\tmOp_o(\vec{x}_1.\, t_1; \ldots;\vec{x}_n.\, t_n)}^\isDir{d}}\;(\text{op})
-    \and \text{for $o \colon \Xi \rhd [\Delta_1]A_{1}^{d_1}, \ldots, [\Delta_{n}] A^{d_n}_{n} \to A^{d}$ in $\Omega$}
-  \end{mathpar}
-  \caption{Raw terms with a mode $d$ for a bidirectional type system $(\Sigma, \Omega)$}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \small
-  \begin{mathpar}
-    \boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A^\isDir{d}}
-    \quad \text{where $\vdash_{\Sigma, \Omega} \isTerm{t}^\isDir{d}$} 
-    \\
-    \inferrule{(x : A) \in \Gamma}{\Gamma \vdash_{\Sigma, \Omega} \isTerm{x} : A^\syn}\;(\var)
-    \and
-    \inferrule{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : B^\chk \\ A = B}{\mid \Gamma \vdash_{\Sigma, \Omega} (\isTerm{t \annote B}):  A^\syn}\;(\text{anno})
-    \and
-    \inferrule{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : B^{\syn} \\ A = B}{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t\uparrow} : A^\chk}\;(\text{sub})
-    \\
-    \inferrule{\rho\colon \Sub{\Xi}{\emptyset} \\ \Gamma, \simsub{\isTerm{\vec{x}_{1}} : \Delta_{1}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_{1}} : \simsub{A_{1}}{\rho}^{\isDir{d_1}} \\
-      \cdots \\
-    \Gamma, \simsub{\isTerm{\vec{x}_{n}} : \Delta_{n}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_n} : \simsub{A_{n}}{\rho}^{\isDir{d_n}}}
-  {\Gamma \vdash_{\Sigma, \Omega} \isTerm{\tmOp_o(\vec{x}_1 .\, t_{1}; \ldots; \vec{x}_{n}.\, t_{n})} : \simsub{A}{\rho}^{\isDir{d}}} \; (\tmOp)
-    \and \text{for $o \colon \Xi \rhd [\Delta_1]A_{1}^{d_1}, \ldots, [\Delta_{n}] A^{d_n}_{n} \to A^{d}$ in $\Omega$}
-    \\
-    \text{Abbreviations: $\boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Rightarrow A} \defeq \boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A^\syn}$ and $\boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Leftarrow A} \defeq \boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A^\chk}$}
-  \end{mathpar}
-  \caption{Typing rules for a bidirectional system $(\Sigma, \Omega)$}
-\end{figure}
-
-\begin{example}
-
-\end{example}
-
-\section{Soundness, Completeness, and Bidirectionalisation} \label{sec:annotatability}
-
-\subsection{Soundness and Completeness}
-\begin{theorem}[Soundness]\label{thm:term-soundness}
-
-\end{theorem}
-
-\begin{theorem}[Completeness]\label{thm:term-completeness}
-
-\end{theorem}
-
-\subsection{Bidirectionalisation}
-\begin{theorem}[Decidability for Bidirectionalisation] \label{thm:bidirectionalisation}
-  For every raw term $t$ and a mode $d$, it is decidable that $t$ has a mode $d$, i.e.\ $\vdash_{\Sigma, \Omega} \isTerm{t}^\isDir{d}$ is derivable.
-\end{theorem}
-
-\subsection{Completeness versus Annotatability}
-
-\section{Bidirectional Type Synthesis and Checking} \label{sec:type-synthesis}
-
-\subsection{Mode Correctness}
-
-\begin{theorem}[Uniqueness of Synthesised Types]\label{thm:unique-syn}
-  Let $(\Sigma, \Omega)$ be a bidirectional type system where $\Omega$ is mode-correct.
-  Then, for every raw $t$ with a synthesis mode and two derivations of 
-  \[
-    \Gamma \vdash_{\Sigma, \Omega} t \Rightarrow A
-    \quad\text{and}\quad
-    \Gamma \vdash_{\Sigma, \Omega} t \Rightarrow B
-  \]
-  for some types $A$ and $B$ respectively, the synthesised types must be equal, i.e.\ $A = B$.
-\end{theorem}
-
-\subsection{Trichotomy of Raw Terms by Type Synthesis}
-
-\begin{theorem}[Decidability of Bidirectional Typing Synthesis] \label{thm:bidirectional-type-synthesis}
-  For every type signature $\Sigma$ and bidirectional binding signature $\Omega$ for $\Sigma$ that is mode-correct, the following two statements hold.
-
-  \begin{enumerate}
-    \item For every context $\Gamma$ and raw term $\isTerm{t}$ with a synthesis mode, it is decidable that there exists a type $A$ and a derivation of
-      \[
-        \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Rightarrow A.
-      \]
-    \item For every context $\Gamma$, raw term $\isTerm{t}$ with a checking mode, and type $A$, it is decidable that there exists a derivation of
-      \[
-        \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Leftarrow A.
-      \]
-  \end{enumerate}
-\end{theorem}
-
-\begin{corollary}[Trichotomy of Raw Terms]
-  \LT{trichotomy over decidability?}
-  Let $\Sigma$ be a type signature and $\Omega$ a bidirectional binding signature.
-  Then, for every context $\Gamma$ and raw term $t$, exactly one of the following statements holds:
-  \begin{enumerate}
-    \item there exists a typing derivation of $\Gamma \vdash_{\Sigma, \Omega} t : A$ for some type $A$.
-    \item there is no typing derivation of $\Gamma \vdash_{\Sigma, \Omega} t : A$ for any type $A$.
-    \item $t$ cannot have a synthesis mode.
-  \end{enumerate}
-\end{corollary}
-\begin{proof}
-  Combine  \Cref{thm:term-soundness,thm:term-completeness,thm:bidirectionalisation} with \Cref{thm:bidirectional-type-synthesis},.
-
-\end{proof}
-
-\subsection{Decidability of Bidirectional Type Synthesis}
-
-\begin{proof}[Proof of {\Cref{thm:bidirectional-type-synthesis}}]
-  By induction on raw terms $t$ with a mode and for the (op) rule on the list of arguments.
-Moreover, if the type $A$ of $t$ is synthesised and checked against $B$ but $A \neq B$, then \Cref{thm:unique-syn} is applied to derive a contradiction.
-
-  \begin{enumerate}
-    \item 
-        \LT{explain all cases but operation here}
-%    \item For every mode-correct operation $o \colon \Xi \rhd [\Delta_1]A_{1}^{d_1}, \ldots, [\Delta_{n}] A^{d_n}_{n} \to A^{d}$, a context $\Gamma$, and a raw term
-%      \[
-%        \isTerm{\tmOp_{o}(\vec{x}_1.t_1,\ldots, \vec{x}_n.t_n)}, 
-%      \]
-%      with a mode $d$, it is decidable that there exists a substitution $\rho$ from $\Xi$ to $\emptyset$ and a typing derivation of
-%      \[
-%        \Gamma \vdash_{\Sigma, \Omega} \isTerm{\tmOp_{o}(\vec{x}_1.t_1,\ldots, \vec{x}_n.t_n)} : \simsub{A}{\rho}^\isDir{d}.
-%      \]
-    \item For every set $\Xi$ of type variables, list of
-      \[
-        \left(\Delta_i, A_i, d_i\right) : \Cxt_{\Sigma}(\Xi) \times \Type_{\Sigma}(\Xi) \times \Mode, 
-      \]
-      \emph{partial} substitution $\rho$ from $\Xi$ to $\emptyset$, context $\Gamma$, and raw terms $t_i$'s with a mode $d_i$,
-      \begin{enumerate}
-        \item either there is a minimal extension $\ext{\rho}$ of $\rho$ such that all     of following judgements are derivable
-          \[
-            \Gamma, \simsub{\Delta_i}{\ext{\rho}} \vdash_{\Sigma, \Omega} t_i \colon \simsub{A_i}{\ext{\rho}}^{d_i}
-          \]
-
-        \item or there is no extension $\ext{\rho}$ such that all $\Gamma, \simsub{\Delta_i}{\ext{\rho}} \vdash_{\Sigma, \Omega} t_i \colon \simsub{A_i}{\ext{\rho}}^{d_i}$ has a typing derivation. 
-      \end{enumerate}
-  \end{enumerate}
-\end{proof}
-
-\section{Formalisation and More Examples} \label{sec:formalisation}
-\LT{how to type Agda code without lhs2tex?}
-
-\begin{enumerate}
-  \item Bidirectional binding signature, functor, and terms (extrinsic typing, raw terms, raw terms with a mode, i.e.\ pre-typed terms).
-  \item type synthesis and checking.
-  \item examples. 
-\end{enumerate}
+%! TEX root = BiSig.tex
 
@@
 
-\begin{acks}
-We thank Nathanael Arkor for the useful conversation and to thank the Programming Languages and Formal Methods Laboratory at Academia Sinica for the opportunity of sharing earlier ideas.
-
-The work is supported by the Ministry of Science and Technology of Taiwan under grant MOST 109-2222-E-001-002-MY3.
-\end{acks}
-
-\bibliographystyle{ACM-Reference-Format}
-
-\IfFileExists{reference.bib}{%
-  \bibliography{reference}
-}{\bibliography{ref}}
-
-\end{document}
+\paragraph{Polymorphic Algebraic Theories}
+\cite{Fiore2013,Hamana2011,Hamana2022}

### 2023-06-13 tex/sec5-synthesis.tex
[ tex ] Update notation

@@
       \]
-      \emph{partial} substitution $\rho$ from $\Xi$ to $\emptyset$, context $\Gamma$, and raw terms $t_i$'s in mode $d_i$,
+      for $i = 1, \ldots, n$, \emph{partial} substitution $\rho$ from $\Xi$ to $\emptyset$, context $\Gamma$, and raw terms $t_i$'s in mode $d_i$,
       \begin{enumerate}
@@
 
-        \item or there is no extension $\ext{\rho}$ such that all $\Gamma, \simsub{\Delta_i}{\ext{\rho}} \vdash_{\Sigma, \Omega} t_i \colon \simsub{A_i}{\ext{\rho}}^{d_i}$ has a typing derivation. 
+        \item or there is no extension $\sigma$ such that all $\Gamma, \simsub{\Delta_i}{\sigma} \vdash_{\Sigma, \Omega} t_i \colon \simsub{A_i}{\sigma}^{d_i}$ has a typing derivation. 
       \end{enumerate}

### 2023-06-13 tex/BiSig.tex
[ tex ] Update bib; remove old.bib
[ tex ] Merge tex

@@
 \newcommand{\Josh}[1]{\todo[author=Josh,inline,color=green!40,caption={}]{{#1}}}
-
-\usepackage{subfiles}
 
@@
 
-\section{Introduction}\label{sec:intro}
-
-\begin{enumerate}
-  \item Type checking as a transition from (parsed) abstract syntax trees to well-typed terms.
-  \item Bidirectional typing is the only solution that works that scales well.
-  \item Not yet a general theory but by now a design principle.
-    No `type checker generator' as each type checker is still designed and implemented on a case-by-case basis, but can be derived from a syntax-directed type theory `easily'.
-  \item The existing work of bidirectional typing does not take account of missing annotations carefully enough. 
-  \item A meta-problem: existing work is based on (or at least left unspecified) classical logic.
-    One has to state soundness of completeness of their type checking algorithm in a less straightforward way.
-  \item A review of what has been done towards this line of research.
-  \item Finally, our responses (contributions of this paper).
-\end{enumerate}
-
-\subsection{Related work}
-
-\citep{Xie2018}
-\subsubsection{Language formalization and its frameworks}
-\cite{Wadler2022}
-PoplMark Challenge~\citep{Aydemir2005}
-
-\cite{Cimini2020,Cimini2022}
-\citep{Fiore2022}
-\cite{Allais2021,Gheri2020,Ahrens2022}
-
-\subsubsection{Algebraic theory of type theories}
-\cite{Fiore1999,Fiore2022,Ahrens2021,Arkor2020,Hamana2011}
-
-\subsubsection{Type checker generation}
-\cite{Gast2004,Grewe2015,Pacak2020,Cimini2020}
-
-\subsubsection{Bidirectional typing}
-
-\cite{Pierce2000}
-\cite{Dunfield2021}
-
-\subsection{Plan and Contributions of this paper}
-
-Our assumptions and contributions:
-\begin{enumerate}
-  \item Assumption: Syntax-directed typing rules
-  \item Practical: bidirectionalisation over annotatability, decidability over soundness and completeness.
-  \item Generic: Type checking once and for all bidirectional type systems for simple types.
-  \item Correct and Constructive: Fully formalised in Martin-Löf type theory (with Axiom K).
-    LEM is not assumed, so a decidable statement has both computational and non-trivial logical readings.
-\end{enumerate}
-
-Plan of this paper:
-\begin{enumerate}
-  \item Key ideas (\Cref{sec:key-ideas})
-  \item Definitions for simple type theories and for bidirectional type systems (signature erasure and mode annotation) (\Cref{sec:defs})
-  \item Soundness, completeness, bidirectionalisation, and annotatability (\Cref{sec:annotatability})
-    \Josh{Erasure from bidirectional typing derivation to raw terms with a mode}
-  \item Bidirectional type inference (\Cref{sec:type-synthesis})
-  \item Formalisation with further examples (\Cref{sec:formalisation})
-  \item Future work (\Cref{sec:future})
-\end{enumerate}
-  \LT{better title?}
-
-\section{Core Ideas with Bidirectional Simply Typed \texorpdfstring{$\lambda$}{lambda}-Calculus} \label{sec:key-ideas}
-\begin{figure}
-  \centering
-  \small
-  \begin{mathpar}
-    \boxed{\Gamma \vdash \isTerm{t} \Rightarrow A}
-    \quad\text{and}\quad \boxed{\Gamma \vdash \isTerm{t} \Leftarrow A}
-    \\
-    \inferrule{(x : A) \in \Gamma}{\Gamma \vdash \isTerm{x} \Rightarrow A}\;(\text{var}^\Rightarrow)
-    \and
-    \inferrule{\Gamma \vdash \isTerm{t} \Leftarrow B \\ A = B}{\Gamma \vdash \isTerm{t : B}\Rightarrow A}\;(\text{anno})
-    \and
-    \inferrule{\Gamma \vdash \isTerm{t} \Rightarrow B \\ A = B}{\Gamma \vdash \isTerm{t\subsum} \Leftarrow A}\;(\text{sub})
-    \and
-    \inferrule{\Gamma \vdash \isTerm{t} \Rightarrow A \bto B \\ \Gamma \vdash \isTerm{u} \Leftarrow A}{\Gamma \vdash \isTerm{t\;u} \Rightarrow B}\;(\text{app})
-    \and
-    \inferrule{\Gamma, \isTerm{x} : A \vdash \isTerm{t} \Leftarrow B}{\Gamma \vdash \isTerm{\lam{x}t} \Leftarrow A \bto B}\;(\text{abs})
-  \end{mathpar}
-  \caption{Bidirectional simply typed $\lambda$-calculus $\Lambda^{\Leftrightarrow}_{\bto}$}
-  \label{fig:bi-stlc}
-\end{figure}
-\begin{remark}
-  We avoid using the term `function type' and its conventional notation $\to$ at the object level on purpose, as it may be confused with function types in our type-theoretic meta-language.
-\end{remark}
-
-\section{Definitions for Simple Type Theories and Bidirectional Type Systems}\label{sec:defs}
-
-\subsection{Simple Types}
-\begin{definition}
-  A \emph{(simple type) signature} $\Sigma$ consists of a set $I$ and an \emph{arity} function $\arity\colon I \to \mathbb{N}$.
-  An inhabitant $i : I$ is meant to represent the \emph{$i$-th operation} $\tyOp_i$ and its \emph{number} $ \arity(i)$ of arguments.
-
-  The judgements of \emph{$\Sigma$-types} and \emph{$\Sigma$-contexts} over a variable set $\Theta$ are defined inductively by \Cref{fig:simple-type,fig:simple-context} respectively.
-We write $A : \Type_{\Sigma}(\Theta)$ for $\Theta \vdash_{\Sigma} A$ and $\Gamma \colon \Cxt_{\Sigma}(\Theta)$ for $\Theta \vdash_{\Sigma} \Gamma$ to disambiguate $\Sigma$-types and $\Sigma$-contexts in the case of any confusion.
-\end{definition}
-
-\begin{figure}
-  \begin{minipage}[b]{.55\textwidth}
-    \centering
-    \small
-    \begin{mathpar}
-      \boxed{\Theta\vdash_{\Sigma} A} \\
-      \inferrule{\Theta \ni X_i}{\Theta \vdash_\Sigma X_i} \and
-    \inferrule{\Theta \vdash_{\Sigma} A_1 \\ \cdots \\ \Theta \vdash_{\Sigma} A_{\arity(o)}}{\Theta \vdash_{\Sigma} \tyOp_o(A_1, \ldots, A_{\arity(o)})}
-    \end{mathpar}
-    \caption{Type formation}
-    \label{fig:simple-type}
-  \end{minipage}
-  \begin{minipage}[b]{.4\textwidth}
-    \centering
-    \small
-    \begin{mathpar}
-      \boxed{\Theta \vdash_{\Sigma} \Gamma} \\
-      \inferrule{ }{\Theta \vdash_{\Sigma} \cdot }\and
-      \inferrule{\Theta \vdash_{\Sigma} A \\ \Theta\vdash_{\Sigma} \Gamma}{\Theta \vdash_{\Sigma} \Gamma, A}
-    \end{mathpar}
-    \caption{Context formation}
-  \label{fig:simple-context}
-  \end{minipage}
-\end{figure}
-
-\begin{example}\label{ex:implication}
-  Simply typed $\lambda$-calculus $\Lambda_{\bto}$ typically includes a binary implication type denoted by $\bto$ and some base types to ensure that the set of all types is non-empty.
-  The type signature $\Sigma_{\bto}$ of simply typed $\lambda$-calculus consists of a binary operation $\bto$ and a base type (nullary operation) $N$.
-  In the case of simply typed $\lambda$-calculus $\Lambda_{\bto, \btimes}$ with binary products, we can extend $\Sigma_{\bto}$ by adding a binary operation $\btimes$ to represent the binary product type, denoted as $\Sigma_{\bto, \btimes}$.
-\end{example}
-
-\subsection{Binding Signatures} \label{subsec:binding-sig}
-\begin{definition}\label{def:binding-signature}
-  For a type signature $\Sigma$, a \emph{binding signature} $\Omega$ consists of a set $O$ and a function
-  \[
-    \mathit{ar}\colon O \to \sum_{\Xi : \N} \left(\Cxt_\Sigma(\Xi) \times \Type_\Sigma(\Xi)\right)^* \times \Type_\Sigma(\Xi).
-  \]
-  Each inhabitant $o: O$ is meant to represent a term construct $\tmOp_o$ in a simple type theory with a triple $\left(\Xi, \left[\left(\Delta_1; A_1\right), \ldots, \left(\Delta_{n}; A_{n}\right) \right], A\right)$
-  instead of a number as its arity $\arity_o$ where
-  \begin{enumerate}
-    \item $\Xi$ is the number of type variables, 
-    \item $A : \Type_\Sigma(\Xi)$ is the target type of $\tmOp_o$, and
-    \item $\left[\left(\Delta_1; A_{1}\right), \ldots, \left(\Delta_{n}; A_{n}\right) \right]$ is a list of pairs where
-    \item the $i$-th pair $(\Delta_i, A_i) :\Cxt_{\Sigma}(\Xi) \times \Term_{\Sigma}(\Xi)$ are types of the binding variables and the type of the $i$-th argument of $\tmOp_{o}$.
-  \end{enumerate}
-  For brevity, we write $o \colon \Xi \rhd [\Delta_1]A_{1}, \ldots, \left[\Delta_{n}\right] A_{n} \to A$ to indicate an operation with its arity. 
-\end{definition}
-\begin{remark}[Terminology on binding signature]
-  \cite{Aczel1978,Fiore2010}
-\end{remark}
-
-\begin{example} \label{ex:STLC-sig}
-  Given the type signature $\Sigma_{\bto}$ for the implication type, the term signature $\Lambda_{\bto}$ for simply typed $\lambda$-calculus can be described by operations
-  \begin{align*}
-    \mathsf{app}\colon A, B \rhd (A \bto B), A \to B && \mathsf{abs}\colon A , B \rhd [A]B \to (A \bto B)
-  \end{align*}
-    or, verbosely, the signature $\Lambda_{\bto}$ consists of the type $O_\Lambda = \{\app, \abs\}$ with arities
-  \begin{align*}
-    \arity(\app) = (\{A, B\}, [(\cdot; A \bto B), (\cdot; A)], B)
-    && 
-    \arity(\abs) = (\{A, B\}, [(\cdot, A; B)], A \bto B)
-  \end{align*}
-  For the type signature $\Sigma_{\bto, \btimes}$, the term signature $\Omega_{\bto, \btimes}$ for simply typed $\lambda$-calculus with finite products has three additional operations
-to $\Omega_{\bto}$:
-  \begin{align*}
-    \mathsf{pair}\colon A, B \rhd A, B \to A \btimes B
-    && \fst \colon A, B \rhd A \btimes B \to A
-    && \snd \colon A, B \rhd A \btimes B \to B
-  \end{align*}
-\end{example}
-
-First, it is noteworthy that the set of term constructs in a type theory need not be finite.
-For instance, a type theory may adopt a spine application which takes indefinitely many arguments---for each number $n$ of arguments an $(n+1)$-ary application construct can be introduced, so even if each operation has a definite number of arguments by definition a spine application is still expressible.
-
-More importantly, the inclusion of the set of variables used in an operation is a salient feature of our binding signatures.
-Instead of treating application as a family of constructs $\app_{A, B}$ indexed by all types $A$ and $B$, as done by \citet{Fiore2022}, we are able to identify them as a single construct $\app$.
-This is not only for brevity but also necessary to compare the type equality during type synthesis and checking.
-
-%\subsection{Intrinsically Typed Terms}
-%
-%Intrinsically typed terms for a type signature $\Sigma$ and a term signature $\Omega$ are nothing more than derivations of the intrinsic typing judgement $\Gamma \vdash_{\Sigma, \Omega} A$ constructed with only the rule $(\var)$ and the rule scheme $(\tmOp)$ displayed in \Cref{fig:intrinsic-typing}.
-%
-%The substitution $\rho\colon \Sub{\Xi}{\emptyset}$ is used to instantiate variables in the local context $\Xi$ with concrete types.
-%Accordingly, $\rho$ has to be applied to types that appear in the arity to construct a term by $\tmOp$ to ensure all types are well-formed without any use of type variables.
-%
-%\begin{figure}
-%  \centering
-%  \small
-%  \begin{mathpar}
-%    \boxed{\Gamma \vdash_{\Sigma, \Omega} A} \\
-%    \inferrule{A \in \Gamma}{\Gamma \vdash_{\Sigma, \Omega} A}\;(\var)
-%    \and
-%    \inferrule{\rho\colon \Sub{\Xi}{\emptyset}  \\ \Gamma, \simsub{\Delta_{1}}{\rho} \vdash_{\Sigma, \Omega} \simsub{A_{1}}{\rho} \quad\cdots\quad \Gamma, \simsub{\Delta_{n}}{\rho} \vdash_{\Sigma, \Omega} \simsub{A_{n}}{\rho}}
-%    {\Gamma \vdash_{\Sigma, \Omega} \simsub{A}{\rho}}\;(\tmOp)
-%    \and \text{for $o\colon \Xi \rhd [\Delta_1]A_1, \ldots, [\Delta_{n}]A_{n} \to A$ in $\Omega$}
-%  \end{mathpar}
-%  \caption{Intrinsic typing rules for a simple type theory $(\Sigma, \Omega)$}
-%  \label{fig:intrinsic-typing}
-%\end{figure}
-
-\subsection{Simple Type Theory for a Signature}
-
-\begin{figure}
-  \centering
-  \small
-  \begin{mathpar}
-    \boxed{\vdash_{\Sigma, \Omega} \isTerm{t}}
-    \\
-    \inferrule{x : \Identifier}{\vdash_{\Sigma, \Omega} \isTerm{x}}\;(\text{var})
-    \and
-    \inferrule{\cdot \vdash_{\Sigma} A \\ \vdash_{\Sigma, \Omega}\isTerm{t}}{\vdash_{\Sigma, \Omega} \isTerm{t \annote A}}\;(\text{anno})
-    \\
-    \inferrule{\vdash_{\Sigma, \Omega} \isTerm{t_1} \quad \cdots \quad \vdash_{\Sigma, \Omega} \isTerm{t_n}}
-    {\vdash_{\Sigma, \Omega} \isTerm{\tmOp_o(t_1, \ldots, t_n)}}\;(\text{op}) 
-    \and \text{for $o \colon \Xi \rhd [\Delta_1]A_{1}, \ldots, [\Delta_{n}] A_{n} \to A$ in $\Omega$}
-  \end{mathpar}
-  \caption{Raw terms for $(\Sigma, \Omega)$ with annotation}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \small
-  \begin{mathpar}
-    \boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A} \quad \text{where $\vdash_{\Sigma, \Omega} t$} \\
-    \inferrule{(x : A) \in \Gamma}{\Gamma \vdash_{\Sigma, \Omega} \isTerm{x} : A}\;(\var)
-    \and
-    \inferrule{\Gamma \vdash \isTerm{t} : A}{\Gamma \vdash (\isTerm{t \annote A}) : A}\;(\text{anno})
-    \and
-    \inferrule{\rho : \Sub{\Xi}{\emptyset} \\ \Gamma, \isTerm{\vec{x}_1} : \simsub{\Delta_{1}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_1} : \simsub{A_{1}}{\rho} \quad\cdots\quad \Gamma, \isTerm{\vec{x}_n} : \simsub{\Delta_{n}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_n} : \simsub{A_{n}}{\rho}}
-    {\Gamma \vdash_{\Sigma, \Omega} \isTerm{\tmOp_o(\vec{x}_1.\,t_1; \ldots; \vec{x}_n.\,t_n)} : \simsub{A}{\rho}}\;(\tmOp)
-    \and \text{for $o\colon \Xi \rhd [\Delta_1]A_1, \ldots, [\Delta_{n}]A_{n} \to A$ in $\Omega$}
-  \end{mathpar}
-  \caption{Typing rules for a simple type theory $(\Sigma, \Omega)$ with annotation}
-  \label{fig:extrinsic-typing}
-\end{figure}
-
-\subsection{Bidirectional Binding Signatures and their Bidirectional Type Systems}
-
-\begin{definition}
-  For a type signature $\Sigma$, a \emph{bidirectional binding signature} $\Omega$ is a set $O$ with a function
-  \[
-    \mathit{ar}\colon O \to \sum_{\Xi : \N} \left(\Cxt_{\Sigma}(\Xi) \times \Type_{\Sigma}(\Xi) \times \isDir{\Mode}\right)^* \times \Type_{\Sigma}(\Xi) \times \isDir{\Mode}.
-  \]
-  where $\Mode$ consists of two inhabitants $\chk$ for checking and $\syn$ for synthesis.
-  Bidirectional binding signatures are just binding signatures (\Cref{def:binding-signature}) augmented with a mode for each argument and its target of a construct $\tmOp_o$ in a bidirectional system, i.e.
-  an arity is a $4$-tuple
-  \[
-    \left(\Xi, \left[\left(\Delta_1; A_1; d_1\right), \ldots, \left(\Delta_{n}; A_{n}; d_n\right) \right], A, d\right)
-  \]
-  where $d$ and $d_i$'s indicate the modes of a construct and its arguments respectively.
-
-  For brevity, we write $o \colon \Xi \rhd [\Delta_1]A_{1}^{d_1}, \ldots, [\Delta_{n}] A^{d_n}_{n} \to A^{d}$ to indicate an operation with its arity. 
-\end{definition}
-
-\begin{example}
-  The term signature $\Lambda_{\bto}^{\leftrightarrows}$ for bidirectional simply typed $\lambda$-calculus introduced in \Cref{subsec:binding-sig} can be specified by operations 
-  \begin{align*}
-    \mathsf{app}\colon A, B \rhd (A \bto B)^{\Rightarrow}, A^{\Leftarrow} \to B^{\Rightarrow} &&
-    \mathsf{abs}\colon A , B \rhd [A]B^{\Leftarrow} \to (A \bto B)^{\Leftarrow}
-  \end{align*}
-  extending $\Lambda_{\bto}$ (\Cref{ex:STLC-sig}) with the mode information.
-\end{example}
-
-\LT{Some remark about signature erasure and annotation; introduce the notation $\erase{\Omega}$}
-
-\begin{figure}
-  \centering
-  \small
-  \begin{mathpar}
-    \boxed{\vdash_{\Sigma, \Omega} \isTerm{t}^\isDir{d}}
-    \quad\text{where $\vdash_{\Sigma, \erase{\Omega}} \isTerm{t}$}
-    \\
-    \inferrule{x : \Identifier}{\vdash_{\Sigma, \Omega} \isTerm{x}^\isDir{\syn}}\;(\text{var})
-    \and
-    \inferrule{\cdot \vdash_{\Sigma} A \\ \vdash_{\Sigma, \Omega}\isTerm{t}^\isDir{\chk}}{\vdash_{\Sigma, \Omega} (\isTerm{t \annote A})^\isDir{\syn}}\;(\text{anno})
-    \and
-    \inferrule{\vdash_{\Sigma, \Omega} \isTerm{t}^\isDir{\syn}}{\vdash_{\Sigma, \Omega} {\isTerm{t\subsum}}^\isDir{\chk}}\;(\text{ sub})
-  \end{mathpar}
-  \begin{mathpar}
-    \inferrule{\vdash_{\Sigma, \Omega} \isTerm{t_1}^\isDir{d_1} \quad \cdots \quad \vdash_{\Sigma, \Omega} \isTerm{t_n}^\isDir{d_n}}
-    {\vdash_{\Sigma, \Omega} \isTerm{\tmOp_o(\vec{x}_1.\, t_1; \ldots;\vec{x}_n.\, t_n)}^\isDir{d}}\;(\text{op})
-    \and \text{for $o \colon \Xi \rhd [\Delta_1]A_{1}^{d_1}, \ldots, [\Delta_{n}] A^{d_n}_{n} \to A^{d}$ in $\Omega$}
-  \end{mathpar}
-  \caption{Raw terms in some mode $d$ for a bidirectional type system $(\Sigma, \Omega)$}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \small
-  \begin{mathpar}
-    \boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A^\isDir{d}}
-    \quad \text{where $\vdash_{\Sigma, \erase{\Omega}} \isTerm{t}$ and $\vdash_{\Sigma, \Omega} \isTerm{t}^\isDir{d}$} 
-    \\
-    \inferrule{(x : A) \in \Gamma}{\Gamma \vdash_{\Sigma, \Omega} \isTerm{x} : A^\isDir{\syn}}\;(\var)
-    \and
-    \inferrule{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A^\isDir{\chk}}{\Gamma \vdash_{\Sigma, \Omega} (\isTerm{t \annote B}):  A^\isDir{\syn}}\;(\text{anno})
-    \and
-    \inferrule{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : B^{\isDir{\syn}} \\ A = B}{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t\subsum} : A^\isDir{\chk}}\;(\text{sub})
-    \\
-    \inferrule{\rho\colon \Sub{\Xi}{\emptyset} \\ \Gamma, \simsub{\isTerm{\vec{x}_{1}} : \Delta_{1}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_{1}} : \simsub{A_{1}}{\rho}^{\isDir{d_1}} \\
-      \cdots \\
-    \Gamma, \simsub{\isTerm{\vec{x}_{n}} : \Delta_{n}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_n} : \simsub{A_{n}}{\rho}^{\isDir{d_n}}}
-  {\Gamma \vdash_{\Sigma, \Omega} \isTerm{\tmOp_o(\vec{x}_1 .\, t_{1}; \ldots; \vec{x}_{n}.\, t_{n})} : \simsub{A}{\rho}^{\isDir{d}}} \; (\tmOp)
-    \and \text{for $o \colon \Xi \rhd [\Delta_1]A_{1}^{d_1}, \ldots, [\Delta_{n}] A^{d_n}_{n} \to A^{d}$ in $\Omega$}
-    \\
-    \text{Abbreviations: $\boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Rightarrow A} \defeq \boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A^\isDir{\syn}}$ and $\boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Leftarrow A} \defeq \boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A^\isDir{\chk}}$}
-  \end{mathpar}
-  \caption{Typing rules for a bidirectional system $(\Sigma, \Omega)$}
-\end{figure}
-
-\begin{example}
-
-\end{example}
-
-\section{Soundness, Completeness, and Bidirectionalisation} \label{sec:annotatability}
-
-\subsection{Soundness and Completeness}
-\begin{theorem}[Soundness]\label{thm:term-soundness}
-
-\end{theorem}
-
-\begin{theorem}[Completeness]\label{thm:term-completeness}
-
-\end{theorem}
-
-\subsection{Bidirectionalisation}
-\begin{proposition}[Decidability for Bidirectionalisation] \label{thm:bidirectionalisation}
-  For every raw term $t$ and a mode $d$, it is decidable that $t$ is in mode $d$, i.e.\ $\vdash_{\Sigma, \Omega} \isTerm{t}^\isDir{d}$ is derivable.
-\end{proposition}
-
-\subsection{Completeness versus Annotatability}
-
-\section{Bidirectional Type Synthesis and Checking} \label{sec:type-synthesis}
-
-\subsection{Mode Correctness}
-
-\begin{theorem}[Uniqueness of Synthesised Types]\label{thm:unique-syn}
-  Let $(\Sigma, \Omega)$ be a bidirectional type system where $\Omega$ is mode-correct.
-  Then, for every raw term $t$ in synthesising mode and two derivations of 
-  \[
-    \Gamma \vdash_{\Sigma, \Omega} t \Rightarrow A
-    \quad\text{and}\quad
-    \Gamma \vdash_{\Sigma, \Omega} t \Rightarrow B
-  \]
-  for some types $A$ and $B$ respectively, the synthesised types must be equal, i.e.\ $A = B$.
-\end{theorem}
-
-\subsection{Trichotomy of Raw Terms by Type Synthesis}
-
-\begin{theorem}[Decidability of Bidirectional Typing Synthesis] \label{thm:bidirectional-type-synthesis}
-  For every type signature $\Sigma$ and bidirectional binding signature $\Omega$ for $\Sigma$ that is mode-correct, the following two statements hold.
-
-  \begin{enumerate}
-    \item For every context $\Gamma$ and raw term $\isTerm{t}$ in synthesising mode, it is decidable that there is a type~$A$ and a derivation of
-      \[
-        \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Rightarrow A.
-      \]
-    \item For every context $\Gamma$, raw term $\isTerm{t}$ in checking mode, and type $A$, it is decidable that there is a derivation of
-      \[
-        \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Leftarrow A.
-      \]
-  \end{enumerate}
-\end{theorem}
-
-\begin{corollary}[Trichotomy of Raw Terms]
-  Let $\Sigma$ be a type signature and $\Omega$ a bidirectional binding signature.
-  Then, for every context $\Gamma$ and raw term $t$, exactly one of the following statements holds:
-  \begin{enumerate}
-    \item $t$ is synthesising and there exists a typing derivation of\, $\Gamma \vdash_{\Sigma, \Omega} t : A$ for some type $A$.
-    \item $t$ is synthesising but there is no typing derivation of\, $\Gamma \vdash_{\Sigma, \Omega} t : A$ for any type $A$.
-    \item $t$ is not synthesising.
-  \end{enumerate}
-\end{corollary}
-\begin{proof}
-  Combine  \Cref{thm:term-soundness,thm:term-completeness,thm:bidirectionalisation} with \Cref{thm:bidirectional-type-synthesis}.
-
-\end{proof}
-
-\subsection{Decidability of Bidirectional Type Synthesis}
-
-\begin{proof}[Proof of {\Cref{thm:bidirectional-type-synthesis}}]
-  By induction on raw terms $t$ in some mode and for the $(\text{op})$ rule on the list of arguments.
-  Moreover, if the type $A$ of $t$ is synthesised and checked against $B$ but $A \neq B$, then \Cref{thm:unique-syn} is applied to derive a contradiction.
-
-  \begin{enumerate}
-    \item 
-        \LT{explain all cases but operation here}
-%    \item For every mode-correct operation $o \colon \Xi \rhd [\Delta_1]A_{1}^{d_1}, \ldots, [\Delta_{n}] A^{d_n}_{n} \to A^{d}$, a context $\Gamma$, and a raw term
-%      \[
-%        \isTerm{\tmOp_{o}(\vec{x}_1.t_1,\ldots, \vec{x}_n.t_n)}, 
-%      \]
-%      with a mode $d$, it is decidable that there exists a substitution $\rho$ from $\Xi$ to $\emptyset$ and a typing derivation of
-%      \[
-%        \Gamma \vdash_{\Sigma, \Omega} \isTerm{\tmOp_{o}(\vec{x}_1.t_1,\ldots, \vec{x}_n.t_n)} : \simsub{A}{\rho}^\isDir{d}.
-%      \]
-    \item For every set $\Xi$ of type variables, list of
-      \[
-        \left(\Delta_i, A_i, d_i\right) : \Cxt_{\Sigma}(\Xi) \times \Type_{\Sigma}(\Xi) \times \Mode, 
-      \]
-      \emph{partial} substitution $\rho$ from $\Xi$ to $\emptyset$, context $\Gamma$, and raw terms $t_i$'s in mode $d_i$,
-      \begin{enumerate}
-        \item either there is a minimal extension $\ext{\rho}$ of $\rho$ such that all     of following judgements are derivable
-          \[
-            \Gamma, \simsub{\Delta_i}{\ext{\rho}} \vdash_{\Sigma, \Omega} t_i \colon \simsub{A_i}{\ext{\rho}}^{d_i}
-          \]
-
-        \item or there is no extension $\ext{\rho}$ such that all $\Gamma, \simsub{\Delta_i}{\ext{\rho}} \vdash_{\Sigma, \Omega} t_i \colon \simsub{A_i}{\ext{\rho}}^{d_i}$ has a typing derivation. 
-      \end{enumerate}
-  \end{enumerate}
-\end{proof}
-
-\section{Formalisation and More Examples} \label{sec:formalisation}
-\LT{how to type Agda code without lhs2tex?}
-
-\begin{enumerate}
-  \item Bidirectional binding signature, functor, and terms (extrinsic typing, raw terms, raw terms with a mode, i.e.\ pre-typed terms).
-  \item type synthesis and checking.
-  \item examples. 
-\end{enumerate}
-
-\section{Future Work} \label{sec:future}
-
-\Josh{Getting rid of the assumption of being syntax-directed (the triangular picture)}
+\input{sec1-intro}
+\input{sec2-outline}
+\input{sec3-defs.tex}
+\input{sec4-bidirectionalisation.tex}
+\input{sec5-synthesis.tex}
+\input{sec6-formalisation.tex}
+\input{sec7-future.tex}
 

### 2023-06-13 tex/sec2-outline.tex
[ tex ] Merge tex

%! TEX root = BiSig.tex

\section{Core Ideas with Bidirectional Simply Typed \texorpdfstring{$\lambda$}{lambda}-Calculus} \label{sec:key-ideas}
\begin{figure}
  \centering
  \small
  \begin{mathpar}
    \boxed{\Gamma \vdash \isTerm{t} \Rightarrow A}
    \quad\text{and}\quad \boxed{\Gamma \vdash \isTerm{t} \Leftarrow A}
    \\
    \inferrule{(x : A) \in \Gamma}{\Gamma \vdash \isTerm{x} \Rightarrow A}\;(\text{var}^\Rightarrow)
    \and
    \inferrule{\Gamma \vdash \isTerm{t} \Leftarrow B \\ A = B}{\Gamma \vdash \isTerm{t : B}\Rightarrow A}\;(\text{anno})
    \and
    \inferrule{\Gamma \vdash \isTerm{t} \Rightarrow B \\ A = B}{\Gamma \vdash \isTerm{t\subsum} \Leftarrow A}\;(\text{sub})
    \and
    \inferrule{\Gamma \vdash \isTerm{t} \Rightarrow A \bto B \\ \Gamma \vdash \isTerm{u} \Leftarrow A}{\Gamma \vdash \isTerm{t\;u} \Rightarrow B}\;(\text{app})
    \and
    \inferrule{\Gamma, \isTerm{x} : A \vdash \isTerm{t} \Leftarrow B}{\Gamma \vdash \isTerm{\lam{x}t} \Leftarrow A \bto B}\;(\text{abs})
  \end{mathpar}
  \caption{Bidirectional simply typed $\lambda$-calculus $\Lambda^{\Leftrightarrow}_{\bto}$}
  \label{fig:bi-stlc}
\end{figure}
\begin{remark}
  We avoid using the term `function type' and its conventional notation $\to$ at the object level on purpose, as it may be confused with function types in our type-theoretic meta-language.
\end{remark}

### 2023-06-13 tex/sec5-synthesis.tex
[ tex ] Add the pen-and-paper proof of bidirectional type synthesis

@@
 \subsection{Mode Correctness}
+\begin{definition}[Mode-Correctness]
+\end{definition}
 
-\begin{theorem}[Uniqueness of Synthesised Types]\label{thm:unique-syn}
+\begin{theorem}[Uniqueness of the Synthesised Types]\label{thm:unique-syn}
   Let $(\Sigma, \Omega)$ be a bidirectional type system where $\Omega$ is mode-correct.
@@
   \[
-    \Gamma \vdash_{\Sigma, \Omega} t \Rightarrow A
+    \Gamma \vdash_{\Sigma, \Omega} t \syn A
     \quad\text{and}\quad
-    \Gamma \vdash_{\Sigma, \Omega} t \Rightarrow B
+    \Gamma \vdash_{\Sigma, \Omega} t \syn B
   \]
@@
 \begin{theorem}[Decidability of Bidirectional Typing Synthesis] \label{thm:bidirectional-type-synthesis}
-  For every type signature $\Sigma$ and bidirectional binding signature $\Omega$ for $\Sigma$ that is mode-correct, the following two statements hold.
-
+  Let $(\Sigma, \Omega)$ be a mode-correct bidirectional type system.
+  For every context $\Gamma$ and raw term $\isTerm{t}$ in mode $d$, 
   \begin{enumerate}
-    \item For every context $\Gamma$ and raw term $\isTerm{t}$ in synthesising mode, it is decidable that there is a type~$A$ and a derivation of
+    \item if $\isTerm{t}$ is synthesising, then it is decidable that there is a type~$A$ and a derivation of
       \[
-        \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Rightarrow A.
+        \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \syn A.
       \]
-    \item For every context $\Gamma$, raw term $\isTerm{t}$ in checking mode, and type $A$, it is decidable that there is a derivation of
+    \item if $\isTerm{t}$ is checking, then for every type $A$ it is decidable that there is a derivation of
       \[
-        \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Leftarrow A.
+        \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \chk A.
       \]
@@
 \begin{corollary}[Trichotomy of Raw Terms]
-  Let $\Sigma$ be a type signature and $\Omega$ a bidirectional binding signature.
+  Let $(\Sigma, \Omega)$ be a mode-correct bidirectional type system.
   Then, for every context $\Gamma$ and raw term $t$, exactly one of the following statements holds:
   \begin{enumerate}
-    \item $t$ is synthesising and there exists a typing derivation of\, $\Gamma \vdash_{\Sigma, \Omega} t : A$ for some type $A$.
-    \item $t$ is synthesising but there is no typing derivation of\, $\Gamma \vdash_{\Sigma, \Omega} t : A$ for any type $A$.
+    \item $t$ is synthesising and there is a typing derivation of\, $\Gamma \vdash_{\Sigma, \Omega} t : A$ for some type $A$.
+    \item $t$ is synthesising but there exists no typing derivation of\, $\Gamma \vdash_{\Sigma, \Omega} t : A$ for any type $A$.
     \item $t$ is not synthesising.
@@
 
+\begin{lemma}\label{lem:args-induction}
+  Assume \Cref{thm:bidirectional-type-synthesis}.
+  Then, for every set $\Xi$ of type variables, list 
+  \[
+    \mathit{as} = \left([\Delta_i] A_{i}^{d_i}\right)_{i = 1}^{n}
+  \]
+  list of raw terms $t_i$'s in mode $d_i$, context $\Gamma$, and \emph{partial} substitution $\rho$ from $\Xi$ to $\emptyset$
+  \begin{enumerate}
+    \item either there is a minimal extension $\ext{\rho}$ of $\rho$ defined on $\synvar(\mathit{as})$ such that all of following judgements are derivable
+      \[
+        \Gamma, \simsub{\Delta_i}{\ext{\rho}} \vdash_{\Sigma, \Omega} t_i \colon \simsub{A_i}{\ext{\rho}}^{d_i}
+      \]
+
+    \item or there is no extension $\sigma$ such that all $\Gamma, \simsub{\Delta_i}{\sigma} \vdash_{\Sigma, \Omega} t_i \colon \simsub{A_i}{\sigma}^{d_i}$ has a typing derivation. 
+  \end{enumerate}
+\end{lemma}
+
 \begin{proof}[Proof of {\Cref{thm:bidirectional-type-synthesis}}]
-  We prove this statement by induction on $t$.
+  We prove this statement by induction on the derivation of $\vdash_{\Sigma, \Omega} \isTerm{t}^\isDir{d}$ (instead of $t$ alone).
+  The first two cases (var) and (anno) are straightforward and have nothing to do with mode-correctness directly.
+  The third case (sub) invokes the uniqueness of synthesised types to refute the case that $\Gamma \vdash t \syn B$ but $A \neq B$ for a given type $A$ to check.
+  The last case (op) is the interesting part of this theorem, as we need \Cref{lem:args-induction}.
+  \begin{description}
+    \item[(var)] If $t$ is a variable $x$ in the context $\Gamma$, then there exists a type $A$ such that $\Gamma \vdash_{\Sigma, \Omega} x \syn A$.
+      Otherwise, if $x$ is not in $\Gamma$ but $\Gamma \vdash_{\Sigma, \Omega} x \syn A$ is derivable, then by inversion $x$ must be in $\Gamma$, a contradiction.
+    \item[(anno)] For a raw term of the form $(t : A)$ in synthesising mode, it is decidable that $\Gamma \vdash t \chk A$ is derivable or not by induction hypothesis.
+      \begin{enumerate}
+        \item If $\Gamma \vdash t \chk A$ is derivable, then $\Gamma \vdash (t : A) \syn A$ is derivable.
+        \item If $\Gamma \vdash t \chk A$ is not derivable but $\Gamma \vdash (t : A) \syn$ is derivable, then by inversion $\Gamma \vdash t \chk A$ is derivable, thus a contradiction.
+      \end{enumerate}
 
-  \begin{description}
-    \item[(var)] If $x : A$ is in the context $\Gamma$, then there exists a type $A$ such that $\Gamma \vdash_{\Sigma, \Omega} x \Rightarrow A$.
-      Otherwise, if $\Gamma \vdash_{\Sigma, \Omega} x \Rightarrow A$ is derivable, then by inversion $(x : A)$ must be in $\Gamma$, a contradiction.
-    \item[(anno)]
-    \item[(sub)]
+    \item[(sub)] If $t$ is of the form $t\subsum$ in checking mode, then similarly by induction it is decidable that $\Gamma \vdash t \syn B$ for some $B$:
+      \begin{enumerate}
+        \item If $\Gamma \vdash t \syn B$ is derivable for some $B$, then we have to consider any type $A$ and it is decidable if $A$ is equal to $B$:
+          \begin{enumerate}
+            \item if $A = B$ then we have arrived a derivation of $\Gamma \vdash t\subsum \chk A$.
+            \item if $A \neq B$ but $\Gamma \vdash t \subsum \chk A$ is derivable, then by inversion $\Gamma \vdash t \syn A$.
+              However, by \Cref{thm:unique-syn}, synthesised types $A$ and $B$ must be equal, so it is a contradiction.
+          \end{enumerate}
+        \item if $\Gamma \vdash t \syn C$ is not derivable for any type $C$ but $\Gamma \vdash t\subsum \chk A$ is derivable, then by inversion $\Gamma \vdash t \syn B$ for some $B = A$, thus a contradiction.
+      \end{enumerate}
+
 %    \item For every mode-correct operation $o \colon \Xi \rhd [\Delta_1]A_{1}^{d_1}, \ldots, [\Delta_{n}] A^{d_n}_{n} \to A^{d}$, a context $\Gamma$, and a raw term
@@
 %      \]
-    \item For every set $\Xi$ of type variables, list of
-      \[
-        \left(\Delta_i, A_i, d_i\right) : \Cxt_{\Sigma}(\Xi) \times \Type_{\Sigma}(\Xi) \times \Mode, 
-      \]
-      for $i = 1, \ldots, n$, \emph{partial} substitution $\rho$ from $\Xi$ to $\emptyset$, context $\Gamma$, and raw terms $t_i$'s in mode $d_i$,
-      \begin{enumerate}
-        \item either there is a minimal extension $\ext{\rho}$ of $\rho$ such that all     of following judgements are derivable
-          \[
-            \Gamma, \simsub{\Delta_i}{\ext{\rho}} \vdash_{\Sigma, \Omega} t_i \colon \simsub{A_i}{\ext{\rho}}^{d_i}
-          \]
+    \item[(op)]
+      Suppose that $t$ is of them $\tmOp_o(\vec{x}_1.t_1; \ldots; \vec{x}_n.t_n)$ for some mode-correct operation $o$ in $\Omega$.
+      No matter $t$ is checking or synthesising, we need to build derivations for its arguments incrementally and maintain a set of unifiers, i.e.\ a partial substitution $\rho\colon \PSub(\Xi, \emptyset)$ incrementally which is defined nowhere initially if $t$ is synthesising or on variables of the target $A$ if $t$ is checking.
 
-        \item or there is no extension $\sigma$ such that all $\Gamma, \simsub{\Delta_i}{\sigma} \vdash_{\Sigma, \Omega} t_i \colon \simsub{A_i}{\sigma}^{d_i}$ has a typing derivation. 
-      \end{enumerate}
   \end{description}

### 2023-06-13 tex/sec1-intro.tex
[ tex ] Add points to address; finalise the paper title

@@
   \item Finally, our responses (contributions of this paper).
+\end{enumerate}
+
+\begin{enumerate}
+  \item Use application in spin form to justify our design choice: type expression of arbitrary rank and the number of constructs in a calculus.
+  \item Argue that syntax-directed form is essential and practical.
+  \item Pfenning's recipe is only of design guide not a technical requirement (there are many exceptions).
 \end{enumerate}
@@
 \cite{Dunfield2021}
+\cite{Jones2007}
 

### 2023-06-13 tex/sec3-defs.tex
[ tex ] Merge tex
[ tex ] Add the pen-and-paper proof of bidirectional type synthesis
[ tex ] Add points to address; finalise the paper title

@@
 
-\subsection{Simple Type Theory for a Signature}
+\subsection{Simple Type Theory}
 
@@
 
-\subsection{Bidirectional Binding Signatures and their Bidirectional Type Systems}
+\subsection{Bidirectional Binding Signatures and Bidirectional Type Systems}
 
@@
   \begin{align*}
-    \mathsf{app}\colon A, B \rhd (A \bto B)^{\Rightarrow}, A^{\Leftarrow} \to B^{\Rightarrow} &&
-    \mathsf{abs}\colon A , B \rhd [A]B^{\Leftarrow} \to (A \bto B)^{\Leftarrow}
+    \mathsf{app}\colon A, B \rhd (A \bto B)^{\syn}, A^{\chk} \to B^{\syn} &&
+    \mathsf{abs}\colon A , B \rhd [A]B^{\chk} \to (A \bto B)^{\chk}
   \end{align*}
@@
   \end{mathpar}
-  \caption{Raw terms in some mode $d$ for a bidirectional type system $(\Sigma, \Omega)$}
+  \caption{Raw terms in mode $d$ for a bidirectional type system $(\Sigma, \Omega)$}
 \end{figure}
@@
     \\
-    \text{Abbreviations: $\boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Rightarrow A} \defeq \boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A^\isDir{\syn}}$ and $\boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Leftarrow A} \defeq \boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A^\isDir{\chk}}$}
+    \text{Abbreviations: $\boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \syn A} \defeq \boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A^\isDir{\syn}}$ and $\boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \chk A} \defeq \boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A^\isDir{\chk}}$}
   \end{mathpar}

### 2023-06-14 tex/sec3-defs.tex
[ tex ] Added acmart files; separate compilation for TeXShop
[ tex ] Using the same color as by Dunfield et al.

@@
-%! TEX root = BiSig.tex
+%!TEX root = BiSig.tex
 
-\section{Definitions for Simple Type Theories and Bidirectional Type Systems}\label{sec:defs}
+\section{Simple Type Theories and Bidirectional Type Systems}\label{sec:defs}
 
@@
   The judgements of \emph{$\Sigma$-types} and \emph{$\Sigma$-contexts} over a variable set $\Theta$ are defined inductively by \Cref{fig:simple-type,fig:simple-context} respectively.
-We write $A : \Type_{\Sigma}(\Theta)$ for $\Theta \vdash_{\Sigma} A$ and $\Gamma \colon \Cxt_{\Sigma}(\Theta)$ for $\Theta \vdash_{\Sigma} \Gamma$ to disambiguate $\Sigma$-types and $\Sigma$-contexts in the case of any confusion.
+We write $A : \Type_{\Sigma}(\Theta)$ for $\Theta |-_{\Sigma} A$ and $\Gamma \colon \Cxt_{\Sigma}(\Theta)$ for $\Theta |-_{\Sigma} \Gamma$ to disambiguate $\Sigma$-types and $\Sigma$-contexts in the case of any confusion.
 \end{definition}
@@
     \small
+    \judgbox{\Theta|-_{\Sigma} A}{$A$ is well-formed with type variables in $\Theta$}
     \begin{mathpar}
-      \boxed{\Theta\vdash_{\Sigma} A} \\
-      \inferrule{\Theta \ni X_i}{\Theta \vdash_\Sigma X_i} \and
-    \inferrule{\Theta \vdash_{\Sigma} A_1 \\ \cdots \\ \Theta \vdash_{\Sigma} A_{\arity(o)}}{\Theta \vdash_{\Sigma} \tyOp_o(A_1, \ldots, A_{\arity(o)})}
+      \inferrule{\Theta \ni X_i}{\Theta |-_\Sigma X_i} \and
+    \inferrule{\Theta |-_{\Sigma} A_1 \\ \cdots \\ \Theta |-_{\Sigma} A_{\arity(o)}}{\Theta |-_{\Sigma} \tyOp_o(A_1, \ldots, A_{\arity(o)})}
     \end{mathpar}
-    \caption{Type formation}
+    \caption{Type formation rules}
     \label{fig:simple-type}
@@
     \small
+    \judgbox{\Theta |-_{\Sigma} \Gamma}{}
     \begin{mathpar}
-      \boxed{\Theta \vdash_{\Sigma} \Gamma} \\
-      \inferrule{ }{\Theta \vdash_{\Sigma} \cdot }\and
-      \inferrule{\Theta \vdash_{\Sigma} A \\ \Theta\vdash_{\Sigma} \Gamma}{\Theta \vdash_{\Sigma} \Gamma, A}
+      \inferrule{ }{\Theta |-_{\Sigma} \cdot }\and
+      \inferrule{\Theta |-_{\Sigma} A \\ \Theta|-_{\Sigma} \Gamma}{\Theta |-_{\Sigma} \Gamma, A}
     \end{mathpar}
-    \caption{Context formation}
+    \caption{Context formation rules}
   \label{fig:simple-context}
@@
   \end{align*}
-    or, verbosely, the signature $\Lambda_{\bto}$ consists of the type $O_\Lambda = \{\app, \abs\}$ with arities
+  or, verbosely, the signature $\Lambda_{\bto}$ consists of the type $O_\Lambda = \{\abs, \app\}$ with arities
   \begin{align*}
@@
 %
-%Intrinsically typed terms for a type signature $\Sigma$ and a term signature $\Omega$ are nothing more than derivations of the intrinsic typing judgement $\Gamma \vdash_{\Sigma, \Omega} A$ constructed with only the rule $(\var)$ and the rule scheme $(\tmOp)$ displayed in \Cref{fig:intrinsic-typing}.
+%Intrinsically typed terms for a type signature $\Sigma$ and a term signature $\Omega$ are nothing more than derivations of the intrinsic typing judgement $\Gamma |-_{\Sigma, \Omega} A$ constructed with only the rule $(\var)$ and the rule scheme $(\tmOp)$ displayed in \Cref{fig:intrinsic-typing}.
 %
@@
 %  \begin{mathpar}
-%    \boxed{\Gamma \vdash_{\Sigma, \Omega} A} \\
-%    \inferrule{A \in \Gamma}{\Gamma \vdash_{\Sigma, \Omega} A}\;(\var)
+%    \boxed{\Gamma |-_{\Sigma, \Omega} A} \\
+%    \inferrule{A \in \Gamma}{\Gamma |-_{\Sigma, \Omega} A}\;(\var)
 %    \and
-%    \inferrule{\rho\colon \Sub{\Xi}{\emptyset}  \\ \Gamma, \simsub{\Delta_{1}}{\rho} \vdash_{\Sigma, \Omega} \simsub{A_{1}}{\rho} \quad\cdots\quad \Gamma, \simsub{\Delta_{n}}{\rho} \vdash_{\Sigma, \Omega} \simsub{A_{n}}{\rho}}
-%    {\Gamma \vdash_{\Sigma, \Omega} \simsub{A}{\rho}}\;(\tmOp)
+%    \inferrule{\rho\colon \Sub{\Xi}{\emptyset}  \\ \Gamma, \simsub{\Delta_{1}}{\rho} |-_{\Sigma, \Omega} \simsub{A_{1}}{\rho} \quad\cdots\quad \Gamma, \simsub{\Delta_{n}}{\rho} |-_{\Sigma, \Omega} \simsub{A_{n}}{\rho}}
+%    {\Gamma |-_{\Sigma, \Omega} \simsub{A}{\rho}}\;(\tmOp)
 %    \and \text{for $o\colon \Xi \rhd [\Delta_1]A_1, \ldots, [\Delta_{n}]A_{n} \to A$ in $\Omega$}
@@
 
-\subsection{Simple Type Theory}
+\subsection{Simple Type Theories}
 
@@
   \small
+  \judgbox{|-_{\Sigma, \Omega} \isTerm{t}}{$t$ is a raw term for the signature $(\Sigma, \Omega)$}
   \begin{mathpar}
-    \boxed{\vdash_{\Sigma, \Omega} \isTerm{t}}
+    \inferrule{x : \Identifier}{|-_{\Sigma, \Omega} \isTerm{x}}\,\Rule{Var}
+    \and
+    \inferrule{\cdot |-_{\Sigma} A \\ |-_{\Sigma, \Omega}\isTerm{t}}{|-_{\Sigma, \Omega} \isTerm{t \annote A}}\,\Rule{Anno}
     \\
-    \inferrule{x : \Identifier}{\vdash_{\Sigma, \Omega} \isTerm{x}}\;(\text{var})
-    \and
-    \inferrule{\cdot \vdash_{\Sigma} A \\ \vdash_{\Sigma, \Omega}\isTerm{t}}{\vdash_{\Sigma, \Omega} \isTerm{t \annote A}}\;(\text{anno})
-    \\
-    \inferrule{\vdash_{\Sigma, \Omega} \isTerm{t_1} \quad \cdots \quad \vdash_{\Sigma, \Omega} \isTerm{t_n}}
-    {\vdash_{\Sigma, \Omega} \isTerm{\tmOp_o(t_1, \ldots, t_n)}}\;(\text{op}) 
+    \inferrule{|-_{\Sigma, \Omega} \isTerm{t_1} \quad \cdots \quad |-_{\Sigma, \Omega} \isTerm{t_n}}
+    {|-_{\Sigma, \Omega} \isTerm{\tmOp_o(t_1, \ldots, t_n)}}\,\Rule{Op}_o
     \and \text{for $o \colon \Xi \rhd [\Delta_1]A_{1}, \ldots, [\Delta_{n}] A_{n} \to A$ in $\Omega$}
   \end{mathpar}
-  \caption{Raw terms for $(\Sigma, \Omega)$ with annotation}
+  \caption{Raw term formation rules}
 \end{figure}
@@
   \small
+  \judgbox{\Gamma |-_{\Sigma, \Omega} \isTerm{t} : A}{A raw term $t$ has a type $A$ (without type variables) under $\Gamma$ for the signature $(\Sigma, \Omega)$}
   \begin{mathpar}
-    \boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A} \quad \text{where $\vdash_{\Sigma, \Omega} t$} \\
-    \inferrule{(x : A) \in \Gamma}{\Gamma \vdash_{\Sigma, \Omega} \isTerm{x} : A}\;(\var)
+    \inferrule{(x : A) \in \Gamma}{\Gamma |-_{\Sigma, \Omega} \isTerm{x} : A}\,\Rule{Var}
     \and
-    \inferrule{\Gamma \vdash \isTerm{t} : A}{\Gamma \vdash (\isTerm{t \annote A}) : A}\;(\text{anno})
+    \inferrule{\Gamma |- \isTerm{t} : A}{\Gamma |- (\isTerm{t \annote A}) : A}\,\Rule{Anno}
     \and
-    \inferrule{\rho : \Sub{\Xi}{\emptyset} \\ \Gamma, \isTerm{\vec{x}_1} : \simsub{\Delta_{1}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_1} : \simsub{A_{1}}{\rho} \quad\cdots\quad \Gamma, \isTerm{\vec{x}_n} : \simsub{\Delta_{n}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_n} : \simsub{A_{n}}{\rho}}
-    {\Gamma \vdash_{\Sigma, \Omega} \isTerm{\tmOp_o(\vec{x}_1.\,t_1; \ldots; \vec{x}_n.\,t_n)} : \simsub{A}{\rho}}\;(\tmOp)
+    \inferrule{\rho : \Sub{\Xi}{\emptyset} \\ \Gamma, \isTerm{\vec{x}_1} : \simsub{\Delta_{1}}{\rho} |-_{\Sigma, \Omega} \isTerm{t_1} : \simsub{A_{1}}{\rho} \quad\cdots\quad \Gamma, \isTerm{\vec{x}_n} : \simsub{\Delta_{n}}{\rho} |-_{\Sigma, \Omega} \isTerm{t_n} : \simsub{A_{n}}{\rho}}
+    {\Gamma |-_{\Sigma, \Omega} \isTerm{\tmOp_o(\vec{x}_1.\,t_1; \ldots; \vec{x}_n.\,t_n)} : \simsub{A}{\rho}}\,\Rule{Op}
     \and \text{for $o\colon \Xi \rhd [\Delta_1]A_1, \ldots, [\Delta_{n}]A_{n} \to A$ in $\Omega$}
   \end{mathpar}
-  \caption{Typing rules for a simple type theory $(\Sigma, \Omega)$ with annotation}
+  \caption{Typing rules}
   \label{fig:extrinsic-typing}
@@
   \[
-    \mathit{ar}\colon O \to \sum_{\Xi : \N} \left(\Cxt_{\Sigma}(\Xi) \times \Type_{\Sigma}(\Xi) \times \isDir{\Mode}\right)^* \times \Type_{\Sigma}(\Xi) \times \isDir{\Mode}.
+    \mathit{ar}\colon O \to \sum_{\Xi : \N} \left(\Cxt_{\Sigma}(\Xi) \times \Type_{\Sigma}(\Xi) \times {\Mode}\right)^* \times \Type_{\Sigma}(\Xi) \times {\Mode}.
   \]
@@
 \begin{example}
-  The term signature $\Lambda_{\bto}^{\leftrightarrows}$ for bidirectional simply typed $\lambda$-calculus introduced in \Cref{subsec:binding-sig} can be specified by operations 
+  The term signature $\Lambda_{\bto}^{\Leftrightarrow}$ for bidirectional simply typed $\lambda$-calculus introduced in \Cref{subsec:binding-sig} can be specified by operations 
   \begin{align*}
@@
   \small
+  \judgbox{|-_{\Sigma, \Omega} \isTerm{t}^\dir{d}}{$t$ is a raw term for $(\Sigma, \erase{\Omega})$ in mode $d$ for  $(\Sigma, \Omega)$}
   \begin{mathpar}
-    \boxed{\vdash_{\Sigma, \Omega} \isTerm{t}^\isDir{d}}
-    \quad\text{where $\vdash_{\Sigma, \erase{\Omega}} \isTerm{t}$}
-    \\
-    \inferrule{x : \Identifier}{\vdash_{\Sigma, \Omega} \isTerm{x}^\isDir{\syn}}\;(\text{var})
+    \inferrule{x : \Identifier}{|-_{\Sigma, \Omega} \isTerm{x}^{\syn}}\,\SynRule{Var}
     \and
-    \inferrule{\cdot \vdash_{\Sigma} A \\ \vdash_{\Sigma, \Omega}\isTerm{t}^\isDir{\chk}}{\vdash_{\Sigma, \Omega} (\isTerm{t \annote A})^\isDir{\syn}}\;(\text{anno})
+    \inferrule{\cdot |-_{\Sigma} A \\ |-_{\Sigma, \Omega}\isTerm{t}^{\chk}}{|-_{\Sigma, \Omega} (\isTerm{t \annote A})^{\syn}}\,\SynRule{Anno}
     \and
-    \inferrule{\vdash_{\Sigma, \Omega} \isTerm{t}^\isDir{\syn}}{\vdash_{\Sigma, \Omega} {\isTerm{t\subsum}}^\isDir{\chk}}\;(\text{sub})
+    \inferrule{|-_{\Sigma, \Omega} \isTerm{t}^{\syn}}{|-_{\Sigma, \Omega} \isTerm{t}^{\chk}}\,\ChkRule{Sub}
   \end{mathpar}
   \begin{mathpar}
-    \inferrule{\vdash_{\Sigma, \Omega} \isTerm{t_1}^\isDir{d_1} \quad \cdots \quad \vdash_{\Sigma, \Omega} \isTerm{t_n}^\isDir{d_n}}
-    {\vdash_{\Sigma, \Omega} \isTerm{\tmOp_o(\vec{x}_1.\, t_1; \ldots;\vec{x}_n.\, t_n)}^\isDir{d}}\;(\text{op})
-    \and \text{for $o \colon \Xi \rhd [\Delta_1]A_{1}^{d_1}, \ldots, [\Delta_{n}] A^{d_n}_{n} \to A^{d}$ in $\Omega$}
+    \inferrule{|-_{\Sigma, \Omega} \isTerm{t_1}^\dir{d_1} \quad \cdots \quad |-_{\Sigma, \Omega} \isTerm{t_n}^\dir{d_n}}
+    {|-_{\Sigma, \Omega} \isTerm{\tmOp_o(\vec{x}_1.\, t_1; \ldots;\vec{x}_n.\, t_n)}^\dir{d}}\,\Rule{Op}
+    \and \text{for $o \colon \Xi \rhd [\Delta_1]A_{1}^\dir{d_1}, \ldots, [\Delta_{n}] A_{n}^\dir{d_n} \to A^\dir{d}$ in $\Omega$}
   \end{mathpar}
-  \caption{Raw terms in mode $d$ for a bidirectional type system $(\Sigma, \Omega)$}
+  \caption{Preprocessed Raw terms}
 \end{figure}
@@
   \small
+  \judgbox{\Gamma |-_{\Sigma, \Omega} \isTerm{t} :^\dir{d} A}{A raw term $t$ in mode $d$ has a type $A$ under $\Gamma$}
   \begin{mathpar}
-    \boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A^\isDir{d}}
-    \quad \text{where $\vdash_{\Sigma, \erase{\Omega}} \isTerm{t}$ and $\vdash_{\Sigma, \Omega} \isTerm{t}^\isDir{d}$} 
+    \inferrule{(x : A) \in \Gamma}{\Gamma |-_{\Sigma, \Omega} \isTerm{x} :^{\syn} A}\,\SynRule{Var}
+    \and
+    \inferrule{\Gamma |-_{\Sigma, \Omega} \isTerm{t} :^{\chk} A}{\Gamma |-_{\Sigma, \Omega} (\isTerm{t \annote B}) :^{\syn}  A}\,\SynRule{Anno}
+    \and
+    \inferrule{\Gamma |-_{\Sigma, \Omega} \isTerm{t} :^{{\syn}} B \\ A = B}{\Gamma |-_{\Sigma, \Omega} \isTerm{t} :^{\chk} A}\,\ChkRule{Sub}
     \\
-    \inferrule{(x : A) \in \Gamma}{\Gamma \vdash_{\Sigma, \Omega} \isTerm{x} : A^\isDir{\syn}}\;(\var)
-    \and
-    \inferrule{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A^\isDir{\chk}}{\Gamma \vdash_{\Sigma, \Omega} (\isTerm{t \annote B}):  A^\isDir{\syn}}\;(\text{anno})
-    \and
-    \inferrule{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : B^{\isDir{\syn}} \\ A = B}{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t\subsum} : A^\isDir{\chk}}\;(\text{sub})
-    \\
-    \inferrule{\rho\colon \Sub{\Xi}{\emptyset} \\ \Gamma, \simsub{\isTerm{\vec{x}_{1}} : \Delta_{1}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_{1}} : \simsub{A_{1}}{\rho}^{\isDir{d_1}} \\
+    \inferrule{\rho\colon \Sub{\Xi}{\emptyset} \\ \Gamma, \simsub{\isTerm{\vec{x}_{1}} : \Delta_{1}}{\rho} |-_{\Sigma, \Omega} \isTerm{t_{1}} :^\dir{d_1} \simsub{A_1}{\rho} \\
       \cdots \\
-    \Gamma, \simsub{\isTerm{\vec{x}_{n}} : \Delta_{n}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_n} : \simsub{A_{n}}{\rho}^{\isDir{d_n}}}
-  {\Gamma \vdash_{\Sigma, \Omega} \isTerm{\tmOp_o(\vec{x}_1 .\, t_{1}; \ldots; \vec{x}_{n}.\, t_{n})} : \simsub{A}{\rho}^{\isDir{d}}} \; (\tmOp)
-    \and \text{for $o \colon \Xi \rhd [\Delta_1]A_{1}^{d_1}, \ldots, [\Delta_{n}] A^{d_n}_{n} \to A^{d}$ in $\Omega$}
-    \\
-    \text{Abbreviations: $\boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \syn A} \defeq \boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A^\isDir{\syn}}$ and $\boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \chk A} \defeq \boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A^\isDir{\chk}}$}
+    \Gamma, \simsub{\isTerm{\vec{x}_{n}} : \Delta_{n}}{\rho} |-_{\Sigma, \Omega} \isTerm{t_n} :^{\dir{d_n}} \simsub{A_{n}}{\rho}}
+    {\Gamma |-_{\Sigma, \Omega} \isTerm{\tmOp_o(\vec{x}_1 .\, t_{1}; \ldots; \vec{x}_{n}.\, t_{n})} :^{\dir{d}} \simsub{A}{\rho}} \,\mathsf{Op}
+    \and \text{for $o \colon \Xi \rhd [\Delta_1]A_{1}^\dir{d_1}, \ldots, [\Delta_{n}] A_{n}^\dir{d_n} \to A^\dir{d}$ in $\Omega$}
   \end{mathpar}
-  \caption{Typing rules for a bidirectional system $(\Sigma, \Omega)$}
+  \begin{minipage}{.15\linewidth}
+  Abbreviations: 
+  \end{minipage}
+  \begin{minipage}{.8\linewidth}
+    \begin{flushleft}
+    $\boxed{\Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn A} \defeq \boxed{\Gamma |-_{\Sigma, \Omega} \isTerm{t} :^{\syn} A}$ \quad A raw term $t$ synthesises a type $A$ under $\Gamma$ \\
+    $\boxed{\Gamma |-_{\Sigma, \Omega} \isTerm{t} \chk A} \defeq \boxed{\Gamma |-_{\Sigma, \Omega} \isTerm{t} :^{\chk} A}$ \quad A raw term $t$ checks against a type $A$ under $\Gamma$
+    \end{flushleft}
+  \end{minipage}
+  \caption{Bidirectional typing rules}
 \end{figure}

### 2023-06-14 tex/sec4-bidirectionalisation.tex
[ tex ] Added acmart files; separate compilation for TeXShop
[ tex ] Using the same color as by Dunfield et al.

@@
-%! TEX root = BiSig.tex
+%!TEX root = BiSig.tex
 
@@
 \begin{proposition}[Decidability for Bidirectionalisation] \label{thm:bidirectionalisation}
-  For every raw term $t$ and a mode $d$, it is decidable that $t$ is in mode $d$, i.e.\ $\vdash_{\Sigma, \Omega} \isTerm{t}^\isDir{d}$ is derivable.
+  For every raw term $t$ and a mode $d$, it is decidable whether $t$ is in mode $d$.
 \end{proposition}
 
-\subsection{Completeness versus Annotatability}
+\subsection{Annotatability is not Completeness}

### 2023-06-14 tex/sec5-synthesis.tex
[ tex ] Added acmart files; separate compilation for TeXShop
[ tex ] Using the same color as by Dunfield et al.

@@
   \[
-    \Gamma \vdash_{\Sigma, \Omega} t \syn A
+    \Gamma |-_{\Sigma, \Omega} t \syn A
     \quad\text{and}\quad
-    \Gamma \vdash_{\Sigma, \Omega} t \syn B
+    \Gamma |-_{\Sigma, \Omega} t \syn B
   \]
@@
   Let $(\Sigma, \Omega)$ be a mode-correct bidirectional type system.
-  For every context $\Gamma$ and raw term $\isTerm{t}$ in mode $d$, 
-  \begin{enumerate}
-    \item if $\isTerm{t}$ is synthesising, then it is decidable that there is a type~$A$ and a derivation of
+  For every context $\Gamma$ and raw term $\isTerm{t}$ in synthesising mode, it is decidable whether
   \[
-        \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \syn A.
+    \Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn A
   \]
-    \item if $\isTerm{t}$ is checking, then for every type $A$ it is decidable that there is a derivation of
-      \[
-        \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \chk A.
-      \]
-  \end{enumerate}
+  has a derivation for some type $A$.
 \end{theorem}
@@
   \begin{enumerate}
-    \item $t$ is synthesising and there is a typing derivation of\, $\Gamma \vdash_{\Sigma, \Omega} t : A$ for some type $A$.
-    \item $t$ is synthesising but there exists no typing derivation of\, $\Gamma \vdash_{\Sigma, \Omega} t : A$ for any type $A$.
+    \item $t$ is synthesising and the judgement $\Gamma |-_{\Sigma, \Omega} t : A$ has a derivation for some type $A$.
+    \item $t$ is synthesising but the judgement $\Gamma |-_{\Sigma, \Omega} t : A$ has no derivation for any type $A$.
     \item $t$ is not synthesising.
@@
 
-\subsection{Decidability of Bidirectional Type Synthesis}
+\subsection{Decidability of Bidirectional Type Synthesis and Checking}
 
+\LT{What induction principle are we using?}
+
+\begin{theorem}[Decidability of Bidirectional Typing Synthesis] \label{thm:bidirectional-type-synthesis-checking}
+  Let $(\Sigma, \Omega)$ be a mode-correct bidirectional type system.
+  For every context $\Gamma$ and raw term $\isTerm{t}$ in mode $d$, 
+  \begin{enumerate}
+    \item if $\isTerm{t}$ is synthesising, then it is decidable whether there is a type~$A$ and a derivation of
+      \[
+        \Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn A.
+      \]
+    \item if $\isTerm{t}$ is checking, then for every type $A$ it is decidable whether there is a derivation of
+      \[
+        \Gamma |-_{\Sigma, \Omega} \isTerm{t} \chk A.
+      \]
+  \end{enumerate}
+\end{theorem}
 \begin{lemma}\label{lem:args-induction}
@@
       \[
-        \Gamma, \simsub{\Delta_i}{\ext{\rho}} \vdash_{\Sigma, \Omega} t_i \colon \simsub{A_i}{\ext{\rho}}^{d_i}
+        \Gamma, \simsub{\Delta_i}{\ext{\rho}} |-_{\Sigma, \Omega} t_i \colon \simsub{A_i}{\ext{\rho}}^{d_i}
       \]
 
-    \item or there is no extension $\sigma$ such that all $\Gamma, \simsub{\Delta_i}{\sigma} \vdash_{\Sigma, \Omega} t_i \colon \simsub{A_i}{\sigma}^{d_i}$ has a typing derivation. 
+    \item or there is no extension $\sigma$ of $\rho$ such that all $\Gamma, \simsub{\Delta_i}{\sigma} |-_{\Sigma, \Omega} t_i \colon \simsub{A_i}{\sigma}^{d_i}$ has a typing derivation. 
   \end{enumerate}
@@
 \begin{proof}[Proof of {\Cref{thm:bidirectional-type-synthesis}}]
-  We prove this statement by induction on the derivation of $\vdash_{\Sigma, \Omega} \isTerm{t}^\isDir{d}$ (instead of $t$ alone).
+  We prove this statement by induction on the derivation of $|-_{\Sigma, \Omega} \isTerm{t}^{d}$ (instead of $t$ alone).
   The first two cases (var) and (anno) are straightforward and have nothing to do with mode-correctness directly.
-  The third case (sub) invokes the uniqueness of synthesised types to refute the case that $\Gamma \vdash t \syn B$ but $A \neq B$ for a given type $A$ to check.
+  The third case (sub) invokes the uniqueness of synthesised types to refute the case that $\Gamma |- t \syn B$ but $A \neq B$ for a given type $A$ to check.
   The last case (op) is the interesting part of this theorem, as we need \Cref{lem:args-induction}.
   \begin{description}
-    \item[(var)] If $t$ is a variable $x$ in the context $\Gamma$, then there exists a type $A$ such that $\Gamma \vdash_{\Sigma, \Omega} x \syn A$.
-      Otherwise, if $x$ is not in $\Gamma$ but $\Gamma \vdash_{\Sigma, \Omega} x \syn A$ is derivable, then by inversion $x$ must be in $\Gamma$, a contradiction.
-    \item[(anno)] For a raw term of the form $(t : A)$ in synthesising mode, it is decidable that $\Gamma \vdash t \chk A$ is derivable or not by induction hypothesis.
+    \item[\SynRule{Var}:] If $t$ is a variable $x$ in the context $\Gamma$, then there exists a type $A$ such that $\Gamma |-_{\Sigma, \Omega} x \syn A$.
+      Otherwise, if $x$ is not in $\Gamma$ but $\Gamma |-_{\Sigma, \Omega} x \syn A$ is derivable, then by inversion $x$ must be in $\Gamma$, a contradiction.
+    \item[\SynRule{Anno}:] For a raw term of the form $(t : A)$ in synthesising mode, it is decidable that $\Gamma |- t \chk A$ is derivable or not by induction hypothesis.
       \begin{enumerate}
-        \item If $\Gamma \vdash t \chk A$ is derivable, then $\Gamma \vdash (t : A) \syn A$ is derivable.
-        \item If $\Gamma \vdash t \chk A$ is not derivable but $\Gamma \vdash (t : A) \syn$ is derivable, then by inversion $\Gamma \vdash t \chk A$ is derivable, thus a contradiction.
+        \item If $\Gamma |- t \chk A$ is derivable, then $\Gamma |- (t : A) \syn A$ is derivable.
+        \item If $\Gamma |- t \chk A$ is not derivable but $\Gamma |- (t : A) \syn$ is derivable, then by inversion $\Gamma |- t \chk A$ is derivable, thus a contradiction.
       \end{enumerate}
 
-    \item[(sub)] If $t$ is of the form $t\subsum$ in checking mode, then similarly by induction it is decidable that $\Gamma \vdash t \syn B$ for some $B$:
+    \item[\ChkRule{Sub}:] If $t$ is in checking mode because of the subsumption rule, then by induction hypothesis it is decidable whether $\Gamma |- t \syn B$ for some $B$:
       \begin{enumerate}
-        \item If $\Gamma \vdash t \syn B$ is derivable for some $B$, then we have to consider any type $A$ and it is decidable if $A$ is equal to $B$:
+        \item If $\Gamma |- t \syn B$ is derivable for some $B$, then we have to consider any type $A$ and it is decidable if $A$ is equal to $B$:
           \begin{enumerate}
-            \item if $A = B$ then we have arrived a derivation of $\Gamma \vdash t\subsum \chk A$.
-            \item if $A \neq B$ but $\Gamma \vdash t \subsum \chk A$ is derivable, then by inversion $\Gamma \vdash t \syn A$.
+            \item if $A = B$ then we have arrived a derivation of $\Gamma |- t\subsum \chk A$.
+            \item if $A \neq B$ but $\Gamma |- t \subsum \chk A$ is derivable, then by inversion $\Gamma |- t \syn A$.
               However, by \Cref{thm:unique-syn}, synthesised types $A$ and $B$ must be equal, so it is a contradiction.
           \end{enumerate}
-        \item if $\Gamma \vdash t \syn C$ is not derivable for any type $C$ but $\Gamma \vdash t\subsum \chk A$ is derivable, then by inversion $\Gamma \vdash t \syn B$ for some $B = A$, thus a contradiction.
+        \item if $\Gamma |- t \syn C$ is not derivable for any type $C$ but $\Gamma |- t\subsum \chk A$ is derivable, then by inversion $\Gamma |- t \syn B$ for some $B = A$, thus a contradiction.
       \end{enumerate}
@@
 %      \[
-%        \Gamma \vdash_{\Sigma, \Omega} \isTerm{\tmOp_{o}(\vec{x}_1.t_1,\ldots, \vec{x}_n.t_n)} : \simsub{A}{\rho}^\isDir{d}.
+%        \Gamma |-_{\Sigma, \Omega} \isTerm{\tmOp_{o}(\vec{x}_1.t_1,\ldots, \vec{x}_n.t_n)} : \simsub{A}{\rho}^{d}.
 %      \]
-    \item[(op)]
+    \item[\Rule{Op}:]
       Suppose that $t$ is of them $\tmOp_o(\vec{x}_1.t_1; \ldots; \vec{x}_n.t_n)$ for some mode-correct operation $o$ in $\Omega$.

### 2023-06-14 tex/sec4-bidirectionalisation.tex
[ tex ] Fix the colour of modes; use the verb form of decidable.
For example, instead of saying “for every type A it is decidable whether …”, we can say “every type A decides whether …”

@@
 \begin{proposition}[Decidability for Bidirectionalisation] \label{thm:bidirectionalisation}
-  For every raw term $t$ and a mode $d$, it is decidable whether $t$ is in mode $d$.
+  It is decidable whether a raw term~$t$ is in mode~$d$ for any $t$ and $d$.
 \end{proposition}

### 2023-06-14 tex/sec5-synthesis.tex
[ tex ] Fix the colour of modes; use the verb form of decidable.
For example, instead of saying “for every type A it is decidable whether …”, we can say “every type A decides whether …”

@@
 \begin{theorem}[Uniqueness of the Synthesised Types]\label{thm:unique-syn}
-  Let $(\Sigma, \Omega)$ be a bidirectional type system where $\Omega$ is mode-correct.
+  Let $(\Sigma, \Omega)$ be a mode-correct bidirectional type system.
   Then, for every raw term $t$ in synthesising mode and two derivations of 
@@
 
-\begin{theorem}[Decidability of Bidirectional Typing Synthesis] \label{thm:bidirectional-type-synthesis}
+\begin{theorem}[Decidability of Bidirectional Type Synthesis] \label{thm:bidirectional-type-synthesis}
   Let $(\Sigma, \Omega)$ be a mode-correct bidirectional type system.
-  For every context $\Gamma$ and raw term $\isTerm{t}$ in synthesising mode, it is decidable whether
+  Then, every context $\Gamma$ and raw term $\isTerm{t}$ in synthesising mode decide whether
   \[
@@
       \]
-    \item if $\isTerm{t}$ is checking, then for every type $A$ it is decidable whether there is a derivation of
+    \item if $\isTerm{t}$ is checking, then every type $A$ decides whether there is a derivation of
       \[
@@
   \[
-    \mathit{as} = \left([\Delta_i] A_{i}^{d_i}\right)_{i = 1}^{n}
+    \mathit{as} = \left([\Delta_i] A_{i}^{\dir{d_i}}\right)_{i = 1}^{n}
   \]
@@
       \[
-        \Gamma, \simsub{\Delta_i}{\ext{\rho}} |-_{\Sigma, \Omega} t_i \colon \simsub{A_i}{\ext{\rho}}^{d_i}
+        \Gamma, \simsub{\Delta_i}{\ext{\rho}} |-_{\Sigma, \Omega} \isTerm{t_i} \colon \simsub{A_i}{\ext{\rho}}^{\dir{d_i}}
       \]
 
-    \item or there is no extension $\sigma$ of $\rho$ such that all $\Gamma, \simsub{\Delta_i}{\sigma} |-_{\Sigma, \Omega} t_i \colon \simsub{A_i}{\sigma}^{d_i}$ has a typing derivation. 
+    \item or there is no extension $\sigma$ of $\rho$ such that all $\Gamma, \simsub{\Delta_i}{\sigma} |-_{\Sigma, \Omega} \isTerm{t_i} \colon \simsub{A_i}{\sigma}^{\dir{d_i}}$ has a typing derivation.
   \end{enumerate}
@@
   We prove this statement by induction on the derivation of $|-_{\Sigma, \Omega} \isTerm{t}^{d}$ (instead of $t$ alone).
-  The first two cases (var) and (anno) are straightforward and have nothing to do with mode-correctness directly.
-  The third case (sub) invokes the uniqueness of synthesised types to refute the case that $\Gamma |- t \syn B$ but $A \neq B$ for a given type $A$ to check.
-  The last case (op) is the interesting part of this theorem, as we need \Cref{lem:args-induction}.
+  The first two cases \SynRule{Var} and \SynRule{Anno} are straightforward and have nothing to do with mode-correctness directly.
+  The third case \ChkRule{Sub} invokes the uniqueness of synthesised types to refute the case that $\Gamma |- t \syn B$ but $A \neq B$ for a given type $A$ to check.
+  The last case \Rule{Op} is the interesting part of this theorem.
   \begin{description}

### 2023-06-14 tex/sec3-defs.tex
[ tex ] Fix the colour of modes; use the verb form of decidable.
For example, instead of saying “for every type A it is decidable whether …”, we can say “every type A decides whether …”
[ tex ] Fix typos and remove old requirements

@@
 
-  For brevity, we write $o \colon \Xi \rhd [\Delta_1]A_{1}^{d_1}, \ldots, [\Delta_{n}] A^{d_n}_{n} \to A^{d}$ to indicate an operation with its arity. 
+  For brevity, we write $o \colon \Xi \rhd [\Delta_1]A_{1}^{\dir{d_1}}, \ldots, [\Delta_{n}] A^{\dir{d_n}}_{n} \to A^{\dir{d}}$ to indicate an operation with its arity. 
 \end{definition}
@@
     {|-_{\Sigma, \Omega} \isTerm{\tmOp_o(\vec{x}_1.\, t_1; \ldots;\vec{x}_n.\, t_n)}^\dir{d}}\,\Rule{Op}
-    \and \text{for $o \colon \Xi \rhd [\Delta_1]A_{1}^\dir{d_1}, \ldots, [\Delta_{n}] A_{n}^\dir{d_n} \to A^\dir{d}$ in $\Omega$}
+    \and \text{for $o \colon \Xi \rhd [\Delta_1]A_{1}^{\dir{d_1}}, \ldots, [\Delta_{n}] A_{n}^{\dir{d_n}} \to A^{\dir{d}}$ in $\Omega$}
   \end{mathpar}
-  \caption{Preprocessed Raw terms}
+  \caption{Preprocessed raw terms}
 \end{figure}
@@
   \small
-  \judgbox{\Gamma |-_{\Sigma, \Omega} \isTerm{t} :^\dir{d} A}{A raw term $t$ in mode $d$ has a type $A$ under $\Gamma$}
+  \judgbox{\Gamma |-_{\Sigma, \Omega} \isTerm{t} :^\dir{d} A}{A raw term $t$ has a type $A$ (without type variables) under $\Gamma$}
   \begin{mathpar}
@@
     {\Gamma |-_{\Sigma, \Omega} \isTerm{\tmOp_o(\vec{x}_1 .\, t_{1}; \ldots; \vec{x}_{n}.\, t_{n})} :^{\dir{d}} \simsub{A}{\rho}} \,\mathsf{Op}
-    \and \text{for $o \colon \Xi \rhd [\Delta_1]A_{1}^\dir{d_1}, \ldots, [\Delta_{n}] A_{n}^\dir{d_n} \to A^\dir{d}$ in $\Omega$}
+    \and \text{for $o \colon \Xi \rhd [\Delta_1]A_{1}^{\dir{d_1}}, \ldots, [\Delta_{n}] A_{n}^{\dir{d_n}} \to A^\dir{d}$ in $\Omega$}
   \end{mathpar}

### 2023-06-14 tex/sec5-synthesis.tex
[ tex ] Fix typos and remove old requirements

@@
 
-\begin{theorem}[Decidability of Bidirectional Typing Synthesis] \label{thm:bidirectional-type-synthesis-checking}
+\begin{theorem}[Decidability of Bidirectional Type Synthesis and Checking] \label{thm:bidirectional-type-synthesis-checking}
   Let $(\Sigma, \Omega)$ be a mode-correct bidirectional type system.

### 2023-06-15 tex/sec5-synthesis.tex
[ tex ] Minor

@@
 
-\LT{What induction principle are we using?}
-
 \begin{theorem}[Decidability of Bidirectional Type Synthesis and Checking] \label{thm:bidirectional-type-synthesis-checking}
@@
   \begin{description}
-    \item[\SynRule{Var}:] If $t$ is a variable $x$ in the context $\Gamma$, then there exists a type $A$ such that $\Gamma |-_{\Sigma, \Omega} x \syn A$.
-      Otherwise, if $x$ is not in $\Gamma$ but $\Gamma |-_{\Sigma, \Omega} x \syn A$ is derivable, then by inversion $x$ must be in $\Gamma$, a contradiction.
+    \item[\SynRule{Var}:] If $t$ is a variable $x$ in the context $\Gamma$, then there exists a type $A$ such that $\Gamma |- x \syn A$.
+      Otherwise, if $x$ is not in $\Gamma$ but $\Gamma |- x \syn A$ is derivable, then by inversion $x$ must be in $\Gamma$, a contradiction.
     \item[\SynRule{Anno}:] For a raw term of the form $(t : A)$ in synthesising mode, it is decidable that $\Gamma |- t \chk A$ is derivable or not by induction hypothesis.
@@
           \begin{enumerate}
-            \item if $A = B$ then we have arrived a derivation of $\Gamma |- t\subsum \chk A$.
-            \item if $A \neq B$ but $\Gamma |- t \subsum \chk A$ is derivable, then by inversion $\Gamma |- t \syn A$.
+            \item if $A = B$ then we have arrived a derivation of $\Gamma |- t \chk A$.
+            \item if $A \neq B$ but $\Gamma |- t \chk A$ is derivable, then by inversion $\Gamma |- t \syn A$.
               However, by \Cref{thm:unique-syn}, synthesised types $A$ and $B$ must be equal, so it is a contradiction.
           \end{enumerate}
-        \item if $\Gamma |- t \syn C$ is not derivable for any type $C$ but $\Gamma |- t\subsum \chk A$ is derivable, then by inversion $\Gamma |- t \syn B$ for some $B = A$, thus a contradiction.
+        \item if $\Gamma |- t \syn C$ is not derivable for any type $C$ but $\Gamma |- t \chk A$ is derivable, then by inversion $\Gamma |- t \syn B$ for some $B = A$, thus a contradiction.
       \end{enumerate}
@@
       Suppose that $t$ is of them $\tmOp_o(\vec{x}_1.t_1; \ldots; \vec{x}_n.t_n)$ for some mode-correct operation $o$ in $\Omega$.
-      No matter $t$ is checking or synthesising, we need to build derivations for its arguments incrementally and maintain a set of unifiers, i.e.\ a partial substitution $\rho\colon \PSub(\Xi, \emptyset)$ incrementally which is defined nowhere initially if $t$ is synthesising or on variables of the target $A$ if $t$ is checking.
+      No matter $t$ is checking or synthesising, we need to build derivations for its arguments incrementally and maintain a set of unifiers, i.e.\ a partial substitution $\rho\colon \PSub{\Xi}{\emptyset}$ is incrementally defined which is initially defined nowhere if $t$ is synthesising or on variables of the target $A$ if $t$ is checking.
 

### 2023-06-15 tex/sec2-outline.tex
[ tex ] Add the pen-and-paper proof of bidirectional type synthesis
[ tex ] Add points to address; finalise the paper title
[ tex ] Added acmart files; separate compilation for TeXShop
[ tex ] Using the same color as by Dunfield et al.
[ tex ] Fix some figures; add stubs for typesetting rules

@@
-%! TEX root = BiSig.tex
+%!TEX root = BiSig.tex
 
-\section{Core Ideas with Bidirectional Simply Typed \texorpdfstring{$\lambda$}{lambda}-Calculus} \label{sec:key-ideas}
+\section{The Key Ideas} \label{sec:key-ideas}
 \begin{figure}
+  \small
+  \bgroup
+  \renewcommand{\arraystretch}{1.5}
+  \begin{tabular}{ r l }
+    $\boxed{\Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn A}$ & A raw term $t$ synthesises a type $A$ under $\Gamma$ \\
+    $\boxed{\Gamma |-_{\Sigma, \Omega} \isTerm{t} \chk A}$ &A raw term $t$ checks against a type $A$ under $\Gamma$
+  \end{tabular}
+  \egroup
   \centering
-  \small
   \begin{mathpar}
-    \boxed{\Gamma \vdash \isTerm{t} \Rightarrow A}
-    \quad\text{and}\quad \boxed{\Gamma \vdash \isTerm{t} \Leftarrow A}
-    \\
-    \inferrule{(x : A) \in \Gamma}{\Gamma \vdash \isTerm{x} \Rightarrow A}\;(\text{var}^\Rightarrow)
+    \inferrule{(x : A) \in \Gamma}{\Gamma \vdash \isTerm{x} \syn A}\,\SynRule{Var}
     \and
-    \inferrule{\Gamma \vdash \isTerm{t} \Leftarrow B \\ A = B}{\Gamma \vdash \isTerm{t : B}\Rightarrow A}\;(\text{anno})
+    \inferrule{\Gamma \vdash \isTerm{t} \chk B \\ A = B}{\Gamma \vdash \isTerm{t : B}\syn A}\,\SynRule{Anno}
     \and
-    \inferrule{\Gamma \vdash \isTerm{t} \Rightarrow B \\ A = B}{\Gamma \vdash \isTerm{t\subsum} \Leftarrow A}\;(\text{sub})
+    \inferrule{\Gamma \vdash \isTerm{t} \syn B \\ A = B}{\Gamma \vdash \isTerm{t} \chk A}\,\ChkRule{Sub}
     \and
-    \inferrule{\Gamma \vdash \isTerm{t} \Rightarrow A \bto B \\ \Gamma \vdash \isTerm{u} \Leftarrow A}{\Gamma \vdash \isTerm{t\;u} \Rightarrow B}\;(\text{app})
+    \inferrule{\Gamma \vdash \isTerm{t} \syn A \bto B \\ \Gamma \vdash \isTerm{u} \chk A}{\Gamma \vdash \isTerm{t\;u} \syn B}\,\SynRule{App}
     \and
-    \inferrule{\Gamma, \isTerm{x} : A \vdash \isTerm{t} \Leftarrow B}{\Gamma \vdash \isTerm{\lam{x}t} \Leftarrow A \bto B}\;(\text{abs})
+    \inferrule{\Gamma, \isTerm{x} : A \vdash \isTerm{t} \chk B}{\Gamma \vdash \isTerm{\lam{x}t} \chk A \bto B}\,\ChkRule{Abs}
   \end{mathpar}

### 2023-06-15 tex/sec3-defs.tex
[ tex ] Minor
[ tex ] Fix some figures; add stubs for typesetting rules

@@
 \begin{figure}
-  \begin{minipage}[b]{.55\textwidth}
+  \begin{minipage}[b]{.6\textwidth}
     \centering
     \small
-    \judgbox{\Theta|-_{\Sigma} A}{$A$ is well-formed with type variables in $\Theta$}
+    \judgbox{\Theta|-_{\Sigma} A}{$A$ is a well-formed type with type variables in $\Theta$}
     \begin{mathpar}
@@
     \end{mathpar}
-    \caption{Type formation rules}
+    \caption{Simple types}
     \label{fig:simple-type}
   \end{minipage}
-  \begin{minipage}[b]{.4\textwidth}
+  \begin{minipage}[b]{.35\textwidth}
     \centering
@@
     \end{mathpar}
-    \caption{Context formation rules}
+    \caption{Contexts}
   \label{fig:simple-context}
@@
     \inferrule{|-_{\Sigma, \Omega} \isTerm{t_1} \quad \cdots \quad |-_{\Sigma, \Omega} \isTerm{t_n}}
-    {|-_{\Sigma, \Omega} \isTerm{\tmOp_o(t_1, \ldots, t_n)}}\,\Rule{Op}_o
+    {|-_{\Sigma, \Omega} \isTerm{\tmOp_o(\vec{x}_1.t_1; \ldots; \vec{x}_n.t_n)}}\,\Rule{Op}
     \and \text{for $o \colon \Xi \rhd [\Delta_1]A_{1}, \ldots, [\Delta_{n}] A_{n} \to A$ in $\Omega$}
   \end{mathpar}
-  \caption{Raw term formation rules}
+  \caption{Raw terms}
 \end{figure}
@@
   \end{mathpar}
-  \caption{Preprocessed raw terms}
+  \caption{Preprocessed raw terms, I}
 \end{figure}
@@
   \small
-  \judgbox{\Gamma |-_{\Sigma, \Omega} \isTerm{t} :^\dir{d} A}{A raw term $t$ has a type $A$ (without type variables) under $\Gamma$}
+  \judgbox{\Gamma |-_{\Sigma, \Omega} \isTerm{t} :^\dir{d} A}{A raw term $t$ is of type $A$ (without type variables) in mode $d$ under $\Gamma$}
   \begin{mathpar}
@@
   \end{mathpar}
-  \begin{minipage}{.15\linewidth}
-  Abbreviations: 
-  \end{minipage}
-  \begin{minipage}{.8\linewidth}
-    \begin{flushleft}
-    $\boxed{\Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn A} \defeq \boxed{\Gamma |-_{\Sigma, \Omega} \isTerm{t} :^{\syn} A}$ \quad A raw term $t$ synthesises a type $A$ under $\Gamma$ \\
-    $\boxed{\Gamma |-_{\Sigma, \Omega} \isTerm{t} \chk A} \defeq \boxed{\Gamma |-_{\Sigma, \Omega} \isTerm{t} :^{\chk} A}$ \quad A raw term $t$ checks against a type $A$ under $\Gamma$
-    \end{flushleft}
-  \end{minipage}
+Abbreviations: $\boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \syn A} \defeq \boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A^{\syn}}$ and $\boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \chk A} \defeq \boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A^{\chk}}$
   \caption{Bidirectional typing rules}
@@
 
-\begin{example}
+\LT{What induction principle are we using?}
+\begin{theorem}[Mutual Induction Principles]
 
-\end{example}
+\end{theorem}

### 2023-06-15 tex/sec4-bidirectionalisation.tex
[ tex ] Fix some figures; add stubs for typesetting rules

@@
 \subsection{Bidirectionalisation}
+
 \begin{proposition}[Decidability for Bidirectionalisation] \label{thm:bidirectionalisation}
@@
 \subsection{Annotatability is not Completeness}
+
+\begin{figure}
+  \centering
+  \small
+  \bgroup
+  \renewcommand{\arraystretch}{1.5}
+  \begin{tabular}{ r l }
+    $\boxed{|-_{\Sigma, \Omega} \isTerm{t}\;\dir{d}}$ & $t$ is a raw term possibly with missing annotation or redundant casts in mode $d$ \\
+    $\boxed{|-_{\Sigma, \Omega} \isTerm{t}\;\dir{d}}$ & $t$ is a raw term without missing annotation \\
+    $\boxed{|-_{\Sigma, \Omega} \isTerm{t}\;\dir{d}}$ & $t$ is a raw term without redundant casts \\
+  \end{tabular}
+  \egroup
+  \LT{Design a notation for \textsf{Pre?} so that a raw term without missing annotation or redundant casts is exactly the intersection of these two cases.}
+  \begin{mathpar}
+    \inferrule{x : \Identifier}{|-_{\Sigma, \Omega} \isTerm{x}^{\syn}}\,\SynRule{Var}
+    \and
+    \inferrule{|-_{\Sigma, \Omega}\isTerm{t}^{\chk}}{|-_{\Sigma, \Omega} (\isTerm{t \annote A})^{\syn}}\,\SynRule{Anno}
+    \and
+    \inferrule{|-_{\Sigma, \Omega}\isTerm{t}^{\chk}}{|-_{\Sigma, \Omega} \isTerm{t}^{\syn}}\,\SynRule{Anno?}
+    \and
+    \inferrule{|-_{\Sigma, \Omega} \isTerm{t}^{\syn}}{|-_{\Sigma, \Omega} \isTerm{t}^{\chk}}\,\ChkRule{Sub}
+    \and
+    \inferrule{|-_{\Sigma, \Omega} \isTerm{t_1}^\dir{d_1} \quad \cdots \quad |-_{\Sigma, \Omega} \isTerm{t_n}^\dir{d_n}}
+    {|-_{\Sigma, \Omega} \isTerm{\tmOp_o(\vec{x}_1.\, t_1; \ldots;\vec{x}_n.\, t_n)}^\dir{d}}\,\Rule{Op}
+  \end{mathpar}
+  \caption{Preprocessed raw terms, II}
+  \label{fig:raw-with-missing-annotations}
+\end{figure}
+
+\begin{figure}
+  \centering\small
+  \judgbox{\isTerm{t} \sqsupseteq \isTerm{u}}{A raw term $t$ is more annotated than $u$ (for some bidirectional type system $(\Sigma, \Omega)$)}
+  \begin{mathpar}
+    \inferrule{\isTerm{t} \sqsupseteq \isTerm{u}}{(\isTerm{t} : A) \sqsupseteq \isTerm{u}}\;\Rule{More}
+    \and
+    \inferrule{\vphantom{x : \Identifier}}{\isTerm{x} \sqsupseteq \isTerm{x}}
+    \and
+    \inferrule{\isTerm{t} \sqsupseteq \isTerm{u}}{(\isTerm{t} : A) \sqsupseteq (\isTerm{u} : A)}
+    \and
+    \inferrule{\isTerm{t_1} \sqsupseteq \isTerm{u_1} \quad \cdots \quad \isTerm{t_n} \sqsupseteq{u_n}}
+    {\isTerm{\tmOp_o(\vec{x}_1. t_1; \ldots; \vec{x}_n. t_n)} \sqsupseteq \isTerm{\tmOp_o(\vec{x}_1.u_1; \ldots; \vec{x}_n.u_n)}}
+  \end{mathpar}
+
+  \caption{Annotation ordering between raw terms}
+  \label{fig:annotation-order}
+\end{figure}

### 2023-06-15 tex/BiSig.tex
[ tex ] Comment out some unused packages
[ tex ] Change title
[ tex ] Add points to address; finalise the paper title
[ tex ] Using the same color as by Dunfield et al.
[ tex ] Minor
[ tex ] Draft the abstract

@@
 \documentclass[acmsmall,screen]{acmart}
+
+\ifPDFTeX
 \usepackage[T1]{fontenc}
@@
 \usepackage[british]{babel}
+\else
+\usepackage{polyglossia}
+\setdefaultlanguage{british}
+\fi
 
-\usepackage{bm,braket,mathtools,mathpartir,bussproofs}
-\EnableBpAbbreviations
-\usepackage[all,cmtip]{xy}
+\usepackage{bm}
+\usepackage{mathtools}
+\usepackage{mathpartir}
+\input{mathlig}
+%\usepackage{xfrac}
+\usepackage{xspace}
+%\usepackage{hyperref}
+\usepackage[capitalise,noabbrev]{cleveref}
 
-\usepackage{subcaption,xfrac,xspace,xcolor,xspace}
-\usepackage{hyperref}
-\usepackage[capitalise,noabbrev]{cleveref}
-\usepackage[inline]{enumitem} % for environment enumerate*
-\setlist[enumerate]{mode=unboxed}
+%\usepackage[inline]{enumitem} % for environment enumerate*
+%\setlist[enumerate]{mode=unboxed}
 
@@
 \usepackage[color=yellow,textsize=small]{todonotes}
-\newcommand{\LT}[1]{\todo[author=LT,inline,color=green!40,caption={}]{{#1}}}
-\newcommand{\Josh}[1]{\todo[author=Josh,inline,color=green!40,caption={}]{{#1}}}
 
@@
 \citestyle{acmauthoryear}
-
 \AtEndPreamble{%
@@
 \input{macros}
+
+\newcommand{\LT}[1]{\todo[author=LT,inline,color=green!40,caption={}]{{#1}}}
+\newcommand{\Josh}[1]{\todo[author=Josh,inline,color=green!40,caption={}]{{#1}}}
 
@@
 
-\title{A Constructive Metatheory of Bidirectional Type Synthesis for Simple Types}
+\title{A Theory of Bidirectional Type Synthesis for Simple Types}
 
 \begin{abstract}
-  This paper proposes a formal theory of bidirectional typing.
+%  Type checking (and inference) can be seen as the transition from parsed abstract syntax trees to well-typed programs.
+%  Many techniques and principles have been developed to type check advanced type theories and systems.
+%  However, as theory and practice of parsing are able to address common grammars uniformly and parser generators have existed, type checkers by now are still designed and implemented on a case-by-case basis, not to mention a general theory of type checking and a robust `type checker generator' that can handle languages uniformly.
+%
+%  This paper is a step towards a \emph{general theory} of type checking and inference.
+%  Inspired by recent language-formalisation frameworks, we theroise bidirectional typing---a design discipline of type synthesis mutually defining type checking and inference---for \emph{arbitrary} bidirectional type systems specified by a bidirectional binding signature and also formalise the theory in vanilla \Agda with Axiom K.
+%  Prominently, our theorems including decidability of type synthesis are sound and complete type synthesisers which are correct by construction.
+  Type checking and inference serve as the passage from parsed abstract syntax trees to well-typed programs.
+  While significant progress has been made in developing techniques for type checking advanced type systems on a case-by-case basis, the lack of a general theory and a robust `type checker generator' poses a challenge to achieving uniformity across diverse programming languages.
+
+  Motivated by this challenge, this paper contributes to the development of a general and practical theory of type checking and inference.
+  We present a \emph{theory} of bidirectional typing, which had only been a \emph{design discipline}, for synthesising types to handle \emph{syntax-directed} bidirectional type systems of simple types that can be specified by a bidirectional binding signature.
+  Here, \emph{type synthesis} refers to the mutual use of type checking and inference to synthesise a type from programs possibly with type annotations.
+
+  To establish the correctness and effectiveness of our proposal, we formalize our theory in vanilla \Agda.
+  We prove logical decidability of type synthesis, subsuming soundness and completeness of type synthesis algorithm.
+  By Curry-Howard correspondence, our theorems are actually correct-by-construction synthesisers.
+  We also discuss another essential characteristic of bidirectional typing called \emph{annotability} and propose a related but simpler and practical notion called \emph{bidirectionalisation}.
+  Combining with decidability of type synthesis, we derive the trichotomy of raw terms by type synthesis.
+  Put differently, the trichotomy is a type synthesiser which distinguishes programs with missing type annotations from programs that cannot be checked.
+
+  In summary, this paper represents a step towards a general and practical theory of type checking and inference.
+  By formalizing our theory of bidirectional type synthesise in \Agda, we provide a framework capable of handling bidirectional type systems of simple types.
+  Our work addresses the problem with existing methodologies, setting the stage for further advances towards a general theory of type synthesis.
 \end{abstract}
+
+\keywords{\Agda, abstract syntax, bidirectional typing, type synthesis, type checking, type inference, language formalisation}
 

### 2023-06-15 tex/sec1-intro.tex
[ tex ] Added acmart files; separate compilation for TeXShop
[ tex ] Fix bib (Jones -> Peyton-Jones)
[ tex ] Draft the abstract

@@
 
+Points to elaborate:
 \begin{enumerate}
-  \item Type checking as a transition from (parsed) abstract syntax trees to well-typed terms.
-  \item Bidirectional typing is the only solution that works that scales well.
+  \item Type inference and checking as a transition from (parsed) abstract syntax trees to well-typed terms.
+  \item Bidirectional typing is by now the only solution that scales well.
   \item Not yet a general theory but by now a design principle.
-    No `type checker generator' as each type checker is still designed and implemented on a case-by-case basis, but can be derived from a syntax-directed type theory `easily'.
-  \item The existing work of bidirectional typing does not take account of missing annotations carefully enough. 
-  \item A meta-problem: existing work is based on (or at least left unspecified) classical logic.
-    One has to state soundness of completeness of their type checking algorithm in a less straightforward way.
+    No `type checker generator' as each type checker is implemented on a case-by-case basis, but an algorithm can usually be read off from syntax-directed typing rules.
+    A hint to a general pattern in syntax-directedness.
+  \item The existing work of bidirectional typing does not take account of practical aspects such as missing annotation.
+    Lightly or rarely discussed.
+  \item A meta-problem: existing work is based on classical logic (or left unspecified at least).
+    Soundness of completeness of a type checking algorithm in a less straightforward way.
   \item A review of what has been done towards this line of research.
@@
 \cite{Pierce2000}
+\cite{Peyton-Jones2007}
 \cite{Dunfield2021}
-\cite{Jones2007}
 
@@
   \item Key ideas (\Cref{sec:key-ideas})
-  \item Definitions for simple type theories and for bidirectional type systems (signature erasure and mode annotation) (\Cref{sec:defs})
+  \item Definitions for bidirectional type systems (bidirectional binding signature, bidirectional type systems, signature erasure and mode annotation) (\Cref{sec:defs})
   \item Soundness, completeness, bidirectionalisation, and annotatability (\Cref{sec:annotatability})

### 2023-06-15 tex/BiSig.tex
[ tex ] Emphaise generality and practicality of our theory

@@
   Motivated by this challenge, this paper contributes to the development of a general and practical theory of type checking and inference.
-  We present a \emph{theory} of bidirectional typing, which had only been a \emph{design discipline}, for synthesising types to handle \emph{syntax-directed} bidirectional type systems of simple types that can be specified by a bidirectional binding signature.
-  Here, \emph{type synthesis} refers to the mutual use of type checking and inference to synthesise a type from programs possibly with type annotations.
+  First, we present a \emph{general} theory of bidirectional typing, which had only been a design discipline, to synthesise types for \emph{syntax-directed} bidirectional type systems of \emph{simple types}.
+  Here, \emph{(bidirectional) type synthesis} refers to the mutual use of type checking and inference to synthesise the type of a programs possibly with type annotations.
+  Second, our theory is \emph{practical} not only because we formalize our theory in \Agda but also we have formulated our results from the practical aspect to provide useful computational meaning.
+  Specifically, we prove \emph{logical decidability} of type synthesis, subsuming soundness, completeness, and decidability of type synthesis algorithm; by Curry-Howard correspondence, our proof is a correct-by-construction synthesiser.
+  Also we discuss an essential characteristic of bidirectional typing called \emph{annotability} and propose a related but simpler and practical replacement called \emph{bidirectionalisation}.
+  Combining with decidability of type synthesis, we derive the trichotomy of raw terms whose proof is a type synthesiser which distinguishes raw terms with missing type annotations from those that cannot be checked.
 
-  To establish the correctness and effectiveness of our proposal, we formalize our theory in vanilla \Agda.
-  We prove logical decidability of type synthesis, subsuming soundness and completeness of type synthesis algorithm.
-  By Curry-Howard correspondence, our theorems are actually correct-by-construction synthesisers.
-  We also discuss another essential characteristic of bidirectional typing called \emph{annotability} and propose a related but simpler and practical notion called \emph{bidirectionalisation}.
-  Combining with decidability of type synthesis, we derive the trichotomy of raw terms by type synthesis.
-  Put differently, the trichotomy is a type synthesiser which distinguishes programs with missing type annotations from programs that cannot be checked.
-
-  In summary, this paper represents a step towards a general and practical theory of type checking and inference.
+  In short, this paper is a first step towards a general and practical theory of type checking and inference.
   By formalizing our theory of bidirectional type synthesise in \Agda, we provide a framework capable of handling bidirectional type systems of simple types.

### 2023-06-15 tex/BiSig.tex
[ tex ] Refined abstract

--- 2023-06-13 tex/BiSig.tex
@@
 
-\title{A Constructive Metatheory of Bidirectional Type Synthesis for Simple Types}
+\title{A Theory of Bidirectional Type Synthesis for Simple Types}
 
 \begin{abstract}
-  This paper proposes a formal theory of bidirectional typing.
+  Type checking and inference serve as the passage from parsed abstract syntax trees to well-typed programs.
+  While significant progress has been made in developing techniques for typechecking advanced type systems on a case-by-case basis, the lack of a general theory poses a difficulty of achieving a robust `type checker generator' for diverse programming languages.
+
+  Motivated by this problem, this paper develops a \emph{general} and \emph{practical} theory of bidirectional type synthesis where type synthesis refers to the mutual use of type checking and inference for synthesising the type of a programs possibly with type annotations, based on bidirectional typing.
+%  Bidirectional typing had only been a set of principles and characteristics to design typing rules so that a type synthesis algorithm can be `read off' easily
+  First, our theory is general as it deals with any simple type theory and its bidirectional type system that can be specified by some bidirectional binding signature, similar to the role of grammar in parsing.
+  Second, our theory is practical not only because our results are formalised in Martin-L\"{o}f's type theory, \Agda, but also properties have been formulated from the practical aspect in order to compute by Curry-Howard correspondence.
+  Specifically, we prove logical decidability of type synthesis for any mode-correct signature, subsuming soundness, completeness, and decidability of type synthesis algorithm.
+  For a given (mode-correct) signature, its proof is instantiated to a synthesiser specific to a simple type theory, i.e.\ a type synthesiser generator.
+  A simple notion called bidirectionalisation is proposed to refine an essential characteristic of bidirectional typing called annotability.
+  This notion leads to the trichotomy of raw terms whose proof models a practical type synthesiser that handles ill-formed inputs, including raw terms with missing type annotations.
+
+  In short, this paper is a first step towards a general and practical theory of type checking and inference.
+  By formalizing our theory of bidirectional type synthesise in \Agda, we provide a framework capable of handling bidirectional type systems of simple types uniformly.
+  Our work addresses the problem with existing methodologies, setting the stage for further advances towards a more general theory of type synthesis.
 \end{abstract}
+
+\keywords{\Agda, abstract syntax, bidirectional typing, type synthesis, type checking, type inference, language formalisation}
 

### 2023-06-15 tex/BiSig.tex
[ tex ] Use "constructive" intead of "practical" (to be more specific and humble)

@@
 
-  Motivated by this problem, this paper develops a \emph{general} and \emph{practical} theory of bidirectional type synthesis where type synthesis refers to the mutual use of type checking and inference for synthesising the type of a programs possibly with type annotations, based on bidirectional typing.
+  Motivated by this problem, this paper develops a \emph{general} and \emph{constructive} theory of bidirectional type synthesis where bidirectional type synthesis refers to the mutual use of type checking and inference based on bidirectional typing rules for synthesising the type of a programs possibly with type annotations.
 %  Bidirectional typing had only been a set of principles and characteristics to design typing rules so that a type synthesis algorithm can be `read off' easily
   First, our theory is general as it deals with any simple type theory and its bidirectional type system that can be specified by some bidirectional binding signature, similar to the role of grammar in parsing.
-  Second, our theory is practical not only because our results are formalised in Martin-L\"{o}f's type theory, \Agda, but also properties have been formulated from the practical aspect in order to compute by Curry-Howard correspondence.
+  Second, our theory is constructive not only because our results are formalised in Martin-L\"{o}f's type theory, \Agda, but also properties have been formulated from the practical aspect in order to compute by Curry-Howard correspondence.
   Specifically, we prove logical decidability of type synthesis for any mode-correct signature, subsuming soundness, completeness, and decidability of type synthesis algorithm.
@@
 
-  In short, this paper is a first step towards a general and practical theory of type checking and inference.
+  In short, this paper is a first step towards a general and constructive theory of type checking and inference.
   By formalizing our theory of bidirectional type synthesise in \Agda, we provide a framework capable of handling bidirectional type systems of simple types uniformly.

### 2023-06-15 tex/BiSig.tex
[ tex ] Add CCS concepts

@@
 
+\begin{CCSXML}
+<ccs2012>
+   <concept>
+       <concept_id>10011007.10011006.10011041</concept_id>
+       <concept_desc>Software and its engineering~Compilers</concept_desc>
+       <concept_significance>500</concept_significance>
+       </concept>
+   <concept>
+       <concept_id>10011007.10011006.10011039.10011040</concept_id>
+       <concept_desc>Software and its engineering~Syntax</concept_desc>
+       <concept_significance>500</concept_significance>
+       </concept>
+   <concept>
+       <concept_id>10011007.10011074.10011099.10011692</concept_id>
+       <concept_desc>Software and its engineering~Formal software verification</concept_desc>
+       <concept_significance>300</concept_significance>
+       </concept>
+   <concept>
+       <concept_id>10003752.10003790.10011740</concept_id>
+       <concept_desc>Theory of computation~Type theory</concept_desc>
+       <concept_significance>500</concept_significance>
+       </concept>
+ </ccs2012>
+\end{CCSXML}
+
+\ccsdesc[500]{Software and its engineering~Compilers}
+\ccsdesc[500]{Software and its engineering~Syntax}
+\ccsdesc[500]{Theory of computation~Type theory}
+\ccsdesc[300]{Software and its engineering~Formal software verification}
+
 \keywords{\Agda, abstract syntax, bidirectional typing, type synthesis, type checking, type inference, language formalisation}

### 2023-06-15 tex/BiSig.tex
[ tex ] Downplay the significance in abstract

@@
   First, our theory is general as it deals with any simple type theory and its bidirectional type system that can be specified by some bidirectional binding signature, similar to the role of grammar in parsing.
-  Second, our theory is constructive not only because our results are formalised in Martin-L\"{o}f's type theory, \Agda, but also properties have been formulated from the practical aspect in order to compute by Curry-Howard correspondence.
+  Second, our theory is constructive not only because our results are formalised in Martin-L\"{o}f's type theory, \Agda, but also properties have been formulated from the practical aspect in order to compute.
   Specifically, we prove logical decidability of type synthesis for any mode-correct signature, subsuming soundness, completeness, and decidability of type synthesis algorithm.
-  For a given (mode-correct) signature, its proof is instantiated to a synthesiser specific to a simple type theory, i.e.\ a type synthesiser generator.
+  For a given (mode-correct) signature, its proof is instantiated to a synthesiser specific to the corresponding simple type theory, similar to a parser generator.
   A simple notion called bidirectionalisation is proposed to refine an essential characteristic of bidirectional typing called annotability.
@@
 
-  In short, this paper is a first step towards a general and constructive theory of type checking and inference.
+  In short, this paper is a small but first step towards a general theory of type checking and inference.
   By formalizing our theory of bidirectional type synthesise in \Agda, we provide a framework capable of handling bidirectional type systems of simple types uniformly.

### 2023-06-15 tex/BiSig.tex
[ tex ] Fix grammatical mistakes

@@
   First, our theory is general as it deals with any simple type theory and its bidirectional type system that can be specified by some bidirectional binding signature, similar to the role of grammar in parsing.
-  Second, our theory is constructive not only because our results are formalised in Martin-L\"{o}f's type theory, \Agda, but also properties have been formulated from the practical aspect in order to compute.
+  Second, our theory is constructive because properties are not only formalised in Martin-L\"{o}f's type theory, \Agda, but also formulated from the practical aspect in order to compute.
   Specifically, we prove logical decidability of type synthesis for any mode-correct signature, subsuming soundness, completeness, and decidability of type synthesis algorithm.

### 2023-06-16 tex/BiSig.tex
[ tex ] Shorten the abstract

@@
   Type checking and inference serve as the passage from parsed abstract syntax trees to well-typed programs.
-  While significant progress has been made in developing techniques for typechecking advanced type systems on a case-by-case basis, the lack of a general theory poses a difficulty of achieving a robust `type checker generator' for diverse programming languages.
+  While progress has been made in developing techniques for typechecking type systems on a case-by-case basis, the lack of a general theory poses a difficulty of achieving a `type checker generator' for diverse languages.
 
-  Motivated by this problem, this paper develops a \emph{general} and \emph{constructive} theory of bidirectional type synthesis where bidirectional type synthesis refers to the mutual use of type checking and inference based on bidirectional typing rules for synthesising the type of a programs possibly with type annotations.
-%  Bidirectional typing had only been a set of principles and characteristics to design typing rules so that a type synthesis algorithm can be `read off' easily
-  First, our theory is general as it deals with any simple type theory and its bidirectional type system that can be specified by some bidirectional binding signature, similar to the role of grammar in parsing.
-  Second, our theory is constructive because properties are not only formalised in Martin-L\"{o}f's type theory, \Agda, but also formulated from the practical aspect in order to compute.
-  Specifically, we prove logical decidability of type synthesis for any mode-correct signature, subsuming soundness, completeness, and decidability of type synthesis algorithm.
-  For a given (mode-correct) signature, its proof is instantiated to a synthesiser specific to the corresponding simple type theory, similar to a parser generator.
-  A simple notion called bidirectionalisation is proposed to refine an essential characteristic of bidirectional typing called annotability.
-  This notion leads to the trichotomy of raw terms whose proof models a practical type synthesiser that handles ill-formed inputs, including raw terms with missing type annotations.
-
-  In short, this paper is a small but first step towards a general theory of type checking and inference.
-  By formalizing our theory of bidirectional type synthesise in \Agda, we provide a framework capable of handling bidirectional type systems of simple types uniformly.
-  Our work addresses the problem with existing methodologies, setting the stage for further advances towards a more general theory of type synthesis.
+  Motivated by this issue, this paper develops a \emph{simple} but \emph{general} and \emph{constructive} theory of bidirectional type synthesis where bidirectional type synthesis refers to the mutual use of type checking and inference based on bidirectional typing for synthesising the type of a program with type annotations.
+  First, our theory is simple but general as it deals with any simple type theory and its bidirectional type system that can be specified by a signature, similar to the role of grammar in parsing.
+  Second, our theory is constructive because it is formalised in \Agda and formulated positively in order to compute.
+  Specifically, for any mode-correct signature, logical decidability of type synthesis is proved and its proof is instantiated to a synthesiser specific to the corresponding simple type theory, similar to a parser generator.
+  The notion of bidirectionalisation is introduced to refine annotability in bidirectional typing and it leads to the trichotomy of raw terms which models a practical type synthesiser that triggers an exception when its input has missing annotations.
+%  By formalizing our theory in \Agda,
 \end{abstract}
@@
 
-\keywords{\Agda, abstract syntax, bidirectional typing, type synthesis, type checking, type inference, language formalisation}
+\keywords{\Agda, abstract syntax, bidirectional typing, type synthesis, type checking, type inference, language formalisation, datatype-generic programming}
 

### 2023-06-16 tex/BiSig.tex
[ tex ] abstract

@@
   Type checking and inference serve as the passage from parsed abstract syntax trees to well-typed programs.
-  While progress has been made in developing techniques for typechecking type systems on a case-by-case basis, the lack of a general theory poses a difficulty of achieving a `type checker generator' for diverse languages.
+  While progress has been made in developing techniques for typechecking on a case-by-case basis, the lack of a general theory poses a difficulty of implementing a robust `type checker generator' for diverse languages.
 
-  Motivated by this issue, this paper develops a \emph{simple} but \emph{general} and \emph{constructive} theory of bidirectional type synthesis where bidirectional type synthesis refers to the mutual use of type checking and inference based on bidirectional typing for synthesising the type of a program with type annotations.
-  First, our theory is simple but general as it deals with any simple type theory and its bidirectional type system that can be specified by a signature, similar to the role of grammar in parsing.
-  Second, our theory is constructive because it is formalised in \Agda and formulated positively in order to compute.
-  Specifically, for any mode-correct signature, logical decidability of type synthesis is proved and its proof is instantiated to a synthesiser specific to the corresponding simple type theory, similar to a parser generator.
-  The notion of bidirectionalisation is introduced to refine annotability in bidirectional typing and it leads to the trichotomy of raw terms which models a practical type synthesiser that triggers an exception when its input has missing annotations.
-%  By formalizing our theory in \Agda,
+  Motivated by this issue, we develop a \emph{simple} but \emph{general} and \emph{constructive} theory of bidirectional type synthesis where bidirectional type synthesis refers to the mutual use of type checking and inference based on bidirectional typing rules for synthesising the type of a program with type annotations.
+  First, our theory is simple but general as it deals with any simple type theory and its bidirectional type system that can be specified by a signature, analogous to the role of grammar in parsing.
+  Second, our theory is constructive because it is not only formalised in a proof assistant based on constructive type theory but also formulated positively in order to compute.
+  Specifically, a type synthesiser for a simple type theory specified by some mode-correct signature is the proof of logical decidability of synthesis, analogous to a parser generator.
+  The notion of bidirectionalisation is introduced to refine annotability in bidirectional typing.
+  Type synthesis and bidirectionalisation lead to the trichotomy of raw terms which models a practical type synthesiser that throws an exception when its input does not have enough annotations potentially with a proof to locate missing annotations.
+  Hence, by formalizing our general theory of bidirectional type synthesis constructively, we provide a provably correct `type checker generator' for simple type theories, setting the stage for further advances towards a more general theory of type synthesis.
 \end{abstract}

### 2023-06-16 tex/BiSig.tex
[ tex ] Writing job distribution

@@
 
-\newcommand{\LT}[1]{\todo[author=LT,inline,color=green!40,caption={}]{{#1}}}
-\newcommand{\Josh}[1]{\todo[author=Josh,inline,color=green!40,caption={}]{{#1}}}
+\newcommand{\LT}[1]{\todo[author=LT,inline,color=orange!40,caption={}]{{#1}}}
+\newcommand{\Josh}[1]{\todo[author=Josh,inline,color=yellow!40,caption={}]{{#1}}}
 
@@
   Type checking and inference serve as the passage from parsed abstract syntax trees to well-typed programs.
-  While progress has been made in developing techniques for typechecking on a case-by-case basis, the lack of a general theory poses a difficulty of implementing a robust `type checker generator' for diverse languages.
+  While progress has been made in developing techniques for type checking on a case-by-case basis, the lack of a general theory poses a difficulty of implementing a robust `type-checker generator' for diverse languages.
 
-  Motivated by this issue, we develop a \emph{simple} but \emph{general} and \emph{constructive} theory of bidirectional type synthesis where bidirectional type synthesis refers to the mutual use of type checking and inference based on bidirectional typing rules for synthesising the type of a program with type annotations.
+  Motivated by this issue, we develop a \emph{simple} but \emph{general} and \emph{constructive} theory of \emph{bidirectional type synthesis} where bidirectional type synthesis refers to the mutual use of type checking and inference based on bidirectional typing rules for synthesising the type of a program with type annotations.
   First, our theory is simple but general as it deals with any simple type theory and its bidirectional type system that can be specified by a signature, analogous to the role of grammar in parsing.
@@
   Specifically, a type synthesiser for a simple type theory specified by some mode-correct signature is the proof of logical decidability of synthesis, analogous to a parser generator.
-  The notion of bidirectionalisation is introduced to refine annotability in bidirectional typing.
+  The notion of bidirectionalisation is introduced to refine annotatability in bidirectional typing.
   Type synthesis and bidirectionalisation lead to the trichotomy of raw terms which models a practical type synthesiser that throws an exception when its input does not have enough annotations potentially with a proof to locate missing annotations.
-  Hence, by formalizing our general theory of bidirectional type synthesis constructively, we provide a provably correct `type checker generator' for simple type theories, setting the stage for further advances towards a more general theory of type synthesis.
+  Hence, by formalizing our general theory of bidirectional type synthesis constructively, we provide a provably correct `type-checker generator' for simple type theories, setting the stage for further advances towards a more general theory of type synthesis.
 \end{abstract}

### 2023-06-16 tex/sec6-formalisation.tex
[ tex ] Added acmart files; separate compilation for TeXShop
[ tex ] Minor
[ tex ] Writing job distribution

@@
-%! TEX root = BiSig.tex
+%!TEX root = BiSig.tex
 
-\section{Formalisation and More Examples} \label{sec:formalisation}
+\section{Formalisation} \label{sec:formalisation}
 \LT{how to type Agda code without lhs2tex?}
 \begin{enumerate}
-  \item Bidirectional binding signature, functor, and terms (extrinsic typing, raw terms, raw terms in some mode).
-  \item type synthesis and checking.
-  \item examples including STLC, STLC with products, PCF, Computational Calculi, Partial Differentiation
+  \item (Bidirectional) binding signature, functor, and terms (extrinsic typing, raw terms, raw terms in some mode) --- 3 pp
+  \item Compare the induction principle and the \Agda proof of Soundness --- 2 pp
+  \item type synthesis and checking -- 0.5 p
+  \item Examples including STLC (PCF), application in spine form --- 1p
 \end{enumerate}
+
+\LT{7 pages in total}

### 2023-06-17 tex/BiSig.tex
[ tex ] Use blackboard font for annotation (in raw terms)
[ tex ] Clarify the abstract

@@
 
+\usepackage{bbold}
+\newcommand{\hmmax}{0}
+\newcommand{\bmmax}{0}
 \usepackage{bm}
 \usepackage{mathtools}
+\input{mathlig}
 \usepackage{mathpartir}
-\input{mathlig}
-%\usepackage{xfrac}
 \usepackage{xspace}
-%\usepackage{hyperref}
 \usepackage[capitalise,noabbrev]{cleveref}
@@
 \begin{abstract}
-  Type checking and inference serve as the passage from parsed abstract syntax trees to well-typed programs.
-  While progress has been made in developing techniques for type checking on a case-by-case basis, the lack of a general theory poses a difficulty of implementing a robust `type-checker generator' for diverse languages.
+  Type checking and inference serve as the transition from parsed abstract syntax trees to well-typed programs.
+  While progress has been made in developing type-checking techniques on a case-by-case basis, the lack of a general theory poses a difficulty in implementing a robust `type-checker generator' for diverse languages.
 
-  Motivated by this issue, we develop a \emph{simple} but \emph{general} and \emph{constructive} theory of \emph{bidirectional type synthesis} where bidirectional type synthesis refers to the mutual use of type checking and inference based on bidirectional typing rules for synthesising the type of a program with type annotations.
+  In response to this difficulty, we develop a \emph{simple} yet \emph{general} and \emph{constructive} theory of bidirectional type synthesis where \emph{bidirectional type synthesis} refers to the mutual use of type checking and inference based on bidirectional typing rules to determine the type of a program with type annotations.
   First, our theory is simple but general as it deals with any simple type theory and its bidirectional type system that can be specified by a signature, analogous to the role of grammar in parsing.
   Second, our theory is constructive because it is not only formalised in a proof assistant based on constructive type theory but also formulated positively in order to compute.
-  Specifically, a type synthesiser for a simple type theory specified by some mode-correct signature is the proof of logical decidability of synthesis, analogous to a parser generator.
-  The notion of bidirectionalisation is introduced to refine annotatability in bidirectional typing.
-  Type synthesis and bidirectionalisation lead to the trichotomy of raw terms which models a practical type synthesiser that throws an exception when its input does not have enough annotations potentially with a proof to locate missing annotations.
-  Hence, by formalizing our general theory of bidirectional type synthesis constructively, we provide a provably correct `type-checker generator' for simple type theories, setting the stage for further advances towards a more general theory of type synthesis.
+  Specifically, `the proof' of logical decidability of type synthesis is instantiated to a type synthesiser for the simple type theory specified by a valid signature, analogous to a parser generator.
+  We introduce the notion of bidirectionalisation to refine annotatability in bidirectional typing.
+  Type synthesis and bidirectionalisation result in the trichotomy of raw terms, which models a practical type synthesiser that may throw an exception to locate missing annotations.
+  By formalizing our general theory of bidirectional type synthesis constructively, we provide a correct-by-construction `type-checker generator' for simple type theories, taking the first step towards a more general theory of bidirectional type synthesis.
 \end{abstract}

### 2023-06-17 tex/sec1-intro.tex
[ tex ] Use '\paragraph' instead of '\subsubsection' for related work
[ tex ] Shorten the abstract
[ tex ] Writing job distribution
[ tex ] Clarify the abstract

@@
 
+%  Bidirectional typing had only been a set of principles and characteristics to design typing rules so that a type synthesis algorithm can be `read off' easily
+  %, subsuming soundness, completeness, and decidability of type synthesis algorithm.
+%  Our work addresses the problem with existing methodologies, setting the stage for further advances towards a more general theory of type synthesis.
+
+  %In short, this paper is a small but first step towards a general theory of type checking and inference.
+  %By formalizing our theory of bidirectional type synthesise in \Agda, we provide a framework capable of handling bidirectional type systems of simple types uniformly.
+  %Our work addresses the problem with existing methodologies, 
 Points to elaborate:
@@
   \item A review of what has been done towards this line of research.
-  \item Finally, our responses (contributions of this paper).
+  \item Finally, our responses (contributions of this paper) and some example.
 \end{enumerate}
@@
 \citep{Xie2018}
-\subsubsection{Language formalization and its frameworks}
+\paragraph{Language formalization and its frameworks}
 \cite{Wadler2022}
@@
 
-\subsubsection{Theories of abstract syntax with variable binding}
+\paragraph{Theories of abstract syntax with variable binding}
 \cite{Fiore1999,Hirschowitz2010,Ahrens2018,Fiore2022,Ahrens2021,Arkor2020,Hirschowitz2022}
@@
 
-\subsubsection{Type checker generation}
+\paragraph{Type checker generation}
 \cite{Gast2004,Grewe2015,Pacak2020,Cimini2020}
 
-\subsubsection{Bidirectional typing}
+\paragraph{Bidirectional typing}
 
@@
 \begin{enumerate}
+  \item Introduction (\Cref{sec:intro}) 
+    \LT{1+3 pp}
   \item Key ideas (\Cref{sec:key-ideas})
+    \Josh{2.5 pp}
   \item Definitions for bidirectional type systems (bidirectional binding signature, bidirectional type systems, signature erasure and mode annotation) (\Cref{sec:defs})
+    \LT{4 pp}
   \item Soundness, completeness, bidirectionalisation, and annotatability (\Cref{sec:annotatability})
     \Josh{Erasure from bidirectional typing derivation to raw terms with a mode}
+    \Josh{3 pp}
   \item Bidirectional type inference (\Cref{sec:type-synthesis})
-  \item Formalisation with further examples (\Cref{sec:formalisation})
-  \item Future work (\Cref{sec:future})
+    \LT{4 pp}
+  \item Formalisation and further examples (\Cref{sec:formalisation})
+    \LT{5.5pp}
+  \item Concluding remarks (\Cref{sec:future})
+    \LT{1 p} \Josh{1p}
 \end{enumerate}

### 2023-06-17 tex/sec2-outline.tex
[ tex ] Use blackboard font for annotation (in raw terms)
[ tex ] Fix mistakes in Anno
[ tex ] Style raw terms

@@
   \begin{tabular}{ r l }
-    $\boxed{\Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn A}$ & A raw term $t$ synthesises a type $A$ under $\Gamma$ \\
-    $\boxed{\Gamma |-_{\Sigma, \Omega} \isTerm{t} \chk A}$ &A raw term $t$ checks against a type $A$ under $\Gamma$
+    $\boxed{\Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn A}$ & A raw term $\isTerm{t}$ synthesises a type $A$ under $\Gamma$ \\
+    $\boxed{\Gamma |-_{\Sigma, \Omega} \isTerm{t} \chk A}$ & A raw term $\isTerm{t}$ checks against a type $A$ under $\Gamma$
   \end{tabular}
@@
   \begin{mathpar}
-    \inferrule{(x : A) \in \Gamma}{\Gamma \vdash \isTerm{x} \syn A}\,\SynRule{Var}
+    \inferrule{(\isTerm{x} : A) \in \Gamma}{\Gamma \vdash \isTerm{x} \syn A}\,\SynRule{Var}
     \and
-    \inferrule{\Gamma \vdash \isTerm{t} \chk B \\ A = B}{\Gamma \vdash \isTerm{t : B}\syn A}\,\SynRule{Anno}
+    \inferrule{\Gamma \vdash \isTerm{t} \chk A}{\Gamma \vdash \isTerm{t \annotate A}\syn A}\,\SynRule{Anno}
     \and
-    \inferrule{\Gamma \vdash \isTerm{t} \syn B \\ A = B}{\Gamma \vdash \isTerm{t} \chk A}\,\ChkRule{Sub}
+    \inferrule{\Gamma \vdash \isTerm{t} \syn B \\ B = A}{\Gamma \vdash \isTerm{t} \chk A}\,\ChkRule{Sub}
     \and

### 2023-06-17 tex/sec3-defs.tex
[ tex ] Make some figures more compact
[ tex ] Make raw term formation rules in one line
[ tex ] Use blackboard font for annotation (in raw terms)
[ tex ] Fix mistakes in Anno
[ tex ] Style raw terms

@@
       \inferrule{\Theta \ni X_i}{\Theta |-_\Sigma X_i} \and
-    \inferrule{\Theta |-_{\Sigma} A_1 \\ \cdots \\ \Theta |-_{\Sigma} A_{\arity(o)}}{\Theta |-_{\Sigma} \tyOp_o(A_1, \ldots, A_{\arity(o)})}
+      \inferrule{\Theta |-_{\Sigma} A_1 \\ \cdots \\ \Theta |-_{\Sigma} A_n}{\Theta |-_{\Sigma} \tyOp_o(A_1, \ldots, A_n)}\;\text{where $n = \arity(o)$}
     \end{mathpar}
@@
   \small
-  \judgbox{|-_{\Sigma, \Omega} \isTerm{t}}{$t$ is a raw term for the signature $(\Sigma, \Omega)$}
+  \judgbox{|-_{\Sigma, \Omega} \isTerm{t}}{$\isTerm{t}$ is a raw term for the signature $(\Sigma, \Omega)$}
   \begin{mathpar}
-    \inferrule{x : \Identifier}{|-_{\Sigma, \Omega} \isTerm{x}}\,\Rule{Var}
+    \inferrule{\isTerm{x} : \Identifier}{|-_{\Sigma, \Omega} \isTerm{x}}\,\Rule{Var}
     \and
-    \inferrule{\cdot |-_{\Sigma} A \\ |-_{\Sigma, \Omega}\isTerm{t}}{|-_{\Sigma, \Omega} \isTerm{t \annote A}}\,\Rule{Anno}
-    \\
+    \inferrule{\cdot |-_{\Sigma} A \\ |-_{\Sigma, \Omega}\isTerm{t}}{|-_{\Sigma, \Omega} \isTerm{t \annotate A}}\,\Rule{Anno}
+    \and
     \inferrule{|-_{\Sigma, \Omega} \isTerm{t_1} \quad \cdots \quad |-_{\Sigma, \Omega} \isTerm{t_n}}
-    {|-_{\Sigma, \Omega} \isTerm{\tmOp_o(\vec{x}_1.t_1; \ldots; \vec{x}_n.t_n)}}\,\Rule{Op}
-    \and \text{for $o \colon \Xi \rhd [\Delta_1]A_{1}, \ldots, [\Delta_{n}] A_{n} \to A$ in $\Omega$}
+    {|-_{\Sigma, \Omega} \tmOpts }\,\Rule{Op}
   \end{mathpar}
+  \begin{flushright}
+    for $o \colon \Xi \rhd [\Delta_1]A_{1}, \ldots, [\Delta_{n}] A_{n} \to A$ in $\Omega$
+  \end{flushright}%
   \caption{Raw terms}
@@
   \small
-  \judgbox{\Gamma |-_{\Sigma, \Omega} \isTerm{t} : A}{A raw term $t$ has a type $A$ (without type variables) under $\Gamma$ for the signature $(\Sigma, \Omega)$}
+  \judgbox{\Gamma |-_{\Sigma, \Omega} \isTerm{t} : A}{A raw term $\isTerm{t}$ has a type $A$ (without type variables) under $\Gamma$ for the signature $(\Sigma, \Omega)$}
   \begin{mathpar}
-    \inferrule{(x : A) \in \Gamma}{\Gamma |-_{\Sigma, \Omega} \isTerm{x} : A}\,\Rule{Var}
+    \inferrule{(\isTerm{x} : A) \in \Gamma}{\Gamma |-_{\Sigma, \Omega} \isTerm{x} : A}\,\Rule{Var}
     \and
-    \inferrule{\Gamma |- \isTerm{t} : A}{\Gamma |- (\isTerm{t \annote A}) : A}\,\Rule{Anno}
+    \inferrule{\Gamma |- \isTerm{t} : A}{\Gamma |- (\isTerm{t \annotate A}) : A}\,\Rule{Anno}
     \and
-    \inferrule{\rho : \Sub{\Xi}{\emptyset} \\ \Gamma, \isTerm{\vec{x}_1} : \simsub{\Delta_{1}}{\rho} |-_{\Sigma, \Omega} \isTerm{t_1} : \simsub{A_{1}}{\rho} \quad\cdots\quad \Gamma, \isTerm{\vec{x}_n} : \simsub{\Delta_{n}}{\rho} |-_{\Sigma, \Omega} \isTerm{t_n} : \simsub{A_{n}}{\rho}}
-    {\Gamma |-_{\Sigma, \Omega} \isTerm{\tmOp_o(\vec{x}_1.\,t_1; \ldots; \vec{x}_n.\,t_n)} : \simsub{A}{\rho}}\,\Rule{Op}
+    \inferrule{\rho : \Sub{\Xi}{\emptyset} \\ \Gamma, \vec{\isTerm{x}}_\isTerm{1} : \simsub{\Delta_{1}}{\rho} |-_{\Sigma, \Omega} \isTerm{t_1} : \simsub{A_{1}}{\rho} \quad\cdots\quad \Gamma, \vec{\isTerm{x}}_\isTerm{n} : \simsub{\Delta_{n}}{\rho} |-_{\Sigma, \Omega} \isTerm{t_n} : \simsub{A_{n}}{\rho}}
+    {\Gamma |-_{\Sigma, \Omega} \tmOpts : \simsub{A}{\rho}}\,\Rule{Op}
     \and \text{for $o\colon \Xi \rhd [\Delta_1]A_1, \ldots, [\Delta_{n}]A_{n} \to A$ in $\Omega$}
@@
   \small
-  \judgbox{|-_{\Sigma, \Omega} \isTerm{t}^\dir{d}}{$t$ is a raw term for $(\Sigma, \erase{\Omega})$ in mode $d$ for  $(\Sigma, \Omega)$}
+  \judgbox{|-_{\Sigma, \Omega} \isTerm{t}^\dir{d}}{$\isTerm{t}$ is a raw term for $(\Sigma, \erase{\Omega})$ in mode $d$ for  $(\Sigma, \Omega)$}
   \begin{mathpar}
@@
     \and
-    \inferrule{\cdot |-_{\Sigma} A \\ |-_{\Sigma, \Omega}\isTerm{t}^{\chk}}{|-_{\Sigma, \Omega} (\isTerm{t \annote A})^{\syn}}\,\SynRule{Anno}
+    \inferrule{\cdot |-_{\Sigma} A \\ |-_{\Sigma, \Omega}\isTerm{t}^{\chk}}{|-_{\Sigma, \Omega} (\isTerm{t \annotate A})^{\syn}}\,\SynRule{Anno}
     \and
     \inferrule{|-_{\Sigma, \Omega} \isTerm{t}^{\syn}}{|-_{\Sigma, \Omega} \isTerm{t}^{\chk}}\,\ChkRule{Sub}
-  \end{mathpar}
-  \begin{mathpar}
+    \and
     \inferrule{|-_{\Sigma, \Omega} \isTerm{t_1}^\dir{d_1} \quad \cdots \quad |-_{\Sigma, \Omega} \isTerm{t_n}^\dir{d_n}}
-    {|-_{\Sigma, \Omega} \isTerm{\tmOp_o(\vec{x}_1.\, t_1; \ldots;\vec{x}_n.\, t_n)}^\dir{d}}\,\Rule{Op}
-    \and \text{for $o \colon \Xi \rhd [\Delta_1]A_{1}^{\dir{d_1}}, \ldots, [\Delta_{n}] A_{n}^{\dir{d_n}} \to A^{\dir{d}}$ in $\Omega$}
+    {|-_{\Sigma, \Omega} \tmOpts^\dir{d}}\,\Rule{Op}
   \end{mathpar}
@@
   \small
-  \judgbox{\Gamma |-_{\Sigma, \Omega} \isTerm{t} :^\dir{d} A}{A raw term $t$ is of type $A$ (without type variables) in mode $d$ under $\Gamma$}
+  \judgbox{\Gamma |-_{\Sigma, \Omega} \isTerm{t} :^\dir{d} A}{A raw term $\isTerm{t}$ is of type $A$ (without type variables) in mode $\dir{d}$ under $\Gamma$}
   \begin{mathpar}
@@
     \and
-    \inferrule{\Gamma |-_{\Sigma, \Omega} \isTerm{t} :^{\chk} A}{\Gamma |-_{\Sigma, \Omega} (\isTerm{t \annote B}) :^{\syn}  A}\,\SynRule{Anno}
+    \inferrule{\Gamma |-_{\Sigma, \Omega} \isTerm{t} :^{\chk} A}{\Gamma |-_{\Sigma, \Omega} (\isTerm{t \annotate A}) :^{\syn}  A}\,\SynRule{Anno}
     \and
-    \inferrule{\Gamma |-_{\Sigma, \Omega} \isTerm{t} :^{{\syn}} B \\ A = B}{\Gamma |-_{\Sigma, \Omega} \isTerm{t} :^{\chk} A}\,\ChkRule{Sub}
+    \inferrule{\Gamma |-_{\Sigma, \Omega} \isTerm{t} :^{{\syn}} B \\ B = A}{\Gamma |-_{\Sigma, \Omega} \isTerm{t} :^{\chk} A}\,\ChkRule{Sub}
     \\
-    \inferrule{\rho\colon \Sub{\Xi}{\emptyset} \\ \Gamma, \simsub{\isTerm{\vec{x}_{1}} : \Delta_{1}}{\rho} |-_{\Sigma, \Omega} \isTerm{t_{1}} :^\dir{d_1} \simsub{A_1}{\rho} \\
+    \inferrule{\rho\colon \Sub{\Xi}{\emptyset} \\ \Gamma, \simsub{\vars{x}_\isTerm{1} : \Delta_{1}}{\rho} |-_{\Sigma, \Omega} \isTerm{t_{1}} :^\dir{d_1} \simsub{A_1}{\rho} \\
       \cdots \\
-    \Gamma, \simsub{\isTerm{\vec{x}_{n}} : \Delta_{n}}{\rho} |-_{\Sigma, \Omega} \isTerm{t_n} :^{\dir{d_n}} \simsub{A_{n}}{\rho}}
-    {\Gamma |-_{\Sigma, \Omega} \isTerm{\tmOp_o(\vec{x}_1 .\, t_{1}; \ldots; \vec{x}_{n}.\, t_{n})} :^{\dir{d}} \simsub{A}{\rho}} \,\mathsf{Op}
+    \Gamma, \simsub{\vars{x}_\isTerm{n} : \Delta_{n}}{\rho} |-_{\Sigma, \Omega} \isTerm{t_n} :^{\dir{d_n}} \simsub{A_{n}}{\rho}}
+    {\Gamma |-_{\Sigma, \Omega} \tmOpts :^{\dir{d}} \simsub{A}{\rho}} \,\mathsf{Op}
     \and \text{for $o \colon \Xi \rhd [\Delta_1]A_{1}^{\dir{d_1}}, \ldots, [\Delta_{n}] A_{n}^{\dir{d_n}} \to A^\dir{d}$ in $\Omega$}

### 2023-06-17 tex/sec4-bidirectionalisation.tex
[ tex ] Make some figures more compact
[ tex ] Writing job distribution
[ tex ] Use blackboard font for annotation (in raw terms)
[ tex ] Style raw terms

--- 2023-06-14 tex/sec4-bidirectionalisation.tex
@@
 \subsection{Bidirectionalisation}
+
 \begin{proposition}[Decidability for Bidirectionalisation] \label{thm:bidirectionalisation}
-  It is decidable whether a raw term~$t$ is in mode~$d$ for any $t$ and $d$.
+  It is decidable whether a raw term~$\isTerm{t}$ is in mode~$d$ for any $\isTerm{t}$ and $d$.
 \end{proposition}
@@
 \subsection{Annotatability is not Completeness}
+
+\begin{figure}
+  \centering
+  \small
+  \bgroup
+  \renewcommand{\arraystretch}{1.5}
+  \begin{tabular}{ r l }
+    $\boxed{|-_{\Sigma, \Omega} \isTerm{t}\;\dir{d}}$ & $\isTerm{t}$ is a raw term possibly with missing annotation or redundant casts in mode $d$ \\
+    $\boxed{|-_{\Sigma, \Omega} \isTerm{t}\;\dir{d}}$ & $\isTerm{t}$ is a raw term without missing annotation \\
+    $\boxed{|-_{\Sigma, \Omega} \isTerm{t}\;\dir{d}}$ & $\isTerm{t}$ is a raw term without redundant casts \\
+  \end{tabular}
+  \egroup
+  \LT{Design a notation for \textsf{Pre?} so that a raw term without missing annotation or redundant casts is exactly the intersection of these two cases.}
+  \begin{mathpar}
+    \inferrule{|-_{\Sigma, \Omega}\isTerm{t}^{\chk}}{|-_{\Sigma, \Omega} \isTerm{t}^{\syn}}\,\SynRule{Anno?}
+    \and
+    \inferrule{\isTerm{x} : \Identifier}{|-_{\Sigma, \Omega} \isTerm{x}^{\syn}}
+    \and
+    \inferrule{|-_{\Sigma, \Omega}\isTerm{t}^{\chk}}{|-_{\Sigma, \Omega} (\isTerm{t \annotate A})^{\syn}}
+    \and
+    \inferrule{|-_{\Sigma, \Omega} \isTerm{t}^{\syn}}{|-_{\Sigma, \Omega} \isTerm{t}^{\chk}}
+    \and
+    \inferrule{|-_{\Sigma, \Omega} \isTerm{t_1}^\dir{d_1} \quad \cdots \quad |-_{\Sigma, \Omega} \isTerm{t_n}^\dir{d_n}}
+    {|-_{\Sigma, \Omega} \tmOpts^\dir{d}}
+  \end{mathpar}
+  \caption{Preprocessed raw terms, II}
+  \label{fig:raw-with-missing-annotations}
+\end{figure}
+
+\begin{figure}
+  \centering\small
+  \judgbox{\isTerm{t} \sqsupseteq \isTerm{u}}{A raw term $t$ is more annotated than $u$ (for some bidirectional type system $(\Sigma, \Omega)$)}
+  \begin{mathpar}
+    \inferrule{\isTerm{t} \sqsupseteq \isTerm{u}}
+              {(\isTerm{t} \annotate A) \sqsupseteq \isTerm{u}}\;\Rule{More}
+    \and
+    \inferrule{\vphantom{x : \Identifier}}
+              {\isTerm{x} \sqsupseteq \isTerm{x}}
+    \and
+    \inferrule{\isTerm{t} \sqsupseteq \isTerm{u}}
+              {(\isTerm{t} \annotate A) \sqsupseteq (\isTerm{u} \annotate A)}
+    \and
+    \inferrule{\isTerm{t_1} \sqsupseteq \isTerm{u_1} \quad \cdots \quad \isTerm{t_n} \sqsupseteq{u_n}}
+              {\tmOpts \sqsupseteq \tmOpus}
+  \end{mathpar}
+
+  \caption{Annotation ordering between raw terms}
+  \label{fig:annotation-order}
+\end{figure}

### 2023-06-17 tex/sec5-synthesis.tex
[ tex ] Style raw terms

@@
   Let $(\Sigma, \Omega)$ be a mode-correct bidirectional type system.
-  Then, for every raw term $t$ in synthesising mode and two derivations of 
+  Then, for every raw term $\isTerm{t}$ in synthesising mode and two derivations of 
   \[
-    \Gamma |-_{\Sigma, \Omega} t \syn A
+    \Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn A
     \quad\text{and}\quad
-    \Gamma |-_{\Sigma, \Omega} t \syn B
+    \Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn B
   \]
@@
   Let $(\Sigma, \Omega)$ be a mode-correct bidirectional type system.
-  Then, for every context $\Gamma$ and raw term $t$, exactly one of the following statements holds:
+  Then, for every context $\Gamma$ and raw term $\isTerm{t}$, exactly one of the following statements holds:
   \begin{enumerate}
-    \item $t$ is synthesising and the judgement $\Gamma |-_{\Sigma, \Omega} t : A$ has a derivation for some type $A$.
-    \item $t$ is synthesising but the judgement $\Gamma |-_{\Sigma, \Omega} t : A$ has no derivation for any type $A$.
-    \item $t$ is not synthesising.
+    \item $\isTerm{t}$ is synthesising and the judgement $\Gamma |-_{\Sigma, \Omega} \isTerm{t} : A$ has a derivation for some type $A$.
+    \item $\isTerm{t}$ is synthesising but the judgement $\Gamma |-_{\Sigma, \Omega} \isTerm{t} : A$ has no derivation for any type $A$.
+    \item $\isTerm{t}$ is not synthesising.
   \end{enumerate}
@@
       \[
-        \Gamma, \simsub{\Delta_i}{\ext{\rho}} |-_{\Sigma, \Omega} \isTerm{t_i} \colon \simsub{A_i}{\ext{\rho}}^{\dir{d_i}}
+        \Gamma, \vars{x}_\isTerm{i} : \simsub{\Delta_i}{\ext{\rho}} |-_{\Sigma, \Omega} \isTerm{t_i} \colon \simsub{A_i}{\ext{\rho}}^{\dir{d_i}}
       \]
 
-    \item or there is no extension $\sigma$ of $\rho$ such that all $\Gamma, \simsub{\Delta_i}{\sigma} |-_{\Sigma, \Omega} \isTerm{t_i} \colon \simsub{A_i}{\sigma}^{\dir{d_i}}$ has a typing derivation.
+    \item or there is no extension $\sigma$ of $\rho$ such that all $\Gamma, \vars{x}_\isTerm{i} : \simsub{\Delta_i}{\sigma} |-_{\Sigma, \Omega} \isTerm{t_i} \colon \simsub{A_i}{\sigma}^{\dir{d_i}}$ has a typing derivation.
   \end{enumerate}
@@
 \begin{proof}[Proof of {\Cref{thm:bidirectional-type-synthesis}}]
-  We prove this statement by induction on the derivation of $|-_{\Sigma, \Omega} \isTerm{t}^{d}$ (instead of $t$ alone).
+  We prove this statement by induction on the derivation of $|-_{\Sigma, \Omega} \isTerm{t}^{d}$ (instead of $\isTerm{t}$ alone).
   The first two cases \SynRule{Var} and \SynRule{Anno} are straightforward and have nothing to do with mode-correctness directly.
-  The third case \ChkRule{Sub} invokes the uniqueness of synthesised types to refute the case that $\Gamma |- t \syn B$ but $A \neq B$ for a given type $A$ to check.
+  The third case \ChkRule{Sub} invokes the uniqueness of synthesised types to refute the case that $\Gamma |- \isTerm{t} \syn B$ but $A \neq B$ for a given type $A$ to check.
   The last case \Rule{Op} is the interesting part of this theorem.
   \begin{description}
-    \item[\SynRule{Var}:] If $t$ is a variable $x$ in the context $\Gamma$, then there exists a type $A$ such that $\Gamma |- x \syn A$.
-      Otherwise, if $x$ is not in $\Gamma$ but $\Gamma |- x \syn A$ is derivable, then by inversion $x$ must be in $\Gamma$, a contradiction.
-    \item[\SynRule{Anno}:] For a raw term of the form $(t : A)$ in synthesising mode, it is decidable that $\Gamma |- t \chk A$ is derivable or not by induction hypothesis.
+    \item[\SynRule{Var}:] If $\isTerm{t}$ is a variable $\isTerm{x}$ in the context $\Gamma$, then there exists a type $A$ such that $\Gamma |- \isTerm{x} \syn A$.
+      Otherwise, if $\isTerm{x}$ is not in $\Gamma$ but $\Gamma |- \isTerm{x} \syn A$ is derivable, then by inversion $\isTerm{x}$ must be in $\Gamma$, a contradiction.
+    \item[\SynRule{Anno}:] For a raw term of the form $\isTerm{t} \annotate A$ in synthesising mode, it is decidable that $\Gamma |- \isTerm{t} \chk A$ is derivable or not by induction hypothesis.
       \begin{enumerate}
-        \item If $\Gamma |- t \chk A$ is derivable, then $\Gamma |- (t : A) \syn A$ is derivable.
-        \item If $\Gamma |- t \chk A$ is not derivable but $\Gamma |- (t : A) \syn$ is derivable, then by inversion $\Gamma |- t \chk A$ is derivable, thus a contradiction.
+        \item If $\Gamma |- \isTerm{t} \chk A$ is derivable, then $\Gamma |- \isTerm{t \annotate A} \syn A$ is derivable.
+        \item If $\Gamma |- t \chk A$ is not derivable but $\Gamma |- \isTerm{t \annotate A} \syn$ is derivable, then by inversion $\Gamma |- t \chk A$ is derivable, thus a contradiction.
       \end{enumerate}
@@
     \item[\Rule{Op}:]
-      Suppose that $t$ is of them $\tmOp_o(\vec{x}_1.t_1; \ldots; \vec{x}_n.t_n)$ for some mode-correct operation $o$ in $\Omega$.
+      Suppose that $t$ is of them $\tmOpts$ for some mode-correct operation $o$ in $\Omega$.
       No matter $t$ is checking or synthesising, we need to build derivations for its arguments incrementally and maintain a set of unifiers, i.e.\ a partial substitution $\rho\colon \PSub{\Xi}{\emptyset}$ is incrementally defined which is initially defined nowhere if $t$ is synthesising or on variables of the target $A$ if $t$ is checking.

### 2023-06-17 tex/BiSig.tex
[ tex ] Fix mistakes in Anno
[ tex ] Minor

@@
 \email{joshko@iis.sinica.edu.tw}
-
 \affiliation{%
@@
 \title{A Theory of Bidirectional Type Synthesis for Simple Types}
-
 \begin{abstract}
@@
   In response to this difficulty, we develop a \emph{simple} yet \emph{general} and \emph{constructive} theory of bidirectional type synthesis where \emph{bidirectional type synthesis} refers to the mutual use of type checking and inference based on bidirectional typing rules to determine the type of a program with type annotations.
-  First, our theory is simple but general as it deals with any simple type theory and its bidirectional type system that can be specified by a signature, analogous to the role of grammar in parsing.
+  First, our theory is simple yet general as it deals with any simple type theory and its bidirectional type system that can be specified by a signature, analogous to the role of grammar in parsing.
   Second, our theory is constructive because it is not only formalised in a proof assistant based on constructive type theory but also formulated positively in order to compute.
-  Specifically, `the proof' of logical decidability of type synthesis is instantiated to a type synthesiser for the simple type theory specified by a valid signature, analogous to a parser generator.
+  Specifically, the `proof' of logical decidability of type synthesis is instantiated to a synthesiser for the simple type theory specified by a valid signature, analogous to a parser generator.
   We introduce the notion of bidirectionalisation to refine annotatability in bidirectional typing.

### 2023-06-19 tex/sec2-outline.tex
[ tex ] A bit of S2

@@
 
-\section{The Key Ideas} \label{sec:key-ideas}
+\section{Bidirectional Type Synthesis for Simply Typed $\lambda$-Calculus} \label{sec:key-ideas}
+
 \begin{figure}
@@
   \begin{tabular}{ r l }
-    $\boxed{\Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn A}$ & A raw term $\isTerm{t}$ synthesises a type $A$ under $\Gamma$ \\
-    $\boxed{\Gamma |-_{\Sigma, \Omega} \isTerm{t} \chk A}$ & A raw term $\isTerm{t}$ checks against a type $A$ under $\Gamma$
+    $\boxed{\Gamma' |- \isTerm{t}}$ & Given a list~$\Gamma'$ of variables, $t$~is a raw term with free variables in~$\Gamma'$
+  \end{tabular}
+  \egroup
+  \centering
+  \begin{mathpar}
+    \inferrule{x \in \Gamma'}{\Gamma' \vdash \isTerm{x}}\,\Rule{Var}
+    \and
+    \inferrule{\Gamma' \vdash \isTerm{t} \\ \isTerm{A}\text{ is a closed type}}{\Gamma' \vdash \isTerm{(t \annotate A)}}\,\Rule{Anno}
+    \and
+    \inferrule{\Gamma', \isTerm{x} \vdash \isTerm{t}}{\Gamma' \vdash \isTerm{\lam{x}t}}\,\Rule{Abs}
+    \and
+    \inferrule{\Gamma' \vdash \isTerm{t} \\ \Gamma' \vdash \isTerm{u}}{\Gamma' \vdash \isTerm{t\;u}}\,\Rule{App}
+  \end{mathpar}
+  \caption{Raw terms for simply typed $\lambda$-calculus}
+  \label{fig:STLC-raw-terms}
+\end{figure}
+
+\begin{figure}
+  \small
+  \bgroup
+  \renewcommand{\arraystretch}{1.5}
+  \begin{tabular}{ r l }
+    $\boxed{\Gamma |- \isTerm{t} : A}$ & A raw term~$t$ has type~$A$ under context~$\Gamma$
+  \end{tabular}
+  \egroup
+  \centering
+  \begin{mathpar}
+    \inferrule{(x : A) \in \Gamma}{\Gamma \vdash \isTerm{x} : A}\,\Rule{Var}
+    \and
+    \inferrule{\Gamma \vdash \isTerm{t} : A}{\Gamma \vdash \isTerm{(t \annotate A)} : A}\,\Rule{Anno}
+    \and
+    \inferrule{\Gamma, \isTerm{x} : A \vdash \isTerm{t} : B}{\Gamma \vdash \isTerm{\lam{x}t} : A \bto B}\,\Rule{Abs}
+    \and
+    \inferrule{\Gamma \vdash \isTerm{t} : A \bto B \\ \Gamma \vdash \isTerm{u} : A}{\Gamma \vdash \isTerm{t\;u} : B}\,\Rule{App}
+  \end{mathpar}
+  \caption{Typing derivations for simply typed $\lambda$-calculus}
+  \label{fig:STLC-typing-derivations}
+\end{figure}
+
+To start off, we give an overview of our theory by instantiating it to simply typed $\lambda$-calculus~($\Lambda_{\bto}$).
+Roughly speaking, the problem of type synthesis requires us to take an untyped abstract syntax tree ---which we call a \emph{raw term}--- as input, and produce a typing derivation for the term if possible.
+To give more precise definitions:
+The raw terms for~$\Lambda_{\bto}$ are defined in \cref{fig:STLC-raw-terms}; besides the standard constructs, there is an \textsc{Anno} rule that allows the user to insert type annotations to help with type synthesis.
+Correspondingly, the definition of typing derivations for~$\Lambda_{\bto}$ in \cref{fig:STLC-typing-derivations} also includes an \textsc{Anno} rule enforcing that the type of an annotated term does match the annotation.
+Now we can define what it means to solve the type synthesis problem.
+
+\begin{definition}
+\label{def:STLC-type-synthesisers}
+Parametrised by a (possibly proof-relevant) predicate~$E$ on raw terms, a \emph{type synthesiser} (for $\Lambda_{\bto}$) takes a context~$\Gamma$ and a raw term $\Gamma' \vdash t$ (where $\Gamma'$~is the list of variables appearing in~$\Gamma$) as input, and establishes one of the following outcomes:
+\begin{enumerate}
+\item a derivation of $\Gamma \vdash t : A$ for some type~$A$,
+\item there does not exist a derivation $\Gamma \vdash t : A$ for any type~$A$, or
+\item $E$~holds for~$t$.
+\end{enumerate}
+\end{definition}
+
+It is crucial to allow the third outcome, without which we would be requiring the type synthesis problem to be decidable, but this requirement would quickly become impossible to meet when the theory is extended to handle more complex types.
+If a type synthesiser cannot decide whether there is a typing derivation, it is allowed to give an `excuse' instead of an answer.
+Acceptable excuses are captured by the predicate~$E$, which describes what is wrong with an input term, for example not having enough type annotations.
+
+\begin{figure}
+  \small
+  \bgroup
+  \renewcommand{\arraystretch}{1.5}
+  \begin{tabular}{ r l }
+    $\boxed{\Gamma |- \isTerm{t} \syn A}$ & A raw term $t$ synthesises a type $A$ under $\Gamma$ \\
+    $\boxed{\Gamma |- \isTerm{t} \chk A}$ &A raw term $t$ checks against a type $A$ under $\Gamma$
   \end{tabular}
@@
     \and
-    \inferrule{\Gamma \vdash \isTerm{t} \chk A}{\Gamma \vdash \isTerm{t \annotate A}\syn A}\,\SynRule{Anno}
+    \inferrule{\Gamma \vdash \isTerm{t} \chk A}{\Gamma \vdash \isTerm{(t \annotate A)}\syn A}\,\SynRule{Anno}
     \and
@@
     \and
+    \inferrule{\Gamma, \isTerm{x} : A \vdash \isTerm{t} \chk B}{\Gamma \vdash \isTerm{\lam{x}t} \chk A \bto B}\,\ChkRule{Abs}
+    \and
     \inferrule{\Gamma \vdash \isTerm{t} \syn A \bto B \\ \Gamma \vdash \isTerm{u} \chk A}{\Gamma \vdash \isTerm{t\;u} \syn B}\,\SynRule{App}
-    \and
-    \inferrule{\Gamma, \isTerm{x} : A \vdash \isTerm{t} \chk B}{\Gamma \vdash \isTerm{\lam{x}t} \chk A \bto B}\,\ChkRule{Abs}
   \end{mathpar}
-  \caption{Bidirectional simply typed $\lambda$-calculus $\Lambda^{\Leftrightarrow}_{\bto}$}
-  \label{fig:bi-stlc}
+  \caption{Bidirectional typing derivations for simply typed $\lambda$-calculus}
+  \label{fig:STLC-bidirectional-typing-derivations}
 \end{figure}
+
+Now our goal is to use \cref{def:STLC-type-synthesisers} as a specification and implement it using a \emph{bidirectional} type synthesiser, which attempts to produce \emph{bidirectional} typing derivations defined in \cref{fig:STLC-bidirectional-typing-derivations}.
+While type synthesis is not decidable in general, for certain kinds of terms it is still possible to synthesise their types (for example, the type of a variable can be looked up in the context), and for other kinds of terms we can switch to the simpler problem of type \emph{checking}, where the expected type of a term is also given so that the type checker has more information to work with.
+If a bidirectional type system is designed well, then synthesised type information will be enough for figuring out the expected types for checking.
+
 \begin{remark}
@@
 \end{remark}
+
+\Josh{
+
+• Figures: raw terms, standard typing, bidirectional typing, and mode derivations ($\frac{2}{3}$~pages)
+
+• Bidirectional typing (general idea and specific STLC rules); the goal is to implement a type synthesiser using a bidirectional one
+
+• Beneficial to separate \emph{bidirectionalisation} from type synthesis (syntactic characterisation of the domain of type synthesis; making type synthesisers syntax- (or more precisely, derivation-) directed)
+
+• Definitions of mode derivations, bidirectionalisers, and bidirectional type synthesisers
+
+• Back to the goal; soundness and completeness ($1\frac{1}{2}$~pages)
+}

### 2023-06-20 tex/BiSig.tex
[ tex ] Add another point to elaborate on simple types

@@
   We introduce the notion of bidirectionalisation to refine annotatability in bidirectional typing.
-  Type synthesis and bidirectionalisation result in the trichotomy of raw terms, which models a practical type synthesiser that may throw an exception to locate missing annotations.
+  Type synthesis and bidirectionalisation result in the trichotomy of raw terms, which models a practical type synthesiser that may throw an exception and locate missing annotations.
   By formalizing our general theory of bidirectional type synthesis constructively, we provide a correct-by-construction `type-checker generator' for simple type theories, taking the first step towards a more general theory of bidirectional type synthesis.

### 2023-06-20 tex/sec1-intro.tex
[ tex ] Add another point to elaborate on simple types

@@
 
-Our assumptions and contributions:
+Our restrictions:
 \begin{enumerate}
+  \item Assumptions: Why only simple types? Various ways to deal with polymorphism; consistent on simple types.
   \item Assumption: Syntax-directed typing rules
-  \item Practical: bidirectionalisation over annotatability, decidability over soundness and completeness.
-  \item Generic: Type checking once and for all bidirectional type systems for simple types.
-  \item Correct and Constructive: Fully formalised in Martin-Löf type theory (with Axiom K).
+\end{enumerate}
+
+Our contributions:
+\begin{enumerate}
+  \item Generic: Type checking for simple type theories;
+  \item Computational formulation: Bidirectionalisation over annotatability; Decidability over soundness and completeness;
+  \item Verified and Constructive: Fully formalised in Martin-Löf type theory (with Axiom K).
     LEM is not assumed, so a decidable statement has both computational and non-trivial logical readings.

### 2023-06-21 tex/sec2-outline.tex
[ tex ] Use `\texorpdfstring` to substitute latex with Unicode characters in PDF
[ tex ] Most of S2

--- 2023-06-17 tex/sec2-outline.tex
@@
 
-\section{The Key Ideas} \label{sec:key-ideas}
+\section{Bidirectional Type Synthesis for Simply Typed \texorpdfstring{$\lambda$}{λ}-Calculus} \label{sec:key-ideas}
+
 \begin{figure}
@@
   \begin{tabular}{ r l }
-    $\boxed{\Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn A}$ & A raw term $\isTerm{t}$ synthesises a type $A$ under $\Gamma$ \\
-    $\boxed{\Gamma |-_{\Sigma, \Omega} \isTerm{t} \chk A}$ & A raw term $\isTerm{t}$ checks against a type $A$ under $\Gamma$
+    $\boxed{V |- \isTerm{t}}$ & Given a list~$V$ of variables, $t$~is a raw term with free variables in~$V$
+  \end{tabular}
+  \egroup
+  \centering
+  \begin{mathpar}
+    \inferrule{x \in V}{V \vdash \isTerm{x}}\,\Rule{Var}
+    \and
+    \inferrule{V \vdash \isTerm{t} \\ \isTerm{A}\text{ is a closed type}}{V \vdash \isTerm{(t \annotate A)}}\,\Rule{Anno}
+    \and
+    \inferrule{V, \isTerm{x} \vdash \isTerm{t}}{V \vdash \isTerm{\lam{x}t}}\,\Rule{Abs}
+    \and
+    \inferrule{V \vdash \isTerm{t} \\ V \vdash \isTerm{u}}{V \vdash \isTerm{t\;u}}\,\Rule{App}
+  \end{mathpar}
+  \caption{Raw terms for simply typed $\lambda$-calculus}
+  \label{fig:STLC-raw-terms}
+\end{figure}
+
+\begin{figure}
+  \small
+  \bgroup
+  \renewcommand{\arraystretch}{1.5}
+  \begin{tabular}{ r l }
+    $\boxed{\Gamma |- \isTerm{t} : A}$ & A raw term~$t$ has type~$A$ under context~$\Gamma$
+  \end{tabular}
+  \egroup
+  \centering
+  \begin{mathpar}
+    \inferrule{(x : A) \in \Gamma}{\Gamma \vdash \isTerm{x} : A}\,\Rule{Var}
+    \and
+    \inferrule{\Gamma \vdash \isTerm{t} : A}{\Gamma \vdash \isTerm{(t \annotate A)} : A}\,\Rule{Anno}
+    \and
+    \inferrule{\Gamma, \isTerm{x} : A \vdash \isTerm{t} : B}{\Gamma \vdash \isTerm{\lam{x}t} : A \bto B}\,\Rule{Abs}
+    \and
+    \inferrule{\Gamma \vdash \isTerm{t} : A \bto B \\ \Gamma \vdash \isTerm{u} : A}{\Gamma \vdash \isTerm{t\;u} : B}\,\Rule{App}
+  \end{mathpar}
+  \caption{Typing derivations for simply typed $\lambda$-calculus}
+  \label{fig:STLC-typing-derivations}
+\end{figure}
+
+We start with an overview of our theory by instantiating it to simply typed $\lambda$-calculus.
+Roughly speaking, the problem of type synthesis requires us to take an untyped abstract syntax tree ---which we call a \emph{raw term}--- as input, and produce a typing derivation for the term if possible.
+To give more precise definitions:
+The raw terms for simply typed $\lambda$-calculus are defined in \cref{fig:STLC-raw-terms};%
+\footnote{We omit the usual conditions about named representations of variables throughout this paper.}
+besides the standard constructs, there is an \textsc{Anno} rule that allows the user to insert type annotations to help with type synthesis.
+Correspondingly, the definition of typing derivations in \cref{fig:STLC-typing-derivations} also includes an \textsc{Anno} rule enforcing that the type of an annotated term does match the annotation.
+Now we can define what it means to solve the type synthesis problem.
+
+\begin{definition}
+\label{def:STLC-type-synthesiser}
+Parametrised by an `excuse' predicate~$E$ on raw terms, a \emph{type synthesiser} takes a context~$\Gamma$ and a raw term $|\Gamma| \vdash t$ (where $|\Gamma|$~is the list of variables appearing in~$\Gamma$) as input, and establishes one of the following outcomes:
+\begin{enumerate}
+\item there exists a derivation of $\Gamma \vdash t : A$ for some type~$A$,
+\item there does not exist a derivation $\Gamma \vdash t : A$ for any type~$A$, or
+\item $E$~holds for~$t$.
+\end{enumerate}
+\end{definition}
+
+It is crucial to allow the third outcome, without which we would be requiring the type synthesis problem to be decidable, but this requirement would quickly become impossible to meet when the theory is extended to handle more complex types.
+If a type synthesiser cannot decide whether there is a typing derivation, it is allowed to give an excuse instead of an answer.
+Acceptable excuses are defined by the predicate~$E$, which describes what is wrong with an input term, for example not having enough type annotations.
+
+\Josh{Checking is subsumed by synthesis because of type annotations}
+
+\begin{figure}
+  \small
+  \bgroup
+  \renewcommand{\arraystretch}{1.5}
+  \begin{tabular}{ r l }
+    $\boxed{\Gamma |- \isTerm{t} \syn A}$ & A raw term $t$ synthesises a type $A$ under $\Gamma$ \\
+    $\boxed{\Gamma |- \isTerm{t} \chk A}$ &A raw term $t$ checks against a type $A$ under $\Gamma$
   \end{tabular}
@@
     \and
-    \inferrule{\Gamma \vdash \isTerm{t} \chk A}{\Gamma \vdash \isTerm{t \annotate A}\syn A}\,\SynRule{Anno}
+    \inferrule{\Gamma \vdash \isTerm{t} \chk A}{\Gamma \vdash \isTerm{(t \annotate A)}\syn A}\,\SynRule{Anno}
     \and
-    \inferrule{\Gamma \vdash \isTerm{t} \syn B \\ B = A}{\Gamma \vdash \isTerm{t} \chk A}\,\ChkRule{Sub}
+    \inferrule{\Gamma \vdash \isTerm{t} \syn A \\ A = B}{\Gamma \vdash \isTerm{t} \chk B}\,\ChkRule{Sub}
+    \and
+    \inferrule{\Gamma, \isTerm{x} : A \vdash \isTerm{t} \chk B}{\Gamma \vdash \isTerm{\lam{x}t} \chk A \bto B}\,\ChkRule{Abs}
     \and
     \inferrule{\Gamma \vdash \isTerm{t} \syn A \bto B \\ \Gamma \vdash \isTerm{u} \chk A}{\Gamma \vdash \isTerm{t\;u} \syn B}\,\SynRule{App}
+  \end{mathpar}
+  \caption{Bidirectional typing derivations for simply typed $\lambda$-calculus}
+  \label{fig:STLC-bidirectional-typing-derivations}
+\end{figure}
+
+Now our goal is to use \cref{def:STLC-type-synthesiser} as a specification and implement it using a \emph{bidirectional} type synthesiser, which attempts to produce \emph{bidirectional} typing derivations defined in \cref{fig:STLC-bidirectional-typing-derivations}.
+Here we briefly recap the basic ideas of bidirectional type synthesis.
+While the problem of type synthesis is not decidable in general, for certain kinds of terms it is still possible to synthesise their types --- for example, the $\SynRule{Var}$ rule says that the type of a variable can be looked up in the context.
+For other kinds of terms, we can switch to the simpler problem of type \emph{checking}, where the expected type of a term is given as input so that the type checker has more information to work with.
+A bidirectional type synthesiser switches between the two \emph{modes}: synthesising wherever possible, and checking otherwise.
+If a bidirectional type system is designed well (i.e.~\emph{mode-correct}, which will be defined in \cref{sec:type-synthesis}), then there will be a flow of type information in each rule that allows us to determine unknown types (e.g.~types to be checked) from known ones (e.g.~types previously synthesised), and suggests how bidirectional type synthesis is performed --- for example, the $\SynRule{App}$ rule says that to synthesise the type of an application $\isTerm{t\;u}$, we first synthesise the type of~$\isTerm{t}$, which should be an implication $A \bto B$, from which we can extract the expected type of~$\isTerm{u}$, namely~$A$, and perform checking; then the type of the whole application, namely~$B$, can also be extracted from the previously synthesised type $A \bto B$.
+%The $\ChkRule{Abs}$ rule checks whether the type of an abstraction $\isTerm{\lam{x}t}$ is $A \bto B$ by checking whether $\isTerm{t}$~has type~$B$ assuming $\isTerm{x}$~has type~$A$ in the context, where both types $A$~and~$B$ can be determined from the type $A \bto B$ given initially.
+%For an annotated term $t \bbcolon A$, the $\SynRule{Anno}$ rule simply synthesises~$A$ as its type, provided that $t$~can be successfully checked to have type~$A$.
+%Finally, the subsumption rule $\ChkRule{Sub}$ says that a term synthesising a type~$A$ can be checked to have type~$B$ if $A = B$ --- in general this type equality can be replaced by some subtyping relation.
+
+\Josh{There are other ways to correctly bidirectionalise the type system}
+
+\begin{figure}
+  \centering
+  \small
+  \judgbox{|- \isTerm{t}^\dir{d}}{A raw term~$\isTerm{t}$ is in mode~$d$}
+  \begin{mathpar}
+    \inferrule{x \in V}{V |- \isTerm{x}^{\syn}}\,\SynRule{Var}
     \and
-    \inferrule{\Gamma, \isTerm{x} : A \vdash \isTerm{t} \chk B}{\Gamma \vdash \isTerm{\lam{x}t} \chk A \bto B}\,\ChkRule{Abs}
+    \inferrule{V |- \isTerm{t}^{\chk}}{V |- (\isTerm{t \annotate A})^{\syn}}\,\SynRule{Anno}
+    \and
+    \inferrule{V |- \isTerm{t}^{\syn}}{V |- \isTerm{t}^{\chk}}\,\ChkRule{Sub}
+    \\
+    \inferrule{V, x |- \isTerm{t}^{\chk}}
+    {V |- (\isTerm{\lam{x}t})^{\chk}}\,\ChkRule{Abs}
+    \and
+    \inferrule{V |- \isTerm{t}^{\syn} \quad V |- \isTerm{u}^{\chk}}
+    {V |- (\isTerm{t\;u})^{\syn}}\,\SynRule{App}
   \end{mathpar}
-  \caption{Bidirectional simply typed $\lambda$-calculus $\Lambda^{\Leftrightarrow}_{\bto}$}
-  \label{fig:bi-stlc}
+  \caption{Mode derivations for simply typed $\lambda$-calculus}
+  \label{fig:STLC-mode-derivations}
 \end{figure}
-\begin{remark}
-  We avoid using the term `function type' and its conventional notation $\to$ at the object level on purpose, as it may be confused with function types in our type-theoretic meta-language.
-\end{remark}
+
+While it is possible for a bidirectional type synthesiser to do its job in one go, which can be thought of as adding both mode and typing information to a raw term and arriving at a bidirectional typing derivation, it is beneficial to have a preprocessing step which adds only mode information, based on which the synthesiser then continues to add typing information.
+More precisely, the preprocessing step attempts to produce \emph{mode derivations} as defined in \cref{fig:STLC-mode-derivations}, where the rules are exactly the mode part of the bidirectional typing rules~(\cref{fig:STLC-bidirectional-typing-derivations}).
+
+\begin{definition}
+\label{def:STLC-mode-preprocessor}
+A \emph{mode preprocessor} decides for any raw term~$V \vdash \isTerm{t}$ and mode~$\dir{d}$ whether $V \vdash \isTerm{t}^\dir{d}$.
+\end{definition}
+
+One (smaller) benefit of mode preprocessing is that it helps to simplify the synthesiser, whose computation can be partly directed by a mode derivation.
+More importantly, whether there is a mode derivation for a term is actually very useful information to the user, because it corresponds to whether the term has enough type annotations:
+Observe that the $\SynRule{Anno}$ and $\ChkRule{Sub}$ rules allow us to switch between the synthesising and checking modes;
+the switch from synthesising to checking is free, whereas the opposite direction requires a type annotation.
+That is, any term in synthesising mode is also in checking mode, but not necessarily vice versa.
+A type annotation is required wherever a term that can only be in checking mode is required to be in synthesising mode, and a term does not have a mode derivation if and only if type annotations are missing in such places.
+(We will give a more rigorous treatment in \cref{sec:pre-synthesis}.)
+For example, an abstraction is strictly in checking mode, but the left sub-term of an application has to be synthesising, so the term $\isTerm{(\lam{x}t)\;u}$ does not have a mode derivation unless we annotate the abstraction.
+
+\Josh{Determining whether there is enough type annotation with mode information only (not involving typing)}
+
+Perhaps most importantly, mode derivations enable us to give bidirectional type synthesisers a tight definition: if we restrict the domain of a synthesiser to terms in synthesising mode (i.e.~having enough type annotations for synthesis), then it is possible for the synthesiser to \emph{decide} whether there is a suitable bidirectional typing derivation.
+
+\begin{definition}
+\label{def:STLC-bidirectional-type-synthesiser}
+A \emph{bidirectional type synthesiser} decides for any context~$\Gamma$ and raw term $|\Gamma| \vdash \isTerm{t}$ such that $|\Gamma| \vdash \isTerm{t}^{\syn}$ whether $\Gamma \vdash \isTerm{t} \syn A$ for some type~$A$.
+\end{definition}
+
+Now we can get back to our goal of implementing a type synthesiser~(\cref{def:STLC-type-synthesiser}).
+
+\begin{theorem}
+A type synthesiser that uses `not in synthesising mode' as its excuse can be constructed from a mode preprocessor and a bidirectional type synthesiser.
+\end{theorem}
+
+The construction is straightforward:
+Run the mode preprocessor on the input term~$|\Gamma| \vdash t$.
+If there is no synthesising mode derivation, report that $t$~is not in synthesising mode (the third outcome).
+Otherwise $|\Gamma| \vdash t^{\syn}$, and we can run the bidirectional type synthesiser.
+If it finds a derivation of $\Gamma \vdash t \syn A$ for some type~$A$, return a derivation of $\Gamma \vdash t : A$ (the first outcome), which is possible because the bidirectional typing~(\cref{fig:STLC-bidirectional-typing-derivations}) is \emph{sound} with respect to the original typing~(\cref{fig:STLC-typing-derivations}).
+
+\begin{lemma}[Soundness]
+If\/ $\Gamma \vdash t :^\dir{d} A$, then $\Gamma \vdash t : A$.
+\end{lemma}
+
+If there is no derivation of $\Gamma \vdash t \syn A$ for any type~$A$, report that there is no derivation of $\Gamma \vdash t : A$ for any~$A$ either (the second outcome), because the bidirectional typing is \emph{complete} with respect to the original typing.
+
+\begin{lemma}[Completeness]
+If\/ $|\Gamma| \vdash t^\dir{d}$ and\/ $\Gamma \vdash t : A$, then $\Gamma \vdash t :^\dir{d} A$.
+\end{lemma}
+
+\LT{We avoid using the term `function type' and its conventional notation $\to$ at the object level on purpose, as it may be confused with function types in our type-theoretic meta-language.}

### 2023-06-22 tex/sec2-outline.tex
[ tex ] Done with S2 for now

@@
 Correspondingly, the definition of typing derivations in \cref{fig:STLC-typing-derivations} also includes an \textsc{Anno} rule enforcing that the type of an annotated term does match the annotation.
+%\footnote{The presence of type annotations allows type synthesis to subsume type checking: to check whether a term~$t$ has type~$A$, synthesise the type of $t \bbcolon A$.}
 Now we can define what it means to solve the type synthesis problem.
@@
 Acceptable excuses are defined by the predicate~$E$, which describes what is wrong with an input term, for example not having enough type annotations.
-
-\Josh{Checking is subsumed by synthesis because of type annotations}
 
@@
 While the problem of type synthesis is not decidable in general, for certain kinds of terms it is still possible to synthesise their types --- for example, the $\SynRule{Var}$ rule says that the type of a variable can be looked up in the context.
-For other kinds of terms, we can switch to the simpler problem of type \emph{checking}, where the expected type of a term is given as input so that the type checker has more information to work with.
-A bidirectional type synthesiser switches between the two \emph{modes}: synthesising wherever possible, and checking otherwise.
+For other kinds of terms, we can switch to the simpler problem of type \emph{checking}, where the expected type of a term is given as input so that there is more information to work with.
+When a bidirectional type synthesiser traverses a raw term, it switches between the two \emph{modes}, synthesising wherever possible, and checking otherwise.
 If a bidirectional type system is designed well (i.e.~\emph{mode-correct}, which will be defined in \cref{sec:type-synthesis}), then there will be a flow of type information in each rule that allows us to determine unknown types (e.g.~types to be checked) from known ones (e.g.~types previously synthesised), and suggests how bidirectional type synthesis is performed --- for example, the $\SynRule{App}$ rule says that to synthesise the type of an application $\isTerm{t\;u}$, we first synthesise the type of~$\isTerm{t}$, which should be an implication $A \bto B$, from which we can extract the expected type of~$\isTerm{u}$, namely~$A$, and perform checking; then the type of the whole application, namely~$B$, can also be extracted from the previously synthesised type $A \bto B$.
+
 %The $\ChkRule{Abs}$ rule checks whether the type of an abstraction $\isTerm{\lam{x}t}$ is $A \bto B$ by checking whether $\isTerm{t}$~has type~$B$ assuming $\isTerm{x}$~has type~$A$ in the context, where both types $A$~and~$B$ can be determined from the type $A \bto B$ given initially.
@@
 %Finally, the subsumption rule $\ChkRule{Sub}$ says that a term synthesising a type~$A$ can be checked to have type~$B$ if $A = B$ --- in general this type equality can be replaced by some subtyping relation.
-
-\Josh{There are other ways to correctly bidirectionalise the type system}
 
@@
 A type annotation is required wherever a term that can only be in checking mode is required to be in synthesising mode, and a term does not have a mode derivation if and only if type annotations are missing in such places.
-(We will give a more rigorous treatment in \cref{sec:pre-synthesis}.)
+(We will treat all these more rigorously in \cref{sec:pre-synthesis}.)
 For example, an abstraction is strictly in checking mode, but the left sub-term of an application has to be synthesising, so the term $\isTerm{(\lam{x}t)\;u}$ does not have a mode derivation unless we annotate the abstraction.
 
-\Josh{Determining whether there is enough type annotation with mode information only (not involving typing)}
-
-Perhaps most importantly, mode derivations enable us to give bidirectional type synthesisers a tight definition: if we restrict the domain of a synthesiser to terms in synthesising mode (i.e.~having enough type annotations for synthesis), then it is possible for the synthesiser to \emph{decide} whether there is a suitable bidirectional typing derivation.
+Perhaps most importantly, mode derivations enable us to give bidirectional type synthesisers a tight definition: if we restrict the domain of a synthesiser to terms in synthesising mode (i.e.~having enough type annotations for performing synthesis), then it is possible for the synthesiser to \emph{decide} whether there is a suitable typing derivation.
 
@@
 \label{def:STLC-bidirectional-type-synthesiser}
-A \emph{bidirectional type synthesiser} decides for any context~$\Gamma$ and raw term $|\Gamma| \vdash \isTerm{t}$ such that $|\Gamma| \vdash \isTerm{t}^{\syn}$ whether $\Gamma \vdash \isTerm{t} \syn A$ for some type~$A$.
+A \emph{bidirectional type synthesiser} decides for any context~$\Gamma$ and synthesising term $|\Gamma| \vdash \isTerm{t}^{\syn}$ whether $\Gamma \vdash \isTerm{t} \syn A$ for some type~$A$.
 \end{definition}
@@
 
+We will construct a mode preprocessor~(\cref{sec:mode-preprocessing}) and a bidirectional type synthesiser~(\cref{sec:type-synthesis}) and prove both lemmas~(\cref{sec:soundness-and-completeness}) for all syntax-directed bidirectional simple type systems.
+To quantify over all such systems, we need their generic definitions, which we formulate in \cref{sec:defs}.
+
 \LT{We avoid using the term `function type' and its conventional notation $\to$ at the object level on purpose, as it may be confused with function types in our type-theoretic meta-language.}

### 2023-06-22 tex/sec2-outline.tex
[ tex ] Minor revisions in S2

@@
     \and
-    \inferrule{V \vdash \isTerm{t} \\ \isTerm{A}\text{ is a closed type}}{V \vdash \isTerm{(t \annotate A)}}\,\Rule{Anno}
+    \inferrule{V \vdash \isTerm{t}}{V \vdash \isTerm{(t \annotate A)}}\,\Rule{Anno}
     \and
@@
 To give more precise definitions:
-The raw terms for simply typed $\lambda$-calculus are defined in \cref{fig:STLC-raw-terms};%
+The raw terms%
 \footnote{We omit the usual conditions about named representations of variables throughout this paper.}
+for simply typed $\lambda$-calculus are defined in \cref{fig:STLC-raw-terms};
 besides the standard constructs, there is an \textsc{Anno} rule that allows the user to insert type annotations to help with type synthesis.
-Correspondingly, the definition of typing derivations in \cref{fig:STLC-typing-derivations} also includes an \textsc{Anno} rule enforcing that the type of an annotated term does match the annotation.
+Correspondingly, the definition of typing derivations%
+\footnote{We write~`$\bto$' instead of~`$\to$' for the function types of simply typed $\lambda$-calculus to avoid confusion with the function types in our type-theoretic meta-language.}
+in \cref{fig:STLC-typing-derivations} also includes an \textsc{Anno} rule enforcing that the type of an annotated term does match the annotation.
 %\footnote{The presence of type annotations allows type synthesis to subsume type checking: to check whether a term~$t$ has type~$A$, synthesise the type of $t \bbcolon A$.}
@@
 When a bidirectional type synthesiser traverses a raw term, it switches between the two \emph{modes}, synthesising wherever possible, and checking otherwise.
-If a bidirectional type system is designed well (i.e.~\emph{mode-correct}, which will be defined in \cref{sec:type-synthesis}), then there will be a flow of type information in each rule that allows us to determine unknown types (e.g.~types to be checked) from known ones (e.g.~types previously synthesised), and suggests how bidirectional type synthesis is performed --- for example, the $\SynRule{App}$ rule says that to synthesise the type of an application $\isTerm{t\;u}$, we first synthesise the type of~$\isTerm{t}$, which should be an implication $A \bto B$, from which we can extract the expected type of~$\isTerm{u}$, namely~$A$, and perform checking; then the type of the whole application, namely~$B$, can also be extracted from the previously synthesised type $A \bto B$.
+If a bidirectional type system is designed well (i.e.~\emph{mode-correct}, which will be defined in \cref{sec:type-synthesis}), then there will be a flow of type information in each rule that allows us to determine unknown types (e.g.~types to be checked) from known ones (e.g.~types previously synthesised), and suggests how bidirectional type synthesis is performed --- for example, the $\SynRule{App}$ rule says that to synthesise the type of an application $\isTerm{t\;u}$, we first synthesise the type of~$\isTerm{t}$, which should have the form $A \bto B$, from which we can extract the expected type of~$\isTerm{u}$, namely~$A$, and perform checking; then the type of the whole application, namely~$B$, can also be extracted from the previously synthesised type $A \bto B$.
 
@@
     \and
-    \inferrule{V |- \isTerm{t}^{\syn} \quad V |- \isTerm{u}^{\chk}}
+    \inferrule{V |- \isTerm{t}^{\syn} \\ V |- \isTerm{u}^{\chk}}
     {V |- (\isTerm{t\;u})^{\syn}}\,\SynRule{App}
@@
 To quantify over all such systems, we need their generic definitions, which we formulate in \cref{sec:defs}.
-
-\LT{We avoid using the term `function type' and its conventional notation $\to$ at the object level on purpose, as it may be confused with function types in our type-theoretic meta-language.}

### 2023-06-22 tex/sec1-intro.tex
[ tex ] Done with S2 for now
[ tex ] S1

@@
 
-%  Bidirectional typing had only been a set of principles and characteristics to design typing rules so that a type synthesis algorithm can be `read off' easily
-  %, subsuming soundness, completeness, and decidability of type synthesis algorithm.
-%  Our work addresses the problem with existing methodologies, setting the stage for further advances towards a more general theory of type synthesis.
+Type checking and inference serve as the transition from parsed abstract syntax trees, referred to as \emph{raw terms}, to well-typed programs, known as \emph{well-typed terms}.
+Initially, type inference techniques focused on complete type inference, aiming to infer the type of any unannotated raw term.
+However, it was soon realised that once polymorphism is introduced, complete type inference becomes undecidable~\citep{Wells1999}.
 
-  %In short, this paper is a small but first step towards a general theory of type checking and inference.
-  %By formalizing our theory of bidirectional type synthesise in \Agda, we provide a framework capable of handling bidirectional type systems of simple types uniformly.
-  %Our work addresses the problem with existing methodologies, 
-Points to elaborate:
-\begin{enumerate}
-  \item Type inference and checking as a transition from (parsed) abstract syntax trees to well-typed terms.
-  \item Bidirectional typing is by now the only solution that scales well.
-  \item Not yet a general theory but by now a design principle.
-    No `type checker generator' as each type checker is implemented on a case-by-case basis, but an algorithm can usually be read off from syntax-directed typing rules.
-    A hint to a general pattern in syntax-directedness.
-  \item The existing work of bidirectional typing does not take account of practical aspects such as missing annotation.
-    Lightly or rarely discussed.
-  \item A meta-problem: existing work is based on classical logic (or left unspecified at least).
-    Soundness of completeness of a type checking algorithm in a less straightforward way.
-  \item A review of what has been done towards this line of research.
-  \item Finally, our responses (contributions of this paper) and some example.
-\end{enumerate}
+In light of this, \citet{Pierce2000} argued that type annotations are valuable for programmers in understanding the purpose of a program.
+The prevalence of type annotations in current software engineering practices supports this claim.
+Acknowledging the undecidability of complete type inference, the concept of bidirectional type synthesis emerged as a decidable algorithm to infer the type of suitably annotated programs in polymorphic programming languages.
 
-\begin{enumerate}
-  \item Use application in spin form to justify our design choice: type expression of arbitrary rank and the number of constructs in a calculus.
-  \item Argue that syntax-directed form is essential and practical.
-  \item Pfenning's recipe is only of design guide not a technical requirement (there are many exceptions).
-\end{enumerate}
+Bidirectional typing~\citep{Pierce2000} extends the notion of typing rules by introducing two modes that govern the flow of types, thus earning its name.
+This shift from (complete) type inference to type synthesis allows the use of existing type annotations to guide the flow of type information and determine the type of a term.
+This is achieved through typing judgements in two modes:
+  \[
+    (\text{synthesising})\quad \Gamma \vdash \isTerm{t} \syn A
+    \quad\text{and}\quad
+    (\text{checking})\quad \Gamma \vdash \isTerm{t} \chk A
+  \]
+  In bidirectional type synthesis, the type of a term can be synthesized from the information in its context and the term itself or it can be checked against a given type depending on \emph{the mode of a raw term}.
+For example, it may involve looking up the variable in its context; using a type annotation provided in the term after checking against the given type first, ensuring the validity of the provided annotation.
+An algorithm for bidirectional type synthesis can often be `read off' directly from a calculus presented in bidirectional typing, provided that it is syntax-directed and mode-correct, as further explained later.
+
+Bidirectional type synthesis is not limited to polymorphic languages; it is also applicable to dependently typed languages and has become an integral part of type checkers for modern programming languages.
+A recent survey has summarised the design principles of bidirectional typing and progress so far on bidirectional typing.
+ In fact, bidirectional type synthesis has demonstrated such significance that certain traditional techniques like unification are no longer as essential in comparison.
+ As a result, bidirectional type synthesis can be regarded as the cornerstone of type-checking technology.
+
+Bidirectional typing has proven to be an valuable set of design principles.
+However, unlike parsing, which has both theoretical foundations and practical implementations, bidirectional typing has been developed on a case-by-case basis, lacking a general theory.
+While it is rather easy to implement a specific type synthesizer using bidirectional typing, a theory that provides explicit specifications with rigorously proven properties is still absent.
+Instead, we can only offer general expectations and considerations when designing bidirectional typing for a specific system, as summarised by~\citet{Dunfield2021}.
+Consequently, there is no readily available `type-checker generator', and each type checker based on bidirectional typing must be implemented independently.
+
+In light of the extensive adoption and scalability of bidirectional typing in programming languages, this paper aims to initiate a general theory of bidirectional typing and type synthesis. 
+We will explore the key characteristics of bidirectional typing~\citep{Dunfield2021}, which are sufficient to derive a sound, complete, and decidable type synthesis algorithm for simple type theories.
+Our approach involves constructing the theory within Martin-L\"of type theory and formalising it using the \Agda proof assistant.
+This formalisation provides a generic type synthesis algorithm that is applicable to all calculi employing bidirectional typing for simple types, hence essentially a verified type-checker generator.
+By rigorously defining its core principles, we aim to lay the foundation for further advancements in bidirectional typing.
 
 \subsection{Related work}
+Before proceeding with the main development, let us review what has been done towards this line of research.
 
@@
 
+%  Our work addresses the problem with existing methodologies, setting the stage for further advances towards a more general theory of type synthesis.
+
+  %In short, this paper is a small but first step towards a general theory of type checking and inference.
+  %By formalizing our theory of bidirectional type synthesise in \Agda, we provide a framework capable of handling bidirectional type systems of simple types uniformly.
+  %Our work addresses the problem with existing methodologies,
+
+\LT{
+Points to elaborate:
+\begin{enumerate}
+  \item The existing work of bidirectional typing does not take account of practical aspects such as missing annotation.
+    Lightly or rarely discussed.
+  \item A meta-problem: existing work is based on classical logic (or left unspecified at least).
+    Soundness of completeness of a type checking algorithm in a less straightforward way.
+  \item A review of what has been done towards this line of research.
+  \item Finally, our responses (contributions of this paper) and some example.
+\end{enumerate}
+Assumption to justify:
+\begin{enumerate}
+  \item Use application in spin form to justify our design choice: type expression of arbitrary rank and the number of constructs in a calculus.
+  \item Argue that syntax-directed form is essential and practical.
+  \item Pfenning's recipe is only of design guide not a technical requirement (there are many exceptions).
+\end{enumerate}
+}
 Our restrictions:
@@
     \LT{4 pp}
-  \item Soundness, completeness, bidirectionalisation, and annotatability (\Cref{sec:annotatability})
+  \item Soundness, completeness, bidirectionalisation, and annotatability (\Cref{sec:pre-synthesis})
     \Josh{Erasure from bidirectional typing derivation to raw terms with a mode}

### 2023-06-23 tex/BiSig.tex
[ tex ] Most of S2
[ tex ] S1: related work on language-formalisation frameworks

@@
 
-%\usepackage[inline]{enumitem} % for environment enumerate*
+\usepackage[inline]{enumitem} % for environment enumerate*
 %\setlist[enumerate]{mode=unboxed}
@@
 
-\newcommand{\LT}[1]{\todo[author=LT,inline,color=orange!40,caption={}]{{#1}}}
-\newcommand{\Josh}[1]{\todo[author=Josh,inline,color=yellow!40,caption={}]{{#1}}}
+\newcommand{\LT}[1]{\todo[author=LT,inline,color=orange!40]{{#1}}}
+\newcommand{\Josh}[1]{\todo[author=Josh,inline,color=yellow!40]{{#1}}}
 
@@
 
-  In response to this difficulty, we develop a \emph{simple} yet \emph{general} and \emph{constructive} theory of bidirectional type synthesis where \emph{bidirectional type synthesis} refers to the mutual use of type checking and inference based on bidirectional typing rules to determine the type of a program with type annotations.
+  In response to this difficulty, we develop a simple yet general and constructive theory of bidirectional type synthesis where bidirectional type synthesis refers to the mutual use of type checking and inference based on bidirectional typing rules to determine the type of a program with type annotations.
   First, our theory is simple yet general as it deals with any simple type theory and its bidirectional type system that can be specified by a signature, analogous to the role of grammar in parsing.
@@
 \input{sec3-defs.tex}
-\input{sec4-bidirectionalisation.tex}
+\input{sec4-pre-synthesis.tex}
 \input{sec5-synthesis.tex}

### 2023-06-23 tex/sec1-intro.tex
[ tex ] S1: related work on language-formalisation frameworks

--- 2023-06-20 tex/sec1-intro.tex
@@
 
-%  Bidirectional typing had only been a set of principles and characteristics to design typing rules so that a type synthesis algorithm can be `read off' easily
-  %, subsuming soundness, completeness, and decidability of type synthesis algorithm.
-%  Our work addresses the problem with existing methodologies, setting the stage for further advances towards a more general theory of type synthesis.
+Type checking and inference serve as the transition from parsed abstract syntax trees, referred to as \emph{raw terms}, to well-typed programs, known as \emph{well-typed terms}.
+Type inference algorithms were developed for determining the type of any unannotated raw term, but it was discovered that full parametric polymorphism makes the type inference problem undecidable~\citep{Wells1999} as well as dependent types~\citep{Dowek1993}.
+Recognising the limitations of type inference, the concept of bidirectional type synthesis based on bidirectional typing emerged as an alternative approach, offering decidable algorithms for determining the types of \emph{suitably annotated} programs in programming languages with polymorphic types, dependent types, gradual types, and more.
 
-  %In short, this paper is a small but first step towards a general theory of type checking and inference.
-  %By formalizing our theory of bidirectional type synthesise in \Agda, we provide a framework capable of handling bidirectional type systems of simple types uniformly.
-  %Our work addresses the problem with existing methodologies, 
-Points to elaborate:
-\begin{enumerate}
-  \item Type inference and checking as a transition from (parsed) abstract syntax trees to well-typed terms.
-  \item Bidirectional typing is by now the only solution that scales well.
-  \item Not yet a general theory but by now a design principle.
-    No `type checker generator' as each type checker is implemented on a case-by-case basis, but an algorithm can usually be read off from syntax-directed typing rules.
-    A hint to a general pattern in syntax-directedness.
-  \item The existing work of bidirectional typing does not take account of practical aspects such as missing annotation.
-    Lightly or rarely discussed.
-  \item A meta-problem: existing work is based on classical logic (or left unspecified at least).
-    Soundness of completeness of a type checking algorithm in a less straightforward way.
-  \item A review of what has been done towards this line of research.
-  \item Finally, our responses (contributions of this paper) and some example.
-\end{enumerate}
+The core idea of bidirectional type synthesis is simple.
+First we start with bidirectional typing which extends typing judgements with two modes that govern the flow of types, hence its name.
+That is, typing judgements are classified into two forms:
+  \[
+    (\text{synthesising})\quad \Gamma \vdash \isTerm{t} \syn A
+    \quad\text{and}\quad
+    (\text{checking})\quad \Gamma \vdash \isTerm{t} \chk A
+  \]
+where the type of a term can be synthesised as the output from the information in its context and the term itself as inputs; it can be checked against a type additionally as an input, depending on \emph{the mode of a raw term}.
+Then, an algorithm of bidirectional type synthesis can be `read off' based on how type information flows, if the output can be determined by inputs without guessing by design.
+  For example, the type of a raw term $\isTerm{t}$ may be synthesised by looking up the variable in its context if $\isTerm{t}$ is a variable or by using a type annotation $A$ provided in the term if $\isTerm{t}$ is of the form $t \annotate A$ after checking $\isTerm{t}$ against the given type~$A$ first, ensuring the validity of the provided annotation.
 
-\begin{enumerate}
-  \item Use application in spin form to justify our design choice: type expression of arbitrary rank and the number of constructs in a calculus.
-  \item Argue that syntax-directed form is essential and practical.
-  \item Pfenning's recipe is only of design guide not a technical requirement (there are many exceptions).
-\end{enumerate}
+Type annotations may be seen burdensome at first, but in fact they are useful for programmers to understand the purpose of a program.
+This claim was put forth by~\citet{Pierce2000} and supported by software engineering practices, since type annotations are often included as part of top-level definitions and of the documentation even for languages such as variants of Hindley--Damas--Milner type system where (complete) type inference is possible.
+As a result, the focus of type-checking technology has shifted from type inference to bidirectional type synthesis, considering both theoretical limitations and practical considerations.
+
+  A recent survey paper by~\citet{Dunfield2021} summarised the design principles of bidirectional typing and its coverage of applications so far.
+  Bidirectional type synthesis does not require unification---the traditional technique used in type inference.
+Also argued by~\citet{Pierce2000}, missing annotations in bidirectional type synthesis can be recovered using the local information only, so introducing long-distance unification constraints to bidirectional type synthesis actually breaks its premise of being local.
+ We may rightly claim that bidirectional type synthesis is the only approach that works in theory and in practice for a variety of languages.
+
+Unlike parsing, however, which has theory and practice that work for a wide class of languages, bidirectional typing has only been developed on a case-by-case basis, lacking a general theory.
+Although it is rather easy to `read off' an algorithm of bidirectional type synthesis, a theory that provides explicit specifications with rigorously proven properties is still absent.
+By now, we can only offer informal design principles and considerations for individual systems~\citep{Dunfield2021}.
+Likewise, there is no off-the-shelf `type-checker generator', and each type checker based on bidirectional typing has to be implemented independently.
+
+To address the aforementioned issue, this paper aims to develop a theory of bidirectional type synthesis and a type-checker generator.
+While competing ways exist for designing typing rules, such as those related to polymorphic types by~\citet{Pierce2000,Peyton-Jones2007,Dunfield2013,Xie2018} to name a few, the derivation of an algorithm is relatively standard once a set of bidirectional typing rules are in place.
+Therefore, the goal of our theory is not to formulate bidirectional typing for different language features, but rather to define the `grammar' of bidirectional typing rules, referred to as the \emph{bidirectional binding signature} in this paper, along with its essential criteria including \emph{mode-correctness}, \emph{completeness}, and \emph{annotatability} in bidirectional typing that are adequate for deriving a type synthesis algorithm as outlined informally \emph{op.\ cit}.
+To illustrate the core idea, we confine our discussion to simple type theories and their bidirectional type systems that have syntax-directed typing rules for simplicity, while more advanced type theories are left for future work.
 
 \subsection{Related work}
+As we will mechanise our theory not only for all simple type theories that can be specified by some signature but also with perspectives and techniques different from existing approaches in bidirectional typing, it is timely to take a review of the related work upon which this work is built.
 
-\citep{Xie2018}
 \paragraph{Language formalization and its frameworks}
-\cite{Wadler2022}
-PoplMark Challenge~\citep{Aydemir2005}
+Formalising the metatheory of a programming language was initiated by the \PoplMark challenge~\citep{Aydemir2005} and this objective has been manifested in a textbook by~\citet{Wadler2022} where programming language concepts are all discussed formally in the proof assistant~\Agda, including bidirectional type synthesis for~\PCF.
 
-\cite{Cimini2020,Cimini2022}
+It was soon observed that rudimentary operations and meta-properties for variable binding and substitution constitute the bulk of language formalisation, so it is rather unwise to tackle language formalisation on a case-by-case basis.
+Hence, there are language-formalisation frameworks by~\citet{Ahrens2018,Fiore2022,Gheri2020,Ahrens2022,Allais2021} to establish substitution, term traversal, and their meta-properties once and for all supported abstract syntax with variable binding.
+One of the core ideas of these frameworks is the concept of \emph{descriptions} (coined by~\citet{Chapman2010} for datatype-generic programming) or \emph{binding signatures} (coined by~\citet{Aczel1978} in line with the term in universal algebra) for specifying the set of typing rules of a language.
+It is noteworthy that these frameworks only work for simple types at most and none of them can express polymorphic types nor dependent types.
 
-\citep{Ahrens2018,Fiore2022,Gheri2020,Ahrens2022}
-\cite{Allais2021}
+\begin{remark}\label{re:type-synthesis-as-decidability-proof}
+It is interesting to note that the algorithm of bidirectional type synthesis by~\citeauthor{Wadler2022} is formulated very differently from the literature:
+\begin{enumerate*}
+  \item the former algorithm is presented as a (logical) decidability proof of whether a bidirectionally decorated raw term can be bidirectionally typed or equivalently a dependently typed program that takes a raw term $\isTerm{t}$ and returns a typing derivation for $\isTerm{t}$ if there is one or the evidence of a contradiction otherwise;
+
+  \item an algorithm in the latter form is presented as \emph{algorithmic system} relations such as $\Gamma |- \isTerm{t} \syn A \mapsto \isTerm{t}'$, which is read that type annotations can be added to $\isTerm{t}$ in the external language to yield $\isTerm{t}'$ of type $A$ in the internal language, and accompanied with the \emph{soundness} and \emph{completeness} statements.
+\end{enumerate*}
+\end{remark}
 
@@
 
-\paragraph{Bidirectional typing}
-
-\cite{Pierce2000}
-\cite{Peyton-Jones2007}
-\cite{Dunfield2021}
-
 \subsection{Plan and Contributions of this paper}
 
-Our restrictions:
+Moreover, given this opportunity of re-examining elements of bidirectional typing, we begin our theory from a general idea of type synthesis independent of bidirectional type synthesis and spot subtleties that are rarely, if any, discussed in the literature on bidirectional typing.
+
+Our approach involves constructing the theory within Martin-L\"of type theory and formalising it using the \Agda proof assistant.
+This formalisation provides a generic type synthesis algorithm that is applicable to all calculi employing bidirectional typing for simple types, hence essentially a verified type-checker generator.
+By rigorously defining its core principles, we aim to lay the foundation for further advancements in bidirectional typing.
+
+%  Our work addresses the problem with existing methodologies, setting the stage for further advances towards a more general theory of type synthesis.
+
+  %By formalizing our theory of bidirectional type synthesise in \Agda, we provide a framework capable of handling bidirectional type systems of simple types uniformly.
+  %Our work addresses the problem with existing methodologies,
+
+Points to elaborate:
 \begin{enumerate}
-  \item Assumptions: Why only simple types? Various ways to deal with polymorphism; consistent on simple types.
-  \item Assumption: Syntax-directed typing rules
+  \item The existing work of bidirectional typing does not take account of practical aspects such as missing annotation.
+    Lightly or rarely discussed.
+  \item A meta-problem: existing work is based on classical logic (or left unspecified at least).
+    Soundness of completeness of a type checking algorithm in a less straightforward way.
+\end{enumerate}
+Assumption to justify:
+\begin{enumerate}
+  \item Use application in spin form to justify our design choice: type expression of arbitrary rank and the number of constructs in a calculus.
+  \item Argue that syntax-directed form is essential and practical.
+  \item Pfenning's recipe is only of design guide not a technical requirement (there are many exceptions).
 \end{enumerate}

### 2023-06-24 tex/sec4-pre-synthesis.tex
[ tex ] Most of S2
[ tex ] Done with S2 for now
[ tex ] S4 in progress

--- 2023-06-14 tex/sec4-bidirectionalisation.tex
@@
 
-\section{Soundness, Completeness, and Bidirectionalisation} \label{sec:annotatability}
+\section{Soundness, Completeness, and Mode Preprocessing}\label{sec:pre-synthesis}
 
 \subsection{Soundness and Completeness}
-\begin{theorem}[Soundness]\label{thm:term-soundness}
+\label{sec:soundness-and-completeness}
 
+\Josh{Throughout this sub-section, we quantify over any bidirectional type system~$(\Sigma, \Omega)$, context $\Gamma : \Cxt_\Sigma(\emptyset)$, raw term $\erase\Gamma \vdash_{\Sigma, \erase\Omega} t$, mode~$d$, and type $A : \Type_\Sigma(\emptyset)$.}
+
+Erasure of a bidirectional binding signature removes mode information and keeps everything else intact; this can be straightforwardly extended by induction to remove mode information from a bidirectional typing derivation and arrive at an ordinary typing derivation.
+
+\begin{lemma}[Soundness]\label{thm:term-soundness}
+If\/ $\Gamma \vdash_{\Sigma, \Omega} t :^\dir{d} A$, then $\Gamma \vdash_{\Sigma, \erase\Omega} t : A$.
+\end{lemma}
+
+\begin{proof}
+Induction on the given derivation, mapping every bidirectional typing rule to its mode-less counterpart except $\ChkRule{Sub}$, in which case the induction hypothesis $\Gamma |-_{\Sigma,\erase\Omega} t : B$ suffices due to the premise $B = A$.
+\end{proof}
+
+We can also remove typing and retain mode information, arriving at a mode derivation instead.
+
+\begin{proposition}\label{thm:typing-removal}
+If\/ $\Gamma \vdash_{\Sigma, \Omega} t :^\dir{d} A$, then $\erase\Gamma \vdash_{\Sigma, \Omega} t^\dir{d}$.
+\end{proposition}
+
+\begin{proof}
+Induction on the given derivation, mapping every rule to its counterpart.
+\end{proof}
+
+Conversely, if we have both mode and typing derivations for the same term, we can combine them and obtain a bidirectional typing derivation.
+
+\begin{lemma}[Completeness]\label{thm:term-completeness}
+If\/ $\erase\Gamma |-_{\Sigma, \Omega} t^\dir{d}$ and\/ $\Gamma |-_{\Sigma, \erase\Omega} t : A$, then $\Gamma |-_{\Sigma, \Omega} t :^\dir{d} A$.
+\end{lemma}
+
+\begin{proof}
+Induction on the given mode derivation.
+For \SynRule{Var}, \SynRule{Anno}, and \Rule{Op}, the outermost rule used in the given typing derivation must be the corresponding typing rule, so by the induction hypotheses we have bidirectional typing derivations for all the sub-terms, to which we can then apply the corresponding bidirectional typing rule.
+The \ChkRule{Sub} case is similar but slightly simpler: the induction hypothesis directly gives us a derivation of $\Gamma |-_{\Sigma,\Omega} t \syn A$, to which we apply $\ChkRule{Sub}$.
+\end{proof}
+
+In short, soundness and completeness are no more than the separation and combination of mode and typing information carried by the three kinds of derivations while keeping their basic structure, which is directed by the same raw term.
+
+\subsection{Mode Preprocessing}
+\label{sec:mode-preprocessing}
+
+\newcommand{\True}{\mathbf{T}}
+\newcommand{\False}{\mathbf{F}}
+
+\begin{figure}
+  \centering
+  \small
+  \begin{tabular}{ r r l }
+    & & is in (generalised) mode~$d$, \\
+    $\smash{\boxed{V |-_{\Sigma, \Omega} \isTerm{t}^{\dir{d}\,v\,e}}}$
+    & The raw term~$\isTerm{t}$\hspace{-.6em}
+    & misses some type annotation iff $v = \False$, and \\
+    & & has an outermost mode cast iff $e = \False$
+  \end{tabular}
+  \begin{mathpar}
+    \inferrule{x \in V}{V |-_{\Sigma, \Omega} \isTerm{x}^{\syn\,\True\,\True}}\,\SynRule{Var}
+    \and
+    \inferrule{\cdot |-_{\Sigma} A \\ V |-_{\Sigma, \Omega}\isTerm{t}^{\chk\,v\,e}}{V |-_{\Sigma, \Omega} (\isTerm{t \annotate A})^{\syn\,v\,\True}}\,\SynRule{Anno}
+    \and
+    \inferrule{V |-_{\Sigma, \Omega}\isTerm{t}^{\chk\,v\,\True}}{V |-_{\Sigma, \Omega} \isTerm{t}^{\syn\,\False\,\False}}\,\SynRule{Missing}
+    \and
+    \inferrule{V |-_{\Sigma, \Omega} \isTerm{t}^{\syn\,v\,\True}}{V |-_{\Sigma, \Omega} \isTerm{t}^{\chk\,v\,\False}}\,\ChkRule{Sub}
+    \and
+    \inferrule{V, \vec x_1 |-_{\Sigma, \Omega} \isTerm{t_1}^{\dir{d_1}\,v_1\,e_1} \\ \cdots \\ V, \vec x_n |-_{\Sigma, \Omega} \isTerm{t_n}^{\dir{d_n}\,v_n\,e_n}}
+    {V |-_{\Sigma, \Omega} \tmOpts^{\dir{d}\,(\bigwedge_i v_i)\,\True}}\,\Rule{Op}
+  \end{mathpar}
+  \caption{Generalised mode derivations}
+\end{figure}
+
+\begin{theorem}[Generalised Mode Preprocessing]\label{thm:generalised-mode-preprocessing}
+For any raw term $V |-_{\Sigma,\erase\Omega} t$ and mode~$\dir{d}$, there is a derivation of\/ $V |-_{\Sigma,\Omega} t^{\dir{d}\,v\,e}$ for some boolean $v$~and~$e$.
 \end{theorem}
 
-\begin{theorem}[Completeness]\label{thm:term-completeness}
+\begin{proof}
+By induction on~$t$, we prove the rest of the theorem statement
+\begin{equation}\label{eq:preprocess}
+\forall d.~\exists v, e.~V |-_{\Sigma,\Omega} t^{\dir{d}\,v\,e}
+\end{equation}
+in conjunction with
+\begin{equation}\label{eq:preprocess'}
+V |- t^{\syn\,\True\,\True} ~~\vee~~ V |- t^{\chk\,\True\,\True} ~~\vee~~ \exists e.~V |- t^{\chk\,\False\,e}
+\end{equation}
+Intuitively, statement~(\ref{eq:preprocess'}) gives a precise classification of a term as either a `real' synthesising term (i.e.~without missing annotations or an outermost mode cast), a `real' checking term, or a term that has missing annotations.
+After a derivation is constructed for statement~(\ref{eq:preprocess'}), we can perform some slight adjustment to obtain a desired derivation for statement~(\ref{eq:preprocess}) depending on~$d$:
+\begin{itemize}
+\item For the first disjunct with $\dir{d} = {\syn}$, the second disjunct with $\dir{d} = {\chk}$, and the third disjunct with $\dir{d} = {\chk}$:
+The mode we have matches the mode to be established, so no adjustment is needed.
+\item For the first disjunct with $\dir{d} = {\chk}$:
+Apply $\ChkRule{Sub}$.
+\item For the second disjunct with $\dir{d} = {\syn}$:
+Apply $\SynRule{Missing}$.
+\item For the third disjunct with $\dir{d} = {\syn}$:
+If $e = \False$, then the outermost rule of the derivation must be $\ChkRule{Sub}$, removing which we get a derivation of $V |-_{\Sigma,\Omega} t^{\syn\,\False\,\True}$;
+otherwise $e = \True$, in which case we can apply $\SynRule{Missing}$.
+\end{itemize}
+Now we prove statement~(\ref{eq:preprocess'}) with a case analysis on~$t$.
+\begin{itemize}
+\item $\Rule{Var}$:
+Establish the first disjunct using $\SynRule{Var}$.
+\item $\Rule{Anno}$ ($t \bbcolon A$):
+By statement~(\ref{eq:preprocess}) in the induction hypothesis, there is a derivation of $V |-_{\Sigma,\Omega} t^{\chk\,v\,e}$ for some boolean $v$~and~$e$.
+If $v = \False$, establish the third conjunct using $\SynRule{Anno}$ and $\ChkRule{Sub}$; otherwise, establish the first conjunct using $\ChkRule{Sub}$.
+\item $\Rule{Op}$:
+By statement~(\ref{eq:preprocess}) in the induction hypothesis, there is a derivation of $V |-_{\Sigma,\Omega} t_i^{\dir{d_i}\,v_i\,e_i}$ for all the sub-terms~$t_i$.
+\end{itemize}
+\end{proof}
 
-\end{theorem}
+\begin{lemma}\label{thm:Pre?-true}
+If\/ $V |-_{\Sigma,\Omega} t^{\dir{d}\,\True\,e}$, then $V |-_{\Sigma,\Omega} t^\dir{d}$.
+\end{lemma}
 
-\subsection{Bidirectionalisation}
-\begin{proposition}[Decidability for Bidirectionalisation] \label{thm:bidirectionalisation}
-  It is decidable whether a raw term~$t$ is in mode~$d$ for any $t$ and $d$.
-\end{proposition}
+\begin{lemma}\label{thm:Pre?-false}
+If\/ $V |-_{\Sigma,\Omega} t^{\dir{d}\,\False\,e}$, then $V \not|-_{\Sigma,\Omega} t^\dir{d}$.
+\end{lemma}
 
-\subsection{Annotatability is not Completeness}
+\begin{corollary}[Mode Preprocessing]\label{thm:mode-preprocessing}
+  It is decidable for any raw term $V |-_{\Sigma,\erase\Omega} t$ and mode~$\dir{d}$ whether $V |-_{\Sigma,\Omega} \isTerm{t}^\dir{d}$.
+\end{corollary}
+
+\begin{proof}
+By \cref{thm:generalised-mode-preprocessing}, there is a derivation of $V |-_{\Sigma,\erase\Omega} t^{\dir{d}\,v\,e}$ for some boolean $v$~and~$e$.
+If $v = \True$, then $V |-_{\Sigma,\Omega} t^\dir{d}$ by \cref{thm:Pre?-true}; otherwise $V \not|-_{\Sigma,\Omega} t^\dir{d}$ by \cref{thm:Pre?-false}.
+\end{proof}
+
+\begin{corollary}
+If\/ $V |-_{\Sigma,\Omega} t^\dir{d}$, then $V |-_{\Sigma,\Omega} t^{\dir{d}\,\True\,e}$ for some boolean~$e$.
+\end{corollary}
+
+\begin{corollary}
+If\/ $V \not|-_{\Sigma,\Omega} t^\dir{d}$, then $V |-_{\Sigma,\Omega} t^{\dir{d}\,\False\,e}$ for some boolean~$e$.
+\end{corollary}
+
+\begin{proof}
+By \cref{thm:generalised-mode-preprocessing}, there is a derivation of $V |-_{\Sigma,\erase\Omega} t^{\dir{d}\,v\,e}$ for some boolean $v$~and~$e$.
+Either $v = \True$ or $v = \False$.
+If $V |-_{\Sigma,\Omega} t^\dir{d}$, then $v$ has to be $\True$ since $v = \False$ leads to a contradiction with \cref{thm:Pre?-false}.
+Symmetrically, if $V \not|-_{\Sigma,\Omega} t^\dir{d}$, then $v$ has to be $\False$ so as not to contradict \cref{thm:Pre?-true}.
+\end{proof}
+
+\Josh{($1\frac{1}{2}$~pages)
+
+• Generalised mode derivations and bidirectionalisation
+
+• Intuitive meaning of $v$~and~$e$
+
+• Decidability of the original mode judgements
+
+}
+
+\subsection{Annotatability Is Not Completeness}
+\label{sec:annotatability}
+
+\Josh{$\frac{1}{2}$~pages}
+
+\begin{figure}
+  \centering\small
+  \judgbox{\isTerm{t} \sqsupseteq \isTerm{u}}{A raw term $t$ is more annotated than $u$ (for some bidirectional type system $(\Sigma, \Omega)$)}
+  \begin{mathpar}
+    \inferrule{\isTerm{t} \sqsupseteq \isTerm{u}}
+              {(\isTerm{t} \annotate A) \sqsupseteq \isTerm{u}}\;\Rule{More}
+    \and
+    \inferrule{\vphantom{x : \Identifier}}
+              {\isTerm{x} \sqsupseteq \isTerm{x}}
+    \and
+    \inferrule{\isTerm{t} \sqsupseteq \isTerm{u}}
+              {(\isTerm{t} \annotate A) \sqsupseteq (\isTerm{u} \annotate A)}
+    \and
+    \inferrule{\isTerm{t_1} \sqsupseteq \isTerm{u_1} \quad \cdots \quad \isTerm{t_n} \sqsupseteq{u_n}}
+              {\tmOpts \sqsupseteq \tmOpus}
+  \end{mathpar}
+
+  \caption{Annotation ordering between raw terms}
+  \label{fig:annotation-order}
+\end{figure}

### 2023-06-24 tex/sec4-pre-synthesis.tex
[ tex ] A bit more S4

@@
 
-\begin{theorem}[Generalised Mode Preprocessing]\label{thm:generalised-mode-preprocessing}
-For any raw term $V |-_{\Sigma,\erase\Omega} t$ and mode~$\dir{d}$, there is a derivation of\/ $V |-_{\Sigma,\Omega} t^{\dir{d}\,v\,e}$ for some boolean $v$~and~$e$.
-\end{theorem}
+\begin{lemma}\label{thm:adjustment}
+For any raw term $V |-_{\Sigma,\erase\Omega} t$, if
+\[ V |- t^{\syn\,\True\,\True} ~~\vee~~ V |- t^{\chk\,\True\,\True} ~~\vee~~ \exists e.~V |- t^{\chk\,\False\,e} \]
+then for any mode~$\dir{d}$, there is a derivation of\/ $V |-_{\Sigma,\Omega} t^{\dir{d}\,v\,e}$ for some boolean $v$~and~$e$.
+\end{lemma}
 
-\begin{proof}
-By induction on~$t$, we prove the rest of the theorem statement
-\begin{equation}\label{eq:preprocess}
-\forall d.~\exists v, e.~V |-_{\Sigma,\Omega} t^{\dir{d}\,v\,e}
-\end{equation}
-in conjunction with
-\begin{equation}\label{eq:preprocess'}
-V |- t^{\syn\,\True\,\True} ~~\vee~~ V |- t^{\chk\,\True\,\True} ~~\vee~~ \exists e.~V |- t^{\chk\,\False\,e}
-\end{equation}
-Intuitively, statement~(\ref{eq:preprocess'}) gives a precise classification of a term as either a `real' synthesising term (i.e.~without missing annotations or an outermost mode cast), a `real' checking term, or a term that has missing annotations.
-After a derivation is constructed for statement~(\ref{eq:preprocess'}), we can perform some slight adjustment to obtain a desired derivation for statement~(\ref{eq:preprocess}) depending on~$d$:
+Intuitively, the antecedent gives a precise classification of a term as either a `real' synthesising term (i.e.~without missing annotations or an outermost mode cast), a `real' checking term, or a term that has missing annotations.
+The lemma only performs some slight adjustment to obtain a derivation with any required mode.
+
+\begin{proof}[Proof of \cref{thm:adjustment}]
+Case analysis on the antecedent and~$\dir{d}$.
 \begin{itemize}
@@
 otherwise $e = \True$, in which case we can apply $\SynRule{Missing}$.
+\vspace{-\topsep-\baselineskip}
 \end{itemize}
-Now we prove statement~(\ref{eq:preprocess'}) with a case analysis on~$t$.
+\end{proof}
+
+\begin{theorem}[Generalised Mode Preprocessing]\label{thm:generalised-mode-preprocessing}
+For any raw term $V |-_{\Sigma,\erase\Omega} t$ and mode~$\dir{d}$, there is a derivation of\/ $V |-_{\Sigma,\Omega} t^{\dir{d}\,v\,e}$ for some boolean $v$~and~$e$.
+\end{theorem}
+
+\begin{proof}
+The conclusion is the same as that of \cref{thm:adjustment}, so it suffices to prove the antecedent of \cref{thm:adjustment} by induction on~$t$.
 \begin{itemize}
@@
 \item $\Rule{Anno}$ ($t \bbcolon A$):
-By statement~(\ref{eq:preprocess}) in the induction hypothesis, there is a derivation of $V |-_{\Sigma,\Omega} t^{\chk\,v\,e}$ for some boolean $v$~and~$e$.
+By the induction hypothesis and \cref{thm:adjustment}, there is a derivation of $V |-_{\Sigma,\Omega} t^{\chk\,v\,e}$ for some boolean $v$~and~$e$.
 If $v = \False$, establish the third conjunct using $\SynRule{Anno}$ and $\ChkRule{Sub}$; otherwise, establish the first conjunct using $\ChkRule{Sub}$.
-\item $\Rule{Op}$:
-By statement~(\ref{eq:preprocess}) in the induction hypothesis, there is a derivation of $V |-_{\Sigma,\Omega} t_i^{\dir{d_i}\,v_i\,e_i}$ for all the sub-terms~$t_i$.
+\item $\Rule{Op}$ ($\tmOpts$):
+By the induction hypotheses and \cref{thm:adjustment}, there is a derivation of $V, \vec x_i |-_{\Sigma,\Omega} t_i^{\dir{d_i}\,v_i\,e_i}$ for each sub-term~$t_i$, and by applying $\Rule{Op}$ we get a derivation of $V |-_{\Sigma,\Omega} \tmOpts^{\dir d\,v\,\True}$ where $v = \bigwedge_i v_i$.
+If $v = \True$, establish the first or second disjunct depending on~$\dir d$; otherwise, establish the third disjunct, applying $\ChkRule{Sub}$ if $\dir d = {\syn}$.
+\vspace{-\topsep-\baselineskip}
 \end{itemize}

### 2023-06-25 tex/sec4-pre-synthesis.tex
[ tex ] More proofs in S4

--- 2023-06-24 tex/sec4-pre-synthesis.tex
@@
 
+\Josh{Overview of the section}
+
 \subsection{Soundness and Completeness}
@@
 
-\Josh{Throughout this sub-section, we quantify over any bidirectional type system~$(\Sigma, \Omega)$, context $\Gamma : \Cxt_\Sigma(\emptyset)$, raw term $\erase\Gamma \vdash_{\Sigma, \erase\Omega} t$, mode~$d$, and type $A : \Type_\Sigma(\emptyset)$.}
+%\Josh{Throughout this sub-section, we quantify over any bidirectional type system~$(\Sigma, \Omega)$, context $\Gamma : \Cxt_\Sigma(\emptyset)$, raw term $\erase\Gamma \vdash_{\Sigma, \erase\Omega} t$, mode~$d$, and type $A : \Type_\Sigma(\emptyset)$.}
 
@@
     & misses some type annotation iff $v = \False$, and \\
-    & & has an outermost mode cast iff $e = \False$
+    & & is in (generalised) mode~$d$ due to an outermost mode cast iff $e = \False$
   \end{tabular}
@@
 
+\begin{lemma}\label{thm:adjustment}
+For any raw term $V |-_{\Sigma,\erase\Omega} t$, if
+\begin{equation}\label{eq:classification}
+V |- t^{\syn\,\True\,\True} ~~\vee~~ V |- t^{\chk\,\True\,\True} ~~\vee~~ \exists e.~V |- t^{\chk\,\False\,e}
+\end{equation}
+then for any mode~$\dir{d}$, there is a derivation of\/ $V |-_{\Sigma,\Omega} t^{\dir{d}\,v\,e}$ for some booleans $v$~and~$e$.
+\end{lemma}
+
+\begin{proof}
+Case analysis on statement~(\ref{eq:classification}) and~$\dir{d}$.
+%\begin{itemize}
+%\item For the first disjunct with $\dir{d} = {\syn}$, the second disjunct with $\dir{d} = {\chk}$, and the third disjunct with $\dir{d} = {\chk}$:
+%The mode we have matches the mode to be established, so no adjustment is needed.
+%\item For the first disjunct with $\dir{d} = {\chk}$:
+%Apply $\ChkRule{Sub}$.
+%\item For the second disjunct with $\dir{d} = {\syn}$:
+%Apply $\SynRule{Missing}$.
+%\item For the third disjunct with $\dir{d} = {\syn}$:
+%If $e = \False$, then the outermost rule of the derivation must be $\ChkRule{Sub}$, removing which we get a derivation of $V |-_{\Sigma,\Omega} t^{\syn\,\False\,\True}$;
+%otherwise $e = \True$, in which case we can apply $\SynRule{Missing}$.
+%\vspace{-\topsep-\baselineskip}
+%\end{itemize}
+\end{proof}
+
+Intuitively, statement~(\ref{eq:classification}) gives a precise classification of a term as either a `real' synthesising term (i.e.~without missing annotations or an outermost mode cast), a real checking term, or a term that has missing annotations.
+The lemma performs some slight adjustment that involves only outermost mode casts to switch an input derivation to any required mode.
+
 \begin{theorem}[Generalised Mode Preprocessing]\label{thm:generalised-mode-preprocessing}
-For any raw term $V |-_{\Sigma,\erase\Omega} t$ and mode~$\dir{d}$, there is a derivation of\/ $V |-_{\Sigma,\Omega} t^{\dir{d}\,v\,e}$ for some boolean $v$~and~$e$.
+For any raw term $V |-_{\Sigma,\erase\Omega} t$ and mode~$\dir{d}$, there is a derivation of\/ $V |-_{\Sigma,\Omega} t^{\dir{d}\,v\,e}$ for some booleans $v$~and~$e$.
 \end{theorem}
@@
 \begin{proof}
-By induction on~$t$, we prove the rest of the theorem statement
-\begin{equation}\label{eq:preprocess}
-\forall d.~\exists v, e.~V |-_{\Sigma,\Omega} t^{\dir{d}\,v\,e}
-\end{equation}
-in conjunction with
-\begin{equation}\label{eq:preprocess'}
-V |- t^{\syn\,\True\,\True} ~~\vee~~ V |- t^{\chk\,\True\,\True} ~~\vee~~ \exists e.~V |- t^{\chk\,\False\,e}
-\end{equation}
-Intuitively, statement~(\ref{eq:preprocess'}) gives a precise classification of a term as either a `real' synthesising term (i.e.~without missing annotations or an outermost mode cast), a `real' checking term, or a term that has missing annotations.
-After a derivation is constructed for statement~(\ref{eq:preprocess'}), we can perform some slight adjustment to obtain a desired derivation for statement~(\ref{eq:preprocess}) depending on~$d$:
-\begin{itemize}
-\item For the first disjunct with $\dir{d} = {\syn}$, the second disjunct with $\dir{d} = {\chk}$, and the third disjunct with $\dir{d} = {\chk}$:
-The mode we have matches the mode to be established, so no adjustment is needed.
-\item For the first disjunct with $\dir{d} = {\chk}$:
-Apply $\ChkRule{Sub}$.
-\item For the second disjunct with $\dir{d} = {\syn}$:
-Apply $\SynRule{Missing}$.
-\item For the third disjunct with $\dir{d} = {\syn}$:
-If $e = \False$, then the outermost rule of the derivation must be $\ChkRule{Sub}$, removing which we get a derivation of $V |-_{\Sigma,\Omega} t^{\syn\,\False\,\True}$;
-otherwise $e = \True$, in which case we can apply $\SynRule{Missing}$.
-\end{itemize}
-Now we prove statement~(\ref{eq:preprocess'}) with a case analysis on~$t$.
+It suffices to prove statement~(\ref{eq:classification}) by induction on~$t$ and then apply \cref{thm:adjustment}.
 \begin{itemize}
 \item $\Rule{Var}$:
-Establish the first disjunct using $\SynRule{Var}$.
-\item $\Rule{Anno}$ ($t \bbcolon A$):
-By statement~(\ref{eq:preprocess}) in the induction hypothesis, there is a derivation of $V |-_{\Sigma,\Omega} t^{\chk\,v\,e}$ for some boolean $v$~and~$e$.
-If $v = \False$, establish the third conjunct using $\SynRule{Anno}$ and $\ChkRule{Sub}$; otherwise, establish the first conjunct using $\ChkRule{Sub}$.
+A variable is a real synthesising term, so we establish the first disjunct using $\SynRule{Var}$.
+\item $\Rule{Anno}$:
+Let $t = (t' \bbcolon A)$.
+By the induction hypothesis and \cref{thm:adjustment}, there is a derivation of $V |-_{\Sigma,\Omega} {t'}^{\chk\,v\,e}$ for some booleans $v$~and~$e$.
+If $v = \False$, then $t'$, and therefore~$t$, have missing annotations, and we establish the third disjunct using $\SynRule{Anno}$ and $\ChkRule{Sub}$; otherwise, $t$~is a real synthesising term, and we establish the first disjunct using $\ChkRule{Sub}$.
 \item $\Rule{Op}$:
-By statement~(\ref{eq:preprocess}) in the induction hypothesis, there is a derivation of $V |-_{\Sigma,\Omega} t_i^{\dir{d_i}\,v_i\,e_i}$ for all the sub-terms~$t_i$.
+Let $t = \tmOpts$.
+By the induction hypotheses and \cref{thm:adjustment}, there is a derivation of $V, \vec x_i |-_{\Sigma,\Omega} {t_i}^{\dir{d_i}\,v_i\,e_i}$ for each sub-term~$t_i$; apply $\Rule{Op}$ and we get a derivation of $V |-_{\Sigma,\Omega} t^{\dir d\,v\,\True}$ where $v = \bigwedge_i v_i$.
+If $v = \True$, then $t$~is a real synthesising or checking term, and we establish the first or second disjunct depending on~$\dir d$; otherwise, some~$t_i$, and therefore~$t$, have missing annotations, and we establish the third disjunct, applying $\ChkRule{Sub}$ to get the mode right if $\dir d = {\syn}$.
+\vspace{-\topsep-\baselineskip}
 \end{itemize}
 \end{proof}
+
+\Josh{Separating generalised mode preprocessing into a lemma and a theorem to simplify the case analyses}
 
@@
 
+\begin{proof}
+Induction on the given derivation.
+The $\SynRule{Missing}$ rule cannot appear, and the other rules are mapped to their counterparts.
+\end{proof}
+
 \begin{lemma}\label{thm:Pre?-false}
@@
 \end{lemma}
+
+\begin{proof}
+Induction on the generalised mode derivation, in each case analysing an arbitrary mode derivation and showing that it cannot exist.
+The key case is $\SynRule{Missing}$, where we have a sub-derivation of $V |-_{\Sigma,\Omega} t^{\chk\,v\,\True}$ for some boolean~$v$.
+We do not have an induction hypothesis and seem to get stuck, because $v$~is not necessarily~$\False$ --- indeed, $t$~may well be a real checking term, and this is where it is required to (but cannot) be synthesising.
+But all that matters here is that $t$~is \emph{exactly} in checking mode: if we continue to analyse the sub-derivation, the outermost rule must be $\Rule{Op}$ with $\dir d = {\chk}$, implying that $t$~has to be an operation in checking mode.
+Then a case analysis shows that it is impossible to have a (synthesising) mode derivation of $V |-_{\Sigma,\Omega} t^{\syn}$.
+%\begin{itemize}
+%\item $\SynRule{Anno}$:
+%Any mode derivation must also use $\SynRule{Anno}$ as the outermost rule, and then the induction hypothesis suffices.
+%\item $\ChkRule{Sub}$:
+%We have a sub-derivation of $V |-_{\Sigma,\Omega} t^{\syn\,\False\,\True}$.
+%Any mode derivation of $V |-_{\Sigma,\Omega} t^{\chk}$ must use $\ChkRule{Chk}$ as the outermost rule.
+%(The $\Rule{Op}$ rule with $\dir d = {\chk}$ is impossible because the sub-derivation shows that $t$~is exactly a synthesising term.)
+%We then have a sub-derivation of $V |-_{\Sigma,\Omega} t^{\syn}$, and can use the induction hypothesis.
+%\end{itemize}
+\end{proof}
 
@@
 \begin{proof}
-By \cref{thm:generalised-mode-preprocessing}, there is a derivation of $V |-_{\Sigma,\erase\Omega} t^{\dir{d}\,v\,e}$ for some boolean $v$~and~$e$.
-If $v = \True$, then $V |-_{\Sigma,\Omega} t^\dir{d}$ by \cref{thm:Pre?-true}; otherwise $V \not|-_{\Sigma,\Omega} t^\dir{d}$ by \cref{thm:Pre?-false}.
+By \cref{thm:generalised-mode-preprocessing}, there is a derivation of $V |-_{\Sigma,\erase\Omega} t^{\dir{d}\,v\,e}$ for some booleans $v$~and~$e$, and then simply check~$v$ and apply either \cref{thm:Pre?-true} or \cref{thm:Pre?-false} to obtain $V |-_{\Sigma,\Omega} t^\dir{d}$ or $V \not|-_{\Sigma,\Omega} t^\dir{d}$.
 \end{proof}
 
-\begin{corollary}
+\begin{corollary}\label{thm:toPre?-true}
 If\/ $V |-_{\Sigma,\Omega} t^\dir{d}$, then $V |-_{\Sigma,\Omega} t^{\dir{d}\,\True\,e}$ for some boolean~$e$.
@@
 
-\begin{corollary}
+\begin{corollary}\label{thm:toPre?-false}
 If\/ $V \not|-_{\Sigma,\Omega} t^\dir{d}$, then $V |-_{\Sigma,\Omega} t^{\dir{d}\,\False\,e}$ for some boolean~$e$.
@@
 
-\begin{proof}
-By \cref{thm:generalised-mode-preprocessing}, there is a derivation of $V |-_{\Sigma,\erase\Omega} t^{\dir{d}\,v\,e}$ for some boolean $v$~and~$e$.
-Either $v = \True$ or $v = \False$.
+Whereas it is possible to prove \cref{thm:toPre?-true} directly by induction, converting a mode derivation to a generalised one, the negated antecedent of \cref{thm:toPre?-false} does not provide useful information.
+To prove \cref{thm:toPre?-false}, basically we have to construct a generalised mode derivation from scratch for a raw term, but this is exactly generalised mode preprocessing, so we can simply reuse \cref{thm:generalised-mode-preprocessing}.
+And \cref{thm:toPre?-true} can also be proved easily in the same way.
+
+\begin{proof}[Proofs of \cref{thm:toPre?-true,thm:toPre?-false}]
+
+By \cref{thm:generalised-mode-preprocessing}, there is a derivation of $V |-_{\Sigma,\Omega} t^{\dir{d}\,v\,e}$ for some booleans $v$~and~$e$.
 If $V |-_{\Sigma,\Omega} t^\dir{d}$, then $v$ has to be $\True$ since $v = \False$ leads to a contradiction with \cref{thm:Pre?-false}.

### 2023-06-25 tex/sec1-intro.tex
[ tex ] Related work in S1

--- 2023-06-20 tex/sec1-intro.tex
@@
 
-%  Bidirectional typing had only been a set of principles and characteristics to design typing rules so that a type synthesis algorithm can be `read off' easily
-  %, subsuming soundness, completeness, and decidability of type synthesis algorithm.
+Type checking and inference serve as the transition to type-checked programs, also known as \emph{well-typed terms}, from parsed abstract syntax trees, referred as \emph{raw terms} by contrast.
+Type inference algorithms were conceived to ascertain the type of any raw term without any type annotations.
+However, it was later found that full parametric polymorphism leads to undecidability in type inference, as do dependent types~\citep{Wells1999,Dowek1993}.
+In light of these limitations, bidirectional type synthesis based on bidirectional typing emerged as a viable alternative.
+It presents decidable algorithms for determining the types of \emph{suitably annotated} programs in languages with simple types, polymorphic types, dependent types, gradual types, among others.
+
+The core idea of bidirectional type synthesis is simple.
+It begins with bidirectional typing, extending typing judgements with two modes that direct the flow of types --- the origin of its name.
+Thus, typing judgements come in two forms:
+\begin{enumerate*}
+  \item $\Gamma |- \isTerm{t} \syn A$ for synthesis, and 
+  \item $\Gamma |- \isTerm{t} \chk A$ for checking.
+\end{enumerate*}
+The former indicates that the type of a raw term is synthesized, using both the term and its context as input.
+The latter allows a term to be checked against a given type.
+For instance, the type of a variable $x$ is synthesized via its context, and the type of an annotated term $t \annotate A$ is synthesized as $A$ after $t$ has been checked against $A$.
+From this, a bidirectional type synthesis algorithm can be `read off' based on the type information flow, provided that all outputs are definitively determined by any inputs.
+\citeauthor{Dunfield2021}'s recent survey~\citeyearpar{Dunfield2021} provides a summary of its essential criteria, design principles, and applications that we know so far.
+
+%Type annotations may be seen burdensome at first, but in fact they are useful for programmers to understand the purpose of a program.
+%This claim was put forth by~\citet{Pierce2000} and supported by software engineering practices, since type annotations are often included as part of top-level definitions and of the documentation even for languages such as variants of Hindley--Damas--Milner type system where (complete) type inference is possible.
+%As a result, the focus of type-checking technology has shifted from type inference to bidirectional type synthesis, considering both theoretical limitations and practical considerations.
+
+Distinct from the traditional technique of type inference such as Damas--Milner type inference, bidirectional type synthesis does \emph{not} need unification.
+As argued by~\citet{Pierce2000}, missing annotations are recovered only using the information within adjacent nodes of a syntax tree.
+Imposing long-distance unification constraints on bidirectional type synthesis undermines its essence of locality.
+Further, adding type annotations for top-level definitions can improve clarity, making the purpose of a program more comprehensible, so this process should not be seen as a burden but rather as a practical compromise.
+It is fair to say that bidirectional type synthesis is the foundation of type-checking technology across diverse languages.
+
+Nevertheless, unlike parsing, which has a comprehensive theory and widely applicable practical techniques, bidirectional typing has been developed on a case-by-case basis without a general theory.
+While it is relatively easy to derive a bidirectional type synthesis algorithm, a theory providing logical specifications with rigorously proven properties is still lacking.
+As a result, we can only present informal design principles and considerations that have been discovered from individual systems.
+Moreover, unlike the plethora of available parser generators, no `type-checker generators' exist, and each type checker must be independently built from scratch.
+
+To tackle these challenges, this paper presents a theory of bidirectional typing and a verified implementation of bidirectional type synthesis for simple type theories.
+There are multiple methods and extensions for designing bidirectional typing rules, such as those related to polymorphic types~\citep{Pierce2000,Peyton-Jones2007,Dunfield2013,Xie2018}.
+Yet, once the definition of a bidirectional type system is established, deriving an algorithm becomes uniform.
+Accordingly, our theory's objective is not to formulate rules for various language features, but to introduce 'grammars' for bidirectional typing, called \emph{bidirectional binding signatures}.
+We also characterise essential criteria including \emph{completeness}, \emph{annotatability}, and \emph{mode-correctness} that are sufficient to derive a bidirectional type synthesis algorithm, as informally outlined by~\citet{Dunfield2021}.
+We confine our discussion to simple type theories and their bidirectional type systems that have \emph{syntax-directed} typing rules for simplicity and because of the technical difficulty of presenting more advanced type theories in general.
+
+\subsection{Related work}
+As we will formalise our theory for not only simple type theories but also in a manner distinct from existing formulations in bidirectional typing, it is pertinent to discuss related work, apart from bidirectional typing, upon which this work is built.
+
+\paragraph{Language formalisation and its frameworks}
+The vision of formalising the metatheory of every programming language was initiated by the \PoplMark challenge~\citep{Aydemir2005}.
+This is has been further exemplified in a textbook by~\citet{Wadler2022}, where language concepts are formally addressed using the proof assistant~\Agda, including bidirectional type synthesis for~\PCF.
+
+It was observed~\citep{Altenkirch1993} that rudimentary meta-operations and meta-properties for variable binding and substitution constitute the bulk of language formalisation, inspiring a number of language-formalisation frameworks~\citep{Ahrens2018,Fiore2022,Gheri2020,Ahrens2022,Allais2021} to establish substitution, term traversal, and their meta-properties once and for a range of abstract syntax with variable binding.
+One of the core gadgets of these frameworks is the concept of \emph{descriptions} (coined by~\citet{Chapman2010} for datatype-generic programming) or \emph{binding signatures} (coined by~\citet{Aczel1978} in line with the term \emph{signature} in universal algebra) for specifying typing rules of a language.
+It is noteworthy that these state-of-the-art frameworks cannot fully define polymorphic types nor dependent types yet.
+
+\begin{remark}\label{re:type-synthesis-as-decidability-proof}
+The algorithm of bidirectional type synthesis as provided by~\citeauthor{Wadler2022} differs significantly from existing literature:
+\begin{enumerate*}
+  \item The algorithm is presented as a proof of \emph{(logical) decidability} as to whether a bidirectionally decorated raw term can be bidirectionally typed, equivalent to a program that returns a typing derivation for $\isTerm{t}$, if one exists, or the evidence of a contradiction otherwise.
+  \item In contrast, an algorithm typically found in the literature is presented as \emph{algorithmic system} relations such as $\Gamma |- \isTerm{t} \syn A \mapsto \isTerm{t}'$, denoting that type annotations can be added to $\isTerm{t}$ in the external language to produce $\isTerm{t}'$ of type $A$ in the internal language.
+    Such an algorithm is often accompanied with \emph{soundness} and \emph{completeness} assertions.
+\end{enumerate*}
+\end{remark}
+
+\paragraph{Theories of abstract syntax with variable binding}
+The aforementioned frameworks except \citeauthor{Gheri2020}'s are based on or inspired by the categorical analysis of initial semantics for abstract syntax with variable binding~\cite{Fiore1999}.
+Its main idea is that the set of (untyped) abstract syntax trees for a given language can be interpreted as the \emph{initial algebra} for its \emph{signature} endofunctor on a family of sets $X_{\Gamma}$, indexed by contexts~$\Gamma$.
+Simultaneous renaming of variables for some function $\sigma\colon \Gamma \to \Delta$ acts as a functorial map, mapping elements $t$ of $X_{\Gamma}$ to $\simsub{t}{\rho}$ of $X_{\Delta}$, forming a (covariant) \emph{presheaf}.
+The signature endofunctor encapsulates the collection of language constructs, excluding the variable rule, while the initiality amounts to structural recursion, i.e.\ \emph{term traversal}.
+
+Substitution can also be modelled categorically, but it does not play a role in this paper, though.
+
+\begin{remark}
+Constructing the initial algebra of syntax trees can be as simple as defining an inductive type with a few constructors that align with the variable rule and the rule scheme for algebraic operations specified by a signature~\citep{Allais2021,Fiore2022} (without taking the colimit of a sequence of objects iteratively constructed by the signature functor).
+\end{remark}
+
+\begin{remark} \label{re:type-signature}
+The majority of existing theories treat types independently of terms, thereby excluding them from binding signatures.
+To the best of our knowledge, the only exception to this approach is found in the work of~\citet{Arkor2020}, which incorporates signatures and equations for both terms and types.
+Interestingly, this inclusion is critical for type synthesis when comparing a concrete type $N \bto N$ with an abstract type $A \bto B$, where $A$ and $B$ are type variables in a typing rule.
+\end{remark}
+
+While algebraic approaches for polymorphic type theories have been developed~\citep{Fiore2013,Hamana2011}, these approaches do not take subtyping into account.
+Subtyping is essential for formulating important concepts in type synthesis such as \emph{principal types}.
+On the other hand, in the realm of dependent types, \citeauthor{Cartmell1986}'s generalized algebraic theories~\citeyearpar{Cartmell1986} can handle a wide variety of dependent type theories.
+\citet{Bezem2021} investigate the notion of presentation (extending signature) in the context of generalized algebraic theories.
+Nonetheless, type synthesis for dependent types requires normalisation (or a form of conversion) to check type equality.
+Normalisation in its generic form remains elusive and recent advances in this line of research are discussed in the doctoral thesis by~\citet{Valliappan2023}.
+
+\paragraph{Type Checker Generation}
+While there have been some efforts to generate type checkers grounded in unification~\citep{Gast2004,Grewe2015}, it should be noted that approaches solely based on unification are not suited to handling more complex type theories.
+
+\cite{Pacak2020,Cimini2020}.
+
+\subsection{Plan and Contributions of this paper}
+
+Moreover, given this opportunity of re-examining elements of bidirectional typing, we begin our theory from a general idea of type synthesis independent of bidirectional type synthesis and spot subtleties that are rarely, if any, discussed in the literature on bidirectional typing.
+
+Our approach involves constructing the theory within Martin-L\"of type theory and formalising it using the \Agda proof assistant.
+This formalisation provides a generic type synthesis algorithm that is applicable to all calculi employing bidirectional typing for simple types, hence essentially a verified type-checker generator.
+By rigorously defining its core principles, we aim to lay the foundation for further advancements in bidirectional typing.
+
 %  Our work addresses the problem with existing methodologies, setting the stage for further advances towards a more general theory of type synthesis.
 
-  %In short, this paper is a small but first step towards a general theory of type checking and inference.
   %By formalizing our theory of bidirectional type synthesise in \Agda, we provide a framework capable of handling bidirectional type systems of simple types uniformly.
   %Our work addresses the problem with existing methodologies,
+
 Points to elaborate:
 \begin{enumerate}
-  \item Type inference and checking as a transition from (parsed) abstract syntax trees to well-typed terms.
-  \item Bidirectional typing is by now the only solution that scales well.
-  \item Not yet a general theory but by now a design principle.
-    No `type checker generator' as each type checker is implemented on a case-by-case basis, but an algorithm can usually be read off from syntax-directed typing rules.
-    A hint to a general pattern in syntax-directedness.
   \item The existing work of bidirectional typing does not take account of practical aspects such as missing annotation.
@@
     Soundness of completeness of a type checking algorithm in a less straightforward way.
-  \item A review of what has been done towards this line of research.
-  \item Finally, our responses (contributions of this paper) and some example.
 \end{enumerate}
-
+Assumption to justify:
 \begin{enumerate}
@@
   \item Pfenning's recipe is only of design guide not a technical requirement (there are many exceptions).
-\end{enumerate}
-
-\subsection{Related work}
-
-\citep{Xie2018}
-\paragraph{Language formalization and its frameworks}
-\cite{Wadler2022}
-PoplMark Challenge~\citep{Aydemir2005}
-
-\cite{Cimini2020,Cimini2022}
-
-\citep{Ahrens2018,Fiore2022,Gheri2020,Ahrens2022}
-\cite{Allais2021}
-
-\paragraph{Theories of abstract syntax with variable binding}
-\cite{Fiore1999,Hirschowitz2010,Ahrens2018,Fiore2022,Ahrens2021,Arkor2020,Hirschowitz2022}
-\cite{Fiore2013,Hamana2011,Hamana2022}
-
-\paragraph{Type checker generation}
-\cite{Gast2004,Grewe2015,Pacak2020,Cimini2020}
-
-\paragraph{Bidirectional typing}
-
-\cite{Pierce2000}
-\cite{Peyton-Jones2007}
-\cite{Dunfield2021}
-
-\subsection{Plan and Contributions of this paper}
-
-Our restrictions:
-\begin{enumerate}
-  \item Assumptions: Why only simple types? Various ways to deal with polymorphism; consistent on simple types.
-  \item Assumption: Syntax-directed typing rules
 \end{enumerate}

### 2023-06-25 tex/sec1-intro.tex
[ tex ] Some minor fixes

@@
 \begin{remark} \label{re:type-signature}
-The majority of existing theories treat types independently of terms, thereby excluding them from binding signatures.
-To the best of our knowledge, the only exception to this approach is found in the work of~\citet{Arkor2020}, which incorporates signatures and equations for both terms and types.
-Interestingly, this inclusion is critical for type synthesis when comparing a concrete type $N \bto N$ with an abstract type $A \bto B$, where $A$ and $B$ are type variables in a typing rule.
+Most of existing theories treat types independently of terms, thereby excluding them from signatures.
+To the best of our knowledge, the only exception to this approach is found in the work of~\citet{Arkor2020}, which incorporates signatures for both terms and types.
+Interestingly, this inclusion is critical for type synthesis for comparing a concrete type $N \bto N$ with an abstract type $A \bto B$, where $A$ and $B$ are type variables in a typing rule.
 \end{remark}
@@
 While there have been some efforts to generate type checkers grounded in unification~\citep{Gast2004,Grewe2015}, it should be noted that approaches solely based on unification are not suited to handling more complex type theories.
-
-\cite{Pacak2020,Cimini2020}.
+\LT{Check literature \cite{Pacak2020,Cimini2020}.}
 

### 2023-06-25 tex/sec4-pre-synthesis.tex
[ tex ] S4.2 nearly finished

--- 2023-06-24 tex/sec4-pre-synthesis.tex
@@
 In short, soundness and completeness are no more than the separation and combination of mode and typing information carried by the three kinds of derivations while keeping their basic structure, which is directed by the same raw term.
+Formulating the notion of mode derivation helps to complete this clean picture.
 
@@
   \begin{tabular}{ r r l }
-    & & is in (generalised) mode~$d$, \\
-    $\smash{\boxed{V |-_{\Sigma, \Omega} \isTerm{t}^{\dir{d}\,v\,e}}}$
+    & & is in mode~$d$, \\
+    $\smash{\boxed{V |-_{\Sigma, \Omega} \isTerm{t}^{\dir{d}\,g\,s}}}$
     & The raw term~$\isTerm{t}$\hspace{-.6em}
-    & misses some type annotation iff $v = \False$, and \\
-    & & has an outermost mode cast iff $e = \False$
+    & misses some type annotation iff $g = \False$, and \\
+    & & is in mode~$d$ due to an outermost mode cast iff $s = \False$
   \end{tabular}
@@
     \and
-    \inferrule{\cdot |-_{\Sigma} A \\ V |-_{\Sigma, \Omega}\isTerm{t}^{\chk\,v\,e}}{V |-_{\Sigma, \Omega} (\isTerm{t \annotate A})^{\syn\,v\,\True}}\,\SynRule{Anno}
+    \inferrule{\cdot |-_{\Sigma} A \\ V |-_{\Sigma, \Omega}\isTerm{t}^{\chk\,g\,s}}{V |-_{\Sigma, \Omega} (\isTerm{t \annotate A})^{\syn\,g\,\True}}\,\SynRule{Anno}
     \and
-    \inferrule{V |-_{\Sigma, \Omega}\isTerm{t}^{\chk\,v\,\True}}{V |-_{\Sigma, \Omega} \isTerm{t}^{\syn\,\False\,\False}}\,\SynRule{Missing}
+    \inferrule{V |-_{\Sigma, \Omega}\isTerm{t}^{\chk\,g\,\True}}{V |-_{\Sigma, \Omega} \isTerm{t}^{\syn\,\False\,\False}}\,\SynRule{Missing}
     \and
-    \inferrule{V |-_{\Sigma, \Omega} \isTerm{t}^{\syn\,v\,\True}}{V |-_{\Sigma, \Omega} \isTerm{t}^{\chk\,v\,\False}}\,\ChkRule{Sub}
+    \inferrule{V |-_{\Sigma, \Omega} \isTerm{t}^{\syn\,g\,\True}}{V |-_{\Sigma, \Omega} \isTerm{t}^{\chk\,g\,\False}}\,\ChkRule{Sub}
     \and
-    \inferrule{V, \vec x_1 |-_{\Sigma, \Omega} \isTerm{t_1}^{\dir{d_1}\,v_1\,e_1} \\ \cdots \\ V, \vec x_n |-_{\Sigma, \Omega} \isTerm{t_n}^{\dir{d_n}\,v_n\,e_n}}
-    {V |-_{\Sigma, \Omega} \tmOpts^{\dir{d}\,(\bigwedge_i v_i)\,\True}}\,\Rule{Op}
+    \inferrule{V, \vec x_1 |-_{\Sigma, \Omega} \isTerm{t_1}^{\dir{d_1}\,g_1\,s_1} \\ \cdots \\ V, \vec x_n |-_{\Sigma, \Omega} \isTerm{t_n}^{\dir{d_n}\,g_n\,s_n}}
+    {V |-_{\Sigma, \Omega} \tmOpts^{\dir{d}\,(\bigwedge_i g_i)\,\True}}\,\Rule{Op}
   \end{mathpar}
   \caption{Generalised mode derivations}
+  \label{fig:generalised-mode-derivations}
 \end{figure}
 
-\begin{theorem}[Generalised Mode Preprocessing]\label{thm:generalised-mode-preprocessing}
-For any raw term $V |-_{\Sigma,\erase\Omega} t$ and mode~$\dir{d}$, there is a derivation of\/ $V |-_{\Sigma,\Omega} t^{\dir{d}\,v\,e}$ for some boolean $v$~and~$e$.
+The goal of this sub-section is to construct a mode preprocessor, which decides for any raw term $V |-_{\Sigma,\erase\Omega} t$ and mode~$\dir d$ whether $V |-_{\Sigma,\Omega} t^\dir{d}$ or not.
+In fact we will do better:
+If a mode preprocessor returns a proof that no mode derivation exists, that proof (of a negation) does not provide useful information for the user.
+It will be more helpful if a preprocessor can produce an explanation of why no mode derivation exists, and even how to fix the input term to have a mode derivation.
+We will construct such a \emph{generalised mode preprocessor}~(\cref{thm:generalised-mode-preprocessing}), which can be weakened to an ordinary mode preprocessor~(\cref{thm:mode-preprocessing}) if the additional explanation is not needed.
+
+Intuitively, a term does not have a mode derivation exactly when there are not enough type annotations.
+To formulate this more concretely, we can point out the places in the term that require annotations.
+For a bidirectional type system, an annotation is required wherever a term is `strictly' (which we will explain shortly) in checking mode but required to be in synthesising mode, in which case there is no rule for switching from checking to synthesising, and thus there is no way to construct a mode derivation.
+We can, however, consider \emph{generalised mode derivations} defined in \cref{fig:generalised-mode-derivations} that allow the use of an additional $\SynRule{Missing}$ rule for such switching, so that a derivation can always be constructed.
+Given a generalised mode derivation, if it uses $\SynRule{Missing}$ in some places, then those places are exactly where annotations should be supplied, which is helpful information to the user; if it does not use $\SynRule{Missing}$, then the derivation is \emph{genuine} in the sense that it corresponds directly to an original mode derivation.
+This can be succinctly formulated as \cref{thm:Pre?-true} below by encoding genuineness as a boolean~$g$ in the generalised mode judgement, which is set to~$\False$ only by the $\SynRule{Missing}$ rule.
+(Ignore the boolean~$s$ for now.)
+
+\begin{lemma}\label{thm:Pre?-true}
+If\/ $V |-_{\Sigma,\Omega} t^{\dir{d}\,\True\,s}$, then $V |-_{\Sigma,\Omega} t^\dir{d}$.
+\end{lemma}
+
+\begin{proof}
+Induction on the given derivation.
+The $\SynRule{Missing}$ rule cannot appear because $g = \True$, and the other rules are mapped to their counterparts.
+\end{proof}
+
+We also want a lemma that covers the case where $g = \False$.
+
+\begin{lemma}\label{thm:Pre?-false}
+If\/ $V |-_{\Sigma,\Omega} t^{\dir{d}\,\False\,s}$, then $V \not|-_{\Sigma,\Omega} t^\dir{d}$.
+\end{lemma}
+
+This lemma would be wrong if the `strictness' boolean~$s$ was left out of the rules:
+Having both $\ChkRule{Sub}$ and $\SynRule{Missing}$, which we call \emph{mode casts}, it would be possible to switch between the two modes freely, which unfortunately means that we could insert a pair of $\ChkRule{Sub}$ and $\SynRule{Missing}$ anywhere, constructing a non-genuine derivation even when there is in fact a genuine one.
+The `strictness' boolean~$s$ can be thought of as disrupting the formation of such pairs of mode casts:
+Every rule other than the mode casts sets~$s$ to~$\True$, meaning that a term is \emph{strictly} in the mode assigned by the rule (i.e.~not altered by a mode cast), whereas the mode casts set~$s$ to~$\False$.
+Furthermore, the sub-derivation of a mode cast has to be strict, so it is impossible to have consecutive mode casts.
+Another way to understand the role of~$s$ is that it makes the $\SynRule{Missing}$ rule precise: an annotation is truly missing only when a term is \emph{strictly} in checking mode but is required to be in synthesising mode.
+Now we can show that non-genuine derivations are `truly non-genuine'.
+
+\begin{proof}[Proof of \cref{thm:Pre?-false}]
+Induction on the generalised mode derivation, in each case analysing an arbitrary mode derivation and showing that it cannot exist.
+The key case is $\SynRule{Missing}$, where we have a sub-derivation of $V |-_{\Sigma,\Omega} t^{\chk\,g\,\True}$ for some boolean~$g$.
+We do not have an induction hypothesis and seem to get stuck, because $g$~is not necessarily~$\False$.
+But what matters here is that $t$~is \emph{strictly} in checking mode: if we continue to analyse the sub-derivation, the outermost rule must be $\Rule{Op}$ with $\dir d = {\chk}$, implying that $t$~has to be an operation in checking mode.
+Then a case analysis shows that it is impossible to have a (synthesising) mode derivation of $V |-_{\Sigma,\Omega} t^{\syn}$.
+%\begin{itemize}
+%\item $\SynRule{Anno}$:
+%Any mode derivation must also use $\SynRule{Anno}$ as the outermost rule, and then the induction hypothesis suffices.
+%\item $\ChkRule{Sub}$:
+%We have a sub-derivation of $V |-_{\Sigma,\Omega} t^{\syn\,\False\,\True}$.
+%Any mode derivation of $V |-_{\Sigma,\Omega} t^{\chk}$ must use $\ChkRule{Chk}$ as the outermost rule.
+%(The $\Rule{Op}$ rule with $\dir d = {\chk}$ is impossible because the sub-derivation shows that $t$~is exactly a synthesising term.)
+%We then have a sub-derivation of $V |-_{\Sigma,\Omega} t^{\syn}$, and can use the induction hypothesis.
+%\end{itemize}
+\end{proof}
+
+Now we are almost ready to construct a generalised mode preprocessor.
+
+\begin{theorem}\label{thm:generalised-mode-preprocessing}
+For any raw term $V |-_{\Sigma,\erase\Omega} t$ and mode~$\dir{d}$, there is a derivation of\/ $V |-_{\Sigma,\Omega} t^{\dir{d}\,g\,s}$ for some booleans $g$~and~$s$.
 \end{theorem}
 
+The theorem could be proved directly, but that would take some messy case analyses that inspect the input term and apply mode casts depending on what modes are required all at once.
+Instead, we distill the case analysis that deals with mode casts into the following \cref{thm:adjustment}, whose antecedent~(\ref{eq:classification}) is then established by induction in the proof of \cref{thm:generalised-mode-preprocessing}, which does not need to deal with mode casts anymore.
+
+\begin{lemma}\label{thm:adjustment}
+For any raw term $V |-_{\Sigma,\erase\Omega} t$, if
+\begin{equation}\label{eq:classification}
+\exists\dir{d'}.\quad V |- t^{\dir{d'}\,\True\,\True} ~~\vee~~ \exists s.~V |- t^{\dir{d'}\,\False\,s}
+\end{equation}
+then for any mode~$\dir{d}$, there is a derivation of\/ $V |-_{\Sigma,\Omega} t^{\dir{d}\,g\,s}$ for some booleans $g$~and~$s$.
+\end{lemma}
+
 \begin{proof}
-By induction on~$t$, we prove the rest of the theorem statement
-\begin{equation}\label{eq:preprocess}
-\forall d.~\exists v, e.~V |-_{\Sigma,\Omega} t^{\dir{d}\,v\,e}
-\end{equation}
-in conjunction with
-\begin{equation}\label{eq:preprocess'}
-V |- t^{\syn\,\True\,\True} ~~\vee~~ V |- t^{\chk\,\True\,\True} ~~\vee~~ \exists e.~V |- t^{\chk\,\False\,e}
-\end{equation}
-Intuitively, statement~(\ref{eq:preprocess'}) gives a precise classification of a term as either a `real' synthesising term (i.e.~without missing annotations or an outermost mode cast), a `real' checking term, or a term that has missing annotations.
-After a derivation is constructed for statement~(\ref{eq:preprocess'}), we can perform some slight adjustment to obtain a desired derivation for statement~(\ref{eq:preprocess}) depending on~$d$:
-\begin{itemize}
-\item For the first disjunct with $\dir{d} = {\syn}$, the second disjunct with $\dir{d} = {\chk}$, and the third disjunct with $\dir{d} = {\chk}$:
-The mode we have matches the mode to be established, so no adjustment is needed.
-\item For the first disjunct with $\dir{d} = {\chk}$:
-Apply $\ChkRule{Sub}$.
-\item For the second disjunct with $\dir{d} = {\syn}$:
-Apply $\SynRule{Missing}$.
-\item For the third disjunct with $\dir{d} = {\syn}$:
-If $e = \False$, then the outermost rule of the derivation must be $\ChkRule{Sub}$, removing which we get a derivation of $V |-_{\Sigma,\Omega} t^{\syn\,\False\,\True}$;
-otherwise $e = \True$, in which case we can apply $\SynRule{Missing}$.
-\end{itemize}
-Now we prove statement~(\ref{eq:preprocess'}) with a case analysis on~$t$.
+Exhaustive case analysis on $\dir{d}$, $\dir{d'}$, the disjunction, and~$s$, adding or removing an outermost mode cast to change the given derivation to a different mode if required.
+\end{proof}
+
+Statement~(\ref{eq:classification}) is a slightly strengthened classification of a generalised mode derivation based on its genuineness: when the derivation is genuine, the mode~$\dir{d'}$ should be strict.
+
+\begin{proof}[Proof of \cref{thm:generalised-mode-preprocessing}]
+By \cref{thm:adjustment}, it suffices to prove statement~(\ref{eq:classification}) by induction on~$t$.
 \begin{itemize}
 \item $\Rule{Var}$:
-Establish the first disjunct using $\SynRule{Var}$.
-\item $\Rule{Anno}$ ($t \bbcolon A$):
-By statement~(\ref{eq:preprocess}) in the induction hypothesis, there is a derivation of $V |-_{\Sigma,\Omega} t^{\chk\,v\,e}$ for some boolean $v$~and~$e$.
-If $v = \False$, establish the third conjunct using $\SynRule{Anno}$ and $\ChkRule{Sub}$; otherwise, establish the first conjunct using $\ChkRule{Sub}$.
+A variable is valid, so we establish the first disjunct using $\SynRule{Var}$.
+\item $\Rule{Anno}$:
+Let $t = (t' \bbcolon A)$.
+By the induction hypothesis and \cref{thm:adjustment}, there is a derivation of $V |-_{\Sigma,\Omega} {t'}^{\chk\,g\,s}$ for some booleans $g$~and~$s$, to which we can apply $\Rule{Anno}$, and then we establish the first or second disjunct depending on whether $g$~is $\True$~or~$\False$ (that is, whether $t'$, and therefore~$t$, are valid).
 \item $\Rule{Op}$:
-By statement~(\ref{eq:preprocess}) in the induction hypothesis, there is a derivation of $V |-_{\Sigma,\Omega} t_i^{\dir{d_i}\,v_i\,e_i}$ for all the sub-terms~$t_i$.
+Let $t = \tmOpts$.
+By the induction hypotheses and \cref{thm:adjustment}, there is a derivation of $V, \vec x_i |-_{\Sigma,\Omega} {t_i}^{\dir{d_i}\,g_i\,s_i}$ for each sub-term~$t_i$; apply $\Rule{Op}$ and we get a derivation of $V |-_{\Sigma,\Omega} t^{\dir d\,g\,\True}$ where $g = \bigwedge_i g_i$, depending on which we establish the appropriate disjunct.
+\vspace{-\topsep-\baselineskip}
 \end{itemize}
 \end{proof}
-
-\begin{lemma}\label{thm:Pre?-true}
-If\/ $V |-_{\Sigma,\Omega} t^{\dir{d}\,\True\,e}$, then $V |-_{\Sigma,\Omega} t^\dir{d}$.
-\end{lemma}
-
-\begin{lemma}\label{thm:Pre?-false}
-If\/ $V |-_{\Sigma,\Omega} t^{\dir{d}\,\False\,e}$, then $V \not|-_{\Sigma,\Omega} t^\dir{d}$.
-\end{lemma}
 
@@
 \begin{proof}
-By \cref{thm:generalised-mode-preprocessing}, there is a derivation of $V |-_{\Sigma,\erase\Omega} t^{\dir{d}\,v\,e}$ for some boolean $v$~and~$e$.
-If $v = \True$, then $V |-_{\Sigma,\Omega} t^\dir{d}$ by \cref{thm:Pre?-true}; otherwise $V \not|-_{\Sigma,\Omega} t^\dir{d}$ by \cref{thm:Pre?-false}.
+By \cref{thm:generalised-mode-preprocessing}, there is a derivation of $V |-_{\Sigma,\erase\Omega} t^{\dir{d}\,g\,s}$ for some booleans $g$~and~$s$, and then simply check~$g$ and apply either \cref{thm:Pre?-true} or \cref{thm:Pre?-false} to obtain $V |-_{\Sigma,\Omega} t^\dir{d}$ or $V \not|-_{\Sigma,\Omega} t^\dir{d}$.
 \end{proof}
 
-\begin{corollary}
-If\/ $V |-_{\Sigma,\Omega} t^\dir{d}$, then $V |-_{\Sigma,\Omega} t^{\dir{d}\,\True\,e}$ for some boolean~$e$.
+\begin{corollary}\label{thm:toPre?-true}
+If\/ $V |-_{\Sigma,\Omega} t^\dir{d}$, then $V |-_{\Sigma,\Omega} t^{\dir{d}\,\True\,s}$ for some boolean~$s$.
 \end{corollary}
 
-\begin{corollary}
-If\/ $V \not|-_{\Sigma,\Omega} t^\dir{d}$, then $V |-_{\Sigma,\Omega} t^{\dir{d}\,\False\,e}$ for some boolean~$e$.
+\begin{corollary}\label{thm:toPre?-false}
+If\/ $V \not|-_{\Sigma,\Omega} t^\dir{d}$, then $V |-_{\Sigma,\Omega} t^{\dir{d}\,\False\,s}$ for some boolean~$s$.
 \end{corollary}
 
-\begin{proof}
-By \cref{thm:generalised-mode-preprocessing}, there is a derivation of $V |-_{\Sigma,\erase\Omega} t^{\dir{d}\,v\,e}$ for some boolean $v$~and~$e$.
-Either $v = \True$ or $v = \False$.
-If $V |-_{\Sigma,\Omega} t^\dir{d}$, then $v$ has to be $\True$ since $v = \False$ leads to a contradiction with \cref{thm:Pre?-false}.
-Symmetrically, if $V \not|-_{\Sigma,\Omega} t^\dir{d}$, then $v$ has to be $\False$ so as not to contradict \cref{thm:Pre?-true}.
+Whereas it is possible to prove \cref{thm:toPre?-true} directly by induction, converting a mode derivation to a generalised one, the negated antecedent of \cref{thm:toPre?-false} provides little information for constructing a required derivation, and basically we have to construct such a derivation from scratch for a raw term.
+But this is exactly what generalised mode preprocessing does, so we can simply reuse \cref{thm:generalised-mode-preprocessing}.
+And \cref{thm:toPre?-true} can also be proved easily in the same way.
+
+\begin{proof}[Proofs of \cref{thm:toPre?-true,thm:toPre?-false}]
+
+By \cref{thm:generalised-mode-preprocessing}, there is a derivation of $V |-_{\Sigma,\Omega} t^{\dir{d}\,g\,s}$ for some booleans $g$~and~$s$.
+If $V |-_{\Sigma,\Omega} t^\dir{d}$, then $g$~has to be $\True$ since $g = \False$ leads to a contradiction with \cref{thm:Pre?-false}.
+Symmetrically, if $V \not|-_{\Sigma,\Omega} t^\dir{d}$, then $g$~has to be~$\False$ so as not to contradict \cref{thm:Pre?-true}.
 \end{proof}
-
-\Josh{($1\frac{1}{2}$~pages)
-
-• Generalised mode derivations and bidirectionalisation
-
-• Intuitive meaning of $v$~and~$e$
-
-• Decidability of the original mode judgements
-
-}
 

### 2023-06-25 tex/sec4-pre-synthesis.tex
[ tex ] Done with S4.2 for now

@@
 Another way to understand the role of~$s$ is that it makes the $\SynRule{Missing}$ rule precise: an annotation is truly missing only when a term is \emph{strictly} in checking mode but is required to be in synthesising mode.
-Now we can show that non-genuine derivations are `truly non-genuine'.
+With strictness coming into play, non-genuine derivations are now `truly non-genuine'.
 
@@
 
-The theorem could be proved directly, but that would take some messy case analyses that inspect the input term and apply mode casts depending on what modes are required all at once.
-Instead, we distill the case analysis that deals with mode casts into the following \cref{thm:adjustment}, whose antecedent~(\ref{eq:classification}) is then established by induction in the proof of \cref{thm:generalised-mode-preprocessing}, which does not need to deal with mode casts anymore.
+The theorem could be proved directly, but that would mix up two case analyses which respectively inspect the input term and apply mode casts depending on what modes are required.
+Instead, we distill the case analysis that deals with mode casts into the following \cref{thm:adjustment}, whose antecedent~(\ref{eq:strict-derivation}) is then established by induction on the input term in the proof of \cref{thm:generalised-mode-preprocessing}.
 
@@
 For any raw term $V |-_{\Sigma,\erase\Omega} t$, if
-\begin{equation}\label{eq:classification}
-\exists\dir{d'}.\quad V |- t^{\dir{d'}\,\True\,\True} ~~\vee~~ \exists s.~V |- t^{\dir{d'}\,\False\,s}
+\begin{equation}\label{eq:strict-derivation}
+V |-_{\Sigma,\Omega} t^{\dir{d'}\,g'\,\True} \quad\text{for some mode~$\dir{d'}$ and boolean~$g'$}
 \end{equation}
@@
 \begin{proof}
-Exhaustive case analysis on $\dir{d}$, $\dir{d'}$, the disjunction, and~$s$, adding or removing an outermost mode cast to change the given derivation to a different mode if required.
+Case analysis on $\dir{d}$ and $\dir{d'}$, adding an outermost mode cast to change the given derivation to a different mode if $d \neq d'$.
+Note that it is permissible to add an outermost mode cast because the antecedent~(\ref{eq:strict-derivation}) requires the given derivation to be strict.
 \end{proof}
 
-Statement~(\ref{eq:classification}) is a slightly strengthened classification of a generalised mode derivation based on its genuineness: when the derivation is genuine, the mode~$\dir{d'}$ should be strict.
+\begin{proof}[Proof of \cref{thm:generalised-mode-preprocessing}]
+By \cref{thm:adjustment}, it suffices to prove statement~(\ref{eq:strict-derivation}) by induction on~$t$.
 
-\begin{proof}[Proof of \cref{thm:generalised-mode-preprocessing}]
-By \cref{thm:adjustment}, it suffices to prove statement~(\ref{eq:classification}) by induction on~$t$.
 \begin{itemize}
-\item $\Rule{Var}$:
-A variable is valid, so we establish the first disjunct using $\SynRule{Var}$.
+\item $\Rule{Var}$ is mapped to $\SynRule{Var}$.
 \item $\Rule{Anno}$:
 Let $t = (t' \bbcolon A)$.
-By the induction hypothesis and \cref{thm:adjustment}, there is a derivation of $V |-_{\Sigma,\Omega} {t'}^{\chk\,g\,s}$ for some booleans $g$~and~$s$, to which we can apply $\Rule{Anno}$, and then we establish the first or second disjunct depending on whether $g$~is $\True$~or~$\False$ (that is, whether $t'$, and therefore~$t$, are valid).
+By the induction hypothesis and \cref{thm:adjustment}, there is a derivation of $V |-_{\Sigma,\Omega} {t'}^{\chk\,g\,s}$ for some booleans $g$~and~$s$, to which we apply $\SynRule{Anno}$.
 \item $\Rule{Op}$:
 Let $t = \tmOpts$.
-By the induction hypotheses and \cref{thm:adjustment}, there is a derivation of $V, \vec x_i |-_{\Sigma,\Omega} {t_i}^{\dir{d_i}\,g_i\,s_i}$ for each sub-term~$t_i$; apply $\Rule{Op}$ and we get a derivation of $V |-_{\Sigma,\Omega} t^{\dir d\,g\,\True}$ where $g = \bigwedge_i g_i$, depending on which we establish the appropriate disjunct.
+By the induction hypotheses and \cref{thm:adjustment}, there is a derivation of $V, \vec x_i |-_{\Sigma,\Omega} {t_i}^{\dir{d_i}\,g_i\,s_i}$ for each sub-term~$t_i$; apply $\Rule{Op}$ to all the derivations.
 \vspace{-\topsep-\baselineskip}
@@
 \end{proof}
+
+Finally, having constructed a generalised mode preprocessor, it is easy to derive an ordinary one.
 
@@
 
-\begin{corollary}\label{thm:toPre?-true}
-If\/ $V |-_{\Sigma,\Omega} t^\dir{d}$, then $V |-_{\Sigma,\Omega} t^{\dir{d}\,\True\,s}$ for some boolean~$s$.
-\end{corollary}
-
-\begin{corollary}\label{thm:toPre?-false}
-If\/ $V \not|-_{\Sigma,\Omega} t^\dir{d}$, then $V |-_{\Sigma,\Omega} t^{\dir{d}\,\False\,s}$ for some boolean~$s$.
-\end{corollary}
-
-Whereas it is possible to prove \cref{thm:toPre?-true} directly by induction, converting a mode derivation to a generalised one, the negated antecedent of \cref{thm:toPre?-false} provides little information for constructing a required derivation, and basically we have to construct such a derivation from scratch for a raw term.
-But this is exactly what generalised mode preprocessing does, so we can simply reuse \cref{thm:generalised-mode-preprocessing}.
-And \cref{thm:toPre?-true} can also be proved easily in the same way.
-
-\begin{proof}[Proofs of \cref{thm:toPre?-true,thm:toPre?-false}]
-
-By \cref{thm:generalised-mode-preprocessing}, there is a derivation of $V |-_{\Sigma,\Omega} t^{\dir{d}\,g\,s}$ for some booleans $g$~and~$s$.
-If $V |-_{\Sigma,\Omega} t^\dir{d}$, then $g$~has to be $\True$ since $g = \False$ leads to a contradiction with \cref{thm:Pre?-false}.
-Symmetrically, if $V \not|-_{\Sigma,\Omega} t^\dir{d}$, then $g$~has to be~$\False$ so as not to contradict \cref{thm:Pre?-true}.
-\end{proof}
+%We can also establish the converse of \cref{thm:Pre?-true,thm:Pre?-false}.
+%
+%\begin{corollary}\label{thm:toPre?-true}
+%If\/ $V |-_{\Sigma,\Omega} t^\dir{d}$, then $V |-_{\Sigma,\Omega} t^{\dir{d}\,\True\,s}$ for some boolean~$s$.
+%\end{corollary}
+%
+%\begin{corollary}\label{thm:toPre?-false}
+%If\/ $V \not|-_{\Sigma,\Omega} t^\dir{d}$, then $V |-_{\Sigma,\Omega} t^{\dir{d}\,\False\,s}$ for some boolean~$s$.
+%\end{corollary}
+%
+%Whereas it is possible to prove \cref{thm:toPre?-true} directly by induction, converting a mode derivation to a generalised one, the negated antecedent of \cref{thm:toPre?-false} provides little information for constructing a required derivation, and basically we have to construct such a derivation from scratch for a raw term.
+%But this is exactly what generalised mode preprocessing does, so we can simply reuse \cref{thm:generalised-mode-preprocessing}.
+%And \cref{thm:toPre?-true} can also be proved easily in the same way.
+%
+%\begin{proof}[Proofs of \cref{thm:toPre?-true,thm:toPre?-false}]
+%
+%By \cref{thm:generalised-mode-preprocessing}, there is a derivation of $V |-_{\Sigma,\Omega} t^{\dir{d}\,g\,s}$ for some booleans $g$~and~$s$.
+%If $V |-_{\Sigma,\Omega} t^\dir{d}$, then $g$~has to be $\True$ since $g = \False$ leads to a contradiction with \cref{thm:Pre?-false}.
+%Symmetrically, if $V \not|-_{\Sigma,\Omega} t^\dir{d}$, then $g$~has to be~$\False$ so as not to contradict \cref{thm:Pre?-true}.
+%\end{proof}
 

### 2023-06-26 tex/BiSig.tex
[ tex ] Contributions in S1

@@
   Specifically, the `proof' of logical decidability of type synthesis is instantiated to a synthesiser for the simple type theory specified by a valid signature, analogous to a parser generator.
-  We introduce the notion of bidirectionalisation to refine annotatability in bidirectional typing.
-  Type synthesis and bidirectionalisation result in the trichotomy of raw terms, which models a practical type synthesiser that may throw an exception and locate missing annotations.
+  We introduce the notion of mode preprocessing to refine annotatability in bidirectional typing.
+  Type synthesis and mode preprocessing result in the trichotomy of raw terms, which models a practical type synthesiser that may throw an exception and locate missing annotations.
   By formalizing our general theory of bidirectional type synthesis constructively, we provide a correct-by-construction `type-checker generator' for simple type theories, taking the first step towards a more general theory of bidirectional type synthesis.

### 2023-06-26 tex/sec1-intro.tex
[ tex ] Contributions in S1

@@
 However, it was later found that full parametric polymorphism leads to undecidability in type inference, as do dependent types~\citep{Wells1999,Dowek1993}.
-In light of these limitations, bidirectional type synthesis based on bidirectional typing emerged as a viable alternative.
-It presents decidable algorithms for determining the types of \emph{suitably annotated} programs in languages with simple types, polymorphic types, dependent types, gradual types, among others.
+In light of these limitations, bidirectional type synthesis emerged as a viable alternative, providing decidable algorithms for determining the types of \emph{suitably annotated} programs in languages with more complex types.
+\citet{Dunfield2021} summarised its design principles and applications that we know so far.
 
 The core idea of bidirectional type synthesis is simple.
-It begins with bidirectional typing, extending typing judgements with two modes that direct the flow of types --- the origin of its name.
-Thus, typing judgements come in two forms:
+Briefly speaking, it begins with bidirectional typing, extending typing judgements with two modes that direct the flow of types:
 \begin{enumerate*}
@@
 \end{enumerate*}
-The former indicates that the type of a raw term is synthesized, using both the term and its context as input.
-The latter allows a term to be checked against a given type.
-For instance, the type of a variable $x$ is synthesized via its context, and the type of an annotated term $t \annotate A$ is synthesized as $A$ after $t$ has been checked against $A$.
-From this, a bidirectional type synthesis algorithm can be `read off' based on the type information flow, provided that all outputs are definitively determined by any inputs.
-\citeauthor{Dunfield2021}'s recent survey~\citeyearpar{Dunfield2021} provides a summary of its essential criteria, design principles, and applications that we know so far.
+The former means that the type of a raw term is synthesised, using both the term and its context as input, while for the latter a raw term is to be checked against a given type.
+For instance, the type of a variable $x$ is synthesised by looking it up in its context; the type of an annotated term $t \annotate A$ is synthesised as $A$ after $t$ has been checked against $A$.
+From this, a bidirectional type synthesis algorithm can be `read off' provided that a conditional called \emph{mode-correctness} is met for every typing rule---every synthesised type in a typing rule is determined by previously synthesised types from its premises and its input if checking.
 
-%Type annotations may be seen burdensome at first, but in fact they are useful for programmers to understand the purpose of a program.
-%This claim was put forth by~\citet{Pierce2000} and supported by software engineering practices, since type annotations are often included as part of top-level definitions and of the documentation even for languages such as variants of Hindley--Damas--Milner type system where (complete) type inference is possible.
-%As a result, the focus of type-checking technology has shifted from type inference to bidirectional type synthesis, considering both theoretical limitations and practical considerations.
-
-Distinct from the traditional technique of type inference such as Damas--Milner type inference, bidirectional type synthesis does \emph{not} need unification.
-As argued by~\citet{Pierce2000}, missing annotations are recovered only using the information within adjacent nodes of a syntax tree.
-Imposing long-distance unification constraints on bidirectional type synthesis undermines its essence of locality.
-Further, adding type annotations for top-level definitions can improve clarity, making the purpose of a program more comprehensible, so this process should not be seen as a burden but rather as a practical compromise.
+Distinct from the classic Damas--Milner type inference, bidirectional type synthesis does \emph{not} require unification.
+As noted by~\citet{Pierce2000}, annotations are propagated locally within adjacent nodes of a syntax tree, so introducing long-distance unification constraints in bidirectional type synthesis undermines its essence of locality.
+Further, adding annotations for top-level definitions improves clarity, making the purpose of a program easier to understand, so it should be considered useful for the sake of documentation.
 It is fair to say that bidirectional type synthesis is the foundation of type-checking technology across diverse languages.
 
-Nevertheless, unlike parsing, which has a comprehensive theory and widely applicable practical techniques, bidirectional typing has been developed on a case-by-case basis without a general theory.
-While it is relatively easy to derive a bidirectional type synthesis algorithm, a theory providing logical specifications with rigorously proven properties is still lacking.
-As a result, we can only present informal design principles and considerations that have been discovered from individual systems.
-Moreover, unlike the plethora of available parser generators, no `type-checker generators' exist, and each type checker must be independently built from scratch.
+Unlike parsing, nevertheless, which has a comprehensive theory and widely applicable practical techniques, bidirectional typing has only been developed on a case-by-case basis without a theory.
+While it is easy to derive a bidirectional type synthesis algorithm, a general theory providing logical specifications with rigorously proven properties for a broad range of systems is still lacking.
+As a result, we can only present design principles that we learned from individual systems loosely.
+Moreover, unlike the plethora of available parser generators, `type-checker generators' rarely exist, so each type checker has to be independently built from scratch.
 
 To tackle these challenges, this paper presents a theory of bidirectional typing and a verified implementation of bidirectional type synthesis for simple type theories.
-There are multiple methods and extensions for designing bidirectional typing rules, such as those related to polymorphic types~\citep{Pierce2000,Peyton-Jones2007,Dunfield2013,Xie2018}.
-Yet, once the definition of a bidirectional type system is established, deriving an algorithm becomes uniform.
-Accordingly, our theory's objective is not to formulate rules for various language features, but to introduce 'grammars' for bidirectional typing, called \emph{bidirectional binding signatures}.
-We also characterise essential criteria including \emph{completeness}, \emph{annotatability}, and \emph{mode-correctness} that are sufficient to derive a bidirectional type synthesis algorithm, as informally outlined by~\citet{Dunfield2021}.
-We confine our discussion to simple type theories and their bidirectional type systems that have \emph{syntax-directed} typing rules for simplicity and because of the technical difficulty of presenting more advanced type theories in general.
+There are many ways and extensions for designing bidirectional typing rules, such as those related to polymorphic types~\citep{Pierce2000,Peyton-Jones2007,Dunfield2013,Xie2018}.
+Yet, once the definition of a bidirectional type system is in place, deriving an algorithm becomes straightforward.
+Accordingly, our theory's objective is not to formulate rules for various language features, but to introduce a formalism for bidirectional typing, analogous to grammar in parsing.
+We introduce the notion of \emph{bidirectional binding signatures} to specify bidirectional type systems and characterise essential criteria including \emph{completeness}, \emph{annotatability}, and \emph{mode-correctness} that are sufficient to derive a type synthesis algorithm, as informally outlined by~\citet{Dunfield2021}.
+For simplicity, we confine our discussion to bidirectional simple type systems that have \emph{syntax-directed} typing rules and leave the problem of presenting bidirectional typing for more advanced type theories in general as future work.
 
 \subsection{Related work}
-As we will formalise our theory for not only simple type theories but also in a manner distinct from existing formulations in bidirectional typing, it is pertinent to discuss related work, apart from bidirectional typing, upon which this work is built.
+As we will formalise our theory for not only simple type theories but also in a manner distinct from conventional formulations in bidirectional typing, it is pertinent to discuss related work upon which this work is built, apart from bidirectional typing.
 
-\paragraph{Language formalisation and its frameworks}
+\subsubsection{Language formalisation and its frameworks} \label{sec:language-formalisation}
 The vision of formalising the metatheory of every programming language was initiated by the \PoplMark challenge~\citep{Aydemir2005}.
@@
 
-It was observed~\citep{Altenkirch1993} that rudimentary meta-operations and meta-properties for variable binding and substitution constitute the bulk of language formalisation, inspiring a number of language-formalisation frameworks~\citep{Ahrens2018,Fiore2022,Gheri2020,Ahrens2022,Allais2021} to establish substitution, term traversal, and their meta-properties once and for a range of abstract syntax with variable binding.
+Earlier than the \PoplMark challenge, \citet{Altenkirch1993} already commented that rudimentary meta-operations and meta-properties for substitution constitute the bulk of language formalisation, motivating a number of language-formalisation frameworks~\citep{Ahrens2018,Fiore2022,Gheri2020,Ahrens2022,Allais2021} to establish substitution, term traversal, and their meta-properties universally.
+% once and for a range of abstract syntax with variable binding.
 One of the core gadgets of these frameworks is the concept of \emph{descriptions} (coined by~\citet{Chapman2010} for datatype-generic programming) or \emph{binding signatures} (coined by~\citet{Aczel1978} in line with the term \emph{signature} in universal algebra) for specifying typing rules of a language.
-It is noteworthy that these state-of-the-art frameworks cannot fully define polymorphic types nor dependent types yet.
+It is noteworthy that these state-of-the-art frameworks cannot fully define polymorphic type theories or dependent type theories yet.
 
 \begin{remark}\label{re:type-synthesis-as-decidability-proof}
-The algorithm of bidirectional type synthesis as provided by~\citeauthor{Wadler2022} differs significantly from existing literature:
+The algorithm of bidirectional type synthesis formulated by~\citeauthor{Wadler2022} differs from existing literature:
 \begin{enumerate*}
-  \item The algorithm is presented as a proof of \emph{(logical) decidability} as to whether a bidirectionally decorated raw term can be bidirectionally typed, equivalent to a program that returns a typing derivation for $\isTerm{t}$, if one exists, or the evidence of a contradiction otherwise.
-  \item In contrast, an algorithm typically found in the literature is presented as \emph{algorithmic system} relations such as $\Gamma |- \isTerm{t} \syn A \mapsto \isTerm{t}'$, denoting that type annotations can be added to $\isTerm{t}$ in the external language to produce $\isTerm{t}'$ of type $A$ in the internal language.
-    Such an algorithm is often accompanied with \emph{soundness} and \emph{completeness} assertions.
+  \item The algorithm is presented as a proof of \emph{(logical) decidability} as to whether a `bidirectionally decorated' raw term~$\isTerm{t}$ can be bidirectionally typed, equivalent to a program that returns a typing derivation for $\isTerm{t}$, if one exists, or the evidence of a contradiction otherwise.
+  \item In contrast, an algorithm typically found in the literature is presented as \emph{algorithmic system} relations such as $\Gamma |- \isTerm{t} \syn A \mapsto \isTerm{t}'$, denoting that annotations can be added to $\isTerm{t}$ in the external language to produce $\isTerm{t}'$ of type $A$ in the internal language.
+    Such an algorithm is often accompanied with \emph{soundness} and \emph{completeness} assertions in the sense that the algorithm correctly synthesises the type of a raw term and every typable term can be synthesised if there are enough annotations.
 \end{enumerate*}
@@
 
-\paragraph{Theories of abstract syntax with variable binding}
+\subsubsection{Theories of abstract syntax with variable binding}\label{sec:theory-of-syntax}
 The aforementioned frameworks except \citeauthor{Gheri2020}'s are based on or inspired by the categorical analysis of initial semantics for abstract syntax with variable binding~\cite{Fiore1999}.
-Its main idea is that the set of (untyped) abstract syntax trees for a given language can be interpreted as the \emph{initial algebra} for its \emph{signature} endofunctor on a family of sets $X_{\Gamma}$, indexed by contexts~$\Gamma$.
-Simultaneous renaming of variables for some function $\sigma\colon \Gamma \to \Delta$ acts as a functorial map, mapping elements $t$ of $X_{\Gamma}$ to $\simsub{t}{\rho}$ of $X_{\Delta}$, forming a (covariant) \emph{presheaf}.
-The signature endofunctor encapsulates the collection of language constructs, excluding the variable rule, while the initiality amounts to structural recursion, i.e.\ \emph{term traversal}.
+Its main idea is that the set of (untyped) abstract syntax trees for a given language can be interpreted as the \emph{initial algebra} for its \emph{signature endofunctor} over a family $\Term_{\Gamma}$ of well-scoped terms under the context~$\Gamma$.
+Variable renaming for a function $\sigma\colon \Gamma \to \Delta$ between variables acts as a functorial map from $X_{\Gamma}$ to $X_{\Delta}$, giving rise to a presheaf.
+The signature (endofunctor) encapsulates the collection of language constructs while the initiality amounts to structural recursion, i.e.\ \emph{term traversal}.
 
@@
 \begin{remark}
-Constructing the initial algebra of syntax trees can be as simple as defining an inductive type with a few constructors that align with the variable rule and the rule scheme for algebraic operations specified by a signature~\citep{Allais2021,Fiore2022} (without taking the colimit of a sequence of objects iteratively constructed by the signature functor).
+Constructing the initial algebra of syntax trees boils down to defining an inductive type with a few constructors that align with the variable rule and the rule scheme for language constructs specified by a signature~\citep{Allais2021,Fiore2022} (without explicitly taking the colimit of a sequence of objects iteratively defined by the signature functor).
 \end{remark}
@@
 To the best of our knowledge, the only exception to this approach is found in the work of~\citet{Arkor2020}, which incorporates signatures for both terms and types.
-Interestingly, this inclusion is critical for type synthesis for comparing a concrete type $N \bto N$ with an abstract type $A \bto B$, where $A$ and $B$ are type variables in a typing rule.
+Interestingly, this inclusion is also critical for type synthesis for comparing a concrete type $N \bto N$ with an abstract type $A \bto B$, where $A$ and $B$ are type variables in a typing rule.
 \end{remark}
 
-While algebraic approaches for polymorphic type theories have been developed~\citep{Fiore2013,Hamana2011}, these approaches do not take subtyping into account.
+While algebraic approaches to polymorphic type theories have been developed~\citep{Fiore2013,Hamana2011}, these approaches do not take subtyping into account.
 Subtyping is essential for formulating important concepts in type synthesis such as \emph{principal types}.
 On the other hand, in the realm of dependent types, \citeauthor{Cartmell1986}'s generalized algebraic theories~\citeyearpar{Cartmell1986} can handle a wide variety of dependent type theories.
-\citet{Bezem2021} investigate the notion of presentation (extending signature) in the context of generalized algebraic theories.
-Nonetheless, type synthesis for dependent types requires normalisation (or a form of conversion) to check type equality.
-Normalisation in its generic form remains elusive and recent advances in this line of research are discussed in the doctoral thesis by~\citet{Valliappan2023}.
+\citet{Bezem2021} investigate the notion of presentation (extending the notion of signature) in the context of generalized algebraic theories.
+Nonetheless, type synthesis for dependent types requires normalisation or some form of conversion to check type equality.
+Normalisation in its generic form still remains out of reach and recent advances in this topic are discussed in the doctoral thesis by~\citet{Valliappan2023}.
 
-\paragraph{Type Checker Generation}
-While there have been some efforts to generate type checkers grounded in unification~\citep{Gast2004,Grewe2015}, it should be noted that approaches solely based on unification are not suited to handling more complex type theories.
-\LT{Check literature \cite{Pacak2020,Cimini2020}.}
+\subsubsection{Type Checker Generation}
+While there are some efforts to generate type checkers grounded in unification~\citep{Gast2004,Grewe2015}, it should be noted that unification-based approaches are not suited to more complex type theories.
+Moreover, their algorithms are not proved complete.
 
-\subsection{Plan and Contributions of this paper}
+\subsection{Contributions and Plan of this paper}
+The core contributions of our theory of bidirectional typing, extending upon discussions in previous sections, are as follows.
+First, our theory can be viewed as a counterpart to those theories in \Cref{sec:theory-of-syntax}, with the difference being its specific design for \emph{extrinsically-typed} and \emph{syntax-directed} bidirectional type systems of simple types specified by a \emph{type signature} and a \emph{bidirectional binding signature} for terms.
+The assumption of being syntax-directed is motivated by the standard idea of re-casting non-syntax-directed typing rules of a system into syntax-directed form for deriving a type synthesis algorithm~\cite{Peyton-Jones2007}.
+Since operational semantics are not needed for type synthesis for simple types, we do not address substitution. Instead, our focus is on:
+\begin{enumerate*}
+  \item establishing the connection between the specified bidirectional type system and its associated simple type theory, i.e.\ soundness and completeness; 
+  \item formulating the mode-correctness condition on signatures, sufficient to derive a decidable bidirectional type synthesis.
+\end{enumerate*}
 
-Moreover, given this opportunity of re-examining elements of bidirectional typing, we begin our theory from a general idea of type synthesis independent of bidirectional type synthesis and spot subtleties that are rarely, if any, discussed in the literature on bidirectional typing.
+Our theory is based on Martin-L\"of type theory and has been formalised using the proof assistant \Agda, akin to those language-formalisation frameworks in \Cref{sec:language-formalisation}.
+The natural of constructive logic implies that the law of excluded middle does not stand as an axiom, thus making the proof of a \emph{decidable statement} $P \vee \neg P$ non-trivial.
+Given that all proofs as programs terminate, the notion of algorithmic decidability is implied by the notion of logical decidability.
+The algorithm of our bidirectional type synthesis can be (roughly) understood as a proof of the statement:
+\begin{quote}
+  \emph{For a given raw term $t$, a typing derivation of $t$ either exists or it does not}
+\end{quote}
+This follows from the Curry-Howard correspondence and echoes the algorithm formulated by \citet{Wadler2022} as noted in \Cref{re:type-synthesis-as-decidability-proof}.
+It is important to mention that both algorithmic soundness and completeness are also implied: the proof either returns a typing derivation, if any, for the given raw term~$t$ or a proof that such a derivation cannot exist.
+In essence, the former case is algorithmic soundness, while the latter is computationally stronger than algorithmic completeness.
+Therefore, our algorithm of bidirectional type synthesis, posed as a proof of logical decidability, encapsulates algorithmic soundness, completeness, and decidability all at once.
 
-Our approach involves constructing the theory within Martin-L\"of type theory and formalising it using the \Agda proof assistant.
-This formalisation provides a generic type synthesis algorithm that is applicable to all calculi employing bidirectional typing for simple types, hence essentially a verified type-checker generator.
-By rigorously defining its core principles, we aim to lay the foundation for further advancements in bidirectional typing.
+However, diverging from \citet{Wadler2022}, our methodology does not begin with `bidirectionally decorated' raw terms, but rather with raw terms and a \emph{general type synthesis problem} that is independent of bidirectional type synthesis.
+This divergence enables us to uncover subtleties concerning missing annotations that have been largely overlooked in the literature, and to differentiate between completeness (with respect to a type assignment system) and the concept of annotatability, which have been conflated by \citet{Dunfield2021}.
+We propose the concept of \emph{mode preprocessing} to allocate a mode to a raw term, clarifying the aforementioned difference.
 
-%  Our work addresses the problem with existing methodologies, setting the stage for further advances towards a more general theory of type synthesis.
-
-  %By formalizing our theory of bidirectional type synthesise in \Agda, we provide a framework capable of handling bidirectional type systems of simple types uniformly.
-  %Our work addresses the problem with existing methodologies,
-
-Points to elaborate:
+In summary, we contribute a theory of bidirectional type synthesis that is:
 \begin{enumerate}
-  \item The existing work of bidirectional typing does not take account of practical aspects such as missing annotation.
-    Lightly or rarely discussed.
-  \item A meta-problem: existing work is based on classical logic (or left unspecified at least).
-    Soundness of completeness of a type checking algorithm in a less straightforward way.
+  \item \emph{generic}, applicable to all bidirectional type systems that can be specified by some signature;
+  \item \emph{computationally significant}, in terms of: 
+    \begin{enumerate*}
+      \item prioritising logical decidability over algorithmic soundness and completeness,
+      \item preferring mode preprocessing over annotatability; 
+    \end{enumerate*}
 \end{enumerate}
-Assumption to justify:
-\begin{enumerate}
-  \item Use application in spin form to justify our design choice: type expression of arbitrary rank and the number of constructs in a calculus.
-  \item Argue that syntax-directed form is essential and practical.
-  \item Pfenning's recipe is only of design guide not a technical requirement (there are many exceptions).
-\end{enumerate}
-
-Our contributions:
-\begin{enumerate}
-  \item Generic: Type checking for simple type theories;
-  \item Computational formulation: Bidirectionalisation over annotatability; Decidability over soundness and completeness;
-  \item Verified and Constructive: Fully formalised in Martin-Löf type theory (with Axiom K).
-    LEM is not assumed, so a decidable statement has both computational and non-trivial logical readings.
-\end{enumerate}
+Moreover, all these concepts have been rigorously proved in the proof assistant \Agda.
 

### 2023-06-26 tex/sec2-outline.tex
[ tex ] S4 draft

@@
 
-\begin{theorem}
+\begin{theorem}\label{thm:implementation}
 A type synthesiser that uses `not in synthesising mode' as its excuse can be constructed from a mode preprocessor and a bidirectional type synthesiser.

### 2023-06-26 tex/sec4-pre-synthesis.tex
[ tex ] S4 draft

@@
 
-\begin{lemma}[Soundness]\label{thm:term-soundness}
+\begin{lemma}[Soundness]\label{thm:soundness}
 If\/ $\Gamma \vdash_{\Sigma, \Omega} t :^\dir{d} A$, then $\Gamma \vdash_{\Sigma, \erase\Omega} t : A$.
@@
 
-\begin{lemma}[Completeness]\label{thm:term-completeness}
+\begin{lemma}[Completeness]\label{thm:completeness}
 If\/ $\erase\Gamma |-_{\Sigma, \Omega} t^\dir{d}$ and\/ $\Gamma |-_{\Sigma, \erase\Omega} t : A$, then $\Gamma |-_{\Sigma, \Omega} t :^\dir{d} A$.
@@
 In short, soundness and completeness are no more than the separation and combination of mode and typing information carried by the three kinds of derivations while keeping their basic structure, which is directed by the same raw term.
-Formulating the notion of mode derivation helps to complete this clean picture.
+%Formulating the notion of mode derivation helps to complete this clean picture.
 
@@
 \label{sec:mode-preprocessing}
+
+The goal of this sub-section is to construct a mode preprocessor, which decides for any raw term $V |-_{\Sigma,\erase\Omega} t$ and mode~$\dir d$ whether $V |-_{\Sigma,\Omega} t^\dir{d}$ or not.
+In fact we will do better:
+If a mode preprocessor returns a proof that no mode derivation exists, that proof (of a negation) does not provide useful information for the user.
+It will be more helpful if a preprocessor can produce an explanation of why no mode derivation exists, and even how to fix the input term to have a mode derivation.
+We will construct such a \emph{generalised mode preprocessor}~(\cref{thm:generalised-mode-preprocessing}), which can be weakened to an ordinary mode preprocessor~(\cref{thm:mode-preprocessing}) if the additional explanation is not needed.
 
@@
 \end{figure}
-
-The goal of this sub-section is to construct a mode preprocessor, which decides for any raw term $V |-_{\Sigma,\erase\Omega} t$ and mode~$\dir d$ whether $V |-_{\Sigma,\Omega} t^\dir{d}$ or not.
-In fact we will do better:
-If a mode preprocessor returns a proof that no mode derivation exists, that proof (of a negation) does not provide useful information for the user.
-It will be more helpful if a preprocessor can produce an explanation of why no mode derivation exists, and even how to fix the input term to have a mode derivation.
-We will construct such a \emph{generalised mode preprocessor}~(\cref{thm:generalised-mode-preprocessing}), which can be weakened to an ordinary mode preprocessor~(\cref{thm:mode-preprocessing}) if the additional explanation is not needed.
 
@@
 
-Now we are almost ready to construct a generalised mode preprocessor.
+Now we are ready to construct a generalised mode preprocessor.
 
@@
 
-\subsection{Annotatability Is Not Completeness}
+\subsection{Annotatability}
 \label{sec:annotatability}
-
-\Josh{$\frac{1}{2}$~pages}
 
@@
   \caption{Annotation ordering between raw terms}
-  \label{fig:annotation-order}
+  \label{fig:annotation-ordering}
 \end{figure}
+
+\citet{Dunfield2021} formulated completeness differently from ours (\cref{thm:completeness}) and proposed \emph{annotatability} as a more suitable name.
+Here we reformulate annotatability in our theory and discuss its relationship with our completeness and generalised mode preprocessing.
+
+\begin{proposition}[Annotatability]\label{thm:annotatability}
+If\/ $\Gamma |-_{\Sigma,\erase\Omega} t : A$, then there exists~$t'$ such that\/ $t' \sqsupseteq t$ and $\Gamma |-_{\Sigma,\Omega} t' :^\dir{d} A$ for some~$\dir{d}$.
+\end{proposition}
+
+Defined in \cref{fig:annotation-ordering}, the `annotation ordering' $t' \sqsupseteq t$ means that $t'$~has the same or more annotations than~$t$.
+In a sense, annotatability is a reasonable form of completeness: if a term is typable in $(\Sigma, \erase\Omega)$, it may not be typable in $(\Sigma, \Omega)$ directly, but will be if some annotations are added.
+Indeed, \cref{thm:annotatability} can be easily proved in our theory from generalised mode preprocessing.
+
+\begin{proof}[Proof of \cref{thm:annotatability}]
+By \cref{thm:generalised-mode-preprocessing}, there is a generalised mode derivation of $V |-_{\Sigma,\erase\Omega} t^{\dir{d}\,g\,s}$ for some mode~$\dir{d}$ and booleans $g$~and~$s$.
+Perform induction on this generalised mode derivation to construct a bidirectional typing derivation in the same mode (and also the term~$t'$ and a proof of $t' \sqsupseteq t$, which are straightforward and omitted here).
+\begin{itemize}
+\item $\SynRule{Var}$, $\SynRule{Anno}$, and $\Rule{Op}$:
+The outermost rule of the given typing derivation must be the corresponding one; apply the corresponding bidirectional typing rule to the induction hypotheses.
+\item $\ChkRule{Sub}$:
+Apply $\ChkRule{Sub}$ to the induction hypothesis.
+\item $\SynRule{Missing}$:
+This is the interesting case where we map a $\SynRule{Missing}$ rule to an $\SynRule{Anno}$ rule and add to the term a type annotation, which comes from the given typing derivation.
+\vspace{-\topsep-\baselineskip}
+\end{itemize}
+\end{proof}
+
+On the other hand, when using a bidirectional type synthesiser to implement type synthesis~(\cref{thm:implementation}), annotatability cannot replace our completeness since it relates typing derivations for possibly differently annotated terms, but a bidirectional type synthesiser is concerned about properties of \emph{one} given term throughout the process.
+Annotatability also does not help the user to deal with missing annotations like generalised mode preprocessing does: even though annotatability seems to construct a term with missing annotations filled in, its antecedent requires a typing derivation, which is what the user is trying to construct and does not have yet.
+It seems that as we focus on properties that serve practical purposes, it becomes unclear how annotatability comes into the picture.
+In particular, annotatability mixes up the aspects of annotating and typing, and therefore does not fit into our theory, where the two aspects are separated for better usability.

### 2023-06-26 tex/sec1-intro.tex
[ tex ] Finish contributions in S1

@@
 Accordingly, our theory's objective is not to formulate rules for various language features, but to introduce a formalism for bidirectional typing, analogous to grammar in parsing.
-We introduce the notion of \emph{bidirectional binding signatures} to specify bidirectional type systems and characterise essential criteria including \emph{completeness}, \emph{annotatability}, and \emph{mode-correctness} that are sufficient to derive a type synthesis algorithm, as informally outlined by~\citet{Dunfield2021}.
+We introduce the notion of bidirectional binding signatures to specify bidirectional type systems and characterise essential criteria including \emph{completeness}, \emph{annotatability}, and \emph{mode-correctness} that are sufficient to derive a type synthesis algorithm, as informally outlined by~\citet{Dunfield2021}.
 For simplicity, we confine our discussion to bidirectional simple type systems that have \emph{syntax-directed} typing rules and leave the problem of presenting bidirectional typing for more advanced type theories in general as future work.
 
-\subsection{Related work}
-As we will formalise our theory for not only simple type theories but also in a manner distinct from conventional formulations in bidirectional typing, it is pertinent to discuss related work upon which this work is built, apart from bidirectional typing.
+\subsection{Related work}\label{sec:related-work}
+As we will formalise our theory for not only simple type theories but also in a manner distinct from conventional formulations in bidirectional typing, it is pertinent to discuss related work upon which this work is built, apart from bidirectional typing, in order to articulate our contributions.
 
@@
 
-Earlier than the \PoplMark challenge, \citet{Altenkirch1993} already commented that rudimentary meta-operations and meta-properties for substitution constitute the bulk of language formalisation, motivating a number of language-formalisation frameworks~\citep{Ahrens2018,Fiore2022,Gheri2020,Ahrens2022,Allais2021} to establish substitution, term traversal, and their meta-properties universally.
-% once and for a range of abstract syntax with variable binding.
-One of the core gadgets of these frameworks is the concept of \emph{descriptions} (coined by~\citet{Chapman2010} for datatype-generic programming) or \emph{binding signatures} (coined by~\citet{Aczel1978} in line with the term \emph{signature} in universal algebra) for specifying typing rules of a language.
-It is noteworthy that these state-of-the-art frameworks cannot fully define polymorphic type theories or dependent type theories yet.
+Earlier than the \PoplMark challenge already, \citet{Altenkirch1993} commented that rudimentary meta-operations and meta-properties for substitution constitute the bulk of language formalisation, motivating a number of language-formalisation frameworks~\citep{Ahrens2018,Fiore2022,Gheri2020,Ahrens2022,Allais2021} to define the type of well-scoped/typed terms, substitution, term traversal and their meta-properties universally.
+One of the core gadgets of these frameworks is the concept of \emph{descriptions} or \emph{binding signatures} (coined by~\citet{Aczel1978} in line with the term \emph{signature} in universal algebra) for specifying typing rules of a language.
+It is noteworthy that these state-of-the-art frameworks cannot specify polymorphic type theories or dependent type theories yet and only \citeauthor{Allais2021} discussed generic programs beyond substitution.
 
@@
 \subsubsection{Theories of abstract syntax with variable binding}\label{sec:theory-of-syntax}
-The aforementioned frameworks except \citeauthor{Gheri2020}'s are based on or inspired by the categorical analysis of initial semantics for abstract syntax with variable binding~\cite{Fiore1999}.
-Its main idea is that the set of (untyped) abstract syntax trees for a given language can be interpreted as the \emph{initial algebra} for its \emph{signature endofunctor} over a family $\Term_{\Gamma}$ of well-scoped terms under the context~$\Gamma$.
-Variable renaming for a function $\sigma\colon \Gamma \to \Delta$ between variables acts as a functorial map from $X_{\Gamma}$ to $X_{\Delta}$, giving rise to a presheaf.
-The signature (endofunctor) encapsulates the collection of language constructs while the initiality amounts to structural recursion, i.e.\ \emph{term traversal}.
+The aforementioned frameworks except \citeauthor{Gheri2020}'s are at least inspired by \citet{Fiore1999}'s initial semantics for abstract syntax with variable binding using category theory.
+The main idea is that the set of (untyped) abstract syntax trees for a language can be interpreted as the initial algebra for its signature endofunctor $\mathbb{\Sigma}$ on a family $\Term_{\Gamma}$ of well-scoped terms under the context~$\Gamma$ with
+variable renaming for a function $\sigma\colon \Gamma \to \Delta$ between variables acting as a functorial map from $X_{\Gamma}$ to $X_{\Delta}$, equivalently a free $\mathbb{\Sigma}$-algebra on the presheaf of variables $V\colon \mathbb{F} \hookrightarrow \mathsf{Set}$.
+The endofunctor $\mathbb{\Sigma}$ encapsulates language constructs (except variables) while the initiality amounts to structural recursion, i.e.\ \emph{term traversal}.
 
@@
 \begin{remark}
-Constructing the initial algebra of syntax trees boils down to defining an inductive type with a few constructors that align with the variable rule and the rule scheme for language constructs specified by a signature~\citep{Allais2021,Fiore2022} (without explicitly taking the colimit of a sequence of objects iteratively defined by the signature functor).
+Constructing the initial algebra of terms boils down to defining an inductive type with a few constructors that align with the variable rule and the rule scheme for language constructs specified by a signature~\citep{Allais2021,Fiore2022}, without the need to explicitly define the colimit of a sequence of objects iteratively defined by the signature functor.
 \end{remark}
@@
 \subsection{Contributions and Plan of this paper}
-The core contributions of our theory of bidirectional typing, extending upon discussions in previous sections, are as follows.
-First, our theory can be viewed as a counterpart to those theories in \Cref{sec:theory-of-syntax}, with the difference being its specific design for \emph{extrinsically-typed} and \emph{syntax-directed} bidirectional type systems of simple types specified by a \emph{type signature} and a \emph{bidirectional binding signature} for terms.
-The assumption of being syntax-directed is motivated by the standard idea of re-casting non-syntax-directed typing rules of a system into syntax-directed form for deriving a type synthesis algorithm~\cite{Peyton-Jones2007}.
-Since operational semantics are not needed for type synthesis for simple types, we do not address substitution. Instead, our focus is on:
+The major contributions of paper, extending upon previous discussions, are explained as follows.
+
+First, our theory can be viewed as a counterpart to those theories in \Cref{sec:theory-of-syntax} and is tailored towards \emph{extrinsically-typed}, \emph{syntax-directed} bidirectional type systems of simple types (\Cref{fig:raw-terms,fig:bidirectional-typing-derivations}) specified by a \emph{signature} for simple types and a \emph{bidirectional binding signature} for terms (\Cref{def:bidirectional-binding-signature}).
+The extrinsic typing is needed for type synthesis and our adherence to syntax-directed systems aligns with the standard idea of re-casting non-syntax-directed typing rules into their syntax-directed form to derive type synthesis algorithm~\citep{Peyton-Jones2007}.
+As operational semantics are not needed for synthesising simple types, we do not address substitution.
+Instead, our focus is on:
 \begin{enumerate*}
-  \item establishing the connection between the specified bidirectional type system and its associated simple type theory, i.e.\ soundness and completeness; 
-  \item formulating the mode-correctness condition on signatures, sufficient to derive a decidable bidirectional type synthesis.
+  \item establishing the connection between the specified bidirectional type system and its simple type theory, i.e.\ \emph{soundness} and \emph{completeness} (\Cref{sec:soundness-and-completeness}); 
+  \item formulating the \emph{mode-correctness} condition (\Cref{def:mode-correctness}) which suffices to establish the uniqueness of synthesised types (\Cref{thm:unique-syn}) and a decidable bidirectional type synthesis and checking (\Cref{thm:bidirectional-type-synthesis-checking}).
 \end{enumerate*}
 
-Our theory is based on Martin-L\"of type theory and has been formalised using the proof assistant \Agda, akin to those language-formalisation frameworks in \Cref{sec:language-formalisation}.
-The natural of constructive logic implies that the law of excluded middle does not stand as an axiom, thus making the proof of a \emph{decidable statement} $P \vee \neg P$ non-trivial.
-Given that all proofs as programs terminate, the notion of algorithmic decidability is implied by the notion of logical decidability.
-The algorithm of our bidirectional type synthesis can be (roughly) understood as a proof of the statement:
+Our theory is based on Martin-L\"of type theory and formalised in \Agda (\Cref{sec:formalisation}), akin to those frameworks in \Cref{sec:language-formalisation}.
+Moreover, we exploit the nature of constructive logic to formulate algorithmic soundness, completeness, and decidability. 
+To be more specific, recall that the law of excluded middle does not hold universally as an axiom, thus making a \emph{decidable statement} $P \vee \neg P$ non-trivial to prove.
+Given that all proofs as programs terminate, logical decidability inherently implies algorithmic decidability.
+Further, a proof of the following decidable statement:
 \begin{quote}
-  \emph{For a given raw term $t$, a typing derivation of $t$ either exists or it does not}
+  `For a context $\Gamma$ and a raw term $t$, either a typing derivation of\, $\Gamma |- t : A$ exists for some $A$ or any derivation of\, $\Gamma |- t : A$ for some $A$ leads to a contradiction'
 \end{quote}
-This follows from the Curry-Howard correspondence and echoes the algorithm formulated by \citet{Wadler2022} as noted in \Cref{re:type-synthesis-as-decidability-proof}.
-It is important to mention that both algorithmic soundness and completeness are also implied: the proof either returns a typing derivation, if any, for the given raw term~$t$ or a proof that such a derivation cannot exist.
-In essence, the former case is algorithmic soundness, while the latter is computationally stronger than algorithmic completeness.
-Therefore, our algorithm of bidirectional type synthesis, posed as a proof of logical decidability, encapsulates algorithmic soundness, completeness, and decidability all at once.
+or rephrased succinctly as 
+\begin{quote}
+  `A context $\Gamma$ and a raw term $t$ \emph{decide} whether $\Gamma |- t : A$ has a derivation for some $A$'
+\end{quote}
+computes a type and a typing derivation for a context and a raw term, echoing the algorithm given by \citeauthor{Wadler2022} as noted in \Cref{re:type-synthesis-as-decidability-proof}.
+Both algorithmic soundness and completeness are implicitly stated:
+the proof either yields a typing derivation for the given raw term~$t$ or a proof that such a derivation is impossible.
+Accordingly, the former case is algorithmic soundness, while the latter is computationally stronger than completeness.
+Hence, our decidable bidirectional type synthesis (\Cref{thm:bidirectional-type-synthesis-checking}) encapsulates algorithmic soundness, completeness, and decidability simultaneously.
 
-However, diverging from \citet{Wadler2022}, our methodology does not begin with `bidirectionally decorated' raw terms, but rather with raw terms and a \emph{general type synthesis problem} that is independent of bidirectional type synthesis.
-This divergence enables us to uncover subtleties concerning missing annotations that have been largely overlooked in the literature, and to differentiate between completeness (with respect to a type assignment system) and the concept of annotatability, which have been conflated by \citet{Dunfield2021}.
-We propose the concept of \emph{mode preprocessing} to allocate a mode to a raw term, clarifying the aforementioned difference.
+In contrast to \citeauthor{Wadler2022}, our theory starts with raw terms and a general type synthesis problem, independent of bidirectional typing.
+This angle uncovers nuances regarding the lack of enough type annotations in bidirectional typing, generally overlooked in literature, and distinguishes proof-theoretic completeness (with respect to a type assignment system) from annotatability, concepts conflated by \citet{Dunfield2021}.
+To clarify the difference (\Cref{sec:annotatability}), we propose \emph{mode preprocessing} (\Cref{sec:mode-preprocessing}), which assigns a mode to a raw term with enough annotations, or pinpoints missing type annotations otherwise.
+By combining bidirectional type synthesis, soundness, completeness, and mode preprocessing, we arrive our main result (\Cref{cor:trichotomy}) modelling a type synthesiser which checks if a given raw term is \emph{suitably annotated} and if a suitably annotated raw term has a typing derivation with respect to the specified simple type theory, instead of its bidirectional type system.
+Our bidirectional type synthesis works for any mode-correct signature and allows us to instantiate it to any specified system, effectively a verified type-checker generator.
 
@@
 \begin{enumerate}
-  \item \emph{generic}, applicable to all bidirectional type systems that can be specified by some signature;
-  \item \emph{computationally significant}, in terms of: 
+  \item \emph{simple yet general} for any system that can be specified by a bidirectional binding signature;
+  \item \emph{constructive}, in terms of: 
     \begin{enumerate*}
-      \item prioritising logical decidability over algorithmic soundness and completeness,
-      \item preferring mode preprocessing over annotatability; 
+      \item prioritising logical decidability over algorithmic soundness, completeness, and decidability,
+      \item preferring mode preprocessing over annotatability.
     \end{enumerate*}
 \end{enumerate}
-Moreover, all these concepts have been rigorously proved in the proof assistant \Agda.
+Moreover, all these concepts have been formally proved in \Agda.
 

### 2023-06-26 tex/sec3-defs.tex
[ tex ] Finish contributions in S1

@@
 \subsection{Simple Types}
-\begin{definition}
+\begin{definition} \label{def:simple-signature}
   A \emph{(simple type) signature} $\Sigma$ consists of a set $I$ and an \emph{arity} function $\arity\colon I \to \mathbb{N}$.
@@
   \caption{Raw terms}
+  \label{fig:raw-terms}
 \end{figure}
@@
   \end{mathpar}
-  \caption{Typing rules}
+  \caption{Typing derivations}
   \label{fig:extrinsic-typing}
@@
 
-\subsection{Bidirectional Binding Signatures and Bidirectional Type Systems}
+\subsection{Bidirectional Binding Signatures and Bidirectional Type Systems} \label{sec:bidirectional-system}
 
-\begin{definition}
+\begin{definition} \label{def:bidirectional-binding-signature}
   For a type signature $\Sigma$, a \emph{bidirectional binding signature} $\Omega$ is a set $O$ with a function
@@
   \end{mathpar}
-  \caption{Preprocessed raw terms, I}
+  \caption{Mode derivations}
 \end{figure}
@@
 Abbreviations: $\boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \syn A} \defeq \boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A^{\syn}}$ and $\boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \chk A} \defeq \boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A^{\chk}}$
-  \caption{Bidirectional typing rules}
+  \caption{Bidirectional typing derivations}
+  \label{fig:bidirectional-typing-derivations}
 \end{figure}

### 2023-06-26 tex/sec5-synthesis.tex
[ tex ] S4 in progress
[ tex ] Finish contributions in S1

@@
 \subsection{Mode Correctness}
-\begin{definition}[Mode-Correctness]
+\begin{definition}[Mode-Correctness] \label{def:mode-correctness}
 \end{definition}
@@
 
-\begin{corollary}[Trichotomy of Raw Terms]
+\begin{corollary}[Trichotomy of Raw Terms]\label{cor:trichotomy}
   Let $(\Sigma, \Omega)$ be a mode-correct bidirectional type system.
@@
 \begin{proof}
-  Combine  \Cref{thm:term-soundness,thm:term-completeness,thm:bidirectionalisation} with \Cref{thm:bidirectional-type-synthesis}.
+  Combine  \Cref{thm:term-soundness,thm:term-completeness,thm:mode-preprocessing} with \Cref{thm:bidirectional-type-synthesis}.
 

### 2023-06-26 tex/sec1-intro.tex
[ tex ] Fix undefined references

@@
 
-First, our theory can be viewed as a counterpart to those theories in \Cref{sec:theory-of-syntax} and is tailored towards \emph{extrinsically-typed}, \emph{syntax-directed} bidirectional type systems of simple types (\Cref{fig:raw-terms,fig:bidirectional-typing-derivations}) specified by a \emph{signature} for simple types and a \emph{bidirectional binding signature} for terms (\Cref{def:bidirectional-binding-signature}).
+First, our theory can be viewed as a counterpart to those theories in \Cref{sec:theory-of-syntax} and is tailored towards \emph{extrinsically-typed}, \emph{syntax-directed} bidirectional type systems of simple types (\Cref{fig:raw-terms,fig:bidirectional-typing-derivations}) specified by a \emph{signature} for simple types (c.f.\ \Cref{re:type-signature}) and a \emph{bidirectional binding signature} for terms (\Cref{def:bidirectional-binding-signature}).
 The extrinsic typing is needed for type synthesis and our adherence to syntax-directed systems aligns with the standard idea of re-casting non-syntax-directed typing rules into their syntax-directed form to derive type synthesis algorithm~\citep{Peyton-Jones2007}.
@@
 In contrast to \citeauthor{Wadler2022}, our theory starts with raw terms and a general type synthesis problem, independent of bidirectional typing.
-This angle uncovers nuances regarding the lack of enough type annotations in bidirectional typing, generally overlooked in literature, and distinguishes proof-theoretic completeness (with respect to a type assignment system) from annotatability, concepts conflated by \citet{Dunfield2021}.
+This angle uncovers nuances regarding the lack of enough type annotations in bidirectional typing, generally overlooked in literature, and distinguishes completeness (with respect to a type assignment system) from annotatability, concepts conflated by \citet{Dunfield2021}.
 To clarify the difference (\Cref{sec:annotatability}), we propose \emph{mode preprocessing} (\Cref{sec:mode-preprocessing}), which assigns a mode to a raw term with enough annotations, or pinpoints missing type annotations otherwise.

### 2023-06-26 tex/sec1-intro.tex
[ tex ] Add a reference to (generalised) mode derivations

@@
 To be more specific, recall that the law of excluded middle does not hold universally as an axiom, thus making a \emph{decidable statement} $P \vee \neg P$ non-trivial to prove.
-Given that all proofs as programs terminate, logical decidability inherently implies algorithmic decidability.
+Given that all proofs as programs terminate, \emph{logical decidability} inherently implies \emph{algorithmic decidability}.
 Further, a proof of the following decidable statement:
@@
 Accordingly, the former case is algorithmic soundness, while the latter is computationally stronger than completeness.
-Hence, our decidable bidirectional type synthesis (\Cref{thm:bidirectional-type-synthesis-checking}) encapsulates algorithmic soundness, completeness, and decidability simultaneously.
+Hence, our decidable bidirectional type synthesis (\Cref{thm:bidirectional-type-synthesis-checking}) encapsulates algorithmic soundness, completeness, and decidability \emph{simultaneously}.
 
 In contrast to \citeauthor{Wadler2022}, our theory starts with raw terms and a general type synthesis problem, independent of bidirectional typing.
-This angle uncovers nuances regarding the lack of enough type annotations in bidirectional typing, generally overlooked in literature, and distinguishes completeness (with respect to a type assignment system) from annotatability, concepts conflated by \citet{Dunfield2021}.
-To clarify the difference (\Cref{sec:annotatability}), we propose \emph{mode preprocessing} (\Cref{sec:mode-preprocessing}), which assigns a mode to a raw term with enough annotations, or pinpoints missing type annotations otherwise.
-By combining bidirectional type synthesis, soundness, completeness, and mode preprocessing, we arrive our main result (\Cref{cor:trichotomy}) modelling a type synthesiser which checks if a given raw term is \emph{suitably annotated} and if a suitably annotated raw term has a typing derivation with respect to the specified simple type theory, instead of its bidirectional type system.
-Our bidirectional type synthesis works for any mode-correct signature and allows us to instantiate it to any specified system, effectively a verified type-checker generator.
+This angle uncovers nuances regarding the lack of enough type annotations in bidirectional typing, often overlooked in literature, and distinguishes completeness (with respect to a type assignment system) from annotatability, concepts conflated by \citet{Dunfield2021}.
+To clarify the difference (\Cref{sec:annotatability}), we propose \emph{mode derivations} (\Cref{fig:mode-derivations,fig:generalised-mode-derivations}) and \emph{mode preprocessing} (\Cref{sec:mode-preprocessing}) assigning a mode to a raw term with enough annotations or otherwise pinpoints missing annotations.
+By combining bidirectional type synthesis, soundness, completeness, and mode preprocessing, we arrive our main result (\Cref{cor:trichotomy}) modelling a type synthesiser which checks if a given raw term is \emph{suitably annotated} and if a suitably annotated raw term has a typing derivation with respect to the specified simple type theory---a \emph{trichotomy} of raw terms.
+Our bidirectional type synthesis can be instantiated for any system specified by a mode-correct signature, effectively a verified type-checker generator.
 
@@
 \end{enumerate}
-Moreover, all these concepts have been formally proved in \Agda.
+Moreover, all these concepts have been formally developed in \Agda.
 
@@
   \item Introduction (\Cref{sec:intro}) 
-    \LT{1+3 pp}
+    \LT{4.5 pp}
   \item Key ideas (\Cref{sec:key-ideas})

### 2023-06-26 tex/sec1-intro.tex
[ tex ] Remove annotatability from abstract; add soundness instead.

@@
 Accordingly, our theory's objective is not to formulate rules for various language features, but to introduce a formalism for bidirectional typing, analogous to grammar in parsing.
-We introduce the notion of bidirectional binding signatures to specify bidirectional type systems and characterise essential criteria including \emph{completeness}, \emph{annotatability}, and \emph{mode-correctness} that are sufficient to derive a type synthesis algorithm, as informally outlined by~\citet{Dunfield2021}.
+We introduce the notion of bidirectional binding signatures to specify bidirectional type systems and characterise essential criteria including \emph{soundness}, \emph{completeness}, and \emph{mode-correctness} that are sufficient to derive a type synthesis algorithm, as informally outlined by~\citet{Dunfield2021}.
 For simplicity, we confine our discussion to bidirectional simple type systems that have \emph{syntax-directed} typing rules and leave the problem of presenting bidirectional typing for more advanced type theories in general as future work.

### 2023-06-26 tex/sec1-intro.tex
[ tex ] Move the discussion on polymorphic type theories and dependent type theories to Concluding Remarks

@@
 
-The core idea of bidirectional type synthesis is simple.
-Briefly speaking, it begins with bidirectional typing, extending typing judgements with two modes that direct the flow of types:
+The idea of bidirectional type synthesis begins with extending typing judgements with two modes that direct the flow of types:
 \begin{enumerate*}
-  \item $\Gamma |- \isTerm{t} \syn A$ for synthesis, and 
+  \item $\Gamma |- \isTerm{t} \syn A$ for synthesis and 
   \item $\Gamma |- \isTerm{t} \chk A$ for checking.
@@
 The former means that the type of a raw term is synthesised, using both the term and its context as input, while for the latter a raw term is to be checked against a given type.
-For instance, the type of a variable $x$ is synthesised by looking it up in its context; the type of an annotated term $t \annotate A$ is synthesised as $A$ after $t$ has been checked against $A$.
-From this, a bidirectional type synthesis algorithm can be `read off' provided that a conditional called \emph{mode-correctness} is met for every typing rule---every synthesised type in a typing rule is determined by previously synthesised types from its premises and its input if checking.
+Then a type synthesis algorithm can be `read off' from a bidirectional type system, provided that a conditional called \emph{mode-correctness} -- every synthesised type in a typing rule is determined by previously synthesised types from its premises and its input if checking --- is satisfied for every bidirectional typing rule.
 
@@
 Further, adding annotations for top-level definitions improves clarity, making the purpose of a program easier to understand, so it should be considered useful for the sake of documentation.
-It is fair to say that bidirectional type synthesis is the foundation of type-checking technology across diverse languages.
+It is fair to say that bidirectional type synthesis is the modern foundation of type-checking technology across diverse languages.
 
@@
 
-Earlier than the \PoplMark challenge already, \citet{Altenkirch1993} commented that rudimentary meta-operations and meta-properties for substitution constitute the bulk of language formalisation, motivating a number of language-formalisation frameworks~\citep{Ahrens2018,Fiore2022,Gheri2020,Ahrens2022,Allais2021} to define the type of well-scoped/typed terms, substitution, term traversal and their meta-properties universally.
-One of the core gadgets of these frameworks is the concept of \emph{descriptions} or \emph{binding signatures} (coined by~\citet{Aczel1978} in line with the term \emph{signature} in universal algebra) for specifying typing rules of a language.
-It is noteworthy that these state-of-the-art frameworks cannot specify polymorphic type theories or dependent type theories yet and only \citeauthor{Allais2021} discussed generic programs beyond substitution.
-
 \begin{remark}\label{re:type-synthesis-as-decidability-proof}
@@
   \item In contrast, an algorithm typically found in the literature is presented as \emph{algorithmic system} relations such as $\Gamma |- \isTerm{t} \syn A \mapsto \isTerm{t}'$, denoting that annotations can be added to $\isTerm{t}$ in the external language to produce $\isTerm{t}'$ of type $A$ in the internal language.
-    Such an algorithm is often accompanied with \emph{soundness} and \emph{completeness} assertions in the sense that the algorithm correctly synthesises the type of a raw term and every typable term can be synthesised if there are enough annotations.
+    Such an algorithm is often accompanied with \emph{soundness} and \emph{completeness} assertions such that the algorithm correctly synthesises the type of a raw term and every typable term can be synthesised if there are enough annotations.
 \end{enumerate*}
 \end{remark}
+
+Earlier than the \PoplMark Challenge, \citet{Altenkirch1993} commented that rudimentary meta-operations and meta-properties for substitution constitute the bulk of language formalisation, motivating a number of language-formalisation frameworks~\citep{Ahrens2018,Fiore2022,Gheri2020,Ahrens2022,Allais2021} to define the type of well-scoped/typed terms, substitution, term traversal and their meta-properties universally.
+One of the core gadgets of these frameworks is the concept of \emph{descriptions} or \emph{binding signatures} (coined by~\citet{Aczel1978} in line with the term \emph{signature} in universal algebra) for specifying typing rules of a language.
+It is noteworthy that these state-of-the-art frameworks cannot specify polymorphic type theories or dependent type theories yet and only \citeauthor{Allais2021} discussed meta-operations beyond substitution.
 
@@
 
+Constructing the initial algebra of terms boils down to defining an inductive type with a few constructors that align with the variable rule and the rule scheme for language constructs specified by a signature~\citep{Allais2021,Fiore2022}, without the need to explicitly define the colimit of a sequence of objects iteratively defined by the signature functor.
+
 Substitution can also be modelled categorically, but it does not play a role in this paper, though.
-
-\begin{remark}
-Constructing the initial algebra of terms boils down to defining an inductive type with a few constructors that align with the variable rule and the rule scheme for language constructs specified by a signature~\citep{Allais2021,Fiore2022}, without the need to explicitly define the colimit of a sequence of objects iteratively defined by the signature functor.
-\end{remark}
 
@@
 \end{remark}
-
-While algebraic approaches to polymorphic type theories have been developed~\citep{Fiore2013,Hamana2011}, these approaches do not take subtyping into account.
-Subtyping is essential for formulating important concepts in type synthesis such as \emph{principal types}.
-On the other hand, in the realm of dependent types, \citeauthor{Cartmell1986}'s generalized algebraic theories~\citeyearpar{Cartmell1986} can handle a wide variety of dependent type theories.
-\citet{Bezem2021} investigate the notion of presentation (extending the notion of signature) in the context of generalized algebraic theories.
-Nonetheless, type synthesis for dependent types requires normalisation or some form of conversion to check type equality.
-Normalisation in its generic form still remains out of reach and recent advances in this topic are discussed in the doctoral thesis by~\citet{Valliappan2023}.
 
@@
 Moreover, we exploit the nature of constructive logic to formulate algorithmic soundness, completeness, and decidability. 
-To be more specific, recall that the law of excluded middle does not hold universally as an axiom, thus making a \emph{decidable statement} $P \vee \neg P$ non-trivial to prove.
-Given that all proofs as programs terminate, \emph{logical decidability} inherently implies \emph{algorithmic decidability}.
+Recall that the law of excluded middle does not hold universally as an axiom in Martin-L\"of type theory, thus making a \emph{decidable statement} $P \vee \neg P$ non-trivial to prove.
+Given that all proofs as programs terminate, logical decidability implies \emph{algorithmic decidability}.
 Further, a proof of the following decidable statement:
@@
 \end{quote}
-computes a type and a typing derivation for a context and a raw term, echoing the algorithm given by \citeauthor{Wadler2022} as noted in \Cref{re:type-synthesis-as-decidability-proof}.
-Both algorithmic soundness and completeness are implicitly stated:
-the proof either yields a typing derivation for the given raw term~$t$ or a proof that such a derivation is impossible.
-Accordingly, the former case is algorithmic soundness, while the latter is computationally stronger than completeness.
-Hence, our decidable bidirectional type synthesis (\Cref{thm:bidirectional-type-synthesis-checking}) encapsulates algorithmic soundness, completeness, and decidability \emph{simultaneously}.
+computes a type and a derivation from a context and a raw term, echoing the algorithm given by \citeauthor{Wadler2022} as noted in \Cref{re:type-synthesis-as-decidability-proof}.
+Both algorithmic soundness and completeness are implied:
+the proof either yields a typing derivation for the given raw term~$t$ or a proof that such a derivation is impossible where the former case is algorithmic soundness and the latter is the contrapositive form of completeness.
+That is, our decidable bidirectional type synthesis (\Cref{thm:bidirectional-type-synthesis-checking}) provides algorithmic soundness, completeness, and decidability all at once.
 
 In contrast to \citeauthor{Wadler2022}, our theory starts with raw terms and a general type synthesis problem, independent of bidirectional typing.
-This angle uncovers nuances regarding the lack of enough type annotations in bidirectional typing, often overlooked in literature, and distinguishes completeness (with respect to a type assignment system) from annotatability, concepts conflated by \citet{Dunfield2021}.
+This starting point uncovers nuances regarding the lack of enough type annotations in bidirectional typing, often overlooked in literature, and thus distinguishes completeness (with respect to a type assignment system) from annotatability, concepts conflated by \citet{Dunfield2021}.
 To clarify the difference (\Cref{sec:annotatability}), we propose \emph{mode derivations} (\Cref{fig:mode-derivations,fig:generalised-mode-derivations}) and \emph{mode preprocessing} (\Cref{sec:mode-preprocessing}) assigning a mode to a raw term with enough annotations or otherwise pinpoints missing annotations.
@@
   \item \emph{simple yet general} for any system that can be specified by a bidirectional binding signature;
-  \item \emph{constructive}, in terms of: 
+  \item \emph{constructive}, preferring
     \begin{enumerate*}
-      \item prioritising logical decidability over algorithmic soundness, completeness, and decidability,
-      \item preferring mode preprocessing over annotatability.
+      \item logical decidability over algorithmic soundness, completeness, and decidability;
+      \item mode preprocessing over annotatability.
     \end{enumerate*}
@@
   \item Introduction (\Cref{sec:intro}) 
-    \LT{4.5 pp}
+    \LT{4 pp}
   \item Key ideas (\Cref{sec:key-ideas})
@@
 \end{enumerate}
-  \LT{better title?}

### 2023-06-26 tex/sec7-future.tex
[ tex ] Add points to address; finalise the paper title
[ tex ] Added acmart files; separate compilation for TeXShop
[ tex ] Writing job distribution
[ tex ] Move the discussion on polymorphic type theories and dependent type theories to Concluding Remarks

@@
-%! TEX root = BiSig.tex
+%!TEX root = BiSig.tex
 
-\section{Future Work} \label{sec:future}
+\section{Concluding Remarks} \label{sec:future}
 
-\Josh{Getting rid of the assumption of being syntax-directed (the triangular picture)}
+\Josh{Getting rid of the assumption of being syntax-directed (the triangular picture) --- 0.5p}
+\Josh{Ornaments -- 0.25p}
+\Josh{NDGP -- 0.25p (optional?), \citep{Ko2022} }
 
-\paragraph{Polymorphic Algebraic Theories}
-\cite{Fiore2013,Hamana2011,Hamana2022}
+\LT{polymorphic algebraic theories, dependent signatures --- 0.5p}
+
+\paragraph{Beyond simple types}
+While algebraic approaches to polymorphic type theories have been developed~\citep{Fiore2013,Hamana2011}, these approaches do not take subtyping into account.
+Subtyping is essential for formulating important concepts in type synthesis such as \emph{principal types}.
+On the other hand, in the realm of dependent types, \citeauthor{Cartmell1986}'s generalized algebraic theories~\citeyearpar{Cartmell1986} can handle a wide variety of dependent type theories.
+\citet{Bezem2021} investigate the notion of presentation (extending the notion of signature) in the context of generalized algebraic theories.
+Nonetheless, type synthesis for dependent types requires normalisation or some form of conversion to check type equality.
+Normalisation in its generic form still remains out of reach and recent advances in this topic are discussed in the doctoral thesis by~\citet{Valliappan2023}.

### 2023-06-26 tex/sec1-intro.tex
[ tex ] Minor changes in S1

@@
 \subsection{Related work}\label{sec:related-work}
-As we will formalise our theory for not only simple type theories but also in a manner distinct from conventional formulations in bidirectional typing, it is pertinent to discuss related work upon which this work is built, apart from bidirectional typing, in order to articulate our contributions.
+As we will formalise our theory for not only simple type theories but also in a manner distinct from typical formulations in bidirectional typing, it is pertinent to discuss related work upon which this work is built, apart from bidirectional typing, in order to articulate our contributions.
 
@@
 The vision of formalising the metatheory of every programming language was initiated by the \PoplMark challenge~\citep{Aydemir2005}.
-This is has been further exemplified in a textbook by~\citet{Wadler2022}, where language concepts are formally addressed using the proof assistant~\Agda, including bidirectional type synthesis for~\PCF.
+This is has been exemplified in a textbook by~\citet{Wadler2022}, where language concepts are formally addressed using the proof assistant~\Agda, including bidirectional type synthesis for~\PCF.
 
 \begin{remark}\label{re:type-synthesis-as-decidability-proof}
-The algorithm of bidirectional type synthesis formulated by~\citeauthor{Wadler2022} differs from existing literature:
+The bidirectional type synthesis formulated by~\citeauthor{Wadler2022} differs from existing literature:
 \begin{enumerate*}
-  \item The algorithm is presented as a proof of \emph{(logical) decidability} as to whether a `bidirectionally decorated' raw term~$\isTerm{t}$ can be bidirectionally typed, equivalent to a program that returns a typing derivation for $\isTerm{t}$, if one exists, or the evidence of a contradiction otherwise.
+  \item Its algorithm is presented as a proof of \emph{logical decidability} as to whether a `bidirectionally decorated' raw term~$\isTerm{t}$ can be bidirectionally typed, equivalent to a program that returns a typing derivation for $\isTerm{t}$ or otherwise the proof that such derivation is impossible.
   \item In contrast, an algorithm typically found in the literature is presented as \emph{algorithmic system} relations such as $\Gamma |- \isTerm{t} \syn A \mapsto \isTerm{t}'$, denoting that annotations can be added to $\isTerm{t}$ in the external language to produce $\isTerm{t}'$ of type $A$ in the internal language.
-    Such an algorithm is often accompanied with \emph{soundness} and \emph{completeness} assertions such that the algorithm correctly synthesises the type of a raw term and every typable term can be synthesised if there are enough annotations.
+    Such an algorithm is then accompanied with \emph{soundness} and \emph{completeness} assertions such that the algorithm correctly synthesises the type of a raw term and every typable term can be synthesised if there are enough annotations.
 \end{enumerate*}
@@
 Given that all proofs as programs terminate, logical decidability implies \emph{algorithmic decidability}.
-Further, a proof of the following decidable statement:
+Further, suppose that we have a proof of the following statement:
 \begin{quote}
@@
 \begin{quote}
-  `A context $\Gamma$ and a raw term $t$ \emph{decide} whether $\Gamma |- t : A$ has a derivation for some $A$'
+  `A context $\Gamma$ and a raw term $t$ \emph{decide} whether $\Gamma |- t : A$ has a derivation for some $A$',
 \end{quote}
-computes a type and a derivation from a context and a raw term, echoing the algorithm given by \citeauthor{Wadler2022} as noted in \Cref{re:type-synthesis-as-decidability-proof}.
+echoing the algorithm given by \citeauthor{Wadler2022} as noted in \Cref{re:type-synthesis-as-decidability-proof}.
 Both algorithmic soundness and completeness are implied:

### 2023-06-27 tex/sec4-pre-synthesis.tex
[ tex ] Update the discussion about annotatability

@@
 \citet{Dunfield2021} formulated completeness differently from ours (\cref{thm:completeness}) and proposed \emph{annotatability} as a more suitable name.
-Here we reformulate annotatability in our theory and discuss its relationship with our completeness and generalised mode preprocessing.
+Here we formulate annotatability in our theory and discuss its relationship with our completeness and generalised mode preprocessing.
 
@@
 Defined in \cref{fig:annotation-ordering}, the `annotation ordering' $t' \sqsupseteq t$ means that $t'$~has the same or more annotations than~$t$.
-In a sense, annotatability is a reasonable form of completeness: if a term is typable in $(\Sigma, \erase\Omega)$, it may not be typable in $(\Sigma, \Omega)$ directly, but will be if some annotations are added.
-Indeed, \cref{thm:annotatability} can be easily proved in our theory from generalised mode preprocessing.
+In a sense, annotatability is a reasonable form of completeness: if a term is typable in $(\Sigma, \erase\Omega)$, it may not be typable in $(\Sigma, \Omega)$ directly due to some missing annotations, but will be if those annotations are added correctly.
+In our theory, \cref{thm:annotatability} can be easily proved from generalised mode preprocessing.
 
@@
 By \cref{thm:generalised-mode-preprocessing}, there is a generalised mode derivation of $V |-_{\Sigma,\erase\Omega} t^{\dir{d}\,g\,s}$ for some mode~$\dir{d}$ and booleans $g$~and~$s$.
-Perform induction on this generalised mode derivation to construct a bidirectional typing derivation in the same mode (and also the term~$t'$ and a proof of $t' \sqsupseteq t$, which are straightforward and omitted here).
+Perform induction on this generalised mode derivation to construct a bidirectional typing derivation in the same mode (and also the term~$t'$ and a proof of $t' \sqsupseteq t$, which are determined by the derivations and omitted here).
 \begin{itemize}
@@
 
-On the other hand, when using a bidirectional type synthesiser to implement type synthesis~(\cref{thm:implementation}), annotatability cannot replace our completeness since it relates typing derivations for possibly differently annotated terms, but a bidirectional type synthesiser is concerned about properties of \emph{one} given term throughout the process.
-Annotatability also does not help the user to deal with missing annotations like generalised mode preprocessing does: even though annotatability seems to construct a term with missing annotations filled in, its antecedent requires a typing derivation, which is what the user is trying to construct and does not have yet.
-It seems that as we focus on properties that serve practical purposes, it becomes unclear how annotatability comes into the picture.
-In particular, annotatability mixes up the aspects of annotating and typing, and therefore does not fit into our theory, where the two aspects are separated for better usability.
+On the other hand, when using a bidirectional type synthesiser to implement type synthesis~(\cref{thm:implementation}), our completeness is much simpler to use than annotatability, which requires the bidirectional type synthesiser to produce more complex evidence when the synthesis fails.
+Annotatability also does not help the user to deal with missing annotations like (generalised) mode preprocessing does: although annotatability seems capable of determining where annotations are missing and even filling them in correctly, its antecedent requires a typing derivation, which is what the user is trying to construct and does not have yet.
+Therefore we believe that the notion of annotatability is too complex for what it achieves, and our theory offers simpler and more useful alternatives.

### 2023-06-27 tex/BiSig.tex
[ tex ] Minor changes in S1
[ tex ] Done with S1 for now

@@
 \setlength {\marginparwidth }{2cm}
-\usepackage[color=yellow,textsize=small]{todonotes}
+\usepackage[obeyFinal,color=yellow,textsize=small]{todonotes}
 
@@
   Type synthesis and mode preprocessing result in the trichotomy of raw terms, which models a practical type synthesiser that may throw an exception and locate missing annotations.
-  By formalizing our general theory of bidirectional type synthesis constructively, we provide a correct-by-construction `type-checker generator' for simple type theories, taking the first step towards a more general theory of bidirectional type synthesis.
+  By formalising our general theory of bidirectional type synthesis constructively, we provide a correct-by-construction `type-checker generator' for simple type theories, taking the first step towards a more general theory of bidirectional type synthesis.
 \end{abstract}
@@
 
-\keywords{\Agda, abstract syntax, bidirectional typing, type synthesis, type checking, type inference, language formalisation, datatype-generic programming}
+\keywords{bidirectional typing, type synthesis, type checking, type inference, language formalisation, \Agda}
 

### 2023-06-27 tex/sec1-intro.tex
[ tex ] Done with S1 for now

@@
 However, it was later found that full parametric polymorphism leads to undecidability in type inference, as do dependent types~\citep{Wells1999,Dowek1993}.
-In light of these limitations, bidirectional type synthesis emerged as a viable alternative, providing decidable algorithms for determining the types of \emph{suitably annotated} programs in languages with more complex types.
+In light of these limitations, bidirectional type synthesis emerged as a viable alternative, providing decidable algorithms for determining the types of \emph{suitably annotated} programs in languages with simple types, polymorphic types, dependent types, among others.
 \citet{Dunfield2021} summarised its design principles and applications that we know so far.
@@
 \end{enumerate*}
-The former means that the type of a raw term is synthesised, using both the term and its context as input, while for the latter a raw term is to be checked against a given type.
-Then a type synthesis algorithm can be `read off' from a bidirectional type system, provided that a conditional called \emph{mode-correctness} -- every synthesised type in a typing rule is determined by previously synthesised types from its premises and its input if checking --- is satisfied for every bidirectional typing rule.
+The former means that the type of a term is synthesised, using both the term and its context as input, while for the latter a term is to be checked against a given type.
+Then, a type synthesis algorithm can be `read off' from a bidirectional type system, provided that a conditional called \emph{mode-correctness} -- every synthesised type in a typing rule is determined by previously synthesised types from its premises and its input if checking --- is satisfied for every bidirectional typing rule.
 
-Distinct from the classic Damas--Milner type inference, bidirectional type synthesis does \emph{not} require unification.
-As noted by~\citet{Pierce2000}, annotations are propagated locally within adjacent nodes of a syntax tree, so introducing long-distance unification constraints in bidirectional type synthesis undermines its essence of locality.
-Further, adding annotations for top-level definitions improves clarity, making the purpose of a program easier to understand, so it should be considered useful for the sake of documentation.
-It is fair to say that bidirectional type synthesis is the modern foundation of type-checking technology across diverse languages.
+Contrary to the classic Damas--Milner type inference, bidirectional type synthesis does \emph{not} require unification.
+\citet{Pierce2000} noted that bidirectional type synthesis propagates annotations locally within adjacent nodes of a syntax tree, so introducing long-distance unification constraints undermines the essence of locality in bidirectional type synthesis.
+Further, annotations for, say, top-level definitions improve clarity, making the purpose of a program easier to understand, so they should not be considered redundant.
+In light of these considerations, bidirectional type synthesis can be deemed as a technique that is more fundamental than unification and is a general design paradigm capable of handling a broad spectrum of programming languages.
 
 Unlike parsing, nevertheless, which has a comprehensive theory and widely applicable practical techniques, bidirectional typing has only been developed on a case-by-case basis without a theory.
-While it is easy to derive a bidirectional type synthesis algorithm, a general theory providing logical specifications with rigorously proven properties for a broad range of systems is still lacking.
+While it is straightforward to derive a type synthesis algorithm, a general theory providing logical specifications and rigorously proven properties for a class of systems is still lacking.
 As a result, we can only present design principles that we learned from individual systems loosely.
-Moreover, unlike the plethora of available parser generators, `type-checker generators' rarely exist, so each type checker has to be independently built from scratch.
+Moreover, unlike the plethora of available parser generators, `type-checker generators' rarely exist, so each type checker, based on bidirectional typing or not, has to be independently built from scratch.
 
@@
 We introduce the notion of bidirectional binding signatures to specify bidirectional type systems and characterise essential criteria including \emph{soundness}, \emph{completeness}, and \emph{mode-correctness} that are sufficient to derive a type synthesis algorithm, as informally outlined by~\citet{Dunfield2021}.
-For simplicity, we confine our discussion to bidirectional simple type systems that have \emph{syntax-directed} typing rules and leave the problem of presenting bidirectional typing for more advanced type theories in general as future work.
+For simplicity, we confine our discussion to bidirectional simple type systems that have \emph{syntax-directed} typing rules and leave the problem of presenting bidirectional typing for more general and advanced type theories as future work.
 
 \subsection{Related work}\label{sec:related-work}
-As we will formalise our theory for not only simple type theories but also in a manner distinct from typical formulations in bidirectional typing, it is pertinent to discuss related work upon which this work is built, apart from bidirectional typing, in order to articulate our contributions.
+As our theory is developed formally not only for a class of simple type theories but also in a manner distinct from typical formulations in bidirectional typing, it is pertinent to discuss related work upon which this work is built, apart from bidirectional typing, in order to explain our contributions.
 
@@
 The vision of formalising the metatheory of every programming language was initiated by the \PoplMark challenge~\citep{Aydemir2005}.
-This is has been exemplified in a textbook by~\citet{Wadler2022}, where language concepts are formally addressed using the proof assistant~\Agda, including bidirectional type synthesis for~\PCF.
+This has been exemplified in the textbook by~\citet{Wadler2022} where language concepts are formally addressed using the proof assistant~\Agda, including bidirectional type synthesis for~\PCF.
 
@@
 
-Earlier than the \PoplMark Challenge, \citet{Altenkirch1993} commented that rudimentary meta-operations and meta-properties for substitution constitute the bulk of language formalisation, motivating a number of language-formalisation frameworks~\citep{Ahrens2018,Fiore2022,Gheri2020,Ahrens2022,Allais2021} to define the type of well-scoped/typed terms, substitution, term traversal and their meta-properties universally.
+Earlier than the \PoplMark challenge, \citet{Altenkirch1993} commented that rudimentary meta-operations and their meta-properties constitute the bulk of formalisation, motivating a number of frameworks~\citep{Ahrens2018,Fiore2022,Gheri2020,Ahrens2022,Allais2021} to define the type of well-scoped/typed terms, substitution, term traversal and their meta-properties universally.
 One of the core gadgets of these frameworks is the concept of \emph{descriptions} or \emph{binding signatures} (coined by~\citet{Aczel1978} in line with the term \emph{signature} in universal algebra) for specifying typing rules of a language.
@@
 The aforementioned frameworks except \citeauthor{Gheri2020}'s are at least inspired by \citet{Fiore1999}'s initial semantics for abstract syntax with variable binding using category theory.
-The main idea is that the set of (untyped) abstract syntax trees for a language can be interpreted as the initial algebra for its signature endofunctor $\mathbb{\Sigma}$ on a family $\Term_{\Gamma}$ of well-scoped terms under the context~$\Gamma$ with
-variable renaming for a function $\sigma\colon \Gamma \to \Delta$ between variables acting as a functorial map from $X_{\Gamma}$ to $X_{\Delta}$, equivalently a free $\mathbb{\Sigma}$-algebra on the presheaf of variables $V\colon \mathbb{F} \hookrightarrow \mathsf{Set}$.
-The endofunctor $\mathbb{\Sigma}$ encapsulates language constructs (except variables) while the initiality amounts to structural recursion, i.e.\ \emph{term traversal}.
+The main idea is that the set of (untyped) abstract syntax trees for a language consists of
+\begin{enumerate*}
+  \item a family $\Term_{\Gamma}$ of well-scoped terms under the context~$\Gamma$ with
+  \item variable renaming for a function $\sigma\colon \Gamma \to \Delta$ between variables acting as a functorial map from $X_{\Gamma}$ to $X_{\Delta}$, thus a presheaf $\Term\colon \mathbb{F} \to \mathsf{Set}$, and
+  \item an initial algebra structure given by the variable rule as a map from the embedding $V\colon \mathbb{F} \hookrightarrow \mathsf{Set}$ to $\Term$ and other language constructs as $\mathbb{\Sigma}\Term \to \Term$ where $\mathbb{F}$ is the category of contexts, the functor $\mathbb{\Sigma}\colon \mathsf{Set}^\mathbb{F} \to \mathsf{Set}^\mathbb{F}$ encapsulates language constructs (except variables), and the initiality amounts to structural recursion, i.e.\ \emph{term traversal}.
+\end{enumerate*}
+To put it succinctly, it is the free $\mathbb{\Sigma}$-algebra over the presheaf of variables $V\colon \mathbb{F} \hookrightarrow \mathsf{Set}$.
 
-Constructing the initial algebra of terms boils down to defining an inductive type with a few constructors that align with the variable rule and the rule scheme for language constructs specified by a signature~\citep{Allais2021,Fiore2022}, without the need to explicitly define the colimit of a sequence of objects iteratively defined by the signature functor.
+Fortunately, constructing the initial algebra of terms in type theory boils down to defining an inductive type with a few constructors that align with the variable rule and a rule scheme for language constructs specified by a signature~\citep{Allais2021,Fiore2022}.
 
-Substitution can also be modelled categorically, but it does not play a role in this paper, though.
+Substitution is also modelled categorically, but it does not play a role in this paper, though.
 
@@
 
-First, our theory can be viewed as a counterpart to those theories in \Cref{sec:theory-of-syntax} and is tailored towards \emph{extrinsically-typed}, \emph{syntax-directed} bidirectional type systems of simple types (\Cref{fig:raw-terms,fig:bidirectional-typing-derivations}) specified by a \emph{signature} for simple types (c.f.\ \Cref{re:type-signature}) and a \emph{bidirectional binding signature} for terms (\Cref{def:bidirectional-binding-signature}).
+First, our theory can be viewed as a counterpart to those theories in \Cref{sec:theory-of-syntax} and is tailored towards \emph{extrinsically-typed}, \emph{syntax-directed} bidirectional type systems of simple types (\Cref{fig:raw-terms,fig:bidirectional-typing-derivations}) specified by a \emph{signature} for simple types and a \emph{bidirectional binding signature} for terms (\Cref{def:bidirectional-binding-signature} and c.f.\ \Cref{re:type-signature}).
 The extrinsic typing is needed for type synthesis and our adherence to syntax-directed systems aligns with the standard idea of re-casting non-syntax-directed typing rules into their syntax-directed form to derive type synthesis algorithm~\citep{Peyton-Jones2007}.
-As operational semantics are not needed for synthesising simple types, we do not address substitution.
+As substitution is not needed for synthesising simple types, it is not addressed.
 Instead, our focus is on:
 \begin{enumerate*}
-  \item establishing the connection between the specified bidirectional type system and its simple type theory, i.e.\ \emph{soundness} and \emph{completeness} (\Cref{sec:soundness-and-completeness}); 
-  \item formulating the \emph{mode-correctness} condition (\Cref{def:mode-correctness}) which suffices to establish the uniqueness of synthesised types (\Cref{thm:unique-syn}) and a decidable bidirectional type synthesis and checking (\Cref{thm:bidirectional-type-synthesis-checking}).
+  \item establishing the link between the specified bidirectional type system and its simple type theory, i.e.\ \emph{soundness} and \emph{completeness} (\Cref{sec:soundness-and-completeness}); 
+  \item defining the \emph{mode-correctness} condition (\Cref{def:mode-correctness}) which suffices to establish the uniqueness of synthesised types (\Cref{thm:unique-syn}) and a decidable bidirectional type synthesis and checking (\Cref{thm:bidirectional-type-synthesis-checking}).
 \end{enumerate*}
@@
 Our theory is based on Martin-L\"of type theory and formalised in \Agda (\Cref{sec:formalisation}), akin to those frameworks in \Cref{sec:language-formalisation}.
-Moreover, we exploit the nature of constructive logic to formulate algorithmic soundness, completeness, and decidability. 
-Recall that the law of excluded middle does not hold universally as an axiom in Martin-L\"of type theory, thus making a \emph{decidable statement} $P \vee \neg P$ non-trivial to prove.
+More importantly, we exploit the computational and logical aspects of type theory to formulate algorithmic soundness, completeness, and decidability. 
+Recall that the law of excluded middle does not hold universally as an axiom in Martin-L\"of type theory, so a \emph{decidable statement} $P \vee \neg P$ is non-trivial.
 Given that all proofs as programs terminate, logical decidability implies \emph{algorithmic decidability}.
@@
 Both algorithmic soundness and completeness are implied:
-the proof either yields a typing derivation for the given raw term~$t$ or a proof that such a derivation is impossible where the former case is algorithmic soundness and the latter is the contrapositive form of completeness.
-That is, our decidable bidirectional type synthesis (\Cref{thm:bidirectional-type-synthesis-checking}) provides algorithmic soundness, completeness, and decidability all at once.
+the proof either yields a typing derivation for the given raw term~$t$ or a proof that such a derivation is impossible where the former case is algorithmic soundness and the latter is algorithmic completeness in contrapositive form.
+That is, our decidable bidirectional type synthesis (\Cref{thm:bidirectional-type-synthesis-checking}) provides algorithmic soundness, completeness, and decidability.
 
@@
 This starting point uncovers nuances regarding the lack of enough type annotations in bidirectional typing, often overlooked in literature, and thus distinguishes completeness (with respect to a type assignment system) from annotatability, concepts conflated by \citet{Dunfield2021}.
-To clarify the difference (\Cref{sec:annotatability}), we propose \emph{mode derivations} (\Cref{fig:mode-derivations,fig:generalised-mode-derivations}) and \emph{mode preprocessing} (\Cref{sec:mode-preprocessing}) assigning a mode to a raw term with enough annotations or otherwise pinpoints missing annotations.
-By combining bidirectional type synthesis, soundness, completeness, and mode preprocessing, we arrive our main result (\Cref{cor:trichotomy}) modelling a type synthesiser which checks if a given raw term is \emph{suitably annotated} and if a suitably annotated raw term has a typing derivation with respect to the specified simple type theory---a \emph{trichotomy} of raw terms.
+To clarify the difference (\Cref{sec:annotatability}), we propose \emph{mode derivations} (\Cref{fig:mode-derivations,fig:generalised-mode-derivations}) and \emph{mode preprocessing} (\Cref{sec:mode-preprocessing}) assigning a mode first to a raw term with enough annotations or otherwise pinpoints missing annotations.
+By combining bidirectional type synthesis, soundness, completeness, and mode preprocessing, we achieve our main result (\Cref{cor:trichotomy}) --- a \emph{trichotomy} of raw terms: a type synthesiser which checks if a given raw term is \emph{suitably annotated} and if a suitably annotated raw term has a typing derivation with respect to the specified simple type theory.
 Our bidirectional type synthesis can be instantiated for any system specified by a mode-correct signature, effectively a verified type-checker generator.
@@
   \item \emph{simple yet general} for any system that can be specified by a bidirectional binding signature;
-  \item \emph{constructive}, preferring
-    \begin{enumerate*}
-      \item logical decidability over algorithmic soundness, completeness, and decidability;
-      \item mode preprocessing over annotatability.
-    \end{enumerate*}
+  \item \emph{constructive}, based on Martin-L\"of theory and preferring logical decidability over algorithmic soundness, completeness, and decidability; mode preprocessing over annotatability.
 \end{enumerate}
@@
 
-Plan of this paper:
+This paper is structured in logical order as follows.
+We first present an overview of our theory using simply typed $\lambda$-calculus in \Cref{sec:key-ideas}, prior to developing a general framework for specifying bidirectional type systems in \Cref{sec:defs}.
+Following this, we discuss the connection between a specified bidirectional type system and its associated simple type theory in \Cref{sec:pre-synthesis}.
+In \Cref{sec:type-synthesis}, we introduce mode-correctness and bidirectional type synthesis.
+We conclude by reflecting on related topics and potential challenges in extending our theory to a more general setting.
+
+\todo[inline,caption={}]{
 \begin{enumerate}
@@
 \end{enumerate}
+}

### 2023-06-27 tex/sec1-intro.tex
[ tex ] Minor changes

@@
 However, it was later found that full parametric polymorphism leads to undecidability in type inference, as do dependent types~\citep{Wells1999,Dowek1993}.
-In light of these limitations, bidirectional type synthesis emerged as a viable alternative, providing decidable algorithms for determining the types of \emph{suitably annotated} programs in languages with simple types, polymorphic types, dependent types, among others.
+In light of these limitations, bidirectional type synthesis emerged as a viable alternative, providing decidable algorithms for determining the types of \emph{suitably annotated} programs in languages with simple types, polymorphic types, dependent types, gradual types, among others.
 \citet{Dunfield2021} summarised its design principles and applications that we know so far.
@@
 Further, annotations for, say, top-level definitions improve clarity, making the purpose of a program easier to understand, so they should not be considered redundant.
-In light of these considerations, bidirectional type synthesis can be deemed as a technique that is more fundamental than unification and is a general design paradigm capable of handling a broad spectrum of programming languages.
+In light of these considerations, bidirectional type synthesis can be deemed as a technique that is more fundamental than unification and is a design paradigm capable of handling a broad spectrum of programming languages.
 
@@
 \subsection{Related work}\label{sec:related-work}
-As our theory is developed formally not only for a class of simple type theories but also in a manner distinct from typical formulations in bidirectional typing, it is pertinent to discuss related work upon which this work is built, apart from bidirectional typing, in order to explain our contributions.
+As our theory is developed formally not only for simple type theories but also in a manner distinct from typical formulations in bidirectional typing, it is pertinent to discuss related work upon which this work is built, apart from bidirectional typing, in order to explain our contributions.
 
@@
 
-First, our theory can be viewed as a counterpart to those theories in \Cref{sec:theory-of-syntax} and is tailored towards \emph{extrinsically-typed}, \emph{syntax-directed} bidirectional type systems of simple types (\Cref{fig:raw-terms,fig:bidirectional-typing-derivations}) specified by a \emph{signature} for simple types and a \emph{bidirectional binding signature} for terms (\Cref{def:bidirectional-binding-signature} and c.f.\ \Cref{re:type-signature}).
+Our theory can be viewed as a counterpart to those theories in \Cref{sec:theory-of-syntax} and is tailored towards \emph{extrinsically-typed}, \emph{syntax-directed} bidirectional type systems of simple types (\Cref{fig:raw-terms,fig:bidirectional-typing-derivations}) specified by a \emph{signature} for simple types and a \emph{bidirectional binding signature} for terms (\Cref{def:bidirectional-binding-signature} and c.f.\ \Cref{re:type-signature}).
 The extrinsic typing is needed for type synthesis and our adherence to syntax-directed systems aligns with the standard idea of re-casting non-syntax-directed typing rules into their syntax-directed form to derive type synthesis algorithm~\citep{Peyton-Jones2007}.
@@
 \begin{quote}
-  `A context $\Gamma$ and a raw term $t$ \emph{decide} whether $\Gamma |- t : A$ has a derivation for some $A$',
+  `A context $\Gamma$ and a raw term $t$ \underline{\emph{decide}} whether $\Gamma |- t : A$ has a derivation for some $A$',
 \end{quote}
@@
 \end{enumerate}
-Moreover, all these concepts have been formally developed in \Agda.
+Moreover, all these concepts have been formally developed with \Agda.
 
@@
 In \Cref{sec:type-synthesis}, we introduce mode-correctness and bidirectional type synthesis.
-We conclude by reflecting on related topics and potential challenges in extending our theory to a more general setting.
+In \Cref{sec:formalisation}, we sketch the formalisation of our theory in \Agda and some examples other than simply typed $\lambda$-calculus.
+We conclude in \Cref{sec:future}, where we reflect on related techniques that can be used to improve the formal development and potential challenges in extending our theory to a more general setting.
 

### 2023-06-27 tex/sec1-intro.tex
[ tex ] Fix em dash in S1

@@
 The former means that the type of a term is synthesised, using both the term and its context as input, while for the latter a term is to be checked against a given type.
-Then, a type synthesis algorithm can be `read off' from a bidirectional type system, provided that a conditional called \emph{mode-correctness} -- every synthesised type in a typing rule is determined by previously synthesised types from its premises and its input if checking --- is satisfied for every bidirectional typing rule.
+Then, a type synthesis algorithm can be `read off' from a bidirectional type system, provided that a conditional called \emph{mode-correctness} --- every synthesised type in a typing rule is determined by previously synthesised types from its premises and its input if checking --- is satisfied for every bidirectional typing rule.
 

### 2023-06-27 tex/sec2-outline.tex
[ tex ] Add an overview figure in S2

@@
 To quantify over all such systems, we need their generic definitions, which we formulate in \cref{sec:defs}.
+
+\begin{figure}
+  \begin{tikzpicture}
+  \matrix (m) [matrix of math nodes, row sep=3em, column sep=10em]
+  {
+    \boxed{\erase{\Gamma} \vdash \isTerm{t}}   & \boxed{\erase{\Gamma} \vdash \isTerm{t}^{\syn}} \\
+    \boxed{\Gamma \vdash \isTerm{t} : A} & \boxed{\Gamma \vdash \isTerm{t} \syn A} \\
+  };
+  \path[->]
+      (m-1-1) edge node [above,align=center] {mode preprocessing} (m-1-2)
+              edge node [left,align=left]  {type\\synthesis} (m-2-1)
+      (m-1-2) edge node [right,align=right] {bidirectional\\type synthesis} (m-2-2)
+      (m-2-1) edge[bend right=8] node [below] {completeness} (m-2-2);
+  \path[->]
+      (m-2-2) edge[bend right=8] node [above] {soundness} (m-2-1);
+  \end{tikzpicture}
+  \caption{An implementation of type synthesis via bidirectional typing}
+\end{figure}

### 2023-06-28 tex/BiSig.tex
[ tex ] (Really) done with S1 for now

@@
 \input{mathlig}
+\usepackage{tikz}
+\usetikzlibrary{matrix,arrows}
 \usepackage{mathpartir}
@@
 \title{A Theory of Bidirectional Type Synthesis for Simple Types}
+
 \begin{abstract}
@@
   First, our theory is simple yet general as it deals with any simple type theory and its bidirectional type system that can be specified by a signature, analogous to the role of grammar in parsing.
-  Second, our theory is constructive because it is not only formalised in a proof assistant based on constructive type theory but also formulated positively in order to compute.
+  Second, our theory is constructive because it is not only formalised in a proof assistant based on constructive type theory but also formulated towards computation.
   Specifically, the `proof' of logical decidability of type synthesis is instantiated to a synthesiser for the simple type theory specified by a valid signature, analogous to a parser generator.

### 2023-06-28 tex/sec1-intro.tex
[ tex ] (Really) done with S1 for now

--- 2023-06-27 tex/sec1-intro.tex
@@
 Type checking and inference serve as the transition to type-checked programs, also known as \emph{well-typed terms}, from parsed abstract syntax trees, referred as \emph{raw terms} by contrast.
-Type inference algorithms were conceived to ascertain the type of any raw term without any type annotations.
+Type inference algorithms were conceived to ascertain the type of any \emph{unannotated} raw term.
 However, it was later found that full parametric polymorphism leads to undecidability in type inference, as do dependent types~\citep{Wells1999,Dowek1993}.
-In light of these limitations, bidirectional type synthesis emerged as a viable alternative, providing decidable algorithms for determining the types of \emph{suitably annotated} programs in languages with simple types, polymorphic types, dependent types, gradual types, among others.
-\citet{Dunfield2021} summarised its design principles and applications that we know so far.
+In light of these limitations, bidirectional type synthesis emerged as a viable alternative, providing decidable algorithms for determining the type of a \emph{suitably annotated} raw term.
+\citet{Dunfield2021} summarised its design principles and its wide coverage of languages with simple types, polymorphic types, dependent types, gradual types, among others.
 
@@
 The former means that the type of a term is synthesised, using both the term and its context as input, while for the latter a term is to be checked against a given type.
-Then, a type synthesis algorithm can be `read off' from a bidirectional type system, provided that a conditional called \emph{mode-correctness} -- every synthesised type in a typing rule is determined by previously synthesised types from its premises and its input if checking --- is satisfied for every bidirectional typing rule.
+Then, a type synthesis algorithm can be `read off' from a bidirectional type system, provided that a condition called \emph{mode-correctness} --- every synthesised type in a typing rule is determined by previously synthesised types from its premises and its input if checking --- is satisfied for every bidirectional typing rule.
 
-Contrary to the classic Damas--Milner type inference, bidirectional type synthesis does \emph{not} require unification.
-\citet{Pierce2000} noted that bidirectional type synthesis propagates annotations locally within adjacent nodes of a syntax tree, so introducing long-distance unification constraints undermines the essence of locality in bidirectional type synthesis.
-Further, annotations for, say, top-level definitions improve clarity, making the purpose of a program easier to understand, so they should not be considered redundant.
-In light of these considerations, bidirectional type synthesis can be deemed as a technique that is more fundamental than unification and is a design paradigm capable of handling a broad spectrum of programming languages.
+Contrary to Damas--Milner type inference, unification is not required for bidirectional type synthesis.
+Moreover, \citet{Pierce2000} noted that bidirectional type synthesis propagates annotations locally within adjacent nodes of a syntax tree, so introducing long-distance unification constraints undermines the essence of locality in bidirectional type synthesis.
+Also, annotations for, say, top-level definitions improve clarity, making the purpose of a program easier to understand, so they should be considered a practical compromise at worst.
+In light of these considerations, bidirectional type synthesis can be deemed as a type-checking technique that is more fundamental than unification, as it is capable of handling a broad spectrum of programming languages.
 
@@
 Accordingly, our theory's objective is not to formulate rules for various language features, but to introduce a formalism for bidirectional typing, analogous to grammar in parsing.
-We introduce the notion of bidirectional binding signatures to specify bidirectional type systems and characterise essential criteria including \emph{soundness}, \emph{completeness}, and \emph{mode-correctness} that are sufficient to derive a type synthesis algorithm, as informally outlined by~\citet{Dunfield2021}.
+We introduce the notion of \emph{bidirectional binding signatures} to specify bidirectional type systems and characterise essential criteria including \emph{soundness}, \emph{completeness}, and \emph{mode-correctness} that are sufficient to derive a type synthesis algorithm, as informally outlined by~\citet{Dunfield2021}.
 For simplicity, we confine our discussion to bidirectional simple type systems that have \emph{syntax-directed} typing rules and leave the problem of presenting bidirectional typing for more general and advanced type theories as future work.
@@
 
-Earlier than the \PoplMark challenge, \citet{Altenkirch1993} commented that rudimentary meta-operations and their meta-properties constitute the bulk of formalisation, motivating a number of frameworks~\citep{Ahrens2018,Fiore2022,Gheri2020,Ahrens2022,Allais2021} to define the type of well-scoped/typed terms, substitution, term traversal and their meta-properties universally.
-One of the core gadgets of these frameworks is the concept of \emph{descriptions} or \emph{binding signatures} (coined by~\citet{Aczel1978} in line with the term \emph{signature} in universal algebra) for specifying typing rules of a language.
-It is noteworthy that these state-of-the-art frameworks cannot specify polymorphic type theories or dependent type theories yet and only \citeauthor{Allais2021} discussed meta-operations beyond substitution.
+Earlier than the \PoplMark challenge, \citet{Altenkirch1993} commented that rudimentary meta-operations and their meta-properties constitute the bulk of formalisation, motivating a number of frameworks~\citep{Ahrens2018,Fiore2022,Gheri2020,Ahrens2022,Allais2021} to define well-scoped/typed terms, substitution, term traversal and their meta-properties universally.
+One of the core gadgets of these frameworks is the notion of \emph{binding signatures} (coined by~\citet{Aczel1978} in line with the term \emph{signature} in universal algebra) for specifying typing rules
+It is noteworthy that these state-of-the-art frameworks cannot specify polymorphic types or dependent types yet and only \citeauthor{Allais2021} discussed meta-operations beyond substitution.
 
@@
 \begin{enumerate*}
-  \item a family $\Term_{\Gamma}$ of well-scoped terms under the context~$\Gamma$ with
-  \item variable renaming for a function $\sigma\colon \Gamma \to \Delta$ between variables acting as a functorial map from $X_{\Gamma}$ to $X_{\Delta}$, thus a presheaf $\Term\colon \mathbb{F} \to \mathsf{Set}$, and
-  \item an initial algebra structure given by the variable rule as a map from the embedding $V\colon \mathbb{F} \hookrightarrow \mathsf{Set}$ to $\Term$ and other language constructs as $\mathbb{\Sigma}\Term \to \Term$ where $\mathbb{F}$ is the category of contexts, the functor $\mathbb{\Sigma}\colon \mathsf{Set}^\mathbb{F} \to \mathsf{Set}^\mathbb{F}$ encapsulates language constructs (except variables), and the initiality amounts to structural recursion, i.e.\ \emph{term traversal}.
+  \item a family of sets $\Term_{\Gamma}$ of well-scoped terms under a context~$\Gamma$ with
+  \item variable renaming for a function $\sigma\colon \Gamma \to \Delta$ between variables acting as a functorial map from $X_{\Gamma}$ to $X_{\Delta}$, i.e.\ a presheaf $\Term\colon \mathbb{F} \to \mathsf{Set}$, and
+  \item an initial algebra $[\mathsf{v}, \mathsf{op}]$ on~$\Term$ given by the variable rule as a map $\mathsf{v}$ from the presheaf~$V$ of variables (i.e.\ the embedding $V\colon \mathbb{F} \hookrightarrow \mathsf{Set}$) to $\Term$ and other constructs as $\mathsf{op}\colon \mathbb{\Sigma}\Term \to \Term$ where $\mathbb{F}$ is the category of contexts, the functor $\mathbb{\Sigma}\colon \mathsf{Set}^\mathbb{F} \to \mathsf{Set}^\mathbb{F}$ encodes the arities of constructs, and the initiality amounts to structural recursion, i.e.\ \emph{term traversal}.
 \end{enumerate*}
-To put it succinctly, it is the free $\mathbb{\Sigma}$-algebra over the presheaf of variables $V\colon \mathbb{F} \hookrightarrow \mathsf{Set}$.
+To put it succinctly, it is the free $\mathbb{\Sigma}$-algebra over the presheaf $V$ of variables.
 
-Fortunately, constructing the initial algebra of terms in type theory boils down to defining an inductive type with a few constructors that align with the variable rule and a rule scheme for language constructs specified by a signature~\citep{Allais2021,Fiore2022}.
+Fortunately, constructing the initial algebra of terms in type theory boils down to defining an inductive type with a few constructors that align with the variable rule and a rule scheme for language constructs specified by a signature~\citep{Fiore2022}.
 
@@
 
-Our theory can be viewed as a counterpart to those theories in \Cref{sec:theory-of-syntax} and is tailored towards \emph{extrinsically-typed}, \emph{syntax-directed} bidirectional type systems of simple types (\Cref{fig:raw-terms,fig:bidirectional-typing-derivations}) specified by a \emph{signature} for simple types and a \emph{bidirectional binding signature} for terms (\Cref{def:bidirectional-binding-signature} and c.f.\ \Cref{re:type-signature}).
+Our theory can be viewed as a counterpart to those theories in \Cref{sec:theory-of-syntax} and is tailored towards \emph{extrinsically-typed}, \emph{syntax-directed} bidirectional simple type systems (\Cref{fig:raw-terms,fig:bidirectional-typing-derivations}) specified by a \emph{signature} for simple types and a \emph{bidirectional binding signature} for terms (\Cref{def:simple-signature,def:bidirectional-binding-signature}), c.f.\ \Cref{re:type-signature}.
 The extrinsic typing is needed for type synthesis and our adherence to syntax-directed systems aligns with the standard idea of re-casting non-syntax-directed typing rules into their syntax-directed form to derive type synthesis algorithm~\citep{Peyton-Jones2007}.
@@
 \begin{enumerate*}
-  \item establishing the link between the specified bidirectional type system and its simple type theory, i.e.\ \emph{soundness} and \emph{completeness} (\Cref{sec:soundness-and-completeness}); 
+  \item establishing the connection between the specified bidirectional type system and its simple type theory, i.e.\ \emph{soundness} and \emph{completeness} (\Cref{sec:soundness-and-completeness}); 
   \item defining the \emph{mode-correctness} condition (\Cref{def:mode-correctness}) which suffices to establish the uniqueness of synthesised types (\Cref{thm:unique-syn}) and a decidable bidirectional type synthesis and checking (\Cref{thm:bidirectional-type-synthesis-checking}).
@@
 Our theory is based on Martin-L\"of type theory and formalised in \Agda (\Cref{sec:formalisation}), akin to those frameworks in \Cref{sec:language-formalisation}.
-More importantly, we exploit the computational and logical aspects of type theory to formulate algorithmic soundness, completeness, and decidability. 
+More importantly, we exploit the computational and logical aspects of type theory to formulate algorithmic soundness, completeness, and decidability in one go. 
 Recall that the law of excluded middle does not hold universally as an axiom in Martin-L\"of type theory, so a \emph{decidable statement} $P \vee \neg P$ is non-trivial.
 Given that all proofs as programs terminate, logical decidability implies \emph{algorithmic decidability}.
-Further, suppose that we have a proof of the following statement:
+Further, suppose a proof of the following statement:
 \begin{quote}
@@
 \end{quote}
-echoing the algorithm given by \citeauthor{Wadler2022} as noted in \Cref{re:type-synthesis-as-decidability-proof}.
-Both algorithmic soundness and completeness are implied:
+echoing \citeauthor{Wadler2022}'s formulation as noted in \Cref{re:type-synthesis-as-decidability-proof}.
+Then, both algorithmic soundness and completeness are implied:
 the proof either yields a typing derivation for the given raw term~$t$ or a proof that such a derivation is impossible where the former case is algorithmic soundness and the latter is algorithmic completeness in contrapositive form.
@@
 In contrast to \citeauthor{Wadler2022}, our theory starts with raw terms and a general type synthesis problem, independent of bidirectional typing.
-This starting point uncovers nuances regarding the lack of enough type annotations in bidirectional typing, often overlooked in literature, and thus distinguishes completeness (with respect to a type assignment system) from annotatability, concepts conflated by \citet{Dunfield2021}.
+This starting point uncovers nuances regarding the case of not having enough annotations, often overlooked in literature, and thus distinguishes completeness (with respect to a type assignment system) from annotatability, concepts conflated by \citet{Dunfield2021}.
 To clarify the difference (\Cref{sec:annotatability}), we propose \emph{mode derivations} (\Cref{fig:mode-derivations,fig:generalised-mode-derivations}) and \emph{mode preprocessing} (\Cref{sec:mode-preprocessing}) assigning a mode first to a raw term with enough annotations or otherwise pinpoints missing annotations.
@@
 \end{enumerate}
-Moreover, all these concepts have been formally developed with \Agda.
+All concepts were formally developed with \Agda and translated to the mathematical vernacular.
 
@@
 Following this, we discuss the connection between a specified bidirectional type system and its associated simple type theory in \Cref{sec:pre-synthesis}.
-In \Cref{sec:type-synthesis}, we introduce mode-correctness and bidirectional type synthesis.
-In \Cref{sec:formalisation}, we sketch the formalisation of our theory in \Agda and some examples other than simply typed $\lambda$-calculus.
+The core technical contribution lies in \Cref{sec:type-synthesis} where we introduce mode-correctness and bidirectional type synthesis.
+In \Cref{sec:formalisation}, we sketch the formalisation of our theory in \Agda and give some examples other than simply typed $\lambda$-calculus.
 We conclude in \Cref{sec:future}, where we reflect on related techniques that can be used to improve the formal development and potential challenges in extending our theory to a more general setting.

### 2023-06-28 tex/sec2-outline.tex
[ tex ] Fine-tune the overview figure in S2

--- 2023-06-26 tex/sec2-outline.tex
@@
 To quantify over all such systems, we need their generic definitions, which we formulate in \cref{sec:defs}.
+
+\begin{figure}
+  \begin{tikzpicture}[thick]
+  \matrix (m) [matrix of math nodes, row sep=3em, column sep=10em]
+  {
+    \boxed{\erase{\Gamma} \vdash \isTerm{t}}   & \boxed{\erase{\Gamma} \vdash \isTerm{t}^{\syn}} \\
+    \boxed{\Gamma \vdash \isTerm{t} : A} & \boxed{\Gamma \vdash \isTerm{t} \syn A} \\
+  };
+  \path[->]
+      (m-1-1) edge node [above] {mode preprocessing} (m-1-2)
+      (m-1-2) edge node [right,align=center] {bidirectional\\type synthesis} (m-2-2)
+      (m-2-1) edge[bend right=8] node [below] {completeness} (m-2-2)
+      (m-2-2) edge[bend right=7] node [above] {soundness}    (m-2-1);
+  \path[dashed,->]
+      (m-1-1) edge node [left,align=center]  {type\\synthesis} (m-2-1);
+  \end{tikzpicture}
+  \caption{An implementation of type synthesis via bidirectional typing}
+\end{figure}

### 2023-06-29 tex/sec1-intro.tex
[ tex ] Fix a typo in S1

@@
   \item a family of sets $\Term_{\Gamma}$ of well-scoped terms under a context~$\Gamma$ with
-  \item variable renaming for a function $\sigma\colon \Gamma \to \Delta$ between variables acting as a functorial map from $X_{\Gamma}$ to $X_{\Delta}$, i.e.\ a presheaf $\Term\colon \mathbb{F} \to \mathsf{Set}$, and
+  \item variable renaming for a function $\sigma\colon \Gamma \to \Delta$ between variables acting as a functorial map from $\Term_{\Gamma}$ to $\Term_{\Delta}$, i.e.\ a presheaf $\Term\colon \mathbb{F} \to \mathsf{Set}$, and
   \item an initial algebra $[\mathsf{v}, \mathsf{op}]$ on~$\Term$ given by the variable rule as a map $\mathsf{v}$ from the presheaf~$V$ of variables (i.e.\ the embedding $V\colon \mathbb{F} \hookrightarrow \mathsf{Set}$) to $\Term$ and other constructs as $\mathsf{op}\colon \mathbb{\Sigma}\Term \to \Term$ where $\mathbb{F}$ is the category of contexts, the functor $\mathbb{\Sigma}\colon \mathsf{Set}^\mathbb{F} \to \mathsf{Set}^\mathbb{F}$ encodes the arities of constructs, and the initiality amounts to structural recursion, i.e.\ \emph{term traversal}.
@@
 
-Fortunately, constructing the initial algebra of terms in type theory boils down to defining an inductive type with a few constructors that align with the variable rule and a rule scheme for language constructs specified by a signature~\citep{Fiore2022}.
+Fortunately, constructing the initial algebra of terms in type theory boils down to defining an inductive type with a few constructors that align with the variable rule and a rule schema for language constructs specified by a signature~\citep{Fiore2022}.
 

### 2023-06-29 tex/sec3-defs.tex
[ tex ] Add a reference to (generalised) mode derivations
[ tex ] Simple types and simple tyep theories in S3

@@
 \section{Simple Type Theories and Bidirectional Type Systems}\label{sec:defs}
+This section provides generic definitions\footnote{%
+Here's the 'small print': Our definitions cover typical examples and set the scope for bidirectional simple type systems discussed, but do not offer comprehensive coverage of all possibilities.}
+of simple types, simple type theories, and bidirectional type systems.
+Our definitions follow a two-step approach:
+\begin{enumerate*}
+\item We first introduce a notion of signature which includes a set of constructs and an assignment of arities to each construct.
+\item Subsequently, we define abstract syntax trees (for types or terms) as specified by a given signature inductively with some primitive constructs.
+\end{enumerate*}
+Upon moving from simple types to bidirectionally typed terms, the notion of arity, initially as the number of arguments of a construct, is enriched to incorporate an extension context for each argument and a mode denoting the direction of type information flow.
 
-\subsection{Simple Types}
+\subsection{Signatures and Simple Types} \label{subsec:simple-types}
+A simple type is understood intuitively as any collection of types whose formation judgements does not depend on anything such as base types $\mathtt{bool}$, $\mathtt{nat}$, product types $A \times B$, and function types $A \bto B$.
+That is, the only datum needed for specifying a type construct is its number of arguments:
+
 \begin{definition} \label{def:simple-signature}
-  A \emph{(simple type) signature} $\Sigma$ consists of a set $I$ and an \emph{arity} function $\arity\colon I \to \mathbb{N}$.
-  An inhabitant $i : I$ is meant to represent the \emph{$i$-th operation} $\tyOp_i$ and its \emph{number} $ \arity(i)$ of arguments.
-
-  The judgements of \emph{$\Sigma$-types} and \emph{$\Sigma$-contexts} over a variable set $\Theta$ are defined inductively by \Cref{fig:simple-type,fig:simple-context} respectively.
-We write $A : \Type_{\Sigma}(\Theta)$ for $\Theta |-_{\Sigma} A$ and $\Gamma \colon \Cxt_{\Sigma}(\Theta)$ for $\Theta |-_{\Sigma} \Gamma$ to disambiguate $\Sigma$-types and $\Sigma$-contexts in the case of any confusion.
-\end{definition}
-
   \begin{figure}
@@
       \small
-    \judgbox{\Theta|-_{\Sigma} A}{$A$ is a well-formed type with type variables in $\Theta$}
+      \judgbox{\Xi|-_{\Sigma} A}{$A$ is a well-formed type with type variables in $\Xi$}
       \begin{mathpar}
-      \inferrule{\Theta \ni X_i}{\Theta |-_\Sigma X_i} \and
-      \inferrule{\Theta |-_{\Sigma} A_1 \\ \cdots \\ \Theta |-_{\Sigma} A_n}{\Theta |-_{\Sigma} \tyOp_o(A_1, \ldots, A_n)}\;\text{where $n = \arity(o)$}
+        \inferrule{\Xi \ni X_i}{\Xi |-_\Sigma X_i} \and
+        \inferrule{\Xi |-_{\Sigma} A_1 \\ \cdots \\ \Xi |-_{\Sigma} A_n}{\Xi |-_{\Sigma} \tyOp_i(A_1, \ldots, A_n)}\;\text{where $n = \arity(i)$}
       \end{mathpar}
@@
       \small
-    \judgbox{\Theta |-_{\Sigma} \Gamma}{}
+      \judgbox{\Xi |-_{\Sigma} \Gamma}{}
       \begin{mathpar}
-      \inferrule{ }{\Theta |-_{\Sigma} \cdot }\and
-      \inferrule{\Theta |-_{\Sigma} A \\ \Theta|-_{\Sigma} \Gamma}{\Theta |-_{\Sigma} \Gamma, A}
+        \inferrule{ }{\Xi |-_{\Sigma} \cdot }\and
+        \inferrule{\Xi |-_{\Sigma} A \\ \Xi|-_{\Sigma} \Gamma}{\Xi |-_{\Sigma} \Gamma, x : A}
       \end{mathpar}
@@
 
-\begin{example}\label{ex:implication}
-  Simply typed $\lambda$-calculus $\Lambda_{\bto}$ typically includes a binary implication type denoted by $\bto$ and some base types to ensure that the set of all types is non-empty.
-  The type signature $\Sigma_{\bto}$ of simply typed $\lambda$-calculus consists of a binary operation $\bto$ and a base type (nullary operation) $N$.
-  In the case of simply typed $\lambda$-calculus $\Lambda_{\bto, \btimes}$ with binary products, we can extend $\Sigma_{\bto}$ by adding a binary operation $\btimes$ to represent the binary product type, denoted as $\Sigma_{\bto, \btimes}$.
+  A \emph{signature} $\Sigma$ for simple types consists of a set\footnote{%
+    Even though our theory is developed in Martin-L\"of type theory, the term `set' is used instead of `type' to avoid the obvious confusion. 
+    As we work with Martin-L\"of type theory with \AxiomK, all types are actually sets in the sense of homotopy type theory where all inhabitants of its identity type are proved equal.
+  }
+  $I$ with a decidable equality and an \emph{arity} function $\arity\colon I \to \mathbb{N}$.
+  For a signature $\Sigma$, a \emph{type} $A : \Type_{\Sigma}(\Xi)$ over a variable set $\Xi$ is either
+  \begin{enumerate}
+    \item a variable in $\Xi$ or
+    \item $\tyOp_{i}(A_1, \ldots, A_n)$ for some $i:I$ with $\arity(i) = n$ and types $A_1,\ldots, A_n$
+  \end{enumerate}
+  as defined in \Cref{fig:simple-type}.
+  A \emph{context} $\Gamma \colon \Cxt_{\Sigma}(\Xi)$ over $\Xi$ is a list $\Gamma$ of types over $\Xi$ as defined in \Cref{fig:simple-context}. 
+\end{definition}
+Note that the second clause is a rule schema giving rise to as many rules as inhabitants $i$ in $I$.
+
+\begin{example} \label{ex:type-signature-for-function-type}
+  Simply typed $\lambda$-calculus includes function types $A \bto B$ and typically a base type~$\mathtt{b}$ to ensure that the set of all types without type variables, i.e.\ concrete types, is non-empty.
+  The type signature $\Sigma_{\bto}$ used to define types in simply typed $\lambda$-calculus consists of a binary operation $\mathsf{imp}$ and a nullary operation $\mathsf{b}$ where $\arity(\mathsf{imp}) = 2$ and $\arity(\mathsf{b}) = 0$.
+  Then, all types in simply typed $\lambda$-calculus can be given as $\Sigma_{\bto}$-types over the empty set where $A \bto B$ is introduced as $\tyOp_{\mathsf{imp}}(A, B)$ and $\mathtt{b}$ as $\tyOp_{\mathsf{b}}$.
 \end{example}
 
-\subsection{Binding Signatures} \label{subsec:binding-sig}
+For bidirectional type synthesis, we need a decidable equality on $\Type_{\Sigma}(\Xi)$.
+For example, we need to compare the synthesised type $B$ with a given type $A$ for the $\Rule{Sub}$ rule.
+The decidable equality can be given if $\Xi$ has a decidable equality.
+
+As our definition includes type variables, we introduce the typical meta-operations: substitution.
+\begin{definition}
+  The \emph{(simultaneous) substitution} of a type $A$ over $\Xi$ for a function $\rho\colon\Xi \to \Type_{\Sigma}(\Xi')$ mapping $A$ to a type $\simsub{A}{\rho}$ over $\Xi'$ is defined as usual.
+  By abuse of notation, the substitution $\simsub{\Gamma}{\rho}$ of a context $\Gamma$ is defined by applying substitution $\simsub{A}{\rho}$ to each type $A$ in the context.
+\end{definition}
+
+\subsection{Binding Signatures and Simple Type Theories} \label{subsec:binding-sig}
+
+For a type signature $\Sigma$, a simple type theory consists of
+\begin{enumerate*}
+  \item a set of raw terms $\isTerm{t}$ indexed by a set $V$ of untyped variables where each construct is allowed to bind some variables (as $\Rule{Abs}$) and to take multiple arguments (as $\Rule{App}$) like \Cref{fig:STLC-raw-terms};
+  \item a set of typing derivations indexed by a set $\Gamma$ of typed variables, a type~$A$, and a raw term $\isTerm{t}$ to set their type constraints like \Cref{fig:STLC-typing-derivations}. 
+\end{enumerate*}
+
+As variables are not invariant, an extension context for variable binding is only a list of types.
+Also, not only the number of arguments for a term construct but also their types such as $\Rule{App}$ where the type $A$ of second argument must be the domain of the function type $A \bto B$.
+As a result, the notion of signature is now enriched with extension contexts as well as types as follows.
+
 \begin{definition}\label{def:binding-signature}
@@
   \]
+\end{definition}
 Each inhabitant $o: O$ is meant to represent a term construct $\tmOp_o$ in a simple type theory with a triple $\left(\Xi, \left[\left(\Delta_1; A_1\right), \ldots, \left(\Delta_{n}; A_{n}\right) \right], A\right)$
@@
 \begin{enumerate}
-    \item $\Xi$ is the number of type variables, 
-    \item $A : \Type_\Sigma(\Xi)$ is the target type of $\tmOp_o$, and
+  \item $\Xi$ is the number\footnote{%
+    Since the arity of a type construct is always finite, using a finite set for the set of type variables is sufficient.
+    Moreover we can always define inductively a finite set of $n$ many inhabitants canonically with a decidable equality, eliminating the need to require a decidable equality explicitly, so we will always formally use a natural number for a set of type variables.
+    For the convenience of presentation, variables are still named, though.}
+  of type variables, 
+  \item $A : \Type_\Sigma(\Xi)$ is for the type of $\tmOp_o$, and
   \item $\left[\left(\Delta_1; A_{1}\right), \ldots, \left(\Delta_{n}; A_{n}\right) \right]$ is a list of pairs where
-    \item the $i$-th pair $(\Delta_i, A_i) :\Cxt_{\Sigma}(\Xi) \times \Term_{\Sigma}(\Xi)$ are types of the binding variables and the type of the $i$-th argument of $\tmOp_{o}$.
+  \item the $i$-th pair $(\Delta_i, A_i) :\Cxt_{\Sigma}(\Xi) \times \Term_{\Sigma}(\Xi)$ are for types of the binding variables and the type of the $i$-th argument of $\tmOp_{o}$.
 \end{enumerate}
-  For brevity, we write $o \colon \Xi \rhd [\Delta_1]A_{1}, \ldots, \left[\Delta_{n}\right] A_{n} \to A$ to indicate an operation with its arity. 
-\end{definition}
-\begin{remark}[Terminology on binding signature]
-  \cite{Aczel1978,Fiore2010}
-\end{remark}
-
-\begin{example} \label{ex:STLC-sig}
-  Given the type signature $\Sigma_{\bto}$ for the implication type, the term signature $\Lambda_{\bto}$ for simply typed $\lambda$-calculus can be described by operations
-  \begin{align*}
-    \mathsf{app}\colon A, B \rhd (A \bto B), A \to B && \mathsf{abs}\colon A , B \rhd [A]B \to (A \bto B)
-  \end{align*}
-  or, verbosely, the signature $\Lambda_{\bto}$ consists of the type $O_\Lambda = \{\abs, \app\}$ with arities
-  \begin{align*}
-    \arity(\app) = (\{A, B\}, [(\cdot; A \bto B), (\cdot; A)], B)
-    && 
-    \arity(\abs) = (\{A, B\}, [(\cdot, A; B)], A \bto B)
-  \end{align*}
-  For the type signature $\Sigma_{\bto, \btimes}$, the term signature $\Omega_{\bto, \btimes}$ for simply typed $\lambda$-calculus with finite products has three additional operations
-to $\Omega_{\bto}$:
-  \begin{align*}
-    \mathsf{pair}\colon A, B \rhd A, B \to A \btimes B
-    && \fst \colon A, B \rhd A \btimes B \to A
-    && \snd \colon A, B \rhd A \btimes B \to B
-  \end{align*}
-\end{example}
-
-First, it is noteworthy that the set of term constructs in a type theory need not be finite.
-For instance, a type theory may adopt a spine application which takes indefinitely many arguments---for each number $n$ of arguments an $(n+1)$-ary application construct can be introduced, so even if each operation has a definite number of arguments by definition a spine application is still expressible.
-
-More importantly, the inclusion of the set of variables used in an operation is a salient feature of our binding signatures.
-Instead of treating application as a family of constructs $\app_{A, B}$ indexed by all types $A$ and $B$, as done by \citet{Fiore2022}, we are able to identify them as a single construct $\app$.
-This is not only for brevity but also necessary to compare the type equality during type synthesis and checking.
-
-%\subsection{Intrinsically Typed Terms}
-%
-%Intrinsically typed terms for a type signature $\Sigma$ and a term signature $\Omega$ are nothing more than derivations of the intrinsic typing judgement $\Gamma |-_{\Sigma, \Omega} A$ constructed with only the rule $(\var)$ and the rule scheme $(\tmOp)$ displayed in \Cref{fig:intrinsic-typing}.
-%
-%The substitution $\rho\colon \Sub{\Xi}{\emptyset}$ is used to instantiate variables in the local context $\Xi$ with concrete types.
-%Accordingly, $\rho$ has to be applied to types that appear in the arity to construct a term by $\tmOp$ to ensure all types are well-formed without any use of type variables.
-%
-%\begin{figure}
-%  \centering
-%  \small
-%  \begin{mathpar}
-%    \boxed{\Gamma |-_{\Sigma, \Omega} A} \\
-%    \inferrule{A \in \Gamma}{\Gamma |-_{\Sigma, \Omega} A}\;(\var)
-%    \and
-%    \inferrule{\rho\colon \Sub{\Xi}{\emptyset}  \\ \Gamma, \simsub{\Delta_{1}}{\rho} |-_{\Sigma, \Omega} \simsub{A_{1}}{\rho} \quad\cdots\quad \Gamma, \simsub{\Delta_{n}}{\rho} |-_{\Sigma, \Omega} \simsub{A_{n}}{\rho}}
-%    {\Gamma |-_{\Sigma, \Omega} \simsub{A}{\rho}}\;(\tmOp)
-%    \and \text{for $o\colon \Xi \rhd [\Delta_1]A_1, \ldots, [\Delta_{n}]A_{n} \to A$ in $\Omega$}
-%  \end{mathpar}
-%  \caption{Intrinsic typing rules for a simple type theory $(\Sigma, \Omega)$}
-%  \label{fig:intrinsic-typing}
-%\end{figure}
-
-\subsection{Simple Type Theories}
+For brevity, we write $o \colon \Xi \rhd [\Delta_1]A_{1}, \ldots, \left[\Delta_{n}\right] A_{n} \to A_0$ to indicate an operation with its arity and omit $[\Delta]$ if $\Delta$ is empty. 
+For example, the $\Rule{Abs}$ rule can be specified by the operation 
+\[
+  \mathsf{abs}\colon A , B \rhd [A]B \to (A \bto B)
+\]
+indicating that the rule $\Rule{Abs}$ has two type variables $A$ and $B$, only one argument of type $B$ and which binds a variable of type $A$, and the type $A \bto B$ for an abstraction.
+Similarly, the $\Rule{App}$ rule can be specified by $\mathsf{app}\colon A, B \rhd A \bto B, A \to B$ where extension contexts for its two arguments are empty.
 
@@
   \small
-  \judgbox{|-_{\Sigma, \Omega} \isTerm{t}}{$\isTerm{t}$ is a raw term for the signature $(\Sigma, \Omega)$}
+  \judgbox{V |-_{\Sigma, \Omega} \isTerm{t}}{Given a list V of variables, $\isTerm{t}$ is a raw term for the signature $(\Sigma, \Omega)$ with free variables in $V$}
   \begin{mathpar}
-    \inferrule{\isTerm{x} : \Identifier}{|-_{\Sigma, \Omega} \isTerm{x}}\,\Rule{Var}
+    \inferrule{\isTerm{x} \in V}{V |-_{\Sigma, \Omega} \isTerm{x}}\,\Rule{Var}
     \and
-    \inferrule{\cdot |-_{\Sigma} A \\ |-_{\Sigma, \Omega}\isTerm{t}}{|-_{\Sigma, \Omega} \isTerm{t \annotate A}}\,\Rule{Anno}
+    \inferrule{\cdot |-_{\Sigma} A \\ V |-_{\Sigma, \Omega}\isTerm{t}}{V |-_{\Sigma, \Omega} \isTerm{t \annotate A}}\,\Rule{Anno}
     \and
-    \inferrule{|-_{\Sigma, \Omega} \isTerm{t_1} \quad \cdots \quad |-_{\Sigma, \Omega} \isTerm{t_n}}
-    {|-_{\Sigma, \Omega} \tmOpts }\,\Rule{Op}
+    \inferrule{V, \vars{x}_\isTerm{1} |-_{\Sigma, \Omega} \isTerm{t_1} \\ \cdots \\ V, \vars{x}_\isTerm{n} |-_{\Sigma, \Omega} \isTerm{t_n}}
+    {V |-_{\Sigma, \Omega} \tmOpts }\,\Rule{Op}
   \end{mathpar}
-  \begin{flushright}
-    for $o \colon \Xi \rhd [\Delta_1]A_{1}, \ldots, [\Delta_{n}] A_{n} \to A$ in $\Omega$
-  \end{flushright}%
+%  \begin{flushright}
+%    for $o \colon \Xi \rhd [\Delta_1]A_{1}, \ldots, [\Delta_{n}] A_{n} \to A$ in $\Omega$
+%  \end{flushright}%
   \caption{Raw terms}
@@
 
+In the following we define raw terms for a type signature $\Sigma$ and a binding signature $\Omega$.
+\begin{definition}
+  In a simple type theory specified by $(\Sigma, \Omega)$, the set of \emph{raw terms} under a context $V$ of variables consists of
+  \begin{enumerate}
+    \item variables in $V$,
+    \item annotations $\isTerm{t \annotate A}$ for some raw term $t$ and a type $A$, and
+    \item a construct $\tmOpts$ for some $o \colon \Xi \rhd [\Delta_1]A_{1}, \ldots, [\Delta_{n}] A_{n} \to A$ in $O$, where $t_i$'s are raw terms and $\vars{x}_{\isTerm{i}}$'s are lists of variables whose length is equal to the length of~$\Delta_i$,
+  \end{enumerate}
+  corresponding to rules $\Rule{Var}$, $\Rule{Anno}$, and $\Rule{Op}$ in \Cref{fig:raw-terms} respectively.
+\end{definition}
+
+The definition of typing derivations is slightly involved.
+We need information to compare types on the object level during type synthesis and have to substitute those type variables in a typing derivation $\Gamma \vdash \tmOpts : A$ specified by some operation $o$ in $\Omega$ somewhere.
+Hence we include a substitution as part of its typing derivation explicitly.
+
+\begin{definition}
   \begin{figure}
@@
     \small
-  \judgbox{\Gamma |-_{\Sigma, \Omega} \isTerm{t} : A}{A raw term $\isTerm{t}$ has a type $A$ (without type variables) under $\Gamma$ for the signature $(\Sigma, \Omega)$}
+    \judgbox{\Gamma |-_{\Sigma, \Omega} \isTerm{t} : A}{A raw term $\isTerm{t}$ has a type $A$ without type variables under $\Gamma$ for the signature $(\Sigma, \Omega)$}
     \begin{mathpar}
@@
       \and
-    \inferrule{\rho : \Sub{\Xi}{\emptyset} \\ \Gamma, \vec{\isTerm{x}}_\isTerm{1} : \simsub{\Delta_{1}}{\rho} |-_{\Sigma, \Omega} \isTerm{t_1} : \simsub{A_{1}}{\rho} \quad\cdots\quad \Gamma, \vec{\isTerm{x}}_\isTerm{n} : \simsub{\Delta_{n}}{\rho} |-_{\Sigma, \Omega} \isTerm{t_n} : \simsub{A_{n}}{\rho}}
-    {\Gamma |-_{\Sigma, \Omega} \tmOpts : \simsub{A}{\rho}}\,\Rule{Op}
-    \and \text{for $o\colon \Xi \rhd [\Delta_1]A_1, \ldots, [\Delta_{n}]A_{n} \to A$ in $\Omega$}
+      \inferrule{\rho : \Sub{\Xi}{\emptyset} \\  A = \simsub{A_0}{\rho} \\ \Gamma, \vec{\isTerm{x}}_\isTerm{1} : \simsub{\Delta_{1}}{\rho} |-_{\Sigma, \Omega} \isTerm{t_1} : \simsub{A_{1}}{\rho} \quad\cdots\quad \Gamma, \vec{\isTerm{x}}_\isTerm{n} : \simsub{\Delta_{n}}{\rho} |-_{\Sigma, \Omega} \isTerm{t_n} : \simsub{A_{n}}{\rho}}
+      {\Gamma |-_{\Sigma, \Omega} \tmOpts : A}\,\Rule{Op}
     \end{mathpar}
@@
   \end{figure}
+  In a simple type theory specified by $(\Sigma, \Omega)$, the set of \emph{typing derivations} of $\Gamma \vdash \isTerm{t} : A$ for a context $\Gamma$ of typed variables, a type $A$, and a raw term under $\erase{\Gamma}$ consists of
+  \begin{enumerate}
+    \item a derivation of $\Gamma |- x : A$ if $x : A$ is in $\Gamma$,
+    \item a derivation of $\Gamma |- (t \annotate A) : A$ if $\Gamma \vdash \isTerm{t} : A$ is derivable, and
+    \item a derivation of $\Gamma |- \tmOpts : A$ for $o \colon \Xi \rhd [\Delta_1]A_1, \ldots, [\Delta_{n}]A_{n} \to A_0$ if there is a function $\rho\colon \Xi \to \Type_{\Sigma}(\emptyset)$ such that $A = \simsub{A_0}{\rho}$ and $\Gamma, \vars{x}_{\isTerm{i}} : \Delta_i \vdash \isTerm{t_i} : \simsub{A_i}{\rho}$ is derivable for all $i = 1, \ldots, n$,
+  \end{enumerate}
+  corresponding to rules $\Rule{Var}$, $\Rule{Anno}$, and $\Rule{Op}$ in \Cref{fig:extrinsic-typing} respectively.
+\end{definition}
 
-\subsection{Bidirectional Binding Signatures and Bidirectional Type Systems} \label{sec:bidirectional-system}
+\subsection{Bidirectional Binding Signatures and Bidirectional Type Systems} \label{subsec:bidirectional-system}
 
@@
   \end{align*}
-  extending $\Lambda_{\bto}$ (\Cref{ex:STLC-sig}) with the mode information.
 \end{example}
@@
 \LT{Some remark about signature erasure and annotation; introduce the notation $\erase{\Omega}$}
+
+\LT{First, it is noteworthy that the set of term constructs in a type theory need not be finite.
+For instance, a type theory may adopt a spine application which takes indefinitely many arguments---for each number $n$ of arguments an $(n+1)$-ary application construct can be introduced, so even if each operation has a definite number of arguments by definition a spine application is still expressible.}
+
+\LT{More importantly, the inclusion of the set of variables used in an operation is a salient feature of our binding signatures.
+Instead of treating application as a family of constructs $\app_{A, B}$ indexed by all types $A$ and $B$, as done by \citet{Fiore2022}, we are able to identify them as a single construct $\app$.
+This is not only for brevity but also necessary to compare the type equality during type synthesis and checking.
+}
 
@@
   \caption{Mode derivations}
+  \label{fig:mode-derivations}
 \end{figure}

### 2023-06-29 tex/sec2-outline.tex
[ tex ] Fine-tune the figure of overview

--- 2023-06-27 tex/sec2-outline.tex
@@
     \and
-    \inferrule{\Gamma \vdash \isTerm{t} \syn A \\ A = B}{\Gamma \vdash \isTerm{t} \chk B}\,\ChkRule{Sub}
+    \inferrule{\Gamma \vdash \isTerm{t} \syn B \\ B = A}{\Gamma \vdash \isTerm{t} \chk A}\,\ChkRule{Sub}
     \and
@@
   };
-  \path[->]
-      (m-1-1) edge node [above,align=center] {mode preprocessing} (m-1-2)
-              edge node [left,align=left]  {type\\synthesis} (m-2-1)
-      (m-1-2) edge node [right,align=right] {bidirectional\\type synthesis} (m-2-2)
-      (m-2-1) edge[bend right=8] node [below] {completeness} (m-2-2);
-  \path[->]
+  \path[->, line width=0.7pt]
+      (m-1-1) edge node [above] {mode preprocessing} (m-1-2)
+      (m-1-2) edge node [right,align=center] {bidirectional\\type synthesis} (m-2-2)
+      (m-2-1) edge[bend right=8] node [below] {completeness} (m-2-2)
       (m-2-2) edge[bend right=8] node [above] {soundness}    (m-2-1);
+  \path[dashed,->,line width=0.7pt]
+      (m-1-1) edge node [left,align=center]  {type\\synthesis} (m-2-1);
   \end{tikzpicture}

### 2023-06-29 tex/sec2-outline.tex
[ tex ] Done with S3 for now

@@
   \small
-  \judgbox{|- \isTerm{t}^\dir{d}}{A raw term~$\isTerm{t}$ is in mode~$d$}
+  \judgbox{V |- \isTerm{t}^\dir{d}}{A raw term~$\isTerm{t}$ is in mode~$d$ with free variables in $V$}
   \begin{mathpar}
@@
 \begin{figure}
+  \small
   \begin{tikzpicture}
@@
   };
-  \path[->, line width=0.7pt]
+  \path[->, line width=0.5pt]
       (m-1-1) edge node [above] {mode preprocessing} (m-1-2)
@@
       (m-2-2) edge[bend right=8] node [above] {soundness}    (m-2-1);
-  \path[dashed,->,line width=0.7pt]
+  \path[dashed,->,line width=0.5pt]
       (m-1-1) edge node [left,align=center]  {type\\synthesis} (m-2-1);

### 2023-06-29 tex/sec3-defs.tex
[ tex ] Done with S3 for now

@@
 \section{Simple Type Theories and Bidirectional Type Systems}\label{sec:defs}
-This section provides generic definitions\footnote{%
-Here's the 'small print': Our definitions cover typical examples and set the scope for bidirectional simple type systems discussed, but do not offer comprehensive coverage of all possibilities.}
-of simple types, simple type theories, and bidirectional type systems.
-Our definitions follow a two-step approach:
+This section provides generic definitions
+%\footnote{%
+%Here's the 'small print': Our definitions cover typical examples and set the scope for bidirectional simple type systems discussed, but do not offer comprehensive coverage of all possibilities.}
+of simple types, simple type theories, and bidirectional type systems using simply typed $\lambda$-calculus as our running example.
+
+Our definitions are defined in two steps:
 \begin{enumerate*}
-\item We first introduce a notion of signature which includes a set of constructs and an assignment of arities to each construct.
-\item Subsequently, we define abstract syntax trees (for types or terms) as specified by a given signature inductively with some primitive constructs.
+\item we first introduce a notion of signature which includes a set of operation symbols and an assignment of arities to each operation;
+\item then, we define abstract syntax trees inductively with primitive constructs and constructs specified by a given signature.
 \end{enumerate*}
-Upon moving from simple types to bidirectionally typed terms, the notion of arity, initially as the number of arguments of a construct, is enriched to incorporate an extension context for each argument and a mode denoting the direction of type information flow.
+Upon moving from simple types to bidirectionally typed terms, the notion of arity, initially as the number of arguments of an operation, is enriched to incorporate an extension context for each argument and a mode denoting the direction of type information flow.
 
 \subsection{Signatures and Simple Types} \label{subsec:simple-types}
-A simple type is understood intuitively as any collection of types whose formation judgements does not depend on anything such as base types $\mathtt{bool}$, $\mathtt{nat}$, product types $A \times B$, and function types $A \bto B$.
-That is, the only datum needed for specifying a type construct is its number of arguments:
+Simple types of a language is understood intuitively as a collection of types whose formation judgement is not indexed by anything and the only datum needed for specifying a type construct is its number of arguments:
 
@@
 \begin{example} \label{ex:type-signature-for-function-type}
-  Simply typed $\lambda$-calculus includes function types $A \bto B$ and typically a base type~$\mathtt{b}$ to ensure that the set of all types without type variables, i.e.\ concrete types, is non-empty.
+  Simply typed $\lambda$-calculus includes function types $A \bto B$ and typically a base type~$\mathtt{b}$ to ensure that the set of all types without type variables is non-empty.
   The type signature $\Sigma_{\bto}$ used to define types in simply typed $\lambda$-calculus consists of a binary operation $\mathsf{imp}$ and a nullary operation $\mathsf{b}$ where $\arity(\mathsf{imp}) = 2$ and $\arity(\mathsf{b}) = 0$.
@@
 
-For bidirectional type synthesis, we need a decidable equality on $\Type_{\Sigma}(\Xi)$.
-For example, we need to compare the synthesised type $B$ with a given type $A$ for the $\Rule{Sub}$ rule.
-The decidable equality can be given if $\Xi$ has a decidable equality.
+Without the loss of generality, we may use the set $\Fin(n)$ of natural numbers less than $n$ as the set of type variables in this paper, following the inductive definition by \citet{Dybjer1994}.
+In bidirectional type synthesis, it is necessary to have a decidable equality on $\Type_{\Sigma}(\Xi)$, as we need to compare the synthesized type $B$ with a given type $A$ for the $\Rule{Sub}$ rule.
+When $\Xi$ has a decidable equality, the desired decidable equality can be given.
+Since the arity of a type operation is always finite, it is sufficient to use a finite set like $\Fin(n)$.
+Additionally, $\Fin(n)$ has a decidable equality, ensuring that the set of simple types over $\Fin(n)$ also has a decidable quality.
+Therefore, for convenience, we can represent the set of type variables using a natural number.
+However, for illustrative purposes, variables will still be assigned names.
 
-As our definition includes type variables, we introduce the typical meta-operations: substitution.
+As \Cref{def:simple-signature} includes type variables, we introduce the typical meta-operations: substitution.
 \begin{definition}
@@
   \item a set of raw terms $\isTerm{t}$ indexed by a set $V$ of untyped variables where each construct is allowed to bind some variables (as $\Rule{Abs}$) and to take multiple arguments (as $\Rule{App}$) like \Cref{fig:STLC-raw-terms};
-  \item a set of typing derivations indexed by a set $\Gamma$ of typed variables, a type~$A$, and a raw term $\isTerm{t}$ to set their type constraints like \Cref{fig:STLC-typing-derivations}. 
+  \item a set of typing derivations indexed by a list $\Gamma$ of typed variables, a type~$A$, and a raw term $\isTerm{t}$ to set their type constraints like \Cref{fig:STLC-typing-derivations}. 
 \end{enumerate*}
-
-As variables are not invariant, an extension context for variable binding is only a list of types.
-Also, not only the number of arguments for a term construct but also their types such as $\Rule{App}$ where the type $A$ of second argument must be the domain of the function type $A \bto B$.
-As a result, the notion of signature is now enriched with extension contexts as well as types as follows.
-
+Therefore, to specify a construct like $\Rule{Abs}$ or $\Rule{App}$, we need to enrich the notion of signatures with an extension context for variable binding for each argument, a type for each argument, and a type for the operation:
 \begin{definition}\label{def:binding-signature}
@@
 \end{definition}
-Each inhabitant $o: O$ is meant to represent a term construct $\tmOp_o$ in a simple type theory with a triple $\left(\Xi, \left[\left(\Delta_1; A_1\right), \ldots, \left(\Delta_{n}; A_{n}\right) \right], A\right)$
-instead of a number as its arity $\arity_o$ where
+Each inhabitant $o: O$, meant to represent a term construct $\tmOp_o$ in a simple type theory, has a triple $\left(\Xi, \left[\left(\Delta_1; A_1\right), \ldots, \left(\Delta_{n}; A_{n}\right) \right], A_0\right)$ as its arity $\arity_o$ where
 \begin{enumerate}
-  \item $\Xi$ is the number\footnote{%
-    Since the arity of a type construct is always finite, using a finite set for the set of type variables is sufficient.
-    Moreover we can always define inductively a finite set of $n$ many inhabitants canonically with a decidable equality, eliminating the need to require a decidable equality explicitly, so we will always formally use a natural number for a set of type variables.
-    For the convenience of presentation, variables are still named, though.}
-  of type variables, 
+  \item $\Xi$ is the number of type variables, 
   \item $A : \Type_\Sigma(\Xi)$ is for the type of $\tmOp_o$, and
@@
 \]
-indicating that the rule $\Rule{Abs}$ has two type variables $A$ and $B$, only one argument of type $B$ and which binds a variable of type $A$, and the type $A \bto B$ for an abstraction.
+indicating that the $\Rule{Abs}$ rule has two type variables $A$ and $B$, one argument of type $B$ and which binds a variable of type $A$, and the type $A \bto B$ for the construct $\lambda$.
 Similarly, the $\Rule{App}$ rule can be specified by $\mathsf{app}\colon A, B \rhd A \bto B, A \to B$ where extension contexts for its two arguments are empty.
 
+We define raw terms specified by a type signature $\Sigma$ and a binding signature $\Omega$ as follows.
+\begin{definition}
 \begin{figure}
@@
 \end{figure}
-
-In the following we define raw terms for a type signature $\Sigma$ and a binding signature $\Omega$.
-\begin{definition}
-  In a simple type theory specified by $(\Sigma, \Omega)$, the set of \emph{raw terms} under a context $V$ of variables consists of
+  In a simple type theory specified by $(\Sigma, \Omega)$, the set of \emph{raw terms} under a context~$V$ of variables consists of
   \begin{enumerate}
@@
     \item annotations $\isTerm{t \annotate A}$ for some raw term $t$ and a type $A$, and
-    \item a construct $\tmOpts$ for some $o \colon \Xi \rhd [\Delta_1]A_{1}, \ldots, [\Delta_{n}] A_{n} \to A$ in $O$, where $t_i$'s are raw terms and $\vars{x}_{\isTerm{i}}$'s are lists of variables whose length is equal to the length of~$\Delta_i$,
+    \item a construct $\tmOpts$ for some $o \colon \Xi \rhd [\Delta_1]A_{1}, \ldots, [\Delta_{n}] A_{n} \to A_0$ in $O$, where $t_i$'s are raw terms and $\vars{x}_{\isTerm{i}}$'s are lists of variables whose length is equal to the length of~$\Delta_i$,
   \end{enumerate}
@@
 The definition of typing derivations is slightly involved.
-We need information to compare types on the object level during type synthesis and have to substitute those type variables in a typing derivation $\Gamma \vdash \tmOpts : A$ specified by some operation $o$ in $\Omega$ somewhere.
-Hence we include a substitution as part of its typing derivation explicitly.
-
-\begin{definition}
+We need some information to compare types on the object level during type synthesis and substitute those type variables in a typing derivation $\Gamma \vdash \tmOpts : A$ for an operation $o$ in $\Omega$ at some point.
+Hence we choose to include a substitution as part of its typing derivation explicitly:
+\begin{definition}\label{def:typing-derivations}
   \begin{figure}
@@
       \and
-      \inferrule{\rho : \Sub{\Xi}{\emptyset} \\  A = \simsub{A_0}{\rho} \\ \Gamma, \vec{\isTerm{x}}_\isTerm{1} : \simsub{\Delta_{1}}{\rho} |-_{\Sigma, \Omega} \isTerm{t_1} : \simsub{A_{1}}{\rho} \quad\cdots\quad \Gamma, \vec{\isTerm{x}}_\isTerm{n} : \simsub{\Delta_{n}}{\rho} |-_{\Sigma, \Omega} \isTerm{t_n} : \simsub{A_{n}}{\rho}}
-      {\Gamma |-_{\Sigma, \Omega} \tmOpts : A}\,\Rule{Op}
+      \inferrule{\rho : \Sub{\Xi}{\emptyset} \\  \Gamma, \vec{\isTerm{x}}_\isTerm{1} : \simsub{\Delta_{1}}{\rho} |-_{\Sigma, \Omega} \isTerm{t_1} : \simsub{A_{1}}{\rho} \quad\cdots\quad \Gamma, \vec{\isTerm{x}}_\isTerm{n} : \simsub{\Delta_{n}}{\rho} |-_{\Sigma, \Omega} \isTerm{t_n} : \simsub{A_{n}}{\rho}}
+      {\Gamma |-_{\Sigma, \Omega} \tmOpts : \simsub{A_0}{\rho}}\,\Rule{Op}
     \end{mathpar}
@@
   \end{figure}
-  In a simple type theory specified by $(\Sigma, \Omega)$, the set of \emph{typing derivations} of $\Gamma \vdash \isTerm{t} : A$ for a context $\Gamma$ of typed variables, a type $A$, and a raw term under $\erase{\Gamma}$ consists of
+  In a simple type theory specified by $(\Sigma, \Omega)$, the set of \emph{typing derivations} of $\Gamma \vdash \isTerm{t} : A$ for a context $\Gamma$ of typed variables, a raw term under $\erase{\Gamma}$, and a type $A$ consists of
   \begin{enumerate}
@@
     \item a derivation of $\Gamma |- (t \annotate A) : A$ if $\Gamma \vdash \isTerm{t} : A$ is derivable, and
-    \item a derivation of $\Gamma |- \tmOpts : A$ for $o \colon \Xi \rhd [\Delta_1]A_1, \ldots, [\Delta_{n}]A_{n} \to A_0$ if there is a function $\rho\colon \Xi \to \Type_{\Sigma}(\emptyset)$ such that $A = \simsub{A_0}{\rho}$ and $\Gamma, \vars{x}_{\isTerm{i}} : \Delta_i \vdash \isTerm{t_i} : \simsub{A_i}{\rho}$ is derivable for all $i = 1, \ldots, n$,
+    \item a derivation of $\Gamma |- \tmOpts : A$ for $o \colon \Xi \rhd [\Delta_1]A_1, \ldots, [\Delta_{n}]A_{n} \to A_0$ if there is a function $\rho\colon \Xi \to \Type_{\Sigma}(\emptyset)$ such that $A = \simsub{A_0}{\rho}$ and $\Gamma, \vars{x}_{\isTerm{i}} : \Delta_i \vdash \isTerm{t_i} : \simsub{A_i}{\rho}$ is derivable for all $i$,
   \end{enumerate}
   corresponding to rules $\Rule{Var}$, $\Rule{Anno}$, and $\Rule{Op}$ in \Cref{fig:extrinsic-typing} respectively.
+  For better readability, the subscript $(\Sigma, \Omega)$ in the above typing judgements is suppressed.
 \end{definition}
+Note that, by definition, our simple type theory is always syntax-directed, since each rule in \Cref{fig:raw-terms} corresponds exactly to one rule in \Cref{fig:extrinsic-typing}. 
+The $\Rule{Op}$ in both figures is a rule schema and gives rise to as many rules as there are operations in $o : O$.
+
+\begin{example}
+  Raw terms and typing derivations for simply typed $\lambda$-calculus (\Cref{fig:STLC-raw-terms,fig:STLC-typing-derivations}) can be specified by the type signature $\Sigma_{\bto}$ (\Cref{ex:type-signature-for-function-type}) and the binding signature consisting of
+ \[
+   \mathsf{app}\colon A, B \rhd A \bto B, A \to B
+   \quad\text{and}\quad
+   \mathsf{abs}\colon A , B \rhd [A]B \to (A \bto B).
+ \]
+ Rules $\Rule{Abs}$ and $\Rule{App}$ in simply typed $\lambda$-calculus are subsumed by the $\Rule{Op}$ rule schema, as terms $\tmOp_{\mathsf{app}}(t, u)$ and $\tmOp_{\mathsf{abs}}(x.t)$ are exactly an application $t\;u$ and an abstraction $\lam{x}t$, respectively.
+\end{example}
 
 \subsection{Bidirectional Binding Signatures and Bidirectional Type Systems} \label{subsec:bidirectional-system}
+Finally, for a bidirectional type system, we have raw terms and typing derivations like a simple type theory, but its typing judgements come in two forms $\Gamma |- t \syn A$ and $\Gamma |- t \chk A$.
+Observe that these two typing judgements are just one typing judgement $\Gamma |- t :^\dir{d} A$ indexed additionally by a \emph{mode} $d : \Mode$ where $\Mode$ consists of two inhabitants $\syn$ for synthesis and $\chk$ for checking.
+Hence, to specify a bidirectional type system, we need further enrich the notion of binding signatures with a mode for each argument and a mode for an operation itself as follows.
 
@@
   \]
-  where $\Mode$ consists of two inhabitants $\chk$ for checking and $\syn$ for synthesis.
-  Bidirectional binding signatures are just binding signatures (\Cref{def:binding-signature}) augmented with a mode for each argument and its target of a construct $\tmOp_o$ in a bidirectional system, i.e.
-  an arity is a $4$-tuple
-  \[
-    \left(\Xi, \left[\left(\Delta_1; A_1; d_1\right), \ldots, \left(\Delta_{n}; A_{n}; d_n\right) \right], A, d\right)
-  \]
-  where $d$ and $d_i$'s indicate the modes of a construct and its arguments respectively.
+\end{definition}
+Each inhabitant $o:O$ is similarly meant to represent a term construct $\tmOp_o$ in a bidirectional type system but the arity of $o : O$ becomes a $4$-tuple $\left(\Xi, \left[\left(\Delta_1; A_1; d_1\right), \ldots, \left(\Delta_{n}; A_{n}; d_n\right) \right], A_0, d\right)$
+including additionally $d$ and $d_i$'s for the modes of the construct $\tmOp_o$ and its arguments respectively.
 
-  For brevity, we write $o \colon \Xi \rhd [\Delta_1]A_{1}^{\dir{d_1}}, \ldots, [\Delta_{n}] A^{\dir{d_n}}_{n} \to A^{\dir{d}}$ to indicate an operation with its arity. 
-\end{definition}
+For brevity, we write $o \colon \Xi \rhd [\Delta_1]A_{1}^{\dir{d_1}}, \ldots, [\Delta_{n}] A^{\dir{d_n}}_{n} \to A_0^{\dir{d}}$ to indicate an operation with its arity. 
+For example, the $\ChkRule{Abs}$ rule in \Cref{fig:STLC-bidirectional-typing-derivations} can be specified by $\mathsf{abs}\colon A , B \rhd [A]B^{\chk} \to (A \bto B)^{\chk}$ indicating additionally that the construct $\lambda$ and its argument are in checking mode.
+Similarly, the $\SynRule{App}$ rule can be specified by $\mathsf{app}\colon A, B \rhd (A \bto B)^{\syn}, A^{\chk} \to B^{\syn}$.
 
-\begin{example}
-  The term signature $\Lambda_{\bto}^{\Leftrightarrow}$ for bidirectional simply typed $\lambda$-calculus introduced in \Cref{subsec:binding-sig} can be specified by operations 
-  \begin{align*}
-    \mathsf{app}\colon A, B \rhd (A \bto B)^{\syn}, A^{\chk} \to B^{\syn} &&
-    \mathsf{abs}\colon A , B \rhd [A]B^{\chk} \to (A \bto B)^{\chk}
-  \end{align*}
-\end{example}
+\begin{remark}
+  Every bidirectional binding signature $\Omega$ gives rise to a binding signature $\erase{\Omega}$ if we erase modes from $\Omega$, so a bidirectional binding signature $\Omega$ also specifies a simple type theory, including its raw terms and typing derivations.
+  For this notion, we call $\erase{\Omega}$ the \emph{(mode) erasure} of $\Omega$.
+\end{remark}
 
-\LT{Some remark about signature erasure and annotation; introduce the notation $\erase{\Omega}$}
-
-\LT{First, it is noteworthy that the set of term constructs in a type theory need not be finite.
-For instance, a type theory may adopt a spine application which takes indefinitely many arguments---for each number $n$ of arguments an $(n+1)$-ary application construct can be introduced, so even if each operation has a definite number of arguments by definition a spine application is still expressible.}
-
-\LT{More importantly, the inclusion of the set of variables used in an operation is a salient feature of our binding signatures.
-Instead of treating application as a family of constructs $\app_{A, B}$ indexed by all types $A$ and $B$, as done by \citet{Fiore2022}, we are able to identify them as a single construct $\app$.
-This is not only for brevity but also necessary to compare the type equality during type synthesis and checking.
-}
-
+We generalise bidirectional typing rules as well as model derivations for simply typed $\lambda$-calculus in \Cref{fig:STLC-bidirectional-typing-derivations,fig:STLC-mode-derivations} to derivations for a bidirectional type system specified by $(\Sigma, \Omega)$ where $\Sigma$ is still a type signature and $\Omega$ is a bidirectional binding signature, similar to \Cref{def:typing-derivations} for typing derivations but with an additional $\ChkRule{Sub}$ rule:
+\begin{definition}\label{def:bidirectional-typing-derivations}\label{def:mode-derivations}
   \begin{figure}
@@
     \small
-  \judgbox{|-_{\Sigma, \Omega} \isTerm{t}^\dir{d}}{$\isTerm{t}$ is a raw term for $(\Sigma, \erase{\Omega})$ in mode $d$ for  $(\Sigma, \Omega)$}
-  \begin{mathpar}
-    \inferrule{x : \Identifier}{|-_{\Sigma, \Omega} \isTerm{x}^{\syn}}\,\SynRule{Var}
-    \and
-    \inferrule{\cdot |-_{\Sigma} A \\ |-_{\Sigma, \Omega}\isTerm{t}^{\chk}}{|-_{\Sigma, \Omega} (\isTerm{t \annotate A})^{\syn}}\,\SynRule{Anno}
-    \and
-    \inferrule{|-_{\Sigma, \Omega} \isTerm{t}^{\syn}}{|-_{\Sigma, \Omega} \isTerm{t}^{\chk}}\,\ChkRule{Sub}
-    \and
-    \inferrule{|-_{\Sigma, \Omega} \isTerm{t_1}^\dir{d_1} \quad \cdots \quad |-_{\Sigma, \Omega} \isTerm{t_n}^\dir{d_n}}
-    {|-_{\Sigma, \Omega} \tmOpts^\dir{d}}\,\Rule{Op}
-  \end{mathpar}
-  \caption{Mode derivations}
-  \label{fig:mode-derivations}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \small
-  \judgbox{\Gamma |-_{\Sigma, \Omega} \isTerm{t} :^\dir{d} A}{A raw term $\isTerm{t}$ is of type $A$ (without type variables) in mode $\dir{d}$ under $\Gamma$}
+    \judgbox{\Gamma |-_{\Sigma, \Omega} \isTerm{t} :^\dir{d} A}{A raw term $\isTerm{t}$ is of type $A$ without type variables in mode $\dir{d}$ under $\Gamma$}
     \begin{mathpar}
@@
       \Gamma, \simsub{\vars{x}_\isTerm{n} : \Delta_{n}}{\rho} |-_{\Sigma, \Omega} \isTerm{t_n} :^{\dir{d_n}} \simsub{A_{n}}{\rho}}
-    {\Gamma |-_{\Sigma, \Omega} \tmOpts :^{\dir{d}} \simsub{A}{\rho}} \,\mathsf{Op}
-    \and \text{for $o \colon \Xi \rhd [\Delta_1]A_{1}^{\dir{d_1}}, \ldots, [\Delta_{n}] A_{n}^{\dir{d_n}} \to A^\dir{d}$ in $\Omega$}
+      {\Gamma |-_{\Sigma, \Omega} \tmOpts :^{\dir{d}} \simsub{A_0}{\rho}} \,\mathsf{Op}
     \end{mathpar}
-Abbreviations: $\boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \syn A} \defeq \boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A^{\syn}}$ and $\boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \chk A} \defeq \boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A^{\chk}}$
+    Abbreviations: $\boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \syn A} \defeq \boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} :^{\syn} A}$ and $\boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \chk A} \defeq \boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} :^{\chk} A}$
     \caption{Bidirectional typing derivations}
@@
   \end{figure}
+  \begin{figure}
+    \centering
+    \small
+    \judgbox{V |-_{\Sigma, \Omega} \isTerm{t}^\dir{d}}{A raw term $\isTerm{t}$ is in mode $d$}
+    \begin{mathpar}
+      \inferrule{x : V}{V |-_{\Sigma, \Omega} \isTerm{x}^{\syn}}\,\SynRule{Var}
+      \and
+      \inferrule{\cdot |-_{\Sigma} A \\ V |-_{\Sigma, \Omega}\isTerm{t}^{\chk}}{V |-_{\Sigma, \Omega} (\isTerm{t \annotate A})^{\syn}}\,\SynRule{Anno}
+      \and
+      \inferrule{V |-_{\Sigma, \Omega} \isTerm{t}^{\syn}}{V |-_{\Sigma, \Omega} \isTerm{t}^{\chk}}\,\ChkRule{Sub}
+      \and
+      \inferrule{V, \vars{x}_1 |-_{\Sigma, \Omega} \isTerm{t_1}^\dir{d_1} \\ \cdots \\ V, \vars{x}_{n} |-_{\Sigma, \Omega} \isTerm{t_n}^\dir{d_n}}
+      {V |-_{\Sigma, \Omega} \tmOpts^\dir{d}}\,\Rule{Op}
+    \end{mathpar}
+    \caption{Mode derivations}
+    \label{fig:mode-derivations}
+  \end{figure}
+  In a bidirectional type system specified $(\Sigma, \Omega)$,
+  \begin{itemize}
+    \item the set of \emph{bidirectional typing derivations} of $\Gamma \vdash t :^\dir{d} A$ for a context $\Gamma$ of typed variables, a raw term $\isTerm{t}$ under $\erase{\Gamma}$, a mode $\dir{d}$, and a type $A$ is defined by \Cref{fig:bidirectional-typing-derivations};
+    \item the set of \emph{mode derivations} of $V |- t^\dir{d}$ for a list $V$ of variables, a raw term $\isTerm{t}$ under $V$, and mode $d$ is defined by \Cref{fig:mode-derivations}.
+  \end{itemize}
+  The subscript $(\Sigma, \Omega)$ in judgements above is suppressed.
+\end{definition}
 
-\LT{What induction principle are we using?}
-\begin{theorem}[Mutual Induction Principles]
+\begin{example}
+  Bidirectional typing derivations and mode derivations for simply typed $\lambda$-calculus (\Cref{fig:STLC-bidirectional-typing-derivations,fig:STLC-mode-derivations}) can be specified by the type signature $\Sigma_{\bto}$ (\Cref{ex:type-signature-for-function-type}) and the bidirectional binding signature consisting of $\mathsf{abs}\colon A , B \rhd [A]B^{\chk} \to (A \bto B)^{\chk}$ and $\mathsf{app}\colon A, B \rhd (A \bto B)^{\syn}, A^{\chk} \to B^{\syn}$.
+\end{example}
 
-\end{theorem}
+After developing these generic definitions, we are now able to reason about any bidirectional simple type system specified by a pair of signatures $(\Sigma, \Omega)$ in the following two sections.

### 2023-06-29 tex/sec6-formalisation.tex
[ tex ] Clarify the abstract
[ tex ] Done with S3 for now

@@
 \begin{enumerate}
-  \item (Bidirectional) binding signature, functor, and terms (extrinsic typing, raw terms, raw terms in some mode) --- 3 pp
+  \item (Bidirectional) binding signature, functor, and terms (extrinsic typing, raw terms, raw terms in some mode) --- 2.5 pp
   \item Compare the induction principle and the \Agda proof of Soundness --- 2 pp
@@
 
-\LT{7 pages in total}
+\LT{First, it is noteworthy that the set of term constructs in a type theory need not be finite.
+For instance, a type theory may adopt a spine application which takes indefinitely many arguments---for each number $n$ of arguments an $(n+1)$-ary application construct can be introduced, so even if each operation has a definite number of arguments by definition a spine application is still expressible.}
+
+\LT{6 pages in total}

### 2023-06-29 tex/sec3-defs.tex
[ tex ] Use "indexed by" explicitly instead of "for" in S3

@@
 \end{figure}
-  In a simple type theory specified by $(\Sigma, \Omega)$, the set of \emph{raw terms} under a context~$V$ of variables consists of
+  In a simple type theory specified by $(\Sigma, \Omega)$, the set of \emph{raw terms} indexed by a context~$V$ of variables consists of
   \begin{enumerate}
@@
     \item annotations $\isTerm{t \annotate A}$ for some raw term $t$ and a type $A$, and
-    \item a construct $\tmOpts$ for some $o \colon \Xi \rhd [\Delta_1]A_{1}, \ldots, [\Delta_{n}] A_{n} \to A_0$ in $O$, where $t_i$'s are raw terms and $\vars{x}_{\isTerm{i}}$'s are lists of variables whose length is equal to the length of~$\Delta_i$,
+    \item a construct $\tmOpts$ for some $o \colon \Xi \rhd [\Delta_1]A_{1}, \ldots, [\Delta_{n}] A_{n} \to A_0$ in $O$, where $\vars{x}_{\isTerm{i}}$'s are lists of variables whose length is equal to the length of~$\Delta_i$ and $t_i$'s are raw terms under the context $V, \vars{x}_i$,
   \end{enumerate}
@@
   \end{figure}
-  In a simple type theory specified by $(\Sigma, \Omega)$, the set of \emph{typing derivations} of $\Gamma \vdash \isTerm{t} : A$ for a context $\Gamma$ of typed variables, a raw term under $\erase{\Gamma}$, and a type $A$ consists of
+  In a simple type theory specified by $(\Sigma, \Omega)$, the set of \emph{typing derivations} of $\Gamma \vdash \isTerm{t} : A$ indexed by a context $\Gamma$ of typed variables, a raw term under $\erase{\Gamma}$, and a type $A$ consists of
   \begin{enumerate}
@@
   \end{figure}
-  In a bidirectional type system specified $(\Sigma, \Omega)$,
+  In a bidirectional type system specified by $(\Sigma, \Omega)$,
   \begin{itemize}
-    \item the set of \emph{bidirectional typing derivations} of $\Gamma \vdash t :^\dir{d} A$ for a context $\Gamma$ of typed variables, a raw term $\isTerm{t}$ under $\erase{\Gamma}$, a mode $\dir{d}$, and a type $A$ is defined by \Cref{fig:bidirectional-typing-derivations};
-    \item the set of \emph{mode derivations} of $V |- t^\dir{d}$ for a list $V$ of variables, a raw term $\isTerm{t}$ under $V$, and mode $d$ is defined by \Cref{fig:mode-derivations}.
+    \item the set of \emph{bidirectional typing derivations} of $\Gamma \vdash t :^\dir{d} A$ indexed by a context $\Gamma$ of typed variables, a raw term $\isTerm{t}$ under $\erase{\Gamma}$, a mode $\dir{d}$, and a type $A$ is defined by \Cref{fig:bidirectional-typing-derivations};
+    \item the set of \emph{mode derivations} of $V |- t^\dir{d}$ indexed by a list $V$ of variables, a raw term $\isTerm{t}$ under $V$, and a mode $d$ is defined by \Cref{fig:mode-derivations}.
   \end{itemize}

### 2023-06-29 tex/sec3-defs.tex
[ tex ] Minor changes in S3

@@
 \end{figure}
-  In a simple type theory specified by $(\Sigma, \Omega)$, the set of \emph{raw terms} indexed by a context~$V$ of variables consists of
+  In a simple type theory specified by $(\Sigma, \Omega)$, the set of \emph{raw terms} indexed by a context~$V$ of free variables consists of
   \begin{enumerate}
     \item variables in $V$,
-    \item annotations $\isTerm{t \annotate A}$ for some raw term $t$ and a type $A$, and
-    \item a construct $\tmOpts$ for some $o \colon \Xi \rhd [\Delta_1]A_{1}, \ldots, [\Delta_{n}] A_{n} \to A_0$ in $O$, where $\vars{x}_{\isTerm{i}}$'s are lists of variables whose length is equal to the length of~$\Delta_i$ and $t_i$'s are raw terms under the context $V, \vars{x}_i$,
+    \item annotations $\isTerm{t \annotate A}$ for some raw term $t$ in $V$ and a type $A$, and
+    \item a construct $\tmOpts$ for some $o \colon \Xi \rhd [\Delta_1]A_{1}, \ldots, [\Delta_{n}] A_{n} \to A_0$ in $O$, where $\vars{x}_{\isTerm{i}}$'s are lists of variables whose length is equal to the length of~$\Delta_i$ and $t_i$'s are raw terms in the list $V, \vars{x}_i$ of variables,
   \end{enumerate}
@@
   \end{figure}
-  In a simple type theory specified by $(\Sigma, \Omega)$, the set of \emph{typing derivations} of $\Gamma \vdash \isTerm{t} : A$ indexed by a context $\Gamma$ of typed variables, a raw term under $\erase{\Gamma}$, and a type $A$ consists of
+  In a simple type theory specified by $(\Sigma, \Omega)$, the set of \emph{typing derivations} of $\Gamma \vdash \isTerm{t} : A$ indexed by a context $\Gamma$, a raw term with free variables in $\erase{\Gamma}$, and a type $A$ consists of
   \begin{enumerate}
@@
     \item a derivation of $\Gamma |- (t \annotate A) : A$ if $\Gamma \vdash \isTerm{t} : A$ is derivable, and
-    \item a derivation of $\Gamma |- \tmOpts : A$ for $o \colon \Xi \rhd [\Delta_1]A_1, \ldots, [\Delta_{n}]A_{n} \to A_0$ if there is a function $\rho\colon \Xi \to \Type_{\Sigma}(\emptyset)$ such that $A = \simsub{A_0}{\rho}$ and $\Gamma, \vars{x}_{\isTerm{i}} : \Delta_i \vdash \isTerm{t_i} : \simsub{A_i}{\rho}$ is derivable for all $i$,
+    \item a derivation of $\Gamma |- \tmOpts : \simsub{A_0}{\rho}$ for $o \colon \Xi \rhd [\Delta_1]A_1, \ldots, [\Delta_{n}]A_{n} \to A_0$ if there is a function $\rho\colon \Xi \to \Type_{\Sigma}(\emptyset)$ and a derivation of $\Gamma, \vars{x}_{\isTerm{i}} : \Delta_i \vdash \isTerm{t_i} : \simsub{A_i}{\rho}$ for each $i$,
   \end{enumerate}
@@
 \begin{example}
-  Raw terms and typing derivations for simply typed $\lambda$-calculus (\Cref{fig:STLC-raw-terms,fig:STLC-typing-derivations}) can be specified by the type signature $\Sigma_{\bto}$ (\Cref{ex:type-signature-for-function-type}) and the binding signature consisting of
+  Raw terms (\Cref{fig:STLC-raw-terms}) and typing derivations (\Cref{fig:STLC-typing-derivations}) for simply typed $\lambda$-calculus can be specified by the type signature $\Sigma_{\bto}$ (\Cref{ex:type-signature-for-function-type}) and the binding signature consisting of
  \[
@@
 
-We generalise bidirectional typing rules as well as model derivations for simply typed $\lambda$-calculus in \Cref{fig:STLC-bidirectional-typing-derivations,fig:STLC-mode-derivations} to derivations for a bidirectional type system specified by $(\Sigma, \Omega)$ where $\Sigma$ is still a type signature and $\Omega$ is a bidirectional binding signature, similar to \Cref{def:typing-derivations} for typing derivations but with an additional $\ChkRule{Sub}$ rule:
+We generalise bidirectional typing rules (\Cref{fig:STLC-bidirectional-typing-derivations}) as well as model derivations (\Cref{fig:STLC-mode-derivations}) for simply typed $\lambda$-calculus to derivations for a bidirectional type system specified by $(\Sigma, \Omega)$ where $\Sigma$ is still a type signature and $\Omega$ is a bidirectional binding signature, similar to \Cref{def:typing-derivations} for typing derivations but with an additional $\ChkRule{Sub}$ rule:
 \begin{definition}\label{def:bidirectional-typing-derivations}\label{def:mode-derivations}
@@
 \begin{example}
-  Bidirectional typing derivations and mode derivations for simply typed $\lambda$-calculus (\Cref{fig:STLC-bidirectional-typing-derivations,fig:STLC-mode-derivations}) can be specified by the type signature $\Sigma_{\bto}$ (\Cref{ex:type-signature-for-function-type}) and the bidirectional binding signature consisting of $\mathsf{abs}\colon A , B \rhd [A]B^{\chk} \to (A \bto B)^{\chk}$ and $\mathsf{app}\colon A, B \rhd (A \bto B)^{\syn}, A^{\chk} \to B^{\syn}$.
+  Bidirectional typing derivations and mode derivations for simply typed $\lambda$-calculus can be specified by the type signature $\Sigma_{\bto}$ (\Cref{ex:type-signature-for-function-type}) and the bidirectional binding signature consisting of $\mathsf{abs}\colon A , B \rhd [A]B^{\chk} \to (A \bto B)^{\chk}$ and $\mathsf{app}\colon A, B \rhd (A \bto B)^{\syn}, A^{\chk} \to B^{\syn}$.
 \end{example}

### 2023-06-30 tex/sec3-defs.tex
[ tex ] Minor changes in S3

@@
 %Here's the 'small print': Our definitions cover typical examples and set the scope for bidirectional simple type systems discussed, but do not offer comprehensive coverage of all possibilities.}
-of simple types, simple type theories, and bidirectional type systems using simply typed $\lambda$-calculus as our running example.
+of simple types, simple type theories, and bidirectional type systems using simply typed $\lambda$-calculus in \Cref{sec:key-ideas} as our running example.
 
@@
 \item we first introduce a notion of signature which includes a set of operation symbols and an assignment of arities to each operation;
-\item then, we define abstract syntax trees inductively with primitive constructs and constructs specified by a given signature.
+\item then, we define abstract syntax trees inductively with primitive constructs such as $\Rule{Var}$ and constructs specified by a given signature.
 \end{enumerate*}
-Upon moving from simple types to bidirectionally typed terms, the notion of arity, initially as the number of arguments of an operation, is enriched to incorporate an extension context for each argument and a mode denoting the direction of type information flow.
+Upon moving from simple types to bidirectionally typed terms, the notion of arity, initially as the number of arguments of an operation, is enriched to incorporate an extension context for variable binding and the direction of type information flow.
 
 \subsection{Signatures and Simple Types} \label{subsec:simple-types}
-Simple types of a language is understood intuitively as a collection of types whose formation judgement is not indexed by anything and the only datum needed for specifying a type construct is its number of arguments:
+Simple types of a language is understood intuitively as a collection of types whose formation judgement is not indexed by anything but a fixed set of type variables and the only datum needed for specifying a type construct is its number of arguments:
 
@@
   \begin{figure}
-    \begin{minipage}[b]{.6\textwidth}
+%    \begin{minipage}[b]{.6\textwidth}
       \centering
@@
       \label{fig:simple-type}
-    \end{minipage}
-    \begin{minipage}[b]{.35\textwidth}
-      \centering
-      \small
-      \judgbox{\Xi |-_{\Sigma} \Gamma}{}
-      \begin{mathpar}
-        \inferrule{ }{\Xi |-_{\Sigma} \cdot }\and
-        \inferrule{\Xi |-_{\Sigma} A \\ \Xi|-_{\Sigma} \Gamma}{\Xi |-_{\Sigma} \Gamma, x : A}
-      \end{mathpar}
-      \caption{Contexts}
-    \label{fig:simple-context}
-    \end{minipage}
+%    \end{minipage}
+%    \begin{minipage}[b]{.35\textwidth}
+%      \centering
+%      \small
+%      \judgbox{\Xi |-_{\Sigma} \Gamma}{}
+%      \begin{mathpar}
+%        \inferrule{ }{\Xi |-_{\Sigma} \cdot }\and
+%        \inferrule{\Xi |-_{\Sigma} A \\ \Xi|-_{\Sigma} \Gamma}{\Xi |-_{\Sigma} \Gamma, x : A}
+%      \end{mathpar}
+%      \caption{Contexts}
+%    \label{fig:simple-context}
+%    \end{minipage}
   \end{figure}
-
   A \emph{signature} $\Sigma$ for simple types consists of a set\footnote{%
@@
     \item a variable in $\Xi$ or
-    \item $\tyOp_{i}(A_1, \ldots, A_n)$ for some $i:I$ with $\arity(i) = n$ and types $A_1,\ldots, A_n$
+    \item $\tyOp_{i}(A_1, \ldots, A_n)$ for some $i:I$ with $\arity(i) = n$ and types $A_1,\ldots, A_n$.
   \end{enumerate}
-  as defined in \Cref{fig:simple-type}.
-  A \emph{context} $\Gamma \colon \Cxt_{\Sigma}(\Xi)$ over $\Xi$ is a list $\Gamma$ of types over $\Xi$ as defined in \Cref{fig:simple-context}. 
+  %as defined in \Cref{fig:simple-type}.
+  A \emph{context} $\Gamma \colon \Cxt_{\Sigma}(\Xi)$ over $\Xi$ is a list $\Gamma$ of types over $\Xi$. 
 \end{definition}
@@
   Simply typed $\lambda$-calculus includes function types $A \bto B$ and typically a base type~$\mathtt{b}$ to ensure that the set of all types without type variables is non-empty.
-  The type signature $\Sigma_{\bto}$ used to define types in simply typed $\lambda$-calculus consists of a binary operation $\mathsf{imp}$ and a nullary operation $\mathsf{b}$ where $\arity(\mathsf{imp}) = 2$ and $\arity(\mathsf{b}) = 0$.
-  Then, all types in simply typed $\lambda$-calculus can be given as $\Sigma_{\bto}$-types over the empty set where $A \bto B$ is introduced as $\tyOp_{\mathsf{imp}}(A, B)$ and $\mathtt{b}$ as $\tyOp_{\mathsf{b}}$.
+  The type signature $\Sigma_{\bto}$ used to define types in simply typed $\lambda$-calculus consists of a binary operation $\mathsf{fun}$ and a nullary operation $\mathsf{b}$ where $\arity(\mathsf{fun}) = 2$ and $\arity(\mathsf{b}) = 0$.
+  Then, all types in simply typed $\lambda$-calculus can be given as $\Sigma_{\bto}$-types over the empty set with $A \bto B$ introduced as $\tyOp_{\mathsf{fun}}(A, B)$ and $\mathtt{b}$ as $\tyOp_{\mathsf{b}}$.
 \end{example}
 
-Without the loss of generality, we may use the set $\Fin(n)$ of natural numbers less than $n$ as the set of type variables in this paper, following the inductive definition by \citet{Dybjer1994}.
+Without the loss of generality, we may use the family of sets $\Fin(n)$ of natural numbers less than $n$ as sets of type variables in this paper, following the inductive definition by \citet{Dybjer1994}.
 In bidirectional type synthesis, it is necessary to have a decidable equality on $\Type_{\Sigma}(\Xi)$, as we need to compare the synthesized type $B$ with a given type $A$ for the $\Rule{Sub}$ rule.
-When $\Xi$ has a decidable equality, the desired decidable equality can be given.
-Since the arity of a type operation is always finite, it is sufficient to use a finite set like $\Fin(n)$.
-Additionally, $\Fin(n)$ has a decidable equality, ensuring that the set of simple types over $\Fin(n)$ also has a decidable quality.
-Therefore, for convenience, we can represent the set of type variables using a natural number.
-However, for illustrative purposes, variables will still be assigned names.
+If $\Xi$ such as $\Fin(n)$ has a decidable equality , the decidable equality on $\Type_{\Sigma}(\Xi)$ can be given.
+In addition, the arity of a construct is always finite, so it suffices to use natural numbers to denote type variables in a $\Sigma$-type.
+Formally, we restrict the set $\Xi$ of type variables to the family of $\Fin(n)$, but for illustrative purposes we still use named variables for presentation.
 
@@
 \begin{definition}
-  The \emph{(simultaneous) substitution} of a type $A$ over $\Xi$ for a function $\rho\colon\Xi \to \Type_{\Sigma}(\Xi')$ mapping $A$ to a type $\simsub{A}{\rho}$ over $\Xi'$ is defined as usual.
+  The \emph{(simultaneous) type substitution} for a function $\rho\colon\Xi \to \Type_{\Sigma}(\Xi')$ is a map, which sends $A : \Type_{\Sigma}(\Xi)$ to $\simsub{A}{\rho} : \Type_{\Sigma}(\Xi')$, defined as usual.
   By abuse of notation, the substitution $\simsub{\Gamma}{\rho}$ of a context $\Gamma$ is defined by applying substitution $\simsub{A}{\rho}$ to each type $A$ in the context.
@@
 
-For a type signature $\Sigma$, a simple type theory consists of
+A simple type theory consists of
 \begin{enumerate*}
-  \item a set of raw terms $\isTerm{t}$ indexed by a set $V$ of untyped variables where each construct is allowed to bind some variables (as $\Rule{Abs}$) and to take multiple arguments (as $\Rule{App}$) like \Cref{fig:STLC-raw-terms};
-  \item a set of typing derivations indexed by a list $\Gamma$ of typed variables, a type~$A$, and a raw term $\isTerm{t}$ to set their type constraints like \Cref{fig:STLC-typing-derivations}. 
+  \item a set of raw terms $\isTerm{t}$ indexed by a set $V$ of untyped variables where each construct is allowed to bind some variables (as $\Rule{Abs}$) and to take multiple arguments (as $\Rule{App}$);
+  \item a set of typing derivations indexed by a list $\Gamma$ of typed variables, a type~$A$, and a raw term $\isTerm{t}$ to set their type constraints. 
 \end{enumerate*}
-Therefore, to specify a construct like $\Rule{Abs}$ or $\Rule{App}$, we need to enrich the notion of signatures with an extension context for variable binding for each argument, a type for each argument, and a type for the operation:
+Therefore, to specify a construct like $\Rule{Abs}$ or $\Rule{App}$, we need to enrich the notion of signature with a type and an extension context for variable binding for each argument and a type for the operation:
 \begin{definition}\label{def:binding-signature}
@@
   \item $\left[\left(\Delta_1; A_{1}\right), \ldots, \left(\Delta_{n}; A_{n}\right) \right]$ is a list of pairs where
-  \item the $i$-th pair $(\Delta_i, A_i) :\Cxt_{\Sigma}(\Xi) \times \Term_{\Sigma}(\Xi)$ are for types of the binding variables and the type of the $i$-th argument of $\tmOp_{o}$.
+    \begin{enumerate}
+      \item $\Delta_i : \Cxt_{\Sigma}(\Xi)$ is for the types of bound variables and
+      \item $A_i : \Term_{\Sigma}(\Xi)$ for the type of the $i$-th argument of $\tmOp_{o}$.
+    \end{enumerate}
 \end{enumerate}
@@
 
-We define raw terms specified by a type signature $\Sigma$ and a binding signature $\Omega$ as follows.
+Also for brevity, a simple type theory $(\Sigma, \Omega)$ refers to the simple type theory specified by some type signature $\Sigma$ and a binding signature $\Omega$.
+
 \begin{definition}
@@
 \end{figure}
-  In a simple type theory specified by $(\Sigma, \Omega)$, the set of \emph{raw terms} indexed by a context~$V$ of free variables consists of
+  For a simple type theory $(\Sigma, \Omega)$, the set of \emph{raw terms} indexed by a context~$V$ of free variables consists of
   \begin{enumerate}
@@
     \item annotations $\isTerm{t \annotate A}$ for some raw term $t$ in $V$ and a type $A$, and
-    \item a construct $\tmOpts$ for some $o \colon \Xi \rhd [\Delta_1]A_{1}, \ldots, [\Delta_{n}] A_{n} \to A_0$ in $O$, where $\vars{x}_{\isTerm{i}}$'s are lists of variables whose length is equal to the length of~$\Delta_i$ and $t_i$'s are raw terms in the list $V, \vars{x}_i$ of variables,
+    \item a construct $\tmOpts$ for some $o \colon \Xi \rhd [\Delta_1]A_{1}, \ldots, [\Delta_{n}] A_{n} \to A_0$ in $O$, where $\vars{x}_{\isTerm{i}}$'s are lists of variables whose length is equal to the length of~$\Delta_i$ and $t_i$'s are raw terms in the variable list $V, \vars{x}_i$
   \end{enumerate}
@@
       \and
-      \inferrule{\Gamma |- \isTerm{t} : A}{\Gamma |- (\isTerm{t \annotate A}) : A}\,\Rule{Anno}
+      \inferrule{\Gamma |-_{\Sigma, \Omega} \isTerm{t} : A}{\Gamma |-_{\Sigma, \Omega} (\isTerm{t \annotate A}) : A}\,\Rule{Anno}
       \and
@@
   \end{figure}
-  In a simple type theory specified by $(\Sigma, \Omega)$, the set of \emph{typing derivations} of $\Gamma \vdash \isTerm{t} : A$ indexed by a context $\Gamma$, a raw term with free variables in $\erase{\Gamma}$, and a type $A$ consists of
+  For a simple type theory $(\Sigma, \Omega)$, the set of \emph{typing derivations} of $\Gamma \vdash \isTerm{t} : A$ indexed by a context $\Gamma$, a raw term with free variables in $\erase{\Gamma}$, and a type $A$ consists of
   \begin{enumerate}
-    \item a derivation of $\Gamma |- x : A$ if $x : A$ is in $\Gamma$,
-    \item a derivation of $\Gamma |- (t \annotate A) : A$ if $\Gamma \vdash \isTerm{t} : A$ is derivable, and
-    \item a derivation of $\Gamma |- \tmOpts : \simsub{A_0}{\rho}$ for $o \colon \Xi \rhd [\Delta_1]A_1, \ldots, [\Delta_{n}]A_{n} \to A_0$ if there is a function $\rho\colon \Xi \to \Type_{\Sigma}(\emptyset)$ and a derivation of $\Gamma, \vars{x}_{\isTerm{i}} : \Delta_i \vdash \isTerm{t_i} : \simsub{A_i}{\rho}$ for each $i$,
+    \item a derivation of $\Gamma |-_{\Sigma, \Omega} x : A$ if $x : A$ is in $\Gamma$,
+    \item a derivation of $\Gamma |-_{\Sigma, \Omega} (t \annotate A) : A$ if $\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A$ has a derivation, and
+    \item a derivation of $\Gamma |-_{\Sigma, \Omega} \tmOpts : \simsub{A_0}{\rho}$ for $o \colon \Xi \rhd [\Delta_1]A_1, \ldots, [\Delta_{n}]A_{n} \to A_0$ if there is a function $\rho\colon \Xi \to \Type_{\Sigma}(\emptyset)$ and a derivation of $\Gamma, \vars{x}_{\isTerm{i}} : \Delta_i \vdash_{\Sigma, \Omega} \isTerm{t_i} : \simsub{A_i}{\rho}$ for each $i$,
   \end{enumerate}
   corresponding to rules $\Rule{Var}$, $\Rule{Anno}$, and $\Rule{Op}$ in \Cref{fig:extrinsic-typing} respectively.
-  For better readability, the subscript $(\Sigma, \Omega)$ in the above typing judgements is suppressed.
 \end{definition}
 Note that, by definition, our simple type theory is always syntax-directed, since each rule in \Cref{fig:raw-terms} corresponds exactly to one rule in \Cref{fig:extrinsic-typing}. 
-The $\Rule{Op}$ in both figures is a rule schema and gives rise to as many rules as there are operations in $o : O$.
+The $\Rule{Op}$ in both figures is a rule schema and gives rise to as many rules as there are operations $o : O$.
 
@@
  \]
- Rules $\Rule{Abs}$ and $\Rule{App}$ in simply typed $\lambda$-calculus are subsumed by the $\Rule{Op}$ rule schema, as terms $\tmOp_{\mathsf{app}}(t, u)$ and $\tmOp_{\mathsf{abs}}(x.t)$ are exactly an application $t\;u$ and an abstraction $\lam{x}t$, respectively.
+ Rules $\Rule{Abs}$ and $\Rule{App}$ in simply typed $\lambda$-calculus are subsumed by the $\Rule{Op}$ rule schema, as an application $t\;u$ and an abstraction $\lam{x}t$ can be introduced as $\tmOp_{\mathsf{app}}(t, u)$ and $\tmOp_{\mathsf{abs}}(x.t)$, respectively.
 \end{example}
@@
 Finally, for a bidirectional type system, we have raw terms and typing derivations like a simple type theory, but its typing judgements come in two forms $\Gamma |- t \syn A$ and $\Gamma |- t \chk A$.
-Observe that these two typing judgements are just one typing judgement $\Gamma |- t :^\dir{d} A$ indexed additionally by a \emph{mode} $d : \Mode$ where $\Mode$ consists of two inhabitants $\syn$ for synthesis and $\chk$ for checking.
-Hence, to specify a bidirectional type system, we need further enrich the notion of binding signatures with a mode for each argument and a mode for an operation itself as follows.
+Observe that these two typing judgements are just one typing judgement $\Gamma |- t :^\dir{d} A$ indexed additionally by a \emph{mode} $d : \Mode$---which is either $\syn$ or $\chk$---so that $\Gamma |- t \syn A$ and $\Gamma |- t \chk A$ are just abbreviations of $\Gamma |- t :^{\syn} A$ and $\Gamma |- t :^{\chk} A$ respectively.
+Hence, to specify a bidirectional type system, we need further enrich the notion of binding signatures with modes as follows.
 
@@
 
+We also say a bidirectional type system $(\Sigma, \Omega)$ referring the bidirectional type system specified by some type signature $\Sigma$ and a bidirectional binding signature $\Omega$.
+
 \begin{remark}
@@
 
-We generalise bidirectional typing rules (\Cref{fig:STLC-bidirectional-typing-derivations}) as well as model derivations (\Cref{fig:STLC-mode-derivations}) for simply typed $\lambda$-calculus to derivations for a bidirectional type system specified by $(\Sigma, \Omega)$ where $\Sigma$ is still a type signature and $\Omega$ is a bidirectional binding signature, similar to \Cref{def:typing-derivations} for typing derivations but with an additional $\ChkRule{Sub}$ rule:
+We generalise bidirectional typing rules (\Cref{fig:STLC-bidirectional-typing-derivations}) as well as model derivations (\Cref{fig:STLC-mode-derivations}) for simply typed $\lambda$-calculus to derivations for a bidirectional type system $(\Sigma, \Omega)$ where $\Omega$ is some bidirectional binding signature, similar to \Cref{def:typing-derivations} but with an additional $\ChkRule{Sub}$ rule:
 \begin{definition}\label{def:bidirectional-typing-derivations}\label{def:mode-derivations}
@@
     \small
-    \judgbox{V |-_{\Sigma, \Omega} \isTerm{t}^\dir{d}}{A raw term $\isTerm{t}$ is in mode $d$}
+    \judgbox{V |-_{\Sigma, \Omega} \isTerm{t}^\dir{d}}{A raw term $\isTerm{t}$ is in mode $d$ with free variables in $V$}
     \begin{mathpar}
-      \inferrule{x : V}{V |-_{\Sigma, \Omega} \isTerm{x}^{\syn}}\,\SynRule{Var}
+      \inferrule{x \in V}{V |-_{\Sigma, \Omega} \isTerm{x}^{\syn}}\,\SynRule{Var}
       \and
@@
   \end{figure}
-  In a bidirectional type system specified by $(\Sigma, \Omega)$,
+  For a bidirectional type system $(\Sigma, \Omega)$,
   \begin{itemize}
-    \item the set of \emph{bidirectional typing derivations} of $\Gamma \vdash t :^\dir{d} A$ indexed by a context $\Gamma$ of typed variables, a raw term $\isTerm{t}$ under $\erase{\Gamma}$, a mode $\dir{d}$, and a type $A$ is defined by \Cref{fig:bidirectional-typing-derivations};
-    \item the set of \emph{mode derivations} of $V |- t^\dir{d}$ indexed by a list $V$ of variables, a raw term $\isTerm{t}$ under $V$, and a mode $d$ is defined by \Cref{fig:mode-derivations}.
+    \item the set of \emph{bidirectional typing derivations} of $\Gamma \vdash_{\Sigma, \Omega} t :^\dir{d} A$ indexed by a context $\Gamma$ of typed variables, a raw term $\isTerm{t}$ under $\erase{\Gamma}$, a mode $\dir{d}$, and a type $A$ is defined in \Cref{fig:bidirectional-typing-derivations};
+    \item the set of \emph{mode derivations} of $V |-_{\Sigma, \Omega} t^\dir{d}$ indexed by a list $V$ of variables, a raw term $\isTerm{t}$ under $V$, and a mode $d$ is defined in \Cref{fig:mode-derivations}.
   \end{itemize}
-  The subscript $(\Sigma, \Omega)$ in judgements above is suppressed.
 \end{definition}

### 2023-06-30 tex/sec1-intro.tex
[ tex ] No spaces between an em-dash and the adjacent words

@@
 The former means that the type of a term is synthesised, using both the term and its context as input, while for the latter a term is to be checked against a given type.
-Then, a type synthesis algorithm can be `read off' from a bidirectional type system, provided that a condition called \emph{mode-correctness} --- every synthesised type in a typing rule is determined by previously synthesised types from its premises and its input if checking --- is satisfied for every bidirectional typing rule.
+Then, a type synthesis algorithm can be `read off' from a bidirectional type system, provided that a condition called \emph{mode-correctness}---every synthesised type in a typing rule is determined by previously synthesised types from its premises and its input if checking---is satisfied for every bidirectional typing rule.
 
@@
 To clarify the difference (\Cref{sec:annotatability}), we propose \emph{mode derivations} (\Cref{fig:mode-derivations,fig:generalised-mode-derivations}) and \emph{mode preprocessing} (\Cref{sec:mode-preprocessing}) assigning a mode first to a raw term with enough annotations or otherwise pinpoints missing annotations.
-By combining bidirectional type synthesis, soundness, completeness, and mode preprocessing, we achieve our main result (\Cref{cor:trichotomy}) --- a \emph{trichotomy} of raw terms: a type synthesiser which checks if a given raw term is \emph{suitably annotated} and if a suitably annotated raw term has a typing derivation with respect to the specified simple type theory.
+By combining bidirectional type synthesis, soundness, completeness, and mode preprocessing, we achieve our main result (\Cref{cor:trichotomy})---a \emph{trichotomy} of raw terms: a type synthesiser which checks if a given raw term is \emph{suitably annotated} and if a suitably annotated raw term has a typing derivation with respect to the specified simple type theory.
 Our bidirectional type synthesis can be instantiated for any system specified by a mode-correct signature, effectively a verified type-checker generator.

### 2023-06-30 tex/sec2-outline.tex
[ tex ] No spaces between an em-dash and the adjacent words

@@
 We start with an overview of our theory by instantiating it to simply typed $\lambda$-calculus.
-Roughly speaking, the problem of type synthesis requires us to take an untyped abstract syntax tree ---which we call a \emph{raw term}--- as input, and produce a typing derivation for the term if possible.
+Roughly speaking, the problem of type synthesis requires us to take an untyped abstract syntax tree---which we call a \emph{raw term}---as input, and produce a typing derivation for the term if possible.
 To give more precise definitions:

### 2023-06-30 tex/sec3-defs.tex
[ tex ] Minor fixes

@@
 Observe that these two typing judgements are just one typing judgement $\Gamma |- t :^\dir{d} A$ indexed additionally by a \emph{mode} $d : \Mode$---which is either $\syn$ or $\chk$---so that $\Gamma |- t \syn A$ and $\Gamma |- t \chk A$ are just abbreviations of $\Gamma |- t :^{\syn} A$ and $\Gamma |- t :^{\chk} A$ respectively.
-Hence, to specify a bidirectional type system, we need further enrich the notion of binding signatures with modes as follows.
+Hence, to specify a bidirectional type system, we need further enrich the notion of binding signature with modes as follows.
 
@@
   Every bidirectional binding signature $\Omega$ gives rise to a binding signature $\erase{\Omega}$ if we erase modes from $\Omega$, so a bidirectional binding signature $\Omega$ also specifies a simple type theory, including its raw terms and typing derivations.
-  For this notion, we call $\erase{\Omega}$ the \emph{(mode) erasure} of $\Omega$.
+  We call $\erase{\Omega}$ the \emph{(mode) erasure} of $\Omega$.
 \end{remark}
 
-We generalise bidirectional typing rules (\Cref{fig:STLC-bidirectional-typing-derivations}) as well as model derivations (\Cref{fig:STLC-mode-derivations}) for simply typed $\lambda$-calculus to derivations for a bidirectional type system $(\Sigma, \Omega)$ where $\Omega$ is some bidirectional binding signature, similar to \Cref{def:typing-derivations} but with an additional $\ChkRule{Sub}$ rule:
+We generalise bidirectional typing rules (\Cref{fig:STLC-bidirectional-typing-derivations}) as well as mode derivations (\Cref{fig:STLC-mode-derivations}) for simply typed $\lambda$-calculus to derivations for a bidirectional type system $(\Sigma, \Omega)$ where $\Omega$ is some bidirectional binding signature, similar to \Cref{def:typing-derivations} but with an additional $\ChkRule{Sub}$ rule:
 \begin{definition}\label{def:bidirectional-typing-derivations}\label{def:mode-derivations}
@@
 
-After developing these generic definitions, we are now able to reason about any bidirectional simple type system specified by a pair of signatures $(\Sigma, \Omega)$ in the following two sections.
+After developing these generic definitions, we are now able to reason about constructions and properties that hold for any bidirectional simple type system $(\Sigma, \Omega)$.

### 2023-06-30 tex/sec7-future.tex
[ tex ] Minor fixes

@@
 \paragraph{Beyond simple types}
-While algebraic approaches to polymorphic type theories have been developed~\citep{Fiore2013,Hamana2011}, these approaches do not take subtyping into account.
-Subtyping is essential for formulating important concepts in type synthesis such as \emph{principal types}.
+While algebraic approaches to polymorphic types have been developed~\citep{Fiore2013,Hamana2011}, these approaches do not take subtyping into account.
+Subtyping is essential for formulating important concepts such as \emph{principal types} in type synthesis.
 On the other hand, in the realm of dependent types, \citeauthor{Cartmell1986}'s generalized algebraic theories~\citeyearpar{Cartmell1986} can handle a wide variety of dependent type theories.

### 2023-07-05 tex/BiSig.tex
[ tex ] Compile Sec6 separately with lhs2TeX
[ tex ] Revised abstract & S1

@@
 \documentclass[acmsmall,screen]{acmart}
+
+\acmJournal{PACMPL}
+\acmVolume{0}
+\acmNumber{0}
+\acmArticle{0}
+\acmMonth{7}
 
@@
 
+\usepackage{xifthen}
+\newcommand{\varcitet}[3][]{\citeauthor{#2}#3~[\ifthenelse{\isempty{#1}}{\citeyear{#2}}{\citeyear[#1]{#2}}]}
+
 \usepackage[inline]{enumitem} % for environment enumerate*
-%\setlist[enumerate]{mode=unboxed}
 
-\setlength {\marginparwidth }{2cm}
-\usepackage[obeyFinal,color=yellow,textsize=small]{todonotes}
+\setlength{\marginparwidth}{1.25cm}
+\usepackage[obeyFinal,color=yellow,textsize=scriptsize]{todonotes}
 
 \usepackage{microtype}
+\usepackage{subfiles}
 
@@
 \theoremstyle{acmdefinition}
-\newtheorem{remark}[theorem]{Remark}}
+\newtheorem{remark}[theorem]{Remark}
+}
+
+\newcommand{\LT}[1]{\todo[author=LT,inline,color=orange!40]{{#1}}}
+\newcommand{\Josh}[1]{\todo[author=Josh,inline,color=yellow!40]{{#1}}}
 
@@
 \input{macros}
-
-\newcommand{\LT}[1]{\todo[author=LT,inline,color=orange!40]{{#1}}}
-\newcommand{\Josh}[1]{\todo[author=Josh,inline,color=yellow!40]{{#1}}}
+\usepackage{mylhs2tex}
 
@@
 \begin{abstract}
-  Type checking and inference serve as the transition from parsed abstract syntax trees to well-typed programs.
-  While progress has been made in developing type-checking techniques on a case-by-case basis, the lack of a general theory poses a difficulty in implementing a robust `type-checker generator' for diverse languages.
-
-  In response to this difficulty, we develop a simple yet general and constructive theory of bidirectional type synthesis where bidirectional type synthesis refers to the mutual use of type checking and inference based on bidirectional typing rules to determine the type of a program with type annotations.
-  First, our theory is simple yet general as it deals with any simple type theory and its bidirectional type system that can be specified by a signature, analogous to the role of grammar in parsing.
-  Second, our theory is constructive because it is not only formalised in a proof assistant based on constructive type theory but also formulated towards computation.
-  Specifically, the `proof' of logical decidability of type synthesis is instantiated to a synthesiser for the simple type theory specified by a valid signature, analogous to a parser generator.
-  We introduce the notion of mode preprocessing to refine annotatability in bidirectional typing.
-  Type synthesis and mode preprocessing result in the trichotomy of raw terms, which models a practical type synthesiser that may throw an exception and locate missing annotations.
-  By formalising our general theory of bidirectional type synthesis constructively, we provide a correct-by-construction `type-checker generator' for simple type theories, taking the first step towards a more general theory of bidirectional type synthesis.
+%  Type checking and inference serve as the transition from parsed abstract syntax trees to well-typed programs.
+  While there is much progress in designing bidirectional type systems and associated type synthesis algorithms on a case-by-case basis, the lack of a general theory makes it difficult to study the common principles and develop widely applicable techniques and practical tools such as a `type-synthesiser generator'.
+  In response to this difficulty, we propose a simple yet general and constructive theory of bidirectional type synthesis.
+%  which refers to the mutual use of type checking and synthesis based on bidirectional typing rules to determine the type of a program with type annotations.
+  First, our theory is simple yet general as it applies to any syntax-directed bidirectional simple type system that can be specified by a signature.
+%  which is analogous to a grammar in parsing.
+  Second, our theory is constructive because it is not only formalised in a proof assistant based on Martin-L\"of type theory, but also formulated in a way that unifies computation and proof, resulting in a compact and practical theory---for example, a proof of the decidability of the type synthesis problem is also computationally a type synthesiser.
+  Moreover, we introduce `mode preprocessing' to deal with missing type annotations, an aspect often overlooked in the literature.
+  The main theorem is logically a trichotomy on parsed syntax trees and computationally a type synthesiser that may throw an exception due to missing type annotations.
+  Being parametrised by a signature, our theory works as a verified type-synthesiser generator overall.
 \end{abstract}
@@
 \input{sec5-synthesis.tex}
-\input{sec6-formalisation.tex}
+\subfile{sec6-formalisation}
 \input{sec7-future.tex}

### 2023-07-05 tex/sec1-intro.tex
[ tex ] Revised abstract & S1

--- 2023-06-28 tex/sec1-intro.tex
@@
 
-Type checking and inference serve as the transition to type-checked programs, also known as \emph{well-typed terms}, from parsed abstract syntax trees, referred as \emph{raw terms} by contrast.
-Type inference algorithms were conceived to ascertain the type of any \emph{unannotated} raw term.
+Type inference is an important mechanism for the transition to well-typed programs from parsed abstract syntax trees, which we call \emph{raw terms}.
+Here `type inference' refers specifically to algorithms that ascertain the type of any raw term \emph{without type annotations}.
 However, it was later found that full parametric polymorphism leads to undecidability in type inference, as do dependent types~\citep{Wells1999,Dowek1993}.
-In light of these limitations, bidirectional type synthesis emerged as a viable alternative, providing decidable algorithms for determining the type of a \emph{suitably annotated} raw term.
-\citet{Dunfield2021} summarised its design principles and its wide coverage of languages with simple types, polymorphic types, dependent types, gradual types, among others.
+In light of these limitations, bidirectional type synthesis emerged as a viable alternative, providing algorithms for deciding the types of raw terms that meet some syntactic criteria%
+\todo{confirm}
+and usually contain some type annotations.
+\citet{Dunfield2021} summarised the design principles of bidirectional type synthesis and its wide coverage of languages with simple types, polymorphic types, dependent types, gradual types, among others.
 
-The idea of bidirectional type synthesis begins with extending typing judgements with two modes that direct the flow of types:
+The basic idea of bidirectional type synthesis is that while the problem of type inference is not decidable in general, for certain kinds of terms it is still possible to infer their types (for example, the type of a variable can be looked up in the context); for other kinds of terms, we can switch to the simpler problem of type checking, where the expected type of a raw term is also given so that there is more information to work with.
+More formally, judgements in a bidirectional type system are extended with two \emph{modes}:
 \begin{enumerate*}
-  \item $\Gamma |- \isTerm{t} \syn A$ for synthesis and 
-  \item $\Gamma |- \isTerm{t} \chk A$ for checking.
+  \item $\Gamma |- \isTerm{t} \syn A$ for \emph{synthesis} and 
+  \item $\Gamma |- \isTerm{t} \chk A$ for \emph{checking}.
 \end{enumerate*}
-The former means that the type of a term is synthesised, using both the term and its context as input, while for the latter a term is to be checked against a given type.
-Then, a type synthesis algorithm can be `read off' from a bidirectional type system, provided that a condition called \emph{mode-correctness} --- every synthesised type in a typing rule is determined by previously synthesised types from its premises and its input if checking --- is satisfied for every bidirectional typing rule.
+The former indicates that the type~$A$ is computed as output, using both the context~$\Gamma$ and the term~$t$ as input, while for the latter, all three of $\Gamma$, $t$, and~$A$ are given as input.
+The algorithm of a bidirectional type synthesiser can usually be `read off' from a well designed bidirectional type system: as the synthesiser traverses a raw term, it switches between synthesis and checking, following the modes assigned to the judgements in the typing rules.
+%if a condition called \emph{mode-correctness} is satisfied.
+%every synthesised type in a typing rule is determined by previously synthesised types from its premises and its input if checking.
+As \citet{Pierce2000} noted, bidirectional type synthesis propagates type information locally within adjacent nodes of a raw term, and does not require unification as Damas--Milner type inference does---in fact, introducing long-distance unification constraints would undermine the essence of locality in bidirectional type synthesis.
 
-Contrary to Damas--Milner type inference, unification is not required for bidirectional type synthesis.
-Moreover, \citet{Pierce2000} noted that bidirectional type synthesis propagates annotations locally within adjacent nodes of a syntax tree, so introducing long-distance unification constraints undermines the essence of locality in bidirectional type synthesis.
-Also, annotations for, say, top-level definitions improve clarity, making the purpose of a program easier to understand, so they should be considered a practical compromise at worst.
-In light of these considerations, bidirectional type synthesis can be deemed as a type-checking technique that is more fundamental than unification, as it is capable of handling a broad spectrum of programming languages.
+%Also, annotations for, say, top-level definitions make the purpose of a program easier to understand, so they are sometimes beneficial and not necessarily a nuisance.
+%In light of these considerations, bidirectional type synthesis can be deemed as a type-checking technique that is more fundamental than unification, as it is capable of handling a broad spectrum of programming languages.%
+%\todo{Reorganise into two paragraphs, one on old approaches and the other exclusively on bidirectional type synthesis?}
 
-Unlike parsing, nevertheless, which has a comprehensive theory and widely applicable practical techniques, bidirectional typing has only been developed on a case-by-case basis without a theory.
-While it is straightforward to derive a type synthesis algorithm, a general theory providing logical specifications and rigorously proven properties for a class of systems is still lacking.
-As a result, we can only present design principles that we learned from individual systems loosely.
-Moreover, unlike the plethora of available parser generators, `type-checker generators' rarely exist, so each type checker, based on bidirectional typing or not, has to be independently built from scratch.
+Although the basic idea is well known, bidirectional typing has only been developed on a case-by-case basis without a theory; this situation is in stark contrast to parsing, which has a comprehensive theory together with widely applicable techniques and practical tools.%
+\footnote{The same could even be said for type checking in general, not just for bidirectional type systems.
+While there are some efforts to generate type checkers grounded in unification~\citep{Gast2004,Grewe2015}, it should be noted that unification-based approaches are not suited to more complex type theories.
+Moreover, their algorithms are not proved complete.}
+%While it is straightforward to derive a type synthesis algorithm,
+For bidirectional typing, \citet{Dunfield2021} could only present informal design principles that the community learned from individual systems, rather than a general theory providing logical specifications and rigorously proven properties for a class of systems.
+Moreover, unlike the plethora of available parser generators, `type-synthesiser generators' rarely exist, so each type synthesiser has to be independently built from scratch.
+To remedy the situation, we present a first theory of bidirectional typing, which amounts to a verified type-synthesiser generator for syntax-directed bidirectional simple type systems.
 
-To tackle these challenges, this paper presents a theory of bidirectional typing and a verified implementation of bidirectional type synthesis for simple type theories.
-There are many ways and extensions for designing bidirectional typing rules, such as those related to polymorphic types~\citep{Pierce2000,Peyton-Jones2007,Dunfield2013,Xie2018}.
-Yet, once the definition of a bidirectional type system is in place, deriving an algorithm becomes straightforward.
-Accordingly, our theory's objective is not to formulate rules for various language features, but to introduce a formalism for bidirectional typing, analogous to grammar in parsing.
-We introduce the notion of \emph{bidirectional binding signatures} to specify bidirectional type systems and characterise essential criteria including \emph{soundness}, \emph{completeness}, and \emph{mode-correctness} that are sufficient to derive a type synthesis algorithm, as informally outlined by~\citet{Dunfield2021}.
-For simplicity, we confine our discussion to bidirectional simple type systems that have \emph{syntax-directed} typing rules and leave the problem of presenting bidirectional typing for more general and advanced type theories as future work.
+\subsection{Related Work}\label{sec:related-work}
 
-\subsection{Related work}\label{sec:related-work}
-As our theory is developed formally not only for simple type theories but also in a manner distinct from typical formulations in bidirectional typing, it is pertinent to discuss related work upon which this work is built, apart from bidirectional typing, in order to explain our contributions.
+To explain our contributions, first we discuss related work upon which our work is built.
 
-\subsubsection{Language formalisation and its frameworks} \label{sec:language-formalisation}
-The vision of formalising the metatheory of every programming language was initiated by the \PoplMark challenge~\citep{Aydemir2005}.
-This has been exemplified in the textbook by~\citet{Wadler2022} where language concepts are formally addressed using the proof assistant~\Agda, including bidirectional type synthesis for~\PCF.
+%As our theory is developed formally not only for simple type theories%
+%\todo{`and for other theories too?'}
+%but also in a manner distinct from typical formulations in bidirectional typing, it is pertinent to discuss related work upon which this work is built, apart from bidirectional typing, in order to explain our contributions.%
+%\todo{a general description like `the landscape of language formalisation'}
 
-\begin{remark}\label{re:type-synthesis-as-decidability-proof}
-The bidirectional type synthesis formulated by~\citeauthor{Wadler2022} differs from existing literature:
-\begin{enumerate*}
-  \item Its algorithm is presented as a proof of \emph{logical decidability} as to whether a `bidirectionally decorated' raw term~$\isTerm{t}$ can be bidirectionally typed, equivalent to a program that returns a typing derivation for $\isTerm{t}$ or otherwise the proof that such derivation is impossible.
-  \item In contrast, an algorithm typically found in the literature is presented as \emph{algorithmic system} relations such as $\Gamma |- \isTerm{t} \syn A \mapsto \isTerm{t}'$, denoting that annotations can be added to $\isTerm{t}$ in the external language to produce $\isTerm{t}'$ of type $A$ in the internal language.
+\subsubsection{Bidirectional type synthesis in Martin-L\"of Type Theory}
+\label{sec:PLFA}
+
+Our work is partly inspired by \varcitet{Wadler2022}{'s} textbook, where programming language concepts are formally introduced using the proof assistant~\Agda based on Martin-L\"of Type Theory, and one particular topic is bidirectional type synthesis for~\PCF.
+Their treatment deviates from existing literature:
+Traditionally, a type synthesis algorithm is presented as \emph{algorithmic system} relations such as $\Gamma |- \isTerm{t} \syn A \mapsto \isTerm{t}'$, denoting that annotations can be added to $\isTerm{t}$ in the surface language to produce $\isTerm{t}'$ of type $A$ in the core language.
 Such an algorithm is then accompanied with \emph{soundness} and \emph{completeness} assertions such that the algorithm correctly synthesises the type of a raw term and every typable term can be synthesised if there are enough annotations.
-\end{enumerate*}
-\end{remark}
-
-Earlier than the \PoplMark challenge, \citet{Altenkirch1993} commented that rudimentary meta-operations and their meta-properties constitute the bulk of formalisation, motivating a number of frameworks~\citep{Ahrens2018,Fiore2022,Gheri2020,Ahrens2022,Allais2021} to define well-scoped/typed terms, substitution, term traversal and their meta-properties universally.
-One of the core gadgets of these frameworks is the notion of \emph{binding signatures} (coined by~\citet{Aczel1978} in line with the term \emph{signature} in universal algebra) for specifying typing rules
-It is noteworthy that these state-of-the-art frameworks cannot specify polymorphic types or dependent types yet and only \citeauthor{Allais2021} discussed meta-operations beyond substitution.
-
-\subsubsection{Theories of abstract syntax with variable binding}\label{sec:theory-of-syntax}
-The aforementioned frameworks except \citeauthor{Gheri2020}'s are at least inspired by \citet{Fiore1999}'s initial semantics for abstract syntax with variable binding using category theory.
-The main idea is that the set of (untyped) abstract syntax trees for a language consists of
-\begin{enumerate*}
-  \item a family of sets $\Term_{\Gamma}$ of well-scoped terms under a context~$\Gamma$ with
-  \item variable renaming for a function $\sigma\colon \Gamma \to \Delta$ between variables acting as a functorial map from $X_{\Gamma}$ to $X_{\Delta}$, i.e.\ a presheaf $\Term\colon \mathbb{F} \to \mathsf{Set}$, and
-  \item an initial algebra $[\mathsf{v}, \mathsf{op}]$ on~$\Term$ given by the variable rule as a map $\mathsf{v}$ from the presheaf~$V$ of variables (i.e.\ the embedding $V\colon \mathbb{F} \hookrightarrow \mathsf{Set}$) to $\Term$ and other constructs as $\mathsf{op}\colon \mathbb{\Sigma}\Term \to \Term$ where $\mathbb{F}$ is the category of contexts, the functor $\mathbb{\Sigma}\colon \mathsf{Set}^\mathbb{F} \to \mathsf{Set}^\mathbb{F}$ encodes the arities of constructs, and the initiality amounts to structural recursion, i.e.\ \emph{term traversal}.
-\end{enumerate*}
-To put it succinctly, it is the free $\mathbb{\Sigma}$-algebra over the presheaf $V$ of variables.
-
-Fortunately, constructing the initial algebra of terms in type theory boils down to defining an inductive type with a few constructors that align with the variable rule and a rule scheme for language constructs specified by a signature~\citep{Fiore2022}.
-
-Substitution is also modelled categorically, but it does not play a role in this paper, though.
-
-\begin{remark} \label{re:type-signature}
-Most of existing theories treat types independently of terms, thereby excluding them from signatures.
-To the best of our knowledge, the only exception to this approach is found in the work of~\citet{Arkor2020}, which incorporates signatures for both terms and types.
-Interestingly, this inclusion is also critical for type synthesis for comparing a concrete type $N \bto N$ with an abstract type $A \bto B$, where $A$ and $B$ are type variables in a typing rule.
-\end{remark}
-
-\subsubsection{Type Checker Generation}
-While there are some efforts to generate type checkers grounded in unification~\citep{Gast2004,Grewe2015}, it should be noted that unification-based approaches are not suited to more complex type theories.
-Moreover, their algorithms are not proved complete.
-
-\subsection{Contributions and Plan of this paper}
-The major contributions of paper, extending upon previous discussions, are explained as follows.
-
-Our theory can be viewed as a counterpart to those theories in \Cref{sec:theory-of-syntax} and is tailored towards \emph{extrinsically-typed}, \emph{syntax-directed} bidirectional simple type systems (\Cref{fig:raw-terms,fig:bidirectional-typing-derivations}) specified by a \emph{signature} for simple types and a \emph{bidirectional binding signature} for terms (\Cref{def:simple-signature,def:bidirectional-binding-signature}), c.f.\ \Cref{re:type-signature}.
-The extrinsic typing is needed for type synthesis and our adherence to syntax-directed systems aligns with the standard idea of re-casting non-syntax-directed typing rules into their syntax-directed form to derive type synthesis algorithm~\citep{Peyton-Jones2007}.
-As substitution is not needed for synthesising simple types, it is not addressed.
-Instead, our focus is on:
-\begin{enumerate*}
-  \item establishing the connection between the specified bidirectional type system and its simple type theory, i.e.\ \emph{soundness} and \emph{completeness} (\Cref{sec:soundness-and-completeness}); 
-  \item defining the \emph{mode-correctness} condition (\Cref{def:mode-correctness}) which suffices to establish the uniqueness of synthesised types (\Cref{thm:unique-syn}) and a decidable bidirectional type synthesis and checking (\Cref{thm:bidirectional-type-synthesis-checking}).
-\end{enumerate*}
-
-Our theory is based on Martin-L\"of type theory and formalised in \Agda (\Cref{sec:formalisation}), akin to those frameworks in \Cref{sec:language-formalisation}.
-More importantly, we exploit the computational and logical aspects of type theory to formulate algorithmic soundness, completeness, and decidability in one go. 
-Recall that the law of excluded middle does not hold universally as an axiom in Martin-L\"of type theory, so a \emph{decidable statement} $P \vee \neg P$ is non-trivial.
+By contrast, \citeauthor{Wadler2022} exploit the simultaneously computational and logical nature of Martin-L\"of type theory to formulate algorithmic soundness, completeness, and decidability in one go.
+%(although they do not emphasise the difference of their approach from the traditional one)
+Recall that the law of excluded middle $P \vee \neg P$ does not hold universally as an axiom in type theory---it means that whether $P$~holds is \emph{decidable}, and is non-trivial to prove.
+For example, a proof of $\forall x, y.~x = y \mathrel\vee x \neq y$ is a way to compare arbitrary $x$~and~$y$ and decide whether they are equal, which may or may not be possible depending on the domain of $x$~and~$y$.
 Given that all proofs as programs terminate, logical decidability implies \emph{algorithmic decidability}.
@@
 \begin{quote}
-  `For a context $\Gamma$ and a raw term $t$, either a typing derivation of\, $\Gamma |- t : A$ exists for some $A$ or any derivation of\, $\Gamma |- t : A$ for some $A$ leads to a contradiction'
+  `For a context $\Gamma$ and a raw term $t$, either a typing derivation of\, $\Gamma |- t : A$ exists for some type~$A$ or any derivation of\, $\Gamma |- t : A$ for some type~$A$ leads to a contradiction'
 \end{quote}
@@
 \begin{quote}
-  `A context $\Gamma$ and a raw term $t$ \underline{\emph{decide}} whether $\Gamma |- t : A$ has a derivation for some $A$',
+  `It is \underline{\emph{decidable}} for any $\Gamma$~and~$t$ whether $\Gamma |- t : A$ has a derivation for some~$A$'.
 \end{quote}
-echoing \citeauthor{Wadler2022}'s formulation as noted in \Cref{re:type-synthesis-as-decidability-proof}.
-Then, both algorithmic soundness and completeness are implied:
-the proof either yields a typing derivation for the given raw term~$t$ or a proof that such a derivation is impossible where the former case is algorithmic soundness and the latter is algorithmic completeness in contrapositive form.
-That is, our decidable bidirectional type synthesis (\Cref{thm:bidirectional-type-synthesis-checking}) provides algorithmic soundness, completeness, and decidability.
+%echoing \citeauthor{Wadler2022}'s formulation as noted in \cref{re:type-synthesis-as-decidability-proof}.
+Then, both algorithmic soundness and completeness are implied.
+%the proof either yields a typing derivation for the given raw term~$t$, or a proof that such a derivation is impossible; the former case is algorithmic soundness, and the latter is algorithmic completeness in contrapositive form.
+%\todo{traditional completeness is still derivable from this `LEM'}
+Our work takes the same approach but expands the scope significantly.
+%That is, our decidable bidirectional type synthesis (\cref{thm:bidirectional-type-synthesis-checking}) provides algorithmic soundness, completeness, and decidability.
 
-In contrast to \citeauthor{Wadler2022}, our theory starts with raw terms and a general type synthesis problem, independent of bidirectional typing.
-This starting point uncovers nuances regarding the case of not having enough annotations, often overlooked in literature, and thus distinguishes completeness (with respect to a type assignment system) from annotatability, concepts conflated by \citet{Dunfield2021}.
-To clarify the difference (\Cref{sec:annotatability}), we propose \emph{mode derivations} (\Cref{fig:mode-derivations,fig:generalised-mode-derivations}) and \emph{mode preprocessing} (\Cref{sec:mode-preprocessing}) assigning a mode first to a raw term with enough annotations or otherwise pinpoints missing annotations.
-By combining bidirectional type synthesis, soundness, completeness, and mode preprocessing, we achieve our main result (\Cref{cor:trichotomy}) --- a \emph{trichotomy} of raw terms: a type synthesiser which checks if a given raw term is \emph{suitably annotated} and if a suitably annotated raw term has a typing derivation with respect to the specified simple type theory.
-Our bidirectional type synthesis can be instantiated for any system specified by a mode-correct signature, effectively a verified type-checker generator.
+%\begin{remark}\label{re:type-synthesis-as-decidability-proof}
+%\begin{enumerate*}
+%  \item Its algorithm is presented as a proof of \emph{logical decidability} as to whether a `bidirectionally decorated' raw term~$\isTerm{t}$ can be bidirectionally typed, equivalent to a program that returns a typing derivation for $\isTerm{t}$ or otherwise the proof that such derivation is impossible.
+%  \item In contrast, 
+%\end{enumerate*}
+%\end{remark}
 
-In summary, we contribute a theory of bidirectional type synthesis that is:
+%Our theory is based on Martin-L\"of type theory and formalised in \Agda (\cref{sec:formalisation}), akin to those frameworks in \cref{sec:language-formalisation}.
+
+\subsubsection{Language formalisation frameworks}
+\label{sec:language-formalisation}
+
+The vision of formalising the metatheory of every programming language was initiated by the \PoplMark challenge~\citep{Aydemir2005}.
+Earlier than \PoplMark, \citet{Altenkirch1993} commented that basic meta-operations and their meta-properties constitute the bulk of formalisation, motivating a number of frameworks~\citep{Ahrens2018,Fiore2022,Gheri2020,Ahrens2022,Allais2021} to define well-scoped/typed terms, substitution, term traversal, and their meta-properties universally.
+One of the core gadgets of these frameworks is the notion of \emph{binding signatures} (coined by~\citet{Aczel1978} in line with the term \emph{signature} in universal algebra) for specifying type systems.
+It is noteworthy that these state-of-the-art frameworks are basically restricted to simple types%
+\todo{confirm}
+and cannot specify polymorphic types or dependent types yet, and only \citeauthor{Allais2021} discussed meta-operations beyond substitution.
+Like these frameworks, our work is  built on a variant of binding signature for simple types.
+Unlike these frameworks, we target constructions for a different part of programming language theory, namely the transition from a surface language (raw terms) to a core language (well-typed terms), and our type synthesiser is significantly more complex than those meta-operations previously considered.
+Another difference is that several of the frameworks%
+\todo{Which ones precisely?}
+adopt intrinsically typed terms, whereas for our theory it is essential to use an extrinsic formulation to make it possible to relate a synthesised typing derivation with an input term.
+This difference can be easily reconciled by converting extrinsic typing derivations to intrinsically typed terms though.
+
+\subsection{Contributions and Plan of This Paper}
+
+%The major contributions of our paper, extending upon previous discussions, are explained as follows.
+
+Instead of focusing solely on a bidirectional type synthesiser like \citet{Wadler2022}~(\cref{sec:PLFA}), our theory centres around a general specification of type synthesis for an ordinary type system where judgements are not assigned modes, and fills in the gap between language formalisation frameworks based on such type systems~(\cref{sec:language-formalisation}) and raw terms produced by parsers.
+Working similarly to those frameworks, our theory is parametrised by a \emph{bidirectional binding signature}, which describes a language with a bidirectional type system, analogously to a grammar describing a formal language in the theory of parsing.
+As a first step, we confine our theory to \emph{syntax-directed} systems, meaning that for every raw term construct, the ordinary type system has exactly one typing rule, which is assigned modes in exactly one way in the bidirectional type system.%
+\footnote{It is reasonable to assume syntax-directedness (especially as a first step) given that it is a standard idea to re-cast non-syntax-directed typing rules into their syntax-directed form to derive a type synthesiser~\citep{Peyton-Jones2007}.}
+With this assumption, using a single bidirectional binding signature we can specify the raw terms, the ordinary type system, and the bidirectional type system of a language.
+Then, if the signature is \emph{mode-correct}~\citep[Section~3.1]{Dunfield2021}, we can instantiate a type synthesiser that is formulated similarly to \varcitet{Wadler2022}{'s} with respect to the ordinary type system, and performs bidirectional type synthesis internally.
+Our theory thus works as a verified type-synthesiser generator, analogous to a parser generator.
+
+Within the theory, we formulate essential criteria including \emph{soundness}, \emph{completeness}, and \emph{mode-correctness} that are sufficient to derive a bidirectional type synthesiser and ensure that it implements the general specification, formalising what were only informally outlined by~\citet{Dunfield2021}.
+We also pay attention to the problem of whether a raw term has enough type annotations, often overlooked in the literature, by introducing \emph{mode preprocessing}, which assigns a mode to a raw term with enough annotations or otherwise pinpoints missing annotations.
+Practically, by running a mode preprocessor, the user can first make sure that a raw term has enough annotations, and then deal with typing issues without worrying about missing annotations anymore.
+Theoretically, separating the annotation and typing aspects streamlines our theory, in particular allowing us to formulate completeness naturally and differently from \emph{annotatability}, which was proposed and conflated with completeness by \citet{Dunfield2021}.
+By combining bidirectional type synthesis, soundness, completeness, and mode preprocessing, we achieve our main result~(\cref{cor:trichotomy}): a \emph{trichotomy} of raw terms, which is computationally a type synthesiser that checks if a given raw term is \emph{suitably annotated} and if a suitably annotated raw term has an ordinary typing derivation.
+
+In summary, we contribute a theory of bidirectional type synthesis that is
 \begin{enumerate}
-  \item \emph{simple yet general} for any system that can be specified by a bidirectional binding signature;
-  \item \emph{constructive}, based on Martin-L\"of theory and preferring logical decidability over algorithmic soundness, completeness, and decidability; mode preprocessing over annotatability.
+  \item \emph{simple yet general}, working for any syntax-directed bidirectional simple type system that can be specified by a bidirectional binding signature;
+  \item \emph{constructive} based on Martin-L\"of type theory, and compactly formulated in a way that unifies computation and proof, preferring logical decidability over algorithmic soundness, completeness, and decidability; and mode preprocessing over annotatability.
 \end{enumerate}
@@
 This paper is structured in logical order as follows.
-We first present an overview of our theory using simply typed $\lambda$-calculus in \Cref{sec:key-ideas}, prior to developing a general framework for specifying bidirectional type systems in \Cref{sec:defs}.
-Following this, we discuss the connection between a specified bidirectional type system and its associated simple type theory in \Cref{sec:pre-synthesis}.
-The core technical contribution lies in \Cref{sec:type-synthesis} where we introduce mode-correctness and bidirectional type synthesis.
-In \Cref{sec:formalisation}, we sketch the formalisation of our theory in \Agda and give some examples other than simply typed $\lambda$-calculus.
-We conclude in \Cref{sec:future}, where we reflect on related techniques that can be used to improve the formal development and potential challenges in extending our theory to a more general setting.
+We first present a concrete overview of our theory using simply typed $\lambda$-calculus in \cref{sec:key-ideas}, prior to developing a general framework for specifying bidirectional type systems in \cref{sec:defs}.
+Following this, we discuss mode preprocessing and related properties in \cref{sec:pre-synthesis}.
+The most heavyweight technical contribution lies in \cref{sec:type-synthesis}, where we introduce mode-correctness and bidirectional type synthesis.
+In \cref{sec:formalisation}, we sketch the formalisation of our theory in \Agda and give some examples other than simply typed $\lambda$-calculus.
+We conclude in \cref{sec:future}, where we reflect on related techniques that can be used to improve the formal development and potential challenges in extending our theory to a more general setting.
 
-\todo[inline,caption={}]{
-\begin{enumerate}
-  \item Introduction (\Cref{sec:intro}) 
-    \LT{4 pp}
-  \item Key ideas (\Cref{sec:key-ideas})
-    \Josh{2.5 pp}
-  \item Definitions for bidirectional type systems (bidirectional binding signature, bidirectional type systems, signature erasure and mode annotation) (\Cref{sec:defs})
-    \LT{4 pp}
-  \item Soundness, completeness, bidirectionalisation, and annotatability (\Cref{sec:pre-synthesis})
-    \Josh{Erasure from bidirectional typing derivation to raw terms with a mode}
-    \Josh{3 pp}
-  \item Bidirectional type inference (\Cref{sec:type-synthesis})
-    \LT{4 pp}
-  \item Formalisation and further examples (\Cref{sec:formalisation})
-    \LT{5.5pp}
-  \item Concluding remarks (\Cref{sec:future})
-    \LT{1 p} \Josh{1p}
-\end{enumerate}
-}
+%\todo[inline,caption={}]{\begin{enumerate}
+%\item `Input' and `output' of our theory, comparing with grammars and parsing
+%\item A more comprehensive theory than Wadler et al., and more formal and refined than Dunfield et al.'
+%\item Restrictions to simple types (not individual language features) and syntax-directedness
+%\end{enumerate}}
+
+%which together describe an \emph{extrinsically typed} and \emph{syntax-directed} bidirectional simple type system~(\cref{fig:raw-terms,fig:bidirectional-typing-derivations}).
+%
+%There are many ways and extensions for designing bidirectional typing rules, .
+%Yet, once the definition of a bidirectional type system is in place, deriving an algorithm becomes straightforward.
+%Accordingly, our theory's objective is not to formulate rules for various language features, but to introduce a formalism for bidirectional typing.
+%\todo{Looks like we should have axiomatised what types we can deal with; left as future work}
+
+%The extrinsic typing is needed for type synthesis,%
+%\todo{Vague; delete or mention it's easy to go intrinsic?}
+%%As substitution is not needed for synthesising simple types, it is not addressed.
+%Our focus is on:
+%\begin{enumerate*}
+%  \item establishing the connection between the specified bidirectional type system and its simple type theory, i.e.\ \emph{soundness} and \emph{completeness} (\cref{sec:soundness-and-completeness});
+%  \item defining the \emph{mode-correctness} condition (\cref{def:mode-correctness}), which suffices to establish the uniqueness of synthesised types (\cref{thm:unique-syn}) and decidable bidirectional type synthesis and checking (\cref{thm:bidirectional-type-synthesis-checking}).
+%\end{enumerate*}
+
+%\todo[inline,caption={}]{
+%\begin{enumerate}
+%  \item Introduction (\cref{sec:intro}) 
+%    \LT{4 pp}
+%  \item Key ideas (\cref{sec:key-ideas})
+%    \Josh{2.5 pp}
+%  \item Definitions for bidirectional type systems (bidirectional binding signature, bidirectional type systems, signature erasure and mode annotation) (\cref{sec:defs})
+%    \LT{4 pp}
+%  \item Soundness, completeness, bidirectionalisation, and annotatability (\cref{sec:pre-synthesis})
+%    \Josh{Erasure from bidirectional typing derivation to raw terms with a mode}
+%    \Josh{3 pp}
+%  \item Bidirectional type inference (\cref{sec:type-synthesis})
+%    \LT{4 pp}
+%  \item Formalisation and further examples (\cref{sec:formalisation})
+%    \LT{5.5pp}
+%  \item Concluding remarks (\cref{sec:future})
+%    \LT{1 p} \Josh{1p}
+%\end{enumerate}
+%}

### 2023-07-05 tex/sec2-outline.tex
[ tex ] Revised abstract & S1

@@
 We start with an overview of our theory by instantiating it to simply typed $\lambda$-calculus.
-Roughly speaking, the problem of type synthesis requires us to take an untyped abstract syntax tree---which we call a \emph{raw term}---as input, and produce a typing derivation for the term if possible.
+Roughly speaking, the problem of type synthesis requires us to take a raw term (i.e.~an untyped abstract syntax tree) as input, and produce a typing derivation for the term if possible.
 To give more precise definitions:
@@
 Now our goal is to use \cref{def:STLC-type-synthesiser} as a specification and implement it using a \emph{bidirectional} type synthesiser, which attempts to produce \emph{bidirectional} typing derivations defined in \cref{fig:STLC-bidirectional-typing-derivations}.
-Here we briefly recap the basic ideas of bidirectional type synthesis.
-While the problem of type synthesis is not decidable in general, for certain kinds of terms it is still possible to synthesise their types --- for example, the $\SynRule{Var}$ rule says that the type of a variable can be looked up in the context.
-For other kinds of terms, we can switch to the simpler problem of type \emph{checking}, where the expected type of a term is given as input so that there is more information to work with.
-When a bidirectional type synthesiser traverses a raw term, it switches between the two \emph{modes}, synthesising wherever possible, and checking otherwise.
-If a bidirectional type system is designed well (i.e.~\emph{mode-correct}, which will be defined in \cref{sec:type-synthesis}), then there will be a flow of type information in each rule that allows us to determine unknown types (e.g.~types to be checked) from known ones (e.g.~types previously synthesised), and suggests how bidirectional type synthesis is performed --- for example, the $\SynRule{App}$ rule says that to synthesise the type of an application $\isTerm{t\;u}$, we first synthesise the type of~$\isTerm{t}$, which should have the form $A \bto B$, from which we can extract the expected type of~$\isTerm{u}$, namely~$A$, and perform checking; then the type of the whole application, namely~$B$, can also be extracted from the previously synthesised type $A \bto B$.
+
+\Josh{Here we briefly recap the basic ideas of bidirectional type synthesis.
+If a bidirectional type system is designed well (i.e.~\emph{mode-correct}, which will be defined in \cref{sec:type-synthesis}), then there will be a flow of type information in each rule that allows us to determine unknown types (e.g.~types to be checked) from known ones (e.g.~types previously synthesised), and suggests how bidirectional type synthesis is performed --- for example, the $\SynRule{App}$ rule says that to synthesise the type of an application $\isTerm{t\;u}$, we first synthesise the type of~$\isTerm{t}$, which should have the form $A \bto B$, from which we can extract the expected type of~$\isTerm{u}$, namely~$A$, and perform checking; then the type of the whole application, namely~$B$, can also be extracted from the previously synthesised type $A \bto B$.}
 
@@
 
-One (smaller) benefit of mode preprocessing is that it helps to simplify the synthesiser, whose computation can be partly directed by a mode derivation.
+One (less important) benefit of mode preprocessing is that it helps to simplify the synthesiser, whose computation can be partly directed by a mode derivation.
 More importantly, whether there is a mode derivation for a term is actually very useful information to the user, because it corresponds to whether the term has enough type annotations:
@@
 Otherwise $|\Gamma| \vdash t^{\syn}$, and we can run the bidirectional type synthesiser.
-If it finds a derivation of $\Gamma \vdash t \syn A$ for some type~$A$, return a derivation of $\Gamma \vdash t : A$ (the first outcome), which is possible because the bidirectional typing~(\cref{fig:STLC-bidirectional-typing-derivations}) is \emph{sound} with respect to the original typing~(\cref{fig:STLC-typing-derivations}).
+If it finds a derivation of $\Gamma \vdash t \syn A$ for some type~$A$, return a derivation of $\Gamma \vdash t : A$ (the first outcome), which is possible because the bidirectional typing~(\cref{fig:STLC-bidirectional-typing-derivations}) is \emph{sound}%
+\todo{Explain the notation `$:^\dir{d}$'}
+with respect to the original typing~(\cref{fig:STLC-typing-derivations}).
 

### 2023-07-05 tex/sec4-pre-synthesis.tex
[ tex ] Compile Sec6 separately with lhs2TeX
[ tex ] Revised abstract & S1

@@
 
-\section{Soundness, Completeness, and Mode Preprocessing}\label{sec:pre-synthesis}
+\section{Mode Preprocessing and Related Properties}\label{sec:pre-synthesis}
 
@@
 %\Josh{Throughout this sub-section, we quantify over any bidirectional type system~$(\Sigma, \Omega)$, context $\Gamma : \Cxt_\Sigma(\emptyset)$, raw term $\erase\Gamma \vdash_{\Sigma, \erase\Omega} t$, mode~$d$, and type $A : \Type_\Sigma(\emptyset)$.}
+
+\Josh{One theorem to rule them all}
 
@@
 
-\subsection{Mode Preprocessing}
+\subsection{Generalised Mode Preprocessing}
 \label{sec:mode-preprocessing}
@@
 
-\newcommand{\True}{\mathbf{T}}
-\newcommand{\False}{\mathbf{F}}
+\renewcommand{\True}{\mathbf{T}}
+\renewcommand{\False}{\mathbf{F}}
 
@@
 
-Intuitively, a term does not have a mode derivation exactly when there are not enough type annotations.
-To formulate this more concretely, we can point out the places in the term that require annotations.
+Intuitively, a term does not have a mode derivation exactly when there are not enough type annotations, but such negative formulations convey little information.%
+\todo{still default in the rest of the paper}
+Instead, we can provide more information by pointing out the places in the term that require annotations.
 For a bidirectional type system, an annotation is required wherever a term is `strictly' (which we will explain shortly) in checking mode but required to be in synthesising mode, in which case there is no rule for switching from checking to synthesising, and thus there is no way to construct a mode derivation.
@@
 
-\begin{theorem}\label{thm:generalised-mode-preprocessing}
+\begin{theorem}[Generalised Mode Preprocessing]\label{thm:generalised-mode-preprocessing}
 For any raw term $V |-_{\Sigma,\erase\Omega} t$ and mode~$\dir{d}$, there is a derivation of\/ $V |-_{\Sigma,\Omega} t^{\dir{d}\,g\,s}$ for some booleans $g$~and~$s$.

### 2023-07-05 tex/sec1-intro.tex
[ tex ] Minor correction

@@
 Theoretically, separating the annotation and typing aspects streamlines our theory, in particular allowing us to formulate completeness naturally and differently from \emph{annotatability}, which was proposed and conflated with completeness by \citet{Dunfield2021}.
-By combining bidirectional type synthesis, soundness, completeness, and mode preprocessing, we achieve our main result~(\cref{cor:trichotomy}): a \emph{trichotomy} of raw terms, which is computationally a type synthesiser that checks if a given raw term is \emph{suitably annotated} and if a suitably annotated raw term has an ordinary typing derivation.
+By combining bidirectional type synthesis, soundness, completeness, and mode preprocessing, we achieve our main result~(\cref{cor:trichotomy}): a \emph{trichotomy} on raw terms, which is computationally a type synthesiser that checks if a given raw term is \emph{suitably annotated} and if a suitably annotated raw term has an ordinary typing derivation.
 

### 2023-07-06 tex/BiSig.tex
[ tex ] Rename `mylhs2tex` to `agda-lhs2tex`
[ tex ] Fix the dangling end in abstract

@@
 \usepackage{microtype}
+\microtypesetup{babel}
 \usepackage{subfiles}
@@
 \input{macros}
-\usepackage{mylhs2tex}
+\usepackage{agda-lhs2tex}
 
@@
 %  which refers to the mutual use of type checking and synthesis based on bidirectional typing rules to determine the type of a program with type annotations.
-  First, our theory is simple yet general as it applies to any syntax-directed bidirectional simple type system that can be specified by a signature.
+  First, it is simple yet general as it applies to any syntax-directed bidirectional simple type system that can be specified by a signature.
 %  which is analogous to a grammar in parsing.
-  Second, our theory is constructive because it is not only formalised in a proof assistant based on Martin-L\"of type theory, but also formulated in a way that unifies computation and proof, resulting in a compact and practical theory---for example, a proof of the decidability of the type synthesis problem is also computationally a type synthesiser.
+  Second, it is constructive because it is not only formalised in a proof assistant based on Martin-L\"of type theory, but also formulated in a way that unifies computation and proof, resulting in a compact and practical theory---for example, a proof of the decidability of the type synthesis problem is computationally a type synthesiser.
   Moreover, we introduce `mode preprocessing' to deal with missing type annotations, an aspect often overlooked in the literature.
-  The main theorem is logically a trichotomy on parsed syntax trees and computationally a type synthesiser that may throw an exception due to missing type annotations.
-  Being parametrised by a signature, our theory works as a verified type-synthesiser generator overall.
+  The main theorem is logically a trichotomy on parsed syntax trees and computationally a synthesiser that may throw an exception due to missing type annotations.
+  By being parametrised by a signature, our theory works as a verified type-synthesiser generator overall.
 \end{abstract}
@@
 
-\keywords{bidirectional typing, type synthesis, type checking, type inference, language formalisation, \Agda}
+\keywords{type checking, type inference, language formalisation, \Agda}
 

### 2023-07-06 tex/sec1-intro.tex
[ tex ] Include the term `decidable equality`

@@
 %(although they do not emphasise the difference of their approach from the traditional one)
-Recall that the law of excluded middle $P \vee \neg P$ does not hold universally as an axiom in type theory---it means that whether $P$~holds is \emph{decidable}, and is non-trivial to prove.
-For example, a proof of $\forall x, y.~x = y \mathrel\vee x \neq y$ is a way to compare arbitrary $x$~and~$y$ and decide whether they are equal, which may or may not be possible depending on the domain of $x$~and~$y$.
+Recall that the law of excluded middle $P + \neg P$ does not hold universally as an axiom in type theory---it means that whether $P$~holds is \emph{decidable}, and is non-trivial to prove.
+For example, a proof of \emph{decidable equality}, i.e.
+$\forall x, y.~(x = y) + (x \neq y)$, is a way to compare arbitrary $x$~and~$y$ and decide whether they are equal, which may or may not be possible depending on the domain of $x$~and~$y$.
 Given that all proofs as programs terminate, logical decidability implies \emph{algorithmic decidability}.
@@
 \end{enumerate}
-All concepts were formally developed with \Agda and translated to the mathematical vernacular.
+Our theory was first developed with \Agda and then translated to the mathematical vernacular.
 

### 2023-07-06 tex/sec3-defs.tex
[ tex ] Use `\aritysymbol` in S3; explain the notation more carefully

@@
 
-\section{Simple Type Theories and Bidirectional Type Systems}\label{sec:defs}
+\section{Simple Type Systems and Bidirectional Type Systems}\label{sec:defs}
 This section provides generic definitions
@@
 %Here's the 'small print': Our definitions cover typical examples and set the scope for bidirectional simple type systems discussed, but do not offer comprehensive coverage of all possibilities.}
-of simple types, simple type theories, and bidirectional type systems using simply typed $\lambda$-calculus in \Cref{sec:key-ideas} as our running example.
+of simple types, simple type systems, and bidirectional type systems using simply typed $\lambda$-calculus in \Cref{sec:key-ideas} as our running example.
 
@@
 \begin{definition} \label{def:simple-signature}
-  \begin{figure}
-%    \begin{minipage}[b]{.6\textwidth}
-      \centering
-      \small
-      \judgbox{\Xi|-_{\Sigma} A}{$A$ is a well-formed type with type variables in $\Xi$}
-      \begin{mathpar}
-        \inferrule{\Xi \ni X_i}{\Xi |-_\Sigma X_i} \and
-        \inferrule{\Xi |-_{\Sigma} A_1 \\ \cdots \\ \Xi |-_{\Sigma} A_n}{\Xi |-_{\Sigma} \tyOp_i(A_1, \ldots, A_n)}\;\text{where $n = \arity(i)$}
-      \end{mathpar}
-      \caption{Simple types}
-      \label{fig:simple-type}
-%    \end{minipage}
-%    \begin{minipage}[b]{.35\textwidth}
+%  \begin{figure}
+%%    \begin{minipage}[b]{.6\textwidth}
 %      \centering
 %      \small
-%      \judgbox{\Xi |-_{\Sigma} \Gamma}{}
+%      \judgbox{\Xi|-_{\Sigma} A}{$A$ is a well-formed type with type variables in $\Xi$}
 %      \begin{mathpar}
-%        \inferrule{ }{\Xi |-_{\Sigma} \cdot }\and
-%        \inferrule{\Xi |-_{\Sigma} A \\ \Xi|-_{\Sigma} \Gamma}{\Xi |-_{\Sigma} \Gamma, x : A}
+%        \inferrule{\Xi \ni X_i}{\Xi |-_\Sigma X_i} \and
+%        \inferrule{\Xi |-_{\Sigma} A_1 \\ \cdots \\ \Xi |-_{\Sigma} A_n}{\Xi |-_{\Sigma} \tyOp_i(A_1, \ldots, A_n)}\;\text{where $n = \arity(i)$}
 %      \end{mathpar}
-%      \caption{Contexts}
-%    \label{fig:simple-context}
-%    \end{minipage}
-  \end{figure}
+%      \caption{Simple types}
+%      \label{fig:simple-type}
+%%    \end{minipage}
+%%    \begin{minipage}[b]{.35\textwidth}
+%%      \centering
+%%      \small
+%%      \judgbox{\Xi |-_{\Sigma} \Gamma}{}
+%%      \begin{mathpar}
+%%        \inferrule{ }{\Xi |-_{\Sigma} \cdot }\and
+%%        \inferrule{\Xi |-_{\Sigma} A \\ \Xi|-_{\Sigma} \Gamma}{\Xi |-_{\Sigma} \Gamma, x : A}
+%%      \end{mathpar}
+%%      \caption{Contexts}
+%%    \label{fig:simple-context}
+%%    \end{minipage}
+%  \end{figure}
   A \emph{signature} $\Sigma$ for simple types consists of a set\footnote{%
     Even though our theory is developed in Martin-L\"of type theory, the term `set' is used instead of `type' to avoid the obvious confusion. 
-    As we work with Martin-L\"of type theory with \AxiomK, all types are actually sets in the sense of homotopy type theory where all inhabitants of its identity type are proved equal.
+    As we work in Martin-L\"of type theory with \AxiomK, all types are legitimately sets in the sense of homotopy type theory where all inhabitants of its identity type are proved equal.
   }
@@
 \end{definition}
-Note that the second clause is a rule schema giving rise to as many rules as inhabitants $i$ in $I$.
 
@@
 
-Without the loss of generality, we may use the family of sets $\Fin(n)$ of natural numbers less than $n$ as sets of type variables in this paper, following the inductive definition by \citet{Dybjer1994}.
-In bidirectional type synthesis, it is necessary to have a decidable equality on $\Type_{\Sigma}(\Xi)$, as we need to compare the synthesized type $B$ with a given type $A$ for the $\Rule{Sub}$ rule.
-If $\Xi$ such as $\Fin(n)$ has a decidable equality , the decidable equality on $\Type_{\Sigma}(\Xi)$ can be given.
-In addition, the arity of a construct is always finite, so it suffices to use natural numbers to denote type variables in a $\Sigma$-type.
-Formally, we restrict the set $\Xi$ of type variables to the family of $\Fin(n)$, but for illustrative purposes we still use named variables for presentation.
-
 As \Cref{def:simple-signature} includes type variables, we introduce the typical meta-operations: substitution.
 \begin{definition}
-  The \emph{(simultaneous) type substitution} for a function $\rho\colon\Xi \to \Type_{\Sigma}(\Xi')$ is a map, which sends $A : \Type_{\Sigma}(\Xi)$ to $\simsub{A}{\rho} : \Type_{\Sigma}(\Xi')$, defined as usual.
+  The \emph{(simultaneous) type substitution} for a function $\rho\colon\Xi \to \Type_{\Sigma}(\Xi')$ is a map which sends $A : \Type_{\Sigma}(\Xi)$ to $\simsub{A}{\rho} : \Type_{\Sigma}(\Xi')$ and is defined as usual.
   By abuse of notation, the substitution $\simsub{\Gamma}{\rho}$ of a context $\Gamma$ is defined by applying substitution $\simsub{A}{\rho}$ to each type $A$ in the context.
@@
 
-\subsection{Binding Signatures and Simple Type Theories} \label{subsec:binding-sig}
+We may confine arbitrary variable set $\Xi$ to the family of sets $\Fin(n)$ of natural numbers less than $n$ as sets of type variables, following the inductive definition by \citet{Dybjer1994}, for the reasons below.
+In bidirectional type synthesis, it is necessary to have a decidable equality on $\Type_{\Sigma}(\Xi)$, as we need to compare the synthesized type $B$ with a given type $A$ for the $\Rule{Sub}$ rule.
+If $\Xi$ such as $\Fin(n)$ has a decidable equality, the decidable equality on $\Type_{\Sigma}(\Xi)$ can be given.
+In addition, the arity of a construct is always finite, so it suffices to use natural numbers to denote type variables in a type.
+Therefore, we formally restrict the set $\Xi$ of type variables to the family of $\Fin(n)$, but for illustrative purposes we still use named variables for presentation.
+\subsection{Binding Signatures and Simple Type Systems} \label{subsec:binding-sig}
 
-A simple type theory consists of
+A simple type system consists of
 \begin{enumerate*}
-  \item a set of raw terms $\isTerm{t}$ indexed by a set $V$ of untyped variables where each construct is allowed to bind some variables (as $\Rule{Abs}$) and to take multiple arguments (as $\Rule{App}$);
+  \item a set of raw terms $\isTerm{t}$ indexed by a set $V$ of untyped variables where each construct is allowed to bind some variables like $\Rule{Abs}$ and to take multiple arguments like $\Rule{App}$;
   \item a set of typing derivations indexed by a list $\Gamma$ of typed variables, a type~$A$, and a raw term $\isTerm{t}$ to set their type constraints. 
 \end{enumerate*}
-Therefore, to specify a construct like $\Rule{Abs}$ or $\Rule{App}$, we need to enrich the notion of signature with a type and an extension context for variable binding for each argument and a type for the operation:
+Therefore, to specify a construct like $\Rule{Abs}$ and $\Rule{App}$, we need to enrich the notion of signature with a type and an extension context for variable binding for each argument and a type for the operation:
 \begin{definition}\label{def:binding-signature}
@@
   \[
-    \mathit{ar}\colon O \to \sum_{\Xi : \N} \left(\Cxt_\Sigma(\Xi) \times \Type_\Sigma(\Xi)\right)^* \times \Type_\Sigma(\Xi).
+    \mathit{ar}\colon O \to \sm{\Xi : \N} \left(\Cxt_\Sigma(\Xi) \times \Type_\Sigma(\Xi)\right)^* \times \Type_\Sigma(\Xi).
   \]
 \end{definition}
-Each inhabitant $o: O$, meant to represent a term construct $\tmOp_o$ in a simple type theory, has a triple $\left(\Xi, \left[\left(\Delta_1; A_1\right), \ldots, \left(\Delta_{n}; A_{n}\right) \right], A_0\right)$ as its arity $\arity_o$ where
+Each inhabitant $o: O$, meant to represent a term construct $\tmOp_o$ in a simple type system, has a triple $\left(\Xi, \left[\left(\Delta_1; A_1\right), \ldots, \left(\Delta_{n}; A_{n}\right) \right], A_0\right)$ as its arity $\arity_o$ where
 \begin{enumerate}
-  \item $\Xi$ is the number of type variables, 
-  \item $A : \Type_\Sigma(\Xi)$ is for the type of $\tmOp_o$, and
-  \item $\left[\left(\Delta_1; A_{1}\right), \ldots, \left(\Delta_{n}; A_{n}\right) \right]$ is a list of pairs where
+  \item $\Xi$ is the number of type variables that are allowed to appear in a typing rule, 
+  \item $\left[\left(\Delta_1; A_{1}\right), \ldots, \left(\Delta_{n}; A_{n}\right) \right]$ is a list of pairs for specifying arguments where
     \begin{enumerate}
-      \item $\Delta_i : \Cxt_{\Sigma}(\Xi)$ is for the types of bound variables and
-      \item $A_i : \Term_{\Sigma}(\Xi)$ for the type of the $i$-th argument of $\tmOp_{o}$.
+      \item $\Delta_i : \Cxt_{\Sigma}(\Xi)$ is the \emph{extension context} consisting of types for bound variables and
+      \item $A_i : \Term_{\Sigma}(\Xi)$ is for the type of the $i$-th argument (i.e.\ premise) of $\tmOp_{o}$;
     \end{enumerate}
+  \item $A_0 : \Type_\Sigma(\Xi)$ is for the type of the construct $\tmOp_o$ in the conclusion.
 \end{enumerate}
-For brevity, we write $o \colon \Xi \rhd [\Delta_1]A_{1}, \ldots, \left[\Delta_{n}\right] A_{n} \to A_0$ to indicate an operation with its arity and omit $[\Delta]$ if $\Delta$ is empty. 
-For example, the $\Rule{Abs}$ rule can be specified by the operation 
+For brevity, we write $\aritysymbol{o}{\Xi}{[\Delta_1]A_{1}, \ldots, \left[\Delta_{n}\right] A_{n}}{A_0}$ to indicate an operation with its arity and omit $[\Delta]$ if $\Delta$ is empty. 
+For example, the $\Rule{Abs}$ rule in \Cref{fig:STLC-typing-derivations} can be specified by the operation 
 \[
-  \mathsf{abs}\colon A , B \rhd [A]B \to (A \bto B)
+  \aritysymbol{\mathsf{abs}}{(A, B)}{[A]B}{(A \bto B)}
 \]
-indicating that the $\Rule{Abs}$ rule has two type variables $A$ and $B$, one argument of type $B$ and which binds a variable of type $A$, and the type $A \bto B$ for the construct $\lambda$.
-Similarly, the $\Rule{App}$ rule can be specified by $\mathsf{app}\colon A, B \rhd A \bto B, A \to B$ where extension contexts for its two arguments are empty.
+indicating that in the $\Rule{Abs}$ rule for a raw term $\lam{x}t$
+\begin{enumerate*}
+  \item there are \emph{two} type variables named $A$ and $B$;
+  \item an argument (i.e.\ a typing derivation of $\Gamma, x : A |- t : B$) of type $B$ in a context extended by a variable $x$ of type $A$;
+  \item the type $A\bto B$ for the conclusion.
+\end{enumerate*}
+Similarly, the $\Rule{App}$ rule for $t\;u$ can be specified by $\aritysymbol{\mathsf{app}}{(A, B)}{(A \bto B), A}{B}$ which takes two arguments (i.e.\ typing derivations for $t$ and $u$) under the same context so that their extension contexts are both empty.
 
-Also for brevity, a simple type theory $(\Sigma, \Omega)$ refers to the simple type theory specified by some type signature $\Sigma$ and a binding signature $\Omega$.
+Also for brevity, a simple type system $(\Sigma, \Omega)$ refers to the simple type system specified by some type signature $\Sigma$ and a binding signature $\Omega$.
 
@@
     \inferrule{\cdot |-_{\Sigma} A \\ V |-_{\Sigma, \Omega}\isTerm{t}}{V |-_{\Sigma, \Omega} \isTerm{t \annotate A}}\,\Rule{Anno}
-    \and
-    \inferrule{V, \vars{x}_\isTerm{1} |-_{\Sigma, \Omega} \isTerm{t_1} \\ \cdots \\ V, \vars{x}_\isTerm{n} |-_{\Sigma, \Omega} \isTerm{t_n}}
-    {V |-_{\Sigma, \Omega} \tmOpts }\,\Rule{Op}
+    \\
+    \inferrule{V, \vars{x}_\isTerm{1} |-_{\Sigma, \Omega} \isTerm{t_1} \\ \cdots \\ V, \vars{x}_\isTerm{n} |-_{\Sigma, \Omega} \isTerm{t_n} } {V |-_{\Sigma, \Omega} \tmOpts }\,\Rule{Op} \and
+    \text{for $\aritysymbol{o}{\Xi}{[\Delta_1]A_{1}, \ldots, [\Delta_{n}] A_{n}}{A_0}$ in $\Omega$}
   \end{mathpar}
-%  \begin{flushright}
-%    for $o \colon \Xi \rhd [\Delta_1]A_{1}, \ldots, [\Delta_{n}] A_{n} \to A$ in $\Omega$
-%  \end{flushright}%
   \caption{Raw terms}
@@
 \end{figure}
-  For a simple type theory $(\Sigma, \Omega)$, the set of \emph{raw terms} indexed by a context~$V$ of free variables consists of
+  For a simple type system $(\Sigma, \Omega)$, the set of \emph{raw terms} indexed by a context~$V$ of free variables consists of
   \begin{enumerate}
@@
     \item annotations $\isTerm{t \annotate A}$ for some raw term $t$ in $V$ and a type $A$, and
-    \item a construct $\tmOpts$ for some $o \colon \Xi \rhd [\Delta_1]A_{1}, \ldots, [\Delta_{n}] A_{n} \to A_0$ in $O$, where $\vars{x}_{\isTerm{i}}$'s are lists of variables whose length is equal to the length of~$\Delta_i$ and $t_i$'s are raw terms in the variable list $V, \vars{x}_i$
+    \item a construct $\tmOpts$ for some $\aritysymbol{o}{\Xi}{[\Delta_1]A_{1}, \ldots, [\Delta_{n}] A_{n}}{A_0}$ in $O$, where $\vars{x}_{\isTerm{i}}$'s are lists of variables whose length is equal to the length of~$\Delta_i$ and $t_i$'s are raw terms in the variable list $V, \vars{x}_i$
   \end{enumerate}
@@
 
-The definition of typing derivations is slightly involved.
+The definition of typing derivations is a bit more involved.
 We need some information to compare types on the object level during type synthesis and substitute those type variables in a typing derivation $\Gamma \vdash \tmOpts : A$ for an operation $o$ in $\Omega$ at some point.
@@
       {\Gamma |-_{\Sigma, \Omega} \tmOpts : \simsub{A_0}{\rho}}\,\Rule{Op}
+      \\
+    \text{for $\aritysymbol{o}{\Xi}{[\Delta_1]A_{1}, \ldots, [\Delta_{n}] A_{n}}{A_0}$ in $\Omega$}
     \end{mathpar}
@@
   \end{figure}
-  For a simple type theory $(\Sigma, \Omega)$, the set of \emph{typing derivations} of $\Gamma \vdash \isTerm{t} : A$ indexed by a context $\Gamma$, a raw term with free variables in $\erase{\Gamma}$, and a type $A$ consists of
+  For a simple type system $(\Sigma, \Omega)$, the set of \emph{typing derivations} of $\Gamma \vdash \isTerm{t} : A$ indexed by a context $\Gamma$, a raw term with free variables in $\erase{\Gamma}$, and a type $A$ consists of
   \begin{enumerate}
@@
     \item a derivation of $\Gamma |-_{\Sigma, \Omega} (t \annotate A) : A$ if $\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A$ has a derivation, and
-    \item a derivation of $\Gamma |-_{\Sigma, \Omega} \tmOpts : \simsub{A_0}{\rho}$ for $o \colon \Xi \rhd [\Delta_1]A_1, \ldots, [\Delta_{n}]A_{n} \to A_0$ if there is a function $\rho\colon \Xi \to \Type_{\Sigma}(\emptyset)$ and a derivation of $\Gamma, \vars{x}_{\isTerm{i}} : \Delta_i \vdash_{\Sigma, \Omega} \isTerm{t_i} : \simsub{A_i}{\rho}$ for each $i$,
+    \item a derivation of $\Gamma |-_{\Sigma, \Omega} \tmOpts : \simsub{A_0}{\rho}$ for $\aritysymbol{o}{\Xi}{[\Delta_1]A_1, \ldots, [\Delta_{n}]A_{n}}{A_0}$ if there is a function $\rho\colon \Xi \to \Type_{\Sigma}(\emptyset)$ and a derivation of $\Gamma, \vars{x}_{\isTerm{i}} : \Delta_i \vdash_{\Sigma, \Omega} \isTerm{t_i} : \simsub{A_i}{\rho}$ for each $i$,
   \end{enumerate}
@@
 \end{definition}
-Note that, by definition, our simple type theory is always syntax-directed, since each rule in \Cref{fig:raw-terms} corresponds exactly to one rule in \Cref{fig:extrinsic-typing}. 
-The $\Rule{Op}$ in both figures is a rule schema and gives rise to as many rules as there are operations $o : O$.
+Note that our simple type system is always syntax-directed, since each rule in \Cref{fig:raw-terms} corresponds exactly to one rule in \Cref{fig:extrinsic-typing}. 
+The $\Rule{Op}$ rule in both figures is a rule schema and gives rise to as many rules as there are operations $o : O$.
 
@@
  \[
-   \mathsf{app}\colon A, B \rhd A \bto B, A \to B
+   \aritysymbol{\mathsf{app}}{(A, B)}{(A \bto B), A}{B}
    \quad\text{and}\quad
-   \mathsf{abs}\colon A , B \rhd [A]B \to (A \bto B).
+   \aritysymbol{\mathsf{abs}}{(A , B)}{[A]B}{(A \bto B)}.
  \]
@@
 \subsection{Bidirectional Binding Signatures and Bidirectional Type Systems} \label{subsec:bidirectional-system}
-Finally, for a bidirectional type system, we have raw terms and typing derivations like a simple type theory, but its typing judgements come in two forms $\Gamma |- t \syn A$ and $\Gamma |- t \chk A$.
+Finally, for a bidirectional type system, we have raw terms and typing derivations like a simple type system, but its typing judgements come in two forms $\Gamma |- t \syn A$ and $\Gamma |- t \chk A$.
 Observe that these two typing judgements are just one typing judgement $\Gamma |- t :^\dir{d} A$ indexed additionally by a \emph{mode} $d : \Mode$---which is either $\syn$ or $\chk$---so that $\Gamma |- t \syn A$ and $\Gamma |- t \chk A$ are just abbreviations of $\Gamma |- t :^{\syn} A$ and $\Gamma |- t :^{\chk} A$ respectively.
@@
 
-For brevity, we write $o \colon \Xi \rhd [\Delta_1]A_{1}^{\dir{d_1}}, \ldots, [\Delta_{n}] A^{\dir{d_n}}_{n} \to A_0^{\dir{d}}$ to indicate an operation with its arity. 
-For example, the $\ChkRule{Abs}$ rule in \Cref{fig:STLC-bidirectional-typing-derivations} can be specified by $\mathsf{abs}\colon A , B \rhd [A]B^{\chk} \to (A \bto B)^{\chk}$ indicating additionally that the construct $\lambda$ and its argument are in checking mode.
-Similarly, the $\SynRule{App}$ rule can be specified by $\mathsf{app}\colon A, B \rhd (A \bto B)^{\syn}, A^{\chk} \to B^{\syn}$.
+For brevity, we write $\aritysymbol{o}{\Xi}{[\Delta_1]A_{1}^{\dir{d_1}}, \ldots, [\Delta_{n}] A^{\dir{d_n}}_{n}}{A_0^{\dir{d}}}$ to indicate an operation with its arity. 
+For example, the $\ChkRule{Abs}$ rule in \Cref{fig:STLC-bidirectional-typing-derivations} can be specified by $\aritysymbol{\mathsf{abs}}{A , B}{[A]B^{\chk}}{(A \bto B)^{\chk}}$ indicating additionally that $\lam{x}t$ and its argument $t$ are in checking mode.
+Similarly, the $\SynRule{App}$ rule can be specified by $\aritysymbol{\mathsf{app}}{A, B}{(A \bto B)^{\syn}, A^{\chk}}{B^{\syn}}$.
 
-We also say a bidirectional type system $(\Sigma, \Omega)$ referring the bidirectional type system specified by some type signature $\Sigma$ and a bidirectional binding signature $\Omega$.
+We also say a bidirectional type system $(\Sigma, \Omega)$ referring to the bidirectional type system specified by some type signature $\Sigma$ and a bidirectional binding signature $\Omega$.
 
 \begin{remark}
-  Every bidirectional binding signature $\Omega$ gives rise to a binding signature $\erase{\Omega}$ if we erase modes from $\Omega$, so a bidirectional binding signature $\Omega$ also specifies a simple type theory, including its raw terms and typing derivations.
+  Every bidirectional binding signature $\Omega$ gives rise to a binding signature $\erase{\Omega}$ if we erase modes from $\Omega$, so a bidirectional binding signature $\Omega$ also specifies a simple type system, including its raw terms and typing derivations.
   We call $\erase{\Omega}$ the \emph{(mode) erasure} of $\Omega$.
@@
       {\Gamma |-_{\Sigma, \Omega} \tmOpts :^{\dir{d}} \simsub{A_0}{\rho}} \,\mathsf{Op}
+      \\ \text{for $\aritysymbol{o}{\Xi}{[\Delta_1]A_{1}^{\dir{d_1}}, \ldots, [\Delta_{n}] A_{n}^{\dir{d_n}}}{A_0^{\dir{d}}}$ in $\Omega$}
     \end{mathpar}
@@
       {V |-_{\Sigma, \Omega} \tmOpts^\dir{d}}\,\Rule{Op}
+      \and \text{for $\aritysymbol{o}{\Xi}{[\Delta_1]A_{1}^{\dir{d_1}}, \ldots, [\Delta_{n}] A^{\dir{d_n}}_{n}}{A_0^{\dir{d}}}$}
     \end{mathpar}
@@
   \begin{itemize}
-    \item the set of \emph{bidirectional typing derivations} of $\Gamma \vdash_{\Sigma, \Omega} t :^\dir{d} A$ indexed by a context $\Gamma$ of typed variables, a raw term $\isTerm{t}$ under $\erase{\Gamma}$, a mode $\dir{d}$, and a type $A$ is defined in \Cref{fig:bidirectional-typing-derivations};
+    \item the set of \emph{bidirectional typing derivations} of $\Gamma \vdash_{\Sigma, \Omega} t :^\dir{d} A$ indexed by a context $\Gamma$ of typed variables, a raw term $\isTerm{t}$ under $\erase{\Gamma}$, a mode $\dir{d}$, and a type $A$ is defined in \Cref{fig:bidirectional-typing-derivations} and
+      \begin{itemize}
+        \item in particular, a derivation of
+          \[
+            \Gamma |-_{\Sigma, \Omega} \tmOpts :^{\dir{d}} \simsub{A_0}{\rho}
+          \]
+          for $\aritysymbol{o}{\Xi}{[\Delta_1]A_{1}^{\dir{d_1}}, \ldots, [\Delta_{n}] A_{n}^{\dir{d_n}}}{A_0^{\dir{d}}}$ in $\Omega$ if there is a function $\rho\colon \Xi \to \Type_{\Sigma}(\emptyset)$ and a derivation of $\Gamma, \vars{x}_{\isTerm{i}} : \Delta_i \vdash_{\Sigma, \Omega} \isTerm{t_i} :^{\dir{d_i}} \simsub{A_i}{\rho}$ for each $i$;
+      \end{itemize}
     \item the set of \emph{mode derivations} of $V |-_{\Sigma, \Omega} t^\dir{d}$ indexed by a list $V$ of variables, a raw term $\isTerm{t}$ under $V$, and a mode $d$ is defined in \Cref{fig:mode-derivations}.
@@
 \begin{example}
-  Bidirectional typing derivations and mode derivations for simply typed $\lambda$-calculus can be specified by the type signature $\Sigma_{\bto}$ (\Cref{ex:type-signature-for-function-type}) and the bidirectional binding signature consisting of $\mathsf{abs}\colon A , B \rhd [A]B^{\chk} \to (A \bto B)^{\chk}$ and $\mathsf{app}\colon A, B \rhd (A \bto B)^{\syn}, A^{\chk} \to B^{\syn}$.
+  Bidirectional typing derivations and mode derivations for simply typed $\lambda$-calculus can be specified by the type signature $\Sigma_{\bto}$ (\Cref{ex:type-signature-for-function-type}) and the bidirectional binding signature consisting of $\aritysymbol{\mathsf{abs}}{A , B}{[A]B^{\chk}}{(A \bto B)^{\chk}}$ and $\aritysymbol{\mathsf{app}}{A, B}{(A \bto B)^{\syn}, A^{\chk}}{B^{\syn}}$.
 \end{example}

### 2023-07-06 tex/sec1-intro.tex
[ tex ] Minor in S1

@@
 However, it was later found that full parametric polymorphism leads to undecidability in type inference, as do dependent types~\citep{Wells1999,Dowek1993}.
-In light of these limitations, bidirectional type synthesis emerged as a viable alternative, providing algorithms for deciding the types of raw terms that meet some syntactic criteria%
-\todo{confirm}
-and usually contain some type annotations.
+In light of these limitations, bidirectional type synthesis emerged as a viable alternative, providing algorithms for deciding the types of raw terms that meet some syntactic criteria and usually contain some type annotations.
 \citet{Dunfield2021} summarised the design principles of bidirectional type synthesis and its wide coverage of languages with simple types, polymorphic types, dependent types, gradual types, among others.

### 2023-07-06 tex/sec2-outline.tex
[ tex ] Display the overview eariler
[ tex ] Do not explain the general form of bidirectional typing judgment

@@
 
-The construction is straightforward:
-Run the mode preprocessor on the input term~$|\Gamma| \vdash t$.
-If there is no synthesising mode derivation, report that $t$~is not in synthesising mode (the third outcome).
-Otherwise $|\Gamma| \vdash t^{\syn}$, and we can run the bidirectional type synthesiser.
-If it finds a derivation of $\Gamma \vdash t \syn A$ for some type~$A$, return a derivation of $\Gamma \vdash t : A$ (the first outcome), which is possible because the bidirectional typing~(\cref{fig:STLC-bidirectional-typing-derivations}) is \emph{sound}%
-\todo{Explain the notation `$:^\dir{d}$'}
-with respect to the original typing~(\cref{fig:STLC-typing-derivations}).
-
-\begin{lemma}[Soundness]
-If\/ $\Gamma \vdash t :^\dir{d} A$, then $\Gamma \vdash t : A$.
-\end{lemma}
-
-If there is no derivation of $\Gamma \vdash t \syn A$ for any type~$A$, report that there is no derivation of $\Gamma \vdash t : A$ for any~$A$ either (the second outcome), because the bidirectional typing is \emph{complete} with respect to the original typing.
-
-\begin{lemma}[Completeness]
-If\/ $|\Gamma| \vdash t^\dir{d}$ and\/ $\Gamma \vdash t : A$, then $\Gamma \vdash t :^\dir{d} A$.
-\end{lemma}
-
-We will construct a mode preprocessor~(\cref{sec:mode-preprocessing}) and a bidirectional type synthesiser~(\cref{sec:type-synthesis}) and prove both lemmas~(\cref{sec:soundness-and-completeness}) for all syntax-directed bidirectional simple type systems.
-To quantify over all such systems, we need their generic definitions, which we formulate in \cref{sec:defs}.
-
 \begin{figure}
@@
 \end{figure}
+
+The construction is straightforward:
+Run the mode preprocessor on the input term~$|\Gamma| \vdash t$.
+If there is no synthesising mode derivation, report that $t$~is not in synthesising mode (the third outcome).
+Otherwise $|\Gamma| \vdash t^{\syn}$, and we can run the bidirectional type synthesiser.
+If it finds a derivation of $\Gamma \vdash t \syn A$ for some type~$A$, return a derivation of $\Gamma \vdash t : A$ (the first outcome), which is possible because the bidirectional typing~(\cref{fig:STLC-bidirectional-typing-derivations}) is \emph{sound}%
+with respect to the original typing~(\cref{fig:STLC-typing-derivations}).
+
+\begin{lemma}[Soundness]
+If\/ $\Gamma \vdash t \syn A$, then $\Gamma \vdash t : A$.
+\end{lemma}
+
+If there is no derivation of $\Gamma \vdash t \syn A$ for any type~$A$, report that there is no derivation of $\Gamma \vdash t : A$ for any~$A$ either (the second outcome), because the bidirectional typing is \emph{complete} with respect to the original typing.
+
+\begin{lemma}[Completeness]
+If\/ $|\Gamma| \vdash t^\dir{\syn}$ and\/ $\Gamma \vdash t : A$, then $\Gamma \vdash t \syn A$.
+\end{lemma}
+
+We will construct a mode preprocessor~(\cref{sec:mode-preprocessing}) and a bidirectional type synthesiser~(\cref{sec:type-synthesis}) and prove both lemmas~(\cref{sec:soundness-and-completeness}) for all syntax-directed bidirectional simple type systems.
+To quantify over all such systems, we need their generic definitions, which we formulate in \cref{sec:defs}.

### 2023-07-06 tex/BiSig.tex
[ tex ] Use `\aritysymbol` in S3; explain the notation more carefully
[ tex ] Minor revisions up to S4

@@
 
+\usepackage{subfiles}
 \usepackage{bbold}
@@
 
+% Used for typesetting Agda code
+\usepackage{agda-lhs2tex}
+
 \usepackage{xifthen}
@@
 \microtypesetup{babel}
-\usepackage{subfiles}
-
-\citestyle{acmauthoryear}
-\AtEndPreamble{%
-\theoremstyle{acmdefinition}
-\newtheorem{remark}[theorem]{Remark}
-}
 
@@
 \input{macros}
-\usepackage{agda-lhs2tex}
+
+\citestyle{acmauthoryear}
+
+\AtEndPreamble{%
+\theoremstyle{acmdefinition}
+\newtheorem{remark}[theorem]{Remark}
+\newtheorem*{claim}{Claim}
+\newtheorem{claimproof}{Proof of Claim}
+}
 
@@
 %  which is analogous to a grammar in parsing.
-  Second, it is constructive because it is not only formalised in a proof assistant based on Martin-L\"of type theory, but also formulated in a way that unifies computation and proof, resulting in a compact and practical theory---for example, a proof of the decidability of the type synthesis problem is computationally a type synthesiser.
+  Second, it is constructive because it is not only formalised in a proof assistant based on Martin-L\"of type theory, but also compactly formulated in a way that unifies computation and proof---for example, a proof of the decidability of the type synthesis problem is computationally a type synthesiser.
   Moreover, we introduce `mode preprocessing' to deal with missing type annotations, an aspect often overlooked in the literature.
-  The main theorem is logically a trichotomy on parsed syntax trees and computationally a synthesiser that may throw an exception due to missing type annotations.
+  The main theorem is logically a trichotomy on untyped abstract syntax trees and computationally a type synthesiser that may throw an exception due to missing type annotations.
   By being parametrised by a signature, our theory works as a verified type-synthesiser generator overall.
@@
 \begin{acks}
-We thank Nathanael Arkor for the useful conversation and to thank the Programming Languages and Formal Methods Laboratory at Academia Sinica for the opportunity of sharing earlier ideas.
-
+%We thank Nathanael Arkor for the useful conversation.
 The work is supported by the Ministry of Science and Technology of Taiwan under grant MOST 109-2222-E-001-002-MY3.

### 2023-07-06 tex/sec1-intro.tex
[ tex ] Minor revisions up to S4

@@
 
-Type inference is an important mechanism for the transition to well-typed programs from parsed abstract syntax trees, which we call \emph{raw terms}.
+Type inference is an important mechanism for the transition to well-typed programs from untyped abstract syntax trees, which we call \emph{raw terms}.
 Here `type inference' refers specifically to algorithms that ascertain the type of any raw term \emph{without type annotations}.
@@
 Theoretically, separating the annotation and typing aspects streamlines our theory, in particular allowing us to formulate completeness naturally and differently from \emph{annotatability}, which was proposed and conflated with completeness by \citet{Dunfield2021}.
-By combining bidirectional type synthesis, soundness, completeness, and mode preprocessing, we achieve our main result~(\cref{cor:trichotomy}): a \emph{trichotomy} on raw terms, which is computationally a type synthesiser that checks if a given raw term is \emph{suitably annotated} and if a suitably annotated raw term has an ordinary typing derivation.
+By combining bidirectional type synthesis, soundness, completeness, and mode preprocessing, we achieve our main result~(\cref{cor:trichotomy}): a \emph{trichotomy} on raw terms, which is computationally a type synthesiser that checks if a given raw term is sufficiently annotated and if a sufficiently annotated raw term has an ordinary typing derivation.
 

### 2023-07-06 tex/sec2-outline.tex
[ tex ] Minor revisions up to S4

@@
 To give more precise definitions:
-The raw terms%
+The raw terms for simply typed $\lambda$-calculus are defined%
 \footnote{We omit the usual conditions about named representations of variables throughout this paper.}
-for simply typed $\lambda$-calculus are defined in \cref{fig:STLC-raw-terms};
-besides the standard constructs, there is an \textsc{Anno} rule that allows the user to insert type annotations to help with type synthesis.
+in \cref{fig:STLC-raw-terms};
+besides the standard constructs, there is an \textsc{Anno} rule that allows the user to insert type annotations to facilitate type synthesis.
 Correspondingly, the definition of typing derivations%
@@
 \label{def:STLC-type-synthesiser}
-Parametrised by an `excuse' predicate~$E$ on raw terms, a \emph{type synthesiser} takes a context~$\Gamma$ and a raw term $|\Gamma| \vdash t$ (where $|\Gamma|$~is the list of variables appearing in~$\Gamma$) as input, and establishes one of the following outcomes:
+Parametrised by an `excuse' predicate~$E$ on raw terms, a \emph{type synthesiser} takes a context~$\Gamma$ and a raw term $\erase\Gamma \vdash t$ (where $\erase\Gamma$~is the domain of~$\Gamma$) as input, and establishes one of the following outcomes:
 \begin{enumerate}
@@
 Now our goal is to use \cref{def:STLC-type-synthesiser} as a specification and implement it using a \emph{bidirectional} type synthesiser, which attempts to produce \emph{bidirectional} typing derivations defined in \cref{fig:STLC-bidirectional-typing-derivations}.
-
-\Josh{Here we briefly recap the basic ideas of bidirectional type synthesis.
-If a bidirectional type system is designed well (i.e.~\emph{mode-correct}, which will be defined in \cref{sec:type-synthesis}), then there will be a flow of type information in each rule that allows us to determine unknown types (e.g.~types to be checked) from known ones (e.g.~types previously synthesised), and suggests how bidirectional type synthesis is performed --- for example, the $\SynRule{App}$ rule says that to synthesise the type of an application $\isTerm{t\;u}$, we first synthesise the type of~$\isTerm{t}$, which should have the form $A \bto B$, from which we can extract the expected type of~$\isTerm{u}$, namely~$A$, and perform checking; then the type of the whole application, namely~$B$, can also be extracted from the previously synthesised type $A \bto B$.}
+It is often said that a type synthesis algorithm can be `read off' from well-designed bidirectional typing rules.
+Take the $\SynRule{App}$ rule as an example:
+to synthesise the type of an application $\isTerm{t\;u}$, we first synthesise the type of~$\isTerm{t}$, which should have the form $A \bto B$, from which we can extract the expected type of~$\isTerm{u}$, namely~$A$, and perform checking; then the type of the whole application, namely~$B$, can also be extracted from the type $A \bto B$.
+Note that the synthesiser is able to figure out the type~$A$ for checking~$u$ and the type~$B$ to be synthesised for $t\;u$ because they have been computed when synthesising the type $A \bto B$ of~$t$.
+In general, there should be a flow of type information in each rule that allows us to determine unknown types (e.g.~types to be checked) from known ones (e.g.~types previously synthesised).
+This is called \emph{mode-correctness}, which we will formally define in \cref{sec:mode-correctness}.
 
@@
   \small
-  \judgbox{V |- \isTerm{t}^\dir{d}}{A raw term~$\isTerm{t}$ is in mode~$d$ with free variables in $V$}
+  \bgroup
+  \renewcommand{\arraystretch}{1.5}
+  \begin{tabular}{ r l }
+    $\boxed{V |- \isTerm{t}^{\syn}}$ & A raw term~$\isTerm{t}$ (with free variables in~$V$) is in synthesising mode \\
+    $\boxed{V |- \isTerm{t}^{\chk}}$ & A raw term~$\isTerm{t}$ (with free variables in~$V$) is in checking mode
+  \end{tabular}
+  \egroup
   \begin{mathpar}
@@
 \label{def:STLC-mode-preprocessor}
-A \emph{mode preprocessor} decides for any raw term~$V \vdash \isTerm{t}$ and mode~$\dir{d}$ whether $V \vdash \isTerm{t}^\dir{d}$.
+A \emph{mode preprocessor} decides for any raw term~$V \vdash \isTerm{t}$ whether $V \vdash \isTerm{t}^{\syn}$.
 \end{definition}
@@
 (We will treat all these more rigorously in \cref{sec:pre-synthesis}.)
-For example, an abstraction is strictly in checking mode, but the left sub-term of an application has to be synthesising, so the term $\isTerm{(\lam{x}t)\;u}$ does not have a mode derivation unless we annotate the abstraction.
+For example, an abstraction is strictly in checking mode, but the left sub-term of an application has to be synthesising, so a term of the form $\isTerm{(\lam{x}t)\;u}$ does not have a mode derivation unless we annotate the abstraction.
 
@@
 \label{def:STLC-bidirectional-type-synthesiser}
-A \emph{bidirectional type synthesiser} decides for any context~$\Gamma$ and synthesising term $|\Gamma| \vdash \isTerm{t}^{\syn}$ whether $\Gamma \vdash \isTerm{t} \syn A$ for some type~$A$.
+A \emph{bidirectional type synthesiser} decides for any context~$\Gamma$ and synthesising term $\erase\Gamma \vdash \isTerm{t}^{\syn}$ whether $\Gamma \vdash \isTerm{t} \syn A$ for some type~$A$.
 \end{definition}
@@
 
-\begin{figure}
-  \small
-  \begin{tikzpicture}
-  \matrix (m) [matrix of math nodes, row sep=3em, column sep=10em]
-  {
-    \boxed{\erase{\Gamma} \vdash \isTerm{t}}   & \boxed{\erase{\Gamma} \vdash \isTerm{t}^{\syn}} \\
-    \boxed{\Gamma \vdash \isTerm{t} : A} & \boxed{\Gamma \vdash \isTerm{t} \syn A} \\
-  };
-  \path[->, line width=0.5pt]
-      (m-1-1) edge node [above] {mode preprocessing} (m-1-2)
-      (m-1-2) edge node [right,align=center] {bidirectional\\type synthesis} (m-2-2)
-      (m-2-1) edge[bend right=8] node [below] {completeness} (m-2-2)
-      (m-2-2) edge[bend right=8] node [above] {soundness}    (m-2-1);
-  \path[dashed,->,line width=0.5pt]
-      (m-1-1) edge node [left,align=center]  {type\\synthesis} (m-2-1);
-  \end{tikzpicture}
-  \caption{An implementation of type synthesis via bidirectional typing}
-\end{figure}
-
 The construction is straightforward:
-Run the mode preprocessor on the input term~$|\Gamma| \vdash t$.
+Run the mode preprocessor on the input term~$\erase\Gamma \vdash t$.
 If there is no synthesising mode derivation, report that $t$~is not in synthesising mode (the third outcome).
-Otherwise $|\Gamma| \vdash t^{\syn}$, and we can run the bidirectional type synthesiser.
-If it finds a derivation of $\Gamma \vdash t \syn A$ for some type~$A$, return a derivation of $\Gamma \vdash t : A$ (the first outcome), which is possible because the bidirectional typing~(\cref{fig:STLC-bidirectional-typing-derivations}) is \emph{sound}%
-with respect to the original typing~(\cref{fig:STLC-typing-derivations}).
+Otherwise $\erase\Gamma \vdash t^{\syn}$, and we can run the bidirectional type synthesiser.
+If it finds a derivation of $\Gamma \vdash t \syn A$ for some type~$A$, return a derivation of $\Gamma \vdash t : A$ (the first outcome), which is possible because the bidirectional typing~(\cref{fig:STLC-bidirectional-typing-derivations}) is \emph{sound} with respect to the original typing~(\cref{fig:STLC-typing-derivations}).
 
@@
 
-\begin{lemma}[Completeness]
-If\/ $|\Gamma| \vdash t^\dir{\syn}$ and\/ $\Gamma \vdash t : A$, then $\Gamma \vdash t \syn A$.
+\begin{lemma}[Completeness] \label{thm:STLC-completeness}
+If\/ $\erase\Gamma \vdash t^{\syn}$ and\/ $\Gamma \vdash t : A$, then $\Gamma \vdash t \syn A$.
 \end{lemma}
@@
 We will construct a mode preprocessor~(\cref{sec:mode-preprocessing}) and a bidirectional type synthesiser~(\cref{sec:type-synthesis}) and prove both lemmas~(\cref{sec:soundness-and-completeness}) for all syntax-directed bidirectional simple type systems.
-To quantify over all such systems, we need their generic definitions, which we formulate in \cref{sec:defs}.
+To quantify over all such systems, we need their generic definitions, which we formulate next.
+
+%\begin{figure}
+%  \small
+%  \begin{tikzpicture}
+%  \matrix (m) [matrix of math nodes, row sep=3em, column sep=10em]
+%  {
+%    \boxed{\erase{\Gamma} \vdash \isTerm{t}}   & \boxed{\erase{\Gamma} \vdash \isTerm{t}^{\syn}} \\
+%    \boxed{\Gamma \vdash \isTerm{t} : A} & \boxed{\Gamma \vdash \isTerm{t} \syn A} \\
+%  };
+%  \path[->, line width=0.5pt]
+%      (m-1-1) edge node [above] {mode preprocessing} (m-1-2)
+%      (m-1-2) edge node [right,align=center] {bidirectional\\type synthesis} (m-2-2)
+%      (m-2-1) edge[bend right=8] node [below] {completeness} (m-2-2)
+%      (m-2-2) edge[bend right=8] node [above] {soundness}    (m-2-1);
+%  \path[dashed,->,line width=0.5pt]
+%      (m-1-1) edge node [left,align=center]  {type\\synthesis} (m-2-1);
+%  \end{tikzpicture}
+%  \caption{An implementation of type synthesis via bidirectional typing}
+%\end{figure}

### 2023-07-06 tex/sec3-defs.tex
[ tex ] Minor revisions up to S4

@@
 %Here's the 'small print': Our definitions cover typical examples and set the scope for bidirectional simple type systems discussed, but do not offer comprehensive coverage of all possibilities.}
-of simple types, simple type systems, and bidirectional type systems using simply typed $\lambda$-calculus in \Cref{sec:key-ideas} as our running example.
+of simple types, simply typed languages, and bidirectional type systems, using simply typed $\lambda$-calculus in \cref{sec:key-ideas} as our running example.
+All these definitions are inductive, which we will not emphasise in the individual definitions.
 
-Our definitions are defined in two steps:
+Our definitions are formulated in two steps:
 \begin{enumerate*}
@@
 \subsection{Signatures and Simple Types} \label{subsec:simple-types}
-Simple types of a language is understood intuitively as a collection of types whose formation judgement is not indexed by anything but a fixed set of type variables and the only datum needed for specifying a type construct is its number of arguments:
+Simple types are understood intuitively as a collection of types whose formation judgement is not indexed by anything but a fixed set of type variables, and the only datum needed for specifying a type construct is its number of arguments:
 
@@
     Even though our theory is developed in Martin-L\"of type theory, the term `set' is used instead of `type' to avoid the obvious confusion. 
-    As we work in Martin-L\"of type theory with \AxiomK, all types are legitimately sets in the sense of homotopy type theory where all inhabitants of its identity type are proved equal.
-  }
+    As we work in Martin-L\"of type theory with \AxiomK, all types are legitimately sets in the sense of homotopy type theory where all inhabitants of its identity type are identified.}
   $I$ with a decidable equality and an \emph{arity} function $\arity\colon I \to \mathbb{N}$.
@@
 
-As \Cref{def:simple-signature} includes type variables, we introduce the typical meta-operations: substitution.
+As \Cref{def:simple-signature} includes type variables, we introduce the typical meta-operation: substitution.
 \begin{definition}
@@
 
-Also for brevity, a simple type system $(\Sigma, \Omega)$ refers to the simple type system specified by some type signature $\Sigma$ and a binding signature $\Omega$.
+Also for brevity, a simply typed language $(\Sigma, \Omega)$ refers to the simply typed language specified by some type signature $\Sigma$ and a binding signature $\Omega$.
 
@@
   \small
-  \judgbox{V |-_{\Sigma, \Omega} \isTerm{t}}{Given a list V of variables, $\isTerm{t}$ is a raw term for the signature $(\Sigma, \Omega)$ with free variables in $V$}
+  \judgbox{V |-_{\Sigma, \Omega} \isTerm{t}}{Given a list~$V$ of variables, $\isTerm{t}$ is a raw term for the signature $(\Sigma, \Omega)$ with free variables in~$V$}
   \begin{mathpar}

### 2023-07-06 tex/sec4-pre-synthesis.tex
[ tex ] Minor revisions up to S4

@@
 
-\Josh{Overview of the section}
+Our first important construction is mode preprocessing in \cref{sec:mode-preprocessing}, which is in fact generalised to pinpoint any missing type annotations in a given raw term.
+We also discuss some related properties:
+By bringing mode derivations into the picture, we are able to give a natural formulation of soundness and completeness of a bidirectional type system with respect to its erasure to an ordinary type system in \cref{sec:soundness-and-completeness}.
+We also formulate \varcitet{Dunfield2021}{'s} annotatability and discuss its relationship with our completeness and generalised mode preprocessing in \cref{sec:annotatability}.
 
@@
 
-\Josh{One theorem to rule them all}
+Erasure of a bidirectional binding signature removes mode information and keeps everything else intact; this can be straightforwardly extended by induction to remove mode information from a bidirectional typing derivation and arrive at an ordinary typing derivation, which is soundness.
+We can also remove typing and retain mode information, arriving at a mode derivation instead.
+Conversely, if we have both mode and typing derivations for the same term, we can combine them and obtain a bidirectional typing derivation, which is completeness.
+In short, soundness and completeness are no more than the separation and combination of mode and typing information carried by the three kinds of derivations while keeping their basic structure, which is directed by the same raw term.
+%Formulating the notion of mode derivation helps to complete this clean picture.
+Technically, all these can be summarised in one lemma.
 
-Erasure of a bidirectional binding signature removes mode information and keeps everything else intact; this can be straightforwardly extended by induction to remove mode information from a bidirectional typing derivation and arrive at an ordinary typing derivation.
-
-\begin{lemma}[Soundness]\label{thm:soundness}
-If\/ $\Gamma \vdash_{\Sigma, \Omega} t :^\dir{d} A$, then $\Gamma \vdash_{\Sigma, \erase\Omega} t : A$.
+\begin{lemma}
+There is a bidirectional typing derivation of\/ $\Gamma |-_{\Sigma, \Omega} t :^\dir{d} A$ if and only if there are mode and typing derivations of\/ $\erase\Gamma |-_{\Sigma, \Omega} t^\dir{d}$ and\/ $\Gamma |-_{\Sigma, \erase\Omega} t : A$.
 \end{lemma}
@@
 \begin{proof}
-Induction on the given derivation, mapping every bidirectional typing rule to its mode-less counterpart except $\ChkRule{Sub}$, in which case the induction hypothesis $\Gamma |-_{\Sigma,\erase\Omega} t : B$ suffices due to the premise $B = A$.
-\end{proof}
+From left to right:
+Induction on the bidirectional typing derivation, mapping every rule to its counterpart except when constructing a typing derivation from the $\ChkRule{Sub}$ rule, in which case the induction hypothesis $\Gamma |-_{\Sigma,\erase\Omega} t : B$ suffices due to the premise $B = A$.
 
-We can also remove typing and retain mode information, arriving at a mode derivation instead.
-
-\begin{proposition}\label{thm:typing-removal}
-If\/ $\Gamma \vdash_{\Sigma, \Omega} t :^\dir{d} A$, then $\erase\Gamma \vdash_{\Sigma, \Omega} t^\dir{d}$.
-\end{proposition}
-
-\begin{proof}
-Induction on the given derivation, mapping every rule to its counterpart.
-\end{proof}
-
-Conversely, if we have both mode and typing derivations for the same term, we can combine them and obtain a bidirectional typing derivation.
-
-\begin{lemma}[Completeness]\label{thm:completeness}
-If\/ $\erase\Gamma |-_{\Sigma, \Omega} t^\dir{d}$ and\/ $\Gamma |-_{\Sigma, \erase\Omega} t : A$, then $\Gamma |-_{\Sigma, \Omega} t :^\dir{d} A$.
-\end{lemma}
-
-\begin{proof}
-Induction on the given mode derivation.
-For \SynRule{Var}, \SynRule{Anno}, and \Rule{Op}, the outermost rule used in the given typing derivation must be the corresponding typing rule, so by the induction hypotheses we have bidirectional typing derivations for all the sub-terms, to which we can then apply the corresponding bidirectional typing rule.
+From right to left:
+Induction on the mode derivation.
+For \SynRule{Var}, \SynRule{Anno}, and \Rule{Op}, the outermost rule used in the typing derivation must be the corresponding typing rule, so by the induction hypotheses we have bidirectional typing derivations for all the sub-terms, to which we can then apply the corresponding bidirectional typing rule.
 The \ChkRule{Sub} case is similar but slightly simpler: the induction hypothesis directly gives us a derivation of $\Gamma |-_{\Sigma,\Omega} t \syn A$, to which we apply $\ChkRule{Sub}$.
 \end{proof}
-
-In short, soundness and completeness are no more than the separation and combination of mode and typing information carried by the three kinds of derivations while keeping their basic structure, which is directed by the same raw term.
-%Formulating the notion of mode derivation helps to complete this clean picture.
 
@@
 It will be more helpful if a preprocessor can produce an explanation of why no mode derivation exists, and even how to fix the input term to have a mode derivation.
-We will construct such a \emph{generalised mode preprocessor}~(\cref{thm:generalised-mode-preprocessing}), which can be weakened to an ordinary mode preprocessor~(\cref{thm:mode-preprocessing}) if the additional explanation is not needed.
-
-\renewcommand{\True}{\mathbf{T}}
-\renewcommand{\False}{\mathbf{F}}
+We will construct such a \emph{generalised mode preprocessor}~(\cref{thm:generalised-mode-preprocessing}), which can be weakened to an ordinary mode preprocessor~(\cref{thm:mode-preprocessing}) if the additional explanation is not needed.%
+\footnote{For clarity, we use ordinary mode preprocessing elsewhere in this paper.}
 
@@
     & The raw term~$\isTerm{t}$\hspace{-.6em}
-    & misses some type annotation iff $g = \False$, and \\
-    & & is in mode~$d$ due to an outermost mode cast iff $s = \False$
+    & misses some type annotation iff $g = \BFalse$, and \\
+    & & is in mode~$d$ due to an outermost mode cast iff $s = \BFalse$
   \end{tabular}
   \begin{mathpar}
-    \inferrule{x \in V}{V |-_{\Sigma, \Omega} \isTerm{x}^{\syn\,\True\,\True}}\,\SynRule{Var}
+    \inferrule{x \in V}{V |-_{\Sigma, \Omega} \isTerm{x}^{\syn\,\BTrue\,\BTrue}}\,\SynRule{Var}
     \and
-    \inferrule{\cdot |-_{\Sigma} A \\ V |-_{\Sigma, \Omega}\isTerm{t}^{\chk\,g\,s}}{V |-_{\Sigma, \Omega} (\isTerm{t \annotate A})^{\syn\,g\,\True}}\,\SynRule{Anno}
+    \inferrule{\cdot |-_{\Sigma} A \\ V |-_{\Sigma, \Omega}\isTerm{t}^{\chk\,g\,s}}{V |-_{\Sigma, \Omega} (\isTerm{t \annotate A})^{\syn\,g\,\BTrue}}\,\SynRule{Anno}
     \and
-    \inferrule{V |-_{\Sigma, \Omega}\isTerm{t}^{\chk\,g\,\True}}{V |-_{\Sigma, \Omega} \isTerm{t}^{\syn\,\False\,\False}}\,\SynRule{Missing}
+    \inferrule{V |-_{\Sigma, \Omega}\isTerm{t}^{\chk\,g\,\BTrue}}{V |-_{\Sigma, \Omega} \isTerm{t}^{\syn\,\BFalse\,\BFalse}}\,\SynRule{Missing}
     \and
-    \inferrule{V |-_{\Sigma, \Omega} \isTerm{t}^{\syn\,g\,\True}}{V |-_{\Sigma, \Omega} \isTerm{t}^{\chk\,g\,\False}}\,\ChkRule{Sub}
+    \inferrule{V |-_{\Sigma, \Omega} \isTerm{t}^{\syn\,g\,\BTrue}}{V |-_{\Sigma, \Omega} \isTerm{t}^{\chk\,g\,\BFalse}}\,\ChkRule{Sub}
     \and
     \inferrule{V, \vec x_1 |-_{\Sigma, \Omega} \isTerm{t_1}^{\dir{d_1}\,g_1\,s_1} \\ \cdots \\ V, \vec x_n |-_{\Sigma, \Omega} \isTerm{t_n}^{\dir{d_n}\,g_n\,s_n}}
-    {V |-_{\Sigma, \Omega} \tmOpts^{\dir{d}\,(\bigwedge_i g_i)\,\True}}\,\Rule{Op}
+    {V |-_{\Sigma, \Omega} \tmOpts^{\dir{d}\,(\bigwedge_i g_i)\,\BTrue}}\,\Rule{Op}
   \end{mathpar}
@@
 
-Intuitively, a term does not have a mode derivation exactly when there are not enough type annotations, but such negative formulations convey little information.%
-\todo{still default in the rest of the paper}
+Intuitively, a term does not have a mode derivation exactly when there are not enough type annotations, but such negative formulations convey little information.
 Instead, we can provide more information by pointing out the places in the term that require annotations.
@@
 Given a generalised mode derivation, if it uses $\SynRule{Missing}$ in some places, then those places are exactly where annotations should be supplied, which is helpful information to the user; if it does not use $\SynRule{Missing}$, then the derivation is \emph{genuine} in the sense that it corresponds directly to an original mode derivation.
-This can be succinctly formulated as \cref{thm:Pre?-true} below by encoding genuineness as a boolean~$g$ in the generalised mode judgement, which is set to~$\False$ only by the $\SynRule{Missing}$ rule.
+This can be succinctly formulated as \cref{thm:Pre?-true} below by encoding genuineness as a boolean~$g$ in the generalised mode judgement, which is set to~$\BFalse$ only by the $\SynRule{Missing}$ rule.
 (Ignore the boolean~$s$ for now.)
@@
 \begin{lemma}\label{thm:Pre?-true}
-If\/ $V |-_{\Sigma,\Omega} t^{\dir{d}\,\True\,s}$, then $V |-_{\Sigma,\Omega} t^\dir{d}$.
+If\/ $V |-_{\Sigma,\Omega} t^{\dir{d}\,\BTrue\,s}$, then $V |-_{\Sigma,\Omega} t^\dir{d}$.
 \end{lemma}
@@
 Induction on the given derivation.
-The $\SynRule{Missing}$ rule cannot appear because $g = \True$, and the other rules are mapped to their counterparts.
+The $\SynRule{Missing}$ rule cannot appear because $g = \BTrue$, and the other rules are mapped to their counterparts.
 \end{proof}
 
-We also want a lemma that covers the case where $g = \False$.
+We also want a lemma that covers the case where $g = \BFalse$.
 
 \begin{lemma}\label{thm:Pre?-false}
-If\/ $V |-_{\Sigma,\Omega} t^{\dir{d}\,\False\,s}$, then $V \not|-_{\Sigma,\Omega} t^\dir{d}$.
+If\/ $V |-_{\Sigma,\Omega} t^{\dir{d}\,\BFalse\,s}$, then $V \not|-_{\Sigma,\Omega} t^\dir{d}$.
 \end{lemma}
@@
 The `strictness' boolean~$s$ can be thought of as disrupting the formation of such pairs of mode casts:
-Every rule other than the mode casts sets~$s$ to~$\True$, meaning that a term is \emph{strictly} in the mode assigned by the rule (i.e.~not altered by a mode cast), whereas the mode casts set~$s$ to~$\False$.
+Every rule other than the mode casts sets~$s$ to~$\BTrue$, meaning that a term is \emph{strictly} in the mode assigned by the rule (i.e.~not altered by a mode cast), whereas the mode casts set~$s$ to~$\BFalse$.
 Furthermore, the sub-derivation of a mode cast has to be strict, so it is impossible to have consecutive mode casts.
@@
 Induction on the generalised mode derivation, in each case analysing an arbitrary mode derivation and showing that it cannot exist.
-The key case is $\SynRule{Missing}$, where we have a sub-derivation of $V |-_{\Sigma,\Omega} t^{\chk\,g\,\True}$ for some boolean~$g$.
-We do not have an induction hypothesis and seem to get stuck, because $g$~is not necessarily~$\False$.
+The key case is $\SynRule{Missing}$, where we have a sub-derivation of $V |-_{\Sigma,\Omega} t^{\chk\,g\,\BTrue}$ for some boolean~$g$.
+We do not have an induction hypothesis and seem to get stuck, because $g$~is not necessarily~$\BFalse$.
 But what matters here is that $t$~is \emph{strictly} in checking mode: if we continue to analyse the sub-derivation, the outermost rule must be $\Rule{Op}$ with $\dir d = {\chk}$, implying that $t$~has to be an operation in checking mode.
@@
 %\item $\ChkRule{Sub}$:
-%We have a sub-derivation of $V |-_{\Sigma,\Omega} t^{\syn\,\False\,\True}$.
+%We have a sub-derivation of $V |-_{\Sigma,\Omega} t^{\syn\,\BFalse\,\BTrue}$.
 %Any mode derivation of $V |-_{\Sigma,\Omega} t^{\chk}$ must use $\ChkRule{Chk}$ as the outermost rule.
@@
 \begin{equation}\label{eq:strict-derivation}
-V |-_{\Sigma,\Omega} t^{\dir{d'}\,g'\,\True} \quad\text{for some mode~$\dir{d'}$ and boolean~$g'$}
+V |-_{\Sigma,\Omega} t^{\dir{d'}\,g'\,\BTrue} \quad\text{for some mode~$\dir{d'}$ and boolean~$g'$}
 \end{equation}
@@
 %\begin{corollary}\label{thm:toPre?-true}
-%If\/ $V |-_{\Sigma,\Omega} t^\dir{d}$, then $V |-_{\Sigma,\Omega} t^{\dir{d}\,\True\,s}$ for some boolean~$s$.
+%If\/ $V |-_{\Sigma,\Omega} t^\dir{d}$, then $V |-_{\Sigma,\Omega} t^{\dir{d}\,\BTrue\,s}$ for some boolean~$s$.
 %\end{corollary}
@@
 %\begin{corollary}\label{thm:toPre?-false}
-%If\/ $V \not|-_{\Sigma,\Omega} t^\dir{d}$, then $V |-_{\Sigma,\Omega} t^{\dir{d}\,\False\,s}$ for some boolean~$s$.
+%If\/ $V \not|-_{\Sigma,\Omega} t^\dir{d}$, then $V |-_{\Sigma,\Omega} t^{\dir{d}\,\BFalse\,s}$ for some boolean~$s$.
 %\end{corollary}
@@
 %By \cref{thm:generalised-mode-preprocessing}, there is a derivation of $V |-_{\Sigma,\Omega} t^{\dir{d}\,g\,s}$ for some booleans $g$~and~$s$.
-%If $V |-_{\Sigma,\Omega} t^\dir{d}$, then $g$~has to be $\True$ since $g = \False$ leads to a contradiction with \cref{thm:Pre?-false}.
-%Symmetrically, if $V \not|-_{\Sigma,\Omega} t^\dir{d}$, then $g$~has to be~$\False$ so as not to contradict \cref{thm:Pre?-true}.
+%If $V |-_{\Sigma,\Omega} t^\dir{d}$, then $g$~has to be $\BTrue$ since $g = \BFalse$ leads to a contradiction with \cref{thm:Pre?-false}.
+%Symmetrically, if $V \not|-_{\Sigma,\Omega} t^\dir{d}$, then $g$~has to be~$\BFalse$ so as not to contradict \cref{thm:Pre?-true}.
 %\end{proof}
@@
 
-\citet{Dunfield2021} formulated completeness differently from ours (\cref{thm:completeness}) and proposed \emph{annotatability} as a more suitable name.
-Here we formulate annotatability in our theory and discuss its relationship with our completeness and generalised mode preprocessing.
+\citet{Dunfield2021} formulated completeness differently from ours (\cref{thm:STLC-completeness,sec:soundness-and-completeness}) and proposed \emph{annotatability} as a more suitable name.
+In our theory, we may formulate annotatability as follows.
 

### 2023-07-06 tex/sec5-synthesis.tex
[ tex ] Fix undefined references
[ tex ] Minor revisions up to S4
[ tex ] Some progress in S5

--- 2023-06-17 tex/sec5-synthesis.tex
@@
 \section{Bidirectional Type Synthesis and Checking} \label{sec:type-synthesis}
+This section is devoted to defining mode-correctness and deriving bidirectional type synthesis for any mode-correct bidirectional type system $(\Sigma, \Omega)$.
+\Cref{subsec:mode-correctness} formalises the concept of mode-correctness and establishes the uniqueness of synthesised types.
+This indicates that for a given raw term and context, any two synthesised types must be equal.
+This characteristic is also used in the proof of the logical decidability of bidirectional type synthesis.
+Later, in \Cref{subsec:trichotomy}, we introduce the decidability of bidirectional synthesis but postpone its proof until \Cref{subsec:bidirectional-synthesis-checking}.
+At that point, we will combine mode preprocessing, soundness, and completeness from \Cref{sec:pre-synthesis} to show the trichotomy on raw terms as a simple corollary.
+Note that the result is applicable to an arbitrary mode-correct bidirectional type system, rather than being restricted to the simply typed $\lambda$-calculus as we did in \Cref{sec:key-ideas}.
 
-\subsection{Mode Correctness}
-\begin{definition}[Mode-Correctness]
+\subsection{Mode Correctness}\label{sec:mode-correctness}
+As \citet[Section~3.1]{Dunfield2021} outlined, mode-correctness for a bidirectional type system means that for every bidirectional typing rule:
+\begin{enumerate*}
+\item each `input' type variable in a premise must be synthesised from `earlier' premises or, if the mode of the conclusion is checking, provided by the conclusion;
+\item each `output' type variable in the typing judgement of the conclusion should have been synthesised from premises already, if the mode of the conclusion is synthesis.
+\end{enumerate*}
+An `input' type variable refers to types in an extension context and the type of a premise in the checking mode.
+It is important to note that the order of premises in a bidirectional typing rule matters, as it also sets the order of bidirectional type synthesis.
+
+Consider the rule $\ChkRule{Abs}$ (\Cref{fig:STLC-bidirectional-typing-derivations}) as an example.
+This rule is mode-correct, as the type variables $A$ and $B$ in its only premise are already provided by its conclusion $A \bto B$.
+Likewise, the rule $\SynRule{App}$ for an application term $\isTerm{t\;u}$ is mode-correct because:
+\begin{enumerate*}
+\item the type $A \bto B$ of the first argument $t$ is synthesised, thereby ensuring type variables $A$ and $B$ must be known if successfully synthesised;
+\item the type of the second argument $u$ is checked against $A$, which was synthesised earlier;
+\item as a result, the type of an application $t\;u$ can be synthesised.
+\end{enumerate*}
+
+As this is a condition for \emph{typing rules} and not for raw terms or typing derivations, this property \emph{cannot} be formalised without the concept of bidirectional binding signature presented in \Cref{sec:defs}.
+This contrasts with the properties in \Cref{sec:pre-synthesis}, which are properties that can still be expressed for individual systems even without the general definition of bidirectional type systems.
+Now that we have the notion of signature available, we can articulate mode-correctness more precisely.
+
+First, let us introduce the finite subset\footnote{%
+  There are various definitions for finite subsets of a set within Martin-L\"{o}f type theory.
+  However, for our purposes, the choice among these definitions is not a matter of concern.
+}
+$\fv(A)$ and $\fv(\Gamma)$ of (free) variables in a type and in a context, respectively, which are defined in the usual way:
+\begin{enumerate*}
+\item For a type $A : \Type_{\Sigma}(\Xi)$ define the subset $\fv(A)$ of $\Xi$ inductively by $\fv(x) = {x}$ and $\fv(\tyOp_{i}(A_1, \ldots, A_n)) = \bigcup_{i = 1}^{n} \fv(A_i)$;
+\item by reusing this notation, for a context $\Gamma : \Type_{\Sigma}(\Xi)$ define the set $\fv(\Gamma)$ as $\fv(\emptyset) = \emptyset$ and $\fv(\Gamma, A) = \fv(\Gamma) \cup \fv(A)$.
+\end{enumerate*}
+As we have sketched informally, the mode-correctness condition for a synthesis rule is different from a checking rule
+\begin{definition}[Mode-Correctness] \label{def:mode-correctness}
+  An operation $\aritysymbol{o}{\Xi}{[\Delta_1]A_{1}^{\dir{d_1}}, \ldots, [\Delta_{n}] A_{n}^{\dir{d_n}}}{A_0^{\dir{d}}}$ is \emph{mode-correct} whenever
+  \begin{enumerate}
+    \item if $d$ is $\syn$ (synthesis);
+    \item if $d$ is $\chk$ (checking);
+  \end{enumerate}
+
 \end{definition}
 
-\begin{theorem}[Uniqueness of the Synthesised Types]\label{thm:unique-syn}
+\begin{lemma}[Uniqueness of the Synthesised Types]\label{thm:unique-syn}
   Let $(\Sigma, \Omega)$ be a mode-correct bidirectional type system.
@@
   for some types $A$ and $B$ respectively, the synthesised types must be equal, i.e.\ $A = B$.
-\end{theorem}
+\end{lemma}
 
-\subsection{Trichotomy of Raw Terms by Type Synthesis}
+\subsection{Trichotomy on Raw Terms by Type Synthesis} \label{subsec:trichotomy}
 
@@
 
-\begin{corollary}[Trichotomy of Raw Terms]
+\begin{corollary}[Trichotomy on Raw Terms]\label{cor:trichotomy}
   Let $(\Sigma, \Omega)$ be a mode-correct bidirectional type system.
@@
 \begin{proof}
-  Combine  \Cref{thm:term-soundness,thm:term-completeness,thm:bidirectionalisation} with \Cref{thm:bidirectional-type-synthesis}.
+  Combine  \Cref{thm:soundness,thm:completeness,thm:mode-preprocessing} with \Cref{thm:bidirectional-type-synthesis}.
 
@@
 
-\subsection{Decidability of Bidirectional Type Synthesis and Checking}
+\subsection{Decidability of Bidirectional Type Synthesis and Checking}\label{subsec:bidirectional-synthesis-checking}
 
@@
 \end{theorem}
-\begin{lemma}\label{lem:args-induction}
-  Assume \Cref{thm:bidirectional-type-synthesis}.
-  Then, for every set $\Xi$ of type variables, list 
-  \[
-    \mathit{as} = \left([\Delta_i] A_{i}^{\dir{d_i}}\right)_{i = 1}^{n}
-  \]
-  list of raw terms $t_i$'s in mode $d_i$, context $\Gamma$, and \emph{partial} substitution $\rho$ from $\Xi$ to $\emptyset$
-  \begin{enumerate}
-    \item either there is a minimal extension $\ext{\rho}$ of $\rho$ defined on $\synvar(\mathit{as})$ such that all of following judgements are derivable
-      \[
-        \Gamma, \vars{x}_\isTerm{i} : \simsub{\Delta_i}{\ext{\rho}} |-_{\Sigma, \Omega} \isTerm{t_i} \colon \simsub{A_i}{\ext{\rho}}^{\dir{d_i}}
-      \]
-
-    \item or there is no extension $\sigma$ of $\rho$ such that all $\Gamma, \vars{x}_\isTerm{i} : \simsub{\Delta_i}{\sigma} |-_{\Sigma, \Omega} \isTerm{t_i} \colon \simsub{A_i}{\sigma}^{\dir{d_i}}$ has a typing derivation.
-  \end{enumerate}
-\end{lemma}
 
@@
 
+  \begin{claim}\label{lem:args-induction}
+    For every set $\Xi$ of type variables, list 
+    \[
+      \mathit{as} = \left([\Delta_i] A_{i}^{\dir{d_i}}\right)_{i = 1}^{n}
+    \]
+    list of raw terms $t_i$'s in mode $d_i$, context $\Gamma$, and \emph{partial} substitution $\rho$ from $\Xi$ to $\emptyset$
+    \begin{enumerate}
+      \item either there is a minimal extension $\ext{\rho}$ of $\rho$ defined on $\synvar(\mathit{as})$ such that all of following judgements are derivable
+        \[
+          \Gamma, \vars{x}_\isTerm{i} : \simsub{\Delta_i}{\ext{\rho}} |-_{\Sigma, \Omega} \isTerm{t_i} \colon \simsub{A_i}{\ext{\rho}}^{\dir{d_i}}
+        \]
+
+      \item or there is no extension $\sigma$ of $\rho$ such that all $\Gamma, \vars{x}_\isTerm{i} : \simsub{\Delta_i}{\sigma} |-_{\Sigma, \Omega} \isTerm{t_i} \colon \simsub{A_i}{\sigma}^{\dir{d_i}}$ has a typing derivation.
+    \end{enumerate}
+  \end{claim}
   \end{description}

### 2023-07-06 tex/sec3-defs.tex
[ tex ] Revise S3

@@
 \begin{enumerate*}
-\item we first introduce a notion of signature which includes a set of operation symbols and an assignment of arities to each operation;
+  \item we first introduce a notion of signature which includes a set of operation symbols and an assignment of \emph{arities} to each operation;
 \item then, we define abstract syntax trees inductively with primitive constructs such as $\Rule{Var}$ and constructs specified by a given signature.
@@
     Even though our theory is developed in Martin-L\"of type theory, the term `set' is used instead of `type' to avoid the obvious confusion. 
-    As we work in Martin-L\"of type theory with \AxiomK, all types are legitimately sets in the sense of homotopy type theory where all inhabitants of its identity type are identified.}
+    Also, as we assume \AxiomK, all types are legitimate sets in the sense of homotopy type theory.
+  }
   $I$ with a decidable equality and an \emph{arity} function $\arity\colon I \to \mathbb{N}$.
@@
 \begin{definition}
-  The \emph{(simultaneous) type substitution} for a function $\rho\colon\Xi \to \Type_{\Sigma}(\Xi')$ is a map which sends $A : \Type_{\Sigma}(\Xi)$ to $\simsub{A}{\rho} : \Type_{\Sigma}(\Xi')$ and is defined as usual.
-  By abuse of notation, the substitution $\simsub{\Gamma}{\rho}$ of a context $\Gamma$ is defined by applying substitution $\simsub{A}{\rho}$ to each type $A$ in the context.
+  The \emph{substitution} for a function $\rho\colon \Xi to \Type_{\Sigma}(\Xi')$, denoted by $\rho\colon \Sub{\Xi}{\Xi'}$, is a map which sends a type $A$ of $\Type_{\Sigma}(\Xi)$ to $\simsub{A}{\rho}$ of $\Type_{\Sigma}(\Xi')$ and is defined as usual.
+  By abuse of notation, the substitution $\simsub{\Gamma}{\rho}$ of a context $\Gamma$ is defined by applying substitution to each $A$ in $\Gamma$.
 \end{definition}
 
-We may confine arbitrary variable set $\Xi$ to the family of sets $\Fin(n)$ of natural numbers less than $n$ as sets of type variables, following the inductive definition by \citet{Dybjer1994}, for the reasons below.
-In bidirectional type synthesis, it is necessary to have a decidable equality on $\Type_{\Sigma}(\Xi)$, as we need to compare the synthesized type $B$ with a given type $A$ for the $\Rule{Sub}$ rule.
-If $\Xi$ such as $\Fin(n)$ has a decidable equality, the decidable equality on $\Type_{\Sigma}(\Xi)$ can be given.
-In addition, the arity of a construct is always finite, so it suffices to use natural numbers to denote type variables in a type.
-Therefore, we formally restrict the set $\Xi$ of type variables to the family of $\Fin(n)$, but for illustrative purposes we still use named variables for presentation.
+We may confine the use of an arbitrary variable set $\Xi$ to some set in the family of sets $\Fin(n)$ of natural numbers less than $n$, as defined by \citet{Dybjer1994}, for the following reasons.
+In bidirectional type synthesis, a decidable equality on $\Type_{\Sigma}(\Xi)$ is necessary, as we need to compare the synthesised type $B$ with a given type $A$ for the $\Rule{Sub}$ rule.
+If $\Xi$ has a decidable equality as $\Fin(n)$ does, then $\Type_{\Sigma}(\Xi)$ also has a decidable equality.
+In addition, the arity of a construct is finite, so it suffices to use natural numbers to denote type variables in a type.
+As we will only consider a finite collection of types (in a typing rule or in a signature), we can formally restrict the set $\Xi$ to the family of $\Fin(n)$ in later definitions.
+For illustrative purposes, though, we still use named variables for presentation.
 \subsection{Binding Signatures and Simple Type Systems} \label{subsec:binding-sig}
@@
 \end{enumerate*}
-Therefore, to specify a construct like $\Rule{Abs}$ and $\Rule{App}$, we need to enrich the notion of signature with a type and an extension context for variable binding for each argument and a type for the operation:
+Therefore, to specify a term construct, we enrich the notion of arity with sorts and extension context for variable binding:
+\begin{definition}\label{def:binding-arity}
+  A \emph{binding arity} with a sort $T$ is an inhabitant of $\left(T^* \times T\right)^* \times T$ where $T^*$ is the set of lists over $T$.
+  In a binding arity ${[\Delta_1]A_{1}, \ldots, \left[\Delta_{n}\right] A_{n}} \to {A_0}$, every $\Delta_i$ and $A_i$ refers to the \emph{extension context} and the sort of the $i$-th argument, respectively; for brevity, we denote it as ${[\Delta_1]A_{1}, \ldots, \left[\Delta_{n}\right] A_{n}} \to {A_0}$ and omit $[\Delta_i]$ if $\Delta_i$ is empty.
+\end{definition}
+
+For example, the $\Rule{Abs}$ rule (\Cref{fig:STLC-typing-derivations}) has the binding arity with $T = \Type_{\Sigma}(\{A, B\})$:
+\[
+  {[A]B} \to {(A \bto B)}
+\]
+indicating that the $\Rule{Abs}$ rule for an abstraction $\lam{x}t$ has
+\begin{enumerate*}
+  \item an argument (i.e.\ a typing derivation of $\Gamma, x : A |- t : B$) of type $B$ in a context extended by a variable $x$ of type $A$;
+  \item the type $A\bto B$ for the conclusion $\Gamma |- \lam{x}t : A \bto B$ itself.
+\end{enumerate*}
+Similarly, the arity of the $\Rule{App}$ rule for an application $t\;u$ is ${(A \bto B), A} \to {B}$, indicating that it takes two arguments (i.e.\ typing derivations for $t$ and $u$) of types $A \bto B$ and $A$, respectively, with their extension contexts empty and the resulting type is $B$.
+
+Next, akin to a signature, our binding signature $\Omega$ consists of a set of operation symbols along with their respective binding arities.
+It also includes additionally the number $\Xi$ of type variables that appear in the binding arity with the sort $\Type_{\Sigma}(\Xi)$ for each operation:
 \begin{definition}\label{def:binding-signature}
@@
   \[
-    \mathit{ar}\colon O \to \sm{\Xi : \N} \left(\Cxt_\Sigma(\Xi) \times \Type_\Sigma(\Xi)\right)^* \times \Type_\Sigma(\Xi).
+    \arity(o) \colon O \to \sm{\Xi : \N} \left(\Cxt_\Sigma(\Xi) \times \Type_\Sigma(\Xi)\right)^* \times \Type_\Sigma(\Xi).
   \]
 \end{definition}
-Each inhabitant $o: O$, meant to represent a term construct $\tmOp_o$ in a simple type system, has a triple $\left(\Xi, \left[\left(\Delta_1; A_1\right), \ldots, \left(\Delta_{n}; A_{n}\right) \right], A_0\right)$ as its arity $\arity_o$ where
-\begin{enumerate}
-  \item $\Xi$ is the number of type variables that are allowed to appear in a typing rule, 
-  \item $\left[\left(\Delta_1; A_{1}\right), \ldots, \left(\Delta_{n}; A_{n}\right) \right]$ is a list of pairs for specifying arguments where
-    \begin{enumerate}
-      \item $\Delta_i : \Cxt_{\Sigma}(\Xi)$ is the \emph{extension context} consisting of types for bound variables and
-      \item $A_i : \Term_{\Sigma}(\Xi)$ is for the type of the $i$-th argument (i.e.\ premise) of $\tmOp_{o}$;
-    \end{enumerate}
-  \item $A_0 : \Type_\Sigma(\Xi)$ is for the type of the construct $\tmOp_o$ in the conclusion.
-\end{enumerate}
-For brevity, we write $\aritysymbol{o}{\Xi}{[\Delta_1]A_{1}, \ldots, \left[\Delta_{n}\right] A_{n}}{A_0}$ to indicate an operation with its arity and omit $[\Delta]$ if $\Delta$ is empty. 
-For example, the $\Rule{Abs}$ rule in \Cref{fig:STLC-typing-derivations} can be specified by the operation 
-\[
-  \aritysymbol{\mathsf{abs}}{(A, B)}{[A]B}{(A \bto B)}
-\]
-indicating that in the $\Rule{Abs}$ rule for a raw term $\lam{x}t$
-\begin{enumerate*}
-  \item there are \emph{two} type variables named $A$ and $B$;
-  \item an argument (i.e.\ a typing derivation of $\Gamma, x : A |- t : B$) of type $B$ in a context extended by a variable $x$ of type $A$;
-  \item the type $A\bto B$ for the conclusion.
-\end{enumerate*}
-Similarly, the $\Rule{App}$ rule for $t\;u$ can be specified by $\aritysymbol{\mathsf{app}}{(A, B)}{(A \bto B), A}{B}$ which takes two arguments (i.e.\ typing derivations for $t$ and $u$) under the same context so that their extension contexts are both empty.
-
-Also for brevity, a simply typed language $(\Sigma, \Omega)$ refers to the simply typed language specified by some type signature $\Sigma$ and a binding signature $\Omega$.
+Each inhabitant $o: O$ is meant to represent a term construct except the rules $\Rule{Var}$ and $\Rule{Anno}$ in a simple type system and is associated with a number $\Xi$ of type variables and a binding arity with sort $T = \Type_{\Sigma}(\Xi)$.
+As we will always consider binding arities with a number of type variables that appear in a typing rule, we simply call $\arity(o)$ the arity of $o$, denoted by $\aritysymbol{o}{\Xi}{[\Delta_1]A_1, \ldots, [\Delta_{n}]A_{n}}{A_0}$.
+By a \emph{simply typed language} $(\Sigma, \Omega)$, we mean a pair of a type signature $\Sigma$ and a binding signature $\Omega$.
 
@@
 \end{definition}
-Note that our simple type system is always syntax-directed, since each rule in \Cref{fig:raw-terms} corresponds exactly to one rule in \Cref{fig:extrinsic-typing}. 
-The $\Rule{Op}$ rule in both figures is a rule schema and gives rise to as many rules as there are operations $o : O$.
 
@@
 \subsection{Bidirectional Binding Signatures and Bidirectional Type Systems} \label{subsec:bidirectional-system}
-Finally, for a bidirectional type system, we have raw terms and typing derivations like a simple type system, but its typing judgements come in two forms $\Gamma |- t \syn A$ and $\Gamma |- t \chk A$.
-Observe that these two typing judgements are just one typing judgement $\Gamma |- t :^\dir{d} A$ indexed additionally by a \emph{mode} $d : \Mode$---which is either $\syn$ or $\chk$---so that $\Gamma |- t \syn A$ and $\Gamma |- t \chk A$ are just abbreviations of $\Gamma |- t :^{\syn} A$ and $\Gamma |- t :^{\chk} A$ respectively.
-Hence, to specify a bidirectional type system, we need further enrich the notion of binding signature with modes as follows.
+
+Finally, for a bidirectional type system, we have raw terms and typing derivations like a simple type system. 
+However, its typing judgements appear in two forms: $\Gamma |- t \syn A$ and $\Gamma |- t \chk A$.
+It is noteworthy that these two typing judgements can be considered as a single typing judgement $\Gamma |- t :^\dir{d} A$, additionally indexed by a \emph{mode} $d : \Mode$---which can either be $\syn$ or $\chk$.
+Therefore, to define a bidirectional type system, we need to further enrich the concept of binding arity by incorporating modes, as described below.
 
 \begin{definition} \label{def:bidirectional-binding-signature}
-  For a type signature $\Sigma$, a \emph{bidirectional binding signature} $\Omega$ is a set $O$ with a function
+  A \emph{bidirectional binding arity} with a set $T$ of sorts is an inhabitant of
+  \[
+    \left(T^* \times T \times \Mode \right)^* \times T \times \Mode.
+  \]
+  Similarly, a bidirectional binding arity is denoted by ${[\Delta_1]A_{1}^{\dir{d_1}}, \ldots, [\Delta_{n}] A^{\dir{d_n}}_{n}} \to {A_0^{\dir{d}}}$.
+\end{definition}
+Compared to a binding arity, a bidirectional binding arity specifies additionally the mode of each of its argument and the resulting mode.
+For example, the $\ChkRule{Abs}$ rule (\Cref{fig:STLC-bidirectional-typing-derivations}) has the arity ${[A]B^{\chk}}\to{(A \bto B)^{\chk}}$ with $T = \Type_{\Sigma}(\{A, B\})$ indicating additionally that $\lam{x}t$ and its argument $t$ are in checking mode.
+Similarly, the $\SynRule{App}$ rule has the arity ${(A \bto B)^{\syn}, A^{\chk}} \to {B^{\syn}}$.
+
+\begin{definition}
+  For a type signature $\Sigma$, a \emph{bidirectional binding signature} $\Omega$ is a set $O$ with
   \[
@@
 \end{definition}
-Each inhabitant $o:O$ is similarly meant to represent a term construct $\tmOp_o$ in a bidirectional type system but the arity of $o : O$ becomes a $4$-tuple $\left(\Xi, \left[\left(\Delta_1; A_1; d_1\right), \ldots, \left(\Delta_{n}; A_{n}; d_n\right) \right], A_0, d\right)$
-including additionally $d$ and $d_i$'s for the modes of the construct $\tmOp_o$ and its arguments respectively.
+Similar to a binding signature, each $o:O$ is meant to represent a term construct except $\SynRule{Var}$, $\SynRule{Anno}$, and $\ChkRule{Sub}$ in a bidirectional type system.
+We write $\aritysymbol{o}{\Xi}{[\Delta_1]A_{1}^{\dir{d_1}}, \ldots, [\Delta_{n}] A^{\dir{d_n}}_{n}}{A_0^{\dir{d}}}$ to indicate an operation $o$ with the number $\Xi$ of type variables and its arity with sort $\Type_{\Sigma}(\Xi)$.
+Finally, a bidirectional type system $(\Sigma, \Omega)$ refers to a pair of $\Sigma$ and a bidirectional binding signature $\Omega$.
 
-For brevity, we write $\aritysymbol{o}{\Xi}{[\Delta_1]A_{1}^{\dir{d_1}}, \ldots, [\Delta_{n}] A^{\dir{d_n}}_{n}}{A_0^{\dir{d}}}$ to indicate an operation with its arity. 
-For example, the $\ChkRule{Abs}$ rule in \Cref{fig:STLC-bidirectional-typing-derivations} can be specified by $\aritysymbol{\mathsf{abs}}{A , B}{[A]B^{\chk}}{(A \bto B)^{\chk}}$ indicating additionally that $\lam{x}t$ and its argument $t$ are in checking mode.
-Similarly, the $\SynRule{App}$ rule can be specified by $\aritysymbol{\mathsf{app}}{A, B}{(A \bto B)^{\syn}, A^{\chk}}{B^{\syn}}$.
-
-We also say a bidirectional type system $(\Sigma, \Omega)$ referring to the bidirectional type system specified by some type signature $\Sigma$ and a bidirectional binding signature $\Omega$.
-
-\begin{remark}
 Every bidirectional binding signature $\Omega$ gives rise to a binding signature $\erase{\Omega}$ if we erase modes from $\Omega$, so a bidirectional binding signature $\Omega$ also specifies a simple type system, including its raw terms and typing derivations.
 We call $\erase{\Omega}$ the \emph{(mode) erasure} of $\Omega$.
-\end{remark}
 
-We generalise bidirectional typing rules (\Cref{fig:STLC-bidirectional-typing-derivations}) as well as mode derivations (\Cref{fig:STLC-mode-derivations}) for simply typed $\lambda$-calculus to derivations for a bidirectional type system $(\Sigma, \Omega)$ where $\Omega$ is some bidirectional binding signature, similar to \Cref{def:typing-derivations} but with an additional $\ChkRule{Sub}$ rule:
 \begin{definition}\label{def:bidirectional-typing-derivations}\label{def:mode-derivations}

### 2023-07-06 tex/sec3-defs.tex
[ tex ] Revise S3; exactly 4 pages

@@
 
-\section{Simple Type Systems and Bidirectional Type Systems}\label{sec:defs}
+\section{Simple Typed Languages and Bidirectional Type Systems}\label{sec:defs}
 This section provides generic definitions
@@
 For illustrative purposes, though, we still use named variables for presentation.
-\subsection{Binding Signatures and Simple Type Systems} \label{subsec:binding-sig}
+\subsection{Binding Signatures and Simple Typed Languages} \label{subsec:binding-sig}
 
-A simple type system consists of
+A simple typed language consists of
 \begin{enumerate*}
@@
 \end{definition}
-Each inhabitant $o: O$ is meant to represent a term construct except the rules $\Rule{Var}$ and $\Rule{Anno}$ in a simple type system and is associated with a number $\Xi$ of type variables and a binding arity with sort $T = \Type_{\Sigma}(\Xi)$.
+Each inhabitant $o: O$ is meant to represent a term construct except the rules $\Rule{Var}$ and $\Rule{Anno}$ in a simply typed language and is associated with a number $\Xi$ of type variables and a binding arity with sort $\Type_{\Sigma}(\Xi)$.
 As we will always consider binding arities with a number of type variables that appear in a typing rule, we simply call $\arity(o)$ the arity of $o$, denoted by $\aritysymbol{o}{\Xi}{[\Delta_1]A_1, \ldots, [\Delta_{n}]A_{n}}{A_0}$.
@@
 \end{figure}
-  For a simple type system $(\Sigma, \Omega)$, the set of \emph{raw terms} indexed by a context~$V$ of free variables consists of
+  For a simply typed language $(\Sigma, \Omega)$, the set of \emph{raw terms} indexed by a context~$V$ of free variables consists of
   \begin{enumerate}
@@
   \end{figure}
-  For a simple type system $(\Sigma, \Omega)$, the set of \emph{typing derivations} of $\Gamma \vdash \isTerm{t} : A$ indexed by a context $\Gamma$, a raw term with free variables in $\erase{\Gamma}$, and a type $A$ consists of
+  For a simply typed language $(\Sigma, \Omega)$, the set of \emph{typing derivations} of $\Gamma \vdash \isTerm{t} : A$ indexed by a context $\Gamma$, a raw term with free variables in $\erase{\Gamma}$, and a type $A$ consists of
   \begin{enumerate}
@@
 \subsection{Bidirectional Binding Signatures and Bidirectional Type Systems} \label{subsec:bidirectional-system}
-
-Finally, for a bidirectional type system, we have raw terms and typing derivations like a simple type system. 
-However, its typing judgements appear in two forms: $\Gamma |- t \syn A$ and $\Gamma |- t \chk A$.
-It is noteworthy that these two typing judgements can be considered as a single typing judgement $\Gamma |- t :^\dir{d} A$, additionally indexed by a \emph{mode} $d : \Mode$---which can either be $\syn$ or $\chk$.
-Therefore, to define a bidirectional type system, we need to further enrich the concept of binding arity by incorporating modes, as described below.
+For a bidirectional type system, typing judgements appear in two forms: $\Gamma |- t \syn A$ and $\Gamma |- t \chk A$, but these two typing judgements can be considered as a single typing judgement $\Gamma |- t :^\dir{d} A$, additionally indexed by a \emph{mode} $d : \Mode$---which can either be $\syn$ or $\chk$.
+Therefore, to define a bidirectional type system, we further enrich the concept of binding arity by incorporating modes:
 
@@
 
-Every bidirectional binding signature $\Omega$ gives rise to a binding signature $\erase{\Omega}$ if we erase modes from $\Omega$, so a bidirectional binding signature $\Omega$ also specifies a simple type system, including its raw terms and typing derivations.
+Every bidirectional binding signature $\Omega$ gives rise to a binding signature $\erase{\Omega}$ if we erase modes from $\Omega$, so a bidirectional binding signature $\Omega$ also specifies a simply typed language, including its raw terms and typing derivations.
 We call $\erase{\Omega}$ the \emph{(mode) erasure} of $\Omega$.
@@
     \end{mathpar}
-    Abbreviations: $\boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \syn A} \defeq \boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} :^{\syn} A}$ and $\boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \chk A} \defeq \boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} :^{\chk} A}$
     \caption{Bidirectional typing derivations}
@@
   \begin{itemize}
-    \item the set of \emph{bidirectional typing derivations} of $\Gamma \vdash_{\Sigma, \Omega} t :^\dir{d} A$ indexed by a context $\Gamma$ of typed variables, a raw term $\isTerm{t}$ under $\erase{\Gamma}$, a mode $\dir{d}$, and a type $A$ is defined in \Cref{fig:bidirectional-typing-derivations} and
-      \begin{itemize}
-        \item in particular, a derivation of
+    \item the set of \emph{bidirectional typing derivations} of $\Gamma \vdash_{\Sigma, \Omega} t :^\dir{d} A$, indexed by a context $\Gamma$ of typed variables, a raw term $\isTerm{t}$ under $\erase{\Gamma}$, a mode $\dir{d}$, and a type $A$, is defined in \Cref{fig:bidirectional-typing-derivations} and particularly
           \[
@@
           \]
-          for $\aritysymbol{o}{\Xi}{[\Delta_1]A_{1}^{\dir{d_1}}, \ldots, [\Delta_{n}] A_{n}^{\dir{d_n}}}{A_0^{\dir{d}}}$ in $\Omega$ if there is a function $\rho\colon \Xi \to \Type_{\Sigma}(\emptyset)$ and a derivation of $\Gamma, \vars{x}_{\isTerm{i}} : \Delta_i \vdash_{\Sigma, \Omega} \isTerm{t_i} :^{\dir{d_i}} \simsub{A_i}{\rho}$ for each $i$;
+          has a derivation for $\aritysymbol{o}{\Xi}{[\Delta_1]A_{1}^{\dir{d_1}}, \ldots, [\Delta_{n}] A_{n}^{\dir{d_n}}}{A_0^{\dir{d}}}$ in $\Omega$ if there is a function $\rho\colon \Xi \to \Type_{\Sigma}(\emptyset)$ and a derivation of $\Gamma, \vars{x}_{\isTerm{i}} : \Delta_i \vdash_{\Sigma, \Omega} \isTerm{t_i} :^{\dir{d_i}} \simsub{A_i}{\rho}$ for each $i$;
+    \item the set of \emph{mode derivations} of $V |-_{\Sigma, \Omega} t^\dir{d}$, indexed by a list $V$ of variables, a raw term $\isTerm{t}$ under $V$, and a mode $d$, is defined in \Cref{fig:mode-derivations}.
   \end{itemize}
-    \item the set of \emph{mode derivations} of $V |-_{\Sigma, \Omega} t^\dir{d}$ indexed by a list $V$ of variables, a raw term $\isTerm{t}$ under $V$, and a mode $d$ is defined in \Cref{fig:mode-derivations}.
-  \end{itemize}
+  $\boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \syn A}$ and $\boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \chk A}$
+  are shorthands for ${\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} :^{\syn} A}$ and ${\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} :^{\chk} A}$ respectively.
 \end{definition}

### 2023-07-06 tex/sec4-pre-synthesis.tex
[ tex ] Revise S3; exactly 4 pages

@@
 
-\begin{lemma}
+\begin{lemma}\label{lem:soundness-completeness}
 There is a bidirectional typing derivation of\/ $\Gamma |-_{\Sigma, \Omega} t :^\dir{d} A$ if and only if there are mode and typing derivations of\/ $\erase\Gamma |-_{\Sigma, \Omega} t^\dir{d}$ and\/ $\Gamma |-_{\Sigma, \erase\Omega} t : A$.

### 2023-07-06 tex/sec5-synthesis.tex
[ tex ] Revise S3; exactly 4 pages

@@
 This section is devoted to defining mode-correctness and deriving bidirectional type synthesis for any mode-correct bidirectional type system $(\Sigma, \Omega)$.
-\Cref{subsec:mode-correctness} formalises the concept of mode-correctness and establishes the uniqueness of synthesised types.
+\Cref{sec:mode-correctness} formalises the concept of mode-correctness and establishes the uniqueness of synthesised types.
 This indicates that for a given raw term and context, any two synthesised types must be equal.
@@
 \begin{proof}
-  Combine  \Cref{thm:soundness,thm:completeness,thm:mode-preprocessing} with \Cref{thm:bidirectional-type-synthesis}.
+  Combine  \Cref{lem:soundness-completeness,thm:mode-preprocessing} with \Cref{thm:bidirectional-type-synthesis}.
 

### 2023-07-06 tex/sec1-intro.tex
[ tex ] Emphasise the philosophical implication of trichotomy in S1

@@
 Theoretically, separating the annotation and typing aspects streamlines our theory, in particular allowing us to formulate completeness naturally and differently from \emph{annotatability}, which was proposed and conflated with completeness by \citet{Dunfield2021}.
-By combining bidirectional type synthesis, soundness, completeness, and mode preprocessing, we achieve our main result~(\cref{cor:trichotomy}): a \emph{trichotomy} on raw terms, which is computationally a type synthesiser that checks if a given raw term is sufficiently annotated and if a sufficiently annotated raw term has an ordinary typing derivation.
+By combining bidirectional type synthesis, soundness, completeness, and mode preprocessing, we achieve our main result~(\cref{cor:trichotomy}): a \emph{trichotomy} on raw terms, which is computationally a type synthesiser that checks if a given raw term is sufficiently annotated and if a sufficiently annotated raw term has an ordinary typing derivation, suggesting that type synthesis should not be viewed as a binary division of raw terms between typable and non-typable but a trichotomy in general.
 

### 2023-07-06 tex/sec4-pre-synthesis.tex
[ tex ] Call Soundness and Completeness a theorem

--- 2023-07-06 tex/sec4-pre-synthesis.tex
@@
 %Formulating the notion of mode derivation helps to complete this clean picture.
-Technically, all these can be summarised in one lemma.
+Technically, all these can be summarised in one theorem.
 
-\begin{lemma}
+\begin{theorem}\label{lem:soundness-completeness}
 There is a bidirectional typing derivation of\/ $\Gamma |-_{\Sigma, \Omega} t :^\dir{d} A$ if and only if there are mode and typing derivations of\/ $\erase\Gamma |-_{\Sigma, \Omega} t^\dir{d}$ and\/ $\Gamma |-_{\Sigma, \erase\Omega} t : A$.
-\end{lemma}
+\end{theorem}
 

### 2023-07-06 tex/sec1-intro.tex
[ tex ] Minor changes

--- 2023-07-06 tex/sec1-intro.tex
@@
 Theoretically, separating the annotation and typing aspects streamlines our theory, in particular allowing us to formulate completeness naturally and differently from \emph{annotatability}, which was proposed and conflated with completeness by \citet{Dunfield2021}.
-By combining bidirectional type synthesis, soundness, completeness, and mode preprocessing, we achieve our main result~(\cref{cor:trichotomy}): a \emph{trichotomy} on raw terms, which is computationally a type synthesiser that checks if a given raw term is sufficiently annotated and if a sufficiently annotated raw term has an ordinary typing derivation.
+By combining bidirectional type synthesis, soundness, completeness, and mode preprocessing, we achieve our main result~(\cref{cor:trichotomy}): a \emph{trichotomy} on raw terms, which is computationally a type synthesiser that checks if a given raw term is sufficiently annotated and if a sufficiently annotated raw term has an ordinary typing derivation, suggesting that type synthesis should not be viewed as a bisection between typable and untypable raw terms but a trichotomy in general.
 

### 2023-07-06 tex/sec3-defs.tex
[ tex ] Minor changes

@@
 
-After developing these generic definitions, we are now able to reason about constructions and properties that hold for any bidirectional simple type system $(\Sigma, \Omega)$.
+After developing these general definitions, we are now able to reason about constructions and properties that hold for any bidirectional simple type system $(\Sigma, \Omega)$.

### 2023-07-07 tex/sec1-intro.tex
[ tex ] Revise S3; Use an aribtrary set for variables anyway

@@
 
-Within the theory, we formulate essential criteria including \emph{soundness}, \emph{completeness}, and \emph{mode-correctness} that are sufficient to derive a bidirectional type synthesiser and ensure that it implements the general specification, formalising what were only informally outlined by~\citet{Dunfield2021}.
+Within the theory, we formulate essential conditions including \emph{soundness}, \emph{completeness}, and \emph{mode-correctness} that are sufficient to derive a bidirectional type synthesiser and ensure that it implements the general specification, formalising what were only informally outlined by~\citet{Dunfield2021}.
 We also pay attention to the problem of whether a raw term has enough type annotations, often overlooked in the literature, by introducing \emph{mode preprocessing}, which assigns a mode to a raw term with enough annotations or otherwise pinpoints missing annotations.

### 2023-07-07 tex/sec3-defs.tex
[ tex ] Revise S3; Use an aribtrary set for variables anyway

@@
 
-\section{Simple Typed Languages and Bidirectional Type Systems}\label{sec:defs}
+\section{Simply Typed Languages and Bidirectional Type Systems}\label{sec:defs}
 This section provides generic definitions
@@
 \begin{enumerate*}
-  \item we first introduce a notion of signature which includes a set of operation symbols and an assignment of \emph{arities} to each operation;
-\item then, we define abstract syntax trees inductively with primitive constructs such as $\Rule{Var}$ and constructs specified by a given signature.
+  \item first we introduce a notion of arity and a notion of signature which includes a set of operation symbols and an assignment of arities to symbols;
+\item then, we define sets of terms or derivations inductively by primitives such as $\Rule{Var}$ and constructs $\tmOp_o$ for each symbol $o$ specified by a given signature.
 \end{enumerate*}
@@
 \begin{definition} \label{def:simple-signature}
-%  \begin{figure}
-%%    \begin{minipage}[b]{.6\textwidth}
+  \begin{figure}
+%    \begin{minipage}[b]{.6\textwidth}
+      \centering
+      \small
+      \judgbox{\Xi|-_{\Sigma} A}{$A$ is a well-formed type with type variables in $\Xi$}
+      \begin{mathpar}
+        \inferrule{\Xi \ni X_i}{\Xi |-_\Sigma X_i} \and
+        \inferrule{\Xi |-_{\Sigma} A_1 \\ \cdots \\ \Xi |-_{\Sigma} A_n}{\Xi |-_{\Sigma} \tyOp_i(A_1, \ldots, A_n)}\;\text{where $n = \arity(i)$}
+      \end{mathpar}
+      \caption{Simple types}
+      \label{fig:simple-type}
+%    \end{minipage}
+%    \begin{minipage}[b]{.35\textwidth}
 %      \centering
 %      \small
-%      \judgbox{\Xi|-_{\Sigma} A}{$A$ is a well-formed type with type variables in $\Xi$}
+%      \judgbox{\Xi |-_{\Sigma} \Gamma}{}
 %      \begin{mathpar}
-%        \inferrule{\Xi \ni X_i}{\Xi |-_\Sigma X_i} \and
-%        \inferrule{\Xi |-_{\Sigma} A_1 \\ \cdots \\ \Xi |-_{\Sigma} A_n}{\Xi |-_{\Sigma} \tyOp_i(A_1, \ldots, A_n)}\;\text{where $n = \arity(i)$}
+%        \inferrule{ }{\Xi |-_{\Sigma} \cdot }\and
+%        \inferrule{\Xi |-_{\Sigma} A \\ \Xi|-_{\Sigma} \Gamma}{\Xi |-_{\Sigma} \Gamma, x : A}
 %      \end{mathpar}
-%      \caption{Simple types}
-%      \label{fig:simple-type}
-%%    \end{minipage}
-%%    \begin{minipage}[b]{.35\textwidth}
-%%      \centering
-%%      \small
-%%      \judgbox{\Xi |-_{\Sigma} \Gamma}{}
-%%      \begin{mathpar}
-%%        \inferrule{ }{\Xi |-_{\Sigma} \cdot }\and
-%%        \inferrule{\Xi |-_{\Sigma} A \\ \Xi|-_{\Sigma} \Gamma}{\Xi |-_{\Sigma} \Gamma, x : A}
-%%      \end{mathpar}
-%%      \caption{Contexts}
-%%    \label{fig:simple-context}
-%%    \end{minipage}
-%  \end{figure}
+%      \caption{Contexts}
+%    \label{fig:simple-context}
+%    \end{minipage}
+  \end{figure}
   A \emph{signature} $\Sigma$ for simple types consists of a set\footnote{%
@@
     \item a variable in $\Xi$ or
-    \item $\tyOp_{i}(A_1, \ldots, A_n)$ for some $i:I$ with $\arity(i) = n$ and types $A_1,\ldots, A_n$.
+    \item $\tyOp_{i}(A_1, \ldots, A_n)$ for some $i:I$ with $\arity(i) = n$ and types $A_1,\ldots, A_n$,
   \end{enumerate}
-  %as defined in \Cref{fig:simple-type}.
+  as shown in \Cref{fig:simple-type}.
   A \emph{context} $\Gamma \colon \Cxt_{\Sigma}(\Xi)$ over $\Xi$ is a list $\Gamma$ of types over $\Xi$. 
@@
 \begin{definition}
-  The \emph{substitution} for a function $\rho\colon \Xi to \Type_{\Sigma}(\Xi')$, denoted by $\rho\colon \Sub{\Xi}{\Xi'}$, is a map which sends a type $A$ of $\Type_{\Sigma}(\Xi)$ to $\simsub{A}{\rho}$ of $\Type_{\Sigma}(\Xi')$ and is defined as usual.
+  The \emph{substitution} for a function $\rho\colon \Xi \to \Type_{\Sigma}(\Xi')$, denoted by $\rho\colon \Sub{\Xi}{\Xi'}$, is a map which sends a type $A$ of $\Type_{\Sigma}(\Xi)$ to $\simsub{A}{\rho}$ of $\Type_{\Sigma}(\Xi')$ and is defined as usual.
   By abuse of notation, the substitution $\simsub{\Gamma}{\rho}$ of a context $\Gamma$ is defined by applying substitution to each $A$ in $\Gamma$.
@@
 
-We may confine the use of an arbitrary variable set $\Xi$ to some set in the family of sets $\Fin(n)$ of natural numbers less than $n$, as defined by \citet{Dybjer1994}, for the following reasons.
-In bidirectional type synthesis, a decidable equality on $\Type_{\Sigma}(\Xi)$ is necessary, as we need to compare the synthesised type $B$ with a given type $A$ for the $\Rule{Sub}$ rule.
-If $\Xi$ has a decidable equality as $\Fin(n)$ does, then $\Type_{\Sigma}(\Xi)$ also has a decidable equality.
-In addition, the arity of a construct is finite, so it suffices to use natural numbers to denote type variables in a type.
-As we will only consider a finite collection of types (in a typing rule or in a signature), we can formally restrict the set $\Xi$ to the family of $\Fin(n)$ in later definitions.
-For illustrative purposes, though, we still use named variables for presentation.
-\subsection{Binding Signatures and Simple Typed Languages} \label{subsec:binding-sig}
+% We may confine the use of an arbitrary variable set $\Xi$ to some set in the family of sets $\Fin_n$ of natural numbers less than $n$, as defined by \citet{Dybjer1994}, for the following reasons.
+% In bidirectional type synthesis, a decidable equality on $\Type_{\Sigma}(\Xi)$ is necessary, as we need to compare the synthesised type $B$ with a given type $A$ for the $\Rule{Sub}$ rule.
+% If $\Xi$ has a decidable equality as $\Fin_n$ does, then $\Type_{\Sigma}(\Xi)$ also has a decidable equality.
+% In addition, the arity of a construct is finite, so it suffices to use natural numbers to denote type variables in a type.
+% As we will only consider a finite collection of types (in a typing rule or in a signature), we can formally restrict $\Xi$ to the family of $\Fin_n$ in later definitions.
+% For illustrative purposes, though, we still use named variables for presentation.
 
-A simple typed language consists of
+\subsection{Binding Signatures and Simply Typed Languages} \label{subsec:binding-sig}
+
+A simply typed language consists of
 \begin{enumerate*}
@@
 \end{enumerate*}
-Therefore, to specify a term construct, we enrich the notion of arity with sorts and extension context for variable binding:
+Therefore, to specify a term construct, we enrich the notion of arity with some sort and extension context for variable binding:
 \begin{definition}\label{def:binding-arity}
@@
 
-For example, the $\Rule{Abs}$ rule (\Cref{fig:STLC-typing-derivations}) has the binding arity with $T = \Type_{\Sigma}(\{A, B\})$:
+For example, the $\Rule{Abs}$ rule (\Cref{fig:STLC-typing-derivations}) has the binding arity with the sort $\Type_{\Sigma_{\bto}}\{A, B\}$:
 \[
@@
 Next, akin to a signature, our binding signature $\Omega$ consists of a set of operation symbols along with their respective binding arities.
-It also includes additionally the number $\Xi$ of type variables that appear in the binding arity with the sort $\Type_{\Sigma}(\Xi)$ for each operation:
+It also includes additionally the set $\Xi$ of type variables that appear in the binding arity with the sort $\Type_{\Sigma}(\Xi)$ for each operation:
 \begin{definition}\label{def:binding-signature}
@@
   \[
-    \arity(o) \colon O \to \sm{\Xi : \N} \left(\Cxt_\Sigma(\Xi) \times \Type_\Sigma(\Xi)\right)^* \times \Type_\Sigma(\Xi).
+    \arity(o) \colon O \to \sm{\Xi : \UU} \left(\Cxt_\Sigma(\Xi) \times \Type_\Sigma(\Xi)\right)^* \times \Type_\Sigma(\Xi).
   \]
 \end{definition}
-Each inhabitant $o: O$ is meant to represent a term construct except the rules $\Rule{Var}$ and $\Rule{Anno}$ in a simply typed language and is associated with a number $\Xi$ of type variables and a binding arity with sort $\Type_{\Sigma}(\Xi)$.
-As we will always consider binding arities with a number of type variables that appear in a typing rule, we simply call $\arity(o)$ the arity of $o$, denoted by $\aritysymbol{o}{\Xi}{[\Delta_1]A_1, \ldots, [\Delta_{n}]A_{n}}{A_0}$.
+That is, each inhabitant $o: O$  is associated with a set $\Xi$ of type variables and a binding arity with sort $\Type_{\Sigma}(\Xi)$.
+We may simply call $\arity(o)$ the arity of $o$, denoted by $\aritysymbol{o}{\Xi}{[\Delta_1]A_1, \ldots, [\Delta_{n}]A_{n}}{A_0}$.
 By a \emph{simply typed language} $(\Sigma, \Omega)$, we mean a pair of a type signature $\Sigma$ and a binding signature $\Omega$.
+
+The inclusion of a variable set $\Xi$ for each operation, called its \emph{local context}, plays a crucial role in type synthesis.
+For a typing rule like $\ChkRule{Abs}$, we will need to substitute type variables $A, B$ for types without any variables, but firstly, we must identify which type variables to substitute.
+As such, this information forms part of the arity of an operation, and typing derivations, defined subsequently, will include an instantiation used to assign concrete types to type variables.
 
@@
   For a simply typed language $(\Sigma, \Omega)$, the set of \emph{raw terms} indexed by a context~$V$ of free variables consists of
-  \begin{enumerate}
+  \begin{enumerate*}
     \item variables in $V$,
@@
     \item a construct $\tmOpts$ for some $\aritysymbol{o}{\Xi}{[\Delta_1]A_{1}, \ldots, [\Delta_{n}] A_{n}}{A_0}$ in $O$, where $\vars{x}_{\isTerm{i}}$'s are lists of variables whose length is equal to the length of~$\Delta_i$ and $t_i$'s are raw terms in the variable list $V, \vars{x}_i$
-  \end{enumerate}
+  \end{enumerate*}
   corresponding to rules $\Rule{Var}$, $\Rule{Anno}$, and $\Rule{Op}$ in \Cref{fig:raw-terms} respectively.
@@
 We need some information to compare types on the object level during type synthesis and substitute those type variables in a typing derivation $\Gamma \vdash \tmOpts : A$ for an operation $o$ in $\Omega$ at some point.
-Hence we choose to include a substitution as part of its typing derivation explicitly:
+Hence we choose to include a substitution $\rho$ from the local context $\Xi$ to $\emptyset$ as part of its typing derivation explicitly:
 \begin{definition}\label{def:typing-derivations}
@@
   \end{figure}
-  For a simply typed language $(\Sigma, \Omega)$, the set of \emph{typing derivations} of $\Gamma \vdash \isTerm{t} : A$ indexed by a context $\Gamma$, a raw term with free variables in $\erase{\Gamma}$, and a type $A$ consists of
+  For a simply typed language $(\Sigma, \Omega)$ the set of \emph{typing derivations} of $\Gamma \vdash \isTerm{t} : A$, indexed by a context $\Gamma$, a raw term with free variables in $\erase{\Gamma}$, and a type $A$, consists of
   \begin{enumerate}
@@
  \[
-   \aritysymbol{\mathsf{app}}{(A, B)}{(A \bto B), A}{B}
+   \aritysymbol{\mathsf{app}}{A, B}{(A \bto B), A}{B}
    \quad\text{and}\quad
-   \aritysymbol{\mathsf{abs}}{(A , B)}{[A]B}{(A \bto B)}.
+   \aritysymbol{\mathsf{abs}}{A , B}{[A]B}{(A \bto B)}.
  \]
@@
 \end{definition}
-Compared to a binding arity, a bidirectional binding arity specifies additionally the mode of each of its argument and the resulting mode.
-For example, the $\ChkRule{Abs}$ rule (\Cref{fig:STLC-bidirectional-typing-derivations}) has the arity ${[A]B^{\chk}}\to{(A \bto B)^{\chk}}$ with $T = \Type_{\Sigma}(\{A, B\})$ indicating additionally that $\lam{x}t$ and its argument $t$ are in checking mode.
+Compared to a binding arity, a bidirectional binding arity specifies additionally the mode of each of its argument and the mode of its conclusion.
+For example, the $\ChkRule{Abs}$ rule for an abstraction $\lam{x}t$ (\Cref{fig:STLC-bidirectional-typing-derivations}) has the arity ${[A]B^{\chk}}\to{(A \bto B)^{\chk}}$ with the sort $\Type_{\Sigma_{\bto}}\{A, B\}$ indicating additionally that $\lam{x}t$ and its argument $t$ are checking.
 Similarly, the $\SynRule{App}$ rule has the arity ${(A \bto B)^{\syn}, A^{\chk}} \to {B^{\syn}}$.
@@
   \[
-    \mathit{ar}\colon O \to \sum_{\Xi : \N} \left(\Cxt_{\Sigma}(\Xi) \times \Type_{\Sigma}(\Xi) \times {\Mode}\right)^* \times \Type_{\Sigma}(\Xi) \times {\Mode}.
+    \mathit{ar}\colon O \to \sum_{\Xi : \UU} \left(\Cxt_{\Sigma}(\Xi) \times \Type_{\Sigma}(\Xi) \times {\Mode}\right)^* \times \Type_{\Sigma}(\Xi) \times {\Mode}.
   \]
 \end{definition}
-Similar to a binding signature, each $o:O$ is meant to represent a term construct except $\SynRule{Var}$, $\SynRule{Anno}$, and $\ChkRule{Sub}$ in a bidirectional type system.
-We write $\aritysymbol{o}{\Xi}{[\Delta_1]A_{1}^{\dir{d_1}}, \ldots, [\Delta_{n}] A^{\dir{d_n}}_{n}}{A_0^{\dir{d}}}$ to indicate an operation $o$ with the number $\Xi$ of type variables and its arity with sort $\Type_{\Sigma}(\Xi)$.
-Finally, a bidirectional type system $(\Sigma, \Omega)$ refers to a pair of $\Sigma$ and a bidirectional binding signature $\Omega$.
-
-Every bidirectional binding signature $\Omega$ gives rise to a binding signature $\erase{\Omega}$ if we erase modes from $\Omega$, so a bidirectional binding signature $\Omega$ also specifies a simply typed language, including its raw terms and typing derivations.
-We call $\erase{\Omega}$ the \emph{(mode) erasure} of $\Omega$.
+We write $\aritysymbol{o}{\Xi}{[\Delta_1]A_{1}^{\dir{d_1}}, \ldots, [\Delta_{n}] A^{\dir{d_n}}_{n}}{A_0^{\dir{d}}}$ for an operation $o$ with a variable set $\Xi$ and its bidirectional binding arity with sort $\Type_{\Sigma}(\Xi)$.
+A bidirectional type system $(\Sigma, \Omega)$ refers to a pair of $\Sigma$ and a bidirectional binding signature $\Omega$.
+An operation $\aritysymbol{o}{\Xi}{[\Delta_1]A_{1}^{\dir{d_1}}, \ldots, [\Delta_{n}] A^{\dir{d_n}}_{n}}{A_0^{\dir{d}}}$ is checking if $d$ is ${\chk}$ or synthesis if $d$ is ${\syn}$. 
+Every bidirectional binding signature $\Omega$ gives rise to a binding signature $\erase{\Omega}$ if we erase modes from $\Omega$, called the \emph{(mode) erasure} of $\Omega$.
+Hence a bidirectional type system $(\Sigma, \Omega)$ also specifies a simply typed language $(\Sigma, \erase{\Omega})$, including raw terms and typing derivations.
 
@@
   \end{itemize}
-  $\boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \syn A}$ and $\boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \chk A}$
-  are shorthands for ${\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} :^{\syn} A}$ and ${\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} :^{\chk} A}$ respectively.
+  {\small$\boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \syn A}$} and {\small$\boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \chk A}$} are shorthands for ${\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} :^{\syn} A}$ and ${\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} :^{\chk} A}$ respectively.
+  A typing rule is \emph{checking} if its conclusion mode is $\chk$ and \emph{synthesis} if the conclusion mode is $\syn$.
 \end{definition}
 
-\begin{example}
-  Bidirectional typing derivations and mode derivations for simply typed $\lambda$-calculus can be specified by the type signature $\Sigma_{\bto}$ (\Cref{ex:type-signature-for-function-type}) and the bidirectional binding signature consisting of $\aritysymbol{\mathsf{abs}}{A , B}{[A]B^{\chk}}{(A \bto B)^{\chk}}$ and $\aritysymbol{\mathsf{app}}{A, B}{(A \bto B)^{\syn}, A^{\chk}}{B^{\syn}}$.
-\end{example}
-
-After developing these general definitions, we are now able to reason about constructions and properties that hold for any bidirectional simple type system $(\Sigma, \Omega)$.
+Having established generic definitions, we can now specify the simply typed $\lambda$-calculus and its bidirectional type system in \Cref{sec:key-ideas}---including raw terms, (bidirectional) typing derivations, and mode derivations---using just a pair of signatures $\Sigma_{\bto}$ and $\Omega^{\Leftrightarrow}_\Lambda$ where $\Omega^{\Leftrightarrow}_{\Lambda}$ consists of
+\[
+  \aritysymbol{\mathsf{abs}}{A , B}{[A]B^{\chk}}{(A \bto B)^{\chk}} \quad\text{and}\quad \aritysymbol{\mathsf{app}}{A, B}{(A \bto B)^{\syn}, A^{\chk}}{B^{\syn}}
+\]
+and $\Sigma_{\bto}$ is from \Cref{ex:type-signature-for-function-type}.
+Moreover, we are able to reason about constructions and properties that hold for any bidirectional simple type system $(\Sigma, \Omega)$.

### 2023-07-07 tex/sec4-pre-synthesis.tex
[ tex ] Shorten Thm 4.1

@@
 \begin{theorem}\label{lem:soundness-completeness}
-There is a bidirectional typing derivation of\/ $\Gamma |-_{\Sigma, \Omega} t :^\dir{d} A$ if and only if there are mode and typing derivations of\/ $\erase\Gamma |-_{\Sigma, \Omega} t^\dir{d}$ and\/ $\Gamma |-_{\Sigma, \erase\Omega} t : A$.
+$\Gamma |-_{\Sigma, \Omega} t :^\dir{d} A$ if and only if\/ $\erase\Gamma |-_{\Sigma, \Omega} t^\dir{d}$ and\/ $\Gamma |-_{\Sigma, \erase\Omega} t : A$.
 \end{theorem}

### 2023-07-07 tex/sec3-defs.tex
[ tex ] Explain mode-correctness

@@
 
-% We may confine the use of an arbitrary variable set $\Xi$ to some set in the family of sets $\Fin_n$ of natural numbers less than $n$, as defined by \citet{Dybjer1994}, for the following reasons.
-% In bidirectional type synthesis, a decidable equality on $\Type_{\Sigma}(\Xi)$ is necessary, as we need to compare the synthesised type $B$ with a given type $A$ for the $\Rule{Sub}$ rule.
-% If $\Xi$ has a decidable equality as $\Fin_n$ does, then $\Type_{\Sigma}(\Xi)$ also has a decidable equality.
-% In addition, the arity of a construct is finite, so it suffices to use natural numbers to denote type variables in a type.
-% As we will only consider a finite collection of types (in a typing rule or in a signature), we can formally restrict $\Xi$ to the family of $\Fin_n$ in later definitions.
-% For illustrative purposes, though, we still use named variables for presentation.
-
 \subsection{Binding Signatures and Simply Typed Languages} \label{subsec:binding-sig}
@@
   A \emph{binding arity} with a sort $T$ is an inhabitant of $\left(T^* \times T\right)^* \times T$ where $T^*$ is the set of lists over $T$.
-  In a binding arity ${[\Delta_1]A_{1}, \ldots, \left[\Delta_{n}\right] A_{n}} \to {A_0}$, every $\Delta_i$ and $A_i$ refers to the \emph{extension context} and the sort of the $i$-th argument, respectively; for brevity, we denote it as ${[\Delta_1]A_{1}, \ldots, \left[\Delta_{n}\right] A_{n}} \to {A_0}$ and omit $[\Delta_i]$ if $\Delta_i$ is empty.
+  In a binding arity ${[\Delta_1]A_{1}, \ldots, \left[\Delta_{n}\right] A_{n}} \to {A_0}$, every $\Delta_i$ and $A_i$ refers to the \emph{extension context} and the sort of the $i$-th argument, respectively; for brevity, we denote it as ${[\Delta_1]A_{1}, \ldots, \left[\Delta_{n}\right] A_{n}} \to {A_0}$ and omit $[\Delta_i]$ if the extension context $\Delta_i$ is empty.
 \end{definition}
@@
 Next, akin to a signature, our binding signature $\Omega$ consists of a set of operation symbols along with their respective binding arities.
-It also includes additionally the set $\Xi$ of type variables that appear in the binding arity with the sort $\Type_{\Sigma}(\Xi)$ for each operation:
+It also includes additionally the set $\Xi$ of type variables that may appear in the binding arity with the sort $\Type_{\Sigma}(\Xi)$ for each operation:
 \begin{definition}\label{def:binding-signature}

### 2023-07-07 tex/sec4-pre-synthesis.tex
[ tex ] Explain mode-correctness

@@
 \begin{theorem}\label{lem:soundness-completeness}
-$\Gamma |-_{\Sigma, \Omega} t :^\dir{d} A$ if and only if\/ $\erase\Gamma |-_{\Sigma, \Omega} t^\dir{d}$ and\/ $\Gamma |-_{\Sigma, \erase\Omega} t : A$.
+$\Gamma |-_{\Sigma, \Omega} t :^\dir{d} A$ if and only if both\/ $\erase\Gamma |-_{\Sigma, \Omega} t^\dir{d}$ and\/ $\Gamma |-_{\Sigma, \erase\Omega} t : A$.
 \end{theorem}

### 2023-07-07 tex/sec5-synthesis.tex
[ tex ] Explain mode-correctness

@@
 This indicates that for a given raw term and context, any two synthesised types must be equal.
-This characteristic is also used in the proof of the logical decidability of bidirectional type synthesis.
+This property will also be used in the proof of the logical decidability of bidirectional type synthesis.
 Later, in \Cref{subsec:trichotomy}, we introduce the decidability of bidirectional synthesis but postpone its proof until \Cref{subsec:bidirectional-synthesis-checking}.
 At that point, we will combine mode preprocessing, soundness, and completeness from \Cref{sec:pre-synthesis} to show the trichotomy on raw terms as a simple corollary.
-Note that the result is applicable to an arbitrary mode-correct bidirectional type system, rather than being restricted to the simply typed $\lambda$-calculus as we did in \Cref{sec:key-ideas}.
 
 \subsection{Mode Correctness}\label{sec:mode-correctness}
-As \citet[Section~3.1]{Dunfield2021} outlined, mode-correctness for a bidirectional type system means that for every bidirectional typing rule:
+As \citet{Dunfield2021} outlined, mode-correctness for a bidirectional type system means that for every typing rule:
 \begin{enumerate*}
-\item each `input' type variable in a premise must be synthesised from `earlier' premises or, if the mode of the conclusion is checking, provided by the conclusion;
-\item each `output' type variable in the typing judgement of the conclusion should have been synthesised from premises already, if the mode of the conclusion is synthesis.
+\item each `input' type variable in a premise must be synthesised from `earlier' premises or provided by the conclusion, if the rule is checking;
+\item each `output' type variable in the conclusion should have been synthesised from premises already, if the rule is synthesis.
 \end{enumerate*}
-An `input' type variable refers to types in an extension context and the type of a premise in the checking mode.
+Here an `input' type variable refers to types in an extension context and the type of a premise in the checking mode.
 It is important to note that the order of premises in a bidirectional typing rule matters, as it also sets the order of bidirectional type synthesis.
@@
 \item the type $A \bto B$ of the first argument $t$ is synthesised, thereby ensuring type variables $A$ and $B$ must be known if successfully synthesised;
-\item the type of the second argument $u$ is checked against $A$, which was synthesised earlier;
+\item the type of the second argument $u$ is checked against $A$, which has been synthesised earlier;
 \item as a result, the type of an application $t\;u$ can be synthesised.
@@
 As this is a condition for \emph{typing rules} and not for raw terms or typing derivations, this property \emph{cannot} be formalised without the concept of bidirectional binding signature presented in \Cref{sec:defs}.
-This contrasts with the properties in \Cref{sec:pre-synthesis}, which are properties that can still be expressed for individual systems even without the general definition of bidirectional type systems.
+This contrasts with the properties in \Cref{sec:pre-synthesis}, which are properties that can still be expressed for individual systems even without a general definition of bidirectional type systems.
 Now that we have the notion of signature available, we can articulate mode-correctness more precisely.
 
-First, let us introduce the finite subset\footnote{%
+But, first, we assume that the set $\Xi$ of variables has a \emph{decidable equality}, rather than being an arbitrary set, so that the set $\Type_{\Sigma}(\Xi)$ of types also has a decidable equality.
+This requirement for decidable equality is necessary, as type comparisons are ubiquitous throughout type synthesis.
+To simplify our choice, we could just restrict $\Xi$ to any set within the family of sets $\Fin_n$ of natural numbers less than $n$, as defined by \citet{Dybjer1994}.
+Given that this set has a decidable equality and the arity of a type construct is always finite, it would be a viable choice.
+For clarity in presentation, however, we will continue to use named variables.
+
+Now we proceed with the definition of mode-correctness.
+As we have outlined at the beginning, the condition of mode-correctness for a synthesis rule is different from that of a checking rule, and the argument order also matters.
+Defining the condition directly for a signature can be somewhat intricate.
+Hence, we instead choose to define the condition in a few more steps, based on the list of arguments---more specifically, triples of an extension context $\Delta_i$, a type $A_i$, and a mode $\dir{d_i}$---pertaining to an operation, and subsequently on a signature.
+We also need some auxiliary definitions for the subset of variables of a type and the set of variables that have been synthesised:
+\begin{definition}
+  The finite subset\footnote{%
   There are various definitions for finite subsets of a set within Martin-L\"{o}f type theory.
-  However, for our purposes, the choice among these definitions is not a matter of concern.
-}
-$\fv(A)$ and $\fv(\Gamma)$ of (free) variables in a type and in a context, respectively, which are defined in the usual way:
-\begin{enumerate*}
-\item For a type $A : \Type_{\Sigma}(\Xi)$ define the subset $\fv(A)$ of $\Xi$ inductively by $\fv(x) = {x}$ and $\fv(\tyOp_{i}(A_1, \ldots, A_n)) = \bigcup_{i = 1}^{n} \fv(A_i)$;
-\item by reusing this notation, for a context $\Gamma : \Type_{\Sigma}(\Xi)$ define the set $\fv(\Gamma)$ as $\fv(\emptyset) = \emptyset$ and $\fv(\Gamma, A) = \fv(\Gamma) \cup \fv(A)$.
-\end{enumerate*}
-As we have sketched informally, the mode-correctness condition for a synthesis rule is different from a checking rule
-\begin{definition}[Mode-Correctness] \label{def:mode-correctness}
-  An operation $\aritysymbol{o}{\Xi}{[\Delta_1]A_{1}^{\dir{d_1}}, \ldots, [\Delta_{n}] A_{n}^{\dir{d_n}}}{A_0^{\dir{d}}}$ is \emph{mode-correct} whenever
+  However, for our purposes, the choice among these definitions is not a matter of concern.}
+  of \emph{variables} which occur in a type $A$ is denoted by $\fv(A)$.
+  The finite subset $\fv(\Gamma)$ of variables in a context $\Gamma$ is defined inductively by $\fv(\cdot) = \emptyset$ and $\fv(\Gamma, A) = \fv(\Gamma) \cup \fv(A)$.
+  For a list of \emph{arguments} $\overrightarrow{[\Delta_i]A_i^{d_i}}$, define the set of \emph{synthesis type variables} inductively as 
+  \begin{align*}
+    \synvar(\cdot)                                  & = \emptyset  \\
+    \synvar(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}, [\Delta_{n}] A_{n}^{\dir{\chk}}) & = 
+    \synvar(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}) \\
+    \synvar(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}, [\Delta_{n}] A_{n}^{\dir{\syn}}) & = 
+    \synvar(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}) \cup \fv(A_{n}).
+  \end{align*}
+\end{definition}
+This subset accumulates type variables of a synthesising argument and they are exactly those type variables that would be synthesised during type synthesis.
+
+\begin{definition}\label{def:mode-correctness}
+  The \emph{mode-correctness for an argument list} $\Xi \rhd [\Delta_{1}] A_{1}^{\dir{d}}, \ldots, [\Delta_n]A_{n}^{\dir{d_n}}$  with respect to a subset $S$ of $\Xi$ is defined inductively as
+  \begin{align*}
+    \MC_{\mathit{as}}( \cdot ) ={} & \top \\
+    \MC_{\mathit{as}}(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}, [\Delta_{n}]A_{n}^{\syn}) 
+    & = \MC_{\mathit{as}}(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}})
+    \land \fv(\Delta) \subseteq \left( S \cup \synvar(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}) \right)\\
+    \MC_{\mathit{as}}(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}, [\Delta_{n}]A_{n}^{\chk})
+    & = \MC_{\mathit{as}}(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}})
+    \land \fv(\Delta, A_n) \subseteq \left( S \cup \synvar(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}})\right).
+  \end{align*}
+\end{definition}
+This definition encapsulates the idea that every `input' type variable, possibly deriving from a context $\Delta$ or a checking argument $A_n$, must be an 'output' variable from $\synvar(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}})$ or, if the rule is checking, belong to the set $S$ of 'input' variables in its conclusion.
+This condition must be met for every tail of the argument list to ensure that `output' variables accessible at each argument position are from preceding arguments only.
+\begin{definition}
+  An operation $o\colon \Xi \rhd [\Delta_{1}] A_{1}^{\dir{d}}, \ldots, [\Delta_n]A_{n}^{\dir{d_n}} \to A_0^{\dir{d}}$ is \emph{mode-correct} provided that
   \begin{enumerate}
-    \item if $d$ is $\syn$ (synthesis);
-    \item if $d$ is $\chk$ (checking);
+    \item $d$ is $\chk$, its argument list is mode-correct w.r.t.\ $\fv(A_0)$, and the union $\fv(A_0) \cup \synvar(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}})$ contains every inhabitant of $\Xi$;
+    \item $d$ is $\syn$, its argument list is mode-correct w.r.t.\ $\emptyset$, and the finite subset $\synvar(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}})$ contains every inhabitant of $\Xi$ and, particularly, $\fv(A_0)$.
   \end{enumerate}
-
+  A bidirectional binding signature $\Omega$ is \emph{mode-correct} if its operations are all mode-correct.
 \end{definition}
+Finally, for a checking operation, an 'input' variable of an argument could be derived from $A_0$, as these are known during type checking as an input.
+Since every inhabitant of $\Xi$ can be located in either $A_0$ or synthesis variables, we can determine a concrete type for each instance of $\Xi$ during type synthesis.
+On the other hand, for a synthesis operation, we do not have any known variables at the onset of type synthesis, so the argument list should be mode-correct with respect to $\emptyset$.
+Also, the set of synthesis variables alone should include every type variable in $\Xi$ and particularly in $A_n$.
 

### 2023-07-08 tex/sec5-synthesis.tex
[ tex ] Swap S5.1 and S5.2

@@
 \section{Bidirectional Type Synthesis and Checking} \label{sec:type-synthesis}
-This section is devoted to defining mode-correctness and deriving bidirectional type synthesis for any mode-correct bidirectional type system $(\Sigma, \Omega)$.
-\Cref{sec:mode-correctness} formalises the concept of mode-correctness and establishes the uniqueness of synthesised types.
-This indicates that for a given raw term and context, any two synthesised types must be equal.
-This property will also be used in the proof of the logical decidability of bidirectional type synthesis.
-Later, in \Cref{subsec:trichotomy}, we introduce the decidability of bidirectional synthesis but postpone its proof until \Cref{subsec:bidirectional-synthesis-checking}.
-At that point, we will combine mode preprocessing, soundness, and completeness from \Cref{sec:pre-synthesis} to show the trichotomy on raw terms as a simple corollary.
+This section focuses on defining mode-correctness and deriving bidirectional type synthesis for any mode-correct bidirectional type system $(\Sigma, \Omega)$.
+We start with presenting the statement of the decidability of bidirectional synthesis (whose proof will be deferred until \Cref{subsec:bidirectional-synthesis-checking}) and the trichotomy on raw terms in \Cref{subsec:trichotomy}.
+We will combine mode preprocessing, soundness, and completeness from \Cref{sec:pre-synthesis} to exhibit the trichotomy as a corollary.
+We will then formalise the mode-correctness condition and show the uniqueness of synthesized types in \Cref{sec:mode-correctness}.
+This uniqueness property means that any two synthesised types for the same raw term $t$ under the same context $\Gamma$ have to be equal.
+This will be used later in the proof of the decidability of bidirectional type synthesis and checking in \Cref{subsec:bidirectional-synthesis-checking}.
+
+\subsection{Trichotomy on Raw Terms by Type Synthesis} \label{subsec:trichotomy}
+
+\begin{theorem}[Decidability of Bidirectional Type Synthesis] \label{thm:bidirectional-type-synthesis}
+  In a mode-correct bidirectional type system $(\Sigma, \Omega)$,
+  it is decidable whether there exists some $A$ such that
+  \[
+    \Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn A.
+  \]
+\end{theorem}
+
+\begin{corollary}[Trichotomy on Raw Terms]\label{cor:trichotomy}
+  Exactly one of the following statements holds:
+  \begin{enumerate}
+    \item $\isTerm{t}$ is synthesising and the judgement $\Gamma |-_{\Sigma, \Omega} \isTerm{t} : A$ has a derivation for some type $A$.
+    \item $\isTerm{t}$ is synthesising but the judgement $\Gamma |-_{\Sigma, \Omega} \isTerm{t} : A$ has no derivation for any type $A$.
+    \item $\isTerm{t}$ is not synthesising.
+  \end{enumerate}
+\end{corollary}
+\begin{proof}
+  Combine  \Cref{lem:soundness-completeness,thm:mode-preprocessing} with \Cref{thm:bidirectional-type-synthesis}.
+
+\end{proof}
 
@@
 Here an `input' type variable refers to types in an extension context and the type of a premise in the checking mode.
-It is important to note that the order of premises in a bidirectional typing rule matters, as it also sets the order of bidirectional type synthesis.
+It is important to note that the order of premises in a bidirectional typing rule also matters, since synthesised types are accumulated during type synthesis.
 
@@
 \end{enumerate*}
-
-As this is a condition for \emph{typing rules} and not for raw terms or typing derivations, this property \emph{cannot} be formalised without the concept of bidirectional binding signature presented in \Cref{sec:defs}.
-This contrasts with the properties in \Cref{sec:pre-synthesis}, which are properties that can still be expressed for individual systems even without a general definition of bidirectional type systems.
-Now that we have the notion of signature available, we can articulate mode-correctness more precisely.
-
-But, first, we assume that the set $\Xi$ of variables has a \emph{decidable equality}, rather than being an arbitrary set, so that the set $\Type_{\Sigma}(\Xi)$ of types also has a decidable equality.
-This requirement for decidable equality is necessary, as type comparisons are ubiquitous throughout type synthesis.
-To simplify our choice, we could just restrict $\Xi$ to any set within the family of sets $\Fin_n$ of natural numbers less than $n$, as defined by \citet{Dybjer1994}.
-Given that this set has a decidable equality and the arity of a type construct is always finite, it would be a viable choice.
-For clarity in presentation, however, we will continue to use named variables.
 
@@
   of \emph{variables} which occur in a type $A$ is denoted by $\fv(A)$.
-  The finite subset $\fv(\Gamma)$ of variables in a context $\Gamma$ is defined inductively by $\fv(\cdot) = \emptyset$ and $\fv(\Gamma, A) = \fv(\Gamma) \cup \fv(A)$.
+  The subset $\fv(\Gamma)$ of variables in a context $\Gamma$ is defined by $\fv(\cdot) = \emptyset$ and $\fv(\Gamma, A) = \fv(\Gamma) \cup \fv(A)$.
   For a list of \emph{arguments} $\overrightarrow{[\Delta_i]A_i^{d_i}}$, define the set of \emph{synthesis type variables} inductively as 
@@
 \end{definition}
-This subset accumulates type variables of a synthesising argument and they are exactly those type variables that would be synthesised during type synthesis.
+This subset contains type variables of a synthesising argument and they are exactly those type variables that would be synthesised during type synthesis.
 
@@
     \MC_{\mathit{as}}( \cdot ) ={} & \top \\
-    \MC_{\mathit{as}}(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}, [\Delta_{n}]A_{n}^{\syn}) 
-    & = \MC_{\mathit{as}}(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}})
-    \land \fv(\Delta) \subseteq \left( S \cup \synvar(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}) \right)\\
-    \MC_{\mathit{as}}(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}, [\Delta_{n}]A_{n}^{\chk})
-    & = \MC_{\mathit{as}}(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}})
-    \land \fv(\Delta, A_n) \subseteq \left( S \cup \synvar(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}})\right).
+    \MC_{\mathit{as}}\left(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}, [\Delta_{n}]A_{n}^{\syn}\right) 
+    & = \MC_{\mathit{as}}\left(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}\right)
+    \land \fv(\Delta) \subseteq \left( S \cup \synvar\left(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}\right) \right)\\
+    \MC_{\mathit{as}}\left(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}, [\Delta_{n}]A_{n}^{\chk}\right)
+    & = \MC_{\mathit{as}}\left(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}\right)
+    \land \fv(\Delta, A_n) \subseteq \left( S \cup \synvar\left(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}\right)\right).
   \end{align*}
 \end{definition}
-This definition encapsulates the idea that every `input' type variable, possibly deriving from a context $\Delta$ or a checking argument $A_n$, must be an 'output' variable from $\synvar(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}})$ or, if the rule is checking, belong to the set $S$ of 'input' variables in its conclusion.
-This condition must be met for every tail of the argument list to ensure that `output' variables accessible at each argument position are from preceding arguments only.
+This definition encapsulates the idea that every `input' type variable, possibly deriving from a context $\Delta$ or a checking argument $A_n$, must be an `output' variable from $\synvar(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}})$ or, if the rule is checking, belong to the set $S$ of `input' variables in its conclusion.
+This condition must be met for every tail of the argument list as well to ensure that `output' variables accessible at each argument position are from preceding arguments only, hence an inductive definition.
 \begin{definition}
@@
 \end{definition}
-Finally, for a checking operation, an 'input' variable of an argument could be derived from $A_0$, as these are known during type checking as an input.
-Since every inhabitant of $\Xi$ can be located in either $A_0$ or synthesis variables, we can determine a concrete type for each instance of $\Xi$ during type synthesis.
+Finally, for a checking operation, an `input' variable of an argument could be derived from $A_0$, as these are known during type checking as an input.
+Since every inhabitant of $\Xi$ can be located in either $A_0$ or synthesis variables, we can determine a concrete type for each inhabitant of $\Xi$ during type synthesis.
 On the other hand, for a synthesis operation, we do not have any known variables at the onset of type synthesis, so the argument list should be mode-correct with respect to $\emptyset$.
@@
 
+\begin{remark}
+  As this is a condition for \emph{typing rules} and not for raw terms or typing derivations, this property cannot be formalised without the concept of bidirectional binding signature presented in \Cref{sec:defs}.
+  This contrasts with the properties in \Cref{sec:pre-synthesis}, which are properties that can still be expressed for individual systems even without a general definition of bidirectional type systems.
+  Now that we have the notion of signature available, we can articulate mode-correctness above.
+\end{remark}
+
 \begin{lemma}[Uniqueness of the Synthesised Types]\label{thm:unique-syn}
-  Let $(\Sigma, \Omega)$ be a mode-correct bidirectional type system.
-  Then, for every raw term $\isTerm{t}$ in synthesising mode and two derivations of 
+  In a mode-correct bidirectional type system $(\Sigma, \Omega)$, the synthesised types of any two derivations
   \[
@@
   \]
-  for some types $A$ and $B$ respectively, the synthesised types must be equal, i.e.\ $A = B$.
+  for the same term $t$ must be equal, i.e.\ $A = B$.
 \end{lemma}
 
-\subsection{Trichotomy on Raw Terms by Type Synthesis} \label{subsec:trichotomy}
+\subsection{Decidability of Bidirectional Type Synthesis and Checking}\label{subsec:bidirectional-synthesis-checking}
 
-\begin{theorem}[Decidability of Bidirectional Type Synthesis] \label{thm:bidirectional-type-synthesis}
-  Let $(\Sigma, \Omega)$ be a mode-correct bidirectional type system.
-  Then, every context $\Gamma$ and raw term $\isTerm{t}$ in synthesising mode decide whether
-  \[
-    \Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn A
-  \]
-  has a derivation for some type $A$.
-\end{theorem}
-
-\begin{corollary}[Trichotomy on Raw Terms]\label{cor:trichotomy}
-  Let $(\Sigma, \Omega)$ be a mode-correct bidirectional type system.
-  Then, for every context $\Gamma$ and raw term $\isTerm{t}$, exactly one of the following statements holds:
-  \begin{enumerate}
-    \item $\isTerm{t}$ is synthesising and the judgement $\Gamma |-_{\Sigma, \Omega} \isTerm{t} : A$ has a derivation for some type $A$.
-    \item $\isTerm{t}$ is synthesising but the judgement $\Gamma |-_{\Sigma, \Omega} \isTerm{t} : A$ has no derivation for any type $A$.
-    \item $\isTerm{t}$ is not synthesising.
-  \end{enumerate}
-\end{corollary}
-\begin{proof}
-  Combine  \Cref{lem:soundness-completeness,thm:mode-preprocessing} with \Cref{thm:bidirectional-type-synthesis}.
-
-\end{proof}
-
-\subsection{Decidability of Bidirectional Type Synthesis and Checking}\label{subsec:bidirectional-synthesis-checking}
+We assume further that the set $\Xi$ of variables has a \emph{decidable equality}, rather than being an arbitrary set, so that the set $\Type_{\Sigma}(\Xi)$ of types also has a decidable equality.
+This requirement for decidable equality is necessary, as type comparisons are ubiquitous throughout type synthesis.
+To simplify our choice, we could just restrict $\Xi$ to any set within the family of sets $\Fin_n$ of natural numbers less than $n$, as defined by \citet{Dybjer1994}.
+Given that this set has a decidable equality and the arity of a type construct is always finite, it would be a viable choice.
+For clarity in presentation, however, we will continue to use named variables.
 

### 2023-07-08 tex/sec5-synthesis.tex
[ tex ] S5

@@
 This section focuses on defining mode-correctness and deriving bidirectional type synthesis for any mode-correct bidirectional type system $(\Sigma, \Omega)$.
-We start with presenting the statement of the decidability of bidirectional synthesis (whose proof will be deferred until \Cref{subsec:bidirectional-synthesis-checking}) and the trichotomy on raw terms in \Cref{subsec:trichotomy}.
-We will combine mode preprocessing, soundness, and completeness from \Cref{sec:pre-synthesis} to exhibit the trichotomy as a corollary.
-We will then formalise the mode-correctness condition and show the uniqueness of synthesized types in \Cref{sec:mode-correctness}.
+We start with the statement of the decidability of bidirectional synthesis and the trichotomy on raw terms for mode-correct bidirectional type systems in \Cref{subsec:trichotomy}, but defer the definition of mode-correctness until \Cref{sec:mode-correctness} and the decidability proof until \Cref{subsec:bidirectional-synthesis-checking}.
+We will then formalise mode-correctness and show the uniqueness of synthesized types in \Cref{sec:mode-correctness}.
 This uniqueness property means that any two synthesised types for the same raw term $t$ under the same context $\Gamma$ have to be equal.

### 2023-07-08 tex/sec5-synthesis.tex
[ tex ] Shorten the statement of trichotomy

@@
 \begin{corollary}[Trichotomy on Raw Terms]\label{cor:trichotomy}
-  Exactly one of the following statements holds:
+  Exactly one of the following holds:
   \begin{enumerate}
-    \item $\isTerm{t}$ is synthesising and the judgement $\Gamma |-_{\Sigma, \Omega} \isTerm{t} : A$ has a derivation for some type $A$.
-    \item $\isTerm{t}$ is synthesising but the judgement $\Gamma |-_{\Sigma, \Omega} \isTerm{t} : A$ has no derivation for any type $A$.
+    \item $\isTerm{t}$ is synthesising and\/ $\Gamma |-_{\Sigma, \Omega} \isTerm{t} : A$ for some type $A$.
+    \item $\isTerm{t}$ is synthesising but\/ $\Gamma \not|-_{\Sigma, \Omega} \isTerm{t} : A$ for any type $A$.
     \item $\isTerm{t}$ is not synthesising.
@@
 Now we proceed with the definition of mode-correctness.
-As we have outlined at the beginning, the condition of mode-correctness for a synthesis rule is different from that of a checking rule, and the argument order also matters.
+As we have outlined, the condition of mode-correctness for a synthesis rule is different from that of a checking rule, and the argument order also matters.
 Defining the condition directly for a signature can be somewhat intricate.
-Hence, we instead choose to define the condition in a few more steps, based on the list of arguments---more specifically, triples of an extension context $\Delta_i$, a type $A_i$, and a mode $\dir{d_i}$---pertaining to an operation, and subsequently on a signature.
-We also need some auxiliary definitions for the subset of variables of a type and the set of variables that have been synthesised:
+Instead, we choose to define the condition in a few more steps, based on the list of arguments---more specifically, triples of an extension context $\Delta_i$, a type $A_i$, and a mode $\dir{d_i}$---pertaining to an operation, and subsequently on a signature.
+We also need some auxiliary definitions for the subset of variables of a type and of a context and the set of variables that have been synthesised:
 \begin{definition}
@@
 \end{lemma}
+
+\begin{proof}
+
+\end{proof}
 

### 2023-07-08 tex/sec5-synthesis.tex
[ tex ] Revise the statement of Trichotomy again

@@
 \begin{corollary}[Trichotomy on Raw Terms]\label{cor:trichotomy}
-  Exactly one of the following holds:
+  For any mode-correct bidirectional type system $(\Sigma, \Omega)$, 
+  exactly one of the following holds:
   \begin{enumerate}
-    \item $\isTerm{t}$ is synthesising and\/ $\Gamma |-_{\Sigma, \Omega} \isTerm{t} : A$ for some type $A$.
-    \item $\isTerm{t}$ is synthesising but\/ $\Gamma \not|-_{\Sigma, \Omega} \isTerm{t} : A$ for any type $A$.
-    \item $\isTerm{t}$ is not synthesising.
+    \item $\erase{\Gamma} |-_{\Sigma, \Omega} \isTerm{t}^{\syn}$ and\/ $\Gamma |-_{\Sigma, \Omega} \isTerm{t} : A$ for some type $A$.
+    \item $\erase{\Gamma} |-_{\Sigma, \Omega} \isTerm{t}^{\syn}$ but\/ $\Gamma \not|-_{\Sigma, \Omega} \isTerm{t} : A$.
+    \item $\erase{\Gamma} \not|-_{\Sigma, \Omega} \isTerm{t}^{\syn}$.
   \end{enumerate}

### 2023-07-08 tex/sec1-intro.tex
[ tex ] Remove some todos in S1

@@
 One of the core gadgets of these frameworks is the notion of \emph{binding signatures} (coined by~\citet{Aczel1978} in line with the term \emph{signature} in universal algebra) for specifying type systems.
-It is noteworthy that these state-of-the-art frameworks are basically restricted to simple types%
-\todo{confirm}
-and cannot specify polymorphic types or dependent types yet, and only \citeauthor{Allais2021} discussed meta-operations beyond substitution.
+It is noteworthy that these state-of-the-art frameworks are basically restricted to simple types and cannot specify polymorphic types or dependent types yet, and only \citeauthor{Allais2021} discussed meta-operations beyond substitution.
 Like these frameworks, our work is  built on a variant of binding signature for simple types.
 Unlike these frameworks, we target constructions for a different part of programming language theory, namely the transition from a surface language (raw terms) to a core language (well-typed terms), and our type synthesiser is significantly more complex than those meta-operations previously considered.
-Another difference is that several of the frameworks%
-\todo{Which ones precisely?}
-adopt intrinsically typed terms, whereas for our theory it is essential to use an extrinsic formulation to make it possible to relate a synthesised typing derivation with an input term.
+Another difference is that several of the frameworks~\citep{Fiore2022,Allais2021,Ahrens2022} focus on intrinsically typed terms, whereas for our theory it is essential to use an extrinsic formulation to make it possible to relate a synthesised typing derivation with an input term.
 This difference can be easily reconciled by converting extrinsic typing derivations to intrinsically typed terms though.

### 2023-07-08 tex/sec4-pre-synthesis.tex
[ tex ] Simplify the statement of mode preprocessing

@@
 \begin{corollary}[Mode Preprocessing]\label{thm:mode-preprocessing}
-  It is decidable for any raw term $V |-_{\Sigma,\erase\Omega} t$ and mode~$\dir{d}$ whether $V |-_{\Sigma,\Omega} \isTerm{t}^\dir{d}$.
+  It is decidable whether $V |-_{\Sigma,\Omega} \isTerm{t}^\dir{d}$.%for any raw term $V |-_{\Sigma,\erase\Omega} t$ and mode~$\dir{d}$
 \end{corollary}

### 2023-07-08 tex/sec3-defs.tex
[ tex ] Merging from the old master branch

@@
 of simple types, simply typed languages, and bidirectional type systems, using simply typed $\lambda$-calculus in \cref{sec:key-ideas} as our running example.
-All these definitions are inductive, which we will not emphasise in the individual definitions.
 
@@
   \item first we introduce a notion of arity and a notion of signature which includes a set of operation symbols and an assignment of arities to symbols;
-\item then, we define sets of terms or derivations inductively by primitives such as $\Rule{Var}$ and constructs $\tmOp_o$ for each symbol $o$ specified by a given signature.
+\item then, given a signature, we define sets of raw terms and derivations by primitive rules such as $\Rule{Var}$ and a rule schema for constructs $\tmOp_o$ indexed by a operational symbol~$o$.
 \end{enumerate*}
-Upon moving from simple types to bidirectionally typed terms, the notion of arity, initially as the number of arguments of an operation, is enriched to incorporate an extension context for variable binding and the direction of type information flow.
+All these definitions are inductive as usual but include a \emph{rule schema} giving rise to as many rules as there are operation symbols in a signature.
+Upon moving from simple types to bidirectional typing, the notion of arity, initially as the number of arguments of an operation, is enriched to incorporate an extension context for variable binding and the mode for the direction of type information flow.
 
 \subsection{Signatures and Simple Types} \label{subsec:simple-types}
-Simple types are understood intuitively as a collection of types whose formation judgement is not indexed by anything but a fixed set of type variables, and the only datum needed for specifying a type construct is its number of arguments:
 
+For simple types, the only datum needed for specifying a type construct is its number of arguments:
 \begin{definition} \label{def:simple-signature}
@@
 \end{enumerate*}
-Therefore, to specify a term construct, we enrich the notion of arity with some sort and extension context for variable binding:
+Therefore, to specify a term construct, we enrich the notion of arity with some sort for typing and extension context for variable binding:
 \begin{definition}\label{def:binding-arity}
   A \emph{binding arity} with a sort $T$ is an inhabitant of $\left(T^* \times T\right)^* \times T$ where $T^*$ is the set of lists over $T$.
-  In a binding arity ${[\Delta_1]A_{1}, \ldots, \left[\Delta_{n}\right] A_{n}} \to {A_0}$, every $\Delta_i$ and $A_i$ refers to the \emph{extension context} and the sort of the $i$-th argument, respectively; for brevity, we denote it as ${[\Delta_1]A_{1}, \ldots, \left[\Delta_{n}\right] A_{n}} \to {A_0}$ and omit $[\Delta_i]$ if the extension context $\Delta_i$ is empty.
+  In a binding arity ${[\Delta_1]A_{1}, \ldots, \left[\Delta_{n}\right] A_{n}} \to {A_0}$, every $\Delta_i$ and $A_i$ refers to the \emph{extension context} and the sort of the $i$-th argument, respectively; for brevity, it is denoted by ${[\Delta_1]A_{1}, \ldots, \left[\Delta_{n}\right] A_{n}} \to {A_0}$ and omit $[\Delta_i]$ if the extension context $\Delta_i$ is empty.
 \end{definition}
 
-For example, the $\Rule{Abs}$ rule (\Cref{fig:STLC-typing-derivations}) has the binding arity with the sort $\Type_{\Sigma_{\bto}}\{A, B\}$:
+For example, consider the $\Rule{Abs}$ rule (\Cref{fig:STLC-typing-derivations}).
+Its binding arity has the sort $\Type_{\Sigma_{\bto}}\{A, B\}$ and is
 \[
-  {[A]B} \to {(A \bto B)}
+  {[A]B} \to {(A \bto B)}.
 \]
-indicating that the $\Rule{Abs}$ rule for an abstraction $\lam{x}t$ has
+This means that the $\Rule{Abs}$ rule for derivations of\/ $\Gamma |- \lam{x}t : A \bto B$, contains:
 \begin{enumerate*}
-  \item an argument (i.e.\ a typing derivation of $\Gamma, x : A |- t : B$) of type $B$ in a context extended by a variable $x$ of type $A$;
-  \item the type $A\bto B$ for the conclusion $\Gamma |- \lam{x}t : A \bto B$ itself.
+  \item a derivation of\/ $\Gamma, x : A |- t : B$ as an argument of type $B$ in a \emph{context extended} by variable $x$ of type $A$;
+  \item the type $A\bto B$ for itself.
 \end{enumerate*}
-Similarly, the arity of the $\Rule{App}$ rule for an application $t\;u$ is ${(A \bto B), A} \to {B}$, indicating that it takes two arguments (i.e.\ typing derivations for $t$ and $u$) of types $A \bto B$ and $A$, respectively, with their extension contexts empty and the resulting type is $B$.
+In a similar vein, the arity of the $\Rule{App}$ rule is denoted as ${(A \bto B), A} \to {B}$, saying that any derivation of\/ $\Gamma \vdash t\;u : B$ accepts derivations $\Gamma \vdash t : A \bto B$ and $\Gamma \vdash u : A$ as its arguments.
+These arguments have types $A \bto B$ and $A$ respectively and their extension contexts are empty.
 
-Next, akin to a signature, our binding signature $\Omega$ consists of a set of operation symbols along with their respective binding arities.
-It also includes additionally the set $\Xi$ of type variables that may appear in the binding arity with the sort $\Type_{\Sigma}(\Xi)$ for each operation:
+Next, akin to a signature, a binding signature $\Omega$ consists of a set of operation symbols along with their respective binding arities.
+This also includes the set $\Xi$ of type variables that might be present in the binding arity, each operation is assigned the sort $\Type_{\Sigma}(\Xi)$:
 \begin{definition}\label{def:binding-signature}
@@
 \end{definition}
-That is, each inhabitant $o: O$  is associated with a set $\Xi$ of type variables and a binding arity with sort $\Type_{\Sigma}(\Xi)$.
+That is, each inhabitant $o: O$  is associated with a set $\Xi$ of type variables and a binding arity with the sort $\Type_{\Sigma}(\Xi)$.
 We may simply call $\arity(o)$ the arity of $o$, denoted by $\aritysymbol{o}{\Xi}{[\Delta_1]A_1, \ldots, [\Delta_{n}]A_{n}}{A_0}$.
@@
 The inclusion of a variable set $\Xi$ for each operation, called its \emph{local context}, plays a crucial role in type synthesis.
-For a typing rule like $\ChkRule{Abs}$, we will need to substitute type variables $A, B$ for types without any variables, but firstly, we must identify which type variables to substitute.
-As such, this information forms part of the arity of an operation, and typing derivations, defined subsequently, will include an instantiation used to assign concrete types to type variables.
+For a typing rule like $\ChkRule{Abs}$, we will need to substitute concrete types (i.e.\ types without any type variables) for variables $A, B$, but firstly, we must identify for which type variables to substitute.
+As such, this information forms part of the arity of an operation, and typing derivations, defined subsequently, will include an instantiation to assign concrete types to variables.
 
@@
 
-\begin{example}
 Raw terms (\Cref{fig:STLC-raw-terms}) and typing derivations (\Cref{fig:STLC-typing-derivations}) for simply typed $\lambda$-calculus can be specified by the type signature $\Sigma_{\bto}$ (\Cref{ex:type-signature-for-function-type}) and the binding signature consisting of
@@
  \]
- Rules $\Rule{Abs}$ and $\Rule{App}$ in simply typed $\lambda$-calculus are subsumed by the $\Rule{Op}$ rule schema, as an application $t\;u$ and an abstraction $\lam{x}t$ can be introduced as $\tmOp_{\mathsf{app}}(t, u)$ and $\tmOp_{\mathsf{abs}}(x.t)$, respectively.
-\end{example}
+Rules $\Rule{Abs}$ and $\Rule{App}$ in simply typed $\lambda$-calculus are subsumed by the $\Rule{Op}$ rule schema, as applications $t\;u$ and abstractions $\lam{x}t$ can be introduced uniformly as $\tmOp_{\mathsf{app}}(t, u)$ and $\tmOp_{\mathsf{abs}}(x.t)$, respectively.
 
@@
 \begin{definition} \label{def:bidirectional-binding-signature}
-  A \emph{bidirectional binding arity} with a set $T$ of sorts is an inhabitant of
+  A \emph{bidirectional binding arity} with a sort $T$ is an inhabitant of
   \[
@@
 \end{definition}
-Compared to a binding arity, a bidirectional binding arity specifies additionally the mode of each of its argument and the mode of its conclusion.
-For example, the $\ChkRule{Abs}$ rule for an abstraction $\lam{x}t$ (\Cref{fig:STLC-bidirectional-typing-derivations}) has the arity ${[A]B^{\chk}}\to{(A \bto B)^{\chk}}$ with the sort $\Type_{\Sigma_{\bto}}\{A, B\}$ indicating additionally that $\lam{x}t$ and its argument $t$ are checking.
-Similarly, the $\SynRule{App}$ rule has the arity ${(A \bto B)^{\syn}, A^{\chk}} \to {B^{\syn}}$.
+When compared to a binding arity, a bidirectional binding arity provides further details, specifying the mode for each of its arguments and for the conclusion.
+Take the $\ChkRule{Abs}$ rule (\Cref{fig:STLC-bidirectional-typing-derivations}) for $\lam{x}t$ as an example.
+It has the arity ${[A]B^{\chk}}\to{(A \bto B)^{\chk}}$, indicating additionally that both $\lam{x}t$ and its argument $t$ are checking.
+Likewise, the $\SynRule{App}$ rule has the arity ${(A \bto B)^{\syn}, A^{\chk}} \to {B^{\syn}}$.
 

### 2023-07-08 tex/sec5-synthesis.tex
[ tex ] Merging from the old master branch

@@
   For a list of \emph{arguments} $\overrightarrow{[\Delta_i]A_i^{d_i}}$, define the set of \emph{synthesis type variables} inductively as 
-  \begin{align*}
-    \synvar(\cdot)                                  & = \emptyset  \\
-    \synvar(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}, [\Delta_{n}] A_{n}^{\dir{\chk}}) & = 
+  \begin{align}
+    \label{eq:synvar1}\synvar(\cdot)                                  & = \emptyset  \\
+    \label{eq:synvar2}\synvar(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}, [\Delta_{n}] A_{n}^{\dir{\chk}}) & = 
     \synvar(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}) \\
-    \synvar(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}, [\Delta_{n}] A_{n}^{\dir{\syn}}) & = 
+    \label{eq:synvar3}\synvar(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}, [\Delta_{n}] A_{n}^{\dir{\syn}}) & = 
     \synvar(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}) \cup \fv(A_{n}).
-  \end{align*}
+  \end{align}
 \end{definition}
@@
   The \emph{mode-correctness for an argument list} $\Xi \rhd [\Delta_{1}] A_{1}^{\dir{d}}, \ldots, [\Delta_n]A_{n}^{\dir{d_n}}$  with respect to a subset $S$ of $\Xi$ is defined inductively as
-  \begin{align*}
-    \MC_{\mathit{as}}( \cdot ) ={} & \top \\
-    \MC_{\mathit{as}}\left(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}, [\Delta_{n}]A_{n}^{\syn}\right) 
+  \begin{align}
+    \label{eq:MC1} \MC_{\mathit{as}}( \cdot ) ={} & \top \\
+    \label{eq:MC2} \MC_{\mathit{as}}\left(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}, [\Delta_{n}]A_{n}^{\syn}\right) 
     & = \MC_{\mathit{as}}\left(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}\right)
     \land \fv(\Delta) \subseteq \left( S \cup \synvar\left(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}\right) \right)\\
-    \MC_{\mathit{as}}\left(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}, [\Delta_{n}]A_{n}^{\chk}\right)
+    \label{eq:MC3} \MC_{\mathit{as}}\left(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}, [\Delta_{n}]A_{n}^{\chk}\right)
     & = \MC_{\mathit{as}}\left(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}\right)
     \land \fv(\Delta, A_n) \subseteq \left( S \cup \synvar\left(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}\right)\right).
-  \end{align*}
+  \end{align}
 \end{definition}
@@
 
-\begin{remark}
-  As this is a condition for \emph{typing rules} and not for raw terms or typing derivations, this property cannot be formalised without the concept of bidirectional binding signature presented in \Cref{sec:defs}.
-  This contrasts with the properties in \Cref{sec:pre-synthesis}, which are properties that can still be expressed for individual systems even without a general definition of bidirectional type systems.
-  Now that we have the notion of signature available, we can articulate mode-correctness above.
-\end{remark}
-
 \begin{lemma}[Uniqueness of the Synthesised Types]\label{thm:unique-syn}
@@
 \end{lemma}
-
-\begin{proof}
-
+\begin{proof}%[Proof of \Cref{thm:unique-syn}]
+  We prove the statement by induction on derivations $d_1$ and $d_2$ for $\Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn A$ and $\Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn B$.
+  Our system is syntax-directed, so $d_1$ and $d_2$ must be derived from the same rule: 
+  \begin{itemize}
+    \item $\SynRule{Var}$ follows from the requirement that each variable in a context is unique.
+    \item $\SynRule{Anno}$ holds trivially, since the synthesised type $A$ is from the term $t \annotate A$ in question.
+    \item $\Rule{Op}$: Recall here that a derivation of\/ $\Gamma |- \tmOpts : A$ contains a substitution $\rho$ to concrete types from the local context $\Xi$ of an operation $o\colon \Xi \rhd [\Delta_{1}] A_{1}^{\dir{d}}, \ldots, [\Delta_n]A_{n}^{\dir{d_n}} \to A_0^{\syn}$.
+      To prove that any two derivations for the same $\tmOpts$ has the same synthesised type, it suffices to show that those substitutions $\rho_1$ and $\rho_2$ agree on $X = \synvar\left([\Delta_{1}] A_{1}^{\dir{d}}, \ldots, [\Delta_n]A_{n}^{\dir{d_n}}\right)$ by induction on the list:
+      \begin{itemize}
+        \item For the empty list, the statement is vacuously true by~\eqref{eq:synvar1}.
+        \item If $\dir{d_{i+1}}$ is checking, then it holds by induction hypothesis.
+        \item If $\dir{d_{i+1}}$ is synthesis, then $\simsub{\Delta_{i+1}}{\rho_1} = \simsub{\Delta_{i+1}}{\rho_2}$ by mode-correctness~\eqref{eq:MC2} and induction hypothesis (of the list).
+          Therefore, under the same context $\Gamma, \simsub{\Delta_{i+1}}{\rho_1} = \Gamma, \simsub{\Delta_{i+1}}{\rho_2}$ the term $t_{i+1}$ must have the same synthesised type $\simsub{A_{i+1}}{\rho_1} = \simsub{A_{i+1}}{\rho_1}$ by induction hypothesis (of the typing derivation), so $\rho_1$ and $\rho_2$ agree on $\fv(A_{i+1})$ in addition to $\synvar\left([\Delta_{1}] A_{1}^{\dir{d}}, \ldots, [\Delta_i]A_{i}^{\dir{d_i}}\right)$, as required for the case \eqref{eq:synvar3} in the definition of $\synvar$.
+      \end{itemize}
+  \end{itemize}
 \end{proof}
 
+Uniqueness above is a prevalent property in bidirectional type systems, although the specific proofs can vary depending on the constructs in the system, which can involve selecting synthesis arguments whose type variables include the variables of the conclusion.
+For instance, for a derivation of $\Gamma |- t\;u \syn B$, one would apply the hypothesis to the sub-derivation $\Gamma |- t \syn A \bto B$.
+On the other hand, our proof is based on mode-correctness and need not consider specific sub-derivations.
+
+\begin{remark}
+  Mode-correctness is fundamentally a condition for typing \emph{rules} themselves, not for derivations.
+  Thus, this property cannot be formalised without treating rules as some mathematical object, such as the notion of bidirectional binding signature presented in \Cref{sec:defs}.
+  This contrasts with the properties in \Cref{sec:pre-synthesis}, which can still be articulated for individual systems even in the absence of a general definition of bidirectional type systems.
+  With the concept of signature now at our disposal, we can express mode-correctness as defined formally above.
+\end{remark}
 \subsection{Decidability of Bidirectional Type Synthesis and Checking}\label{subsec:bidirectional-synthesis-checking}
 
-We assume further that the set $\Xi$ of variables has a \emph{decidable equality}, rather than being an arbitrary set, so that the set $\Type_{\Sigma}(\Xi)$ of types also has a decidable equality.
-This requirement for decidable equality is necessary, as type comparisons are ubiquitous throughout type synthesis.
+For bidirectional type synthesis, we assume further that the set $\Xi$ of variables has a \emph{decidable equality}, rather than being an arbitrary set, so that the set $\Type_{\Sigma}(\Xi)$ of types also has a decidable equality.
+This assumption is necessary, as type comparisons are ubiquitous throughout type synthesis. 
 To simplify our choice, we could just restrict $\Xi$ to any set within the family of sets $\Fin_n$ of natural numbers less than $n$, as defined by \citet{Dybjer1994}.
@@
 \begin{theorem}[Decidability of Bidirectional Type Synthesis and Checking] \label{thm:bidirectional-type-synthesis-checking}
-  Let $(\Sigma, \Omega)$ be a mode-correct bidirectional type system.
-  For every context $\Gamma$ and raw term $\isTerm{t}$ in mode $d$, 
+  In a mode-correct bidirectional type system $(\Sigma, \Omega)$,
   \begin{enumerate}
-    \item if $\isTerm{t}$ is synthesising, then it is decidable whether there is a type~$A$ and a derivation of
+    \item if $\erase{\Gamma} |-_{\Sigma, \erase{\Omega}} \isTerm{t}^{\syn}$, then it is decidable whether there is a type~$A$ such that
       \[
-        \Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn A.
+        \Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn A;
       \]
-    \item if $\isTerm{t}$ is checking, then every type $A$ decides whether there is a derivation of
+    \item if $\erase{\Gamma} |-_{\Sigma, \erase{\Omega}} \isTerm{t}^{\chk}$, then it is decidable for any $A$ whether
       \[
@@
 \begin{proof}[Proof of {\Cref{thm:bidirectional-type-synthesis}}]
-  We prove this statement by induction on the derivation of $|-_{\Sigma, \Omega} \isTerm{t}^{d}$ (instead of $\isTerm{t}$ alone).
+  We prove this statement by induction on the derivation of $|-_{\Sigma, \Omega} \isTerm{t}^{\dir{d}}$ (instead of $\isTerm{t}$ alone).
   The first two cases \SynRule{Var} and \SynRule{Anno} are straightforward and have nothing to do with mode-correctness directly.
@@
   The last case \Rule{Op} is the interesting part of this theorem.
-  \begin{description}
-    \item[\SynRule{Var}:] If $\isTerm{t}$ is a variable $\isTerm{x}$ in the context $\Gamma$, then there exists a type $A$ such that $\Gamma |- \isTerm{x} \syn A$.
+  \begin{itemize}
+    \item \SynRule{Var}: If $\isTerm{t}$ is a variable $\isTerm{x}$ in the context $\Gamma$, then there exists a type $A$ such that $\Gamma |- \isTerm{x} \syn A$.
       Otherwise, if $\isTerm{x}$ is not in $\Gamma$ but $\Gamma |- \isTerm{x} \syn A$ is derivable, then by inversion $\isTerm{x}$ must be in $\Gamma$, a contradiction.
-    \item[\SynRule{Anno}:] For a raw term of the form $\isTerm{t} \annotate A$ in synthesising mode, it is decidable that $\Gamma |- \isTerm{t} \chk A$ is derivable or not by induction hypothesis.
+    \item \SynRule{Anno}: For a raw term of the form $\isTerm{t} \annotate A$ in synthesising mode, it is decidable that $\Gamma |- \isTerm{t} \chk A$ is derivable or not by induction hypothesis.
       \begin{enumerate}
@@
 
-    \item[\ChkRule{Sub}:] If $t$ is in checking mode because of the subsumption rule, then by induction hypothesis it is decidable whether $\Gamma |- t \syn B$ for some $B$:
+    \item \ChkRule{Sub}: If $t$ is in checking mode because of the subsumption rule, then by induction hypothesis it is decidable whether $\Gamma |- t \syn B$ for some $B$:
       \begin{enumerate}
@@
 %      \]
-    \item[\Rule{Op}:]
+    \item \Rule{Op}:
       Suppose that $t$ is of them $\tmOpts$ for some mode-correct operation $o$ in $\Omega$.
@@
   \end{claim}
-  \end{description}
+  \end{itemize}
 \end{proof}

### 2023-07-08 tex/sec7-future.tex
[ tex ] Merging from the old master branch

@@
 Normalisation in its generic form still remains out of reach and recent advances in this topic are discussed in the doctoral thesis by~\citet{Valliappan2023}.
+
+\todo[inline]{We leave the problem of presenting bidirectional typing for more general and advanced language features as future work (such as those related to polymorphic types~\citep{Pierce2000,Peyton-Jones2007,Dunfield2013,Xie2018}), which requires advancement in language formalisation~(\cref{sec:language-formalisation}) that is orthogonal to our work.}
+
+%\subsubsection{Theories of abstract syntax with variable binding}
+%\label{sec:theory-of-syntax}
+%
+%The aforementioned frameworks except \citeauthor{Gheri2020}'s are at least inspired by \varcitet{Fiore1999}{'s} initial semantics for abstract syntax with variable binding using category theory.
+%The main idea is that the set of (untyped) abstract syntax trees for a language consists of
+%\begin{enumerate*}
+%  \item a family of sets $\Term_{\Gamma}$ of well-scoped terms under a context~$\Gamma$ with
+%  \item variable renaming for a function $\sigma\colon \Gamma \to \Delta$ between variables acting as a functorial map from $\Term_{\Gamma}$ to $\Term_{\Delta}$, i.e.\ a presheaf $\Term\colon \mathbb{F} \to \mathsf{Set}$, and
+%  \item an initial algebra $[\mathsf{v}, \mathsf{op}]$ on~$\Term$ given by the variable rule as a map $\mathsf{v}$ from the presheaf~$V$ of variables (i.e.\ the embedding $V\colon \mathbb{F} \hookrightarrow \mathsf{Set}$) to $\Term$ and other constructs as $\mathsf{op}\colon \mathbb{\Sigma}\Term \to \Term$ where $\mathbb{F}$ is the category of contexts, the functor $\mathbb{\Sigma}\colon \mathsf{Set}^\mathbb{F} \to \mathsf{Set}^\mathbb{F}$ encodes the arities of constructs, and the initiality amounts to structural recursion, i.e.\ \emph{term traversal}.
+%\end{enumerate*}
+%To put it succinctly, it is the free $\mathbb{\Sigma}$-algebra over the presheaf~$V$ of variables.
+%
+%Fortunately, constructing the initial algebra of terms in type theory boils down to defining an inductive type with a few constructors that align with the variable rule and a rule schema for language constructs specified by a signature~\citep{Fiore2022}.
+
+%Substitution is also modelled categorically, but it does not play a role in this paper.%
+%\todo{Reveal a bit about this paper?}
+
+%\begin{remark} \label{re:type-signature}
+%Most of existing theories treat types independently of terms, thereby excluding them from signatures.
+%To the best of our knowledge, the only exception to this approach is found in the work of~\citet{Arkor2020}, which incorporates signatures for both terms and types.
+%Interestingly, this inclusion is also critical for type synthesis for comparing a concrete type $N \bto N$ with an abstract type $A \bto B$, where $A$ and $B$ are type variables in a typing rule.
+%\end{remark}

### 2023-07-08 tex/sec3-defs.tex
[ tex ] Minor in S3

@@
 
-Having established generic definitions, we can now specify the simply typed $\lambda$-calculus and its bidirectional type system in \Cref{sec:key-ideas}---including raw terms, (bidirectional) typing derivations, and mode derivations---using just a pair of signatures $\Sigma_{\bto}$ and $\Omega^{\Leftrightarrow}_\Lambda$ where $\Omega^{\Leftrightarrow}_{\Lambda}$ consists of
+\begin{example}\label{ex:signature-simply-typed-lambda}
+Having established generic definitions, we can now specify the simply typed $\lambda$-calculus and its bidirectional type system---including raw terms, (bidirectional) typing derivations, and mode derivations---using just a pair of signatures $\Sigma_{\bto}$ (\Cref{ex:type-signature-for-function-type}) and $\Omega^{\Leftrightarrow}_\Lambda$ which consists of
 \[
-  \aritysymbol{\mathsf{abs}}{A , B}{[A]B^{\chk}}{(A \bto B)^{\chk}} \quad\text{and}\quad \aritysymbol{\mathsf{app}}{A, B}{(A \bto B)^{\syn}, A^{\chk}}{B^{\syn}}
+  \aritysymbol{\mathsf{abs}}{A , B}{[A]B^{\chk}}{(A \bto B)^{\chk}} \quad\text{and}\quad \aritysymbol{\mathsf{app}}{A, B}{(A \bto B)^{\syn}, A^{\chk}}{B^{\syn}}.
 \]
-and $\Sigma_{\bto}$ is from \Cref{ex:type-signature-for-function-type}.
+\end{example}
 Moreover, we are able to reason about constructions and properties that hold for any bidirectional simple type system $(\Sigma, \Omega)$.

### 2023-07-08 tex/sec6-formalisation.tex
[ add ] agda-lhs2tex.sty and sec6 (compilable even without lhs2tex)

--- 2023-06-16 tex/sec6-formalisation.tex
@@
 
+\documentclass[BiSig.tex]{subfiles}
+
+%% ODER: format ==         = "\mathrel{==}"
+%% ODER: format /=         = "\neq "
+%
+%
+\makeatletter
+\@ifundefined{lhs2tex.lhs2tex.sty.read}%
+  {\@namedef{lhs2tex.lhs2tex.sty.read}{}%
+   \newcommand\SkipToFmtEnd{}%
+   \newcommand\EndFmtInput{}%
+   \long\def\SkipToFmtEnd#1\EndFmtInput{}%
+  }\SkipToFmtEnd
+
+\newcommand\ReadOnlyOnce[1]{\@ifundefined{#1}{\@namedef{#1}{}}\SkipToFmtEnd}
+\usepackage{amstext}
+\usepackage{amssymb}
+\usepackage{stmaryrd}
+\DeclareFontFamily{OT1}{cmtex}{}
+\DeclareFontShape{OT1}{cmtex}{m}{n}
+  {<5><6><7><8>cmtex8
+   <9>cmtex9
+   <10><10.95><12><14.4><17.28><20.74><24.88>cmtex10}{}
+\DeclareFontShape{OT1}{cmtex}{m}{it}
+  {<-> ssub * cmtt/m/it}{}
+\newcommand{\texfamily}{\fontfamily{cmtex}\selectfont}
+\DeclareFontShape{OT1}{cmtt}{bx}{n}
+  {<5><6><7><8>cmtt8
+   <9>cmbtt9
+   <10><10.95><12><14.4><17.28><20.74><24.88>cmbtt10}{}
+\DeclareFontShape{OT1}{cmtex}{bx}{n}
+  {<-> ssub * cmtt/bx/n}{}
+\newcommand{\tex}[1]{\text{\texfamily#1}}	% NEU
+
+\newcommand{\Sp}{\hskip.33334em\relax}
+
+\newcommand{\Conid}[1]{\mathit{#1}}
+\newcommand{\Varid}[1]{\mathit{#1}}
+\newcommand{\anonymous}{\kern0.06em \vbox{\hrule\@width.5em}}
+\newcommand{\plus}{\mathbin{+\!\!\!+}}
+\newcommand{\bind}{\mathbin{>\!\!\!>\mkern-6.7mu=}}
+\newcommand{\rbind}{\mathbin{=\mkern-6.7mu<\!\!\!<}}% suggested by Neil Mitchell
+\newcommand{\sequ}{\mathbin{>\!\!\!>}}
+\renewcommand{\leq}{\leqslant}
+\renewcommand{\geq}{\geqslant}
+\usepackage{polytable}
+
+%mathindent has to be defined
+\@ifundefined{mathindent}%
+  {\newdimen\mathindent\mathindent\leftmargini}%
+  {}%
+
+\def\resethooks{%
+  \global\let\SaveRestoreHook\empty
+  \global\let\ColumnHook\empty}
+\newcommand*{\savecolumns}[1][default]%
+  {\g@addto@macro\SaveRestoreHook{\savecolumns[#1]}}
+\newcommand*{\restorecolumns}[1][default]%
+  {\g@addto@macro\SaveRestoreHook{\restorecolumns[#1]}}
+\newcommand*{\aligncolumn}[2]%
+  {\g@addto@macro\ColumnHook{\column{#1}{#2}}}
+
+\resethooks
+
+\newcommand{\onelinecommentchars}{\quad-{}- }
+\newcommand{\commentbeginchars}{\enskip\{-}
+\newcommand{\commentendchars}{-\}\enskip}
+
+\newcommand{\visiblecomments}{%
+  \let\onelinecomment=\onelinecommentchars
+  \let\commentbegin=\commentbeginchars
+  \let\commentend=\commentendchars}
+
+\newcommand{\invisiblecomments}{%
+  \let\onelinecomment=\empty
+  \let\commentbegin=\empty
+  \let\commentend=\empty}
+
+\visiblecomments
+
+\newlength{\blanklineskip}
+\setlength{\blanklineskip}{0.66084ex}
+
+\newcommand{\hsindent}[1]{\quad}% default is fixed indentation
+\let\hspre\empty
+\let\hspost\empty
+\newcommand{\NB}{\textbf{NB}}
+\newcommand{\Todo}[1]{$\langle$\textbf{To do:}~#1$\rangle$}
+
+\EndFmtInput
+\makeatother
+%
+%
+%
+%
+%
+%
+% This package provides two environments suitable to take the place
+% of hscode, called "plainhscode" and "arrayhscode". 
+%
+% The plain environment surrounds each code block by vertical space,
+% and it uses \abovedisplayskip and \belowdisplayskip to get spacing
+% similar to formulas. Note that if these dimensions are changed,
+% the spacing around displayed math formulas changes as well.
+% All code is indented using \leftskip.
+%
+% Changed 19.08.2004 to reflect changes in colorcode. Should work with
+% CodeGroup.sty.
+%
+\ReadOnlyOnce{polycode.fmt}%
+\makeatletter
+
+\newcommand{\hsnewpar}[1]%
+  {{\parskip=0pt\parindent=0pt\par\vskip #1\noindent}}
+
+% can be used, for instance, to redefine the code size, by setting the
+% command to \small or something alike
+\newcommand{\hscodestyle}{}
+
+% The command \sethscode can be used to switch the code formatting
+% behaviour by mapping the hscode environment in the subst directive
+% to a new LaTeX environment.
+
+\newcommand{\sethscode}[1]%
+  {\expandafter\let\expandafter\hscode\csname #1\endcsname
+   \expandafter\let\expandafter\endhscode\csname end#1\endcsname}
+
+% "compatibility" mode restores the non-polycode.fmt layout.
+
+\newenvironment{compathscode}%
+  {\par\noindent
+   \advance\leftskip\mathindent
+   \hscodestyle
+   \let\\=\@normalcr
+   \let\hspre\(\let\hspost\)%
+   \pboxed}%
+  {\endpboxed\)%
+   \par\noindent
+   \ignorespacesafterend}
+
+\newcommand{\compaths}{\sethscode{compathscode}}
+
+% "plain" mode is the proposed default.
+% It should now work with \centering.
+% This required some changes. The old version
+% is still available for reference as oldplainhscode.
+
+\newenvironment{plainhscode}%
+  {\hsnewpar\abovedisplayskip
+   \advance\leftskip\mathindent
+   \hscodestyle
+   \let\hspre\(\let\hspost\)%
+   \pboxed}%
+  {\endpboxed%
+   \hsnewpar\belowdisplayskip
+   \ignorespacesafterend}
+
+\newenvironment{oldplainhscode}%
+  {\hsnewpar\abovedisplayskip
+   \advance\leftskip\mathindent
+   \hscodestyle
+   \let\\=\@normalcr
+   \(\pboxed}%
+  {\endpboxed\)%
+   \hsnewpar\belowdisplayskip
+   \ignorespacesafterend}
+
+% Here, we make plainhscode the default environment.
+
+\newcommand{\plainhs}{\sethscode{plainhscode}}
+\newcommand{\oldplainhs}{\sethscode{oldplainhscode}}
+\plainhs
+
+% The arrayhscode is like plain, but makes use of polytable's
+% parray environment which disallows page breaks in code blocks.
+
+\newenvironment{arrayhscode}%
+  {\hsnewpar\abovedisplayskip
+   \advance\leftskip\mathindent
+   \hscodestyle
+   \let\\=\@normalcr
+   \(\parray}%
+  {\endparray\)%
+   \hsnewpar\belowdisplayskip
+   \ignorespacesafterend}
+
+\newcommand{\arrayhs}{\sethscode{arrayhscode}}
+
+% The mathhscode environment also makes use of polytable's parray 
+% environment. It is supposed to be used only inside math mode 
+% (I used it to typeset the type rules in my thesis).
+
+\newenvironment{mathhscode}%
+  {\parray}{\endparray}
+
+\newcommand{\mathhs}{\sethscode{mathhscode}}
+
+% texths is similar to mathhs, but works in text mode.
+
+\newenvironment{texthscode}%
+  {\(\parray}{\endparray\)}
+
+\newcommand{\texths}{\sethscode{texthscode}}
+
+% The framed environment places code in a framed box.
+
+\def\codeframewidth{\arrayrulewidth}
+\RequirePackage{calc}
+
+\newenvironment{framedhscode}%
+  {\parskip=\abovedisplayskip\par\noindent
+   \hscodestyle
+   \arrayrulewidth=\codeframewidth
+   \tabular{@{}|p{\linewidth-2\arraycolsep-2\arrayrulewidth-2pt}|@{}}%
+   \hline\framedhslinecorrect\\{-1.5ex}%
+   \let\endoflinesave=\\
+   \let\\=\@normalcr
+   \(\pboxed}%
+  {\endpboxed\)%
+   \framedhslinecorrect\endoflinesave{.5ex}\hline
+   \endtabular
+   \parskip=\belowdisplayskip\par\noindent
+   \ignorespacesafterend}
+
+\newcommand{\framedhslinecorrect}[2]%
+  {#1[#2]}
+
+\newcommand{\framedhs}{\sethscode{framedhscode}}
+
+% The inlinehscode environment is an experimental environment
+% that can be used to typeset displayed code inline.
+
+\newenvironment{inlinehscode}%
+  {\(\def\column##1##2{}%
+   \let\>\undefined\let\<\undefined\let\\\undefined
+   \newcommand\>[1][]{}\newcommand\<[1][]{}\newcommand\\[1][]{}%
+   \def\fromto##1##2##3{##3}%
+   \def\nextline{}}{\) }%
+
+\newcommand{\inlinehs}{\sethscode{inlinehscode}}
+
+% The joincode environment is a separate environment that
+% can be used to surround and thereby connect multiple code
+% blocks.
+
+\newenvironment{joincode}%
+  {\let\orighscode=\hscode
+   \let\origendhscode=\endhscode
+   \def\endhscode{\def\hscode{\endgroup\def\@currenvir{hscode}\\}\begingroup}
+   %\let\SaveRestoreHook=\empty
+   %\let\ColumnHook=\empty
+   %\let\resethooks=\empty
+   \orighscode\def\hscode{\endgroup\def\@currenvir{hscode}}}%
+  {\origendhscode
+   \global\let\hscode=\orighscode
+   \global\let\endhscode=\origendhscode}%
+
+\makeatother
+\EndFmtInput
+%
+%
+\ReadOnlyOnce{agda.fmt}%
+
+\RequirePackage[T1]{fontenc}
+\RequirePackage[utf8]{inputenc}
+\RequirePackage{amsfonts}
+
+\providecommand\mathbbm{\mathbb}
+
+% TODO: Define more of these ...
+\DeclareUnicodeCharacter{737}{\textsuperscript{l}}
+\DeclareUnicodeCharacter{8718}{\ensuremath{\blacksquare}}
+\DeclareUnicodeCharacter{8759}{::}
+\DeclareUnicodeCharacter{9669}{\ensuremath{\triangleleft}}
+\DeclareUnicodeCharacter{8799}{\ensuremath{\stackrel{\scriptscriptstyle ?}{=}}}
+\DeclareUnicodeCharacter{10214}{\ensuremath{\llbracket}}
+\DeclareUnicodeCharacter{10215}{\ensuremath{\rrbracket}}
+\DeclareUnicodeCharacter{27E6}{\ensuremath{\llbracket}}
+\DeclareUnicodeCharacter{27E7}{\ensuremath{\rrbracket}}
+\DeclareUnicodeCharacter{2200}{\ensuremath{\forall}}
+
+\DeclareUnicodeCharacter{2294}{\ensuremath{\sqcup}}
+\DeclareUnicodeCharacter{2080}{\ensuremath{_0}}
+\DeclareUnicodeCharacter{2081}{\ensuremath{_1}}
+\DeclareUnicodeCharacter{2082}{\ensuremath{_2}}
+\DeclareUnicodeCharacter{2083}{\ensuremath{_3}}
+\DeclareUnicodeCharacter{2084}{\ensuremath{_4}}
+
+\DeclareUnicodeCharacter{2115}{\ensuremath{\mathbb{N}}}
+\DeclareUnicodeCharacter{2236}{:}
+\DeclareUnicodeCharacter{2237}{\ensuremath{\mathrel{::}}}
+\DeclareUnicodeCharacter{03A3}{\ensuremath{\Sigma}}
+\DeclareUnicodeCharacter{039B}{\ensuremath{\Lambda}}
+\DeclareUnicodeCharacter{039E}{\ensuremath{\Xi}}
+
+\DeclareUnicodeCharacter{03B9}{\ensuremath{\iota}}
+\DeclareUnicodeCharacter{03BB}{\ensuremath{\lambda}}
+\DeclareUnicodeCharacter{03C0}{\ensuremath{\pi}}
+\DeclareUnicodeCharacter{03C3}{\ensuremath{\sigma}}
+\DeclareUnicodeCharacter{03C9}{\ensuremath{\omega}}
+
+\DeclareUnicodeCharacter{2032}{\ensuremath{\prime}}
+\DeclareUnicodeCharacter{2113}{\ensuremath{\ell}}
+\DeclareUnicodeCharacter{2207}{\ensuremath{\nabla}}
+\DeclareUnicodeCharacter{220B}{\ensuremath{\ni}}
+\DeclareUnicodeCharacter{2264}{\ensuremath{\leq}}
+\DeclareUnicodeCharacter{21D2}{\ensuremath{\Rightarrow}}
+\DeclareUnicodeCharacter{22A2}{\ensuremath{\vdash}}
+\DeclareUnicodeCharacter{22A4}{\ensuremath{\top}}
+\DeclareUnicodeCharacter{22A5}{\ensuremath{\bot}}
+
+\DeclareUnicodeCharacter{1D57}{\ensuremath{^t}}
+
+% TODO: This is in general not a good idea.
+\providecommand\textepsilon{$\epsilon$}
+\providecommand\textmu{$\mu$}
+
+%Actually, varsyms should not occur in Agda output.
+
+% TODO: Make this configurable. IMHO, italics doesn't work well
+% for Agda code.
+
+\renewcommand\Varid[1]{\mathord{\textsf{#1}}}
+\let\Conid\Varid
+\newcommand\Keyword[1]{\textsf{\textbf{#1}}}
+
+\EndFmtInput
+
+\begin{document}
+
 \section{Formalisation} \label{sec:formalisation}
-\LT{how to type Agda code without lhs2tex?}
 \begin{enumerate}
-  \item (Bidirectional) binding signature, functor, and terms (extrinsic typing, raw terms, raw terms in some mode) --- 3 pp
+  \item (Bidirectional) binding signature, functor, and terms (extrinsic typing, raw terms, raw terms in some mode) --- 2.5 pp
   \item Compare the induction principle and the \Agda proof of Soundness --- 2 pp
@@
   \item Examples including STLC (PCF), application in spine form --- 1p
+
 \end{enumerate}
 
-\LT{7 pages in total}
+\begin{figure}
+  \small
+  \begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{7}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{14}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{16}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{19}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{24}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[3]{}\Keyword{record}\;\Conid{ArgD}\;(\Conid{Ξ}\;\mathbin{:}\;\Conid{ℕ})\;\mathbin{:}\;\Conid{Set}\;\Keyword{where}{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\Keyword{field}{}\<[E]%
+\\
+\>[5]{}\hsindent{2}{}\<[7]%
+\>[7]{}\Varid{cxt}\;{}\<[14]%
+\>[14]{}\mathbin{:}\;\Conid{Cxt}\;\Conid{Ξ}{}\<[E]%
+\\
+\>[5]{}\hsindent{2}{}\<[7]%
+\>[7]{}\Varid{mode}\;{}\<[14]%
+\>[14]{}\mathbin{:}\;\Conid{Mode}{}\<[E]%
+\\
+\>[5]{}\hsindent{2}{}\<[7]%
+\>[7]{}\Varid{type}\;{}\<[14]%
+\>[14]{}\mathbin{:}\;\Conid{TExp}\;\Conid{Ξ}{}\<[E]%
+\\[\blanklineskip]%
+\>[3]{}\Conid{ArgsD}\;\mathbin{:}\;\Conid{ℕ}\;\Varid{→}\;\Conid{Set}{}\<[E]%
+\\
+\>[3]{}\Conid{ArgsD}\;\Conid{Ξ}\;\mathrel{=}\;\Conid{List}\;(\Conid{ArgD}\;\Conid{Ξ}){}\<[E]%
+\\[\blanklineskip]%
+\>[3]{}\Keyword{record}\;\Conid{ConD}\;\mathbin{:}\;\Conid{Set}\;\Keyword{where}{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\Keyword{constructor}\;\Varid{ι}{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\Keyword{field}{}\<[E]%
+\\
+\>[5]{}\hsindent{2}{}\<[7]%
+\>[7]{}\{\mskip1.5mu \Varid{vars}\mskip1.5mu\}\;{}\<[16]%
+\>[16]{}\mathbin{:}\;\Conid{ℕ}{}\<[E]%
+\\
+\>[5]{}\hsindent{2}{}\<[7]%
+\>[7]{}\Varid{mode}\;{}\<[16]%
+\>[16]{}\mathbin{:}\;\Conid{Mode}{}\<[E]%
+\\
+\>[5]{}\hsindent{2}{}\<[7]%
+\>[7]{}\Varid{type}\;{}\<[16]%
+\>[16]{}\mathbin{:}\;\Conid{TExp}\;{}\<[24]%
+\>[24]{}\Varid{vars}{}\<[E]%
+\\
+\>[5]{}\hsindent{2}{}\<[7]%
+\>[7]{}\Varid{args}\;{}\<[16]%
+\>[16]{}\mathbin{:}\;\Conid{ArgsD}\;\Varid{vars}{}\<[E]%
+\\[\blanklineskip]%
+\>[3]{}\Keyword{record}\;\Conid{Desc}\;\mathbin{:}\;\Conid{Set₁}\;\Keyword{where}{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\Keyword{constructor}\;\Varid{desc}{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\Keyword{field}{}\<[E]%
+\\
+\>[5]{}\hsindent{2}{}\<[7]%
+\>[7]{}\Conid{Op}\;{}\<[19]%
+\>[19]{}\mathbin{:}\;\Conid{Set}{}\<[E]%
+\\
+\>[5]{}\hsindent{2}{}\<[7]%
+\>[7]{}\{\kern-.9pt\vrule width .75pt height 7.125pt depth 1.975pt\kern-1.5pt\;\Varid{decOp}\;\kern-1.5pt\vrule width .75pt height 7.125pt depth 1.975pt\kern-.9pt\}\;{}\<[19]%
+\>[19]{}\mathbin{:}\;\Conid{DecEq}\;\Conid{Op}{}\<[E]%
+\\
+\>[5]{}\hsindent{2}{}\<[7]%
+\>[7]{}\Varid{rules}\;{}\<[19]%
+\>[19]{}\mathbin{:}\;\Conid{Op}\;\Varid{→}\;\Conid{ConD}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+  \caption{Definition of Bidirectional Binding Signature}  
+\end{figure}
+
+\end{document}

### 2023-07-08 tex/sec5-synthesis.tex
[ tex ] Decidiability proof for bidirectional type synthesis and checking before Op

@@
 This section focuses on defining mode-correctness and deriving bidirectional type synthesis for any mode-correct bidirectional type system $(\Sigma, \Omega)$.
+\LT{trichotomy first or last?}
 We start with the statement of the decidability of bidirectional synthesis and the trichotomy on raw terms for mode-correct bidirectional type systems in \Cref{subsec:trichotomy}, but defer the definition of mode-correctness until \Cref{sec:mode-correctness} and the decidability proof until \Cref{subsec:bidirectional-synthesis-checking}.
@@
 This will be used later in the proof of the decidability of bidirectional type synthesis and checking in \Cref{subsec:bidirectional-synthesis-checking}.
-
-\subsection{Trichotomy on Raw Terms by Type Synthesis} \label{subsec:trichotomy}
-
-\begin{theorem}[Decidability of Bidirectional Type Synthesis] \label{thm:bidirectional-type-synthesis}
-  In a mode-correct bidirectional type system $(\Sigma, \Omega)$,
-  it is decidable whether there exists some $A$ such that
-  \[
-    \Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn A.
-  \]
-\end{theorem}
-
-\begin{corollary}[Trichotomy on Raw Terms]\label{cor:trichotomy}
-  For any mode-correct bidirectional type system $(\Sigma, \Omega)$, 
-  exactly one of the following holds:
-  \begin{enumerate}
-    \item $\erase{\Gamma} |-_{\Sigma, \Omega} \isTerm{t}^{\syn}$ and\/ $\Gamma |-_{\Sigma, \Omega} \isTerm{t} : A$ for some type $A$.
-    \item $\erase{\Gamma} |-_{\Sigma, \Omega} \isTerm{t}^{\syn}$ but\/ $\Gamma \not|-_{\Sigma, \Omega} \isTerm{t} : A$.
-    \item $\erase{\Gamma} \not|-_{\Sigma, \Omega} \isTerm{t}^{\syn}$.
-  \end{enumerate}
-\end{corollary}
-\begin{proof}
-  Combine  \Cref{lem:soundness-completeness,thm:mode-preprocessing} with \Cref{thm:bidirectional-type-synthesis}.
-
-\end{proof}
 
@@
 
-Now we proceed with the definition of mode-correctness.
+Now we proceed with the formal definition of mode-correctness.
 As we have outlined, the condition of mode-correctness for a synthesis rule is different from that of a checking rule, and the argument order also matters.
@@
   \begin{align}
-    \label{eq:MC1} \MC_{\mathit{as}}( \cdot ) ={} & \top \\
+    \label{eq:MC1} \MC_{\mathit{as}}( \cdot ) & = \top \\
     \label{eq:MC2} \MC_{\mathit{as}}\left(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}, [\Delta_{n}]A_{n}^{\syn}\right) 
@@
     & = \MC_{\mathit{as}}\left(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}\right)
-    \land \fv(\Delta, A_n) \subseteq \left( S \cup \synvar\left(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}\right)\right).
+    \land \fv(\Delta, A_n) \subseteq \left( S \cup \synvar\left(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}\right)\right)
   \end{align}
+  where \eqref{eq:MC1} means an empty list is always mode-correct.
 \end{definition}
@@
 \end{definition}
-Finally, for a checking operation, an `input' variable of an argument could be derived from $A_0$, as these are known during type checking as an input.
+For a checking operation, an `input' variable of an argument could be derived from $A_0$, as these are known during type checking as an input.
 Since every inhabitant of $\Xi$ can be located in either $A_0$ or synthesis variables, we can determine a concrete type for each inhabitant of $\Xi$ during type synthesis.
@@
 Also, the set of synthesis variables alone should include every type variable in $\Xi$ and particularly in $A_n$.
+It is also easy to check the bidirectional type system $(\Sigma_{\bto}, \Omega^{\Leftrightarrow}_{\Lambda}$) in \Cref{ex:signature-simply-typed-lambda} for simply typed $\lambda$-calculus is mode-correct according to our definition.
+\LT{perhaps we shall prove that mode-correctness is decidable}
 
+\begin{remark}
+  Mode-correctness is fundamentally a condition for typing \emph{rules} themselves, not for derivations.
+  Thus, this property cannot be formalised without treating rules as some mathematical object, such as the notion of bidirectional binding signature presented in \Cref{sec:defs}.
+  This contrasts with the properties in \Cref{sec:pre-synthesis}, which can still be articulated for individual systems even in the absence of a general definition of bidirectional type systems.
+\end{remark}
+
+Now, we set out to show the uniqueness of the synthesised types for a mode-correct bidirectional type system.
+For a specific system, its proof is typically a straightforward induction.
+However, given the mode-correctness---which incorporates an inductive definition on the argument list---we establish our proof through induction, both on typing derivations and on the argument list:
 \begin{lemma}[Uniqueness of the Synthesised Types]\label{thm:unique-syn}
@@
   \begin{itemize}
-    \item $\SynRule{Var}$ follows from the requirement that each variable in a context is unique.
+    \item $\SynRule{Var}$ follows from that each variable as a raw term refers to the same variable in its context.
     \item $\SynRule{Anno}$ holds trivially, since the synthesised type $A$ is from the term $t \annotate A$ in question.
-    \item $\Rule{Op}$: Recall here that a derivation of\/ $\Gamma |- \tmOpts : A$ contains a substitution $\rho$ to concrete types from the local context $\Xi$ of an operation $o\colon \Xi \rhd [\Delta_{1}] A_{1}^{\dir{d}}, \ldots, [\Delta_n]A_{n}^{\dir{d_n}} \to A_0^{\syn}$.
-      To prove that any two derivations for the same $\tmOpts$ has the same synthesised type, it suffices to show that those substitutions $\rho_1$ and $\rho_2$ agree on $X = \synvar\left([\Delta_{1}] A_{1}^{\dir{d}}, \ldots, [\Delta_n]A_{n}^{\dir{d_n}}\right)$ by induction on the list:
-      \begin{itemize}
+    \item $\Rule{Op}$: Recall here that a derivation of\/ $\Gamma |- \tmOpts \syn A$ for a synthesis operation $o\colon \Xi \rhd [\Delta_{1}] A_{1}^{\dir{d}}, \ldots, [\Delta_n]A_{n}^{\dir{d_n}} \to A_0^{\syn}$ contains a substitution $\rho$ from the local context $\Xi$ to concrete types \Cref{fig:extrinsic-typing}.
+      To prove that any two typing derivations has the same synthesised type, it suffices to show that those substitutions $\rho_1$ and $\rho_2$ agree on variables in $\synvar\left([\Delta_{1}] A_{1}^{\dir{d}}, \ldots, [\Delta_n]A_{n}^{\dir{d_n}}\right)$ so that $\simsub{A_0}{\rho_1} = \simsub{A_0}{\rho_2}$.
+      We prove it by induction on the argument list:
+      \begin{enumerate}
         \item For the empty list, the statement is vacuously true by~\eqref{eq:synvar1}.
-        \item If $\dir{d_{i+1}}$ is checking, then it holds by induction hypothesis.
+        \item If $\dir{d_{i+1}}$ is checking, then the statement holds for~\eqref{eq:synvar2} by induction hypothesis.
         \item If $\dir{d_{i+1}}$ is synthesis, then $\simsub{\Delta_{i+1}}{\rho_1} = \simsub{\Delta_{i+1}}{\rho_2}$ by mode-correctness~\eqref{eq:MC2} and induction hypothesis (of the list).
           Therefore, under the same context $\Gamma, \simsub{\Delta_{i+1}}{\rho_1} = \Gamma, \simsub{\Delta_{i+1}}{\rho_2}$ the term $t_{i+1}$ must have the same synthesised type $\simsub{A_{i+1}}{\rho_1} = \simsub{A_{i+1}}{\rho_1}$ by induction hypothesis (of the typing derivation), so $\rho_1$ and $\rho_2$ agree on $\fv(A_{i+1})$ in addition to $\synvar\left([\Delta_{1}] A_{1}^{\dir{d}}, \ldots, [\Delta_i]A_{i}^{\dir{d_i}}\right)$, as required for the case \eqref{eq:synvar3} in the definition of $\synvar$.
-      \end{itemize}
+      \end{enumerate}
   \end{itemize}
@@
 
-Uniqueness above is a prevalent property in bidirectional type systems, although the specific proofs can vary depending on the constructs in the system, which can involve selecting synthesis arguments whose type variables include the variables of the conclusion.
-For instance, for a derivation of $\Gamma |- t\;u \syn B$, one would apply the hypothesis to the sub-derivation $\Gamma |- t \syn A \bto B$.
-On the other hand, our proof is based on mode-correctness and need not consider specific sub-derivations.
+%Uniqueness of the synthesised types is a prevalent property in bidirectional type systems, although the specific proofs can vary depending on the constructs in the system.
+%For instance, for derivations of $\Gamma |- t\;u \syn B_i$ for $i = 1, 2$ in simply typed $\lambda$-calculus, the hypothesis is applied to their sub-derivations $\Gamma |- t \syn A_i \bto B_i$ to conclude that $A_1 \bto B_1 = A_2 \bto B_2$ and thus $B_1 = B_2$.
+%On the other hand, our proof is based on mode-correctness and need not consider specific sub-derivations.
 
-\begin{remark}
-  Mode-correctness is fundamentally a condition for typing \emph{rules} themselves, not for derivations.
-  Thus, this property cannot be formalised without treating rules as some mathematical object, such as the notion of bidirectional binding signature presented in \Cref{sec:defs}.
-  This contrasts with the properties in \Cref{sec:pre-synthesis}, which can still be articulated for individual systems even in the absence of a general definition of bidirectional type systems.
-  With the concept of signature now at our disposal, we can express mode-correctness as defined formally above.
-\end{remark}
 \subsection{Decidability of Bidirectional Type Synthesis and Checking}\label{subsec:bidirectional-synthesis-checking}
-
-For bidirectional type synthesis, we assume further that the set $\Xi$ of variables has a \emph{decidable equality}, rather than being an arbitrary set, so that the set $\Type_{\Sigma}(\Xi)$ of types also has a decidable equality.
-This assumption is necessary, as type comparisons are ubiquitous throughout type synthesis. 
-To simplify our choice, we could just restrict $\Xi$ to any set within the family of sets $\Fin_n$ of natural numbers less than $n$, as defined by \citet{Dybjer1994}.
-Given that this set has a decidable equality and the arity of a type construct is always finite, it would be a viable choice.
-For clarity in presentation, however, we will continue to use named variables.
-
+We are ready for the last technical result of this paper---the decidability of bidirectional type synthesis and checking for any mode correct bidirectional type system:
 \begin{theorem}[Decidability of Bidirectional Type Synthesis and Checking] \label{thm:bidirectional-type-synthesis-checking}
@@
   \begin{enumerate}
-    \item if $\erase{\Gamma} |-_{\Sigma, \erase{\Omega}} \isTerm{t}^{\syn}$, then it is decidable whether there is a type~$A$ such that
+    \item if $\erase{\Gamma} |-_{\Sigma, \Omega} \isTerm{t}^{\syn}$, then it is decidable whether there is a type~$A$ such that
       \[
@@
       \]
-    \item if $\erase{\Gamma} |-_{\Sigma, \erase{\Omega}} \isTerm{t}^{\chk}$, then it is decidable for any $A$ whether
+    \item if $\erase{\Gamma} |-_{\Sigma, \Omega} \isTerm{t}^{\chk}$, then it is decidable for any $A$ whether
       \[
@@
 
-\begin{proof}[Proof of {\Cref{thm:bidirectional-type-synthesis}}]
-  We prove this statement by induction on the derivation of $|-_{\Sigma, \Omega} \isTerm{t}^{\dir{d}}$ (instead of $\isTerm{t}$ alone).
-  The first two cases \SynRule{Var} and \SynRule{Anno} are straightforward and have nothing to do with mode-correctness directly.
-  The third case \ChkRule{Sub} invokes the uniqueness of synthesised types to refute the case that $\Gamma |- \isTerm{t} \syn B$ but $A \neq B$ for a given type $A$ to check.
-  The last case \Rule{Op} is the interesting part of this theorem.
+The interesting part of the proof is the case for the $\Rule{Op}$ rule, so we shall give the insight first before jumping into details.
+Recall that a typing derivation for $\tmOpts$ for an operation $o\colon \Xi \rhd [\Delta_{1}] A_{1}^{\dir{d}}, \ldots, [\Delta_n]A_{n}^{\dir{d_n}} \to A_0^{\dir{d}}$ contains a substitution $\rho\colon \Xi \to \Type_{\Sigma}(\emptyset)$.
+The goal of type synthesis is exactly to define such a substitution $\rho$ for $\tmOpts$ while synthesising or checking each argument $t_i$, so we have to start with a \emph{partial} substitution $\rho_0$ defined on $\fv(A_0)$ if $d$ is $\chk$ or otherwise nowhere, as an `accumulating argument'.
+Then, as we visit each synthesis argument of type $A_{i+1}$, we may be able to extend the domain of a partial substitution $\rho_i$ to include the synthesis variables $\fv(A_{i + 1})$ if type synthesis is successful and also that the synthesised type can be \emph{unified with $A_{i+ 1}$} and \emph{extending} $\rho_i$ to $\bar{\rho_i} = \rho_{i+1}$.
+By mode-correctness, the accumulating substitution~$\rho_i$ would be defined on enough synthesis variables so that each type synthesis and checking can be performed on $t_i$ based on its mode derivation $\erase{\Gamma}, \vars{x}_i |-_{\Sigma, \Omega} t_i^{\dir{d_i}}$.
+Then, once we go through every argument $t_i$ successfully, we will have a partial substitution $\rho_n$ defined on every variable $x : \Xi$ by mode-correctness (\Cref{def:mode-correctness}), i.e.\ a substitution $\rho_n$ from $\Xi$ to $\Type_{\Sigma}(\emptyset)$.
+
+To make the aforementioned argument holds, it is necessary to compare types and solve a unification problem.
+Hence, we assume that the set $\Xi$ of variables has a decidable equality, rather than being an arbitrary set, thereby ensuring that the set $\Type_{\Sigma}(\Xi)$ of types also has a decidable equality.\footnote{%
+To simplify our choice, we could simply confine $\Xi$ to any set within the family of sets $\Fin_n$ of naturals less than~$n$.
+Given that these sets have a decidable equality and the arity of a type construct is finite, this would be a viable choice.
+Indeed, in our formalisation, we adopt $\Fin_n$ as the set of type variables in the definition of $\Type_{\Sigma}$ (see \Cref{sec:formalisation} for details).
+However, for the sake of clarity in presentation, we keep using named variables and just assume that $\Xi$ has a decidable equality.}
+The unification problem is as follows:
+\begin{lemma}
+  For any $A$ of\/ $\Type_{\Sigma}(\Xi)$, $B$ of\/ $\Type_{\Sigma}(\emptyset)$, and a partial substitution $\rho \colon \Xi \to \Type_{\Sigma}(\emptyset)$, 
+  \begin{enumerate}
+    \item either there is a \emph{minimal extension} $\bar{\rho}$ of $\rho$ such that $\simsub{A}{\bar{\rho}} = B$, or 
+    \item there is no extension $\sigma$ of $\rho$ such that $\simsub{A}{\sigma} = B$
+  \end{enumerate}
+  where by an extension $\sigma \geq \rho$ we mean that the domain of $\sigma$ contains the domain $\rho$ and $\sigma(x) = \rho(x)$ for every $x$ in the domain of $\rho$; by a minimal extension $\bar{\rho}$ of $\rho$ we mean an extension $\bar{\rho}$ of $\rho$ such that $\sigma \geq \bar{\rho}$ whenever $\sigma$ is an extension of $\rho$.
+\end{lemma}
+This lemma can be inferred from the correctness of first-order unification as proved by \citet{McBride2003,McBride2003a}, or be directly proved without unification.
+We now proceed with the decidability proof.
+
+\begin{proof}[Proof of {\Cref{thm:bidirectional-type-synthesis-checking}}]
+  We prove this statement by induction on the mode derivation\/ $\erase{\Gamma} |-_{\Sigma, \Omega} \isTerm{t}^{\dir{d}}$.
+  The two cases \SynRule{Var} and \SynRule{Anno} are straightforward and have nothing to do with mode-correctness.
+  The case \ChkRule{Sub} invokes the uniqueness of synthesised types to refute the case that $\Gamma |- \isTerm{t} \syn B$ but $A \neq B$ for a given type $A$.
+  The first three cases follow essentially the same argument provided by \citet{Wadler2022}, but we still give the arguments for the sake of completeness.
+  The last case \Rule{Op} is new and has been discussed above.
   \begin{itemize}
-    \item \SynRule{Var}: If $\isTerm{t}$ is a variable $\isTerm{x}$ in the context $\Gamma$, then there exists a type $A$ such that $\Gamma |- \isTerm{x} \syn A$.
-      Otherwise, if $\isTerm{x}$ is not in $\Gamma$ but $\Gamma |- \isTerm{x} \syn A$ is derivable, then by inversion $\isTerm{x}$ must be in $\Gamma$, a contradiction.
-    \item \SynRule{Anno}: For a raw term of the form $\isTerm{t} \annotate A$ in synthesising mode, it is decidable that $\Gamma |- \isTerm{t} \chk A$ is derivable or not by induction hypothesis.
-      \begin{enumerate}
-        \item If $\Gamma |- \isTerm{t} \chk A$ is derivable, then $\Gamma |- \isTerm{t \annotate A} \syn A$ is derivable.
-        \item If $\Gamma |- t \chk A$ is not derivable but $\Gamma |- \isTerm{t \annotate A} \syn$ is derivable, then by inversion $\Gamma |- t \chk A$ is derivable, thus a contradiction.
-      \end{enumerate}
+    \item \SynRule{Var}: As $\erase{\Gamma} |-_{\Sigma, \Omega} \isTerm{t}$, there exists a type $A$ such that $\Gamma |- \isTerm{x} \syn A$.
 
-    \item \ChkRule{Sub}: If $t$ is in checking mode because of the subsumption rule, then by induction hypothesis it is decidable whether $\Gamma |- t \syn B$ for some $B$:
-      \begin{enumerate}
-        \item If $\Gamma |- t \syn B$ is derivable for some $B$, then we have to consider any type $A$ and it is decidable if $A$ is equal to $B$:
-          \begin{enumerate}
-            \item if $A = B$ then we have arrived a derivation of $\Gamma |- t \chk A$.
-            \item if $A \neq B$ but $\Gamma |- t \chk A$ is derivable, then by inversion $\Gamma |- t \syn A$.
-              However, by \Cref{thm:unique-syn}, synthesised types $A$ and $B$ must be equal, so it is a contradiction.
-          \end{enumerate}
-        \item if $\Gamma |- t \syn C$ is not derivable for any type $C$ but $\Gamma |- t \chk A$ is derivable, then by inversion $\Gamma |- t \syn B$ for some $B = A$, thus a contradiction.
-      \end{enumerate}
+    \item \SynRule{Anno}: For $\erase{\Gamma} |-_{\Sigma, \Omega} (\isTerm{t} \annotate A)^{\syn}$, it is decidable whether $\Gamma |- \isTerm{t} \chk A$ by induction hypothesis.
+      \begin{itemize}
+        \item If $\Gamma |- \isTerm{t} \chk A$, then $\Gamma |- \isTerm{t \annotate A} \syn A$.
+        \item If $\Gamma \not|- t \chk A$ but $\Gamma |- \isTerm{t \annotate A} \syn$, then by inversion $\Gamma |- t \chk A$, leading to a contradiction.
+      \end{itemize}
 
-%    \item For every mode-correct operation $o \colon \Xi \rhd [\Delta_1]A_{1}^{d_1}, \ldots, [\Delta_{n}] A^{d_n}_{n} \to A^{d}$, a context $\Gamma$, and a raw term
+    \item \ChkRule{Sub}: For $\erase{\Gamma} |-_{\Sigma, \Omega}t^{\chk}$ by the rule $\ChkRule{Sub}$, we show that for any type $A$ it is decidable whether $\Gamma |-_{\Sigma, \Omega} t \chk A$.
+      First, by induction hypothesis, it is decidable whether $\Gamma |- t \syn B$ for some $B$:
+      \begin{itemize}
+        \item If $\Gamma \not|- t \syn C$ for any type $C$ but $\Gamma |- t \chk A$ is derivable, then by inversion $\Gamma |- t \syn B$ for some $B = A$, thus a contradiction.
+        \item If $\Gamma |- t \syn B$ for some $B$, then either $A = B$ or $A \neq B$ by the decidable equality on $\Type_{\Sigma}(\Xi)$: 
+          \begin{itemize}
+            \item if $A = B$ then we have\/ $\Gamma |- t \chk A$;
+            \item if $A \neq B$ but $\Gamma |- t \chk A$, then by inversion $\Gamma |- t \syn A$.
+              However, by \Cref{thm:unique-syn}, synthesised types $A$ and $B$ must be equal, so we derive a contradiction.
+          \end{itemize}
+      \end{itemize}
+    \item \Rule{Op}:
+      Consider the mode derivation of\/ $\erase{\Gamma} |-_{\Sigma, \Omega} \tmOpts^{\dir{d}}$ for an operation $o\colon \Xi \rhd [\Delta_{1}] A_{1}^{\dir{d}}, \ldots, [\Delta_n]A_{n}^{\dir{d_n}} \to A_0^{\dir{d}}$.
+%
+%  \begin{claim}\label{lem:args-induction}
+%    For the argument list $\overrightarrow{[\Delta_i] A_{i}^{\dir{d_i}}}$, \emph{partial} substitution $\rho$ from $\Xi$ to $\emptyset$
+%    \begin{enumerate}
+%      \item either there is a minimal extension $\ext{\rho}$ of $\rho$ defined on $\synvar(\mathit{as})$ such that all of following judgements are derivable
 %        \[
-%        \isTerm{\tmOp_{o}(\vec{x}_1.t_1,\ldots, \vec{x}_n.t_n)}, 
+%          \Gamma, \vars{x}_\isTerm{i} : \simsub{\Delta_i}{\ext{\rho}} |-_{\Sigma, \Omega} \isTerm{t_i} \colon \simsub{A_i}{\ext{\rho}}^{\dir{d_i}}
 %        \]
-%      with a mode $d$, it is decidable that there exists a substitution $\rho$ from $\Xi$ to $\emptyset$ and a typing derivation of
-%      \[
-%        \Gamma |-_{\Sigma, \Omega} \isTerm{\tmOp_{o}(\vec{x}_1.t_1,\ldots, \vec{x}_n.t_n)} : \simsub{A}{\rho}^{d}.
-%      \]
-    \item \Rule{Op}:
-      Suppose that $t$ is of them $\tmOpts$ for some mode-correct operation $o$ in $\Omega$.
-      No matter $t$ is checking or synthesising, we need to build derivations for its arguments incrementally and maintain a set of unifiers, i.e.\ a partial substitution $\rho\colon \PSub{\Xi}{\emptyset}$ is incrementally defined which is initially defined nowhere if $t$ is synthesising or on variables of the target $A$ if $t$ is checking.
-
-  \begin{claim}\label{lem:args-induction}
-    For every set $\Xi$ of type variables, list 
-    \[
-      \mathit{as} = \left([\Delta_i] A_{i}^{\dir{d_i}}\right)_{i = 1}^{n}
-    \]
-    list of raw terms $t_i$'s in mode $d_i$, context $\Gamma$, and \emph{partial} substitution $\rho$ from $\Xi$ to $\emptyset$
-    \begin{enumerate}
-      \item either there is a minimal extension $\ext{\rho}$ of $\rho$ defined on $\synvar(\mathit{as})$ such that all of following judgements are derivable
-        \[
-          \Gamma, \vars{x}_\isTerm{i} : \simsub{\Delta_i}{\ext{\rho}} |-_{\Sigma, \Omega} \isTerm{t_i} \colon \simsub{A_i}{\ext{\rho}}^{\dir{d_i}}
-        \]
-
-      \item or there is no extension $\sigma$ of $\rho$ such that all $\Gamma, \vars{x}_\isTerm{i} : \simsub{\Delta_i}{\sigma} |-_{\Sigma, \Omega} \isTerm{t_i} \colon \simsub{A_i}{\sigma}^{\dir{d_i}}$ has a typing derivation.
-    \end{enumerate}
-  \end{claim}
+%
+%      \item or there is no extension $\sigma$ of $\rho$ such that all $\Gamma, \vars{x}_\isTerm{i} : \simsub{\Delta_i}{\sigma} |-_{\Sigma, \Omega} \isTerm{t_i} \colon \simsub{A_i}{\sigma}^{\dir{d_i}}$ has a typing derivation.
+%    \end{enumerate}
+%  \end{claim}
   \end{itemize}
 \end{proof}
+
+\subsection{Trichotomy on Raw Terms by Type Synthesis} \label{subsec:trichotomy}
+
+%\begin{corollary}[Decidability of Bidirectional Type Synthesis] \label{thm:bidirectional-type-synthesis}
+%  In a mode-correct bidirectional type system $(\Sigma, \Omega)$,
+%  it is decidable whether there exists some $A$ such that
+%  \[
+%    \Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn A.
+%  \]
+%\end{corollary}
+
+\begin{corollary}[Trichotomy on Raw Terms]\label{cor:trichotomy}
+  For any mode-correct bidirectional type system $(\Sigma, \Omega)$, 
+  exactly one of the following holds:
+  \begin{enumerate}
+    \item $\erase{\Gamma} |-_{\Sigma, \Omega} \isTerm{t}^{\syn}$ and\/ $\Gamma |-_{\Sigma, \Omega} \isTerm{t} : A$ for some type $A$.
+    \item $\erase{\Gamma} |-_{\Sigma, \Omega} \isTerm{t}^{\syn}$ but\/ $\Gamma \not|-_{\Sigma, \Omega} \isTerm{t} : A$.
+    \item $\erase{\Gamma} \not|-_{\Sigma, \Omega} \isTerm{t}^{\syn}$.
+  \end{enumerate}
+\end{corollary}
+\begin{proof}
+  Combine \Cref{lem:soundness-completeness,thm:mode-preprocessing} with \Cref{thm:bidirectional-type-synthesis-checking}.
+
+\end{proof}

### 2023-07-09 tex/BiSig.tex
[ tex ] Revise S3
[ tex ] Minor changes
[ tex ] Merging from the old master branch
[ tex ] Minor in S3; adjust the todonote configuration
[ tex ] Fix typos and revise S5

@@
 
-\newcommand{\LT}[1]{\todo[author=LT,inline,color=orange!40]{{#1}}}
-\newcommand{\Josh}[1]{\todo[author=Josh,inline,color=yellow!40]{{#1}}}
+\newcommand{\LT}[2][]{\todo[author=LT,color=orange!40,#1]{#2}}
+\newcommand{\Josh}[2][]{\todo[author=Josh,color=yellow!40,#1]{#2}}
 
@@
 \begin{abstract}
+  \LT{Still not coherent enough}
 %  Type checking and inference serve as the transition from parsed abstract syntax trees to well-typed programs.
@@
   The main theorem is logically a trichotomy on untyped abstract syntax trees and computationally a type synthesiser that may throw an exception due to missing type annotations.
-  By being parametrised by a signature, our theory works as a verified type-synthesiser generator overall.
+  By being parametrised by a signature, our theory works overall as a verified type-synthesiser generator.
 \end{abstract}
@@
 \ccsdesc[500]{Software and its engineering~Syntax}
-\ccsdesc[500]{Theory of computation~Type theory}
 \ccsdesc[300]{Software and its engineering~Formal software verification}
+\ccsdesc[300]{Theory of computation~Type theory}
 
-\keywords{type checking, type inference, language formalisation, \Agda}
+\keywords{datatype-generic programming, language formalisation, \Agda}
 
@@
 %We thank Nathanael Arkor for the useful conversation.
+We thank Kuen-Bang {Hou (Favonia)} for his comments and suggestions.
 The work is supported by the Ministry of Science and Technology of Taiwan under grant MOST 109-2222-E-001-002-MY3.

### 2023-07-09 tex/sec3-defs.tex
[ tex ] Fix typos and revise S5

@@
   A \emph{binding arity} with a sort $T$ is an inhabitant of $\left(T^* \times T\right)^* \times T$ where $T^*$ is the set of lists over $T$.
-  In a binding arity ${[\Delta_1]A_{1}, \ldots, \left[\Delta_{n}\right] A_{n}} \to {A_0}$, every $\Delta_i$ and $A_i$ refers to the \emph{extension context} and the sort of the $i$-th argument, respectively; for brevity, it is denoted by ${[\Delta_1]A_{1}, \ldots, \left[\Delta_{n}\right] A_{n}} \to {A_0}$ and omit $[\Delta_i]$ if the extension context $\Delta_i$ is empty.
+  In a binding arity $(((\Delta_1, A_1), \ldots, (\Delta_n, A_n)), A)$, every $\Delta_i$ and $A_i$ refers to the \emph{extension context} and the sort of the $i$-th argument, respectively, and $A$ the target sort.
+  For brevity, a binding arity is denoted by $\bargs \to {A_0}$ and omit $[\Delta_i]$ if $\Delta_i$ is empty.
 \end{definition}
@@
 \end{definition}
-That is, each inhabitant $o: O$  is associated with a set $\Xi$ of type variables and a binding arity with the sort $\Type_{\Sigma}(\Xi)$.
-We may simply call $\arity(o)$ the arity of $o$, denoted by $\aritysymbol{o}{\Xi}{[\Delta_1]A_1, \ldots, [\Delta_{n}]A_{n}}{A_0}$.
+That is, each inhabitant $o: O$  is associated with a set $\Xi$ of type variables and a binding arity $\arity(o)$ with the sort $\Type_{\Sigma}(\Xi)$, denoted by $\bop$.
 By a \emph{simply typed language} $(\Sigma, \Omega)$, we mean a pair of a type signature $\Sigma$ and a binding signature $\Omega$.
@@
     \inferrule{V, \vars{x}_\isTerm{1} |-_{\Sigma, \Omega} \isTerm{t_1} \\ \cdots \\ V, \vars{x}_\isTerm{n} |-_{\Sigma, \Omega} \isTerm{t_n} } {V |-_{\Sigma, \Omega} \tmOpts }\,\Rule{Op} \and
-    \text{for $\aritysymbol{o}{\Xi}{[\Delta_1]A_{1}, \ldots, [\Delta_{n}] A_{n}}{A_0}$ in $\Omega$}
+    \text{for $\bop$ in $\Omega$}
   \end{mathpar}
@@
     \item annotations $\isTerm{t \annotate A}$ for some raw term $t$ in $V$ and a type $A$, and
-    \item a construct $\tmOpts$ for some $\aritysymbol{o}{\Xi}{[\Delta_1]A_{1}, \ldots, [\Delta_{n}] A_{n}}{A_0}$ in $O$, where $\vars{x}_{\isTerm{i}}$'s are lists of variables whose length is equal to the length of~$\Delta_i$ and $t_i$'s are raw terms in the variable list $V, \vars{x}_i$
+    \item a construct $\tmOpts$ for some $\bop$ in $O$, where $\vars{x}_{\isTerm{i}}$'s are lists of variables whose length is equal to the length of~$\Delta_i$ and $t_i$'s are raw terms in the variable list $V, \vars{x}_i$
   \end{enumerate*}
@@
       \\
-    \text{for $\aritysymbol{o}{\Xi}{[\Delta_1]A_{1}, \ldots, [\Delta_{n}] A_{n}}{A_0}$ in $\Omega$}
+    \text{for $\bop$ in $\Omega$}
     \end{mathpar}
@@
     \item a derivation of $\Gamma |-_{\Sigma, \Omega} (t \annotate A) : A$ if $\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A$ has a derivation, and
-    \item a derivation of $\Gamma |-_{\Sigma, \Omega} \tmOpts : \simsub{A_0}{\rho}$ for $\aritysymbol{o}{\Xi}{[\Delta_1]A_1, \ldots, [\Delta_{n}]A_{n}}{A_0}$ if there is a function $\rho\colon \Xi \to \Type_{\Sigma}(\emptyset)$ and a derivation of $\Gamma, \vars{x}_{\isTerm{i}} : \Delta_i \vdash_{\Sigma, \Omega} \isTerm{t_i} : \simsub{A_i}{\rho}$ for each $i$,
+    \item a derivation of $\Gamma |-_{\Sigma, \Omega} \tmOpts : \simsub{A_0}{\rho}$ for $\bop$ if there is a function $\rho\colon \Xi \to \Type_{\Sigma}(\emptyset)$ and a derivation of $\Gamma, \vars{x}_{\isTerm{i}} : \Delta_i |-_{\Sigma, \Omega} \isTerm{t_i} : \simsub{A_i}{\rho}$ for each $i$,
   \end{enumerate}
@@
   \]
-  Similarly, a bidirectional binding arity is denoted by ${[\Delta_1]A_{1}^{\dir{d_1}}, \ldots, [\Delta_{n}] A^{\dir{d_n}}_{n}} \to {A_0^{\dir{d}}}$.
+  and is denoted by $\biarity$.
 \end{definition}
@@
 \end{definition}
-We write $\aritysymbol{o}{\Xi}{[\Delta_1]A_{1}^{\dir{d_1}}, \ldots, [\Delta_{n}] A^{\dir{d_n}}_{n}}{A_0^{\dir{d}}}$ for an operation $o$ with a variable set $\Xi$ and its bidirectional binding arity with sort $\Type_{\Sigma}(\Xi)$.
-A bidirectional type system $(\Sigma, \Omega)$ refers to a pair of $\Sigma$ and a bidirectional binding signature $\Omega$.
-An operation $\aritysymbol{o}{\Xi}{[\Delta_1]A_{1}^{\dir{d_1}}, \ldots, [\Delta_{n}] A^{\dir{d_n}}_{n}}{A_0^{\dir{d}}}$ is checking if $d$ is ${\chk}$ or synthesis if $d$ is ${\syn}$. 
+We write $\biop$ for an operation $o$ with a variable set $\Xi$ and its bidirectional binding arity with sort $\Type_{\Sigma}(\Xi)$.
+We call it \emph{checking} if $d$ is ${\chk}$ or \emph{synthesis} if $d$ is ${\syn}$; similarly its $i$-th argument is checking if $d_i$ is $\chk$ and synthesis if $d_i$ is $\syn$.
+A bidirectional type system $(\Sigma, \Omega)$ refers to a pair of a type signature $\Sigma$ and a bidirectional binding signature $\Omega$.
 Every bidirectional binding signature $\Omega$ gives rise to a binding signature $\erase{\Omega}$ if we erase modes from $\Omega$, called the \emph{(mode) erasure} of $\Omega$.
@@
       {\Gamma |-_{\Sigma, \Omega} \tmOpts :^{\dir{d}} \simsub{A_0}{\rho}} \,\mathsf{Op}
-      \\ \text{for $\aritysymbol{o}{\Xi}{[\Delta_1]A_{1}^{\dir{d_1}}, \ldots, [\Delta_{n}] A_{n}^{\dir{d_n}}}{A_0^{\dir{d}}}$ in $\Omega$}
+      \\ \text{for $\biop$ in $\Omega$}
     \end{mathpar}
@@
       {V |-_{\Sigma, \Omega} \tmOpts^\dir{d}}\,\Rule{Op}
-      \and \text{for $\aritysymbol{o}{\Xi}{[\Delta_1]A_{1}^{\dir{d_1}}, \ldots, [\Delta_{n}] A^{\dir{d_n}}_{n}}{A_0^{\dir{d}}}$}
+      \and \text{for $\biop$}
     \end{mathpar}
@@
           \]
-          has a derivation for $\aritysymbol{o}{\Xi}{[\Delta_1]A_{1}^{\dir{d_1}}, \ldots, [\Delta_{n}] A_{n}^{\dir{d_n}}}{A_0^{\dir{d}}}$ in $\Omega$ if there is a function $\rho\colon \Xi \to \Type_{\Sigma}(\emptyset)$ and a derivation of $\Gamma, \vars{x}_{\isTerm{i}} : \Delta_i \vdash_{\Sigma, \Omega} \isTerm{t_i} :^{\dir{d_i}} \simsub{A_i}{\rho}$ for each $i$;
+          has a derivation for $\biop$ in $\Omega$ if there is a function $\rho\colon \Xi \to \Type_{\Sigma}(\emptyset)$ and a derivation of $\Gamma, \vars{x}_{\isTerm{i}} : \Delta_i \vdash_{\Sigma, \Omega} \isTerm{t_i} :^{\dir{d_i}} \simsub{A_i}{\rho}$ for each $i$;
     \item the set of \emph{mode derivations} of $V |-_{\Sigma, \Omega} t^\dir{d}$, indexed by a list $V$ of variables, a raw term $\isTerm{t}$ under $V$, and a mode $d$, is defined in \Cref{fig:mode-derivations}.
@@
 \[
-  \aritysymbol{\mathsf{abs}}{A , B}{[A]B^{\chk}}{(A \bto B)^{\chk}} \quad\text{and}\quad \aritysymbol{\mathsf{app}}{A, B}{(A \bto B)^{\syn}, A^{\chk}}{B^{\syn}}.
+  \aritysymbol{\mathsf{abs}}{A , B}{[A]B^{\chk}}{(A \bto B)^{\chk}}
+  \quad\text{and}\quad
+  \aritysymbol{\mathsf{app}}{A, B}{(A \bto B)^{\syn}, A^{\chk}}{B^{\syn}}.
 \]
 \end{example}
-Moreover, we are able to reason about constructions and properties that hold for any bidirectional simple type system $(\Sigma, \Omega)$.
+More importantly, we are able to reason about constructions and properties that hold for any bidirectional simple type system $(\Sigma, \Omega)$ once and for all.

### 2023-07-09 tex/sec4-pre-synthesis.tex
[ tex ] Fix typos and revise S5

@@
 \begin{corollary}[Mode Preprocessing]\label{thm:mode-preprocessing}
-  It is decidable whether $V |-_{\Sigma,\Omega} \isTerm{t}^\dir{d}$.%for any raw term $V |-_{\Sigma,\erase\Omega} t$ and mode~$\dir{d}$
+  It is decidable whether\/ $V |-_{\Sigma,\Omega} \isTerm{t}^\dir{d}$.%for any raw term $V |-_{\Sigma,\erase\Omega} t$ and mode~$\dir{d}$
 \end{corollary}

### 2023-07-09 tex/sec5-synthesis.tex
[ tex ] Fix typos and revise S5

@@
 \subsection{Mode Correctness}\label{sec:mode-correctness}
-As \citet{Dunfield2021} outlined, mode-correctness for a bidirectional type system means that for every typing rule:
+As \citet{Dunfield2021} outlined, mode-correctness for a bidirectional typing rule means that 
 \begin{enumerate*}
-\item each `input' type variable in a premise must be synthesised from `earlier' premises or provided by the conclusion, if the rule is checking;
-\item each `output' type variable in the conclusion should have been synthesised from premises already, if the rule is synthesis.
+\item each `input' type variable in a premise must be an `output` variable in `earlier' premises or provided by the conclusion, if the rule is checking;
+\item each `output' type variable in the conclusion should be some `output' variable in premises, if the rule is synthesis.
 \end{enumerate*}
-Here an `input' type variable refers to types in an extension context and the type of a premise in the checking mode.
-It is important to note that the order of premises in a bidirectional typing rule also matters, since synthesised types are accumulated during type synthesis.
+Here `input' variables refers to variables in an extension context and in the type of a premise in the checking mode.
+It is important to note that the order of premises in a bidirectional typing rule also matters, since synthesis type variables are instantiated incrementally during type synthesis.
 
@@
 
-Now we proceed with the formal definition of mode-correctness.
+Now let us define mode-correctness rigorously.
 As we have outlined, the condition of mode-correctness for a synthesis rule is different from that of a checking rule, and the argument order also matters.
-Defining the condition directly for a signature can be somewhat intricate.
-Instead, we choose to define the condition in a few more steps, based on the list of arguments---more specifically, triples of an extension context $\Delta_i$, a type $A_i$, and a mode $\dir{d_i}$---pertaining to an operation, and subsequently on a signature.
+Defining the condition directly for a rule, and thus in our setting for an operation, can be somewhat intricate.
+Instead, we choose to define the conditions for the argument list---more specifically, triples $\biargvec$ of an extension context $\Delta_i$, a type $A_i$, and a mode $\dir{d_i}$---pertaining to an operation, for an operation, and subsequently for a signature.
 We also need some auxiliary definitions for the subset of variables of a type and of a context and the set of variables that have been synthesised:
@@
   However, for our purposes, the choice among these definitions is not a matter of concern.}
-  of \emph{variables} which occur in a type $A$ is denoted by $\fv(A)$.
-  The subset $\fv(\Gamma)$ of variables in a context $\Gamma$ is defined by $\fv(\cdot) = \emptyset$ and $\fv(\Gamma, A) = \fv(\Gamma) \cup \fv(A)$.
-  For a list of \emph{arguments} $\overrightarrow{[\Delta_i]A_i^{d_i}}$, define the set of \emph{synthesis type variables} inductively as 
+  of \emph{(free) variables} of a type $A$ is denoted by $\fv(A)$.
+  The subset $\fv(\Gamma)$ of variables in a context $\Gamma$ is defined by\/ $\fv(\cdot) = \emptyset$ and\/ $\fv(\Gamma, A) = \fv(\Gamma) \cup \fv(A)$.
+  For an argument list $\biargvec$,\LT{Eww...} define the set of \emph{synthesis type variables} inductively as 
   \begin{align}
     \label{eq:synvar1}\synvar(\cdot)                   & = \emptyset  \\
-    \label{eq:synvar2}\synvar(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}, [\Delta_{n}] A_{n}^{\dir{\chk}}) & = 
-    \synvar(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}) \\
-    \label{eq:synvar3}\synvar(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}, [\Delta_{n}] A_{n}^{\dir{\syn}}) & = 
-    \synvar(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}) \cup \fv(A_{n}).
+    \label{eq:synvar2}\synvar(\biargvec, \chkbiarg[n]) & = \phantom{\fv(A_{n}) \cup {}} \synvar(\biargvec) \\
+    \label{eq:synvar3}\synvar(\biargvec, \synbiarg[n]) & = \fv(A_{n}) \cup           \synvar(\biargvec).
   \end{align}
 \end{definition}
-This subset contains type variables of a synthesising argument and they are exactly those type variables that would be synthesised during type synthesis.
+This subset contains type variables of a synthesis argument and they are exactly those type variables that would be synthesised during type synthesis.
 
-\begin{definition}\label{def:mode-correctness}
-  The \emph{mode-correctness for an argument list} $\Xi \rhd [\Delta_{1}] A_{1}^{\dir{d}}, \ldots, [\Delta_n]A_{n}^{\dir{d_n}}$  with respect to a subset $S$ of $\Xi$ is defined inductively as
+\begin{definition}\label{def:mode-correctness-args}
+  The \emph{mode-correctness} $\MCas\left(\biargvec\right)$ for an argument list $\biargs$ with respect to a subset $S$ of $\Xi$ is defined as
   \begin{align}
-    \label{eq:MC1} \MC_{\mathit{as}}( \cdot ) & = \top \\
-    \label{eq:MC2} \MC_{\mathit{as}}\left(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}, [\Delta_{n}]A_{n}^{\syn}\right) 
-    & = \MC_{\mathit{as}}\left(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}\right)
-    \land \fv(\Delta) \subseteq \left( S \cup \synvar\left(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}\right) \right)\\
-    \label{eq:MC3} \MC_{\mathit{as}}\left(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}, [\Delta_{n}]A_{n}^{\chk}\right)
-    & = \MC_{\mathit{as}}\left(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}\right)
-    \land \fv(\Delta, A_n) \subseteq \left( S \cup \synvar\left(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}\right)\right)
+    \label{eq:MC1} \MCas( \cdot ) & = \top \\
+    \label{eq:MC3} \MCas\left(\biargvec, \chkbiarg[n]\right)
+                                  & = \fv(\Delta, A_n) \subseteq \left( S \cup \synvar\left(\biargvec\right)\right) \land \MCas\left(\biargvec\right) \\
+    \label{eq:MC2} \MCas\left(\biargvec, \synbiarg[n]\right) 
+                                  & = \phantom{, A_n} \fv(\Delta) \subseteq \left( S \cup \synvar\left(\biargvec\right)\right) \land  \MCas\left(\biargvec\right)
   \end{align}
@@
 \end{definition}
-This definition encapsulates the idea that every `input' type variable, possibly deriving from a context $\Delta$ or a checking argument $A_n$, must be an `output' variable from $\synvar(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}})$ or, if the rule is checking, belong to the set $S$ of `input' variables in its conclusion.
+This definition encapsulates the idea that every `input' type variable, possibly deriving from a context $\Delta$ or a checking argument $A_n$, must be an `output' variable from $\synvar(\biargvec)$ or, if the rule is checking, belong to the set $S$ of `input' variables in its conclusion.
 This condition must be met for every tail of the argument list as well to ensure that `output' variables accessible at each argument position are from preceding arguments only, hence an inductive definition.
-\begin{definition}
-  An operation $o\colon \Xi \rhd [\Delta_{1}] A_{1}^{\dir{d}}, \ldots, [\Delta_n]A_{n}^{\dir{d_n}} \to A_0^{\dir{d}}$ is \emph{mode-correct} provided that
+\begin{definition}\label{def:mode-correctness}
+  An operation $\biop$ is \emph{mode-correct} if either
   \begin{enumerate}
-    \item $d$ is $\chk$, its argument list is mode-correct w.r.t.\ $\fv(A_0)$, and the union $\fv(A_0) \cup \synvar(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}})$ contains every inhabitant of $\Xi$;
-    \item $d$ is $\syn$, its argument list is mode-correct w.r.t.\ $\emptyset$, and the finite subset $\synvar(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}})$ contains every inhabitant of $\Xi$ and, particularly, $\fv(A_0)$.
+    \item $d$ is $\chk$, its argument list is mode-correct w.r.t.\ $\fv(A_0)$, and the union $\fv(A_0) \cup \synvar(\biargvec)$ contains every inhabitant of $\Xi$; or
+    \item $d$ is $\syn$, its argument list is mode-correct w.r.t.\ $\emptyset$, and the finite subset $\synvar(\biargvec)$ contains every inhabitant of $\Xi$ and, particularly, $\fv(A_0)$.
   \end{enumerate}
@@
   Thus, this property cannot be formalised without treating rules as some mathematical object, such as the notion of bidirectional binding signature presented in \Cref{sec:defs}.
-  This contrasts with the properties in \Cref{sec:pre-synthesis}, which can still be articulated for individual systems even in the absence of a general definition of bidirectional type systems.
+  This contrasts with the properties in \Cref{sec:pre-synthesis}, which can still be specified for individual systems even in the absence of a generic definition of bidirectional type systems.
 \end{remark}
@@
     \item $\SynRule{Anno}$ holds trivially, since the synthesised type $A$ is from the term $t \annotate A$ in question.
-    \item $\Rule{Op}$: Recall here that a derivation of\/ $\Gamma |- \tmOpts \syn A$ for a synthesis operation $o\colon \Xi \rhd [\Delta_{1}] A_{1}^{\dir{d}}, \ldots, [\Delta_n]A_{n}^{\dir{d_n}} \to A_0^{\syn}$ contains a substitution $\rho$ from the local context $\Xi$ to concrete types \Cref{fig:extrinsic-typing}.
-      To prove that any two typing derivations has the same synthesised type, it suffices to show that those substitutions $\rho_1$ and $\rho_2$ agree on variables in $\synvar\left([\Delta_{1}] A_{1}^{\dir{d}}, \ldots, [\Delta_n]A_{n}^{\dir{d_n}}\right)$ so that $\simsub{A_0}{\rho_1} = \simsub{A_0}{\rho_2}$.
+    \item $\Rule{Op}$: Recall that a derivation of\/ $\Gamma |- \tmOpts \syn A$ contains a substitution $\rho$ from the local context $\Xi$ to concrete types.
+      To prove that any two typing derivations has the same synthesised type, it suffices to show that those substitutions $\rho_1$ and $\rho_2$ of $d_1$ and $d_2$, respectively, agree on variables in $\synvar\left(\biargs\right)$ so that $\simsub{A_0}{\rho_1} = \simsub{A_0}{\rho_2}$.
       We prove it by induction on the argument list:
@@
         \item If $\dir{d_{i+1}}$ is synthesis, then $\simsub{\Delta_{i+1}}{\rho_1} = \simsub{\Delta_{i+1}}{\rho_2}$ by mode-correctness~\eqref{eq:MC2} and induction hypothesis (of the list).
-          Therefore, under the same context $\Gamma, \simsub{\Delta_{i+1}}{\rho_1} = \Gamma, \simsub{\Delta_{i+1}}{\rho_2}$ the term $t_{i+1}$ must have the same synthesised type $\simsub{A_{i+1}}{\rho_1} = \simsub{A_{i+1}}{\rho_1}$ by induction hypothesis (of the typing derivation), so $\rho_1$ and $\rho_2$ agree on $\fv(A_{i+1})$ in addition to $\synvar\left([\Delta_{1}] A_{1}^{\dir{d}}, \ldots, [\Delta_i]A_{i}^{\dir{d_i}}\right)$, as required for the case \eqref{eq:synvar3} in the definition of $\synvar$.
+          Therefore, under the same context $\Gamma, \simsub{\Delta_{i+1}}{\rho_1} = \Gamma, \simsub{\Delta_{i+1}}{\rho_2}$ the term $t_{i+1}$ must have the same synthesised type $\simsub{A_{i+1}}{\rho_1} = \simsub{A_{i+1}}{\rho_1}$ by induction hypothesis (of the typing derivation), so $\rho_1$ and $\rho_2$ agree on $\fv(A_{i+1})$ in addition to $\synvar\left(\biargs\right)$, as required for the case \eqref{eq:synvar3} in the definition of $\synvar$.
       \end{enumerate}
@@
 \subsection{Decidability of Bidirectional Type Synthesis and Checking}\label{subsec:bidirectional-synthesis-checking}
-We are ready for the last technical result of this paper---the decidability of bidirectional type synthesis and checking for any mode correct bidirectional type system:
+We are here for the main technical contribution of this paper---the decidability of bidirectional type synthesis and checking for any mode correct bidirectional type system:
 \begin{theorem}[Decidability of Bidirectional Type Synthesis and Checking] \label{thm:bidirectional-type-synthesis-checking}
@@
   \begin{enumerate}
-    \item if $\erase{\Gamma} |-_{\Sigma, \Omega} \isTerm{t}^{\syn}$, then it is decidable whether there is a type~$A$ such that
+    \item if\/ $\erase{\Gamma} |-_{\Sigma, \Omega} \isTerm{t}^{\syn}$, then it is decidable whether there is~$A$ such that
       \[
@@
       \]
-    \item if $\erase{\Gamma} |-_{\Sigma, \Omega} \isTerm{t}^{\chk}$, then it is decidable for any $A$ whether
+    \item if\/ $\erase{\Gamma} |-_{\Sigma, \Omega} \isTerm{t}^{\chk}$, then it is decidable for any~$A$ whether
       \[
@@
 The interesting part of the proof is the case for the $\Rule{Op}$ rule, so we shall give the insight first before jumping into details.
-Recall that a typing derivation for $\tmOpts$ for an operation $o\colon \Xi \rhd [\Delta_{1}] A_{1}^{\dir{d}}, \ldots, [\Delta_n]A_{n}^{\dir{d_n}} \to A_0^{\dir{d}}$ contains a substitution $\rho\colon \Xi \to \Type_{\Sigma}(\emptyset)$.
-The goal of type synthesis is exactly to define such a substitution $\rho$ for $\tmOpts$ while synthesising or checking each argument $t_i$, so we have to start with a \emph{partial} substitution $\rho_0$ defined on $\fv(A_0)$ if $d$ is $\chk$ or otherwise nowhere, as an `accumulating argument'.
-Then, as we visit each synthesis argument of type $A_{i+1}$, we may be able to extend the domain of a partial substitution $\rho_i$ to include the synthesis variables $\fv(A_{i + 1})$ if type synthesis is successful and also that the synthesised type can be \emph{unified with $A_{i+ 1}$} and \emph{extending} $\rho_i$ to $\bar{\rho_i} = \rho_{i+1}$.
-By mode-correctness, the accumulating substitution~$\rho_i$ would be defined on enough synthesis variables so that each type synthesis and checking can be performed on $t_i$ based on its mode derivation $\erase{\Gamma}, \vars{x}_i |-_{\Sigma, \Omega} t_i^{\dir{d_i}}$.
-Then, once we go through every argument $t_i$ successfully, we will have a partial substitution $\rho_n$ defined on every variable $x : \Xi$ by mode-correctness (\Cref{def:mode-correctness}), i.e.\ a substitution $\rho_n$ from $\Xi$ to $\Type_{\Sigma}(\emptyset)$.
+Recall that a typing derivation for $\tmOpts$ contains a substitution $\rho\colon \Xi \to \Type_{\Sigma}(\emptyset)$.
+The goal of type synthesis is exactly to define such a substitution $\rho$, so we have to start with a \emph{partial} substitution $\rho_0$ defined on $\fv(A_0)$ if $d$ is $\chk$ or otherwise nowhere, as an `accumulating argument'.
+Then, as we visit each synthesis argument $\synbiarg[i+1]$, we may be able to extend the domain of $\rho_i$ to include the synthesis variables $\fv(A_{i + 1})$ if type synthesis is successful and also that the synthesised type can be \emph{unified with $A_{i+ 1}$} and \emph{extending} $\rho_i$ to $\bar{\rho_i} = \rho_{i+1}$.
+By mode-correctness, the accumulating substitution~$\rho_i$ would be defined on enough synthesis variables so that each type synthesis and checking can be performed on $t_{i}$ with the context $\Gamma, \simsub{\Delta_{i+1}}{\rho_i}$ based on its mode derivation $\erase{\Gamma}, \vars{x}_i |-_{\Sigma, \Omega} t_i^{\dir{d_i}}$.
+Then, if we go through every argument $t_i$ successfully, we will have a substitution $\rho_n$ from $\Xi$ to $\Type_{\Sigma}(\emptyset)$ defined on every variable $x : \Xi$ by mode-correctness.
 
@@
 Hence, we assume that the set $\Xi$ of variables has a decidable equality, rather than being an arbitrary set, thereby ensuring that the set $\Type_{\Sigma}(\Xi)$ of types also has a decidable equality.\footnote{%
-To simplify our choice, we could simply confine $\Xi$ to any set within the family of sets $\Fin_n$ of naturals less than~$n$.
+To simplify our choice, we could simply confine $\Xi$ to any set within the family of sets $\Fin(n)$ of naturals less than~$n$.
 Given that these sets have a decidable equality and the arity of a type construct is finite, this would be a viable choice.
-Indeed, in our formalisation, we adopt $\Fin_n$ as the set of type variables in the definition of $\Type_{\Sigma}$ (see \Cref{sec:formalisation} for details).
+Indeed, in our formalisation, we adopt $\Fin(n)$ as the set of type variables in the definition of $\Type_{\Sigma}$ (see \Cref{sec:formalisation} for details).
 However, for the sake of clarity in presentation, we keep using named variables and just assume that $\Xi$ has a decidable equality.}
@@
 \begin{lemma}
-  For any $A$ of\/ $\Type_{\Sigma}(\Xi)$, $B$ of\/ $\Type_{\Sigma}(\emptyset)$, and a partial substitution $\rho \colon \Xi \to \Type_{\Sigma}(\emptyset)$, 
+  For any\/ $A$ of\/ $\Type_{\Sigma}(\Xi)$, $B$ of\/ $\Type_{\Sigma}(\emptyset)$, and a partial substitution $\rho \colon \Xi \to \Type_{\Sigma}(\emptyset)$, 
   \begin{enumerate}
@@
   \end{enumerate}
-  where by an extension $\sigma \geq \rho$ we mean that the domain of $\sigma$ contains the domain $\rho$ and $\sigma(x) = \rho(x)$ for every $x$ in the domain of $\rho$; by a minimal extension $\bar{\rho}$ of $\rho$ we mean an extension $\bar{\rho}$ of $\rho$ such that $\sigma \geq \bar{\rho}$ whenever $\sigma$ is an extension of $\rho$.
+  where by an extension $\sigma \geq \rho$ we mean that the domain of $\sigma$ contains the domain $\rho$ and $\sigma(x) = \rho(x)$ for every $x$ in the domain of $\rho$; by a minimal extension $\bar{\rho}$ of $\rho$ satisfying $P$ we mean an extension $\bar{\rho}$ of $\rho$ with $P(\bar{\rho})$ such that $\sigma \geq \bar{\rho}$ whenever $\sigma$ is an extension of $\rho$ with $P(\sigma)$.
 \end{lemma}
-This lemma can be inferred from the correctness of first-order unification as proved by \citet{McBride2003,McBride2003a}, or be directly proved without unification.
-We now proceed with the decidability proof.
+This lemma can be inferred from the correctness of first-order unification~\citep{McBride2003,McBride2003a}, or be proved directly without unification.
+We are now ready for the decidability proof.
 
@@
   The case \ChkRule{Sub} invokes the uniqueness of synthesised types to refute the case that $\Gamma |- \isTerm{t} \syn B$ but $A \neq B$ for a given type $A$.
-  The first three cases follow essentially the same argument provided by \citet{Wadler2022}, but we still give the arguments for the sake of completeness.
-  The last case \Rule{Op} is new and has been discussed above.
+  The first three cases follow essentially the same argument provided by \citet{Wadler2022}, but we still give the arguments here for the sake of completeness.
+  The last case \Rule{Op} is new and has been discussed informally above.
   \begin{itemize}
@@
 
-    \item \ChkRule{Sub}: For $\erase{\Gamma} |-_{\Sigma, \Omega}t^{\chk}$ by the rule $\ChkRule{Sub}$, we show that for any type $A$ it is decidable whether $\Gamma |-_{\Sigma, \Omega} t \chk A$.
-      First, by induction hypothesis, it is decidable whether $\Gamma |- t \syn B$ for some $B$:
+    \item \ChkRule{Sub}: If $\erase{\Gamma} |-_{\Sigma, \Omega}t^{\chk}$ by the rule $\ChkRule{Sub}$, then $\Gamma |-_{\Sigma, \Omega} t^{\syn}$ by inversion.
+      We show that it is decidable for any $A$ whether $\Gamma |-_{\Sigma, \Omega} t \chk A$.
+      By induction hypothesis, it is decidable whether $\Gamma |- t \syn B$ for some $B$:
       \begin{itemize}
@@
     \item \Rule{Op}:
-      Consider the mode derivation of\/ $\erase{\Gamma} |-_{\Sigma, \Omega} \tmOpts^{\dir{d}}$ for an operation $o\colon \Xi \rhd [\Delta_{1}] A_{1}^{\dir{d}}, \ldots, [\Delta_n]A_{n}^{\dir{d_n}} \to A_0^{\dir{d}}$.
+      Consider the mode derivation of\/ $\erase{\Gamma} |-_{\Sigma, \Omega} \tmOpts^{\dir{d}}$.
 %

### 2023-07-09 tex/sec5-synthesis.tex
[ tex ] Trichotomy in the end of S5

@@
 This section focuses on defining mode-correctness and deriving bidirectional type synthesis for any mode-correct bidirectional type system $(\Sigma, \Omega)$.
-\LT{trichotomy first or last?}
-We start with the statement of the decidability of bidirectional synthesis and the trichotomy on raw terms for mode-correct bidirectional type systems in \Cref{subsec:trichotomy}, but defer the definition of mode-correctness until \Cref{sec:mode-correctness} and the decidability proof until \Cref{subsec:bidirectional-synthesis-checking}.
-We will then formalise mode-correctness and show the uniqueness of synthesized types in \Cref{sec:mode-correctness}.
-This uniqueness property means that any two synthesised types for the same raw term $t$ under the same context $\Gamma$ have to be equal.
-This will be used later in the proof of the decidability of bidirectional type synthesis and checking in \Cref{subsec:bidirectional-synthesis-checking}.
+We start with \Cref{sec:mode-correctness} by defining mode-correctness and showing the uniqueness of synthesized types.
+This uniqueness means that any two synthesised types for the same raw term $t$ under the same context $\Gamma$ have to be equal.
+It will be used especially in \Cref{subsec:bidirectional-synthesis-checking} for the proof of the decidability of bidirectional type synthesis and checking.
+Then, we conclude this section with the trichotomy on raw terms that has been illustrated at the beginning in \Cref{subsec:trichotomy}.
 

### 2023-07-09 tex/sec4-pre-synthesis.tex
[ tex ] Use \nvdash instead of \not\vdash

@@
 
-%\Josh{Throughout this sub-section, we quantify over any bidirectional type system~$(\Sigma, \Omega)$, context $\Gamma : \Cxt_\Sigma(\emptyset)$, raw term $\erase\Gamma \vdash_{\Sigma, \erase\Omega} t$, mode~$d$, and type $A : \Type_\Sigma(\emptyset)$.}
+%\Josh{Throughout this sub-section, we quantify over any bidirectional type system~$(\Sigma, \Omega)$, context $\Gamma : \Cxt_\Sigma(\emptyset)$, raw term $\erase\Gamma |-_{\Sigma, \erase\Omega} t$, mode~$d$, and type $A : \Type_\Sigma(\emptyset)$.}
 
@@
 \begin{lemma}\label{thm:Pre?-false}
-If\/ $V |-_{\Sigma,\Omega} t^{\dir{d}\,\BFalse\,s}$, then $V \not|-_{\Sigma,\Omega} t^\dir{d}$.
+If\/ $V |-_{\Sigma,\Omega} t^{\dir{d}\,\BFalse\,s}$, then $V |/-_{\Sigma,\Omega} t^\dir{d}$.
 \end{lemma}
@@
 \begin{proof}
-By \cref{thm:generalised-mode-preprocessing}, there is a derivation of $V |-_{\Sigma,\erase\Omega} t^{\dir{d}\,g\,s}$ for some booleans $g$~and~$s$, and then simply check~$g$ and apply either \cref{thm:Pre?-true} or \cref{thm:Pre?-false} to obtain $V |-_{\Sigma,\Omega} t^\dir{d}$ or $V \not|-_{\Sigma,\Omega} t^\dir{d}$.
+By \cref{thm:generalised-mode-preprocessing}, there is a derivation of $V |-_{\Sigma,\erase\Omega} t^{\dir{d}\,g\,s}$ for some booleans $g$~and~$s$, and then simply check~$g$ and apply either \cref{thm:Pre?-true} or \cref{thm:Pre?-false} to obtain $V |-_{\Sigma,\Omega} t^\dir{d}$ or $V |/-_{\Sigma,\Omega} t^\dir{d}$.
 \end{proof}
@@
 %\begin{corollary}\label{thm:toPre?-false}
-%If\/ $V \not|-_{\Sigma,\Omega} t^\dir{d}$, then $V |-_{\Sigma,\Omega} t^{\dir{d}\,\BFalse\,s}$ for some boolean~$s$.
+%If\/ $V |/-_{\Sigma,\Omega} t^\dir{d}$, then $V |-_{\Sigma,\Omega} t^{\dir{d}\,\BFalse\,s}$ for some boolean~$s$.
 %\end{corollary}
@@
 %If $V |-_{\Sigma,\Omega} t^\dir{d}$, then $g$~has to be $\BTrue$ since $g = \BFalse$ leads to a contradiction with \cref{thm:Pre?-false}.
-%Symmetrically, if $V \not|-_{\Sigma,\Omega} t^\dir{d}$, then $g$~has to be~$\BFalse$ so as not to contradict \cref{thm:Pre?-true}.
+%Symmetrically, if $V |/-_{\Sigma,\Omega} t^\dir{d}$, then $g$~has to be~$\BFalse$ so as not to contradict \cref{thm:Pre?-true}.
 %\end{proof}

### 2023-07-09 tex/sec5-synthesis.tex
[ tex ] Use \nvdash instead of \not\vdash

@@
         \item If $\Gamma |- \isTerm{t} \chk A$, then $\Gamma |- \isTerm{t \annotate A} \syn A$.
-        \item If $\Gamma \not|- t \chk A$ but $\Gamma |- \isTerm{t \annotate A} \syn$, then by inversion $\Gamma |- t \chk A$, leading to a contradiction.
+        \item If $\Gamma |/- t \chk A$ but $\Gamma |- \isTerm{t \annotate A} \syn$, then by inversion $\Gamma |- t \chk A$, leading to a contradiction.
       \end{itemize}
@@
       \begin{itemize}
-        \item If $\Gamma \not|- t \syn C$ for any type $C$ but $\Gamma |- t \chk A$ is derivable, then by inversion $\Gamma |- t \syn B$ for some $B = A$, thus a contradiction.
+        \item If $\Gamma |/- t \syn C$ for any type $C$ but $\Gamma |- t \chk A$ is derivable, then by inversion $\Gamma |- t \syn B$ for some $B = A$, thus a contradiction.
         \item If $\Gamma |- t \syn B$ for some $B$, then either $A = B$ or $A \neq B$ by the decidable equality on $\Type_{\Sigma}(\Xi)$: 
@@
     \item $\erase{\Gamma} |-_{\Sigma, \Omega} \isTerm{t}^{\syn}$ and\/ $\Gamma |-_{\Sigma, \Omega} \isTerm{t} : A$ for some type $A$.
-    \item $\erase{\Gamma} |-_{\Sigma, \Omega} \isTerm{t}^{\syn}$ but\/ $\Gamma \not|-_{\Sigma, \Omega} \isTerm{t} : A$.
-    \item $\erase{\Gamma} \not|-_{\Sigma, \Omega} \isTerm{t}^{\syn}$.
+    \item $\erase{\Gamma} |-_{\Sigma, \Omega} \isTerm{t}^{\syn}$ but\/ $\Gamma |/-_{\Sigma, \Omega} \isTerm{t} : A$.
+    \item $\erase{\Gamma} |/-_{\Sigma, \Omega} \isTerm{t}^{\syn}$.
   \end{enumerate}

### 2023-07-09 tex/sec7-future.tex
[ tex ] Minor in S3; adjust the todonote configuration
[ tex ] ‘Beyond syntax-directed type systems’ in S7

@@
 
-\Josh{Getting rid of the assumption of being syntax-directed (the triangular picture) --- 0.5p}
-\Josh{Ornaments -- 0.25p}
-\Josh{NDGP -- 0.25p (optional?), \citep{Ko2022} }
+\paragraph{Beyond syntax-directed type systems}
 
-\LT{polymorphic algebraic theories, dependent signatures --- 0.5p}
+As a first theory of bidirectional type synthesis, our work focuses on syntax-directed type systems.
+To explore more general settings, an extreme possibility is to independently specify an ordinary type system and a bidirectional one over the same raw terms and then investigate possible relationships between the two systems, but a more manageable and practical setting suitable for a next step is probably one that only mildly generalises ours:
+for each raw term construct, there can be several ordinary typing rules, and each typing rule can be refined to several bidirectional typing rules with different mode assignments.
+Even for this mildly generalised setting, there is already a lot more work to do:
+Both the mode preprocessor and the bidirectional type synthesiser will have to backtrack, and become significantly more complex.
+For soundness and completeness, it should still be possible to treat them as the separation and combination of mode and type information, but the completeness direction will pose a problem---for every node of a raw term, a mode derivation chooses a mode assignment while a typing derivation chooses a typing rule, but there may not be a bidirectional typing rule for this particular combination.
+There should be ways to fix this problem while retaining the mode preprocessing phase---for example, one idea is to make the mode preprocessor produce all possible mode derivations, and refine completeness to say that any typing derivation can be combined with one of these mode derivations into a bidirectional typing derivation.
+All these considerations are rather too complex for a first theory, however, and should best be left as future work.
+
+%\Josh[inline, caption={}]{Getting rid of the assumption of being syntax-directed (the triangular picture) --- 0.5p
+%\begin{enumerate}
+%\item In general: two independently formulated type systems over raw terms
+%\item More sensible next step: a `refinement chain'
+%\item Backtracking algorithms are required. for both mode preprocessing and bidirectional type synthesis.
+%\item The idea of separating and combining mode and type information should still work, but requires more care in the completeness direction.
+%\item Too complex for a first theory, and left as future work
+%\end{enumerate}}
+
+\LT[inline]{polymorphic algebraic theories, dependent signatures --- 0.5p}
 
@@
 Subtyping is essential for formulating important concepts such as \emph{principal types} in type synthesis.
-On the other hand, in the realm of dependent types, \citeauthor{Cartmell1986}'s generalized algebraic theories~\citeyearpar{Cartmell1986} can handle a wide variety of dependent type theories.
-\citet{Bezem2021} investigate the notion of presentation (extending the notion of signature) in the context of generalized algebraic theories.
+On the other hand, in the realm of dependent types, \citeauthor{Cartmell1986}'s generalised algebraic theories~\citeyearpar{Cartmell1986} can handle a wide variety of dependent type theories.
+\citet{Bezem2021} investigate the notion of presentation (extending the notion of signature) in the context of generalised algebraic theories.
 Nonetheless, type synthesis for dependent types requires normalisation or some form of conversion to check type equality.
@@
 \todo[inline]{We leave the problem of presenting bidirectional typing for more general and advanced language features as future work (such as those related to polymorphic types~\citep{Pierce2000,Peyton-Jones2007,Dunfield2013,Xie2018}), which requires advancement in language formalisation~(\cref{sec:language-formalisation}) that is orthogonal to our work.}
+
+\Josh[inline]{Optional: ornaments (0.25p), NDGP (0.25p)}
 

### 2023-07-09 tex/sec5-synthesis.tex
[ tex ] Done with S5 for now

@@
 It is also easy to check the bidirectional type system $(\Sigma_{\bto}, \Omega^{\Leftrightarrow}_{\Lambda}$) in \Cref{ex:signature-simply-typed-lambda} for simply typed $\lambda$-calculus is mode-correct according to our definition.
-\LT{perhaps we shall prove that mode-correctness is decidable}
+\LT{perhaps we shall prove it decidable}
 
@@
 
-The interesting part of the proof is the case for the $\Rule{Op}$ rule, so we shall give the insight first before jumping into details.
+The interesting part of the proof is the case for the $\Rule{Op}$ rule, so we shall give the insight first instead of jumping into the details.
 Recall that a typing derivation for $\tmOpts$ contains a substitution $\rho\colon \Xi \to \Type_{\Sigma}(\emptyset)$.
-The goal of type synthesis is exactly to define such a substitution $\rho$, so we have to start with a \emph{partial} substitution $\rho_0$ defined on $\fv(A_0)$ if $d$ is $\chk$ or otherwise nowhere, as an `accumulating argument'.
+The goal of type synthesis is exactly to define such a substitution $\rho$, so we have to start with a \emph{partial} substitution $\rho_0$, as an `accumulating argument', defined on $\fv(A_0)$ if $d$ is $\chk$ or otherwise nowhere.
 Then, as we visit each synthesis argument $\synbiarg[i+1]$, we may be able to extend the domain of $\rho_i$ to include the synthesis variables $\fv(A_{i + 1})$ if type synthesis is successful and also that the synthesised type can be \emph{unified with $A_{i+ 1}$} and \emph{extending} $\rho_i$ to $\bar{\rho_i} = \rho_{i+1}$.
-By mode-correctness, the accumulating substitution~$\rho_i$ would be defined on enough synthesis variables so that each type synthesis and checking can be performed on $t_{i}$ with the context $\Gamma, \simsub{\Delta_{i+1}}{\rho_i}$ based on its mode derivation $\erase{\Gamma}, \vars{x}_i |-_{\Sigma, \Omega} t_i^{\dir{d_i}}$.
-Then, if we go through every argument $t_i$ successfully, we will have a substitution $\rho_n$ from $\Xi$ to $\Type_{\Sigma}(\emptyset)$ defined on every variable $x : \Xi$ by mode-correctness.
+By mode-correctness, the accumulating substitution~$\rho_i$ would be defined on enough synthesis variables so that type synthesis or checking can be performed on $t_{i}$ with the context $\Gamma, \simsub{\Delta_{i+1}}{\rho_i}$ based on its mode derivation $\erase{\Gamma}, \vars{x}_i |-_{\Sigma, \Omega} t_i^{\dir{d_i}}$.
+Then, if we go through every argument $t_i$ successfully, we will have a substitution $\rho_n$ defined everywhere by mode-correctness.
 
-To make the aforementioned argument holds, it is necessary to compare types and solve a unification problem.
-Hence, we assume that the set $\Xi$ of variables has a decidable equality, rather than being an arbitrary set, thereby ensuring that the set $\Type_{\Sigma}(\Xi)$ of types also has a decidable equality.\footnote{%
-To simplify our choice, we could simply confine $\Xi$ to any set within the family of sets $\Fin(n)$ of naturals less than~$n$.
-Given that these sets have a decidable equality and the arity of a type construct is finite, this would be a viable choice.
+\begin{remark}
+To make the argument above sound, it is necessary to compare types and solve a unification problem.
+Hence, we assume that the set\/ $\Xi$ of variables has a decidable equality, thereby ensuring that the set $\Type_{\Sigma}(\Xi)$ of types also has a decidable equality.\footnote{%
+To simplify our choice, we could simply confine $\Xi$ to any set within the family of sets $\Fin(n)$ of naturals less than~$n$, given that these sets have a decidable equality and the arity of a type construct is finite.
 Indeed, in our formalisation, we adopt $\Fin(n)$ as the set of type variables in the definition of $\Type_{\Sigma}$ (see \Cref{sec:formalisation} for details).
-However, for the sake of clarity in presentation, we keep using named variables and just assume that $\Xi$ has a decidable equality.}
-The unification problem is as follows:
-\begin{lemma}
-  For any\/ $A$ of\/ $\Type_{\Sigma}(\Xi)$, $B$ of\/ $\Type_{\Sigma}(\emptyset)$, and a partial substitution $\rho \colon \Xi \to \Type_{\Sigma}(\emptyset)$, 
+For the sake of clarity in presentation, though, we keep using named variables and just assume that $\Xi$ has a decidable equality.}
+\end{remark}
+We need some auxiliary definitions for the notion of extension to state the unification problem:
+\begin{definition}
+  Let $\rho$ be a partial substitution.
+  Then, by an \emph{extension}\/ $\sigma \geq \rho$ we mean that the domain of $\sigma$ contains the domain $\rho$ and\/ $\sigma(x) = \rho(x)$ for every\/ $x$ in the domain of $\rho$; by a \emph{minimal extension}\/ $\bar{\rho}$ of $\rho$ satisfying $P$ we mean an extension $\bar{\rho} \geq \rho$ with $P(\bar{\rho})$ such that $\sigma \geq \bar{\rho}$ whenever $\sigma \geq \rho$ and $P(\sigma)$.
+\end{definition}
+\begin{lemma}\label{lem:unify}
+  For any\/ $A$ of\/ $\Type_{\Sigma}(\Xi)$, $B$ of\/ $\Type_{\Sigma}(\emptyset)$, and a partial substitution\/ $\rho \colon \Xi \to \Type_{\Sigma}(\emptyset)$, 
   \begin{enumerate}
-    \item either there is a \emph{minimal extension} $\bar{\rho}$ of $\rho$ such that $\simsub{A}{\bar{\rho}} = B$, or 
-    \item there is no extension $\sigma$ of $\rho$ such that $\simsub{A}{\sigma} = B$
+    \item either there is a minimal extension\/ $\bar{\rho}$ of\/ $\rho$ such that\/ $\simsub{A}{\bar{\rho}} = B$, or 
+    \item there is no extension\/ $\sigma$ of\/ $\rho$ such that\/ $\simsub{A}{\sigma} = B$
   \end{enumerate}
-  where by an extension $\sigma \geq \rho$ we mean that the domain of $\sigma$ contains the domain $\rho$ and $\sigma(x) = \rho(x)$ for every $x$ in the domain of $\rho$; by a minimal extension $\bar{\rho}$ of $\rho$ satisfying $P$ we mean an extension $\bar{\rho}$ of $\rho$ with $P(\bar{\rho})$ such that $\sigma \geq \bar{\rho}$ whenever $\sigma$ is an extension of $\rho$ with $P(\sigma)$.
 \end{lemma}
@@
   The two cases \SynRule{Var} and \SynRule{Anno} are straightforward and have nothing to do with mode-correctness.
-  The case \ChkRule{Sub} invokes the uniqueness of synthesised types to refute the case that $\Gamma |- \isTerm{t} \syn B$ but $A \neq B$ for a given type $A$.
+  The case \ChkRule{Sub} invokes the uniqueness of synthesised types to refute the case that $\Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn B$ but $A \neq B$ for a given type $A$.
   The first three cases follow essentially the same argument provided by \citet{Wadler2022}, but we still give the arguments here for the sake of completeness.
-  The last case \Rule{Op} is new and has been discussed informally above.
+  The last case \Rule{Op} is new and has been discussed above.
+  (For brevity we omit the subscript $(\Sigma, \Omega)$.)
   \begin{itemize}
-    \item \SynRule{Var}: As $\erase{\Gamma} |-_{\Sigma, \Omega} \isTerm{t}$, there exists a type $A$ such that $\Gamma |- \isTerm{x} \syn A$.
+    \item \SynRule{Var}: As $\erase{\Gamma} |- \isTerm{t}$, there exists a type $A$ such that $\Gamma |- \isTerm{x} \syn A$.
 
-    \item \SynRule{Anno}: For $\erase{\Gamma} |-_{\Sigma, \Omega} (\isTerm{t} \annotate A)^{\syn}$, it is decidable whether $\Gamma |- \isTerm{t} \chk A$ by induction hypothesis.
+    \item \SynRule{Anno}: For $\erase{\Gamma} |- (\isTerm{t} \annotate A)^{\syn}$, it is decidable whether $\Gamma |- \isTerm{t} \chk A$ by induction hypothesis.
       \begin{itemize}
@@
 
-    \item \ChkRule{Sub}: If $\erase{\Gamma} |-_{\Sigma, \Omega}t^{\chk}$ by the rule $\ChkRule{Sub}$, then $\Gamma |-_{\Sigma, \Omega} t^{\syn}$ by inversion.
-      We show that it is decidable for any $A$ whether $\Gamma |-_{\Sigma, \Omega} t \chk A$.
+    \item \ChkRule{Sub}: If $\erase{\Gamma} |-t^{\chk}$ by the rule $\ChkRule{Sub}$, then $\Gamma |- t^{\syn}$ by inversion.
+      We show that it is decidable for any $A$ whether $\Gamma |- t \chk A$.
       By induction hypothesis, it is decidable whether $\Gamma |- t \syn B$ for some $B$:
       \begin{itemize}
-        \item If $\Gamma |/- t \syn C$ for any type $C$ but $\Gamma |- t \chk A$ is derivable, then by inversion $\Gamma |- t \syn B$ for some $B = A$, thus a contradiction.
+        \item If $\Gamma |/- t \syn C$ for any $C$ but $\Gamma |- t \chk A$, then by inversion $\Gamma |- t \syn B$ for some $B = A$, thus a contradiction.
         \item If $\Gamma |- t \syn B$ for some $B$, then either $A = B$ or $A \neq B$ by the decidable equality on $\Type_{\Sigma}(\Xi)$: 
@@
     \item \Rule{Op}:
-      Consider the mode derivation of\/ $\erase{\Gamma} |-_{\Sigma, \Omega} \tmOpts^{\dir{d}}$.
-%
-%  \begin{claim}\label{lem:args-induction}
-%    For the argument list $\overrightarrow{[\Delta_i] A_{i}^{\dir{d_i}}}$, \emph{partial} substitution $\rho$ from $\Xi$ to $\emptyset$
-%    \begin{enumerate}
-%      \item either there is a minimal extension $\ext{\rho}$ of $\rho$ defined on $\synvar(\mathit{as})$ such that all of following judgements are derivable
-%        \[
-%          \Gamma, \vars{x}_\isTerm{i} : \simsub{\Delta_i}{\ext{\rho}} |-_{\Sigma, \Omega} \isTerm{t_i} \colon \simsub{A_i}{\ext{\rho}}^{\dir{d_i}}
-%        \]
-%
-%      \item or there is no extension $\sigma$ of $\rho$ such that all $\Gamma, \vars{x}_\isTerm{i} : \simsub{\Delta_i}{\sigma} |-_{\Sigma, \Omega} \isTerm{t_i} \colon \simsub{A_i}{\sigma}^{\dir{d_i}}$ has a typing derivation.
-%    \end{enumerate}
-%  \end{claim}
+      Consider the mode derivation of $\erase{\Gamma} |- \tmOpts^{\dir{d}}$.
+      First we claim:
+      \begin{claim}\label{lem:args-induction}
+        For an argument list $\as = \biargvec$ and any partial substitution $\rho$ from $\Xi$ to $\emptyset$
+        \begin{enumerate}
+          \item either there is a minimal extension $\ext{\rho}$ of $\rho$ such that 
+            \begin{equation} \label{eq:claim}
+              \text{$\ext{\rho}$ is defined on $\synvar(\as)$ and\/} \quad
+              \Gamma, \vars{x}_\isTerm{i} : \simsub{\Delta_i}{\ext{\rho}} |- \isTerm{t_i} \colon \simsub{A_i}{\ext{\rho}}^{\dir{d_i}}
+              \quad\text{for all $i$};
+            \end{equation}
+          \item or there is no extension $\sigma$ of $\rho$ such that \eqref{eq:claim} holds.
+        \end{enumerate}
+      \end{claim}
+      Then, case analysis on $\dir{d}$:
+      \begin{itemize}
+        \item $\dir{d}$ is $\syn$: We apply our claim with the partial substitution $\rho_0$ defined nowhere.
+          \begin{enumerate}
+            \item If there is no $\sigma \geq \rho$ such that \eqref{eq:claim} holds but for some $A$ we have $\Gamma |- \tmOpts \syn A$, then by inversion we have a substitution $\rho\colon \Sub{\Xi}{\emptyset}$ such that ${\Gamma, \vars{x}_\isTerm{i} : \simsub{\Delta_i}{\rho} |- \isTerm{t_i} \colon \simsub{A_i}{\rho}^{\dir{d_i}}}$ for every $i$.
+              Obviously, $\rho$ is an extension of $\rho_0$ such that ${\Gamma, \vars{x}_\isTerm{i} : \simsub{\Delta_i}{\rho} |- \isTerm{t_i} \colon \simsub{A_i}{\rho}^{\dir{d_i}}}$ for every $i$ and it contradicts the assumption that no such an extension exists.
+
+            \item If there exists a minimal $\ext{\rho} \geq \rho_0$ defined on $\synvar(\biargs)$ such that \eqref{eq:claim} holds, then by mode-correctness we have $\ext{\rho}$ is total and
+              \[
+                \Gamma |- \tmOpts \syn \simsub{A_0}{\ext{\rho}}.
+              \]
+          \end{enumerate}
+
+        \item $\dir{d}$ is $\chk$: Let $A$ be a type and apply \Cref{lem:unify} with $\rho_0$ defined nowhere.
+          \begin{enumerate}
+            \item If there is no $\sigma \geq \rho_0$ such that $\simsub{A_0}{\sigma} = A$ but $\Gamma |- \tmOpts \chk A$, then by inversion there is a substitution $\rho$ such that $A = \simsub{A_0}{\rho}$, thus a contradiction.
+            \item If there is a minimal $\ext{\rho} \geq \rho_0$ such that $\simsub{A_0}{\ext{\rho}} = A$, then apply our claim with $\ext{\rho}$:
+              \begin{enumerate}
+                \item If there is no $\sigma \geq \ext{\rho}$ satisfying \eqref{eq:claim} but $\Gamma |- \tmOpts \chk A$, then by inversion there is $\gamma$ such that $\simsub{A_0}{\gamma} = A$ and also ${\Gamma, \vars{x}_\isTerm{i} : \simsub{\Delta_i}{\gamma} |- \isTerm{t_i} \colon \simsub{A_i}{\gamma}^{\dir{d_i}}}$ for every $i$.
+                  Given that $\ext{\rho} \geq \rho$ is minimal such that $\simsub{A_0}{\ext{\rho}} = A$, then $\gamma$ is an extension of $\ext{\rho}$ but by assumption no such an extension satisfying ${\Gamma, \vars{x}_\isTerm{i} : \simsub{\Delta_i}{\gamma} |- \isTerm{t_i} \colon \simsub{A_i}{\gamma}^{\dir{d_i}}}$ exists, thus a contradiction.
+
+                \item If there is a minimal $\ext{\ext{\rho}} \geq \ext{\rho}$ such that \eqref{eq:claim}, then by mode-correctness $\ext{\ext{\rho}}$ is total and
+                  \[
+                    \Gamma |- \tmOpts \chk \simsub{A_0}{\ext{\ext{\rho}}}
+                  \]
+                  where $\simsub{A_0}{\ext{\ext{\rho}}} = \simsub{A_0}{\ext{\rho}} = A$ since $\ext{\ext{\rho}}(x) = \ext{\rho}$ for every $x$ in the domain of $\ext{\rho}$.
+              \end{enumerate}
+          \end{enumerate}
+      \end{itemize}
+      \begin{claimproof}
+        For a partial substitution $\rho$, we prove it by induction on the list $\biargvec$:
+        \begin{enumerate}
+          \item For the empty list, $\rho$ is the minimal extension of $\rho$ itself satisfying \eqref{eq:claim} trivially. 
+          \item For $\biargvec, \biarg[m+1]$, by induction hypothesis on the list, we have two cases:
+            \begin{enumerate}
+              \item If there is no $\sigma \geq \rho$ such that \eqref{eq:claim} holds for all $1 \leq i \leq m$ but a minimal $\gamma \geq \rho$ such that~\eqref{eq:claim} holds for all $1 \leq i \leq m + 1$, then we clearly have a contradiction.
+              \item There is a minimal $\ext{\rho} \geq \rho$ such that \eqref{eq:claim} holds for all $1 \leq i \leq m$.
+                By case analysis on $\dir{d_{m+1}}$:
+                \begin{itemize}
+                  \item $\dir{d_{m+1}}$ is $\chk$: By mode-correctness, $\simsub{\Delta_{m+1}}{\ext{\rho}}$ and $\simsub{A_{m+1}}{\ext{\rho}}$ are defined and by the induction hypothesis it is decidable whether $\Gamma, \vars{x}_\isTerm{m+1} : \simsub{\Delta_{m+1}}{\ext{\rho}} |- \isTerm{t_{m+1}} \chk \simsub{A_{m+1}}{\ext{\rho}}^{\dir{d_{m+1}}}$.
+                    Clearly, if $\Gamma, \vars{x}_\isTerm{m+1} : \simsub{\Delta_{m+1}}{\ext{\rho}} |- \isTerm{t_{m+1}} \chk \simsub{A_{m+1}}{\ext{\rho}}^{\dir{d_{m+1}}}$ then the desired statement is proved; otherwise we can also easily derive a contradiction.
+
+                  \item $\dir{d_{m+1}}$ is $\syn$: By mode-correctness, $\simsub{\Delta_{m+1}}{\ext{\rho}}$ is defined and by the induction hypothesis, it is decidable that $\Gamma, \vars{x}_\isTerm{m+1} : \simsub{\Delta_{m+1}}{\ext{\rho}} |- \isTerm{t_{m+1}} \syn A$ for some $A$.
+                    \begin{enumerate}
+                      \item If $\Gamma, \vars{x}_\isTerm{m+1} : \simsub{\Delta_{m+1}}{\ext{\rho}} |/- \isTerm{t_{m+1}} \syn A$ for any $A$ but there is $\gamma \geq \rho$ s.t.\ \eqref{eq:claim} holds for $1 \leq i \leq m+1$, then $\gamma \geq \ext{\rho}$.
+                        Therefore $\simsub{\Delta_{m+1}}{\ext{\rho}} = \simsub{\Delta_{m+1}}{\gamma}$ and we derive a contradiction because $\Gamma, \vars{x}_\isTerm{m+1} : \simsub{\Delta_{m+1}}{\ext{\rho}} |- \isTerm{t_{m+1}} \syn \simsub{A_{m+1}}{\gamma}$.
+                      \item If $\Gamma, \vars{x}_\isTerm{m+1} : \simsub{\Delta_{m+1}}{\ext{\rho}} |- \isTerm{t_{m+1}} \syn A$ for some $A$, then by \Cref{lem:unify} we have two cases: %we unify $A$ with $A_{m+1}$ extending $\ext{\rho}$.
+                        \begin{itemize}
+                          \item Suppose no $\sigma \geq \ext{\rho}$ such that $\simsub{A_{m+1}}{\sigma} = A$ but an extension $\gamma \geq \rho$ such that \eqref{eq:claim} holds for $1 \leq i \leq m + 1$.
+                            Then, $\gamma \geq \ext{\rho}$ by the minimality of $\ext{\rho}$ and thus
+                            $\Gamma, \vars{x}_\isTerm{m+1} : \simsub{\Delta_{m+1}}{\ext{\rho}} |- \isTerm{t_{m+1}} \syn \simsub{A_{m+1}}{\gamma}$.
+                            However, by \Cref{thm:unique-syn}, the synthesised type $\simsub{A_{m+1}}{\gamma}$ must be unique, so $\gamma$ is an extension of $\ext{\rho}$ such that $\simsub{A_{m+1}}{\gamma} = A$, leading to a contradiction.
+                          \item If there is a minimal $\ext{\ext{\rho}}$ such that $\simsub{A_{m+1}}{\ext{\ext{\rho}}} = A$, then it is now hard to show that $\ext{\ext{\rho}}$ is the minimal extension of $\rho$ such that \eqref{eq:claim} holds for all $1 \leq i \leq m + 1$.
+                        \end{itemize}
+                    \end{enumerate}
+                \end{itemize}
+            \end{enumerate}
+        \end{enumerate}
+        \hfill{\footnotesize$\blacksquare$}
+      \end{claimproof}
+      We now complete the decidability proof.
   \end{itemize}
 \end{proof}
+While the proof presented above is complex, its formal counterpart essentially serves as two programs for checking and synthesis.
+These programs either compute the typing derivation or provide a proof of contradiction.
+Each case analysis simply branches based on the outcomes of bidirectional type synthesis and checking for each subterm, as well as the unification.
+If a proof of contradiction is not of interest for implementation, the programs can be simplified by disregarding the cases that provide such contradiction proofs.
+On the other hand, the contradiction proof contains the information as to why the construction of a derivation failed.
+This can potentially be restructured further to provide informative error messages, assisting users in resolving issues with ill-typed terms, instead of merely answering a blatant `no'.
 
 \subsection{Trichotomy on Raw Terms by Type Synthesis} \label{subsec:trichotomy}
-
-%\begin{corollary}[Decidability of Bidirectional Type Synthesis] \label{thm:bidirectional-type-synthesis}
-%  In a mode-correct bidirectional type system $(\Sigma, \Omega)$,
-%  it is decidable whether there exists some $A$ such that
-%  \[
-%    \Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn A.
-%  \]
-%\end{corollary}
 

### 2023-07-09 tex/sec6-formalisation.tex
[ tex ] Some code in S6

@@
 \DeclareUnicodeCharacter{2294}{\ensuremath{\sqcup}}
+
+\DeclareUnicodeCharacter{1D43}{\ensuremath{^a}}
+\DeclareUnicodeCharacter{1D9C}{\ensuremath{^c}}
+\DeclareUnicodeCharacter{02E2}{\ensuremath{^s}}
+
 \DeclareUnicodeCharacter{2080}{\ensuremath{_0}}
@@
 \DeclareUnicodeCharacter{2115}{\ensuremath{\mathbb{N}}}
+\DeclareUnicodeCharacter{2208}{\ensuremath{\in}}
 \DeclareUnicodeCharacter{2236}{:}
+\DeclareUnicodeCharacter{2261}{\ensuremath{\equiv}}
 \DeclareUnicodeCharacter{2237}{\ensuremath{\mathrel{::}}}
+\DeclareUnicodeCharacter{0393}{\ensuremath{\Gamma}}
+\DeclareUnicodeCharacter{0394}{\ensuremath{\Delta}}
 \DeclareUnicodeCharacter{03A3}{\ensuremath{\Sigma}}
@@
   \small
+  \begin{minipage}[t]{.3\linewidth}
   \begin{hscode}\SaveRestoreHook
@@
 \column{5}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{7}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{14}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{16}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{19}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{24}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{12}{@{}>{\hspre}l<{\hspost}@{}}%
 \column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[3]{}\Keyword{record}\;\Conid{ArgD}\;(\Conid{Ξ}\;\mathbin{:}\;\Conid{ℕ})\;\mathbin{:}\;\Conid{Set}\;\Keyword{where}{}\<[E]%
+\>[B]{}\Keyword{record}\;\Conid{ArgD}\;(\Conid{Ξ}\;\mathbin{:}\;\Conid{ℕ})\;\mathbin{:}\;\Conid{Set}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Keyword{where}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Keyword{field}{}\<[E]%
 \\
 \>[3]{}\hsindent{2}{}\<[5]%
-\>[5]{}\Keyword{field}{}\<[E]%
-\\
-\>[5]{}\hsindent{2}{}\<[7]%
-\>[7]{}\Varid{cxt}\;{}\<[14]%
-\>[14]{}\mathbin{:}\;\Conid{Cxt}\;\Conid{Ξ}{}\<[E]%
-\\
-\>[5]{}\hsindent{2}{}\<[7]%
-\>[7]{}\Varid{mode}\;{}\<[14]%
-\>[14]{}\mathbin{:}\;\Conid{Mode}{}\<[E]%
-\\
-\>[5]{}\hsindent{2}{}\<[7]%
-\>[7]{}\Varid{type}\;{}\<[14]%
-\>[14]{}\mathbin{:}\;\Conid{TExp}\;\Conid{Ξ}{}\<[E]%
-\\[\blanklineskip]%
-\>[3]{}\Conid{ArgsD}\;\mathbin{:}\;\Conid{ℕ}\;\Varid{→}\;\Conid{Set}{}\<[E]%
-\\
-\>[3]{}\Conid{ArgsD}\;\Conid{Ξ}\;\mathrel{=}\;\Conid{List}\;(\Conid{ArgD}\;\Conid{Ξ}){}\<[E]%
-\\[\blanklineskip]%
-\>[3]{}\Keyword{record}\;\Conid{ConD}\;\mathbin{:}\;\Conid{Set}\;\Keyword{where}{}\<[E]%
+\>[5]{}\Varid{cxt}\;{}\<[12]%
+\>[12]{}\mathbin{:}\;\Conid{Cxt}\;\Conid{Ξ}{}\<[E]%
 \\
 \>[3]{}\hsindent{2}{}\<[5]%
-\>[5]{}\Keyword{constructor}\;\Varid{ι}{}\<[E]%
+\>[5]{}\Varid{type}\;{}\<[12]%
+\>[12]{}\mathbin{:}\;\Conid{TExp}\;\Conid{Ξ}{}\<[E]%
 \\
 \>[3]{}\hsindent{2}{}\<[5]%
-\>[5]{}\Keyword{field}{}\<[E]%
+\>[5]{}\Varid{mode}\;{}\<[12]%
+\>[12]{}\mathbin{:}\;\Conid{Mode}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+  \end{minipage}
+  \begin{minipage}[t]{.32\linewidth}
+  \begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{14}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{22}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Keyword{record}\;\Conid{ConD}\;\mathbin{:}\;\Conid{Set}\;\Keyword{where}{}\<[E]%
 \\
-\>[5]{}\hsindent{2}{}\<[7]%
-\>[7]{}\{\mskip1.5mu \Varid{vars}\mskip1.5mu\}\;{}\<[16]%
-\>[16]{}\mathbin{:}\;\Conid{ℕ}{}\<[E]%
-\\
-\>[5]{}\hsindent{2}{}\<[7]%
-\>[7]{}\Varid{mode}\;{}\<[16]%
-\>[16]{}\mathbin{:}\;\Conid{Mode}{}\<[E]%
-\\
-\>[5]{}\hsindent{2}{}\<[7]%
-\>[7]{}\Varid{type}\;{}\<[16]%
-\>[16]{}\mathbin{:}\;\Conid{TExp}\;{}\<[24]%
-\>[24]{}\Varid{vars}{}\<[E]%
-\\
-\>[5]{}\hsindent{2}{}\<[7]%
-\>[7]{}\Varid{args}\;{}\<[16]%
-\>[16]{}\mathbin{:}\;\Conid{ArgsD}\;\Varid{vars}{}\<[E]%
-\\[\blanklineskip]%
-\>[3]{}\Keyword{record}\;\Conid{Desc}\;\mathbin{:}\;\Conid{Set₁}\;\Keyword{where}{}\<[E]%
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Keyword{field}{}\<[E]%
 \\
 \>[3]{}\hsindent{2}{}\<[5]%
-\>[5]{}\Keyword{constructor}\;\Varid{desc}{}\<[E]%
+\>[5]{}\{\mskip1.5mu \Varid{vars}\mskip1.5mu\}\;{}\<[14]%
+\>[14]{}\mathbin{:}\;\Conid{ℕ}{}\<[E]%
 \\
 \>[3]{}\hsindent{2}{}\<[5]%
-\>[5]{}\Keyword{field}{}\<[E]%
+\>[5]{}\Varid{args}\;{}\<[14]%
+\>[14]{}\mathbin{:}\;\Conid{List}\;(\Conid{ArgsD}\;\Varid{vars}){}\<[E]%
 \\
-\>[5]{}\hsindent{2}{}\<[7]%
-\>[7]{}\Conid{Op}\;{}\<[19]%
-\>[19]{}\mathbin{:}\;\Conid{Set}{}\<[E]%
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\Varid{type}\;{}\<[14]%
+\>[14]{}\mathbin{:}\;\Conid{TExp}\;{}\<[22]%
+\>[22]{}\Varid{vars}{}\<[E]%
 \\
-\>[5]{}\hsindent{2}{}\<[7]%
-\>[7]{}\{\kern-.9pt\vrule width .75pt height 7.125pt depth 1.975pt\kern-1.5pt\;\Varid{decOp}\;\kern-1.5pt\vrule width .75pt height 7.125pt depth 1.975pt\kern-.9pt\}\;{}\<[19]%
-\>[19]{}\mathbin{:}\;\Conid{DecEq}\;\Conid{Op}{}\<[E]%
-\\
-\>[5]{}\hsindent{2}{}\<[7]%
-\>[7]{}\Varid{rules}\;{}\<[19]%
-\>[19]{}\mathbin{:}\;\Conid{Op}\;\Varid{→}\;\Conid{ConD}{}\<[E]%
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\Varid{mode}\;{}\<[14]%
+\>[14]{}\mathbin{:}\;\Conid{Mode}{}\<[E]%
 \ColumnHook
 \end{hscode}\resethooks
-  \caption{Definition of Bidirectional Binding Signature}  
+  \end{minipage}
+  \begin{minipage}[t]{.3\linewidth}
+   \begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{17}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Keyword{record}\;\Conid{Desc}\;\mathbin{:}\;\Conid{Set₁}\;\Keyword{where}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Keyword{field}{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\Conid{Op}\;{}\<[17]%
+\>[17]{}\mathbin{:}\;\Conid{Set}{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\Varid{rules}\;{}\<[17]%
+\>[17]{}\mathbin{:}\;\Conid{Op}\;\Varid{→}\;\Conid{ConD}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+  \end{minipage}
+
+  \caption{Definition of bidirectional binding signature}  
 \end{figure}
 
+{\small\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{7}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{8}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{10}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{20}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{24}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{27}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{30}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{33}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{36}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{46}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Conid{Fam}\;\mathbin{:}\;(\Varid{ℓ′}\;\mathbin{:}\;\Conid{Level})\;(\Conid{X}\;\mathbin{:}\;\Conid{R.Fam}\;\Varid{ℓ})\;\Varid{→}\;\Conid{Set}\;(\Varid{ℓ}\;\Varid{⊔}\;\Varid{lsuc}\;\Varid{ℓ′}){}\<[E]%
+\\
+\>[B]{}\Conid{Fam}\;\Varid{ℓ′}\;\Conid{X}\;\mathrel{=}\;(\Conid{Γ}\;\mathbin{:}\;\Conid{Cxt}\;\Varid{0})\;\Varid{→}\;\Conid{X}\;(\Varid{length}\;\Conid{Γ})\;\Varid{→}\;\Conid{Mode}\;\Varid{→}\;\Conid{Ty}\;\Varid{→}\;\Conid{Set}\;\Varid{ℓ′}{}\<[E]%
+\\[\blanklineskip]%
+\>[B]{}\Varid{⟦\char95 ⟧ᵃ}\;{}\<[7]%
+\>[7]{}\mathbin{:}\;(\Conid{Δ}\;\mathbin{:}\;\Conid{TExps}\;\Conid{Ξ})\;(\Conid{X}\;\mathbin{:}\;\Conid{R.Fam}\;\Varid{ℓ})\;(\Conid{Y}\;\mathbin{:}\;(\Conid{Γ}\;\mathbin{:}\;\Conid{Cxt}\;\Varid{0})\;\Varid{→}\;\Conid{X}\;(\Varid{length}\;\Conid{Γ})\;\Varid{→}\;\Conid{Set}\;\Varid{ℓ′})\;{}\<[E]%
+\\
+\>[7]{}\Varid{→}\;\Conid{TSub}\;\Conid{Ξ}\;\Varid{0}\;\Varid{→}\;(\Conid{Γ}\;\mathbin{:}\;\Conid{Cxt}\;\Varid{0})\;\Varid{→}\;\Conid{R.⟦}\;\Conid{Δ}\;\Varid{⟧ᵃ}\;\Conid{X}\;(\Varid{length}\;\Conid{Γ})\;\Varid{→}\;\Conid{Set}\;\Varid{ℓ′}{}\<[E]%
+\\
+\>[B]{}\Varid{⟦}\;\Varid{[]}\;{}\<[10]%
+\>[10]{}\Varid{⟧ᵃ}\;\Conid{X}\;\Conid{Y}\;\Varid{σ}\;\Conid{Γ}\;\Varid{t}\;{}\<[24]%
+\>[24]{}\mathrel{=}\;\Conid{Y}\;\Conid{Γ}\;\Varid{t}{}\<[E]%
+\\
+\>[B]{}\Varid{⟦}\;\Conid{A}\;\Varid{∷}\;\Conid{Δ}\;{}\<[10]%
+\>[10]{}\Varid{⟧ᵃ}\;\Conid{X}\;\Conid{Y}\;\Varid{σ}\;\Conid{Γ}\;\Varid{t}\;{}\<[24]%
+\>[24]{}\mathrel{=}\;\Varid{⟦}\;\Conid{Δ}\;\Varid{⟧ᵃ}\;\Conid{X}\;\Conid{Y}\;\Varid{σ}\;((\Conid{A}\;\Varid{⟨}\;\Varid{σ}\;\Varid{⟩})\;\Varid{∷}\;\Conid{Γ})\;\Varid{t}{}\<[E]%
+\\[\blanklineskip]%
+\>[B]{}\Varid{⟦\char95 ⟧ᵃˢ}\;{}\<[8]%
+\>[8]{}\mathbin{:}\;(\Conid{D}\;\mathbin{:}\;\Conid{ArgsD}\;\Conid{Ξ})\;(\Conid{X}\;\mathbin{:}\;\Conid{R.Fam}\;\Varid{ℓ})\;(\Conid{Y}\;\mathbin{:}\;\Conid{Fam}\;\Varid{ℓ′}\;\Conid{X})\;{}\<[E]%
+\\
+\>[8]{}\Varid{→}\;\Conid{TSub}\;\Conid{Ξ}\;\Varid{0}\;\Varid{→}\;(\Conid{Γ}\;\mathbin{:}\;\Conid{Cxt}\;\Varid{0})\;\Varid{→}\;\Conid{R.⟦}\;\Varid{eraseᵃˢ}\;\Conid{D}\;\Varid{⟧ᵃˢ}\;\Conid{X}\;(\Varid{length}\;\Conid{Γ})\;\Varid{→}\;\Conid{Set}\;\Varid{ℓ′}{}\<[E]%
+\\
+\>[B]{}\Varid{⟦}\;\Varid{[]}\;{}\<[20]%
+\>[20]{}\Varid{⟧ᵃˢ}\;\anonymous \;{}\<[27]%
+\>[27]{}\anonymous \;{}\<[30]%
+\>[30]{}\anonymous \;{}\<[33]%
+\>[33]{}\anonymous \;{}\<[36]%
+\>[36]{}\anonymous \;{}\<[46]%
+\>[46]{}\mathrel{=}\;\Varid{⊤}{}\<[E]%
+\\
+\>[B]{}\Varid{⟦}\;\Conid{Δ}\;\Varid{⊢[}\;\Varid{d}\;\mskip1.5mu]\;\Conid{A}\;\Varid{∷}\;\Conid{Ds}\;{}\<[20]%
+\>[20]{}\Varid{⟧ᵃˢ}\;\Conid{X}\;{}\<[27]%
+\>[27]{}\Conid{Y}\;{}\<[30]%
+\>[30]{}\Varid{σ}\;{}\<[33]%
+\>[33]{}\Conid{Γ}\;{}\<[36]%
+\>[36]{}(\Varid{x}\;\Varid{,}\;\Varid{xs})\;{}\<[46]%
+\>[46]{}\mathrel{=}\;{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{⟦}\;\Conid{Δ}\;\Varid{⟧ᵃ}\;\Conid{X}\;(\Varid{λ}\;\Conid{Γ'}\;\Varid{x'}\;\Varid{→}\;\Conid{Y}\;\Conid{Γ'}\;\Varid{x'}\;\Varid{d}\;(\Conid{A}\;\Varid{⟨}\;\Varid{σ}\;\Varid{⟩}))\;\Varid{σ}\;\Conid{Γ}\;\Varid{x}\;\Varid{×}\;\Varid{⟦}\;\Conid{Ds}\;\Varid{⟧ᵃˢ}\;\Conid{X}\;\Conid{Y}\;\Varid{σ}\;\Conid{Γ}\;\Varid{xs}{}\<[E]%
+\\[\blanklineskip]%
+\>[B]{}\Varid{⟦\char95 ⟧ᶜ}\;\mathbin{:}\;(\Conid{D}\;\mathbin{:}\;\Conid{ConD})\;(\Conid{X}\;\mathbin{:}\;\Conid{R.Fam}\;\Varid{ℓ})\;(\Conid{Y}\;\mathbin{:}\;\Conid{Fam}\;\Varid{ℓ′}\;\Conid{X})\;\Varid{→}\;\Conid{Fam}\;\Varid{ℓ′}\;(\Conid{R.⟦}\;\Varid{eraseᶜ}\;\Conid{D}\;\Varid{⟧ᶜ}\;\Conid{X}){}\<[E]%
+\\
+\>[B]{}\Varid{⟦}\;\Varid{ι}\;\{\mskip1.5mu \Conid{Ξ}\mskip1.5mu\}\;\Varid{d}\;\Conid{B}\;\Conid{D}\;\Varid{⟧ᶜ}\;\Conid{X}\;\Conid{Y}\;\Conid{Γ}\;\Varid{xs}\;\Varid{d′}\;\Conid{A}\;\mathrel{=}\;{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{d}\;\Varid{≡}\;\Varid{d′}\;\Varid{×}\;\Conid{Σ[}\;\Varid{σ}\;\Varid{∈}\;\Conid{TSub}\;\Conid{Ξ}\;\Varid{0}\;\mskip1.5mu]\;\Conid{B}\;\Varid{⟨}\;\Varid{σ}\;\Varid{⟩}\;\Varid{≡}\;\Conid{A}\;\Varid{×}\;\Varid{⟦}\;\Conid{D}\;\Varid{⟧ᵃˢ}\;\Conid{X}\;\Conid{Y}\;\Varid{σ}\;\Conid{Γ}\;\Varid{xs}{}\<[E]%
+\\[\blanklineskip]%
+\>[B]{}\Varid{⟦\char95 ⟧}\;\mathbin{:}\;(\Conid{D}\;\mathbin{:}\;\Conid{Desc})\;(\Conid{X}\;\mathbin{:}\;\Conid{R.Fam}\;\Varid{ℓ})\;(\Conid{Y}\;\mathbin{:}\;\Conid{Fam}\;\Varid{ℓ′}\;\Conid{X})\;\Varid{→}\;\Conid{Fam}\;\Varid{ℓ′}\;(\Conid{R.⟦}\;\Varid{erase}\;\Conid{D}\;\Varid{⟧}\;\Conid{X}){}\<[E]%
+\\
+\>[B]{}\Varid{⟦}\;\Conid{D}\;\Varid{⟧}\;\Conid{X}\;\Conid{Y}\;\Conid{Γ}\;(\Varid{i}\;\Varid{,}\;\Varid{xs})\;\Varid{d}\;\Conid{A}\;\mathrel{=}\;\Varid{⟦}\;\Conid{D}\;\Varid{.rules}\;\Varid{i}\;\Varid{⟧ᶜ}\;\Conid{X}\;\Conid{Y}\;\Conid{Γ}\;\Varid{xs}\;\Varid{d}\;\Conid{A}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+}
+
 \end{document}

### 2023-07-09 tex/sec5-synthesis.tex
[ tex ] Revise the remark on the decidability proof for bidirectional type synthesis and checking

@@
 \end{proof}
-While the proof presented above is complex, its formal counterpart essentially serves as two programs for checking and synthesis.
+While the proof presented above is intricate, its formal counterpart essentially functions as two programs for type checking and synthesis.
 These programs either compute the typing derivation or provide a proof of contradiction.
-Each case analysis simply branches based on the outcomes of bidirectional type synthesis and checking for each subterm, as well as the unification.
-If a proof of contradiction is not of interest for implementation, the programs can be simplified by disregarding the cases that provide such contradiction proofs.
-On the other hand, the contradiction proof contains the information as to why the construction of a derivation failed.
-This can potentially be restructured further to provide informative error messages, assisting users in resolving issues with ill-typed terms, instead of merely answering a blatant `no'.
+Each case analysis simply branches depending on the outcomes of bidirectional type synthesis and checking for each subterm, as well as the unification process.
+If a contradiction proof is not of interest for implementation, the programs can be simplified by disregarding the cases that yield such contradiction proofs.
+Alternatively, we could consider generalising typing derivations like our generalised mode derivations (\Cref{fig:generalised-mode-derivations}).
+This could accommodate ill-typed cases and improve contradiction proofs to deliver more informative error messages.
+This would assist users in resolving issues with ill-typed terms, rather than simply returning a blatant 'no'.
 

### 2023-07-10 tex/BiSig.tex
[ tex ] ‘Beyond syntax-directed type systems’ in S7
[ tex ] Done with S5 for now
[ tex ] Turn on `anonymous` and `review`

@@
-\documentclass[acmsmall,screen]{acmart}
+\documentclass[acmsmall,screen,review,anonymous,draft]{acmart}
 
@@
 \AtEndPreamble{%
+\newtheorem*{claim}{Claim}
 \theoremstyle{acmdefinition}
+\newtheorem*{claimproof}{Proof of Claim}
 \newtheorem{remark}[theorem]{Remark}
-\newtheorem*{claim}{Claim}
-\newtheorem{claimproof}{Proof of Claim}
 }
@@
 We thank Kuen-Bang {Hou (Favonia)} for his comments and suggestions.
-The work is supported by the Ministry of Science and Technology of Taiwan under grant MOST 109-2222-E-001-002-MY3.
+The work is supported by the \grantsponsor{NSTC}{National Science and Technology Council of Taiwan}{https://www.nstc.gov.tw/} under grant \grantnum{NSTC}{NSTC 112-2221-E-001-003-MY3}.
 \end{acks}

### 2023-07-10 tex/sec1-intro.tex
[ tex ] Minor changes

@@
 Instead of focusing solely on a bidirectional type synthesiser like \citet{Wadler2022}~(\cref{sec:PLFA}), our theory centres around a general specification of type synthesis for an ordinary type system where judgements are not assigned modes, and fills in the gap between language formalisation frameworks based on such type systems~(\cref{sec:language-formalisation}) and raw terms produced by parsers.
-Working similarly to those frameworks, our theory is parametrised by a \emph{bidirectional binding signature}, which describes a language with a bidirectional type system, analogously to a grammar describing a formal language in the theory of parsing.
+Working similarly to those frameworks, our theory is parametrised by a \emph{bidirectional binding signature}, which extends \citeauthor*{Aczel1978}'s binding signature and describes a language with a bidirectional type system, analogously to a grammar describing a formal language in the theory of parsing.
 As a first step, we confine our theory to \emph{syntax-directed} systems, meaning that for every raw term construct, the ordinary type system has exactly one typing rule, which is assigned modes in exactly one way in the bidirectional type system.%

### 2023-07-10 tex/sec3-defs.tex
[ tex ] Minor changes

@@
   \item first we introduce a notion of arity and a notion of signature which includes a set of operation symbols and an assignment of arities to symbols;
-\item then, given a signature, we define sets of raw terms and derivations by primitive rules such as $\Rule{Var}$ and a rule schema for constructs $\tmOp_o$ indexed by a operational symbol~$o$.
+\item then, given a signature, we define sets of raw terms and derivations by primitive rules such as $\Rule{Var}$ and a rule schema for constructs $\tmOp_o$ indexed by an operational symbol~$o$.
 \end{enumerate*}
@@
 We need some information to compare types on the object level during type synthesis and substitute those type variables in a typing derivation $\Gamma \vdash \tmOpts : A$ for an operation $o$ in $\Omega$ at some point.
-Hence we choose to include a substitution $\rho$ from the local context $\Xi$ to $\emptyset$ as part of its typing derivation explicitly:
+Here we choose to include a substitution $\rho$ from the local context $\Xi$ to $\emptyset$ as part of its typing derivation explicitly:
 \begin{definition}\label{def:typing-derivations}
@@
   \]
-  and is denoted by $\biarity$.
+  For clarity, an arity is denoted by $\biarity$.
 \end{definition}

### 2023-07-10 tex/sec5-synthesis.tex
[ tex ] Minor changes

@@
 \begin{definition}\label{def:mode-correctness}
-  An operation $\biop$ is \emph{mode-correct} if either
+  An operation $\biop$ is \emph{mode-correct} if 
   \begin{enumerate}
-    \item $d$ is $\chk$, its argument list is mode-correct w.r.t.\ $\fv(A_0)$, and the union $\fv(A_0) \cup \synvar(\biargvec)$ contains every inhabitant of $\Xi$; or
-    \item $d$ is $\syn$, its argument list is mode-correct w.r.t.\ $\emptyset$, and the finite subset $\synvar(\biargvec)$ contains every inhabitant of $\Xi$ and, particularly, $\fv(A_0)$.
+    \item either $d$ is $\chk$, its argument list is mode-correct w.r.t.\ $\fv(A_0)$, and the union $\fv(A_0) \cup \synvar(\biargvec)$ contains every inhabitant of $\Xi$;
+    \item or $d$ is $\syn$, its argument list is mode-correct w.r.t.\ $\emptyset$, and the finite subset $\synvar(\biargvec)$ contains every inhabitant of $\Xi$ and, particularly, $\fv(A_0)$.
   \end{enumerate}
@@
 It is also easy to check the bidirectional type system $(\Sigma_{\bto}, \Omega^{\Leftrightarrow}_{\Lambda}$) in \Cref{ex:signature-simply-typed-lambda} for simply typed $\lambda$-calculus is mode-correct according to our definition.
-\LT{perhaps we shall prove it decidable}
 
@@
 \subsection{Decidability of Bidirectional Type Synthesis and Checking}\label{subsec:bidirectional-synthesis-checking}
-We are here for the main technical contribution of this paper---the decidability of bidirectional type synthesis and checking for any mode correct bidirectional type system:
+
+We are here for the main technical contribution of this paper.
+
 \begin{theorem}[Decidability of Bidirectional Type Synthesis and Checking] \label{thm:bidirectional-type-synthesis-checking}
@@
 \begin{definition}
-  Let $\rho$ be a partial substitution.
-  Then, by an \emph{extension}\/ $\sigma \geq \rho$ we mean that the domain of $\sigma$ contains the domain $\rho$ and\/ $\sigma(x) = \rho(x)$ for every\/ $x$ in the domain of $\rho$; by a \emph{minimal extension}\/ $\bar{\rho}$ of $\rho$ satisfying $P$ we mean an extension $\bar{\rho} \geq \rho$ with $P(\bar{\rho})$ such that $\sigma \geq \bar{\rho}$ whenever $\sigma \geq \rho$ and $P(\sigma)$.
+  By an \emph{extension}\/ $\sigma \geq \rho$ of a partial substitution $\rho$ we mean that the domain of $\sigma$ contains the domain $\rho$ and\/ $\sigma(x) = \rho(x)$ for every\/ $x$ in the domain of $\rho$.
+  By a \emph{minimal extension}\/ $\bar{\rho}$ of $\rho$ satisfying $P$ we mean an extension $\bar{\rho} \geq \rho$ with $P(\bar{\rho})$ such that $\sigma \geq \bar{\rho}$ whenever $\sigma \geq \rho$ and $P(\sigma)$.
 \end{definition}
@@
     \item \ChkRule{Sub}: If $\erase{\Gamma} |-t^{\chk}$ by the rule $\ChkRule{Sub}$, then $\Gamma |- t^{\syn}$ by inversion.
-      We show that it is decidable for any $A$ whether $\Gamma |- t \chk A$.
       By induction hypothesis, it is decidable whether $\Gamma |- t \syn B$ for some $B$:
@@
       \begin{claim}\label{lem:args-induction}
-        For an argument list $\as = \biargvec$ and any partial substitution $\rho$ from $\Xi$ to $\emptyset$
+        For an argument list $\as = \biargs$ and any partial substitution $\rho$ from $\Xi$ to $\emptyset$
         \begin{enumerate}
@@
               \Gamma, \vars{x}_\isTerm{i} : \simsub{\Delta_i}{\ext{\rho}} |- \isTerm{t_i} \colon \simsub{A_i}{\ext{\rho}}^{\dir{d_i}}
-              \quad\text{for all $i$};
+              \quad\text{for $i = 1, \ldots, n$};
             \end{equation}
@@
       \end{claim}
-      Then, case analysis on $\dir{d}$:
+      Then, we proceed with a case analysis on $\dir{d}$:
       \begin{itemize}
@@
           \begin{enumerate}
-            \item If there is no $\sigma \geq \rho$ such that \eqref{eq:claim} holds but for some $A$ we have $\Gamma |- \tmOpts \syn A$, then by inversion we have a substitution $\rho\colon \Sub{\Xi}{\emptyset}$ such that ${\Gamma, \vars{x}_\isTerm{i} : \simsub{\Delta_i}{\rho} |- \isTerm{t_i} \colon \simsub{A_i}{\rho}^{\dir{d_i}}}$ for every $i$.
-              Obviously, $\rho$ is an extension of $\rho_0$ such that ${\Gamma, \vars{x}_\isTerm{i} : \simsub{\Delta_i}{\rho} |- \isTerm{t_i} \colon \simsub{A_i}{\rho}^{\dir{d_i}}}$ for every $i$ and it contradicts the assumption that no such an extension exists.
+            \item If there is no $\sigma \geq \rho$ such that \eqref{eq:claim} holds but $\Gamma |- \tmOpts \syn A$ for some $A$, then by inversion we have a substitution $\rho\colon \Sub{\Xi}{\emptyset}$ such that
+              \[
+                \Gamma, \vars{x}_\isTerm{i} : \simsub{\Delta_i}{\rho} |- \isTerm{t_i} \colon \simsub{A_i}{\rho}^{\dir{d_i}}
+              \]
+              for every $i$.
+              Obviously, $\rho \geq \rho_0$ and ${\Gamma, \vars{x}_\isTerm{i} : \simsub{\Delta_i}{\rho} |- \isTerm{t_i} \colon \simsub{A_i}{\rho}^{\dir{d_i}}}$ for every $i$ and it contradicts the assumption that no such an extension exists.
 
-            \item If there exists a minimal $\ext{\rho} \geq \rho_0$ defined on $\synvar(\biargs)$ such that \eqref{eq:claim} holds, then by mode-correctness we have $\ext{\rho}$ is total and
+            \item If there exists a minimal $\ext{\rho} \geq \rho_0$ defined on $\synvar(\biargs)$ such that~\eqref{eq:claim} holds, then by mode-correctness $\ext{\rho}$ is total and thus
               \[
@@
       \begin{claimproof}
-        For a partial substitution $\rho$, we prove it by induction on the list $\biargvec$:
+        We prove it by induction on the list $\biargs$:
         \begin{enumerate}
@@
                             However, by \Cref{thm:unique-syn}, the synthesised type $\simsub{A_{m+1}}{\gamma}$ must be unique, so $\gamma$ is an extension of $\ext{\rho}$ such that $\simsub{A_{m+1}}{\gamma} = A$, leading to a contradiction.
-                          \item If there is a minimal $\ext{\ext{\rho}}$ such that $\simsub{A_{m+1}}{\ext{\ext{\rho}}} = A$, then it is now hard to show that $\ext{\ext{\rho}}$ is the minimal extension of $\rho$ such that \eqref{eq:claim} holds for all $1 \leq i \leq m + 1$.
+                          \item If there is a minimal $\ext{\ext{\rho}} \geq \ext{\rho}$ such that $\simsub{A_{m+1}}{\ext{\ext{\rho}}} = A$, then it is not hard to show that $\ext{\ext{\rho}}$ is the minimal extension of $\rho$ such that \eqref{eq:claim} holds for all $1 \leq i \leq m + 1$.
                         \end{itemize}
@@
       \end{claimproof}
-      We now complete the decidability proof.
+      We now have completed the decidability proof.
   \end{itemize}
 \end{proof}
-While the proof presented above is intricate, its formal counterpart essentially functions as two programs for type checking and synthesis.
+
+The formal counterpart of the above proof in \Agda essentially functions as two programs for type checking and synthesis.
 These programs either compute the typing derivation or provide a proof of contradiction.
 Each case analysis simply branches depending on the outcomes of bidirectional type synthesis and checking for each subterm, as well as the unification process.
-If a contradiction proof is not of interest for implementation, the programs can be simplified by disregarding the cases that yield such contradiction proofs.
+If a contradiction proof is not of interest for implementation, these programs can be simplified by disregarding the cases that yield such contradiction proofs.
 Alternatively, we could consider generalising typing derivations like our generalised mode derivations (\Cref{fig:generalised-mode-derivations}).

### 2023-07-10 tex/BiSig.tex
[ tex ] Fix the claimproof environment
[ tex ] Mode decoractor?

@@
 \theoremstyle{acmdefinition}
-\newtheorem*{claimproof}{Proof of Claim}
+%\newtheorem*{claimproof}{Proof of Claim}
+\newenvironment{claimproof}[1][\proofname]{\par
+  \renewcommand{\qedsymbol}{\footnotesize$\blacksquare$}%
+    \begin{proof}[#1]%
+}{%
+  \end{proof}%
+}
 \newtheorem{remark}[theorem]{Remark}
@@
   Moreover, we introduce `mode preprocessing' to deal with missing type annotations, an aspect often overlooked in the literature.
+  \LT{mode decorator?}
   The main theorem is logically a trichotomy on untyped abstract syntax trees and computationally a type synthesiser that may throw an exception due to missing type annotations.

### 2023-07-10 tex/sec1-intro.tex
[ tex ] Explain decidability a bit further in S1

@@
 For example, a proof of \emph{decidable equality}, i.e.
-$\forall x, y.~(x = y) + (x \neq y)$, is a way to compare arbitrary $x$~and~$y$ and decide whether they are equal, which may or may not be possible depending on the domain of $x$~and~$y$.
+$\forall x, y.~(x = y) + (x \neq y)$, decides whether they are equal and gives an identity proof explicitly if they are, which may or may not be possible depending on the domain of $x$~and~$y$.
 Given that all proofs as programs terminate, logical decidability implies \emph{algorithmic decidability}.
@@
 \end{quote}
+It yields either a typing derivation for the given raw term~$t$ or a contradiction proof that such a derivation is impossible where the former case is algorithmic soundness and the latter algorithmic completeness in contrapositive form.
+Then, both algorithmic soundness and completeness in original form are implied.
 %echoing \citeauthor{Wadler2022}'s formulation as noted in \cref{re:type-synthesis-as-decidability-proof}.
-Then, both algorithmic soundness and completeness are implied.
-%the proof either yields a typing derivation for the given raw term~$t$, or a proof that such a derivation is impossible; the former case is algorithmic soundness, and the latter is algorithmic completeness in contrapositive form.
 %\todo{traditional completeness is still derivable from this `LEM'}

### 2023-07-10 tex/sec3-defs.tex
[ tex ] Emphasie the domain of discourse of \Omega

@@
 The inclusion of a variable set $\Xi$ for each operation, called its \emph{local context}, plays a crucial role in type synthesis.
-For a typing rule like $\ChkRule{Abs}$, we will need to substitute concrete types (i.e.\ types without any type variables) for variables $A, B$, but firstly, we must identify for which type variables to substitute.
+For a typing rule like $\ChkRule{Abs}$, we will need to substitute \emph{concrete types}, i.e.\ types without any type variables, for variables $A, B$.
+But, we must first identify for which type variables to substitute.
 As such, this information forms part of the arity of an operation, and typing derivations, defined subsequently, will include an instantiation to assign concrete types to variables.
@@
   \small
-  \judgbox{V |-_{\Sigma, \Omega} \isTerm{t}}{Given a list~$V$ of variables, $\isTerm{t}$ is a raw term for the signature $(\Sigma, \Omega)$ with free variables in~$V$}
+  \judgbox{V |-_{\Sigma, \Omega} \isTerm{t}}{$\isTerm{t}$ is a raw term for a simply typed language $(\Sigma, \Omega)$ with free variables in~$V$}
   \begin{mathpar}
@@
     \small
-    \judgbox{\Gamma |-_{\Sigma, \Omega} \isTerm{t} : A}{A raw term $\isTerm{t}$ has a type $A$ without type variables under $\Gamma$ for the signature $(\Sigma, \Omega)$}
+    \judgbox{\Gamma |-_{\Sigma, \Omega} \isTerm{t} : A}{A raw term $\isTerm{t}$ has a concrete type $A$ under $\Gamma$ for a simply typed language $(\Sigma, \Omega)$}
     \begin{mathpar}
@@
     \small
-    \judgbox{\Gamma |-_{\Sigma, \Omega} \isTerm{t} :^\dir{d} A}{A raw term $\isTerm{t}$ is of type $A$ without type variables in mode $\dir{d}$ under $\Gamma$}
+    \judgbox{\Gamma |-_{\Sigma, \Omega} \isTerm{t} :^\dir{d} A}{$\isTerm{t}$ has a concrete type $A$ in mode $\dir{d}$ under $\Gamma$ for a bidirectional type system $(\Sigma, \Omega)$} 
     \begin{mathpar}
@@
     \small
-    \judgbox{V |-_{\Sigma, \Omega} \isTerm{t}^\dir{d}}{A raw term $\isTerm{t}$ is in mode $d$ with free variables in $V$}
+    \judgbox{V |-_{\Sigma, \Omega} \isTerm{t}^\dir{d}}{$\isTerm{t}$ is in mode $d$ with free variables in $V$ for a bidirectional type system $(\Sigma, \Omega)$}
     \begin{mathpar}

### 2023-07-10 tex/sec5-synthesis.tex
[ tex ] Fix the claimproof environment
[ tex ] Revise S5 further

@@
   The subset $\fv(\Gamma)$ of variables in a context $\Gamma$ is defined by\/ $\fv(\cdot) = \emptyset$ and\/ $\fv(\Gamma, A) = \fv(\Gamma) \cup \fv(A)$.
-  For an argument list $\biargvec$,\LT{Eww...} define the set of \emph{synthesis type variables} inductively as 
+  For an argument list $\biargs$,\LT{Eww...} define its set of \emph{synthesis type variables} inductively as 
   \begin{align}
     \label{eq:synvar1}\synvar(\cdot)                   & = \emptyset  \\
-    \label{eq:synvar2}\synvar(\biargvec, \chkbiarg[n]) & = \phantom{\fv(A_{n}) \cup {}} \synvar(\biargvec) \\
-    \label{eq:synvar3}\synvar(\biargvec, \synbiarg[n]) & = \fv(A_{n}) \cup           \synvar(\biargvec).
+    \label{eq:synvar2}\synvar(\biargs, \chkbiarg[n+1]) & = \phantom{\fv(A_{n+1}) \cup {}} \synvar(\biargs) \\
+    \label{eq:synvar3}\synvar(\biargs, \synbiarg[n+1]) & = \fv(A_{n+1}) \cup           \synvar(\biargs).
   \end{align}
@@
     \item $\Rule{Op}$: Recall that a derivation of\/ $\Gamma |- \tmOpts \syn A$ contains a substitution $\rho$ from the local context $\Xi$ to concrete types.
-      To prove that any two typing derivations has the same synthesised type, it suffices to show that those substitutions $\rho_1$ and $\rho_2$ of $d_1$ and $d_2$, respectively, agree on variables in $\synvar\left(\biargs\right)$ so that $\simsub{A_0}{\rho_1} = \simsub{A_0}{\rho_2}$.
+      To prove that any two typing derivations has the same synthesised type, it suffices to show that those substitutions $\rho_1$ and $\rho_2$ of $d_1$ and $d_2$, respectively, agree on variables in $\synvar(\biargs)$ so that $\simsub{A_0}{\rho_1} = \simsub{A_0}{\rho_2}$.
       We prove it by induction on the argument list:
@@
         \item If $\dir{d_{i+1}}$ is synthesis, then $\simsub{\Delta_{i+1}}{\rho_1} = \simsub{\Delta_{i+1}}{\rho_2}$ by mode-correctness~\eqref{eq:MC2} and induction hypothesis (of the list).
-          Therefore, under the same context $\Gamma, \simsub{\Delta_{i+1}}{\rho_1} = \Gamma, \simsub{\Delta_{i+1}}{\rho_2}$ the term $t_{i+1}$ must have the same synthesised type $\simsub{A_{i+1}}{\rho_1} = \simsub{A_{i+1}}{\rho_1}$ by induction hypothesis (of the typing derivation), so $\rho_1$ and $\rho_2$ agree on $\fv(A_{i+1})$ in addition to $\synvar\left(\biargs\right)$, as required for the case \eqref{eq:synvar3} in the definition of $\synvar$.
+          Therefore, under the same context $\Gamma, \simsub{\Delta_{i+1}}{\rho_1} = \Gamma, \simsub{\Delta_{i+1}}{\rho_2}$ the term $t_{i+1}$ must have the same synthesised type $\simsub{A_{i+1}}{\rho_1} = \simsub{A_{i+1}}{\rho_1}$ by induction hypothesis (of the typing derivation), so $\rho_1$ and $\rho_2$ agree on $\fv(A_{i+1})$ in addition to $\synvar(\biargs)$, as required for the case \eqref{eq:synvar3} in the definition of $\synvar$.
       \end{enumerate}
@@
 \begin{definition}
-  By an \emph{extension}\/ $\sigma \geq \rho$ of a partial substitution $\rho$ we mean that the domain of $\sigma$ contains the domain $\rho$ and\/ $\sigma(x) = \rho(x)$ for every\/ $x$ in the domain of $\rho$.
+By an \emph{extension}\/ $\sigma \geq \rho$ of a partial substitution $\rho$ we mean that the domain $\dom(\sigma)$ of $\sigma$ contains the domain of $\rho$ and $\sigma(x) = \rho(x)$ for every\/ $x$ in $\dom(\rho)$.
   By a \emph{minimal extension}\/ $\bar{\rho}$ of $\rho$ satisfying $P$ we mean an extension $\bar{\rho} \geq \rho$ with $P(\bar{\rho})$ such that $\sigma \geq \bar{\rho}$ whenever $\sigma \geq \rho$ and $P(\sigma)$.
@@
   The last case \Rule{Op} is new and has been discussed above.
-  (For brevity we omit the subscript $(\Sigma, \Omega)$.)
+  For brevity we omit the subscript $(\Sigma, \Omega)$.
   \begin{itemize}
@@
     \item \Rule{Op}:
-      Consider the mode derivation of $\erase{\Gamma} |- \tmOpts^{\dir{d}}$.
-      First we claim:
+      For a mode derivation of $\erase{\Gamma} |- \tmOpts^{\dir{d}}$, we first claim:
       \begin{claim}\label{lem:args-induction}
-        For an argument list $\as = \biargs$ and any partial substitution $\rho$ from $\Xi$ to $\emptyset$
+        For an argument list $\biargs$ and any partial substitution $\rho$ from $\Xi$ to $\emptyset$
         \begin{enumerate}
@@
             \begin{equation} \label{eq:claim}
-              \text{$\ext{\rho}$ is defined on $\synvar(\as)$ and\/} \quad
+              \dom(\ext{\rho}) \supseteq \synvar(\biargs)
+              \quad\text{and}\quad
+%              \text{the domain of $\ext{\rho}$ contains $\synvar(\biargs)$ and\/} \quad
               \Gamma, \vars{x}_\isTerm{i} : \simsub{\Delta_i}{\ext{\rho}} |- \isTerm{t_i} \colon \simsub{A_i}{\ext{\rho}}^{\dir{d_i}}
@@
       \end{claim}
-      Then, we proceed with a case analysis on $\dir{d}$:
+
+      Then, we proceed with a case analysis on $\dir{d}$ in the mode derivation:
       \begin{itemize}
@@
         \end{enumerate}
-        \hfill{\footnotesize$\blacksquare$}
+        Therefore, we have proved our claim for any argument list.
       \end{claimproof}
+  \end{itemize}
   We now have completed the decidability proof.
-  \end{itemize}
 \end{proof}
 
-The formal counterpart of the above proof in \Agda essentially functions as two programs for type checking and synthesis.
+The formal counterpart of the above proof in \Agda functions as two top-level programs for type checking and synthesis.
 These programs either compute the typing derivation or provide a proof of contradiction.

### 2023-07-10 tex/sec5-synthesis.tex
[ tex ] A bit more on S5

@@
         \end{enumerate}
-        Therefore, we have proved our claim for any argument list.
+        Therefore, we have proved our claim for any argument list by induction.
       \end{claimproof}
   \end{itemize}
-  We now have completed the decidability proof.
+  We now have completed the decidability proof by induction on the mode derivation $\erase{\Gamma} |-_{\Sigma, \Omega} t^{\dir{d}}$.
 \end{proof}

### 2023-07-10 tex/BiSig.tex
[ tex ] Revise S5 further
[ tex ] Revised the abstract; changed ‘mode preprocessing’ to ‘mode decoration’; addressed some of Favonia’s ‘complaints’

@@
 \input{mathlig}
-\usepackage{tikz}
-\usetikzlibrary{matrix,arrows}
+%\usepackage{tikz}
+%\usetikzlibrary{matrix,arrows}
 \usepackage{mathpartir}
+
 \usepackage{xspace}
@@
 %\newtheorem*{claimproof}{Proof of Claim}
-\newenvironment{claimproof}[1][\proofname]{\par
+\newenvironment{claimproof}{\par
   \renewcommand{\qedsymbol}{\footnotesize$\blacksquare$}%
-    \begin{proof}[#1]%
+  \begin{proof}[{Proof of Claim}]%
 }{%
@@
 \begin{abstract}
-  \LT{Still not coherent enough}
-%  Type checking and inference serve as the transition from parsed abstract syntax trees to well-typed programs.
-  While there is much progress in designing bidirectional type systems and associated type synthesis algorithms on a case-by-case basis, the lack of a general theory makes it difficult to study the common principles and develop widely applicable techniques and practical tools such as a `type-synthesiser generator'.
-  In response to this difficulty, we propose a simple yet general and constructive theory of bidirectional type synthesis.
-%  which refers to the mutual use of type checking and synthesis based on bidirectional typing rules to determine the type of a program with type annotations.
-  First, it is simple yet general as it applies to any syntax-directed bidirectional simple type system that can be specified by a signature.
-%  which is analogous to a grammar in parsing.
-  Second, it is constructive because it is not only formalised in a proof assistant based on Martin-L\"of type theory, but also compactly formulated in a way that unifies computation and proof---for example, a proof of the decidability of the type synthesis problem is computationally a type synthesiser.
-  Moreover, we introduce `mode preprocessing' to deal with missing type annotations, an aspect often overlooked in the literature.
-  \LT{mode decorator?}
-  The main theorem is logically a trichotomy on untyped abstract syntax trees and computationally a type synthesiser that may throw an exception due to missing type annotations.
-  By being parametrised by a signature, our theory works overall as a verified type-synthesiser generator.
+There has been much progress in designing bidirectional type systems and associated type synthesis algorithms, but only on a case-by-case basis.
+This situation is in stark contrast to parsing, for which there have been general theories and widely applicable tools such as parser generators.
+To remedy the situation, this paper develops a first theory of bidirectional type synthesis.
+The whole theory works as a verified type-synthesiser generator for syntax-directed simple type systems:
+From a signature that specifies a bidirectionally typed language, the theory produces a type synthesiser that decides whether an input abstract syntax tree has a typing derivation or reports that the input does not have enough type annotations.
+Within the theory, we formally define soundness, completeness, and mode-correctness, which are sufficient conditions for deriving a correct type synthesiser.
+We also propose a preprocessing step called `mode decoration', which helps the user to deal with missing type annotations and streamlines the theory.
+The entire theory is formulated constructively, and has been formalised in the \Agda proof assistant.
+
+%\LT{Still not coherent enough}
+%Type checking and inference serve as the transition from parsed abstract syntax trees to well-typed programs.
+%In response to this difficulty, we propose a simple yet general and constructive theory of bidirectional type synthesis.
+%which refers to the mutual use of type checking and synthesis based on bidirectional typing rules to determine the type of a program with type annotations.
+%First, it is simple yet general as it applies to any syntax-directed bidirectional simple type system that can be specified by a signature.
+%which is analogous to a grammar in parsing.
+%Second, it is constructive because it is not only formalised in a proof assistant based on Martin-L\"of type theory, but also compactly formulated in a way that unifies computation and proof---for example, a proof of the decidability of the type synthesis problem is computationally a type synthesiser.
+%Moreover, we introduce `mode decoration' to deal with missing type annotations, an aspect often overlooked in the literature.
+%The main theorem is logically a trichotomy on untyped abstract syntax trees and computationally a type synthesiser that may throw an exception due to missing type annotations.
+%By being parametrised by a signature, our theory works overall as a verified type-synthesiser generator.
 \end{abstract}

### 2023-07-10 tex/sec1-intro.tex
[ tex ] Revised the abstract; changed ‘mode preprocessing’ to ‘mode decoration’; addressed some of Favonia’s ‘complaints’

@@
 The basic idea of bidirectional type synthesis is that while the problem of type inference is not decidable in general, for certain kinds of terms it is still possible to infer their types (for example, the type of a variable can be looked up in the context); for other kinds of terms, we can switch to the simpler problem of type checking, where the expected type of a raw term is also given so that there is more information to work with.
-More formally, judgements in a bidirectional type system are extended with two \emph{modes}:
+More formally, every judgement in a bidirectional type system is assigned one of two \emph{modes}:
 \begin{enumerate*}
@@
 
-Although the basic idea is well known, bidirectional typing has only been developed on a case-by-case basis without a theory; this situation is in stark contrast to parsing, which has a comprehensive theory together with widely applicable techniques and practical tools.%
+Despite sharing the same basic idea, bidirectional typing has only been developed on a case-by-case basis without a theory; this situation is in stark contrast to parsing, which has a comprehensive theory together with widely applicable techniques and practical tools, notably parser generators.%
 \footnote{The same could even be said for type checking in general, not just for bidirectional type systems.
@@
 
-\subsection{Related Work}\label{sec:related-work}
+\subsection{Background and Related Work}\label{sec:related-work}
 
@@
 Within the theory, we formulate essential conditions including \emph{soundness}, \emph{completeness}, and \emph{mode-correctness} that are sufficient to derive a bidirectional type synthesiser and ensure that it implements the general specification, formalising what were only informally outlined by~\citet{Dunfield2021}.
-We also pay attention to the problem of whether a raw term has enough type annotations, often overlooked in the literature, by introducing \emph{mode preprocessing}, which assigns a mode to a raw term with enough annotations or otherwise pinpoints missing annotations.
-Practically, by running a mode preprocessor, the user can first make sure that a raw term has enough annotations, and then deal with typing issues without worrying about missing annotations anymore.
+We also pay attention to the problem of whether a raw term has enough type annotations, often overlooked in the literature, by introducing \emph{mode decoration}, which assigns a mode to a raw term with enough annotations or otherwise pinpoints missing annotations.
+Practically, by running a mode decorator, the user can first make sure that a raw term has enough annotations, and then deal with typing issues without worrying about missing annotations anymore.
 Theoretically, separating the annotation and typing aspects streamlines our theory, in particular allowing us to formulate completeness naturally and differently from \emph{annotatability}, which was proposed and conflated with completeness by \citet{Dunfield2021}.
-By combining bidirectional type synthesis, soundness, completeness, and mode preprocessing, we achieve our main result~(\cref{cor:trichotomy}): a \emph{trichotomy} on raw terms, which is computationally a type synthesiser that checks if a given raw term is sufficiently annotated and if a sufficiently annotated raw term has an ordinary typing derivation, suggesting that type synthesis should not be viewed as a bisection between typable and untypable raw terms but a trichotomy in general.
+By combining bidirectional type synthesis, soundness, completeness, and mode decoration, we achieve our main result~(\cref{cor:trichotomy}): a \emph{trichotomy} on raw terms, which is computationally a type synthesiser that checks if a given raw term is sufficiently annotated and if a sufficiently annotated raw term has an ordinary typing derivation, suggesting that type synthesis should not be viewed as a bisection between typable and untypable raw terms but a trichotomy in general.
 
@@
   \item \emph{simple yet general}, working for any syntax-directed bidirectional simple type system that can be specified by a bidirectional binding signature;
-  \item \emph{constructive} based on Martin-L\"of type theory, and compactly formulated in a way that unifies computation and proof, preferring logical decidability over algorithmic soundness, completeness, and decidability; and mode preprocessing over annotatability.
+  \item \emph{constructive} based on Martin-L\"of type theory, and compactly formulated in a way that unifies computation and proof, preferring logical decidability over algorithmic soundness, completeness, and decidability; and mode decoration over annotatability.
 \end{enumerate}
@@
 We first present a concrete overview of our theory using simply typed $\lambda$-calculus in \cref{sec:key-ideas}, prior to developing a general framework for specifying bidirectional type systems in \cref{sec:defs}.
-Following this, we discuss mode preprocessing and related properties in \cref{sec:pre-synthesis}.
+Following this, we discuss mode decoration and related properties in \cref{sec:pre-synthesis}.
 The most heavyweight technical contribution lies in \cref{sec:type-synthesis}, where we introduce mode-correctness and bidirectional type synthesis.

### 2023-07-10 tex/sec2-outline.tex
[ tex ] Revised the abstract; changed ‘mode preprocessing’ to ‘mode decoration’; addressed some of Favonia’s ‘complaints’

@@
 
-\begin{definition}
-\label{def:STLC-mode-preprocessor}
-A \emph{mode preprocessor} decides for any raw term~$V \vdash \isTerm{t}$ whether $V \vdash \isTerm{t}^{\syn}$.
+\begin{definition} \label{def:STLC-mode-decorator}
+A \emph{mode decorator} decides for any raw term~$V \vdash \isTerm{t}$ whether $V \vdash \isTerm{t}^{\syn}$.
 \end{definition}
 
-One (less important) benefit of mode preprocessing is that it helps to simplify the synthesiser, whose computation can be partly directed by a mode derivation.
+One (less important) benefit of mode decoration is that it helps to simplify the synthesiser, whose computation can be partly directed by a mode derivation.
 More importantly, whether there is a mode derivation for a term is actually very useful information to the user, because it corresponds to whether the term has enough type annotations:
@@
 \begin{theorem}\label{thm:implementation}
-A type synthesiser that uses `not in synthesising mode' as its excuse can be constructed from a mode preprocessor and a bidirectional type synthesiser.
+A type synthesiser that uses `not in synthesising mode' as its excuse can be constructed from a mode decorator and a bidirectional type synthesiser.
 \end{theorem}
@@
 The construction is straightforward:
-Run the mode preprocessor on the input term~$\erase\Gamma \vdash t$.
+Run the mode decorator on the input term~$\erase\Gamma \vdash t$.
 If there is no synthesising mode derivation, report that $t$~is not in synthesising mode (the third outcome).
@@
 
-We will construct a mode preprocessor~(\cref{sec:mode-preprocessing}) and a bidirectional type synthesiser~(\cref{sec:type-synthesis}) and prove both lemmas~(\cref{sec:soundness-and-completeness}) for all syntax-directed bidirectional simple type systems.
+We will construct a mode decorator~(\cref{sec:mode-decoration}) and a bidirectional type synthesiser~(\cref{sec:type-synthesis}) and prove both lemmas~(\cref{sec:soundness-and-completeness}) for all syntax-directed bidirectional simple type systems.
 To quantify over all such systems, we need their generic definitions, which we formulate next.
@@
 %  \path[->, line width=0.5pt]
-%      (m-1-1) edge node [above] {mode preprocessing} (m-1-2)
+%      (m-1-1) edge node [above] {mode decoration} (m-1-2)
 %      (m-1-2) edge node [right,align=center] {bidirectional\\type synthesis} (m-2-2)

### 2023-07-10 tex/sec3-defs.tex
[ tex ] Revised the abstract; changed ‘mode preprocessing’ to ‘mode decoration’; addressed some of Favonia’s ‘complaints’

@@
       \begin{mathpar}
-        \inferrule{\Xi \ni X_i}{\Xi |-_\Sigma X_i} \and
+        \inferrule{X_i \in \Xi}{\Xi |-_\Sigma X_i} \and
         \inferrule{\Xi |-_{\Sigma} A_1 \\ \cdots \\ \Xi |-_{\Sigma} A_n}{\Xi |-_{\Sigma} \tyOp_i(A_1, \ldots, A_n)}\;\text{where $n = \arity(i)$}
@@
   as shown in \Cref{fig:simple-type}.
-  A \emph{context} $\Gamma \colon \Cxt_{\Sigma}(\Xi)$ over $\Xi$ is a list $\Gamma$ of types over $\Xi$. 
+  A \emph{context} $\Gamma \colon \Cxt_{\Sigma}(\Xi)$ over $\Xi$ is a list $\Gamma$ of types over $\Xi$.%
+\todo{Name-type pairs?}
 \end{definition}
@@
   In a binding arity $(((\Delta_1, A_1), \ldots, (\Delta_n, A_n)), A)$, every $\Delta_i$ and $A_i$ refers to the \emph{extension context} and the sort of the $i$-th argument, respectively, and $A$ the target sort.
-  For brevity, a binding arity is denoted by $\bargs \to {A_0}$ and omit $[\Delta_i]$ if $\Delta_i$ is empty.
+  For brevity, a binding arity is denoted by $\bargs \to A$ where $[\Delta_i]$ is omitted if $\Delta_i$ is empty.
 \end{definition}
@@
 \]
-This means that the $\Rule{Abs}$ rule for derivations of\/ $\Gamma |- \lam{x}t : A \bto B$, contains:
+This means that the $\Rule{Abs}$ rule for derivations of\/ $\Gamma |- \lam{x}t : A \bto B$ contains:
 \begin{enumerate*}
@@
 \end{enumerate*}
-In a similar vein, the arity of the $\Rule{App}$ rule is denoted as ${(A \bto B), A} \to {B}$, saying that any derivation of\/ $\Gamma \vdash t\;u : B$ accepts derivations $\Gamma \vdash t : A \bto B$ and $\Gamma \vdash u : A$ as its arguments.
-These arguments have types $A \bto B$ and $A$ respectively and their extension contexts are empty.
+In a similar vein, the arity of the $\Rule{App}$ rule is denoted as
+\[
+  {(A \bto B), A} \to {B}
+\]
+saying that any derivation of\/ $\Gamma \vdash t\;u : B$ accepts derivations $\Gamma \vdash t : A \bto B$ and $\Gamma \vdash u : A$ as its arguments.
+These arguments have types $A \bto B$ and $A$ respectively and empty extension contexts.
 

### 2023-07-10 tex/sec4-pre-synthesis.tex
[ tex ] Revised the abstract; changed ‘mode preprocessing’ to ‘mode decoration’; addressed some of Favonia’s ‘complaints’

@@
 
-\section{Mode Preprocessing and Related Properties}\label{sec:pre-synthesis}
+\section{Mode Decoration and Related Properties}\label{sec:pre-synthesis}
 
-Our first important construction is mode preprocessing in \cref{sec:mode-preprocessing}, which is in fact generalised to pinpoint any missing type annotations in a given raw term.
+Our first important construction is mode decoration in \cref{sec:mode-decoration}, which is in fact generalised to pinpoint any missing type annotations in a given raw term.
 We also discuss some related properties:
 By bringing mode derivations into the picture, we are able to give a natural formulation of soundness and completeness of a bidirectional type system with respect to its erasure to an ordinary type system in \cref{sec:soundness-and-completeness}.
-We also formulate \varcitet{Dunfield2021}{'s} annotatability and discuss its relationship with our completeness and generalised mode preprocessing in \cref{sec:annotatability}.
+We also formulate \varcitet{Dunfield2021}{'s} annotatability and discuss its relationship with our completeness and generalised mode decoration in \cref{sec:annotatability}.
 
@@
 
-\subsection{Generalised Mode Preprocessing}
-\label{sec:mode-preprocessing}
+\subsection{Generalised Mode Decoration}
+\label{sec:mode-decoration}
 
-The goal of this sub-section is to construct a mode preprocessor, which decides for any raw term $V |-_{\Sigma,\erase\Omega} t$ and mode~$\dir d$ whether $V |-_{\Sigma,\Omega} t^\dir{d}$ or not.
+The goal of this sub-section is to construct a mode decorator, which decides for any raw term $V |-_{\Sigma,\erase\Omega} t$ and mode~$\dir d$ whether $V |-_{\Sigma,\Omega} t^\dir{d}$ or not.
 In fact we will do better:
-If a mode preprocessor returns a proof that no mode derivation exists, that proof (of a negation) does not provide useful information for the user.
-It will be more helpful if a preprocessor can produce an explanation of why no mode derivation exists, and even how to fix the input term to have a mode derivation.
-We will construct such a \emph{generalised mode preprocessor}~(\cref{thm:generalised-mode-preprocessing}), which can be weakened to an ordinary mode preprocessor~(\cref{thm:mode-preprocessing}) if the additional explanation is not needed.%
-\footnote{For clarity, we use ordinary mode preprocessing elsewhere in this paper.}
+If a mode decorator returns a proof that no mode derivation exists, that proof (of a negation) does not provide useful information for the user.
+It will be more helpful if a decorator can produce an explanation of why no mode derivation exists, and even how to fix the input term to have a mode derivation.
+We will construct such a \emph{generalised mode decorator}~(\cref{thm:generalised-mode-decoration}), which can be weakened to an ordinary mode decorator~(\cref{thm:mode-decoration}) if the additional explanation is not needed.%
+\footnote{For brevity, we use ordinary mode decoration elsewhere in this paper.}
 
@@
 
-Now we are ready to construct a generalised mode preprocessor.
+Now we are ready to construct a generalised mode decorator.
 
-\begin{theorem}[Generalised Mode Preprocessing]\label{thm:generalised-mode-preprocessing}
+\begin{theorem}[Generalised Mode Preprocessing]\label{thm:generalised-mode-decoration}
 For any raw term $V |-_{\Sigma,\erase\Omega} t$ and mode~$\dir{d}$, there is a derivation of\/ $V |-_{\Sigma,\Omega} t^{\dir{d}\,g\,s}$ for some booleans $g$~and~$s$.
@@
 The theorem could be proved directly, but that would mix up two case analyses which respectively inspect the input term and apply mode casts depending on what modes are required.
-Instead, we distill the case analysis that deals with mode casts into the following \cref{thm:adjustment}, whose antecedent~(\ref{eq:strict-derivation}) is then established by induction on the input term in the proof of \cref{thm:generalised-mode-preprocessing}.
+Instead, we distill the case analysis that deals with mode casts into the following \cref{thm:adjustment}, whose antecedent~(\ref{eq:strict-derivation}) is then established by induction on the input term in the proof of \cref{thm:generalised-mode-decoration}.
 
@@
 
-\begin{proof}[Proof of \cref{thm:generalised-mode-preprocessing}]
+\begin{proof}[Proof of \cref{thm:generalised-mode-decoration}]
 By \cref{thm:adjustment}, it suffices to prove statement~(\ref{eq:strict-derivation}) by induction on~$t$.
@@
 
-Finally, having constructed a generalised mode preprocessor, it is easy to derive an ordinary one.
+Finally, having constructed a generalised mode decorator, it is easy to derive an ordinary one.
 
-\begin{corollary}[Mode Preprocessing]\label{thm:mode-preprocessing}
+\begin{corollary}[Mode Preprocessing]\label{thm:mode-decoration}
   It is decidable whether\/ $V |-_{\Sigma,\Omega} \isTerm{t}^\dir{d}$.%for any raw term $V |-_{\Sigma,\erase\Omega} t$ and mode~$\dir{d}$
@@
 \begin{proof}
-By \cref{thm:generalised-mode-preprocessing}, there is a derivation of $V |-_{\Sigma,\erase\Omega} t^{\dir{d}\,g\,s}$ for some booleans $g$~and~$s$, and then simply check~$g$ and apply either \cref{thm:Pre?-true} or \cref{thm:Pre?-false} to obtain $V |-_{\Sigma,\Omega} t^\dir{d}$ or $V |/-_{\Sigma,\Omega} t^\dir{d}$.
+By \cref{thm:generalised-mode-decoration}, there is a derivation of $V |-_{\Sigma,\erase\Omega} t^{\dir{d}\,g\,s}$ for some booleans $g$~and~$s$, and then simply check~$g$ and apply either \cref{thm:Pre?-true} or \cref{thm:Pre?-false} to obtain $V |-_{\Sigma,\Omega} t^\dir{d}$ or $V |/-_{\Sigma,\Omega} t^\dir{d}$.
 \end{proof}
@@
 %Whereas it is possible to prove \cref{thm:toPre?-true} directly by induction, converting a mode derivation to a generalised one, the negated antecedent of \cref{thm:toPre?-false} provides little information for constructing a required derivation, and basically we have to construct such a derivation from scratch for a raw term.
-%But this is exactly what generalised mode preprocessing does, so we can simply reuse \cref{thm:generalised-mode-preprocessing}.
+%But this is exactly what generalised mode decoration does, so we can simply reuse \cref{thm:generalised-mode-decoration}.
 %And \cref{thm:toPre?-true} can also be proved easily in the same way.
@@
 %
-%By \cref{thm:generalised-mode-preprocessing}, there is a derivation of $V |-_{\Sigma,\Omega} t^{\dir{d}\,g\,s}$ for some booleans $g$~and~$s$.
+%By \cref{thm:generalised-mode-decoration}, there is a derivation of $V |-_{\Sigma,\Omega} t^{\dir{d}\,g\,s}$ for some booleans $g$~and~$s$.
 %If $V |-_{\Sigma,\Omega} t^\dir{d}$, then $g$~has to be $\BTrue$ since $g = \BFalse$ leads to a contradiction with \cref{thm:Pre?-false}.
@@
 In a sense, annotatability is a reasonable form of completeness: if a term is typable in $(\Sigma, \erase\Omega)$, it may not be typable in $(\Sigma, \Omega)$ directly due to some missing annotations, but will be if those annotations are added correctly.
-In our theory, \cref{thm:annotatability} can be easily proved from generalised mode preprocessing.
+In our theory, \cref{thm:annotatability} can be easily proved from generalised mode decoration.
 
 \begin{proof}[Proof of \cref{thm:annotatability}]
-By \cref{thm:generalised-mode-preprocessing}, there is a generalised mode derivation of $V |-_{\Sigma,\erase\Omega} t^{\dir{d}\,g\,s}$ for some mode~$\dir{d}$ and booleans $g$~and~$s$.
+By \cref{thm:generalised-mode-decoration}, there is a generalised mode derivation of $V |-_{\Sigma,\erase\Omega} t^{\dir{d}\,g\,s}$ for some mode~$\dir{d}$ and booleans $g$~and~$s$.
 Perform induction on this generalised mode derivation to construct a bidirectional typing derivation in the same mode (and also the term~$t'$ and a proof of $t' \sqsupseteq t$, which are determined by the derivations and omitted here).
@@
 On the other hand, when using a bidirectional type synthesiser to implement type synthesis~(\cref{thm:implementation}), our completeness is much simpler to use than annotatability, which requires the bidirectional type synthesiser to produce more complex evidence when the synthesis fails.
-Annotatability also does not help the user to deal with missing annotations like (generalised) mode preprocessing does: although annotatability seems capable of determining where annotations are missing and even filling them in correctly, its antecedent requires a typing derivation, which is what the user is trying to construct and does not have yet.
+Annotatability also does not help the user to deal with missing annotations like (generalised) mode decoration does: although annotatability seems capable of determining where annotations are missing and even filling them in correctly, its antecedent requires a typing derivation, which is what the user is trying to construct and does not have yet.
 Therefore we believe that the notion of annotatability is too complex for what it achieves, and our theory offers simpler and more useful alternatives.

### 2023-07-10 tex/sec7-future.tex
[ tex ] Revised the abstract; changed ‘mode preprocessing’ to ‘mode decoration’; addressed some of Favonia’s ‘complaints’

@@
 Even for this mildly generalised setting, there is already a lot more work to do:
-Both the mode preprocessor and the bidirectional type synthesiser will have to backtrack, and become significantly more complex.
+Both the mode decorator and the bidirectional type synthesiser will have to backtrack, and become significantly more complex.
 For soundness and completeness, it should still be possible to treat them as the separation and combination of mode and type information, but the completeness direction will pose a problem---for every node of a raw term, a mode derivation chooses a mode assignment while a typing derivation chooses a typing rule, but there may not be a bidirectional typing rule for this particular combination.
-There should be ways to fix this problem while retaining the mode preprocessing phase---for example, one idea is to make the mode preprocessor produce all possible mode derivations, and refine completeness to say that any typing derivation can be combined with one of these mode derivations into a bidirectional typing derivation.
-All these considerations are rather too complex for a first theory, however, and should best be left as future work.
+There should be ways to fix this problem while retaining the mode decoration phase---for example, one idea is to make the mode decorator produce all possible mode derivations, and refine completeness to say that any typing derivation can be combined with one of these mode derivations into a bidirectional typing derivation.
+All these considerations are rather too complex for a first theory, however, and we decide to leave them for future work.
 
@@
 %\item More sensible next step: a `refinement chain'
-%\item Backtracking algorithms are required. for both mode preprocessing and bidirectional type synthesis.
+%\item Backtracking algorithms are required. for both mode decoration and bidirectional type synthesis.
 %\item The idea of separating and combining mode and type information should still work, but requires more care in the completeness direction.

### 2023-07-10 tex/sec5-synthesis.tex
[ tex ] Revised the abstract; changed ‘mode preprocessing’ to ‘mode decoration’; addressed some of Favonia’s ‘complaints’
[ tex ] Clean up code in S6

@@
   The subset $\fv(\Gamma)$ of variables in a context $\Gamma$ is defined by\/ $\fv(\cdot) = \emptyset$ and\/ $\fv(\Gamma, A) = \fv(\Gamma) \cup \fv(A)$.
-  For an argument list $\biargs$,\LT{Eww...} define its set of \emph{synthesis type variables} inductively as 
+  For an argument list $\biargs$, define its set of \emph{synthesis type variables} inductively as 
   \begin{align}
@@
 \begin{proof}
-  Combine \Cref{lem:soundness-completeness,thm:mode-preprocessing} with \Cref{thm:bidirectional-type-synthesis-checking}.
+  Combine \Cref{lem:soundness-completeness,thm:mode-decoration} with \Cref{thm:bidirectional-type-synthesis-checking}.
 

### 2023-07-10 tex/sec6-formalisation.tex
[ tex ] Clean up code in S6

@@
 \DeclareUnicodeCharacter{1D9C}{\ensuremath{^c}}
+\DeclareUnicodeCharacter{02B3}{\ensuremath{^r}}
 \DeclareUnicodeCharacter{02E2}{\ensuremath{^s}}
@@
 \DeclareUnicodeCharacter{2237}{\ensuremath{\mathrel{::}}}
+\DeclareUnicodeCharacter{2982}{\ensuremath{\bbcolon}}
+
 \DeclareUnicodeCharacter{0393}{\ensuremath{\Gamma}}
 \DeclareUnicodeCharacter{0394}{\ensuremath{\Delta}}
+\DeclareUnicodeCharacter{0398}{\ensuremath{\Theta}}
 \DeclareUnicodeCharacter{03A3}{\ensuremath{\Sigma}}
@@
 
-\DeclareUnicodeCharacter{2032}{\ensuremath{\prime}}
+\DeclareUnicodeCharacter{2032}{\ensuremath{{}^\prime}}
 \DeclareUnicodeCharacter{2113}{\ensuremath{\ell}}
@@
 \DeclareUnicodeCharacter{2264}{\ensuremath{\leq}}
+\DeclareUnicodeCharacter{21D0}{\ensuremath{\Leftarrow}}
 \DeclareUnicodeCharacter{21D2}{\ensuremath{\Rightarrow}}
@@
 
-\section{Formalisation} \label{sec:formalisation}
-\begin{enumerate}
-  \item (Bidirectional) binding signature, functor, and terms (extrinsic typing, raw terms, raw terms in some mode) --- 2.5 pp
-  \item Compare the induction principle and the \Agda proof of Soundness --- 2 pp
-  \item type synthesis and checking -- 0.5 p
-  \item Examples including STLC (PCF), application in spine form --- 1p
+\section{Formalisation and Examples} \label{sec:formalisation}
+%\begin{enumerate}
+%  \item (Bidirectional) binding signature, functor, and terms (extrinsic typing, raw terms, raw terms in some mode) --- 2.5 pp
+%  \item Compare the induction principle and the \Agda proof of Soundness --- 2 pp
+%  \item type synthesis and checking -- 0.5 p
+%  \item Examples including STLC (PCF), application in spine form --- 1p
+%
+%\end{enumerate}
+Our theory, initially developed in \Agda, has been later translated into the mathematical vernacular.
+While the translation is reasonably straightforward, understanding the design of the formalisation itself could pose some difficulty.
+Indeed, as discussed in \Cref{sec:language-formalisation}, designing a language formalisation framework is still a challenging task and our formalisation is no exception.
+Unlike prior frameworks that have primarily focused on meta-properties centred around substitution and term traversal for intrinsically-typed terms, our theory of bidirectional type synthesis does not require term substitution but structural induction for extrinsically-typed terms.
+The formal definitions of extrinsic typing in bidirectional type systems, including bidirectional typing derivations and mode derivations, are more complex than their intrinsic counterparts.
+For a specific language such as PCF, it is noted by \citet{Wadler2022} already that `extrinsically-typed terms require about 1.6 times as much code as intrinsically-typed' for their formalisation of type safety.
+As for our generic definitions, the formal type-theoretic definition for extrinsic typing derivations $\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A$ (\Cref{fig:extrinsic-typing}), for example, is a generic families of sets, also known as a \emph{proof-relevant predicate}, indexed by a generic raw \emph{term} $t$ in additional to a type $A$ and a context $\Gamma$ rather than just one family of sets of intrinsically-typed terms indexed by $A$ and $\Gamma$.
 
-\end{enumerate}
+A categorical analysis of extrinsic typing requires not only \varcitet{Fiore1999}{'s} theory of abstract syntax with variable binding which backs \citet{Fiore2022}'s framework but also \varcitet{Hermida1998}{'s} interpretation of inductive proofs as algebras on the category of predicates.
+Thus, a faithful understanding of our formal constructions requires certain proficiency in category theory, which falls outside the scope of this paper.
 
-\begin{figure}
+As such, in this section, we will merely provide an overview of our design and draw an analogy between the construction of simple types in \Cref{subsec:formal-simple-types} and bidirectional typing rules in \Cref{subsec:formal-extrinsic-typing} in plain terms, avoiding category theory deliberately.
+On the other hand, the formal proofs in our theory closely mirror the informal versions presented so far, thus we will only illustrate the soundness part of \Cref{lem:soundness-completeness} as an example in \Cref{subsec:formal-proofs}.
+We conclude the technical content of this paper with some mode-correct simply typed languages in \Cref{subsec:formal-examples}.
+
+\subsection{Defining Simple Types}\label{subsec:formal-simple-types}
+
+% simple types
+\begin{figure}[h]
   \small
-  \begin{minipage}[t]{.3\linewidth}
+  \begin{minipage}[t]{.45\textwidth}
+    \begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{16}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{19}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Keyword{record}\;\Conid{Desc}\;\mathbin{:}\;\Conid{Set₁}\;\Keyword{where}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Keyword{field}{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\Conid{Op}\;{}\<[16]%
+\>[16]{}\mathbin{:}\;{}\<[19]%
+\>[19]{}\Conid{Set}{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\{\kern-.9pt\vrule width .75pt height 7.125pt depth 1.975pt\kern-1.5pt\;\Varid{decEq}\;\kern-1.5pt\vrule width .75pt height 7.125pt depth 1.975pt\kern-.9pt\}\;{}\<[16]%
+\>[16]{}\mathbin{:}\;{}\<[19]%
+\>[19]{}\Conid{DecEq}\;\Conid{Op}{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\Varid{rules}\;{}\<[16]%
+\>[16]{}\mathbin{:}\;{}\<[19]%
+\>[19]{}\Conid{Op}\;\Varid{→}\;\Conid{ℕ}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+  \end{minipage}
+  \begin{minipage}[t]{.45\textwidth}
+    \begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{8}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{24}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{⟦\char95 ⟧}\;\mathbin{:}\;\Conid{Desc}\;\Varid{→}\;\Conid{Set}\;\Varid{→}\;\Conid{Set}{}\<[E]%
+\\
+\>[B]{}\Varid{⟦}\;\Conid{D}\;\Varid{⟧}\;\Conid{X}\;\mathrel{=}\;\Conid{Σ[}\;\Varid{i}\;\Varid{∈}\;\Conid{D}\;\Varid{.Op}\;\mskip1.5mu]\;\Conid{X}\;\hat{}\;(\Conid{D}\;\Varid{.rules}\;\Varid{i}){}\<[E]%
+\\[\blanklineskip]%
+\>[B]{}\Keyword{data}\;\Conid{Tm}\;(\Conid{Θ}\;\mathbin{:}\;\Conid{ℕ})\;\mathbin{:}\;\Conid{Set}\;\Keyword{where}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\text{\textasciigrave}\anonymous \;{}\<[8]%
+\>[8]{}\mathbin{:}\;\Conid{Fin}\;\Conid{Θ}\;{}\<[24]%
+\>[24]{}\Varid{→}\;\Conid{Tm}\;\Conid{Θ}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{op}\;{}\<[8]%
+\>[8]{}\mathbin{:}\;\Varid{⟦}\;\Conid{D}\;\Varid{⟧}\;(\Conid{Tm}\;\Conid{Θ})\;{}\<[24]%
+\>[24]{}\Varid{→}\;\Conid{Tm}\;\Conid{Θ}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+  \end{minipage}
+\end{figure}
+\subsection{Defining Raw Terms and Extrinsic Typing Derivations}\label{subsec:formal-extrinsic-typing}
+
+\begin{figure}[H]
+  \small
+  \begin{minipage}[t]{.3\textwidth}
   \begin{hscode}\SaveRestoreHook
@@
   \end{minipage}
-  \begin{minipage}[t]{.32\linewidth}
+  \begin{minipage}[t]{.33\textwidth}
   \begin{hscode}\SaveRestoreHook
@@
 \column{14}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{22}{@{}>{\hspre}l<{\hspost}@{}}%
 \column{E}{@{}>{\hspre}l<{\hspost}@{}}%
 \>[B]{}\Keyword{record}\;\Conid{ConD}\;\mathbin{:}\;\Conid{Set}\;\Keyword{where}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Keyword{constructor}\;\Varid{ι}{}\<[E]%
 \\
@@
 \>[5]{}\Varid{args}\;{}\<[14]%
-\>[14]{}\mathbin{:}\;\Conid{List}\;(\Conid{ArgsD}\;\Varid{vars}){}\<[E]%
+\>[14]{}\mathbin{:}\;\Conid{List}\;(\Conid{ArgD}\;\Varid{vars}){}\<[E]%
 \\
@@
 \>[5]{}\Varid{type}\;{}\<[14]%
-\>[14]{}\mathbin{:}\;\Conid{TExp}\;{}\<[22]%
-\>[22]{}\Varid{vars}{}\<[E]%
+\>[14]{}\mathbin{:}\;\Conid{TExp}\;\Varid{vars}{}\<[E]%
 \\
@@
   \end{minipage}
-  \begin{minipage}[t]{.3\linewidth}
+  \begin{minipage}[t]{.3\textwidth}
    \begin{hscode}\SaveRestoreHook
@@
   \end{minipage}
-
-  \caption{Definition of bidirectional binding signature}  
 \end{figure}
 
-{\small\begin{hscode}\SaveRestoreHook
+% functors for raw terms
+\begin{figure}[H]
+  \small
+  \begin{minipage}[t]{.57\textwidth}
+%Fam : (ℓ : Level) → Set (lsuc ℓ)
+%Fam ℓ = ℕ → Set ℓ
+%
+%Fam₀ : Set₁
+%Fam₀ = Fam 0ℓ
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{8}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{17}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{19}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{22}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{27}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{⟦\char95 ⟧ᵃ}\;{}\<[8]%
+\>[8]{}\mathbin{:}\;\Conid{TExps}\;\Conid{Ξ}\;{}\<[19]%
+\>[19]{}\Varid{→}\;\Conid{Fam}\;\Varid{→}\;\Conid{Fam}{}\<[E]%
+\\
+\>[B]{}\Varid{⟦}\;\Conid{Δ}\;{}\<[17]%
+\>[17]{}\Varid{⟧ᵃ}\;{}\<[22]%
+\>[22]{}\Conid{X}\;\Varid{n}\;{}\<[27]%
+\>[27]{}\mathrel{=}\;\Conid{X}\;(\Varid{length}\;\Conid{Δ}\;\Varid{ʳ+}\;\Varid{n}){}\<[E]%
+\\[\blanklineskip]%
+\>[B]{}\Varid{⟦\char95 ⟧ᵃˢ}\;{}\<[8]%
+\>[8]{}\mathbin{:}\;\Conid{ArgsD}\;\Conid{Ξ}\;{}\<[19]%
+\>[19]{}\Varid{→}\;\Conid{Fam}\;\Varid{→}\;\Conid{Fam}{}\<[E]%
+\\
+\>[B]{}\Varid{⟦}\;\Varid{[]}\;{}\<[17]%
+\>[17]{}\Varid{⟧ᵃˢ}\;{}\<[22]%
+\>[22]{}\anonymous \;\anonymous \;{}\<[27]%
+\>[27]{}\mathrel{=}\;\Varid{⊤}{}\<[E]%
+\\
+\>[B]{}\Varid{⟦}\;(\Conid{Δ}\;\Varid{⊢}\;\Conid{A})\;\Varid{∷}\;\Conid{Ds}\;{}\<[17]%
+\>[17]{}\Varid{⟧ᵃˢ}\;{}\<[22]%
+\>[22]{}\Conid{X}\;\Varid{n}\;{}\<[27]%
+\>[27]{}\mathrel{=}\;\Varid{⟦}\;\Conid{Δ}\;\Varid{⟧ᵃ}\;\Conid{X}\;\Varid{n}\;\Varid{×}\;\Varid{⟦}\;\Conid{Ds}\;\Varid{⟧ᵃˢ}\;\Conid{X}\;\Varid{n}{}\<[E]%
+\\[\blanklineskip]%
+\>[B]{}\Varid{⟦\char95 ⟧ᶜ}\;{}\<[8]%
+\>[8]{}\mathbin{:}\;\Conid{ConD}\;{}\<[19]%
+\>[19]{}\Varid{→}\;\Conid{Fam}\;\Varid{→}\;\Conid{Fam}{}\<[E]%
+\\
+\>[B]{}\Varid{⟦}\;\Varid{ι}\;\anonymous \;\Conid{D}\;{}\<[17]%
+\>[17]{}\Varid{⟧ᶜ}\;{}\<[22]%
+\>[22]{}\Conid{X}\;{}\<[27]%
+\>[27]{}\mathrel{=}\;\Varid{⟦}\;\Conid{D}\;\Varid{⟧ᵃˢ}\;\Conid{X}{}\<[E]%
+\\[\blanklineskip]%
+\>[B]{}\Varid{⟦\char95 ⟧}\;{}\<[8]%
+\>[8]{}\mathbin{:}\;\Conid{Desc}\;{}\<[19]%
+\>[19]{}\Varid{→}\;\Conid{Fam}\;\Varid{→}\;\Conid{Fam}{}\<[E]%
+\\
+\>[B]{}\Varid{⟦}\;\Conid{D}\;{}\<[17]%
+\>[17]{}\Varid{⟧}\;{}\<[22]%
+\>[22]{}\Conid{X}\;\Varid{n}\;{}\<[27]%
+\>[27]{}\mathrel{=}\;\Conid{Σ[}\;\Varid{i}\;\Varid{∈}\;\Conid{D}\;\Varid{.Op}\;\mskip1.5mu]\;\Varid{⟦}\;\Conid{D}\;\Varid{.rules}\;\Varid{i}\;\Varid{⟧ᶜ}\;\Conid{X}\;\Varid{n}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+  \end{minipage}
+  \begin{minipage}[t]{.4\textwidth}
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{7}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{32}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Keyword{infix}\;\Varid{5}\;\Varid{\char95 ∋\char95 }{}\<[E]%
+\\[\blanklineskip]%
+\>[B]{}\Keyword{data}\;\Conid{Raw}\;\mathbin{:}\;\Conid{ℕ}\;\Varid{→}\;\Conid{Set}\;\Keyword{where}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\text{\textasciigrave}\anonymous \;{}\<[7]%
+\>[7]{}\mathbin{:}\;\Conid{Fin}\;\Varid{n}\;{}\<[32]%
+\>[32]{}\Varid{→}\;\Conid{Raw}\;\Varid{n}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{\char95 ∋\char95 }\;\mathbin{:}\;(\Conid{A}\;\mathbin{:}\;\Conid{Ty})\;(\Varid{t}\;\mathbin{:}\;\Conid{Raw}\;\Varid{n})\;{}\<[32]%
+\>[32]{}\Varid{→}\;\Conid{Raw}\;\Varid{n}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{op}\;{}\<[7]%
+\>[7]{}\mathbin{:}\;\Varid{⟦}\;\Conid{D}\;\Varid{⟧}\;\Conid{Raw}\;\Varid{n}\;{}\<[32]%
+\>[32]{}\Varid{→}\;\Conid{Raw}\;\Varid{n}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+  \end{minipage}
+\end{figure}
+
+\begin{figure}[H]
+  \small
+%Fam : (ℓ′ : Level) → (X : R.Fam ℓ) → Set (ℓ ⊔ lsuc ℓ′)
+%Fam ℓ′ X = (Γ : Cxt 0) → X (length Γ) → Ty → Set ℓ′
+%
+%Fam₀ : (X : R.Fam ℓ) → Set (ℓ ⊔ lsuc 0ℓ)
+%Fam₀ X = Fam 0ℓ X
+%
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{7}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{17}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{39}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{⟦\char95 ⟧ᵃ}\;{}\<[7]%
+\>[7]{}\mathbin{:}\;(\Conid{Δ}\;\mathbin{:}\;\Conid{TExps}\;\Conid{Ξ})\;(\Conid{X}\;\mathbin{:}\;\Conid{R.Fam})\;(\Conid{Y}\;\mathbin{:}\;(\Conid{Γ}\;\mathbin{:}\;\Conid{Cxt}\;\Varid{0}){}\<[E]%
+\\
+\>[7]{}\Varid{→}\;\Conid{X}\;(\Varid{length}\;\Conid{Γ})\;\Varid{→}\;\Conid{Set})\;{}\<[E]%
+\\
+\>[7]{}\Varid{→}\;\Conid{TSub}\;\Conid{Ξ}\;\Varid{0}\;\Varid{→}\;(\Conid{Γ}\;\mathbin{:}\;\Conid{Cxt}\;\Varid{0})\;\Varid{→}\;\Conid{R.⟦}\;\Conid{Δ}\;\Varid{⟧ᵃ}\;\Conid{X}\;(\Varid{length}\;\Conid{Γ})\;{}\<[E]%
+\\
+\>[7]{}\Varid{→}\;\Conid{Set}{}\<[E]%
+\\
+\>[B]{}\Varid{⟦}\;\Varid{[]}\;{}\<[17]%
+\>[17]{}\Varid{⟧ᵃ}\;\Conid{X}\;\Conid{Y}\;\Varid{σ}\;\Conid{Γ}\;\Varid{x}\;{}\<[39]%
+\>[39]{}\mathrel{=}\;\Conid{Y}\;\Conid{Γ}\;\Varid{x}{}\<[E]%
+\\
+\>[B]{}\Varid{⟦}\;\Conid{A}\;\Varid{∷}\;\Conid{Δ}\;{}\<[17]%
+\>[17]{}\Varid{⟧ᵃ}\;\Conid{X}\;\Conid{Y}\;\Varid{σ}\;\Conid{Γ}\;\Varid{x}\;{}\<[39]%
+\>[39]{}\mathrel{=}\;\Varid{⟦}\;\Conid{Δ}\;\Varid{⟧ᵃ}\;\Conid{X}\;\Conid{Y}\;\Varid{σ}\;((\Conid{A}\;\Varid{⟨}\;\Varid{σ}\;\Varid{⟩})\;\Varid{∷}\;\Conid{Γ})\;\Varid{x}{}\<[E]%
+\\[\blanklineskip]%
+\>[B]{}\Varid{⟦\char95 ⟧ᵃˢ}\;\mathbin{:}\;(\Conid{D}\;\mathbin{:}\;\Conid{ArgsD}\;\Conid{Ξ})\;(\Conid{X}\;\mathbin{:}\;\Conid{R.Fam})\;(\Conid{Y}\;\mathbin{:}\;\Conid{Fam}\;\Conid{X})\;{}\<[E]%
+\\
+\>[B]{}\hsindent{7}{}\<[7]%
+\>[7]{}\Varid{→}\;\Conid{TSub}\;\Conid{Ξ}\;\Varid{0}\;\Varid{→}\;(\Conid{Γ}\;\mathbin{:}\;\Conid{Cxt}\;\Varid{0})\;\Varid{→}\;\Conid{R.⟦}\;\Conid{D}\;\Varid{⟧ᵃˢ}\;\Conid{X}\;(\Varid{length}\;\Conid{Γ})\;\Varid{→}\;\Conid{Set}{}\<[E]%
+\\
+\>[B]{}\Varid{⟦}\;\Varid{[]}\;{}\<[17]%
+\>[17]{}\Varid{⟧ᵃˢ}\;\anonymous \;\anonymous \;\anonymous \;\anonymous \;\anonymous \;{}\<[39]%
+\>[39]{}\mathrel{=}\;\Varid{⊤}{}\<[E]%
+\\
+\>[B]{}\Varid{⟦}\;(\Conid{Δ}\;\Varid{⊢}\;\Conid{A})\;\Varid{∷}\;\Conid{Ds}\;{}\<[17]%
+\>[17]{}\Varid{⟧ᵃˢ}\;\Conid{X}\;\Conid{Y}\;\Varid{σ}\;\Conid{Γ}\;(\Varid{x}\;\Varid{,}\;\Varid{xs})\;{}\<[39]%
+\>[39]{}\mathrel{=}\;{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{⟦}\;\Conid{Δ}\;\Varid{⟧ᵃ}\;\Conid{X}\;(\Varid{λ}\;\iden{\Conid{Γ}^\prime}\;\iden{\Varid{x}^\prime}\;\Varid{→}\;\Conid{Y}\;\iden{\Conid{Γ}^\prime}\;\iden{\Varid{x}^\prime}\;(\Conid{A}\;\Varid{⟨}\;\Varid{σ}\;\Varid{⟩}))\;\Varid{σ}\;\Conid{Γ}\;\Varid{x}\;\Varid{×}\;\Varid{⟦}\;\Conid{Ds}\;\Varid{⟧ᵃˢ}\;\Conid{X}\;\Conid{Y}\;\Varid{σ}\;\Conid{Γ}\;\Varid{xs}{}\<[E]%
+\\[\blanklineskip]%
+\>[B]{}\Varid{⟦\char95 ⟧ᶜ}\;\mathbin{:}\;(\Conid{D}\;\mathbin{:}\;\Conid{ConD})\;(\Conid{X}\;\mathbin{:}\;\Conid{R.Fam})\;(\Conid{Y}\;\mathbin{:}\;\Conid{Fam}\;\Conid{X})\;\Varid{→}\;\Conid{Fam}\;\iden{\Varid{ℓ}^\prime}\;(\Conid{R.⟦}\;\Conid{D}\;\Varid{⟧ᶜ}\;\Conid{X}){}\<[E]%
+\\
+\>[B]{}\Varid{⟦}\;\Varid{ι}\;\{\mskip1.5mu \Conid{Ξ}\mskip1.5mu\}\;\Conid{B}\;\Conid{D}\;{}\<[17]%
+\>[17]{}\Varid{⟧ᶜ}\;\Conid{X}\;\Conid{Y}\;\Conid{Γ}\;\Varid{xs}\;\Conid{A}\;{}\<[39]%
+\>[39]{}\mathrel{=}\;{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Conid{Σ[}\;\Varid{σ}\;\Varid{∈}\;\Conid{TSub}\;\Conid{Ξ}\;\Varid{0}\;\mskip1.5mu]\;\Conid{B}\;\Varid{⟨}\;\Varid{σ}\;\Varid{⟩}\;\Varid{≡}\;\Conid{A}\;\Varid{×}\;\Varid{⟦}\;\Conid{D}\;\Varid{⟧ᵃˢ}\;\Conid{X}\;\Conid{Y}\;\Varid{σ}\;\Conid{Γ}\;\Varid{xs}{}\<[E]%
+\\[\blanklineskip]%
+\>[B]{}\Varid{⟦\char95 ⟧}\;\mathbin{:}\;(\Conid{D}\;\mathbin{:}\;\Conid{Desc})\;(\Conid{X}\;\mathbin{:}\;\Conid{R.Fam})\;(\Conid{Y}\;\mathbin{:}\;\Conid{Fam}\;\Conid{X})\;\Varid{→}\;\Conid{Fam}\;(\Conid{R.⟦}\;\Conid{D}\;\Varid{⟧}\;\Conid{X}){}\<[E]%
+\\
+\>[B]{}\Varid{⟦}\;\Conid{D}\;{}\<[17]%
+\>[17]{}\Varid{⟧}\;\Conid{X}\;\Conid{Y}\;\Conid{Γ}\;(\Varid{i}\;\Varid{,}\;\Varid{xs})\;\Conid{A}\;{}\<[39]%
+\>[39]{}\mathrel{=}\;\Varid{⟦}\;\Conid{D}\;\Varid{.rules}\;\Varid{i}\;\Varid{⟧ᶜ}\;\Conid{X}\;\Conid{Y}\;\Conid{Γ}\;\Varid{xs}\;\Conid{A}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+\end{figure}
+\begin{figure}[H]
+  \small
+\begin{minipage}[t]{.47\textwidth}
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{8}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{10}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Keyword{data}\;\Varid{\char95 ⊢\char95 ⦂\char95 }\;\mathbin{:}\;\Conid{Fam₀}\;\Conid{Raw}\;\Keyword{where}{}\<[E]%
+\\[\blanklineskip]%
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{var}\;{}\<[8]%
+\>[8]{}\mathbin{:}\;(\Varid{i}\;\mathbin{:}\;\Conid{A}\;\Varid{∈}\;\Conid{Γ})\;{}\<[E]%
+\\
+\>[8]{}\Varid{→}\;\{\mskip1.5mu \Varid{j}\;\mathbin{:}\;\Conid{Fin}\;(\Varid{length}\;\Conid{Γ})\mskip1.5mu\}\;{}\<[E]%
+\\
+\>[8]{}\Varid{→}\;\Conid{L.index}\;\Varid{i}\;\Varid{≡}\;\Varid{j}\;{}\<[E]%
+\\
+\>[8]{}\Varid{→}\;\mbox{\onelinecomment }{}\<[E]%
+\\
+\>[8]{}\hsindent{2}{}\<[10]%
+\>[10]{}\Conid{Γ}\;\Varid{⊢}\;(\text{\textasciigrave}\;\Varid{j})\;\Varid{⦂}\;\Conid{A}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+  \end{minipage}
+\begin{minipage}[t]{.47\textwidth}
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{8}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{10}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[3]{}\Varid{\char95 ∋\char95 }\;{}\<[8]%
+\>[8]{}\mathbin{:}\;(\Conid{A}\;\mathbin{:}\;\Conid{Ty})\;{}\<[E]%
+\\
+\>[8]{}\Varid{→}\;\Conid{Γ}\;\Varid{⊢}\;\Varid{r}\;\Varid{⦂}\;\Conid{A}\;{}\<[E]%
+\\
+\>[8]{}\Varid{→}\;\mbox{\onelinecomment }{}\<[E]%
+\\
+\>[8]{}\hsindent{2}{}\<[10]%
+\>[10]{}\Conid{Γ}\;\Varid{⊢}\;(\Conid{A}\;\Varid{∋}\;\Varid{r})\;\Varid{⦂}\;\Conid{A}{}\<[E]%
+\\
+\>[3]{}\Varid{op}\;{}\<[8]%
+\>[8]{}\mathbin{:}\;\Varid{⟦}\;\Conid{D}\;\Varid{⟧}\;\Conid{Raw}\;\Varid{\char95 ⊢\char95 ⦂\char95 }\;\Conid{Γ}\;\Varid{rs}\;\Conid{A}\;{}\<[E]%
+\\
+\>[8]{}\Varid{→}\;\mbox{\onelinecomment }{}\<[E]%
+\\
+\>[8]{}\hsindent{2}{}\<[10]%
+\>[10]{}\Conid{Γ}\;\Varid{⊢}\;\Varid{op}\;\Varid{rs}\;\Varid{⦂}\;\Conid{A}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+  \end{minipage}
+\end{figure}
+
+\begin{figure}[H]
+  \small
+% bidirectionally typed terms
+%Fam : (ℓ′ : Level) (X : R.Fam ℓ) → Set (ℓ ⊔ lsuc ℓ′)
+%Fam ℓ′ X = (Γ : Cxt 0) → X (length Γ) → Mode → Ty → Set ℓ′
+%
+\begin{hscode}\SaveRestoreHook
 \column{B}{@{}>{\hspre}l<{\hspost}@{}}%
@@
 \column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\Conid{Fam}\;\mathbin{:}\;(\Varid{ℓ′}\;\mathbin{:}\;\Conid{Level})\;(\Conid{X}\;\mathbin{:}\;\Conid{R.Fam}\;\Varid{ℓ})\;\Varid{→}\;\Conid{Set}\;(\Varid{ℓ}\;\Varid{⊔}\;\Varid{lsuc}\;\Varid{ℓ′}){}\<[E]%
+\>[B]{}\Varid{⟦\char95 ⟧ᵃ}\;{}\<[7]%
+\>[7]{}\mathbin{:}\;(\Conid{Δ}\;\mathbin{:}\;\Conid{TExps}\;\Conid{Ξ})\;(\Conid{X}\;\mathbin{:}\;\Conid{R.Fam})\;(\Conid{Y}\;\mathbin{:}\;(\Conid{Γ}\;\mathbin{:}\;\Conid{Cxt}\;\Varid{0})\;\Varid{→}\;\Conid{X}\;(\Varid{length}\;\Conid{Γ})\;\Varid{→}\;\Conid{Set})\;{}\<[E]%
 \\
-\>[B]{}\Conid{Fam}\;\Varid{ℓ′}\;\Conid{X}\;\mathrel{=}\;(\Conid{Γ}\;\mathbin{:}\;\Conid{Cxt}\;\Varid{0})\;\Varid{→}\;\Conid{X}\;(\Varid{length}\;\Conid{Γ})\;\Varid{→}\;\Conid{Mode}\;\Varid{→}\;\Conid{Ty}\;\Varid{→}\;\Conid{Set}\;\Varid{ℓ′}{}\<[E]%
-\\[\blanklineskip]%
-\>[B]{}\Varid{⟦\char95 ⟧ᵃ}\;{}\<[7]%
-\>[7]{}\mathbin{:}\;(\Conid{Δ}\;\mathbin{:}\;\Conid{TExps}\;\Conid{Ξ})\;(\Conid{X}\;\mathbin{:}\;\Conid{R.Fam}\;\Varid{ℓ})\;(\Conid{Y}\;\mathbin{:}\;(\Conid{Γ}\;\mathbin{:}\;\Conid{Cxt}\;\Varid{0})\;\Varid{→}\;\Conid{X}\;(\Varid{length}\;\Conid{Γ})\;\Varid{→}\;\Conid{Set}\;\Varid{ℓ′})\;{}\<[E]%
-\\
-\>[7]{}\Varid{→}\;\Conid{TSub}\;\Conid{Ξ}\;\Varid{0}\;\Varid{→}\;(\Conid{Γ}\;\mathbin{:}\;\Conid{Cxt}\;\Varid{0})\;\Varid{→}\;\Conid{R.⟦}\;\Conid{Δ}\;\Varid{⟧ᵃ}\;\Conid{X}\;(\Varid{length}\;\Conid{Γ})\;\Varid{→}\;\Conid{Set}\;\Varid{ℓ′}{}\<[E]%
+\>[7]{}\Varid{→}\;\Conid{TSub}\;\Conid{Ξ}\;\Varid{0}\;\Varid{→}\;(\Conid{Γ}\;\mathbin{:}\;\Conid{Cxt}\;\Varid{0})\;\Varid{→}\;\Conid{R.⟦}\;\Conid{Δ}\;\Varid{⟧ᵃ}\;\Conid{X}\;(\Varid{length}\;\Conid{Γ})\;\Varid{→}\;\Conid{Set}{}\<[E]%
 \\
@@
 \>[B]{}\Varid{⟦\char95 ⟧ᵃˢ}\;{}\<[8]%
-\>[8]{}\mathbin{:}\;(\Conid{D}\;\mathbin{:}\;\Conid{ArgsD}\;\Conid{Ξ})\;(\Conid{X}\;\mathbin{:}\;\Conid{R.Fam}\;\Varid{ℓ})\;(\Conid{Y}\;\mathbin{:}\;\Conid{Fam}\;\Varid{ℓ′}\;\Conid{X})\;{}\<[E]%
+\>[8]{}\mathbin{:}\;(\Conid{D}\;\mathbin{:}\;\Conid{ArgsD}\;\Conid{Ξ})\;(\Conid{X}\;\mathbin{:}\;\Conid{R.Fam})\;(\Conid{Y}\;\mathbin{:}\;\Conid{Fam}\;\Conid{X})\;{}\<[E]%
 \\
-\>[8]{}\Varid{→}\;\Conid{TSub}\;\Conid{Ξ}\;\Varid{0}\;\Varid{→}\;(\Conid{Γ}\;\mathbin{:}\;\Conid{Cxt}\;\Varid{0})\;\Varid{→}\;\Conid{R.⟦}\;\Varid{eraseᵃˢ}\;\Conid{D}\;\Varid{⟧ᵃˢ}\;\Conid{X}\;(\Varid{length}\;\Conid{Γ})\;\Varid{→}\;\Conid{Set}\;\Varid{ℓ′}{}\<[E]%
+\>[8]{}\Varid{→}\;\Conid{TSub}\;\Conid{Ξ}\;\Varid{0}\;\Varid{→}\;(\Conid{Γ}\;\mathbin{:}\;\Conid{Cxt}\;\Varid{0})\;\Varid{→}\;\Conid{R.⟦}\;\Varid{eraseᵃˢ}\;\Conid{D}\;\Varid{⟧ᵃˢ}\;\Conid{X}\;(\Varid{length}\;\Conid{Γ})\;\Varid{→}\;\Conid{Set}{}\<[E]%
 \\
@@
 \>[B]{}\hsindent{3}{}\<[3]%
-\>[3]{}\Varid{⟦}\;\Conid{Δ}\;\Varid{⟧ᵃ}\;\Conid{X}\;(\Varid{λ}\;\Conid{Γ'}\;\Varid{x'}\;\Varid{→}\;\Conid{Y}\;\Conid{Γ'}\;\Varid{x'}\;\Varid{d}\;(\Conid{A}\;\Varid{⟨}\;\Varid{σ}\;\Varid{⟩}))\;\Varid{σ}\;\Conid{Γ}\;\Varid{x}\;\Varid{×}\;\Varid{⟦}\;\Conid{Ds}\;\Varid{⟧ᵃˢ}\;\Conid{X}\;\Conid{Y}\;\Varid{σ}\;\Conid{Γ}\;\Varid{xs}{}\<[E]%
+\>[3]{}\Varid{⟦}\;\Conid{Δ}\;\Varid{⟧ᵃ}\;\Conid{X}\;(\Varid{λ}\;\iden{\Conid{Γ}^\prime}\;\iden{\Varid{x}^\prime}\;\Varid{→}\;\Conid{Y}\;\iden{\Conid{Γ}^\prime}\;\iden{\Varid{x}^\prime}\;\Varid{d}\;(\Conid{A}\;\Varid{⟨}\;\Varid{σ}\;\Varid{⟩}))\;\Varid{σ}\;\Conid{Γ}\;\Varid{x}\;\Varid{×}\;\Varid{⟦}\;\Conid{Ds}\;\Varid{⟧ᵃˢ}\;\Conid{X}\;\Conid{Y}\;\Varid{σ}\;\Conid{Γ}\;\Varid{xs}{}\<[E]%
 \\[\blanklineskip]%
-\>[B]{}\Varid{⟦\char95 ⟧ᶜ}\;\mathbin{:}\;(\Conid{D}\;\mathbin{:}\;\Conid{ConD})\;(\Conid{X}\;\mathbin{:}\;\Conid{R.Fam}\;\Varid{ℓ})\;(\Conid{Y}\;\mathbin{:}\;\Conid{Fam}\;\Varid{ℓ′}\;\Conid{X})\;\Varid{→}\;\Conid{Fam}\;\Varid{ℓ′}\;(\Conid{R.⟦}\;\Varid{eraseᶜ}\;\Conid{D}\;\Varid{⟧ᶜ}\;\Conid{X}){}\<[E]%
+\>[B]{}\Varid{⟦\char95 ⟧ᶜ}\;\mathbin{:}\;(\Conid{D}\;\mathbin{:}\;\Conid{ConD})\;(\Conid{X}\;\mathbin{:}\;\Conid{R.Fam})\;(\Conid{Y}\;\mathbin{:}\;\Conid{Fam}\;\Conid{X})\;\Varid{→}\;\Conid{Fam}\;(\Conid{R.⟦}\;\Varid{eraseᶜ}\;\Conid{D}\;\Varid{⟧ᶜ}\;\Conid{X}){}\<[E]%
 \\
@@
 \\[\blanklineskip]%
-\>[B]{}\Varid{⟦\char95 ⟧}\;\mathbin{:}\;(\Conid{D}\;\mathbin{:}\;\Conid{Desc})\;(\Conid{X}\;\mathbin{:}\;\Conid{R.Fam}\;\Varid{ℓ})\;(\Conid{Y}\;\mathbin{:}\;\Conid{Fam}\;\Varid{ℓ′}\;\Conid{X})\;\Varid{→}\;\Conid{Fam}\;\Varid{ℓ′}\;(\Conid{R.⟦}\;\Varid{erase}\;\Conid{D}\;\Varid{⟧}\;\Conid{X}){}\<[E]%
+\>[B]{}\Varid{⟦\char95 ⟧}\;\mathbin{:}\;(\Conid{D}\;\mathbin{:}\;\Conid{Desc})\;(\Conid{X}\;\mathbin{:}\;\Conid{R.Fam})\;(\Conid{Y}\;\mathbin{:}\;\Conid{Fam}\;\Conid{X})\;\Varid{→}\;\Conid{Fam}\;(\Conid{R.⟦}\;\Varid{erase}\;\Conid{D}\;\Varid{⟧}\;\Conid{X}){}\<[E]%
 \\
@@
 \end{hscode}\resethooks
-}
+\end{figure}
+
+\begin{figure}[H]
+  \small 
+\begin{minipage}[t]{.48\textwidth}
+%
+%  _⊢_⇐_ _⊢_⇒_
+%    : (Γ : Cxt 0) → Raw (length Γ) → Ty → Set
+%  Γ ⊢ r ⇐ A  = Γ ⊢ r [ Chk ] A
+%  Γ ⊢ r ⇒ A  = Γ ⊢ r [ Syn ] A
+%
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{11}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Keyword{mutual}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Keyword{data}\;\Varid{\char95 ⊢\char95 [\char95 ]\char95 }\;\mathbin{:}\;\Conid{Fam₀}\;\Conid{Raw}\;\Keyword{where}{}\<[E]%
+\\[\blanklineskip]%
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\Varid{var}\;\mathbin{:}\;(\Varid{i}\;\mathbin{:}\;\Conid{A}\;\Varid{∈}\;\Conid{Γ})\;{}\<[E]%
+\\
+\>[5]{}\hsindent{4}{}\<[9]%
+\>[9]{}\Varid{→}\;\Conid{L.index}\;\Varid{i}\;\Varid{≡}\;\Varid{j}\;{}\<[E]%
+\\
+\>[5]{}\hsindent{4}{}\<[9]%
+\>[9]{}\Varid{→}\;\mbox{\onelinecomment }{}\<[E]%
+\\
+\>[9]{}\hsindent{2}{}\<[11]%
+\>[11]{}\Conid{Γ}\;\Varid{⊢}\;(\text{\textasciigrave}\;\Varid{j})\;\Varid{⇒}\;\Conid{A}{}\<[E]%
+\\[\blanklineskip]%
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\Varid{\char95 ∋\char95 }\;\mathbin{:}\;(\Conid{A}\;\mathbin{:}\;\Conid{Ty})\;{}\<[E]%
+\\
+\>[5]{}\hsindent{4}{}\<[9]%
+\>[9]{}\Varid{→}\;\Conid{Γ}\;\Varid{⊢}\;\Varid{r}\;\Varid{⇐}\;\Conid{A}\;{}\<[E]%
+\\
+\>[5]{}\hsindent{4}{}\<[9]%
+\>[9]{}\Varid{→}\;\Conid{Γ}\;\Varid{⊢}\;(\Conid{A}\;\Varid{∋}\;\Varid{r})\;\Varid{⇒}\;\Conid{A}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+\end{minipage}
+\begin{minipage}[t]{.48\textwidth}
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{11}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[5]{}\Varid{\char95 ↑\char95 }\;\mathbin{:}\;\Conid{Γ}\;\Varid{⊢}\;\Varid{r}\;\Varid{⇒}\;\Conid{B}\;{}\<[E]%
+\\
+\>[5]{}\hsindent{4}{}\<[9]%
+\>[9]{}\Varid{→}\;\Conid{A}\;\Varid{≡}\;\Conid{B}\;{}\<[E]%
+\\
+\>[5]{}\hsindent{4}{}\<[9]%
+\>[9]{}\Varid{→}\;\mbox{\onelinecomment }{}\<[E]%
+\\
+\>[9]{}\hsindent{2}{}\<[11]%
+\>[11]{}\Conid{Γ}\;\Varid{⊢}\;\Varid{r}\;\Varid{⇐}\;\Conid{A}{}\<[E]%
+\\[\blanklineskip]%
+\>[5]{}\Varid{op}\;{}\<[9]%
+\>[9]{}\mathbin{:}\;\Varid{⟦}\;\Conid{D}\;\Varid{⟧}\;\Conid{Raw}\;\Varid{\char95 ⊢\char95 [\char95 ]\char95 }\;\Conid{Γ}\;\Varid{rs}\;\Varid{d}\;\Conid{A}\;{}\<[E]%
+\\
+\>[9]{}\Varid{→}\;\mbox{\onelinecomment }{}\<[E]%
+\\
+\>[9]{}\hsindent{2}{}\<[11]%
+\>[11]{}\Conid{Γ}\;\Varid{⊢}\;\Varid{op}\;\Varid{rs}\;[\mskip1.5mu \;\Varid{d}\;\mskip1.5mu]\;\Conid{A}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+
+\end{minipage}
+\end{figure}
+\subsection{Formal Proofs: Soundness as an Example}\label{subsec:formal-proofs}
+
+\begin{figure}[H]
+  \small
+% soundness proof
+\begin{minipage}{.48\textwidth}
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{26}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{30}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{33}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{46}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Keyword{mutual}{}\<[E]%
+\\[\blanklineskip]%
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{soundness}\;\mathbin{:}\;\Conid{Γ}\;\Varid{⊢}\;\Varid{r}\;[\mskip1.5mu \;\Varid{d}\;\mskip1.5mu]\;\Conid{A}\;{}\<[30]%
+\>[30]{}\Varid{→}\;{}\<[33]%
+\>[33]{}\Conid{Γ}\;\Varid{⊢}\;\Varid{r}\;\Varid{⦂}\;\Conid{A}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{soundness}\;(\Varid{var}\;\Varid{i}\;\Varid{eq})\;{}\<[26]%
+\>[26]{}\mathrel{=}\;\Varid{var}\;\Varid{i}\;\Varid{eq}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{soundness}\;(\Conid{A}\;\Varid{∋}\;\Varid{t})\;{}\<[26]%
+\>[26]{}\mathrel{=}\;\Conid{A}\;\Varid{∋}\;\Varid{soundness}\;\Varid{t}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{soundness}\;(\Varid{t}\;\Varid{↑}\;\Varid{refl})\;{}\<[26]%
+\>[26]{}\mathrel{=}\;\Varid{soundness}\;\Varid{t}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{soundness}\;(\Varid{op}\;\Varid{ts})\;{}\<[26]%
+\>[26]{}\mathrel{=}\;{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\Varid{op}\;(\Varid{soundnessᶜ}\;(\Conid{BD}\;\Varid{.rules}\;\anonymous )\;\Varid{ts}){}\<[E]%
+\\[\blanklineskip]%
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{soundnessᶜ}\;{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\mathbin{:}\;(\Conid{D}\;\mathbin{:}\;\Conid{ConD})\;{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\Varid{→}\;\Varid{⟦}\;\Conid{D}\;\Varid{⟧ᶜ}\;\Conid{Raw}\;\Varid{\char95 ⊢\char95 [\char95 ]\char95 }\;\Conid{Γ}\;\Varid{rs}\;\Varid{d}\;\Conid{A}\;{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\Varid{→}\;\Conid{T.⟦}\;\Varid{eraseᶜ}\;\Conid{D}\;\Varid{⟧ᶜ}\;\Conid{Raw}\;\Varid{\char95 ⊢\char95 ⦂\char95 }\;\Conid{Γ}\;\Varid{rs}\;\Conid{A}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{soundnessᶜ}\;(\Varid{ι}\;\anonymous \;\anonymous \;\Conid{Ds})\;(\anonymous \;\Varid{,}\;\Varid{σ}\;\Varid{,}\;\Varid{σ-eq}\;\Varid{,}\;\Varid{ts})\;{}\<[46]%
+\>[46]{}\mathrel{=}\;{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\Varid{σ}\;\Varid{,}\;\Varid{σ-eq}\;\Varid{,}\;\Varid{soundnessᵃˢ}\;\Conid{Ds}\;\Varid{ts}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+\end{minipage}
+\begin{minipage}{.5\textwidth}
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{17}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{23}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{26}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{27}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{35}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{40}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{41}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{45}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{50}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[3]{}\Varid{soundnessᵃˢ}\;{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\mathbin{:}\;(\Conid{Ds}\;\mathbin{:}\;\Conid{ArgsD}\;\Conid{Ξ})\;{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\Varid{→}\;\Varid{⟦}\;\Conid{Ds}\;\Varid{⟧ᵃˢ}\;{}\<[27]%
+\>[27]{}\Conid{Raw}\;\Varid{\char95 ⊢\char95 [\char95 ]\char95 }\;{}\<[40]%
+\>[40]{}\Varid{σ}\;\Conid{Γ}\;\Varid{rs}\;{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\Varid{→}\;\Conid{T.⟦}\;\Varid{eraseᵃˢ}\;\Conid{Ds}\;\Varid{⟧ᵃˢ}\;{}\<[27]%
+\>[27]{}\Conid{Raw}\;\Varid{\char95 ⊢\char95 ⦂\char95 }\;{}\<[40]%
+\>[40]{}\Varid{σ}\;\Conid{Γ}\;\Varid{rs}{}\<[E]%
+\\
+\>[3]{}\Varid{soundnessᵃˢ}\;\Varid{[]}\;{}\<[35]%
+\>[35]{}\anonymous \;{}\<[45]%
+\>[45]{}\mathrel{=}\;{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\Varid{tt}{}\<[E]%
+\\
+\>[3]{}\Varid{soundnessᵃˢ}\;((\Conid{Δ}\;\Varid{⊢[}\;\anonymous \;\mskip1.5mu]\;\anonymous )\;\Varid{∷}\;\Conid{Ds})\;(\Varid{t}\;\Varid{,}\;\Varid{ts})\;{}\<[45]%
+\>[45]{}\mathrel{=}\;{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\Varid{soundnessᵃ}\;\Conid{Δ}\;\Varid{t}\;\Varid{,}\;\Varid{soundnessᵃˢ}\;\Conid{Ds}\;\Varid{ts}{}\<[E]%
+\\[\blanklineskip]%
+\>[3]{}\Varid{soundnessᵃ}\;{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\mathbin{:}\;(\Conid{Δ}\;\mathbin{:}\;\Conid{TExps}\;\Conid{Ξ})\;{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\Varid{→}\;\Varid{⟦}\;\Conid{Δ}\;\Varid{⟧ᵃ}\;{}\<[17]%
+\>[17]{}\Conid{Raw}\;(\Varid{λ}\;\iden{\Conid{Γ}^\prime}\;\iden{\Varid{r}^\prime}\;\Varid{→}\;\iden{\Conid{Γ}^\prime}\;\Varid{⊢}\;\iden{\Varid{r}^\prime}\;[\mskip1.5mu \;\Varid{d}\;\mskip1.5mu]\;\Conid{A})\;{}\<[50]%
+\>[50]{}\Varid{σ}\;\Conid{Γ}\;\Varid{r}\;{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\Varid{→}\;\Conid{T.⟦}\;\Conid{Δ}\;\Varid{⟧ᵃ}\;{}\<[17]%
+\>[17]{}\Conid{Raw}\;(\Varid{λ}\;\iden{\Conid{Γ}^\prime}\;\iden{\Varid{r}^\prime}\;\Varid{→}\;\iden{\Conid{Γ}^\prime}\;\Varid{⊢}\;\iden{\Varid{r}^\prime}\;\Varid{⦂}\;\Conid{A})\;{}\<[50]%
+\>[50]{}\Varid{σ}\;\Conid{Γ}\;\Varid{r}{}\<[E]%
+\\
+\>[3]{}\Varid{soundnessᵃ}\;\Varid{[]}\;{}\<[23]%
+\>[23]{}\Varid{t}\;{}\<[26]%
+\>[26]{}\mathrel{=}\;\Varid{soundness}\;{}\<[41]%
+\>[41]{}\Varid{t}{}\<[E]%
+\\
+\>[3]{}\Varid{soundnessᵃ}\;(\anonymous \;\Varid{∷}\;\Conid{Δ})\;{}\<[23]%
+\>[23]{}\Varid{t}\;{}\<[26]%
+\>[26]{}\mathrel{=}\;\Varid{soundnessᵃ}\;\Conid{Δ}\;\Varid{t}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+\end{minipage}
+
+\end{figure}
+
+\subsection{Examples in \Agda} \label{subsec:formal-examples}
 

### 2023-07-11 tex/example.tex
[ tex ] Add a separate section on Examples

%!TEX root = BiSig.tex

\section{Examples}\label{sec:example}
Since our theory is not confined to a specific simply-typed language, we can employ it with any language that can be specified by a bidirectional binding signature.
To showcase its general applicability, we present two additional examples:
\begin{enumerate}
\item \PCF with products and let bindings
\item spine calculus: a simply-typed $\lambda$-calculus with applications in spine form.
\end{enumerate}
The first serves as a standard exercise in implementing a type synthesiser and the second for an example of a system with infinitely many operations.

\subsection{PCF with Products and Let Bindings}\label{subsec:PCF}
Although we have developed a theory for presenting a bidirectional type system and a condition for the decidability of bidirectional type synthesis, designing a bidirectional type system, or its signature, remains a creative activity.\footnote{This also applies to parsing: while there are grammar and parser generators, one still has to design a grammar to parse or disambiguate an ambiguous grammar.}
While the focus of this paper is not to design bidirectional type systems, we adopt the example of a bidirectional \PCF from \citet{Wadler2022} which extends bidirectional simply typed $\lambda$-calculus in \Cref{sec:key-ideas} by adding following rules:

\bgroup
\small
  \centering
  \begin{mathpar}
    \inferrule{ }{\Gamma \vdash \mathtt{z} \chk \mathtt{nat}}
    \and
    \inferrule{\Gamma \vdash t \chk \mathtt{nat}}{\Gamma \vdash \mathtt{s}(t) \chk \mathtt{nat}}
    \and
    \inferrule{\Gamma \vdash t \syn \mathtt{nat} \and \Gamma \vdash t_0 \chk A \and \Gamma, x : \mathtt{nat} \vdash t_1 \chk A}{\Gamma \vdash \mathtt{ifz}(t_0; x.t_1)(t) \chk A}
    \and
    \inferrule{\Gamma \vdash t \chk A \and \Gamma \vdash u \chk B}{\Gamma \vdash (t, u) \chk A \times B}
    \and
    \inferrule{\Gamma \vdash t \syn A \times B}{\Gamma \vdash \mathtt{proj}_1(t) \syn A}
    \and
    \inferrule{\Gamma \vdash u \syn A \times B}{\Gamma \vdash \mathtt{proj}_2(u) \syn A}
    \and
    \inferrule{\Gamma, x : A \vdash t \chk A}{\Gamma \vdash \mu x.\, x \chk A}
    \and
    \inferrule{\Gamma \vdash t \syn A \and \Gamma, x : A \vdash u \chk B}{\Gamma \vdash \mathtt{let}\;x = t\;\mathtt{in}\;u\chk B}
  \end{mathpar}
\egroup

In this language, there are additional type constructs other than $\bto$, so first we extend $\Sigma_{\bto}$ (\Cref{ex:type-signature-for-function-type}) with operations
In this language, we introduce additional type constructs apart from $\bto$.
To do so, we extend $\Sigma_{\bto}$ (as shown in \Cref{ex:type-signature-for-function-type}) with the operations $\mathsf{nat}$ and $\mathsf{times}$ such that
\[
  \arity({\mathsf{nat}}) = 0 \quad\text{and}\quad \arity({\mathsf{times}}) = 2
\]
enabling the introduction of $\mathtt{nat}$ and $A \times B$ as $\tyOp_{\mathsf{nat}}$ and $\tyOp_{\mathsf{times}}(A, B)$, respectively.
Following this, we also extend the bidirectional binding signature $\Omega_{\Lambda}^{\Leftrightarrow}$ with the following operations to correspond with each rule:
\begin{align*}
  \aritysymbol{\mathsf{z}}{\cdot}{\cdot}{\mathtt{nat}^{\chk}}
  && \aritysymbol{\mathsf{s}}{\cdot}{\mathtt{nat}^{\chk}}{\mathtt{nat}^{\chk}}
  && \aritysymbol{\mathsf{ifz}}{A}{\mathtt{nat}^{\syn}, A^{\chk}, A^{\chk}}{A^{\chk}} \\
  \aritysymbol{\mathsf{pair}}{A, B}{A^{\chk}, B^{\chk}}{A \times B^{\chk}}
  && \aritysymbol{\mathsf{proj}_1}{A, B}{A \times B^{\syn}}{A^{\syn}}
  && \aritysymbol{\mathsf{proj}_2}{A, B}{A \times B^{\syn}}{B^{\syn}} \\
  \aritysymbol{\mathsf{mu}}{A}{[A]A^{\chk}}{A^{\chk}}
  && \aritysymbol{\mathsf{let}}{A, B}{A^{\syn},[A]B^{\chk}}{B^{\chk}}
\end{align*}
With these operations, a straightforward check shows that every rule is mode-correct, therefore this as a mode-correct bidirectional type system.
As a result, we can apply the type synthesiser presented in \Cref{sec:type-synthesis} directly to this system, eliminating the need for re-implementation.

\subsection{Spine Calculus}\label{subsec:spine}

A spine calculus is a calculus with applications in spine form $t\;u_1\;\ldots\;u_n$ which consists of a head $t$ and an indefinite number of arguments $u_1\;u_2\;\dots\;u_n$.
This example might initially appear impossible, since the number of arguments a term construct specified by a signature can take  is finite.
However, we do not have any restriction on the number of symbols that may be inhabited in a signature, enabling us to have a corresponding type construct and term construct for each number $n$ of arguments.
This underscores the importance and necessity of having an arbitrary set (with decidable equality) as a set of operation symbols.
To be precise, consider the following rule

\bgroup
\small
  \centering
  \begin{mathpar}
    \inferrule{\Gamma \vdash t \syn \textstyle{\prod_{i = 1}^{n}} A_i \bto B \and \Gamma \vdash u_1 \chk A_1 \and \dots \and \Gamma \vdash u_n \chk A_n}{\Gamma \vdash t\;u_1\;\ldots\;u_n \syn B}
  \end{mathpar}
\egroup
extending the type signature $\Sigma_{\bto}$ with $\mathsf{fun}_n$ such that $\arity({\mathsf{fun}_n}) = n$ and the bidirectional binding signature $\Omega_{\Lambda}^{\Leftrightarrow}$ with $\mathtt{app}_n$ such that 
\[
  \aritysymbol{\mathsf{app}_n}{A_1,\ldots, A_n}{\tyOp_{\mathsf{fun}_n}(A_1, \ldots, A_n) \bto B^{\syn}, A_1^{\chk}, \ldots, A_n^{\chk}}{B}.
\]
Then each spine application $t\;u_1\;\ldots\;u_n$ of the length $n$ can be introduced as $\tmOp_{\mathsf{app}_n}(t; u_1; \ldots; u_n)$.

### 2023-07-11 tex/sec1-intro.tex
[ tex ] Update paper plan

@@
 The basic idea of bidirectional type synthesis is that while the problem of type inference is not decidable in general, for certain kinds of terms it is still possible to infer their types (for example, the type of a variable can be looked up in the context); for other kinds of terms, we can switch to the simpler problem of type checking, where the expected type of a raw term is also given so that there is more information to work with.
-More formally, every judgement in a bidirectional type system is assigned one of two \emph{modes}:
+More formally, every judgement in a bidirectional type system is extended with a \emph{mode}:
 \begin{enumerate*}
@@
 %every synthesised type in a typing rule is determined by previously synthesised types from its premises and its input if checking.
-As \citet{Pierce2000} noted, bidirectional type synthesis propagates type information locally within adjacent nodes of a raw term, and does not require unification as Damas--Milner type inference does---in fact, introducing long-distance unification constraints would undermine the essence of locality in bidirectional type synthesis.
+As \citet{Pierce2000} noted, bidirectional type synthesis propagates type information locally within adjacent nodes of a raw term, and does not require unification as Damas--Milner type inference does.
+Moreover, introducing long-distance unification constraints would even undermine the essence of locality in bidirectional type synthesis.
 
@@
 
-This paper is structured in logical order as follows.
+This paper is structured as follows.
 We first present a concrete overview of our theory using simply typed $\lambda$-calculus in \cref{sec:key-ideas}, prior to developing a general framework for specifying bidirectional type systems in \cref{sec:defs}.
@@
 The most heavyweight technical contribution lies in \cref{sec:type-synthesis}, where we introduce mode-correctness and bidirectional type synthesis.
-In \cref{sec:formalisation}, we sketch the formalisation of our theory in \Agda and give some examples other than simply typed $\lambda$-calculus.
+Some examples other than simply typed $\lambda$-calculus are given in \Cref{sec:example}.
+We sketch the design of the \Agda formalisation of our theory in \cref{sec:formalisation} briefly.
 We conclude in \cref{sec:future}, where we reflect on related techniques that can be used to improve the formal development and potential challenges in extending our theory to a more general setting.

### 2023-07-11 tex/sec3-defs.tex
[ tex ] Minor fix

@@
     \item a variable in $\Xi$ or
-    \item $\tyOp_{i}(A_1, \ldots, A_n)$ for some $i:I$ with $\arity(i) = n$ and types $A_1,\ldots, A_n$,
+    \item $\tyOp_{i}(A_1, \ldots, A_n)$ for some $i:I$ with $\arity(i) = n$ and types $A_1,\ldots, A_n$.
   \end{enumerate}
-  as shown in \Cref{fig:simple-type}.
-  A \emph{context} $\Gamma \colon \Cxt_{\Sigma}(\Xi)$ over $\Xi$ is a list $\Gamma$ of types over $\Xi$.%
-\todo{Name-type pairs?}
+  A \emph{context} $\Gamma \colon \Cxt_{\Sigma}(\Xi)$ over $\Xi$ is formed by $\cdot$ for the empty context and $\Gamma, A$ for an additional type $A$.
 \end{definition}
@@
 
-As \Cref{def:simple-signature} includes type variables, we introduce the typical meta-operation: substitution.
-\begin{definition}
 The \emph{substitution} for a function $\rho\colon \Xi \to \Type_{\Sigma}(\Xi')$, denoted by $\rho\colon \Sub{\Xi}{\Xi'}$, is a map which sends a type $A$ of $\Type_{\Sigma}(\Xi)$ to $\simsub{A}{\rho}$ of $\Type_{\Sigma}(\Xi')$ and is defined as usual.
 By abuse of notation, the substitution $\simsub{\Gamma}{\rho}$ of a context $\Gamma$ is defined by applying substitution to each $A$ in $\Gamma$.
-\end{definition}
 

### 2023-07-11 tex/sec5-synthesis.tex
[ tex ] Minor fix

@@
 It will be used especially in \Cref{subsec:bidirectional-synthesis-checking} for the proof of the decidability of bidirectional type synthesis and checking.
-Then, we conclude this section with the trichotomy on raw terms that has been illustrated at the beginning in \Cref{subsec:trichotomy}.
+Then, we conclude this section with the trichotomy on raw terms in \Cref{subsec:trichotomy}.
 
@@
       \begin{enumerate}
-        \item For the empty list, the statement is vacuously true by~\eqref{eq:synvar1}.
-        \item If $\dir{d_{i+1}}$ is checking, then the statement holds for~\eqref{eq:synvar2} by induction hypothesis.
-        \item If $\dir{d_{i+1}}$ is synthesis, then $\simsub{\Delta_{i+1}}{\rho_1} = \simsub{\Delta_{i+1}}{\rho_2}$ by mode-correctness~\eqref{eq:MC2} and induction hypothesis (of the list).
-          Therefore, under the same context $\Gamma, \simsub{\Delta_{i+1}}{\rho_1} = \Gamma, \simsub{\Delta_{i+1}}{\rho_2}$ the term $t_{i+1}$ must have the same synthesised type $\simsub{A_{i+1}}{\rho_1} = \simsub{A_{i+1}}{\rho_1}$ by induction hypothesis (of the typing derivation), so $\rho_1$ and $\rho_2$ agree on $\fv(A_{i+1})$ in addition to $\synvar(\biargs)$, as required for the case \eqref{eq:synvar3} in the definition of $\synvar$.
+        \item For the empty list, the statement is vacuously true by \Cref{eq:synvar1}.
+        \item If $\dir{d_{i+1}}$ is checking, then the statement holds for \Cref{eq:synvar2} by induction hypothesis.
+        \item If $\dir{d_{i+1}}$ is synthesis, then $\simsub{\Delta_{i+1}}{\rho_1} = \simsub{\Delta_{i+1}}{\rho_2}$ by mode-correctness \Cref{eq:MC2} and induction hypothesis (of the list).
+          Therefore, under the same context $\Gamma, \simsub{\Delta_{i+1}}{\rho_1} = \Gamma, \simsub{\Delta_{i+1}}{\rho_2}$ the term $t_{i+1}$ must have the same synthesised type $\simsub{A_{i+1}}{\rho_1} = \simsub{A_{i+1}}{\rho_1}$ by induction hypothesis (of the typing derivation), so $\rho_1$ and $\rho_2$ agree on $\fv(A_{i+1})$ in addition to $\synvar(\biargs)$, as required for the case \Cref{eq:synvar3} in the definition of $\synvar$.
       \end{enumerate}
@@
 
-The interesting part of the proof is the case for the $\Rule{Op}$ rule, so we shall give the insight first instead of jumping into the details.
+The interesting part of the theorem is the case for the $\Rule{Op}$ rule and it is rather complex.
+Here we shall give its insight first instead of jumping into the details.
 Recall that a typing derivation for $\tmOpts$ contains a substitution $\rho\colon \Xi \to \Type_{\Sigma}(\emptyset)$.
-The goal of type synthesis is exactly to define such a substitution $\rho$, so we have to start with a \emph{partial} substitution $\rho_0$, as an `accumulating argument', defined on $\fv(A_0)$ if $d$ is $\chk$ or otherwise nowhere.
-Then, as we visit each synthesis argument $\synbiarg[i+1]$, we may be able to extend the domain of $\rho_i$ to include the synthesis variables $\fv(A_{i + 1})$ if type synthesis is successful and also that the synthesised type can be \emph{unified with $A_{i+ 1}$} and \emph{extending} $\rho_i$ to $\bar{\rho_i} = \rho_{i+1}$.
-By mode-correctness, the accumulating substitution~$\rho_i$ would be defined on enough synthesis variables so that type synthesis or checking can be performed on $t_{i}$ with the context $\Gamma, \simsub{\Delta_{i+1}}{\rho_i}$ based on its mode derivation $\erase{\Gamma}, \vars{x}_i |-_{\Sigma, \Omega} t_i^{\dir{d_i}}$.
-Then, if we go through every argument $t_i$ successfully, we will have a substitution $\rho_n$ defined everywhere by mode-correctness.
+The goal of type synthesis is exactly to define such a substitution $\rho$, so we have to start with an `accumulating parameter': a substitution $\rho_0$ \emph{partially defined} on $\fv(A_0)$ if $d$ is $\chk$ or otherwise nowhere.
+By mode-correctness, the accumulating substitution~$\rho_i$ will be defined on enough synthesis variables so that type synthesis or checking can be performed on $t_{i}$ with the context $\Gamma, \vars{x}_{i} : \simsub{\Delta_{i}}{\rho_{i}}$ based on its mode derivation $\erase{\Gamma}, \vars{x}_i |-_{\Sigma, \Omega} t_i^{\dir{d_i}}$.
+If we visit a synthesis argument $\synbiarg[i+1]$, then we may extend the domain of $\rho_i$ to include the synthesis variables $\fv(A_{i + 1})$ if type synthesis is successful and also that the synthesised type can be \emph{unified with $A_{i+ 1}$} and thereby \emph{extend} $\rho_i$ to $\bar{\rho_i} = \rho_{i+1}$ with the unifier.
+Then, if we go through every $t_i$ successfully, we will have a total substitution $\rho_n$ by mode-correctness and a derivation of $\Gamma, \vars{x}_i : \Delta_i |-_{\Sigma, \Omega} t_i :^{\dir{d_i}} \simsub{A}{\rho_n}$ for each subterm $t_i$.
 
@@
   \begin{enumerate}
-    \item either there is a minimal extension\/ $\bar{\rho}$ of\/ $\rho$ such that\/ $\simsub{A}{\bar{\rho}} = B$, or 
-    \item there is no extension\/ $\sigma$ of\/ $\rho$ such that\/ $\simsub{A}{\sigma} = B$
+    \item either there is a minimal extension\/ $\bar{\rho}$ of\/ $\rho$ such that\/ $\simsub{A}{\bar{\rho}} = B$,
+    \item or there is no extension\/ $\sigma$ of\/ $\rho$ such that\/ $\simsub{A}{\sigma} = B$
   \end{enumerate}
@@
   \begin{itemize}
-    \item \SynRule{Var}: As $\erase{\Gamma} |- \isTerm{t}$, there exists a type $A$ such that $\Gamma |- \isTerm{x} \syn A$.
+    \item \SynRule{Var}: If $\erase{\Gamma} |- \isTerm{t}^{\syn}$, then $(x : A) \in \Gamma$ and thus $\Gamma |- \isTerm{x} \syn A$.
 
@@
         \item If $\Gamma |/- t \syn C$ for any $C$ but $\Gamma |- t \chk A$, then by inversion $\Gamma |- t \syn B$ for some $B = A$, thus a contradiction.
-        \item If $\Gamma |- t \syn B$ for some $B$, then either $A = B$ or $A \neq B$ by the decidable equality on $\Type_{\Sigma}(\Xi)$: 
+        \item If $\Gamma |- t \syn B$ for some $B$, then by decidable equality on $\Type_{\Sigma}(\Xi)$ either $A = B$ or $A \neq B$: 
           \begin{itemize}
@@
 
-            \item If there exists a minimal $\ext{\rho} \geq \rho_0$ defined on $\synvar(\biargs)$ such that~\eqref{eq:claim} holds, then by mode-correctness $\ext{\rho}$ is total and thus
+            \item If there exists a minimal $\ext{\rho} \geq \rho_0$ defined on $\synvar(\biargs)$ such that \eqref{eq:claim} holds, then by mode-correctness $\ext{\rho}$ is total and thus
               \[
@@
 
-                \item If there is a minimal $\ext{\ext{\rho}} \geq \ext{\rho}$ such that \eqref{eq:claim}, then by mode-correctness $\ext{\ext{\rho}}$ is total and
+                \item If there is a minimal $\ext{\ext{\rho}} \geq \ext{\rho}$ s.t.\ \eqref{eq:claim}, then by mode-correctness $\ext{\ext{\rho}}$ is total and
                   \[
@@
               \item If there is no $\sigma \geq \rho$ such that \eqref{eq:claim} holds for all $1 \leq i \leq m$ but a minimal $\gamma \geq \rho$ such that~\eqref{eq:claim} holds for all $1 \leq i \leq m + 1$, then we clearly have a contradiction.
-              \item There is a minimal $\ext{\rho} \geq \rho$ such that \eqref{eq:claim} holds for all $1 \leq i \leq m$.
+              \item There is a minimal $\ext{\rho} \geq \rho$ s.t.\ \eqref{eq:claim} holds for $1 \leq i \leq m$.
                 By case analysis on $\dir{d_{m+1}}$:
                 \begin{itemize}
-                  \item $\dir{d_{m+1}}$ is $\chk$: By mode-correctness, $\simsub{\Delta_{m+1}}{\ext{\rho}}$ and $\simsub{A_{m+1}}{\ext{\rho}}$ are defined and by the induction hypothesis it is decidable whether $\Gamma, \vars{x}_\isTerm{m+1} : \simsub{\Delta_{m+1}}{\ext{\rho}} |- \isTerm{t_{m+1}} \chk \simsub{A_{m+1}}{\ext{\rho}}^{\dir{d_{m+1}}}$.
-                    Clearly, if $\Gamma, \vars{x}_\isTerm{m+1} : \simsub{\Delta_{m+1}}{\ext{\rho}} |- \isTerm{t_{m+1}} \chk \simsub{A_{m+1}}{\ext{\rho}}^{\dir{d_{m+1}}}$ then the desired statement is proved; otherwise we can also easily derive a contradiction.
+                  \item $\dir{d_{m+1}}$ is $\chk$: By mode-correctness, $\simsub{\Delta_{m+1}}{\ext{\rho}}$ and $\simsub{A_{m+1}}{\ext{\rho}}$ are defined.
+                    By the induction hypothesis $ \Gamma, \vars{x}_\isTerm{m+1} : \simsub{\Delta_{m+1}}{\ext{\rho}} |- \isTerm{t_{m+1}} \chk \simsub{A_{m+1}}{\ext{\rho}}$ is decidable.
+                    Clearly, if $\Gamma, \vars{x}_\isTerm{m+1} : \simsub{\Delta_{m+1}}{\ext{\rho}} |- \isTerm{t_{m+1}} \chk \simsub{A_{m+1}}{\ext{\rho}}$ then the desired statement is proved; otherwise we can also easily derive a contradiction.
 
-                  \item $\dir{d_{m+1}}$ is $\syn$: By mode-correctness, $\simsub{\Delta_{m+1}}{\ext{\rho}}$ is defined and by the induction hypothesis, it is decidable that $\Gamma, \vars{x}_\isTerm{m+1} : \simsub{\Delta_{m+1}}{\ext{\rho}} |- \isTerm{t_{m+1}} \syn A$ for some $A$.
+                  \item $\dir{d_{m+1}}$ is $\syn$: By mode-correctness, $\simsub{\Delta_{m+1}}{\ext{\rho}}$ is defined.
+                    By the induction hypothesis, it is decidable that $\Gamma, \vars{x}_\isTerm{m+1} : \simsub{\Delta_{m+1}}{\ext{\rho}} |- \isTerm{t_{m+1}} \syn A$ for some $A$.
                     \begin{enumerate}
@@
                             However, by \Cref{thm:unique-syn}, the synthesised type $\simsub{A_{m+1}}{\gamma}$ must be unique, so $\gamma$ is an extension of $\ext{\rho}$ such that $\simsub{A_{m+1}}{\gamma} = A$, leading to a contradiction.
-                          \item If there is a minimal $\ext{\ext{\rho}} \geq \ext{\rho}$ such that $\simsub{A_{m+1}}{\ext{\ext{\rho}}} = A$, then it is not hard to show that $\ext{\ext{\rho}}$ is the minimal extension of $\rho$ such that \eqref{eq:claim} holds for all $1 \leq i \leq m + 1$.
+                          \item If there is a minimal $\ext{\ext{\rho}} \geq \ext{\rho}$ such that $\simsub{A_{m+1}}{\ext{\ext{\rho}}} = A$, then it is not hard to show that $\ext{\ext{\rho}}$ is also the minimal extension of $\rho$ such that \eqref{eq:claim} holds for all $1 \leq i \leq m + 1$.
                         \end{itemize}
@@
 If a contradiction proof is not of interest for implementation, these programs can be simplified by disregarding the cases that yield such contradiction proofs.
-Alternatively, we could consider generalising typing derivations like our generalised mode derivations (\Cref{fig:generalised-mode-derivations}).
-This could accommodate ill-typed cases and improve contradiction proofs to deliver more informative error messages.
-This would assist users in resolving issues with ill-typed terms, rather than simply returning a blatant 'no'.
+Alternatively, we could consider generalising typing derivations instead, like our generalised mode derivations (\Cref{fig:generalised-mode-derivations}).
+This could accommodate ill-typed cases and reformulate contradiction proofs positively to deliver more informative error messages.
+This would assist programmers in resolving issues with ill-typed terms, rather than simply returning a blatant 'no'.
 

### 2023-07-11 tex/sec1-intro.tex
[ tex ] Minor changes

@@
 Some examples other than simply typed $\lambda$-calculus are given in \Cref{sec:example}.
-We sketch the design of the \Agda formalisation of our theory in \cref{sec:formalisation} briefly.
-We conclude in \cref{sec:future}, where we reflect on related techniques that can be used to improve the formal development and potential challenges in extending our theory to a more general setting.
+We sketch the design of our formalisation in \cref{sec:formalisation} briefly.
+We conclude in \cref{sec:future}, where we reflect on techniques that may improve the formal development and potential challenges in extending our theory to a more general setting.
 

### 2023-07-11 tex/example.tex
[ tex ] Minor changes

@@
 \end{enumerate}
-The first serves as a standard exercise in implementing a type synthesiser and the second for an example of a system with infinitely many operations.
+The first is a standard exercise in implementing a type synthesiser and the second is an example of a system with infinitely many operations.
 
@@
 While the focus of this paper is not to design bidirectional type systems, we adopt the example of a bidirectional \PCF from \citet{Wadler2022} which extends bidirectional simply typed $\lambda$-calculus in \Cref{sec:key-ideas} by adding following rules:
-
 \bgroup
 \small
-  \centering
   \begin{mathpar}
@@
     \inferrule{\Gamma \vdash u \syn A \times B}{\Gamma \vdash \mathtt{proj}_2(u) \syn A}
-    \and
+    \\
     \inferrule{\Gamma, x : A \vdash t \chk A}{\Gamma \vdash \mu x.\, x \chk A}
@@
 enabling the introduction of $\mathtt{nat}$ and $A \times B$ as $\tyOp_{\mathsf{nat}}$ and $\tyOp_{\mathsf{times}}(A, B)$, respectively.
-Following this, we also extend the bidirectional binding signature $\Omega_{\Lambda}^{\Leftrightarrow}$ with the following operations to correspond with each rule:
-\begin{align*}
-  \aritysymbol{\mathsf{z}}{\cdot}{\cdot}{\mathtt{nat}^{\chk}}
+Following this, we also extend the bidirectional binding signature $\Omega_{\Lambda}^{\Leftrightarrow}$ with an operation for each rule above:
+\begin{flalign*}
+  &  \aritysymbol{\mathsf{z}}{\cdot}{\cdot}{\mathtt{nat}^{\chk}}
   && \aritysymbol{\mathsf{s}}{\cdot}{\mathtt{nat}^{\chk}}{\mathtt{nat}^{\chk}}
   && \aritysymbol{\mathsf{ifz}}{A}{\mathtt{nat}^{\syn}, A^{\chk}, A^{\chk}}{A^{\chk}} \\
-  \aritysymbol{\mathsf{pair}}{A, B}{A^{\chk}, B^{\chk}}{A \times B^{\chk}}
+  &  \aritysymbol{\mathsf{pair}}{A, B}{A^{\chk}, B^{\chk}}{A \times B^{\chk}}
   && \aritysymbol{\mathsf{proj}_1}{A, B}{A \times B^{\syn}}{A^{\syn}}
   && \aritysymbol{\mathsf{proj}_2}{A, B}{A \times B^{\syn}}{B^{\syn}} \\
-  \aritysymbol{\mathsf{mu}}{A}{[A]A^{\chk}}{A^{\chk}}
+  &  \aritysymbol{\mathsf{mu}}{A}{[A]A^{\chk}}{A^{\chk}}
   && \aritysymbol{\mathsf{let}}{A, B}{A^{\syn},[A]B^{\chk}}{B^{\chk}}
-\end{align*}
+\end{flalign*}
 With these operations, a straightforward check shows that every rule is mode-correct, therefore this as a mode-correct bidirectional type system.
@@
 \subsection{Spine Calculus}\label{subsec:spine}
+A spine application, denoted as $t\;u_1\;\ldots\;u_n$, is a variant of application that consists of a head term $t$ and an indefinite number of arguments $u_1\;u_2\;\dots\;u_n$.
+This arrangement allows direct access to the head term, making it practical in various applications.
+For instance, \Agda's core language employs this form of applications, as does its reflected syntax used for metaprogramming.
 
-A spine calculus is a calculus with applications in spine form $t\;u_1\;\ldots\;u_n$ which consists of a head $t$ and an indefinite number of arguments $u_1\;u_2\;\dots\;u_n$.
-This example might initially appear impossible, since the number of arguments a term construct specified by a signature can take  is finite.
-However, we do not have any restriction on the number of symbols that may be inhabited in a signature, enabling us to have a corresponding type construct and term construct for each number $n$ of arguments.
-This underscores the importance and necessity of having an arbitrary set (with decidable equality) as a set of operation symbols.
-To be precise, consider the following rule
+At first glance, accommodating this form of application may seem impossible, given the number of arguments a construct can accept is finite as specified by a signature.
+Nonetheless, there is no constraint on the total number of operation symbols a signature can have, allowing us to establish corresponding constructs for each number $n$ of arguments, thereby exhibiting the necessity of having an arbitrary set (with decidable equality) for operation symbols.
 
+To illustrate this, let us consider the following rule:
 \bgroup
 \small
-  \centering
   \begin{mathpar}
@@
 \egroup
-extending the type signature $\Sigma_{\bto}$ with $\mathsf{fun}_n$ such that $\arity({\mathsf{fun}_n}) = n$ and the bidirectional binding signature $\Omega_{\Lambda}^{\Leftrightarrow}$ with $\mathtt{app}_n$ such that
+The type signature $\Sigma_{\bto}$ can be extended with $\mathsf{fun}_n$ where $\arity({\mathsf{fun}_n}) = n$ for each $n : \N$.
+Similarly, the bidirectional binding signature $\Omega_{\Lambda}^{\Leftrightarrow}$ can be extended with 
 \[
-  \aritysymbol{\mathsf{app}_n}{A_1,\ldots, A_n}{\tyOp_{\mathsf{fun}_n}(A_1, \ldots, A_n) \bto B^{\syn}, A_1^{\chk}, \ldots, A_n^{\chk}}{B}.
+  \aritysymbol{\mathsf{app}_n}{A_1,\ldots, A_n, B}{\tyOp_{\mathsf{fun}_n}(A_1, \ldots, A_n) \bto B^{\syn}, A_1^{\chk}, \ldots, A_n^{\chk}}{B}
 \]
-Then each spine application $t\;u_1\;\ldots\;u_n$ of the length $n$ can be introduced as $\tmOp_{\mathsf{app}_n}(t; u_1; \ldots; u_n)$.
+also for each $n : \N$.
+Then each spine application $t\;u_1\;\ldots\;u_n$ can be introduced as $\tmOp_{\mathsf{app}_n}(t; u_1; \ldots; u_n)$.

### 2023-07-11 tex/sec1-intro.tex
[ tex ] Minor changes

@@
 The basic idea of bidirectional type synthesis is that while the problem of type inference is not decidable in general, for certain kinds of terms it is still possible to infer their types (for example, the type of a variable can be looked up in the context); for other kinds of terms, we can switch to the simpler problem of type checking, where the expected type of a raw term is also given so that there is more information to work with.
-More formally, every judgement in a bidirectional type system is extended with a \emph{mode}:
+More formally, every judgement in a bidirectional type system is extended with a \emph{mode:}
 \begin{enumerate*}
@@
 Moreover, unlike the plethora of available parser generators, `type-synthesiser generators' rarely exist, so each type synthesiser has to be independently built from scratch.
-To remedy the situation, we present a first theory of bidirectional typing, which amounts to a verified type-synthesiser generator for syntax-directed bidirectional simple type systems.
+To remedy the situation, we initiate a theory of bidirectional typing, which amounts to a verified type-synthesiser generator for syntax-directed bidirectional simple type systems.
 
@@
 %(although they do not emphasise the difference of their approach from the traditional one)
-Recall that the law of excluded middle $P + \neg P$ does not hold universally as an axiom in type theory---it means that whether $P$~holds is \emph{decidable}, and is non-trivial to prove.
+Recall that the law of excluded middle $P + \neg P$ does not hold as an axiom for every $P$ in Martin-L\"of type theory and we say $P$ is \emph{decidable} if it holds for $P$.
 For example, a proof of \emph{decidable equality}, i.e.

### 2023-07-11 tex/sec5-synthesis.tex
[ tex ] Minor changes

@@
 Recall that a typing derivation for $\tmOpts$ contains a substitution $\rho\colon \Xi \to \Type_{\Sigma}(\emptyset)$.
-The goal of type synthesis is exactly to define such a substitution $\rho$, so we have to start with an `accumulating parameter': a substitution $\rho_0$ \emph{partially defined} on $\fv(A_0)$ if $d$ is $\chk$ or otherwise nowhere.
+The goal of type synthesis is exactly to define such a substitution $\rho$, so we have to start with an `accumulating' substitution: a substitution $\rho_0$ that is partially defined on $\fv(A_0)$ if $d$ is $\chk$ or otherwise nowhere.
 By mode-correctness, the accumulating substitution~$\rho_i$ will be defined on enough synthesis variables so that type synthesis or checking can be performed on $t_{i}$ with the context $\Gamma, \vars{x}_{i} : \simsub{\Delta_{i}}{\rho_{i}}$ based on its mode derivation $\erase{\Gamma}, \vars{x}_i |-_{\Sigma, \Omega} t_i^{\dir{d_i}}$.

### 2023-07-11 tex/sec3-defs.tex
[ tex ] Put the context formation rules back in S3
[ tex ] Revised S3

--- 2023-07-10 tex/sec3-defs.tex
@@
   \item first we introduce a notion of arity and a notion of signature which includes a set of operation symbols and an assignment of arities to symbols;
-\item then, given a signature, we define sets of raw terms and derivations by primitive rules such as $\Rule{Var}$ and a rule schema for constructs $\tmOp_o$ indexed by an operational symbol~$o$.
+\item then, given a signature, we define raw terms and typing derivations by primitive rules such as $\Rule{Var}$ and a rule schema for constructs $\tmOp_o$ indexed by an operational symbol~$o$.
 \end{enumerate*}
 All these definitions are inductive as usual but include a \emph{rule schema} giving rise to as many rules as there are operation symbols in a signature.
-Upon moving from simple types to bidirectional typing, the notion of arity, initially as the number of arguments of an operation, is enriched to incorporate an extension context for variable binding and the mode for the direction of type information flow.
+As we move from simple types to bidirectional typing, the notion of arity, initially as the number of arguments of an operation, is enriched to incorporate an extension context for variable binding and the mode for the direction of type information flow.
 
@@
   \begin{figure}
-%    \begin{minipage}[b]{.6\textwidth}
       \centering
       \small
-      \judgbox{\Xi|-_{\Sigma} A}{$A$ is a well-formed type with type variables in $\Xi$}
+      \judgbox{\Xi|-_{\Sigma} A}{$A$ is a type with type variables in $\Xi$ for a signature $\Sigma$}
       \begin{mathpar}
@@
       \label{fig:simple-type}
-%    \end{minipage}
-%    \begin{minipage}[b]{.35\textwidth}
-%      \centering
-%      \small
-%      \judgbox{\Xi |-_{\Sigma} \Gamma}{}
-%      \begin{mathpar}
-%        \inferrule{ }{\Xi |-_{\Sigma} \cdot }\and
-%        \inferrule{\Xi |-_{\Sigma} A \\ \Xi|-_{\Sigma} \Gamma}{\Xi |-_{\Sigma} \Gamma, x : A}
-%      \end{mathpar}
-%      \caption{Contexts}
-%    \label{fig:simple-context}
-%    \end{minipage}
   \end{figure}
@@
     \item a variable in $\Xi$ or
-    \item $\tyOp_{i}(A_1, \ldots, A_n)$ for some $i:I$ with $\arity(i) = n$ and types $A_1,\ldots, A_n$,
+    \item $\tyOp_{i}(A_1, \ldots, A_n)$ for some $i:I$ with $\arity(i) = n$ and types $A_1,\ldots, A_n$.
   \end{enumerate}
-  as shown in \Cref{fig:simple-type}.
-  A \emph{context} $\Gamma \colon \Cxt_{\Sigma}(\Xi)$ over $\Xi$ is a list $\Gamma$ of types over $\Xi$.%
-\todo{Name-type pairs?}
 \end{definition}
@@
 
-As \Cref{def:simple-signature} includes type variables, we introduce the typical meta-operation: substitution.
 \begin{definition}
 The \emph{substitution} for a function $\rho\colon \Xi \to \Type_{\Sigma}(\Xi')$, denoted by $\rho\colon \Sub{\Xi}{\Xi'}$, is a map which sends a type $A$ of $\Type_{\Sigma}(\Xi)$ to $\simsub{A}{\rho}$ of $\Type_{\Sigma}(\Xi')$ and is defined as usual.
-  By abuse of notation, the substitution $\simsub{\Gamma}{\rho}$ of a context $\Gamma$ is defined by applying substitution to each $A$ in $\Gamma$.
 \end{definition}
@@
 
-A simply typed language consists of
+A simply typed language specifies
 \begin{enumerate*}
-  \item a set of raw terms $\isTerm{t}$ indexed by a set $V$ of untyped variables where each construct is allowed to bind some variables like $\Rule{Abs}$ and to take multiple arguments like $\Rule{App}$;
-  \item a set of typing derivations indexed by a list $\Gamma$ of typed variables, a type~$A$, and a raw term $\isTerm{t}$ to set their type constraints. 
+  \item a family of sets of raw terms $\isTerm{t}$ indexed by a list~$V$ of variables (that are currently in scope) where each construct is allowed to bind some variables like $\Rule{Abs}$ and to take multiple arguments like $\Rule{App}$;
+  \item a family of sets of typing derivations indexed by a typing context~$\Gamma$, a raw term~$\isTerm{t}$, and a type~$A$.
 \end{enumerate*}
-Therefore, to specify a term construct, we enrich the notion of arity with some sort for typing and extension context for variable binding:
+Therefore, to specify a term construct, we enrich the notion of arity with some sort for typing and extension context for variable binding.
+
 \begin{definition}\label{def:binding-arity}
@@
 
-For example, consider the $\Rule{Abs}$ rule (\Cref{fig:STLC-typing-derivations}).
-Its binding arity has the sort $\Type_{\Sigma_{\bto}}\{A, B\}$ and is
-\[
-  {[A]B} \to {(A \bto B)}.
-\]
+\begin{example}
+Consider the $\Rule{Abs}$ rule (\Cref{fig:STLC-typing-derivations}).
+Its binding arity has the sort $\Type_{\Sigma_{\bto}}\{A, B\}$ and is ${[A]B} \to {(A \bto B)}$.
 This means that the $\Rule{Abs}$ rule for derivations of\/ $\Gamma |- \lam{x}t : A \bto B$ contains:
 \begin{enumerate*}
-  \item a derivation of\/ $\Gamma, x : A |- t : B$ as an argument of type $B$ in a \emph{context extended} by variable $x$ of type $A$;
+  \item a derivation of\/ $\Gamma, x : A |- t : B$ as an argument of type $B$ in a typing context extended by a variable~$x$ of type~$A$;
   \item the type $A\bto B$ for itself.
 \end{enumerate*}
-In a similar vein, the arity of the $\Rule{App}$ rule is denoted as
-\[
-  {(A \bto B), A} \to {B}
-\]
-saying that any derivation of\/ $\Gamma \vdash t\;u : B$ accepts derivations $\Gamma \vdash t : A \bto B$ and $\Gamma \vdash u : A$ as its arguments.
-These arguments have types $A \bto B$ and $A$ respectively and empty extension contexts.
+In a similar vein, the arity of the $\Rule{App}$ rule is denoted as ${(A \bto B), A} \to {B}$.
+This specifies that any derivation of\/ $\Gamma \vdash t\;u : B$ accepts derivations $\Gamma \vdash t : A \bto B$ and $\Gamma \vdash u : A$ as its arguments.
+These arguments have types $A \bto B$ and $A$ respectively and their extension contexts are empty.
+\end{example}
 
@@
   \[
-    \arity(o) \colon O \to \sm{\Xi : \UU} \left(\Cxt_\Sigma(\Xi) \times \Type_\Sigma(\Xi)\right)^* \times \Type_\Sigma(\Xi).
+    \arity(o) \colon O \to \sm{\Xi : \UU} \left(\Type_\Sigma(\Xi)^* \times \Type_\Sigma(\Xi)\right)^* \times \Type_\Sigma(\Xi).
   \]
@@
 
-The inclusion of a variable set $\Xi$ for each operation, called its \emph{local context}, plays a crucial role in type synthesis.
+The inclusion of a set~$\Xi$ of type variables for each operation, called its \emph{local context}, plays a crucial role in type synthesis.
 For a typing rule like $\ChkRule{Abs}$, we will need to substitute \emph{concrete types}, i.e.\ types without any type variables, for variables $A, B$.
@@
 As such, this information forms part of the arity of an operation, and typing derivations, defined subsequently, will include an instantiation to assign concrete types to variables.
+
+For a simply typed language, first we define its raw terms.
 
@@
 \end{figure}
-  For a simply typed language $(\Sigma, \Omega)$, the set of \emph{raw terms} indexed by a context~$V$ of free variables consists of
+  For a simply typed language $(\Sigma, \Omega)$, the family of sets of \emph{raw terms} indexed by a list~$V$ of variables consists of
   \begin{enumerate*}
@@
     \item annotations $\isTerm{t \annotate A}$ for some raw term $t$ in $V$ and a type $A$, and
-    \item a construct $\tmOpts$ for some $\bop$ in $O$, where $\vars{x}_{\isTerm{i}}$'s are lists of variables whose length is equal to the length of~$\Delta_i$ and $t_i$'s are raw terms in the variable list $V, \vars{x}_i$
+    \item a construct $\tmOpts$ for some $\bop$ in~$O$, where $\vars{x}_{\isTerm{i}}$'s are lists of variables whose length is equal to the length of~$\Delta_i$, and $t_i$'s are raw terms in the variable list $V, \vars{x}_i$.
   \end{enumerate*}
-  corresponding to rules $\Rule{Var}$, $\Rule{Anno}$, and $\Rule{Op}$ in \Cref{fig:raw-terms} respectively.
+  These correspond to rules $\Rule{Var}$, $\Rule{Anno}$, and $\Rule{Op}$ in \Cref{fig:raw-terms} respectively.
+\end{definition}
+
+Before defining typing derivations, we need a definition of typing contexts.
+
+\begin{definition}
+A \emph{typing context} $\Gamma \colon \Cxt_{\Sigma}$ is formed by $\cdot$ for the empty context and $\Gamma, x : A$ for an additional variable~$x$ associated with a type $A : \Type_{\Sigma}(\emptyset)$.
+The list of variables in~$\Gamma$ is denoted~$\erase\Gamma$.
+%The substitution $\simsub{\Gamma}{\rho}$ of a context $\Gamma$ is defined by applying substitution to each $A$ in $\Gamma$.
 \end{definition}
@@
 Here we choose to include a substitution $\rho$ from the local context $\Xi$ to $\emptyset$ as part of its typing derivation explicitly:
+
 \begin{definition}\label{def:typing-derivations}
@@
   \end{figure}
-  For a simply typed language $(\Sigma, \Omega)$ the set of \emph{typing derivations} of $\Gamma \vdash \isTerm{t} : A$, indexed by a context $\Gamma$, a raw term with free variables in $\erase{\Gamma}$, and a type $A$, consists of
+  For a simply typed language $(\Sigma, \Omega)$, the family of sets of \emph{typing derivations} of $\Gamma \vdash \isTerm{t} : A$, indexed by a typing context $\Gamma : \Cxt_\Sigma$, a raw term~$\isTerm{t}$ with free variables in $\erase{\Gamma}$, and a type $A : \Type_\Sigma(\emptyset)$, consists of
   \begin{enumerate}
@@
 
+\begin{example}
 Raw terms (\Cref{fig:STLC-raw-terms}) and typing derivations (\Cref{fig:STLC-typing-derivations}) for simply typed $\lambda$-calculus can be specified by the type signature $\Sigma_{\bto}$ (\Cref{ex:type-signature-for-function-type}) and the binding signature consisting of
@@
 Rules $\Rule{Abs}$ and $\Rule{App}$ in simply typed $\lambda$-calculus are subsumed by the $\Rule{Op}$ rule schema, as applications $t\;u$ and abstractions $\lam{x}t$ can be introduced uniformly as $\tmOp_{\mathsf{app}}(t, u)$ and $\tmOp_{\mathsf{abs}}(x.t)$, respectively.
+\end{example}
 
 \subsection{Bidirectional Binding Signatures and Bidirectional Type Systems} \label{subsec:bidirectional-system}
-For a bidirectional type system, typing judgements appear in two forms: $\Gamma |- t \syn A$ and $\Gamma |- t \chk A$, but these two typing judgements can be considered as a single typing judgement $\Gamma |- t :^\dir{d} A$, additionally indexed by a \emph{mode} $d : \Mode$---which can either be $\syn$ or $\chk$.
-Therefore, to define a bidirectional type system, we further enrich the concept of binding arity by incorporating modes:
+For a bidirectional type system, typing judgements appear in two forms: $\Gamma |- t \syn A$ and $\Gamma |- t \chk A$, but these two typing judgements can be considered as a single typing judgement $\Gamma |- t :^\dir{d} A$, additionally indexed by a \emph{mode} $d : \Mode$---which can be either $\syn$ or $\chk$.
+Therefore, to define a bidirectional type system, we enrich the concept of binding arity to \emph{bidirectional binding arity}, which further specifies the mode for each of its arguments and for the conclusion:
 
@@
 \end{definition}
-When compared to a binding arity, a bidirectional binding arity provides further details, specifying the mode for each of its arguments and for the conclusion.
-Take the $\ChkRule{Abs}$ rule (\Cref{fig:STLC-bidirectional-typing-derivations}) for $\lam{x}t$ as an example.
+
+\begin{example}
+Consider the $\ChkRule{Abs}$ rule (\Cref{fig:STLC-bidirectional-typing-derivations}) for $\lam{x}t$.
 It has the arity ${[A]B^{\chk}}\to{(A \bto B)^{\chk}}$, indicating additionally that both $\lam{x}t$ and its argument $t$ are checking.
 Likewise, the $\SynRule{App}$ rule has the arity ${(A \bto B)^{\syn}, A^{\chk}} \to {B^{\syn}}$.
+\end{example}
 
@@
   \[
-    \mathit{ar}\colon O \to \sum_{\Xi : \UU} \left(\Cxt_{\Sigma}(\Xi) \times \Type_{\Sigma}(\Xi) \times {\Mode}\right)^* \times \Type_{\Sigma}(\Xi) \times {\Mode}.
+    \mathit{ar}\colon O \to \sum_{\Xi : \UU} \left(\Type_{\Sigma}(\Xi)^* \times \Type_{\Sigma}(\Xi) \times {\Mode}\right)^* \times \Type_{\Sigma}(\Xi) \times {\Mode}.
   \]
@@
 We write $\biop$ for an operation $o$ with a variable set $\Xi$ and its bidirectional binding arity with sort $\Type_{\Sigma}(\Xi)$.
-We call it \emph{checking} if $d$ is ${\chk}$ or \emph{synthesis} if $d$ is ${\syn}$; similarly its $i$-th argument is checking if $d_i$ is $\chk$ and synthesis if $d_i$ is $\syn$.
+We call it \emph{checking} if $d$ is ${\chk}$ or \emph{synthesising} if $d$ is ${\syn}$; similarly its $i$-th argument is checking if $d_i$ is $\chk$ and synthesising if $d_i$ is $\syn$.
 A bidirectional type system $(\Sigma, \Omega)$ refers to a pair of a type signature $\Sigma$ and a bidirectional binding signature $\Omega$.
@@
   \begin{itemize}
-    \item the set of \emph{bidirectional typing derivations} of $\Gamma \vdash_{\Sigma, \Omega} t :^\dir{d} A$, indexed by a context $\Gamma$ of typed variables, a raw term $\isTerm{t}$ under $\erase{\Gamma}$, a mode $\dir{d}$, and a type $A$, is defined in \Cref{fig:bidirectional-typing-derivations} and particularly
+    \item the set of \emph{bidirectional typing derivations} of $\Gamma \vdash_{\Sigma, \Omega} t :^\dir{d} A$, indexed by a typing context $\Gamma$, a raw term $\isTerm{t}$ under $\erase{\Gamma}$, a mode $\dir{d}$, and a type $A$, is defined in \Cref{fig:bidirectional-typing-derivations} and particularly
           \[
@@
   {\small$\boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \syn A}$} and {\small$\boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \chk A}$} are shorthands for ${\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} :^{\syn} A}$ and ${\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} :^{\chk} A}$ respectively.
-  A typing rule is \emph{checking} if its conclusion mode is $\chk$ and \emph{synthesis} if the conclusion mode is $\syn$.
+  A typing rule is \emph{checking} if its conclusion mode is $\chk$ and \emph{synthesising} if its conclusion mode is $\syn$.
 \end{definition}

### 2023-07-11 tex/sec6-formalisation.tex
[ tex ] Add a separate section on Examples
[ tex ] Minor changes
[ tex ] Two more subsections to go in Formalisation

@@
 
-\section{Formalisation and Examples} \label{sec:formalisation}
-%\begin{enumerate}
-%  \item (Bidirectional) binding signature, functor, and terms (extrinsic typing, raw terms, raw terms in some mode) --- 2.5 pp
-%  \item Compare the induction principle and the \Agda proof of Soundness --- 2 pp
-%  \item type synthesis and checking -- 0.5 p
-%  \item Examples including STLC (PCF), application in spine form --- 1p
-%
-%\end{enumerate}
-Our theory, initially developed in \Agda, has been later translated into the mathematical vernacular.
+\section{Formalisation} \label{sec:formalisation}
+As we have mentioned in \Cref{sec:intro}, our theory was initially developed with \Agda using \AxiomK and has been later translated into the mathematical vernacular.
 While the translation is reasonably straightforward, understanding the design of the formalisation itself could pose some difficulty.
-Indeed, as discussed in \Cref{sec:language-formalisation}, designing a language formalisation framework is still a challenging task and our formalisation is no exception.
-Unlike prior frameworks that have primarily focused on meta-properties centred around substitution and term traversal for intrinsically-typed terms, our theory of bidirectional type synthesis does not require term substitution but structural induction for extrinsically-typed terms.
+If the reader is already comfortable with the informal theory presented so far and assured by the existence of its formalisation, they may choose to skip this section.
+
+\paragraph{Revisiting language formalisation frameworks}
+Unlike prior frameworks~\citep{Allais2021,Fiore2022,Ahrens2022} that have primarily focused on meta-properties centred around substitution and term traversal for intrinsically-typed terms, our theory of bidirectional type synthesis does not require term substitution but structural induction for extrinsically-typed terms.
 The formal definitions of extrinsic typing in bidirectional type systems, including bidirectional typing derivations and mode derivations, are more complex than their intrinsic counterparts.
-For a specific language such as PCF, it is noted by \citet{Wadler2022} already that `extrinsically-typed terms require about 1.6 times as much code as intrinsically-typed' for their formalisation of type safety.
-As for our generic definitions, the formal type-theoretic definition for extrinsic typing derivations $\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A$ (\Cref{fig:extrinsic-typing}), for example, is a generic families of sets, also known as a \emph{proof-relevant predicate}, indexed by a generic raw \emph{term} $t$ in additional to a type $A$ and a context $\Gamma$ rather than just one family of sets of intrinsically-typed terms indexed by $A$ and $\Gamma$.
+For a specific language such as \PCF, \citet{Wadler2022} noted that `extrinsically-typed terms require about 1.6 times as much code as intrinsically-typed' for their formalisation of type safety.
+Take the formal type-theoretic definition of typing derivations $\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A$ (\Cref{fig:extrinsic-typing}) as an example.
+Its intrinsic definition consists of of just one family of sets of intrinsically-typed terms indexed by $A$ and $\Gamma$, but its extrinsic counterpart is a generic family of sets indexed by additionally a generic raw term $t$, involving constructions of two different layers.
 
-A categorical analysis of extrinsic typing requires not only \varcitet{Fiore1999}{'s} theory of abstract syntax with variable binding which backs \citet{Fiore2022}'s framework but also \varcitet{Hermida1998}{'s} interpretation of inductive proofs as algebras on the category of predicates.
-Thus, a faithful understanding of our formal constructions requires certain proficiency in category theory, which falls outside the scope of this paper.
+\paragraph{Category-theoretic analysis of well-typed terms}
+\citet{Fiore1999}{'s} theory of abstract syntax and variable binding forms the foundation of \citet{Fiore2022}'s framework and inspires frameworks by \citet{Allais2021,Ahrens2022}. 
+We may sketch the idea of their analysis as follows.
+The set of (untyped) abstract syntax trees for a language can be understood as
+\begin{enumerate*}
+  \item a family of sets $\Term_{\Gamma}$ of well-scoped terms under a context~$\Gamma$ with
+  \item variable renaming for a function $\sigma\colon \Gamma \to \Delta$ between variables acting as a functorial map from $\Term_{\Gamma}$ to $\Term_{\Delta}$, i.e.\ a presheaf $\Term\colon \mathbb{F} \to \mathsf{Set}$, and
+  \item an \emph{initial} algebra $[\mathsf{v}, \mathsf{op}]$ on~$\Term$ given by the variable rule as a map $\mathsf{v}$ from the presheaf $V\colon \mathbb{F} \hookrightarrow \mathsf{Set}$ of variables to $\Term$ and other constructs as $\mathsf{op}\colon \mathbb{\Sigma}\Term \to \Term$ where $\mathbb{F}$ is the category of contexts, the functor $\mathbb{\Sigma}\colon \mathsf{Set}^\mathbb{F} \to \mathsf{Set}^\mathbb{F}$ encodes the binding arities of constructs, and the initiality amounts to structural recursion, i.e.\ \emph{term traversal}.
+\end{enumerate*}
+Substitution is modelled as some monoid multiplication with the notion of strength and a suitable monoidal structural on the category of presheaves.
+To put it succinctly, it is the free $\mathbb{\Sigma}$-monoid over the presheaf~$V$ of variables.
 
-As such, in this section, we will merely provide an overview of our design and draw an analogy between the construction of simple types in \Cref{subsec:formal-simple-types} and bidirectional typing rules in \Cref{subsec:formal-extrinsic-typing} in plain terms, avoiding category theory deliberately.
-On the other hand, the formal proofs in our theory closely mirror the informal versions presented so far, thus we will only illustrate the soundness part of \Cref{lem:soundness-completeness} as an example in \Cref{subsec:formal-proofs}.
-We conclude the technical content of this paper with some mode-correct simply typed languages in \Cref{subsec:formal-examples}.
+\paragraph{Type-theoretic construction of well-typed terms}
+Fortunately, in type theory, constructing the initial algebra of well-typed terms boils down to defining an inductive type with a few constructors that align primitive rules (such as $\Rule{Var}$) and a rule schema; term traversal can be defined by as usual pattern matching for the inductive type~\citep{Fiore2022}.
+
+\paragraph{Whither a theory of extrinsic typing?}
+In \citep{Fiore1999} or existing frameworks, extrinsic typing is not studied.
+Our formalisation is inspired by \varcitet{Hermida1998}{'s} interpretation of structural induction as algebras on the category of predicates, viewing extrinsic definitions as endofunctors on the category of predicates.
+However, a category-theoretic analysis of our formal constructions in line with \varcitet{Fiore1999}{'s} theory requires proficiency in category theory, which falls outside the scope of this paper.
+
+As such, in this section, we will merely provide an overview of our design and explain the construction of simple types in \Cref{subsec:formal-simple-types} and bidirectional typing rules in \Cref{subsec:formal-extrinsic-typing} intuitively.
+On the other hand, the formal proofs closely mirror their informal description presented so far, thus we will only illustrate the `if' part of \Cref{lem:soundness-completeness} as an example in \Cref{subsec:formal-proofs}.
 
@@
 
-% simple types
-\begin{figure}[h]
+We begin the formal definition of signatures and simple types (\Cref{def:simple-signature}):
+\begin{figure}[H]
   \small
@@
 \\[\blanklineskip]%
-\>[B]{}\Keyword{data}\;\Conid{Tm}\;(\Conid{Θ}\;\mathbin{:}\;\Conid{ℕ})\;\mathbin{:}\;\Conid{Set}\;\Keyword{where}{}\<[E]%
+\>[B]{}\Keyword{data}\;\Conid{Ty}\;(\Conid{Ξ}\;\mathbin{:}\;\Conid{ℕ})\;\mathbin{:}\;\Conid{Set}\;\Keyword{where}{}\<[E]%
 \\
@@
 \>[3]{}\text{\textasciigrave}\anonymous \;{}\<[8]%
-\>[8]{}\mathbin{:}\;\Conid{Fin}\;\Conid{Θ}\;{}\<[24]%
-\>[24]{}\Varid{→}\;\Conid{Tm}\;\Conid{Θ}{}\<[E]%
+\>[8]{}\mathbin{:}\;\Conid{Fin}\;\Conid{Ξ}\;{}\<[24]%
+\>[24]{}\Varid{→}\;\Conid{Ty}\;\Conid{Ξ}{}\<[E]%
 \\
@@
 \>[3]{}\Varid{op}\;{}\<[8]%
-\>[8]{}\mathbin{:}\;\Varid{⟦}\;\Conid{D}\;\Varid{⟧}\;(\Conid{Tm}\;\Conid{Θ})\;{}\<[24]%
-\>[24]{}\Varid{→}\;\Conid{Tm}\;\Conid{Θ}{}\<[E]%
+\>[8]{}\mathbin{:}\;\Varid{⟦}\;\Conid{D}\;\Varid{⟧}\;(\Conid{Ty}\;\Conid{Ξ})\;{}\<[24]%
+\>[24]{}\Varid{→}\;\Conid{Ty}\;\Conid{Ξ}{}\<[E]%
 \ColumnHook
@@
 \end{figure}
-\subsection{Defining Raw Terms and Extrinsic Typing Derivations}\label{subsec:formal-extrinsic-typing}
+This definition mirrors our informal counterpart, with the exception of \ensuremath{\Varid{⟦}\;\Conid{D}\;\Varid{⟧}} for a signature~\ensuremath{\Conid{D}}.
+As we are defining simple types inductively, the arity \ensuremath{(\Varid{rules}\;\Varid{o})} is intended to denote the number of \ensuremath{\Conid{Ty}} arguments, referring back to what we are in the process of defining.
+To circumvent this self-reference, we first define \ensuremath{\Varid{⟦}\;\Conid{D}\;\Varid{⟧}} on an arbitrary \ensuremath{\Conid{Set}} rather than \ensuremath{\Conid{Ty}} itself.
+Consequently, the inductive type of types specified by a signature can be defined using two constructors, \ensuremath{\text{\textasciigrave}\anonymous } and \ensuremath{\Varid{op}}.
+A variable is represented by \ensuremath{\text{\textasciigrave}\Varid{n}} for some inhabitant \ensuremath{\Varid{n}} of the type \ensuremath{\Conid{Fin}\;\Conid{Ξ}} of natural numbers less than~\ensuremath{\Conid{Ξ}}.
+Each inhabitant \ensuremath{\Varid{op}\;(\Varid{i}\;\Varid{,}\;\Varid{ts})} consists of an operation symbol \ensuremath{\Varid{i}} and a $n$-tuple \ensuremath{\Varid{ts}}.
 
+From a categorical perspective, \ensuremath{\Varid{⟦}\;\Conid{D}\;\Varid{⟧}} can be understood as the functor from \ensuremath{\Conid{Set}} to \ensuremath{\Conid{Set}} which maps $X$ to a $O$-indexed coproduct $\sum_{i \in O} X ^ {\arity(o)}$ of products.
+The type \ensuremath{\Conid{Ty}} is then the free \ensuremath{\Varid{⟦}\;\Conid{D}\;\Varid{⟧}}-algebra over the type \ensuremath{\Conid{Fin}\;\Conid{Ξ}} or the initial algebra for the functor \ensuremath{\Conid{Fin}\;\Conid{Ξ}\;\Varid{+}\;\Varid{⟦}\;\Conid{D}\;\Varid{⟧}}.
+
+In short, to define terms by a signature, we define additionally a \emph{signature functor} \ensuremath{\Varid{⟦}\;\Conid{D}\;\Varid{⟧}}.
+
+\subsection{Defining Raw Terms}\label{subsec:formal-extrinsic-typing}
+
+The following record types represent the type \ensuremath{\Conid{ArgD}} for arguments $[\Delta]A^{\dir{d}}$, the type \ensuremath{\Conid{ConD}} for operations $\biop$, and the type \ensuremath{\Conid{Desc}} for bidirectional binding signatures, respectively, verbatim.
+Removing \ensuremath{\Varid{mode}} from these definitions recover the definitions of binding arguments, arities, and signatures.
 \begin{figure}[H]
@@
 \column{5}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{12}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{19}{@{}>{\hspre}l<{\hspost}@{}}%
 \column{E}{@{}>{\hspre}l<{\hspost}@{}}%
@@
 \>[3]{}\hsindent{2}{}\<[5]%
-\>[5]{}\Varid{cxt}\;{}\<[12]%
-\>[12]{}\mathbin{:}\;\Conid{Cxt}\;\Conid{Ξ}{}\<[E]%
+\>[5]{}\Varid{cxt}\;{}\<[19]%
+\>[19]{}\mathbin{:}\;\Conid{Cxt}\;\Conid{Ξ}{}\<[E]%
 \\
 \>[3]{}\hsindent{2}{}\<[5]%
-\>[5]{}\Varid{type}\;{}\<[12]%
-\>[12]{}\mathbin{:}\;\Conid{TExp}\;\Conid{Ξ}{}\<[E]%
+\>[5]{}\Varid{type}\;{}\<[19]%
+\>[19]{}\mathbin{:}\;\Conid{Ty}\;\Conid{Ξ}{}\<[E]%
 \\
 \>[3]{}\hsindent{2}{}\<[5]%
-\>[5]{}\Varid{mode}\;{}\<[12]%
-\>[12]{}\mathbin{:}\;\Conid{Mode}{}\<[E]%
+\>[5]{}\dir{\Varid{mode}}\;{}\<[19]%
+\>[19]{}\mathbin{:}\;\Conid{Mode}{}\<[E]%
 \ColumnHook
@@
 \column{5}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{14}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{18}{@{}>{\hspre}l<{\hspost}@{}}%
 \column{E}{@{}>{\hspre}l<{\hspost}@{}}%
@@
 \>[3]{}\hsindent{2}{}\<[5]%
-\>[5]{}\{\mskip1.5mu \Varid{vars}\mskip1.5mu\}\;{}\<[14]%
-\>[14]{}\mathbin{:}\;\Conid{ℕ}{}\<[E]%
+\>[5]{}\{\mskip1.5mu \Varid{vars}\mskip1.5mu\}\;{}\<[18]%
+\>[18]{}\mathbin{:}\;\Conid{ℕ}{}\<[E]%
 \\
 \>[3]{}\hsindent{2}{}\<[5]%
-\>[5]{}\Varid{args}\;{}\<[14]%
-\>[14]{}\mathbin{:}\;\Conid{List}\;(\Conid{ArgD}\;\Varid{vars}){}\<[E]%
+\>[5]{}\Varid{args}\;{}\<[18]%
+\>[18]{}\mathbin{:}\;\Conid{List}\;(\Conid{ArgD}\;\Varid{vars}){}\<[E]%
 \\
 \>[3]{}\hsindent{2}{}\<[5]%
-\>[5]{}\Varid{type}\;{}\<[14]%
-\>[14]{}\mathbin{:}\;\Conid{TExp}\;\Varid{vars}{}\<[E]%
+\>[5]{}\Varid{type}\;{}\<[18]%
+\>[18]{}\mathbin{:}\;\Conid{Ty}\;\Varid{vars}{}\<[E]%
 \\
 \>[3]{}\hsindent{2}{}\<[5]%
-\>[5]{}\Varid{mode}\;{}\<[14]%
-\>[14]{}\mathbin{:}\;\Conid{Mode}{}\<[E]%
+\>[5]{}\dir{\Varid{mode}}\;{}\<[18]%
+\>[18]{}\mathbin{:}\;\Conid{Mode}{}\<[E]%
 \ColumnHook
@@
 \end{figure}
-
-% functors for raw terms
+Our raw terms, being well-scoped and indexed by a list $V$ of free variables, requires another definition of a signature functor for their construction.
+This involves mapping \ensuremath{\Conid{Fam}} to \ensuremath{\Conid{Fam}}, where \ensuremath{\Conid{Fam}\;\mathrel{=}\;\Conid{ℕ}\;\Varid{→}\;\Conid{Set}} represents the family of sets indexed by a number (representing the number of free variables).
+The signature endofunctor for constructing raw terms shares similarities with the signature functor used for defining simple types, but rather than using a single functor, we employ four distinct functors that are defined separately for the extension context, arguments, constructs, and the coproduct indexed by constructs.
+Each of these corresponds to a variant of \ensuremath{\Varid{⟦\char95 ⟧}}:
 \begin{figure}[H]
   \small 
-  \begin{minipage}[t]{.57\textwidth}
-%Fam : (ℓ : Level) → Set (lsuc ℓ)
-%Fam ℓ = ℕ → Set ℓ
-%
-%Fam₀ : Set₁
-%Fam₀ = Fam 0ℓ
+  \begin{minipage}[t]{.52\textwidth}
 \begin{hscode}\SaveRestoreHook
@@
 \>[27]{}\mathrel{=}\;\Varid{⟦}\;\Conid{Δ}\;\Varid{⟧ᵃ}\;\Conid{X}\;\Varid{n}\;\Varid{×}\;\Varid{⟦}\;\Conid{Ds}\;\Varid{⟧ᵃˢ}\;\Conid{X}\;\Varid{n}{}\<[E]%
-\\[\blanklineskip]%
-\>[B]{}\Varid{⟦\char95 ⟧ᶜ}\;{}\<[8]%
-\>[8]{}\mathbin{:}\;\Conid{ConD}\;{}\<[19]%
-\>[19]{}\Varid{→}\;\Conid{Fam}\;\Varid{→}\;\Conid{Fam}{}\<[E]%
-\\
-\>[B]{}\Varid{⟦}\;\Varid{ι}\;\anonymous \;\Conid{D}\;{}\<[17]%
-\>[17]{}\Varid{⟧ᶜ}\;{}\<[22]%
-\>[22]{}\Conid{X}\;{}\<[27]%
-\>[27]{}\mathrel{=}\;\Varid{⟦}\;\Conid{D}\;\Varid{⟧ᵃˢ}\;\Conid{X}{}\<[E]%
-\\[\blanklineskip]%
-\>[B]{}\Varid{⟦\char95 ⟧}\;{}\<[8]%
-\>[8]{}\mathbin{:}\;\Conid{Desc}\;{}\<[19]%
-\>[19]{}\Varid{→}\;\Conid{Fam}\;\Varid{→}\;\Conid{Fam}{}\<[E]%
-\\
-\>[B]{}\Varid{⟦}\;\Conid{D}\;{}\<[17]%
-\>[17]{}\Varid{⟧}\;{}\<[22]%
-\>[22]{}\Conid{X}\;\Varid{n}\;{}\<[27]%
-\>[27]{}\mathrel{=}\;\Conid{Σ[}\;\Varid{i}\;\Varid{∈}\;\Conid{D}\;\Varid{.Op}\;\mskip1.5mu]\;\Varid{⟦}\;\Conid{D}\;\Varid{.rules}\;\Varid{i}\;\Varid{⟧ᶜ}\;\Conid{X}\;\Varid{n}{}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-  \end{minipage}
-  \begin{minipage}[t]{.4\textwidth}
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{7}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{32}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\Keyword{infix}\;\Varid{5}\;\Varid{\char95 ∋\char95 }{}\<[E]%
-\\[\blanklineskip]%
-\>[B]{}\Keyword{data}\;\Conid{Raw}\;\mathbin{:}\;\Conid{ℕ}\;\Varid{→}\;\Conid{Set}\;\Keyword{where}{}\<[E]%
-\\
-\>[B]{}\hsindent{3}{}\<[3]%
-\>[3]{}\text{\textasciigrave}\anonymous \;{}\<[7]%
-\>[7]{}\mathbin{:}\;\Conid{Fin}\;\Varid{n}\;{}\<[32]%
-\>[32]{}\Varid{→}\;\Conid{Raw}\;\Varid{n}{}\<[E]%
-\\
-\>[B]{}\hsindent{3}{}\<[3]%
-\>[3]{}\Varid{\char95 ∋\char95 }\;\mathbin{:}\;(\Conid{A}\;\mathbin{:}\;\Conid{Ty})\;(\Varid{t}\;\mathbin{:}\;\Conid{Raw}\;\Varid{n})\;{}\<[32]%
-\>[32]{}\Varid{→}\;\Conid{Raw}\;\Varid{n}{}\<[E]%
-\\
-\>[B]{}\hsindent{3}{}\<[3]%
-\>[3]{}\Varid{op}\;{}\<[7]%
-\>[7]{}\mathbin{:}\;\Varid{⟦}\;\Conid{D}\;\Varid{⟧}\;\Conid{Raw}\;\Varid{n}\;{}\<[32]%
-\>[32]{}\Varid{→}\;\Conid{Raw}\;\Varid{n}{}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-  \end{minipage}
-\end{figure}
-
-\begin{figure}[H]
-  \small
-%Fam : (ℓ′ : Level) → (X : R.Fam ℓ) → Set (ℓ ⊔ lsuc ℓ′)
-%Fam ℓ′ X = (Γ : Cxt 0) → X (length Γ) → Ty → Set ℓ′
-%
-%Fam₀ : (X : R.Fam ℓ) → Set (ℓ ⊔ lsuc 0ℓ)
-%Fam₀ X = Fam 0ℓ X
-%
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{7}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{17}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{39}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\Varid{⟦\char95 ⟧ᵃ}\;{}\<[7]%
-\>[7]{}\mathbin{:}\;(\Conid{Δ}\;\mathbin{:}\;\Conid{TExps}\;\Conid{Ξ})\;(\Conid{X}\;\mathbin{:}\;\Conid{R.Fam})\;(\Conid{Y}\;\mathbin{:}\;(\Conid{Γ}\;\mathbin{:}\;\Conid{Cxt}\;\Varid{0}){}\<[E]%
-\\
-\>[7]{}\Varid{→}\;\Conid{X}\;(\Varid{length}\;\Conid{Γ})\;\Varid{→}\;\Conid{Set})\;{}\<[E]%
-\\
-\>[7]{}\Varid{→}\;\Conid{TSub}\;\Conid{Ξ}\;\Varid{0}\;\Varid{→}\;(\Conid{Γ}\;\mathbin{:}\;\Conid{Cxt}\;\Varid{0})\;\Varid{→}\;\Conid{R.⟦}\;\Conid{Δ}\;\Varid{⟧ᵃ}\;\Conid{X}\;(\Varid{length}\;\Conid{Γ})\;{}\<[E]%
-\\
-\>[7]{}\Varid{→}\;\Conid{Set}{}\<[E]%
-\\
-\>[B]{}\Varid{⟦}\;\Varid{[]}\;{}\<[17]%
-\>[17]{}\Varid{⟧ᵃ}\;\Conid{X}\;\Conid{Y}\;\Varid{σ}\;\Conid{Γ}\;\Varid{x}\;{}\<[39]%
-\>[39]{}\mathrel{=}\;\Conid{Y}\;\Conid{Γ}\;\Varid{x}{}\<[E]%
-\\
-\>[B]{}\Varid{⟦}\;\Conid{A}\;\Varid{∷}\;\Conid{Δ}\;{}\<[17]%
-\>[17]{}\Varid{⟧ᵃ}\;\Conid{X}\;\Conid{Y}\;\Varid{σ}\;\Conid{Γ}\;\Varid{x}\;{}\<[39]%
-\>[39]{}\mathrel{=}\;\Varid{⟦}\;\Conid{Δ}\;\Varid{⟧ᵃ}\;\Conid{X}\;\Conid{Y}\;\Varid{σ}\;((\Conid{A}\;\Varid{⟨}\;\Varid{σ}\;\Varid{⟩})\;\Varid{∷}\;\Conid{Γ})\;\Varid{x}{}\<[E]%
-\\[\blanklineskip]%
-\>[B]{}\Varid{⟦\char95 ⟧ᵃˢ}\;\mathbin{:}\;(\Conid{D}\;\mathbin{:}\;\Conid{ArgsD}\;\Conid{Ξ})\;(\Conid{X}\;\mathbin{:}\;\Conid{R.Fam})\;(\Conid{Y}\;\mathbin{:}\;\Conid{Fam}\;\Conid{X})\;{}\<[E]%
-\\
-\>[B]{}\hsindent{7}{}\<[7]%
-\>[7]{}\Varid{→}\;\Conid{TSub}\;\Conid{Ξ}\;\Varid{0}\;\Varid{→}\;(\Conid{Γ}\;\mathbin{:}\;\Conid{Cxt}\;\Varid{0})\;\Varid{→}\;\Conid{R.⟦}\;\Conid{D}\;\Varid{⟧ᵃˢ}\;\Conid{X}\;(\Varid{length}\;\Conid{Γ})\;\Varid{→}\;\Conid{Set}{}\<[E]%
-\\
-\>[B]{}\Varid{⟦}\;\Varid{[]}\;{}\<[17]%
-\>[17]{}\Varid{⟧ᵃˢ}\;\anonymous \;\anonymous \;\anonymous \;\anonymous \;\anonymous \;{}\<[39]%
-\>[39]{}\mathrel{=}\;\Varid{⊤}{}\<[E]%
-\\
-\>[B]{}\Varid{⟦}\;(\Conid{Δ}\;\Varid{⊢}\;\Conid{A})\;\Varid{∷}\;\Conid{Ds}\;{}\<[17]%
-\>[17]{}\Varid{⟧ᵃˢ}\;\Conid{X}\;\Conid{Y}\;\Varid{σ}\;\Conid{Γ}\;(\Varid{x}\;\Varid{,}\;\Varid{xs})\;{}\<[39]%
-\>[39]{}\mathrel{=}\;{}\<[E]%
-\\
-\>[B]{}\hsindent{3}{}\<[3]%
-\>[3]{}\Varid{⟦}\;\Conid{Δ}\;\Varid{⟧ᵃ}\;\Conid{X}\;(\Varid{λ}\;\iden{\Conid{Γ}^\prime}\;\iden{\Varid{x}^\prime}\;\Varid{→}\;\Conid{Y}\;\iden{\Conid{Γ}^\prime}\;\iden{\Varid{x}^\prime}\;(\Conid{A}\;\Varid{⟨}\;\Varid{σ}\;\Varid{⟩}))\;\Varid{σ}\;\Conid{Γ}\;\Varid{x}\;\Varid{×}\;\Varid{⟦}\;\Conid{Ds}\;\Varid{⟧ᵃˢ}\;\Conid{X}\;\Conid{Y}\;\Varid{σ}\;\Conid{Γ}\;\Varid{xs}{}\<[E]%
-\\[\blanklineskip]%
-\>[B]{}\Varid{⟦\char95 ⟧ᶜ}\;\mathbin{:}\;(\Conid{D}\;\mathbin{:}\;\Conid{ConD})\;(\Conid{X}\;\mathbin{:}\;\Conid{R.Fam})\;(\Conid{Y}\;\mathbin{:}\;\Conid{Fam}\;\Conid{X})\;\Varid{→}\;\Conid{Fam}\;\iden{\Varid{ℓ}^\prime}\;(\Conid{R.⟦}\;\Conid{D}\;\Varid{⟧ᶜ}\;\Conid{X}){}\<[E]%
-\\
-\>[B]{}\Varid{⟦}\;\Varid{ι}\;\{\mskip1.5mu \Conid{Ξ}\mskip1.5mu\}\;\Conid{B}\;\Conid{D}\;{}\<[17]%
-\>[17]{}\Varid{⟧ᶜ}\;\Conid{X}\;\Conid{Y}\;\Conid{Γ}\;\Varid{xs}\;\Conid{A}\;{}\<[39]%
-\>[39]{}\mathrel{=}\;{}\<[E]%
-\\
-\>[B]{}\hsindent{3}{}\<[3]%
-\>[3]{}\Conid{Σ[}\;\Varid{σ}\;\Varid{∈}\;\Conid{TSub}\;\Conid{Ξ}\;\Varid{0}\;\mskip1.5mu]\;\Conid{B}\;\Varid{⟨}\;\Varid{σ}\;\Varid{⟩}\;\Varid{≡}\;\Conid{A}\;\Varid{×}\;\Varid{⟦}\;\Conid{D}\;\Varid{⟧ᵃˢ}\;\Conid{X}\;\Conid{Y}\;\Varid{σ}\;\Conid{Γ}\;\Varid{xs}{}\<[E]%
-\\[\blanklineskip]%
-\>[B]{}\Varid{⟦\char95 ⟧}\;\mathbin{:}\;(\Conid{D}\;\mathbin{:}\;\Conid{Desc})\;(\Conid{X}\;\mathbin{:}\;\Conid{R.Fam})\;(\Conid{Y}\;\mathbin{:}\;\Conid{Fam}\;\Conid{X})\;\Varid{→}\;\Conid{Fam}\;(\Conid{R.⟦}\;\Conid{D}\;\Varid{⟧}\;\Conid{X}){}\<[E]%
-\\
-\>[B]{}\Varid{⟦}\;\Conid{D}\;{}\<[17]%
-\>[17]{}\Varid{⟧}\;\Conid{X}\;\Conid{Y}\;\Conid{Γ}\;(\Varid{i}\;\Varid{,}\;\Varid{xs})\;\Conid{A}\;{}\<[39]%
-\>[39]{}\mathrel{=}\;\Varid{⟦}\;\Conid{D}\;\Varid{.rules}\;\Varid{i}\;\Varid{⟧ᶜ}\;\Conid{X}\;\Conid{Y}\;\Conid{Γ}\;\Varid{xs}\;\Conid{A}{}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-\end{figure}
-\begin{figure}[H]
-  \small
-\begin{minipage}[t]{.47\textwidth}
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{8}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{10}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\Keyword{data}\;\Varid{\char95 ⊢\char95 ⦂\char95 }\;\mathbin{:}\;\Conid{Fam₀}\;\Conid{Raw}\;\Keyword{where}{}\<[E]%
-\\[\blanklineskip]%
-\>[B]{}\hsindent{3}{}\<[3]%
-\>[3]{}\Varid{var}\;{}\<[8]%
-\>[8]{}\mathbin{:}\;(\Varid{i}\;\mathbin{:}\;\Conid{A}\;\Varid{∈}\;\Conid{Γ})\;{}\<[E]%
-\\
-\>[8]{}\Varid{→}\;\{\mskip1.5mu \Varid{j}\;\mathbin{:}\;\Conid{Fin}\;(\Varid{length}\;\Conid{Γ})\mskip1.5mu\}\;{}\<[E]%
-\\
-\>[8]{}\Varid{→}\;\Conid{L.index}\;\Varid{i}\;\Varid{≡}\;\Varid{j}\;{}\<[E]%
-\\
-\>[8]{}\Varid{→}\;\mbox{\onelinecomment }{}\<[E]%
-\\
-\>[8]{}\hsindent{2}{}\<[10]%
-\>[10]{}\Conid{Γ}\;\Varid{⊢}\;(\text{\textasciigrave}\;\Varid{j})\;\Varid{⦂}\;\Conid{A}{}\<[E]%
 \ColumnHook
@@
 \column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
 \column{8}{@{}>{\hspre}l<{\hspost}@{}}%
 \column{10}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{14}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{16}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{19}{@{}>{\hspre}l<{\hspost}@{}}%
 \column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[3]{}\Varid{\char95 ∋\char95 }\;{}\<[8]%
-\>[8]{}\mathbin{:}\;(\Conid{A}\;\mathbin{:}\;\Conid{Ty})\;{}\<[E]%
+\>[B]{}\Varid{⟦\char95 ⟧ᶜ}\;{}\<[8]%
+\>[8]{}\mathbin{:}\;\Conid{ConD}\;{}\<[16]%
+\>[16]{}\Varid{→}\;\Conid{Fam}\;\Varid{→}\;\Conid{Fam}{}\<[E]%
 \\
-\>[8]{}\Varid{→}\;\Conid{Γ}\;\Varid{⊢}\;\Varid{r}\;\Varid{⦂}\;\Conid{A}\;{}\<[E]%
+\>[B]{}\Varid{⟦}\;\Varid{ι}\;\Conid{D}\;\anonymous \;{}\<[10]%
+\>[10]{}\Varid{⟧ᶜ}\;{}\<[14]%
+\>[14]{}\Conid{X}\;{}\<[19]%
+\>[19]{}\mathrel{=}\;\Varid{⟦}\;\Conid{D}\;\Varid{⟧ᵃˢ}\;\Conid{X}{}\<[E]%
+\\[\blanklineskip]%
+\>[B]{}\Varid{⟦\char95 ⟧}\;{}\<[8]%
+\>[8]{}\mathbin{:}\;\Conid{Desc}\;{}\<[16]%
+\>[16]{}\Varid{→}\;\Conid{Fam}\;\Varid{→}\;\Conid{Fam}{}\<[E]%
 \\
-\>[8]{}\Varid{→}\;\mbox{\onelinecomment }{}\<[E]%
-\\
-\>[8]{}\hsindent{2}{}\<[10]%
-\>[10]{}\Conid{Γ}\;\Varid{⊢}\;(\Conid{A}\;\Varid{∋}\;\Varid{r})\;\Varid{⦂}\;\Conid{A}{}\<[E]%
-\\
-\>[3]{}\Varid{op}\;{}\<[8]%
-\>[8]{}\mathbin{:}\;\Varid{⟦}\;\Conid{D}\;\Varid{⟧}\;\Conid{Raw}\;\Varid{\char95 ⊢\char95 ⦂\char95 }\;\Conid{Γ}\;\Varid{rs}\;\Conid{A}\;{}\<[E]%
-\\
-\>[8]{}\Varid{→}\;\mbox{\onelinecomment }{}\<[E]%
-\\
-\>[8]{}\hsindent{2}{}\<[10]%
-\>[10]{}\Conid{Γ}\;\Varid{⊢}\;\Varid{op}\;\Varid{rs}\;\Varid{⦂}\;\Conid{A}{}\<[E]%
+\>[B]{}\Varid{⟦}\;\Conid{D}\;{}\<[10]%
+\>[10]{}\Varid{⟧}\;{}\<[14]%
+\>[14]{}\Conid{X}\;\Varid{n}\;{}\<[19]%
+\>[19]{}\mathrel{=}\;\Conid{Σ[}\;\Varid{i}\;\Varid{∈}\;\Conid{D}\;\Varid{.Op}\;\mskip1.5mu]\;\Varid{⟦}\;\Conid{D}\;\Varid{.rules}\;\Varid{i}\;\Varid{⟧ᶜ}\;\Conid{X}\;\Varid{n}{}\<[E]%
 \ColumnHook
@@
 \end{figure}
+The inductive type of raw terms indexed by a list of free variables can now be defined by 
+\begin{figure}[H]
+  \small
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{8}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{23}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{33}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Keyword{data}\;\Conid{Raw}\;\mathbin{:}\;\Conid{ℕ}\;\Varid{→}\;\Conid{Set}\;\Keyword{where}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\text{\textasciigrave}\anonymous \;{}\<[8]%
+\>[8]{}\mathbin{:}\;\Conid{Fin}\;\Varid{n}\;{}\<[33]%
+\>[33]{}\Varid{→}\;\Conid{Raw}\;\Varid{n}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{\char95 ∋\char95 }\;{}\<[8]%
+\>[8]{}\mathbin{:}\;\Conid{Ty}\;{}\<[23]%
+\>[23]{}\Varid{→}\;\Conid{Raw}\;\Varid{n}\;{}\<[33]%
+\>[33]{}\Varid{→}\;\Conid{Raw}\;\Varid{n}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{op}\;{}\<[8]%
+\>[8]{}\mathbin{:}\;\Varid{⟦}\;\Conid{D}\;\Varid{⟧}\;\Conid{Raw}\;\Varid{n}\;{}\<[33]%
+\>[33]{}\Varid{→}\;\Conid{Raw}\;\Varid{n}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+\end{figure}
+and mirrors our informal definition (\Cref{fig:raw-terms}) where \ensuremath{\Conid{A}\;\Varid{∋}\;\Varid{t}} corresponds to the raw term $t \annotate A$.
 
+\subsection{Defining Extrinsic Bidirectional Typing Derivations}
 \begin{figure}[H]
@@
 \end{figure}
-
+%
 \begin{figure}[H]
@@
 \column{9}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{11}{@{}>{\hspre}l<{\hspost}@{}}%
 \column{E}{@{}>{\hspre}l<{\hspost}@{}}%
@@
 \>[5]{}\hsindent{4}{}\<[9]%
-\>[9]{}\Varid{→}\;\mbox{\onelinecomment }{}\<[E]%
-\\
-\>[9]{}\hsindent{2}{}\<[11]%
-\>[11]{}\Conid{Γ}\;\Varid{⊢}\;(\text{\textasciigrave}\;\Varid{j})\;\Varid{⇒}\;\Conid{A}{}\<[E]%
+\>[9]{}\Varid{→}\;\Conid{Γ}\;\Varid{⊢}\;(\text{\textasciigrave}\;\Varid{j})\;\Varid{⇒}\;\Conid{A}{}\<[E]%
 \\[\blanklineskip]%
@@
 \>[5]{}\hsindent{4}{}\<[9]%
-\>[9]{}\Varid{→}\;\mbox{\onelinecomment }{}\<[E]%
-\\
-\>[9]{}\hsindent{2}{}\<[11]%
-\>[11]{}\Conid{Γ}\;\Varid{⊢}\;\Varid{r}\;\Varid{⇐}\;\Conid{A}{}\<[E]%
+\>[9]{}\Varid{→}\;\Conid{Γ}\;\Varid{⊢}\;\Varid{r}\;\Varid{⇐}\;\Conid{A}{}\<[E]%
 \\[\blanklineskip]%
@@
 \\
-\>[9]{}\Varid{→}\;\mbox{\onelinecomment }{}\<[E]%
-\\
-\>[9]{}\hsindent{2}{}\<[11]%
-\>[11]{}\Conid{Γ}\;\Varid{⊢}\;\Varid{op}\;\Varid{rs}\;[\mskip1.5mu \;\Varid{d}\;\mskip1.5mu]\;\Conid{A}{}\<[E]%
+\>[9]{}\Varid{→}\;\Conid{Γ}\;\Varid{⊢}\;\Varid{op}\;\Varid{rs}\;[\mskip1.5mu \;\Varid{d}\;\mskip1.5mu]\;\Conid{A}{}\<[E]%
 \ColumnHook
@@
 
-\subsection{Examples in \Agda} \label{subsec:formal-examples}
-
 \end{document}

### 2023-07-11 tex/example.tex
[ tex ] Revisions here and there

@@
 \section{Examples}\label{sec:example}
-Since our theory is not confined to a specific simply-typed language, we can employ it with any language that can be specified by a bidirectional binding signature.
+Our theory works for any simply typed language that can be specified in a syntax-directed manner by a bidirectional binding signature, and is not confined to a specific language.
 To showcase its general applicability, we present two additional examples:
 \begin{enumerate}
-\item \PCF with products and let bindings
-\item spine calculus: a simply-typed $\lambda$-calculus with applications in spine form.
+\item \PCF with products and let bindings, and
+\item spine calculus, a simply typed $\lambda$-calculus with applications in spine form.
 \end{enumerate}
@@
 \subsection{PCF with Products and Let Bindings}\label{subsec:PCF}
-Although we have developed a theory for presenting a bidirectional type system and a condition for the decidability of bidirectional type synthesis, designing a bidirectional type system, or its signature, remains a creative activity.\footnote{This also applies to parsing: while there are grammar and parser generators, one still has to design a grammar to parse or disambiguate an ambiguous grammar.}
-While the focus of this paper is not to design bidirectional type systems, we adopt the example of a bidirectional \PCF from \citet{Wadler2022} which extends bidirectional simply typed $\lambda$-calculus in \Cref{sec:key-ideas} by adding following rules:
+Although we have developed a theory for presenting a bidirectional type system and a condition for the decidability of bidirectional type synthesis, designing a bidirectional type system, or its signature, remains a creative activity.\footnote{This has also been the case for parsing: while there are grammar and parser generators, one still has to design a grammar to parse or disambiguate an ambiguous grammar.}
+While the focus of this paper is not to design bidirectional type systems, we adopt the example of a bidirectional \PCF from \citet{Wadler2022}, which extends bidirectional simply typed $\lambda$-calculus in \Cref{sec:key-ideas} by adding the following rules:
 \bgroup
@@
 
-In this language, there are additional type constructs other than $\bto$, so first we extend $\Sigma_{\bto}$ (\Cref{ex:type-signature-for-function-type}) with operations
+%In this language, there are additional type constructs other than $\bto$, so first we extend $\Sigma_{\bto}$ (\Cref{ex:type-signature-for-function-type}) with operations
 In this language, we introduce additional type constructs apart from $\bto$.
@@
 With these operations, a straightforward check shows that every rule is mode-correct, therefore this as a mode-correct bidirectional type system.
-As a result, we can apply the type synthesiser presented in \Cref{sec:type-synthesis} directly to this system, eliminating the need for re-implementation.
+As a result, we can instantiate the type synthesiser presented in \Cref{sec:type-synthesis} for this system, eliminating the need for re-implementation.
 
 \subsection{Spine Calculus}\label{subsec:spine}
-A spine application, denoted as $t\;u_1\;\ldots\;u_n$, is a variant of application that consists of a head term $t$ and an indefinite number of arguments $u_1\;u_2\;\dots\;u_n$.
+A spine application, denoted as $t\;u_1\;\ldots\;u_n$, is a variant of application that consists of a head term $t$ and an indeterminate number of arguments $u_1\;u_2\;\dots\;u_n$.
 This arrangement allows direct access to the head term, making it practical in various applications.
-For instance, \Agda's core language employs this form of applications, as does its reflected syntax used for metaprogramming.
+For instance, \Agda's core language employs this form of application, as does its reflected syntax used for metaprogramming.
 
-At first glance, accommodating this form of application may seem impossible, given the number of arguments a construct can accept is finite as specified by a signature.
+At first glance, accommodating this form of application may seem impossible, given that the number of arguments a construct can accept has to be fixed in a signature.
 Nonetheless, there is no constraint on the total number of operation symbols a signature can have, allowing us to establish corresponding constructs for each number $n$ of arguments, thereby exhibiting the necessity of having an arbitrary set (with decidable equality) for operation symbols.

### 2023-07-11 tex/sec4-pre-synthesis.tex
[ tex ] Revisions here and there

@@
 
-The goal of this sub-section is to construct a mode decorator, which decides for any raw term $V |-_{\Sigma,\erase\Omega} t$ and mode~$\dir d$ whether $V |-_{\Sigma,\Omega} t^\dir{d}$ or not.
+The major goal of this section is to construct a mode decorator, which decides for any raw term $V |-_{\Sigma,\erase\Omega} t$ and mode~$\dir d$ whether $V |-_{\Sigma,\Omega} t^\dir{d}$ or not.
 In fact we will do better:
@@
 Instead, we can provide more information by pointing out the places in the term that require annotations.
-For a bidirectional type system, an annotation is required wherever a term is `strictly' (which we will explain shortly) in checking mode but required to be in synthesising mode, in which case there is no rule for switching from checking to synthesising, and thus there is no way to construct a mode derivation.
+For a bidirectional type system, an annotation is required wherever a term is `strictly' (which we will define shortly) in checking mode but required to be in synthesising mode, in which case there is no rule for switching from checking to synthesising, and thus there is no way to construct a mode derivation.
 We can, however, consider \emph{generalised mode derivations} defined in \cref{fig:generalised-mode-derivations} that allow the use of an additional $\SynRule{Missing}$ rule for such switching, so that a derivation can always be constructed.
@@
 \begin{proof}
-Case analysis on $\dir{d}$ and $\dir{d'}$, adding an outermost mode cast to change the given derivation to a different mode if $d \neq d'$.
+Case analysis on $\dir{d}$ and $\dir{d'}$, adding an outermost mode cast to change the given derivation to a different mode if $\dir{d} \neq \dir{d'}$.
 Note that it is permissible to add an outermost mode cast because the antecedent~(\ref{eq:strict-derivation}) requires the given derivation to be strict.
@@
 Defined in \cref{fig:annotation-ordering}, the `annotation ordering' $t' \sqsupseteq t$ means that $t'$~has the same or more annotations than~$t$.
-In a sense, annotatability is a reasonable form of completeness: if a term is typable in $(\Sigma, \erase\Omega)$, it may not be typable in $(\Sigma, \Omega)$ directly due to some missing annotations, but will be if those annotations are added correctly.
+In a sense, annotatability is a reasonable form of completeness: if a term of a simply typed language $(\Sigma, \erase\Omega)$ is typable in the ordinary type system, it may not be directly typable in the bidirectional type system $(\Sigma, \Omega)$ due to some missing annotations, but will be if those annotations are added correctly.
 In our theory, \cref{thm:annotatability} can be easily proved from generalised mode decoration.

### 2023-07-11 tex/sec5-synthesis.tex
[ tex ] Revisions here and there

@@
 This section focuses on defining mode-correctness and deriving bidirectional type synthesis for any mode-correct bidirectional type system $(\Sigma, \Omega)$.
-We start with \Cref{sec:mode-correctness} by defining mode-correctness and showing the uniqueness of synthesized types.
+We start with \Cref{sec:mode-correctness} by defining mode-correctness and showing the uniqueness of synthesised types.
 This uniqueness means that any two synthesised types for the same raw term $t$ under the same context $\Gamma$ have to be equal.
@@
 \begin{enumerate*}
-\item each `input' type variable in a premise must be an `output` variable in `earlier' premises or provided by the conclusion, if the rule is checking;
-\item each `output' type variable in the conclusion should be some `output' variable in premises, if the rule is synthesis.
+\item each `input' type variable in a premise must be an `output' variable in `earlier' premises, or provided by the conclusion if the rule is checking;
+\item each `output' type variable in the conclusion should be some `output' variable in a premise if the rule is synthesising.
 \end{enumerate*}
-Here `input' variables refers to variables in an extension context and in the type of a premise in the checking mode.
-It is important to note that the order of premises in a bidirectional typing rule also matters, since synthesis type variables are instantiated incrementally during type synthesis.
+Here `input' variables refer to variables in an extension context and in the type of a premise in checking mode.
+It is important to note that the order of premises in a bidirectional typing rule also matters, since synthesised type variables are instantiated incrementally during type synthesis.
 
@@
 Now let us define mode-correctness rigorously.
-As we have outlined, the condition of mode-correctness for a synthesis rule is different from that of a checking rule, and the argument order also matters.
+As we have outlined, the condition of mode-correctness for a synthesising rule is different from that of a checking rule, and the argument order also matters.
 Defining the condition directly for a rule, and thus in our setting for an operation, can be somewhat intricate.
@@
   of \emph{(free) variables} of a type $A$ is denoted by $\fv(A)$.
-  The subset $\fv(\Gamma)$ of variables in a context $\Gamma$ is defined by\/ $\fv(\cdot) = \emptyset$ and\/ $\fv(\Gamma, A) = \fv(\Gamma) \cup \fv(A)$.
-  For an argument list $\biargs$, define its set of \emph{synthesis type variables} inductively as 
+  The subset $\fv(\Gamma)$ of variables in an extension context~$\Delta$ is defined by\/ $\fv(\cdot) = \emptyset$ and\/ $\fv(\Delta, A) = \fv(\Delta) \cup \fv(A)$.
+  For an argument list $\biargs$, define its set of \emph{synthesised type variables} inductively by 
   \begin{align}
@@
 \end{definition}
-This subset contains type variables of a synthesis argument and they are exactly those type variables that would be synthesised during type synthesis.
+This subset contains type variables of a synthesising argument and they are exactly those type variables that will be synthesised during type synthesis.
 
@@
 \end{definition}
-This definition encapsulates the idea that every `input' type variable, possibly deriving from a context $\Delta$ or a checking argument $A_n$, must be an `output' variable from $\synvar(\biargvec)$ or, if the rule is checking, belong to the set $S$ of `input' variables in its conclusion.
+This definition encapsulates the idea that every `input' type variable, possibly derived from a context $\Delta$ or a checking argument $A_n$, must be an `output' variable from $\synvar(\biargvec)$ or, if the rule is checking, belong to the set $S$ of `input' variables in its conclusion.
 This condition must be met for every tail of the argument list as well to ensure that `output' variables accessible at each argument position are from preceding arguments only, hence an inductive definition.
@@
 For a checking operation, an `input' variable of an argument could be derived from $A_0$, as these are known during type checking as an input.
-Since every inhabitant of $\Xi$ can be located in either $A_0$ or synthesis variables, we can determine a concrete type for each inhabitant of $\Xi$ during type synthesis.
-On the other hand, for a synthesis operation, we do not have any known variables at the onset of type synthesis, so the argument list should be mode-correct with respect to $\emptyset$.
-Also, the set of synthesis variables alone should include every type variable in $\Xi$ and particularly in $A_n$.
+Since every inhabitant of $\Xi$ can be located in either $A_0$ or synthesised variables, we can determine a concrete type for each inhabitant of $\Xi$ during type synthesis.
+On the other hand, for a synthesising operation, we do not have any known variables at the onset of type synthesis, so the argument list should be mode-correct with respect to $\emptyset$.
+Also, the set of synthesised variables alone should include every type variable in $\Xi$ and particularly in $A_n$.
 It is also easy to check the bidirectional type system $(\Sigma_{\bto}, \Omega^{\Leftrightarrow}_{\Lambda}$) in \Cref{ex:signature-simply-typed-lambda} for simply typed $\lambda$-calculus is mode-correct according to our definition.
@@
 \begin{remark}
-  Mode-correctness is fundamentally a condition for typing \emph{rules} themselves, not for derivations.
+  Mode-correctness is fundamentally a condition for bidirectional typing \emph{rules} themselves, not for derivations.
   Thus, this property cannot be formalised without treating rules as some mathematical object, such as the notion of bidirectional binding signature presented in \Cref{sec:defs}.
@@
 
-Now, we set out to show the uniqueness of the synthesised types for a mode-correct bidirectional type system.
+Now, we set out to show the uniqueness of synthesised types for a mode-correct bidirectional type system.
 For a specific system, its proof is typically a straightforward induction.
 However, given the mode-correctness---which incorporates an inductive definition on the argument list---we establish our proof through induction, both on typing derivations and on the argument list:
-\begin{lemma}[Uniqueness of the Synthesised Types]\label{thm:unique-syn}
+\begin{lemma}[Uniqueness of Synthesised Types]\label{thm:unique-syn}
   In a mode-correct bidirectional type system $(\Sigma, \Omega)$, the synthesised types of any two derivations
@@
     \item $\Rule{Op}$: Recall that a derivation of\/ $\Gamma |- \tmOpts \syn A$ contains a substitution $\rho$ from the local context $\Xi$ to concrete types.
-      To prove that any two typing derivations has the same synthesised type, it suffices to show that those substitutions $\rho_1$ and $\rho_2$ of $d_1$ and $d_2$, respectively, agree on variables in $\synvar(\biargs)$ so that $\simsub{A_0}{\rho_1} = \simsub{A_0}{\rho_2}$.
+      To prove that any two typing derivations has the same synthesised type, it suffices to show that those substitutions $\rho_1$~and~$\rho_2$ of $d_1$~and~$d_2$, respectively, agree on variables in $\synvar(\biargs)$ so that $\simsub{A_0}{\rho_1} = \simsub{A_0}{\rho_2}$.
       We prove it by induction on the argument list:
@@
         \item For the empty list, the statement is vacuously true by \Cref{eq:synvar1}.
-        \item If $\dir{d_{i+1}}$ is checking, then the statement holds for \Cref{eq:synvar2} by induction hypothesis.
-        \item If $\dir{d_{i+1}}$ is synthesis, then $\simsub{\Delta_{i+1}}{\rho_1} = \simsub{\Delta_{i+1}}{\rho_2}$ by mode-correctness \Cref{eq:MC2} and induction hypothesis (of the list).
-          Therefore, under the same context $\Gamma, \simsub{\Delta_{i+1}}{\rho_1} = \Gamma, \simsub{\Delta_{i+1}}{\rho_2}$ the term $t_{i+1}$ must have the same synthesised type $\simsub{A_{i+1}}{\rho_1} = \simsub{A_{i+1}}{\rho_1}$ by induction hypothesis (of the typing derivation), so $\rho_1$ and $\rho_2$ agree on $\fv(A_{i+1})$ in addition to $\synvar(\biargs)$, as required for the case \Cref{eq:synvar3} in the definition of $\synvar$.
+        \item If $\dir{d_{i+1}}$ is~$\chk$, then the statement holds for \Cref{eq:synvar2} by induction hypothesis.
+        \item If $\dir{d_{i+1}}$ is~$\syn$, then $\simsub{\Delta_{i+1}}{\rho_1} = \simsub{\Delta_{i+1}}{\rho_2}$ by mode-correctness \Cref{eq:MC2} and induction hypothesis (of the list).
+          Therefore, under the same context $\Gamma, \simsub{\Delta_{i+1}}{\rho_1} = \Gamma, \simsub{\Delta_{i+1}}{\rho_2}$ the term $t_{i+1}$ must have the same synthesised type $\simsub{A_{i+1}}{\rho_1} = \simsub{A_{i+1}}{\rho_1}$ by induction hypothesis (of the typing derivation), so $\rho_1$ and $\rho_2$ agree on $\fv(A_{i+1})$ in addition to $\synvar(\biargs)$, as required for \Cref{eq:synvar3} in the definition of $\synvar$.
       \end{enumerate}
+    \vspace{-\topsep-\baselineskip}
   \end{itemize}
@@
 
-We are here for the main technical contribution of this paper.
+We have arrived at the main technical contribution of this paper.
 
@@
 The goal of type synthesis is exactly to define such a substitution $\rho$, so we have to start with an `accumulating' substitution: a substitution $\rho_0$ that is partially defined on $\fv(A_0)$ if $d$ is $\chk$ or otherwise nowhere.
-By mode-correctness, the accumulating substitution~$\rho_i$ will be defined on enough synthesis variables so that type synthesis or checking can be performed on $t_{i}$ with the context $\Gamma, \vars{x}_{i} : \simsub{\Delta_{i}}{\rho_{i}}$ based on its mode derivation $\erase{\Gamma}, \vars{x}_i |-_{\Sigma, \Omega} t_i^{\dir{d_i}}$.
-If we visit a synthesis argument $\synbiarg[i+1]$, then we may extend the domain of $\rho_i$ to include the synthesis variables $\fv(A_{i + 1})$ if type synthesis is successful and also that the synthesised type can be \emph{unified with $A_{i+ 1}$} and thereby \emph{extend} $\rho_i$ to $\bar{\rho_i} = \rho_{i+1}$ with the unifier.
+By mode-correctness, the accumulating substitution~$\rho_i$ will be defined on enough synthesised variables so that type synthesis or checking can be performed on $t_{i}$ with the context $\Gamma, \vars{x}_{i} : \simsub{\Delta_{i}}{\rho_{i}}$ based on its mode derivation $\erase{\Gamma}, \vars{x}_i |-_{\Sigma, \Omega} t_i^{\dir{d_i}}$.
+If we visit a synthesising argument $\synbiarg[i+1]$, then we may extend the domain of $\rho_i$ to include the synthesised variables $\fv(A_{i + 1})$ if type synthesis is successful and also that the synthesised type can be \emph{unified with $A_{i+ 1}$} and thereby \emph{extend} $\rho_i$ to $\bar{\rho_i} = \rho_{i+1}$ with the unifier.
 Then, if we go through every $t_i$ successfully, we will have a total substitution $\rho_n$ by mode-correctness and a derivation of $\Gamma, \vars{x}_i : \Delta_i |-_{\Sigma, \Omega} t_i :^{\dir{d_i}} \simsub{A}{\rho_n}$ for each subterm $t_i$.
@@
   The case \ChkRule{Sub} invokes the uniqueness of synthesised types to refute the case that $\Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn B$ but $A \neq B$ for a given type $A$.
-  The first three cases follow essentially the same argument provided by \citet{Wadler2022}, but we still give the arguments here for the sake of completeness.
+  The first three cases follow essentially the same reasoning provided by \citet{Wadler2022}, but we still present the reasoning here for the sake of completeness.
   The last case \Rule{Op} is new and has been discussed above.
@@
 This could accommodate ill-typed cases and reformulate contradiction proofs positively to deliver more informative error messages.
-This would assist programmers in resolving issues with ill-typed terms, rather than simply returning a blatant 'no'.
+This would assist programmers in resolving issues with ill-typed terms, rather than simply returning a blatant `no'.
 
 \subsection{Trichotomy on Raw Terms by Type Synthesis} \label{subsec:trichotomy}
+
+Combining the bidirectional synthesiser in \cref{subsec:bidirectional-synthesis-checking} with the mode decorator, soundness, and completeness in \cref{sec:pre-synthesis}, we can now construct a type synthesiser, which generalises \cref{thm:implementation}.
 
@@
   \begin{enumerate}
-    \item $\erase{\Gamma} |-_{\Sigma, \Omega} \isTerm{t}^{\syn}$ and\/ $\Gamma |-_{\Sigma, \Omega} \isTerm{t} : A$ for some type $A$.
-    \item $\erase{\Gamma} |-_{\Sigma, \Omega} \isTerm{t}^{\syn}$ but\/ $\Gamma |/-_{\Sigma, \Omega} \isTerm{t} : A$.
+    \item $\erase{\Gamma} |-_{\Sigma, \Omega} \isTerm{t}^{\syn}$ and\/ $\Gamma |-_{\Sigma, \Omega} \isTerm{t} : A$ for some type~$A$.
+    \item $\erase{\Gamma} |-_{\Sigma, \Omega} \isTerm{t}^{\syn}$ but\/ $\Gamma |/-_{\Sigma, \Omega} \isTerm{t} : A$ for any type~$A$.
     \item $\erase{\Gamma} |/-_{\Sigma, \Omega} \isTerm{t}^{\syn}$.
@@
 \begin{proof}
-  Combine \Cref{lem:soundness-completeness,thm:mode-decoration} with \Cref{thm:bidirectional-type-synthesis-checking}.
-
+  Combine \cref{thm:bidirectional-type-synthesis-checking} with \cref{lem:soundness-completeness,thm:mode-decoration}.
 \end{proof}

### 2023-07-12 tex/example.tex
[ tex ] Fix some typos and minor changes

@@
 \section{Examples}\label{sec:example}
-Our theory works for any simply typed language that can be specified in a syntax-directed manner by a bidirectional binding signature, and is not confined to a specific language.
+Our theory works for any language that can be specified by a bidirectional binding signature.
 To showcase its general applicability, we present two additional examples:
@@
 \subsection{PCF with Products and Let Bindings}\label{subsec:PCF}
-Although we have developed a theory for presenting a bidirectional type system and a condition for the decidability of bidirectional type synthesis, designing a bidirectional type system, or its signature, remains a creative activity.\footnote{This has also been the case for parsing: while there are grammar and parser generators, one still has to design a grammar to parse or disambiguate an ambiguous grammar.}
+Although we have developed a theory for presenting a bidirectional type system and a condition for the decidability of bidirectional type synthesis, designing a bidirectional type system, or its signature, remains a creative activity.\footnote{This has also been the case for parsing: while there are grammars and parser generators, one still has to design a grammar to parse or disambiguate an ambiguous grammar.}
 While the focus of this paper is not to design bidirectional type systems, we adopt the example of a bidirectional \PCF from \citet{Wadler2022}, which extends bidirectional simply typed $\lambda$-calculus in \Cref{sec:key-ideas} by adding the following rules:
@@
 \end{flalign*}
-With these operations, a straightforward check shows that every rule is mode-correct, therefore this as a mode-correct bidirectional type system.
+With these operations, a straightforward check shows that every rule is mode-correct, therefore this is a mode-correct bidirectional type system.
 As a result, we can instantiate the type synthesiser presented in \Cref{sec:type-synthesis} for this system, eliminating the need for re-implementation.

### 2023-07-12 tex/sec1-intro.tex
[ tex ] Fix some typos and minor changes

@@
 
-The basic idea of bidirectional type synthesis is that while the problem of type inference is not decidable in general, for certain kinds of terms it is still possible to infer their types (for example, the type of a variable can be looked up in the context); for other kinds of terms, we can switch to the simpler problem of type checking, where the expected type of a raw term is also given so that there is more information to work with.
+The basic idea of bidirectional type synthesis is that while the problem of type inference is not decidable in general, for certain kinds of terms it is still possible to infer their types (for example, the type of a variable can be looked up in the context); for other kinds of terms, we can switch to the simpler problem of type checking, where the expected type of a term is also given so that there is more information to work with.
 More formally, every judgement in a bidirectional type system is extended with a \emph{mode:}
@@
 The former indicates that the type~$A$ is computed as output, using both the context~$\Gamma$ and the term~$t$ as input, while for the latter, all three of $\Gamma$, $t$, and~$A$ are given as input.
-The algorithm of a bidirectional type synthesiser can usually be `read off' from a well designed bidirectional type system: as the synthesiser traverses a raw term, it switches between synthesis and checking, following the modes assigned to the judgements in the typing rules.
+The algorithm of a bidirectional type synthesiser can usually be `read off' from a well-designed bidirectional type system: as the synthesiser traverses a raw term, it switches between synthesis and checking, following the modes assigned to the judgements in the typing rules.
 %if a condition called \emph{mode-correctness} is satisfied.
 %every synthesised type in a typing rule is determined by previously synthesised types from its premises and its input if checking.
-As \citet{Pierce2000} noted, bidirectional type synthesis propagates type information locally within adjacent nodes of a raw term, and does not require unification as Damas--Milner type inference does.
+As \citet{Pierce2000} noted, bidirectional type synthesis propagates type information locally within adjacent nodes of a term, and does not require unification as Damas--Milner type inference does.
 Moreover, introducing long-distance unification constraints would even undermine the essence of locality in bidirectional type synthesis.
@@
 
-Despite sharing the same basic idea, bidirectional typing has only been developed on a case-by-case basis without a theory; this situation is in stark contrast to parsing, which has a comprehensive theory together with widely applicable techniques and practical tools, notably parser generators.%
+Despite sharing the same basic idea, bidirectional typing has only been developed on a case-by-case basis without a theory; this situation is in stark contrast to parsing, which has general theories together with widely applicable techniques and practical tools, notably parser generators.%
 \footnote{The same could even be said for type checking in general, not just for bidirectional type systems.
@@
 Moreover, unlike the plethora of available parser generators, `type-synthesiser generators' rarely exist, so each type synthesiser has to be independently built from scratch.
-To remedy the situation, we initiate a theory of bidirectional typing, which amounts to a verified type-synthesiser generator for syntax-directed bidirectional simple type systems.
+To remedy the situation, we initiate a theory of bidirectional typing, which amounts to a verified type-synthesiser generator for syntax-directed simple type systems.
 
@@
 %(although they do not emphasise the difference of their approach from the traditional one)
-Recall that the law of excluded middle $P + \neg P$ does not hold as an axiom for every $P$ in Martin-L\"of type theory and we say $P$ is \emph{decidable} if it holds for $P$.
+Recall that the law of excluded middle $P + \neg P$ does not hold as an axiom for every $P$ in Martin-L\"of type theory, and we say $P$~is \emph{decidable} if the law holds for~$P$.
 For example, a proof of \emph{decidable equality}, i.e.
-$\forall x, y.~(x = y) + (x \neq y)$, decides whether they are equal and gives an identity proof explicitly if they are, which may or may not be possible depending on the domain of $x$~and~$y$.
+$\forall x, y.~(x = y) + (x \neq y)$, decides whether $x$~and~$y$ are equal and accordingly gives an identity proof or a refutation explicitly; such a decidability proof may or may not be possible depending on the domain of $x$~and~$y$, and is non-trivial.
 Given that all proofs as programs terminate, logical decidability implies \emph{algorithmic decidability}.
@@
 The vision of formalising the metatheory of every programming language was initiated by the \PoplMark challenge~\citep{Aydemir2005}.
-Earlier than \PoplMark, \citet{Altenkirch1993} commented that basic meta-operations and their meta-properties constitute the bulk of formalisation, motivating a number of frameworks~\citep{Ahrens2018,Fiore2022,Gheri2020,Ahrens2022,Allais2021} to define well-scoped/typed terms, substitution, term traversal, and their meta-properties universally.
+Earlier than \PoplMark, \citet{Altenkirch1993} commented that basic meta-operations and their meta-properties constitute the bulk of formalisation, motivating a number of frameworks~\citep{Ahrens2018,Fiore2022,Gheri2020,Ahrens2022,Allais2021} to define well-scoped/typed terms, substitution, term traversal, and their meta-properties universally (in the sense of universal algebra).
 One of the core gadgets of these frameworks is the notion of \emph{binding signatures} (coined by~\citet{Aczel1978} in line with the term \emph{signature} in universal algebra) for specifying type systems.
@@
 \end{enumerate}
-Our theory was first developed with \Agda and then translated to the mathematical vernacular.
+Our theory was first developed in \Agda and then translated to the mathematical vernacular.
 
@@
 Some examples other than simply typed $\lambda$-calculus are given in \Cref{sec:example}.
-We sketch the design of our formalisation in \cref{sec:formalisation} briefly.
-We conclude in \cref{sec:future}, where we reflect on techniques that may improve the formal development and potential challenges in extending our theory to a more general setting.
+We briefly sketch the design of our formalisation in \cref{sec:formalisation} and conclude in \cref{sec:future} with potential challenges in extending our theory to a more general setting.
 

### 2023-07-12 tex/sec3-defs.tex
[ tex ] Revisions here and there
[ tex ] Fix some typos and minor changes

@@
 of simple types, simply typed languages, and bidirectional type systems, using simply typed $\lambda$-calculus in \cref{sec:key-ideas} as our running example.
+(These definitions may look dense, especially on first reading.
+The reader may choose to skim through this section, in particular the figures, and still get some rough ideas from later sections.)
 
-Our definitions are formulated in two steps:
+The definitions are formulated in two steps:
 \begin{enumerate*}
   \item first we introduce a notion of arity and a notion of signature which includes a set of operation symbols and an assignment of arities to symbols;
-\item then, given a signature, we define raw terms and typing derivations by primitive rules such as $\Rule{Var}$ and a rule schema for constructs $\tmOp_o$ indexed by an operational symbol~$o$.
+\item then, given a signature, we define raw terms and typing derivations by primitive rules such as $\Rule{Var}$ and a rule schema for constructs $\tmOp_o$ indexed by an operation symbol~$o$.
 \end{enumerate*}
@@
 
-\begin{definition}
-The \emph{substitution} for a function $\rho\colon \Xi \to \Type_{\Sigma}(\Xi')$, denoted by $\rho\colon \Sub{\Xi}{\Xi'}$, is a map which sends a type $A$ of $\Type_{\Sigma}(\Xi)$ to $\simsub{A}{\rho}$ of $\Type_{\Sigma}(\Xi')$ and is defined as usual.
+\begin{definition} \label{def:substitution}
+The \emph{substitution} for a function $\rho\colon \Xi \to \Type_{\Sigma}(\Xi')$, denoted by $\rho\colon \Sub{\Xi}{\Xi'}$, is a map which sends a type $A : \Type_{\Sigma}(\Xi)$ to $\simsub{A}{\rho} : \Type_{\Sigma}(\Xi')$ and is defined as usual.
 \end{definition}
@@
 
-Next, akin to a signature, a binding signature $\Omega$ consists of a set of operation symbols along with their respective binding arities.
-This also includes the set $\Xi$ of type variables that might be present in the binding arity, each operation is assigned the sort $\Type_{\Sigma}(\Xi)$:
+Next, akin to a signature, a binding signature $\Omega$ consists of a set of operation symbols along with their respective binding arities:
+%Also associated with every operation is the set~$\Xi$ of type variables that might be present in the operation's binding arity, and the sort of the binding arity is $\Type_{\Sigma}(\Xi)$:
 \begin{definition}\label{def:binding-signature}
@@
 That is, each inhabitant $o: O$  is associated with a set $\Xi$ of type variables and a binding arity $\arity(o)$ with the sort $\Type_{\Sigma}(\Xi)$, denoted by $\bop$.
-By a \emph{simply typed language} $(\Sigma, \Omega)$, we mean a pair of a type signature $\Sigma$ and a binding signature $\Omega$.
 
-The inclusion of a set~$\Xi$ of type variables for each operation, called its \emph{local context}, plays a crucial role in type synthesis.
-For a typing rule like $\ChkRule{Abs}$, we will need to substitute \emph{concrete types}, i.e.\ types without any type variables, for variables $A, B$.
-But, we must first identify for which type variables to substitute.
-As such, this information forms part of the arity of an operation, and typing derivations, defined subsequently, will include an instantiation to assign concrete types to variables.
+The set~$\Xi$ of type variables for each operation, called its \emph{local context}, plays a somewhat important role:
+To use a rule like $\Rule{Abs}$ in an actual typing derivation, we will need to substitute \emph{concrete types}, i.e.\ types without any type variables, for variables $A, B$.
+In our formulation of substitution~(\cref{def:substitution}), we must first identify for which type variables to substitute.
+As such, this information forms part of the arity of an operation, and typing derivations, defined subsequently, will include functions from~$\Xi$ to concrete types specifying how to instantiate typing rules by substitution.
 
-For a simply typed language, first we define its raw terms.
+By a \emph{simply typed language} $(\Sigma, \Omega)$, we mean a pair of a type signature~$\Sigma$ and a binding signature~$\Omega$.
+Given a simply typed language, first we define its raw terms.
 
@@
 \begin{definition}
-A \emph{typing context} $\Gamma \colon \Cxt_{\Sigma}$ is formed by $\cdot$ for the empty context and $\Gamma, x : A$ for an additional variable~$x$ associated with a type $A : \Type_{\Sigma}(\emptyset)$.
+A \emph{typing context} $\Gamma \colon \Cxt_{\Sigma}$ is formed by $\cdot$ for the empty context and $\Gamma, x : A$ for an additional variable~$x$ with a concrete type $A : \Type_{\Sigma}(\emptyset)$.
 The list of variables in~$\Gamma$ is denoted~$\erase\Gamma$.
@@
     \item a derivation of $\Gamma |-_{\Sigma, \Omega} (t \annotate A) : A$ if $\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A$ has a derivation, and
-    \item a derivation of $\Gamma |-_{\Sigma, \Omega} \tmOpts : \simsub{A_0}{\rho}$ for $\bop$ if there is a function $\rho\colon \Xi \to \Type_{\Sigma}(\emptyset)$ and a derivation of $\Gamma, \vars{x}_{\isTerm{i}} : \Delta_i |-_{\Sigma, \Omega} \isTerm{t_i} : \simsub{A_i}{\rho}$ for each $i$,
+    \item a derivation of $\Gamma |-_{\Sigma, \Omega} \tmOpts : \simsub{A_0}{\rho}$ for $\bop$ if there is $\rho\colon \Xi \to \Type_{\Sigma}(\emptyset)$ and a derivation of $\Gamma, \vars{x}_{\isTerm{i}} : \simsub{\Delta_i}{\rho} |-_{\Sigma, \Omega} \isTerm{t_i} : \simsub{A_i}{\rho}$ for each $i$,
   \end{enumerate}
@@
   \]
-  For clarity, an arity is denoted by $\biarity$.
+  For clarity, a bidirectional binding arity is denoted by $\biarity$.
 \end{definition}
@@
 A bidirectional type system $(\Sigma, \Omega)$ refers to a pair of a type signature $\Sigma$ and a bidirectional binding signature $\Omega$.
-Every bidirectional binding signature $\Omega$ gives rise to a binding signature $\erase{\Omega}$ if we erase modes from $\Omega$, called the \emph{(mode) erasure} of $\Omega$.
-Hence a bidirectional type system $(\Sigma, \Omega)$ also specifies a simply typed language $(\Sigma, \erase{\Omega})$, including raw terms and typing derivations.
 
@@
           \]
-          has a derivation for $\biop$ in $\Omega$ if there is a function $\rho\colon \Xi \to \Type_{\Sigma}(\emptyset)$ and a derivation of $\Gamma, \vars{x}_{\isTerm{i}} : \Delta_i \vdash_{\Sigma, \Omega} \isTerm{t_i} :^{\dir{d_i}} \simsub{A_i}{\rho}$ for each $i$;
+          has a derivation for $\biop$ in $\Omega$ if there is $\rho\colon \Xi \to \Type_{\Sigma}(\emptyset)$ and a derivation of $\Gamma, \vars{x}_{\isTerm{i}} : \simsub{\Delta_i}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_i} :^{\dir{d_i}} \simsub{A_i}{\rho}$ for each $i$;
     \item the set of \emph{mode derivations} of $V |-_{\Sigma, \Omega} t^\dir{d}$, indexed by a list $V$ of variables, a raw term $\isTerm{t}$ under $V$, and a mode $d$, is defined in \Cref{fig:mode-derivations}.
   \end{itemize}
-  {\small$\boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \syn A}$} and {\small$\boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \chk A}$} are shorthands for ${\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} :^{\syn} A}$ and ${\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} :^{\chk} A}$ respectively.
-  A typing rule is \emph{checking} if its conclusion mode is $\chk$ and \emph{synthesising} if its conclusion mode is $\syn$.
+  The judgements {\small$\boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \syn A}$} and {\small$\boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \chk A}$} are shorthands for ${\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} :^{\syn} A}$ and ${\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} :^{\chk} A}$, respectively.
+  A typing rule is \emph{checking} if its conclusion mode is $\chk$ or \emph{synthesising} otherwise.
 \end{definition}
+
+Every bidirectional binding signature $\Omega$ gives rise to a binding signature $\erase{\Omega}$ if we erase modes from $\Omega$, called the \emph{(mode) erasure} of $\Omega$.
+Hence a bidirectional type system $(\Sigma, \Omega)$ also specifies a simply typed language $(\Sigma, \erase{\Omega})$, including raw terms and typing derivations.
 
@@
 \end{example}
-More importantly, we are able to reason about constructions and properties that hold for any bidirectional simple type system $(\Sigma, \Omega)$ once and for all.
+More importantly, we are able to reason about constructions and properties that hold for any simply typed language with a bidirectional type system once and for all by quantifying over $(\Sigma, \Omega)$.

### 2023-07-12 tex/sec5-synthesis.tex
[ tex ] Fix some typos and minor changes

@@
 \end{enumerate*}
-Here `input' variables refer to variables in an extension context and in the type of a premise in checking mode.
+Here `input' variables refer to variables in an extension context and in a checking premise.
 It is important to note that the order of premises in a bidirectional typing rule also matters, since synthesised type variables are instantiated incrementally during type synthesis.
@@
 Instead, we choose to define the conditions for the argument list---more specifically, triples $\biargvec$ of an extension context $\Delta_i$, a type $A_i$, and a mode $\dir{d_i}$---pertaining to an operation, for an operation, and subsequently for a signature.
-We also need some auxiliary definitions for the subset of variables of a type and of a context and the set of variables that have been synthesised:
+We also need some auxiliary definitions for the subset of variables of a type and of an extension context, and the set of variables that have been synthesised:
 \begin{definition}
@@
   of \emph{(free) variables} of a type $A$ is denoted by $\fv(A)$.
-  The subset $\fv(\Gamma)$ of variables in an extension context~$\Delta$ is defined by\/ $\fv(\cdot) = \emptyset$ and\/ $\fv(\Delta, A) = \fv(\Delta) \cup \fv(A)$.
+  The subset $\fv(\Delta)$ of variables in an extension context~$\Delta$ is defined by\/ $\fv(\cdot) = \emptyset$ and\/ $\fv(\Delta, A) = \fv(\Delta) \cup \fv(A)$.
   For an argument list $\biargs$, define its set of \emph{synthesised type variables} inductively by 
@@
 \begin{definition}\label{def:mode-correctness-args}
-  The \emph{mode-correctness} $\MCas\left(\biargvec\right)$ for an argument list $\biargs$ with respect to a subset $S$ of $\Xi$ is defined as
+  The \emph{mode-correctness} $\MCas\left(\biargvec\right)$ for an argument list $\biargs$ with respect to a subset $S$ of $\Xi$ is defined by
   \begin{align}
@@
     \label{eq:MC3} \MCas\left(\biargvec, \chkbiarg[n]\right)
-                                  & = \fv(\Delta, A_n) \subseteq \left( S \cup \synvar\left(\biargvec\right)\right) \land \MCas\left(\biargvec\right) \\
+                                  & = \fv(\Delta_n, A_n) \subseteq \left( S \cup \synvar\left(\biargvec\right)\right) \land \MCas\left(\biargvec\right) \\
     \label{eq:MC2} \MCas\left(\biargvec, \synbiarg[n]\right) 
-                                  & = \phantom{, A_n} \fv(\Delta) \subseteq \left( S \cup \synvar\left(\biargvec\right)\right) \land  \MCas\left(\biargvec\right)
+                                  & = \phantom{, A_n} \fv(\Delta_n) \subseteq \left( S \cup \synvar\left(\biargvec\right)\right) \land  \MCas\left(\biargvec\right)
   \end{align}
@@
 \end{definition}
-This definition encapsulates the idea that every `input' type variable, possibly derived from a context $\Delta$ or a checking argument $A_n$, must be an `output' variable from $\synvar(\biargvec)$ or, if the rule is checking, belong to the set $S$ of `input' variables in its conclusion.
+This definition encapsulates the idea that every `input' type variable, possibly derived from an extension context~$\Delta_n$ or a checking argument~$A_n$, must be an `output' variable from $\synvar(\biargvec)$ or, if the rule is checking, belong to the set $S$ of `input' variables in its conclusion.
 This condition must be met for every tail of the argument list as well to ensure that `output' variables accessible at each argument position are from preceding arguments only, hence an inductive definition.
@@
   \begin{enumerate}
-    \item either $d$ is $\chk$, its argument list is mode-correct w.r.t.\ $\fv(A_0)$, and the union $\fv(A_0) \cup \synvar(\biargvec)$ contains every inhabitant of $\Xi$;
-    \item or $d$ is $\syn$, its argument list is mode-correct w.r.t.\ $\emptyset$, and the finite subset $\synvar(\biargvec)$ contains every inhabitant of $\Xi$ and, particularly, $\fv(A_0)$.
+    \item either $d$ is $\chk$, its argument list is mode-correct with respect to $\fv(A_0)$, and the union $\fv(A_0) \cup \synvar(\biargvec)$ contains every inhabitant of $\Xi$;
+    \item or $d$ is $\syn$, its argument list is mode-correct with respect to $\emptyset$, and $\synvar(\biargvec)$ contains every inhabitant of $\Xi$ and, particularly, $\fv(A_0)$.
   \end{enumerate}
@@
 \end{definition}
-For a checking operation, an `input' variable of an argument could be derived from $A_0$, as these are known during type checking as an input.
-Since every inhabitant of $\Xi$ can be located in either $A_0$ or synthesised variables, we can determine a concrete type for each inhabitant of $\Xi$ during type synthesis.
-On the other hand, for a synthesising operation, we do not have any known variables at the onset of type synthesis, so the argument list should be mode-correct with respect to $\emptyset$.
-Also, the set of synthesised variables alone should include every type variable in $\Xi$ and particularly in $A_n$.
-It is also easy to check the bidirectional type system $(\Sigma_{\bto}, \Omega^{\Leftrightarrow}_{\Lambda}$) in \Cref{ex:signature-simply-typed-lambda} for simply typed $\lambda$-calculus is mode-correct according to our definition.
+For a checking operation, an `input' variable of an argument could be derived from~$A_0$, as these are known during type checking as an input.
+Since every inhabitant of~$\Xi$ can be located in either~$A_0$ or synthesised variables, we can determine a concrete type for each inhabitant of~$\Xi$ during type synthesis.
+On the other hand, for a synthesising operation, we do not have any known variables at the onset of type synthesis, so the argument list should be mode-correct with respect to~$\emptyset$.
+Also, the set of synthesised variables alone should include every type variable in~$\Xi$ and particularly in~$A_n$.
+
+It is easy to check the bidirectional type system $(\Sigma_{\bto}, \Omega^{\Leftrightarrow}_{\Lambda}$) in \Cref{ex:signature-simply-typed-lambda} for simply typed $\lambda$-calculus is mode-correct according to our definition.
 
@@
 Now, we set out to show the uniqueness of synthesised types for a mode-correct bidirectional type system.
-For a specific system, its proof is typically a straightforward induction.
-However, given the mode-correctness---which incorporates an inductive definition on the argument list---we establish our proof through induction, both on typing derivations and on the argument list:
+For a specific system, its proof is typically a straightforward induction on the typing derivations.
+However, since mode-correctness is inductively defined on the argument list, our proof proceeds by induction on both the typing derivations and the argument list:
 \begin{lemma}[Uniqueness of Synthesised Types]\label{thm:unique-syn}
@@
         \item If $\dir{d_{i+1}}$ is~$\chk$, then the statement holds for \Cref{eq:synvar2} by induction hypothesis.
-        \item If $\dir{d_{i+1}}$ is~$\syn$, then $\simsub{\Delta_{i+1}}{\rho_1} = \simsub{\Delta_{i+1}}{\rho_2}$ by mode-correctness \Cref{eq:MC2} and induction hypothesis (of the list).
+        \item If $\dir{d_{i+1}}$ is~$\syn$, then $\simsub{\Delta_{i+1}}{\rho_1} = \simsub{\Delta_{i+1}}{\rho_2}$ by \Cref{eq:MC2} and induction hypothesis (of the list).
           Therefore, under the same context $\Gamma, \simsub{\Delta_{i+1}}{\rho_1} = \Gamma, \simsub{\Delta_{i+1}}{\rho_2}$ the term $t_{i+1}$ must have the same synthesised type $\simsub{A_{i+1}}{\rho_1} = \simsub{A_{i+1}}{\rho_1}$ by induction hypothesis (of the typing derivation), so $\rho_1$ and $\rho_2$ agree on $\fv(A_{i+1})$ in addition to $\synvar(\biargs)$, as required for \Cref{eq:synvar3} in the definition of $\synvar$.
@@
 If we visit a synthesising argument $\synbiarg[i+1]$, then we may extend the domain of $\rho_i$ to include the synthesised variables $\fv(A_{i + 1})$ if type synthesis is successful and also that the synthesised type can be \emph{unified with $A_{i+ 1}$} and thereby \emph{extend} $\rho_i$ to $\bar{\rho_i} = \rho_{i+1}$ with the unifier.
-Then, if we go through every $t_i$ successfully, we will have a total substitution $\rho_n$ by mode-correctness and a derivation of $\Gamma, \vars{x}_i : \Delta_i |-_{\Sigma, \Omega} t_i :^{\dir{d_i}} \simsub{A}{\rho_n}$ for each subterm $t_i$.
+Then, if we go through every $t_i$ successfully, we will have a total substitution $\rho_n$ by mode-correctness and a derivation of $\Gamma, \vars{x}_i : \Delta_i |-_{\Sigma, \Omega} t_i :^{\dir{d_i}} \simsub{A}{\rho_n}$ for each sub-term $t_i$.
 
@@
 To make the argument above sound, it is necessary to compare types and solve a unification problem.
-Hence, we assume that the set\/ $\Xi$ of variables has a decidable equality, thereby ensuring that the set $\Type_{\Sigma}(\Xi)$ of types also has a decidable equality.\footnote{%
+Hence, we assume that the set~$\Xi$ of type variables has a decidable equality, thereby ensuring that the set $\Type_{\Sigma}(\Xi)$ of types also has a decidable equality.\footnote{%
 To simplify our choice, we could simply confine $\Xi$ to any set within the family of sets $\Fin(n)$ of naturals less than~$n$, given that these sets have a decidable equality and the arity of a type construct is finite.
@@
   The two cases \SynRule{Var} and \SynRule{Anno} are straightforward and have nothing to do with mode-correctness.
-  The case \ChkRule{Sub} invokes the uniqueness of synthesised types to refute the case that $\Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn B$ but $A \neq B$ for a given type $A$.
+  The case \ChkRule{Sub} invokes the uniqueness of synthesised types~(\cref{thm:unique-syn}) to refute the case that $\Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn B$ but $A \neq B$ for a given type $A$.
   The first three cases follow essentially the same reasoning provided by \citet{Wadler2022}, but we still present the reasoning here for the sake of completeness.
@@
 These programs either compute the typing derivation or provide a proof of contradiction.
-Each case analysis simply branches depending on the outcomes of bidirectional type synthesis and checking for each subterm, as well as the unification process.
+Each case analysis simply branches depending on the outcomes of bidirectional type synthesis and checking for each sub-term, as well as the unification process.
 If a contradiction proof is not of interest for implementation, these programs can be simplified by disregarding the cases that yield such contradiction proofs.
@@
 
-Combining the bidirectional synthesiser in \cref{subsec:bidirectional-synthesis-checking} with the mode decorator, soundness, and completeness in \cref{sec:pre-synthesis}, we can now construct a type synthesiser, which generalises \cref{thm:implementation}.
+Combining the bidirectional type synthesiser with the mode decorator, soundness, and completeness from \cref{sec:pre-synthesis}, we get a type synthesiser parameterised by $(\Sigma, \Omega)$, generalising \cref{thm:implementation}.
 
@@
   \begin{enumerate}
-    \item $\erase{\Gamma} |-_{\Sigma, \Omega} \isTerm{t}^{\syn}$ and\/ $\Gamma |-_{\Sigma, \Omega} \isTerm{t} : A$ for some type~$A$.
-    \item $\erase{\Gamma} |-_{\Sigma, \Omega} \isTerm{t}^{\syn}$ but\/ $\Gamma |/-_{\Sigma, \Omega} \isTerm{t} : A$ for any type~$A$.
-    \item $\erase{\Gamma} |/-_{\Sigma, \Omega} \isTerm{t}^{\syn}$.
+    \item (\/$\erase{\Gamma} |-_{\Sigma, \Omega} \isTerm{t}^{\syn}$ and)\/ $\Gamma |-_{\Sigma, \Omega} \isTerm{t} : A$ for some type~$A$.
+    \item (\/$\erase{\Gamma} |-_{\Sigma, \Omega} \isTerm{t}^{\syn}$ but)\/ $\Gamma |/-_{\Sigma, \Omega} \isTerm{t} : A$ for any type~$A$.
+    \item \phantom{(\/$\erase{\Gamma} |-_{\Sigma, \Omega} \isTerm{t}^{\syn}$}\llap{$\erase{\Gamma} |/-_{\Sigma, \Omega} \isTerm{t}^{\syn}$\kern.5pt}.
   \end{enumerate}

### 2023-07-12 tex/sec3-defs.tex
[ tex ] Use "stand for" instead of "are shorthands" (they are not shorter anyway)

@@
   \end{itemize}
-  The judgements {\small$\boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \syn A}$} and {\small$\boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \chk A}$} are shorthands for ${\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} :^{\syn} A}$ and ${\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} :^{\chk} A}$, respectively.
+  The two judgements {\small$\boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \syn A}$} and {\small$\boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \chk A}$} stand for ${\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} :^{\syn} A}$ and ${\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} :^{\chk} A}$, respectively.
   A typing rule is \emph{checking} if its conclusion mode is $\chk$ or \emph{synthesising} otherwise.

### 2023-07-12 tex/BiSig.tex
[ tex ] Clean up code in S6
[ tex ] Add a separate section on Examples
[ tex ] Minor changes
[ tex ] Revisions up to S6

@@
-\documentclass[acmsmall,screen,review,anonymous,draft]{acmart}
+\documentclass[acmsmall,screen,review,anonymous]{acmart}
 
@@
 \usepackage{agda-lhs2tex}
+\newcommand{\cons}[1]{\mathbf{#1}}
+\newcommand{\iden}{\mathit}
 
@@
 
-\begin{document}
-
 \author{Liang-Ting Chen}
@@
 The whole theory works as a verified type-synthesiser generator for syntax-directed simple type systems:
-From a signature that specifies a bidirectionally typed language, the theory produces a type synthesiser that decides whether an input abstract syntax tree has a typing derivation or reports that the input does not have enough type annotations.
+From a signature that specifies a simply typed language with a bidirectional type system, the theory produces a type synthesiser that decides whether an input abstract syntax tree has a typing derivation or reports that the input does not have enough type annotations.
 Within the theory, we formally define soundness, completeness, and mode-correctness, which are sufficient conditions for deriving a correct type synthesiser.
 We also propose a preprocessing step called `mode decoration', which helps the user to deal with missing type annotations and streamlines the theory.
-The entire theory is formulated constructively, and has been formalised in the \Agda proof assistant.
+The entire theory is formulated constructively, and has been formalised in the proof assistant \Agda with \AxiomK.
 
-%\LT{Still not coherent enough}
-%Type checking and inference serve as the transition from parsed abstract syntax trees to well-typed programs.
-%In response to this difficulty, we propose a simple yet general and constructive theory of bidirectional type synthesis.
-%which refers to the mutual use of type checking and synthesis based on bidirectional typing rules to determine the type of a program with type annotations.
-%First, it is simple yet general as it applies to any syntax-directed bidirectional simple type system that can be specified by a signature.
-%which is analogous to a grammar in parsing.
-%Second, it is constructive because it is not only formalised in a proof assistant based on Martin-L\"of type theory, but also compactly formulated in a way that unifies computation and proof---for example, a proof of the decidability of the type synthesis problem is computationally a type synthesiser.
-%Moreover, we introduce `mode decoration' to deal with missing type annotations, an aspect often overlooked in the literature.
-%The main theorem is logically a trichotomy on untyped abstract syntax trees and computationally a type synthesiser that may throw an exception due to missing type annotations.
-%By being parametrised by a signature, our theory works overall as a verified type-synthesiser generator.
 \end{abstract}
+\begin{document}
 
@@
 \input{sec5-synthesis.tex}
+\input{example.tex}
 \subfile{sec6-formalisation}

### 2023-07-12 tex/example.tex
[ tex ] Revisions up to S6

@@
 \end{flalign*}
-With these operations, a straightforward check shows that every rule is mode-correct, therefore this is a mode-correct bidirectional type system.
+With these operations, a straightforward check shows that every rule is mode-correct, therefore this as a mode-correct bidirectional type system.
 As a result, we can instantiate the type synthesiser presented in \Cref{sec:type-synthesis} for this system, eliminating the need for re-implementation.

### 2023-07-12 tex/sec1-intro.tex
[ tex ] Revisions up to S6

@@
 
-Despite sharing the same basic idea, bidirectional typing has only been developed on a case-by-case basis without a theory; this situation is in stark contrast to parsing, which has general theories together with widely applicable techniques and practical tools, notably parser generators.%
+Despite sharing the same basic idea, bidirectional typing has only been developed on a case-by-case basis without a theory; this situation is in stark contrast to parsing, which has a comprehensive theory together with widely applicable techniques and practical tools, notably parser generators.%
 \footnote{The same could even be said for type checking in general, not just for bidirectional type systems.
@@
 Some examples other than simply typed $\lambda$-calculus are given in \Cref{sec:example}.
-We briefly sketch the design of our formalisation in \cref{sec:formalisation} and conclude in \cref{sec:future} with potential challenges in extending our theory to a more general setting.
+We briefly sketch our \Agda formalisation in \cref{sec:formalisation}.
+Finally, in \cref{sec:future} we reflect on techniques that may improve the formal development and potential challenges in extending our theory to a more general setting.
 

### 2023-07-12 tex/sec2-outline.tex
[ tex ] Revisions up to S6

@@
 \label{def:STLC-type-synthesiser}
-Parametrised by an `excuse' predicate~$E$ on raw terms, a \emph{type synthesiser} takes a context~$\Gamma$ and a raw term $\erase\Gamma \vdash t$ (where $\erase\Gamma$~is the domain of~$\Gamma$) as input, and establishes one of the following outcomes:
+Parametrised by an `excuse' predicate~$E$ on raw terms, a \emph{type synthesiser} takes a context~$\Gamma$ and a raw term $\erase\Gamma \vdash t$ (where $\erase\Gamma$~is the list of variables in~$\Gamma$) as input, and establishes one of the following outcomes:
 \begin{enumerate}
@@
 While it is possible for a bidirectional type synthesiser to do its job in one go, which can be thought of as adding both mode and typing information to a raw term and arriving at a bidirectional typing derivation, it is beneficial to have a preprocessing step which adds only mode information, based on which the synthesiser then continues to add typing information.
-More precisely, the preprocessing step attempts to produce \emph{mode derivations} as defined in \cref{fig:STLC-mode-derivations}, where the rules are exactly the mode part of the bidirectional typing rules~(\cref{fig:STLC-bidirectional-typing-derivations}).
+More precisely, the preprocessing step, which we call \emph{mode decoration}, attempts to produce \emph{mode derivations} as defined in \cref{fig:STLC-mode-derivations}, where the rules are exactly the mode part of the bidirectional typing rules~(\cref{fig:STLC-bidirectional-typing-derivations}).
 

### 2023-07-12 tex/sec3-defs.tex
[ tex ] Revisions up to S6

@@
     \item a derivation of $\Gamma |-_{\Sigma, \Omega} (t \annotate A) : A$ if $\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A$ has a derivation, and
-    \item a derivation of $\Gamma |-_{\Sigma, \Omega} \tmOpts : \simsub{A_0}{\rho}$ for $\bop$ if there is $\rho\colon \Xi \to \Type_{\Sigma}(\emptyset)$ and a derivation of $\Gamma, \vars{x}_{\isTerm{i}} : \simsub{\Delta_i}{\rho} |-_{\Sigma, \Omega} \isTerm{t_i} : \simsub{A_i}{\rho}$ for each $i$,
+    \item a derivation of $\Gamma |-_{\Sigma, \Omega} \tmOpts : \simsub{A_0}{\rho}$ for $\bop$ if there is a function $\rho\colon \Xi \to \Type_{\Sigma}(\emptyset)$ and a derivation of $\Gamma, \vars{x}_{\isTerm{i}} : \Delta_i |-_{\Sigma, \Omega} \isTerm{t_i} : \simsub{A_i}{\rho}$ for each $i$,
   \end{enumerate}
@@
           \]
-          has a derivation for $\biop$ in $\Omega$ if there is $\rho\colon \Xi \to \Type_{\Sigma}(\emptyset)$ and a derivation of $\Gamma, \vars{x}_{\isTerm{i}} : \simsub{\Delta_i}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_i} :^{\dir{d_i}} \simsub{A_i}{\rho}$ for each $i$;
+          has a derivation for $\biop$ in $\Omega$ if there is a function $\rho\colon \Xi \to \Type_{\Sigma}(\emptyset)$ and a derivation of $\Gamma, \vars{x}_{\isTerm{i}} : \Delta_i \vdash_{\Sigma, \Omega} \isTerm{t_i} :^{\dir{d_i}} \simsub{A_i}{\rho}$ for each $i$;
     \item the set of \emph{mode derivations} of $V |-_{\Sigma, \Omega} t^\dir{d}$, indexed by a list $V$ of variables, a raw term $\isTerm{t}$ under $V$, and a mode $d$, is defined in \Cref{fig:mode-derivations}.
   \end{itemize}
-  The two judgements {\small$\boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \syn A}$} and {\small$\boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \chk A}$} stand for ${\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} :^{\syn} A}$ and ${\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} :^{\chk} A}$, respectively.
-  A typing rule is \emph{checking} if its conclusion mode is $\chk$ or \emph{synthesising} otherwise.
+  {\small$\boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \syn A}$} and {\small$\boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \chk A}$} are shorthands for ${\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} :^{\syn} A}$ and ${\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} :^{\chk} A}$ respectively.
+  A typing rule is \emph{checking} if its conclusion mode is $\chk$ and \emph{synthesising} if its conclusion mode is $\syn$.
 \end{definition}

### 2023-07-12 tex/sec4-pre-synthesis.tex
[ tex ] Revisions up to S6

@@
 We will construct such a \emph{generalised mode decorator}~(\cref{thm:generalised-mode-decoration}), which can be weakened to an ordinary mode decorator~(\cref{thm:mode-decoration}) if the additional explanation is not needed.%
-\footnote{For brevity, we use ordinary mode decoration elsewhere in this paper.}
+\footnote{To simplify the presentation, we use ordinary mode decoration elsewhere in this paper.}
 
@@
 
-The theorem could be proved directly, but that would mix up two case analyses which respectively inspect the input term and apply mode casts depending on what modes are required.
-Instead, we distill the case analysis that deals with mode casts into the following \cref{thm:adjustment}, whose antecedent~(\ref{eq:strict-derivation}) is then established by induction on the input term in the proof of \cref{thm:generalised-mode-decoration}.
+The theorem could be proved directly, but that would mix up two case analyses which respectively inspect the input term~$\isTerm{t}$ and apply mode casts depending on which mode~$\dir{d}$ is required.
+Instead, we distill the case analysis on~$\dir{d}$ that deals with mode casts into the following \cref{thm:adjustment}, whose antecedent~(\ref{eq:strict-derivation}) is then established by induction on~$\isTerm{t}$ in the proof of \cref{thm:generalised-mode-decoration}.
 
@@
 
-\citet{Dunfield2021} formulated completeness differently from ours (\cref{thm:STLC-completeness,sec:soundness-and-completeness}) and proposed \emph{annotatability} as a more suitable name.
+\citet[Section~3.2]{Dunfield2021} formulated completeness differently from ours (\cref{thm:STLC-completeness,sec:soundness-and-completeness}) and proposed \emph{annotatability} as a more suitable name.
 In our theory, we may formulate annotatability as follows.
@@
 In a sense, annotatability is a reasonable form of completeness: if a term of a simply typed language $(\Sigma, \erase\Omega)$ is typable in the ordinary type system, it may not be directly typable in the bidirectional type system $(\Sigma, \Omega)$ due to some missing annotations, but will be if those annotations are added correctly.
-In our theory, \cref{thm:annotatability} can be easily proved from generalised mode decoration.
+In our theory, \cref{thm:annotatability} can be easily proved with generalised mode decoration.
 

### 2023-07-12 tex/sec5-synthesis.tex
[ tex ] Use "stand for" instead of "are shorthands" (they are not shorter anyway)
[ tex ] Revisions up to S6

@@
 \end{enumerate*}
-Here `input' variables refer to variables in an extension context and in a checking premise.
+Here `input' variables refer to variables in an extension context and in the type of a premise in checking mode.
 It is important to note that the order of premises in a bidirectional typing rule also matters, since synthesised type variables are instantiated incrementally during type synthesis.
@@
   \begin{enumerate}
-    \item either $d$ is $\chk$, its argument list is mode-correct with respect to $\fv(A_0)$, and the union $\fv(A_0) \cup \synvar(\biargvec)$ contains every inhabitant of $\Xi$;
-    \item or $d$ is $\syn$, its argument list is mode-correct with respect to $\emptyset$, and $\synvar(\biargvec)$ contains every inhabitant of $\Xi$ and, particularly, $\fv(A_0)$.
+    \item either $d$ is $\chk$, its argument list is mode-correct w.r.t.\ $\fv(A_0)$, and the union $\fv(A_0) \cup \synvar(\biargvec)$ contains every inhabitant of $\Xi$;
+    \item or $d$ is $\syn$, its argument list is mode-correct w.r.t.\ $\emptyset$, and the finite subset $\synvar(\biargvec)$ contains every inhabitant of $\Xi$ and, particularly, $\fv(A_0)$.
   \end{enumerate}
@@
         \item If $\dir{d_{i+1}}$ is~$\chk$, then the statement holds for \Cref{eq:synvar2} by induction hypothesis.
-        \item If $\dir{d_{i+1}}$ is~$\syn$, then $\simsub{\Delta_{i+1}}{\rho_1} = \simsub{\Delta_{i+1}}{\rho_2}$ by \Cref{eq:MC2} and induction hypothesis (of the list).
+        \item If $\dir{d_{i+1}}$ is~$\syn$, then $\simsub{\Delta_{i+1}}{\rho_1} = \simsub{\Delta_{i+1}}{\rho_2}$ by mode-correctness \Cref{eq:MC2} and induction hypothesis (of the list).
           Therefore, under the same context $\Gamma, \simsub{\Delta_{i+1}}{\rho_1} = \Gamma, \simsub{\Delta_{i+1}}{\rho_2}$ the term $t_{i+1}$ must have the same synthesised type $\simsub{A_{i+1}}{\rho_1} = \simsub{A_{i+1}}{\rho_1}$ by induction hypothesis (of the typing derivation), so $\rho_1$ and $\rho_2$ agree on $\fv(A_{i+1})$ in addition to $\synvar(\biargs)$, as required for \Cref{eq:synvar3} in the definition of $\synvar$.

### 2023-07-12 tex/sec6-formalisation.tex
[ tex ] Done with formalisation (for now ...)

--- 2023-07-09 tex/sec6-formalisation.tex
@@
 \DeclareUnicodeCharacter{2200}{\ensuremath{\forall}}
+\DeclareUnicodeCharacter{2218}{\ensuremath{\circ}}
 
@@
 \section{Formalisation} \label{sec:formalisation}
-\begin{enumerate}
-  \item (Bidirectional) binding signature, functor, and terms (extrinsic typing, raw terms, raw terms in some mode) --- 2.5 pp
-  \item Compare the induction principle and the \Agda proof of Soundness --- 2 pp
-  \item type synthesis and checking -- 0.5 p
-  \item Examples including STLC (PCF), application in spine form --- 1p
+As we have mentioned in \Cref{sec:intro}, our theory was initially developed with \Agda.
+While the translation to the natural language is reasonably straightforward, understanding the formalisation itself could pose some difficulty.
+If the reader is comfortable with the informal presentation and assured by the existence of its formalisation, they may feel free to skip this section.
 
-\end{enumerate}
+\paragraph{Revisiting language formalisation frameworks}
+Unlike prior frameworks~\citep{Allais2021,Fiore2022,Ahrens2022} that have primarily focused on meta-properties centred around substitution for intrinsically-typed terms, our theory of bidirectional type synthesis does not require term substitution but structural induction for extrinsically-typed terms.
+The formal definitions of extrinsic typing are more complex than their intrinsic counterparts.
+For a specific language such as \PCF, \citet{Wadler2022} noted that `extrinsically-typed terms require about 1.6 times as much code as intrinsically-typed' for their formalisation of type safety.
+For our generic definitions, take the formal type-theoretic definition of typing derivations $\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A$ as an example.
+Its intrinsic definition consists of of just one family of sets of intrinsically-typed terms indexed by $A$ and $\Gamma$, but its extrinsic counterpart is a generic family of sets indexed by additionally a generic raw term $t$, involving constructions of two different layers.
+
+\paragraph{Category-theoretic analysis of well-typed terms}
+The difference between intrinsic typing and extrinsic typing could be partly manifested by its categorical analysis.
+\citet{Fiore1999}{'s} theory of abstract syntax and variable binding forms the foundation of \citet{Fiore2022}'s framework and inspires frameworks by \citet{Allais2021,Ahrens2022}.
+Let us take a closer look of their idea.
+The set of (untyped) abstract syntax trees for a language can be understood as
+\begin{enumerate*}
+  \item a family of sets $\Term_{\Gamma}$ of well-scoped terms under a context~$\Gamma$ with
+  \item variable renaming for a function $\sigma\colon \Gamma \to \Delta$ between variables acting as a functorial map from $\Term_{\Gamma}$ to $\Term_{\Delta}$, i.e.\ a presheaf $\Term\colon \mathbb{F} \to \mathsf{Set}$, with
+  \item an \emph{initial} algebra $[\mathsf{v}, \mathsf{op}]$ on~$\Term$ given by the variable rule as a map $\mathsf{v}$ from the presheaf $V\colon \mathbb{F} \hookrightarrow \mathsf{Set}$ of variables to $\Term$ and other constructs as $\mathsf{op}\colon \mathbb{\Sigma}\Term \to \Term$ where $\mathbb{F}$ is the category of contexts and the functor $\mathbb{\Sigma}\colon \mathsf{Set}^\mathbb{F} \to \mathsf{Set}^\mathbb{F}$ encodes the binding arities of constructs.
+  The initiality amounts to structural recursion on terms, namely \emph{term traversal}.
+\end{enumerate*}
+Substitution is modelled as a monoid multiplication with the notion of strength and a suitable monoidal structure on the category of presheaves.
+To put it succinctly, it is the free $\mathbb{\Sigma}$-monoid over the presheaf~$V$ of variables.
+
+Such an initial algebra can be constructed as the least fixpoint of a sequence of objects defined via iteratively applying the signature functor by the Initial Algebra Theorem~\cite{Trnkova1975}.
+
+\paragraph{Type-theoretic construction of well-typed terms}
+Fortunately, in type theory, constructing the initial algebra of well-typed terms boils down to defining an inductive type with a few constructors that align primitive rules (such as $\Rule{Var}$) and a rule schema; term traversal can be defined by pattern matching as usual for the inductive type~\citep{Fiore2022}.
+
+\paragraph{Lack of a theory for extrinsic typing}
+While the theory of \citet{Fiore1999} and others provide significant insights, they do not consider extrinsic typing.
+We find inspiration in the interpretation of structural induction as algebras for an endofunctor on the category of predicates over a base category, as put forward by \citet{Hermida1998}.
+From this perspective, we interpret extrinsic typing as a signature `functor' on the category of predicates consisting of a $\N$-indexed family $X$ of sets and another family $Y$ of sets over $X$; constructing typing derivations as its initial algebra over well-scoped terms.
+Our formal construction appears novel and we believe that a theoretic understanding of our construction and its connection to intrinsic typing as a bridge between raw terms and well-typed terms deserve some attention.
+However, carrying out a category-theoretic analysis requires expertise in category theory, which goes beyond the scope of this paper.
+
+Therefore, in this section, we will primarily provide an overview of our design and offer intuitive explanations.
+We will discuss the construction of simple types in \Cref{subsec:formal-simple-types}, raw terms in~\Cref{subsec:formal-raw-terms}, and bidirectional typing rules in \Cref{subsec:formal-bidirectional-typing}.
+Since the formal proofs largely reflect their informal presentation, we will limit our illustration to the 'if' part of \Cref{lem:soundness-completeness} as an example in \Cref{subsec:formal-proofs}.
+
+\subsection{Defining Simple Types}\label{subsec:formal-simple-types}
 
 \begin{figure}
+  \codefigure
   \small
-  \begin{minipage}[t]{.3\linewidth}
+  \begin{minipage}[t]{.45\textwidth}
     \begin{hscode}\SaveRestoreHook
@@
 \column{5}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{12}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{13}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{16}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Keyword{record}\;\Conid{SigD}\;\mathbin{:}\;\Conid{Set₁}\;\Keyword{where}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Keyword{field}{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\Conid{Op}\;{}\<[13]%
+\>[13]{}\mathbin{:}\;{}\<[16]%
+\>[16]{}\Conid{Set}{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\Varid{decEq}\;{}\<[13]%
+\>[13]{}\mathbin{:}\;{}\<[16]%
+\>[16]{}\Conid{DecEq}\;\Conid{Op}{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\Varid{ar}\;{}\<[13]%
+\>[13]{}\mathbin{:}\;{}\<[16]%
+\>[16]{}\Conid{Op}\;\Varid{→}\;\Conid{ℕ}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+  \end{minipage}
+  \begin{minipage}[t]{.45\textwidth}
+    \begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{8}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{24}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{⟦\char95 ⟧}\;\mathbin{:}\;\Conid{SigD}\;\Varid{→}\;\Conid{Set}\;\Varid{→}\;\Conid{Set}{}\<[E]%
+\\
+\>[B]{}\Varid{⟦}\;\Conid{D}\;\Varid{⟧}\;\Conid{X}\;\mathrel{=}\;\Conid{Σ[}\;\Varid{i}\;\Varid{∈}\;\Conid{D}\;\Varid{.Op}\;\mskip1.5mu]\;\Conid{X}\;\hat{}\;(\Conid{D}\;\Varid{.ar}\;\Varid{i}){}\<[E]%
+\\[\blanklineskip]%
+\>[B]{}\Keyword{data}\;\Conid{Ty}\;(\Conid{Ξ}\;\mathbin{:}\;\Conid{ℕ})\;\mathbin{:}\;\Conid{Set}\;\Keyword{where}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\text{\textasciigrave}\anonymous \;{}\<[8]%
+\>[8]{}\mathbin{:}\;\Conid{Fin}\;\Conid{Ξ}\;{}\<[24]%
+\>[24]{}\Varid{→}\;\Conid{Ty}\;\Conid{Ξ}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{op}\;{}\<[8]%
+\>[8]{}\mathbin{:}\;\Varid{⟦}\;\Conid{D}\;\Varid{⟧}\;(\Conid{Ty}\;\Conid{Ξ})\;{}\<[24]%
+\>[24]{}\Varid{→}\;\Conid{Ty}\;\Conid{Ξ}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+  \end{minipage}
+  \caption{Simple types in \Agda}
+  \label{fig:formal-simple-type}
+\end{figure}
+We start with the formal definition of signatures and simple types (\Cref{fig:formal-simple-type}), which parallels its informal counterpart (\Cref{def:simple-signature}), with the exception of \ensuremath{\Varid{⟦\char95 ⟧}}.
+To prevent circular references during the inductive definition of simple types, we initially define \ensuremath{\Varid{⟦}\;\Conid{D}\;\Varid{⟧}} on an arbitrary \ensuremath{\Conid{Set}} instead of \ensuremath{\Conid{Ty}}.
+Thus, types specified by a signature can be defined using two constructors--- \ensuremath{\text{\textasciigrave}\anonymous }\ and \ensuremath{\Varid{op}}---where a variable is represented by \ensuremath{\text{\textasciigrave}\;\Varid{n}} for some \ensuremath{\Varid{n}} less than \ensuremath{\Conid{Ξ}}, and each inhabitant \ensuremath{\Varid{op}\;(\Varid{i}\;\Varid{,}\;\Varid{ts})} is a pair of an operation symbol \ensuremath{\Varid{i}} and a list \ensuremath{\Varid{ts}} of length $n$.
+
+From a categorical standpoint, \ensuremath{\Varid{⟦}\;\Conid{D}\;\Varid{⟧}} acts as a functor from \ensuremath{\Conid{Set}} to \ensuremath{\Conid{Set}}, mapping $X$ to an $\Conid{Op}$-indexed coproduct $\sum_{i \in \Conid{Op}} X ^ {\Conid{ar}(i)}$ of products of $X$.
+The type \ensuremath{\Conid{Ty}} is the free \ensuremath{\Varid{⟦}\;\Conid{D}\;\Varid{⟧}}-algebra over the type \ensuremath{\Conid{Fin}\;\Conid{Ξ}} or the initial algebra for the functor \ensuremath{\Conid{Fin}\;\Conid{Ξ}\;\Varid{+}\;\Varid{⟦}\;\Conid{D}\;\Varid{⟧}}.
+
+Finally, to account for indices when defining raw terms and bidirectional typing derivations by a signature, we define a corresponding signature functor \ensuremath{\Varid{⟦}\;\Conid{D}\;\Varid{⟧}} for a (bidirectional) binding signature.
+
+\subsection{Defining Raw Terms}\label{subsec:formal-raw-terms}
+As shown in \Cref{fig:formal-bibisig}, we establish \ensuremath{\Conid{ArgD}}, \ensuremath{\Conid{OpD}}, and \ensuremath{\Conid{SigD}} to represent arguments $[\Delta]A^{\dir{d}}$, bidirectional binding operations $\biop$, and bidirectional binding signatures, respectively, in line with their informal definitions.
+By removing \ensuremath{\Varid{mode}} from these definitions, we retrieve the definitions of binding arguments, arities, and signatures.
+\begin{figure}
+  \codefigure
+  \begin{minipage}[t]{.3\textwidth}
+  \begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{19}{@{}>{\hspre}l<{\hspost}@{}}%
 \column{E}{@{}>{\hspre}l<{\hspost}@{}}%
@@
 \>[3]{}\hsindent{2}{}\<[5]%
-\>[5]{}\Varid{cxt}\;{}\<[12]%
-\>[12]{}\mathbin{:}\;\Conid{Cxt}\;\Conid{Ξ}{}\<[E]%
+\>[5]{}\Varid{cxt}\;{}\<[19]%
+\>[19]{}\mathbin{:}\;\Conid{Cxt}\;\Conid{Ξ}{}\<[E]%
 \\
 \>[3]{}\hsindent{2}{}\<[5]%
-\>[5]{}\Varid{type}\;{}\<[12]%
-\>[12]{}\mathbin{:}\;\Conid{TExp}\;\Conid{Ξ}{}\<[E]%
+\>[5]{}\Varid{type}\;{}\<[19]%
+\>[19]{}\mathbin{:}\;\Conid{Ty}\;\Conid{Ξ}{}\<[E]%
 \\
 \>[3]{}\hsindent{2}{}\<[5]%
-\>[5]{}\Varid{mode}\;{}\<[12]%
-\>[12]{}\mathbin{:}\;\Conid{Mode}{}\<[E]%
+\>[5]{}\dir{\Varid{mode}}\;{}\<[19]%
+\>[19]{}\mathbin{:}\;\Conid{Mode}{}\<[E]%
+\\[\blanklineskip]%
+\>[B]{}\Conid{ArgsD}\;\mathrel{=}\;\Conid{List}\;\Varid{∘}\;\Conid{ArgD}{}\<[E]%
 \ColumnHook
@@
 \column{5}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{14}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{22}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{21}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{29}{@{}>{\hspre}l<{\hspost}@{}}%
 \column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\Keyword{record}\;\Conid{ConD}\;\mathbin{:}\;\Conid{Set}\;\Keyword{where}{}\<[E]%
+\>[B]{}\Keyword{record}\;\Conid{OpD}\;\mathbin{:}\;\Conid{Set}\;\Keyword{where}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Keyword{constructor}\;\Varid{ι}{}\<[E]%
 \\
@@
 \>[3]{}\hsindent{2}{}\<[5]%
-\>[5]{}\{\mskip1.5mu \Varid{vars}\mskip1.5mu\}\;{}\<[14]%
-\>[14]{}\mathbin{:}\;\Conid{ℕ}{}\<[E]%
+\>[5]{}\{\mskip1.5mu \Varid{tvars}\mskip1.5mu\}\;{}\<[21]%
+\>[21]{}\mathbin{:}\;\Conid{ℕ}{}\<[E]%
 \\
 \>[3]{}\hsindent{2}{}\<[5]%
-\>[5]{}\Varid{args}\;{}\<[14]%
-\>[14]{}\mathbin{:}\;\Conid{List}\;(\Conid{ArgsD}\;\Varid{vars}){}\<[E]%
+\>[5]{}\dir{\Varid{mode}}\;{}\<[21]%
+\>[21]{}\mathbin{:}\;\Conid{Mode}{}\<[E]%
 \\
 \>[3]{}\hsindent{2}{}\<[5]%
-\>[5]{}\Varid{type}\;{}\<[14]%
-\>[14]{}\mathbin{:}\;\Conid{TExp}\;{}\<[22]%
-\>[22]{}\Varid{vars}{}\<[E]%
+\>[5]{}\Varid{type}\;{}\<[21]%
+\>[21]{}\mathbin{:}\;\Conid{TExp}\;{}\<[29]%
+\>[29]{}\Varid{tvars}{}\<[E]%
 \\
 \>[3]{}\hsindent{2}{}\<[5]%
-\>[5]{}\Varid{mode}\;{}\<[14]%
-\>[14]{}\mathbin{:}\;\Conid{Mode}{}\<[E]%
+\>[5]{}\Varid{args}\;{}\<[21]%
+\>[21]{}\mathbin{:}\;\Conid{ArgsD}\;\Varid{tvars}{}\<[E]%
 \ColumnHook
@@
 \column{5}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{17}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{15}{@{}>{\hspre}l<{\hspost}@{}}%
 \column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\Keyword{record}\;\Conid{Desc}\;\mathbin{:}\;\Conid{Set₁}\;\Keyword{where}{}\<[E]%
+\>[B]{}\Keyword{record}\;\Conid{SigD}\;\mathbin{:}\;\Conid{Set₁}\;\Keyword{where}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Keyword{constructor}\;\Varid{sigd}{}\<[E]%
 \\
@@
 \>[3]{}\hsindent{2}{}\<[5]%
-\>[5]{}\Conid{Op}\;{}\<[17]%
-\>[17]{}\mathbin{:}\;\Conid{Set}{}\<[E]%
+\>[5]{}\Conid{Op}\;{}\<[15]%
+\>[15]{}\mathbin{:}\;\Conid{Set}{}\<[E]%
 \\
 \>[3]{}\hsindent{2}{}\<[5]%
-\>[5]{}\Varid{rules}\;{}\<[17]%
-\>[17]{}\mathbin{:}\;\Conid{Op}\;\Varid{→}\;\Conid{ConD}{}\<[E]%
+\>[5]{}\Varid{ar}\;{}\<[15]%
+\>[15]{}\mathbin{:}\;\Conid{Op}\;\Varid{→}\;\Conid{OpD}{}\<[E]%
 \ColumnHook
@@
   \end{minipage}
-
-  \caption{Definition of bidirectional binding signature}  
+  \caption{Bidirectional binding signature in \Agda}
+  \label{fig:formal-bibisig}
+\end{figure}
+To construct well-scoped raw terms indexed by a list $V$ of free variables, we introduce another signature functor mapping \ensuremath{\Conid{Fam}} to \ensuremath{\Conid{Fam}}, where \ensuremath{\Conid{Fam}} is the family of sets defined as \ensuremath{\Conid{ℕ}\;\Varid{→}\;\Conid{Set}}, indexed by a natural number indicating the number of free variables.
+This signature endofunctor for raw term construction resembles the one for defining simple types.
+However, we define four distinct functors instead of defining one directly.
+These are \ensuremath{\Varid{⟦}\;\Conid{Δ}\;\Varid{⟧ᵃ}} for an extension context \ensuremath{\Conid{Δ}}, \ensuremath{\Varid{⟦}\;\Conid{Ds}\;\Varid{⟧ᵃˢ}} for an argument list \ensuremath{\Conid{Ds}}, \ensuremath{\Varid{⟦}\;\Varid{ι}\;\Conid{Ds}\;\anonymous \;\Varid{⟧ᶜ}} for an operation (\ensuremath{\Varid{ι}\;\Conid{Ds}\;\Conid{A}}), and \ensuremath{\Varid{⟦}\;\Conid{D}\;\Varid{⟧ᶜ}} for a binding signature $D$.
+Notably, the functor \ensuremath{\Varid{⟦}\;\Conid{Δ}\;\Varid{⟧ᵃ}} extends the context by $\erase{\Delta}$ variables by mapping the family of sets \ensuremath{\Conid{X}\;\Varid{n}} to \ensuremath{\Conid{X}\;(\Varid{length}\;\Conid{Δ}\;\Varid{+}\;\Varid{n})}.
+\begin{figure}
+  \codefigure
+\begin{minipage}[t]{.33\textwidth}
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{8}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{17}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{19}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{22}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{27}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{⟦\char95 ⟧ᵃ}\;{}\<[8]%
+\>[8]{}\mathbin{:}\;\Conid{TExps}\;\Conid{Ξ}\;{}\<[19]%
+\>[19]{}\Varid{→}\;\Conid{Fam}\;\Varid{→}\;\Conid{Fam}{}\<[E]%
+\\
+\>[B]{}\Varid{⟦}\;\Conid{Δ}\;{}\<[17]%
+\>[17]{}\Varid{⟧ᵃ}\;{}\<[22]%
+\>[22]{}\Conid{X}\;\Varid{n}\;{}\<[27]%
+\>[27]{}\mathrel{=}\;{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Conid{X}\;(\Varid{length}\;\Conid{Δ}\;\Varid{+}\;\Varid{n}){}\<[E]%
+\\[\blanklineskip]%
+\>[B]{}\Varid{⟦\char95 ⟧ᵃˢ}\;{}\<[8]%
+\>[8]{}\mathbin{:}\;\Conid{ArgsD}\;\Conid{Ξ}\;{}\<[19]%
+\>[19]{}\Varid{→}\;\Conid{Fam}\;\Varid{→}\;\Conid{Fam}{}\<[E]%
+\\
+\>[B]{}\Varid{⟦}\;\Varid{[]}\;{}\<[17]%
+\>[17]{}\Varid{⟧ᵃˢ}\;{}\<[22]%
+\>[22]{}\anonymous \;\anonymous \;{}\<[27]%
+\>[27]{}\mathrel{=}\;\Varid{⊤}{}\<[E]%
+\\
+\>[B]{}\Varid{⟦}\;(\Conid{Δ}\;\Varid{⊢}\;\Conid{A})\;\Varid{∷}\;\Conid{Ds}\;{}\<[17]%
+\>[17]{}\Varid{⟧ᵃˢ}\;{}\<[22]%
+\>[22]{}\Conid{X}\;\Varid{n}\;{}\<[27]%
+\>[27]{}\mathrel{=}\;{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{⟦}\;\Conid{Δ}\;\Varid{⟧ᵃ}\;\Conid{X}\;\Varid{n}\;\Varid{×}\;\Varid{⟦}\;\Conid{Ds}\;\Varid{⟧ᵃˢ}\;\Conid{X}\;\Varid{n}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+\end{minipage}
+\begin{minipage}[t]{.3\textwidth}
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{8}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{10}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{11}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{15}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{16}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{20}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{⟦\char95 ⟧ᶜ}\;{}\<[8]%
+\>[8]{}\mathbin{:}\;\Conid{OpD}\;{}\<[15]%
+\>[15]{}\Varid{→}\;\Conid{Fam}\;\Varid{→}\;\Conid{Fam}{}\<[E]%
+\\
+\>[B]{}\Varid{⟦}\;\Varid{ι}\;\Conid{Ds}\;\anonymous \;{}\<[11]%
+\>[11]{}\Varid{⟧ᶜ}\;{}\<[15]%
+\>[15]{}\Conid{X}\;{}\<[20]%
+\>[20]{}\mathrel{=}\;\Varid{⟦}\;\Conid{Ds}\;\Varid{⟧ᵃˢ}\;\Conid{X}{}\<[E]%
+\\[\blanklineskip]%
+\>[B]{}\Varid{⟦\char95 ⟧}\;{}\<[8]%
+\>[8]{}\mathbin{:}\;\Conid{SigD}\;{}\<[16]%
+\>[16]{}\Varid{→}\;\Conid{Fam}\;\Varid{→}\;\Conid{Fam}{}\<[E]%
+\\
+\>[B]{}\Varid{⟦}\;\Conid{D}\;{}\<[10]%
+\>[10]{}\Varid{⟧}\;{}\<[15]%
+\>[15]{}\Conid{X}\;\Varid{n}\;{}\<[20]%
+\>[20]{}\mathrel{=}\;{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Conid{Σ[}\;\Varid{i}\;\Varid{∈}\;\Conid{D}\;\Varid{.Op}\;\mskip1.5mu]\;\Varid{⟦}\;\Conid{D}\;\Varid{.ar}\;\Varid{i}\;\Varid{⟧ᶜ}\;\Conid{X}\;\Varid{n}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+\end{minipage}
+\begin{minipage}[t]{.33\textwidth}
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{8}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{23}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{33}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Keyword{data}\;\Conid{Raw}\;\mathbin{:}\;\Conid{ℕ}\;\Varid{→}\;\Conid{Set}\;\Keyword{where}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\text{\textasciigrave}\anonymous \;{}\<[8]%
+\>[8]{}\mathbin{:}\;\Conid{Fin}\;\Varid{n}\;{}\<[33]%
+\>[33]{}\Varid{→}\;\Conid{Raw}\;\Varid{n}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{\char95 ∋\char95 }\;{}\<[8]%
+\>[8]{}\mathbin{:}\;\Conid{Ty}\;{}\<[23]%
+\>[23]{}\Varid{→}\;\Conid{Raw}\;\Varid{n}\;{}\<[33]%
+\>[33]{}\Varid{→}\;\Conid{Raw}\;\Varid{n}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{op}\;{}\<[8]%
+\>[8]{}\mathbin{:}\;\Varid{⟦}\;\Conid{D}\;\Varid{⟧}\;\Conid{Raw}\;\Varid{n}\;{}\<[33]%
+\>[33]{}\Varid{→}\;\Conid{Raw}\;\Varid{n}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+\end{minipage}
+\caption{Signature functor for raw terms and raw terms in \Agda}
+\label{fig:formal-raw-term-functor}
+\label{fig:formal-raw-terms}
 \end{figure}
 
+The inductive type of raw terms, indexed by a list of free variables, is defined as shown in \Cref{fig:formal-raw-terms}.
+This mirrors our informal definition (\Cref{fig:raw-terms}), where \ensuremath{\Conid{A}\;\Varid{∋}\;\Varid{t}} corresponds to the raw term $t \annotate A$.
+These definitions align closely with those found in other referenced frameworks.
+
+\subsection{Defining Extrinsic Bidirectional Typing Derivations}\label{subsec:formal-bidirectional-typing}
+Formally defining extrinsic derivations is arguably the most complex aspect of our formalisation.
+This requires defining a family of sets, indexed by context, type, mode, and notably, a raw term.
+This definition involves another signature endofunctor, requiring consistency with the functor used for constructing raw terms.
+
+We begin with the family of sets on which the functor acts.
+A typing judgement $\Gamma \vdash_{\Sigma, \Omega} t :^{\dir{d}} A$ can be viewed as a predicate over raw terms with four indices: a context, a $\N$-indexed family of sets, a mode, and a type.
+The type of this predicate can be generalised to a family of sets over the $\N$-indexed family of sets, as follows:
+{\small
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{8}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Conid{Fam}\;\mathbin{:}\;\Conid{R.Fam}\;\Varid{→}\;\Conid{Set}\;\anonymous {}\<[E]%
+\\
+\>[B]{}\Conid{Fam}\;\Conid{X}\;{}\<[8]%
+\>[8]{}\mathrel{=}\;(\Conid{Γ}\;\mathbin{:}\;\Conid{Cxt}\;\Varid{0})\;\Varid{→}\;\Conid{X}\;(\Varid{length}\;\Conid{Γ})\;\Varid{→}\;\Conid{Mode}\;\Varid{→}\;\Conid{Ty}\;\Varid{→}\;\Conid{Set}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+}
+Observe that the second index \ensuremath{\Conid{X}} depends on the length of the first index \ensuremath{\Conid{Γ}}.
+This is because a typing derivation for a raw term $V \vdash_{\Sigma, \erase{\Omega}} t$ requires the same number of variables as in $V$.
+
+In a similar vein, we define four functors for the construction of bidirectional typing derivations, albeit in a top-down manner.
+For an endofunctor on predicates, each functor must act as a \emph{lifting} of the functor on its domain, illustrated below.
+The signature endofunctor \ensuremath{\Varid{⟦\char95 ⟧}} for a bidirectional binding signature is defined as follows:
+{\small
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{6}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{⟦\char95 ⟧}\;{}\<[6]%
+\>[6]{}\mathbin{:}\;(\Conid{D}\;\mathbin{:}\;\Conid{SigD})\;(\Conid{X}\;\mathbin{:}\;\Conid{R.Fam})\;(\Conid{Y}\;\mathbin{:}\;\Conid{Fam}\;\Conid{X})\;\Varid{→}\;\Conid{Fam}\;(\Conid{R.⟦}\;\Varid{erase}\;\Conid{D}\;\Varid{⟧}\;\Conid{X}){}\<[E]%
+\\
+\>[B]{}\Varid{⟦}\;\Conid{D}\;\Varid{⟧}\;\Conid{X}\;\Conid{Y}\;\Conid{Γ}\;(\Varid{o}\;\Varid{,}\;\Varid{ts})\;\Varid{d}\;\Conid{A}\;\mathrel{=}\;\Varid{⟦}\;\Conid{D}\;\Varid{.ar}\;\Varid{o}\;\Varid{⟧ᶜ}\;\Conid{X}\;\Conid{Y}\;\Conid{Γ}\;\Varid{ts}\;\Varid{d}\;\Conid{A}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+}
+This functor maps a predicate \ensuremath{\Conid{Y}} over \ensuremath{\Conid{X}} into a predicate \ensuremath{\Varid{⟦}\;\Conid{D}\;\Varid{⟧}\;\Conid{X}\;\Conid{Y}} over \ensuremath{\Conid{R.⟦}\;\Varid{erase}\;\Conid{D}\;\Varid{⟧}\;\Conid{X}}.
+Here, \ensuremath{\Varid{erase}\;\Conid{D}} represents the mode erasure of \ensuremath{\Conid{D}}, and the prefix \ensuremath{\Conid{R}} in \ensuremath{\Conid{R.⟦\char95 ⟧}} indicates the signature endofunctor used in raw term construction.
+
+The functor \ensuremath{\Varid{⟦}\;\Conid{D}\;\Varid{.ar}\;\Varid{o}\;\Varid{⟧ᶜ}} is defined for an operation symbol $\biop$:
 {\small\begin{hscode}\SaveRestoreHook
@@
 \column{7}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{⟦\char95 ⟧ᶜ}\;{}\<[7]%
+\>[7]{}\mathbin{:}\;(\Conid{D}\;\mathbin{:}\;\Conid{OpD})\;(\Conid{X}\;\mathbin{:}\;\Conid{R.Fam}\;\Varid{ℓ})\;(\Conid{Y}\;\mathbin{:}\;\Conid{Fam}\;\iden{\Varid{ℓ}^\prime}\;\Conid{X})\;{}\<[E]%
+\\
+\>[7]{}\Varid{→}\;\Conid{Fam}\;\iden{\Varid{ℓ}^\prime}\;(\Conid{R.⟦}\;\Varid{eraseᶜ}\;\Conid{D}\;\Varid{⟧ᶜ}\;\Conid{X}){}\<[E]%
+\\
+\>[B]{}\Varid{⟦}\;\Varid{ι}\;\{\mskip1.5mu \Conid{Ξ}\mskip1.5mu\}\;\Varid{d}\;\Conid{A₀}\;\Conid{D}\;\Varid{⟧ᶜ}\;\Conid{X}\;\Conid{Y}\;\Conid{Γ}\;\Varid{xs}\;\Varid{d′}\;\Conid{A}\;\mathrel{=}\;{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{d}\;\Varid{≡}\;\Varid{d′}\;\Varid{×}\;\Conid{Σ[}\;\Varid{p}\;\Varid{∈}\;\Conid{TSub}\;\Conid{Ξ}\;\Varid{0}\;\mskip1.5mu]\;\Conid{A₀}\;\Varid{⟨}\;\Varid{p}\;\Varid{⟩}\;\Varid{≡}\;\Conid{A}\;\Varid{×}\;\Varid{⟦}\;\Conid{D}\;\Varid{⟧ᵃˢ}\;\Conid{X}\;\Conid{Y}\;\Varid{σ}\;\Conid{Γ}\;\Varid{xs}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+}
+A set within the family is only inhabited if the indices correspond with the mode and the type as specified by the operation.
+Therefore, in defining the set \ensuremath{\Varid{⟦}\;\Varid{ι}\;\{\mskip1.5mu \Conid{Ξ}\mskip1.5mu\}\;\Varid{d}\;\Conid{A₀}\;\Conid{D}\;\Varid{⟧ᶜ}\;\Conid{X}\;\Conid{Y}\;\Conid{Γ}\;\Varid{xs}\;\Varid{d′}\;\Conid{A}}, it takes identity proofs for \ensuremath{\Varid{d}\;\Varid{≡}\;\Varid{d′}} and \ensuremath{\Conid{A₀}\;\Varid{⟨}\;\Varid{σ}\;\Varid{⟩}\;\Varid{≡}\;\Conid{A}}, with \ensuremath{\Conid{A₀}} instantiated by a substitution \ensuremath{\Varid{σ}} from $\Xi$ to $\emptyset$.
+This ensures that the set is inhabited only if the indices align with what the arity (\ensuremath{\Conid{D}\;\Varid{.ar}\;\Varid{o}}) specifies.
+
+The signature endofunctor for an argument list \ensuremath{\Conid{Ds}\;\mathbin{:}\;\Conid{ArgsD}\;\Conid{Ξ}} is defined in a manner that it maps a predicate to a family of sets indexed by a context and a family of product sets, considering that each premise in a typing derivation shares a common context $\Gamma$ and a term from a list of subterms.
+{\small\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
 \column{8}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{10}{@{}>{\hspre}l<{\hspost}@{}}%
 \column{20}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{24}{@{}>{\hspre}l<{\hspost}@{}}%
 \column{27}{@{}>{\hspre}l<{\hspost}@{}}%
@@
 \column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\Conid{Fam}\;\mathbin{:}\;(\Varid{ℓ′}\;\mathbin{:}\;\Conid{Level})\;(\Conid{X}\;\mathbin{:}\;\Conid{R.Fam}\;\Varid{ℓ})\;\Varid{→}\;\Conid{Set}\;(\Varid{ℓ}\;\Varid{⊔}\;\Varid{lsuc}\;\Varid{ℓ′}){}\<[E]%
+\>[B]{}\Varid{⟦\char95 ⟧ᵃˢ}\;{}\<[8]%
+\>[8]{}\mathbin{:}\;(\Conid{D}\;\mathbin{:}\;\Conid{ArgsD}\;\Conid{Ξ})\;(\Conid{X}\;\mathbin{:}\;\Conid{R.Fam})\;(\Conid{Y}\;\mathbin{:}\;\Conid{Fam}\;\Conid{X})\;{}\<[E]%
 \\
-\>[B]{}\Conid{Fam}\;\Varid{ℓ′}\;\Conid{X}\;\mathrel{=}\;(\Conid{Γ}\;\mathbin{:}\;\Conid{Cxt}\;\Varid{0})\;\Varid{→}\;\Conid{X}\;(\Varid{length}\;\Conid{Γ})\;\Varid{→}\;\Conid{Mode}\;\Varid{→}\;\Conid{Ty}\;\Varid{→}\;\Conid{Set}\;\Varid{ℓ′}{}\<[E]%
-\\[\blanklineskip]%
-\>[B]{}\Varid{⟦\char95 ⟧ᵃ}\;{}\<[7]%
-\>[7]{}\mathbin{:}\;(\Conid{Δ}\;\mathbin{:}\;\Conid{TExps}\;\Conid{Ξ})\;(\Conid{X}\;\mathbin{:}\;\Conid{R.Fam}\;\Varid{ℓ})\;(\Conid{Y}\;\mathbin{:}\;(\Conid{Γ}\;\mathbin{:}\;\Conid{Cxt}\;\Varid{0})\;\Varid{→}\;\Conid{X}\;(\Varid{length}\;\Conid{Γ})\;\Varid{→}\;\Conid{Set}\;\Varid{ℓ′})\;{}\<[E]%
-\\
-\>[7]{}\Varid{→}\;\Conid{TSub}\;\Conid{Ξ}\;\Varid{0}\;\Varid{→}\;(\Conid{Γ}\;\mathbin{:}\;\Conid{Cxt}\;\Varid{0})\;\Varid{→}\;\Conid{R.⟦}\;\Conid{Δ}\;\Varid{⟧ᵃ}\;\Conid{X}\;(\Varid{length}\;\Conid{Γ})\;\Varid{→}\;\Conid{Set}\;\Varid{ℓ′}{}\<[E]%
-\\
-\>[B]{}\Varid{⟦}\;\Varid{[]}\;{}\<[10]%
-\>[10]{}\Varid{⟧ᵃ}\;\Conid{X}\;\Conid{Y}\;\Varid{σ}\;\Conid{Γ}\;\Varid{t}\;{}\<[24]%
-\>[24]{}\mathrel{=}\;\Conid{Y}\;\Conid{Γ}\;\Varid{t}{}\<[E]%
-\\
-\>[B]{}\Varid{⟦}\;\Conid{A}\;\Varid{∷}\;\Conid{Δ}\;{}\<[10]%
-\>[10]{}\Varid{⟧ᵃ}\;\Conid{X}\;\Conid{Y}\;\Varid{σ}\;\Conid{Γ}\;\Varid{t}\;{}\<[24]%
-\>[24]{}\mathrel{=}\;\Varid{⟦}\;\Conid{Δ}\;\Varid{⟧ᵃ}\;\Conid{X}\;\Conid{Y}\;\Varid{σ}\;((\Conid{A}\;\Varid{⟨}\;\Varid{σ}\;\Varid{⟩})\;\Varid{∷}\;\Conid{Γ})\;\Varid{t}{}\<[E]%
-\\[\blanklineskip]%
-\>[B]{}\Varid{⟦\char95 ⟧ᵃˢ}\;{}\<[8]%
-\>[8]{}\mathbin{:}\;(\Conid{D}\;\mathbin{:}\;\Conid{ArgsD}\;\Conid{Ξ})\;(\Conid{X}\;\mathbin{:}\;\Conid{R.Fam}\;\Varid{ℓ})\;(\Conid{Y}\;\mathbin{:}\;\Conid{Fam}\;\Varid{ℓ′}\;\Conid{X})\;{}\<[E]%
-\\
-\>[8]{}\Varid{→}\;\Conid{TSub}\;\Conid{Ξ}\;\Varid{0}\;\Varid{→}\;(\Conid{Γ}\;\mathbin{:}\;\Conid{Cxt}\;\Varid{0})\;\Varid{→}\;\Conid{R.⟦}\;\Varid{eraseᵃˢ}\;\Conid{D}\;\Varid{⟧ᵃˢ}\;\Conid{X}\;(\Varid{length}\;\Conid{Γ})\;\Varid{→}\;\Conid{Set}\;\Varid{ℓ′}{}\<[E]%
+\>[8]{}\Varid{→}\;\Conid{TSub}\;\Conid{Ξ}\;\Varid{0}\;\Varid{→}\;(\Conid{Γ}\;\mathbin{:}\;\Conid{Cxt}\;\Varid{0})\;\Varid{→}\;\Conid{R.⟦}\;\Varid{eraseᵃˢ}\;\Conid{D}\;\Varid{⟧ᵃˢ}\;\Conid{X}\;(\Varid{length}\;\Conid{Γ})\;\Varid{→}\;\Conid{Set}{}\<[E]%
 \\
@@
 \>[36]{}(\Varid{x}\;\Varid{,}\;\Varid{xs})\;{}\<[46]%
-\>[46]{}\mathrel{=}\;{}\<[E]%
-\\
-\>[B]{}\hsindent{3}{}\<[3]%
-\>[3]{}\Varid{⟦}\;\Conid{Δ}\;\Varid{⟧ᵃ}\;\Conid{X}\;(\Varid{λ}\;\Conid{Γ'}\;\Varid{x'}\;\Varid{→}\;\Conid{Y}\;\Conid{Γ'}\;\Varid{x'}\;\Varid{d}\;(\Conid{A}\;\Varid{⟨}\;\Varid{σ}\;\Varid{⟩}))\;\Varid{σ}\;\Conid{Γ}\;\Varid{x}\;\Varid{×}\;\Varid{⟦}\;\Conid{Ds}\;\Varid{⟧ᵃˢ}\;\Conid{X}\;\Conid{Y}\;\Varid{σ}\;\Conid{Γ}\;\Varid{xs}{}\<[E]%
-\\[\blanklineskip]%
-\>[B]{}\Varid{⟦\char95 ⟧ᶜ}\;\mathbin{:}\;(\Conid{D}\;\mathbin{:}\;\Conid{ConD})\;(\Conid{X}\;\mathbin{:}\;\Conid{R.Fam}\;\Varid{ℓ})\;(\Conid{Y}\;\mathbin{:}\;\Conid{Fam}\;\Varid{ℓ′}\;\Conid{X})\;\Varid{→}\;\Conid{Fam}\;\Varid{ℓ′}\;(\Conid{R.⟦}\;\Varid{eraseᶜ}\;\Conid{D}\;\Varid{⟧ᶜ}\;\Conid{X}){}\<[E]%
-\\
-\>[B]{}\Varid{⟦}\;\Varid{ι}\;\{\mskip1.5mu \Conid{Ξ}\mskip1.5mu\}\;\Varid{d}\;\Conid{B}\;\Conid{D}\;\Varid{⟧ᶜ}\;\Conid{X}\;\Conid{Y}\;\Conid{Γ}\;\Varid{xs}\;\Varid{d′}\;\Conid{A}\;\mathrel{=}\;{}\<[E]%
-\\
-\>[B]{}\hsindent{3}{}\<[3]%
-\>[3]{}\Varid{d}\;\Varid{≡}\;\Varid{d′}\;\Varid{×}\;\Conid{Σ[}\;\Varid{σ}\;\Varid{∈}\;\Conid{TSub}\;\Conid{Ξ}\;\Varid{0}\;\mskip1.5mu]\;\Conid{B}\;\Varid{⟨}\;\Varid{σ}\;\Varid{⟩}\;\Varid{≡}\;\Conid{A}\;\Varid{×}\;\Varid{⟦}\;\Conid{D}\;\Varid{⟧ᵃˢ}\;\Conid{X}\;\Conid{Y}\;\Varid{σ}\;\Conid{Γ}\;\Varid{xs}{}\<[E]%
-\\[\blanklineskip]%
-\>[B]{}\Varid{⟦\char95 ⟧}\;\mathbin{:}\;(\Conid{D}\;\mathbin{:}\;\Conid{Desc})\;(\Conid{X}\;\mathbin{:}\;\Conid{R.Fam}\;\Varid{ℓ})\;(\Conid{Y}\;\mathbin{:}\;\Conid{Fam}\;\Varid{ℓ′}\;\Conid{X})\;\Varid{→}\;\Conid{Fam}\;\Varid{ℓ′}\;(\Conid{R.⟦}\;\Varid{erase}\;\Conid{D}\;\Varid{⟧}\;\Conid{X}){}\<[E]%
-\\
-\>[B]{}\Varid{⟦}\;\Conid{D}\;\Varid{⟧}\;\Conid{X}\;\Conid{Y}\;\Conid{Γ}\;(\Varid{i}\;\Varid{,}\;\Varid{xs})\;\Varid{d}\;\Conid{A}\;\mathrel{=}\;\Varid{⟦}\;\Conid{D}\;\Varid{.rules}\;\Varid{i}\;\Varid{⟧ᶜ}\;\Conid{X}\;\Conid{Y}\;\Conid{Γ}\;\Varid{xs}\;\Varid{d}\;\Conid{A}{}\<[E]%
+\>[46]{}\mathrel{=}\;\Varid{⟦}\;\Conid{Δ}\;\Varid{⟧ᵃ}\;\Conid{X}\;(\Varid{λ}\;\iden{\Conid{Γ}^\prime}\;\iden{\Varid{x}^\prime}\;\Varid{→}\;\Conid{Y}\;\iden{\Conid{Γ}^\prime}\;\iden{\Varid{x}^\prime}\;\Varid{d}\;(\Conid{A}\;\Varid{⟨}\;\Varid{σ}\;\Varid{⟩}))\;\Varid{σ}\;\Conid{Γ}\;\Varid{x}\;\Varid{×}\;\Varid{⟦}\;\Conid{Ds}\;\Varid{⟧ᵃˢ}\;\Conid{X}\;\Conid{Y}\;\Varid{σ}\;\Conid{Γ}\;\Varid{xs}{}\<[E]%
 \ColumnHook
@@
 }
+The signature endofunctor \ensuremath{\Varid{⟦}\;\Conid{Δ}\;\Varid{⟧ᵃ}} for an extension context \ensuremath{\Conid{Δ}} maps a predicate $Y$ over a $\N$-indexed family $X$ of sets into a predicate (\ensuremath{\Varid{⟦}\;\Conid{Δ}\;\Varid{⟧ᵃ}\;\Conid{X}\;\Conid{Y}}) whose context is extended by the types in \ensuremath{\Conid{Δ}}.
+{\small\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{7}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{10}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{24}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{⟦\char95 ⟧ᵃ}\;{}\<[7]%
+\>[7]{}\mathbin{:}\;(\Conid{Δ}\;\mathbin{:}\;\Conid{Cxt}\;\Conid{Ξ})\;(\Conid{X}\;\mathbin{:}\;\Conid{R.Fam})\;(\Conid{Y}\;\mathbin{:}\;(\Conid{Γ}\;\mathbin{:}\;\Conid{Cxt}\;\Varid{0})\;\Varid{→}\;\Conid{X}\;(\Varid{length}\;\Conid{Γ})\;\Varid{→}\;\Conid{Set})\;{}\<[E]%
+\\
+\>[7]{}\Varid{→}\;\Conid{TSub}\;\Conid{Ξ}\;\Varid{0}\;\Varid{→}\;(\Conid{Γ}\;\mathbin{:}\;\Conid{Cxt}\;\Varid{0})\;\Varid{→}\;\Conid{R.⟦}\;\Conid{Δ}\;\Varid{⟧ᵃ}\;\Conid{X}\;(\Varid{length}\;\Conid{Γ})\;\Varid{→}\;\Conid{Set}{}\<[E]%
+\\
+\>[B]{}\Varid{⟦}\;\Varid{[]}\;{}\<[10]%
+\>[10]{}\Varid{⟧ᵃ}\;\Conid{X}\;\Conid{Y}\;\Varid{σ}\;\Conid{Γ}\;\Varid{t}\;{}\<[24]%
+\>[24]{}\mathrel{=}\;\Conid{Y}\;\Conid{Γ}\;\Varid{t}{}\<[E]%
+\\
+\>[B]{}\Varid{⟦}\;\Conid{A}\;\Varid{∷}\;\Conid{Δ}\;{}\<[10]%
+\>[10]{}\Varid{⟧ᵃ}\;\Conid{X}\;\Conid{Y}\;\Varid{σ}\;\Conid{Γ}\;\Varid{t}\;{}\<[24]%
+\>[24]{}\mathrel{=}\;\Varid{⟦}\;\Conid{Δ}\;\Varid{⟧ᵃ}\;\Conid{X}\;\Conid{Y}\;\Varid{σ}\;((\Conid{A}\;\Varid{⟨}\;\Varid{σ}\;\Varid{⟩})\;\Varid{∷}\;\Conid{Γ})\;\Varid{t}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+}
+This completes our definition of the signature functor for constructing extrinsic typing derivations.
+
+At last, we can define the type of bidirectional typing derivations for a signature \ensuremath{\Conid{D}} using \ensuremath{\Varid{⟦}\;\Conid{D}\;\Varid{⟧}}, similarly to how we handled simple types and raw terms.
+This involves the following four constructors:
+\LT{help}
+{\small\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{24}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{34}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{55}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{61}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{70}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Keyword{data}\;\Varid{\char95 ⊢\char95 [\char95 ]\char95 }\;\mathbin{:}\;\Conid{Fam}\;\Conid{Raw}\;\Keyword{where}{}\<[E]%
+\\[\blanklineskip]%
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{var}\;{}\<[9]%
+\>[9]{}\mathbin{:}\;(\Varid{i}\;\mathbin{:}\;\Conid{A}\;\Varid{∈}\;\Conid{Γ})\;{}\<[55]%
+\>[55]{}\Varid{\char95 ↑\char95 }\;{}\<[61]%
+\>[61]{}\mathbin{:}\;\Conid{Γ}\;\Varid{⊢}\;\Varid{r}\;{}\<[70]%
+\>[70]{}\Varid{⇒}\;\Conid{B}\;{}\<[E]%
+\\
+\>[9]{}\Varid{→}\;\Conid{L.index}\;\Varid{i}\;\Varid{≡}\;\Varid{j}\;{}\<[61]%
+\>[61]{}\Varid{→}\;\Conid{A}\;\Varid{≡}\;\Conid{B}\;{}\<[E]%
+\\
+\>[9]{}\Varid{→}\;\Conid{Γ}\;\Varid{⊢}\;(\text{\textasciigrave}\;\Varid{j})\;\Varid{⇒}\;\Conid{A}\;{}\<[61]%
+\>[61]{}\Varid{→}\;\Conid{Γ}\;\Varid{⊢}\;\Varid{r}\;{}\<[70]%
+\>[70]{}\Varid{⇐}\;\Conid{A}{}\<[E]%
+\\[\blanklineskip]%
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{\char95 ∋\char95 }\;{}\<[9]%
+\>[9]{}\mathbin{:}\;(\Conid{A}\;\mathbin{:}\;\Conid{Ty})\;{}\<[55]%
+\>[55]{}\Varid{op}\;{}\<[61]%
+\>[61]{}\mathbin{:}\;\Varid{⟦}\;\Conid{D}\;\Varid{⟧}\;\Conid{Raw}\;\Varid{\char95 ⊢\char95 [\char95 ]\char95 }\;\Conid{Γ}\;\Varid{rs}\;\Varid{d}\;\Conid{A}\;{}\<[E]%
+\\
+\>[9]{}\Varid{→}\;\Conid{Γ}\;\Varid{⊢}\;\Varid{r}\;{}\<[24]%
+\>[24]{}\Varid{⇐}\;\Conid{A}\;{}\<[61]%
+\>[61]{}\Varid{→}\;\Conid{Γ}\;\Varid{⊢}\;\Varid{op}\;\Varid{rs}\;[\mskip1.5mu \;\Varid{d}\;\mskip1.5mu]\;\Conid{A}\;{}\<[E]%
+\\
+\>[9]{}\Varid{→}\;\Conid{Γ}\;\Varid{⊢}\;(\Conid{A}\;\Varid{∋}\;\Varid{r})\;{}\<[24]%
+\>[24]{}\Varid{⇒}\;\Conid{A}{}\<[34]%
+\>[34]{}\hspace{2em}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+}
+corresponding to rules $\SynRule{Var}$, $\SynRule{Anno}$, $\ChkRule{Sub}$, and $\Rule{Op}$ in \Cref{fig:extrinsic-typing}.
+
+\subsection{A Formal Proof Example: Soundness}\label{subsec:formal-proofs}
+
+Induction proofs in our formalisation typically consist of three mutual definitions---one for the inductive type of derivations, one for a list of arguments, and one for an extension context.
+
+As an illustration, consider the soundness proof, i.e.\ the `if' part of \Cref{lem:soundness-completeness}.
+Each constructor of bidirectional typing derivations is mapped to a corresponding constructor of typing derivations.
+For the $\SynRule{Sub}$, the induction hypothesis \ensuremath{\Varid{soundness}\;\Varid{t}} is invoked directly from \ensuremath{\Varid{t}\;\Varid{↑}\;\Varid{refl}}.
+\begin{figure}[b]
+  \codefigure
+  \begin{minipage}[b]{.44\textwidth}
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{26}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{30}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{33}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{46}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Keyword{mutual}{}\<[E]%
+\\[\blanklineskip]%
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{soundness}\;\mathbin{:}\;\Conid{Γ}\;\Varid{⊢}\;\Varid{r}\;[\mskip1.5mu \;\Varid{d}\;\mskip1.5mu]\;\Conid{A}\;{}\<[30]%
+\>[30]{}\Varid{→}\;{}\<[33]%
+\>[33]{}\Conid{Γ}\;\Varid{⊢}\;\Varid{r}\;\Varid{⦂}\;\Conid{A}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{soundness}\;(\Varid{var}\;\Varid{i}\;\Varid{eq})\;{}\<[26]%
+\>[26]{}\mathrel{=}\;\Varid{var}\;\Varid{i}\;\Varid{eq}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{soundness}\;(\Conid{A}\;\Varid{∋}\;\Varid{t})\;{}\<[26]%
+\>[26]{}\mathrel{=}\;\Conid{A}\;\Varid{∋}\;\Varid{soundness}\;\Varid{t}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{soundness}\;(\Varid{t}\;\Varid{↑}\;\Varid{refl})\;{}\<[26]%
+\>[26]{}\mathrel{=}\;\Varid{soundness}\;\Varid{t}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{soundness}\;(\Varid{op}\;\Varid{ts})\;{}\<[26]%
+\>[26]{}\mathrel{=}\;{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\Varid{op}\;(\Varid{soundnessᶜ}\;(\Conid{BD}\;\Varid{.rules}\;\anonymous )\;\Varid{ts}){}\<[E]%
+\\[\blanklineskip]%
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{soundnessᶜ}\;{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\mathbin{:}\;(\Conid{D}\;\mathbin{:}\;\Conid{OpD})\;{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\Varid{→}\;\Varid{⟦}\;\Conid{D}\;\Varid{⟧ᶜ}\;\Conid{Raw}\;\Varid{\char95 ⊢\char95 [\char95 ]\char95 }\;\Conid{Γ}\;\Varid{rs}\;\Varid{d}\;\Conid{A}\;{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\Varid{→}\;\Conid{T.⟦}\;\Varid{eraseᶜ}\;\Conid{D}\;\Varid{⟧ᶜ}\;\Conid{Raw}\;\Varid{\char95 ⊢\char95 ⦂\char95 }\;\Conid{Γ}\;\Varid{rs}\;\Conid{A}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{soundnessᶜ}\;(\Varid{ι}\;\anonymous \;\anonymous \;\Conid{Ds})\;(\anonymous \;\Varid{,}\;\Varid{σ}\;\Varid{,}\;\Varid{σ-eq}\;\Varid{,}\;\Varid{ts})\;{}\<[46]%
+\>[46]{}\mathrel{=}\;{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\Varid{σ}\;\Varid{,}\;\Varid{σ-eq}\;\Varid{,}\;\Varid{soundnessᵃˢ}\;\Conid{Ds}\;\Varid{ts}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+\end{minipage}
+\begin{minipage}[b]{.55\textwidth}
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{17}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{23}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{26}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{27}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{35}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{40}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{41}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{45}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{50}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\vphantom{where}{}\<[E]%
+\\[\blanklineskip]%
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{soundnessᵃˢ}\;{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\mathbin{:}\;(\Conid{Ds}\;\mathbin{:}\;\Conid{ArgsD}\;\Conid{Ξ})\;{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\Varid{→}\;\Varid{⟦}\;\Conid{Ds}\;\Varid{⟧ᵃˢ}\;{}\<[27]%
+\>[27]{}\Conid{Raw}\;\Varid{\char95 ⊢\char95 [\char95 ]\char95 }\;{}\<[40]%
+\>[40]{}\Varid{σ}\;\Conid{Γ}\;\Varid{rs}\;{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\Varid{→}\;\Conid{T.⟦}\;\Varid{eraseᵃˢ}\;\Conid{Ds}\;\Varid{⟧ᵃˢ}\;{}\<[27]%
+\>[27]{}\Conid{Raw}\;\Varid{\char95 ⊢\char95 ⦂\char95 }\;{}\<[40]%
+\>[40]{}\Varid{σ}\;\Conid{Γ}\;\Varid{rs}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{soundnessᵃˢ}\;\Varid{[]}\;{}\<[35]%
+\>[35]{}\anonymous \;{}\<[45]%
+\>[45]{}\mathrel{=}\;\Varid{tt}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{soundnessᵃˢ}\;((\Conid{Δ}\;\Varid{⊢[}\;\anonymous \;\mskip1.5mu]\;\anonymous )\;\Varid{∷}\;\Conid{Ds})\;(\Varid{t}\;\Varid{,}\;\Varid{ts})\;{}\<[45]%
+\>[45]{}\mathrel{=}\;{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\Varid{soundnessᵃ}\;\Conid{Δ}\;\Varid{t}\;\Varid{,}\;\Varid{soundnessᵃˢ}\;\Conid{Ds}\;\Varid{ts}{}\<[E]%
+\\[\blanklineskip]%
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{soundnessᵃ}\;{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\mathbin{:}\;(\Conid{Δ}\;\mathbin{:}\;\Conid{TExps}\;\Conid{Ξ})\;{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\Varid{→}\;\Varid{⟦}\;\Conid{Δ}\;\Varid{⟧ᵃ}\;{}\<[17]%
+\>[17]{}\Conid{Raw}\;(\Varid{λ}\;\iden{\Conid{Γ}^\prime}\;\iden{\Varid{r}^\prime}\;\Varid{→}\;\iden{\Conid{Γ}^\prime}\;\Varid{⊢}\;\iden{\Varid{r}^\prime}\;[\mskip1.5mu \;\Varid{d}\;\mskip1.5mu]\;\Conid{A})\;{}\<[50]%
+\>[50]{}\Varid{σ}\;\Conid{Γ}\;\Varid{r}\;{}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\Varid{→}\;\Conid{T.⟦}\;\Conid{Δ}\;\Varid{⟧ᵃ}\;{}\<[17]%
+\>[17]{}\Conid{Raw}\;(\Varid{λ}\;\iden{\Conid{Γ}^\prime}\;\iden{\Varid{r}^\prime}\;\Varid{→}\;\iden{\Conid{Γ}^\prime}\;\Varid{⊢}\;\iden{\Varid{r}^\prime}\;\Varid{⦂}\;\Conid{A})\;{}\<[50]%
+\>[50]{}\Varid{σ}\;\Conid{Γ}\;\Varid{r}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{soundnessᵃ}\;\Varid{[]}\;{}\<[23]%
+\>[23]{}\Varid{t}\;{}\<[26]%
+\>[26]{}\mathrel{=}\;\Varid{soundness}\;{}\<[41]%
+\>[41]{}\Varid{t}{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{soundnessᵃ}\;(\anonymous \;\Varid{∷}\;\Conid{Δ})\;{}\<[23]%
+\>[23]{}\Varid{t}\;{}\<[26]%
+\>[26]{}\mathrel{=}\;\Varid{soundnessᵃ}\;\Conid{Δ}\;\Varid{t}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+\end{minipage}
+\caption{The soundness proof in \Agda}
+\label{fig:formal-soundness}
+\end{figure}
 

### 2023-07-12 tex/sec1-intro.tex
[ tex ] First submitted version

--- 2023-07-12 tex/sec1-intro.tex
@@
 Some examples other than simply typed $\lambda$-calculus are given in \Cref{sec:example}.
-We briefly sketch the design of our formalisation in \cref{sec:formalisation} and conclude in \cref{sec:future} with potential challenges in extending our theory to a more general setting.
+We briefly sketch the design of our formalisation in \cref{sec:formalisation}, and conclude in \cref{sec:future} with potential challenges in further developing our theory.
 

### 2023-07-12 tex/sec6-formalisation.tex
[ tex ] First submitted version

@@
 \>[B]{}\hsindent{3}{}\<[3]%
-\>[3]{}\Keyword{constructor}\;\Varid{ι}{}\<[E]%
+\>[3]{}\Varid{constructor}\;\Varid{ι}{}\<[E]%
 \\
@@
 \>[B]{}\hsindent{3}{}\<[3]%
-\>[3]{}\Keyword{constructor}\;\Varid{sigd}{}\<[E]%
+\>[3]{}\Varid{constructor}\;\Varid{sigd}{}\<[E]%
 \\
@@
 At last, we can define the type of bidirectional typing derivations for a signature \ensuremath{\Conid{D}} using \ensuremath{\Varid{⟦}\;\Conid{D}\;\Varid{⟧}}, similarly to how we handled simple types and raw terms.
-This involves the following four constructors:
-\LT{help}
+This involves the following four constructors, corresponding to rules $\SynRule{Var}$, $\SynRule{Anno}$, $\ChkRule{Sub}$, and $\Rule{Op}$ in \Cref{fig:extrinsic-typing}.
 {\small\begin{hscode}\SaveRestoreHook
@@
 \column{24}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{34}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{55}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{61}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{70}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{32}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{56}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{62}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{71}{@{}>{\hspre}l<{\hspost}@{}}%
 \column{E}{@{}>{\hspre}l<{\hspost}@{}}%
@@
 \>[3]{}\Varid{var}\;{}\<[9]%
-\>[9]{}\mathbin{:}\;(\Varid{i}\;\mathbin{:}\;\Conid{A}\;\Varid{∈}\;\Conid{Γ})\;{}\<[55]%
-\>[55]{}\Varid{\char95 ↑\char95 }\;{}\<[61]%
-\>[61]{}\mathbin{:}\;\Conid{Γ}\;\Varid{⊢}\;\Varid{r}\;{}\<[70]%
-\>[70]{}\Varid{⇒}\;\Conid{B}\;{}\<[E]%
+\>[9]{}\mathbin{:}\;(\Varid{i}\;\mathbin{:}\;\Conid{A}\;\Varid{∈}\;\Conid{Γ})\;{}\<[32]%
+\>[32]{}\hspace{10em}{}\<[56]%
+\>[56]{}\Varid{\char95 ↑\char95 }\;{}\<[62]%
+\>[62]{}\mathbin{:}\;\Conid{Γ}\;\Varid{⊢}\;\Varid{r}\;{}\<[71]%
+\>[71]{}\Varid{⇒}\;\Conid{B}\;{}\<[E]%
 \\
-\>[9]{}\Varid{→}\;\Conid{L.index}\;\Varid{i}\;\Varid{≡}\;\Varid{j}\;{}\<[61]%
-\>[61]{}\Varid{→}\;\Conid{A}\;\Varid{≡}\;\Conid{B}\;{}\<[E]%
+\>[9]{}\Varid{→}\;\Conid{L.index}\;\Varid{i}\;\Varid{≡}\;\Varid{j}\;{}\<[62]%
+\>[62]{}\Varid{→}\;\Conid{A}\;\Varid{≡}\;\Conid{B}\;{}\<[E]%
 \\
-\>[9]{}\Varid{→}\;\Conid{Γ}\;\Varid{⊢}\;(\text{\textasciigrave}\;\Varid{j})\;\Varid{⇒}\;\Conid{A}\;{}\<[61]%
-\>[61]{}\Varid{→}\;\Conid{Γ}\;\Varid{⊢}\;\Varid{r}\;{}\<[70]%
-\>[70]{}\Varid{⇐}\;\Conid{A}{}\<[E]%
+\>[9]{}\Varid{→}\;\Conid{Γ}\;\Varid{⊢}\;(\text{\textasciigrave}\;\Varid{j})\;\Varid{⇒}\;\Conid{A}\;{}\<[62]%
+\>[62]{}\Varid{→}\;\Conid{Γ}\;\Varid{⊢}\;\Varid{r}\;{}\<[71]%
+\>[71]{}\Varid{⇐}\;\Conid{A}{}\<[E]%
 \\[\blanklineskip]%
@@
 \>[3]{}\Varid{\char95 ∋\char95 }\;{}\<[9]%
-\>[9]{}\mathbin{:}\;(\Conid{A}\;\mathbin{:}\;\Conid{Ty})\;{}\<[55]%
-\>[55]{}\Varid{op}\;{}\<[61]%
-\>[61]{}\mathbin{:}\;\Varid{⟦}\;\Conid{D}\;\Varid{⟧}\;\Conid{Raw}\;\Varid{\char95 ⊢\char95 [\char95 ]\char95 }\;\Conid{Γ}\;\Varid{rs}\;\Varid{d}\;\Conid{A}\;{}\<[E]%
+\>[9]{}\mathbin{:}\;(\Conid{A}\;\mathbin{:}\;\Conid{Ty})\;{}\<[56]%
+\>[56]{}\Varid{op}\;{}\<[62]%
+\>[62]{}\mathbin{:}\;\Varid{⟦}\;\Conid{D}\;\Varid{⟧}\;\Conid{Raw}\;\Varid{\char95 ⊢\char95 [\char95 ]\char95 }\;\Conid{Γ}\;\Varid{rs}\;\Varid{d}\;\Conid{A}\;{}\<[E]%
 \\
 \>[9]{}\Varid{→}\;\Conid{Γ}\;\Varid{⊢}\;\Varid{r}\;{}\<[24]%
-\>[24]{}\Varid{⇐}\;\Conid{A}\;{}\<[61]%
-\>[61]{}\Varid{→}\;\Conid{Γ}\;\Varid{⊢}\;\Varid{op}\;\Varid{rs}\;[\mskip1.5mu \;\Varid{d}\;\mskip1.5mu]\;\Conid{A}\;{}\<[E]%
+\>[24]{}\Varid{⇐}\;\Conid{A}\;{}\<[62]%
+\>[62]{}\Varid{→}\;\Conid{Γ}\;\Varid{⊢}\;\Varid{op}\;\Varid{rs}\;[\mskip1.5mu \;\Varid{d}\;\mskip1.5mu]\;\Conid{A}\;{}\<[E]%
 \\
 \>[9]{}\Varid{→}\;\Conid{Γ}\;\Varid{⊢}\;(\Conid{A}\;\Varid{∋}\;\Varid{r})\;{}\<[24]%
-\>[24]{}\Varid{⇒}\;\Conid{A}{}\<[34]%
-\>[34]{}\hspace{2em}{}\<[E]%
+\>[24]{}\Varid{⇒}\;\Conid{A}{}\<[E]%
 \ColumnHook
@@
 }
-corresponding to rules $\SynRule{Var}$, $\SynRule{Anno}$, $\ChkRule{Sub}$, and $\Rule{Op}$ in \Cref{fig:extrinsic-typing}.
 
@@
 
-Induction proofs in our formalisation typically consist of three mutual definitions---one for the inductive type of derivations, one for a list of arguments, and one for an extension context.
-
-As an illustration, consider the soundness proof, i.e.\ the `if' part of \Cref{lem:soundness-completeness}.
-Each constructor of bidirectional typing derivations is mapped to a corresponding constructor of typing derivations.
-For the $\SynRule{Sub}$, the induction hypothesis \ensuremath{\Varid{soundness}\;\Varid{t}} is invoked directly from \ensuremath{\Varid{t}\;\Varid{↑}\;\Varid{refl}}.
-\begin{figure}[b]
+\begin{figure}
   \codefigure
@@
 
+Induction proofs in our formalisation typically consist of three mutual definitions---one for the inductive type of derivations, one for a list of arguments, and one for an extension context.
+
+As an illustration, consider the soundness proof in \cref{fig:formal-soundness}, i.e.\ the `if' part of \cref{lem:soundness-completeness}.
+Each constructor of bidirectional typing derivations is mapped to a corresponding constructor of typing derivations.
+For the $\SynRule{Sub}$ case, the induction hypothesis \ensuremath{\Varid{soundness}\;\Varid{t}} is invoked directly.
+
 \end{document}

### 2023-07-12 tex/sec6-formalisation.tex
[ tex ] Clean up Formalisation

@@
 The formal definitions of extrinsic typing are more complex than their intrinsic counterparts.
-For a specific language such as \PCF, \citet{Wadler2022} noted that `extrinsically-typed terms require about 1.6 times as much code as intrinsically-typed' for their formalisation of type safety.
-For our generic definitions, take the formal type-theoretic definition of typing derivations $\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A$ as an example.
-Its intrinsic definition consists of of just one family of sets of intrinsically-typed terms indexed by $A$ and $\Gamma$, but its extrinsic counterpart is a generic family of sets indexed by additionally a generic raw term $t$, involving constructions of two different layers.
+%For a specific language such as \PCF, \citet{Wadler2022} noted that `extrinsically-typed terms require about 1.6 times as much code as intrinsically-typed' for their formalisation of type safety.
+%For our generic definitions,
+Take our formal definition of typing derivations $\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A$ as an example.
+Its intrinsic definition is just one family of sets of typed terms indexed by $A$ and $\Gamma$, but its extrinsic counterpart is a generic family of sets indexed by additionally a generic raw term $t$, involving constructions of two different layers.
 
 \paragraph{Category-theoretic analysis of well-typed terms}
-The difference between intrinsic typing and extrinsic typing could be partly manifested by its categorical analysis.
-\citet{Fiore1999}{'s} theory of abstract syntax and variable binding forms the foundation of \citet{Fiore2022}'s framework and inspires frameworks by \citet{Allais2021,Ahrens2022}.
+The difference between intrinsic typing and extrinsic typing could be partly manifested by \citet{Fiore1999}{'s} theory of abstract syntax and variable binding which forms the foundation of \citet{Fiore2022}'s framework and inspires other referenced frameworks.
 Let us take a closer look of their idea.
@@
 \end{enumerate*}
-Substitution is modelled as a monoid multiplication with the notion of strength and a suitable monoidal structure on the category of presheaves.
+Substitution for $\Term$ is modelled as a monoid in the category of presheaves.
 To put it succinctly, it is the free $\mathbb{\Sigma}$-monoid over the presheaf~$V$ of variables.
 
-Such an initial algebra can be constructed as the least fixpoint of a sequence of objects defined via iteratively applying the signature functor by the Initial Algebra Theorem~\cite{Trnkova1975}.
+Such an initial algebra can be constructed by the Initial Algebra Theorem~\cite{Trnkova1975}.
 
@@
 We start with the formal definition of signatures and simple types (\Cref{fig:formal-simple-type}), which parallels its informal counterpart (\Cref{def:simple-signature}), with the exception of \ensuremath{\Varid{⟦\char95 ⟧}}.
-To prevent circular references during the inductive definition of simple types, we initially define \ensuremath{\Varid{⟦}\;\Conid{D}\;\Varid{⟧}} on an arbitrary \ensuremath{\Conid{Set}} instead of \ensuremath{\Conid{Ty}}.
-Thus, types specified by a signature can be defined using two constructors--- \ensuremath{\text{\textasciigrave}\anonymous }\ and \ensuremath{\Varid{op}}---where a variable is represented by \ensuremath{\text{\textasciigrave}\;\Varid{n}} for some \ensuremath{\Varid{n}} less than \ensuremath{\Conid{Ξ}}, and each inhabitant \ensuremath{\Varid{op}\;(\Varid{i}\;\Varid{,}\;\Varid{ts})} is a pair of an operation symbol \ensuremath{\Varid{i}} and a list \ensuremath{\Varid{ts}} of length $n$.
+To prevent circular references during the inductive definition of simple types, we initially define \ensuremath{\Varid{⟦}\;\Conid{D}\;\Varid{⟧}} on an arbitrary \ensuremath{\Conid{Set}} instead of\/ \ensuremath{\Conid{Ty}}.
+Thus, types specified by a signature can be defined inductively where each inhabitant \ensuremath{\Varid{op}\;(\Varid{i}\;\Varid{,}\;\Varid{ts})} for the $\Rule{Op}$ rule is a pair of an operation symbol and a list of length $n$.
 
-From a categorical standpoint, \ensuremath{\Varid{⟦}\;\Conid{D}\;\Varid{⟧}} acts as a functor from \ensuremath{\Conid{Set}} to \ensuremath{\Conid{Set}}, mapping $X$ to an $\Conid{Op}$-indexed coproduct $\sum_{i \in \Conid{Op}} X ^ {\Conid{ar}(i)}$ of products of $X$.
+From a categorical view, \ensuremath{\Varid{⟦}\;\Conid{D}\;\Varid{⟧}} is a functor from \ensuremath{\Conid{Set}} to \ensuremath{\Conid{Set}}, mapping $X$ to an $\Conid{Op}$-indexed coproduct $\sum_{i \in \Conid{Op}} X ^ {\Conid{ar}(i)}$ of products of $X$.
 The type \ensuremath{\Conid{Ty}} is the free \ensuremath{\Varid{⟦}\;\Conid{D}\;\Varid{⟧}}-algebra over the type \ensuremath{\Conid{Fin}\;\Conid{Ξ}} or the initial algebra for the functor \ensuremath{\Conid{Fin}\;\Conid{Ξ}\;\Varid{+}\;\Varid{⟦}\;\Conid{D}\;\Varid{⟧}}.
 
-Finally, to account for indices when defining raw terms and bidirectional typing derivations by a signature, we define a corresponding signature functor \ensuremath{\Varid{⟦}\;\Conid{D}\;\Varid{⟧}} for a (bidirectional) binding signature.
+\subsection{Defining Raw Terms}\label{subsec:formal-raw-terms}
 
-\subsection{Defining Raw Terms}\label{subsec:formal-raw-terms}
 As shown in \Cref{fig:formal-bibisig}, we establish \ensuremath{\Conid{ArgD}}, \ensuremath{\Conid{OpD}}, and \ensuremath{\Conid{SigD}} to represent arguments $[\Delta]A^{\dir{d}}$, bidirectional binding operations $\biop$, and bidirectional binding signatures, respectively, in line with their informal definitions.
@@
 \>[B]{}\hsindent{3}{}\<[3]%
-\>[3]{}\Varid{constructor}\;\Varid{ι}{}\<[E]%
+\>[3]{}\Keyword{constructor}\;\Varid{ι}{}\<[E]%
 \\
@@
 \>[B]{}\hsindent{3}{}\<[3]%
-\>[3]{}\Varid{constructor}\;\Varid{sigd}{}\<[E]%
+\>[3]{}\Keyword{constructor}\;\Varid{sigd}{}\<[E]%
 \\
@@
 \end{figure}
+
 To construct well-scoped raw terms indexed by a list $V$ of free variables, we introduce another signature functor mapping \ensuremath{\Conid{Fam}} to \ensuremath{\Conid{Fam}}, where \ensuremath{\Conid{Fam}} is the family of sets defined as \ensuremath{\Conid{ℕ}\;\Varid{→}\;\Conid{Set}}, indexed by a natural number indicating the number of free variables.
-This signature endofunctor for raw term construction resembles the one for defining simple types.
-However, we define four distinct functors instead of defining one directly.
+This functor for raw term construction resembles the one for defining simple types.
+However, we define four distinct functors instead of directly defining one .
 These are \ensuremath{\Varid{⟦}\;\Conid{Δ}\;\Varid{⟧ᵃ}} for an extension context \ensuremath{\Conid{Δ}}, \ensuremath{\Varid{⟦}\;\Conid{Ds}\;\Varid{⟧ᵃˢ}} for an argument list \ensuremath{\Conid{Ds}}, \ensuremath{\Varid{⟦}\;\Varid{ι}\;\Conid{Ds}\;\anonymous \;\Varid{⟧ᶜ}} for an operation (\ensuremath{\Varid{ι}\;\Conid{Ds}\;\Conid{A}}), and \ensuremath{\Varid{⟦}\;\Conid{D}\;\Varid{⟧ᶜ}} for a binding signature $D$.
-Notably, the functor \ensuremath{\Varid{⟦}\;\Conid{Δ}\;\Varid{⟧ᵃ}} extends the context by $\erase{\Delta}$ variables by mapping the family of sets \ensuremath{\Conid{X}\;\Varid{n}} to \ensuremath{\Conid{X}\;(\Varid{length}\;\Conid{Δ}\;\Varid{+}\;\Varid{n})}.
+Notably, the functor \ensuremath{\Varid{⟦}\;\Conid{Δ}\;\Varid{⟧ᵃ}} extends the context by mapping the family of sets \ensuremath{\Conid{X}\;\Varid{n}} to \ensuremath{\Conid{X}\;(\Varid{length}\;\Conid{Δ}\;\Varid{+}\;\Varid{n})}.
 \begin{figure}
@@
 \subsection{Defining Extrinsic Bidirectional Typing Derivations}\label{subsec:formal-bidirectional-typing}
-Formally defining extrinsic derivations is arguably the most complex aspect of our formalisation.
-This requires defining a family of sets, indexed by context, type, mode, and notably, a raw term.
-This definition involves another signature endofunctor, requiring consistency with the functor used for constructing raw terms.
+Formally defining extrinsic derivations is much trickier.
+This requires defining a family of sets, indexed by context, type, mode, and notably, a raw term, involving another signature endofunctor alone with the functor used for constructing raw terms.
 
@@
 In a similar vein, we define four functors for the construction of bidirectional typing derivations, albeit in a top-down manner.
-For an endofunctor on predicates, each functor must act as a \emph{lifting} of the functor on its domain, illustrated below.
-The signature endofunctor \ensuremath{\Varid{⟦\char95 ⟧}} for a bidirectional binding signature is defined as follows:
+An endofunctor on predicates must act as a \emph{lifting} of the functor on its domain, illustrated by the signature endofunctor \ensuremath{\Varid{⟦\char95 ⟧}} for a bidirectional binding signature:
 {\small
@@
 
-The functor \ensuremath{\Varid{⟦}\;\Conid{D}\;\Varid{.ar}\;\Varid{o}\;\Varid{⟧ᶜ}} is defined for an operation symbol $\biop$:
+The functor \ensuremath{\Varid{⟦}\;\Conid{D}\;\Varid{.ar}\;\Varid{o}\;\Varid{⟧ᶜ}} is defined for an operation $\biop$:
 {\small\begin{hscode}\SaveRestoreHook
@@
 
-The signature endofunctor for an argument list \ensuremath{\Conid{Ds}\;\mathbin{:}\;\Conid{ArgsD}\;\Conid{Ξ}} is defined in a manner that it maps a predicate to a family of sets indexed by a context and a family of product sets, considering that each premise in a typing derivation shares a common context $\Gamma$ and a term from a list of subterms.
+The signature endofunctor for an argument list \ensuremath{\Conid{Ds}} is defined in a manner that it maps a predicate to a family of sets indexed by a context and a family of product sets, considering that each premise in a typing derivation shares a common context $\Gamma$ and a term from a list of subterms.
 {\small\begin{hscode}\SaveRestoreHook
 \column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
 \column{8}{@{}>{\hspre}l<{\hspost}@{}}%
@@
 \>[36]{}(\Varid{x}\;\Varid{,}\;\Varid{xs})\;{}\<[46]%
-\>[46]{}\mathrel{=}\;\Varid{⟦}\;\Conid{Δ}\;\Varid{⟧ᵃ}\;\Conid{X}\;(\Varid{λ}\;\iden{\Conid{Γ}^\prime}\;\iden{\Varid{x}^\prime}\;\Varid{→}\;\Conid{Y}\;\iden{\Conid{Γ}^\prime}\;\iden{\Varid{x}^\prime}\;\Varid{d}\;(\Conid{A}\;\Varid{⟨}\;\Varid{σ}\;\Varid{⟩}))\;\Varid{σ}\;\Conid{Γ}\;\Varid{x}\;\Varid{×}\;\Varid{⟦}\;\Conid{Ds}\;\Varid{⟧ᵃˢ}\;\Conid{X}\;\Conid{Y}\;\Varid{σ}\;\Conid{Γ}\;\Varid{xs}{}\<[E]%
+\>[46]{}\mathrel{=}\;{}\<[E]%
+\\
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{⟦}\;\Conid{Δ}\;\Varid{⟧ᵃ}\;\Conid{X}\;(\Varid{λ}\;\iden{\Conid{Γ}^\prime}\;\iden{\Varid{x}^\prime}\;\Varid{→}\;\Conid{Y}\;\iden{\Conid{Γ}^\prime}\;\iden{\Varid{x}^\prime}\;\Varid{d}\;(\Conid{A}\;\Varid{⟨}\;\Varid{σ}\;\Varid{⟩}))\;\Varid{σ}\;\Conid{Γ}\;\Varid{x}\;\Varid{×}\;\Varid{⟦}\;\Conid{Ds}\;\Varid{⟧ᵃˢ}\;\Conid{X}\;\Conid{Y}\;\Varid{σ}\;\Conid{Γ}\;\Varid{xs}{}\<[E]%
 \ColumnHook
@@
 
-At last, we can define the type of bidirectional typing derivations for a signature \ensuremath{\Conid{D}} using \ensuremath{\Varid{⟦}\;\Conid{D}\;\Varid{⟧}}, similarly to how we handled simple types and raw terms.
+At last, we can define the type of bidirectional typing derivations for a signature \ensuremath{\Conid{D}}, similarly to how we have defined simple types and raw terms.
 This involves the following four constructors, corresponding to rules $\SynRule{Var}$, $\SynRule{Anno}$, $\ChkRule{Sub}$, and $\Rule{Op}$ in \Cref{fig:extrinsic-typing}.

### 2023-07-20 tex/example.tex
[ tex ] Conclude Formalisation properly; minor changes to save more space

--- 2023-07-12 tex/example.tex
@@
 \end{enumerate}
-The first is a standard exercise in implementing a type synthesiser and the second is an example of a system with infinitely many operations.
+The first is a standard exercise in implementing a type synthesiser.
+The second example includes not only infinitely many operations but also a type which is not formed by solely one type construct~$\tyOp$.
 
@@
 Although we have developed a theory for presenting a bidirectional type system and a condition for the decidability of bidirectional type synthesis, designing a bidirectional type system, or its signature, remains a creative activity.\footnote{This has also been the case for parsing: while there are grammars and parser generators, one still has to design a grammar to parse or disambiguate an ambiguous grammar.}
-While the focus of this paper is not to design bidirectional type systems, we adopt the example of a bidirectional \PCF from \citet{Wadler2022}, which extends bidirectional simply typed $\lambda$-calculus in \Cref{sec:key-ideas} by adding the following rules:
+While the focus of this paper is not to design bidirectional type systems, we simply adopt the example of a bidirectional \PCF from \citet{Wadler2022}, which extends bidirectional simply typed $\lambda$-calculus in \Cref{sec:key-ideas} by adding the following rules:
 \bgroup
@@
   \begin{mathpar}
-    \inferrule{\Gamma \vdash t \syn \textstyle{\prod_{i = 1}^{n}} A_i \bto B \and \Gamma \vdash u_1 \chk A_1 \and \dots \and \Gamma \vdash u_n \chk A_n}{\Gamma \vdash t\;u_1\;\ldots\;u_n \syn B}
+    \inferrule{\Gamma \vdash t \syn A_1 \bto \left(A_2 \bto \left(\dots \bto \left(A_n \bto B\right)\ldots\right)\right) \and \Gamma \vdash u_1 \chk A_1 \and \dots \and \Gamma \vdash u_n \chk A_n}{\Gamma \vdash t\;u_1\;\ldots\;u_n \syn B}
   \end{mathpar}
 \egroup
-The type signature $\Sigma_{\bto}$ can be extended with $\mathsf{fun}_n$ where $\arity({\mathsf{fun}_n}) = n$ for each $n : \N$.
-Similarly, the bidirectional binding signature $\Omega_{\Lambda}^{\Leftrightarrow}$ can be extended with 
+We use the same type signature $\Sigma_{\bto}$ but extend the bidirectional binding signature $\Omega_{\Lambda}^{\Leftrightarrow}$ with 
 \[
-  \aritysymbol{\mathsf{app}_n}{A_1,\ldots, A_n, B}{\tyOp_{\mathsf{fun}_n}(A_1, \ldots, A_n) \bto B^{\syn}, A_1^{\chk}, \ldots, A_n^{\chk}}{B}
+  \aritysymbol{\mathsf{app}_n}{A_1,\ldots, A_n, B}{A_1 \bto \left(A_2 \bto \left(\dots \bto \left(A_n \bto B\right)\ldots\right)\right)^{\syn}, A_1^{\chk}, \ldots, A_n^{\chk}}{B}
 \]
-also for each $n : \N$.
+for each $n : \N$.
 Then each spine application $t\;u_1\;\ldots\;u_n$ can be introduced as $\tmOp_{\mathsf{app}_n}(t; u_1; \ldots; u_n)$.

### 2023-07-20 tex/sec3-defs.tex
[ tex ] Conclude Formalisation properly; minor changes to save more space

--- 2023-07-12 tex/sec3-defs.tex
@@
 %Here's the 'small print': Our definitions cover typical examples and set the scope for bidirectional simple type systems discussed, but do not offer comprehensive coverage of all possibilities.}
-of simple types, simply typed languages, and bidirectional type systems, using simply typed $\lambda$-calculus in \cref{sec:key-ideas} as our running example.
-(These definitions may look dense, especially on first reading.
-The reader may choose to skim through this section, in particular the figures, and still get some rough ideas from later sections.)
+of simple types, simply typed languages, and bidirectional type systems, and uses simply typed $\lambda$-calculus in \cref{sec:key-ideas} as our running example.
+These definitions may look dense, especially on first reading.
+The reader may choose to skim through this section, in particular the figures, and still get some rough ideas from later sections.
 
@@
 \begin{enumerate*}
-  \item first we introduce a notion of arity and a notion of signature which includes a set of operation symbols and an assignment of arities to symbols;
-\item then, given a signature, we define raw terms and typing derivations by primitive rules such as $\Rule{Var}$ and a rule schema for constructs $\tmOp_o$ indexed by an operation symbol~$o$.
+  \item first we introduce a notion of arity and a notion of signature which includes a set\footnote{%
+    Even though our theory is developed in Martin-L\"of type theory, the term `set' is used instead of `type' to avoid the obvious confusion. 
+    Also, as we assume \AxiomK, all types are legitimately sets in the sense of homotopy type theory.
+  } of operation symbols and an assignment of arities to symbols;
+\item then, given a signature, we define raw terms and typing derivations inductively by primitive rules such as $\Rule{Var}$ and a rule schema for constructs $\tmOp_o$ indexed by an operation symbol~$o$.
 \end{enumerate*}
-All these definitions are inductive as usual but include a \emph{rule schema} giving rise to as many rules as there are operation symbols in a signature.
+%All these definitions are inductive as usual but include a \emph{rule schema} giving rise to as many rules as there are operation symbols in a signature.
 As we move from simple types to bidirectional typing, the notion of arity, initially as the number of arguments of an operation, is enriched to incorporate an extension context for variable binding and the mode for the direction of type information flow.
@@
   \end{figure}
-  A \emph{signature} $\Sigma$ for simple types consists of a set\footnote{%
-    Even though our theory is developed in Martin-L\"of type theory, the term `set' is used instead of `type' to avoid the obvious confusion. 
-    Also, as we assume \AxiomK, all types are legitimate sets in the sense of homotopy type theory.
-  }
-  $I$ with a decidable equality and an \emph{arity} function $\arity\colon I \to \mathbb{N}$.
+  A \emph{signature} $\Sigma$ for simple types consists of a set $I$ with a decidable equality and an \emph{arity} function $\arity\colon I \to \mathbb{N}$.
   For a signature $\Sigma$, a \emph{type} $A : \Type_{\Sigma}(\Xi)$ over a variable set $\Xi$ is either
@@
 \begin{example} \label{ex:type-signature-for-function-type}
-  Simply typed $\lambda$-calculus includes function types $A \bto B$ and typically a base type~$\mathtt{b}$ to ensure that the set of all types without type variables is non-empty.
-  The type signature $\Sigma_{\bto}$ used to define types in simply typed $\lambda$-calculus consists of a binary operation $\mathsf{fun}$ and a nullary operation $\mathsf{b}$ where $\arity(\mathsf{fun}) = 2$ and $\arity(\mathsf{b}) = 0$.
+  Function types $A \bto B$ and typically a base type~$\mathtt{b}$ are included in simply typed $\lambda$-calculus and can be specified by the type signature $\Sigma_{\bto}$ consisting of operations $\mathsf{fun}$ and $\mathsf{b}$ where $\arity(\mathsf{fun}) = 2$ and $\arity(\mathsf{b}) = 0$.
   Then, all types in simply typed $\lambda$-calculus can be given as $\Sigma_{\bto}$-types over the empty set with $A \bto B$ introduced as $\tyOp_{\mathsf{fun}}(A, B)$ and $\mathtt{b}$ as $\tyOp_{\mathsf{b}}$.

### 2023-07-20 tex/sec4-pre-synthesis.tex
[ tex ] Conclude Formalisation properly; minor changes to save more space

@@
 
-\citet[Section~3.2]{Dunfield2021} formulated completeness differently from ours (\cref{thm:STLC-completeness,sec:soundness-and-completeness}) and proposed \emph{annotatability} as a more suitable name.
+\citet[Section~3.2]{Dunfield2021} formulated completeness differently from our \cref{lem:soundness-completeness} and proposed \emph{annotatability} as a more suitable name.
 In our theory, we may formulate annotatability as follows.

### 2023-07-20 tex/sec5-synthesis.tex
[ tex ] Conclude Formalisation properly; minor changes to save more space

--- 2023-07-12 tex/sec5-synthesis.tex
@@
 \begin{definition}\label{def:mode-correctness-args}
-  The \emph{mode-correctness} $\MCas\left(\biargvec\right)$ for an argument list $\biargs$ with respect to a subset $S$ of $\Xi$ is defined by
+  The \emph{mode-correctness} $\MCas\left(\biargvec\right)$ for an argument list $\biargs$ with respect to a subset $S$ of $\Xi$ is a (proof-relevant) predicate defined by \LT{is it a prop?}
   \begin{align}
@@
 \begin{remark}
-  Mode-correctness is fundamentally a condition for bidirectional typing \emph{rules} themselves, not for derivations.
+  Mode-correctness is fundamentally a condition for bidirectional typing \emph{rules}, not for derivations.
   Thus, this property cannot be formalised without treating rules as some mathematical object, such as the notion of bidirectional binding signature presented in \Cref{sec:defs}.
@@
   \begin{enumerate}
-    \item if\/ $\erase{\Gamma} |-_{\Sigma, \Omega} \isTerm{t}^{\syn}$, then it is decidable whether there is~$A$ such that
-      \[
-        \Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn A;
-      \]
-    \item if\/ $\erase{\Gamma} |-_{\Sigma, \Omega} \isTerm{t}^{\chk}$, then it is decidable for any~$A$ whether
-      \[
-        \Gamma |-_{\Sigma, \Omega} \isTerm{t} \chk A.
-      \]
+    \item if\/ $\erase{\Gamma} |-_{\Sigma, \Omega} \isTerm{t}^{\syn}$, then it is decidable whether there is~$A$ such that $\Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn A$;
+    \item if\/ $\erase{\Gamma} |-_{\Sigma, \Omega} \isTerm{t}^{\chk}$, then it is decidable for any~$A$ whether $\Gamma |-_{\Sigma, \Omega} \isTerm{t} \chk A$.
   \end{enumerate}
@@
   We prove this statement by induction on the mode derivation\/ $\erase{\Gamma} |-_{\Sigma, \Omega} \isTerm{t}^{\dir{d}}$.
-  The two cases \SynRule{Var} and \SynRule{Anno} are straightforward and have nothing to do with mode-correctness.
-  The case \ChkRule{Sub} invokes the uniqueness of synthesised types~(\cref{thm:unique-syn}) to refute the case that $\Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn B$ but $A \neq B$ for a given type $A$.
+  The two cases \SynRule{Var} and \SynRule{Anno} are straightforward and independent of mode-correctness.
+  The case \ChkRule{Sub} invokes the uniqueness of synthesised types to refute the case that $\Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn B$ but $A \neq B$ for a given type $A$.
   The first three cases follow essentially the same reasoning provided by \citet{Wadler2022}, but we still present the reasoning here for the sake of completeness.
@@
 
-    \item \ChkRule{Sub}: If $\erase{\Gamma} |-t^{\chk}$ by the rule $\ChkRule{Sub}$, then $\Gamma |- t^{\syn}$ by inversion.
-      By induction hypothesis, it is decidable whether $\Gamma |- t \syn B$ for some $B$:
+    \item \ChkRule{Sub}: If $\erase{\Gamma} |-t^{\chk}$, then $\Gamma |- t^{\syn}$ by inversion.
+      By induction hypothesis, we have two cases: %it is decidable whether $\Gamma |- t \syn B$ for some $B$:
       \begin{itemize}
@@
                         Therefore $\simsub{\Delta_{m+1}}{\ext{\rho}} = \simsub{\Delta_{m+1}}{\gamma}$ and we derive a contradiction because $\Gamma, \vars{x}_\isTerm{m+1} : \simsub{\Delta_{m+1}}{\ext{\rho}} |- \isTerm{t_{m+1}} \syn \simsub{A_{m+1}}{\gamma}$.
-                      \item If $\Gamma, \vars{x}_\isTerm{m+1} : \simsub{\Delta_{m+1}}{\ext{\rho}} |- \isTerm{t_{m+1}} \syn A$ for some $A$, then by \Cref{lem:unify} we have two cases: %we unify $A$ with $A_{m+1}$ extending $\ext{\rho}$.
+                      \item If $\Gamma, \vars{x}_\isTerm{m+1} : \simsub{\Delta_{m+1}}{\ext{\rho}} |- \isTerm{t_{m+1}} \syn A$ for some $A$, then by \Cref{lem:unify}: % we have two cases: %we unify $A$ with $A_{m+1}$ extending $\ext{\rho}$.
                         \begin{itemize}
@@
   exactly one of the following holds:
+  \LT{Why downplay the pre-condition?}
   \begin{enumerate}
@@
 \end{corollary}
-\begin{proof}
-  Combine \cref{thm:bidirectional-type-synthesis-checking} with \cref{lem:soundness-completeness,thm:mode-decoration}.
-\end{proof}

### 2023-07-20 tex/sec6-formalisation.tex
[ tex ] Conclude Formalisation properly; minor changes to save more space

@@
 As we have mentioned in \Cref{sec:intro}, our theory was initially developed with \Agda.
-While the translation to the natural language is reasonably straightforward, understanding the formalisation itself could pose some difficulty.
+While the translation to the natural language is reasonably straightforward, understanding the formalisation itself could be difficult.
 If the reader is comfortable with the informal presentation and assured by the existence of its formalisation, they may feel free to skip this section.
@@
   \item a family of sets $\Term_{\Gamma}$ of well-scoped terms under a context~$\Gamma$ with
-  \item variable renaming for a function $\sigma\colon \Gamma \to \Delta$ between variables acting as a functorial map from $\Term_{\Gamma}$ to $\Term_{\Delta}$, i.e.\ a presheaf $\Term\colon \mathbb{F} \to \mathsf{Set}$, with
+  \item variable renaming for a function $\sigma\colon \Gamma \to \Delta$ between variables acting as a functorial map from $\Term_{\Gamma}$ to $\Term_{\Delta}$, i.e.\ a presheaf $\Term\colon \mathbb{F} \to \mathsf{Set}$, and
   \item an \emph{initial} algebra $[\mathsf{v}, \mathsf{op}]$ on~$\Term$ given by the variable rule as a map $\mathsf{v}$ from the presheaf $V\colon \mathbb{F} \hookrightarrow \mathsf{Set}$ of variables to $\Term$ and other constructs as $\mathsf{op}\colon \mathbb{\Sigma}\Term \to \Term$ where $\mathbb{F}$ is the category of contexts and the functor $\mathbb{\Sigma}\colon \mathsf{Set}^\mathbb{F} \to \mathsf{Set}^\mathbb{F}$ encodes the binding arities of constructs.
@@
 
-As shown in \Cref{fig:formal-bibisig}, we establish \ensuremath{\Conid{ArgD}}, \ensuremath{\Conid{OpD}}, and \ensuremath{\Conid{SigD}} to represent arguments $[\Delta]A^{\dir{d}}$, bidirectional binding operations $\biop$, and bidirectional binding signatures, respectively, in line with their informal definitions.
+As shown in \Cref{fig:formal-bibisig}, we define \ensuremath{\Conid{ArgD}}, \ensuremath{\Conid{OpD}}, and \ensuremath{\Conid{SigD}} to represent arguments $[\Delta]A^{\dir{d}}$, bidirectional binding operations $\biop$, and bidirectional binding signatures, respectively, in line with their informal definitions.
 By removing \ensuremath{\Varid{mode}} from these definitions, we retrieve the definitions of binding arguments, arities, and signatures.
@@
 \>[B]{}\Varid{⟦\char95 ⟧ᶜ}\;{}\<[7]%
-\>[7]{}\mathbin{:}\;(\Conid{D}\;\mathbin{:}\;\Conid{OpD})\;(\Conid{X}\;\mathbin{:}\;\Conid{R.Fam}\;\Varid{ℓ})\;(\Conid{Y}\;\mathbin{:}\;\Conid{Fam}\;\iden{\Varid{ℓ}^\prime}\;\Conid{X})\;{}\<[E]%
-\\
-\>[7]{}\Varid{→}\;\Conid{Fam}\;\iden{\Varid{ℓ}^\prime}\;(\Conid{R.⟦}\;\Varid{eraseᶜ}\;\Conid{D}\;\Varid{⟧ᶜ}\;\Conid{X}){}\<[E]%
+\>[7]{}\mathbin{:}\;(\Conid{D}\;\mathbin{:}\;\Conid{OpD})\;(\Conid{X}\;\mathbin{:}\;\Conid{R.Fam}\;\Varid{ℓ})\;(\Conid{Y}\;\mathbin{:}\;\Conid{Fam}\;\iden{\Varid{ℓ}^\prime}\;\Conid{X})\;\Varid{→}\;\Conid{Fam}\;\iden{\Varid{ℓ}^\prime}\;(\Conid{R.⟦}\;\Varid{eraseᶜ}\;\Conid{D}\;\Varid{⟧ᶜ}\;\Conid{X}){}\<[E]%
 \\
@@
 \column{9}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{15}{@{}>{\hspre}l<{\hspost}@{}}%
 \column{24}{@{}>{\hspre}l<{\hspost}@{}}%
@@
 \column{71}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{85}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{89}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{108}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{113}{@{}>{\hspre}l<{\hspost}@{}}%
 \column{E}{@{}>{\hspre}l<{\hspost}@{}}%
@@
 \>[62]{}\Varid{→}\;\Conid{Γ}\;\Varid{⊢}\;\Varid{r}\;{}\<[71]%
-\>[71]{}\Varid{⇐}\;\Conid{A}{}\<[E]%
+\>[71]{}\Varid{⇐}\;\Conid{A}\;{}\<[E]%
 \\[\blanklineskip]%
-\>[B]{}\hsindent{3}{}\<[3]%
-\>[3]{}\Varid{\char95 ∋\char95 }\;{}\<[9]%
-\>[9]{}\mathbin{:}\;(\Conid{A}\;\mathbin{:}\;\Conid{Ty})\;{}\<[56]%
+\>[9]{}\Varid{\char95 ∋\char95 }\;{}\<[15]%
+\>[15]{}\mathbin{:}\;(\Conid{A}\;\mathbin{:}\;\Conid{Ty})\;{}\<[56]%
 \>[56]{}\Varid{op}\;{}\<[62]%
-\>[62]{}\mathbin{:}\;\Varid{⟦}\;\Conid{D}\;\Varid{⟧}\;\Conid{Raw}\;\Varid{\char95 ⊢\char95 [\char95 ]\char95 }\;\Conid{Γ}\;\Varid{rs}\;\Varid{d}\;\Conid{A}\;{}\<[E]%
+\>[62]{}\mathbin{:}\;\Varid{⟦}\;\Conid{D}\;\Varid{⟧}\;\Conid{Raw}\;\Varid{\char95 ⊢\char95 [\char95 ]\char95 }\;\Conid{Γ}\;{}\<[85]%
+\>[85]{}\Varid{rs}\;{}\<[89]%
+\>[89]{}\phantom{[}{}\<[108]%
+\>[108]{}\Varid{d}\;{}\<[113]%
+\>[113]{}\Conid{A}\;{}\<[E]%
 \\
@@
 \>[24]{}\Varid{⇐}\;\Conid{A}\;{}\<[62]%
-\>[62]{}\Varid{→}\;\Conid{Γ}\;\Varid{⊢}\;\Varid{op}\;\Varid{rs}\;[\mskip1.5mu \;\Varid{d}\;\mskip1.5mu]\;\Conid{A}\;{}\<[E]%
+\>[62]{}\Varid{→}\;\Conid{Γ}\;\Varid{⊢}\;\Varid{op}\;\Varid{rs}\;{}\<[89]%
+\>[89]{}[\mskip1.5mu \;{}\<[108]%
+\>[108]{}\Varid{d}\;\mskip1.5mu]\;{}\<[113]%
+\>[113]{}\Conid{A}\;{}\<[E]%
 \\
@@
 
-\begin{figure}
+Induction proofs in our formalisation typically consist of three mutual definitions---one for the inductive type of derivations, one for a list of arguments, and one for an extension context.
+
+As an illustration, consider the formal proof of soundness below, i.e.\ the `if' part of \cref{lem:soundness-completeness}.
+Its formal proof is defined in a module which has signatures as its parameters:
+{\small\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Keyword{module}\;\Conid{Theory.Soundness}\;\{\mskip1.5mu \Conid{SD}\;\mathbin{:}\;\Conid{S.SigD}\mskip1.5mu\}\;(\Conid{BD}\;\mathbin{:}\;\Conid{B.SigD}\;\Conid{SD})\;\Keyword{where}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+}
+Then, the induction proof is nothing more than a function defined by pattern matching for each case: each constructor of bidirectional typing derivations is mapped to a corresponding constructor of typing derivations.
+(The definition \ensuremath{\Varid{soundnessᵃ}} might be further removed if \Agda's termination checker 
+The only (slightly) interesting case $\ChkRule{Sub}$ is proved by disregarding the identity proof \ensuremath{\Varid{refl}} and use the induction hypothesis (\ensuremath{\Varid{soundness}\;\Varid{t}}) directly:
+\begin{figure}[H]
   \codefigure
-  \begin{minipage}[b]{.44\textwidth}
+  \begin{minipage}[t]{.55\textwidth}
 \begin{hscode}\SaveRestoreHook
@@
 \column{5}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{26}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{8}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{24}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{28}{@{}>{\hspre}l<{\hspost}@{}}%
 \column{30}{@{}>{\hspre}l<{\hspost}@{}}%
 \column{33}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{46}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{35}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{41}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{45}{@{}>{\hspre}l<{\hspost}@{}}%
 \column{E}{@{}>{\hspre}l<{\hspost}@{}}%
@@
 \>[B]{}\hsindent{3}{}\<[3]%
-\>[3]{}\Varid{soundness}\;(\Varid{var}\;\Varid{i}\;\Varid{eq})\;{}\<[26]%
-\>[26]{}\mathrel{=}\;\Varid{var}\;\Varid{i}\;\Varid{eq}{}\<[E]%
+\>[3]{}\Varid{soundness}\;(\Varid{var}\;\Varid{i}\;\Varid{eq})\;\mathrel{=}\;\Varid{var}\;\Varid{i}\;\Varid{eq}{}\<[E]%
 \\
 \>[B]{}\hsindent{3}{}\<[3]%
-\>[3]{}\Varid{soundness}\;(\Conid{A}\;\Varid{∋}\;\Varid{t})\;{}\<[26]%
-\>[26]{}\mathrel{=}\;\Conid{A}\;\Varid{∋}\;\Varid{soundness}\;\Varid{t}{}\<[E]%
+\>[3]{}\Varid{soundness}\;(\Conid{A}\;\Varid{∋}\;\Varid{t})\;{}\<[24]%
+\>[24]{}\mathrel{=}\;\Conid{A}\;\Varid{∋}\;\Varid{soundness}\;\Varid{t}{}\<[E]%
 \\
 \>[B]{}\hsindent{3}{}\<[3]%
-\>[3]{}\Varid{soundness}\;(\Varid{t}\;\Varid{↑}\;\Varid{refl})\;{}\<[26]%
-\>[26]{}\mathrel{=}\;\Varid{soundness}\;\Varid{t}{}\<[E]%
+\>[3]{}\Varid{soundness}\;(\Varid{t}\;\Varid{↑}\;\Varid{refl})\;\mathrel{=}\;\Varid{soundness}\;\Varid{t}{}\<[E]%
 \\
 \>[B]{}\hsindent{3}{}\<[3]%
-\>[3]{}\Varid{soundness}\;(\Varid{op}\;\Varid{ts})\;{}\<[26]%
-\>[26]{}\mathrel{=}\;{}\<[E]%
+\>[3]{}\Varid{soundness}\;(\Varid{op}\;\{\mskip1.5mu \Varid{rs}\;\mathrel{=}\;\Varid{i}\;\Varid{,}\;\anonymous \mskip1.5mu\}\;(\anonymous \;\Varid{,}\;\Varid{σ}\;\Varid{,}\;\Varid{σ-eq}\;\Varid{,}\;\Varid{ts}))\;\mathrel{=}\;{}\<[E]%
 \\
 \>[3]{}\hsindent{2}{}\<[5]%
-\>[5]{}\Varid{op}\;(\Varid{soundnessᶜ}\;(\Conid{BD}\;\Varid{.rules}\;\anonymous )\;\Varid{ts}){}\<[E]%
-\\[\blanklineskip]%
-\>[B]{}\hsindent{3}{}\<[3]%
-\>[3]{}\Varid{soundnessᶜ}\;{}\<[E]%
-\\
-\>[3]{}\hsindent{2}{}\<[5]%
-\>[5]{}\mathbin{:}\;(\Conid{D}\;\mathbin{:}\;\Conid{OpD})\;{}\<[E]%
-\\
-\>[3]{}\hsindent{2}{}\<[5]%
-\>[5]{}\Varid{→}\;\Varid{⟦}\;\Conid{D}\;\Varid{⟧ᶜ}\;\Conid{Raw}\;\Varid{\char95 ⊢\char95 [\char95 ]\char95 }\;\Conid{Γ}\;\Varid{rs}\;\Varid{d}\;\Conid{A}\;{}\<[E]%
-\\
-\>[3]{}\hsindent{2}{}\<[5]%
-\>[5]{}\Varid{→}\;\Conid{T.⟦}\;\Varid{eraseᶜ}\;\Conid{D}\;\Varid{⟧ᶜ}\;\Conid{Raw}\;\Varid{\char95 ⊢\char95 ⦂\char95 }\;\Conid{Γ}\;\Varid{rs}\;\Conid{A}{}\<[E]%
-\\
-\>[B]{}\hsindent{3}{}\<[3]%
-\>[3]{}\Varid{soundnessᶜ}\;(\Varid{ι}\;\anonymous \;\anonymous \;\Conid{Ds})\;(\anonymous \;\Varid{,}\;\Varid{σ}\;\Varid{,}\;\Varid{σ-eq}\;\Varid{,}\;\Varid{ts})\;{}\<[46]%
-\>[46]{}\mathrel{=}\;{}\<[E]%
-\\
-\>[3]{}\hsindent{2}{}\<[5]%
-\>[5]{}\Varid{σ}\;\Varid{,}\;\Varid{σ-eq}\;\Varid{,}\;\Varid{soundnessᵃˢ}\;\Conid{Ds}\;\Varid{ts}{}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-\end{minipage}
-\begin{minipage}[b]{.55\textwidth}
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{17}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{23}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{26}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{27}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{35}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{40}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{41}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{45}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{50}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\vphantom{where}{}\<[E]%
+\>[5]{}\Varid{op}\;(\Varid{σ}\;\Varid{,}\;\Varid{σ-eq}\;\Varid{,}\;\Varid{soundnessᵃˢ}\;(\Conid{BD}\;\Varid{.ar}\;\Varid{i}\;\Varid{.args})\;\Varid{ts}){}\<[E]%
 \\[\blanklineskip]%
@@
 \>[3]{}\hsindent{2}{}\<[5]%
-\>[5]{}\mathbin{:}\;(\Conid{Ds}\;\mathbin{:}\;\Conid{ArgsD}\;\Conid{Ξ})\;{}\<[E]%
+\>[5]{}\mathbin{:}\;{}\<[8]%
+\>[8]{}(\Conid{Ds}\;\mathbin{:}\;\Conid{ArgsD}\;\Conid{Ξ})\;{}\<[E]%
 \\
 \>[3]{}\hsindent{2}{}\<[5]%
-\>[5]{}\Varid{→}\;\Varid{⟦}\;\Conid{Ds}\;\Varid{⟧ᵃˢ}\;{}\<[27]%
-\>[27]{}\Conid{Raw}\;\Varid{\char95 ⊢\char95 [\char95 ]\char95 }\;{}\<[40]%
-\>[40]{}\Varid{σ}\;\Conid{Γ}\;\Varid{rs}\;{}\<[E]%
+\>[5]{}\Varid{→}\;{}\<[8]%
+\>[8]{}\Varid{⟦}\;\Conid{Ds}\;\Varid{⟧ᵃˢ}\;{}\<[28]%
+\>[28]{}\Conid{Raw}\;\Varid{\char95 ⊢\char95 [\char95 ]\char95 }\;{}\<[41]%
+\>[41]{}\Varid{σ}\;\Conid{Γ}\;\Varid{rs}\;{}\<[E]%
 \\
 \>[3]{}\hsindent{2}{}\<[5]%
-\>[5]{}\Varid{→}\;\Conid{T.⟦}\;\Varid{eraseᵃˢ}\;\Conid{Ds}\;\Varid{⟧ᵃˢ}\;{}\<[27]%
-\>[27]{}\Conid{Raw}\;\Varid{\char95 ⊢\char95 ⦂\char95 }\;{}\<[40]%
-\>[40]{}\Varid{σ}\;\Conid{Γ}\;\Varid{rs}{}\<[E]%
+\>[5]{}\Varid{→}\;{}\<[8]%
+\>[8]{}\Conid{T.⟦}\;\Varid{eraseᵃˢ}\;\Conid{Ds}\;\Varid{⟧ᵃˢ}\;{}\<[28]%
+\>[28]{}\Conid{Raw}\;\Varid{\char95 ⊢\char95 ⦂\char95 }\;{}\<[41]%
+\>[41]{}\Varid{σ}\;\Conid{Γ}\;\Varid{rs}{}\<[E]%
 \\
@@
 \>[5]{}\Varid{soundnessᵃ}\;\Conid{Δ}\;\Varid{t}\;\Varid{,}\;\Varid{soundnessᵃˢ}\;\Conid{Ds}\;\Varid{ts}{}\<[E]%
-\\[\blanklineskip]%
-\>[B]{}\hsindent{3}{}\<[3]%
+\ColumnHook
+\end{hscode}\resethooks
+\end{minipage}
+\begin{minipage}[t]{.44\textwidth}
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{8}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{18}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{23}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{26}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{29}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{34}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{38}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{41}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[3]{} \phantom{mutual} {}\<[E]%
+\\
 \>[3]{}\Varid{soundnessᵃ}\;{}\<[E]%
@@
 \>[3]{}\hsindent{2}{}\<[5]%
-\>[5]{}\mathbin{:}\;(\Conid{Δ}\;\mathbin{:}\;\Conid{TExps}\;\Conid{Ξ})\;{}\<[E]%
+\>[5]{}\mathbin{:}\;{}\<[8]%
+\>[8]{}(\Conid{Δ}\;\mathbin{:}\;\Conid{TExps}\;\Conid{Ξ})\;{}\<[E]%
 \\
 \>[3]{}\hsindent{2}{}\<[5]%
-\>[5]{}\Varid{→}\;\Varid{⟦}\;\Conid{Δ}\;\Varid{⟧ᵃ}\;{}\<[17]%
-\>[17]{}\Conid{Raw}\;(\Varid{λ}\;\iden{\Conid{Γ}^\prime}\;\iden{\Varid{r}^\prime}\;\Varid{→}\;\iden{\Conid{Γ}^\prime}\;\Varid{⊢}\;\iden{\Varid{r}^\prime}\;[\mskip1.5mu \;\Varid{d}\;\mskip1.5mu]\;\Conid{A})\;{}\<[50]%
-\>[50]{}\Varid{σ}\;\Conid{Γ}\;\Varid{r}\;{}\<[E]%
+\>[5]{}\Varid{→}\;{}\<[8]%
+\>[8]{}\Varid{⟦}\;\Conid{Δ}\;\Varid{⟧ᵃ}\;{}\<[18]%
+\>[18]{}\Conid{Raw}\;(\Varid{\char95 ⊢\char95 [}\;{}\<[29]%
+\>[29]{}\Varid{d}\;\mskip1.5mu]\;{}\<[34]%
+\>[34]{}\Conid{A})\;{}\<[38]%
+\>[38]{}\Varid{σ}\;\Conid{Γ}\;\Varid{r}\;{}\<[E]%
 \\
 \>[3]{}\hsindent{2}{}\<[5]%
-\>[5]{}\Varid{→}\;\Conid{T.⟦}\;\Conid{Δ}\;\Varid{⟧ᵃ}\;{}\<[17]%
-\>[17]{}\Conid{Raw}\;(\Varid{λ}\;\iden{\Conid{Γ}^\prime}\;\iden{\Varid{r}^\prime}\;\Varid{→}\;\iden{\Conid{Γ}^\prime}\;\Varid{⊢}\;\iden{\Varid{r}^\prime}\;\Varid{⦂}\;\Conid{A})\;{}\<[50]%
-\>[50]{}\Varid{σ}\;\Conid{Γ}\;\Varid{r}{}\<[E]%
+\>[5]{}\Varid{→}\;{}\<[8]%
+\>[8]{}\Conid{T.⟦}\;\Conid{Δ}\;\Varid{⟧ᵃ}\;{}\<[18]%
+\>[18]{}\Conid{Raw}\;(\Varid{\char95 ⊢\char95 ⦂}\;{}\<[34]%
+\>[34]{}\Conid{A})\;{}\<[38]%
+\>[38]{}\Varid{σ}\;\Conid{Γ}\;\Varid{r}{}\<[E]%
 \\
-\>[B]{}\hsindent{3}{}\<[3]%
 \>[3]{}\Varid{soundnessᵃ}\;\Varid{[]}\;{}\<[23]%
@@
 \\
-\>[B]{}\hsindent{3}{}\<[3]%
 \>[3]{}\Varid{soundnessᵃ}\;(\anonymous \;\Varid{∷}\;\Conid{Δ})\;{}\<[23]%
@@
 \end{minipage}
-\caption{The soundness proof in \Agda}
-\label{fig:formal-soundness}
+%\caption{The soundness proof in \Agda}
+%\label{fig:formal-soundness}
 \end{figure}
-
-Induction proofs in our formalisation typically consist of three mutual definitions---one for the inductive type of derivations, one for a list of arguments, and one for an extension context.
-
-As an illustration, consider the soundness proof in \cref{fig:formal-soundness}, i.e.\ the `if' part of \cref{lem:soundness-completeness}.
-Each constructor of bidirectional typing derivations is mapped to a corresponding constructor of typing derivations.
-For the $\SynRule{Sub}$ case, the induction hypothesis \ensuremath{\Varid{soundness}\;\Varid{t}} is invoked directly.
-
+Compared to an induction proof tailored for a specific system, the size of a generic induction proof remains the same, regardless of the signature.
+This makes it manageable even in a proof assistant like \Agda, which lacks extensive tactics for automating theorem proving.
+Hence, we achieve more by doing less.
 \end{document}

### 2023-07-20 tex/sec6-formalisation.tex
[ refactor ] Shorten soundness and completness

@@
 
-Induction proofs in our formalisation typically consist of three mutual definitions---one for the inductive type of derivations, one for a list of arguments, and one for an extension context.
+Our formal induction proofs typically consist of mutual definitions for the inductive type of derivations, for a list of arguments, for an extension context, and sometimes for an operation separately.
 
@@
 }
-Then, the induction proof is nothing more than a function defined by pattern matching for each case: each constructor of bidirectional typing derivations is mapped to a corresponding constructor of typing derivations.
-(The definition \ensuremath{\Varid{soundnessᵃ}} might be further removed if \Agda's termination checker 
-The only (slightly) interesting case $\ChkRule{Sub}$ is proved by disregarding the identity proof \ensuremath{\Varid{refl}} and use the induction hypothesis (\ensuremath{\Varid{soundness}\;\Varid{t}}) directly:
+The induction proof is nothing more than a function defined by pattern matching for each case: each constructor of bidirectional typing derivations is mapped to a corresponding constructor of typing derivations.
+The only (slightly) interesting case $\ChkRule{Sub}$ is proved by disregarding the identity proof \ensuremath{\Varid{refl}} and use the induction hypothesis \ensuremath{\Varid{soundness}\;\Varid{t}} directly:
 \begin{figure}[H]

### 2023-07-24 tex/sec5-synthesis.tex
[ tex ] Fix a typo

@@
 Each case analysis simply branches depending on the outcomes of bidirectional type synthesis and checking for each sub-term, as well as the unification process.
-If a contradiction proof is not of interest for implementation, these programs can be simplified by disregarding the cases that yield such contradiction proofs.
+If a contradiction proof is not of interest for implementation, these programs can be simplified by discarding the cases that yield such contradiction proofs.
 Alternatively, we could consider generalising typing derivations instead, like our generalised mode derivations (\Cref{fig:generalised-mode-derivations}).

### 2023-07-24 tex/sec1-intro.tex
[ tex ] Remove defensive phrases; a comment on annotatability

@@
 As a first step, we confine our theory to \emph{syntax-directed} systems, meaning that for every raw term construct, the ordinary type system has exactly one typing rule, which is assigned modes in exactly one way in the bidirectional type system.%
-\footnote{It is reasonable to assume syntax-directedness (especially as a first step) given that it is a standard idea to re-cast non-syntax-directed typing rules into their syntax-directed form to derive a type synthesiser~\citep{Peyton-Jones2007}.}
+\footnote{It is reasonable to assume syntax-directedness given that it is a standard idea to re-cast non-syntax-directed typing rules into their syntax-directed form to derive a type synthesiser~\citep{Peyton-Jones2007}.}
 With this assumption, using a single bidirectional binding signature we can specify the raw terms, the ordinary type system, and the bidirectional type system of a language.

### 2023-07-24 tex/sec4-pre-synthesis.tex
[ tex ] Remove defensive phrases; a comment on annotatability

@@
 Annotatability also does not help the user to deal with missing annotations like (generalised) mode decoration does: although annotatability seems capable of determining where annotations are missing and even filling them in correctly, its antecedent requires a typing derivation, which is what the user is trying to construct and does not have yet.
-Therefore we believe that the notion of annotatability is too complex for what it achieves, and our theory offers simpler and more useful alternatives.
+Therefore we believe that the notion of annotatability is too complex\LT{not clear what it means} for what it achieves, and our theory offers simpler and more useful alternatives.

### 2023-07-24 tex/sec5-synthesis.tex
[ tex ] Remove defensive phrases; a comment on annotatability

@@
   \begin{enumerate}
-    \item if\/ $\erase{\Gamma} |-_{\Sigma, \Omega} \isTerm{t}^{\syn}$, then it is decidable whether there is~$A$ such that $\Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn A$;
+    \item if\/ $\erase{\Gamma} |-_{\Sigma, \Omega} \isTerm{t}^{\syn}$, then it is decidable whether $\Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn A$ for some $A$;
     \item if\/ $\erase{\Gamma} |-_{\Sigma, \Omega} \isTerm{t}^{\chk}$, then it is decidable for any~$A$ whether $\Gamma |-_{\Sigma, \Omega} \isTerm{t} \chk A$.

### 2023-08-18 tex/sec6-formalisation.tex
[ tex ] Fix a typo
[ tex ] Fix an alignment issue

@@
 \column{9}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{15}{@{}>{\hspre}l<{\hspost}@{}}%
 \column{24}{@{}>{\hspre}l<{\hspost}@{}}%
@@
 \>[62]{}\Varid{→}\;\Conid{Γ}\;\Varid{⊢}\;\Varid{r}\;{}\<[71]%
-\>[71]{}\Varid{⇐}\;\Conid{A}\;{}\<[E]%
+\>[71]{}\Varid{⇐}\;\Conid{A}{}\<[E]%
 \\[\blanklineskip]%
-\>[9]{}\Varid{\char95 ∋\char95 }\;{}\<[15]%
-\>[15]{}\mathbin{:}\;(\Conid{A}\;\mathbin{:}\;\Conid{Ty})\;{}\<[56]%
+\>[B]{}\hsindent{3}{}\<[3]%
+\>[3]{}\Varid{\char95 ∋\char95 }\;{}\<[9]%
+\>[9]{}\mathbin{:}\;(\Conid{A}\;\mathbin{:}\;\Conid{Ty})\;{}\<[56]%
 \>[56]{}\Varid{op}\;{}\<[62]%
@@
 \end{figure}
-Compared to an induction proof tailored for a specific system, the size of a generic induction proof remains the same, regardless of the signature.
-This makes it manageable even in a proof assistant like \Agda, which lacks extensive tactics for automating theorem proving.
-Hence, we achieve more by doing less.
+Each construct in an induction proof tailored for a specific system such as \PCF has to be scrutinised, so its size increases as the system in question has more constructs.
+On the other hand, the size of a generic induction proof remains the same and is manageable in a proof assistant that lacks extensive tactics for automating theorem proving, like \Agda.
+We have achieved more by doing no ad hoc proofs.
 \end{document}

### 2023-09-13 tex/sec6-formalisation.tex
[ tex ] Fix the mishap in References

@@
 
-Our formal induction proofs typically consist of mutual definitions for the inductive type of derivations, for a list of arguments, for an extension context, and sometimes for an operation separately.
+Our formal proofs of induction typically consist of mutual proofs for the inductive type of derivations, for a list of arguments, for an extension context, and sometimes for the case of operations for clarity.
 

### 2023-10-03 tex/POPL24/sec2-outline.tex
[ tex ] Add pdf

--- 2023-07-12 tex/sec2-outline.tex
@@
 Otherwise $\erase\Gamma \vdash t^{\syn}$, and we can run the bidirectional type synthesiser.
-If it finds a derivation of $\Gamma \vdash t \syn A$ for some type~$A$, return a derivation of $\Gamma \vdash t : A$ (the first outcome), which is possible because the bidirectional typing~(\cref{fig:STLC-bidirectional-typing-derivations}) is \emph{sound} with respect to the original typing~(\cref{fig:STLC-typing-derivations}).
-
-\begin{lemma}[Soundness]
-If\/ $\Gamma \vdash t \syn A$, then $\Gamma \vdash t : A$.
-\end{lemma}
-
-If there is no derivation of $\Gamma \vdash t \syn A$ for any type~$A$, report that there is no derivation of $\Gamma \vdash t : A$ for any~$A$ either (the second outcome), because the bidirectional typing is \emph{complete} with respect to the original typing.
-
-\begin{lemma}[Completeness] \label{thm:STLC-completeness}
-If\/ $\erase\Gamma \vdash t^{\syn}$ and\/ $\Gamma \vdash t : A$, then $\Gamma \vdash t \syn A$.
-\end{lemma}
+If it finds a derivation of $\Gamma \vdash t \syn A$ for some type~$A$, return a derivation of $\Gamma \vdash t : A$ (the first outcome), which is possible because the bidirectional typing~(\cref{fig:STLC-bidirectional-typing-derivations}) is \emph{sound} with respect to the original typing~(\cref{fig:STLC-typing-derivations}); if there is no derivation of $\Gamma \vdash t \syn A$ for any type~$A$, then there is no derivation of $\Gamma \vdash t : A$ for any~$A$ either (the second outcome), because the bidirectional typing is \emph{complete}:
+\begin{theorem}[Soundness and Completeness]
+$\Gamma \vdash t \syn A$ if and only if\/ $\erase\Gamma \vdash t^{\syn}$ and\/ $\Gamma \vdash t : A$.
+\end{theorem}
+%
+%
+%\begin{lemma}[Completeness] \label{thm:STLC-completeness}
+%If\/ $\Gamma \vdash t : A$, then $\Gamma \vdash t \syn A$.
+%\end{lemma}
 

### 2023-10-03 tex/POPL24/sec7-future.tex
[ tex ] Add pdf

--- 2023-07-10 tex/sec7-future.tex
@@
 
-\section{Concluding Remarks} \label{sec:future}
+\section{Future Work} \label{sec:future}
+
+\paragraph{More characteristics of bidirectional typing}
+
+We have formalised some of the most important concepts discussed by \citet{Dunfield2021}.
+There are other concepts that may be formalisable with some more effort, for example Pfenning's recipe for bidirectionalising typing rules.
+We believe that our theory lays the foundation for further formalising these concepts.
+There are some other concepts that may be more difficult to pin down, notably `annotation character', which is roughly about how easy it is for the user to write annotated programs.
+These may continue to require creativity from the designer of type systems.
 
@@
 There should be ways to fix this problem while retaining the mode decoration phase---for example, one idea is to make the mode decorator produce all possible mode derivations, and refine completeness to say that any typing derivation can be combined with one of these mode derivations into a bidirectional typing derivation.
-All these considerations are rather too complex for a first theory, however, and we decide to leave them for future work.
+All these considerations are rather too complex for a first theory, however.
 
@@
 
-\LT[inline]{polymorphic algebraic theories, dependent signatures --- 0.5p}
-
 \paragraph{Beyond simple types}
+We leave the problem of presenting bidirectional typing for more general and advanced language features as future work (such as those related to polymorphic types~\citep{Pierce2000,Peyton-Jones2007,Dunfield2013,Xie2018}).
 While algebraic approaches to polymorphic types have been developed~\citep{Fiore2013,Hamana2011}, these approaches do not take subtyping into account.
 Subtyping is essential for formulating important concepts such as \emph{principal types} in type synthesis.
-On the other hand, in the realm of dependent types, \citeauthor{Cartmell1986}'s generalised algebraic theories~\citeyearpar{Cartmell1986} can handle a wide variety of dependent type theories.
+On the other hand, in the realm of dependent types, \varcitet{Cartmell1986}{'s} generalised algebraic theories can handle a wide variety of dependent type theories.
 \citet{Bezem2021} investigate the notion of presentation (extending the notion of signature) in the context of generalised algebraic theories.
 Nonetheless, type synthesis for dependent types requires normalisation or some form of conversion to check type equality.
-Normalisation in its generic form still remains out of reach and recent advances in this topic are discussed in the doctoral thesis by~\citet{Valliappan2023}.
+Normalisation in its generic form still remains out of reach, and advances in this topic are only recently discussed in the doctoral thesis by~\citet{Valliappan2023}.
+Before we can embark on a richer theory of bidirectional type synthesis, language formalisation frameworks for these more advanced language features have to be developed.
 
-\todo[inline]{We leave the problem of presenting bidirectional typing for more general and advanced language features as future work (such as those related to polymorphic types~\citep{Pierce2000,Peyton-Jones2007,Dunfield2013,Xie2018}), which requires advancement in language formalisation~(\cref{sec:language-formalisation}) that is orthogonal to our work.}
+%\paragraph{Whither extrinsic typing?}
 
-\Josh[inline]{Optional: ornaments (0.25p), NDGP (0.25p)}
+%\todo[inline]{We leave the problem of presenting bidirectional typing for more general and advanced language features as future work (such as those related to polymorphic types~\citep{Pierce2000,Peyton-Jones2007,Dunfield2013,Xie2018}), which requires advancement in language formalisation~(\cref{sec:language-formalisation}) that is orthogonal to our work.}
+
+%\Josh[inline]{Optional: ornaments (0.25p), NDGP (0.25p)}
 
@@
 %
-%The aforementioned frameworks except \citeauthor{Gheri2020}'s are at least inspired by \varcitet{Fiore1999}{'s} initial semantics for abstract syntax with variable binding using category theory.
-%The main idea is that the set of (untyped) abstract syntax trees for a language consists of
-%\begin{enumerate*}
-%  \item a family of sets $\Term_{\Gamma}$ of well-scoped terms under a context~$\Gamma$ with
-%  \item variable renaming for a function $\sigma\colon \Gamma \to \Delta$ between variables acting as a functorial map from $\Term_{\Gamma}$ to $\Term_{\Delta}$, i.e.\ a presheaf $\Term\colon \mathbb{F} \to \mathsf{Set}$, and
-%  \item an initial algebra $[\mathsf{v}, \mathsf{op}]$ on~$\Term$ given by the variable rule as a map $\mathsf{v}$ from the presheaf~$V$ of variables (i.e.\ the embedding $V\colon \mathbb{F} \hookrightarrow \mathsf{Set}$) to $\Term$ and other constructs as $\mathsf{op}\colon \mathbb{\Sigma}\Term \to \Term$ where $\mathbb{F}$ is the category of contexts, the functor $\mathbb{\Sigma}\colon \mathsf{Set}^\mathbb{F} \to \mathsf{Set}^\mathbb{F}$ encodes the arities of constructs, and the initiality amounts to structural recursion, i.e.\ \emph{term traversal}.
-%\end{enumerate*}
-%To put it succinctly, it is the free $\mathbb{\Sigma}$-algebra over the presheaf~$V$ of variables.
-%
-%Fortunately, constructing the initial algebra of terms in type theory boils down to defining an inductive type with a few constructors that align with the variable rule and a rule schema for language constructs specified by a signature~\citep{Fiore2022}.
 

### 2023-10-11 tex/POPL24/response.md
[ update ] Add ESOP submission draft

# Main response

We thank all the reviewers for the time and effort they put into this paper.

(At least) reviewers B & C agree with us that developing a general theory of bidirectional type synthesis is an important direction; if so, there should be a paper at POPL that says so! There are, however, two main concerns raised by reviewers A & B (which we had ourselves actually):

1. There doesn’t seem to be enough ‘delta’ from previous work on bidirectional typing, in particular Dunfield & Krishnaswami’s [2021].

2. The work should’ve treated (or at least discussed) type systems beyond simple types.

We will address these concerns below, but first here’s a summary:

1. A general theory is important for further scientific development, and is more difficult to work out than individual instances. We offer mathematical definitions, theorems, and proofs (which we believe the POPL community values and sometimes even requires) in contrast to Dunfield & Krishnaswami’s informal treatment. And the delta from existing language formalisation frameworks (where the most complex operation is usually substitution) should be taken into account too.

2. A theory about simple types is the best we can do now, because the development hinges on the state-of-the-art techniques used in language formalisation frameworks, which basically deal with simple types only. It is conceivable to extend the theory to more expressive types, but that will have to start from extending language formalisation frameworks, which is substantial work by itself.

We will revise the paper to make the delta clearer, and include more discussion on how we may extend the theory to more expressive types — see the detailed response section below.

Regarding the concern about delta, first allow us to point out that developing a general theory (even if it doesn’t do much at the beginning) is a significant step forward from having knowledge and experience about individual instances. Being able to write a type-synthesiser (or parser) generator means that the knowledge and experience gained from writing individual type synthesisers (or parsers) has matured to the point that general patterns have been identified and precisely formulated, and can be studied further. For example, when writing individual type synthesisers, it suffices to have just a vague idea of mode-correctness, but without a mathematically precise formulation of mode-correctness, it’s impossible to explore its consequences rigorously, for example the fact that the uniqueness of synthesised types follows from mode-correctness (Lemma 5.5 at L746), which was not made clear before. And general theorems are more difficult to prove: a case in point is the op case in the proof of Theorem 5.6 (L828), which is significantly more complex than the concrete cases for STLC or PCF, for example (but captures the general pattern once and for all).

For a comparison with Dunfield & Krishnaswami’s paper in particular: Their paper is more like reflection from experience working with individual type systems (there are intuitive phrases like type system, but no rigorous definitions or general theorems etc). That’s nice to have, but we shouldn’t let our knowledge about type synthesis remain in that state — it is important to consolidate that knowledge and experience into a general theory to enable further scientific development. And it’s not that we’re just formalising Dunfield & Krishnaswami’s paper: mode decoration is new and leads to a nicer specification of type synthesis (the trichotomy). We expect the theory of bidirectional typing to evolve further, but the improvements should start to be accumulated and studied within general theories rather than scattered in a myriad of type systems.

The delta from previous work on language formalisation frameworks should be taken into account too. We mentioned in Section 1.1.2 that most of the existing language formalisation frameworks (including Fiore and Szamozvancev’s distinguished paper at POPL 2022) do not provide operations beyond substitution, offering few incentives to use or extend these frameworks. This is an unfortunate situation that we should change by demonstrating that we can do significantly larger-scale constructions, such as a generic bidirectional type synthesiser, with these frameworks, and that there is a clear need to extend these frameworks to deal with more expressive types (see the discussion about simple types below).

Regarding the concern about simple types being too simple: we also mentioned in Section 1.1.2 that the existing language formalisation frameworks basically support simple types only; our development depends on, and is restricted by, the progress there, but there is no reason to wait for such frameworks to cover polymorphic types with subtyping etc before initiating a general theory of bidirectional typing — there are orthogonal notions such as mode-correctness and mode decoration that we can treat now. It is conceivable that the theory can be scaled up to more expressive type systems, but that’s ‘highly non-trivial’ as reviewer A suspects — in particular, that would have to start with the development of a more expressive language formalisation framework, which is too much work for a paper about bidirectional typing. On the other hand, we agree with reviewer A that there should be more discussion about a ‘path towards extending the framework’; a sample is provided at the end of the detailed response (the ‘Path’ section).

# Detailed response

## Review A

> How hard do you expect it would be to support a wider range of languages? In particular, what about languages with polymorphism or dependent types?

See the ‘Path’ section below.

> 25-26: "However, it was later found that full" -> "However, full" (later?)

Will fix it!  
> I'm not sure section 7 adds much value to the paper, given that the formalization is available.

We admit that it does not add much to the theory itself, but it might be worthwhile to share with the interested reader how the formal definition is inspired through a category-theoretic analysis (especially the idea of Hermida and Jacobs), especially that our definitions contain a rule schema which is unusual in a typical definition for a single language (as the set of term constructs are usually fixed a priori). Defining this rule schema actually took us some effort to engineer working definitions, e.g. a lifting of a signature functor.

> Section 8 also does not feel super useful: … Alternatively, you could try to explain why the extensions you discuss are actually challenging, highlighting, e.g., what part of your formalization would break.

Yes, thanks for the suggestion. It is clear that further developments are necessary, so we’d like to outline the path towards extensions further, as detailed below.

## Review B

> Can you provide some more intuition on how you would expect to scale the approach to richer type systems, such as System F or, especially, F<:?

See the ‘Path’ section below.

> "bidirectional typing has only been developed on a case-by-case basis without a theory" I found it reasonably clear what is meant in context, but statements like this can come across as uninformed and perhaps inflammatory

Thanks for pointing this out. We will revise it to avoid unnecessary misunderstanding. A slightly revised sentence below might do:

‘… on a case-by-case basis without a general theory’

or an even longer version

‘the design of bidirectional typing has been developed based on the experience and properties we have accumulated from individual systems, without the support of a mathematical theory that works generally for all languages like parsing.’

## Review C

> Can your formalization be used to actually synthesize types, albeit inefficiently?

Yes! Please normalise the term `⊢S?` defined at L122 in the file `src/Example/STLC.agda` in our supplementary material.

> In the supplementary material, README.agda promises an HTML rendering of the code that does not seem to be present.

Sorry, we forgot. We’ve now attached it as part of the response.

> L320: Why don't you assert here that $\Xi$ has decidable equality? Is it because this hypothesis isn't needed until much later (L793, L829)?

Yes, you’re right. $\Xi$ doesn’t need to have decidable equality until Section 5.2.

> L1021: Our formal construction appears novel" To me this idea is strongly reminiscent of the idea of Melliès and Zeilberger that …

Interesting, we will have a look.

We will address other detailed comments in our next revision.

## ‘Path towards extending the framework’ (Reviews A&B)

> How hard do you expect it would be to support a wider range of languages?

> Can you provide some more intuition on how you would expect to scale the approach to richer type systems, such as System F or, especially, F<:?

We lightly touched this question in Section 7 and will outline our plan further in the revision. In short, to extend our theory with additional language features, we have (at least) three key steps to follow:

1. Extending algebraic theories: We need to extend the algebraic theory of abstract syntax with variable binding to accommodate more language features or complex types.

2. Transforming theory to framework: the next objective is to transform the extended algebraic theory into a language-formalisation framework that is formally verified and implemented within a proof assistant.

3. Identifying conditions for bidirectional type synthesis: it is necessary to identify the  conditions, beside mode-correctness, to establish a decidable bidirectional type synthesiser for more complex languages.

The first step involves the development of more advanced algebraic theories capable of expressing a wider range of languages. Such an algebraic theory alone could already be worthy of publication in reputable conferences in theoretical computer science. For instance, Fiore et al. [1999], Fiore & Hamana [2013], Arkor & Fiore's Algebraic models of simple type theories [2020] have explored algebraic theories of abstract syntax with variable binding in various directions and published their results all in LICS.

The second task involves making the mathematical arguments of the chosen algebraic theory constructive and amenable to formalisation within a type-theoretic proof assistant. Being constructive and type-theoretic allows theorems compute like ours, but this step can be highly non-trivial. Some statements may need substantial reformulation for the purpose of formalisation. For example, Fiore & Szamozvancev [2022] wrote a whole paper presenting a mathematical reformulation of the syntax/representation-independent presheaf model of variable binding for formalisation in Agda, based on Fiore et al. [1999].

The third step involves identifying the mathematical conditions for bidirectional type synthesis. This is the main focus of this paper.

Each step towards a more general theory of bidirectional type synthesis is further exemplified below by polymorphic and dependent types separately.

### Polymorphism and subtyping

When it comes to polymorphic types and subtyping, the current polymorphic algebraic theory, as presented by Fiore & Hamana [2013], is sufficient in expressing polymorphic typed languages like System F. However, it does not include subtyping (F<:), which is far more interesting. To address this, we intend to expand the algebraic theory by introducing a partial order between types. This extension can be achieved by augmenting the concept of type signature with an ordering apart from type variable binding. Here's what we require:

1. Extending the notion of arity for type constructs: The extended arity is represented as a finite list of numbers, denoting the binding of type variables. This is already in the Fiore & Hamana’s theory.

2. Specifying ordering between type constructs: The type signature must also include a subtyping ordering for type constructs.

3. Establishing decidability conditions (optional): It is crucial to identify and establish the sufficient conditions to ensure that the induced ordering between types is decidable. This step generalises the decidable equality check between types in our case, but we may simply assume it.

Several mathematical concepts need to be formulated subsequently. These include a construction of its syntactic model, (generic) renaming/substitution, a syntactic construction of terms, and term recursion/induction. Additionally, we must ensure that the extended theory includes intended languages, such as F<:, as instances. The notion of term signatures will need to be enriched with modes, which is expected to be relatively straightforward.

However, it remains uncertain how challenging the formalisation of the extended theory within a type-theoretic framework will be.

Once we establish an algebraic theory of polymorphic languages with subtyping, we will be well-equipped to study their bidirectional typing. This foundation will enable us to define, say, the concept of principal types for those formally defined language specifications.

One of the primary challenges in designing a bidirectional type synthesis for polymorphic typed languages is tackling the ‘instantiation problem’. This problem involves solving for $B$ in expressions of the form $\forall \alpha. A \mathrel{<:} A[B/\alpha]$, as discussed in Section 5 of Dunfield & Krishnaswami's paper. Unfortunately, there is no universally accepted, canonical solution to this problem. Instead, various competing approaches exist. Therefore, our task is to formulate a set of conditions for language specifications. These conditions will serve as criteria, ensuring that any solution to the instantiation problem aligns with them. Meeting these conditions will be a prerequisite for the derivation of bidirectional typing synthesis.

Following these steps shall allow us to have a verified type-synthesiser generator for bidirectionally and polymorphically typed languages with subtyping like ours for simple types.

### Dependent types

Developing an algebraic theory for dependently typed languages with a notion of signature is more challenging. It is known that Cartmell's generalized algebraic theory and categories with families can model dependent type theories like Martin-Löf type theory and variants of modal type theories, as demonstrated in Daniel Gratzer's recent PhD thesis, among others (as initial models in the corresponding categories of algebraic models). However, unlike simple types and polymorphic types, type and term signatures have to be conflated. In dependent type theory, contexts, types, terms, and equations are interdependent, making mere ‘signatures’ insufficient. Instead, we need to embrace the concept of a ‘stratified presentation’ as shown by Bezem et al. [2021].

Bezem et al. offer a syntax-independent development and a syntactic construction, in line with the stratified presentation. It's already challenging to ascertain its direct correspondence to a type-theoretic construction. Given the inherent complexity of formalising a dependent type theory, transforming such an algebraic theory into a language-formalisation framework becomes even more daunting.

An alternative in this research direction is Kaposi et al.'s POPL'19 paper 'Constructing quotient inductive-inductive types'. They explore the concept of signatures for quotient inductive-inductive types (QIIT) instead and construct corresponding languages from given (dependent) signatures. Notably, Altenkirch & Kaposi's POPL'16 paper previously showed that QIIT is also able to express Martin-Löf type theory. However, progress beyond (dependent) term recursion/induction and substitution in this formalisation has been limited. We have not evaluated these two approaches, but nevertheless, we believe that both frameworks hold potential for expressing bidirectional dependent type theories.

To synthesis types for terms in a dependent type theory, though, a crucial requirement is a decidable type conversion mechanism, which is a substantial undertaking. One common implementation is achieved through normalisation by evaluation (NbE). NbE has been developed for various type theories, but pattern haven’t yet been formally stated. It's currently unclear how a decidable NbE can be derived in a generic manner, as some rules such as Equality Reflection in extensional type theory are clearly taboos. That is, we shall find out what properties of typing rules make the type equality check decidable. A general theory of bidirectional type synthesis for dependent types is possible, only if generic NbE is established.

### 2023-10-11 tex/ESOP24/BiSig.tex
[ update ] Add ESOP submission draft

\documentclass[orivec,envcountsame,envcountsect,envcountreset,oribibl]{llncs}

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage[british]{babel}
\usepackage[numbers,sort&compress]{natbib}

\usepackage{soul}
\usepackage{subfiles}
\usepackage{xurl} % allow line breaks anywhere in a URL string
\usepackage[colorlinks,urlcolor=blue,citecolor=blue]{hyperref}

\usepackage{bbold}
\newcommand{\hmmax}{0}
\newcommand{\bmmax}{0}
\usepackage{bm}
\usepackage{mathtools}
\input{mathlig}
\usepackage{mathpartir}

\usepackage{xspace}
\usepackage[capitalise,noabbrev]{cleveref}

% Used for typesetting Agda code
\usepackage{agda-lhs2tex}
\newcommand{\cons}[1]{\mathbf{#1}}
\newcommand{\iden}{\mathit}

\usepackage{xifthen}
\newcommand{\varcitet}[3][]{\citeauthor{#2}#3~\ifthenelse{\isempty{#1}}{\cite{#2}}{\cite[#1]{#2}}}

\usepackage[inline]{enumitem} % for environment enumerate*
\newlist{inlineenum}{enumerate*}{1}
\setlist[inlineenum]{label=(\roman*)}

\setlength{\marginparwidth}{2cm}
\usepackage[obeyFinal,color=yellow,textsize=scriptsize]{todonotes}

\usepackage{microtype}
\microtypesetup{babel}

\newcommand{\LT}[2][]{\todo[author=LT,color=orange!40,#1]{#2}}
\newcommand{\Josh}[2][]{\todo[author=Josh,color=yellow!40,#1]{#2}}
\newcommand{\codefigure}{\small\setlength{\parindent}{0em}\setlength{\mathindent}{0em}\setlength{\abovedisplayskip}{0ex}\setlength{\belowdisplayskip}{0ex}}

\input{type-notation} %% for type theory notation copied from HoTT Book
\input{macros}

%\author{Liang-Ting Chen
%\orcidID{0000-0002-3250-1331}
%\and 
%Hsiang-Shang Ko
%\orcidID{0000-0002-2439-1048}}
%
%\institute{Institute of Information Science, Academia Sinica, Taiwan}

\title{A formal treatment of bidirectional typing}

\begin{document}

\maketitle

\begin{abstract}
There has been much progress in designing bidirectional type systems and associated type synthesis algorithms, but mostly on a case-by-case basis.
This situation is in stark contrast to parsing, for which there have been general theories and widely applicable tools such as parser generators.
To remedy the situation, this paper develops a \emph{general} and \emph{formal} theory of bidirectional typing, and provides a \emph{verified} generator of \emph{proof-relevant} type synthesisers for syntax-directed simply typed languages: For every signature that specifies a bidirectionally typed language, there exists a proof-relevant type synthesiser that for an input abstract syntax tree constructs a typing derivation if any, gives its refutation if not, or reports that the input does not have enough type annotations.
Soundness, completeness, and mode-correctness are studied universally for any simply typed language, which are sufficient conditions for deriving a correct type synthesiser.
We propose a preprocessing step called `mode decoration', which helps the user to deal with missing type annotations.
The entire development is formalised in \Agda and can be further integrated with other language-formalisation frameworks.
\end{abstract}

\input{S1-intro}
\input{S2-outline}
\input{S3-defs.tex}
\input{S4-pre-synthesis}
\input{S5-synthesis}
\input{S6-example}
%\subfile{S7-formalisation}
\input{S8-future}
\appendix
%\input{appendix}

%\begin{acks}
%%We thank Nathanael Arkor for the useful conversation.
%We thank Kuen-Bang {Hou (Favonia)} for his comments and suggestions.
%The work is supported by the \grantsponsor{NSTC}{National Science and Technology Council of Taiwan}{https://www.nstc.gov.tw/} under grant \grantnum{NSTC}{NSTC 112-2221-E-001-003-MY3}.
%\end{acks}

\bibliographystyle{splncs04nat}

\bibliography{../reference.bib}
%\IfFileExists{../reference.bib}{%
%}{\bibliography{ref}}

\end{document}

### 2023-10-11 tex/ESOP24/S1-intro.tex
[ update ] Add ESOP submission draft

%!TEX root = BiSig.tex

\section{Introduction}\label{sec:intro}

Type inference is an important mechanism for the transition to well-typed programs from untyped abstract syntax trees, which we call \emph{raw terms}.
Here `type inference' refers specifically to algorithms that ascertain the type of any raw term \emph{without type annotations}.
However, full parametric polymorphism leads to undecidability in type inference, as do dependent types~\cite{Wells1999,Dowek1993}.
In light of these limitations, \emph{bidirectional type synthesis} emerged as a viable alternative, providing algorithms for deciding the types of raw terms that meet some syntactic criteria and usually contain type annotations.
\citet{Dunfield2021} summarised the design principles of bidirectional type synthesis and its wide coverage of languages with simple types, polymorphic types, dependent types, gradual types, among others.

The basic idea of bidirectional type synthesis is that while the problem of type inference is not decidable in general, for certain kinds of terms it is still possible to infer their types (for example, the type of a variable can be looked up in the context); for other kinds of terms, we can switch to the simpler problem of type checking, where the expected type of a term is also given so that there is more information to work with.
More formally, every judgement in a bidirectional type system is extended with a \emph{mode:}
\begin{inlineenum}
  \item $\Gamma |- \isTerm{t} \syn A$ for \emph{synthesis} and 
  \item $\Gamma |- \isTerm{t} \chk A$ for \emph{checking}.
\end{inlineenum}
The former indicates that the type~$A$ is computed as output, using both the context~$\Gamma$ and the term~$t$ as input, while for the latter, all three of $\Gamma$, $t$, and~$A$ are given as input.
The algorithm of a bidirectional type synthesiser can usually be `read off' from a well-designed bidirectional type system: as the synthesiser traverses a raw term, it switches between synthesis and checking, following the modes assigned to the judgements in the typing rules.
%if a condition called \emph{mode-correctness} is satisfied.
%every synthesised type in a typing rule is determined by previously synthesised types from its premises and its input if checking.
%As \citet{Pierce2000} noted, bidirectional type synthesis propagates type information locally within adjacent nodes of a term, and does not require unification as Damas--Milner type inference does.
%Moreover, introducing long-distance unification constraints would even undermine the essence of locality in bidirectional type synthesis.

%Also, annotations for, say, top-level definitions make the purpose of a program easier to understand, so they are sometimes beneficial and not necessarily a nuisance.
%In light of these considerations, bidirectional type synthesis can be deemed as a type-checking technique that is more fundamental than unification, as it is capable of handling a broad spectrum of programming languages.%
%\todo{Reorganise into two paragraphs, one on old approaches and the other exclusively on bidirectional type synthesis?}

Despite sharing the same basic idea, bidirectional typing has been mostly developed on a case-by-case basis; this situation is in stark contrast to parsing, which has general theories together with widely applicable techniques and practical tools, notably parser generators.
%\footnote{The same could be said for type checking in general, not just for bidirectional type systems.
%While there are type checker generators grounded in unification~\citep{Gast2004,Grewe2015}, it should be noted that unification-based approaches are not suited to more complex type systems.}
While it is straightforward to derive a type synthesis algorithm,
For bidirectional typing, \citeauthor{Dunfield2021} only present informal design principles that the community learned from individual systems, rather than a general treatment providing logical specifications and rigorously proven properties for a class of systems.
Moreover, unlike the plethora of available parser generators, `type-synthesiser generators' rarely exist, so each type synthesiser has to be independently built from scratch.
\LT{emphasise that formalisation and verification are important}

To remedy the situation, we develop a general and formal treatment of bidirectional typing, providing an out-of-the-box \emph{verified generator} of \emph{proof-relevant} type synthesisers developed with the proof assistant \Agda:

\subsubsection{Proof-relevance in (bidirectional) type synthesis}
\label{sec:PLFA}

Our work adapts proof-relevant (bidirectional) type synthesis illustrated by \citet{Wadler2022}.
Their treatment deviates from the common formulation: 
Traditionally, a type synthesis algorithm is presented as \emph{algorithmic system} relations such as $\Gamma |- \isTerm{t} \syn A \mapsto \isTerm{t}'$, denoting that annotations can be added to $\isTerm{t}$ in the surface language to produce $\isTerm{t}'$ of type $A$ in the core language.
Such an algorithm is then accompanied with soundness and completeness assertions such that the algorithm correctly synthesises the type of a raw term and every typable term can be synthesised if there are enough annotations.
By contrast, \citeauthor{Wadler2022} exploit the computational and logical nature of Martin-L\"of type theory and formulate \emph{algorithmic soundness, completeness, and decidability in one go}.
%(although they do not emphasise the difference of their approach from the traditional one)

Recall that the law of excluded middle $P + \neg P$ does not hold as an axiom for every $P$ in type theory, and we say $P$~is logically \emph{decidable} if the law holds for~$P$.
For example, a proof of \emph{decidable equality}, i.e. $\forall x, y.~(x = y) + (x \neq y)$, decides whether $x$~and~$y$ are equal and accordingly gives an identity proof or a refutation explicitly; such a decidability proof may or may not be possible depending on the domain of $x$~and~$y$, and is non-trivial.
Given that all proofs as programs terminate, logical decidability implies algorithmic decidability.
Further, suppose a proof of the following statement:
\begin{quote}
  `For a context $\Gamma$ and a raw term $t$, either a typing derivation of\, $\Gamma |- t : A$ exists for some type~$A$ or any derivation of\, $\Gamma |- t : A$ for some type~$A$ leads to a contradiction'
\end{quote}
or rephrased succinctly as 
\begin{quote}
  `It is \underline{\emph{decidable}} for any $\Gamma$~and~$t$ whether $\Gamma |- t : A$ is derivable for some~$A$'.
\end{quote}
It yields either a typing derivation for the given raw term~$t$ or a negation proof that such a derivation is impossible where the former case is algorithmic soundness and the latter algorithmic completeness in contrapositive form.
Then, both algorithmic soundness and completeness in original form are implied.
The proof of this statement in type theory is a \emph{proof-relevant} type synthesiser, because it does not only return yes/no but construct a proof---either a typing derivation or its refutation. 
\LT{emphasise that our proof is an actual program}
%For a type synthesiser being proof-relevant, all of its intermediate programs have to be proof-relevant, including bidirectional type synthesiser and its mode decorator.

\Josh{emphasise the scope of our approach}
\subsubsection{Mode decoration and bidirectional type synthesis}
\Josh{A missing step between bidirectional type synthesis and parsing}
As raw terms may not be sufficiently annotated, we propose a preprocessing step called \emph{mode decoration} to determine whether a raw term can be processed by a bidirectional type synthesiser later.
A mode decorator constructs a mode derivation for a raw term or pinpoints missing annotations.
A bidirectional type synthesiser and a type synthesiser differ in their domain---bidirectional type synthesis only works for mode-decorated raw terms instead of all raw terms.
Soundness and completeness of bidirectional typing is reformulated as a one-to-one correspondence between \emph{bidirectional typing derivations} and \emph{typing derivations} with \emph{mode derivations} (of a raw term), which is a more refined and useful formulation than annotatability described by \citet[Section~3.2]{Dunfield2021}.
Our proof-relevant bidirectional type synthesiser is complete with respect to mode-decorated terms.

We also prove that a raw term with missing annotations is exactly a raw term without a mode derivation, so our mode decorator is indeed proof-relevant.

By combining mode-decoration and bidirectional type synthesis, we show a trichotomy on raw terms, which is computationally a type synthesiser that checks if a given raw term is sufficiently annotated and if a sufficiently annotated raw term has an ordinary typing derivation, suggesting that type synthesis should not be viewed as a bisection between typable and untypable raw terms but a trichotomy in general.

\subsubsection{Mode-correctness and a general definition of typed languages}
\label{sec:language-formalisation}
The most essential characteristics of bidirectional typing is \emph{mode-correctness}, since an algorithm can often be `read off' from the definition of a bidirectionally typed language if mode-correct.
It seems that the implications of mode-correctness have only been treated informally so far as exhibited in the survey paper~\cite{Dunfield2021}, since the notion of mode-correctness is not formally defined for \emph{languages} there.
%But, what is a language anyway?

To make the notion of mode-correctness precise, we exhibit a general definition of bidirectional simple type systems which extends the sorted version of \varcitet{Aczel1978}{'s binding signature} with bidirectionality.
Such a general definition allows us to define mode-correctness and to investigate its consequences rigorously, including the uniqueness of synthesised types and the logical decidability of bidirectional type synthesis for \emph{all} mode-correct \emph{bidirectional binding signatures}.
The latter theorem effectively amounts to a verified proof-relevant bidirectional-type-synthesiser generator for all syntax-directed simply typed languages.

To make our exposition accessible we have chosen to work with simply typed languages so that bidirectional type synthesis is established without any other technical assumptions.
%the idea of extending a signature with bidirectionality should be clear enough for further generalisation.
Nevertheless, we briefly discuss another definition which includes polymorphically typed languages such as \SystemF and additional assumptions we need in \cref{sec:future}.

\subsubsection{Contributions and plan of this paper}

%
In short, we develop a general and formal theory of bidirectional type synthesis for simply typed languages, including 
\begin{enumerate}
  \item general definitions for bidirectional type systems and mode-correctness,
  \item the concept of mode decoration and the comparison between our completeness and \varcitet{Dunfield2021}{'s annotatability}, 
    \Josh{generalisation mode decoration?}
  \item rigorously proven consequences of mode-correctness---the uniqueness of synthesised types and the decidability of bidirectional type synthesis, which amounts to
  \item a fully verified generator for proof-relevant type-synthesisers.
\end{enumerate}
Our theory was first developed in \Agda and then translated to the mathematical vernacular.
\LT{methodology for combining proofs and programs?}

%\begin{enumerate}
%  \item \emph{simple yet general}, working for any syntax-directed bidirectional simple type system that can be specified by a bidirectional binding signature;
%  \item \emph{constructive} based on Martin-L\"of type theory, and compactly formulated in a way that unifies computation and proof, preferring logical decidability over algorithmic soundness, completeness, and decidability; and mode decoration over annotatability.
%\end{enumerate}

This paper is structured as follows.
We first present a concrete overview of our theory using simply typed $\lambda$-calculus in \cref{sec:key-ideas}, prior to developing a general framework for specifying bidirectional type systems in \cref{sec:defs}.
Following this, we discuss mode decoration and related properties in \cref{sec:pre-synthesis}.
The main technical contribution lies in \cref{sec:type-synthesis}, where we introduce mode-correctness and bidirectional type synthesis.
Some examples other than simply typed $\lambda$-calculus are given in \Cref{sec:example}.
We briefly demonstrate the use of our \Agda formalisation as programs in \cref{sec:formalisation} and conclude in \cref{sec:future} with further developments.

### 2023-10-11 tex/ESOP24/S2-outline.tex
[ update ] Add ESOP submission draft

--- 2023-10-03 tex/POPL24/sec2-outline.tex
@@
 
-\section{Bidirectional Type Synthesis for Simply Typed \texorpdfstring{$\lambda$}{λ}-Calculus} \label{sec:key-ideas}
+\section{Bidirectional type synthesis for simply typed \texorpdfstring{$\lambda$}{λ}-calculus} \label{sec:key-ideas}
 
@@
 We will construct a mode decorator~(\cref{sec:mode-decoration}) and a bidirectional type synthesiser~(\cref{sec:type-synthesis}) and prove both lemmas~(\cref{sec:soundness-and-completeness}) for all syntax-directed bidirectional simple type systems.
-To quantify over all such systems, we need their generic definitions, which we formulate next.
+To quantify over all such systems, we need their general definitions, which we formulate next.
 

### 2023-10-11 tex/ESOP24/S3-defs.tex
[ update ] Add ESOP submission draft

--- 2023-07-20 tex/sec3-defs.tex
@@
 
-\section{Simply Typed Languages and Bidirectional Type Systems}\label{sec:defs}
-This section provides generic definitions
+\section{Bidirectionally simply typed languages}\label{sec:defs}
+This section provides general definitions
 %\footnote{%
@@
 
-\subsection{Signatures and Simple Types} \label{subsec:simple-types}
+\subsection{Signatures and simple types} \label{subsec:simple-types}
 
@@
 \begin{definition} \label{def:simple-signature}
-  \begin{figure}
-      \centering
-      \small
-      \judgbox{\Xi|-_{\Sigma} A}{$A$ is a type with type variables in $\Xi$ for a signature $\Sigma$}
-      \begin{mathpar}
-        \inferrule{X_i \in \Xi}{\Xi |-_\Sigma X_i} \and
-        \inferrule{\Xi |-_{\Sigma} A_1 \\ \cdots \\ \Xi |-_{\Sigma} A_n}{\Xi |-_{\Sigma} \tyOp_i(A_1, \ldots, A_n)}\;\text{where $n = \arity(i)$}
-      \end{mathpar}
-      \caption{Simple types}
-      \label{fig:simple-type}
-  \end{figure}
+%  \begin{figure}
+%      \centering
+%      \small
+%      \judgbox{\Xi|-_{\Sigma} A}{$A$ is a type with type variables in $\Xi$ for a signature $\Sigma$}
+%      \begin{mathpar}
+%        \inferrule{X_i \in \Xi}{\Xi |-_\Sigma X_i} \and
+%        \inferrule{\Xi |-_{\Sigma} A_1 \\ \cdots \\ \Xi |-_{\Sigma} A_n}{\Xi |-_{\Sigma} \tyOp_i(A_1, \ldots, A_n)}\;\text{where $n = \arity(i)$}
+%      \end{mathpar}
+%      \caption{Simple types}
+%      \label{fig:simple-type}
+%  \end{figure}
   A \emph{signature} $\Sigma$ for simple types consists of a set $I$ with a decidable equality and an \emph{arity} function $\arity\colon I \to \mathbb{N}$.
@@
 
-\subsection{Binding Signatures and Simply Typed Languages} \label{subsec:binding-sig}
+\subsection{Binding signatures and simply typed languages} \label{subsec:binding-sig}
 
@@
 This means that the $\Rule{Abs}$ rule for derivations of\/ $\Gamma |- \lam{x}t : A \bto B$ contains:
-\begin{enumerate*}
+\begin{inlineenum}
   \item a derivation of\/ $\Gamma, x : A |- t : B$ as an argument of type $B$ in a typing context extended by a variable~$x$ of type~$A$;
   \item the type $A\bto B$ for itself.
-\end{enumerate*}
+\end{inlineenum}
 In a similar vein, the arity of the $\Rule{App}$ rule is denoted as ${(A \bto B), A} \to {B}$.
@@
 \end{definition}
-That is, each inhabitant $o: O$  is associated with a set $\Xi$ of type variables and a binding arity $\arity(o)$ with the sort $\Type_{\Sigma}(\Xi)$, denoted by $\bop$.
+That is, each inhabitant $o: O$  is associated with a set $\Xi$ of type variables and an arity $\arity(o)$ with the sort $\Type_{\Sigma}(\Xi)$ denoted by $\bop$.
 
@@
   \small
-  \judgbox{V |-_{\Sigma, \Omega} \isTerm{t}}{$\isTerm{t}$ is a raw term for a simply typed language $(\Sigma, \Omega)$ with free variables in~$V$}
+  \judgbox{V |-_{\Sigma, \Omega} \isTerm{t}}{$\isTerm{t}$ is a raw term for a language $(\Sigma, \Omega)$ with free variables in~$V$}
   \begin{mathpar}
@@
     \small
-    \judgbox{\Gamma |-_{\Sigma, \Omega} \isTerm{t} : A}{A raw term $\isTerm{t}$ has a concrete type $A$ under $\Gamma$ for a simply typed language $(\Sigma, \Omega)$}
+    \judgbox{\Gamma |-_{\Sigma, \Omega} \isTerm{t} : A}{$\isTerm{t}$ has a concrete type $A$ under $\Gamma$ for a language $(\Sigma, \Omega)$}
     \begin{mathpar}
@@
 
-\subsection{Bidirectional Binding Signatures and Bidirectional Type Systems} \label{subsec:bidirectional-system}
+\subsection{Bidirectional binding signatures and bidirectional type systems} \label{subsec:bidirectional-system}
 For a bidirectional type system, typing judgements appear in two forms: $\Gamma |- t \syn A$ and $\Gamma |- t \chk A$, but these two typing judgements can be considered as a single typing judgement $\Gamma |- t :^\dir{d} A$, additionally indexed by a \emph{mode} $d : \Mode$---which can be either $\syn$ or $\chk$.
@@
   \]
-  For clarity, a bidirectional binding arity is denoted by $\biarity$.
+  For clarity, an arity is denoted by $\biarity$.
 \end{definition}
@@
     \small
-    \judgbox{\Gamma |-_{\Sigma, \Omega} \isTerm{t} :^\dir{d} A}{$\isTerm{t}$ has a concrete type $A$ in mode $\dir{d}$ under $\Gamma$ for a bidirectional type system $(\Sigma, \Omega)$} 
+    \judgbox{\Gamma |-_{\Sigma, \Omega} \isTerm{t} :^\dir{d} A}{$\isTerm{t}$ has a type $A$ in mode $\dir{d}$ under $\Gamma$ for a bidirectional system $(\Sigma, \Omega)$} 
     \begin{mathpar}
@@
     \small
-    \judgbox{V |-_{\Sigma, \Omega} \isTerm{t}^\dir{d}}{$\isTerm{t}$ is in mode $d$ with free variables in $V$ for a bidirectional type system $(\Sigma, \Omega)$}
+    \judgbox{V |-_{\Sigma, \Omega} \isTerm{t}^\dir{d}}{$\isTerm{t}$ is in mode $d$ with free variables in $V$ for $(\Sigma, \Omega)$}
     \begin{mathpar}

### 2023-10-11 tex/ESOP24/S4-pre-synthesis.tex
[ update ] Add ESOP submission draft

--- 2023-07-20 tex/sec4-pre-synthesis.tex
@@
 
-\section{Mode Decoration and Related Properties}\label{sec:pre-synthesis}
+\section{Mode decoration and related properties}\label{sec:pre-synthesis}
 
@@
 
-\subsection{Soundness and Completeness}
+\subsection{Soundness and completeness}
 \label{sec:soundness-and-completeness}
@@
 
-\begin{proof}
-From left to right:
-Induction on the bidirectional typing derivation, mapping every rule to its counterpart except when constructing a typing derivation from the $\ChkRule{Sub}$ rule, in which case the induction hypothesis $\Gamma |-_{\Sigma,\erase\Omega} t : B$ suffices due to the premise $B = A$.
+%\begin{proof}
+%From left to right:
+%Induction on the bidirectional typing derivation, mapping every rule to its counterpart except when constructing a typing derivation from the $\ChkRule{Sub}$ rule, in which case the induction hypothesis $\Gamma |-_{\Sigma,\erase\Omega} t : B$ suffices due to the premise $B = A$.
+%
+%From right to left:
+%Induction on the mode derivation.
+%For \SynRule{Var}, \SynRule{Anno}, and \Rule{Op}, the outermost rule used in the typing derivation must be the corresponding typing rule, so by the induction hypotheses we have bidirectional typing derivations for all the sub-terms, to which we can then apply the corresponding bidirectional typing rule.
+%The \ChkRule{Sub} case is similar but slightly simpler: the induction hypothesis directly gives us a derivation of $\Gamma |-_{\Sigma,\Omega} t \syn A$, to which we apply $\ChkRule{Sub}$.
+%\end{proof}
 
-From right to left:
-Induction on the mode derivation.
-For \SynRule{Var}, \SynRule{Anno}, and \Rule{Op}, the outermost rule used in the typing derivation must be the corresponding typing rule, so by the induction hypotheses we have bidirectional typing derivations for all the sub-terms, to which we can then apply the corresponding bidirectional typing rule.
-The \ChkRule{Sub} case is similar but slightly simpler: the induction hypothesis directly gives us a derivation of $\Gamma |-_{\Sigma,\Omega} t \syn A$, to which we apply $\ChkRule{Sub}$.
-\end{proof}
-
-\subsection{Generalised Mode Decoration}
+\subsection{Generalised mode decoration}
 \label{sec:mode-decoration}
@@
     $\smash{\boxed{V |-_{\Sigma, \Omega} \isTerm{t}^{\dir{d}\,g\,s}}}$
-    & The raw term~$\isTerm{t}$\hspace{-.6em}
+    & $\isTerm{t}$ \hspace{.1em}
     & misses some type annotation iff $g = \BFalse$, and \\
@@
 
-\begin{proof}
-Induction on the given derivation.
-The $\SynRule{Missing}$ rule cannot appear because $g = \BTrue$, and the other rules are mapped to their counterparts.
-\end{proof}
+%\begin{proof}
+%Induction on the given derivation.
+%The $\SynRule{Missing}$ rule cannot appear because $g = \BTrue$, and the other rules are mapped to their counterparts.
+%\end{proof}
 
@@
 
-\begin{proof}[Proof of \cref{thm:Pre?-false}]
-Induction on the generalised mode derivation, in each case analysing an arbitrary mode derivation and showing that it cannot exist.
-The key case is $\SynRule{Missing}$, where we have a sub-derivation of $V |-_{\Sigma,\Omega} t^{\chk\,g\,\BTrue}$ for some boolean~$g$.
-We do not have an induction hypothesis and seem to get stuck, because $g$~is not necessarily~$\BFalse$.
-But what matters here is that $t$~is \emph{strictly} in checking mode: if we continue to analyse the sub-derivation, the outermost rule must be $\Rule{Op}$ with $\dir d = {\chk}$, implying that $t$~has to be an operation in checking mode.
-Then a case analysis shows that it is impossible to have a (synthesising) mode derivation of $V |-_{\Sigma,\Omega} t^{\syn}$.
-%\begin{itemize}
-%\item $\SynRule{Anno}$:
-%Any mode derivation must also use $\SynRule{Anno}$ as the outermost rule, and then the induction hypothesis suffices.
-%\item $\ChkRule{Sub}$:
-%We have a sub-derivation of $V |-_{\Sigma,\Omega} t^{\syn\,\BFalse\,\BTrue}$.
-%Any mode derivation of $V |-_{\Sigma,\Omega} t^{\chk}$ must use $\ChkRule{Chk}$ as the outermost rule.
-%(The $\Rule{Op}$ rule with $\dir d = {\chk}$ is impossible because the sub-derivation shows that $t$~is exactly a synthesising term.)
-%We then have a sub-derivation of $V |-_{\Sigma,\Omega} t^{\syn}$, and can use the induction hypothesis.
-%\end{itemize}
-\end{proof}
+%\begin{proof}[Proof of \cref{thm:Pre?-false}]
+%Induction on the generalised mode derivation, in each case analysing an arbitrary mode derivation and showing that it cannot exist.
+%The key case is $\SynRule{Missing}$, where we have a sub-derivation of $V |-_{\Sigma,\Omega} t^{\chk\,g\,\BTrue}$ for some boolean~$g$.
+%We do not have an induction hypothesis and seem to get stuck, because $g$~is not necessarily~$\BFalse$.
+%But what matters here is that $t$~is \emph{strictly} in checking mode: if we continue to analyse the sub-derivation, the outermost rule must be $\Rule{Op}$ with $\dir d = {\chk}$, implying that $t$~has to be an operation in checking mode.
+%Then a case analysis shows that it is impossible to have a (synthesising) mode derivation of $V |-_{\Sigma,\Omega} t^{\syn}$.
+%%\begin{itemize}
+%%\item $\SynRule{Anno}$:
+%%Any mode derivation must also use $\SynRule{Anno}$ as the outermost rule, and then the induction hypothesis suffices.
+%%\item $\ChkRule{Sub}$:
+%%We have a sub-derivation of $V |-_{\Sigma,\Omega} t^{\syn\,\BFalse\,\BTrue}$.
+%%Any mode derivation of $V |-_{\Sigma,\Omega} t^{\chk}$ must use $\ChkRule{Chk}$ as the outermost rule.
+%%(The $\Rule{Op}$ rule with $\dir d = {\chk}$ is impossible because the sub-derivation shows that $t$~is exactly a synthesising term.)
+%%We then have a sub-derivation of $V |-_{\Sigma,\Omega} t^{\syn}$, and can use the induction hypothesis.
+%%\end{itemize}
+%\end{proof}
 
@@
 
-\begin{proof}
-Case analysis on $\dir{d}$ and $\dir{d'}$, adding an outermost mode cast to change the given derivation to a different mode if $\dir{d} \neq \dir{d'}$.
-Note that it is permissible to add an outermost mode cast because the antecedent~(\ref{eq:strict-derivation}) requires the given derivation to be strict.
-\end{proof}
+%\begin{proof}
+%Case analysis on $\dir{d}$ and $\dir{d'}$, adding an outermost mode cast to change the given derivation to a different mode if $\dir{d} \neq \dir{d'}$.
+%Note that it is permissible to add an outermost mode cast because the antecedent~(\ref{eq:strict-derivation}) requires the given derivation to be strict.
+%\end{proof}
 
-\begin{proof}[Proof of \cref{thm:generalised-mode-decoration}]
-By \cref{thm:adjustment}, it suffices to prove statement~(\ref{eq:strict-derivation}) by induction on~$t$.
+\begin{proof}[\cref{thm:generalised-mode-decoration}]
+By \cref{thm:adjustment}, it suffices to prove~\eqref{eq:strict-derivation} by induction on~$t$.
 
@@
 By the induction hypotheses and \cref{thm:adjustment}, there is a derivation of $V, \vec x_i |-_{\Sigma,\Omega} {t_i}^{\dir{d_i}\,g_i\,s_i}$ for each sub-term~$t_i$; apply $\Rule{Op}$ to all the derivations.
-\vspace{-\topsep-\baselineskip}
 \end{itemize}
@@
 
-\begin{proof}
-By \cref{thm:generalised-mode-decoration}, there is a derivation of $V |-_{\Sigma,\erase\Omega} t^{\dir{d}\,g\,s}$ for some booleans $g$~and~$s$, and then simply check~$g$ and apply either \cref{thm:Pre?-true} or \cref{thm:Pre?-false} to obtain $V |-_{\Sigma,\Omega} t^\dir{d}$ or $V |/-_{\Sigma,\Omega} t^\dir{d}$.
-\end{proof}
+%\begin{proof}
+%By \cref{thm:generalised-mode-decoration}, there is a derivation of $V |-_{\Sigma,\erase\Omega} t^{\dir{d}\,g\,s}$ for some booleans $g$~and~$s$, and then simply check~$g$ and apply either \cref{thm:Pre?-true} or \cref{thm:Pre?-false} to obtain $V |-_{\Sigma,\Omega} t^\dir{d}$ or $V |/-_{\Sigma,\Omega} t^\dir{d}$.
+%\end{proof}
 
@@
   \centering\small
-  \judgbox{\isTerm{t} \sqsupseteq \isTerm{u}}{A raw term $t$ is more annotated than $u$ (for some bidirectional type system $(\Sigma, \Omega)$)}
+  \judgbox{\isTerm{t} \sqsupseteq \isTerm{u}}{A raw term $t$ is more annotated than $u$}
   \begin{mathpar}
@@
 This is the interesting case where we map a $\SynRule{Missing}$ rule to an $\SynRule{Anno}$ rule and add to the term a type annotation, which comes from the given typing derivation.
-\vspace{-\topsep-\baselineskip}
 \end{itemize}
@@
 Annotatability also does not help the user to deal with missing annotations like (generalised) mode decoration does: although annotatability seems capable of determining where annotations are missing and even filling them in correctly, its antecedent requires a typing derivation, which is what the user is trying to construct and does not have yet.
-Therefore we believe that the notion of annotatability is too complex for what it achieves, and our theory offers simpler and more useful alternatives.
+%Therefore we believe that the notion of annotatability is too complex\LT{not clear what it means} for what it achieves, and our theory offers simpler and more useful alternatives.

### 2023-10-11 tex/ESOP24/S5-synthesis.tex
[ update ] Add ESOP submission draft

--- 2023-07-24 tex/sec5-synthesis.tex
@@
 
-\section{Bidirectional Type Synthesis and Checking} \label{sec:type-synthesis}
+\section{Bidirectional type synthesis and checking} \label{sec:type-synthesis}
 This section focuses on defining mode-correctness and deriving bidirectional type synthesis for any mode-correct bidirectional type system $(\Sigma, \Omega)$.
@@
 
-\subsection{Mode Correctness}\label{sec:mode-correctness}
+\subsection{Mode correctness}\label{sec:mode-correctness}
 As \citet{Dunfield2021} outlined, mode-correctness for a bidirectional typing rule means that 
-\begin{enumerate*}
+\begin{inlineenum}
 \item each `input' type variable in a premise must be an `output' variable in `earlier' premises, or provided by the conclusion if the rule is checking;
 \item each `output' type variable in the conclusion should be some `output' variable in a premise if the rule is synthesising.
-\end{enumerate*}
+\end{inlineenum}
 Here `input' variables refer to variables in an extension context and in a checking premise.
@@
 Likewise, the rule $\SynRule{App}$ for an application term $\isTerm{t\;u}$ is mode-correct because:
-\begin{enumerate*}
+\begin{inlineenum}
 \item the type $A \bto B$ of the first argument $t$ is synthesised, thereby ensuring type variables $A$ and $B$ must be known if successfully synthesised;
@@
 \item as a result, the type of an application $t\;u$ can be synthesised.
-\end{enumerate*}
+\end{inlineenum}
 
@@
   The finite subset\footnote{%
-  There are various definitions for finite subsets of a set within Martin-L\"{o}f type theory.
-  However, for our purposes, the choice among these definitions is not a matter of concern.}
+  There are various definitions for finite subsets of a set within type theory, but for our purposes the choice among these definitions is not a matter of concern.}
   of \emph{(free) variables} of a type $A$ is denoted by $\fv(A)$.
@@
 \begin{definition}\label{def:mode-correctness-args}
-  The \emph{mode-correctness} $\MCas\left(\biargvec\right)$ for an argument list $\biargs$ with respect to a subset $S$ of $\Xi$ is a (proof-relevant) predicate defined by \LT{is it a prop?}
+  The \emph{mode-correctness} $\MCas\left(\biargvec\right)$ for an argument list $\biargs$ with respect to a subset $S$ of $\Xi$ is a (proof-relevant) predicate defined by
   \begin{align}
@@
 
+\LT{decidability of mode-correctness}
 Now, we set out to show the uniqueness of synthesised types for a mode-correct bidirectional type system.
@@
       \end{enumerate}
-    \vspace{-\topsep-\baselineskip}
   \end{itemize}
@@
 
-\subsection{Decidability of Bidirectional Type Synthesis and Checking}\label{subsec:bidirectional-synthesis-checking}
+\subsection{Decidability of bidirectional type synthesis and checking}\label{subsec:bidirectional-synthesis-checking}
 
@@
 
-\begin{theorem}[Decidability of Bidirectional Type Synthesis and Checking] \label{thm:bidirectional-type-synthesis-checking}
+\begin{theorem}\label{thm:bidirectional-type-synthesis-checking}
   In a mode-correct bidirectional type system $(\Sigma, \Omega)$,
@@
 This lemma can be inferred from the correctness of first-order unification~\citep{McBride2003,McBride2003a}, or be proved directly without unification.
-We are now ready for the decidability proof.
+We are now ready for the promised proof.
 
-\begin{proof}[Proof of {\Cref{thm:bidirectional-type-synthesis-checking}}]
+\begin{proof}[of {\Cref{thm:bidirectional-type-synthesis-checking}}]
   We prove this statement by induction on the mode derivation\/ $\erase{\Gamma} |-_{\Sigma, \Omega} \isTerm{t}^{\dir{d}}$.
@@
       \end{itemize}
-      \begin{claimproof}
+      \begin{proof}[of Claim]
         We prove it by induction on the list $\biargs$:
@@
         \end{enumerate}
-        Therefore, we have proved our claim for any argument list by induction.
-      \end{claimproof}
+        Therefore, we have proved our claim for any argument list by induction.\hfill
+        $\blacksquare$
+      \end{proof}
   \end{itemize}
-  We now have completed the decidability proof by induction on the mode derivation $\erase{\Gamma} |-_{\Sigma, \Omega} t^{\dir{d}}$.
+  We have completed the proof by induction on the mode derivation $\erase{\Gamma} |-_{\Sigma, \Omega} t^{\dir{d}}$.\qed
 \end{proof}
@@
 Each case analysis simply branches depending on the outcomes of bidirectional type synthesis and checking for each sub-term, as well as the unification process.
-If a contradiction proof is not of interest for implementation, these programs can be simplified by discarding the cases that yield such contradiction proofs.
+If a negation proof is not of interest, these programs can be simplified by discarding the cases that yield such negation proofs.
 Alternatively, we could consider generalising typing derivations instead, like our generalised mode derivations (\Cref{fig:generalised-mode-derivations}).
@@
 
-\subsection{Trichotomy on Raw Terms by Type Synthesis} \label{subsec:trichotomy}
+\subsection{Trichotomy on raw terms by type synthesis} \label{subsec:trichotomy}
 
@@
   exactly one of the following holds:
-  \LT{Why downplay the pre-condition?}
   \begin{enumerate}
-    \item (\/$\erase{\Gamma} |-_{\Sigma, \Omega} \isTerm{t}^{\syn}$ and)\/ $\Gamma |-_{\Sigma, \Omega} \isTerm{t} : A$ for some type~$A$.
-    \item (\/$\erase{\Gamma} |-_{\Sigma, \Omega} \isTerm{t}^{\syn}$ but)\/ $\Gamma |/-_{\Sigma, \Omega} \isTerm{t} : A$ for any type~$A$.
+    \item $\Gamma |-_{\Sigma, \Omega} \isTerm{t} : A$ for some type~$A$.
+    \item $\Gamma |/-_{\Sigma, \Omega} \isTerm{t} : A$ for any type~$A$.
     \item \phantom{(\/$\erase{\Gamma} |-_{\Sigma, \Omega} \isTerm{t}^{\syn}$}\llap{$\erase{\Gamma} |/-_{\Sigma, \Omega} \isTerm{t}^{\syn}$\kern.5pt}.

### 2023-10-11 tex/ESOP24/S8-future.tex
[ update ] Add ESOP submission draft

%!TEX root = BiSig.tex

\section{Discussion and Future work} \label{sec:future}

\subsection{Related work}
\citet{Harper1993a,Harper2007}

Our general definitions are akin to the definitions used in other language formalisation frameworks~\citep{Ahrens2022,Allais2021,Fiore2022} in a proof assistant---the concept of a signature is at object level (i.e.\ an inhabitant of a record type in \Agda).
This allows us to exploit the expressive power of our meta-language while defining a language, so it is possible to express a construct that takes an indeterminate number of arguments such as spine application.
%\todo[inline,caption={}]{\begin{enumerate}
%\item `Input' and `output' of our theory, comparing with grammars and parsing
%\item A more comprehensive theory than Wadler et al., and more formal and refined than Dunfield et al.'
%\item Restrictions to simple types (not individual language features) and syntax-directedness
%\end{enumerate}}

%which together describe an \emph{extrinsically typed} and \emph{syntax-directed} bidirectional simple type system~(\cref{fig:raw-terms,fig:bidirectional-typing-derivations}).
%
%There are many ways and extensions for designing bidirectional typing rules, .
%Yet, once the definition of a bidirectional type system is in place, deriving an algorithm becomes straightforward.
%Accordingly, our theory's objective is not to formulate rules for various language features, but to introduce a formalism for bidirectional typing.
%\todo{Looks like we should have axiomatised what types we can deal with; left as future work}
\paragraph{More characteristics of bidirectional typing}

We have formalised some of the most important concepts discussed by \citet{Dunfield2021}.
There are other concepts that may be formalisable with some more effort, for example Pfenning's recipe for bidirectionalising typing rules.
We believe that our theory lays the foundation for further formalising these concepts.
There are some other concepts that may be more difficult to pin down, notably `annotation character', which is roughly about how easy it is for the user to write annotated programs.
These may continue to require creativity from the designer of type systems.

%\paragraph{Beyond syntax-directed type systems}
%
%As a first theory of bidirectional type synthesis, our work focuses on syntax-directed type systems.
%To explore more general settings, an extreme possibility is to independently specify an ordinary type system and a bidirectional one over the same raw terms and then investigate possible relationships between the two systems, but a more manageable and practical setting suitable for a next step is probably one that only mildly generalises ours: 
%for each raw term construct, there can be several ordinary typing rules, and each typing rule can be refined to several bidirectional typing rules with different mode assignments.
%Even for this mildly generalised setting, there is already a lot more work to do:
%Both the mode decorator and the bidirectional type synthesiser will have to backtrack, and become significantly more complex.
%For soundness and completeness, it should still be possible to treat them as the separation and combination of mode and type information, but the completeness direction will pose a problem---for every node of a raw term, a mode derivation chooses a mode assignment while a typing derivation chooses a typing rule, but there may not be a bidirectional typing rule for this particular combination.
%There should be ways to fix this problem while retaining the mode decoration phase---for example, one idea is to make the mode decorator produce all possible mode derivations, and refine completeness to say that any typing derivation can be combined with one of these mode derivations into a bidirectional typing derivation.
%All these considerations are rather too complex for a first theory, however.
%
%%\Josh[inline, caption={}]{Getting rid of the assumption of being syntax-directed (the triangular picture) --- 0.5p
%%\begin{enumerate}
%%\item In general: two independently formulated type systems over raw terms
%%\item More sensible next step: a `refinement chain'
%%\item Backtracking algorithms are required. for both mode decoration and bidirectional type synthesis.
%%\item The idea of separating and combining mode and type information should still work, but requires more care in the completeness direction.
%%\item Too complex for a first theory, and left as future work
%%\end{enumerate}}
%
\paragraph{Beyond simple types}
We leave the problem of presenting bidirectional typing for more general and advanced language features as future work (such as those related to polymorphic types~\citep{Pierce2000,Peyton-Jones2007,Dunfield2013,Xie2018}).
While algebraic approaches to polymorphic types have been developed~\citep{Fiore2013,Hamana2011}, these approaches do not take subtyping into account.
Subtyping is essential for formulating important concepts such as \emph{principal types} in type synthesis.

%%\paragraph{Whither extrinsic typing?}
%
%%\todo[inline]{We leave the problem of presenting bidirectional typing for more general and advanced language features as future work (such as those related to polymorphic types~\citep{Pierce2000,Peyton-Jones2007,Dunfield2013,Xie2018}), which requires advancement in language formalisation~(\cref{sec:language-formalisation}) that is orthogonal to our work.}
%
%%\Josh[inline]{Optional: ornaments (0.25p), NDGP (0.25p)}
%
%%\subsubsection{Theories of abstract syntax with variable binding}
%%\label{sec:theory-of-syntax}
%%
%
%%Substitution is also modelled categorically, but it does not play a role in this paper.%
%%\todo{Reveal a bit about this paper?}
%
%%\begin{remark} \label{re:type-signature}
%%Most of existing theories treat types independently of terms, thereby excluding them from signatures.
%%To the best of our knowledge, the only exception to this approach is found in the work of~\citet{Arkor2020}, which incorporates signatures for both terms and types.
%%Interestingly, this inclusion is also critical for type synthesis for comparing a concrete type $N \bto N$ with an abstract type $A \bto B$, where $A$ and $B$ are type variables in a typing rule.
%%\end{remark}

### 2023-10-11 tex/ESOP24/appendix.tex
[ update ] Add ESOP submission draft

--- 2023-07-24 tex/sec5-synthesis.tex
@@
 
-\section{Bidirectional Type Synthesis and Checking} \label{sec:type-synthesis}
-This section focuses on defining mode-correctness and deriving bidirectional type synthesis for any mode-correct bidirectional type system $(\Sigma, \Omega)$.
-We start with \Cref{sec:mode-correctness} by defining mode-correctness and showing the uniqueness of synthesised types.
-This uniqueness means that any two synthesised types for the same raw term $t$ under the same context $\Gamma$ have to be equal.
-It will be used especially in \Cref{subsec:bidirectional-synthesis-checking} for the proof of the decidability of bidirectional type synthesis and checking.
-Then, we conclude this section with the trichotomy on raw terms in \Cref{subsec:trichotomy}.
+\section{Detailed definitions and proofs}\label{sec:defs-proofs}
 
@@
 %On the other hand, our proof is based on mode-correctness and need not consider specific sub-derivations.
-
-\subsection{Decidability of Bidirectional Type Synthesis and Checking}\label{subsec:bidirectional-synthesis-checking}
-
-We have arrived at the main technical contribution of this paper.
-
-\begin{theorem}[Decidability of Bidirectional Type Synthesis and Checking] \label{thm:bidirectional-type-synthesis-checking}
-  In a mode-correct bidirectional type system $(\Sigma, \Omega)$,
-  \begin{enumerate}
-    \item if\/ $\erase{\Gamma} |-_{\Sigma, \Omega} \isTerm{t}^{\syn}$, then it is decidable whether $\Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn A$ for some $A$;
-    \item if\/ $\erase{\Gamma} |-_{\Sigma, \Omega} \isTerm{t}^{\chk}$, then it is decidable for any~$A$ whether $\Gamma |-_{\Sigma, \Omega} \isTerm{t} \chk A$.
-  \end{enumerate}
-\end{theorem}
-
-The interesting part of the theorem is the case for the $\Rule{Op}$ rule and it is rather complex.
-Here we shall give its insight first instead of jumping into the details.
-Recall that a typing derivation for $\tmOpts$ contains a substitution $\rho\colon \Xi \to \Type_{\Sigma}(\emptyset)$.
-The goal of type synthesis is exactly to define such a substitution $\rho$, so we have to start with an `accumulating' substitution: a substitution $\rho_0$ that is partially defined on $\fv(A_0)$ if $d$ is $\chk$ or otherwise nowhere.
-By mode-correctness, the accumulating substitution~$\rho_i$ will be defined on enough synthesised variables so that type synthesis or checking can be performed on $t_{i}$ with the context $\Gamma, \vars{x}_{i} : \simsub{\Delta_{i}}{\rho_{i}}$ based on its mode derivation $\erase{\Gamma}, \vars{x}_i |-_{\Sigma, \Omega} t_i^{\dir{d_i}}$.
-If we visit a synthesising argument $\synbiarg[i+1]$, then we may extend the domain of $\rho_i$ to include the synthesised variables $\fv(A_{i + 1})$ if type synthesis is successful and also that the synthesised type can be \emph{unified with $A_{i+ 1}$} and thereby \emph{extend} $\rho_i$ to $\bar{\rho_i} = \rho_{i+1}$ with the unifier.
-Then, if we go through every $t_i$ successfully, we will have a total substitution $\rho_n$ by mode-correctness and a derivation of $\Gamma, \vars{x}_i : \Delta_i |-_{\Sigma, \Omega} t_i :^{\dir{d_i}} \simsub{A}{\rho_n}$ for each sub-term $t_i$.
-
-\begin{remark}
-To make the argument above sound, it is necessary to compare types and solve a unification problem.
-Hence, we assume that the set~$\Xi$ of type variables has a decidable equality, thereby ensuring that the set $\Type_{\Sigma}(\Xi)$ of types also has a decidable equality.\footnote{%
-To simplify our choice, we could simply confine $\Xi$ to any set within the family of sets $\Fin(n)$ of naturals less than~$n$, given that these sets have a decidable equality and the arity of a type construct is finite.
-Indeed, in our formalisation, we adopt $\Fin(n)$ as the set of type variables in the definition of $\Type_{\Sigma}$ (see \Cref{sec:formalisation} for details).
-For the sake of clarity in presentation, though, we keep using named variables and just assume that $\Xi$ has a decidable equality.}
-\end{remark}
-We need some auxiliary definitions for the notion of extension to state the unification problem:
-\begin{definition}
-By an \emph{extension}\/ $\sigma \geq \rho$ of a partial substitution $\rho$ we mean that the domain $\dom(\sigma)$ of $\sigma$ contains the domain of $\rho$ and $\sigma(x) = \rho(x)$ for every\/ $x$ in $\dom(\rho)$.
-  By a \emph{minimal extension}\/ $\bar{\rho}$ of $\rho$ satisfying $P$ we mean an extension $\bar{\rho} \geq \rho$ with $P(\bar{\rho})$ such that $\sigma \geq \bar{\rho}$ whenever $\sigma \geq \rho$ and $P(\sigma)$.
-\end{definition}
-\begin{lemma}\label{lem:unify}
-  For any\/ $A$ of\/ $\Type_{\Sigma}(\Xi)$, $B$ of\/ $\Type_{\Sigma}(\emptyset)$, and a partial substitution\/ $\rho \colon \Xi \to \Type_{\Sigma}(\emptyset)$, 
-  \begin{enumerate}
-    \item either there is a minimal extension\/ $\bar{\rho}$ of\/ $\rho$ such that\/ $\simsub{A}{\bar{\rho}} = B$,
-    \item or there is no extension\/ $\sigma$ of\/ $\rho$ such that\/ $\simsub{A}{\sigma} = B$
-  \end{enumerate}
-\end{lemma}
-This lemma can be inferred from the correctness of first-order unification~\citep{McBride2003,McBride2003a}, or be proved directly without unification.
-We are now ready for the decidability proof.
-
-\begin{proof}[Proof of {\Cref{thm:bidirectional-type-synthesis-checking}}]
-  We prove this statement by induction on the mode derivation\/ $\erase{\Gamma} |-_{\Sigma, \Omega} \isTerm{t}^{\dir{d}}$.
-  The two cases \SynRule{Var} and \SynRule{Anno} are straightforward and independent of mode-correctness.
-  The case \ChkRule{Sub} invokes the uniqueness of synthesised types to refute the case that $\Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn B$ but $A \neq B$ for a given type $A$.
-  The first three cases follow essentially the same reasoning provided by \citet{Wadler2022}, but we still present the reasoning here for the sake of completeness.
-  The last case \Rule{Op} is new and has been discussed above.
-  For brevity we omit the subscript $(\Sigma, \Omega)$.
-  \begin{itemize}
-    \item \SynRule{Var}: If $\erase{\Gamma} |- \isTerm{t}^{\syn}$, then $(x : A) \in \Gamma$ and thus $\Gamma |- \isTerm{x} \syn A$.
-
-    \item \SynRule{Anno}: For $\erase{\Gamma} |- (\isTerm{t} \annotate A)^{\syn}$, it is decidable whether $\Gamma |- \isTerm{t} \chk A$ by induction hypothesis.
-      \begin{itemize}
-        \item If $\Gamma |- \isTerm{t} \chk A$, then $\Gamma |- \isTerm{t \annotate A} \syn A$.
-        \item If $\Gamma |/- t \chk A$ but $\Gamma |- \isTerm{t \annotate A} \syn$, then by inversion $\Gamma |- t \chk A$, leading to a contradiction.
-      \end{itemize}
-
-    \item \ChkRule{Sub}: If $\erase{\Gamma} |-t^{\chk}$, then $\Gamma |- t^{\syn}$ by inversion.
-      By induction hypothesis, we have two cases: %it is decidable whether $\Gamma |- t \syn B$ for some $B$:
-      \begin{itemize}
-        \item If $\Gamma |/- t \syn C$ for any $C$ but $\Gamma |- t \chk A$, then by inversion $\Gamma |- t \syn B$ for some $B = A$, thus a contradiction.
-        \item If $\Gamma |- t \syn B$ for some $B$, then by decidable equality on $\Type_{\Sigma}(\Xi)$ either $A = B$ or $A \neq B$: 
-          \begin{itemize}
-            \item if $A = B$ then we have\/ $\Gamma |- t \chk A$;
-            \item if $A \neq B$ but $\Gamma |- t \chk A$, then by inversion $\Gamma |- t \syn A$.
-              However, by \Cref{thm:unique-syn}, synthesised types $A$ and $B$ must be equal, so we derive a contradiction.
-          \end{itemize}
-      \end{itemize}
-    \item \Rule{Op}:
-      For a mode derivation of $\erase{\Gamma} |- \tmOpts^{\dir{d}}$, we first claim:
-      \begin{claim}\label{lem:args-induction}
-        For an argument list $\biargs$ and any partial substitution $\rho$ from $\Xi$ to $\emptyset$
-        \begin{enumerate}
-          \item either there is a minimal extension $\ext{\rho}$ of $\rho$ such that 
-            \begin{equation} \label{eq:claim}
-              \dom(\ext{\rho}) \supseteq \synvar(\biargs)
-              \quad\text{and}\quad
-%              \text{the domain of $\ext{\rho}$ contains $\synvar(\biargs)$ and\/} \quad
-              \Gamma, \vars{x}_\isTerm{i} : \simsub{\Delta_i}{\ext{\rho}} |- \isTerm{t_i} \colon \simsub{A_i}{\ext{\rho}}^{\dir{d_i}}
-              \quad\text{for $i = 1, \ldots, n$};
-            \end{equation}
-          \item or there is no extension $\sigma$ of $\rho$ such that \eqref{eq:claim} holds.
-        \end{enumerate}
-      \end{claim}
-
-      Then, we proceed with a case analysis on $\dir{d}$ in the mode derivation:
-      \begin{itemize}
-        \item $\dir{d}$ is $\syn$: We apply our claim with the partial substitution $\rho_0$ defined nowhere.
-          \begin{enumerate}
-            \item If there is no $\sigma \geq \rho$ such that \eqref{eq:claim} holds but $\Gamma |- \tmOpts \syn A$ for some $A$, then by inversion we have a substitution $\rho\colon \Sub{\Xi}{\emptyset}$ such that
-              \[
-                \Gamma, \vars{x}_\isTerm{i} : \simsub{\Delta_i}{\rho} |- \isTerm{t_i} \colon \simsub{A_i}{\rho}^{\dir{d_i}}
-              \]
-              for every $i$.
-              Obviously, $\rho \geq \rho_0$ and ${\Gamma, \vars{x}_\isTerm{i} : \simsub{\Delta_i}{\rho} |- \isTerm{t_i} \colon \simsub{A_i}{\rho}^{\dir{d_i}}}$ for every $i$ and it contradicts the assumption that no such an extension exists.
-
-            \item If there exists a minimal $\ext{\rho} \geq \rho_0$ defined on $\synvar(\biargs)$ such that \eqref{eq:claim} holds, then by mode-correctness $\ext{\rho}$ is total and thus
-              \[
-                \Gamma |- \tmOpts \syn \simsub{A_0}{\ext{\rho}}.
-              \]
-          \end{enumerate}
-
-        \item $\dir{d}$ is $\chk$: Let $A$ be a type and apply \Cref{lem:unify} with $\rho_0$ defined nowhere.
-          \begin{enumerate}
-            \item If there is no $\sigma \geq \rho_0$ such that $\simsub{A_0}{\sigma} = A$ but $\Gamma |- \tmOpts \chk A$, then by inversion there is a substitution $\rho$ such that $A = \simsub{A_0}{\rho}$, thus a contradiction.
-            \item If there is a minimal $\ext{\rho} \geq \rho_0$ such that $\simsub{A_0}{\ext{\rho}} = A$, then apply our claim with $\ext{\rho}$:
-              \begin{enumerate}
-                \item If there is no $\sigma \geq \ext{\rho}$ satisfying \eqref{eq:claim} but $\Gamma |- \tmOpts \chk A$, then by inversion there is $\gamma$ such that $\simsub{A_0}{\gamma} = A$ and also ${\Gamma, \vars{x}_\isTerm{i} : \simsub{\Delta_i}{\gamma} |- \isTerm{t_i} \colon \simsub{A_i}{\gamma}^{\dir{d_i}}}$ for every $i$.
-                  Given that $\ext{\rho} \geq \rho$ is minimal such that $\simsub{A_0}{\ext{\rho}} = A$, then $\gamma$ is an extension of $\ext{\rho}$ but by assumption no such an extension satisfying ${\Gamma, \vars{x}_\isTerm{i} : \simsub{\Delta_i}{\gamma} |- \isTerm{t_i} \colon \simsub{A_i}{\gamma}^{\dir{d_i}}}$ exists, thus a contradiction.
-
-                \item If there is a minimal $\ext{\ext{\rho}} \geq \ext{\rho}$ s.t.\ \eqref{eq:claim}, then by mode-correctness $\ext{\ext{\rho}}$ is total and
-                  \[
-                    \Gamma |- \tmOpts \chk \simsub{A_0}{\ext{\ext{\rho}}}
-                  \]
-                  where $\simsub{A_0}{\ext{\ext{\rho}}} = \simsub{A_0}{\ext{\rho}} = A$ since $\ext{\ext{\rho}}(x) = \ext{\rho}$ for every $x$ in the domain of $\ext{\rho}$.
-              \end{enumerate}
-          \end{enumerate}
-      \end{itemize}
-      \begin{claimproof}
-        We prove it by induction on the list $\biargs$:
-        \begin{enumerate}
-          \item For the empty list, $\rho$ is the minimal extension of $\rho$ itself satisfying \eqref{eq:claim} trivially. 
-          \item For $\biargvec, \biarg[m+1]$, by induction hypothesis on the list, we have two cases:
-            \begin{enumerate}
-              \item If there is no $\sigma \geq \rho$ such that \eqref{eq:claim} holds for all $1 \leq i \leq m$ but a minimal $\gamma \geq \rho$ such that~\eqref{eq:claim} holds for all $1 \leq i \leq m + 1$, then we clearly have a contradiction.
-              \item There is a minimal $\ext{\rho} \geq \rho$ s.t.\ \eqref{eq:claim} holds for $1 \leq i \leq m$.
-                By case analysis on $\dir{d_{m+1}}$:
-                \begin{itemize}
-                  \item $\dir{d_{m+1}}$ is $\chk$: By mode-correctness, $\simsub{\Delta_{m+1}}{\ext{\rho}}$ and $\simsub{A_{m+1}}{\ext{\rho}}$ are defined.
-                    By the induction hypothesis $ \Gamma, \vars{x}_\isTerm{m+1} : \simsub{\Delta_{m+1}}{\ext{\rho}} |- \isTerm{t_{m+1}} \chk \simsub{A_{m+1}}{\ext{\rho}}$ is decidable.
-                    Clearly, if $\Gamma, \vars{x}_\isTerm{m+1} : \simsub{\Delta_{m+1}}{\ext{\rho}} |- \isTerm{t_{m+1}} \chk \simsub{A_{m+1}}{\ext{\rho}}$ then the desired statement is proved; otherwise we can also easily derive a contradiction.
-
-                  \item $\dir{d_{m+1}}$ is $\syn$: By mode-correctness, $\simsub{\Delta_{m+1}}{\ext{\rho}}$ is defined.
-                    By the induction hypothesis, it is decidable that $\Gamma, \vars{x}_\isTerm{m+1} : \simsub{\Delta_{m+1}}{\ext{\rho}} |- \isTerm{t_{m+1}} \syn A$ for some $A$.
-                    \begin{enumerate}
-                      \item If $\Gamma, \vars{x}_\isTerm{m+1} : \simsub{\Delta_{m+1}}{\ext{\rho}} |/- \isTerm{t_{m+1}} \syn A$ for any $A$ but there is $\gamma \geq \rho$ s.t.\ \eqref{eq:claim} holds for $1 \leq i \leq m+1$, then $\gamma \geq \ext{\rho}$.
-                        Therefore $\simsub{\Delta_{m+1}}{\ext{\rho}} = \simsub{\Delta_{m+1}}{\gamma}$ and we derive a contradiction because $\Gamma, \vars{x}_\isTerm{m+1} : \simsub{\Delta_{m+1}}{\ext{\rho}} |- \isTerm{t_{m+1}} \syn \simsub{A_{m+1}}{\gamma}$.
-                      \item If $\Gamma, \vars{x}_\isTerm{m+1} : \simsub{\Delta_{m+1}}{\ext{\rho}} |- \isTerm{t_{m+1}} \syn A$ for some $A$, then by \Cref{lem:unify}: % we have two cases: %we unify $A$ with $A_{m+1}$ extending $\ext{\rho}$.
-                        \begin{itemize}
-                          \item Suppose no $\sigma \geq \ext{\rho}$ such that $\simsub{A_{m+1}}{\sigma} = A$ but an extension $\gamma \geq \rho$ such that \eqref{eq:claim} holds for $1 \leq i \leq m + 1$.
-                            Then, $\gamma \geq \ext{\rho}$ by the minimality of $\ext{\rho}$ and thus
-                            $\Gamma, \vars{x}_\isTerm{m+1} : \simsub{\Delta_{m+1}}{\ext{\rho}} |- \isTerm{t_{m+1}} \syn \simsub{A_{m+1}}{\gamma}$.
-                            However, by \Cref{thm:unique-syn}, the synthesised type $\simsub{A_{m+1}}{\gamma}$ must be unique, so $\gamma$ is an extension of $\ext{\rho}$ such that $\simsub{A_{m+1}}{\gamma} = A$, leading to a contradiction.
-                          \item If there is a minimal $\ext{\ext{\rho}} \geq \ext{\rho}$ such that $\simsub{A_{m+1}}{\ext{\ext{\rho}}} = A$, then it is not hard to show that $\ext{\ext{\rho}}$ is also the minimal extension of $\rho$ such that \eqref{eq:claim} holds for all $1 \leq i \leq m + 1$.
-                        \end{itemize}
-                    \end{enumerate}
-                \end{itemize}
-            \end{enumerate}
-        \end{enumerate}
-        Therefore, we have proved our claim for any argument list by induction.
-      \end{claimproof}
-  \end{itemize}
-  We now have completed the decidability proof by induction on the mode derivation $\erase{\Gamma} |-_{\Sigma, \Omega} t^{\dir{d}}$.
-\end{proof}
-
-The formal counterpart of the above proof in \Agda functions as two top-level programs for type checking and synthesis.
-These programs either compute the typing derivation or provide a proof of contradiction.
-Each case analysis simply branches depending on the outcomes of bidirectional type synthesis and checking for each sub-term, as well as the unification process.
-If a contradiction proof is not of interest for implementation, these programs can be simplified by discarding the cases that yield such contradiction proofs.
-Alternatively, we could consider generalising typing derivations instead, like our generalised mode derivations (\Cref{fig:generalised-mode-derivations}).
-This could accommodate ill-typed cases and reformulate contradiction proofs positively to deliver more informative error messages.
-This would assist programmers in resolving issues with ill-typed terms, rather than simply returning a blatant `no'.
-
-\subsection{Trichotomy on Raw Terms by Type Synthesis} \label{subsec:trichotomy}
-
-Combining the bidirectional type synthesiser with the mode decorator, soundness, and completeness from \cref{sec:pre-synthesis}, we get a type synthesiser parameterised by $(\Sigma, \Omega)$, generalising \cref{thm:implementation}.
-
-\begin{corollary}[Trichotomy on Raw Terms]\label{cor:trichotomy}
-  For any mode-correct bidirectional type system $(\Sigma, \Omega)$, 
-  exactly one of the following holds:
-  \LT{Why downplay the pre-condition?}
-  \begin{enumerate}
-    \item (\/$\erase{\Gamma} |-_{\Sigma, \Omega} \isTerm{t}^{\syn}$ and)\/ $\Gamma |-_{\Sigma, \Omega} \isTerm{t} : A$ for some type~$A$.
-    \item (\/$\erase{\Gamma} |-_{\Sigma, \Omega} \isTerm{t}^{\syn}$ but)\/ $\Gamma |/-_{\Sigma, \Omega} \isTerm{t} : A$ for any type~$A$.
-    \item \phantom{(\/$\erase{\Gamma} |-_{\Sigma, \Omega} \isTerm{t}^{\syn}$}\llap{$\erase{\Gamma} |/-_{\Sigma, \Omega} \isTerm{t}^{\syn}$\kern.5pt}.
-  \end{enumerate}
-\end{corollary}

### 2023-10-11 tex/ESOP24/S1-intro.tex
[ update ] Add the missing .bib

@@
 In light of these limitations, \emph{bidirectional type synthesis} emerged as a viable alternative, providing algorithms for deciding the types of raw terms that meet some syntactic criteria and usually contain type annotations.
-\citet{Dunfield2021} summarised the design principles of bidirectional type synthesis and its wide coverage of languages with simple types, polymorphic types, dependent types, gradual types, among others.
+\varcitet{Dunfield2021}{ in their survey paper} summarised the design principles of bidirectional type synthesis and its wide coverage of languages with simple types, polymorphic types, dependent types, gradual types, among others.
 
@@
 While it is straightforward to derive a type synthesis algorithm,
-For bidirectional typing, \citeauthor{Dunfield2021} only present informal design principles that the community learned from individual systems, rather than a general treatment providing logical specifications and rigorously proven properties for a class of systems.
-Moreover, unlike the plethora of available parser generators, `type-synthesiser generators' rarely exist, so each type synthesiser has to be independently built from scratch.
-\LT{emphasise that formalisation and verification are important}
+for bidirectional typing, \citeauthor{Dunfield2021} only present informal design principles that the community learned from individual systems, rather than a general treatment providing logical specifications and rigorously proven properties for a class of systems.
+Moreover, unlike the plethora of available parser generators, `type-synthesiser generators' rarely exist, so each type synthesiser has to be independently built, not to mention their correctness.
 
-To remedy the situation, we develop a general and formal treatment of bidirectional typing, providing an out-of-the-box \emph{verified generator} of \emph{proof-relevant} type synthesisers developed with the proof assistant \Agda:
+Implementing a type synthesiser correctly, especially for advanced type systems, is non-trivial.
+For example, proof assistants in use including \Coq and \Agda have many bugs in their type synthesis such as absurdity, supposedly equal terms which were not normalised to the same canonical term, and so on.
+\LT{some GitHub issues \cite{Felicissimo2023}}
+The gap between the design of type synthesis and its implementation cannot be ignored.
 
-\subsubsection{Proof-relevance in (bidirectional) type synthesis}
+To mitigate the above problem, we demonstrate a general and formal treatment of bidirectional typing, providing an out-of-the-box \emph{verified generator} of \emph{proof-relevant} type synthesisers developed with the proof assistant \Agda:
+
+\subsubsection{Proof-relevance in type synthesis}
 \label{sec:PLFA}
@@
 $\forall x, y.~(x = y) + (x \neq y)$, decides whether $x$~and~$y$ are equal and accordingly gives an identity proof or a refutation explicitly; such a decidability proof may or may not be possible depending on the domain of $x$~and~$y$, and is non-trivial.
-Given that all proofs as programs terminate, logical decidability implies algorithmic decidability.
+In type theory all proofs as programs terminate, so logical decidability implies algorithmic decidability.
 Further, suppose a proof of the following statement:
@@
 It yields either a typing derivation for the given raw term~$t$ or a negation proof that such a derivation is impossible where the former case is algorithmic soundness and the latter algorithmic completeness in contrapositive form.
-Then, both algorithmic soundness and completeness in original form are implied.
-The proof of this statement in type theory is a \emph{proof-relevant} type synthesiser, because it does not only return yes/no but construct a proof---either a typing derivation or its refutation. 
+Hence, both algorithmic soundness and completeness in original form are implied.
+On the other hand, the proof of this statement in type theory is a \emph{proof-relevant} type synthesiser, because it does not only return yes/no but construct a proof---either a typing derivation or its refutation.
 \LT{emphasise that our proof is an actual program}
@@
 
-\subsubsection{Mode-correctness and a general definition of typed languages}
+\subsubsection{Mode-correctness and general definitions of languages}
 \label{sec:language-formalisation}
 The most essential characteristics of bidirectional typing is \emph{mode-correctness}, since an algorithm can often be `read off' from the definition of a bidirectionally typed language if mode-correct.
-It seems that the implications of mode-correctness have only been treated informally so far as exhibited in the survey paper~\cite{Dunfield2021}, since the notion of mode-correctness is not formally defined for \emph{languages} there.
+It seems that the implications of mode-correctness have only been addressed informally as exhibited in the survey paper~\cite{Dunfield2021}, since mode-correctness is not formally defined as a \emph{property of languages}.
 %But, what is a language anyway?
 
-To make the notion of mode-correctness precise, we exhibit a general definition of bidirectional simple type systems which extends the sorted version of \varcitet{Aczel1978}{'s binding signature} with bidirectionality.
-Such a general definition allows us to define mode-correctness and to investigate its consequences rigorously, including the uniqueness of synthesised types and the logical decidability of bidirectional type synthesis for \emph{all} mode-correct \emph{bidirectional binding signatures}.
-The latter theorem effectively amounts to a verified proof-relevant bidirectional-type-synthesiser generator for all syntax-directed simply typed languages.
+To make the notion of mode-correctness precise, we give a general definition of bidirectional simple type systems, called \emph{bidirectional binding signature}, which extends the sorted version of \varcitet{Aczel1978}{'s binding signature} with bidirectionality.
+Such a general definition allows us to define mode-correctness and to investigate its consequences rigorously, including the uniqueness of synthesised types and the decidability of bidirectional type synthesis for all mode-correct bidirectional binding signatures.
+The latter theorem effectively amounts to a verified proof-relevant bidirectional-type-synthesiser generator for all syntax-directed bidirectional simple type systems that are mode-correct.
 
@@
 %the idea of extending a signature with bidirectionality should be clear enough for further generalisation.
-Nevertheless, we briefly discuss another definition which includes polymorphically typed languages such as \SystemF and additional assumptions we need in \cref{sec:future}.
+Nevertheless, we briefly discuss a possible extension which includes polymorphically typed languages such as \SystemF with additional assumptions in \cref{sec:future}.
 
@@
 \end{enumerate}
-Our theory was first developed in \Agda and then translated to the mathematical vernacular.
-\LT{methodology for combining proofs and programs?}
+Our theory was first developed in \Agda and then translated to the mathematical vernacular, filing the gap between theory and implementation.
 

### 2023-10-12 tex/ESOP24/BiSig.tex
[ update ] Add the missing .bib
[ tex ] Revise until S6

@@
-\documentclass[orivec,envcountsame,envcountsect,envcountreset,oribibl]{llncs}
+\documentclass[orivec,envcountsame,envcountsect,oribibl]{llncs}
 
@@
 
+%\let\theorem\relax
+%\let\definition\relax
+%\spnewtheorem{theorem}{Theorem}[section]{\bfseries}{}
+%\spnewtheorem{definition}{Definition}[section]{\bfseries}{}
+
 %\author{Liang-Ting Chen
@@
 This situation is in stark contrast to parsing, for which there have been general theories and widely applicable tools such as parser generators.
-To remedy the situation, this paper develops a \emph{general} and \emph{formal} theory of bidirectional typing, and provides a \emph{verified} generator of \emph{proof-relevant} type synthesisers for syntax-directed simply typed languages: For every signature that specifies a bidirectionally typed language, there exists a proof-relevant type synthesiser that for an input abstract syntax tree constructs a typing derivation if any, gives its refutation if not, or reports that the input does not have enough type annotations.
-Soundness, completeness, and mode-correctness are studied universally for any simply typed language, which are sufficient conditions for deriving a correct type synthesiser.
-We propose a preprocessing step called `mode decoration', which helps the user to deal with missing type annotations.
+To remedy the situation, this paper develops a \emph{general} and \emph{formal} theory of bidirectional typing, and provides a \emph{verified} generator of \emph{proof-relevant} type synthesisers for simply typed languages: For every signature that specifies a mode-correct bidirectionally typed language, there exists a proof-relevant type synthesiser that for an input abstract syntax tree constructs a typing derivation if any, gives its refutation if not, or reports that the input does not have enough type annotations.
+Soundness, completeness, and mode-correctness are studied universally for any simply typed language, which are sufficient conditions for deriving a type synthesiser.
+We propose a preprocessing step called \emph{mode decoration}, which helps the user to deal with missing type annotations in an abstract syntax tree given by programmers.
 The entire development is formalised in \Agda and can be further integrated with other language-formalisation frameworks.
@@
 \input{S6-example}
-%\subfile{S7-formalisation}
+\subfile{S7-formalisation}
 \input{S8-future}
 \appendix
-%\input{appendix}
+\input{appendix}
 
@@
 
-\bibliography{../reference.bib}
-%\IfFileExists{../reference.bib}{%
-%}{\bibliography{ref}}
+\IfFileExists{reference}{
+\bibliography{reference}
+}{\bibliography{ref}}
 

### 2023-10-12 tex/ESOP24/S1-intro.tex
[ tex ] Revise until S6

@@
 
-Implementing a type synthesiser correctly, especially for advanced type systems, is non-trivial.
-For example, proof assistants in use including \Coq and \Agda have many bugs in their type synthesis such as absurdity, supposedly equal terms which were not normalised to the same canonical term, and so on.
-\LT{some GitHub issues \cite{Felicissimo2023}}
-The gap between the design of type synthesis and its implementation cannot be ignored.
-
-To mitigate the above problem, we demonstrate a general and formal treatment of bidirectional typing, providing an out-of-the-box \emph{verified generator} of \emph{proof-relevant} type synthesisers developed with the proof assistant \Agda:
+To mitigate the above problem, we develop with the proof assistant \Agda a general and formal treatment of bidirectional typing and provide a \emph{verified} generator of \emph{proof-relevant} type synthesisers taking a language specification and a (scoped-checked) raw term as input.
 
@@
 Our work adapts proof-relevant (bidirectional) type synthesis illustrated by \citet{Wadler2022}.
-Their treatment deviates from the common formulation: 
-Traditionally, a type synthesis algorithm is presented as \emph{algorithmic system} relations such as $\Gamma |- \isTerm{t} \syn A \mapsto \isTerm{t}'$, denoting that annotations can be added to $\isTerm{t}$ in the surface language to produce $\isTerm{t}'$ of type $A$ in the core language.
-Such an algorithm is then accompanied with soundness and completeness assertions such that the algorithm correctly synthesises the type of a raw term and every typable term can be synthesised if there are enough annotations.
+Their treatment deviates from the usual formulation: 
+a type synthesis algorithm is traditionally presented as \emph{algorithmic rules} such as $\Gamma |- \isTerm{t} \syn A \mapsto \isTerm{t}'$, denoting that annotations can be added to $\isTerm{t}$ in the surface language to produce $\isTerm{t}'$ of type $A$ in the core language.
+Such an algorithm is then followed by soundness and completeness assertions such that the algorithm correctly synthesises the type of a raw term and every typable term can be synthesised if sufficiently annotated.
 By contrast, \citeauthor{Wadler2022} exploit the computational and logical nature of Martin-L\"of type theory and formulate \emph{algorithmic soundness, completeness, and decidability in one go}.
@@
 For example, a proof of \emph{decidable equality}, i.e.
-$\forall x, y.~(x = y) + (x \neq y)$, decides whether $x$~and~$y$ are equal and accordingly gives an identity proof or a refutation explicitly; such a decidability proof may or may not be possible depending on the domain of $x$~and~$y$, and is non-trivial.
+$\forall x, y.~(x = y) + (x \neq y)$, decides whether $x$~and~$y$ are equal and accordingly gives an identity proof or a refutation explicitly; such a decidability proof may or may not be possible depending on the domain of $x$~and~$y$, and is non-trivial in general.
 In type theory all proofs as programs terminate, so logical decidability implies algorithmic decidability.
@@
 \end{quote}
-It yields either a typing derivation for the given raw term~$t$ or a negation proof that such a derivation is impossible where the former case is algorithmic soundness and the latter algorithmic completeness in contrapositive form.
+Computationally, the proof yields either a typing derivation for the given raw term~$t$ or a negation proof that such a derivation is impossible where the former case is algorithmic soundness and the latter algorithmic completeness in contrapositive form.
 Hence, both algorithmic soundness and completeness in original form are implied.
-On the other hand, the proof of this statement in type theory is a \emph{proof-relevant} type synthesiser, because it does not only return yes/no but construct a proof---either a typing derivation or its refutation.
-\LT{emphasise that our proof is an actual program}
+Hence, the formal proof of the statement as a program is a verified \emph{proof-relevant} type synthesiser, because it does not only return yes/no but also constructs a proof. 
 %For a type synthesiser being proof-relevant, all of its intermediate programs have to be proof-relevant, including bidirectional type synthesiser and its mode decorator.
@@
 
-To make the notion of mode-correctness precise, we give a general definition of bidirectional simple type systems, called \emph{bidirectional binding signature}, which extends the sorted version of \varcitet{Aczel1978}{'s binding signature} with bidirectionality.
-Such a general definition allows us to define mode-correctness and to investigate its consequences rigorously, including the uniqueness of synthesised types and the decidability of bidirectional type synthesis for all mode-correct bidirectional binding signatures.
-The latter theorem effectively amounts to a verified proof-relevant bidirectional-type-synthesiser generator for all syntax-directed bidirectional simple type systems that are mode-correct.
+To make the notion of mode-correctness precise, we first give a general definition of bidirectional simple type systems, called \emph{bidirectional binding signature}, extending the sorted version of \varcitet{Aczel1978}{'s binding signature} with bidirectionality.
+A general definition of typed languages allows us to define mode-correctness and to investigate its consequences rigorously, including the uniqueness of synthesised types and the decidability of bidirectional type synthesis for all mode-correct signatures.
+The latter theorem effectively amounts to a generator of proof-relevant bidirectional type synthesiser for all syntax-directed bidirectional simple type systems that are mode-correct.
 
@@
 \end{enumerate}
-Our theory was first developed in \Agda and then translated to the mathematical vernacular, filing the gap between theory and implementation.
+Our theory was first developed in \Agda and then translated to the mathematical vernacular, filling the gap between our informal presentation and its verified implementation.
 

### 2023-10-12 tex/ESOP24/S3-defs.tex
[ tex ] Revise until S6

@@
 The definitions are formulated in two steps:
-\begin{enumerate*}
+\begin{inlineenum}
   \item first we introduce a notion of arity and a notion of signature which includes a set\footnote{%
@@
 \item then, given a signature, we define raw terms and typing derivations inductively by primitive rules such as $\Rule{Var}$ and a rule schema for constructs $\tmOp_o$ indexed by an operation symbol~$o$.
-\end{enumerate*}
+\end{inlineenum}
 %All these definitions are inductive as usual but include a \emph{rule schema} giving rise to as many rules as there are operation symbols in a signature.
@@
 A simply typed language specifies
-\begin{enumerate*}
+\begin{inlineenum}
   \item a family of sets of raw terms $\isTerm{t}$ indexed by a list~$V$ of variables (that are currently in scope) where each construct is allowed to bind some variables like $\Rule{Abs}$ and to take multiple arguments like $\Rule{App}$;
   \item a family of sets of typing derivations indexed by a typing context~$\Gamma$, a raw term~$\isTerm{t}$, and a type~$A$.
-\end{enumerate*}
+\end{inlineenum}
 Therefore, to specify a term construct, we enrich the notion of arity with some sort for typing and extension context for variable binding.
@@
   For a simply typed language $(\Sigma, \Omega)$, the family of sets of \emph{raw terms} indexed by a list~$V$ of variables consists of
-  \begin{enumerate*}
+  \begin{inlineenum}
     \item variables in $V$,
@@
     \item a construct $\tmOpts$ for some $\bop$ in~$O$, where $\vars{x}_{\isTerm{i}}$'s are lists of variables whose length is equal to the length of~$\Delta_i$, and $t_i$'s are raw terms in the variable list $V, \vars{x}_i$.
-  \end{enumerate*}
+  \end{inlineenum}
   These correspond to rules $\Rule{Var}$, $\Rule{Anno}$, and $\Rule{Op}$ in \Cref{fig:raw-terms} respectively.
@@
 
-The definition of typing derivations is a bit more involved.
+The definition of typing derivations is more involved.
 We need some information to compare types on the object level during type synthesis and substitute those type variables in a typing derivation $\Gamma \vdash \tmOpts : A$ for an operation $o$ in $\Omega$ at some point.
@@
 
-\begin{definition}\label{def:bidirectional-typing-derivations}\label{def:mode-derivations}
 \begin{figure}
@@
 \end{figure}
-  \begin{figure}
+
+\begin{definition}\label{def:bidirectional-typing-derivations}\label{def:mode-derivations}
+  For a bidirectional type system $(\Sigma, \Omega)$,
+  \begin{itemize}
+    \item the set of \emph{bidirectional typing derivations} of $\Gamma \vdash_{\Sigma, \Omega} t :^\dir{d} A$, indexed by a typing context $\Gamma$, a raw term $\isTerm{t}$ under $\erase{\Gamma}$, a mode $\dir{d}$, and a type $A$, is defined in \Cref{fig:bidirectional-typing-derivations} and particularly
+          \[
+            \Gamma |-_{\Sigma, \Omega} \tmOpts :^{\dir{d}} \simsub{A_0}{\rho}
+          \]
+          has a derivation for $\biop$ in $\Omega$ if there is $\rho\colon \Xi \to \Type_{\Sigma}(\emptyset)$ and a derivation of $\Gamma, \vars{x}_{\isTerm{i}} : \simsub{\Delta_i}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_i} :^{\dir{d_i}} \simsub{A_i}{\rho}$ for each $i$;
+    \item the set of \emph{mode derivations} of $V |-_{\Sigma, \Omega} t^\dir{d}$, indexed by a list $V$ of variables, a raw term $\isTerm{t}$ under $V$, and a mode $d$, is defined in \Cref{fig:mode-derivations}.
+  \end{itemize}
+  The two judgements {\small$\boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \syn A}$} and {\small$\boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \chk A}$} stand for ${\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} :^{\syn} A}$ and ${\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} :^{\chk} A}$, respectively.
+  A typing rule is \emph{checking} if its conclusion mode is $\chk$ or \emph{synthesising} otherwise.
+\end{definition}
+
+\begin{figure}[b]
   \centering
@@
 \end{figure}
-  For a bidirectional type system $(\Sigma, \Omega)$,
-  \begin{itemize}
-    \item the set of \emph{bidirectional typing derivations} of $\Gamma \vdash_{\Sigma, \Omega} t :^\dir{d} A$, indexed by a typing context $\Gamma$, a raw term $\isTerm{t}$ under $\erase{\Gamma}$, a mode $\dir{d}$, and a type $A$, is defined in \Cref{fig:bidirectional-typing-derivations} and particularly
-          \[
-            \Gamma |-_{\Sigma, \Omega} \tmOpts :^{\dir{d}} \simsub{A_0}{\rho}
-          \]
-          has a derivation for $\biop$ in $\Omega$ if there is $\rho\colon \Xi \to \Type_{\Sigma}(\emptyset)$ and a derivation of $\Gamma, \vars{x}_{\isTerm{i}} : \simsub{\Delta_i}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_i} :^{\dir{d_i}} \simsub{A_i}{\rho}$ for each $i$;
-    \item the set of \emph{mode derivations} of $V |-_{\Sigma, \Omega} t^\dir{d}$, indexed by a list $V$ of variables, a raw term $\isTerm{t}$ under $V$, and a mode $d$, is defined in \Cref{fig:mode-derivations}.
-  \end{itemize}
-  The two judgements {\small$\boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \syn A}$} and {\small$\boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \chk A}$} stand for ${\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} :^{\syn} A}$ and ${\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} :^{\chk} A}$, respectively.
-  A typing rule is \emph{checking} if its conclusion mode is $\chk$ or \emph{synthesising} otherwise.
-\end{definition}
-
 Every bidirectional binding signature $\Omega$ gives rise to a binding signature $\erase{\Omega}$ if we erase modes from $\Omega$, called the \emph{(mode) erasure} of $\Omega$.

### 2023-10-12 tex/ESOP24/S4-pre-synthesis.tex
[ tex ] Revise until S6

@@
 
-\begin{theorem}[Generalised Mode Preprocessing]\label{thm:generalised-mode-decoration}
-For any raw term $V |-_{\Sigma,\erase\Omega} t$ and mode~$\dir{d}$, there is a derivation of\/ $V |-_{\Sigma,\Omega} t^{\dir{d}\,g\,s}$ for some booleans $g$~and~$s$.
+\begin{theorem}[Generalised mode preprocessing]\label{thm:generalised-mode-decoration}
+For any raw term $V |-_{\Sigma,\erase\Omega} t$ and mode~$\dir{d}$, there is a derivation of\/ $V |-_{\Sigma,\Omega} t^{\dir{d}\,g\,s}$ for some $g$~and~$s$.
 \end{theorem}
@@
 \end{equation}
-then for any mode~$\dir{d}$, there is a derivation of\/ $V |-_{\Sigma,\Omega} t^{\dir{d}\,g\,s}$ for some booleans $g$~and~$s$.
+then for any mode~$\dir{d}$, there is a derivation of\/ $V |-_{\Sigma,\Omega} t^{\dir{d}\,g\,s}$ for some $g$~and~$s$.
 \end{lemma}
@@
 
-\begin{proof}[\cref{thm:generalised-mode-decoration}]
+\begin{proof}[of \cref{thm:generalised-mode-decoration}]
 By \cref{thm:adjustment}, it suffices to prove~\eqref{eq:strict-derivation} by induction on~$t$.
@@
 Let $t = (t' \bbcolon A)$.
-By the induction hypothesis and \cref{thm:adjustment}, there is a derivation of $V |-_{\Sigma,\Omega} {t'}^{\chk\,g\,s}$ for some booleans $g$~and~$s$, to which we apply $\SynRule{Anno}$.
+By the induction hypothesis and \cref{thm:adjustment}, there is a derivation of $V |-_{\Sigma,\Omega} {t'}^{\chk\,g\,s}$ for some $g$~and~$s$, to which we apply $\SynRule{Anno}$.
 \item $\Rule{Op}$:
@@
 
-\begin{corollary}[Mode Preprocessing]\label{thm:mode-decoration}
-  It is decidable whether\/ $V |-_{\Sigma,\Omega} \isTerm{t}^\dir{d}$.%for any raw term $V |-_{\Sigma,\erase\Omega} t$ and mode~$\dir{d}$
+\begin{corollary}[Mode preprocessing]\label{thm:mode-decoration}
+  It is decidable whether $V |-_{\Sigma,\Omega} \isTerm{t}^\dir{d}$.%for any raw term $V |-_{\Sigma,\erase\Omega} t$ and mode~$\dir{d}$
 \end{corollary}
@@
 
-\begin{proof}[Proof of \cref{thm:annotatability}]
+\begin{proof}[of \cref{thm:annotatability}]
 By \cref{thm:generalised-mode-decoration}, there is a generalised mode derivation of $V |-_{\Sigma,\erase\Omega} t^{\dir{d}\,g\,s}$ for some mode~$\dir{d}$ and booleans $g$~and~$s$.

### 2023-10-12 tex/ESOP24/S5-synthesis.tex
[ tex ] Revise until S6

@@
   The subset $\fv(\Delta)$ of variables in an extension context~$\Delta$ is defined by\/ $\fv(\cdot) = \emptyset$ and\/ $\fv(\Delta, A) = \fv(\Delta) \cup \fv(A)$.
-  For an argument list $\biargs$, define its set of \emph{synthesised type variables} inductively by 
-  \begin{align}
-    \label{eq:synvar1}\synvar(\cdot)                   & = \emptyset  \\
-    \label{eq:synvar2}\synvar(\biargs, \chkbiarg[n+1]) & = \phantom{\fv(A_{n+1}) \cup {}} \synvar(\biargs) \\
-    \label{eq:synvar3}\synvar(\biargs, \synbiarg[n+1]) & = \fv(A_{n+1}) \cup           \synvar(\biargs).
-  \end{align}
+  For an argument list $\biargs$, its set of \emph{synthesised type variables} is denoted by $\synvar(\biargs)$.
 \end{definition}
@@
 \begin{definition}\label{def:mode-correctness-args}
-  The \emph{mode-correctness} $\MCas\left(\biargvec\right)$ for an argument list $\biargs$ with respect to a subset $S$ of $\Xi$ is a (proof-relevant) predicate defined by
-  \begin{align}
-    \label{eq:MC1} \MCas( \cdot ) & = \top \\
-    \label{eq:MC3} \MCas\left(\biargvec, \chkbiarg[n]\right)
+  The \emph{mode-correctness} $\MCas\left(\biargvec\right)$ for an argument list $\biargs$ with respect to a subset $S$ of $\Xi$ is a predicate defined by
+  \begin{align*}
+    \MCas( \cdot ) & = \top \\
+    \MCas\left(\biargvec, \chkbiarg[n]\right)
                                   & = \fv(\Delta_n, A_n) \subseteq \left( S \cup \synvar\left(\biargvec\right)\right) \land \MCas\left(\biargvec\right) \\
@@
                                   & = \phantom{, A_n} \fv(\Delta_n) \subseteq \left( S \cup \synvar\left(\biargvec\right)\right) \land  \MCas\left(\biargvec\right)
-  \end{align}
-  where \eqref{eq:MC1} means an empty list is always mode-correct.
+  \end{align*}
+  where $\top$ means an empty list is always mode-correct.
 \end{definition}
 This definition encapsulates the idea that every `input' type variable, possibly derived from an extension context~$\Delta_n$ or a checking argument~$A_n$, must be an `output' variable from $\synvar(\biargvec)$ or, if the rule is checking, belong to the set $S$ of `input' variables in its conclusion.
-This condition must be met for every tail of the argument list as well to ensure that `output' variables accessible at each argument position are from preceding arguments only, hence an inductive definition.
+This condition must also be met for every tail of the argument list to ensure that `output' variables accessible at each argument are from preceding arguments only, hence an inductive definition.
 \begin{definition}\label{def:mode-correctness}
@@
 
-It is easy to check the bidirectional type system $(\Sigma_{\bto}, \Omega^{\Leftrightarrow}_{\Lambda}$) in \Cref{ex:signature-simply-typed-lambda} for simply typed $\lambda$-calculus is mode-correct according to our definition.
-
 \begin{remark}
@@
 
-\LT{decidability of mode-correctness}
+It is easy to check the bidirectional type system $(\Sigma_{\bto}, \Omega^{\Leftrightarrow}_{\Lambda}$) in \Cref{ex:signature-simply-typed-lambda} for simply typed $\lambda$-calculus is mode-correct manually by definition or automatically by the following lemma:
+\begin{lemma}
+  It is decidable if an operation $\biop$ is mode-correct or not.
+\end{lemma}
+
 Now, we set out to show the uniqueness of synthesised types for a mode-correct bidirectional type system.
@@
 However, since mode-correctness is inductively defined on the argument list, our proof proceeds by induction on both the typing derivations and the argument list:
-\begin{lemma}[Uniqueness of Synthesised Types]\label{thm:unique-syn}
+\begin{lemma}[Uniqueness of synthesised types]\label{thm:unique-syn}
   In a mode-correct bidirectional type system $(\Sigma, \Omega)$, the synthesised types of any two derivations
@@
       \begin{enumerate}
-        \item For the empty list, the statement is vacuously true by \Cref{eq:synvar1}.
-        \item If $\dir{d_{i+1}}$ is~$\chk$, then the statement holds for \Cref{eq:synvar2} by induction hypothesis.
-        \item If $\dir{d_{i+1}}$ is~$\syn$, then $\simsub{\Delta_{i+1}}{\rho_1} = \simsub{\Delta_{i+1}}{\rho_2}$ by \Cref{eq:MC2} and induction hypothesis (of the list).
-          Therefore, under the same context $\Gamma, \simsub{\Delta_{i+1}}{\rho_1} = \Gamma, \simsub{\Delta_{i+1}}{\rho_2}$ the term $t_{i+1}$ must have the same synthesised type $\simsub{A_{i+1}}{\rho_1} = \simsub{A_{i+1}}{\rho_1}$ by induction hypothesis (of the typing derivation), so $\rho_1$ and $\rho_2$ agree on $\fv(A_{i+1})$ in addition to $\synvar(\biargs)$, as required for \Cref{eq:synvar3} in the definition of $\synvar$.
+        \item For the empty list, the statement is vacuously true.
+        \item If $\dir{d_{i+1}}$ is~$\chk$, then the statement holds by induction hypothesis.
+        \item If $\dir{d_{i+1}}$ is~$\syn$, then $\simsub{\Delta_{i+1}}{\rho_1} = \simsub{\Delta_{i+1}}{\rho_2}$ and induction hypothesis (of the list).
+          Therefore, under the same context $\Gamma, \simsub{\Delta_{i+1}}{\rho_1} = \Gamma, \simsub{\Delta_{i+1}}{\rho_2}$ the term $t_{i+1}$ must have the same synthesised type $\simsub{A_{i+1}}{\rho_1} = \simsub{A_{i+1}}{\rho_1}$ by induction hypothesis (of the typing derivation), so $\rho_1$ and $\rho_2$ agree on $\fv(A_{i+1})$ in addition to $\synvar(\biargs)$.
       \end{enumerate}
@@
 This lemma can be inferred from the correctness of first-order unification~\citep{McBride2003,McBride2003a}, or be proved directly without unification.
-We are now ready for the promised proof.
+We are now ready for \cref{thm:bidirectional-type-synthesis-checking}:
 
@@
   The case \ChkRule{Sub} invokes the uniqueness of synthesised types to refute the case that $\Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn B$ but $A \neq B$ for a given type $A$.
-  The first three cases follow essentially the same reasoning provided by \citet{Wadler2022}, but we still present the reasoning here for the sake of completeness.
-  The last case \Rule{Op} is new and has been discussed above.
+  The first three cases follow essentially the same reasoning provided by \citet{Wadler2022}, so we only detail the last case \Rule{Op} which is new and has been discussed informally above.
   For brevity we omit the subscript $(\Sigma, \Omega)$.
-  \begin{itemize}
-    \item \SynRule{Var}: If $\erase{\Gamma} |- \isTerm{t}^{\syn}$, then $(x : A) \in \Gamma$ and thus $\Gamma |- \isTerm{x} \syn A$.
 
-    \item \SynRule{Anno}: For $\erase{\Gamma} |- (\isTerm{t} \annotate A)^{\syn}$, it is decidable whether $\Gamma |- \isTerm{t} \chk A$ by induction hypothesis.
-      \begin{itemize}
-        \item If $\Gamma |- \isTerm{t} \chk A$, then $\Gamma |- \isTerm{t \annotate A} \syn A$.
-        \item If $\Gamma |/- t \chk A$ but $\Gamma |- \isTerm{t \annotate A} \syn$, then by inversion $\Gamma |- t \chk A$, leading to a contradiction.
-      \end{itemize}
-
-    \item \ChkRule{Sub}: If $\erase{\Gamma} |-t^{\chk}$, then $\Gamma |- t^{\syn}$ by inversion.
-      By induction hypothesis, we have two cases: %it is decidable whether $\Gamma |- t \syn B$ for some $B$:
-      \begin{itemize}
-        \item If $\Gamma |/- t \syn C$ for any $C$ but $\Gamma |- t \chk A$, then by inversion $\Gamma |- t \syn B$ for some $B = A$, thus a contradiction.
-        \item If $\Gamma |- t \syn B$ for some $B$, then by decidable equality on $\Type_{\Sigma}(\Xi)$ either $A = B$ or $A \neq B$: 
-          \begin{itemize}
-            \item if $A = B$ then we have\/ $\Gamma |- t \chk A$;
-            \item if $A \neq B$ but $\Gamma |- t \chk A$, then by inversion $\Gamma |- t \syn A$.
-              However, by \Cref{thm:unique-syn}, synthesised types $A$ and $B$ must be equal, so we derive a contradiction.
-          \end{itemize}
-      \end{itemize}
-    \item \Rule{Op}:
       For a mode derivation of $\erase{\Gamma} |- \tmOpts^{\dir{d}}$, we first claim:
@@
               \dom(\ext{\rho}) \supseteq \synvar(\biargs)
-              \quad\text{and}\quad
+              \;\text{and}\;
 %              \text{the domain of $\ext{\rho}$ contains $\synvar(\biargs)$ and\/} \quad
               \Gamma, \vars{x}_\isTerm{i} : \simsub{\Delta_i}{\ext{\rho}} |- \isTerm{t_i} \colon \simsub{A_i}{\ext{\rho}}^{\dir{d_i}}
-              \quad\text{for $i = 1, \ldots, n$};
             \end{equation}
+            for $i = 1, \ldots, n$
           \item or there is no extension $\sigma$ of $\rho$ such that \eqref{eq:claim} holds.
@@
                   \]
-                  where $\simsub{A_0}{\ext{\ext{\rho}}} = \simsub{A_0}{\ext{\rho}} = A$ since $\ext{\ext{\rho}}(x) = \ext{\rho}$ for every $x$ in the domain of $\ext{\rho}$.
+                  where $\simsub{A_0}{\ext{\ext{\rho}}} = \simsub{A_0}{\ext{\rho}} = A$ since $\ext{\ext{\rho}}(x) = \ext{\rho}$ for every $x$ in $\dom(\ext{\rho})$.
               \end{enumerate}
@@
       \end{proof}
-  \end{itemize}
+%  \end{itemize}
   We have completed the proof by induction on the mode derivation $\erase{\Gamma} |-_{\Sigma, \Omega} t^{\dir{d}}$.\qed
@@
 
-\begin{corollary}[Trichotomy on Raw Terms]\label{cor:trichotomy}
+\begin{corollary}[Trichotomy on raw terms]\label{cor:trichotomy}
   For any mode-correct bidirectional type system $(\Sigma, \Omega)$, 
@@
   \begin{enumerate}
-    \item $\Gamma |-_{\Sigma, \Omega} \isTerm{t} : A$ for some type~$A$.
-    \item $\Gamma |/-_{\Sigma, \Omega} \isTerm{t} : A$ for any type~$A$.
-    \item \phantom{(\/$\erase{\Gamma} |-_{\Sigma, \Omega} \isTerm{t}^{\syn}$}\llap{$\erase{\Gamma} |/-_{\Sigma, \Omega} \isTerm{t}^{\syn}$\kern.5pt}.
+    \item $\Gamma |-_{\Sigma, \Omega} \isTerm{t} : A$ for some type~$A$,
+    \item $\Gamma |/-_{\Sigma, \Omega} \isTerm{t} : A$ for any type~$A$, or 
+    \item $\erase{\Gamma} |/-_{\Sigma, \Omega} \isTerm{t}^{\syn}$.
   \end{enumerate}

### 2023-10-12 tex/ESOP24/appendix.tex
[ tex ] Revise until S6

@@
 
-\subsection{Mode Correctness}\label{sec:mode-correctness}
-As \citet{Dunfield2021} outlined, mode-correctness for a bidirectional typing rule means that 
-\begin{enumerate*}
-\item each `input' type variable in a premise must be an `output' variable in `earlier' premises, or provided by the conclusion if the rule is checking;
-\item each `output' type variable in the conclusion should be some `output' variable in a premise if the rule is synthesising.
-\end{enumerate*}
-Here `input' variables refer to variables in an extension context and in a checking premise.
-It is important to note that the order of premises in a bidirectional typing rule also matters, since synthesised type variables are instantiated incrementally during type synthesis.
-
-Consider the rule $\ChkRule{Abs}$ (\Cref{fig:STLC-bidirectional-typing-derivations}) as an example.
-This rule is mode-correct, as the type variables $A$ and $B$ in its only premise are already provided by its conclusion $A \bto B$.
-Likewise, the rule $\SynRule{App}$ for an application term $\isTerm{t\;u}$ is mode-correct because:
-\begin{enumerate*}
-\item the type $A \bto B$ of the first argument $t$ is synthesised, thereby ensuring type variables $A$ and $B$ must be known if successfully synthesised;
-\item the type of the second argument $u$ is checked against $A$, which has been synthesised earlier;
-\item as a result, the type of an application $t\;u$ can be synthesised.
-\end{enumerate*}
-
-Now let us define mode-correctness rigorously.
-As we have outlined, the condition of mode-correctness for a synthesising rule is different from that of a checking rule, and the argument order also matters.
-Defining the condition directly for a rule, and thus in our setting for an operation, can be somewhat intricate.
-Instead, we choose to define the conditions for the argument list---more specifically, triples $\biargvec$ of an extension context $\Delta_i$, a type $A_i$, and a mode $\dir{d_i}$---pertaining to an operation, for an operation, and subsequently for a signature.
-We also need some auxiliary definitions for the subset of variables of a type and of an extension context, and the set of variables that have been synthesised:
-\begin{definition}
-  The finite subset\footnote{%
-  There are various definitions for finite subsets of a set within Martin-L\"{o}f type theory.
-  However, for our purposes, the choice among these definitions is not a matter of concern.}
-  of \emph{(free) variables} of a type $A$ is denoted by $\fv(A)$.
-  The subset $\fv(\Delta)$ of variables in an extension context~$\Delta$ is defined by\/ $\fv(\cdot) = \emptyset$ and\/ $\fv(\Delta, A) = \fv(\Delta) \cup \fv(A)$.
-  For an argument list $\biargs$, define its set of \emph{synthesised type variables} inductively by 
-  \begin{align}
-    \label{eq:synvar1}\synvar(\cdot)                   & = \emptyset  \\
-    \label{eq:synvar2}\synvar(\biargs, \chkbiarg[n+1]) & = \phantom{\fv(A_{n+1}) \cup {}} \synvar(\biargs) \\
-    \label{eq:synvar3}\synvar(\biargs, \synbiarg[n+1]) & = \fv(A_{n+1}) \cup           \synvar(\biargs).
-  \end{align}
+\begin{definition}\label{def:syn-var}
+  For an argument list $\biargs$, the set of \emph{synthesised type variables} is defined inductively by 
+  \begin{align*}
+    \synvar(\cdot)                   & = \emptyset  \\
+    \synvar(\biargs, \chkbiarg[n+1]) & = \phantom{\fv(A_{n+1}) \cup {}} \synvar(\biargs) \\
+    \synvar(\biargs, \synbiarg[n+1]) & = \fv(A_{n+1}) \cup           \synvar(\biargs).
+  \end{align*}
 \end{definition}
-This subset contains type variables of a synthesising argument and they are exactly those type variables that will be synthesised during type synthesis.
-
-\begin{definition}\label{def:mode-correctness-args}
-  The \emph{mode-correctness} $\MCas\left(\biargvec\right)$ for an argument list $\biargs$ with respect to a subset $S$ of $\Xi$ is a (proof-relevant) predicate defined by
-  \begin{align}
-    \label{eq:MC1} \MCas( \cdot ) & = \top \\
-    \label{eq:MC3} \MCas\left(\biargvec, \chkbiarg[n]\right)
-                                  & = \fv(\Delta_n, A_n) \subseteq \left( S \cup \synvar\left(\biargvec\right)\right) \land \MCas\left(\biargvec\right) \\
-    \label{eq:MC2} \MCas\left(\biargvec, \synbiarg[n]\right) 
-                                  & = \phantom{, A_n} \fv(\Delta_n) \subseteq \left( S \cup \synvar\left(\biargvec\right)\right) \land  \MCas\left(\biargvec\right)
-  \end{align}
-  where \eqref{eq:MC1} means an empty list is always mode-correct.
-\end{definition}
-This definition encapsulates the idea that every `input' type variable, possibly derived from an extension context~$\Delta_n$ or a checking argument~$A_n$, must be an `output' variable from $\synvar(\biargvec)$ or, if the rule is checking, belong to the set $S$ of `input' variables in its conclusion.
-This condition must be met for every tail of the argument list as well to ensure that `output' variables accessible at each argument position are from preceding arguments only, hence an inductive definition.
-\begin{definition}\label{def:mode-correctness}
-  An operation $\biop$ is \emph{mode-correct} if 
-  \begin{enumerate}
-    \item either $d$ is $\chk$, its argument list is mode-correct with respect to $\fv(A_0)$, and the union $\fv(A_0) \cup \synvar(\biargvec)$ contains every inhabitant of $\Xi$;
-    \item or $d$ is $\syn$, its argument list is mode-correct with respect to $\emptyset$, and $\synvar(\biargvec)$ contains every inhabitant of $\Xi$ and, particularly, $\fv(A_0)$.
-  \end{enumerate}
-  A bidirectional binding signature $\Omega$ is \emph{mode-correct} if its operations are all mode-correct.
-\end{definition}
-For a checking operation, an `input' variable of an argument could be derived from~$A_0$, as these are known during type checking as an input.
-Since every inhabitant of~$\Xi$ can be located in either~$A_0$ or synthesised variables, we can determine a concrete type for each inhabitant of~$\Xi$ during type synthesis.
-On the other hand, for a synthesising operation, we do not have any known variables at the onset of type synthesis, so the argument list should be mode-correct with respect to~$\emptyset$.
-Also, the set of synthesised variables alone should include every type variable in~$\Xi$ and particularly in~$A_n$.
-
-It is easy to check the bidirectional type system $(\Sigma_{\bto}, \Omega^{\Leftrightarrow}_{\Lambda}$) in \Cref{ex:signature-simply-typed-lambda} for simply typed $\lambda$-calculus is mode-correct according to our definition.
-
-\begin{remark}
-  Mode-correctness is fundamentally a condition for bidirectional typing \emph{rules}, not for derivations.
-  Thus, this property cannot be formalised without treating rules as some mathematical object, such as the notion of bidirectional binding signature presented in \Cref{sec:defs}.
-  This contrasts with the properties in \Cref{sec:pre-synthesis}, which can still be specified for individual systems even in the absence of a generic definition of bidirectional type systems.
-\end{remark}
-
-Now, we set out to show the uniqueness of synthesised types for a mode-correct bidirectional type system.
-For a specific system, its proof is typically a straightforward induction on the typing derivations.
-However, since mode-correctness is inductively defined on the argument list, our proof proceeds by induction on both the typing derivations and the argument list:
-\begin{lemma}[Uniqueness of Synthesised Types]\label{thm:unique-syn}
-  In a mode-correct bidirectional type system $(\Sigma, \Omega)$, the synthesised types of any two derivations
-  \[
-    \Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn A
-    \quad\text{and}\quad
-    \Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn B
-  \]
-  for the same term $t$ must be equal, i.e.\ $A = B$.
-\end{lemma}
-\begin{proof}%[Proof of \Cref{thm:unique-syn}]
-  We prove the statement by induction on derivations $d_1$ and $d_2$ for $\Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn A$ and $\Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn B$.
-  Our system is syntax-directed, so $d_1$ and $d_2$ must be derived from the same rule: 
-  \begin{itemize}
-    \item $\SynRule{Var}$ follows from that each variable as a raw term refers to the same variable in its context.
-    \item $\SynRule{Anno}$ holds trivially, since the synthesised type $A$ is from the term $t \annotate A$ in question.
-    \item $\Rule{Op}$: Recall that a derivation of\/ $\Gamma |- \tmOpts \syn A$ contains a substitution $\rho$ from the local context $\Xi$ to concrete types.
-      To prove that any two typing derivations has the same synthesised type, it suffices to show that those substitutions $\rho_1$~and~$\rho_2$ of $d_1$~and~$d_2$, respectively, agree on variables in $\synvar(\biargs)$ so that $\simsub{A_0}{\rho_1} = \simsub{A_0}{\rho_2}$.
-      We prove it by induction on the argument list:
-      \begin{enumerate}
-        \item For the empty list, the statement is vacuously true by \Cref{eq:synvar1}.
-        \item If $\dir{d_{i+1}}$ is~$\chk$, then the statement holds for \Cref{eq:synvar2} by induction hypothesis.
-        \item If $\dir{d_{i+1}}$ is~$\syn$, then $\simsub{\Delta_{i+1}}{\rho_1} = \simsub{\Delta_{i+1}}{\rho_2}$ by \Cref{eq:MC2} and induction hypothesis (of the list).
-          Therefore, under the same context $\Gamma, \simsub{\Delta_{i+1}}{\rho_1} = \Gamma, \simsub{\Delta_{i+1}}{\rho_2}$ the term $t_{i+1}$ must have the same synthesised type $\simsub{A_{i+1}}{\rho_1} = \simsub{A_{i+1}}{\rho_1}$ by induction hypothesis (of the typing derivation), so $\rho_1$ and $\rho_2$ agree on $\fv(A_{i+1})$ in addition to $\synvar(\biargs)$, as required for \Cref{eq:synvar3} in the definition of $\synvar$.
-      \end{enumerate}
-  \end{itemize}
-\end{proof}
-
-%Uniqueness of the synthesised types is a prevalent property in bidirectional type systems, although the specific proofs can vary depending on the constructs in the system.
-%For instance, for derivations of $\Gamma |- t\;u \syn B_i$ for $i = 1, 2$ in simply typed $\lambda$-calculus, the hypothesis is applied to their sub-derivations $\Gamma |- t \syn A_i \bto B_i$ to conclude that $A_1 \bto B_1 = A_2 \bto B_2$ and thus $B_1 = B_2$.
-%On the other hand, our proof is based on mode-correctness and need not consider specific sub-derivations.

### 2023-10-13 tex/ESOP24/S1-intro.tex
[ tex ] Update Example section

@@
 
-Type inference is an important mechanism for the transition to well-typed programs from untyped abstract syntax trees, which we call \emph{raw terms}.
+Type inference is an important mechanism for the transition to well typed programs from untyped abstract syntax trees, which we call \emph{raw terms}.
 Here `type inference' refers specifically to algorithms that ascertain the type of any raw term \emph{without type annotations}.
@@
 The former indicates that the type~$A$ is computed as output, using both the context~$\Gamma$ and the term~$t$ as input, while for the latter, all three of $\Gamma$, $t$, and~$A$ are given as input.
-The algorithm of a bidirectional type synthesiser can usually be `read off' from a well-designed bidirectional type system: as the synthesiser traverses a raw term, it switches between synthesis and checking, following the modes assigned to the judgements in the typing rules.
+The algorithm of a bidirectional type synthesiser can usually be `read off' from a well designed bidirectional type system: as the synthesiser traverses a raw term, it switches between synthesis and checking, following the modes assigned to the judgements in the typing rules.
 %if a condition called \emph{mode-correctness} is satisfied.
@@
 %While there are type checker generators grounded in unification~\citep{Gast2004,Grewe2015}, it should be noted that unification-based approaches are not suited to more complex type systems.}
-While it is straightforward to derive a type synthesis algorithm,
-for bidirectional typing, \citeauthor{Dunfield2021} only present informal design principles that the community learned from individual systems, rather than a general treatment providing logical specifications and rigorously proven properties for a class of systems.
+While it is straightforward to derive a bidirectional type synthesis algorithm, \citeauthor{Dunfield2021} only present informal design principles that the community learned from individual systems, rather than a general treatment providing logical specifications and rigorously proven properties for a class of systems.
 Moreover, unlike the plethora of available parser generators, `type-synthesiser generators' rarely exist, so each type synthesiser has to be independently built, not to mention their correctness.
 
-To mitigate the above problem, we develop with the proof assistant \Agda a general and formal treatment of bidirectional typing and provide a \emph{verified} generator of \emph{proof-relevant} type synthesisers taking a language specification and a (scoped-checked) raw term as input.
+To mitigate the above problem, we develop with the proof assistant \Agda a general and formal treatment of bidirectional typing and provide a \emph{verified} generator of \emph{proof-relevant} type synthesisers taking a language specification and a raw term as input.
 
@@
 
-Our work adapts proof-relevant (bidirectional) type synthesis illustrated by \citet{Wadler2022}.
-Their treatment deviates from the usual formulation: 
-a type synthesis algorithm is traditionally presented as \emph{algorithmic rules} such as $\Gamma |- \isTerm{t} \syn A \mapsto \isTerm{t}'$, denoting that annotations can be added to $\isTerm{t}$ in the surface language to produce $\isTerm{t}'$ of type $A$ in the core language.
-Such an algorithm is then followed by soundness and completeness assertions such that the algorithm correctly synthesises the type of a raw term and every typable term can be synthesised if sufficiently annotated.
-By contrast, \citeauthor{Wadler2022} exploit the computational and logical nature of Martin-L\"of type theory and formulate \emph{algorithmic soundness, completeness, and decidability in one go}.
+Our work adopts a proof-relevant approach to (bidirectional) type synthesis, as illustrated by \citet{Wadler2022} for \PCF.
+The proof-relevant formulation deviates from the usual one:
+traditionally, a type synthesis algorithm is presented as \emph{algorithmic rules}, for example in the form $\Gamma |- \isTerm{t} \syn A \mapsto \isTerm{t}'$, which denotes that annotations can be added to~$\isTerm{t}$ in the surface language%
+\Josh{The algorithm adds annotations to the input term?}
+to produce $\isTerm{t}'$ of type~$A$ in the core language.
+Such an algorithm is then accompanied by soundness and completeness assertions that the algorithm correctly synthesises the type of a raw term and every typable term can be synthesised.
+By contrast, the proof-relevant approach exploits the simultaneously computational and logical nature of Martin-L\"of type theory and formulates \emph{algorithmic soundness, completeness, and decidability in one go}.
+We explain this in a bit more detail below.
 %(although they do not emphasise the difference of their approach from the traditional one)
 
-Recall that the law of excluded middle $P + \neg P$ does not hold as an axiom for every $P$ in type theory, and we say $P$~is logically \emph{decidable} if the law holds for~$P$.
-For example, a proof of \emph{decidable equality}, i.e.
-$\forall x, y.~(x = y) + (x \neq y)$, decides whether $x$~and~$y$ are equal and accordingly gives an identity proof or a refutation explicitly; such a decidability proof may or may not be possible depending on the domain of $x$~and~$y$, and is non-trivial in general.
-In type theory all proofs as programs terminate, so logical decidability implies algorithmic decidability.
-Further, suppose a proof of the following statement:
+Recall that the law of excluded middle $P + \neg P$ does not hold as an axiom for every~$P$ constructively, and we say that $P$~is logically \emph{decidable} if the law holds for~$P$.
+Since Martin-L\"of type theory is simultaneously logical and computational, a decidability proof is also a proof-relevant decision procedure that computes not only a yes-or-no answer but also a proof of~$P$ or its refutation, so logical decidability implies algorithmic decidability.
+%For example, a proof of \emph{decidable equality}, i.e.~$\forall x, y.~(x = y) + (x \neq y)$, decides whether $x$~and~$y$ are equal and accordingly gives an identity proof or a refutation explicitly.
+%Such a decidability proof may or may not be possible depending on the domain of $x$~and~$y$, and is non-trivial in general.
+%In type theory all proofs as programs terminate, so logical decidability implies algorithmic decidability.
+More specifically, consider the statement of the type inference problem
 \begin{quote}
-  `For a context $\Gamma$ and a raw term $t$, either a typing derivation of\, $\Gamma |- t : A$ exists for some type~$A$ or any derivation of\, $\Gamma |- t : A$ for some type~$A$ leads to a contradiction'
+  `for a context $\Gamma$ and a raw term $t$, either a typing derivation of\, $\Gamma |- t : A$ exists for some type~$A$ or any derivation of\, $\Gamma |- t : A$ for some type~$A$ leads to a contradiction',
 \end{quote}
-or rephrased succinctly as 
+which can be rephrased more succinctly as
 \begin{quote}
@@
 \end{quote}
-Computationally, the proof yields either a typing derivation for the given raw term~$t$ or a negation proof that such a derivation is impossible where the former case is algorithmic soundness and the latter algorithmic completeness in contrapositive form.
-Hence, both algorithmic soundness and completeness in original form are implied.
-Hence, the formal proof of the statement as a program is a verified \emph{proof-relevant} type synthesiser, because it does not only return yes/no but also constructs a proof. 
+A proof of this statement would also be a program that produces either a typing derivation for the given raw term~$\isTerm{t}$ or a negation proof that such a derivation is impossible.
+The first case is algorithmic soundness, while the second case is algorithmic completeness in contrapositive form (and can be strengthened to the original form).
+Hence, proving the statement is the same as constructing a verified proof-relevant type inferencer, which returns not only an answer but also a proof justifying that answer.
+
+%Hence, both algorithmic soundness and completeness in original form are implied.
 %For a type synthesiser being proof-relevant, all of its intermediate programs have to be proof-relevant, including bidirectional type synthesiser and its mode decorator.
+%\Josh{emphasise the scope of our approach}
 
-\Josh{emphasise the scope of our approach}
-\subsubsection{Mode decoration and bidirectional type synthesis}
-\Josh{A missing step between bidirectional type synthesis and parsing}
-As raw terms may not be sufficiently annotated, we propose a preprocessing step called \emph{mode decoration} to determine whether a raw term can be processed by a bidirectional type synthesiser later.
-A mode decorator constructs a mode derivation for a raw term or pinpoints missing annotations.
-A bidirectional type synthesiser and a type synthesiser differ in their domain---bidirectional type synthesis only works for mode-decorated raw terms instead of all raw terms.
-Soundness and completeness of bidirectional typing is reformulated as a one-to-one correspondence between \emph{bidirectional typing derivations} and \emph{typing derivations} with \emph{mode derivations} (of a raw term), which is a more refined and useful formulation than annotatability described by \citet[Section~3.2]{Dunfield2021}.
-Our proof-relevant bidirectional type synthesiser is complete with respect to mode-decorated terms.
+\subsubsection{Annotations in the type synthesis problem}
+%\Josh{A missing step between bidirectional type synthesis and parsing}
 
-We also prove that a raw term with missing annotations is exactly a raw term without a mode derivation, so our mode decorator is indeed proof-relevant.
+As mentioned in the beginning, with bidirectional typing we avoid the generally undecidable problem of type inference, and instead solve the simpler problem about the typability of `sufficiently annotated' raw terms, which we call the type synthesis problem to distinguish it from type inference.
+Annotations therefore play an important role even in the definition of the problem solved by bidirectional typing, but have not received enough attention.
+In our theory, we define \emph{mode derivations} so that we can explicitly take annotations into account, in particular properly formulating the type synthesis problem.
+We also propose a preprocessing step called \emph{mode decoration} that helps the user to work with annotations in practice.
 
-By combining mode-decoration and bidirectional type synthesis, we show a trichotomy on raw terms, which is computationally a type synthesiser that checks if a given raw term is sufficiently annotated and if a sufficiently annotated raw term has an ordinary typing derivation, suggesting that type synthesis should not be viewed as a bisection between typable and untypable raw terms but a trichotomy in general.
+In a bit more detail:
+The type synthesis problem is not just about deciding whether a raw term is typable---there is a third possibility that the term does not have sufficient annotations.
+To solve the type synthesis problem, before attempting to decide typability (using a bidirectional type synthesiser), we should first decide whether the raw term has sufficient annotations, which corresponds to whether the term has a mode derivation.
+Our theory gives a (proof-relevant) \emph{mode decorator}, which constructs a mode derivation for a raw term or provide information that refutes the existence of any mode derivation and helps the user to pinpoint missing annotations.
+Then a bidirectional type synthesiser decides the typability of only mode-decorated raw terms instead of all raw terms.
+Soundness and completeness of bidirectional typing is reformulated as a one-to-one correspondence between bidirectional typing derivations and pairs of a typing derivation and a mode derivation for the same raw term; this reformulation is more natural and useful than annotatability described by \citet[Section~3.2]{Dunfield2021}.
+
+%As raw terms may not be sufficiently annotated, we propose a preprocessing step called \emph{mode decoration} to determine whether a raw term can be processed by a bidirectional type synthesiser later.
+%A bidirectional type synthesiser and a type synthesiser differ in their domain---bidirectional type synthesis only works for mode-decorated raw terms instead of all raw terms.
+
+%Our proof-relevant bidirectional type synthesiser is complete with respect to mode-decorated terms.
+%We also prove that a raw term with missing annotations is exactly a raw term without a mode derivation, so our mode decorator is indeed proof-relevant.
+
+%By combining mode-decoration and bidirectional type synthesis, we show a trichotomy on raw terms, which is computationally a type synthesiser that checks if a given raw term is sufficiently annotated and if a sufficiently annotated raw term has an ordinary typing derivation, suggesting that type synthesis should not be viewed as a bisection between typable and untypable raw terms but a trichotomy in general.
+%\Josh{elaborate}
 
@@
 \begin{enumerate}
-  \item general definitions for bidirectional type systems and mode-correctness,
-  \item the concept of mode decoration and the comparison between our completeness and \varcitet{Dunfield2021}{'s annotatability}, 
-    \Josh{generalisation mode decoration?}
+  \item general definitions for bidirectional type systems and mode-correctness;
+  \item mode derivations for explicitly dealing with annotations in the theory (in particular leading to a more natural and useful version of completeness than annotatability~\cite{Dunfield2021}), and mode decoration for helping the user to work with annotations in practice;
   \item rigorously proven consequences of mode-correctness---the uniqueness of synthesised types and the decidability of bidirectional type synthesis, which amounts to

### 2023-10-13 tex/ESOP24/S2-outline.tex
[ tex ] Update Example section

@@
 \begin{theorem}\label{thm:implementation}
-A type synthesiser that uses `not in synthesising mode' as its excuse can be constructed from a mode decorator and a bidirectional type synthesiser.
+A type synthesiser using `not in synthesising mode' as its excuse can be constructed from a mode decorator and a bidirectional type synthesiser.
 \end{theorem}

### 2023-10-13 tex/ESOP24/S3-defs.tex
[ tex ] Update Example section

@@
 By a \emph{simply typed language} $(\Sigma, \Omega)$, we mean a pair of a type signature~$\Sigma$ and a binding signature~$\Omega$.
-Given a simply typed language, first we define its raw terms.
+Now, we define raw terms for $(\Sigma, \Omega)$ first.
 
-\begin{definition}
 \begin{figure}
@@
 \end{figure}
+
+\begin{definition}
   For a simply typed language $(\Sigma, \Omega)$, the family of sets of \emph{raw terms} indexed by a list~$V$ of variables consists of
@@
 
-The definition of typing derivations is more involved.
+The definition of typing derivations is a bit more involved.
 We need some information to compare types on the object level during type synthesis and substitute those type variables in a typing derivation $\Gamma \vdash \tmOpts : A$ for an operation $o$ in $\Omega$ at some point.
@@
 \begin{definition}\label{def:typing-derivations}
-  \begin{figure}
-    \centering
-    \small
-    \judgbox{\Gamma |-_{\Sigma, \Omega} \isTerm{t} : A}{$\isTerm{t}$ has a concrete type $A$ under $\Gamma$ for a language $(\Sigma, \Omega)$}
-    \begin{mathpar}
-      \inferrule{(\isTerm{x} : A) \in \Gamma}{\Gamma |-_{\Sigma, \Omega} \isTerm{x} : A}\,\Rule{Var}
-      \and
-      \inferrule{\Gamma |-_{\Sigma, \Omega} \isTerm{t} : A}{\Gamma |-_{\Sigma, \Omega} (\isTerm{t \annotate A}) : A}\,\Rule{Anno}
-      \and
-      \inferrule{\rho : \Sub{\Xi}{\emptyset} \\  \Gamma, \vec{\isTerm{x}}_\isTerm{1} : \simsub{\Delta_{1}}{\rho} |-_{\Sigma, \Omega} \isTerm{t_1} : \simsub{A_{1}}{\rho} \quad\cdots\quad \Gamma, \vec{\isTerm{x}}_\isTerm{n} : \simsub{\Delta_{n}}{\rho} |-_{\Sigma, \Omega} \isTerm{t_n} : \simsub{A_{n}}{\rho}}
-      {\Gamma |-_{\Sigma, \Omega} \tmOpts : \simsub{A_0}{\rho}}\,\Rule{Op}
-      \\
-    \text{for $\bop$ in $\Omega$}
-    \end{mathpar}
-    \caption{Typing derivations}
-    \label{fig:extrinsic-typing}
-  \end{figure}
   For a simply typed language $(\Sigma, \Omega)$, the family of sets of \emph{typing derivations} of $\Gamma \vdash \isTerm{t} : A$, indexed by a typing context $\Gamma : \Cxt_\Sigma$, a raw term~$\isTerm{t}$ with free variables in $\erase{\Gamma}$, and a type $A : \Type_\Sigma(\emptyset)$, consists of
@@
 
+\begin{definition}\label{def:bidirectional-typing-derivations}\label{def:mode-derivations}
+  For a bidirectional type system $(\Sigma, \Omega)$,
+  \begin{itemize}
+    \item the set of \emph{bidirectional typing derivations} of $\Gamma \vdash_{\Sigma, \Omega} t :^\dir{d} A$, indexed by a typing context $\Gamma$, a raw term $\isTerm{t}$ under $\erase{\Gamma}$, a mode $\dir{d}$, and a type $A$, is defined in \Cref{fig:bidirectional-typing-derivations} and particularly
+          \[
+            \Gamma |-_{\Sigma, \Omega} \tmOpts :^{\dir{d}} \simsub{A_0}{\rho}
+          \]
+          has a derivation for $\biop$ in $\Omega$ if there is $\rho\colon \Xi \to \Type_{\Sigma}(\emptyset)$ and a derivation of $\Gamma, \vars{x}_{\isTerm{i}} : \simsub{\Delta_i}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_i} :^{\dir{d_i}} \simsub{A_i}{\rho}$ for each $i$;
+    \item the set of \emph{mode derivations} of $V |-_{\Sigma, \Omega} t^\dir{d}$, indexed by a list $V$ of variables, a raw term $\isTerm{t}$ under $V$, and a mode $d$, is defined in \Cref{fig:mode-derivations}.
+  \end{itemize}
+  The two judgements {\small$\boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \syn A}$} and {\small$\boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \chk A}$} stand for ${\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} :^{\syn} A}$ and ${\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} :^{\chk} A}$, respectively.
+  A typing rule is \emph{checking} if its conclusion mode is $\chk$ or \emph{synthesising} otherwise.
+\end{definition}
+
+Every bidirectional binding signature $\Omega$ gives rise to a binding signature $\erase{\Omega}$ if we erase modes from $\Omega$, called the \emph{(mode) erasure} of $\Omega$.
+Hence a bidirectional type system $(\Sigma, \Omega)$ also specifies a simply typed language $(\Sigma, \erase{\Omega})$, including raw terms and typing derivations.
+
+\begin{example}\label{ex:signature-simply-typed-lambda}
+Having established generic definitions, we can now specify simply typed $\lambda$-calculus and its bidirectional type system---including raw terms, (bidirectional) typing derivations, and mode derivations---using just a pair of signatures $\Sigma_{\bto}$ (\Cref{ex:type-signature-for-function-type}) and $\Omega^{\Leftrightarrow}_\Lambda$ which consists of
+\[
+  \aritysymbol{\mathsf{abs}}{A , B}{[A]B^{\chk}}{(A \bto B)^{\chk}}
+  \quad\text{and}\quad
+  \aritysymbol{\mathsf{app}}{A, B}{(A \bto B)^{\syn}, A^{\chk}}{B^{\syn}}.
+\]
+\end{example}
+More importantly, we are able to reason about constructions and properties that hold for any simply typed language with a bidirectional type system once and for all by quantifying over $(\Sigma, \Omega)$.
+
+\begin{figure}
+  \centering
+  \small
+  \judgbox{\Gamma |-_{\Sigma, \Omega} \isTerm{t} : A}{$\isTerm{t}$ has a concrete type $A$ under $\Gamma$ for a language $(\Sigma, \Omega)$}
+  \begin{mathpar}
+    \inferrule{(\isTerm{x} : A) \in \Gamma}{\Gamma |-_{\Sigma, \Omega} \isTerm{x} : A}\,\Rule{Var}
+    \and
+    \inferrule{\Gamma |-_{\Sigma, \Omega} \isTerm{t} : A}{\Gamma |-_{\Sigma, \Omega} (\isTerm{t \annotate A}) : A}\,\Rule{Anno}
+    \and
+    \inferrule{\rho : \Sub{\Xi}{\emptyset} \\  \Gamma, \vec{\isTerm{x}}_\isTerm{1} : \simsub{\Delta_{1}}{\rho} |-_{\Sigma, \Omega} \isTerm{t_1} : \simsub{A_{1}}{\rho} \quad\cdots\quad \Gamma, \vec{\isTerm{x}}_\isTerm{n} : \simsub{\Delta_{n}}{\rho} |-_{\Sigma, \Omega} \isTerm{t_n} : \simsub{A_{n}}{\rho}}
+    {\Gamma |-_{\Sigma, \Omega} \tmOpts : \simsub{A_0}{\rho}}\,\Rule{Op}
+    \\
+  \text{for $\bop$ in $\Omega$}
+  \end{mathpar}
+  \caption{Typing derivations}
+  \label{fig:extrinsic-typing}
+\end{figure}
 \begin{figure}
@@
 
-\begin{definition}\label{def:bidirectional-typing-derivations}\label{def:mode-derivations}
-  For a bidirectional type system $(\Sigma, \Omega)$,
-  \begin{itemize}
-    \item the set of \emph{bidirectional typing derivations} of $\Gamma \vdash_{\Sigma, \Omega} t :^\dir{d} A$, indexed by a typing context $\Gamma$, a raw term $\isTerm{t}$ under $\erase{\Gamma}$, a mode $\dir{d}$, and a type $A$, is defined in \Cref{fig:bidirectional-typing-derivations} and particularly
-          \[
-            \Gamma |-_{\Sigma, \Omega} \tmOpts :^{\dir{d}} \simsub{A_0}{\rho}
-          \]
-          has a derivation for $\biop$ in $\Omega$ if there is $\rho\colon \Xi \to \Type_{\Sigma}(\emptyset)$ and a derivation of $\Gamma, \vars{x}_{\isTerm{i}} : \simsub{\Delta_i}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_i} :^{\dir{d_i}} \simsub{A_i}{\rho}$ for each $i$;
-    \item the set of \emph{mode derivations} of $V |-_{\Sigma, \Omega} t^\dir{d}$, indexed by a list $V$ of variables, a raw term $\isTerm{t}$ under $V$, and a mode $d$, is defined in \Cref{fig:mode-derivations}.
-  \end{itemize}
-  The two judgements {\small$\boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \syn A}$} and {\small$\boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \chk A}$} stand for ${\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} :^{\syn} A}$ and ${\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} :^{\chk} A}$, respectively.
-  A typing rule is \emph{checking} if its conclusion mode is $\chk$ or \emph{synthesising} otherwise.
-\end{definition}
-
-\begin{figure}[b]
+\begin{figure}
   \centering
@@
 \end{figure}
-Every bidirectional binding signature $\Omega$ gives rise to a binding signature $\erase{\Omega}$ if we erase modes from $\Omega$, called the \emph{(mode) erasure} of $\Omega$.
-Hence a bidirectional type system $(\Sigma, \Omega)$ also specifies a simply typed language $(\Sigma, \erase{\Omega})$, including raw terms and typing derivations.
-
-\begin{example}\label{ex:signature-simply-typed-lambda}
-Having established generic definitions, we can now specify the simply typed $\lambda$-calculus and its bidirectional type system---including raw terms, (bidirectional) typing derivations, and mode derivations---using just a pair of signatures $\Sigma_{\bto}$ (\Cref{ex:type-signature-for-function-type}) and $\Omega^{\Leftrightarrow}_\Lambda$ which consists of
-\[
-  \aritysymbol{\mathsf{abs}}{A , B}{[A]B^{\chk}}{(A \bto B)^{\chk}}
-  \quad\text{and}\quad
-  \aritysymbol{\mathsf{app}}{A, B}{(A \bto B)^{\syn}, A^{\chk}}{B^{\syn}}.
-\]
-\end{example}
-More importantly, we are able to reason about constructions and properties that hold for any simply typed language with a bidirectional type system once and for all by quantifying over $(\Sigma, \Omega)$.

### 2023-10-13 tex/ESOP24/S4-pre-synthesis.tex
[ tex ] Update Example section

@@
 By bringing mode derivations into the picture, we are able to give a natural formulation of soundness and completeness of a bidirectional type system with respect to its erasure to an ordinary type system in \cref{sec:soundness-and-completeness}.
-We also formulate \varcitet{Dunfield2021}{'s} annotatability and discuss its relationship with our completeness and generalised mode decoration in \cref{sec:annotatability}.
+We also formulate annotatability~\cite{Dunfield2021} and discuss its relationship with our completeness and generalised mode decoration in \cref{sec:annotatability}.
 

### 2023-10-13 tex/ESOP24/S5-synthesis.tex
[ tex ] Update Example section

@@
 \begin{definition}
-  The finite subset\footnote{%
+  The finite subsets\footnote{%
   There are various definitions for finite subsets of a set within type theory, but for our purposes the choice among these definitions is not a matter of concern.}
-  of \emph{(free) variables} of a type $A$ is denoted by $\fv(A)$.
-  The subset $\fv(\Delta)$ of variables in an extension context~$\Delta$ is defined by\/ $\fv(\cdot) = \emptyset$ and\/ $\fv(\Delta, A) = \fv(\Delta) \cup \fv(A)$.
-  For an argument list $\biargs$, its set of \emph{synthesised type variables} is denoted by $\synvar(\biargs)$.
+  of \emph{(free) variables} of a type $A$ and of variables in an extension context~$\Delta$ are denoted by $\fv(A)$ and $\fv(\Delta)$ respectively.
+  For an argument list $\biargs$ the set of type variables $A_{i}^{\dir{d_i}}$ with $d_i$ being $\syn$ is denoted by $\synvar(\biargs)$, i.e.\ $\synvar$ gives the set of type variables that will be synthesised during type synthesis.
 \end{definition}
-This subset contains type variables of a synthesising argument and they are exactly those type variables that will be synthesised during type synthesis.
 
@@
 %  \end{itemize}
-  We have completed the proof by induction on the mode derivation $\erase{\Gamma} |-_{\Sigma, \Omega} t^{\dir{d}}$.\qed
+  We have completed the proof by induction on the derivation $\erase{\Gamma} |-_{\Sigma, \Omega} t^{\dir{d}}$.\qed
 \end{proof}

### 2023-10-13 tex/ESOP24/S6-example.tex
[ update ] Add ESOP submission draft
[ tex ] Update Example section

--- 2023-07-20 tex/example.tex
@@
 \section{Examples}\label{sec:example}
-Our theory works for any language that can be specified by a bidirectional binding signature.
-To showcase its general applicability, we present two additional examples:
-\begin{enumerate}
-\item \PCF with products and let bindings, and
-\item spine calculus, a simply typed $\lambda$-calculus with applications in spine form.
-\end{enumerate}
-The first is a standard exercise in implementing a type synthesiser.
-The second example includes not only infinitely many operations but also a type which is not formed by solely one type construct~$\tyOp$.
+In this section, we use two more examples to exhibit the applicability of our approach.
+This first one has infinitely many operations and the second one includes many more constructs than simply typed $\lambda$-calculus with some common features in a programming language.
 
-\subsection{PCF with Products and Let Bindings}\label{subsec:PCF}
-Although we have developed a theory for presenting a bidirectional type system and a condition for the decidability of bidirectional type synthesis, designing a bidirectional type system, or its signature, remains a creative activity.\footnote{This has also been the case for parsing: while there are grammars and parser generators, one still has to design a grammar to parse or disambiguate an ambiguous grammar.}
-While the focus of this paper is not to design bidirectional type systems, we simply adopt the example of a bidirectional \PCF from \citet{Wadler2022}, which extends bidirectional simply typed $\lambda$-calculus in \Cref{sec:key-ideas} by adding the following rules:
-\bgroup
-\small
-  \begin{mathpar}
-    \inferrule{ }{\Gamma \vdash \mathtt{z} \chk \mathtt{nat}}
-    \and
-    \inferrule{\Gamma \vdash t \chk \mathtt{nat}}{\Gamma \vdash \mathtt{s}(t) \chk \mathtt{nat}}
-    \and
-    \inferrule{\Gamma \vdash t \syn \mathtt{nat} \and \Gamma \vdash t_0 \chk A \and \Gamma, x : \mathtt{nat} \vdash t_1 \chk A}{\Gamma \vdash \mathtt{ifz}(t_0; x.t_1)(t) \chk A}
-    \and
-    \inferrule{\Gamma \vdash t \chk A \and \Gamma \vdash u \chk B}{\Gamma \vdash (t, u) \chk A \times B}
-    \and
-    \inferrule{\Gamma \vdash t \syn A \times B}{\Gamma \vdash \mathtt{proj}_1(t) \syn A}
-    \and
-    \inferrule{\Gamma \vdash u \syn A \times B}{\Gamma \vdash \mathtt{proj}_2(u) \syn A}
-    \\
-    \inferrule{\Gamma, x : A \vdash t \chk A}{\Gamma \vdash \mu x.\, x \chk A}
-    \and
-    \inferrule{\Gamma \vdash t \syn A \and \Gamma, x : A \vdash u \chk B}{\Gamma \vdash \mathtt{let}\;x = t\;\mathtt{in}\;u\chk B}
-  \end{mathpar}
-\egroup
-
-%In this language, there are additional type constructs other than $\bto$, so first we extend $\Sigma_{\bto}$ (\Cref{ex:type-signature-for-function-type}) with operations
-In this language, we introduce additional type constructs apart from $\bto$.
-To do so, we extend $\Sigma_{\bto}$ (as shown in \Cref{ex:type-signature-for-function-type}) with the operations $\mathsf{nat}$ and $\mathsf{times}$ such that
-\[
-  \arity({\mathsf{nat}}) = 0 \quad\text{and}\quad \arity({\mathsf{times}}) = 2
-\]
-enabling the introduction of $\mathtt{nat}$ and $A \times B$ as $\tyOp_{\mathsf{nat}}$ and $\tyOp_{\mathsf{times}}(A, B)$, respectively.
-Following this, we also extend the bidirectional binding signature $\Omega_{\Lambda}^{\Leftrightarrow}$ with an operation for each rule above:
-\begin{flalign*}
-  &  \aritysymbol{\mathsf{z}}{\cdot}{\cdot}{\mathtt{nat}^{\chk}}
-  && \aritysymbol{\mathsf{s}}{\cdot}{\mathtt{nat}^{\chk}}{\mathtt{nat}^{\chk}}
-  && \aritysymbol{\mathsf{ifz}}{A}{\mathtt{nat}^{\syn}, A^{\chk}, A^{\chk}}{A^{\chk}} \\
-  &  \aritysymbol{\mathsf{pair}}{A, B}{A^{\chk}, B^{\chk}}{A \times B^{\chk}}
-  && \aritysymbol{\mathsf{proj}_1}{A, B}{A \times B^{\syn}}{A^{\syn}}
-  && \aritysymbol{\mathsf{proj}_2}{A, B}{A \times B^{\syn}}{B^{\syn}} \\
-  &  \aritysymbol{\mathsf{mu}}{A}{[A]A^{\chk}}{A^{\chk}}
-  && \aritysymbol{\mathsf{let}}{A, B}{A^{\syn},[A]B^{\chk}}{B^{\chk}}
-\end{flalign*}
-With these operations, a straightforward check shows that every rule is mode-correct, therefore this is a mode-correct bidirectional type system.
-As a result, we can instantiate the type synthesiser presented in \Cref{sec:type-synthesis} for this system, eliminating the need for re-implementation.
-
-\subsection{Spine Calculus}\label{subsec:spine}
-A spine application, denoted as $t\;u_1\;\ldots\;u_n$, is a variant of application that consists of a head term $t$ and an indeterminate number of arguments $u_1\;u_2\;\dots\;u_n$.
+\subsection{Spine application}\label{subsec:spine}
+A spine application $t\;u_1\;\ldots\;u_n$ is a variant form of application that consists of a head term $t$ and an indeterminate number of arguments $u_1\;u_2\;\dots\;u_n$.
 This arrangement allows direct access to the head term, making it practical in various applications.
@@
 \egroup
-We use the same type signature $\Sigma_{\bto}$ but extend the bidirectional binding signature $\Omega_{\Lambda}^{\Leftrightarrow}$ with 
+We use the same type signature $\Sigma_{\bto}$ but extend the signature $\Omega_{\Lambda}^{\Leftrightarrow}$ with 
 \[
@@
 \]
-for each $n : \N$.
-Then each spine application $t\;u_1\;\ldots\;u_n$ can be introduced as $\tmOp_{\mathsf{app}_n}(t; u_1; \ldots; u_n)$.
+for each $n : \N$, so each application $t\;u_1\;\ldots\;u_n$ can be introduced as $\tmOp_{\mathsf{app}_n}(t; u_1; \ldots; u_n)$.
+\subsection{Computational calculi}\label{subsec:PCF}
+The code size of a type synthesiser grows proportionally to the number of type constructs and the number of arguments for each language construct, making it a tedious task to implement.
+Even for a simply typed language which is generally straightforward to implement a synthesiser individually, having a generator can reduce the effort significantly.
+For illustrative purposes, let us consider the Moggi's computational $\lambda$-calculus~\cite{Moggi1989} with additional constructs.
+
+We first introduce additional type constructs apart from $\bto$ by extending $\Sigma_{\bto}$ (as shown in \Cref{ex:type-signature-for-function-type}) with $\mathsf{nat}$, $\mathsf{prod}$, $\mathsf{sum}$, and $\mathsf{T}$ such that
+\[
+  \arity({\mathsf{nat}}) = 0\qquad
+  \arity({\mathsf{T}}) = 1\qquad
+  \arity({\mathsf{prod}}) = \arity({\mathsf{sum}}) = 2
+\]
+introducing $\mathtt{nat}$, $A \times B$, $A + B$, $T(A)$ as $\tyOp_{\mathsf{nat}}$, $\tyOp_{\mathsf{prod}}(A, B)$, $\tyOp_{\mathsf{sum}}(A, B)$, and $\tyOp_{\mathsf{T}}(A)$ respectively.
+The bidirectional binding signature $\Omega_{\Lambda}^{\Leftrightarrow}$ is then extended with operations in following table.
+
+%\begin{table}
+\bgroup
+  \renewcommand{\arraystretch}{2}
+  \centering\small
+\begin{tabular}{c | l}
+  Rules & Operations \\ \hline\hline
+  $\inferrule{}{\Gamma \vdash \mathtt{z} \chk \mathtt{nat}}$ &  $\aritysymbol{\mathsf{z}}{\cdot}{\cdot}{\mathtt{nat}^{\chk}}$ 
+  \\
+ $\inferrule{\Gamma \vdash t \chk \mathtt{nat}}{\Gamma \vdash \mathtt{s}(t) \chk \mathtt{nat}}$ & $\aritysymbol{\mathsf{s}}{\cdot}{\mathtt{nat}^{\chk}}{\mathtt{nat}^{\chk}}$ \\
+ $\inferrule{\Gamma \vdash t \syn \mathtt{nat} \and \Gamma \vdash t_0 \chk A \and \Gamma, x : \mathtt{nat} \vdash t_1 \chk A}{\Gamma \vdash \mathtt{ifz}(t_0; x.t_1)(t) \chk A}$ & $\aritysymbol{\mathsf{ifz}}{A}{\mathtt{nat}^{\syn}, A^{\chk}, A^{\chk}}{A^{\chk}}$ \\
+ $\inferrule{\Gamma \vdash t \chk A \and \Gamma \vdash u \chk B}{\Gamma \vdash (t, u) \chk A \times B}$ & $\aritysymbol{\mathsf{pair}}{A, B}{A^{\chk}, B^{\chk}}{A \times B^{\chk}}$  \\
+ $\inferrule{\Gamma \vdash t \syn A \times B}{\Gamma \vdash \mathtt{proj}_1(t) \syn A}$ & $\aritysymbol{\mathsf{proj}_1}{A, B}{A \times B^{\syn}}{A^{\syn}}$ \\
+
+ $\inferrule{\Gamma \vdash u \syn A \times B}{\Gamma \vdash \mathtt{proj}_2(u) \syn A}$ & $\aritysymbol{\mathsf{proj}_2}{A, B}{A \times B^{\syn}}{B^{\chk}}$ \\
+
+ $\inferrule{\Gamma \vdash t \chk A}{\Gamma \vdash \mathtt{inj}_1(t) \chk A + B}$ & $\aritysymbol{\mathsf{inj}_1}{A, B}{A^{\chk}}{A + B^{\chk}}$ \\
+
+ $\inferrule{\Gamma \vdash u \chk B}{\Gamma \vdash \mathtt{inj}_2(u) \chk A + B}$ & $\aritysymbol{\mathsf{inj}_2}{A, B}{B^{\chk}}{A + B^{\chk}}$ \\
+
+ $\inferrule{\Gamma \vdash u \syn A + B \and \Gamma, x_1 : A \vdash t_1 \chk C \and \Gamma, x_2 : B \vdash t_2 \chk C}{\Gamma \vdash \mathtt{case}(u; x_1.
+t_1; x_2.
+t_2) \chk C}$ & $\aritysymbol{\mathsf{case}}{A, B, C}{{A + B}^{\syn}, [A]C^{\chk}, [B]C^{\chk}}{C^{\syn}}$ \\
+
+ $\inferrule{\Gamma, x : A \vdash t \chk A}{\Gamma \vdash \mu x.\, t \chk A}$ & $\aritysymbol{\mathsf{mu}}{A}{[A]A^{\chk}}{A^{\chk}}$ \\
+
+ $\inferrule{\Gamma \vdash t \syn A \and \Gamma, x : A \vdash u \chk B}{\Gamma \vdash \mathtt{let}\;x = t\;\mathtt{in}\;u\chk B}$ & 
+  $\aritysymbol{\mathsf{let}}{A, B}{A^{\syn},[A]B^{\chk}}{B^{\chk}}$ \\
+ $\inferrule{\Gamma \vdash t \syn A}{\Gamma \vdash \mathtt{ret}(t) \syn T(A)}$ & $\aritysymbol{\mathsf{ret}}{A}{A^{\syn}}{T(A)^{\syn}}$ \\
+ $\inferrule{\Gamma \vdash t \syn T(A) \and \Gamma, x : A \vdash u \syn T(B)}{\Gamma \vdash \mathtt{bind}(t; x.u) \syn T(B)}$ & $\aritysymbol{\mathsf{bind}}{A, B}{T(A)^{\syn}, [A]T(B)^{\syn}}{T(B)^{\syn}}$
+\end{tabular}
+\egroup
+
+The extended language has `only' 15 constructs, far fewer than what the surface language of a programming language in use would have. 
+However, if we implement a proof-relevant type synthesiser following \citet{Wadler2022}, then there will be nearly 100 cases to check on top of other auxiliary definitions.

### 2023-10-13 tex/ESOP24/appendix.tex
[ tex ] Update Example section

@@
 \begin{definition}\label{def:syn-var}
-  For an argument list $\biargs$, the set of \emph{synthesised type variables} is defined inductively by 
+  The subset $\fv(\Delta)$ of variables in an extension context~$\Delta$ is defined by\/ $\fv(\cdot) = \emptyset$ and\/ $\fv(\Delta, A) = \fv(\Delta) \cup \fv(A)$.
+  For an argument list $\biargs$, the set of \emph{synthesised type variables} is defined by 
   \begin{align*}

### 2023-10-13 tex/ESOP24/S1-intro.tex
[ tex ] Revised S1

@@
 
-Our work adopts a proof-relevant approach to (bidirectional) type synthesis, as illustrated by \citet{Wadler2022} for \PCF.
+Our work adopts a proof-relevant approach to (bidirectional) type synthesis, as illustrated by \citet{Wadler2022} for simply typed $\lambda$-calculus with natural numbers.
 The proof-relevant formulation deviates from the usual one:
@@
   \item general definitions for bidirectional type systems and mode-correctness;
-  \item mode derivations for explicitly dealing with annotations in the theory (in particular leading to a more natural and useful version of completeness than annotatability~\cite{Dunfield2021}), and mode decoration for helping the user to work with annotations in practice;
+  \item mode derivations for explicitly dealing with annotations in the theory (in particular leading to a more natural and useful version of completeness than \varcitet{Dunfield2021}{'s annotatability}), and mode decoration for helping the user to work with annotations in practice;
   \item rigorously proven consequences of mode-correctness---the uniqueness of synthesised types and the decidability of bidirectional type synthesis, which amounts to

### 2023-10-13 tex/ESOP24/BiSig.tex
[ tex ] Update Example section
[ tex ] Rewrite Formalisation

@@
 \begin{abstract}
-There has been much progress in designing bidirectional type systems and associated type synthesis algorithms, but mostly on a case-by-case basis.
+There has been much progress in designing bidirectional type systems and associated type synthesis algorithms, but mainly on a case-by-case basis.
 This situation is in stark contrast to parsing, for which there have been general theories and widely applicable tools such as parser generators.
@@
 Soundness, completeness, and mode-correctness are studied universally for any simply typed language, which are sufficient conditions for deriving a type synthesiser.
-We propose a preprocessing step called \emph{mode decoration}, which helps the user to deal with missing type annotations in an abstract syntax tree given by programmers.
+We propose a preprocessing step called \emph{mode decoration}, which helps the user to deal with missing type annotations in a given abstract syntax tree.
 The entire development is formalised in \Agda and can be further integrated with other language-formalisation frameworks.
@@
 \input{S8-future}
-\appendix
-\input{appendix}
 
@@
 
-\IfFileExists{reference}{
+\IfFileExists{reference.bib}{
 \bibliography{reference}
@@
 
+\appendix
+\input{appendix}
+
 \end{document}

### 2023-10-13 tex/ESOP24/S5-synthesis.tex
[ tex ] Rewrite Formalisation

@@
   Mode-correctness is fundamentally a condition for bidirectional typing \emph{rules}, not for derivations.
-  Thus, this property cannot be formalised without treating rules as some mathematical object, such as the notion of bidirectional binding signature presented in \Cref{sec:defs}.
-  This contrasts with the properties in \Cref{sec:pre-synthesis}, which can still be specified for individual systems even in the absence of a generic definition of bidirectional type systems.
+  Thus, this property cannot be formulated without treating rules as some mathematical object such as those general definitions in \Cref{sec:defs}.
+  This contrasts with the properties in \Cref{sec:pre-synthesis}, which can still be specified for individual systems in the absence of a general definition.
 \end{remark}
 
-It is easy to check the bidirectional type system $(\Sigma_{\bto}, \Omega^{\Leftrightarrow}_{\Lambda}$) in \Cref{ex:signature-simply-typed-lambda} for simply typed $\lambda$-calculus is mode-correct manually by definition or automatically by the following lemma:
-\begin{lemma}
-  It is decidable if an operation $\biop$ is mode-correct or not.
+It is easy to check the bidirectional type system $(\Sigma_{\bto}, \Omega^{\Leftrightarrow}_{\Lambda}$) for simply typed $\lambda$-calculus is mode-correct by definition or by the following lemma:
+\begin{lemma}\label{lem:decidability-mode-correctness}
+  It is decidable whether an operation $\biop$ is mode-correct.
 \end{lemma}

### 2023-10-13 tex/ESOP24/S6-example.tex
[ tex ] Rewrite Formalisation

@@
 \section{Examples}\label{sec:example}
-In this section, we use two more examples to exhibit the applicability of our approach.
-This first one has infinitely many operations and the second one includes many more constructs than simply typed $\lambda$-calculus with some common features in a programming language.
+Now we use two more examples to exhibit the applicability of our approach.
+This first one has infinitely many operations and the second one includes many more constructs than simply typed $\lambda$-calculus with some common features in a programming language to justify the practical needs for a general development.
 
@@
 
-At first glance, accommodating this form of application may seem impossible, given that the number of arguments a construct can accept has to be fixed in a signature.
-Nonetheless, there is no constraint on the total number of operation symbols a signature can have, allowing us to establish corresponding constructs for each number $n$ of arguments, thereby exhibiting the necessity of having an arbitrary set (with decidable equality) for operation symbols.
-
-To illustrate this, let us consider the following rule:
+At first glance, accommodating this form of application may seem impossible, given that the number of arguments for a construct is finite and has to be fixed.
+Nonetheless, there is no constraint on the total number of operation symbols a signature can have, allowing us to establish a corresponding construct for each number $n$ of arguments, i.e.\ viewing the following rule
 \bgroup
@@
 \egroup
-We use the same type signature $\Sigma_{\bto}$ but extend the signature $\Omega_{\Lambda}^{\Leftrightarrow}$ with 
+as a rule schema parametrised by $n$.
+Then, we use the same type signature~$\Sigma_{\bto}$ but extend the signature $\Omega_{\Lambda}^{\Leftrightarrow}$ with 
 \[
@@
 \]
-for each $n : \N$, so each application $t\;u_1\;\ldots\;u_n$ can be introduced as $\tmOp_{\mathsf{app}_n}(t; u_1; \ldots; u_n)$.
+so each application $t\;u_1\;\ldots\;u_n$ can be introduced as $\tmOp_{\mathsf{app}_n}(t; u_1; \ldots; u_n)$, thereby exhibiting the necessity of having an arbitrary set for operation symbols.
+
 \subsection{Computational calculi}\label{subsec:PCF}
 The code size of a type synthesiser grows proportionally to the number of type constructs and the number of arguments for each language construct, making it a tedious task to implement.
-Even for a simply typed language which is generally straightforward to implement a synthesiser individually, having a generator can reduce the effort significantly.
+Even for a simply typed language which is generally straightforward to implement a synthesiser individually, having a generator can still reduce the effort significantly.
 For illustrative purposes, let us consider the Moggi's computational $\lambda$-calculus~\cite{Moggi1989} with additional constructs.
@@
 The extended language has `only' 15 constructs, far fewer than what the surface language of a programming language in use would have. 
-However, if we implement a proof-relevant type synthesiser following \citet{Wadler2022}, then there will be nearly 100 cases to check on top of other auxiliary definitions.
+However, if we implement a proof-relevant type synthesiser in the style of \citet{Wadler2022}, then there will be nearly 100 cases to check on top of other auxiliary definitions.
+On the other hand, similar to a parser generator, a type-synthesiser generator only needs a language specification $(\Sigma, \Omega)$ from the user to produce a corresponding synthesiser. 
+We will see in the next section how the actual formalisation is used.

### 2023-10-15 tex/ESOP24/S6-example.tex
[ tex ] Update Future work

@@
 \section{Examples}\label{sec:example}
-Now we use two more examples to exhibit the applicability of our approach.
-This first one has infinitely many operations and the second one includes many more constructs than simply typed $\lambda$-calculus with some common features in a programming language to justify the practical needs for a general development.
+To exhibit the applicability of our approach, we discuss two more examples:
+one has infinitely many operations and the other includes many more constructs than simply typed $\lambda$-calculus, exhibiting the practical side of a general treatment.
 
 \subsection{Spine application}\label{subsec:spine}
-A spine application $t\;u_1\;\ldots\;u_n$ is a variant form of application that consists of a head term $t$ and an indeterminate number of arguments $u_1\;u_2\;\dots\;u_n$.
-This arrangement allows direct access to the head term, making it practical in various applications.
-For instance, \Agda's core language employs this form of application, as does its reflected syntax used for metaprogramming.
+A spine application $t\;u_1\;\ldots\;u_n$ is a form of application that consists of a head term $t$ and an indeterminate number of arguments $u_1\;u_2\;\dots\;u_n$.
+This arrangement allows direct access to the head term, making it practical in various applications, and has been used by \Agda's core language.
+%employs this form of application, as does its reflected syntax used for metaprogramming.
 
 At first glance, accommodating this form of application may seem impossible, given that the number of arguments for a construct is finite and has to be fixed.
-Nonetheless, there is no constraint on the total number of operation symbols a signature can have, allowing us to establish a corresponding construct for each number $n$ of arguments, i.e.\ viewing the following rule
+Nonetheless, the total number of operation symbols a signature can have need not be finite, allowing us to establish a corresponding construct for each number $n$ of arguments, i.e.\ viewing the following rule
 \bgroup
@@
 \egroup
-as a rule schema parametrised by $n$.
-Then, we use the same type signature~$\Sigma_{\bto}$ but extend the signature $\Omega_{\Lambda}^{\Leftrightarrow}$ with 
+as a rule schema parametrised by $n$, so the signature $\Omega_{\Lambda}^{\Leftrightarrow}$ can be extended with 
 \[
@@
 \]
-so each application $t\;u_1\;\ldots\;u_n$ can be introduced as $\tmOp_{\mathsf{app}_n}(t; u_1; \ldots; u_n)$, thereby exhibiting the necessity of having an arbitrary set for operation symbols.
+Each application $t\;u_1\;\ldots\;u_n$ can be introduced as $\tmOp_{\mathsf{app}_n}(t; u_1; \ldots; u_n)$, thereby exhibiting the necessity of having an arbitrary set for operation symbols.
 
 \subsection{Computational calculi}\label{subsec:PCF}
-The code size of a type synthesiser grows proportionally to the number of type constructs and the number of arguments for each language construct, making it a tedious task to implement.
-Even for a simply typed language which is generally straightforward to implement a synthesiser individually, having a generator can still reduce the effort significantly.
-For illustrative purposes, let us consider the Moggi's computational $\lambda$-calculus~\cite{Moggi1989} with additional constructs.
+Implementing a standalone type synthesizer for a simply typed language is typically a straightforward task.
+However, the code size increases proportionally to the number of type constructs and of arguments associated with each term construct.
+When dealing with a fixed number of $n$ type constructs, for each synthesising construct, there are two cases for a checking argument but $n + 1$ possible cases for each synthesising argument---the successful synthesis of the expected type, instances where it fails, or $n-1$ cases where the expected type does not match; similarly for a checking construct---making the task tedious and error-prone.
+Therefore, having a generator can be helpful and significantly reduce the effort for implementation.
 
-We first introduce additional type constructs apart from $\bto$ by extending $\Sigma_{\bto}$ (as shown in \Cref{ex:type-signature-for-function-type}) with $\mathsf{nat}$, $\mathsf{prod}$, $\mathsf{sum}$, and $\mathsf{T}$ such that
-\[
-  \arity({\mathsf{nat}}) = 0\qquad
-  \arity({\mathsf{T}}) = 1\qquad
-  \arity({\mathsf{prod}}) = \arity({\mathsf{sum}}) = 2
-\]
-introducing $\mathtt{nat}$, $A \times B$, $A + B$, $T(A)$ as $\tyOp_{\mathsf{nat}}$, $\tyOp_{\mathsf{prod}}(A, B)$, $\tyOp_{\mathsf{sum}}(A, B)$, and $\tyOp_{\mathsf{T}}(A)$ respectively.
-The bidirectional binding signature $\Omega_{\Lambda}^{\Leftrightarrow}$ is then extended with operations in following table.
-
-%\begin{table}
-\bgroup
-  \renewcommand{\arraystretch}{2}
-  \centering\small
+For illustrative purposes, consider a computational calculus~\cite{Moggi1989} with additional constructs listed in \cref{tab:computational-calculus}.
+The extended language has `only' 15 constructs, far fewer than a realistic programming language would have, but there are nearly 100 possible cases already to consider during bidirectional type synthesis.
+\begin{table}[h!]
+  \renewcommand{\arraystretch}{2.5}
+  \setlength{\tabcolsep}{3pt}
+  \centering\footnotesize
 \begin{tabular}{c | l}
   Rules & Operations \\ \hline\hline
-  $\inferrule{}{\Gamma \vdash \mathtt{z} \chk \mathtt{nat}}$ &  $\aritysymbol{\mathsf{z}}{\cdot}{\cdot}{\mathtt{nat}^{\chk}}$ 
+  $\inferrule{ }{\Gamma \vdash \mathtt{z} \chk \mathtt{nat}}$ &  $\aritysymbol{\mathsf{z}}{\cdot}{\cdot}{\mathtt{nat}^{\chk}}$ 
   \\
@@
  $\inferrule{\Gamma \vdash t \chk A \and \Gamma \vdash u \chk B}{\Gamma \vdash (t, u) \chk A \times B}$ & $\aritysymbol{\mathsf{pair}}{A, B}{A^{\chk}, B^{\chk}}{A \times B^{\chk}}$  \\
- $\inferrule{\Gamma \vdash t \syn A \times B}{\Gamma \vdash \mathtt{proj}_1(t) \syn A}$ & $\aritysymbol{\mathsf{proj}_1}{A, B}{A \times B^{\syn}}{A^{\syn}}$ \\
+ $\inferrule{\Gamma \vdash t \syn A_1 \times A_2}{\Gamma \vdash \mathtt{proj}_i(t) \syn A_i}$ for $i = 1, 2$ & $\aritysymbol{\mathsf{proj}_i}{A_1, A_2}{A_1 \times A_2^{\syn}}{A_i^{\syn}}$ for $i = 1, 2$ \\
 
- $\inferrule{\Gamma \vdash u \syn A \times B}{\Gamma \vdash \mathtt{proj}_2(u) \syn A}$ & $\aritysymbol{\mathsf{proj}_2}{A, B}{A \times B^{\syn}}{B^{\chk}}$ \\
-
- $\inferrule{\Gamma \vdash t \chk A}{\Gamma \vdash \mathtt{inj}_1(t) \chk A + B}$ & $\aritysymbol{\mathsf{inj}_1}{A, B}{A^{\chk}}{A + B^{\chk}}$ \\
-
- $\inferrule{\Gamma \vdash u \chk B}{\Gamma \vdash \mathtt{inj}_2(u) \chk A + B}$ & $\aritysymbol{\mathsf{inj}_2}{A, B}{B^{\chk}}{A + B^{\chk}}$ \\
+ $\inferrule{\Gamma \vdash t \chk A_i}{\Gamma \vdash \mathtt{inj}_i(t) \chk A_1 + A_2}$ for $i = 1, 2$ & $\aritysymbol{\mathsf{inj}_i}{A_1, A_2}{A_i^{\chk}}{A_1 + A_2^{\chk}}$ for $i = 1, 2$ \\
 
@@
 \end{tabular}
-\egroup
+\caption{A computational calculus with naturals, products, sums, and general recursion}
+\label{tab:computational-calculus}
+\end{table}
 
-The extended language has `only' 15 constructs, far fewer than what the surface language of a programming language in use would have. 
-However, if we implement a proof-relevant type synthesiser in the style of \citet{Wadler2022}, then there will be nearly 100 cases to check on top of other auxiliary definitions.
-On the other hand, similar to a parser generator, a type-synthesiser generator only needs a language specification $(\Sigma, \Omega)$ from the user to produce a corresponding synthesiser. 
-We will see in the next section how the actual formalisation is used.
+On the other hand, similar to a parser generator, a type-synthesiser generator only needs a specification $(\Sigma, \Omega)$ from the user to produce a corresponding synthesiser. 
+In more detail, such a specification can be derived by extending $\Sigma_{\bto}$ with additional type constructs with $\mathsf{nat}$, $\mathsf{prod}$, $\mathsf{sum}$, and $\mathsf{T}$ such that
+\[
+  \arity({\mathsf{nat}}) = 0\qquad
+  \arity({\mathsf{T}}) = 1\qquad
+  \arity({\mathsf{prod}}) = \arity({\mathsf{sum}}) = 2.
+\]
+Types $\mathtt{nat}$, $A \times B$, $A + B$, and $T(A)$ are introduced as $\tyOp_{\mathsf{nat}}$, $\tyOp_{\mathsf{prod}}(A, B)$, $\tyOp_{\mathsf{sum}}(A, B)$, and $\tyOp_{\mathsf{T}}(A)$ respectively.
+The signature $\Omega_{\Lambda}^{\Leftrightarrow}$ is then extended with operations listed in \cref{tab:computational-calculus}.
+Finally, invoking \cref{cor:trichotomy} formally with the specification~$(\Sigma, \Omega)$ gives us a type synthesiser directly.\footnote{%
+See \cref{sec:formalisation} for a demonstration in \Agda.}

### 2023-10-15 tex/ESOP24/S8-future.tex
[ update ] Add the missing .bib
[ tex ] Update Future work

@@
 
-\section{Discussion and Future work} \label{sec:future}
+\section{Discussion} \label{sec:future}
+We believe that our formal treatment lays a foundation for further investigation, as the essential aspects of bidirectional typing have been studied rigorously. 
+While our current development is based on simply typed languages to highlight the core ideas, it is evident that many concepts and extensions remain unexplored.
+We will discuss related work and outline potential directions for future research.
+\subsubsection{Language formalisation frameworks}
+Many general definitions of logics and type theories and frameworks for defining them have been proposed.
 
-\subsection{Related work}
 \citet{Harper1993a,Harper2007}
+\cite{Uemura2021}
+\cite{Bauer2020,Haselwarter2021,Bauer2022a}
+Our general definitions are akin to the definitions used in other language formalisation frameworks~\citep{Ahrens2022,Allais2021,Fiore2022} in a proof assistant---the concept of a signature is at object level.
 
-Our general definitions are akin to the definitions used in other language formalisation frameworks~\citep{Ahrens2022,Allais2021,Fiore2022} in a proof assistant---the concept of a signature is at object level (i.e.\ an inhabitant of a record type in \Agda).
-This allows us to exploit the expressive power of our meta-language while defining a language, so it is possible to express a construct that takes an indeterminate number of arguments such as spine application.
-%\todo[inline,caption={}]{\begin{enumerate}
-%\item `Input' and `output' of our theory, comparing with grammars and parsing
-%\item A more comprehensive theory than Wadler et al., and more formal and refined than Dunfield et al.'
-%\item Restrictions to simple types (not individual language features) and syntax-directedness
-%\end{enumerate}}
+%\subsubsection{Generic bidirectional type synthesis beyond simple types}
+%Clearly, bidirectional type synthesis becomes prominent because it can scale up to deal with more complex types than simple types.
+%As for a general treatment for another class of languages, one has to devise a general definition of languages, which may have been used by existing language-formalisation frameworks.
+%Then, one could enrich the general definition with modes and adopt the definition of mode-correctness accordingly. 
+%Soundness and completeness of our formulation should carry over, since they are only about the separation and combination of mode and typing information for a given raw term (in a syntax-directed formulation).
+%Mode decoration is about annotating a raw term with modes and marking missing annotations, used for guiding the direction during bidirectional type synthesis, so it should also carry over.
+%What makes the general development of bidirectional type synthesis challenging is inherent to the tension between expressiveness and usability for type synthesis and the general treatment for other language facilities.
+%We discusses the cases for dependent types and for polymorphic types below.
+%
+%\paragraph{Dependent types} Type synthesis for a dependent type theory involves term equality check for type equality problem, so a generic decidable bidirectional type synthesis relies on the decidability of the term equality check which in turn requires the ability of specifying computation rules within a language specification.
+%\varcitet{Felicissimo2023}{'s computational LF} is capable of defining dependent type theories with computational rules like \textsc{Dedukti}~\cite{Assaf2016} and generic bidirectional type synthesis.
+%Apart from their more general definition targeting dependent type theories, generic bidirectional type synthesis \cite[Theorem~3.12]{Felicissimo2023} that is decidable, sound, and complete (with respect to decorated terms) is possible under the assumption that a given language specification is mode-correct (or well-typed \cite[Section~3.2]{Felicissimo2023}) and equality check is well-behaved: the specified computation rules are type-preserving, confluent, and strongly normalising.
+%Yet, the last assumption is hard to establish even for a specific language and still lacks a general consideration.
+%
+%\paragraph{Polymorphic types} For \SystemF and other languages where type-level variable binding is allowed, we can start with the notion of polymorphic signature $(\Sigma, \Omega)$ by~\citet{Hamana2011} whereas the arity of each type construct $\sigma$ in $\Sigma$ becomes a binding arity with a single sort $*$ and accordingly a term construct can specify a pair of extension contexts~$\left<\Xi\right>[\Delta]$ for term variables $\Delta$ and type variables $\Xi$.
+%It is not hard to see how general definitions for bidirectional typing rules and mode derivations can be extended from \varcitet[Section~3.4]{Hamana2011}{'s definitions}.
+%For example, the universal type $\forall \alpha.\, A$ and type abstraction in \SystemF can then be specified as operations
+%\[
+%  \mathsf{all}\colon *\rhd [*]* \to *
+%  \quad\text{and}\quad
+%  \mathsf{tabs}\colon [*]A \rhd \left< * \right>A^{\chk} \to
+%  \tyOp_{\mathsf{all}}(\alpha.\, A)^{\chk}
+%\]
+%\Cref{thm:bidirectional-type-synthesis-checking} should also carry over, since no equation is imposed on types and no guess work is needed.
+%However, explicit type application is unusable in practice but implicit application results the \emph{instantiation problem}~\cite[Section~5]{Dunfield2021} which amounts to replacing type equality check with checking the subtyping relation $A <: B$ in rule $\ChkRule{Sub}$ and type synthesis with the guess of $B$ in $\forall \alpha.\, A <: A[B/\alpha]$.
+%A satisfactory theory for bidirectional typing with polymorphism and subtyping should be neutral and thus accommodate various solutions to the instantiation problem, which is still under development.
+%
+\subsubsection{Generic bidirectional type synthesis beyond simple types}
 
-%which together describe an \emph{extrinsically typed} and \emph{syntax-directed} bidirectional simple type system~(\cref{fig:raw-terms,fig:bidirectional-typing-derivations}).
-%
-%There are many ways and extensions for designing bidirectional typing rules, .
-%Yet, once the definition of a bidirectional type system is in place, deriving an algorithm becomes straightforward.
-%Accordingly, our theory's objective is not to formulate rules for various language features, but to introduce a formalism for bidirectional typing.
-%\todo{Looks like we should have axiomatised what types we can deal with; left as future work}
-\paragraph{More characteristics of bidirectional typing}
+Bidirectional type synthesis plays a crucial role in handling more complex types than simple types.
+To provide a formal treatment for a broader class of languages, we need a general definition of languages in question, which may have been used in existing theories or frameworks.
+Then, this general definition can be augmented with modes and the concept of mode-correctness can be introduced.
+Soundness and completeness (\cref{lem:soundness-completeness}) should naturally extend, as they pertain solely to the separation and combination of mode and typing information within a given raw term in a syntax-directed formulation.
+Mode decoration (\cref{sec:mode-decoration}) involves annotating a raw term with modes and marking missing annotations should also work.
+The challenge in developing bidirectional type synthesis on a more general case comes from striking the right balance between expressiveness and usability for type synthesis while considering other language features.
+We discuss cases involving dependent types and polymorphic types below.
 
-We have formalised some of the most important concepts discussed by \citet{Dunfield2021}.
-There are other concepts that may be formalisable with some more effort, for example Pfenning's recipe for bidirectionalising typing rules.
-We believe that our theory lays the foundation for further formalising these concepts.
-There are some other concepts that may be more difficult to pin down, notably `annotation character', which is roughly about how easy it is for the user to write annotated programs.
-These may continue to require creativity from the designer of type systems.
+\paragraph{Dependent Types}
+In the context of dependent type theory, type synthesis entails a term equality check for type equality.
+Achieving generic decidable bidirectional type synthesis depends on the decidability of term equality, which, in turn, relies on the ability to specify computation rules within a language specification.
 
-%\paragraph{Beyond syntax-directed type systems}
-%
-%As a first theory of bidirectional type synthesis, our work focuses on syntax-directed type systems.
-%To explore more general settings, an extreme possibility is to independently specify an ordinary type system and a bidirectional one over the same raw terms and then investigate possible relationships between the two systems, but a more manageable and practical setting suitable for a next step is probably one that only mildly generalises ours:
-%for each raw term construct, there can be several ordinary typing rules, and each typing rule can be refined to several bidirectional typing rules with different mode assignments.
-%Even for this mildly generalised setting, there is already a lot more work to do:
-%Both the mode decorator and the bidirectional type synthesiser will have to backtrack, and become significantly more complex.
-%For soundness and completeness, it should still be possible to treat them as the separation and combination of mode and type information, but the completeness direction will pose a problem---for every node of a raw term, a mode derivation chooses a mode assignment while a typing derivation chooses a typing rule, but there may not be a bidirectional typing rule for this particular combination.
-%There should be ways to fix this problem while retaining the mode decoration phase---for example, one idea is to make the mode decorator produce all possible mode derivations, and refine completeness to say that any typing derivation can be combined with one of these mode derivations into a bidirectional typing derivation.
-%All these considerations are rather too complex for a first theory, however.
-%
-%%\Josh[inline, caption={}]{Getting rid of the assumption of being syntax-directed (the triangular picture) --- 0.5p
-%%\begin{enumerate}
-%%\item In general: two independently formulated type systems over raw terms
-%%\item More sensible next step: a `refinement chain'
-%%\item Backtracking algorithms are required. for both mode decoration and bidirectional type synthesis.
-%%\item The idea of separating and combining mode and type information should still work, but requires more care in the completeness direction.
-%%\item Too complex for a first theory, and left as future work
-%%\end{enumerate}}
-%
-\paragraph{Beyond simple types}
-We leave the problem of presenting bidirectional typing for more general and advanced language features as future work (such as those related to polymorphic types~\citep{Pierce2000,Peyton-Jones2007,Dunfield2013,Xie2018}).
-While algebraic approaches to polymorphic types have been developed~\citep{Fiore2013,Hamana2011}, these approaches do not take subtyping into account.
-Subtyping is essential for formulating important concepts such as \emph{principal types} in type synthesis.
+\varcitet{Felicissimo2023}{'s computation LF} is capable of defining dependent type theories with computational rules, similar to Dedukti, and enabling generic bidirectional type synthesis.
+Generic bidirectional type synthesis, as presented in Felicissimo's work, is decidable, sound, and complete with respect to decorated terms, provided that the underlying language specification is mode-correct, and the equality check is well-behaved, i.e.\ type-preserving, confluent, and strongly normalising.
+However, the last assumption is challenging to establish, even for specific languages and lacks a comprehensive general understanding.
 
-%%\paragraph{Whither extrinsic typing?}
-%
-%%\todo[inline]{We leave the problem of presenting bidirectional typing for more general and advanced language features as future work (such as those related to polymorphic types~\citep{Pierce2000,Peyton-Jones2007,Dunfield2013,Xie2018}), which requires advancement in language formalisation~(\cref{sec:language-formalisation}) that is orthogonal to our work.}
-%
-%%\Josh[inline]{Optional: ornaments (0.25p), NDGP (0.25p)}
-%
-%%\subsubsection{Theories of abstract syntax with variable binding}
-%%\label{sec:theory-of-syntax}
-%%
-%
-%%Substitution is also modelled categorically, but it does not play a role in this paper.%
-%%\todo{Reveal a bit about this paper?}
-%
-%%\begin{remark} \label{re:type-signature}
-%%Most of existing theories treat types independently of terms, thereby excluding them from signatures.
-%%To the best of our knowledge, the only exception to this approach is found in the work of~\citet{Arkor2020}, which incorporates signatures for both terms and types.
-%%Interestingly, this inclusion is also critical for type synthesis for comparing a concrete type $N \bto N$ with an abstract type $A \bto B$, where $A$ and $B$ are type variables in a typing rule.
-%%\end{remark}
+\paragraph{Polymorphic Types}
+In the case of languages like \SystemF and others that permit type-level variable binding, we can start with the notion of a polymorphic signature, as introduced by Hamana---each type construct in a signature is specified by a binding arity with a single sort $*$ and a term construct can employ a pair of extension contexts for term variables and type variables.
+
+Extending general definitions for bidirectional typing and mode derivations from \citeauthor{Hamana2011}'s work is straightforward. 
+For example, the universal type $\forall \alpha.\, A$ and type abstraction in \SystemF can be specified as operations:
+\[
+\mathsf{all} : * \rhd [*] * \to *
+\quad\text{and}\quad
+\mathsf{tabs} : [*] A \rhd \left< * \right> A^{\chk} \to \tyOp_{\mathsf{all}}(\alpha. A)^{\chk}
+\]
+The decidability of bidirectional type synthesis checking (\cref{thm:bidirectional-type-synthesis-checking}) should also carry over, as no equations are imposed on types and no guessing is required.
+
+However, explicit type application is impractical but implicit application results in the \emph{instantiation problem} which amounts to replacing type equality checks $A = B$ with checking subtyping $A <: B$ and guessing $B$ in $\forall \alpha.
+A <: A[B/\alpha]$ in the presence of universal quantification.
+Developing a general theory for bidirectional typing with polymorphism and subtyping that accommodates various solutions to the instantiation problem is still under development.
+
+\subsubsection{More characteristics of bidirectional typing}
+
+\subsubsection{Conclusion}
+There are still many concepts that are possible to articulate such as, for example, Pfenning's recipe for bidirectionalising typing rules.
+In contrast, there are concepts that may be hard to pin down, notably `annotation character'~\cite{Dunfield2021}, which is roughly about how easy it is for the user to write annotated programs.
+These considerations may continue to require creativity from the designer of type systems whereas its implementation could be automated entirely.

### 2023-10-15 tex/ESOP24/S1-intro.tex
[ tex ] Revised S1 & S4

--- 2023-10-13 tex/ESOP24/S1-intro.tex
@@
 In light of these limitations, \emph{bidirectional type synthesis} emerged as a viable alternative, providing algorithms for deciding the types of raw terms that meet some syntactic criteria and usually contain type annotations.
-\varcitet{Dunfield2021}{ in their survey paper} summarised the design principles of bidirectional type synthesis and its wide coverage of languages with simple types, polymorphic types, dependent types, gradual types, among others.
+\varcitet{Dunfield2021}{ summarised in their survey paper} the design principles of bidirectional type synthesis and its wide coverage of languages with simple types, polymorphic types, dependent types, gradual types, among others.
 
@@
 
-Despite sharing the same basic idea, bidirectional typing has been mostly developed on a case-by-case basis; this situation is in stark contrast to parsing, which has general theories together with widely applicable techniques and practical tools, notably parser generators.
+Despite sharing the same basic idea, bidirectional typing has been mostly developed on a case-by-case basis.
+%this situation is in stark contrast to parsing, for which there are general theories and practical tools, notably parser generators.
 %\footnote{The same could be said for type checking in general, not just for bidirectional type systems.
 %While there are type checker generators grounded in unification~\citep{Gast2004,Grewe2015}, it should be noted that unification-based approaches are not suited to more complex type systems.}
-While it is straightforward to derive a bidirectional type synthesis algorithm, \citeauthor{Dunfield2021} only present informal design principles that the community learned from individual systems, rather than a general treatment providing logical specifications and rigorously proven properties for a class of systems.
-Moreover, unlike the plethora of available parser generators, `type-synthesiser generators' rarely exist, so each type synthesiser has to be independently built, not to mention their correctness.
+%While it is usually straightforward to derive a bidirectional type synthesis algorithm,
+\citeauthor{Dunfield2021} present informal design principles learned from individual bidirectional type systems, but in addition to crafting special techniques for individual systems, we should also start to consolidate concepts common to a class of bidirectional type systems into a general and formal theory that gives mathematically precise definitions and proves theorems for the class of systems once and for all.
+%Moreover, by mechanising the theory, we get \emph{verified} type-synthesiser generators analogous to parser generators.
+In this paper, we develop such a theory of bidirectional typing with the proof assistant \Agda.
+%and provide a \emph{verified} generator of \emph{proof-relevant} type synthesisers taking a language specification and a raw term as input.
+%In particular, `type-synthesiser generators' rarely exist, so each type synthesiser has to be independently built, not to mention their correctness.
 
-To mitigate the above problem, we develop with the proof assistant \Agda a general and formal treatment of bidirectional typing and provide a \emph{verified} generator of \emph{proof-relevant} type synthesisers taking a language specification and a raw term as input.
-
-\subsubsection{Proof-relevance in type synthesis}
+\subsubsection{Proof-relevant type synthesis}
 \label{sec:PLFA}
@@
 Recall that the law of excluded middle $P + \neg P$ does not hold as an axiom for every~$P$ constructively, and we say that $P$~is logically \emph{decidable} if the law holds for~$P$.
-Since Martin-L\"of type theory is simultaneously logical and computational, a decidability proof is also a proof-relevant decision procedure that computes not only a yes-or-no answer but also a proof of~$P$ or its refutation, so logical decidability implies algorithmic decidability.
+Since Martin-L\"of type theory is simultaneously logical and computational, a decidability proof is also a proof-relevant decision procedure that computes not only a yes-or-no answer but also a proof of~$P$ or its refutation, so logical decidability is also algorithmic decidability.
 %For example, a proof of \emph{decidable equality}, i.e.~$\forall x, y.~(x = y) + (x \neq y)$, decides whether $x$~and~$y$ are equal and accordingly gives an identity proof or a refutation explicitly.
@@
 A proof of this statement would also be a program that produces either a typing derivation for the given raw term~$\isTerm{t}$ or a negation proof that such a derivation is impossible.
-The first case is algorithmic soundness, while the second case is algorithmic completeness in contrapositive form (and can be strengthened to the original form).
-Hence, proving the statement is the same as constructing a verified proof-relevant type inferencer, which returns not only an answer but also a proof justifying that answer.
+The first case is algorithmic soundness, while the second case is algorithmic completeness in contrapositive form (which implies the original form due to the decidability).
+Hence, proving the statement is the same as constructing a verified proof-relevant type inference algorithm, which returns not only an answer but also a proof justifying that answer.
+This is an economic way to bridge the gap between theory and practice, where proofs double as verified programs, in contrast to separately exhibiting a theory and an implementation that are only loosely related.
 
@@
 
-In a bit more detail:
+To give a bit more detail:
 The type synthesis problem is not just about deciding whether a raw term is typable---there is a third possibility that the term does not have sufficient annotations.
@@
 Our theory gives a (proof-relevant) \emph{mode decorator}, which constructs a mode derivation for a raw term or provide information that refutes the existence of any mode derivation and helps the user to pinpoint missing annotations.
-Then a bidirectional type synthesiser decides the typability of only mode-decorated raw terms instead of all raw terms.
-Soundness and completeness of bidirectional typing is reformulated as a one-to-one correspondence between bidirectional typing derivations and pairs of a typing derivation and a mode derivation for the same raw term; this reformulation is more natural and useful than annotatability described by \citet[Section~3.2]{Dunfield2021}.
+Then a bidirectional type synthesiser is required to decide the typability of only mode-decorated raw terms instead of all raw terms.
+Soundness and completeness of bidirectional typing is reformulated as a one-to-one correspondence between bidirectional typing derivations and pairs of a typing derivation and a mode derivation for the same raw term; this reformulation is more natural and useful than \varcitet[Section~3.2]{Dunfield2021}{'s annotatability}.
 
@@
 The most essential characteristics of bidirectional typing is \emph{mode-correctness}, since an algorithm can often be `read off' from the definition of a bidirectionally typed language if mode-correct.
-It seems that the implications of mode-correctness have only been addressed informally as exhibited in the survey paper~\cite{Dunfield2021}, since mode-correctness is not formally defined as a \emph{property of languages}.
+It seems that the implications of mode-correctness have only been addressed informally by \citet{Dunfield2021}, since mode-correctness is not formally defined as a \emph{property of languages}.
 %But, what is a language anyway?
@@
 A general definition of typed languages allows us to define mode-correctness and to investigate its consequences rigorously, including the uniqueness of synthesised types and the decidability of bidirectional type synthesis for all mode-correct signatures.
-The latter theorem effectively amounts to a generator of proof-relevant bidirectional type synthesiser for all syntax-directed bidirectional simple type systems that are mode-correct.
+The proof of the latter theorem amounts to a generator of proof-relevant bidirectional type synthesisers (analogous to a parser generator working for unambiguous or disambiguated grammars).
 
-To make our exposition accessible we have chosen to work with simply typed languages so that bidirectional type synthesis is established without any other technical assumptions.
+To make our exposition accessible, the theory in this paper focuses on simply typed languages with a syntax-directed bidirectional type system, so that the decidability of bidirectional type synthesis can be established without any other technical assumptions.
 %the idea of extending a signature with bidirectionality should be clear enough for further generalisation.
-Nevertheless, we briefly discuss a possible extension which includes polymorphically typed languages such as \SystemF with additional assumptions in \cref{sec:future}.
+By adding assumptions, it should be possible to extend the theory to deal with more expressive type systems; for instance, we briefly discuss how the theory can handle polymorphically typed languages such as \SystemF with additional assumptions in \cref{sec:future}.
 
@@
   \item general definitions for bidirectional type systems and mode-correctness;
-  \item mode derivations for explicitly dealing with annotations in the theory (in particular leading to a more natural and useful version of completeness than annotatability~\cite{Dunfield2021}), and mode decoration for helping the user to work with annotations in practice;
+  \item mode derivations for explicitly dealing with annotations in the theory, and mode decoration for helping the user to work with annotations in practice;
   \item rigorously proven consequences of mode-correctness---the uniqueness of synthesised types and the decidability of bidirectional type synthesis, which amounts to
-  \item a fully verified generator for proof-relevant type-synthesisers.
+  \item a fully verified generator of proof-relevant type-synthesisers.
 \end{enumerate}
-Our theory was first developed in \Agda and then translated to the mathematical vernacular, filling the gap between our informal presentation and its verified implementation.
+Our theory is fully formally developed in \Agda but translated to the mathematical vernacular for presentation in this paper.
+The formal theory doubles as a verified implementation, with no gap in between.
 

### 2023-10-15 tex/ESOP24/S4-pre-synthesis.tex
[ tex ] Update Future work
[ tex ] Revised S1 & S4

@@
 %Formulating the notion of mode derivation helps to complete this clean picture.
-Technically, all these can be summarised in one theorem.
+Technically, all these can be summarised in one theorem (proved by routine induction).
 
@@
 We will construct such a \emph{generalised mode decorator}~(\cref{thm:generalised-mode-decoration}), which can be weakened to an ordinary mode decorator~(\cref{thm:mode-decoration}) if the additional explanation is not needed.%
-\footnote{To simplify the presentation, we use ordinary mode decoration elsewhere in this paper.}
+\footnote{For the sake of simplicity, we use ordinary mode decoration elsewhere in this paper.}
 
@@
 
-\begin{theorem}[Generalised mode preprocessing]\label{thm:generalised-mode-decoration}
+\begin{theorem}[Generalised mode decoration]\label{thm:generalised-mode-decoration}
 For any raw term $V |-_{\Sigma,\erase\Omega} t$ and mode~$\dir{d}$, there is a derivation of\/ $V |-_{\Sigma,\Omega} t^{\dir{d}\,g\,s}$ for some $g$~and~$s$.
@@
 
-\begin{proof}[of \cref{thm:generalised-mode-decoration}]
-By \cref{thm:adjustment}, it suffices to prove~\eqref{eq:strict-derivation} by induction on~$t$.
-
-\begin{itemize}
-\item $\Rule{Var}$ is mapped to $\SynRule{Var}$.
-\item $\Rule{Anno}$:
-Let $t = (t' \bbcolon A)$.
-By the induction hypothesis and \cref{thm:adjustment}, there is a derivation of $V |-_{\Sigma,\Omega} {t'}^{\chk\,g\,s}$ for some $g$~and~$s$, to which we apply $\SynRule{Anno}$.
-\item $\Rule{Op}$:
-Let $t = \tmOpts$.
-By the induction hypotheses and \cref{thm:adjustment}, there is a derivation of $V, \vec x_i |-_{\Sigma,\Omega} {t_i}^{\dir{d_i}\,g_i\,s_i}$ for each sub-term~$t_i$; apply $\Rule{Op}$ to all the derivations.
-\end{itemize}
-\end{proof}
+%\begin{proof}[of \cref{thm:generalised-mode-decoration}]
+%By \cref{thm:adjustment}, it suffices to prove~\eqref{eq:strict-derivation} by induction on~$t$.
+%
+%\begin{itemize}
+%\item $\Rule{Var}$ is mapped to $\SynRule{Var}$.
+%\item $\Rule{Anno}$:
+%Let $t = (t' \bbcolon A)$.
+%By the induction hypothesis and \cref{thm:adjustment}, there is a derivation of $V |-_{\Sigma,\Omega} {t'}^{\chk\,g\,s}$ for some $g$~and~$s$, to which we apply $\SynRule{Anno}$.
+%\item $\Rule{Op}$:
+%Let $t = \tmOpts$.
+%By the induction hypotheses and \cref{thm:adjustment}, there is a derivation of $V, \vec x_i |-_{\Sigma,\Omega} {t_i}^{\dir{d_i}\,g_i\,s_i}$ for each sub-term~$t_i$; apply $\Rule{Op}$ to all the derivations.
+%\end{itemize}
+%\end{proof}
 
@@
 
-\begin{corollary}[Mode preprocessing]\label{thm:mode-decoration}
+\begin{corollary}[Mode decoration]\label{thm:mode-decoration}
   It is decidable whether $V |-_{\Sigma,\Omega} \isTerm{t}^\dir{d}$.%for any raw term $V |-_{\Sigma,\erase\Omega} t$ and mode~$\dir{d}$
@@
 
-\begin{figure}
+\citet[Section~3.2]{Dunfield2021} formulated completeness differently from our \cref{lem:soundness-completeness} and proposed \emph{annotatability} as a more suitable name.
+In our theory, we may formulate annotatability as follows.
+
+\begin{proposition}[Annotatability]\label{thm:annotatability}
+If\/ $\Gamma |-_{\Sigma,\erase\Omega} t : A$, then there exists~$t'$ such that\/ $t' \sqsupseteq t$ and $\Gamma |-_{\Sigma,\Omega} t' :^\dir{d} A$ for some~$\dir{d}$.
+\end{proposition}
+
+Defined in \cref{fig:annotation-ordering}, the `annotation ordering' $t' \sqsupseteq t$ means that $t'$~has the same or more annotations than~$t$.
+In a sense, annotatability is a reasonable form of completeness: if a term of a simply typed language $(\Sigma, \erase\Omega)$ is typable in the ordinary type system, it may not be directly typable in the bidirectional type system $(\Sigma, \Omega)$ due to some missing annotations, but will be if those annotations are added correctly.
+In our theory, \cref{thm:annotatability} can be straightforwardly proved by induction on the derivation produced by generalised mode decoration~(\cref{thm:generalised-mode-decoration}) to construct a bidirectional typing derivation in the same mode.
+The interesting case is $\SynRule{Missing}$, which is mapped to $\SynRule{Anno}$, adding to the term a type annotation that comes from the given typing derivation.
+
+\begin{figure}[h]
   \centering\small
@@
 
-\citet[Section~3.2]{Dunfield2021} formulated completeness differently from our \cref{lem:soundness-completeness} and proposed \emph{annotatability} as a more suitable name.
-In our theory, we may formulate annotatability as follows.
-
-\begin{proposition}[Annotatability]\label{thm:annotatability}
-If\/ $\Gamma |-_{\Sigma,\erase\Omega} t : A$, then there exists~$t'$ such that\/ $t' \sqsupseteq t$ and $\Gamma |-_{\Sigma,\Omega} t' :^\dir{d} A$ for some~$\dir{d}$.
-\end{proposition}
-
-Defined in \cref{fig:annotation-ordering}, the `annotation ordering' $t' \sqsupseteq t$ means that $t'$~has the same or more annotations than~$t$.
-In a sense, annotatability is a reasonable form of completeness: if a term of a simply typed language $(\Sigma, \erase\Omega)$ is typable in the ordinary type system, it may not be directly typable in the bidirectional type system $(\Sigma, \Omega)$ due to some missing annotations, but will be if those annotations are added correctly.
-In our theory, \cref{thm:annotatability} can be easily proved with generalised mode decoration.
-
-\begin{proof}[of \cref{thm:annotatability}]
-By \cref{thm:generalised-mode-decoration}, there is a generalised mode derivation of $V |-_{\Sigma,\erase\Omega} t^{\dir{d}\,g\,s}$ for some mode~$\dir{d}$ and booleans $g$~and~$s$.
-Perform induction on this generalised mode derivation to construct a bidirectional typing derivation in the same mode (and also the term~$t'$ and a proof of $t' \sqsupseteq t$, which are determined by the derivations and omitted here).
-\begin{itemize}
-\item $\SynRule{Var}$, $\SynRule{Anno}$, and $\Rule{Op}$:
-The outermost rule of the given typing derivation must be the corresponding one; apply the corresponding bidirectional typing rule to the induction hypotheses.
-\item $\ChkRule{Sub}$:
-Apply $\ChkRule{Sub}$ to the induction hypothesis.
-\item $\SynRule{Missing}$:
-This is the interesting case where we map a $\SynRule{Missing}$ rule to an $\SynRule{Anno}$ rule and add to the term a type annotation, which comes from the given typing derivation.
-\end{itemize}
-\end{proof}
-
 On the other hand, when using a bidirectional type synthesiser to implement type synthesis~(\cref{thm:implementation}), our completeness is much simpler to use than annotatability, which requires the bidirectional type synthesiser to produce more complex evidence when the synthesis fails.

### 2023-10-15 tex/ESOP24/S8-future.tex
[ tex ] Revised S1 & S4

@@
 A <: A[B/\alpha]$ in the presence of universal quantification.
-Developing a general theory for bidirectional typing with polymorphism and subtyping that accommodates various solutions to the instantiation problem is still under development.
+A general theory for bidirectional typing with polymorphism and subtyping that accommodates various solutions to the instantiation problem is still under development.
 

### 2023-10-16 tex/ESOP24/S1-intro.tex
[ tex ] Update future work and accordingly S1

@@
 Then a bidirectional type synthesiser is required to decide the typability of only mode-decorated raw terms instead of all raw terms.
-Soundness and completeness of bidirectional typing is reformulated as a one-to-one correspondence between bidirectional typing derivations and pairs of a typing derivation and a mode derivation for the same raw term; this reformulation is more natural and useful than \varcitet[Section~3.2]{Dunfield2021}{'s annotatability}.
+Soundness and completeness of bidirectional typing is reformulated as a one-to-one correspondence between bidirectional typing derivations and pairs of a typing derivation and a mode derivation for the same raw term; this reformulation is more natural and useful than annotatability~\cite[Section~3.2]{Dunfield2021} which is typically formulated in the literature of bidirectional typing.
 
@@
 %the idea of extending a signature with bidirectionality should be clear enough for further generalisation.
-By adding assumptions, it should be possible to extend the theory to deal with more expressive type systems; for instance, we briefly discuss how the theory can handle polymorphically typed languages such as \SystemF with additional assumptions in \cref{sec:future}.
+By adding assumptions, it should be possible to extend the theory to deal with more expressive type systems; for instance, we briefly discuss how the theory can handle polymorphically typed languages such as \SystemF and \SystemFsub with additional assumptions in \cref{sec:future}.
 

### 2023-10-16 tex/ESOP24/S8-future.tex
[ tex ] Update future work and accordingly S1

@@
 While our current development is based on simply typed languages to highlight the core ideas, it is evident that many concepts and extensions remain unexplored.
-We will discuss related work and outline potential directions for future research.
+In this section we discuss related work and outline potential directions for future research.
 \subsubsection{Language formalisation frameworks}
@@
 
-\varcitet{Felicissimo2023}{'s computation LF} is capable of defining dependent type theories with computational rules, similar to Dedukti, and enabling generic bidirectional type synthesis.
+\varcitet{Felicissimo2023}{'s computation LF} is capable of defining dependent type theories with computational rules, similar to \Dedukti, and enabling generic bidirectional type synthesis.
 Generic bidirectional type synthesis, as presented in Felicissimo's work, is decidable, sound, and complete with respect to decorated terms, provided that the underlying language specification is mode-correct, and the equality check is well-behaved, i.e.\ type-preserving, confluent, and strongly normalising.
@@
 \]
-The decidability of bidirectional type synthesis checking (\cref{thm:bidirectional-type-synthesis-checking}) should also carry over, as no equations are imposed on types and no guessing is required.
+The decidability of bidirectional type synthesis (\cref{thm:bidirectional-type-synthesis-checking}) should also carry over, as no equations are imposed on types and no guessing (for type application) is required.
+Adding subtyping $A \mathrel{<:} B$ to languages can be done by replacing type equality with a subtyping relation $\mathrel{<:}$ and type equality check with subtyping check, so polymorphically typed languages with subtyping such as \SystemFsub can be specified.
+The main idea of bidirectional typing does not change, so the formal development should extend without further assumptions as well.
 
-However, explicit type application is impractical but implicit application results in the \emph{instantiation problem} which amounts to replacing type equality checks $A = B$ with checking subtyping $A <: B$ and guessing $B$ in $\forall \alpha.
-A <: A[B/\alpha]$ in the presence of universal quantification.
-A general theory for bidirectional typing with polymorphism and subtyping that accommodates various solutions to the instantiation problem is still under development.
+However, explicit type application is impractical but its implicit counterpart results in the \emph{instantiation problem} which amounts to transforming the implicit type application (a stationary rule) to guessing $B$ in $\forall \alpha.
+A <: A[B/\alpha]$.
+A general theory for bidirectional typing that accommodates various solutions to the instantiation problem as an assumption is still under development.
 

### 2023-10-16 tex/ESOP24/BiSig.tex
[ tex ] Update Future work
[ tex ] Update after sync in person

@@
 There has been much progress in designing bidirectional type systems and associated type synthesis algorithms, but mainly on a case-by-case basis.
-This situation is in stark contrast to parsing, for which there have been general theories and widely applicable tools such as parser generators.
-To remedy the situation, this paper develops a \emph{general} and \emph{formal} theory of bidirectional typing, and provides a \emph{verified} generator of \emph{proof-relevant} type synthesisers for simply typed languages: For every signature that specifies a mode-correct bidirectionally typed language, there exists a proof-relevant type synthesiser that for an input abstract syntax tree constructs a typing derivation if any, gives its refutation if not, or reports that the input does not have enough type annotations.
-Soundness, completeness, and mode-correctness are studied universally for any simply typed language, which are sufficient conditions for deriving a type synthesiser.
+%This situation is in stark contrast to parsing, for which there have been general theories and widely applicable tools such as parser generators.
+To remedy the situation, this paper develops a \emph{general} and \emph{formal} theory of bidirectional typing and as a by-product of our formalism provides a \emph{verified} generator of \emph{proof-relevant} type synthesisers for simply typed languages: For every signature that specifies a mode-correct bidirectionally typed language, there exists a proof-relevant type synthesiser that for an input abstract syntax tree constructs a typing derivation if any, gives its refutation if not, or reports that the input does not have enough type annotations.
+Soundness, completeness, and mode-correctness are studied universally for al signatures, which are sufficient conditions for deriving a type synthesiser.
 We propose a preprocessing step called \emph{mode decoration}, which helps the user to deal with missing type annotations in a given abstract syntax tree.
@@
 \input{S6-example}
-\subfile{S7-formalisation}
 \input{S8-future}
@@
 \appendix
-\input{appendix}
+
+%\input{appendix}
+\subfile{S7-formalisation}
 

### 2023-10-16 tex/ESOP24/S1-intro.tex
[ tex ] Update after sync in person

@@
 The proof-relevant formulation deviates from the usual one:
-traditionally, a type synthesis algorithm is presented as \emph{algorithmic rules}, for example in the form $\Gamma |- \isTerm{t} \syn A \mapsto \isTerm{t}'$, which denotes that annotations can be added to~$\isTerm{t}$ in the surface language%
-\Josh{The algorithm adds annotations to the input term?}
-to produce $\isTerm{t}'$ of type~$A$ in the core language.
+traditionally, a type synthesis algorithm is presented as \emph{algorithmic rules}, for example in the form $\Gamma |- \isTerm{t} \syn A \mapsto \isTerm{t}'$, which denotes that $\isTerm{t}$ in the surface language
+%\Josh{The algorithm adds annotations to the input term?}
+can be transformed to a well typed term $\isTerm{t}'$ of type~$A$ in the core language~\cite{Pierce2000}.
 Such an algorithm is then accompanied by soundness and completeness assertions that the algorithm correctly synthesises the type of a raw term and every typable term can be synthesised.
 By contrast, the proof-relevant approach exploits the simultaneously computational and logical nature of Martin-L\"of type theory and formulates \emph{algorithmic soundness, completeness, and decidability in one go}.
-We explain this in a bit more detail below.
 %(although they do not emphasise the difference of their approach from the traditional one)
 
+In more detail:
 Recall that the law of excluded middle $P + \neg P$ does not hold as an axiom for every~$P$ constructively, and we say that $P$~is logically \emph{decidable} if the law holds for~$P$.
@@
 The most essential characteristics of bidirectional typing is \emph{mode-correctness}, since an algorithm can often be `read off' from the definition of a bidirectionally typed language if mode-correct.
-It seems that the implications of mode-correctness have only been addressed informally by \citet{Dunfield2021}, since mode-correctness is not formally defined as a \emph{property of languages}.
-%But, what is a language anyway?
+As illustrated by \citet{Dunfield2021}, it seems that the implications of mode-correctness have only been addressed informally so far and mode-correctness is not yet formally defined as a \emph{property of languages}.
 
-To make the notion of mode-correctness precise, we first give a general definition of bidirectional simple type systems, called \emph{bidirectional binding signature}, extending the sorted version of \varcitet{Aczel1978}{'s binding signature} with bidirectionality.
+In order to make the notion of mode-correctness precise, we first give a general definition of bidirectional simple type systems, called \emph{bidirectional binding signature}, extending the sorted version of \varcitet{Aczel1978}{'s binding signature} with modes.
 A general definition of typed languages allows us to define mode-correctness and to investigate its consequences rigorously, including the uniqueness of synthesised types and the decidability of bidirectional type synthesis for all mode-correct signatures.
@@
 %the idea of extending a signature with bidirectionality should be clear enough for further generalisation.
-By adding assumptions, it should be possible to extend the theory to deal with more expressive type systems; for instance, we briefly discuss how the theory can handle polymorphically typed languages such as \SystemF and \SystemFsub with additional assumptions in \cref{sec:future}.
+It should be possible to extend the theory to deal with more expressive types and assumptions other than mode-correctness.
+For instance, we briefly discuss how the theory can handle polymorphically typed languages such as \SystemF, \SystemFsub, and those systems using implicit type applications with additional assumptions in \cref{sec:future}.
 
@@
 \end{enumerate}
-Our theory is fully formally developed in \Agda but translated to the mathematical vernacular for presentation in this paper.
+Our theory is fully formally developed with \Agda but translated to the mathematical vernacular for presentation in this paper.
 The formal theory doubles as a verified implementation, with no gap in between.
@@
 Some examples other than simply typed $\lambda$-calculus are given in \Cref{sec:example}.
-We briefly demonstrate the use of our \Agda formalisation as programs in \cref{sec:formalisation} and conclude in \cref{sec:future} with further developments.
+We conclude in \cref{sec:future} with further developments and a demonstration in \cref{sec:formalisation} for the use of our \Agda formalisation as programs.

### 2023-10-16 tex/ESOP24/S8-future.tex
[ tex ] Update after sync in person

@@
 We believe that our formal treatment lays a foundation for further investigation, as the essential aspects of bidirectional typing have been studied rigorously. 
-While our current development is based on simply typed languages to highlight the core ideas, it is evident that many concepts and extensions remain unexplored.
-In this section we discuss related work and outline potential directions for future research.
+While our current development is based on simply typed languages to highlight the core ideas, it is evident that many concepts and aspects remain unexplored.
+In this section we discuss related work and potential directions for future research.
 \subsubsection{Language formalisation frameworks}
@@
 
-%\subsubsection{Generic bidirectional type synthesis beyond simple types}
-%Clearly, bidirectional type synthesis becomes prominent because it can scale up to deal with more complex types than simple types.
-%As for a general treatment for another class of languages, one has to devise a general definition of languages, which may have been used by existing language-formalisation frameworks.
-%Then, one could enrich the general definition with modes and adopt the definition of mode-correctness accordingly. 
-%Soundness and completeness of our formulation should carry over, since they are only about the separation and combination of mode and typing information for a given raw term (in a syntax-directed formulation).
-%Mode decoration is about annotating a raw term with modes and marking missing annotations, used for guiding the direction during bidirectional type synthesis, so it should also carry over.
-%What makes the general development of bidirectional type synthesis challenging is inherent to the tension between expressiveness and usability for type synthesis and the general treatment for other language facilities.
-%We discusses the cases for dependent types and for polymorphic types below.
-%
-%\paragraph{Dependent types} Type synthesis for a dependent type theory involves term equality check for type equality problem, so a generic decidable bidirectional type synthesis relies on the decidability of the term equality check which in turn requires the ability of specifying computation rules within a language specification.
-%\varcitet{Felicissimo2023}{'s computational LF} is capable of defining dependent type theories with computational rules like \textsc{Dedukti}~\cite{Assaf2016} and generic bidirectional type synthesis.
-%Apart from their more general definition targeting dependent type theories, generic bidirectional type synthesis \cite[Theorem~3.12]{Felicissimo2023} that is decidable, sound, and complete (with respect to decorated terms) is possible under the assumption that a given language specification is mode-correct (or well-typed \cite[Section~3.2]{Felicissimo2023}) and equality check is well-behaved: the specified computation rules are type-preserving, confluent, and strongly normalising.
-%Yet, the last assumption is hard to establish even for a specific language and still lacks a general consideration.
-%
-%\paragraph{Polymorphic types} For \SystemF and other languages where type-level variable binding is allowed, we can start with the notion of polymorphic signature $(\Sigma, \Omega)$ by~\citet{Hamana2011} whereas the arity of each type construct $\sigma$ in $\Sigma$ becomes a binding arity with a single sort $*$ and accordingly a term construct can specify a pair of extension contexts~$\left<\Xi\right>[\Delta]$ for term variables $\Delta$ and type variables $\Xi$.
-%It is not hard to see how general definitions for bidirectional typing rules and mode derivations can be extended from \varcitet[Section~3.4]{Hamana2011}{'s definitions}.
-%For example, the universal type $\forall \alpha.\, A$ and type abstraction in \SystemF can then be specified as operations
-%\[
-%  \mathsf{all}\colon *\rhd [*]* \to *
-%  \quad\text{and}\quad
-%  \mathsf{tabs}\colon [*]A \rhd \left< * \right>A^{\chk} \to
-%  \tyOp_{\mathsf{all}}(\alpha.\, A)^{\chk}
-%\]
-%\Cref{thm:bidirectional-type-synthesis-checking} should also carry over, since no equation is imposed on types and no guess work is needed.
-%However, explicit type application is unusable in practice but implicit application results the \emph{instantiation problem}~\cite[Section~5]{Dunfield2021} which amounts to replacing type equality check with checking the subtyping relation $A <: B$ in rule $\ChkRule{Sub}$ and type synthesis with the guess of $B$ in $\forall \alpha.\, A <: A[B/\alpha]$.
-%A satisfactory theory for bidirectional typing with polymorphism and subtyping should be neutral and thus accommodate various solutions to the instantiation problem, which is still under development.
-%
-\subsubsection{Generic bidirectional type synthesis beyond simple types}
+\subsubsection{General theory of bidirectional type synthesis beyond simple types}
 
-Bidirectional type synthesis plays a crucial role in handling more complex types than simple types.
-To provide a formal treatment for a broader class of languages, we need a general definition of languages in question, which may have been used in existing theories or frameworks.
-Then, this general definition can be augmented with modes and the concept of mode-correctness can be introduced.
-Soundness and completeness (\cref{lem:soundness-completeness}) should naturally extend, as they pertain solely to the separation and combination of mode and typing information within a given raw term in a syntax-directed formulation.
-Mode decoration (\cref{sec:mode-decoration}) involves annotating a raw term with modes and marking missing annotations should also work.
-The challenge in developing bidirectional type synthesis on a more general case comes from striking the right balance between expressiveness and usability for type synthesis while considering other language features.
-We discuss cases involving dependent types and polymorphic types below.
+Bidirectional type synthesis plays a crucial role in handling more complex types than simple types, and we sketch how our theory can be extended to treat a broader class of languages.
+We need a general definition of languages in question (\cref{subsec:simple-types,subsec:binding-sig}).
+Then, this general definition can be augmented with modes (\cref{subsec:bidirectional-system}) and the formal definition of mode-correctness can be adapted (\cref{def:mode-correctness}).
+Soundness and completeness (\cref{lem:soundness-completeness}) should still hold, as they amount to the separation and combination of mode and typing information for a given raw term (in a syntax-directed formulation).
+Mode decoration (\cref{sec:mode-decoration}) involves annotating a raw term with modes and marking missing annotations, and should also work.
+As for the decidability of bidirectional type synthesis, we discuss two cases involving polymorphic types and dependent types below.
 
-\paragraph{Dependent Types}
-In the context of dependent type theory, type synthesis entails a term equality check for type equality.
-Achieving generic decidable bidirectional type synthesis depends on the decidability of term equality, which, in turn, relies on the ability to specify computation rules within a language specification.
-
-\varcitet{Felicissimo2023}{'s computation LF} is capable of defining dependent type theories with computational rules, similar to \Dedukti, and enabling generic bidirectional type synthesis.
-Generic bidirectional type synthesis, as presented in Felicissimo's work, is decidable, sound, and complete with respect to decorated terms, provided that the underlying language specification is mode-correct, and the equality check is well-behaved, i.e.\ type-preserving, confluent, and strongly normalising.
-However, the last assumption is challenging to establish, even for specific languages and lacks a comprehensive general understanding.
-
-\paragraph{Polymorphic Types}
-In the case of languages like \SystemF and others that permit type-level variable binding, we can start with the notion of a polymorphic signature, as introduced by Hamana---each type construct in a signature is specified by a binding arity with a single sort $*$ and a term construct can employ a pair of extension contexts for term variables and type variables.
+\paragraph{Polymorphic types}
+In the case of languages like \SystemF and others that permit type-level variable binding, we can start with the notion of a polymorphic signature, as introduced by \citet{Hamana2011}:
+\begin{inlineenum}
+  \item each type construct in a signature is specified by a binding arity with a single sort $*$, and
+  \item a term construct can employ a pair of extension contexts for term variables and type variables.
+\end{inlineenum}
 
@@
 Adding subtyping $A \mathrel{<:} B$ to languages can be done by replacing type equality with a subtyping relation $\mathrel{<:}$ and type equality check with subtyping check, so polymorphically typed languages with subtyping such as \SystemFsub can be specified.
-The main idea of bidirectional typing does not change, so the formal development should extend without further assumptions as well.
+The main idea of bidirectional typing does not change, so it should be possible to extend the formal theory without further assumptions too.
 
@@
 A <: A[B/\alpha]$.
-A general theory for bidirectional typing that accommodates various solutions to the instantiation problem as an assumption is still under development.
+A general theory for bidirectional typing that accommodates various solutions to the instantiation problem as an assumption is left as future work.
+
+\paragraph{Dependent types}
+In the context of dependent type theory, type synthesis entails a term equality check for type equality.
+Achieving decidable bidirectional type synthesis depends on the decidability of term equality, which, in turn, relies on the ability to specify computation rules within a language specification.
+
+Extending the logical framework \Dedukti, \varcitet{Felicissimo2023}{'s \textsf{CompLF}} is capable of defining dependent type theories with computational rules and enables generic bidirectional type synthesis.
+\LT{compare his work with ours}
+Generic bidirectional type synthesis, as presented in Felicissimo's work, is decidable, sound, and complete with respect to mode-decorated terms, provided that the underlying language specification is mode-correct, and the equality check is \emph{well-behaved}, i.e.\ type-preserving, confluent, and strongly normalising.
+However, the last assumption is challenging to establish even for specific languages and lacks a comprehensive general understanding.
 
@@
 
-\subsubsection{Conclusion}
 There are still many concepts that are possible to articulate such as, for example, Pfenning's recipe for bidirectionalising typing rules.
+
 In contrast, there are concepts that may be hard to pin down, notably `annotation character'~\cite{Dunfield2021}, which is roughly about how easy it is for the user to write annotated programs.
 These considerations may continue to require creativity from the designer of type systems whereas its implementation could be automated entirely.
+
+\begin{table}[ht]
+  \renewcommand{\arraystretch}{2.5}
+  \setlength{\tabcolsep}{3pt}
+  \centering\footnotesize
+\begin{tabular}{c | l}
+  Rules & Operations \\ \hline\hline
+  $\inferrule{ }{\Gamma \vdash \mathtt{z} \chk \mathtt{nat}}$ &  $\aritysymbol{\mathsf{z}}{\cdot}{\cdot}{\mathtt{nat}^{\chk}}$ 
+  \\
+ $\inferrule{\Gamma \vdash t \chk \mathtt{nat}}{\Gamma \vdash \mathtt{s}(t) \chk \mathtt{nat}}$ & $\aritysymbol{\mathsf{s}}{\cdot}{\mathtt{nat}^{\chk}}{\mathtt{nat}^{\chk}}$ \\
+ $\inferrule{\Gamma \vdash t \syn \mathtt{nat} \\\\ \Gamma \vdash t_0 \chk A \and \Gamma, x : \mathtt{nat} \vdash t_1 \chk A}{\Gamma \vdash \mathtt{ifz}(t_0; x.t_1)(t) \chk A}$ & $\aritysymbol{\mathsf{ifz}}{A}{\mathtt{nat}^{\syn}, A^{\chk}, A^{\chk}}{A^{\chk}}$ \\
+ $\inferrule{\Gamma \vdash t \chk A \and \Gamma \vdash u \chk B}{\Gamma \vdash (t, u) \chk A \times B}$ & $\aritysymbol{\mathsf{pair}}{A, B}{A^{\chk}, B^{\chk}}{A \times B^{\chk}}$  \\
+ $\inferrule{\Gamma \vdash t \syn A_1 \times A_2}{\Gamma \vdash \mathtt{proj}_i(t) \syn A_i}$ for $i = 1, 2$ & $\aritysymbol{\mathsf{proj}_i}{A_1, A_2}{A_1 \times A_2^{\syn}}{A_i^{\syn}}$ for $i = 1, 2$ \\
+
+ $\inferrule{\Gamma \vdash t \chk A_i}{\Gamma \vdash \mathtt{inj}_i(t) \chk A_1 + A_2}$ for $i = 1, 2$ & $\aritysymbol{\mathsf{inj}_i}{A_1, A_2}{A_i^{\chk}}{A_1 + A_2^{\chk}}$ for $i = 1, 2$ \\
+
+ $\inferrule{\Gamma \vdash u \syn A + B \\\\ \Gamma, x_1 : A \vdash t_1 \chk C \and \Gamma, x_2 : B \vdash t_2 \chk C}{\Gamma \vdash \mathtt{case}(u; x_1.
+t_1; x_2.
+t_2) \chk C}$ & $\aritysymbol{\mathsf{case}}{A, B, C}{{A + B}^{\syn}, [A]C^{\chk}, [B]C^{\chk}}{C^{\syn}}$ \\
+
+ $\inferrule{\Gamma, x : A \vdash t \chk A}{\Gamma \vdash \mu x.\, t \chk A}$ & $\aritysymbol{\mathsf{mu}}{A}{[A]A^{\chk}}{A^{\chk}}$ \\
+
+ $\inferrule{\Gamma \vdash t \syn A \and \Gamma, x : A \vdash u \chk B}{\Gamma \vdash \mathtt{let}\;x = t\;\mathtt{in}\;u\chk B}$ & 
+  $\aritysymbol{\mathsf{let}}{A, B}{A^{\syn},[A]B^{\chk}}{B^{\chk}}$ \\
+ $\inferrule{\Gamma \vdash t \syn A}{\Gamma \vdash \mathtt{ret}(t) \syn T(A)}$ & $\aritysymbol{\mathsf{ret}}{A}{A^{\syn}}{T(A)^{\syn}}$ \\
+ $\inferrule{\Gamma \vdash t \syn T(A) \and \Gamma, x : A \vdash u \syn T(B)}{\Gamma \vdash \mathtt{bind}(t; x.u) \syn T(B)}$ & $\aritysymbol{\mathsf{bind}}{A, B}{T(A)^{\syn}, [A]T(B)^{\syn}}{T(B)^{\syn}}$
+\end{tabular}
+\caption{A computational calculus with naturals, products, sums, and general recursion}
+\label{tab:computational-calculus}
+\end{table}

### 2023-10-16 tex/ESOP24/S6-example.tex
[ tex ] Update future work and accordingly S1
[ tex ] Update after sync in person
[ tex ] Revise Discussion and Examples

@@
 \subsection{Computational calculi}\label{subsec:PCF}
-Implementing a standalone type synthesizer for a simply typed language is typically a straightforward task.
+Implementing a stand-alone type synthesiser for a simply typed language is typically a straightforward task.
 However, the code size increases proportionally to the number of type constructs and of arguments associated with each term construct.
-When dealing with a fixed number of $n$ type constructs, for each synthesising construct, there are two cases for a checking argument but $n + 1$ possible cases for each synthesising argument---the successful synthesis of the expected type, instances where it fails, or $n-1$ cases where the expected type does not match; similarly for a checking construct---making the task tedious and error-prone.
-Therefore, having a generator can be helpful and significantly reduce the effort for implementation.
+When dealing with a fixed number $n$ of type constructs, for each synthesising construct, there are two cases for a checking argument but $n + 1$ cases for each synthesising argument---the successful synthesis of the expected type, an instance where it fails, or $n-1$ cases where the expected type does not match; similarly for a checking construct---making the task tedious.
+Thus, having a generator is helpful and can significantly reduce the effort for implementation.
 
 For illustrative purposes, consider a computational calculus~\cite{Moggi1989} with additional constructs listed in \cref{tab:computational-calculus}.
-The extended language has `only' 15 constructs, far fewer than a realistic programming language would have, but there are nearly 100 possible cases already to consider during bidirectional type synthesis.
-\begin{table}[h!]
-  \renewcommand{\arraystretch}{2.5}
-  \setlength{\tabcolsep}{3pt}
-  \centering\footnotesize
-\begin{tabular}{c | l}
-  Rules & Operations \\ \hline\hline
-  $\inferrule{ }{\Gamma \vdash \mathtt{z} \chk \mathtt{nat}}$ &  $\aritysymbol{\mathsf{z}}{\cdot}{\cdot}{\mathtt{nat}^{\chk}}$ 
-  \\
- $\inferrule{\Gamma \vdash t \chk \mathtt{nat}}{\Gamma \vdash \mathtt{s}(t) \chk \mathtt{nat}}$ & $\aritysymbol{\mathsf{s}}{\cdot}{\mathtt{nat}^{\chk}}{\mathtt{nat}^{\chk}}$ \\
- $\inferrule{\Gamma \vdash t \syn \mathtt{nat} \and \Gamma \vdash t_0 \chk A \and \Gamma, x : \mathtt{nat} \vdash t_1 \chk A}{\Gamma \vdash \mathtt{ifz}(t_0; x.t_1)(t) \chk A}$ & $\aritysymbol{\mathsf{ifz}}{A}{\mathtt{nat}^{\syn}, A^{\chk}, A^{\chk}}{A^{\chk}}$ \\
- $\inferrule{\Gamma \vdash t \chk A \and \Gamma \vdash u \chk B}{\Gamma \vdash (t, u) \chk A \times B}$ & $\aritysymbol{\mathsf{pair}}{A, B}{A^{\chk}, B^{\chk}}{A \times B^{\chk}}$  \\
- $\inferrule{\Gamma \vdash t \syn A_1 \times A_2}{\Gamma \vdash \mathtt{proj}_i(t) \syn A_i}$ for $i = 1, 2$ & $\aritysymbol{\mathsf{proj}_i}{A_1, A_2}{A_1 \times A_2^{\syn}}{A_i^{\syn}}$ for $i = 1, 2$ \\
-
- $\inferrule{\Gamma \vdash t \chk A_i}{\Gamma \vdash \mathtt{inj}_i(t) \chk A_1 + A_2}$ for $i = 1, 2$ & $\aritysymbol{\mathsf{inj}_i}{A_1, A_2}{A_i^{\chk}}{A_1 + A_2^{\chk}}$ for $i = 1, 2$ \\
-
- $\inferrule{\Gamma \vdash u \syn A + B \and \Gamma, x_1 : A \vdash t_1 \chk C \and \Gamma, x_2 : B \vdash t_2 \chk C}{\Gamma \vdash \mathtt{case}(u; x_1.
-t_1; x_2.
-t_2) \chk C}$ & $\aritysymbol{\mathsf{case}}{A, B, C}{{A + B}^{\syn}, [A]C^{\chk}, [B]C^{\chk}}{C^{\syn}}$ \\
-
- $\inferrule{\Gamma, x : A \vdash t \chk A}{\Gamma \vdash \mu x.\, t \chk A}$ & $\aritysymbol{\mathsf{mu}}{A}{[A]A^{\chk}}{A^{\chk}}$ \\
-
- $\inferrule{\Gamma \vdash t \syn A \and \Gamma, x : A \vdash u \chk B}{\Gamma \vdash \mathtt{let}\;x = t\;\mathtt{in}\;u\chk B}$ & 
-  $\aritysymbol{\mathsf{let}}{A, B}{A^{\syn},[A]B^{\chk}}{B^{\chk}}$ \\
- $\inferrule{\Gamma \vdash t \syn A}{\Gamma \vdash \mathtt{ret}(t) \syn T(A)}$ & $\aritysymbol{\mathsf{ret}}{A}{A^{\syn}}{T(A)^{\syn}}$ \\
- $\inferrule{\Gamma \vdash t \syn T(A) \and \Gamma, x : A \vdash u \syn T(B)}{\Gamma \vdash \mathtt{bind}(t; x.u) \syn T(B)}$ & $\aritysymbol{\mathsf{bind}}{A, B}{T(A)^{\syn}, [A]T(B)^{\syn}}{T(B)^{\syn}}$
-\end{tabular}
-\caption{A computational calculus with naturals, products, sums, and general recursion}
-\label{tab:computational-calculus}
-\end{table}
+The extended language has `only' 15 constructs, far fewer than a realistic programming language would have, but there are nearly 100 possible cases to consider in bidirectional type synthesis.
 
@@
 \]
-Types $\mathtt{nat}$, $A \times B$, $A + B$, and $T(A)$ are introduced as $\tyOp_{\mathsf{nat}}$, $\tyOp_{\mathsf{prod}}(A, B)$, $\tyOp_{\mathsf{sum}}(A, B)$, and $\tyOp_{\mathsf{T}}(A)$ respectively.
+Types $\mathtt{nat}$, $A \times B$, $A + B$, and $T(A)$ are given as $\tyOp_{\mathsf{nat}}$, $\tyOp_{\mathsf{prod}}(A, B)$, $\tyOp_{\mathsf{sum}}(A, B)$, and $\tyOp_{\mathsf{T}}(A)$ respectively.
 The signature $\Omega_{\Lambda}^{\Leftrightarrow}$ is then extended with operations listed in \cref{tab:computational-calculus}.
-Finally, invoking \cref{cor:trichotomy} formally with the specification~$(\Sigma, \Omega)$ gives us a type synthesiser directly.\footnote{%
-See \cref{sec:formalisation} for a demonstration in \Agda.}
+Mode-correctness can be proved by applying \cref{lem:decidability-mode-correctness} and its type synthesiser is derived directly by \cref{cor:trichotomy} with the specification~$(\Sigma, \Omega)$.
+For a demonstration in \Agda, see \cref{sec:formalisation}.

### 2023-10-16 tex/ESOP24/S8-future.tex
[ tex ] Revise Discussion and Examples

@@
 We believe that our formal treatment lays a foundation for further investigation, as the essential aspects of bidirectional typing have been studied rigorously. 
-While our current development is based on simply typed languages to highlight the core ideas, it is evident that many concepts and aspects remain unexplored.
+While our current development is based on simply typed languages to highlight the core ideas, it is evident that many concepts and aspects remain untouched.
 In this section we discuss related work and potential directions for future research.
 \subsubsection{Language formalisation frameworks}
-Many general definitions of logics and type theories and frameworks for defining them have been proposed.
 
-\citet{Harper1993a,Harper2007}
-\cite{Uemura2021}
-\cite{Bauer2020,Haselwarter2021,Bauer2022a}
-Our general definitions are akin to the definitions used in other language formalisation frameworks~\citep{Ahrens2022,Allais2021,Fiore2022} in a proof assistant---the concept of a signature is at object level.
+The study of presenting logics universally at least date back to universal algebra and model theory where structures are studied for certain notions of arities and signatures.
+In programming language theory, \varcitet{Aczel1978}{'s binding signature} is an early example which is used to prove a general confluence theorem.
+There have been many more general definitions and frameworks of logics and type theories proposed, and we may classify those into two groups by where signatures are at the meta level or the object level:
+\begin{enumerate}
+\item \varcitet{Harper1993a}{'s logical framework \LF} and its family of variants~\cite{Harper2007,Assaf2016,Felicissimo2023} are \emph{extensions} of Martin-L\"of type theory where the concept of signatures are on the \emph{meta level} and naturally capable of defining dependent type theories;
+\item general dependent type theories~\cite{Bauer2020,Haselwarter2021,Bauer2022a,Uemura2021}, categorical semantics~\cite{Fiore1999,Tanaka2006,Tanaka2006a,Fiore2010,Hamana2011,Fiore2013,Arkor2020,Fiore2022} (which includes the syntactic model as a special case), and frameworks for substructural systems~\cite{Tanaka2006,Tanaka2006a,Wood2022} are developed \emph{within} a meta-theory (set theory or type theory) where signatures are on the \emph{object level} and their expressiveness varies depending on their target languages.
+\end{enumerate}
+
+\LF is naturally expressive but each extension requires a different \LF and thus a different implementation.
+Formalising \LF and its variants is at least as hard as formalising a dependent type theory, so they are mostly implemented separately from their theory.
+To the best of our knowledge, we do not know any verified implementation of \LF in a type theory.
+
+For the second group, theories developed within set theory often can be translated to type theory (with some effort) and thus manageable for formalisation in a type-theoretic proof assistant. 
+Such examples include frameworks developed by \citet{Ahrens2022,Allais2021,Fiore2022} in \Coq or \Agda, which are relatively correct to their meta-language, although there expressiveness is limited to simply typed languages.
+Our development belongs to the second group, as we aim for a formal treatment of bidirectional typing.
 
@@
 
-Bidirectional type synthesis plays a crucial role in handling more complex types than simple types, and we sketch how our theory can be extended to treat a broader class of languages.
+Bidirectional type synthesis plays a crucial role in handling more complex types than simple types, and we sketch how our theory can be (informally) extended to treat a broader class of languages.
 We need a general definition of languages in question (\cref{subsec:simple-types,subsec:binding-sig}).
-Then, this general definition can be augmented with modes (\cref{subsec:bidirectional-system}) and the formal definition of mode-correctness can be adapted (\cref{def:mode-correctness}).
+Then, this general definition can be augmented with modes (\cref{subsec:bidirectional-system}) and the definition of mode-correctness (\cref{def:mode-correctness}) can be adapted by adding the mode information into signatures.
 Soundness and completeness (\cref{lem:soundness-completeness}) should still hold, as they amount to the separation and combination of mode and typing information for a given raw term (in a syntax-directed formulation).
@@
 \paragraph{Polymorphic types}
-In the case of languages like \SystemF and others that permit type-level variable binding, we can start with the notion of a polymorphic signature, as introduced by \citet{Hamana2011}:
-\begin{inlineenum}
+In the case of languages like \SystemF and others that permit type-level variable binding, we can start with the notion of polymorphic signature, as introduced by \citet{Hamana2011}---\begin{inlineenum}
   \item each type construct in a signature is specified by a binding arity with a single sort $*$, and
@@
 A <: A[B/\alpha]$.
-A general theory for bidirectional typing that accommodates various solutions to the instantiation problem as an assumption is left as future work.
+A general theory for bidirectional typing that accommodates various solutions to the instantiation problem (as an assumption) is left as future work.
 
@@
 
-Extending the logical framework \Dedukti, \varcitet{Felicissimo2023}{'s \textsf{CompLF}} is capable of defining dependent type theories with computational rules and enables generic bidirectional type synthesis.
-\LT{compare his work with ours}
-Generic bidirectional type synthesis, as presented in Felicissimo's work, is decidable, sound, and complete with respect to mode-decorated terms, provided that the underlying language specification is mode-correct, and the equality check is \emph{well-behaved}, i.e.\ type-preserving, confluent, and strongly normalising.
+Extending the logical framework \Dedukti, \varcitet{Felicissimo2023}{'s recent \textsf{CompLF}} is capable of defining dependent type theories with computational rules and enables generic bidirectional type synthesis.
+\citeauthor{Felicissimo2023} proposes a notion of \LF signature with modes (called \emph{moded signatures} op.\ cit.) and a definition of mode-correctness (called \emph{well-typed} op.\ cit.) and their algorithm of generic bidirectional type synthesis is decidable, sound, and complete with respect to mode-decorated terms, provided that the specified language is mode-correct, and the equality check is \emph{well-behaved}, i.e.\ type-preserving, confluent, and strongly normalising.
 However, the last assumption is challenging to establish even for specific languages and lacks a comprehensive general understanding.
@@
 
-\begin{table}[ht]
-  \renewcommand{\arraystretch}{2.5}
-  \setlength{\tabcolsep}{3pt}
-  \centering\footnotesize
-\begin{tabular}{c | l}
-  Rules & Operations \\ \hline\hline
-  $\inferrule{ }{\Gamma \vdash \mathtt{z} \chk \mathtt{nat}}$ &  $\aritysymbol{\mathsf{z}}{\cdot}{\cdot}{\mathtt{nat}^{\chk}}$ 
-  \\
- $\inferrule{\Gamma \vdash t \chk \mathtt{nat}}{\Gamma \vdash \mathtt{s}(t) \chk \mathtt{nat}}$ & $\aritysymbol{\mathsf{s}}{\cdot}{\mathtt{nat}^{\chk}}{\mathtt{nat}^{\chk}}$ \\
- $\inferrule{\Gamma \vdash t \syn \mathtt{nat} \\\\ \Gamma \vdash t_0 \chk A \and \Gamma, x : \mathtt{nat} \vdash t_1 \chk A}{\Gamma \vdash \mathtt{ifz}(t_0; x.t_1)(t) \chk A}$ & $\aritysymbol{\mathsf{ifz}}{A}{\mathtt{nat}^{\syn}, A^{\chk}, A^{\chk}}{A^{\chk}}$ \\
- $\inferrule{\Gamma \vdash t \chk A \and \Gamma \vdash u \chk B}{\Gamma \vdash (t, u) \chk A \times B}$ & $\aritysymbol{\mathsf{pair}}{A, B}{A^{\chk}, B^{\chk}}{A \times B^{\chk}}$  \\
- $\inferrule{\Gamma \vdash t \syn A_1 \times A_2}{\Gamma \vdash \mathtt{proj}_i(t) \syn A_i}$ for $i = 1, 2$ & $\aritysymbol{\mathsf{proj}_i}{A_1, A_2}{A_1 \times A_2^{\syn}}{A_i^{\syn}}$ for $i = 1, 2$ \\
-
- $\inferrule{\Gamma \vdash t \chk A_i}{\Gamma \vdash \mathtt{inj}_i(t) \chk A_1 + A_2}$ for $i = 1, 2$ & $\aritysymbol{\mathsf{inj}_i}{A_1, A_2}{A_i^{\chk}}{A_1 + A_2^{\chk}}$ for $i = 1, 2$ \\
-
- $\inferrule{\Gamma \vdash u \syn A + B \\\\ \Gamma, x_1 : A \vdash t_1 \chk C \and \Gamma, x_2 : B \vdash t_2 \chk C}{\Gamma \vdash \mathtt{case}(u; x_1.
-t_1; x_2.
-t_2) \chk C}$ & $\aritysymbol{\mathsf{case}}{A, B, C}{{A + B}^{\syn}, [A]C^{\chk}, [B]C^{\chk}}{C^{\syn}}$ \\
-
- $\inferrule{\Gamma, x : A \vdash t \chk A}{\Gamma \vdash \mu x.\, t \chk A}$ & $\aritysymbol{\mathsf{mu}}{A}{[A]A^{\chk}}{A^{\chk}}$ \\
-
- $\inferrule{\Gamma \vdash t \syn A \and \Gamma, x : A \vdash u \chk B}{\Gamma \vdash \mathtt{let}\;x = t\;\mathtt{in}\;u\chk B}$ & 
-  $\aritysymbol{\mathsf{let}}{A, B}{A^{\syn},[A]B^{\chk}}{B^{\chk}}$ \\
- $\inferrule{\Gamma \vdash t \syn A}{\Gamma \vdash \mathtt{ret}(t) \syn T(A)}$ & $\aritysymbol{\mathsf{ret}}{A}{A^{\syn}}{T(A)^{\syn}}$ \\
- $\inferrule{\Gamma \vdash t \syn T(A) \and \Gamma, x : A \vdash u \syn T(B)}{\Gamma \vdash \mathtt{bind}(t; x.u) \syn T(B)}$ & $\aritysymbol{\mathsf{bind}}{A, B}{T(A)^{\syn}, [A]T(B)^{\syn}}{T(B)^{\syn}}$
-\end{tabular}
-\caption{A computational calculus with naturals, products, sums, and general recursion}
-\label{tab:computational-calculus}
-\end{table}
+%\subsection*{Acknowledgement}
+%We thank Nathanael Arkor for the useful conversation.
+%James Wood
+%We thank Kuen-Bang {Hou (Favonia)} for his comments and suggestions.

### 2023-10-16 tex/ESOP24/S8-future.tex
[ tex ] Revise Discussion

@@
 The study of presenting logics universally at least date back to universal algebra and model theory where structures are studied for certain notions of arities and signatures.
-In programming language theory, \varcitet{Aczel1978}{'s binding signature} is an early example which is used to prove a general confluence theorem.
-There have been many more general definitions and frameworks of logics and type theories proposed, and we may classify those into two groups by where signatures are at the meta level or the object level:
+In programming language theory, \varcitet{Aczel1978}{'s binding signature} is an example which is used to prove a general confluence theorem.
+Many general definitions and frameworks for defining logics and type theories have been proposed, and we classify them into two groups by where signatures reside: the meta level or the object level of a meta-language:
 \begin{enumerate}

### 2023-10-17 tex/ESOP24/S8-future.tex
[ tex ] Revise S7

@@
 \begin{enumerate}
-\item \varcitet{Harper1993a}{'s logical framework \LF} and its family of variants~\cite{Harper2007,Assaf2016,Felicissimo2023} are \emph{extensions} of Martin-L\"of type theory where the concept of signatures are on the \emph{meta level} and naturally capable of defining dependent type theories;
+\item \varcitet{Harper1993a}{'s logical framework \LF} and its family of variants~\cite{Harper2007,Assaf2016,Felicissimo2023} (to name a few) are \emph{extensions} of Martin-L\"of type theory where signatures are on the \emph{meta level} and capable of specifying dependent type theories;
 \item general dependent type theories~\cite{Bauer2020,Haselwarter2021,Bauer2022a,Uemura2021}, categorical semantics~\cite{Fiore1999,Tanaka2006,Tanaka2006a,Fiore2010,Hamana2011,Fiore2013,Arkor2020,Fiore2022} (which includes the syntactic model as a special case), and frameworks for substructural systems~\cite{Tanaka2006,Tanaka2006a,Wood2022} are developed \emph{within} a meta-theory (set theory or type theory) where signatures are on the \emph{object level} and their expressiveness varies depending on their target languages.
@@
 
-\LF is naturally expressive but each extension requires a different \LF and thus a different implementation.
-Formalising \LF and its variants is at least as hard as formalising a dependent type theory, so they are mostly implemented separately from their theory.
-To the best of our knowledge, we do not know any verified implementation of \LF in a type theory.
+\LF is naturally expressive but each extension requires a different \LF and thus a different implementation to check formal \LF proofs.
+Formalising \LF and its variants is at least as hard as formalising a dependent type theory, so they are mostly implemented separately from their theory and unverified.
 
-For the second group, theories developed within set theory often can be translated to type theory (with some effort) and thus manageable for formalisation in a type-theoretic proof assistant. 
-Such examples include frameworks developed by \citet{Ahrens2022,Allais2021,Fiore2022} in \Coq or \Agda, which are relatively correct to their meta-language, although there expressiveness is limited to simply typed languages.
-Our development belongs to the second group, as we aim for a formal treatment of bidirectional typing.
+For the second group, theories developed within set theory often can be translated to type theory and thus manageable for formalisation in a type-theoretic proof assistant. 
+Such examples include frameworks developed by \citet{Ahrens2022,Allais2021,Fiore2022} in \Coq or \Agda, which are relatively correct to their meta-language, although their expressiveness is limited to simply typed theories.
+Our development belongs to the second group, as we aim for a formal treatment of bidirectional typing in a constructive type theory to minimise the gap between theory and implementation.
 
@@
 
-Extending the logical framework \Dedukti, \varcitet{Felicissimo2023}{'s recent \textsf{CompLF}} is capable of defining dependent type theories with computational rules and enables generic bidirectional type synthesis.
+Extending the logical framework \Dedukti~\cite{Assaf2016}, \varcitet{Felicissimo2023}{'s recent \textsf{CompLF}} is capable of defining dependent type theories with computational rules and enables generic bidirectional type synthesis.
 \citeauthor{Felicissimo2023} proposes a notion of \LF signature with modes (called \emph{moded signatures} op.\ cit.) and a definition of mode-correctness (called \emph{well-typed} op.\ cit.) and their algorithm of generic bidirectional type synthesis is decidable, sound, and complete with respect to mode-decorated terms, provided that the specified language is mode-correct, and the equality check is \emph{well-behaved}, i.e.\ type-preserving, confluent, and strongly normalising.
-However, the last assumption is challenging to establish even for specific languages and lacks a comprehensive general understanding.
+However, the last assumption is challenging to establish even for specific languages and lacks a comprehensive general theory.
 

### 2023-10-17 tex/ESOP24/S8-future.tex
[ tex ] Final two paragraphs of future work

@@
 
-\subsubsection{General theory of bidirectional type synthesis beyond simple types}
+\subsubsection{Beyond simple types}
 
@@
 
-\subsubsection{More characteristics of bidirectional typing}
+\subsubsection{Beyond syntax-directed systems}
 
-There are still many concepts that are possible to articulate such as, for example, Pfenning's recipe for bidirectionalising typing rules.
+For non-syntax-directed systems, since there can be multiple moding or typing rules for a language construct, both the mode decorator and the bidirectional type synthesiser will need to backtrack, and the decorator may need to produce all possible mode derivations so as not to limit the synthesiser's choices.
+For soundness and completeness, it should still be possible to treat them as the separation and combination of mode and type information, although bidirectional typing rules and pairs of a moding rule and a typing rule may no longer in perfect correspondence, and we may need to adapt the statement of completeness (for example by taking all possible mode derivations as input to guarantee that at least one can be paired with the input typing derivation).
 
-In contrast, there are concepts that may be hard to pin down, notably `annotation character'~\cite{Dunfield2021}, which is roughly about how easy it is for the user to write annotated programs.
-These considerations may continue to require creativity from the designer of type systems whereas its implementation could be automated entirely.
+\subsubsection{Towards a richer formal theory}
+
+There are more techniques in bidirectional typing that can be formally studied in general, with one notable example being the Pfenning recipe for bidirectionalising typing rules~\citep[Section~4]{Dunfield2021}.
+There are also concepts that may be hard to fully formalise, for example `annotation character'~\citep[Section~3.4]{Dunfield2021}, which is roughly about how easy it is for the user to write annotated programs, but it would be interesting to explore to what extent such concepts can be formalised.
+
+%Such aspects  may continue to require creativity from the designer of type systems whereas its implementation could be automated entirely.
 

### 2023-10-17 tex/ESOP24/S1-intro.tex
[ tex ] Final for ESOP

@@
 Some examples other than simply typed $\lambda$-calculus are given in \Cref{sec:example}.
-We conclude in \cref{sec:future} with further developments and a demonstration in \cref{sec:formalisation} for the use of our \Agda formalisation as programs.
+We discuss further developments in \cref{sec:future} and include a demonstration in \cref{sec:formalisation} for the use of our \Agda formalisation as programs.

### 2023-10-17 tex/ESOP24/S6-example.tex
[ tex ] Final for ESOP

@@
 The signature $\Omega_{\Lambda}^{\Leftrightarrow}$ is then extended with operations listed in \cref{tab:computational-calculus}.
-Mode-correctness can be proved by applying \cref{lem:decidability-mode-correctness} and its type synthesiser is derived directly by \cref{cor:trichotomy} with the specification~$(\Sigma, \Omega)$.
-For a demonstration in \Agda, see \cref{sec:formalisation}.
+Mode-correctness can be derived by \cref{lem:decidability-mode-correctness} and its type synthesiser by \cref{cor:trichotomy} with the specification~$(\Sigma, \Omega)$ directly.%
+\footnote{For a demonstration in \Agda, see \cref{sec:formalisation}.}

### 2023-10-17 tex/ESOP24/S8-future.tex
[ tex ] Final for ESOP

@@
 \begin{enumerate}
-\item \varcitet{Harper1993a}{'s logical framework \LF} and its family of variants~\cite{Harper2007,Assaf2016,Felicissimo2023} (to name a few) are \emph{extensions} of Martin-L\"of type theory where signatures are on the \emph{meta level} and capable of specifying dependent type theories;
+\item \varcitet{Harper1993a}{'s logical framework \LF} and its family of variants~\cite{Harper2007,Assaf2016,Felicissimo2023} are \emph{extensions} of Martin-L\"of type theory where signatures are on the \emph{meta level} and naturally capable of specifying dependent type theories;
 \item general dependent type theories~\cite{Bauer2020,Haselwarter2021,Bauer2022a,Uemura2021}, categorical semantics~\cite{Fiore1999,Tanaka2006,Tanaka2006a,Fiore2010,Hamana2011,Fiore2013,Arkor2020,Fiore2022} (which includes the syntactic model as a special case), and frameworks for substructural systems~\cite{Tanaka2006,Tanaka2006a,Wood2022} are developed \emph{within} a meta-theory (set theory or type theory) where signatures are on the \emph{object level} and their expressiveness varies depending on their target languages.
@@
 
-For the second group, theories developed within set theory often can be translated to type theory and thus manageable for formalisation in a type-theoretic proof assistant. 
-Such examples include frameworks developed by \citet{Ahrens2022,Allais2021,Fiore2022} in \Coq or \Agda, which are relatively correct to their meta-language, although their expressiveness is limited to simply typed theories.
-Our development belongs to the second group, as we aim for a formal treatment of bidirectional typing in a constructive type theory to minimise the gap between theory and implementation.
+For the second group, theories developed in set theory can often be reformulated in type theory and thus manageable for formalisation in a type-theoretic proof assistant. 
+Such examples include frameworks developed by \citet{Allais2021,Fiore2022} in \Agda, which are relatively correct to their meta-language, although their expressiveness is limited to simply typed theories.
+Our work belongs to the second group, as we aim for a formalism in a constructive type theory to minimise the gap between theory and implementation.
 
@@
 
-Bidirectional type synthesis plays a crucial role in handling more complex types than simple types, and we sketch how our theory can be (informally) extended to treat a broader class of languages.
-We need a general definition of languages in question (\cref{subsec:simple-types,subsec:binding-sig}).
-Then, this general definition can be augmented with modes (\cref{subsec:bidirectional-system}) and the definition of mode-correctness (\cref{def:mode-correctness}) can be adapted by adding the mode information into signatures.
+Bidirectional type synthesis plays a crucial role in handling more complex types than simple types, and we sketch how our theory can be extended to treat a broader class of languages.
+First, we need a general definition of languages, discussed above, in question (\cref{subsec:simple-types,subsec:binding-sig}).
+Then, this definition can be augmented with modes (\cref{subsec:bidirectional-system}) and the definition of mode-correctness (\cref{def:mode-correctness}) can be adapted by adding the mode information into signatures.
 Soundness and completeness (\cref{lem:soundness-completeness}) should still hold, as they amount to the separation and combination of mode and typing information for a given raw term (in a syntax-directed formulation).
@@
 
-However, explicit type application is impractical but its implicit counterpart results in the \emph{instantiation problem} which amounts to transforming the implicit type application (a stationary rule) to guessing $B$ in $\forall \alpha.
-A <: A[B/\alpha]$.
-A general theory for bidirectional typing that accommodates various solutions to the instantiation problem (as an assumption) is left as future work.
+However, explicit type application in \SystemF and \SystemFsub is impractical but its implicit version results in a \emph{stationary rule} which is not syntax-directed:
+\[
+  \inferrule{\Gamma \vdash t \syn \forall \alpha. A}{\Gamma \vdash t \syn A[B/\alpha]}
+\]
+By translating the above rule to subtyping, we have the \emph{instantiation problem} which amounts to guessing $B$ in $\forall \alpha. A <: A[B/\alpha]$.
+A general theory that accommodates various solutions to the problem is left as future work.
 
@@
 Extending the logical framework \Dedukti~\cite{Assaf2016}, \varcitet{Felicissimo2023}{'s recent \textsf{CompLF}} is capable of defining dependent type theories with computational rules and enables generic bidirectional type synthesis.
-\citeauthor{Felicissimo2023} proposes a notion of \LF signature with modes (called \emph{moded signatures} op.\ cit.) and a definition of mode-correctness (called \emph{well-typed} op.\ cit.) and their algorithm of generic bidirectional type synthesis is decidable, sound, and complete with respect to mode-decorated terms, provided that the specified language is mode-correct, and the equality check is \emph{well-behaved}, i.e.\ type-preserving, confluent, and strongly normalising.
-However, the last assumption is challenging to establish even for specific languages and lacks a comprehensive general theory.
+They proposes a notion of \LF signature with modes (called \emph{moded signatures} op.\ cit.), a definition of mode-correctness (called \emph{well-typed} op.\ cit.), and an algorithm of generic bidirectional type synthesis which is decidable, sound, and complete with respect to mode-decorated terms, provided that the specified language is mode-correct, and the set of computational rules is \emph{well-behaved}, i.e.\ type-preserving, confluent, and strongly normalising.
+However, the last assumption is challenging to establish even for specific languages and lacks a general understanding.
 
@@
 For non-syntax-directed systems, since there can be multiple moding or typing rules for a language construct, both the mode decorator and the bidirectional type synthesiser will need to backtrack, and the decorator may need to produce all possible mode derivations so as not to limit the synthesiser's choices.
-For soundness and completeness, it should still be possible to treat them as the separation and combination of mode and type information, although bidirectional typing rules and pairs of a moding rule and a typing rule may no longer in perfect correspondence, and we may need to adapt the statement of completeness (for example by taking all possible mode derivations as input to guarantee that at least one can be paired with the input typing derivation).
+For soundness and completeness, it should still be possible to treat them as the separation and combination of mode and typing derivations, although bidirectional typing rules and pairs of a moding rule and a typing rule may no longer be in perfect correspondence.
+Completeness may need to be reformulated to take all possible mode derivations as input to guarantee that at least one can be paired with the input typing derivation.
 
@@
 
-There are more techniques in bidirectional typing that can be formally studied in general, with one notable example being the Pfenning recipe for bidirectionalising typing rules~\citep[Section~4]{Dunfield2021}.
+There are more techniques in bidirectional typing that could be formally studied in general, with one notable example being the Pfenning recipe for bidirectionalising typing rules~\citep[Section~4]{Dunfield2021}.
 There are also concepts that may be hard to fully formalise, for example `annotation character'~\citep[Section~3.4]{Dunfield2021}, which is roughly about how easy it is for the user to write annotated programs, but it would be interesting to explore to what extent such concepts can be formalised.

### 2023-10-17 tex/ESOP24/S4-pre-synthesis.tex
[ tex ] Minor tweaks

@@
 
-\begin{figure}[h]
+\begin{figure}
   \centering\small
@@
   \end{mathpar}
-
   \caption{Annotation ordering between raw terms}
@@
 \end{figure}
-
 On the other hand, when using a bidirectional type synthesiser to implement type synthesis~(\cref{thm:implementation}), our completeness is much simpler to use than annotatability, which requires the bidirectional type synthesiser to produce more complex evidence when the synthesis fails.

### 2023-10-27 tex/ESOP24/BiSig.tex
[ tex ] Update Demonstration
Update BiSig.tex

@@
 
+\numberwithin{table}{section}
+
 %\let\theorem\relax
@@
 
-\title{A formal treatment of bidirectional typing}
+\title{A formal treatment of bidirectional typing%
+%  \thanks{The work is supported by the National Science and Technology Council of Taiwan under grant NSTC 112-2221-E-001-003-MY3.}
+}
 
@@
 To remedy the situation, this paper develops a \emph{general} and \emph{formal} theory of bidirectional typing and as a by-product of our formalism provides a \emph{verified} generator of \emph{proof-relevant} type synthesisers for simply typed languages: For every signature that specifies a mode-correct bidirectionally typed language, there exists a proof-relevant type synthesiser that for an input abstract syntax tree constructs a typing derivation if any, gives its refutation if not, or reports that the input does not have enough type annotations.
-Soundness, completeness, and mode-correctness are studied universally for al signatures, which are sufficient conditions for deriving a type synthesiser.
+Soundness, completeness, and mode-correctness are studied universally for all signatures, which are sufficient conditions for deriving a type synthesiser.
 We propose a preprocessing step called \emph{mode decoration}, which helps the user to deal with missing type annotations in a given abstract syntax tree.
@@
 
-%\begin{acks}
-%%We thank Nathanael Arkor for the useful conversation.
-%We thank Kuen-Bang {Hou (Favonia)} for his comments and suggestions.
-%The work is supported by the \grantsponsor{NSTC}{National Science and Technology Council of Taiwan}{https://www.nstc.gov.tw/} under grant \grantnum{NSTC}{NSTC 112-2221-E-001-003-MY3}.
-%\end{acks}
-
 \bibliographystyle{splncs04nat}
@@
 
-%\input{appendix}
 \subfile{S7-formalisation}
+\input{appendix}
 

### 2023-12-21 tex/ESOP24/BiSig.tex
[ tex ] Address low-level comments incl.
1. British capitalisation
2. hyphenation
3. typos

@@
 \setlength{\marginparwidth}{2cm}
-\usepackage[obeyFinal,color=yellow,textsize=scriptsize]{todonotes}
+\usepackage[color=yellow,textsize=scriptsize]{todonotes}
 
@@
 
+\SetSymbolFont{stmry}{bold}{U}{stmry}{m}{n}
 \newcommand{\LT}[2][]{\todo[author=LT,color=orange!40,#1]{#2}}
@@
 %\let\theorem\relax
+%\spnewtheorem{theorem}{Theorem}[section]{\bfseries}{}
 %\let\definition\relax
-%\spnewtheorem{theorem}{Theorem}[section]{\bfseries}{}
 %\spnewtheorem{definition}{Definition}[section]{\bfseries}{}
 
-%\author{Liang-Ting Chen
-%\orcidID{0000-0002-3250-1331}
-%\and 
-%Hsiang-Shang Ko
-%\orcidID{0000-0002-2439-1048}}
-%
-%\institute{Institute of Information Science, Academia Sinica, Taiwan}
+\author{Liang-Ting Chen
+\orcidID{0000-0002-3250-1331}
+\and 
+Hsiang-Shang Ko
+\orcidID{0000-0002-2439-1048}}
 
-\title{A formal treatment of bidirectional typing%
-%  \thanks{The work is supported by the National Science and Technology Council of Taiwan under grant NSTC 112-2221-E-001-003-MY3.}
-}
+\institute{Institute of Information Science, Academia Sinica, Taiwan}
+
+\title{A formal treatment of bidirectional typing\thanks{The work is supported by the National Science and Technology Council of Taiwan under grant NSTC 112-2221-E-001-003-MY3.}}
 
@@
 %This situation is in stark contrast to parsing, for which there have been general theories and widely applicable tools such as parser generators.
-To remedy the situation, this paper develops a \emph{general} and \emph{formal} theory of bidirectional typing and as a by-product of our formalism provides a \emph{verified} generator of \emph{proof-relevant} type synthesisers for simply typed languages: For every signature that specifies a mode-correct bidirectionally typed language, there exists a proof-relevant type synthesiser that for an input abstract syntax tree constructs a typing derivation if any, gives its refutation if not, or reports that the input does not have enough type annotations.
+To remedy the situation, this paper develops a \emph{general} and \emph{formal} theory of bidirectional typing and as a by-product of our formalism provides a \emph{verified} generator of \emph{proof-relevant} type synthesisers for simply typed languages: for every signature that specifies a mode-correct bidirectionally typed language, there exists a proof-relevant type synthesiser that for an input abstract syntax tree constructs a typing derivation if any, gives its refutation if not, or reports that the input does not have enough type annotations.
 Soundness, completeness, and mode-correctness are studied universally for all signatures, which are sufficient conditions for deriving a type synthesiser.

### 2023-12-21 tex/ESOP24/S1-intro.tex
[ tex ] Address low-level comments incl.
1. British capitalisation
2. hyphenation
3. typos

@@
 
-Type inference is an important mechanism for the transition to well typed programs from untyped abstract syntax trees, which we call \emph{raw terms}.
+Type inference is an important mechanism for the transition to well-typed programs from untyped abstract syntax trees, which we call \emph{raw terms}.
 Here `type inference' refers specifically to algorithms that ascertain the type of any raw term \emph{without type annotations}.
@@
 In light of these limitations, \emph{bidirectional type synthesis} emerged as a viable alternative, providing algorithms for deciding the types of raw terms that meet some syntactic criteria and usually contain type annotations.
-\varcitet{Dunfield2021}{ summarised in their survey paper} the design principles of bidirectional type synthesis and its wide coverage of languages with simple types, polymorphic types, dependent types, gradual types, among others.
+In their survey paper, \varcitet{Dunfield2021}{ summarised} the design principles of bidirectional type synthesis and its wide coverage of languages with simple types, polymorphic types, dependent types, gradual types, among others.
 
-The basic idea of bidirectional type synthesis is that while the problem of type inference is not decidable in general, for certain kinds of terms it is still possible to infer their types (for example, the type of a variable can be looked up in the context); for other kinds of terms, we can switch to the simpler problem of type checking, where the expected type of a term is also given so that there is more information to work with.
+While type inference is not decidable in general, for certain kinds of terms it is still possible to infer their types.
+For example, the type of a variable can be looked up in the context.
+Bidirectional type synthesis combines type \emph{inference} on this subset of terms with type \emph{checking} (based on a given type) on the rest.
 More formally, every judgement in a bidirectional type system is extended with a \emph{mode:}
@@
 \end{inlineenum}
-The former indicates that the type~$A$ is computed as output, using both the context~$\Gamma$ and the term~$t$ as input, while for the latter, all three of $\Gamma$, $t$, and~$A$ are given as input.
-The algorithm of a bidirectional type synthesiser can usually be `read off' from a well designed bidirectional type system: as the synthesiser traverses a raw term, it switches between synthesis and checking, following the modes assigned to the judgements in the typing rules.
+The former indicates that the type~$A$ is an output, using both the context~$\Gamma$ and the term~$t$ as input, while for the latter, all three of $\Gamma$, $t$, and~$A$ are given as input.
+The algorithm of a bidirectional type synthesiser can usually be `read off' from a well-designed bidirectional type system: as the synthesiser traverses a raw term, it switches between synthesis and checking, following the modes assigned to the judgements in the typing rules.
 %if a condition called \emph{mode-correctness} is satisfied.
@@
 %While it is usually straightforward to derive a bidirectional type synthesis algorithm,
-\citeauthor{Dunfield2021} present informal design principles learned from individual bidirectional type systems, but in addition to crafting special techniques for individual systems, we should also start to consolidate concepts common to a class of bidirectional type systems into a general and formal theory that gives mathematically precise definitions and proves theorems for the class of systems once and for all.
+\citeauthor{Dunfield2021} present informal design principles learned from individual bidirectional type systems, but in addition to crafting special techniques for individual systems, we should start to consolidate concepts common to a class of bidirectional type systems into a general and formal theory that gives mathematically precise definitions and proves theorems for the class of systems once and for all.
 %Moreover, by mechanising the theory, we get \emph{verified} type-synthesiser generators analogous to parser generators.
@@
 Our work adopts a proof-relevant approach to (bidirectional) type synthesis, as illustrated by \citet{Wadler2022} for \PCF.
-The proof-relevant formulation deviates from the usual one:
-traditionally, a type synthesis algorithm is presented as \emph{algorithmic rules}, for example in the form $\Gamma |- \isTerm{t} \syn A \mapsto \isTerm{t}'$, which denotes that $\isTerm{t}$ in the surface language
+The proof-relevant formulation deviates from the usual one: traditionally, a type synthesis algorithm is presented as \emph{algorithmic rules}, for example in the form $\Gamma |- \isTerm{t} \syn A \mapsto \isTerm{t}'$, which denotes that $\isTerm{t}$ in the surface language
 %\Josh{The algorithm adds annotations to the input term?}
-can be transformed to a well typed term $\isTerm{t}'$ of type~$A$ in the core language~\cite{Pierce2000}.
-Such an algorithm is then accompanied by soundness and completeness assertions that the algorithm correctly synthesises the type of a raw term and every typable term can be synthesised.
-By contrast, the proof-relevant approach exploits the simultaneously computational and logical nature of Martin-L\"of type theory and formulates \emph{algorithmic soundness, completeness, and decidability in one go}.
+can be transformed to a well-typed term $\isTerm{t}'$ of type~$A$ in the core language~\cite{Pierce2000}.
+Such an algorithm is accompanied by soundness and completeness assertions that the algorithm correctly synthesises the type of a raw term and every typable term can be synthesised.
+By contrast, the proof-relevant approach exploits the simultaneously computational and logical nature of Martin-L\"of type theory and formulates algorithmic soundness, completeness, and decidability \emph{in one go}.
 %(although they do not emphasise the difference of their approach from the traditional one)
 
-In more detail:
 Recall that the law of excluded middle $P + \neg P$ does not hold as an axiom for every~$P$ constructively, and we say that $P$~is logically \emph{decidable} if the law holds for~$P$.
-Since Martin-L\"of type theory is simultaneously logical and computational, a decidability proof is also a proof-relevant decision procedure that computes not only a yes-or-no answer but also a proof of~$P$ or its refutation, so logical decidability is also algorithmic decidability.
+Since Martin-L\"of type theory is simultaneously logical and computational, a decidability proof is a proof-relevant decision procedure that computes not only a yes-or-no answer but also a proof of~$P$ or its refutation, so logical decidability is algorithmic decidability.
 %For example, a proof of \emph{decidable equality}, i.e.~$\forall x, y.~(x = y) + (x \neq y)$, decides whether $x$~and~$y$ are equal and accordingly gives an identity proof or a refutation explicitly.
@@
 Annotations therefore play an important role even in the definition of the problem solved by bidirectional typing, but have not received enough attention.
-In our theory, we define \emph{mode derivations} so that we can explicitly take annotations into account, in particular properly formulating the type synthesis problem.
-We also propose a preprocessing step called \emph{mode decoration} that helps the user to work with annotations in practice.
+In our theory, we define \emph{mode derivations} so that we can explicitly take annotations into account and formulate the type synthesis problem with sufficiently annotated raw terms.
+Accordingly, a preprocessing step called \emph{mode decoration} is proposed to help the user to work with annotations.
 
-To give a bit more detail:
 The type synthesis problem is not just about deciding whether a raw term is typable---there is a third possibility that the term does not have sufficient annotations.
-To solve the type synthesis problem, before attempting to decide typability (using a bidirectional type synthesiser), we should first decide whether the raw term has sufficient annotations, which corresponds to whether the term has a mode derivation.
-Our theory gives a (proof-relevant) \emph{mode decorator}, which constructs a mode derivation for a raw term or provide information that refutes the existence of any mode derivation and helps the user to pinpoint missing annotations.
+Therefore, to solve the type synthesis problem, before attempting to decide typability (using a bidirectional type synthesiser), we should first decide whether the raw term has sufficient annotations, which corresponds to whether the term has a mode derivation.
+Our theory gives a (proof-relevant) \emph{mode decorator}, which constructs a mode derivation for a raw term or provides information that refutes the existence of any mode derivation and helps the user to pinpoint missing annotations.
 Then a bidirectional type synthesiser is required to decide the typability of only mode-decorated raw terms instead of all raw terms.
-Soundness and completeness of bidirectional typing is reformulated as a one-to-one correspondence between bidirectional typing derivations and pairs of a typing derivation and a mode derivation for the same raw term; this reformulation is more natural and useful than annotatability~\cite[Section~3.2]{Dunfield2021} which is typically formulated in the literature of bidirectional typing.
+Soundness and completeness of bidirectional typing is reformulated as a one-to-one correspondence between bidirectional typing derivations and pairs of a typing derivation and a mode derivation for the same raw term; this reformulation is more useful than annotatability~\cite[Section~3.2]{Dunfield2021} which is typically formulated in the literature of bidirectional typing.
 

### 2023-12-21 tex/ESOP24/S2-outline.tex
[ tex ] Update Future work
[ tex ] Address low-level comments incl.
1. British capitalisation
2. hyphenation
3. typos

@@
 \section{Bidirectional type synthesis for simply typed \texorpdfstring{$\lambda$}{λ}-calculus} \label{sec:key-ideas}
+
+We start with an overview of our theory by instantiating it to simply typed $\lambda$-calculus.
+Roughly speaking, the problem of type synthesis requires us to take a raw term (i.e.~an untyped abstract syntax tree) as input, and produce a typing derivation for the term if possible.
+To give more precise definitions:
+The raw terms for simply typed $\lambda$-calculus are defined%
+\footnote{We omit the usual conditions about named representations of variables throughout this paper.}
+in \cref{fig:STLC-raw-terms};
+besides the standard constructs, there is an \textsc{Anno} rule that allows the user to insert type annotations to facilitate type synthesis.
 
@@
 \end{figure}
-
-We start with an overview of our theory by instantiating it to simply typed $\lambda$-calculus.
-Roughly speaking, the problem of type synthesis requires us to take a raw term (i.e.~an untyped abstract syntax tree) as input, and produce a typing derivation for the term if possible.
-To give more precise definitions:
-The raw terms for simply typed $\lambda$-calculus are defined%
-\footnote{We omit the usual conditions about named representations of variables throughout this paper.}
-in \cref{fig:STLC-raw-terms};
-besides the standard constructs, there is an \textsc{Anno} rule that allows the user to insert type annotations to facilitate type synthesis.
 Correspondingly, the definition of typing derivations%
@@
 One (less important) benefit of mode decoration is that it helps to simplify the synthesiser, whose computation can be partly directed by a mode derivation.
-More importantly, whether there is a mode derivation for a term is actually very useful information to the user, because it corresponds to whether the term has enough type annotations:
-Observe that the $\SynRule{Anno}$ and $\ChkRule{Sub}$ rules allow us to switch between the synthesising and checking modes;
+More importantly, whether there is a mode derivation for a term is actually very useful information to the user, because it corresponds to whether the term has enough type annotations: observe that the $\SynRule{Anno}$ and $\ChkRule{Sub}$ rules allow us to switch between the synthesising and checking modes;
 the switch from synthesising to checking is free, whereas the opposite direction requires a type annotation.
@@
 The construction is straightforward:
-Run the mode decorator on the input term~$\erase\Gamma \vdash t$.
+run the mode decorator on the input term~$\erase\Gamma \vdash t$.
 If there is no synthesising mode derivation, report that $t$~is not in synthesising mode (the third outcome).
@@
 
-We will construct a mode decorator~(\cref{sec:mode-decoration}) and a bidirectional type synthesiser~(\cref{sec:type-synthesis}) and prove both lemmas~(\cref{sec:soundness-and-completeness}) for all syntax-directed bidirectional simple type systems.
+We will construct a mode decorator~(\cref{sec:mode-decoration}) and a bidirectional type synthesiser~(\cref{sec:type-synthesis}) and prove the above theorem~(\cref{sec:soundness-and-completeness}) for all syntax-directed bidirectional simple type systems.
 To quantify over all such systems, we need their general definitions, which we formulate next.

### 2023-12-21 tex/ESOP24/S3-defs.tex
[ tex ] Minor tweaks
[ tex ] Address low-level comments incl.
1. British capitalisation
2. hyphenation
3. typos

@@
 %Here's the 'small print': Our definitions cover typical examples and set the scope for bidirectional simple type systems discussed, but do not offer comprehensive coverage of all possibilities.}
-of simple types, simply typed languages, and bidirectional type systems, and uses simply typed $\lambda$-calculus in \cref{sec:key-ideas} as our running example.
+of simple types, simply typed languages, and bidirectional type systems, and uses the simply typed $\lambda$-calculus in \cref{sec:key-ideas} as our running example.
 These definitions may look dense, especially on first reading.
@@
     Even though our theory is developed in Martin-L\"of type theory, the term `set' is used instead of `type' to avoid the obvious confusion. 
-    Also, as we assume \AxiomK, all types are legitimately sets in the sense of homotopy type theory.
-  } of operation symbols and an assignment of arities to symbols;
+    Indeed, as we assume \AxiomK, all types are legitimately sets in the sense of homotopy type theory~\citep[Definition 3.1.1]{UFP2013}.} of operation symbols and an assignment of arities to symbols;
 \item then, given a signature, we define raw terms and typing derivations inductively by primitive rules such as $\Rule{Var}$ and a rule schema for constructs $\tmOp_o$ indexed by an operation symbol~$o$.
@@
 
-\begin{definition} \label{def:substitution}
+\begin{definition}\label{def:substitution}
 The \emph{substitution} for a function $\rho\colon \Xi \to \Type_{\Sigma}(\Xi')$, denoted by $\rho\colon \Sub{\Xi}{\Xi'}$, is a map which sends a type $A : \Type_{\Sigma}(\Xi)$ to $\simsub{A}{\rho} : \Type_{\Sigma}(\Xi')$ and is defined as usual.
@@
 To use a rule like $\Rule{Abs}$ in an actual typing derivation, we will need to substitute \emph{concrete types}, i.e.\ types without any type variables, for variables $A, B$.
-In our formulation of substitution~(\cref{def:substitution}), we must first identify for which type variables to substitute.
+In our formulation of substitution~(\ref{def:substitution}), we must first identify which type variables to substitute for.
 As such, this information forms part of the arity of an operation, and typing derivations, defined subsequently, will include functions from~$\Xi$ to concrete types specifying how to instantiate typing rules by substitution.
@@
     \and
-    \inferrule{\rho : \Sub{\Xi}{\emptyset} \\  \Gamma, \vec{\isTerm{x}}_\isTerm{1} : \simsub{\Delta_{1}}{\rho} |-_{\Sigma, \Omega} \isTerm{t_1} : \simsub{A_{1}}{\rho} \quad\cdots\quad \Gamma, \vec{\isTerm{x}}_\isTerm{n} : \simsub{\Delta_{n}}{\rho} |-_{\Sigma, \Omega} \isTerm{t_n} : \simsub{A_{n}}{\rho}}
+    \inferrule{\rho : \Sub{\Xi}{\emptyset} \and  \Gamma, \vec{\isTerm{x}}_\isTerm{1} : \simsub{\Delta_{1}}{\rho} |-_{\Sigma, \Omega} \isTerm{t_1} : \simsub{A_{1}}{\rho} \and \cdots\and \Gamma, \vec{\isTerm{x}}_\isTerm{n} : \simsub{\Delta_{n}}{\rho} |-_{\Sigma, \Omega} \isTerm{t_n} : \simsub{A_{n}}{\rho}}
     {\Gamma |-_{\Sigma, \Omega} \tmOpts : \simsub{A_0}{\rho}}\,\Rule{Op}
@@
     \\
-    \inferrule{\rho\colon \Sub{\Xi}{\emptyset} \\ \Gamma, \simsub{\vars{x}_\isTerm{1} : \Delta_{1}}{\rho} |-_{\Sigma, \Omega} \isTerm{t_{1}} :^\dir{d_1} \simsub{A_1}{\rho} \\
-      \cdots \\
+    \inferrule{\rho\colon \Sub{\Xi}{\emptyset} \and \Gamma, \simsub{\vars{x}_\isTerm{1} : \Delta_{1}}{\rho} |-_{\Sigma, \Omega} \isTerm{t_{1}} :^\dir{d_1} \simsub{A_1}{\rho} \and
+      \cdots \and
     \Gamma, \simsub{\vars{x}_\isTerm{n} : \Delta_{n}}{\rho} |-_{\Sigma, \Omega} \isTerm{t_n} :^{\dir{d_n}} \simsub{A_{n}}{\rho}}
-    {\Gamma |-_{\Sigma, \Omega} \tmOpts :^{\dir{d}} \simsub{A_0}{\rho}} \,\mathsf{Op}
+    {\Gamma |-_{\Sigma, \Omega} \tmOpts :^{\dir{d}} \simsub{A_0}{\rho}} \,\Rule{Op}
     \\ \text{for $\biop$ in $\Omega$}

### 2023-12-21 tex/ESOP24/S4-pre-synthesis.tex
[ tex ] Address low-level comments incl.
1. British capitalisation
2. hyphenation
3. typos

@@
 Our first important construction is mode decoration in \cref{sec:mode-decoration}, which is in fact generalised to pinpoint any missing type annotations in a given raw term.
-We also discuss some related properties:
-By bringing mode derivations into the picture, we are able to give a natural formulation of soundness and completeness of a bidirectional type system with respect to its erasure to an ordinary type system in \cref{sec:soundness-and-completeness}.
+We also discuss some related properties: by bringing mode derivations into the picture, we are able to give a natural formulation of soundness and completeness of a bidirectional type system with respect to its erasure to an ordinary type system in \cref{sec:soundness-and-completeness}.
 We also formulate annotatability~\cite{Dunfield2021} and discuss its relationship with our completeness and generalised mode decoration in \cref{sec:annotatability}.
@@
 The major goal of this section is to construct a mode decorator, which decides for any raw term $V |-_{\Sigma,\erase\Omega} t$ and mode~$\dir d$ whether $V |-_{\Sigma,\Omega} t^\dir{d}$ or not.
-In fact we will do better:
-If a mode decorator returns a proof that no mode derivation exists, that proof (of a negation) does not provide useful information for the user.
+In fact we will do better: if a mode decorator returns a proof that no mode derivation exists, that proof (of a negation) does not provide useful information for the user.
 It will be more helpful if a decorator can produce an explanation of why no mode derivation exists, and even how to fix the input term to have a mode derivation.
@@
 
-This lemma would be wrong if the `strictness' boolean~$s$ was left out of the rules:
-Having both $\ChkRule{Sub}$ and $\SynRule{Missing}$, which we call \emph{mode casts}, it would be possible to switch between the two modes freely, which unfortunately means that we could insert a pair of $\ChkRule{Sub}$ and $\SynRule{Missing}$ anywhere, constructing a non-genuine derivation even when there is in fact a genuine one.
-The `strictness' boolean~$s$ can be thought of as disrupting the formation of such pairs of mode casts:
-Every rule other than the mode casts sets~$s$ to~$\BTrue$, meaning that a term is \emph{strictly} in the mode assigned by the rule (i.e.~not altered by a mode cast), whereas the mode casts set~$s$ to~$\BFalse$.
+This lemma would be wrong if the `strictness' boolean~$s$ was left out of the rules: having both $\ChkRule{Sub}$ and $\SynRule{Missing}$, which we call \emph{mode casts}, it would be possible to switch between the two modes freely, which unfortunately means that we could insert a pair of $\ChkRule{Sub}$ and $\SynRule{Missing}$ anywhere, constructing a non-genuine derivation even when there is in fact a genuine one.
+The `strictness' boolean~$s$ can be thought of as disrupting the formation of such pairs of mode casts: every rule other than the mode casts sets~$s$ to~$\BTrue$, meaning that a term is \emph{strictly} in the mode assigned by the rule (i.e.~not altered by a mode cast), whereas the mode casts set~$s$ to~$\BFalse$.
 Furthermore, the sub-derivation of a mode cast has to be strict, so it is impossible to have consecutive mode casts.
@@
 \end{proposition}
-
+\LT{the paper is perhaps too modest: (...) I don't think DK annotatability says anything about evidence---a synthesiser that answered "no" could still enjoy annotatability (even if we would not enjoy that answer).}
 Defined in \cref{fig:annotation-ordering}, the `annotation ordering' $t' \sqsupseteq t$ means that $t'$~has the same or more annotations than~$t$.

### 2023-12-21 tex/ESOP24/S5-synthesis.tex
[ tex ] Update Future work
[ tex ] Address low-level comments incl.
1. British capitalisation
2. hyphenation
3. typos

@@
       \end{enumerate}
-  \end{itemize}
+  \end{itemize}\qed
 \end{proof}
@@
 
-Combining the bidirectional type synthesiser with the mode decorator, soundness, and completeness from \cref{sec:pre-synthesis}, we get a type synthesiser parameterised by $(\Sigma, \Omega)$, generalising \cref{thm:implementation}.
+Combining the bidirectional type synthesiser with the mode decorator, soundness, and completeness from \cref{sec:pre-synthesis}, we derive a type synthesiser parameterised by~$(\Sigma, \Omega)$, generalising \cref{thm:implementation}.
 
@@
   \begin{enumerate}
-    \item $\Gamma |-_{\Sigma, \Omega} \isTerm{t} : A$ for some type~$A$,
-    \item $\Gamma |/-_{\Sigma, \Omega} \isTerm{t} : A$ for any type~$A$, or 
+    \item $\erase{\Gamma} |-_{\Sigma, \Omega} \isTerm{t}^{\syn}$ and $\Gamma |-_{\Sigma, \erase{\Omega}} \isTerm{t} : A$ for some type~$A$,
+    \item $\erase{\Gamma} |-_{\Sigma, \Omega} \isTerm{t}^{\syn}$ but $\Gamma |/-_{\Sigma, \erase{\Omega}} \isTerm{t} : A$ for any type~$A$, or 
     \item $\erase{\Gamma} |/-_{\Sigma, \Omega} \isTerm{t}^{\syn}$.

### 2023-12-21 tex/ESOP24/S8-future.tex
[ tex ] Address low-level comments incl.
1. British capitalisation
2. hyphenation
3. typos

@@
 
-However, explicit type application in \SystemF and \SystemFsub is impractical but its implicit version results in a \emph{stationary rule} which is not syntax-directed:
+However, explicit type application in \SystemF and \SystemFsub is impractical but its implicit version results in a \emph{stationary rule}~\cite{Leivant1986}:
 \[
@@
 \]
+which is not syntax-directed.
 By translating the above rule to subtyping, we have the \emph{instantiation problem} which amounts to guessing $B$ in $\forall \alpha. A <: A[B/\alpha]$.
-A general theory that accommodates various solutions to the problem is left as future work.
+A theory that accommodates various solutions to the problem is left as future work.
 
@@
 Extending the logical framework \Dedukti~\cite{Assaf2016}, \varcitet{Felicissimo2023}{'s recent \textsf{CompLF}} is capable of defining dependent type theories with computational rules and enables generic bidirectional type synthesis.
-They proposes a notion of \LF signature with modes (called \emph{moded signatures} op.\ cit.), a definition of mode-correctness (called \emph{well-typed} op.\ cit.), and an algorithm of generic bidirectional type synthesis which is decidable, sound, and complete with respect to mode-decorated terms, provided that the specified language is mode-correct, and the set of computational rules is \emph{well-behaved}, i.e.\ type-preserving, confluent, and strongly normalising.
+They propose a notion of \LF signature with modes (called \emph{moded signatures} op.\ cit.), a definition of mode-correctness (called \emph{well-typed} op.\ cit.), and an algorithm of generic bidirectional type synthesis which is decidable, sound, and complete with respect to mode-decorated terms, provided that the specified language is mode-correct, and the set of computational rules is \emph{well-behaved}, i.e.\ type-preserving, confluent, and strongly normalising.
 However, the last assumption is challenging to establish even for specific languages and lacks a general understanding.

### 2023-12-21 tex/ESOP24/response.md
[ tex ] Include ESOP24 reviews and our response

# Response

We thank the reviewers for their constructive and thoughtful comments to
improve our paper.  Given the limited time for the rebuttal period, we will
mainly answer questions specific for us, discuss common issues raised by
multiple reviewers, and respond to selected detailed comments.

## Questions for us

### Review 1

> Definition 5.2. seems to imply left-to-right argument processing. I can
> imagine rules where the type of an earlier argument is inferable from a later
> argument. Could mode correctness be generalized to only require that some
> correct order of processing exists? Additionally, we could try to find that
> order mechanically during mode correctness checking.

Yes, we can generalise the definition but we do not have to if a bidirectional
type system and its declarative typing are specified separately (with different
argument orders) and connected by an *additional* relation (instead of the mode
erasure). The relation will include a correspondence between the order of
processing and the order of premises.

Therefore, as an order of processing is specified by a bidirectional typing
rule, different orders of processing can be modelled by multiple rules. Thus,
finding that order mechanically may result in multiple bidirectional typing
rules for the same term construct, i.e. a non-syntax-directed system, which are
not dealt with in the current paper.

> In practice it's common that some term former can be both checked and
> inferred. For example, in checking mode we can check both branches of an
> if-then-else, but in inferred mode we have to arbitrarily pick an inferred
> branch. How hard would it be to support such "dual-mode" in the
> formalization?

Similarly, using two signatures to specify a bidirectional type system and a
simple type language separately should suffice to model this case, but the
specified bidirectional system is no longer syntax-directed.

We will discuss non-syntax-directed systems in the next section of our response.

### Review 2

> I wonder whether the one-to-one correspondence between meta-rules and
> operators is really necessary, or you could just reason on arbitrary
> meta-rules following the schema.

No, it is not technically necessary. Rules for raw terms, typing derivations,
and bidirectional typing derivations can all be specified independently from
different signatures and associated by some relations between them.

We will discuss non-syntax-directed systems below separately.

### Review 3

> Cor. 5.11, p. 22: "exactly one of the following holds": the judgement $\Gamma
> \vdash_{\Sigma,\Omega} t : A$ without a direction $d$ is used, which should
> mean the declarative typing judgement? In that case, I believe it is
> impossible that both 1 and 2 are false because the law of excluded middle is
> irrefutable.

Thanks for pointing this out! Indeed, the first two cases miss $\Gamma
\vdash_{\Sigma, \Omega} t \Rightarrow$ and should be the following two
statements

1. $\Gamma \vdash_{\Sigma, \Omega} t \Rightarrow$
   and $\Gamma \vdash_{\Sigma, \Omega} t : A$ 
2. $\Gamma \vdash_{\Sigma, \Omega} t \Rightarrow$
   but $\Gamma \nvdash_{\Sigma, \Omega} t : A$

respectively instead.

The mode decoration step divides raw terms into those satisfying $\Gamma
\vdash_{\Sigma, \Omega} t \Rightarrow$ and $\Gamma \nvdash_{\Sigma, \Omega} t
\Rightarrow$ and the bidirectional type synthesis divides the first class of
terms into terms satisfying $\Gamma \vdash_{\Sigma, \Omega} t : A$ and $\Gamma
\nvdash_{\Sigma, \Omega} t : A$.  The corollary is actually stronger than Def
2.1 as it provides evidence of $\neg E$ (the excuse $E$ does not hold). The
revised formalisation is given as follows:

```agda
synthesise : (Γ : Cxt 0) (r : Raw (length Γ))
           → (Pre Syn r × Dec (∃[ A ] Γ ⊢ r ⦂ A)) ⊎ ¬ Pre Syn r
synthesise Γ r with decorate Syn r
... | no ¬p = inr ¬p
... | yes p with synthesise⇔ Γ p
... | no ¬At      = inl (p , no λ (_ , t) → ¬At (_ , completeness p t))
... | yes (_ , t) = inl (p , yes (_ , soundness t))           
```

Strictly speaking, the above revised formalisation is still not precise, as the
original informal statement says *exactly* one of the three cases holds. That
is, we will need a stronger statement saying that if one of the cases holds
then the other two fail, which can be formalised in Agda as

```agda
data Tri (P : Set) (Q : Set) (R : Set) : Set where
  only-p : P → ¬ Q → ¬ R → Tri P Q R
  only-q : Q → ¬ P → ¬ R → Tri P Q R
  only-r : R → ¬ P → ¬ Q → Tri P Q R
```

Then, the trichotomy can be precisely formalised as

```agda
synthesise′
  : (Γ : Cxt 0) (r : Raw (length Γ))
  → Tri (Pre Syn r × ∃[ A ] Γ ⊢ r ⦂ A)
        (Pre Syn r × ¬ (∃[ A ] Γ ⊢ r ⦂ A))
        (¬ Pre Syn r)
synthesise′ Γ r with decorate Syn r
... | no ¬p = only-r ¬p (λ (p , _) → ¬p p) λ (p , _) → ¬p p 
... | yes p with synthesise⇔ Γ p
... | no ¬At      =
  only-q (p , λ (_ , t) → ¬At (_ , completeness p t)) (λ (_ , _ , At) → ¬At (_ , completeness p At)) (λ ¬p → ¬p p)
... | yes (_ , t) =
  only-p (p , _ , soundness t) (λ (_ , ¬At) → ¬At (_ , soundness t)) (λ ¬p → ¬p p)
```

We will revise our paper and add the above revised proofs in our formal
implementation.

### General concerns

#### Syntax-directedness

Syntax-directedness is a strong assumption, and we definitely want to deal with
non-syntax-directed systems in future work. We discuss this in a paragraph
‘Beyond syntax-directed systems’ on p25, which is rather condensed however, and
probably not easy to understand.

The following is an expanded version of the paragraph:

> To explore more general settings, an extreme possibility is to independently
> specify an ordinary type system and a bidirectional one over the same raw
> terms and then investigate possible relationships between the two systems.
>
> However, a more practical setting suitable for a next step is probably this:
> for each raw term construct, there can be several ordinary typing rules, and
> each typing rule can be refined to several bidirectional typing rules with
> different mode assignments and different orders of premises. Even for this
> setting, there is already more work to do: both the mode decorator and the
> bidirectional type synthesiser will have to backtrack, and become more
> complex. For soundness and completeness, it should still be possible to treat
> them as the separation and combination of mode and type information, but the
> completeness direction will pose a problem—for every node of a raw term, a
> mode derivation chooses a mode assignment while a typing derivation chooses a
> typing rule, but there may not be a bidirectional typing rule for this
> particular combination. There should be ways to fix this problem while
> retaining the mode decoration phase—for example, one idea is to make the mode
> decorator produce all possible mode derivations, and refine completeness to
> say that any typing derivation can be combined with one of these mode
> derivations into a bidirectional typing derivation.
>
> We could also focus on simpler cases first where each construct has exactly
> one typing rule, but each typing rule can be refined to more than one
> bidirectional typing rule, e.g. the dual-mode for `if-then-else`. In such
> cases, the mode decorator would need to find all possible mode derivations,
> but the type synthesiser should still work in a syntax-directed manner on
> each mode derivation as input. And completeness could still take the simple
> form presented in this paper too.

#### Section 3 is dense

We agree that Section 3 is rather dense as we also have admitted at the
beginning of the section.  We will try to squeeze more space for Section 3 to
include more examples and intuitive explanations as suggested by R2 and R3.

## Response to selected comments

> There's just one revision that I'd like to see: Jason Reed's "Redundancy
> Elimination for LF" seems to significantly overlap with the current work, so
> it should be referenced and compared.

Thanks for mentioning this paper to us. We will check this paper and compare it
with our approach in the next revision.

### Review 2

> below Theorem 2.5: which lemmas?

Sorry, it should be Theorem 2.5.

> Then, I do not see the motivation for the definition 3.4 where T is an
> arbitrary set (...).

This makes the Definition 3.6 shorter and allows us to introduce Example 3.5
earlier. In our early drafts, the notion of arities was not defined separately,
making Definition 3.6 longer and (in our opinion) less readable.

The notion of (unsorted) binding arities at least dates back to Aczel's 1978
manuscript (as mentioned in the introduction). The definition, we believe, is
standard and has been used by Fiore and Hur (2010) and Fiore Szamozvancev
(2022). (We will cite Fiore and Hur's paper in our next revision to give them
credit.)

- Marcelo Fiore and Chung-Kil Hur, Second-order equational logic (extended
  abstract), Computer Science Logic. CSL 2010, LNCS, vol. 6247, Springer Berlin
Heidelberg, 2010.

- Marcelo Fiore and Dmitrij Szamozvancev, Formal metatheory of second-order
  abstract syntax, Proceedings of the ACM on Programming Languages 6 (2022),
no. POPL, 1–29.

> Also, you seem to assume that there is exactly one meta-rule for each
> syntactic operator. Is this assumption technically necessary? I am not sure,
> or at least I do not see clearly why.

See our response about syntax-directedness.

> Definition 3.6 what is the set O expected to be? you say "for each
> operation", but the arity seems to be related to a meta-rule ... so it seems
> that you assume that there is exactly one meta-rule for each operator? this
> should be stated explicitly

The set $O$ has been used for the set of term constructs and names for
(bidirectional) typing rules. Indeed, to specify the set of raw terms we only
need an unsorted binding signature which could be introduced first or defined
as a (sorted) binding signature with the singleton set $\top$ as its sort.
Then, we would need to introduce some relations to associate each term
construct with various typing rule and bidirectional typing rules. This would
allow us to formulate a non-syntax-directed system properly at the cost of more
definitions in Section 3.

> Definition 3.4: you should say what is a sort (I guess, a set of type terms
> over some meta-variables?) . Also, I do not see the reason to define a sort
> as the set of type terms and not just the set of type meta-variables. In what
> follows it seems that only the meta-variables matter.

A sort is just a type/set. We will clarify it in our next revision.

> Also, I do not understand the sum in Definition 3.6.

The set of binding arities allowed for each operation $o : O$ consists of pairs
of a type $\Xi$ and a binding arity with a set $Ty(\Xi)$ (which depends on the
first component).

> Also, maybe I missed it, but the $\mathcal{U}$ in the sum was never mentioned
> (I guess it is the universe)

Yes, you are right -- it is the universe of types.

### Review 3

> Fig. 10, p. 16: rules without names

It was intended. We will add names for them, as it seems to cause confusion
unnecessarily.

### Review 4

> I don't think DK annotatability says anything about evidence---a synthesiser
> that answered "no" could still enjoy annotatability (even if we would not
> enjoy that answer)....

We should expand on what we mean by ‘more complex evidence’. When using a
bidirectional type synthesiser to implement a type synthesiser—in Theorem 2.4,
for example—if the bidirectional type synthesiser concludes that there does not
exist a bidirectional typing derivation, we use the contrapositive form of
completeness to establish that there does not exist an ordinary typing
derivation either. Now, annotatability (Proposition 4.7) is a kind of
completeness because (roughly speaking) it also turns an ordinary typing
derivation into a bidirectional typing derivation. Therefore it is conceivable
that we could use annotatability in place of completeness in the proof of
Theorem 2.4. However, in the contrapositive form of annotatability, the
antecedent is ‘there does not exist t’ that’s more annotated than t and has a
bidirectional typing derivation’, which is the ‘more complex evidence’ that the
bidirectional type synthesiser would have to produce.

> . 3: "this reformulation is more natural and useful" I agree it is more
> useful.

We will remove the word "natural".

> On page 9, the hyperlink "(Definition 3.3)" goes to Definition 2.3, but seems
> to intend Definition 3.3.

Thanks for pointing out. It appears to be a bug of the LNCS format, but we will
see if we can fix the link manually. (The link is generated by the command
`\cref` and we have checked that these two definitions have different labels.)

### 2023-12-21 tex/ESOP24/BiSig.tex
[ tex ] Add missing citations

@@
-\documentclass[orivec,envcountsame,envcountsect,oribibl]{llncs}
+\documentclass[orivec,envcountsect,envcountsame]{llncs}
 
@@
 
-\title{A formal treatment of bidirectional typing\thanks{The work is supported by the National Science and Technology Council of Taiwan under grant NSTC 112-2221-E-001-003-MY3.}}
+\title{A formal treatment of bidirectional typing\texorpdfstring{\thanks{The work is supported by the National Science and Technology Council of Taiwan under grant NSTC 112-2221-E-001-003-MY3.}}{}}
 
@@
 
-\IfFileExists{reference.bib}{
-\bibliography{reference}
-}{\bibliography{ref}}
+{\bibliography{ref}
 

### 2023-12-22 tex/ESOP24/S1-intro.tex
[ tex ] sort => type

@@
 
-In order to make the notion of mode-correctness precise, we first give a general definition of bidirectional simple type systems, called \emph{bidirectional binding signature}, extending the sorted version of \varcitet{Aczel1978}{'s binding signature} with modes.
+In order to make the notion of mode-correctness precise, we first give a general definition of bidirectional simple type systems, called \emph{bidirectional binding signature}, extending the typed version of \varcitet{Aczel1978}{'s binding signature} with modes.
 A general definition of typed languages allows us to define mode-correctness and to investigate its consequences rigorously, including the uniqueness of synthesised types and the decidability of bidirectional type synthesis for all mode-correct signatures.

### 2023-12-22 tex/ESOP24/S3-defs.tex
[ tex ] sort => type

@@
 \end{inlineenum}
-Therefore, to specify a term construct, we enrich the notion of arity with some sort for typing and extension context for variable binding.
+Therefore, to specify a term construct, we enrich the notion of arity with some set of types for typing and extension context for variable binding.
 
-\begin{definition}\label{def:binding-arity}
-  A \emph{binding arity} with a sort $T$ is an inhabitant of $\left(T^* \times T\right)^* \times T$ where $T^*$ is the set of lists over $T$.
-  In a binding arity $(((\Delta_1, A_1), \ldots, (\Delta_n, A_n)), A)$, every $\Delta_i$ and $A_i$ refers to the \emph{extension context} and the sort of the $i$-th argument, respectively, and $A$ the target sort.
-  For brevity, a binding arity is denoted by $\bargs \to A$ where $[\Delta_i]$ is omitted if $\Delta_i$ is empty.
+\begin{definition}[{\cite[p.~322]{Fiore2010}}]\label{def:binding-arity}
+  A \emph{binding arity} with a set $T$ of types is an inhabitant of $\left(T^* \times T\right)^* \times T$ where $T^*$ is the set of lists over $T$.
+  In a binding arity $(((\Delta_1, A_1), \ldots, (\Delta_n, A_n)), A)$, every $\Delta_i$ and $A_i$ refers to the \emph{extension context} and the \emph{type of the $i$-th argument}, respectively, and $A$ the target type.
+  For brevity, it is denoted by $\bargs \to A$ where $[\Delta_i]$ is omitted if empty.
 \end{definition}
@@
 Consider the $\Rule{Abs}$ rule (\Cref{fig:STLC-typing-derivations}).
-Its binding arity has the sort $\Type_{\Sigma_{\bto}}\{A, B\}$ and is ${[A]B} \to {(A \bto B)}$.
+Its binding arity has $\Type_{\Sigma_{\bto}}\{A, B\}$ as types and is ${[A]B} \to {(A \bto B)}$.
 This means that the $\Rule{Abs}$ rule for derivations of\/ $\Gamma |- \lam{x}t : A \bto B$ contains:
@@
 Next, akin to a signature, a binding signature $\Omega$ consists of a set of operation symbols along with their respective binding arities:
-%Also associated with every operation is the set~$\Xi$ of type variables that might be present in the operation's binding arity, and the sort of the binding arity is $\Type_{\Sigma}(\Xi)$:
 \begin{definition}\label{def:binding-signature}
@@
 \end{definition}
-That is, each inhabitant $o: O$  is associated with a set $\Xi$ of type variables and an arity $\arity(o)$ with the sort $\Type_{\Sigma}(\Xi)$ denoted by $\bop$.
+That is, each inhabitant $o: O$  is associated with a set $\Xi$ of type variables and an arity $\arity(o)$ with $\Type_{\Sigma}(\Xi)$ as types denoted by $\bop$.
 
@@
 \begin{definition} \label{def:bidirectional-binding-signature}
-  A \emph{bidirectional binding arity} with a sort $T$ is an inhabitant of
+  A \emph{bidirectional binding arity} with a set $T$ of types is an inhabitant of
   \[
@@
 \end{definition}
-We write $\biop$ for an operation $o$ with a variable set $\Xi$ and its bidirectional binding arity with sort $\Type_{\Sigma}(\Xi)$.
+We write $\biop$ for an operation $o$ with a variable set $\Xi$ and its bidirectional binding arity with $\Type_{\Sigma}(\Xi)$ as types.
 We call it \emph{checking} if $d$ is ${\chk}$ or \emph{synthesising} if $d$ is ${\syn}$; similarly its $i$-th argument is checking if $d_i$ is $\chk$ and synthesising if $d_i$ is $\syn$.

### 2023-12-22 tex/ESOP24/S2-outline.tex
[ tex ] Use `\rmfamily' for the body of a definition

@@
 
-\begin{definition}
+\begin{defn}
 \label{def:STLC-type-synthesiser}
@@
 \end{enumerate}
-\end{definition}
+\end{defn}
 
@@
 
-\begin{definition} \label{def:STLC-mode-decorator}
+\begin{defn} \label{def:STLC-mode-decorator}
 A \emph{mode decorator} decides for any raw term~$V \vdash \isTerm{t}$ whether $V \vdash \isTerm{t}^{\syn}$.
-\end{definition}
+\end{defn}
 
@@
 
-\begin{definition}
+\begin{defn}
 \label{def:STLC-bidirectional-type-synthesiser}
 A \emph{bidirectional type synthesiser} decides for any context~$\Gamma$ and synthesising term $\erase\Gamma \vdash \isTerm{t}^{\syn}$ whether $\Gamma \vdash \isTerm{t} \syn A$ for some type~$A$.
-\end{definition}
+\end{defn}
 
@@
 
-We will construct a mode decorator~(\cref{sec:mode-decoration}) and a bidirectional type synthesiser~(\cref{sec:type-synthesis}) and prove the above theorem~(\cref{sec:soundness-and-completeness}) for all syntax-directed bidirectional simple type systems.
+We will construct a mode decorator~(\cref{sec:mode-decoration}) and a bidirectional type synthesiser~(\cref{sec:type-synthesis}) and prove the above theorem for all syntax-directed bidirectional simple type systems~(\cref{sec:soundness-and-completeness}).
 To quantify over all such systems, we need their general definitions, which we formulate next.

### 2023-12-22 tex/ESOP24/S3-defs.tex
[ tex ] Use `\rmfamily' for the body of a definition

@@
 For simple types, the only datum needed for specifying a type construct is its number of arguments:
-\begin{definition} \label{def:simple-signature}
+\begin{defn} \label{def:simple-signature}
 %  \begin{figure}
@@
   \end{enumerate}
-\end{definition}
+\end{defn}
 
@@
 
-\begin{definition}\label{def:substitution}
+\begin{defn}\label{def:substitution}
 The \emph{substitution} for a function $\rho\colon \Xi \to \Type_{\Sigma}(\Xi')$, denoted by $\rho\colon \Sub{\Xi}{\Xi'}$, is a map which sends a type $A : \Type_{\Sigma}(\Xi)$ to $\simsub{A}{\rho} : \Type_{\Sigma}(\Xi')$ and is defined as usual.
-\end{definition}
+\end{defn}
 
@@
 
-\begin{definition}[{\cite[p.~322]{Fiore2010}}]\label{def:binding-arity}
+\begin{defn}[{\cite[p.~322]{Fiore2010}}]\label{def:binding-arity}
   A \emph{binding arity} with a set $T$ of types is an inhabitant of $\left(T^* \times T\right)^* \times T$ where $T^*$ is the set of lists over $T$.
@@
   For brevity, it is denoted by $\bargs \to A$ where $[\Delta_i]$ is omitted if empty.
-\end{definition}
+\end{defn}
 
@@
 Next, akin to a signature, a binding signature $\Omega$ consists of a set of operation symbols along with their respective binding arities:
-\begin{definition}\label{def:binding-signature}
+\begin{defn}\label{def:binding-signature}
   For a type signature $\Sigma$, a \emph{binding signature} $\Omega$ consists of a set $O$ and a function
@@
   \]
-\end{definition}
+\end{defn}
 That is, each inhabitant $o: O$  is associated with a set $\Xi$ of type variables and an arity $\arity(o)$ with $\Type_{\Sigma}(\Xi)$ as types denoted by $\bop$.
@@
 
-\begin{definition}
+\begin{defn}
   For a simply typed language $(\Sigma, \Omega)$, the family of sets of \emph{raw terms} indexed by a list~$V$ of variables consists of
@@
   These correspond to rules $\Rule{Var}$, $\Rule{Anno}$, and $\Rule{Op}$ in \Cref{fig:raw-terms} respectively.
-\end{definition}
+\end{defn}
 
@@
 
-\begin{definition}
+\begin{defn}
 A \emph{typing context} $\Gamma \colon \Cxt_{\Sigma}$ is formed by $\cdot$ for the empty context and $\Gamma, x : A$ for an additional variable~$x$ with a concrete type $A : \Type_{\Sigma}(\emptyset)$.
@@
 %The substitution $\simsub{\Gamma}{\rho}$ of a context $\Gamma$ is defined by applying substitution to each $A$ in $\Gamma$.
-\end{definition}
+\end{defn}
 
@@
 
-\begin{definition}\label{def:typing-derivations}
+\begin{defn}\label{def:typing-derivations}
   For a simply typed language $(\Sigma, \Omega)$, the family of sets of \emph{typing derivations} of $\Gamma \vdash \isTerm{t} : A$, indexed by a typing context $\Gamma : \Cxt_\Sigma$, a raw term~$\isTerm{t}$ with free variables in $\erase{\Gamma}$, and a type $A : \Type_\Sigma(\emptyset)$, consists of
@@
   corresponding to rules $\Rule{Var}$, $\Rule{Anno}$, and $\Rule{Op}$ in \Cref{fig:extrinsic-typing} respectively.
-\end{definition}
+\end{defn}
 
@@
 
-\begin{definition} \label{def:bidirectional-binding-signature}
+\begin{defn} \label{def:bidirectional-binding-signature}
   A \emph{bidirectional binding arity} with a set $T$ of types is an inhabitant of
@@
   For clarity, an arity is denoted by $\biarity$.
-\end{definition}
+\end{defn}
 
@@
 
-\begin{definition}
+\begin{defn}
   For a type signature $\Sigma$, a \emph{bidirectional binding signature} $\Omega$ is a set $O$ with
@@
   \]
-\end{definition}
+\end{defn}
 We write $\biop$ for an operation $o$ with a variable set $\Xi$ and its bidirectional binding arity with $\Type_{\Sigma}(\Xi)$ as types.
@@
 
-\begin{definition}\label{def:bidirectional-typing-derivations}\label{def:mode-derivations}
+\begin{defn}\label{def:bidirectional-typing-derivations}\label{def:mode-derivations}
   For a bidirectional type system $(\Sigma, \Omega)$,
@@
   A typing rule is \emph{checking} if its conclusion mode is $\chk$ or \emph{synthesising} otherwise.
-\end{definition}
+\end{defn}
 

### 2023-12-22 tex/ESOP24/S5-synthesis.tex
[ tex ] Use `\rmfamily' for the body of a definition

@@
 We also need some auxiliary definitions for the subset of variables of a type and of an extension context, and the set of variables that have been synthesised:
-\begin{definition}
+\begin{defn}
   The finite subsets\footnote{%
@@
   For an argument list $\biargs$ the set of type variables $A_{i}^{\dir{d_i}}$ with $d_i$ being $\syn$ is denoted by $\synvar(\biargs)$, i.e.\ $\synvar$ gives the set of type variables that will be synthesised during type synthesis.
-\end{definition}
+\end{defn}
 
-\begin{definition}\label{def:mode-correctness-args}
+\begin{defn}\label{def:mode-correctness-args}
   The \emph{mode-correctness} $\MCas\left(\biargvec\right)$ for an argument list $\biargs$ with respect to a subset $S$ of $\Xi$ is a predicate defined by
@@
   where $\top$ means an empty list is always mode-correct.
-\end{definition}
+\end{defn}
 This definition encapsulates the idea that every `input' type variable, possibly derived from an extension context~$\Delta_n$ or a checking argument~$A_n$, must be an `output' variable from $\synvar(\biargvec)$ or, if the rule is checking, belong to the set $S$ of `input' variables in its conclusion.
 This condition must also be met for every tail of the argument list to ensure that `output' variables accessible at each argument are from preceding arguments only, hence an inductive definition.
-\begin{definition}\label{def:mode-correctness}
+\begin{defn}\label{def:mode-correctness}
   An operation $\biop$ is \emph{mode-correct} if 
@@
   A bidirectional binding signature $\Omega$ is \emph{mode-correct} if its operations are all mode-correct.
-\end{definition}
+\end{defn}
 For a checking operation, an `input' variable of an argument could be derived from~$A_0$, as these are known during type checking as an input.
@@
 We need some auxiliary definitions for the notion of extension to state the unification problem:
-\begin{definition}
+\begin{defn}
 By an \emph{extension}\/ $\sigma \geq \rho$ of a partial substitution $\rho$ we mean that the domain $\dom(\sigma)$ of $\sigma$ contains the domain of $\rho$ and $\sigma(x) = \rho(x)$ for every\/ $x$ in $\dom(\rho)$.
   By a \emph{minimal extension}\/ $\bar{\rho}$ of $\rho$ satisfying $P$ we mean an extension $\bar{\rho} \geq \rho$ with $P(\bar{\rho})$ such that $\sigma \geq \bar{\rho}$ whenever $\sigma \geq \rho$ and $P(\sigma)$.
-\end{definition}
+\end{defn}
 \begin{lemma}\label{lem:unify}

### 2023-12-22 tex/ESOP24/appendix.tex
[ tex ] Update Future work
[ tex ] Update Demonstration
[ tex ] Use `\rmfamily' for the body of a definition

@@
 
-\section{Detailed definitions and proofs}\label{sec:defs-proofs}
+\section{Detailed definitions}\label{sec:defs-proofs}
 
-\begin{definition}\label{def:syn-var}
+\begin{defn}\label{def:syn-var}
   The subset $\fv(\Delta)$ of variables in an extension context~$\Delta$ is defined by\/ $\fv(\cdot) = \emptyset$ and\/ $\fv(\Delta, A) = \fv(\Delta) \cup \fv(A)$.
@@
   \end{align*}
-\end{definition}
+\end{defn}
+
+\begin{table}[b]
+  \renewcommand{\arraystretch}{2.5}
+  \setlength{\tabcolsep}{3pt}
+  \centering\footnotesize
+\begin{tabular}{c | l}
+  Rules & Operations \\ \hline\hline
+  $\inferrule{ }{\Gamma \vdash \mathtt{z} \chk \mathtt{nat}}$ &  $\aritysymbol{\mathsf{z}}{\cdot}{\cdot}{\mathtt{nat}^{\chk}}$ 
+  \\
+ $\inferrule{\Gamma \vdash t \chk \mathtt{nat}}{\Gamma \vdash \mathtt{s}(t) \chk \mathtt{nat}}$ & $\aritysymbol{\mathsf{s}}{\cdot}{\mathtt{nat}^{\chk}}{\mathtt{nat}^{\chk}}$ \\
+ $\inferrule{\Gamma \vdash t \syn \mathtt{nat} \\\\ \Gamma \vdash t_0 \chk A \and \Gamma, x : \mathtt{nat} \vdash t_1 \chk A}{\Gamma \vdash \mathtt{ifz}(t_0; x.t_1)(t) \chk A}$ & $\aritysymbol{\mathsf{ifz}}{A}{\mathtt{nat}^{\syn}, A^{\chk}, A^{\chk}}{A^{\chk}}$ \\
+ $\inferrule{\Gamma \vdash t \chk A \and \Gamma \vdash u \chk B}{\Gamma \vdash (t, u) \chk A \times B}$ & $\aritysymbol{\mathsf{pair}}{A, B}{A^{\chk}, B^{\chk}}{A \times B^{\chk}}$  \\
+ $\inferrule{\Gamma \vdash t \syn A_1 \times A_2}{\Gamma \vdash \mathtt{proj}_i(t) \syn A_i}$ for $i = 1, 2$ & $\aritysymbol{\mathsf{proj}_i}{A_1, A_2}{A_1 \times A_2^{\syn}}{A_i^{\syn}}$ for $i = 1, 2$ \\
+
+ $\inferrule{\Gamma \vdash t \chk A_i}{\Gamma \vdash \mathtt{inj}_i(t) \chk A_1 + A_2}$ for $i = 1, 2$ & $\aritysymbol{\mathsf{inj}_i}{A_1, A_2}{A_i^{\chk}}{A_1 + A_2^{\chk}}$ for $i = 1, 2$ \\
+
+ $\inferrule{\Gamma \vdash u \syn A + B \\\\ \Gamma, x_1 : A \vdash t_1 \chk C \and \Gamma, x_2 : B \vdash t_2 \chk C}{\Gamma \vdash \mathtt{case}(u; x_1.
+t_1; x_2.
+t_2) \chk C}$ & $\aritysymbol{\mathsf{case}}{A, B, C}{{A + B}^{\syn}, [A]C^{\chk}, [B]C^{\chk}}{C^{\syn}}$ \\
+
+ $\inferrule{\Gamma, x : A \vdash t \chk A}{\Gamma \vdash \mu x.\, t \chk A}$ & $\aritysymbol{\mathsf{mu}}{A}{[A]A^{\chk}}{A^{\chk}}$ \\
+
+ $\inferrule{\Gamma \vdash t \syn A \and \Gamma, x : A \vdash u \chk B}{\Gamma \vdash \mathtt{let}\;x = t\;\mathtt{in}\;u\chk B}$ & 
+  $\aritysymbol{\mathsf{let}}{A, B}{A^{\syn},[A]B^{\chk}}{B^{\chk}}$ \\
+ $\inferrule{\Gamma \vdash t \syn A}{\Gamma \vdash \mathtt{ret}(t) \syn T(A)}$ & $\aritysymbol{\mathsf{ret}}{A}{A^{\syn}}{T(A)^{\syn}}$ \\
+ $\inferrule{\Gamma \vdash t \syn T(A) \and \Gamma, x : A \vdash u \syn T(B)}{\Gamma \vdash \mathtt{bind}(t; x.u) \syn T(B)}$ & $\aritysymbol{\mathsf{bind}}{A, B}{T(A)^{\syn}, [A]T(B)^{\syn}}{T(B)^{\syn}}$
+\end{tabular}
+\caption{A computational calculus with naturals, products, sums, and general recursion}
+\label{tab:computational-calculus}
+\end{table}

### 2023-12-25 tex/ESOP24/S4-pre-synthesis.tex
[ tex ] Add rule names for annotation ordering
[ tex ] Add the revised remark about annotatability

@@
     \and
-    \inferrule{\vphantom{x : \Identifier}}
-              {\isTerm{x} \sqsupseteq \isTerm{x}}
+    \inferrule{\vphantom{x : \Identifier}}{\isTerm{x} \sqsupseteq \isTerm{x}}\;\Rule{Var}
     \and
     \inferrule{\isTerm{t} \sqsupseteq \isTerm{u}}
-              {(\isTerm{t} \annotate A) \sqsupseteq (\isTerm{u} \annotate A)}
+      {(\isTerm{t} \annotate A) \sqsupseteq (\isTerm{u} \annotate A)}\;\Rule{Anno}
     \and
     \inferrule{\isTerm{t_1} \sqsupseteq \isTerm{u_1} \quad \cdots \quad \isTerm{t_n} \sqsupseteq{u_n}}
-              {\tmOpts \sqsupseteq \tmOpus}
+      {\tmOpts \sqsupseteq \tmOpus}\;\Rule{Op}
   \end{mathpar}
@@
 \end{figure}
-On the other hand, when using a bidirectional type synthesiser to implement type synthesis~(\cref{thm:implementation}), our completeness is much simpler to use than annotatability, which requires the bidirectional type synthesiser to produce more complex evidence when the synthesis fails.
-Annotatability also does not help the user to deal with missing annotations like (generalised) mode decoration does: although annotatability seems capable of determining where annotations are missing and even filling them in correctly, its antecedent requires a typing derivation, which is what the user is trying to construct and does not have yet.
-%Therefore we believe that the notion of annotatability is too complex\LT{not clear what it means} for what it achieves, and our theory offers simpler and more useful alternatives.
+
+On the other hand, when using a bidirectional type synthesiser to implement a type synthesiser, for example in \cref{thm:implementation}, if the bidirectional type synthesiser concludes that there does not exist a bidirectional typing derivation, we use the contrapositive form of completeness to establish that such an ordinary typing derivation does not exist either.
+Now, annotatability is a kind of completeness because (roughly speaking) it turns an ordinary typing derivation bidirectional.
+Therefore it is conceivable that we could use annotatability in place of completeness in the proof of \cref{thm:implementation}.
+However, in the contrapositive form of annotatability, the antecedent is `there does not exist $t'$ that is more annotated than $t$ and has a bidirectional typing derivation', which is more complex than the bidirectional type synthesiser would have to produce.
+Annotatability also does not help the user to deal with missing annotations: although annotatability seems capable of determining where annotations are missing and even filling them in correctly, its antecedent requires a typing derivation, which is what the user is trying to construct and does not have yet.
+Therefore we believe that our theory offers simpler and more useful alternatives than the notion of annotatability.

### 2023-12-25 tex/ESOP24/S4-pre-synthesis.tex
[ tex ] Squeeze more space

@@
 Our first important construction is mode decoration in \cref{sec:mode-decoration}, which is in fact generalised to pinpoint any missing type annotations in a given raw term.
-We also discuss some related properties: by bringing mode derivations into the picture, we are able to give a natural formulation of soundness and completeness of a bidirectional type system with respect to its erasure to an ordinary type system in \cref{sec:soundness-and-completeness}.
-We also formulate annotatability~\cite{Dunfield2021} and discuss its relationship with our completeness and generalised mode decoration in \cref{sec:annotatability}.
+We discuss some related properties: by bringing mode derivations into the picture, we are able to give a natural formulation of soundness and completeness of a bidirectional type system with respect to its erasure to an ordinary type system in \cref{sec:soundness-and-completeness}.
+We formulate annotatability~\cite{Dunfield2021} and discuss its relationship with our completeness and generalised mode decoration in \cref{sec:annotatability}.
 
@@
 
-Erasure of a bidirectional binding signature removes mode information and keeps everything else intact; this can be straightforwardly extended by induction to remove mode information from a bidirectional typing derivation and arrive at an ordinary typing derivation, which is soundness.
-We can also remove typing and retain mode information, arriving at a mode derivation instead.
-Conversely, if we have both mode and typing derivations for the same term, we can combine them and obtain a bidirectional typing derivation, which is completeness.
-In short, soundness and completeness are no more than the separation and combination of mode and typing information carried by the three kinds of derivations while keeping their basic structure, which is directed by the same raw term.
-%Formulating the notion of mode derivation helps to complete this clean picture.
-Technically, all these can be summarised in one theorem (proved by routine induction).
+Erasure of a bidirectional binding signature removes modes and keeps everything else intact; this can be straightforwardly extended by induction to remove modes from a bidirectional typing derivation and arrive at an ordinary typing derivation, which is soundness.
+Instead we can remove typing and retain modes, arriving at a mode derivation.
+Conversely, if we have both mode and typing derivations for the same term, we can combine them to obtain a bidirectional typing derivation, which is completeness.
+In short, soundness and completeness are no more than the separation and combination of mode and typing information carried by the three kinds of derivations while keeping their basic structure, directed by the same raw term.
+All these can be summarised in one theorem (proved by induction).
 
@@
 
-The major goal of this section is to construct a mode decorator, which decides for any raw term $V |-_{\Sigma,\erase\Omega} t$ and mode~$\dir d$ whether $V |-_{\Sigma,\Omega} t^\dir{d}$ or not.
-In fact we will do better: if a mode decorator returns a proof that no mode derivation exists, that proof (of a negation) does not provide useful information for the user.
-It will be more helpful if a decorator can produce an explanation of why no mode derivation exists, and even how to fix the input term to have a mode derivation.
+The goal of this section is to construct a mode decorator, which decides for any raw term $V |-_{\Sigma,\erase\Omega} t$ and mode~$\dir d$ whether $V |-_{\Sigma,\Omega} t^\dir{d}$ or not.
+In fact we shall do better: if a mode decorator returns a proof that no mode derivation exists, that negation proof does not give useful information for the user.
+It will be helpful if a decorator can produce an explanation of why no mode derivation exists, and even how to fix the input term to have a mode derivation.
 We will construct such a \emph{generalised mode decorator}~(\cref{thm:generalised-mode-decoration}), which can be weakened to an ordinary mode decorator~(\cref{thm:mode-decoration}) if the additional explanation is not needed.%
@@
 For a bidirectional type system, an annotation is required wherever a term is `strictly' (which we will define shortly) in checking mode but required to be in synthesising mode, in which case there is no rule for switching from checking to synthesising, and thus there is no way to construct a mode derivation.
-We can, however, consider \emph{generalised mode derivations} defined in \cref{fig:generalised-mode-derivations} that allow the use of an additional $\SynRule{Missing}$ rule for such switching, so that a derivation can always be constructed.
-Given a generalised mode derivation, if it uses $\SynRule{Missing}$ in some places, then those places are exactly where annotations should be supplied, which is helpful information to the user; if it does not use $\SynRule{Missing}$, then the derivation is \emph{genuine} in the sense that it corresponds directly to an original mode derivation.
+We can, however, consider \emph{generalised mode derivations} (\cref{fig:generalised-mode-derivations}) that allow the use of an additional $\SynRule{Missing}$ rule for such switching, so that a derivation can always be constructed.
+Given a generalised mode derivation, if it uses $\SynRule{Missing}$ in some places, then those places are exactly where annotations should be supplied; if it does not use $\SynRule{Missing}$, then the derivation is \emph{genuine} in the sense that it corresponds directly to an original mode derivation.
 This can be succinctly formulated as \cref{thm:Pre?-true} below by encoding genuineness as a boolean~$g$ in the generalised mode judgement, which is set to~$\BFalse$ only by the $\SynRule{Missing}$ rule.
@@
 
-Finally, having constructed a generalised mode decorator, it is easy to derive an ordinary one.
+With a generalised mode decorator, it is now easy to derive an ordinary one.
 
@@
 
-\subsection{Annotatability}
-\label{sec:annotatability}
-
+\subsection{Annotatability}\label{sec:annotatability}
 \citet[Section~3.2]{Dunfield2021} formulated completeness differently from our \cref{lem:soundness-completeness} and proposed \emph{annotatability} as a more suitable name.
@@
 \end{proposition}
-\LT{the paper is perhaps too modest: (...) I don't think DK annotatability says anything about evidence---a synthesiser that answered "no" could still enjoy annotatability (even if we would not enjoy that answer).}
 Defined in \cref{fig:annotation-ordering}, the `annotation ordering' $t' \sqsupseteq t$ means that $t'$~has the same or more annotations than~$t$.
@@
 The interesting case is $\SynRule{Missing}$, which is mapped to $\SynRule{Anno}$, adding to the term a type annotation that comes from the given typing derivation.
-
 \begin{figure}

### 2023-12-25 tex/ESOP24/S5-synthesis.tex
[ tex ] Squeeze more space

@@
   \end{align*}
-  where $\top$ means an empty list is always mode-correct.
+  where $\MCas( \cdot ) = \top$ means that an empty list is always mode-correct.
 \end{defn}
@@
           \begin{enumerate}
-            \item If there is no $\sigma \geq \rho$ such that \eqref{eq:claim} holds but $\Gamma |- \tmOpts \syn A$ for some $A$, then by inversion we have a substitution $\rho\colon \Sub{\Xi}{\emptyset}$ such that
+            \item If there is no $\sigma \geq \rho$ such that \eqref{eq:claim} holds but $\Gamma |- \tmOpts \syn A$ for some $A$, then by inversion we have $\rho\colon \Sub{\Xi}{\emptyset}$ such that
               \[
@@
             \item If there is no $\sigma \geq \rho_0$ such that $\simsub{A_0}{\sigma} = A$ but $\Gamma |- \tmOpts \chk A$, then by inversion there is a substitution $\rho$ such that $A = \simsub{A_0}{\rho}$, thus a contradiction.
-            \item If there is a minimal $\ext{\rho} \geq \rho_0$ such that $\simsub{A_0}{\ext{\rho}} = A$, then apply our claim with $\ext{\rho}$:
+            \item If there is a minimal $\ext{\rho} \geq \rho_0$ s.t.\ $\simsub{A_0}{\ext{\rho}} = A$, then apply our claim with $\ext{\rho}$:
               \begin{enumerate}
-                \item If there is no $\sigma \geq \ext{\rho}$ satisfying \eqref{eq:claim} but $\Gamma |- \tmOpts \chk A$, then by inversion there is $\gamma$ such that $\simsub{A_0}{\gamma} = A$ and also ${\Gamma, \vars{x}_\isTerm{i} : \simsub{\Delta_i}{\gamma} |- \isTerm{t_i} \colon \simsub{A_i}{\gamma}^{\dir{d_i}}}$ for every $i$.
-                  Given that $\ext{\rho} \geq \rho$ is minimal such that $\simsub{A_0}{\ext{\rho}} = A$, then $\gamma$ is an extension of $\ext{\rho}$ but by assumption no such an extension satisfying ${\Gamma, \vars{x}_\isTerm{i} : \simsub{\Delta_i}{\gamma} |- \isTerm{t_i} \colon \simsub{A_i}{\gamma}^{\dir{d_i}}}$ exists, thus a contradiction.
+                \item If there is no $\sigma \geq \ext{\rho}$ satisfying \eqref{eq:claim} but $\Gamma |- \tmOpts \chk A$, then by inversion there is $\gamma$ s.t.\ $\simsub{A_0}{\gamma} = A$ and ${\Gamma, \vars{x}_\isTerm{i} : \simsub{\Delta_i}{\gamma} |- \isTerm{t_i} \colon \simsub{A_i}{\gamma}^{\dir{d_i}}}$ for every $i$.
+                  Given that $\ext{\rho} \geq \rho$ is minimal s.t.\ $\simsub{A_0}{\ext{\rho}} = A$, then $\gamma$ is an extension of $\ext{\rho}$ but by assumption no such an extension satisfying ${\Gamma, \vars{x}_\isTerm{i} : \simsub{\Delta_i}{\gamma} |- \isTerm{t_i} \colon \simsub{A_i}{\gamma}^{\dir{d_i}}}$ exists, thus a contradiction.
 
@@
             \begin{enumerate}
-              \item If there is no $\sigma \geq \rho$ such that \eqref{eq:claim} holds for all $1 \leq i \leq m$ but a minimal $\gamma \geq \rho$ such that~\eqref{eq:claim} holds for all $1 \leq i \leq m + 1$, then we clearly have a contradiction.
+              \item If there is no $\sigma \geq \rho$ s.t.\ \eqref{eq:claim} holds for all $1 \leq i \leq m$ but a minimal $\gamma \geq \rho$ such that~\eqref{eq:claim} holds for all $1 \leq i \leq m + 1$, then we have a contradiction.
               \item There is a minimal $\ext{\rho} \geq \rho$ s.t.\ \eqref{eq:claim} holds for $1 \leq i \leq m$.
@@
                   \item $\dir{d_{m+1}}$ is $\chk$: By mode-correctness, $\simsub{\Delta_{m+1}}{\ext{\rho}}$ and $\simsub{A_{m+1}}{\ext{\rho}}$ are defined.
-                    By the induction hypothesis $ \Gamma, \vars{x}_\isTerm{m+1} : \simsub{\Delta_{m+1}}{\ext{\rho}} |- \isTerm{t_{m+1}} \chk \simsub{A_{m+1}}{\ext{\rho}}$ is decidable.
-                    Clearly, if $\Gamma, \vars{x}_\isTerm{m+1} : \simsub{\Delta_{m+1}}{\ext{\rho}} |- \isTerm{t_{m+1}} \chk \simsub{A_{m+1}}{\ext{\rho}}$ then the desired statement is proved; otherwise we can also easily derive a contradiction.
+                    By the ind.\ hyp.\ $ \Gamma, \vars{x}_\isTerm{m+1} : \simsub{\Delta_{m+1}}{\ext{\rho}} |- \isTerm{t_{m+1}} \chk \simsub{A_{m+1}}{\ext{\rho}}$ is decidable.
+                    Clearly, if $\Gamma, \vars{x}_\isTerm{m+1} : \simsub{\Delta_{m+1}}{\ext{\rho}} |- \isTerm{t_{m+1}} \chk \simsub{A_{m+1}}{\ext{\rho}}$ then the desired statement is proved; otherwise we easily derive a contradiction.
 
                   \item $\dir{d_{m+1}}$ is $\syn$: By mode-correctness, $\simsub{\Delta_{m+1}}{\ext{\rho}}$ is defined.
-                    By the induction hypothesis, it is decidable that $\Gamma, \vars{x}_\isTerm{m+1} : \simsub{\Delta_{m+1}}{\ext{\rho}} |- \isTerm{t_{m+1}} \syn A$ for some $A$.
+                    By the ind.\ hyp., `$\Gamma, \vars{x}_\isTerm{m+1} : \simsub{\Delta_{m+1}}{\ext{\rho}} |- \isTerm{t_{m+1}} \syn A$ for some $A$' is decidable:
                     \begin{enumerate}
@@
 The formal counterpart of the above proof in \Agda functions as two top-level programs for type checking and synthesis.
-These programs either compute the typing derivation or provide a proof of contradiction.
-Each case analysis simply branches depending on the outcomes of bidirectional type synthesis and checking for each sub-term, as well as the unification process.
-If a negation proof is not of interest, these programs can be simplified by discarding the cases that yield such negation proofs.
-Alternatively, we could consider generalising typing derivations instead, like our generalised mode derivations (\Cref{fig:generalised-mode-derivations}).
-This could accommodate ill-typed cases and reformulate contradiction proofs positively to deliver more informative error messages.
-This would assist programmers in resolving issues with ill-typed terms, rather than simply returning a blatant `no'.
+These programs provide either the typing derivation or its negation proof.
+Each case analysis branches depending on the outcomes of bidirectional type synthesis and checking for each sub-term, as well as the unification process.
+If a negation proof is not of interest in practice, these programs can be simplified by discarding the cases that yield negation proofs.
+Alternatively, we could consider generalising typing derivations instead, like our generalised mode derivations (\Cref{fig:generalised-mode-derivations}), to reformulate negation proofs positively to deliver more informative error messages.
+This would assist programmers in resolving issues with ill-typed terms, rather than returning a blatant `no'.
 

### 2023-12-25 tex/ESOP24/S3-defs.tex
[ tex ] Eliminate most of overfull warnings

@@
     \item a derivation of $\Gamma |-_{\Sigma, \Omega} (t \annotate A) : A$ if $\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A$ has a derivation, and
-    \item a derivation of $\Gamma |-_{\Sigma, \Omega} \tmOpts : \simsub{A_0}{\rho}$ for $\bop$ if there is $\rho\colon \Xi \to \Type_{\Sigma}(\emptyset)$ and a derivation of $\Gamma, \vars{x}_{\isTerm{i}} : \simsub{\Delta_i}{\rho} |-_{\Sigma, \Omega} \isTerm{t_i} : \simsub{A_i}{\rho}$ for each $i$,
+    \item a derivation of $\Gamma |-_{\Sigma, \Omega} \tmOpts : \simsub{A_0}{\rho}$ for an operation $\bop$ if there are $\rho\colon \Xi \to \Type_{\Sigma}(\emptyset)$ and a derivation of $\Gamma, \vars{x}_{\isTerm{i}} : \simsub{\Delta_i}{\rho} |-_{\Sigma, \Omega} \isTerm{t_i} : \simsub{A_i}{\rho}$ for each $i$,
   \end{enumerate}
@@
   \begin{mathpar}
+%    \mprset{sep=0.5em}
     \inferrule{(\isTerm{x} : A) \in \Gamma}{\Gamma |-_{\Sigma, \Omega} \isTerm{x} : A}\,\Rule{Var}
@@
     \and
-    \inferrule{\rho : \Sub{\Xi}{\emptyset} \and  \Gamma, \vec{\isTerm{x}}_\isTerm{1} : \simsub{\Delta_{1}}{\rho} |-_{\Sigma, \Omega} \isTerm{t_1} : \simsub{A_{1}}{\rho} \and \cdots\and \Gamma, \vec{\isTerm{x}}_\isTerm{n} : \simsub{\Delta_{n}}{\rho} |-_{\Sigma, \Omega} \isTerm{t_n} : \simsub{A_{n}}{\rho}}
+    \inferrule{\rho : \Sub{\Xi}{\emptyset} \and  \Gamma, \vec{\isTerm{x}}_\isTerm{1} : \simsub{\Delta_{1}}{\rho} |-_{\Sigma, \Omega} \isTerm{t_1} : \simsub{A_{1}}{\rho} \; \cdots \; \Gamma, \vec{\isTerm{x}}_\isTerm{n} : \simsub{\Delta_{n}}{\rho} |-_{\Sigma, \Omega} \isTerm{t_n} : \simsub{A_{n}}{\rho}}
     {\Gamma |-_{\Sigma, \Omega} \tmOpts : \simsub{A_0}{\rho}}\,\Rule{Op}
@@
     \\
-    \inferrule{\rho\colon \Sub{\Xi}{\emptyset} \and \Gamma, \simsub{\vars{x}_\isTerm{1} : \Delta_{1}}{\rho} |-_{\Sigma, \Omega} \isTerm{t_{1}} :^\dir{d_1} \simsub{A_1}{\rho} \and
-      \cdots \and
-    \Gamma, \simsub{\vars{x}_\isTerm{n} : \Delta_{n}}{\rho} |-_{\Sigma, \Omega} \isTerm{t_n} :^{\dir{d_n}} \simsub{A_{n}}{\rho}}
+    \inferrule{\rho\colon \Sub{\Xi}{\emptyset} \\ \Gamma, \simsub{\vars{x}_\isTerm{1} : \Delta_{1}}{\rho} |-_{\Sigma, \Omega} \isTerm{t_{1}} \mathrel{:^\dir{d_1}} \simsub{A_1}{\rho} \and \cdots \and \Gamma, \simsub{\vars{x}_\isTerm{n} : \Delta_{n}}{\rho} |-_{\Sigma, \Omega} \isTerm{t_n} \mathrel{:^{\dir{d_n}}} \simsub{A_{n}}{\rho}}
     {\Gamma |-_{\Sigma, \Omega} \tmOpts :^{\dir{d}} \simsub{A_0}{\rho}} \,\Rule{Op}

### 2023-12-25 tex/ESOP24/S5-synthesis.tex
[ tex ] Eliminate most of overfull warnings

@@
   The \emph{mode-correctness} $\MCas\left(\biargvec\right)$ for an argument list $\biargs$ with respect to a subset $S$ of $\Xi$ is a predicate defined by
+  {\small
   \begin{align*}
@@
                                   & = \phantom{, A_n} \fv(\Delta_n) \subseteq \left( S \cup \synvar\left(\biargvec\right)\right) \land  \MCas\left(\biargvec\right)
-  \end{align*}
+  \end{align*}}
   where $\MCas( \cdot ) = \top$ means that an empty list is always mode-correct.
@@
 \begin{lemma}\label{lem:decidability-mode-correctness}
-  It is decidable whether an operation $\biop$ is mode-correct.
+  For any operation $\biop$, it is decidable whether $o$ is mode-correct.
 \end{lemma}
@@
           \begin{enumerate}
-            \item If there is no $\sigma \geq \rho_0$ such that $\simsub{A_0}{\sigma} = A$ but $\Gamma |- \tmOpts \chk A$, then by inversion there is a substitution $\rho$ such that $A = \simsub{A_0}{\rho}$, thus a contradiction.
+            \item If there is no $\sigma \geq \rho_0$ s.t.\ $\simsub{A_0}{\sigma} = A$ but $\Gamma |- \tmOpts \chk A$, then by inversion there is a substitution $\rho$ such that $A = \simsub{A_0}{\rho}$, thus a contradiction.
             \item If there is a minimal $\ext{\rho} \geq \rho_0$ s.t.\ $\simsub{A_0}{\ext{\rho}} = A$, then apply our claim with $\ext{\rho}$:
               \begin{enumerate}
-                \item If there is no $\sigma \geq \ext{\rho}$ satisfying \eqref{eq:claim} but $\Gamma |- \tmOpts \chk A$, then by inversion there is $\gamma$ s.t.\ $\simsub{A_0}{\gamma} = A$ and ${\Gamma, \vars{x}_\isTerm{i} : \simsub{\Delta_i}{\gamma} |- \isTerm{t_i} \colon \simsub{A_i}{\gamma}^{\dir{d_i}}}$ for every $i$.
+                \item If there is no $\sigma \geq \ext{\rho}$ satisfying \eqref{eq:claim} but $\Gamma |- \tmOpts \chk A$, then by inversion there is $\gamma$ s.t.\ $\simsub{A_0}{\gamma} = A$ and, for every $i$, ${\Gamma, \vars{x}_\isTerm{i} : \simsub{\Delta_i}{\gamma} |- \isTerm{t_i} \colon \simsub{A_i}{\gamma}^{\dir{d_i}}}$.
                   Given that $\ext{\rho} \geq \rho$ is minimal s.t.\ $\simsub{A_0}{\ext{\rho}} = A$, then $\gamma$ is an extension of $\ext{\rho}$ but by assumption no such an extension satisfying ${\Gamma, \vars{x}_\isTerm{i} : \simsub{\Delta_i}{\gamma} |- \isTerm{t_i} \colon \simsub{A_i}{\gamma}^{\dir{d_i}}}$ exists, thus a contradiction.
@@
                         Therefore $\simsub{\Delta_{m+1}}{\ext{\rho}} = \simsub{\Delta_{m+1}}{\gamma}$ and we derive a contradiction because $\Gamma, \vars{x}_\isTerm{m+1} : \simsub{\Delta_{m+1}}{\ext{\rho}} |- \isTerm{t_{m+1}} \syn \simsub{A_{m+1}}{\gamma}$.
-                      \item If $\Gamma, \vars{x}_\isTerm{m+1} : \simsub{\Delta_{m+1}}{\ext{\rho}} |- \isTerm{t_{m+1}} \syn A$ for some $A$, then by \Cref{lem:unify}: % we have two cases: %we unify $A$ with $A_{m+1}$ extending $\ext{\rho}$.
+                      \item If $\Gamma, \vars{x}_\isTerm{m+1} : \simsub{\Delta_{m+1}}{\ext{\rho}} |- \isTerm{t_{m+1}} \syn A$ for some $A$, then \Cref{lem:unify} gives following two cases:
                         \begin{itemize}
-                          \item Suppose no $\sigma \geq \ext{\rho}$ such that $\simsub{A_{m+1}}{\sigma} = A$ but an extension $\gamma \geq \rho$ such that \eqref{eq:claim} holds for $1 \leq i \leq m + 1$.
+                          \item Suppose no $\sigma \geq \ext{\rho}$ s.t.\ $\simsub{A_{m+1}}{\sigma} = A$ but an extension $\gamma \geq \rho$ s.t.\ \eqref{eq:claim} holds for $1 \leq i \leq m + 1$. 
                             Then, $\gamma \geq \ext{\rho}$ by the minimality of $\ext{\rho}$ and thus
                             $\Gamma, \vars{x}_\isTerm{m+1} : \simsub{\Delta_{m+1}}{\ext{\rho}} |- \isTerm{t_{m+1}} \syn \simsub{A_{m+1}}{\gamma}$.
-                            However, by \Cref{thm:unique-syn}, the synthesised type $\simsub{A_{m+1}}{\gamma}$ must be unique, so $\gamma$ is an extension of $\ext{\rho}$ such that $\simsub{A_{m+1}}{\gamma} = A$, leading to a contradiction.
+                            However, by \Cref{thm:unique-syn}, the synthesised type $\simsub{A_{m+1}}{\gamma}$ must be unique, so $\gamma$ is an extension of $\ext{\rho}$ s.t.\ $\simsub{A_{m+1}}{\gamma} = A$, i.e.\ a contradiction.
                           \item If there is a minimal $\ext{\ext{\rho}} \geq \ext{\rho}$ such that $\simsub{A_{m+1}}{\ext{\ext{\rho}}} = A$, then it is not hard to show that $\ext{\ext{\rho}}$ is also the minimal extension of $\rho$ such that \eqref{eq:claim} holds for all $1 \leq i \leq m + 1$.

### 2023-12-25 tex/ESOP24/S8-future.tex
[ tex ] sort => type
[ tex ] Add Acknowledgement

@@
 In the case of languages like \SystemF and others that permit type-level variable binding, we can start with the notion of polymorphic signature, as introduced by \citet{Hamana2011}---\begin{inlineenum}
-  \item each type construct in a signature is specified by a binding arity with a single sort $*$, and
+  \item each type construct in a signature is specified by a binding arity with only one type $*$, and
   \item a term construct can employ a pair of extension contexts for term variables and type variables.
@@
 
-%Such aspects  may continue to require creativity from the designer of type systems whereas its implementation could be automated entirely.
-
-%\subsection*{Acknowledgement}
-%We thank Nathanael Arkor for the useful conversation.
-%James Wood
-%We thank Kuen-Bang {Hou (Favonia)} for his comments and suggestions.
+\begin{credits}
+\subsubsection{Acknowledgements.}
+We thank Kuen-Bang {Hou (Favonia)} and anonymous reviewers for their comments and suggestions.
+\end{credits}

### 2023-12-25 tex/ESOP24/S2-outline.tex
[ tex ] Minor changes

@@
 Correspondingly, the definition of typing derivations%
-\footnote{We write~`$\bto$' instead of~`$\to$' for the function types of simply typed $\lambda$-calculus to avoid confusion with the function types in our type-theoretic meta-language.}
+\footnote{We write~`$\bto$' instead of~`$\to$' for the function types of the simply typed $\lambda$-calculus to avoid confusion with the function types in our type-theoretic meta-language.}
 in \cref{fig:STLC-typing-derivations} also includes an \textsc{Anno} rule enforcing that the type of an annotated term does match the annotation.

### 2023-12-25 tex/ESOP24/S5-synthesis.tex
[ tex ] Minor changes

@@
 
-It is easy to check the bidirectional type system $(\Sigma_{\bto}, \Omega^{\Leftrightarrow}_{\Lambda}$) for simply typed $\lambda$-calculus is mode-correct by definition or by the following lemma:
+It is easy to check the bidirectional type system $(\Sigma_{\bto}, \Omega^{\Leftrightarrow}_{\Lambda}$) for the simply typed $\lambda$-calculus is mode-correct by definition or by the following lemma:
 \begin{lemma}\label{lem:decidability-mode-correctness}
@@
           \begin{enumerate}
-            \item If there is no $\sigma \geq \rho_0$ s.t.\ $\simsub{A_0}{\sigma} = A$ but $\Gamma |- \tmOpts \chk A$, then by inversion there is a substitution $\rho$ such that $A = \simsub{A_0}{\rho}$, thus a contradiction.
+            \item If there is no $\sigma \geq \rho_0$ s.t.\ $\simsub{A_0}{\sigma} = A$ but $\Gamma |- \tmOpts \chk A$, then inversion gives us a substitution $\rho$ s.t.\ $A = \simsub{A_0}{\rho}$---a contradiction.
             \item If there is a minimal $\ext{\rho} \geq \rho_0$ s.t.\ $\simsub{A_0}{\ext{\rho}} = A$, then apply our claim with $\ext{\rho}$:

### 2023-12-25 tex/ESOP24/S6-example.tex
[ tex ] Eliminate most of overfull warnings
[ tex ] Minor changes

@@
 At first glance, accommodating this form of application may seem impossible, given that the number of arguments for a construct is finite and has to be fixed.
-Nonetheless, the total number of operation symbols a signature can have need not be finite, allowing us to establish a corresponding construct for each number $n$ of arguments, i.e.\ viewing the following rule
+Nonetheless, the total number of operation symbols in a signature need not be finite, allowing us to establish a corresponding construct for each number $n$ of arguments, i.e.\ viewing the following rule
 \bgroup
@@
   \begin{mathpar}
-    \inferrule{\Gamma \vdash t \syn A_1 \bto \left(A_2 \bto \left(\dots \bto \left(A_n \bto B\right)\ldots\right)\right) \and \Gamma \vdash u_1 \chk A_1 \and \dots \and \Gamma \vdash u_n \chk A_n}{\Gamma \vdash t\;u_1\;\ldots\;u_n \syn B}
+    \inferrule{\Gamma \vdash t \syn A_1 \bto \left(A_2 \bto \left(\dots \bto \left(A_n \bto B\right)\ldots\right)\right) \and \Gamma \vdash u_1 \chk A_1 \and \!\!\cdots\!\!
+\and \Gamma \vdash u_n \chk A_n}{\Gamma \vdash t\;u_1\;\ldots\;u_n \syn B}
   \end{mathpar}
@@
 On the other hand, similar to a parser generator, a type-synthesiser generator only needs a specification $(\Sigma, \Omega)$ from the user to produce a corresponding synthesiser. 
-In more detail, such a specification can be derived by extending $\Sigma_{\bto}$ with additional type constructs with $\mathsf{nat}$, $\mathsf{prod}$, $\mathsf{sum}$, and $\mathsf{T}$ such that
+Such a specification can be derived by extending $\Sigma_{\bto}$ with additional type constructs with $\mathsf{nat}$, $\mathsf{prod}$, $\mathsf{sum}$, and $\mathsf{T}$ such that
 \[

### 2023-12-25 tex/ESOP24/S8-future.tex
[ tex ] Expand the discussion on syntax-directed systems

@@
 
-\subsubsection{Beyond syntax-directed systems}
+\subsubsection{Beyond syntax-directedness}
 
-For non-syntax-directed systems, since there can be multiple moding or typing rules for a language construct, both the mode decorator and the bidirectional type synthesiser will need to backtrack, and the decorator may need to produce all possible mode derivations so as not to limit the synthesiser's choices.
-For soundness and completeness, it should still be possible to treat them as the separation and combination of mode and typing derivations, although bidirectional typing rules and pairs of a moding rule and a typing rule may no longer be in perfect correspondence.
-Completeness may need to be reformulated to take all possible mode derivations as input to guarantee that at least one can be paired with the input typing derivation.
+%For non-syntax-directed systems, since there can be multiple moding or typing rules for a language construct, both the mode decorator and the bidirectional type synthesiser will need to backtrack, and the decorator may need to produce all possible mode derivations so as not to limit the synthesiser's choices.
+%For soundness and completeness, it should still be possible to treat them as the separation and combination of mode and typing derivations, although bidirectional typing rules and pairs of a moding rule and a typing rule may no longer be in perfect correspondence.
+%Completeness may need to be reformulated to take all possible mode derivations as input to guarantee that at least one can be paired with the input typing derivation.
+
+To explore general settings, a possibility is to independently specify an ordinary and a bidirectional type system over the same raw terms and then investigate relationships between the two systems.
+
+However, a practical setting suitable for a next step is probably this: for each raw term construct, there can be several ordinary typing rules, and each typing rule can be refined to several bidirectional typing rules with different mode assignments and different orders of premises.
+For this setting, both the mode decorator and the bidirectional type synthesiser will have to backtrack.
+For soundness and completeness, it should still be possible to treat them as the separation and combination of mode and type information, but the completeness direction will pose a problem---for every node of a raw term, a mode derivation chooses a mode assignment while a typing derivation chooses a typing rule, but there may not be a bidirectional typing rule for this particular combination.
+It should be possible to fix this problem while retaining the mode decoration phase.
+For example, one idea is to make the mode decorator produce all possible mode derivations, and refine completeness to say that any typing derivation can be combined with one of mode derivations into a bidirectional typing derivation.
+
+We could instead focus on simpler cases first where each construct has exactly one typing rule, but each typing rule can be refined to more than one bidirectional typing rule.
+In such cases, the mode decorator would need to find all mode derivations, but the type synthesiser should still work in a syntax-directed manner on each mode derivation as input.
+Completeness could still take the simple form presented in this paper too.
 

### 2024-01-04 tex/ESOP24/S1-intro.tex
[ tex ] Add the citation for the artefact on Zenodo

@@
 \end{enumerate}
-Our theory is fully formally developed with \Agda but translated to the mathematical vernacular for presentation in this paper.
+Our theory is fully formally developed~\cite{Chen2024a} with \Agda but translated to the mathematical vernacular for presentation in this paper.
 The formal theory doubles as a verified implementation, with no gap in between.

### 2024-01-05 tex/ESOP24/S8-future.tex
[ tex ] Add the citation for Ahrens et al.'s work

@@
 For the second group, theories developed in set theory can often be reformulated in type theory and thus manageable for formalisation in a type-theoretic proof assistant. 
-Such examples include frameworks developed by \citet{Allais2021,Fiore2022} in \Agda, which are relatively correct to their meta-language, although their expressiveness is limited to simply typed theories.
-Our work belongs to the second group, as we aim for a formalism in a constructive type theory to minimise the gap between theory and implementation.
+Such examples include frameworks developed by \citet{Allais2021,Ahrens2022,Fiore2022} in \Agda, which are relatively correct to their meta-language, although their expressiveness is limited to simply typed theories.
+Our work belongs to the second group, as we aim for a formalism in a type theory to minimise the gap between theory and implementation.
 

### 2024-01-08 tex/ESOP24/BiSig.tex
[ tex ] Minor
[ tex ] Use `\rmfamily' for the body of a definition
[ tex ] Squeeze more space
[ tex ] Rewrite the example for binding arity with colourful annotations

@@
 \usepackage{mathpartir}
+\usepackage{annotate-equations}
 
@@
 \SetSymbolFont{stmry}{bold}{U}{stmry}{m}{n}
+
 \newcommand{\LT}[2][]{\todo[author=LT,color=orange!40,#1]{#2}}
 \newcommand{\Josh}[2][]{\todo[author=Josh,color=yellow!40,#1]{#2}}
+
 \newcommand{\codefigure}{\small\setlength{\parindent}{0em}\setlength{\mathindent}{0em}\setlength{\abovedisplayskip}{0ex}\setlength{\belowdisplayskip}{0ex}}
@@
 
-%\let\theorem\relax
-%\spnewtheorem{theorem}{Theorem}[section]{\bfseries}{}
-%\let\definition\relax
-%\spnewtheorem{definition}{Definition}[section]{\bfseries}{}
+\spnewtheorem{defn}[theorem]{Definition}{\bfseries}{\rmfamily}
+\crefformat{defn}{Definition~#2#1#3}
 
-\author{Liang-Ting Chen
-\orcidID{0000-0002-3250-1331}
-\and 
-Hsiang-Shang Ko
-\orcidID{0000-0002-2439-1048}}
-
+\author{Liang-Ting Chen\orcidID{0000-0002-3250-1331} \and Hsiang-Shang Ko\orcidID{0000-0002-2439-1048}}
 \institute{Institute of Information Science, Academia Sinica, Taiwan}
-
 \title{A formal treatment of bidirectional typing\texorpdfstring{\thanks{The work is supported by the National Science and Technology Council of Taiwan under grant NSTC 112-2221-E-001-003-MY3.}}{}}
@@
 To remedy the situation, this paper develops a \emph{general} and \emph{formal} theory of bidirectional typing and as a by-product of our formalism provides a \emph{verified} generator of \emph{proof-relevant} type synthesisers for simply typed languages: for every signature that specifies a mode-correct bidirectionally typed language, there exists a proof-relevant type synthesiser that for an input abstract syntax tree constructs a typing derivation if any, gives its refutation if not, or reports that the input does not have enough type annotations.
-Soundness, completeness, and mode-correctness are studied universally for all signatures, which are sufficient conditions for deriving a type synthesiser.
+Sufficient conditions for deriving a type synthesiser such as soundness, completeness, and mode-correctness, are studied universally for all signatures.
 We propose a preprocessing step called \emph{mode decoration}, which helps the user to deal with missing type annotations in a given abstract syntax tree.
@@
 \bibliographystyle{splncs04nat}
-
-{\bibliography{ref}
+\bibliography{ref}
 

### 2024-01-08 tex/ESOP24/S1-intro.tex
[ tex ] Rewrite the example for binding arity with colourful annotations

@@
 Bidirectional type synthesis combines type \emph{inference} on this subset of terms with type \emph{checking} (based on a given type) on the rest.
-More formally, every judgement in a bidirectional type system is extended with a \emph{mode:}
+Formally, every judgement in a bidirectional type system is extended with a \emph{mode:}
 \begin{inlineenum}
@@
 The former indicates that the type~$A$ is an output, using both the context~$\Gamma$ and the term~$t$ as input, while for the latter, all three of $\Gamma$, $t$, and~$A$ are given as input.
-The algorithm of a bidirectional type synthesiser can usually be `read off' from a well-designed bidirectional type system: as the synthesiser traverses a raw term, it switches between synthesis and checking, following the modes assigned to the judgements in the typing rules.
+The algorithm of a bidirectional type synthesiser can be `read off' from a well-designed bidirectional type system: as the synthesiser traverses a raw term, it switches between synthesis and checking, following the modes assigned to the judgements in the typing rules.
 %if a condition called \emph{mode-correctness} is satisfied.
@@
 Recall that the law of excluded middle $P + \neg P$ does not hold as an axiom for every~$P$ constructively, and we say that $P$~is logically \emph{decidable} if the law holds for~$P$.
-Since Martin-L\"of type theory is simultaneously logical and computational, a decidability proof is a proof-relevant decision procedure that computes not only a yes-or-no answer but also a proof of~$P$ or its refutation, so logical decidability is algorithmic decidability.
+Since Martin-L\"of type theory is logical and computational, a decidability proof is a proof-relevant decision procedure that computes a yes-or-no answer with a proof of~$P$ or its refutation, so logical decidability is algorithmic decidability.
 %For example, a proof of \emph{decidable equality}, i.e.~$\forall x, y.~(x = y) + (x \neq y)$, decides whether $x$~and~$y$ are equal and accordingly gives an identity proof or a refutation explicitly.
@@
 A proof of this statement would also be a program that produces either a typing derivation for the given raw term~$\isTerm{t}$ or a negation proof that such a derivation is impossible.
-The first case is algorithmic soundness, while the second case is algorithmic completeness in contrapositive form (which implies the original form due to the decidability).
-Hence, proving the statement is the same as constructing a verified proof-relevant type inference algorithm, which returns not only an answer but also a proof justifying that answer.
-This is an economic way to bridge the gap between theory and practice, where proofs double as verified programs, in contrast to separately exhibiting a theory and an implementation that are only loosely related.
+The first case is algorithmic soundness, while the second case is algorithmic completeness in contrapositive form (implying the original form due to the decidability).
+Hence, proving the statement is the same as constructing a verified proof-relevant type inference algorithm, which returns not only an answer but also its justification.
+This is an economic way to bridge the gap between theory and practice, where proofs double as verified programs, in contrast to separately exhibiting a theory and an implementation that are loosely related.
 

### 2024-01-08 tex/ESOP24/S3-defs.tex
[ tex ] Rewrite the example for binding arity with colourful annotations

@@
   A \emph{binding arity} with a set $T$ of types is an inhabitant of $\left(T^* \times T\right)^* \times T$ where $T^*$ is the set of lists over $T$.
-  In a binding arity $(((\Delta_1, A_1), \ldots, (\Delta_n, A_n)), A)$, every $\Delta_i$ and $A_i$ refers to the \emph{extension context} and the \emph{type of the $i$-th argument}, respectively, and $A$ the target type.
+  In a binding arity $(((\Delta_1, A_1), \ldots, (\Delta_n, A_n)), A)$, every $\Delta_i$ and $A_i$ refers to the \emph{extension context} and the \emph{type of the $i$-th argument}, respectively, and $A$ the \emph{target type}.
   For brevity, it is denoted by $\bargs \to A$ where $[\Delta_i]$ is omitted if empty.
@@
 \begin{example}
-Consider the $\Rule{Abs}$ rule (\Cref{fig:STLC-typing-derivations}).
-Its binding arity has $\Type_{\Sigma_{\bto}}\{A, B\}$ as types and is ${[A]B} \to {(A \bto B)}$.
-This means that the $\Rule{Abs}$ rule for derivations of\/ $\Gamma |- \lam{x}t : A \bto B$ contains:
-\begin{inlineenum}
-  \item a derivation of\/ $\Gamma, x : A |- t : B$ as an argument of type $B$ in a typing context extended by a variable~$x$ of type~$A$;
-  \item the type $A\bto B$ for itself.
-\end{inlineenum}
-In a similar vein, the arity of the $\Rule{App}$ rule is denoted as ${(A \bto B), A} \to {B}$.
-This specifies that any derivation of\/ $\Gamma \vdash t\;u : B$ accepts derivations $\Gamma \vdash t : A \bto B$ and $\Gamma \vdash u : A$ as its arguments.
-These arguments have types $A \bto B$ and $A$ respectively and their extension contexts are empty.
+  Observe that $\Rule{Abs}$ and $\Rule{App}$ rules in \Cref{fig:STLC-typing-derivations} can be read as
+  \renewcommand{\eqnhighlightheight}{\vphantom{\Gamma}}
+  \renewcommand{\eqnannotationfont}{\rmfamily\footnotesize} % package default
+  \vspace{2em}
+  \begin{equation*}
+    \frac{\Gamma, \eqnmarkbox[blue]{ext1}{\isTerm{x} : A, \cdot} \vdash \isTerm{t} : \eqnmarkbox[red]{aty1}{B}}{\Gamma \vdash \isTerm{\lam{x}t} : \eqnmarkbox{ty1}{A \bto B}}
+    \qquad\text{and}\qquad
+    \frac{\Gamma, \eqnmarkbox[blue]{ext2}{\cdot} \vdash \isTerm{t} : \eqnmarkbox[red]{aty2}{A \bto B} \qquad \Gamma, \eqnmarkbox[blue]{ext3}{\cdot} \vdash \isTerm{u} : \eqnmarkbox[red]{aty3}{A}}{\Gamma \vdash \isTerm{t\;u} : \eqnmarkbox{ty2}{B}}
+  \end{equation*}
+  \annotate[yshift=1.5em]{above}{ext1,ext2,ext3}{extension context}
+  \annotate[yshift=0.4em]{above,right}{aty1,aty2,aty3}{argument type}
+  \annotate[yshift=0em]{below}{ty1,ty2}{target type}
+  \vspace{.5em}
+  \\
+  if the empty extension context $\cdot$ is added verbosely, so they can be specified by binding arities ${[A]B} \to {(A \bto B)}$ and ${(A \bto B), A} \to {B}$ respectively.
 \end{example}
+
+%\begin{example}
+%Consider the $\Rule{Abs}$ rule (\Cref{fig:STLC-typing-derivations}).
+%Its binding arity has $\Type_{\Sigma_{\bto}}\{A, B\}$ as types and is ${[A]B} \to {(A \bto B)}$.
+%This means that the $\Rule{Abs}$ rule for derivations of\/ $\Gamma |- \lam{x}t : A \bto B$ contains:
+%\begin{inlineenum}
+%  \item a derivation of\/ $\Gamma, x : A |- t : B$ as an argument of type $B$ in a typing context extended by a variable~$x$ of type~$A$;
+%  \item the type $A\bto B$ for itself.
+%\end{inlineenum}
+%In a similar vein, the arity of the $\Rule{App}$ rule is denoted as ${(A \bto B), A} \to {B}$.
+%This specifies that any derivation of\/ $\Gamma \vdash t\;u : B$ accepts derivations $\Gamma \vdash t : A \bto B$ and $\Gamma \vdash u : A$ as its arguments.
+%These arguments have types $A \bto B$ and $A$ respectively and their extension contexts are empty.
+%\end{example}
 
@@
 \begin{defn}\label{def:binding-signature}
-  For a type signature $\Sigma$, a \emph{binding signature} $\Omega$ consists of a set $O$ and a function
+  For a type signature $\Sigma$, a \emph{binding signature} $\Omega$ is a set $O$ with a function
   \[
-    \arity(o) \colon O \to \sm{\Xi : \UU} \left(\Type_\Sigma(\Xi)^* \times \Type_\Sigma(\Xi)\right)^* \times \Type_\Sigma(\Xi).
+    \arity \colon O \to \sm{\Xi : \UU} \left(\Type_\Sigma(\Xi)^* \times \Type_\Sigma(\Xi)\right)^* \times \Type_\Sigma(\Xi).
   \]
@@
 
-The set~$\Xi$ of type variables for each operation, called its \emph{local context}, plays a somewhat important role:
-To use a rule like $\Rule{Abs}$ in an actual typing derivation, we will need to substitute \emph{concrete types}, i.e.\ types without any type variables, for variables $A, B$.
+The set~$\Xi$ of type variables for each operation, called its \emph{local context}, plays an important role.
+To use a rule like $\Rule{Abs}$ in an actual typing derivation, we need to substitute \emph{concrete types}, i.e.\ types without any type variables, for variables $A, B$.
 In our formulation of substitution~(\ref{def:substitution}), we must first identify which type variables to substitute for.
-As such, this information forms part of the arity of an operation, and typing derivations, defined subsequently, will include functions from~$\Xi$ to concrete types specifying how to instantiate typing rules by substitution.
+As such, this information forms part of the arity of an operation, and typing derivations, defined subsequently, will include functions $\rho$ from~$\Xi$ to concrete types specifying how to instantiate typing rules by substitution.
 
@@
     \and
-    \inferrule{\cdot |-_{\Sigma} A \\ V |-_{\Sigma, \Omega}\isTerm{t}}{V |-_{\Sigma, \Omega} \isTerm{t \annotate A}}\,\Rule{Anno}
+    \inferrule{\cdot |-_{\Sigma} A \\ V |-_{\Sigma, \Omega}\isTerm{t}}{V |-_{\Sigma, \Omega} \isTerm{t \annotatecolon A}}\,\Rule{Anno}
     \\
@@
     \item variables in $V$,
-    \item annotations $\isTerm{t \annotate A}$ for some raw term $t$ in $V$ and a type $A$, and
+    \item annotations $\isTerm{t \annotatecolon A}$ for some raw term $t$ in $V$ and a type $A$, and
     \item a construct $\tmOpts$ for some $\bop$ in~$O$, where $\vars{x}_{\isTerm{i}}$'s are lists of variables whose length is equal to the length of~$\Delta_i$, and $t_i$'s are raw terms in the variable list $V, \vars{x}_i$.
@@
     \item a derivation of $\Gamma |-_{\Sigma, \Omega} x : A$ if $x : A$ is in $\Gamma$,
-    \item a derivation of $\Gamma |-_{\Sigma, \Omega} (t \annotate A) : A$ if $\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A$ has a derivation, and
+    \item a derivation of $\Gamma |-_{\Sigma, \Omega} (t \annotatecolon A) : A$ if $\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A$ has a derivation, and
     \item a derivation of $\Gamma |-_{\Sigma, \Omega} \tmOpts : \simsub{A_0}{\rho}$ for an operation $\bop$ if there are $\rho\colon \Xi \to \Type_{\Sigma}(\emptyset)$ and a derivation of $\Gamma, \vars{x}_{\isTerm{i}} : \simsub{\Delta_i}{\rho} |-_{\Sigma, \Omega} \isTerm{t_i} : \simsub{A_i}{\rho}$ for each $i$,
@@
 \begin{example}
-Raw terms (\Cref{fig:STLC-raw-terms}) and typing derivations (\Cref{fig:STLC-typing-derivations}) for simply typed $\lambda$-calculus can be specified by the type signature $\Sigma_{\bto}$ (\Cref{ex:type-signature-for-function-type}) and the binding signature consisting of
- \[
-   \aritysymbol{\mathsf{app}}{A, B}{(A \bto B), A}{B}
-   \quad\text{and}\quad
-   \aritysymbol{\mathsf{abs}}{A , B}{[A]B}{(A \bto B)}.
- \]
+Raw terms (\Cref{fig:STLC-raw-terms}) and typing derivations (\Cref{fig:STLC-typing-derivations}) for simply typed $\lambda$-calculus can be specified by the type signature $\Sigma_{\bto}$ (\Cref{ex:type-signature-for-function-type}) and the binding signature consisting of $\aritysymbol{\mathsf{app}}{A, B}{(A \bto B), A}{B}$ and $\aritysymbol{\mathsf{abs}}{A , B}{[A]B}{(A \bto B)}$.
 Rules $\Rule{Abs}$ and $\Rule{App}$ in simply typed $\lambda$-calculus are subsumed by the $\Rule{Op}$ rule schema, as applications $t\;u$ and abstractions $\lam{x}t$ can be introduced uniformly as $\tmOp_{\mathsf{app}}(t, u)$ and $\tmOp_{\mathsf{abs}}(x.t)$, respectively.
@@
     \and
-    \inferrule{\Gamma |-_{\Sigma, \Omega} \isTerm{t} : A}{\Gamma |-_{\Sigma, \Omega} (\isTerm{t \annotate A}) : A}\,\Rule{Anno}
+    \inferrule{\Gamma |-_{\Sigma, \Omega} \isTerm{t} : A}{\Gamma |-_{\Sigma, \Omega} (\isTerm{t \annotatecolon A}) : A}\,\Rule{Anno}
     \and
@@
     \and
-    \inferrule{\Gamma |-_{\Sigma, \Omega} \isTerm{t} :^{\chk} A}{\Gamma |-_{\Sigma, \Omega} (\isTerm{t \annotate A}) :^{\syn}  A}\,\SynRule{Anno}
+    \inferrule{\Gamma |-_{\Sigma, \Omega} \isTerm{t} :^{\chk} A}{\Gamma |-_{\Sigma, \Omega} (\isTerm{t \annotatecolon A}) :^{\syn}  A}\,\SynRule{Anno}
     \and
@@
     \and
-    \inferrule{\cdot |-_{\Sigma} A \\ V |-_{\Sigma, \Omega}\isTerm{t}^{\chk}}{V |-_{\Sigma, \Omega} (\isTerm{t \annotate A})^{\syn}}\,\SynRule{Anno}
+    \inferrule{\cdot |-_{\Sigma} A \\ V |-_{\Sigma, \Omega}\isTerm{t}^{\chk}}{V |-_{\Sigma, \Omega} (\isTerm{t \annotatecolon A})^{\syn}}\,\SynRule{Anno}
     \and

### 2024-01-11 tex/ESOP24/BiSig.tex
[ tex ] Some revisions before removing appendix for the camera ready version

@@
 \institute{Institute of Information Science, Academia Sinica, Taiwan}
-\title{A formal treatment of bidirectional typing\texorpdfstring{\thanks{The work is supported by the National Science and Technology Council of Taiwan under grant NSTC 112-2221-E-001-003-MY3.}}{}}
+\title{A formal treatment of bidirectional typing}
 
@@
 %This situation is in stark contrast to parsing, for which there have been general theories and widely applicable tools such as parser generators.
-To remedy the situation, this paper develops a \emph{general} and \emph{formal} theory of bidirectional typing and as a by-product of our formalism provides a \emph{verified} generator of \emph{proof-relevant} type synthesisers for simply typed languages: for every signature that specifies a mode-correct bidirectionally typed language, there exists a proof-relevant type synthesiser that for an input abstract syntax tree constructs a typing derivation if any, gives its refutation if not, or reports that the input does not have enough type annotations.
+To remedy the situation, this paper develops a \emph{general} and \emph{formal} theory of bidirectional typing for simply typed languages: for every signature that specifies a mode-correct bidirectionally typed language, there exists a \emph{proof-relevant} type synthesiser that for an input abstract syntax tree constructs a typing derivation if any, gives its refutation if not, or reports that the input does not have enough type annotations.
 Sufficient conditions for deriving a type synthesiser such as soundness, completeness, and mode-correctness, are studied universally for all signatures.
-We propose a preprocessing step called \emph{mode decoration}, which helps the user to deal with missing type annotations in a given abstract syntax tree.
-The entire development is formalised in \Agda and can be further integrated with other language-formalisation frameworks.
+We propose a preprocessing step called \emph{mode decoration}, which helps the user to deal with missing type annotations in a raw term.
+The entire theory is formally implemented in \Agda, so as a by-product of our formalism we provide a \emph{verified} generator of proof-relevant type synthesisers.
 \end{abstract}

### 2024-01-11 tex/ESOP24/S1-intro.tex
[ tex ] Some revisions before removing appendix for the camera ready version

@@
 Here `type inference' refers specifically to algorithms that ascertain the type of any raw term \emph{without type annotations}.
-However, full parametric polymorphism leads to undecidability in type inference, as do dependent types~\cite{Wells1999,Dowek1993}.
-In light of these limitations, \emph{bidirectional type synthesis} emerged as a viable alternative, providing algorithms for deciding the types of raw terms that meet some syntactic criteria and usually contain type annotations.
-In their survey paper, \varcitet{Dunfield2021}{ summarised} the design principles of bidirectional type synthesis and its wide coverage of languages with simple types, polymorphic types, dependent types, gradual types, among others.
+However, full parametric polymorphism entails undecidability in type inference, as do dependent types~\cite{Wells1999,Dowek1993}.
+In light of these limitations, \emph{bidirectional type synthesis} emerged as a viable alternative, deciding the types of raw terms that meet some syntactic criteria and typically contain annotations.
+In their survey paper, \varcitet{Dunfield2021}{ summarised} the principles of bidirectional type synthesis and its wide coverage of languages with simple, polymorphic, dependent, and gradual types, among others.
 
@@
 The former indicates that the type~$A$ is an output, using both the context~$\Gamma$ and the term~$t$ as input, while for the latter, all three of $\Gamma$, $t$, and~$A$ are given as input.
-The algorithm of a bidirectional type synthesiser can be `read off' from a well-designed bidirectional type system: as the synthesiser traverses a raw term, it switches between synthesis and checking, following the modes assigned to the judgements in the typing rules.
+The algorithm of a bidirectional type synthesiser can often be `read off' from a well-designed bidirectional type system: as the synthesiser traverses a raw term, it switches between synthesis and checking, following the modes assigned to the judgements in the typing rules.
 %if a condition called \emph{mode-correctness} is satisfied.
@@
 A proof of this statement would also be a program that produces either a typing derivation for the given raw term~$\isTerm{t}$ or a negation proof that such a derivation is impossible.
-The first case is algorithmic soundness, while the second case is algorithmic completeness in contrapositive form (implying the original form due to the decidability).
-Hence, proving the statement is the same as constructing a verified proof-relevant type inference algorithm, which returns not only an answer but also its justification.
+The first case is algorithmic soundness, while the second case is algorithmic completeness in contrapositive form (which implies the original form due to the decidability).
+Therefore, proving the statement is the same as constructing a verified proof-relevant type inference algorithm, which returns not only an answer but also a proof justifying that answer.
 This is an economic way to bridge the gap between theory and practice, where proofs double as verified programs, in contrast to separately exhibiting a theory and an implementation that are loosely related.
@@
 Annotations therefore play an important role even in the definition of the problem solved by bidirectional typing, but have not received enough attention.
-In our theory, we define \emph{mode derivations} so that we can explicitly take annotations into account and formulate the type synthesis problem with sufficiently annotated raw terms.
+In our theory, we define \emph{mode derivations} to explicitly take annotations into account and formulate the type synthesis problem with sufficiently annotated raw terms.
 Accordingly, a preprocessing step called \emph{mode decoration} is proposed to help the user to work with annotations.
@@
 The type synthesis problem is not just about deciding whether a raw term is typable---there is a third possibility that the term does not have sufficient annotations.
-Therefore, to solve the type synthesis problem, before attempting to decide typability (using a bidirectional type synthesiser), we should first decide whether the raw term has sufficient annotations, which corresponds to whether the term has a mode derivation.
-Our theory gives a (proof-relevant) \emph{mode decorator}, which constructs a mode derivation for a raw term or provides information that refutes the existence of any mode derivation and helps the user to pinpoint missing annotations.
-Then a bidirectional type synthesiser is required to decide the typability of only mode-decorated raw terms instead of all raw terms.
+Thus, before attempting to decide typability (using a bidirectional type synthesiser), we should first decide if the raw term has sufficient annotations, which corresponds to whether the term has a mode derivation.
+Our theory gives a proof-relevant \emph{mode decorator}, which either
+\begin{inlineenum}
+  \item construct a mode derivation for a raw term, or
+  \item provides information that refutes the existence of any mode derivation and pinpoints missing annotations.
+\end{inlineenum}
+Then a bidirectional type synthesiser is only required to decide the typability of mode-decorated raw terms.
 Soundness and completeness of bidirectional typing is reformulated as a one-to-one correspondence between bidirectional typing derivations and pairs of a typing derivation and a mode derivation for the same raw term; this reformulation is more useful than annotatability~\cite[Section~3.2]{Dunfield2021} which is typically formulated in the literature of bidirectional typing.
@@
 In order to make the notion of mode-correctness precise, we first give a general definition of bidirectional simple type systems, called \emph{bidirectional binding signature}, extending the typed version of \varcitet{Aczel1978}{'s binding signature} with modes.
-A general definition of typed languages allows us to define mode-correctness and to investigate its consequences rigorously, including the uniqueness of synthesised types and the decidability of bidirectional type synthesis for all mode-correct signatures.
+A general definition of typed languages allows us to define mode-correctness and to investigate its consequences rigorously, including the uniqueness of synthesised types and the decidability of bidirectional type synthesis for mode-correct signatures.
 The proof of the latter theorem amounts to a generator of proof-relevant bidirectional type synthesisers (analogous to a parser generator working for unambiguous or disambiguated grammars).
 
-To make our exposition accessible, the theory in this paper focuses on simply typed languages with a syntax-directed bidirectional type system, so that the decidability of bidirectional type synthesis can be established without any other technical assumptions.
+To make our exposition accessible, the theory in this paper focuses on simply typed languages with a \emph{syntax-directed} bidirectional type system, so that the decidability of bidirectional type synthesis can be established without any other technical assumptions.
 %the idea of extending a signature with bidirectionality should be clear enough for further generalisation.
 It should be possible to extend the theory to deal with more expressive types and assumptions other than mode-correctness.
-For instance, we briefly discuss how the theory can handle polymorphically typed languages such as \SystemF, \SystemFsub, and those systems using implicit type applications with additional assumptions in \cref{sec:future}.
+For instance, we briefly discuss how the theory could be extended to handle polymorphically typed languages such as \SystemF, \SystemFsub, and those systems using implicit type applications with additional assumptions in \cref{sec:future}.
 

### 2024-01-11 tex/ESOP24/S2-outline.tex
[ tex ] Rewrite the example for binding arity with colourful annotations
[ tex ] Some revisions before removing appendix for the camera ready version

--- 2023-12-22 tex/ESOP24/S2-outline.tex
@@
 We start with an overview of our theory by instantiating it to simply typed $\lambda$-calculus.
-Roughly speaking, the problem of type synthesis requires us to take a raw term (i.e.~an untyped abstract syntax tree) as input, and produce a typing derivation for the term if possible.
-To give more precise definitions:
-The raw terms for simply typed $\lambda$-calculus are defined%
-\footnote{We omit the usual conditions about named representations of variables throughout this paper.}
+Roughly speaking, the problem of type synthesis requires us to take a raw term as input, and produce a typing derivation for the term if possible.
+To give more precise definitions: the raw terms for simply typed $\lambda$-calculus are defined%
+\footnote{The usual condition about named representations of variables is omitted.}
 in \cref{fig:STLC-raw-terms};
@@
     \and
-    \inferrule{V \vdash \isTerm{t}}{V \vdash \isTerm{(t \annotate A)}}\,\Rule{Anno}
+    \inferrule{V \vdash \isTerm{t}}{V \vdash \isTerm{(t \annotatecolon A)}}\,\Rule{Anno}
     \and
@@
     \and
-    \inferrule{\Gamma \vdash \isTerm{t} : A}{\Gamma \vdash \isTerm{(t \annotate A)} : A}\,\Rule{Anno}
+    \inferrule{\Gamma \vdash \isTerm{t} : A}{\Gamma \vdash \isTerm{(t \annotatecolon A)} : A}\,\Rule{Anno}
     \and
@@
 \end{figure}
+
 Correspondingly, the definition of typing derivations%
-\footnote{We write~`$\bto$' instead of~`$\to$' for the function types of simply typed $\lambda$-calculus to avoid confusion with the function types in our type-theoretic meta-language.}
-in \cref{fig:STLC-typing-derivations} also includes an \textsc{Anno} rule enforcing that the type of an annotated term does match the annotation.
+\footnote{We write~`$\bto$' instead of~`$\to$' for the function types of the simply typed $\lambda$-calculus to avoid confusion with the function types in our type-theoretic meta-language.}
+in \cref{fig:STLC-typing-derivations} has an \textsc{Anno} rule enforcing that the type of an annotated term does match the annotation.
+
 %\footnote{The presence of type annotations allows type synthesis to subsume type checking: to check whether a term~$t$ has type~$A$, synthesise the type of $t \bbcolon A$.}
@@
 If a type synthesiser cannot decide whether there is a typing derivation, it is allowed to give an excuse instead of an answer.
-Acceptable excuses are defined by the predicate~$E$, which describes what is wrong with an input term, for example not having enough type annotations.
+Acceptable excuses are defined by the predicate~$E$, which describes what is wrong with an input term, for example, not having enough type annotations.
 
@@
     \and
-    \inferrule{\Gamma \vdash \isTerm{t} \chk A}{\Gamma \vdash \isTerm{(t \annotate A)}\syn A}\,\SynRule{Anno}
+    \inferrule{\Gamma \vdash \isTerm{t} \chk A}{\Gamma \vdash \isTerm{(t \annotatecolon A)}\syn A}\,\SynRule{Anno}
     \and
@@
 \end{figure}
-
-Now our goal is to use \cref{def:STLC-type-synthesiser} as a specification and implement it using a \emph{bidirectional} type synthesiser, which attempts to produce \emph{bidirectional} typing derivations defined in \cref{fig:STLC-bidirectional-typing-derivations}.
-It is often said that a type synthesis algorithm can be `read off' from well-designed bidirectional typing rules.
-Take the $\SynRule{App}$ rule as an example:
-to synthesise the type of an application $\isTerm{t\;u}$, we first synthesise the type of~$\isTerm{t}$, which should have the form $A \bto B$, from which we can extract the expected type of~$\isTerm{u}$, namely~$A$, and perform checking; then the type of the whole application, namely~$B$, can also be extracted from the type $A \bto B$.
-Note that the synthesiser is able to figure out the type~$A$ for checking~$u$ and the type~$B$ to be synthesised for $t\;u$ because they have been computed when synthesising the type $A \bto B$ of~$t$.
-In general, there should be a flow of type information in each rule that allows us to determine unknown types (e.g.~types to be checked) from known ones (e.g.~types previously synthesised).
-This is called \emph{mode-correctness}, which we will formally define in \cref{sec:mode-correctness}.
-
-%The $\ChkRule{Abs}$ rule checks whether the type of an abstraction $\isTerm{\lam{x}t}$ is $A \bto B$ by checking whether $\isTerm{t}$~has type~$B$ assuming $\isTerm{x}$~has type~$A$ in the context, where both types $A$~and~$B$ can be determined from the type $A \bto B$ given initially.
-%For an annotated term $t \bbcolon A$, the $\SynRule{Anno}$ rule simply synthesises~$A$ as its type, provided that $t$~can be successfully checked to have type~$A$.
-%Finally, the subsumption rule $\ChkRule{Sub}$ says that a term synthesising a type~$A$ can be checked to have type~$B$ if $A = B$ --- in general this type equality can be replaced by some subtyping relation.
 
@@
     \and
-    \inferrule{V |- \isTerm{t}^{\chk}}{V |- (\isTerm{t \annotate A})^{\syn}}\,\SynRule{Anno}
+    \inferrule{V |- \isTerm{t}^{\chk}}{V |- (\isTerm{t \annotatecolon A})^{\syn}}\,\SynRule{Anno}
     \and
@@
 \end{figure}
+Now our goal is to use \cref{def:STLC-type-synthesiser} as a specification and implement it using a \emph{bidirectional} type synthesiser, which attempts to produce \emph{bidirectional} typing derivations defined in \cref{fig:STLC-bidirectional-typing-derivations}.
+It is often said that a type synthesis algorithm can be `read off' from well-designed bidirectional typing rules.
+Take the $\SynRule{App}$ rule as an example:
+to synthesise the type of an application $\isTerm{t\;u}$, we first synthesise the type of~$\isTerm{t}$, which should have the form $A \bto B$, from which we can extract the expected type of~$\isTerm{u}$, namely~$A$, and perform checking; then the type of the whole application, namely~$B$, can also be extracted from the type $A \bto B$.
+Note that the synthesiser is able to figure out the type~$A$ for checking~$u$ and the type~$B$ to be synthesised for $t\;u$ because they have been computed when synthesising the type $A \bto B$ of~$t$.
+In general, there should be a flow of type information in each rule that allows us to determine unknown types (e.g.~types to be checked) from known ones (e.g.~types previously synthesised).
+This is called \emph{mode-correctness}, which we will formally define in \cref{sec:mode-correctness}.
+
+%The $\ChkRule{Abs}$ rule checks whether the type of an abstraction $\isTerm{\lam{x}t}$ is $A \bto B$ by checking whether $\isTerm{t}$~has type~$B$ assuming $\isTerm{x}$~has type~$A$ in the context, where both types $A$~and~$B$ can be determined from the type $A \bto B$ given initially.
+%For an annotated term $t \bbcolon A$, the $\SynRule{Anno}$ rule simply synthesises~$A$ as its type, provided that $t$~can be successfully checked to have type~$A$.
+%Finally, the subsumption rule $\ChkRule{Sub}$ says that a term synthesising a type~$A$ can be checked to have type~$B$ if $A = B$ --- in general this type equality can be replaced by some subtyping relation.
 
@@
 \begin{defn} \label{def:STLC-mode-decorator}
-A \emph{mode decorator} decides for any raw term~$V \vdash \isTerm{t}$ whether $V \vdash \isTerm{t}^{\syn}$.
+A \emph{mode decorator} decides for any raw term~$V \vdash \isTerm{t}$ if $V \vdash \isTerm{t}^{\syn}$.
 \end{defn}
@@
 
-Now we can get back to our goal of implementing a type synthesiser~(\cref{def:STLC-type-synthesiser}).
+Now we can get back to implementing a type synthesiser~(\cref{def:STLC-type-synthesiser}).
 
@@
 \begin{theorem}[Soundness and Completeness]
-$\Gamma \vdash t \syn A$ if and only if\/ $\erase\Gamma \vdash t^{\syn}$ and\/ $\Gamma \vdash t : A$.
+$\Gamma \vdash t \syn A$ if and only if $\erase\Gamma \vdash t^{\syn}$ and $\Gamma \vdash t : A$.
 \end{theorem}

### 2024-01-11 tex/ESOP24/S3-defs.tex
[ tex ] Some revisions before removing appendix for the camera ready version

@@
   \begin{equation*}
-    \frac{\Gamma, \eqnmarkbox[blue]{ext1}{\isTerm{x} : A, \cdot} \vdash \isTerm{t} : \eqnmarkbox[red]{aty1}{B}}{\Gamma \vdash \isTerm{\lam{x}t} : \eqnmarkbox{ty1}{A \bto B}}
+    \frac{\Gamma, \isTerm{x} : \eqnmarkbox[blue]{ext1}{A, \cdot} \vdash \isTerm{t} : \eqnmarkbox[red]{aty1}{B}}{\Gamma \vdash \isTerm{\lam{x}t} : \eqnmarkbox{ty1}{A \bto B}}
     \qquad\text{and}\qquad
@@
   \\
-  if the empty extension context $\cdot$ is added verbosely, so they can be specified by binding arities ${[A]B} \to {(A \bto B)}$ and ${(A \bto B), A} \to {B}$ respectively.
+  if the empty context $\cdot$ is added verbosely, so they can be specified by arities ${[A]B} \to {(A \bto B)}$ and ${(A \bto B), A} \to {B}$, respectively, with $\Type_{\Sigma_{\bto}}(A, B)$ as types.
 \end{example}
@@
 
+\begin{defn}
+  For a simply typed language $(\Sigma, \Omega)$, the family of sets of \emph{raw terms} indexed by a list~$V$ of variables consists of
+  \begin{inlineenum}
+    \item variables in $V$,
+    \item annotations $\isTerm{t \annotatecolon A}$ for some raw term $t$ in $V$ and a type $A$, and
+    \item a construct $\tmOpts$ for some $\bop$ in~$O$, where $\vars{x}_{\isTerm{i}}$'s are lists of variables whose length is equal to the length of~$\Delta_i$, and $t_i$'s are raw terms in the variable list $V, \vars{x}_i$.
+  \end{inlineenum}
+  These correspond to rules $\Rule{Var}$, $\Rule{Anno}$, and $\Rule{Op}$ in \Cref{fig:raw-terms} respectively.
+\end{defn}
+
 \begin{figure}
@@
 \end{figure}
-
-\begin{defn}
-  For a simply typed language $(\Sigma, \Omega)$, the family of sets of \emph{raw terms} indexed by a list~$V$ of variables consists of
-  \begin{inlineenum}
-    \item variables in $V$,
-    \item annotations $\isTerm{t \annotatecolon A}$ for some raw term $t$ in $V$ and a type $A$, and
-    \item a construct $\tmOpts$ for some $\bop$ in~$O$, where $\vars{x}_{\isTerm{i}}$'s are lists of variables whose length is equal to the length of~$\Delta_i$, and $t_i$'s are raw terms in the variable list $V, \vars{x}_i$.
-  \end{inlineenum}
-  These correspond to rules $\Rule{Var}$, $\Rule{Anno}$, and $\Rule{Op}$ in \Cref{fig:raw-terms} respectively.
-\end{defn}
-
 Before defining typing derivations, we need a definition of typing contexts.

### 2024-01-11 tex/ESOP24/S5-synthesis.tex
[ tex ] Rewrite the example for binding arity with colourful annotations
[ tex ] Some revisions before removing appendix for the camera ready version

@@
 \begin{defn}\label{def:mode-correctness}
-  An operation $\biop$ is \emph{mode-correct} if 
+  An arity $\biarity$ is \emph{mode-correct} if 
   \begin{enumerate}
@@
   \end{enumerate}
-  A bidirectional binding signature $\Omega$ is \emph{mode-correct} if its operations are all mode-correct.
+  A bidirectional binding signature $\Omega$ is \emph{mode-correct} if every operation's arity is mode-correct.
 \end{defn}
@@
 \begin{lemma}\label{lem:decidability-mode-correctness}
-  For any operation $\biop$, it is decidable whether $o$ is mode-correct.
+  For any bidirectional binding arity $\biarity$, it is decidable whether it is mode-correct.
 \end{lemma}
@@
     \item $\SynRule{Var}$ follows from that each variable as a raw term refers to the same variable in its context.
-    \item $\SynRule{Anno}$ holds trivially, since the synthesised type $A$ is from the term $t \annotate A$ in question.
+    \item $\SynRule{Anno}$ holds trivially, since the synthesised type $A$ is from the term $t \annotatecolon A$ in question.
     \item $\Rule{Op}$: Recall that a derivation of\/ $\Gamma |- \tmOpts \syn A$ contains a substitution $\rho$ from the local context $\Xi$ to concrete types.

### 2024-01-11 tex/ESOP24/S8-future.tex
[ tex ] Some revisions before removing appendix for the camera ready version

--- 2023-12-25 tex/ESOP24/S8-future.tex
@@
 While our current development is based on simply typed languages to highlight the core ideas, it is evident that many concepts and aspects remain untouched.
-In this section we discuss related work and potential directions for future research.
+%In this section we discuss related work and potential directions for future research.
 \subsubsection{Language formalisation frameworks}
@@
 The study of presenting logics universally at least date back to universal algebra and model theory where structures are studied for certain notions of arities and signatures.
-In programming language theory, \varcitet{Aczel1978}{'s binding signature} is an example which is used to prove a general confluence theorem.
+In programming language theory, \varcitet{Aczel1978}{'s binding signature} is an example which has been used to prove a general confluence theorem.
 Many general definitions and frameworks for defining logics and type theories have been proposed, and we classify them into two groups by where signatures reside: the meta level or the object level of a meta-language:
@@
 
-\subsubsection{Beyond syntax-directed systems}
+\subsubsection{Beyond syntax-directedness}
 
-For non-syntax-directed systems, since there can be multiple moding or typing rules for a language construct, both the mode decorator and the bidirectional type synthesiser will need to backtrack, and the decorator may need to produce all possible mode derivations so as not to limit the synthesiser's choices.
-For soundness and completeness, it should still be possible to treat them as the separation and combination of mode and typing derivations, although bidirectional typing rules and pairs of a moding rule and a typing rule may no longer be in perfect correspondence.
-Completeness may need to be reformulated to take all possible mode derivations as input to guarantee that at least one can be paired with the input typing derivation.
+To explore general settings, an extreme possibility is to independently specify an ordinary and a bidirectional type system over the same raw terms and then investigate relationships between the two systems.
+
+We could first focus on simpler cases that the ordinary typing part is still syntax-directed, but each typing rule is allowed to have multiple bidirectional variants, including different orders of their premises.
+In such cases, the mode decorator would need to find all mode derivations, but the type synthesiser should still work in a syntax-directed manner on each mode derivation.
+Completeness could still take the simple form presented in this paper too.
+
+The next step is to consider non-syntax-directed systems where each typing rule has multiple bidirectional variants.
+For this setting, the mode decorator and the bidirectional type synthesiser will have to backtrack.
+It is still possible to treat soundness as the separation of mode and type information, but completeness will pose a problem.
+For every raw term, a mode derivation chooses a mode assignment while a typing derivation chooses a typing rule, but there may not be a bidirectional typing rule for this particular combination.
+To retain the mode decoration, one could make the mode decorator produce all possible mode derivations, and refine completeness to say that any typing derivation can be combined with one of mode derivations into a bidirectional typing derivation.
+%It should be possible to fix this problem while retaining the mode decoration phase.
 
@@
 We thank Kuen-Bang {Hou (Favonia)} and anonymous reviewers for their comments and suggestions.
+The work was supported by the National Science and Technology Council of Taiwan under grant NSTC 112-2221-E-001-003-MY3.
 \end{credits}

### 2024-01-11 tex/ESOP24/S1-intro.tex
[ tex ] Fight with the page limit

@@
 The main technical contribution lies in \cref{sec:type-synthesis}, where we introduce mode-correctness and bidirectional type synthesis.
-Some examples other than simply typed $\lambda$-calculus are given in \Cref{sec:example}.
-We discuss further developments in \cref{sec:future} and include a demonstration in \cref{sec:formalisation} for the use of our \Agda formalisation as programs.
+Some examples other than simply typed $\lambda$-calculus are given in \Cref{sec:example} and further developments are discussed in \cref{sec:future}.
+\ifarxiv
+A demonstration is given in \cref{sec:formalisation} for the use of our \Agda formalisation as programs.
+\fi

### 2024-01-11 tex/ESOP24/S5-synthesis.tex
[ tex ] Fight with the page limit

@@
 Hence, we assume that the set~$\Xi$ of type variables has a decidable equality, thereby ensuring that the set $\Type_{\Sigma}(\Xi)$ of types also has a decidable equality.\footnote{%
-To simplify our choice, we could simply confine $\Xi$ to any set within the family of sets $\Fin(n)$ of naturals less than~$n$, given that these sets have a decidable equality and the arity of a type construct is finite.
-Indeed, in our formalisation, we adopt $\Fin(n)$ as the set of type variables in the definition of $\Type_{\Sigma}$ (see \Cref{sec:formalisation} for details).
-For the sake of clarity in presentation, though, we keep using named variables and just assume that $\Xi$ has a decidable equality.}
+To simplify our choice, we may confine $\Xi$ to any set within the family of sets $\Fin(n)$ of naturals less than~$n$, given that these sets have a decidable equality and the arity of a type construct is finite.
+Indeed, in our formalisation, we adopt $\Fin(n)$ as the set of type variables in the definition of $\Type_{\Sigma}$\ifarxiv (see \Cref{sec:formalisation} for details)\fi.
+For the sake of clarity in presentation, though, we use named variables and just assume that $\Xi$ has a decidable equality.}
 \end{remark}

### 2024-01-11 tex/ESOP24/S6-example.tex
[ tex ] Fight with the page limit

@@
 
+\ifarxiv
 For illustrative purposes, consider a computational calculus~\cite{Moggi1989} with additional constructs listed in \cref{tab:computational-calculus}.
@@
 The signature $\Omega_{\Lambda}^{\Leftrightarrow}$ is then extended with operations listed in \cref{tab:computational-calculus}.
-Mode-correctness can be derived by \cref{lem:decidability-mode-correctness} and its type synthesiser by \cref{cor:trichotomy} with the specification~$(\Sigma, \Omega)$ directly.%
-\footnote{For a demonstration in \Agda, see \cref{sec:formalisation}.}
+Mode-correctness can be derived by \cref{lem:decidability-mode-correctness} and its type synthesiser by \cref{cor:trichotomy} with the specification~$(\Sigma, \Omega)$ directly.
+For a demonstration in \Agda, see \cref{sec:formalisation}.
+\else
+For illustrative purposes, consider a computational calculus~\cite{Moggi1989} with naturals, sums, products, and general recursion.
+The extended language has `only' 15 constructs, including pairing, projections, injections, and so on, whereas there are nearly 100 possible cases to consider in bidirectional type synthesis.
+However, it has far fewer constructs than a realistic programming language would have.
+
+On the other hand, similar to a parser generator, only one specification is needed for a type-synthesiser generator from the user to produce a corresponding synthesiser. 
+Such a specification can be derived by extending $(\Sigma_{\bto}, \Omega_{\Lambda}^{\Leftrightarrow})$ accordingly for additional types and constructs with mode-correctness proved by applying \cref{lem:decidability-mode-correctness}, so its type synthesiser follows from \cref{cor:trichotomy} directly.
+\fi

### 2024-01-11 tex/ESOP24/S8-future.tex
[ tex ] Fight with the page limit

@@
 
-\LF is naturally expressive but each extension requires a different \LF and thus a different implementation to check formal \LF proofs.
-Formalising \LF and its variants is at least as hard as formalising a dependent type theory, so they are mostly implemented separately from their theory and unverified.
+The \LF approach is naturally expressive but each extension requires a different framework and a different implementation to check formal \LF proofs.
+Formalising \LF and its variants is at least as hard as formalising a dependent type theory, and they are mostly implemented separately from their theory and unverified.
 
@@
 Bidirectional type synthesis plays a crucial role in handling more complex types than simple types, and we sketch how our theory can be extended to treat a broader class of languages.
-First, we need a general definition of languages, discussed above, in question (\cref{subsec:simple-types,subsec:binding-sig}).
-Then, this definition can be augmented with modes (\cref{subsec:bidirectional-system}) and the definition of mode-correctness (\cref{def:mode-correctness}) can be adapted by adding the mode information into signatures.
+First, we need a general definition of languages in question (\cref{subsec:simple-types,subsec:binding-sig}).
+Then, this definition can be augmented with modes (\cref{subsec:bidirectional-system}) and the definition of mode-correctness (\cref{def:mode-correctness}) can be adapted accordingly.
 Soundness and completeness (\cref{lem:soundness-completeness}) should still hold, as they amount to the separation and combination of mode and typing information for a given raw term (in a syntax-directed formulation).
-Mode decoration (\cref{sec:mode-decoration}) involves annotating a raw term with modes and marking missing annotations, and should also work.
+Mode decoration (\cref{sec:mode-decoration}), which annotates a raw term with modes and marks missing annotations, should also work.
 As for the decidability of bidirectional type synthesis, we discuss two cases involving polymorphic types and dependent types below.
@@
 Extending general definitions for bidirectional typing and mode derivations from \citeauthor{Hamana2011}'s work is straightforward. 
-For example, the universal type $\forall \alpha.\, A$ and type abstraction in \SystemF can be specified as operations:
+\ifarxiv
+For example, the universal type $\forall \alpha.\, A$ and type abstraction in \SystemF can be specified as operations
 \[
@@
   \quad\text{and}\quad
-\mathsf{tabs} : [*] A \rhd \left< * \right> A^{\chk} \to \tyOp_{\mathsf{all}}(\alpha. A)^{\chk}
+  \mathsf{tabs} : [*] A \rhd \left< * \right> A^{\chk} \to \tyOp_{\mathsf{all}}(\alpha. A)^{\chk}.
 \]
+\else
+For example, the universal type $\forall \alpha.\, A$ and type abstraction in \SystemF can be specified as operations $\mathsf{all} : * \rhd [*] * \to *$ and $\mathsf{tabs} : [*] A \rhd \left< * \right> A^{\chk} \to \tyOp_{\mathsf{all}}(\alpha.
+A)^{\chk}$.
+\fi
 The decidability of bidirectional type synthesis (\cref{thm:bidirectional-type-synthesis-checking}) should also carry over, as no equations are imposed on types and no guessing (for type application) is required.
@@
 
+\ifarxiv
 However, explicit type application in \SystemF and \SystemFsub is impractical but its implicit version results in a \emph{stationary rule}~\cite{Leivant1986}:
@@
 \]
+\else
+However, explicit type application in \SystemF and \SystemFsub is impractical but its implicit version results in a \emph{stationary rule}~\cite{Leivant1986}
+\fi
 which is not syntax-directed.
-By translating the above rule to subtyping, we have the \emph{instantiation problem} which amounts to guessing $B$ in $\forall \alpha. A <: A[B/\alpha]$.
+By translating the rule to subtyping, we have the \emph{instantiation problem} which amounts to guessing $B$ in $\forall \alpha. A <: A[B/\alpha]$.
 A theory that accommodates various solutions to the problem is left as future work.
@@
 Extending the logical framework \Dedukti~\cite{Assaf2016}, \varcitet{Felicissimo2023}{'s recent \textsf{CompLF}} is capable of defining dependent type theories with computational rules and enables generic bidirectional type synthesis.
-They propose a notion of \LF signature with modes (called \emph{moded signatures} op.\ cit.), a definition of mode-correctness (called \emph{well-typed} op.\ cit.), and an algorithm of generic bidirectional type synthesis which is decidable, sound, and complete with respect to mode-decorated terms, provided that the specified language is mode-correct, and the set of computational rules is \emph{well-behaved}, i.e.\ type-preserving, confluent, and strongly normalising.
+They propose a notion of \LF signature with modes, a definition of mode-correctness (called \emph{well-typed} op.\ cit.), and an algorithm of bidirectional type synthesis.
+The algorithm is decidable, sound and complete with respect to mode-decorated terms, provided that the specified language is mode-correct, and that the set of computational rules is type-preserving, confluent, and strongly normalising.
 However, the last assumption is challenging to establish even for specific languages and lacks a general understanding.
@@
 
-There are more techniques in bidirectional typing that could be formally studied in general, with one notable example being the Pfenning recipe for bidirectionalising typing rules~\citep[Section~4]{Dunfield2021}.
+There are more principles and techniques in bidirectional typing that could be formally studied in general, with one notable example being the Pfenning recipe for bidirectionalising typing rules~\citep[Section~4]{Dunfield2021}.
 There are also concepts that may be hard to fully formalise, for example `annotation character'~\citep[Section~3.4]{Dunfield2021}, which is roughly about how easy it is for the user to write annotated programs, but it would be interesting to explore to what extent such concepts can be formalised.

### 2024-01-12 tex/ESOP24/S1-intro.tex
[ tex ] Minor changes

@@
 \end{inlineenum}
-The former indicates that the type~$A$ is an output, using both the context~$\Gamma$ and the term~$t$ as input, while for the latter, all three of $\Gamma$, $t$, and~$A$ are given as input.
+The former indicates that the type~$A$ is an output, using both the context~$\Gamma$ and the term~$t$ as input, while for the latter, all three of $\Gamma$, $t$, and~$A$ are input.
 The algorithm of a bidirectional type synthesiser can often be `read off' from a well-designed bidirectional type system: as the synthesiser traverses a raw term, it switches between synthesis and checking, following the modes assigned to the judgements in the typing rules.

### 2024-01-12 tex/ESOP24/S4-pre-synthesis.tex
[ tex ] Rewrite the example for binding arity with colourful annotations
[ tex ] Minor changes

@@
 Instead we can remove typing and retain modes, arriving at a mode derivation.
-Conversely, if we have both mode and typing derivations for the same term, we can combine them to obtain a bidirectional typing derivation, which is completeness.
-In short, soundness and completeness are no more than the separation and combination of mode and typing information carried by the three kinds of derivations while keeping their basic structure, directed by the same raw term.
-All these can be summarised in one theorem (proved by induction).
+Conversely, if we have both mode and typing derivations for the same term, we can combine them to form a bidirectional typing derivation, which is completeness.
+In short, soundness and completeness are merely the separation and combination of mode and typing information carried by the three kinds of derivations while keeping their basic structure, directed by the same raw term.
+All these can be summarised in one theorem and proved by induction.
 
@@
     \and
-    \inferrule{\cdot |-_{\Sigma} A \\ V |-_{\Sigma, \Omega}\isTerm{t}^{\chk\,g\,s}}{V |-_{\Sigma, \Omega} (\isTerm{t \annotate A})^{\syn\,g\,\BTrue}}\,\SynRule{Anno}
+    \inferrule{\cdot |-_{\Sigma} A \\ V |-_{\Sigma, \Omega}\isTerm{t}^{\chk\,g\,s}}{V |-_{\Sigma, \Omega} (\isTerm{t \annotatecolon A})^{\syn\,g\,\BTrue}}\,\SynRule{Anno}
     \and
@@
     \inferrule{\isTerm{t} \sqsupseteq \isTerm{u}}
-              {(\isTerm{t} \annotate A) \sqsupseteq \isTerm{u}}\;\Rule{More}
+              {(\isTerm{t} \annotatecolon A) \sqsupseteq \isTerm{u}}\;\Rule{More}
     \and
@@
     \inferrule{\isTerm{t} \sqsupseteq \isTerm{u}}
-      {(\isTerm{t} \annotate A) \sqsupseteq (\isTerm{u} \annotate A)}\;\Rule{Anno}
+      {(\isTerm{t} \annotatecolon A) \sqsupseteq (\isTerm{u} \annotatecolon A)}\;\Rule{Anno}
     \and

### 2024-01-12 tex/ESOP24/S5-synthesis.tex
[ tex ] Minor changes

@@
 
-The interesting part of the theorem is the case for the $\Rule{Op}$ rule and it is rather complex.
-Here we shall give its insight first instead of jumping into the details.
+The interesting part of the theorem is the case for the $\Rule{Op}$ rule.
+We shall give its insight first instead of jumping into the details.
 Recall that a typing derivation for $\tmOpts$ contains a substitution $\rho\colon \Xi \to \Type_{\Sigma}(\emptyset)$.
-The goal of type synthesis is exactly to define such a substitution $\rho$, so we have to start with an `accumulating' substitution: a substitution $\rho_0$ that is partially defined on $\fv(A_0)$ if $d$ is $\chk$ or otherwise nowhere.
+The goal of type synthesis for this case is exactly to define such a substitution~$\rho$, and we have to start with an `accumulating' substitution: a substitution $\rho_0$ that is partially defined on $\fv(A_0)$ if $d$ is $\chk$ or otherwise nowhere.
 By mode-correctness, the accumulating substitution~$\rho_i$ will be defined on enough synthesised variables so that type synthesis or checking can be performed on $t_{i}$ with the context $\Gamma, \vars{x}_{i} : \simsub{\Delta_{i}}{\rho_{i}}$ based on its mode derivation $\erase{\Gamma}, \vars{x}_i |-_{\Sigma, \Omega} t_i^{\dir{d_i}}$.
-If we visit a synthesising argument $\synbiarg[i+1]$, then we may extend the domain of $\rho_i$ to include the synthesised variables $\fv(A_{i + 1})$ if type synthesis is successful and also that the synthesised type can be \emph{unified with $A_{i+ 1}$} and thereby \emph{extend} $\rho_i$ to $\bar{\rho_i} = \rho_{i+1}$ with the unifier.
-Then, if we go through every $t_i$ successfully, we will have a total substitution $\rho_n$ by mode-correctness and a derivation of $\Gamma, \vars{x}_i : \Delta_i |-_{\Sigma, \Omega} t_i :^{\dir{d_i}} \simsub{A}{\rho_n}$ for each sub-term $t_i$.
+If we visit a synthesising argument $\synbiarg[i+1]$, then we may extend the domain of $\rho_i$ to $\rho_{i+1}$ with the synthesised variables $\fv(A_{i + 1})$, provided that type synthesis is successful and that the synthesised type can be \emph{unified with $A_{i+ 1}$}.
+If we go through every $t_i$ successfully, then we will have a total substitution $\rho_n$ by mode-correctness and a derivation of $\Gamma, \vars{x}_i : \Delta_i |-_{\Sigma, \Omega} t_i :^{\dir{d_i}} \simsub{A}{\rho_n}$ for each sub-term~$t_i$.
 
@@
 Indeed, in our formalisation, we adopt $\Fin(n)$ as the set of type variables in the definition of $\Type_{\Sigma}$\ifarxiv (see \Cref{sec:formalisation} for details)\fi.
-For the sake of clarity in presentation, though, we use named variables and just assume that $\Xi$ has a decidable equality.}
+For the sake of clarity in presentation, though, we just use named variables and assume that $\Xi$ has a decidable equality.}
 \end{remark}
@@
 \begin{lemma}\label{lem:unify}
-  For any\/ $A$ of\/ $\Type_{\Sigma}(\Xi)$, $B$ of\/ $\Type_{\Sigma}(\emptyset)$, and a partial substitution\/ $\rho \colon \Xi \to \Type_{\Sigma}(\emptyset)$, 
+  For any\/ $A$ of\/ $\Type_{\Sigma}(\Xi)$, $B$ of\/ $\Type_{\Sigma}(\emptyset)$, and a partial substitution\/ $\rho \colon \Xi \to \Type_{\Sigma}(\emptyset)$, either
   \begin{enumerate}
-    \item either there is a minimal extension\/ $\bar{\rho}$ of\/ $\rho$ such that\/ $\simsub{A}{\bar{\rho}} = B$,
-    \item or there is no extension\/ $\sigma$ of\/ $\rho$ such that\/ $\simsub{A}{\sigma} = B$
+    \item there is a minimal extension\/ $\bar{\rho}$ of\/ $\rho$ such that\/ $\simsub{A}{\bar{\rho}} = B$, or
+    \item there is no extension\/ $\sigma$ of\/ $\rho$ such that\/ $\simsub{A}{\sigma} = B$
   \end{enumerate}

### 2024-01-12 tex/ESOP24/S6-example.tex
[ tex ] Minor changes

@@
 
-On the other hand, similar to a parser generator, only one specification is needed for a type-synthesiser generator from the user to produce a corresponding synthesiser. 
+On the other hand, similar to a parser generator, only one specification is needed for a type-synthesiser generator from the user to produce a corresponding type synthesiser. 
 Such a specification can be derived by extending $(\Sigma_{\bto}, \Omega_{\Lambda}^{\Leftrightarrow})$ accordingly for additional types and constructs with mode-correctness proved by applying \cref{lem:decidability-mode-correctness}, so its type synthesiser follows from \cref{cor:trichotomy} directly.

### 2024-01-12 tex/ESOP24/S8-future.tex
[ tex ] Minor changes

@@
 The algorithm is decidable, sound and complete with respect to mode-decorated terms, provided that the specified language is mode-correct, and that the set of computational rules is type-preserving, confluent, and strongly normalising.
-However, the last assumption is challenging to establish even for specific languages and lacks a general understanding.
+The last assumption, however, is challenging to establish even for specific languages and lacks a general understanding.
 
@@
 
-We could first focus on simpler cases that the ordinary typing part is still syntax-directed, but each typing rule is allowed to have multiple bidirectional variants, including different orders of their premises.
+Yet, we could first focus on simpler cases that the ordinary typing part is still syntax-directed, but each typing rule is refined to multiple bidirectional variants, including different orders of its premises.
 In such cases, the mode decorator would need to find all mode derivations, but the type synthesiser should still work in a syntax-directed manner on each mode derivation.

### 2024-01-12 tex/ESOP24/S4-pre-synthesis.tex
[ tex ] Remove \/ (LNCS has enough spacing)

@@
 \begin{theorem}\label{lem:soundness-completeness}
-$\Gamma |-_{\Sigma, \Omega} t :^\dir{d} A$ if and only if both\/ $\erase\Gamma |-_{\Sigma, \Omega} t^\dir{d}$ and\/ $\Gamma |-_{\Sigma, \erase\Omega} t : A$.
+$\Gamma |-_{\Sigma, \Omega} t :^\dir{d} A$ if and only if both $\erase\Gamma |-_{\Sigma, \Omega} t^\dir{d}$ and $\Gamma |-_{\Sigma, \erase\Omega} t : A$.
 \end{theorem}
@@
 \begin{lemma}\label{thm:Pre?-true}
-If\/ $V |-_{\Sigma,\Omega} t^{\dir{d}\,\BTrue\,s}$, then $V |-_{\Sigma,\Omega} t^\dir{d}$.
+If $V |-_{\Sigma,\Omega} t^{\dir{d}\,\BTrue\,s}$, then $V |-_{\Sigma,\Omega} t^\dir{d}$.
 \end{lemma}
@@
 \begin{lemma}\label{thm:Pre?-false}
-If\/ $V |-_{\Sigma,\Omega} t^{\dir{d}\,\BFalse\,s}$, then $V |/-_{\Sigma,\Omega} t^\dir{d}$.
+If $V |-_{\Sigma,\Omega} t^{\dir{d}\,\BFalse\,s}$, then $V |/-_{\Sigma,\Omega} t^\dir{d}$.
 \end{lemma}
@@
 \begin{theorem}[Generalised mode decoration]\label{thm:generalised-mode-decoration}
-For any raw term $V |-_{\Sigma,\erase\Omega} t$ and mode~$\dir{d}$, there is a derivation of\/ $V |-_{\Sigma,\Omega} t^{\dir{d}\,g\,s}$ for some $g$~and~$s$.
+For any raw term $V |-_{\Sigma,\erase\Omega} t$ and mode~$\dir{d}$, there is a derivation of $V |-_{\Sigma,\Omega} t^{\dir{d}\,g\,s}$ for some $g$~and~$s$.
 \end{theorem}
@@
 \end{equation}
-then for any mode~$\dir{d}$, there is a derivation of\/ $V |-_{\Sigma,\Omega} t^{\dir{d}\,g\,s}$ for some $g$~and~$s$.
+then for any mode~$\dir{d}$, there is a derivation of $V |-_{\Sigma,\Omega} t^{\dir{d}\,g\,s}$ for some $g$~and~$s$.
 \end{lemma}
@@
 \begin{proposition}[Annotatability]\label{thm:annotatability}
-If\/ $\Gamma |-_{\Sigma,\erase\Omega} t : A$, then there exists~$t'$ such that\/ $t' \sqsupseteq t$ and $\Gamma |-_{\Sigma,\Omega} t' :^\dir{d} A$ for some~$\dir{d}$.
+If $\Gamma |-_{\Sigma,\erase\Omega} t : A$, then there exists~$t'$ such that $t' \sqsupseteq t$ and $\Gamma |-_{\Sigma,\Omega} t' :^\dir{d} A$ for some~$\dir{d}$.
 \end{proposition}

### 2024-01-12 tex/ESOP24/S5-synthesis.tex
[ tex ] Remove \/ (LNCS has enough spacing)

@@
     \item $\SynRule{Anno}$ holds trivially, since the synthesised type $A$ is from the term $t \annotatecolon A$ in question.
-    \item $\Rule{Op}$: Recall that a derivation of\/ $\Gamma |- \tmOpts \syn A$ contains a substitution $\rho$ from the local context $\Xi$ to concrete types.
+    \item $\Rule{Op}$: Recall that a derivation of $\Gamma |- \tmOpts \syn A$ contains a substitution $\rho$ from the local context $\Xi$ to concrete types.
       To prove that any two typing derivations has the same synthesised type, it suffices to show that those substitutions $\rho_1$~and~$\rho_2$ of $d_1$~and~$d_2$, respectively, agree on variables in $\synvar(\biargs)$ so that $\simsub{A_0}{\rho_1} = \simsub{A_0}{\rho_2}$.
@@
   \begin{enumerate}
-    \item if\/ $\erase{\Gamma} |-_{\Sigma, \Omega} \isTerm{t}^{\syn}$, then it is decidable whether $\Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn A$ for some $A$;
-    \item if\/ $\erase{\Gamma} |-_{\Sigma, \Omega} \isTerm{t}^{\chk}$, then it is decidable for any~$A$ whether $\Gamma |-_{\Sigma, \Omega} \isTerm{t} \chk A$.
+    \item if $\erase{\Gamma} |-_{\Sigma, \Omega} \isTerm{t}^{\syn}$, then it is decidable whether $\Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn A$ for some $A$;
+    \item if $\erase{\Gamma} |-_{\Sigma, \Omega} \isTerm{t}^{\chk}$, then it is decidable for any~$A$ whether $\Gamma |-_{\Sigma, \Omega} \isTerm{t} \chk A$.
   \end{enumerate}
@@
 \begin{defn}
-By an \emph{extension}\/ $\sigma \geq \rho$ of a partial substitution $\rho$ we mean that the domain $\dom(\sigma)$ of $\sigma$ contains the domain of $\rho$ and $\sigma(x) = \rho(x)$ for every\/ $x$ in $\dom(\rho)$.
-  By a \emph{minimal extension}\/ $\bar{\rho}$ of $\rho$ satisfying $P$ we mean an extension $\bar{\rho} \geq \rho$ with $P(\bar{\rho})$ such that $\sigma \geq \bar{\rho}$ whenever $\sigma \geq \rho$ and $P(\sigma)$.
+By an \emph{extension} $\sigma \geq \rho$ of a partial substitution $\rho$ we mean that the domain $\dom(\sigma)$ of $\sigma$ contains the domain of $\rho$ and $\sigma(x) = \rho(x)$ for every $x$ in $\dom(\rho)$.
+  By a \emph{minimal extension} $\bar{\rho}$ of $\rho$ satisfying $P$ we mean an extension $\bar{\rho} \geq \rho$ with $P(\bar{\rho})$ such that $\sigma \geq \bar{\rho}$ whenever $\sigma \geq \rho$ and $P(\sigma)$.
 \end{defn}
 \begin{lemma}\label{lem:unify}
-  For any\/ $A$ of\/ $\Type_{\Sigma}(\Xi)$, $B$ of\/ $\Type_{\Sigma}(\emptyset)$, and a partial substitution\/ $\rho \colon \Xi \to \Type_{\Sigma}(\emptyset)$, either
+  For any $A$ of $\Type_{\Sigma}(\Xi)$, $B$ of $\Type_{\Sigma}(\emptyset)$, and a partial substitution $\rho \colon \Xi \to \Type_{\Sigma}(\emptyset)$, either
   \begin{enumerate}
-    \item there is a minimal extension\/ $\bar{\rho}$ of\/ $\rho$ such that\/ $\simsub{A}{\bar{\rho}} = B$, or
-    \item there is no extension\/ $\sigma$ of\/ $\rho$ such that\/ $\simsub{A}{\sigma} = B$
+    \item there is a minimal extension $\bar{\rho}$ of $\rho$ such that $\simsub{A}{\bar{\rho}} = B$, or
+    \item there is no extension $\sigma$ of $\rho$ such that $\simsub{A}{\sigma} = B$
   \end{enumerate}
@@
 \begin{proof}[of {\Cref{thm:bidirectional-type-synthesis-checking}}]
-  We prove this statement by induction on the mode derivation\/ $\erase{\Gamma} |-_{\Sigma, \Omega} \isTerm{t}^{\dir{d}}$.
+  We prove this statement by induction on the mode derivation $\erase{\Gamma} |-_{\Sigma, \Omega} \isTerm{t}^{\dir{d}}$.
   The two cases \SynRule{Var} and \SynRule{Anno} are straightforward and independent of mode-correctness.

### 2024-01-12 tex/ESOP24/S1-intro.tex
[ tex ] Eliminate most of hyphens the end of every line

@@
 
-As mentioned in the beginning, with bidirectional typing we avoid the generally undecidable problem of type inference, and instead solve the simpler problem about the typability of `sufficiently annotated' raw terms, which we call the type synthesis problem to distinguish it from type inference.
+As we mentioned in the beginning, with bidirectional typing we avoid the generally undecidable problem of type inference, and instead solve the simpler problem about the typability of `sufficiently annotated' raw terms, which we call the type synthesis problem to distinguish it from type inference.
 Annotations therefore play an important role even in the definition of the problem solved by bidirectional typing, but have not received enough attention.
@@
 In order to make the notion of mode-correctness precise, we first give a general definition of bidirectional simple type systems, called \emph{bidirectional binding signature}, extending the typed version of \varcitet{Aczel1978}{'s binding signature} with modes.
-A general definition of typed languages allows us to define mode-correctness and to investigate its consequences rigorously, including the uniqueness of synthesised types and the decidability of bidirectional type synthesis for mode-correct signatures.
+A general definition of typed languages allows us to define mode-correctness and to investigate its consequences rigorously: the uniqueness of synthesised types and the decidability of bidirectional type synthesis for mode-correct signatures.
 The proof of the latter theorem amounts to a generator of proof-relevant bidirectional type synthesisers (analogous to a parser generator working for unambiguous or disambiguated grammars).
 
-To make our exposition accessible, the theory in this paper focuses on simply typed languages with a \emph{syntax-directed} bidirectional type system, so that the decidability of bidirectional type synthesis can be established without any other technical assumptions.
+To make our exposition accessible, the theory in this paper focuses on simply typed languages with a syntax-directed bidirectional type system, so that the decidability of bidirectional type synthesis can be established without any other technical assumptions.
 %the idea of extending a signature with bidirectionality should be clear enough for further generalisation.
 It should be possible to extend the theory to deal with more expressive types and assumptions other than mode-correctness.
-For instance, we briefly discuss how the theory could be extended to handle polymorphically typed languages such as \SystemF, \SystemFsub, and those systems using implicit type applications with additional assumptions in \cref{sec:future}.
+For instance, we briefly discuss how the theory can be extended to handle polymorphically typed languages such as \SystemF, \SystemFsub, and those systems using implicit type applications with additional assumptions in \cref{sec:future}.
 
@@
   \item mode derivations for explicitly dealing with annotations in the theory, and mode decoration for helping the user to work with annotations in practice;
-  \item rigorously proven consequences of mode-correctness---the uniqueness of synthesised types and the decidability of bidirectional type synthesis, which amounts to
+  \item rigorously proven consequences of mode-correctness, including the uniqueness of synthesised types and the decidability of bidirectional type synthesis, which amounts to
   \item a fully verified generator of proof-relevant type-synthesisers.

### 2024-01-12 tex/ESOP24/S3-defs.tex
[ tex ] Eliminate most of hyphens the end of every line

@@
 \begin{example}
-  Observe that $\Rule{Abs}$ and $\Rule{App}$ rules in \Cref{fig:STLC-typing-derivations} can be read as
+  Observe that the $\Rule{Abs}$ and $\Rule{App}$ rules in \Cref{fig:STLC-typing-derivations} can be read as
   \renewcommand{\eqnhighlightheight}{\vphantom{\Gamma}}
@@
   \begin{inlineenum}
-    \item variables in $V$,
+    \item (indices of) variables in $V$,
     \item annotations $\isTerm{t \annotatecolon A}$ for some raw term $t$ in $V$ and a type $A$, and
@@
     \item a derivation of $\Gamma |-_{\Sigma, \Omega} (t \annotatecolon A) : A$ if $\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A$ has a derivation, and
-    \item a derivation of $\Gamma |-_{\Sigma, \Omega} \tmOpts : \simsub{A_0}{\rho}$ for an operation $\bop$ if there are $\rho\colon \Xi \to \Type_{\Sigma}(\emptyset)$ and a derivation of $\Gamma, \vars{x}_{\isTerm{i}} : \simsub{\Delta_i}{\rho} |-_{\Sigma, \Omega} \isTerm{t_i} : \simsub{A_i}{\rho}$ for each $i$,
+    \item a derivation of $\Gamma |-_{\Sigma, \Omega} \tmOpts : \simsub{A_0}{\rho}$ for some operation $\bop$ if there exist $\rho\colon \Xi \to \Type_{\Sigma}(\emptyset)$ and a derivation of $\Gamma, \vars{x}_{\isTerm{i}} : \simsub{\Delta_i}{\rho} |-_{\Sigma, \Omega} \isTerm{t_i} : \simsub{A_i}{\rho}$ for each $i$,
   \end{enumerate}
@@
 \begin{example}\label{ex:signature-simply-typed-lambda}
-Having established generic definitions, we can now specify simply typed $\lambda$-calculus and its bidirectional type system---including raw terms, (bidirectional) typing derivations, and mode derivations---using just a pair of signatures $\Sigma_{\bto}$ (\Cref{ex:type-signature-for-function-type}) and $\Omega^{\Leftrightarrow}_\Lambda$ which consists of
+Having established generic definitions, we can now specify the simply typed $\lambda$-calculus and its bidirectional type system---including raw terms, (bidirectional) typing derivations, and mode derivations---using just a pair of signatures $\Sigma_{\bto}$ (\Cref{ex:type-signature-for-function-type}) and $\Omega^{\Leftrightarrow}_\Lambda$ which consists of
 \[

### 2024-01-12 tex/ESOP24/S4-pre-synthesis.tex
[ tex ] Eliminate most of hyphens the end of every line

@@
 In a sense, annotatability is a reasonable form of completeness: if a term of a simply typed language $(\Sigma, \erase\Omega)$ is typable in the ordinary type system, it may not be directly typable in the bidirectional type system $(\Sigma, \Omega)$ due to some missing annotations, but will be if those annotations are added correctly.
-In our theory, \cref{thm:annotatability} can be straightforwardly proved by induction on the derivation produced by generalised mode decoration~(\cref{thm:generalised-mode-decoration}) to construct a bidirectional typing derivation in the same mode.
+In our theory, \cref{thm:annotatability} can be straightforwardly proved by induction on the derivation given by generalised mode decoration~(\cref{thm:generalised-mode-decoration}) to construct a bidirectional typing derivation in the same mode.
 The interesting case is $\SynRule{Missing}$, which is mapped to $\SynRule{Anno}$, adding to the term a type annotation that comes from the given typing derivation.
@@
 Now, annotatability is a kind of completeness because (roughly speaking) it turns an ordinary typing derivation bidirectional.
-Therefore it is conceivable that we could use annotatability in place of completeness in the proof of \cref{thm:implementation}.
+Hence, it is conceivable that we could use annotatability in place of completeness in the proof of \cref{thm:implementation}.
 However, in the contrapositive form of annotatability, the antecedent is `there does not exist $t'$ that is more annotated than $t$ and has a bidirectional typing derivation', which is more complex than the bidirectional type synthesiser would have to produce.

### 2024-01-12 tex/ESOP24/S8-future.tex
[ tex ] Eliminate most of hyphens the end of every line

@@
 
-The study of presenting logics universally at least date back to universal algebra and model theory where structures are studied for certain notions of arities and signatures.
+The idea of presenting logics universally at least date back to universal algebra and model theory where structures are studied for certain notions of arities and signatures.
 In programming language theory, \varcitet{Aczel1978}{'s binding signature} is an example which has been used to prove a general confluence theorem.
-Many general definitions and frameworks for defining logics and type theories have been proposed, and we classify them into two groups by where signatures reside: the meta level or the object level of a meta-language:
+Many general definitions and frameworks for defining logics and type theories have been proposed and can be classified into two groups by where signatures reside: the meta level or the object level of a meta-language:
 \begin{enumerate}
@@
 
-For the second group, theories developed in set theory can often be reformulated in type theory and thus manageable for formalisation in a type-theoretic proof assistant. 
+For the second group, theories developed in set theory can often be restated in type theory and thus manageable for formalisation in a type-theoretic proof assistant. 
 Such examples include frameworks developed by \citet{Allais2021,Ahrens2022,Fiore2022} in \Agda, which are relatively correct to their meta-language, although their expressiveness is limited to simply typed theories.
@@
 
-To explore general settings, an extreme possibility is to independently specify an ordinary and a bidirectional type system over the same raw terms and then investigate relationships between the two systems.
+To explore general settings, an extreme choice is to independently specify an ordinary and a bidirectional type system over the same raw terms and then investigate relationships between the two systems.
 

### 2024-01-15 tex/ESOP24/S5-synthesis.tex
[ tex ] Fix minor issues

@@
         For an argument list $\biargs$ and any partial substitution $\rho$ from $\Xi$ to $\emptyset$
+        either
         \begin{enumerate}
-          \item either there is a minimal extension $\ext{\rho}$ of $\rho$ such that 
+          \item there is a minimal extension $\ext{\rho}$ of $\rho$ such that 
             \begin{equation} \label{eq:claim}
@@
             \end{equation}
-            for $i = 1, \ldots, n$
-          \item or there is no extension $\sigma$ of $\rho$ such that \eqref{eq:claim} holds.
+            for $i = 1, \ldots, n$, or
+          \item there is no extension $\sigma$ of $\rho$ such that \eqref{eq:claim} holds.
         \end{enumerate}

### 2024-01-15 tex/ESOP24/S8-future.tex
[ tex ] Fix minor issues

@@
 For the second group, theories developed in set theory can often be restated in type theory and thus manageable for formalisation in a type-theoretic proof assistant. 
-Such examples include frameworks developed by \citet{Allais2021,Ahrens2022,Fiore2022} in \Agda, which are relatively correct to their meta-language, although their expressiveness is limited to simply typed theories.
+Such examples include frameworks developed by \citet{Ahrens2022} in \Coq and \citet{Allais2021,Fiore2022} in \Agda, which are relatively correct to their meta-language, although their expressiveness is limited to simply typed theories.
 Our work belongs to the second group, as we aim for a formalism in a type theory to minimise the gap between theory and implementation.

### 2024-01-16 tex/ESOP24/S1-intro.tex
[ tex ] Mention the availability of artefact more explicitly

@@
 \end{enumerate}
-Our theory is fully formally developed~\cite{Chen2024a} with \Agda but translated to the mathematical vernacular for presentation in this paper.
-The formal theory doubles as a verified implementation, with no gap in between.
+Our theory is fully formally developed with \Agda but translated to the mathematical vernacular for presentation in this paper.
+The formal theory doubles as a verified implementation which is available publicly on Zenodo~\cite{Chen2024a}.
 

### 2024-01-16 tex/ESOP24/S5-synthesis.tex
[ tex ] Mention the availability of artefact more explicitly

@@
       \end{itemize}
+      We have proved the decidability by induction on the derivation $\erase{\Gamma} |-_{\Sigma, \Omega} t^{\dir{d}}$, assuming the claim which remains to be proved as follows.
       \begin{proof}[of Claim]
@@
         \end{enumerate}
-        Therefore, we have proved our claim for any argument list by induction.\hfill
+        We have proved our claim for any argument list by induction.\hfill
         $\blacksquare$
@@
 %  \end{itemize}
-  We have completed the proof by induction on the derivation $\erase{\Gamma} |-_{\Sigma, \Omega} t^{\dir{d}}$.\qed
+      We have completed the proof of \cref{thm:bidirectional-type-synthesis-checking}. \qed
 \end{proof}

### 2024-01-18 tex/ESOP24/BiSig.tex
[ tex ] Fight with the page limit
[ tex ] Include the running head
Sprinkled the paper with commas; miscellaneous revisions

@@
-\documentclass[orivec,envcountsect,envcountsame]{llncs}
+\documentclass[runningheads,orivec,envcountsect,envcountsame]{llncs}
 
@@
 \author{Liang-Ting Chen\orcidID{0000-0002-3250-1331} \and Hsiang-Shang Ko\orcidID{0000-0002-2439-1048}}
-\institute{Institute of Information Science, Academia Sinica, Taiwan}
-\title{A formal treatment of bidirectional typing}
+\institute{Institute of Information Science, Academia Sinica, Taipei, Taiwan}
+\title{A Formal Treatment of Bidirectional Typing}
+\authorrunning{L.-T.~Chen and H.-S.~Ko}
 
@@
 %This situation is in stark contrast to parsing, for which there have been general theories and widely applicable tools such as parser generators.
-To remedy the situation, this paper develops a \emph{general} and \emph{formal} theory of bidirectional typing for simply typed languages: for every signature that specifies a mode-correct bidirectionally typed language, there exists a \emph{proof-relevant} type synthesiser that for an input abstract syntax tree constructs a typing derivation if any, gives its refutation if not, or reports that the input does not have enough type annotations.
-Sufficient conditions for deriving a type synthesiser such as soundness, completeness, and mode-correctness, are studied universally for all signatures.
-We propose a preprocessing step called \emph{mode decoration}, which helps the user to deal with missing type annotations in a raw term.
-The entire theory is formally implemented in \Agda, so as a by-product of our formalism we provide a \emph{verified} generator of proof-relevant type synthesisers.
+To remedy the situation, this paper develops a \emph{general} and \emph{formal} theory of bidirectional typing for simply typed languages: for every signature that specifies a mode-correct bidirectionally typed language, there exists a \emph{proof-relevant} type synthesiser which, given an input abstract syntax tree, constructs a typing derivation if any, gives its refutation if not, or reports that the input does not have enough type annotations.
+Sufficient conditions for deriving a type synthesiser such as soundness, completeness, and mode-correctness are studied universally for all signatures.
+We propose a preprocessing step called \emph{mode decoration}, which helps the user to deal with missing type annotations.
+The entire theory is formally implemented in \Agda, so we provide a \emph{verified} generator of proof-relevant type synthesisers as a by-product of our formalism.
 \end{abstract}
+
+% Due to ESOP's page limit, the appendix is only available in the ArXiv version.
+
+\newif\ifarxiv
 
@@
 
+\ifarxiv
 \subfile{S7-formalisation}
+
 \input{appendix}
+\fi
 

### 2024-01-18 tex/ESOP24/S1-intro.tex
Sprinkled the paper with commas; miscellaneous revisions

--- 2024-01-12 tex/ESOP24/S1-intro.tex
@@
 In light of these limitations, \emph{bidirectional type synthesis} emerged as a viable alternative, deciding the types of raw terms that meet some syntactic criteria and typically contain annotations.
-In their survey paper, \varcitet{Dunfield2021}{ summarised} the principles of bidirectional type synthesis and its wide coverage of languages with simple, polymorphic, dependent, and gradual types, among others.
+In their survey paper~\cite{Dunfield2021}, Dunfield and Krishnaswami summarised the principles of bidirectional type synthesis and its wide coverage of languages with simple, polymorphic, dependent, and gradual types, among others.
 
-While type inference is not decidable in general, for certain kinds of terms it is still possible to infer their types.
+While type inference is not decidable in general, for certain kinds of terms it is still possible to synthesise their types.
 For example, the type of a variable can be looked up in the context.
-Bidirectional type synthesis combines type \emph{inference} on this subset of terms with type \emph{checking} (based on a given type) on the rest.
+Bidirectional type synthesis combines type \emph{synthesis} on this subset of terms with type \emph{checking} (based on a given type) on the rest.
 Formally, every judgement in a bidirectional type system is extended with a \emph{mode:}
@@
 %While it is usually straightforward to derive a bidirectional type synthesis algorithm,
-\citeauthor{Dunfield2021} present informal design principles learned from individual bidirectional type systems, but in addition to crafting special techniques for individual systems, we should start to consolidate concepts common to a class of bidirectional type systems into a general and formal theory that gives mathematically precise definitions and proves theorems for the class of systems once and for all.
+Dunfield and Krishnaswami present informal design principles learned from individual bidirectional type systems, but in addition to crafting special techniques for individual systems, we should start to consolidate concepts common to a class of bidirectional type systems into a general and formal theory that gives mathematically precise definitions and proves theorems for the class of systems once and for all.
 %Moreover, by mechanising the theory, we get \emph{verified} type-synthesiser generators analogous to parser generators.
@@
 can be transformed to a well-typed term $\isTerm{t}'$ of type~$A$ in the core language~\cite{Pierce2000}.
-Such an algorithm is accompanied by soundness and completeness assertions that the algorithm correctly synthesises the type of a raw term and every typable term can be synthesised.
-By contrast, the proof-relevant approach exploits the simultaneously computational and logical nature of Martin-L\"of type theory and formulates algorithmic soundness, completeness, and decidability \emph{in one go}.
+Such an algorithm is accompanied by soundness and completeness assertions that the algorithm correctly synthesises the type of a raw term, and every typable term can be synthesised.
+By contrast, the proof-relevant approach exploits the simultaneously computational and logical nature of Martin-L\"of type theory, and formulates algorithmic soundness, completeness, and decidability \emph{in one go}.
 %(although they do not emphasise the difference of their approach from the traditional one)
@@
 Annotations therefore play an important role even in the definition of the problem solved by bidirectional typing, but have not received enough attention.
-In our theory, we define \emph{mode derivations} to explicitly take annotations into account and formulate the type synthesis problem with sufficiently annotated raw terms.
+In our theory, we define \emph{mode derivations} to explicitly take annotations into account, and formulate the type synthesis problem with sufficiently annotated raw terms.
 Accordingly, a preprocessing step called \emph{mode decoration} is proposed to help the user to work with annotations.
@@
 Then a bidirectional type synthesiser is only required to decide the typability of mode-decorated raw terms.
-Soundness and completeness of bidirectional typing is reformulated as a one-to-one correspondence between bidirectional typing derivations and pairs of a typing derivation and a mode derivation for the same raw term; this reformulation is more useful than annotatability~\cite[Section~3.2]{Dunfield2021} which is typically formulated in the literature of bidirectional typing.
+Soundness and completeness of bidirectional typing is reformulated as a one-to-one correspondence between bidirectional typing derivations and pairs of a typing derivation and a mode derivation for the same raw term.
+Our completeness is simpler and more useful than annotatability, which is a typical formulation of completeness in the literature of bidirectional typing~\cite[Section~3.2]{Dunfield2021}.
 
@@
 The most essential characteristics of bidirectional typing is \emph{mode-correctness}, since an algorithm can often be `read off' from the definition of a bidirectionally typed language if mode-correct.
-As illustrated by \citet{Dunfield2021}, it seems that the implications of mode-correctness have only been addressed informally so far and mode-correctness is not yet formally defined as a \emph{property of languages}.
+As illustrated by \citet{Dunfield2021}, it seems that the implications of mode-correctness have only been addressed informally so far, and mode-correctness is not yet formally defined as a \emph{property of languages}.
 
@@
 \end{enumerate}
-Our theory is fully formally developed~\cite{Chen2024a} with \Agda but translated to the mathematical vernacular for presentation in this paper.
-The formal theory doubles as a verified implementation, with no gap in between.
+Our theory is fully formally developed with \Agda, but is translated to the mathematical vernacular for presentation in this paper.
+The formal theory doubles as a verified implementation, which is available publicly on Zenodo~\cite{Chen2024a}.
 
@@
 This paper is structured as follows.
-We first present a concrete overview of our theory using simply typed $\lambda$-calculus in \cref{sec:key-ideas}, prior to developing a general framework for specifying bidirectional type systems in \cref{sec:defs}.
+We present a concrete overview of our theory using simply typed $\lambda$-calculus in \cref{sec:key-ideas}, prior to developing a general framework for specifying bidirectional type systems in \cref{sec:defs}.
 Following this, we discuss mode decoration and related properties in \cref{sec:pre-synthesis}.
 The main technical contribution lies in \cref{sec:type-synthesis}, where we introduce mode-correctness and bidirectional type synthesis.
-Some examples other than simply typed $\lambda$-calculus are given in \Cref{sec:example} and further developments are discussed in \cref{sec:future}.
+Some examples other than simply typed $\lambda$-calculus are given in \Cref{sec:example}, and further developments are discussed in \cref{sec:future}.
 \ifarxiv

### 2024-01-18 tex/ESOP24/S2-outline.tex
[ tex ] Minor changes
[ tex ] Fix minor issues
Sprinkled the paper with commas; miscellaneous revisions

@@
 To give more precise definitions: the raw terms for simply typed $\lambda$-calculus are defined%
-\footnote{The usual condition about named representations of variables is omitted.}
+\footnote{The usual conditions about named representations of variables are omitted.}
 in \cref{fig:STLC-raw-terms};
@@
 Correspondingly, the definition of typing derivations%
-\footnote{We write~`$\bto$' instead of~`$\to$' for the function types of the simply typed $\lambda$-calculus to avoid confusion with the function types in our type-theoretic meta-language.}
+\footnote{We write~`$\bto$' instead of~`$\to$' for the function types of simply typed $\lambda$-calculus to avoid confusion with the function types in our type-theoretic meta-language.}
 in \cref{fig:STLC-typing-derivations} has an \textsc{Anno} rule enforcing that the type of an annotated term does match the annotation.
@@
 \begin{defn} \label{def:STLC-mode-decorator}
-A \emph{mode decorator} decides for any raw term~$V \vdash \isTerm{t}$ if $V \vdash \isTerm{t}^{\syn}$.
+  A \emph{mode decorator} decides for a raw term~$V \vdash \isTerm{t}$ whether $V \vdash \isTerm{t}^{\syn}$.
 \end{defn}

### 2024-01-18 tex/ESOP24/S3-defs.tex
Sprinkled the paper with commas; miscellaneous revisions

@@
 \begin{example} \label{ex:type-signature-for-function-type}
-  Function types $A \bto B$ and typically a base type~$\mathtt{b}$ are included in simply typed $\lambda$-calculus and can be specified by the type signature $\Sigma_{\bto}$ consisting of operations $\mathsf{fun}$ and $\mathsf{b}$ where $\arity(\mathsf{fun}) = 2$ and $\arity(\mathsf{b}) = 0$.
-  Then, all types in simply typed $\lambda$-calculus can be given as $\Sigma_{\bto}$-types over the empty set with $A \bto B$ introduced as $\tyOp_{\mathsf{fun}}(A, B)$ and $\mathtt{b}$ as $\tyOp_{\mathsf{b}}$.
+  Function types $A \bto B$ and typically a base type~$\mathtt{b}$ are included in simply typed $\lambda$-calculus, and can be specified by the type signature $\Sigma_{\bto}$ consisting of operations $\mathsf{fun}$ and $\mathsf{b}$ where $\arity(\mathsf{fun}) = 2$ and $\arity(\mathsf{b}) = 0$.
+  Then, all types in simply typed $\lambda$-calculus can be given as $\Sigma_{\bto}$-types over the empty set, with $A \bto B$ introduced as $\tyOp_{\mathsf{fun}}(A, B)$ and $\mathtt{b}$ as $\tyOp_{\mathsf{b}}$.
 \end{example}
@@
 \begin{inlineenum}
-  \item a family of sets of raw terms $\isTerm{t}$ indexed by a list~$V$ of variables (that are currently in scope) where each construct is allowed to bind some variables like $\Rule{Abs}$ and to take multiple arguments like $\Rule{App}$;
+  \item a family of sets of raw terms $\isTerm{t}$ indexed by a list~$V$ of variables (that are currently in scope), where each construct is allowed to bind some variables like $\Rule{Abs}$ and to take multiple arguments like $\Rule{App}$;
   \item a family of sets of typing derivations indexed by a typing context~$\Gamma$, a raw term~$\isTerm{t}$, and a type~$A$.
@@
 \begin{defn}[{\cite[p.~322]{Fiore2010}}]\label{def:binding-arity}
-  A \emph{binding arity} with a set $T$ of types is an inhabitant of $\left(T^* \times T\right)^* \times T$ where $T^*$ is the set of lists over $T$.
+  A \emph{binding arity} with a set $T$ of types is an inhabitant of $\left(T^* \times T\right)^* \times T$, where $T^*$ is the set of lists over $T$.
   In a binding arity $(((\Delta_1, A_1), \ldots, (\Delta_n, A_n)), A)$, every $\Delta_i$ and $A_i$ refers to the \emph{extension context} and the \emph{type of the $i$-th argument}, respectively, and $A$ the \emph{target type}.
-  For brevity, it is denoted by $\bargs \to A$ where $[\Delta_i]$ is omitted if empty.
+  For brevity, it is denoted by $\bargs \to A$, where $[\Delta_i]$ is omitted if empty.
 \end{defn}
@@
   \end{equation*}
-  \annotate[yshift=1.5em]{above}{ext1,ext2,ext3}{extension context}
-  \annotate[yshift=0.4em]{above,right}{aty1,aty2,aty3}{argument type}
-  \annotate[yshift=0em]{below}{ty1,ty2}{target type}
+  \annotate[yshift=1.5em]{above}{ext1,ext2,ext3}{extension contexts}
+  \annotate[yshift=0.4em]{above,right}{aty1,aty2,aty3}{argument types}
+  \annotate[yshift=0em]{below}{ty1,ty2}{target types}
   \vspace{.5em}
@@
 The definition of typing derivations is a bit more involved.
-We need some information to compare types on the object level during type synthesis and substitute those type variables in a typing derivation $\Gamma \vdash \tmOpts : A$ for an operation $o$ in $\Omega$ at some point.
+We need some information to compare types on the object level during type synthesis and substitute those type variables in a typing derivation of $\Gamma \vdash \tmOpts : A$ for an operation $o$ in $\Omega$ at some point.
 Here we choose to include a substitution $\rho$ from the local context $\Xi$ to $\emptyset$ as part of its typing derivation explicitly:
@@
 \subsection{Bidirectional binding signatures and bidirectional type systems} \label{subsec:bidirectional-system}
-For a bidirectional type system, typing judgements appear in two forms: $\Gamma |- t \syn A$ and $\Gamma |- t \chk A$, but these two typing judgements can be considered as a single typing judgement $\Gamma |- t :^\dir{d} A$, additionally indexed by a \emph{mode} $d : \Mode$---which can be either $\syn$ or $\chk$.
+Typing judgements for a bidirectional type system appear in two forms: $\Gamma |- t \syn A$ and $\Gamma |- t \chk A$.
+These two typing judgements can be considered as a single typing judgement $\Gamma |- t :^\dir{d} A$ indexed by a \emph{mode} $d : \Mode$, which can be either $\syn$ or $\chk$.
 Therefore, to define a bidirectional type system, we enrich the concept of binding arity to \emph{bidirectional binding arity}, which further specifies the mode for each of its arguments and for the conclusion:
@@
 We write $\biop$ for an operation $o$ with a variable set $\Xi$ and its bidirectional binding arity with $\Type_{\Sigma}(\Xi)$ as types.
-We call it \emph{checking} if $d$ is ${\chk}$ or \emph{synthesising} if $d$ is ${\syn}$; similarly its $i$-th argument is checking if $d_i$ is $\chk$ and synthesising if $d_i$ is $\syn$.
+We call it \emph{checking} if $d$ is ${\chk}$ or \emph{synthesising} if $d$ is ${\syn}$; similarly, its $i$-th argument is checking if $d_i$ is $\chk$ and synthesising if $d_i$ is $\syn$.
 A bidirectional type system $(\Sigma, \Omega)$ refers to a pair of a type signature $\Sigma$ and a bidirectional binding signature $\Omega$.
@@
   \begin{itemize}
-    \item the set of \emph{bidirectional typing derivations} of $\Gamma \vdash_{\Sigma, \Omega} t :^\dir{d} A$, indexed by a typing context $\Gamma$, a raw term $\isTerm{t}$ under $\erase{\Gamma}$, a mode $\dir{d}$, and a type $A$, is defined in \Cref{fig:bidirectional-typing-derivations} and particularly
+    \item the set of \emph{bidirectional typing derivations} of $\Gamma \vdash_{\Sigma, \Omega} t :^\dir{d} A$, indexed by a typing context $\Gamma$, a raw term $\isTerm{t}$ under $\erase{\Gamma}$, a mode $\dir{d}$, and a type $A$, is defined in \Cref{fig:bidirectional-typing-derivations}, and particularly
           \[
@@
 \begin{example}\label{ex:signature-simply-typed-lambda}
-Having established generic definitions, we can now specify the simply typed $\lambda$-calculus and its bidirectional type system---including raw terms, (bidirectional) typing derivations, and mode derivations---using just a pair of signatures $\Sigma_{\bto}$ (\Cref{ex:type-signature-for-function-type}) and $\Omega^{\Leftrightarrow}_\Lambda$ which consists of
+Having established generic definitions, we can now specify simply typed $\lambda$-calculus and its bidirectional type system---including raw terms, (bidirectional) typing derivations, and mode derivations---using just a pair of signatures $\Sigma_{\bto}$ (\Cref{ex:type-signature-for-function-type}) and $\Omega^{\Leftrightarrow}_\Lambda$ which consists of
 \[

### 2024-01-18 tex/ESOP24/S4-pre-synthesis.tex
Sprinkled the paper with commas; miscellaneous revisions

@@
 We discuss some related properties: by bringing mode derivations into the picture, we are able to give a natural formulation of soundness and completeness of a bidirectional type system with respect to its erasure to an ordinary type system in \cref{sec:soundness-and-completeness}.
-We formulate annotatability~\cite{Dunfield2021} and discuss its relationship with our completeness and generalised mode decoration in \cref{sec:annotatability}.
+We also reformulate annotatability~\cite[Section~3.2]{Dunfield2021} and compare it with our completeness and generalised mode decoration in \cref{sec:annotatability}.
 
@@
 Erasure of a bidirectional binding signature removes modes and keeps everything else intact; this can be straightforwardly extended by induction to remove modes from a bidirectional typing derivation and arrive at an ordinary typing derivation, which is soundness.
-Instead we can remove typing and retain modes, arriving at a mode derivation.
+Alternatively, we can remove typing and retain modes, arriving at a mode derivation.
 Conversely, if we have both mode and typing derivations for the same term, we can combine them to form a bidirectional typing derivation, which is completeness.
@@
 Another way to understand the role of~$s$ is that it makes the $\SynRule{Missing}$ rule precise: an annotation is truly missing only when a term is \emph{strictly} in checking mode but is required to be in synthesising mode.
-With strictness coming into play, non-genuine derivations are now `truly non-genuine'.
+The explicit formulation of strictness makes non-genuine derivations `truly non-genuine', and \cref{thm:Pre?-false} can be proved.
 
@@
 \citet[Section~3.2]{Dunfield2021} formulated completeness differently from our \cref{lem:soundness-completeness} and proposed \emph{annotatability} as a more suitable name.
-In our theory, we may formulate annotatability as follows.
+In our theory, we may reformulate annotatability as follows.
 

### 2024-01-18 tex/ESOP24/S5-synthesis.tex
Sprinkled the paper with commas; miscellaneous revisions

@@
   of \emph{(free) variables} of a type $A$ and of variables in an extension context~$\Delta$ are denoted by $\fv(A)$ and $\fv(\Delta)$ respectively.
-  For an argument list $\biargs$ the set of type variables $A_{i}^{\dir{d_i}}$ with $d_i$ being $\syn$ is denoted by $\synvar(\biargs)$, i.e.\ $\synvar$ gives the set of type variables that will be synthesised during type synthesis.
+  For an argument list $\biargs$, the set of type variables $A_{i}^{\dir{d_i}}$ with $d_i$ being $\syn$ is denoted by $\synvar(\biargs)$, i.e.\ $\synvar$ gives the set of type variables that will be synthesised during type synthesis.
 \end{defn}
@@
 
-It is easy to check the bidirectional type system $(\Sigma_{\bto}, \Omega^{\Leftrightarrow}_{\Lambda}$) for the simply typed $\lambda$-calculus is mode-correct by definition or by the following lemma:
+It is easy to check the bidirectional type system $(\Sigma_{\bto}, \Omega^{\Leftrightarrow}_{\Lambda}$) for simply typed $\lambda$-calculus is mode-correct by definition or by the following lemma:
 \begin{lemma}\label{lem:decidability-mode-correctness}
@@
   \begin{itemize}
-    \item $\SynRule{Var}$ follows from that each variable as a raw term refers to the same variable in its context.
+    \item $\SynRule{Var}$ follows from the fact that each variable as a raw term refers to the same variable in its context.
     \item $\SynRule{Anno}$ holds trivially, since the synthesised type $A$ is from the term $t \annotatecolon A$ in question.
@@
         \item If $\dir{d_{i+1}}$ is~$\chk$, then the statement holds by induction hypothesis.
-        \item If $\dir{d_{i+1}}$ is~$\syn$, then $\simsub{\Delta_{i+1}}{\rho_1} = \simsub{\Delta_{i+1}}{\rho_2}$ and induction hypothesis (of the list).
+        \item If $\dir{d_{i+1}}$ is~$\syn$, then $\simsub{\Delta_{i+1}}{\rho_1} = \simsub{\Delta_{i+1}}{\rho_2}$ by induction hypothesis (of the list).
           Therefore, under the same context $\Gamma, \simsub{\Delta_{i+1}}{\rho_1} = \Gamma, \simsub{\Delta_{i+1}}{\rho_2}$ the term $t_{i+1}$ must have the same synthesised type $\simsub{A_{i+1}}{\rho_1} = \simsub{A_{i+1}}{\rho_1}$ by induction hypothesis (of the typing derivation), so $\rho_1$ and $\rho_2$ agree on $\fv(A_{i+1})$ in addition to $\synvar(\biargs)$.
@@
 \end{lemma}
-This lemma can be inferred from the correctness of first-order unification~\citep{McBride2003,McBride2003a}, or be proved directly without unification.
+This lemma can be derived from the correctness of first-order unification~\citep{McBride2003,McBride2003a}, or be proved directly without unification.
 We are now ready for \cref{thm:bidirectional-type-synthesis-checking}:
@@
   The case \ChkRule{Sub} invokes the uniqueness of synthesised types to refute the case that $\Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn B$ but $A \neq B$ for a given type $A$.
-  The first three cases follow essentially the same reasoning provided by \citet{Wadler2022}, so we only detail the last case \Rule{Op} which is new and has been discussed informally above.
+  The first three cases follow essentially the same reasoning provided by \citet{Wadler2022}, so we only detail the last case \Rule{Op}, which is new (but has been discussed informally above).
   For brevity we omit the subscript $(\Sigma, \Omega)$.
@@
       \begin{claim}\label{lem:args-induction}
-        For an argument list $\biargs$ and any partial substitution $\rho$ from $\Xi$ to $\emptyset$
+        For an argument list $\biargs$ and any partial substitution $\rho$ from $\Xi$ to $\emptyset$,
         either
@@
               for every $i$.
-              Obviously, $\rho \geq \rho_0$ and ${\Gamma, \vars{x}_\isTerm{i} : \simsub{\Delta_i}{\rho} |- \isTerm{t_i} \colon \simsub{A_i}{\rho}^{\dir{d_i}}}$ for every $i$ and it contradicts the assumption that no such an extension exists.
+              Obviously, $\rho \geq \rho_0$ and ${\Gamma, \vars{x}_\isTerm{i} : \simsub{\Delta_i}{\rho} |- \isTerm{t_i} \colon \simsub{A_i}{\rho}^{\dir{d_i}}}$ for every $i$,  which contradict the assumption that no such extension exists.
 
-            \item If there exists a minimal $\ext{\rho} \geq \rho_0$ defined on $\synvar(\biargs)$ such that \eqref{eq:claim} holds, then by mode-correctness $\ext{\rho}$ is total and thus
+            \item If there exists a minimal $\ext{\rho} \geq \rho_0$ defined on $\synvar(\biargs)$ such that \eqref{eq:claim} holds, then by mode-correctness $\ext{\rho}$ is total, and thus
               \[
@@
               \begin{enumerate}
-                \item If there is no $\sigma \geq \ext{\rho}$ satisfying \eqref{eq:claim} but $\Gamma |- \tmOpts \chk A$, then by inversion there is $\gamma$ s.t.\ $\simsub{A_0}{\gamma} = A$ and, for every $i$, ${\Gamma, \vars{x}_\isTerm{i} : \simsub{\Delta_i}{\gamma} |- \isTerm{t_i} \colon \simsub{A_i}{\gamma}^{\dir{d_i}}}$.
-                  Given that $\ext{\rho} \geq \rho$ is minimal s.t.\ $\simsub{A_0}{\ext{\rho}} = A$, then $\gamma$ is an extension of $\ext{\rho}$ but by assumption no such an extension satisfying ${\Gamma, \vars{x}_\isTerm{i} : \simsub{\Delta_i}{\gamma} |- \isTerm{t_i} \colon \simsub{A_i}{\gamma}^{\dir{d_i}}}$ exists, thus a contradiction.
+                \item If no $\sigma \geq \ext{\rho}$ satisfies \eqref{eq:claim} but $\Gamma |- \tmOpts \chk A$, then by inversion there is $\gamma$ s.t.\ $\simsub{A_0}{\gamma} = A$ and ${\Gamma, \vars{x}_\isTerm{i} : \simsub{\Delta_i}{\gamma} |- \isTerm{t_i} \colon \simsub{A_i}{\gamma}^{\dir{d_i}}}$ for every $i$.
+                  Given that $\ext{\rho} \geq \rho$ is minimal s.t.\ $\simsub{A_0}{\ext{\rho}} = A$, it follows that $\gamma$ is an extension of $\ext{\rho}$, but by assumption no such an extension satisfying ${\Gamma, \vars{x}_\isTerm{i} : \simsub{\Delta_i}{\gamma} |- \isTerm{t_i} \colon \simsub{A_i}{\gamma}^{\dir{d_i}}}$ exists, thus a contradiction.
 
@@
       \end{itemize}
-      We have proved the decidability by induction on the derivation $\erase{\Gamma} |-_{\Sigma, \Omega} t^{\dir{d}}$, assuming the claim which remains to be proved as follows.
+      We have proved the decidability by induction on the derivation $\erase{\Gamma} |-_{\Sigma, \Omega} t^{\dir{d}}$ assuming the claim, which remains to be proved.
       \begin{proof}[of Claim]
@@
                       \item If $\Gamma, \vars{x}_\isTerm{m+1} : \simsub{\Delta_{m+1}}{\ext{\rho}} |/- \isTerm{t_{m+1}} \syn A$ for any $A$ but there is $\gamma \geq \rho$ s.t.\ \eqref{eq:claim} holds for $1 \leq i \leq m+1$, then $\gamma \geq \ext{\rho}$.
-                        Therefore $\simsub{\Delta_{m+1}}{\ext{\rho}} = \simsub{\Delta_{m+1}}{\gamma}$ and we derive a contradiction because $\Gamma, \vars{x}_\isTerm{m+1} : \simsub{\Delta_{m+1}}{\ext{\rho}} |- \isTerm{t_{m+1}} \syn \simsub{A_{m+1}}{\gamma}$.
-                      \item If $\Gamma, \vars{x}_\isTerm{m+1} : \simsub{\Delta_{m+1}}{\ext{\rho}} |- \isTerm{t_{m+1}} \syn A$ for some $A$, then \Cref{lem:unify} gives following two cases:
+                        Therefore $\simsub{\Delta_{m+1}}{\ext{\rho}} = \simsub{\Delta_{m+1}}{\gamma}$, and we derive a contradiction because $\Gamma, \vars{x}_\isTerm{m+1} : \simsub{\Delta_{m+1}}{\ext{\rho}} |- \isTerm{t_{m+1}} \syn \simsub{A_{m+1}}{\gamma}$.
+                      \item If $\Gamma, \vars{x}_\isTerm{m+1} : \simsub{\Delta_{m+1}}{\ext{\rho}} |- \isTerm{t_{m+1}} \syn A$ for some $A$, then \Cref{lem:unify} gives the following two cases:
                         \begin{itemize}

### 2024-01-18 tex/ESOP24/S6-example.tex
Sprinkled the paper with commas; miscellaneous revisions

@@
 \subsection{Spine application}\label{subsec:spine}
-A spine application $t\;u_1\;\ldots\;u_n$ is a form of application that consists of a head term $t$ and an indeterminate number of arguments $u_1\;u_2\;\dots\;u_n$.
+A spine application $t\;u_1\;\ldots\;u_n$ is a form of application that consists of a head term $t$ and an indeterminate number of arguments $u_1$, $u_2$, \ldots, $u_n$.
 This arrangement allows direct access to the head term, making it practical in various applications, and has been used by \Agda's core language.
@@
 However, the code size increases proportionally to the number of type constructs and of arguments associated with each term construct.
-When dealing with a fixed number $n$ of type constructs, for each synthesising construct, there are two cases for a checking argument but $n + 1$ cases for each synthesising argument---the successful synthesis of the expected type, an instance where it fails, or $n-1$ cases where the expected type does not match; similarly for a checking construct---making the task tedious.
+For example, when dealing with a fixed number $n$ of type constructs, for each synthesising construct there are two cases for a checking argument but $n + 1$ cases for each synthesising argument: the successful synthesis of the expected type, an instance where it fails, or $n-1$ cases where the expected type does not match.
 Thus, having a generator is helpful and can significantly reduce the effort for implementation.
@@
 For illustrative purposes, consider a computational calculus~\cite{Moggi1989} with naturals, sums, products, and general recursion.
-The extended language has `only' 15 constructs, including pairing, projections, injections, and so on, whereas there are nearly 100 possible cases to consider in bidirectional type synthesis.
-However, it has far fewer constructs than a realistic programming language would have.
+The extended language has `only' 15 constructs, including pairing, projections, injections, and so on, and this number of constructs is still far fewer than what a realistic programming language would have.
+Even for this small calculus, there are already nearly 100 possible cases to consider in bidirectional type synthesis.
 

### 2024-01-18 tex/ESOP24/S8-future.tex
Sprinkled the paper with commas; miscellaneous revisions

--- 2024-01-12 tex/ESOP24/S8-future.tex
@@
 
-The idea of presenting logics universally at least date back to universal algebra and model theory where structures are studied for certain notions of arities and signatures.
+The idea of presenting logics universally at least date back to universal algebra and model theory, where structures are studied for certain notions of arities and signatures.
 In programming language theory, \varcitet{Aczel1978}{'s binding signature} is an example which has been used to prove a general confluence theorem.
@@
 \begin{enumerate}
-  \item \varcitet{Harper1993a}{'s logical framework \LF} and its family of variants~\cite{Harper2007,Assaf2016,Felicissimo2023} are \emph{extensions} of Martin-L\"of type theory where signatures are on the \emph{meta level} and naturally capable of specifying dependent type theories;
-\item general dependent type theories~\cite{Bauer2020,Haselwarter2021,Bauer2022a,Uemura2021}, categorical semantics~\cite{Fiore1999,Tanaka2006,Tanaka2006a,Fiore2010,Hamana2011,Fiore2013,Arkor2020,Fiore2022} (which includes the syntactic model as a special case), and frameworks for substructural systems~\cite{Tanaka2006,Tanaka2006a,Wood2022} are developed \emph{within} a meta-theory (set theory or type theory) where signatures are on the \emph{object level} and their expressiveness varies depending on their target languages.
+  \item \varcitet{Harper1993a}{'s logical framework \LF} and its family of variants~\cite{Harper2007,Assaf2016,Felicissimo2023} are \emph{extensions} of Martin-L\"of type theory, where signatures are on the \emph{meta level} and naturally capable of specifying dependent type theories;
+\item general dependent type theories~\cite{Bauer2020,Haselwarter2021,Bauer2022a,Uemura2021}, categorical semantics~\cite{Fiore1999,Tanaka2006,Tanaka2006a,Fiore2010,Hamana2011,Fiore2013,Arkor2020,Fiore2022} (which includes the syntactic model as a special case), and frameworks for substructural systems~\cite{Tanaka2006,Tanaka2006a,Wood2022} are developed \emph{within} a meta-theory (set theory or type theory), where signatures are on the \emph{object level} and their expressiveness varies depending on their target languages.
 \end{enumerate}
 
-The \LF approach is naturally expressive but each extension requires a different framework and a different implementation to check formal \LF proofs.
+The \LF approach is naturally expressive, but each extension requires a different framework and a different implementation to check formal \LF proofs.
 Formalising \LF and its variants is at least as hard as formalising a dependent type theory, and they are mostly implemented separately from their theory and unverified.
@@
 For the second group, theories developed in set theory can often be restated in type theory and thus manageable for formalisation in a type-theoretic proof assistant. 
-Such examples include frameworks developed by \citet{Allais2021,Ahrens2022,Fiore2022} in \Agda, which are relatively correct to their meta-language, although their expressiveness is limited to simply typed theories.
+Such examples include frameworks developed by \citet{Ahrens2022} in \Coq and by \citet{Allais2021} and \citet{Fiore2022} in \Agda, which are relatively correct to their meta-language, although their expressiveness is currently limited to simply typed theories.
+
 Our work belongs to the second group, as we aim for a formalism in a type theory to minimise the gap between theory and implementation.
@@
 which is not syntax-directed.
-By translating the rule to subtyping, we have the \emph{instantiation problem} which amounts to guessing $B$ in $\forall \alpha. A <: A[B/\alpha]$.
+By translating the rule to subtyping, we have the \emph{instantiation problem} that requires guessing $B$ in $\forall \alpha. A <: A[B/\alpha]$.
 A theory that accommodates various solutions to the problem is left as future work.
@@
 The algorithm is decidable, sound and complete with respect to mode-decorated terms, provided that the specified language is mode-correct, and that the set of computational rules is type-preserving, confluent, and strongly normalising.
-The last assumption, however, is challenging to establish even for specific languages and lacks a general understanding.
+The last assumption, however, is challenging to establish even for specific languages, and lacks a general treatment.
 
@@
 
-To explore general settings, an extreme choice is to independently specify an ordinary and a bidirectional type system over the same raw terms and then investigate relationships between the two systems.
+%To explore general settings, an extreme choice is to independently specify an ordinary and a bidirectional type system over the same raw terms and then investigate relationships between the two systems.
 
-Yet, we could first focus on simpler cases that the ordinary typing part is still syntax-directed, but each typing rule is refined to multiple bidirectional variants, including different orders of its premises.
-In such cases, the mode decorator would need to find all mode derivations, but the type synthesiser should still work in a syntax-directed manner on each mode derivation.
+To relax the assumption of syntax-directedness, we could start from a simpler case where the ordinary typing part is still syntax-directed, but each typing rule is refined to multiple bidirectional variants, including different orders of its premises.
+In such cases, the mode decorator would need to backtrack and find all mode derivations, but the type synthesiser should still work in a syntax-directed manner on each mode derivation.
 Completeness could still take the simple form presented in this paper too.
 
-The next step is to consider non-syntax-directed systems where each typing rule has multiple bidirectional variants.
-For this setting, the mode decorator and the bidirectional type synthesiser will have to backtrack.
-It is still possible to treat soundness as the separation of mode and type information, but completeness will pose a problem.
-For every raw term, a mode derivation chooses a mode assignment while a typing derivation chooses a typing rule, but there may not be a bidirectional typing rule for this particular combination.
-To retain the mode decoration, one could make the mode decorator produce all possible mode derivations, and refine completeness to say that any typing derivation can be combined with one of mode derivations into a bidirectional typing derivation.
+Next, we could consider systems where each construct can have multiple typing rules, which can have multiple bidirectional variants.
+In this setting, the bidirectional type synthesiser will also need to backtrack.
+It is still possible to treat soundness as the separation of mode and type information, but completeness will pose a problem:
+for every raw term, a mode derivation chooses a mode assignment while a typing derivation chooses a typing rule, but there may not be a bidirectional typing rule for this particular combination.
+A solution might be refining completeness to say that any typing derivation can be combined with one of the possible mode derivations into a bidirectional typing derivation.
 %It should be possible to fix this problem while retaining the mode decoration phase.

### 2024-01-19 tex/ESOP24/S5-synthesis.tex
[ tex ] Minor changes

@@
       \end{itemize}
-      We have proved the decidability by induction on the derivation $\erase{\Gamma} |-_{\Sigma, \Omega} t^{\dir{d}}$ assuming the claim, which remains to be proved.
+      We have proved the decidability by induction on the derivation of $\erase{\Gamma} |-_{\Sigma, \Omega} t^{\dir{d}}$, assuming the claim.
       \begin{proof}[of Claim]

### 2024-01-19 tex/ESOP24/S8-future.tex
[ tex ] Minor changes

@@
 For the second group, theories developed in set theory can often be restated in type theory and thus manageable for formalisation in a type-theoretic proof assistant. 
-Such examples include frameworks developed by \citet{Ahrens2022} in \Coq and by \citet{Allais2021} and \citet{Fiore2022} in \Agda, which are relatively correct to their meta-language, although their expressiveness is currently limited to simply typed theories.
+Such examples include frameworks developed by \citet{Ahrens2022,Allais2021,Fiore2022}, although these formal implementations are limited to simply typed theories for now.
 

### 2024-01-22 tex/ESOP24/S8-future.tex
[ tex ] Add Jason Reed

@@
 In programming language theory, \varcitet{Aczel1978}{'s binding signature} is an example which has been used to prove a general confluence theorem.
-Many general definitions and frameworks for defining logics and type theories have been proposed and can be classified into two groups by where signatures reside: the meta level or the object level of a meta-language:
+Many general definitions and frameworks for defining logics and type theories have been proposed and can be classified into two groups by where signatures reside---the meta level or the object level of a meta-language:
 \begin{enumerate}
-  \item \varcitet{Harper1993a}{'s logical framework \LF} and its family of variants~\cite{Harper2007,Assaf2016,Felicissimo2023} are \emph{extensions} of Martin-L\"of type theory, where signatures are on the \emph{meta level} and naturally capable of specifying dependent type theories;
+  \item \varcitet{Harper1993a}{'s logical framework \LF} and its family of variants~\cite{Harper2007,Reed2008,Assaf2016,Felicissimo2023} are \emph{extensions} of Martin-L\"of type theory, where signatures are on the \emph{meta level} and naturally capable of specifying dependent type theories;
 \item general dependent type theories~\cite{Bauer2020,Haselwarter2021,Bauer2022a,Uemura2021}, categorical semantics~\cite{Fiore1999,Tanaka2006,Tanaka2006a,Fiore2010,Hamana2011,Fiore2013,Arkor2020,Fiore2022} (which includes the syntactic model as a special case), and frameworks for substructural systems~\cite{Tanaka2006,Tanaka2006a,Wood2022} are developed \emph{within} a meta-theory (set theory or type theory), where signatures are on the \emph{object level} and their expressiveness varies depending on their target languages.
@@
 
-The \LF approach is naturally expressive, but each extension requires a different framework and a different implementation to check formal \LF proofs.
-Formalising \LF and its variants is at least as hard as formalising a dependent type theory, and they are mostly implemented separately from their theory and unverified.
+The \LF family is expressive, but each extension is a different metalanguage and requires a different implementation to check formal \LF proofs.
+Formalising \LF and its variants is at least as complicated as formalising a dependent type theory, and they are mostly implemented separately from their theory and unverified.
 
@@
 Extending the logical framework \Dedukti~\cite{Assaf2016}, \varcitet{Felicissimo2023}{'s recent \textsf{CompLF}} is capable of defining dependent type theories with computational rules and enables generic bidirectional type synthesis.
-They propose a notion of \LF signature with modes, a definition of mode-correctness (called \emph{well-typed} op.\ cit.), and an algorithm of bidirectional type synthesis.
-The algorithm is decidable, sound and complete with respect to mode-decorated terms, provided that the specified language is mode-correct, and that the set of computational rules is type-preserving, confluent, and strongly normalising.
-The last assumption, however, is challenging to establish even for specific languages, and lacks a general treatment.
+They propose a notion of \LF signature with modes, a definition of mode-correctness, and an algorithm of bidirectional type synthesis.
+The algorithm is decidable, sound and complete with respect to mode-decorated terms, provided that the specified language is mode-correct, and that the set of computational rules is type-preserving, confluent, and normalising.
+The last assumption, however, lacks a general treatment and is challenging to establish even for specific languages.
 
@@
 
-To relax the assumption of syntax-directedness, we could start from a simpler case where the ordinary typing part is still syntax-directed, but each typing rule is refined to multiple bidirectional variants, including different orders of its premises.
+To relax the assumption of syntax-directedness, we could start from a simple but common case where the ordinary typing part is still syntax-directed, but each typing rule is refined to multiple bidirectional variants, including different orders of its premises.
 In such cases, the mode decorator would need to backtrack and find all mode derivations, but the type synthesiser should still work in a syntax-directed manner on each mode derivation.
@@
 
-Next, we could consider systems where each construct can have multiple typing rules, which can have multiple bidirectional variants.
+Next, we could consider systems where each construct can have multiple typing rules, which can further have multiple bidirectional variants.
 In this setting, the bidirectional type synthesiser will also need to backtrack.

### 2024-01-22 tex/ESOP24/S6-example.tex
[ tex ] Camera ready

@@
 For illustrative purposes, consider a computational calculus~\cite{Moggi1989} with additional constructs listed in \cref{tab:computational-calculus}.
-The extended language has `only' 15 constructs, far fewer than a realistic programming language would have, but there are nearly 100 possible cases to consider in bidirectional type synthesis.
+\else
+For illustrative purposes, consider a computational calculus~\cite{Moggi1989} with naturals, sums, products, and general recursion.
+\fi
+The extended language has `only' 15 constructs, including pairing, projections, injections, and so on, and this number of constructs is still far fewer than what a realistic programming language would have.
+Even for this small calculus, there are already nearly 100 possible cases to consider in bidirectional type synthesis.
 
-On the other hand, similar to a parser generator, a type-synthesiser generator only needs a specification $(\Sigma, \Omega)$ from the user to produce a corresponding synthesiser. 
+On the other hand, similar to a parser generator, only one specification is needed for a type-synthesiser generator from the user to produce a corresponding type synthesiser. 
+\ifarxiv
 Such a specification can be derived by extending $\Sigma_{\bto}$ with additional type constructs with $\mathsf{nat}$, $\mathsf{prod}$, $\mathsf{sum}$, and $\mathsf{T}$ such that
@@
 The signature $\Omega_{\Lambda}^{\Leftrightarrow}$ is then extended with operations listed in \cref{tab:computational-calculus}.
-Mode-correctness can be derived by \cref{lem:decidability-mode-correctness} and its type synthesiser by \cref{cor:trichotomy} with the specification~$(\Sigma, \Omega)$ directly.
+Mode-correctness can be proved by applying \cref{lem:decidability-mode-correctness}, so its type synthesiser follows from \cref{cor:trichotomy} with the extended specification.
 For a demonstration in \Agda, see \cref{sec:formalisation}.
 \else
-For illustrative purposes, consider a computational calculus~\cite{Moggi1989} with naturals, sums, products, and general recursion.
-The extended language has `only' 15 constructs, including pairing, projections, injections, and so on, and this number of constructs is still far fewer than what a realistic programming language would have.
-Even for this small calculus, there are already nearly 100 possible cases to consider in bidirectional type synthesis.
-
-On the other hand, similar to a parser generator, only one specification is needed for a type-synthesiser generator from the user to produce a corresponding type synthesiser. 
 Such a specification can be derived by extending $(\Sigma_{\bto}, \Omega_{\Lambda}^{\Leftrightarrow})$ accordingly for additional types and constructs with mode-correctness proved by applying \cref{lem:decidability-mode-correctness}, so its type synthesiser follows from \cref{cor:trichotomy} directly.

### 2024-01-22 tex/ESOP24/S8-future.tex
[ tex ] Camera ready

@@
 \paragraph{Dependent types}
-In the context of dependent type theory, type synthesis entails a term equality check for type equality.
-Achieving decidable bidirectional type synthesis depends on the decidability of term equality, which, in turn, relies on the ability to specify computation rules within a language specification.
+Logical frameworks with bidirectional typing are proposed by \citet{Reed2008} and \citet{Felicissimo2023}.
+\citeauthor{Felicissimo2023}'s framework is more expressive than \citeauthor{Reed2008}'s, due to its ability to specify rewriting rules.
+Both frameworks extend \LF, enabling generic bidirectional type checking for dependent type theories.
+They also incorporate notions of signatures and mode-correctness (called \emph{strictness} and \emph{validity}, respectively, in their contexts) but differ from ours in several ways.
 
-Extending the logical framework \Dedukti~\cite{Assaf2016}, \varcitet{Felicissimo2023}{'s recent \textsf{CompLF}} is capable of defining dependent type theories with computational rules and enables generic bidirectional type synthesis.
-They propose a notion of \LF signature with modes, a definition of mode-correctness, and an algorithm of bidirectional type synthesis.
-The algorithm is decidable, sound and complete with respect to mode-decorated terms, provided that the specified language is mode-correct, and that the set of computational rules is type-preserving, confluent, and normalising.
-The last assumption, however, lacks a general treatment and is challenging to establish even for specific languages.
+First, the number of operations introduced by a signature in \LF is finite, so constructs like spine application seem impossible to define.
+Second, \citeauthor{Reed2008} and \citeauthor{Felicissimo2023} deal with decorated raw terms only, while our theory bridges the gap between ordinary and decorated raw terms by mode decoration.
+Lastly, \citeauthor{Felicissimo2023} classifies operations \emph{a priori} into introduction and elimination rules, and follows the Pfenning recipe assigning, for example, the synthesising mode to each elimination rule and its principal argument.
+As pointed out by \citeauthor{Dunfield2021} that bidirectional typing is essentially about managing information flow, and that some systems remarkably deviate from this recipe, we do not enforce it but establish our results on any reasonable information flow.
 
